### 2017-01-06 paper/arrows.tex
tex files tmp

\section{Arrows}

John Hughes defined the Arrow typeclass as follows\citHughes:
\begin{lstlisting}[frame=htrbl, caption={Arrow definition}]
class Arrow a where
	arr :: (b -> c) -> a b c
	(>>>) :: a b c -> a c d -> a b d
	first :: a b c -> a (b,d) (c,d)
\end{lstlisting}
\lstinline{arr :: (b -> c) -> a b c} is used to lift an ordinary function to an Arrow type. This can be thought of as analogous to the monadic \lstinline{return}. The \lstinline{>>>} operator, in a similar way, is analogous to the monadic composition operator \lstinline{>>=}. Ant lastly, the \lstinline{first} operator, which takes the input arrow from b to c and converts it into an arrow on pairs with the second argument untouched, is also needed for actual useful code as without it, we wouldn't have a way to save input across arrows.
\\\\
The most prominent instances of this interface are regular functions, \lstinline{(->)}
\begin{lstlisting}[frame=htrbl, caption={Arrow instance for \lstinline{(->)}}]
instance Arrow (->) where
arr f = f
f >>> g = g . f
first f = \(b, d) -> (f b, d) 
\end{lstlisting}
and the Kleisli type.
\begin{lstlisting}[frame=htrbl, caption={Arrow instance for \lstinline{(->)}}]
instance Arrow (Kleisli m) where
arr f = Kleisli $ return . f
f >>> g = Kleisli $ \b -> f b >>= g
first f = Kleisli $ \(b,d) -> f b >>= \c -> return (c,d)
\end{lstlisting}
With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \lstinline{first}, called \lstinline{second},
\begin{lstlisting}[frame=htrbl]
second :: Arrow a => a b c -> a (d, b) (d, c)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{lstlisting}
the *** combinator which combines first and second to handle two inputs in one arrow,
\begin{lstlisting}[frame=htrbl]
(***) :: Arrow a => a b c -> a d e -> a (b, d) (c, e)
f *** g = first f >>> second g
\end{lstlisting}
and the \&\&\& combinator that constructs an arrow which outputs 2 different values like ***, but takes only one input.
\begin{lstlisting}[frame=htrbl]
(&&&) :: Arrow a => a b c -> a b d -> a b (c, d)
f &&& g = arr (\b -> (b, b)) >>> (f *** g)
\end{lstlisting}
A short example given by Hughes on how to use this is \lstinline{add} over arrows:
\begin{lstlisting}[frame=htrbl, caption={add over Arrows}]
add :: Arrow a => a b Int -> a b Int -> a b Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{lstlisting}
The benefit of this interface is now that any type which is shown to be an Arrow can no be used in conjunction with this newly created \lstinline{add} combinator. Even though this example is quite simple, the power of the \lstinline{Arrow} interface immediately is clear: If a type is an \lstinline{Arrow}, it can immediately used together with every library that works on \lstinline{Arrows}. With \lstinline{Monads} this type of extensibility is not possible.

### 2017-01-06 paper/parallelHaskells.tex
tex files tmp

\section{Short into to parallel Haskells}

\subsection{Multicore Haskell}

\subsection{ParMonad}

\subsection{Eden}

\subsection{HdpH}

### 2017-01-06 paper/references.tex
tex files tmp

\begin{thebibliography}{100}

	\bibitem{hughes_arrows_98} Generalising Monads to Arrows
	\url{https://jcp.org/en/jsr/detail?id=220}, November 10, 1998
	%
\end{thebibliography}

### 2017-01-06 Definition/library/Parrows/Definition.hs
minor

@@ -91,10 +91,10 @@ toPar = return
 -- spawns the first n arrows to be evaluated in parallel. this works for infinite lists of arrows as well
 parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) => conf -> [arr a b] -> ChunkSize -> (arr [a] [b])
 parEvalNLazy conf fs chunkSize =
+               -- chunk the functions, feed the function chunks into parEvalN, chunk the input accordingly
                -- evaluate the function chunks in parallel and concat the input to a single list again
                (arr $ chunksOf chunkSize) >>> listApp fchunks >>> (arr $ concat)
                where
-                -- chunk the functions, feed the function chunks into parEvalN, chunk the input accordingly
                 fchunks = map (\x -> parEvalN conf x) $ chunksOf chunkSize fs
 
 -- evaluate two functions with different types in parallel

### 2017-01-06 ParMonad/library/Parrows/ParMonad.hs

@@ -29,13 +29,12 @@ import Parrows.Definition
 import Control.Monad.Par
 import Control.Arrow
 
-zipWithArrM :: (Arrow arr, ArrowApply arr, ArrowChoice arr, Applicative m) => (arr (a, b) (m c)) -> arr ([a], [b]) (m [c])
-zipWithArrM f = (arr $ \abs -> (zipWithArr f, abs)) >>> app >>> arr sequenceA
-
-instance (ArrowApply arr, ArrowChoice arr, NFData b) => ArrowParallel arr a b conf where
+instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b conf where
     parEvalN _ fs = (arr $ \as -> (parEval', (fs, as))) >>> app >>> arr runPar
                   where
-                    parEval' :: (ArrowApply arr, ArrowChoice arr, NFData b) => arr ([arr a b], [a]) (Par [b])
+                    parEval' :: (NFData b, ArrowApply arr, ArrowChoice arr) => arr ([arr a b], [a]) (Par [b])
                     parEval' = (arr $ \fas ->
-                                        (zipWithArrM (app >>> arr return >>> arr Control.Monad.Par.spawn), fas)) >>>
+                                        (zipWithArrA (app >>> arr return >>> arr Control.Monad.Par.spawn), fas)) >>>
                                 app >>> arr (>>= \ibs -> mapM get ibs)
+                    zipWithArrA :: (ArrowApply arr, ArrowChoice arr, Applicative m) => (arr (a, b) (m c)) -> arr ([a], [b]) (m [c])
+                    zipWithArrA f = (arr $ \abs -> (zipWithArr f, abs)) >>> app >>> arr sequenceA

### 2017-01-07 paper/arrows.tex
wip on the paper

@@
 John Hughes defined the Arrow typeclass as follows\citHughes:
-\begin{lstlisting}[frame=htrbl, caption={Arrow definition}]
+\begin{lstlisting}[frame=htrbl]
 class Arrow a where
@@
 The most prominent instances of this interface are regular functions, \lstinline{(->)}
-\begin{lstlisting}[frame=htrbl, caption={Arrow instance for \lstinline{(->)}}]
+\begin{lstlisting}[frame=htrbl]
 instance Arrow (->) where
@@
 and the Kleisli type.
-\begin{lstlisting}[frame=htrbl, caption={Arrow instance for \lstinline{(->)}}]
+\begin{lstlisting}[frame=htrbl]
 instance Arrow (Kleisli m) where
@@
 A short example given by Hughes on how to use this is \lstinline{add} over arrows:
-\begin{lstlisting}[frame=htrbl, caption={add over Arrows}]
+\begin{lstlisting}[frame=htrbl]
 add :: Arrow a => a b Int -> a b Int -> a b Int
@@
 Even though this example is quite simple, the power of the \lstinline{Arrow} interface immediately is clear: If a type is an \lstinline{Arrow}, it can immediately used together with every library that works on \lstinline{Arrows}.
-With \lstinline{Monads} this type of extensibility is not possible.
+With simple \lstinline{Monads} this type of extensibility is not possible.
+\\\\
+\textit{Note: In the definitions above we used the notation \code{a b c} for an arrow from \code{b} to \code{c}.
+From now on we will use the equivalent defintion \code{arr a b} for an arrow from \code{a} to \code{b} to make it easier to find the arrow type in type signatures.
+We kept the original notation \code{a b c} for this section to not change too much from Hughes' original definitions.}

### 2017-01-07 paper/parrows.tex
wip on the paper

\section{Parallel Arrows}
We have seen what Arrows are, so we can take a look at how they can be used to define a general interface not just to computation, but to \textbf{parallel computation} as well, next.
\\\\
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to an parallel arrow \code{arr [a] [b]} (This combinator is similar to listApp from \ref{utilfns}, which does no parallel evaluation of \code{[arr a b]}).
\begin{lstlisting}[frame=htrbl]
parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
As the implementation may a) differ depending on the type of arrow and b) we want this to be an interface for different backends with, we introduce the new typeclass \code{ArrowParallel}.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
As computation in some parallel Haskells involves additional configuration parameters giving info about the environment, we also introduce an additional \code{conf} parameter to the function which can be of different type for each backend, so we add it to the type signature of our class as well.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{lstlisting}
Note that we don't require the conf parameter in every implementation. If it is not needed, we allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN}.

\subsection{Multicore Haskell}
The Multicore Haskell implementation of this class is straightforward using listApp from \ref{utilfns} combined with the \code{using} operator from Multicore Haskell.
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = listApp fs >>> (arr $ flip using $ parList rdeepseq)
\end{lstlisting}
We hardcode the \code{parList rdeepseq} strategy here as in this context it is the only one making sense as we usually want the output list to be fully evaluated to its normal form.

\subsection{ParMonad}
The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: a -> Par (IVar a)} function which forks away the computation of a value and returns an IVar containing the result in the Par monad.
\\\\
We therefore apply each function to its corresponding input value with \code{app} and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call. This yields a list \code{[Par (IVar b)]}, which we then convert into \code{Par [IVar b]} with \code{arr sequenceA}. In order to wait for the computation to finish, we map over the \code{IVar}s inside the ParMonad with \code{arr (>>= mapM get)}. The result of this operation is a \code{Par [b]} from which we can finally remove the monad again by running \code{arr runPar} to get our output of \code{[b]}.
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = 
			(arr $ \as -> (fs, as)) >>>
			zipWithArr (app >>> arr spawnP) >>>
			arr sequenceA >>>
			arr (>>= mapM get) >>>
			arr runPar
\end{lstlisting}

\subsection{Eden}
For the Multicore and ParMonad implementation we could use general implementations that just required the ArrowApply and ArrowChoice interface. With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows. While this could be "fixed" by introducing a typeclass like
\begin{lstlisting}[frame=htrbl]
class (Arrow arr) => ArrowUnwrap arr where
	arr a b -> (a -> b)
\end{lstlisting}
we don't do this in this paper, as this seems too hacky. For now, we just implement \code{ArrowParallel} for normal functions
\begin{lstlisting}[frame=htrbl]
instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
parEvalN _ fs as = spawnF fs as
\end{lstlisting}
and the Kleisli type.
\begin{lstlisting}[frame=htrbl]
instance (Monad m, Trans a, Trans b, Trans (m b)) =>
	ArrowParallel (Kleisli m) a b conf where
parEvalN conf fs =
	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
	(Kleisli $ sequence)
\end{lstlisting}

\subsection{HdpH}

### 2017-01-07 paper/skeletons.tex
wip on the paper

\section{Skeletons}
With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.

\subsection{parEvalNLazy}
\code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. \code{map (arr . (+)) [1..]} or just because we need the arrows evaluated in chunks. \code{parEvalNLazy} fixes this. It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{(arr \$ chunksOf chunkSize)}. These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}. The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
\begin{lstlisting}[frame=htrbl]
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> [arr a b] -> ChunkSize -> (arr [a] [b])
parEvalNLazy conf fs chunkSize =
	(arr $ chunksOf chunkSize) >>>
	listApp fchunks >>>
	arr concat
	where
		fchunks = map (\x -> parEvalN conf x) $ chunksOf chunkSize fs
\end{lstlisting}

\subsection{parEval2}

-- lift the functions to "maybe evaluated" functions
-- so that if they are passed a Nothing they don't compute anything
-- then, make a list of two of these functions evaluated after each other,
-- feed each function the real value and one Nothing for the function they don't have to compute
-- and combine them back to a tuple

\begin{lstlisting}[frame=htrbl]
arrMaybe :: (ArrowApply arr) => (arr a b) -> arr (Maybe a) (Maybe b)
arrMaybe fn = (arr $ go) >>> app
	where 
		go Nothing = (arr $ \Nothing -> Nothing, Nothing)
		go (Just a) = ((arr $ \(Just x) -> (fn, x)) >>> app >>> arr Just, (Just a))
\end{lstlisting}

\begin{lstlisting}[frame=htrbl]
-- evaluate two functions with different types in parallel
parEval2 :: (ArrowParallel arr a b conf,
	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf,
	ArrowApply arr) =>
	conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
parEval2 conf f g =
	(arr $ \(a, c) -> (f_g, [(Just a, Nothing), (Nothing, Just c)])) >>>
	app >>>
	(arr $ \comb -> (fromJust (fst (comb !! 0)), fromJust (snd (comb !! 1))))
where
	f_g = parEvalN conf $ replicate 2 $ arrMaybe f *** arrMaybe g
\end{lstlisting}

\subsection{parMap}

\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf, ArrowApply arr) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f = (arr $ \as -> (parEvalN conf (repeat f), as)) >>> app
\end{lstlisting}

\subsection{parMapStream}

\begin{lstlisting}[frame=htrbl]
parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> arr a b -> ChunkSize -> arr [a] [b]
parMapStream conf f chunkSize =
	(arr $ \as -> (parEvalNLazy conf (repeat f) chunkSize, as)) >>> app
\end{lstlisting}

\subsection{farm}

\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
	ArrowChoice arr, ArrowApply arr) =>
	conf -> arr a b -> NumCores -> arr [a] [b]
farm conf f numCores =  -- chunk the input and create a function that
	-- inside of a chunk, behaves sequentially, transforms the map-chunks into a parallel function
	(arr $ \as -> (parEvalN conf $ repeat $ mapArr f, unshuffle numCores as)) >>>
	-- and then apply that function to the chunked input
	app >>>
	-- [[b]] --> [b]
	(arr shuffle)
\end{lstlisting}

\subsection{farmChunk}

### 2017-01-07 paper/utilityFunctions.tex
wip on the paper

\section{Utility Functions}\label{utilfns}
Before we go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: map and zipWith on arrows.
\\
The map combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]},
\begin{lstlisting}[frame=htrbl]
-- from http://www.cse.chalmers.se/~rjmh/afp-arrows.pdf
mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
mapArr f = arr listcase >>>
			arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
			where
				listcase [] = Left ()
				listcase (x:xs) = Right (x,xs)
\end{lstlisting}
and zipWith lift any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
\begin{lstlisting}[frame=htrbl]
zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
\end{lstlisting}
These two combinators make use of the \code{ArrowChoice} typeclass, which allows us to use the \code{|||} combinator, which takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
\begin{lstlisting}[frame=htrbl]
(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
\end{lstlisting}
With the zipWithArr combinator we can also write a combinator \code{listApp}, which lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]}.
\begin{lstlisting}[frame=htrbl]
listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
\end{lstlisting}
This combinator also makes use of the ArrowApply typeclass which allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
\\
\fixme{Do we need the exact definition of ArrowChoice here?}

### 2017-01-07 paper/benchmarks.tex
paper wip

\section{Benchmarks}
\fixme{needed for this version of the paper? we have to find better benchmarks anyways}

### 2017-01-07 paper/syntacticSugar.tex
paper wip

\section{Syntactic Sugar}
To make using our new API a bit easier, we also introduce some syntactic sugar.
We start with \code{|>>>|}, which is basically \code{>>>} on lists of arrows.
\begin{lstlisting}[frame=htrbl]
(|>>>|) :: (Arrow arr) => [arr a b] -> [arr b c] -> [arr a c]
(|>>>|) = zipWith (>>>)
\end{lstlisting}
For the basic Arrow case, we have the \code{***} combinator which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once. This can easily be lifted into a parallel computation with \code{parEval2}, but requires a backend which has a implementation that does not require any configuration (hence the \code{()} as the conf parameter in the following code snippet).
\begin{lstlisting}[frame=htrbl]
(|***|) :: (ArrowParallel arr a b (),
	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) (),
	ArrowApply arr) =>
	arr a b -> arr c d -> arr (a, c) (b, d)
(|***|) = parEval2 ()
\end{lstlisting}
With this we can analogously to the serial \code{&&&} define the parallel \code{|&&&|}.
\begin{lstlisting}[frame=htrbl]
(|&&&|) :: (ArrowParallel arr a b (),
	ArrowParallel arr (Maybe a, Maybe a) (Maybe b, Maybe c) (),
	ArrowApply arr) =>
	arr a b -> arr a c -> arr a (b, c)
(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
\end{lstlisting}

### 2017-01-09 examples/sudoku/sudoku-par2.hs
added first sudoku benchmark. this is actually a good bench as simon marlow used the exact same code

@@ -1,0 +1,21 @@
+import Sudoku
+import Control.Exception
+import System.Environment
+import Data.Maybe
+import Control.Monad.Par.Scheds.Trace
+
+main :: IO ()
+main = do
+    [f] <- getArgs
+    grids <- fmap lines $ readFile f
+
+    let (as,bs) = splitAt (length grids `div` 2) grids
+
+    print $ length $ filter isJust $ runPar $ do
+       i1 <- new
+       i2 <- new
+       fork $ put i1 (map solve as)
+       fork $ put i2 (map solve bs)
+       as' <- get i1
+       bs' <- get i2
+       return (as' ++ bs')

### 2017-01-09 examples/sudoku/sudoku1.hs

@@ -1,0 +1,16 @@
+-- <<all
+import Sudoku
+import Control.Exception
+import System.Environment
+import Data.Maybe
+
+main :: IO ()
+main = do
+  [f] <- getArgs                           -- <1>
+  file <- readFile f                       -- <2>
+
+  let puzzles   = lines file               -- <3>
+      solutions = map solve puzzles        -- <4>
+
+  print (length (filter isJust solutions)) -- <5>
+-- >>

### 2017-01-09 examples/sudoku/sudoku2.hs

@@ -1,0 +1,27 @@
+import Sudoku
+import Control.Exception
+import System.Environment
+import Control.Parallel.Strategies
+import Control.DeepSeq
+import Data.Maybe
+
+-- <<main
+main :: IO ()
+main = do
+  [f] <- getArgs
+  file <- readFile f
+
+  let puzzles = lines file
+
+      (as,bs) = splitAt (length puzzles `div` 2) puzzles -- <1>
+
+      solutions = runEval $ do
+                    as' <- rpar (force (map solve as))   -- <2>
+                    bs' <- rpar (force (map solve bs))   -- <2>
+                    rseq as'                             -- <3>
+                    rseq bs'                             -- <3>
+                    return (as' ++ bs')                  -- <4>
+
+  print (length (filter isJust solutions))
+-- >>
+

### 2017-01-09 examples/sudoku/sudoku3.hs

@@ -1,0 +1,26 @@
+import Sudoku
+import Control.Exception
+import System.Environment
+import Control.Parallel.Strategies hiding (parMap)
+import Data.Maybe
+
+-- <<main
+main :: IO ()
+main = do
+  [f] <- getArgs
+  file <- readFile f
+
+  let puzzles   = lines file
+      solutions = runEval (parMap solve puzzles)
+
+  print (length (filter isJust solutions))
+-- >>
+
+-- <<parMap
+parMap :: (a -> b) -> [a] -> Eval [b]
+parMap f [] = return []
+parMap f (a:as) = do
+   b <- rpar (f a)
+   bs <- parMap f as
+   return (b:bs)
+-- >>

### 2017-01-09 examples/sudoku/sudoku4.hs

@@ -1,0 +1,25 @@
+import Sudoku
+import Control.Exception
+import System.Environment
+import Control.Parallel.Strategies hiding (parMap)
+import Data.Maybe
+
+-- <<main
+main :: IO ()
+main = do
+  [f] <- getArgs
+  file <- readFile f
+
+  let puzzles   = lines file
+      solutions = runEval (parMap solve puzzles)
+
+  evaluate (length puzzles)
+  print (length (filter isJust solutions))
+-- >>
+
+parMap :: (a -> b) -> [a] -> Eval [b]
+parMap f [] = return []
+parMap f (a:as) = do
+   b <- rpar (f a)
+   bs <- parMap f as
+   return (b:bs)

### 2017-01-09 examples/sudoku/sudoku5.hs

@@ -1,0 +1,17 @@
+import Sudoku
+import Control.Exception
+import System.Environment
+import Control.Parallel.Strategies
+import Data.Maybe
+
+main :: IO ()
+main = do
+  [f] <- getArgs
+  file <- readFile f
+
+  let puzzles   = lines file
+-- <<solutions
+  let solutions = map solve puzzles `using` parList rseq
+-- >>
+
+  print (length (filter isJust solutions))

### 2017-01-09 paper/parrows.tex

@@
 	ArrowParallel arr a b conf where
-		parEvalN _ fs = listApp fs >>> (arr $ flip using $ parList rdeepseq)
+		parEvalN _ fs = listApp fs >>> arr (flip using $ parList rdeepseq)
 \end{lstlisting}

### 2017-01-09 paper/skeletons.tex

@@
 parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> [arr a b] -> ChunkSize -> (arr [a] [b])
-parEvalNLazy conf fs chunkSize =
-	(arr $ chunksOf chunkSize) >>>
+	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
+parEvalNLazy conf chunkSize fs =
+	arr (chunksOf chunkSize) >>>
 	listApp fchunks >>>
@@
 	where
-		fchunks = map (\x -> parEvalN conf x) $ chunksOf chunkSize fs
+		fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
 \end{lstlisting}
@@
 \subsection{parEval2}
-
--- lift the functions to "maybe evaluated" functions
--- so that if they are passed a Nothing they don't compute anything
--- then, make a list of two of these functions evaluated after each other,
--- feed each function the real value and one Nothing for the function they don't have to compute
--- and combine them back to a tuple
-
+We have only talked about parallelizing the computation of arrows of the same type up until now.
+But sometimes we want to paralellize inhomogenous types as well.
+For this, we introduce a helper combinator \code{arrMaybe} first, that converts an arrow \code{arr a b} to an arrow \code{arr (Maybe a) (Maybe b)}.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-
+With this, we can now easily write \code{parEval2} which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)}.
+We start by converting both arrows with \code{arrMaybe}, combining them with \code{***} into a new arrow \code{arr (Maybe a, Maybe c) (Maybe b, Maybe d)}.
+This is then replicated twice and fed into parEvalN to get a \code{arr [(Maybe a, Maybe c)] [(Maybe b, Maybe d)]}.
+We can then apply this arrow to the input \code{[(Just a, Nothing), (Nothing, Just c)]} and then extract the resulting values with \code{fromJust} and the \code{!!} operator on lists in the last step.
 \begin{lstlisting}[frame=htrbl]
--- evaluate two functions with different types in parallel
 parEval2 :: (ArrowParallel arr a b conf,
@@
 \subsection{parMap}
-
+\code{parMap} is probably the most common skeleton for parallel programs.
+We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}.
+Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
 \begin{lstlisting}[frame=htrbl]
@@
 	conf -> (arr a b) -> (arr [a] [b])
-parMap conf f = (arr $ \as -> (parEvalN conf (repeat f), as)) >>> app
+parMap conf f =
+	(arr $ \as -> (f, as)) >>>
+	(first $ arr repeat >>>
+		arr (parEvalN conf)) >>>
+	app
 \end{lstlisting}
@@
 \subsection{parMapStream}
-
+As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
+So it makes sense to also have a \code{parMapStream} which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
 \begin{lstlisting}[frame=htrbl]
 parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> arr a b -> ChunkSize -> arr [a] [b]
-parMapStream conf f chunkSize =
-	(arr $ \as -> (parEvalNLazy conf (repeat f) chunkSize, as)) >>> app
+	conf -> ChunkSize -> arr a b -> arr [a] [b]
+parMapStream conf chunkSize f =
+	(arr $ \as -> (f, as)) >>>
+	(first $ arr repeat >>>
+		arr (parEvalNLazy conf chunkSize)) >>>
+	app
 \end{lstlisting}
@@
 \subsection{farm}
-
+\code{parMap} spawns every single computation in a new thread (at least for the instances of ArrowParallel we gave in this paper).
+This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores > actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
 \begin{lstlisting}[frame=htrbl]
@@
 	ArrowChoice arr, ArrowApply arr) =>
-	conf -> arr a b -> NumCores -> arr [a] [b]
-farm conf f numCores =  -- chunk the input and create a function that
-	-- inside of a chunk, behaves sequentially, transforms the map-chunks into a parallel function
-	(arr $ \as -> (parEvalN conf $ repeat $ mapArr f, unshuffle numCores as)) >>>
-	-- and then apply that function to the chunked input
+	conf -> NumCores -> arr a b -> arr [a] [b]
+farm conf numCores f =
+	(arr $ \as -> (f, as)) >>>
+	(first $ arr mapArr >>> arr repeat >>>
+		arr (parEvalN conf)) >>>
+	(second $ arr (unshuffle numCores)) >>>
 	app >>>
-	-- [[b]] --> [b]
-	(arr shuffle)
+	arr shuffle
 \end{lstlisting}
+The definition of \code{unshuffle} is
+\begin{lstlisting}[frame=htrbl]
+unshuffle :: Int
+	-> [a]
+	-> [[a]]
+unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
+\end{lstlisting}
+, while \code{shuffle} is defined as:
+\begin{lstlisting}[frame=htrbl]
+shuffle :: [[a]]
+	-> [a]
+shuffle = concat . transpose
+\end{lstlisting}
+These were taken from Eden's source code.
+\\
+\fixme{Do we want a farm that works on a list of arrows as well?}
+\\
+\fixme{For this we just have to replace the mapArr with zipWithArr}
 
 \subsection{farmChunk}
+As \code{farm} is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
+So we define \code{farmChunk} which uses \code{parEvalNLazy} instead of \code{parEvalN} like this:
+\begin{lstlisting}[frame=htrbl]
+farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
+	ArrowChoice arr, ArrowApply arr) =>
+	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
+farmChunk conf chunkSize numCores f =
+	(arr $ \as -> (f, as)) >>>
+	(first $ arr mapArr >>> arr repeat >>>
+		arr (parEvalNLazy conf chunkSize)) >>>
+	(second $ arr (unshuffle numCores)) >>>
+	app >>>
+	arr shuffle
+\end{lstlisting}

### 2017-01-12 paper/arrows.tex
paper wip
added benchmarks to the paper

@@
 \begin{lstlisting}[frame=htrbl]
-class Arrow a where
-	arr :: (b -> c) -> a b c
-	(>>>) :: a b c -> a c d -> a b d
-	first :: a b c -> a (b,d) (c,d)
+class Arrow arr where
+	arr :: (a -> b) -> arr a b
+	(>>>) :: arr a b -> arr b c -> arr a c
+	first :: arr a b -> arr (a,c) (b,c)
 \end{lstlisting}
-\lstinline{arr :: (b -> c) -> a b c} is used to lift an ordinary function to an Arrow type.
+\lstinline{arr} is used to lift an ordinary function to an Arrow type.
 This can be thought of as analogous to the monadic \lstinline{return}.
@@
 \\\\
-The most prominent instances of this interface are regular functions, \lstinline{(->)}
+The most prominent instances of this interface are regular functions \lstinline{(->)},
 \begin{lstlisting}[frame=htrbl]
@@
 	f >>> g = g . f
-first f = \(b, d) -> (f b, d) 
+	first f = \(a, c) -> (f a, c) 
 \end{lstlisting}
@@
 	arr f = Kleisli $ return . f
-f >>> g = Kleisli $ \b -> f b >>= g
-first f = Kleisli $ \(b,d) -> f b >>= \c -> return (c,d)
+	f >>> g = Kleisli $ \a -> f a >>= g
+	first f = Kleisli $ \(a,c) -> f a >>= \b -> return (b,c)
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-second :: Arrow a => a b c -> a (d, b) (d, c)
+second :: Arrow arr => arr a b -> arr (c, a) (c, b)
 second f = arr swap >>> first f >>> arr swap
@@
 \begin{lstlisting}[frame=htrbl]
-(***) :: Arrow a => a b c -> a d e -> a (b, d) (c, e)
+(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
 f *** g = first f >>> second g
@@
 \begin{lstlisting}[frame=htrbl]
-(&&&) :: Arrow a => a b c -> a b d -> a b (c, d)
-f &&& g = arr (\b -> (b, b)) >>> (f *** g)
+(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
+f &&& g = arr (\a -> (a, a)) >>> (f *** g)
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-add :: Arrow a => a b Int -> a b Int -> a b Int
+add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
 add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
 \end{lstlisting}
-The benefit of this interface is now that any type which is shown to be an Arrow can no be used in conjunction with this newly created \lstinline{add} combinator.
+The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an Arrow can be used in conjunction with this newly created \lstinline{add} combinator.
 Even though this example is quite simple, the power of the \lstinline{Arrow} interface immediately is clear: If a type is an \lstinline{Arrow}, it can immediately used together with every library that works on \lstinline{Arrows}.
-With simple \lstinline{Monads} this type of extensibility is not possible.
+Compared to simple \lstinline{Monads}, this enables us to write code that is more extensible, without touching the internals of the specific Arrows.
 \\\\
-\textit{Note: In the definitions above we used the notation \code{a b c} for an arrow from \code{b} to \code{c}.
-From now on we will use the equivalent defintion \code{arr a b} for an arrow from \code{a} to \code{b} to make it easier to find the arrow type in type signatures.
+\textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used.
+We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.
 We kept the original notation \code{a b c} for this section to not change too much from Hughes' original definitions.}

### 2017-01-12 paper/parrows.tex
added benchmarks to the paper

@@
 \section{Parallel Arrows}
-We have seen what Arrows are, so we can take a look at how they can be used to define a general interface not just to computation, but to \textbf{parallel computation} as well, next.
-\\\\
+We have seen what Arrows are and how they can be seen as a general interface to computation.
+In the following section we will discuss how Arrows can also be seen as a general interface not only to computation, but to \textbf{parallel computation} as well.
+We start by introducing the interface and explaining the reasonings behind it.
+Then, we discuss some implementations using exisiting parallel Haskells.
+Finally, we explain why using Arrows for expressing parallelism is beneficial.
+\subsection{The ArrowParallel interface}
 In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
@@
 \end{lstlisting}
-Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to an parallel arrow \code{arr [a] [b]} (This combinator is similar to listApp from \ref{utilfns}, which does no parallel evaluation of \code{[arr a b]}).
+Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function listApp, but does parallel instead of serial evaluation).
 \begin{lstlisting}[frame=htrbl]
-parEvalN :: [arr a b] -> arr [a] [b]
+parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
 \end{lstlisting}
-As the implementation may a) differ depending on the type of arrow and b) we want this to be an interface for different backends with, we introduce the new typeclass \code{ArrowParallel}.
+With this definition of \code{parEvalN}, parallel execution can be looked at as yet another arrow combinator.
+But as the implementation may differ depending on the actuall type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-As computation in some parallel Haskells involves additional configuration parameters giving info about the environment, we also introduce an additional \code{conf} parameter to the function which can be of different type for each backend, so we add it to the type signature of our class as well.
+Sometimes parallel Haskells require additional configuration parameters for information about the execution environment.
+This is why we also introduce an additional \code{conf} parameter to the function.
+We also do not want \code{conf} to be of a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}, so we add it to the type signature of the typeclass as well.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-Note that we don't require the conf parameter in every implementation.
-If it is not needed, we allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN}.
+We don't require the conf parameter in every implementation.
+If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN} as we will see in the implementation sections.
 
@@
 \subsection{HdpH}
+
+\subsection{Benefits of parallel Arrows}
+We have seen that we can wrap parallel Haskells inside of the \code{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
+\begin{itemize}
+	\item \textbf{Arrow API benefits}:
+	With the \code{ArrowParallel} typeclass we do not lose any benefits of using arrows as \code{parEvalN} is just yet another arrow combinator.
+The resulting arrow can be used in the same way its serial version can be used.
+This is a big advantage of this approach, especially compared to the monad solutions as all arrow combinators will continue to work instead of requiring special ones for each parallel monad.
+	\item \textbf{Abstraction}:
+	With the \code{ArrowParallel} typeclass, we abstracted all parallel implementation logic away from the business logic.
+This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
+So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster using an Eden built, by just swapping out the actual \code{ArrowParallel} instance.
+\end{itemize}

### 2017-01-12 paper/skeletons.tex
added benchmarks to the paper

@@
 We have only talked about parallelizing the computation of arrows of the same type up until now.
-But sometimes we want to paralellize inhomogenous types as well.
+But sometimes we want to paralellize heterogenous types as well.
 For this, we introduce a helper combinator \code{arrMaybe} first, that converts an arrow \code{arr a b} to an arrow \code{arr (Maybe a) (Maybe b)}.
@@
 \code{parMap} spawns every single computation in a new thread (at least for the instances of ArrowParallel we gave in this paper).
-This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores > actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
+This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-These were taken from Eden's source code.
-\\
-\fixme{Do we want a farm that works on a list of arrows as well?}
-\\
-\fixme{For this we just have to replace the mapArr with zipWithArr}
+(These were taken from Eden's source code.)
 

### 2017-01-12 paper/utilityFunctions.tex
paper wip
added benchmarks to the paper

@@
 mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
-mapArr f = arr listcase >>>
+mapArr f =
+	arr listcase >>>
 	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
@@
 \end{lstlisting}
-These two combinators make use of the \code{ArrowChoice} typeclass, which allows us to use the \code{|||} combinator, which takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
+These two combinators make use of the \code{ArrowChoice} typeclass, which allows us to use the \code{|||} combinator.
+It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
 \begin{lstlisting}[frame=htrbl]
@@
 This combinator also makes use of the ArrowApply typeclass which allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
-\\
-\fixme{Do we need the exact definition of ArrowChoice here?}

### 2017-01-12 paper/arrows.tex
minor

@@
 
-John Hughes defined the Arrow typeclass as follows\citHughes:
+John Hughes defined the arrow typeclass as follows\citHughes:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-\lstinline{arr} is used to lift an ordinary function to an Arrow type.
+\lstinline{arr} is used to lift an ordinary function to an arrow type.
 This can be thought of as analogous to the monadic \lstinline{return}.
@@
 \end{lstlisting}
-The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an Arrow can be used in conjunction with this newly created \lstinline{add} combinator.
-Even though this example is quite simple, the power of the \lstinline{Arrow} interface immediately is clear: If a type is an \lstinline{Arrow}, it can immediately used together with every library that works on \lstinline{Arrows}.
-Compared to simple \lstinline{Monads}, this enables us to write code that is more extensible, without touching the internals of the specific Arrows.
+The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
+Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can immediately used together with every library that works on arrows.
+Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
 \\\\

### 2017-01-12 paper/parrows.tex
minor

@@
 \subsection{Eden}
-For the Multicore and ParMonad implementation we could use general implementations that just required the ArrowApply and ArrowChoice interface.
+For the Multicore and ParMonad implementation we could use general implementations that just required the \code{ArrowApply} and \code{ArrowChoice} typeclasses.
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While this could be "fixed" by introducing a typeclass like
+While we could still manage to have only one class in the module by introducing a typeclass like
 \begin{lstlisting}[frame=htrbl]

### 2017-01-12 paper/utilityFunctions.tex
minor

@@
 \end{lstlisting}
-This combinator also makes use of the ArrowApply typeclass which allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
+This combinator also makes use of the \code{ArrowApply} typeclass which allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.

### 2017-01-12 paper/examples/parallelism-eden.hs
wip

@@ -22,25 +22,11 @@ CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
 TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
 SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
 -}
-{-# LANGUAGE FlexibleInstances, UndecidableInstances, MultiParamTypeClasses #-}
-module Parrows.Eden where
-
-import Parrows.Definition
-
-import Control.Arrow
+module Main where
 
 import Control.Parallel.Eden
 
--- ArrowParallel Instances
+parallel :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
+parallel fs as = spawnF fs as
 
--- FIXME: will this work with (spawnF id bs) with already "computed" bs
--- so that we can write a uniform instance that doesn't require
--- the weird unwrapping of the Kleisli type?
--- Probably not, because we would end up computing the values while
--- sending them over the network, right?
-
-instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
-    parEvalN _ fs as = spawnF fs as
-
-instance (Monad m, Trans a, Trans b, Trans (m b)) => ArrowParallel (Kleisli m) a b conf where
-    parEvalN conf fs = (arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>> (Kleisli $ sequence)
+main = do print $ parallel [(*2), (+1)] [1::Int, 2]

### 2017-01-12 paper/examples/parallelism-multicore.hs

@@ -22,25 +22,11 @@ CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
 TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
 SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
 -}
-{-# LANGUAGE FlexibleInstances, UndecidableInstances, MultiParamTypeClasses #-}
-module Parrows.Eden where
+module Main where
 
-import Parrows.Definition
+import Control.Parallel.Strategies
 
-import Control.Arrow
+parallel :: (NFData b) => [a -> b] -> [a] -> [b]
+parallel fs as = zipWith ($) fs as `using` parList rdeepseq
 
-import Control.Parallel.Eden
-
--- ArrowParallel Instances
-
--- FIXME: will this work with (spawnF id bs) with already "computed" bs
--- so that we can write a uniform instance that doesn't require
--- the weird unwrapping of the Kleisli type?
--- Probably not, because we would end up computing the values while
--- sending them over the network, right?
-
-instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
-    parEvalN _ fs as = spawnF fs as
-
-instance (Monad m, Trans a, Trans b, Trans (m b)) => ArrowParallel (Kleisli m) a b conf where
-    parEvalN conf fs = (arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>> (Kleisli $ sequence)
+main = do print $ parallel [(*2), (+1)] [1::Int, 2]

### 2017-01-12 paper/examples/parallelism-parmonad.hs

@@ -22,25 +22,11 @@ CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
 TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
 SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
 -}
-{-# LANGUAGE FlexibleInstances, UndecidableInstances, MultiParamTypeClasses #-}
-module Parrows.Eden where
+module Main where
 
-import Parrows.Definition
+import Control.Monad.Par
 
-import Control.Arrow
+parallel :: (NFData b) => [a -> b] -> [a] -> [b]
+parallel fs as = runPar $ (sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
 
-import Control.Parallel.Eden
-
--- ArrowParallel Instances
-
--- FIXME: will this work with (spawnF id bs) with already "computed" bs
--- so that we can write a uniform instance that doesn't require
--- the weird unwrapping of the Kleisli type?
--- Probably not, because we would end up computing the values while
--- sending them over the network, right?
-
-instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
-    parEvalN _ fs as = spawnF fs as
-
-instance (Monad m, Trans a, Trans b, Trans (m b)) => ArrowParallel (Kleisli m) a b conf where
-    parEvalN conf fs = (arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>> (Kleisli $ sequence)
+main = do print $ parallel [(*2), (+1)] [1::Int, 2]

### 2017-01-12 paper/benchmarks.tex

@@
 \section{Benchmarks}
-\fixme{needed for this version of the paper? we have to find better benchmarks anyways}
+To check performance of our new parallel arrow API, we conducted some benchmarks.
+These are based on a sudoku solver from the examples for Simon Marlows book "Parallel and Concurrent Programming in Haskell" which can be found on his GitHub \cite{marlow_parconc}.
+The results are displayed in the following graphs which are ordered by problem size.
+\\\\
+The Benchmarks were run on a Core i7-3970X CPU @ 3.5GHz with 6 cores and 12 threads. All versions were run 100 times. The resulting running-times were then averaged.
+\\
+\begin{tikzpicture}
+\begin{axis}[
+xbar, xmin=0,
+width=12cm, height=7cm, enlarge y limits=0.5,
+xlabel={avg. seconds run on sudoku17.1000.txt (100 runs)},
+symbolic y coords={parrows-sudoku-farmChunk, parrows-sudoku-farm, parrows-sudoku-parmapstream, parrows-sudoku-parmap, parmonad-sudoku-parmap, sudoku-sequential},
+ytick=data,
+nodes near coords, nodes near coords align={horizontal},
+]
+\addplot coordinates {(.215,parrows-sudoku-farmChunk) (.213,parrows-sudoku-farm) (.188,parrows-sudoku-parmapstream) (.197,parrows-sudoku-parmap) (.527,parmonad-sudoku-parmap) (1.338,sudoku-sequential)};
+\end{axis}
+\end{tikzpicture}
+\textit{Note: the bad result for parmonad-sudoku-parmap seems to be a artefact as it performs much better in the other benchmarks}
+\\\\
+\begin{tikzpicture}
+\begin{axis}[
+xbar, xmin=0,
+width=12cm, height=7cm, enlarge y limits=0.5,
+xlabel={avg. seconds run on sudoku17.16000.txt (100 runs)},
+symbolic y coords={parrows-sudoku-farmChunk, parrows-sudoku-farm, parrows-sudoku-parmapstream, parrows-sudoku-parmap, parmonad-sudoku-parmap, sudoku-sequential},
+ytick=data,
+nodes near coords, nodes near coords align={horizontal},
+]
+\addplot coordinates {(3.218,parrows-sudoku-farmChunk) (3.148,parrows-sudoku-farm) (2.755,parrows-sudoku-parmapstream) (2.860,parrows-sudoku-parmap) (2.936,parmonad-sudoku-parmap) (21.324,sudoku-sequential)};
+\end{axis}
+\end{tikzpicture}
+\begin{tikzpicture}
+\begin{axis}[
+xbar, xmin=0,
+width=12cm, height=7cm, enlarge y limits=0.5,
+xlabel={avg. seconds run on sudoku17.49151.txt (100 runs)},
+symbolic y coords={parrows-sudoku-farmChunk, parrows-sudoku-farm, parrows-sudoku-parmapstream, parrows-sudoku-parmap, parmonad-sudoku-parmap, sudoku-sequential},
+ytick=data,
+nodes near coords, nodes near coords align={horizontal},
+]
+\addplot coordinates {(10.038,parrows-sudoku-farmChunk) (10.151,parrows-sudoku-farm) (8.776,parrows-sudoku-parmapstream) (8.969,parrows-sudoku-parmap) (8.679,parmonad-sudoku-parmap) (60.931,sudoku-sequential)};
+\end{axis}
+\end{tikzpicture}
+\\\\
+As we can see, the parallel arrow versions of the program all have around the same speedup as the ParMonad based version.
+This means that using the \code{ArrowParallel} typeclass doesn't add any real overhead.
+Also, the slightly worse results for "parrows-sudoku-farm" and "parrows-sudoku-farmChunk" can be explained by either imperfect parameters or by the fact that the benchmarks do not really require the scheduling introduced by the skeletons which in this case introduces unnecessary overhead.
+This would probably look different if the benchmark would use e.g.
+a divide and conquer approach to solve the sudoku puzzles.

### 2017-01-12 paper/syntacticSugar.tex

@@
 \end{lstlisting}
-For the basic Arrow case, we have the \code{***} combinator which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
-This can easily be lifted into a parallel computation with \code{parEval2}, but requires a backend which has a implementation that does not require any configuration (hence the \code{()} as the conf parameter in the following code snippet).
+For basic arrows, we have the \code{***} combinator which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
+This can easily be translated into a parallel version with \code{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \code{()} as the conf parameter in the following code snippet).
 \begin{lstlisting}[frame=htrbl]

### 2017-01-12 paper/parrows.tex
wip

@@
 Finally, we explain why using Arrows for expressing parallelism is beneficial.
-\subsection{The ArrowParallel interface}
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
+\subsection{The ArrowParallel typeclass}
+As we have seen earlier, in its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
 \begin{lstlisting}[frame=htrbl]
@@
 
-\subsection{HdpH}
-
 \subsection{Benefits of parallel Arrows}

### 2017-01-12 paper/examples/parallelism-eden.hs
renamed parallel to parEvalN in the examples for the paper

@@ -26,7 +26,7 @@ module Main where
 
 import Control.Parallel.Eden
 
-parallel :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
-parallel fs as = spawnF fs as
+parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = spawnF fs as
 
-main = do print $ parallel [(*2), (+1)] [1::Int, 2]
+main = do print $ parEvalN [(*2), (+1)] [1::Int, 2]

### 2017-01-12 paper/examples/parallelism-multicore.hs

@@ -26,7 +26,7 @@ module Main where
 
 import Control.Parallel.Strategies
 
-parallel :: (NFData b) => [a -> b] -> [a] -> [b]
-parallel fs as = zipWith ($) fs as `using` parList rdeepseq
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = zipWith ($) fs as `using` parList rdeepseq
 
-main = do print $ parallel [(*2), (+1)] [1::Int, 2]
+main = do print $ parEvalN [(*2), (+1)] [1::Int, 2]

### 2017-01-12 paper/examples/parallelism-parmonad.hs

@@ -26,7 +26,7 @@ module Main where
 
 import Control.Monad.Par
 
-parallel :: (NFData b) => [a -> b] -> [a] -> [b]
-parallel fs as = runPar $ (sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = runPar $ (sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
 
-main = do print $ parallel [(*2), (+1)] [1::Int, 2]
+main = do print $ parEvalN [(*2), (+1)] [1::Int, 2]

### 2017-01-12 paper/abstract.tex

\begin{abstract}
Arrows were introduced in John Hughes paper as an alternative to monads for API design \citHughes. In the paper Hughes describes that arrows have one powerful additional property when compared to monads for API design: Extensibility. In this paper we will show how this property can be used to add parallelism capabilities to different arrows.
\\\\
First, we give an introduction to some of the possible ways to add parallelism to Haskell programs. Then, we give the basic definition of Arrows, which is followed up by the introduction of some utility functions used in this paper. Next, we introduce the \code{ArrowParallel} typeclass together with backends for it written with the parallel Haskells introduced earlier, finishing up with the benefits of this new way of writing parallel programs. After this we give the definition of several parallel skeletons. Then, we introduce some syntactic sugar to mimic the sequential arrow combinators introduced by Hughes in \citHughes but with parallelism added. We also give benchmarks of our newly created parallel Haskell based on arrows. Finally we give a short conclusion of what we managed to achieve.
\end{abstract}

### 2017-01-12 paper/parallelHaskells.tex
wip on the paper

@@
-\section{Short into to parallel Haskells}
+\section{Short introduction to parallel Haskells}
+There are already several ways write parallel programs in Haskell.
+As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
 
@@
 
+\begin{lstlisting}[frame=htrbl]
+parallel :: (NFData b) => [a -> b] -> [a] -> [b]
+parallel fs as = zipWith ($) fs as `using` parList rdeepseq
+\end{lstlisting}
+
 \subsection{ParMonad}
+
+\begin{lstlisting}[frame=htrbl]
+parallel :: (NFData b) => [a -> b] -> [a] -> [b]
+parallel fs as = runPar $ 
+	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+\end{lstlisting}
 
@@
 
-\subsection{HdpH}
+\begin{lstlisting}[frame=htrbl]
+parallel :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
+parallel fs as = spawnF fs as
+\end{lstlisting}

### 2017-01-12 paper/motivation.tex
wip on the paper

--- 2017-01-12 paper/abstract.tex
@@
-\begin{abstract}
+\section{Motivation}
 Arrows were introduced in John Hughes paper as an alternative to monads for API design \citHughes.
 In the paper Hughes describes that arrows have one powerful additional property when compared to monads for API design: Extensibility.
-In this paper we will show how this property can be used to add parallelism capabilities to different arrows.
-\\\\
-First, we give an introduction to some of the possible ways to add parallelism to Haskell programs.
-Then, we give the basic definition of Arrows, which is followed up by the introduction of some utility functions used in this paper.
-Next, we introduce the \code{ArrowParallel} typeclass together with backends for it written with the parallel Haskells introduced earlier, finishing up with the benefits of this new way of writing parallel programs.
-After this we give the definition of several parallel skeletons.
-Then, we introduce some syntactic sugar to mimic the sequential arrow combinators introduced by Hughes in \citHughes but with parallelism added.
-We also give benchmarks of our newly created parallel Haskell based on arrows.
-Finally we give a short conclusion of what we managed to achieve.
-\end{abstract}
+In this paper we will show how this property can be used to add parallelism capabilities to different arrow-types.

### 2017-01-12 paper/parallelHaskells.tex
wip on the paper

--- 2017-01-06 paper/parallelHaskells.tex
@@
-\section{Short into to parallel Haskells}
+\section{Short introduction to parallel Haskells}
+There are already several ways write parallel programs in Haskell.
+As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
+\\\\
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: [a -> b] -> [a] -> [b]
+\end{lstlisting}
 
 \subsection{Multicore Haskell}
+Multicore Haskell \cite{multicore_hackage} is the GHC native way to do parallel processing.
+It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}.
+For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
+This lazy list \code{[b]} is then evaluated in parallel with the \code{using} operator and a strategy \code{Strategy [b]}.
+This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = zipWith ($) fs as `using` parList rdeepseq
+\end{lstlisting}
 
 \subsection{ParMonad}
+The \code{Par} monad introduced by Marlow et al \cite{monad_par_paper}, which can be found in the monad-par package on hackage \cite{monad_par_hackage}, is a monad designed for composition of parallel programs.
+\\\\
+Our parallel evaluation function \code{parEvalN} can be defined by zipping the list of \code{[a -> b]} with the list of inputs \code{[a]} with the application operator \code{\$} just like with Multicore Haskell.
+Then, we map over this not yet evaluated lazy list of results \code{[b]} with \code{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \code{[Par (IVar b)]}, which we convert to \code{Par [IVar b]} with \code{sequenceA}.
+We wait for the computations to finish by mapping over the \code{IVar b}'s inside the \code{Par} monad with \code{get}.
+This results in \code{Par [b]}.
+We finally execute this process with \code{runPar} to finally get \code{[b]} again.
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = runPar $ 
+	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+\end{lstlisting}
 
 \subsection{Eden}
-
-\subsection{HdpH}
+Eden \cite{eden_hackage} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backend.
+This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell.
+\\\\
+While it also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
+\\
+\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
+\\
+This allows us to define \code{parEvalN} quite easily:
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = spawnF fs as
+\end{lstlisting}

### 2017-01-12 paper/parrows.tex
wip on the paper

--- 2017-01-12 paper/parrows.tex
@@
 Finally, we explain why using Arrows for expressing parallelism is beneficial.
-\subsection{The ArrowParallel interface}
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
+\subsection{The ArrowParallel typeclass}
+As we have seen earlier, that, in its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
 \begin{lstlisting}[frame=htrbl]
@@
 \subsection{ParMonad}
-The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: a -> Par (IVar a)} function which forks away the computation of a value and returns an IVar containing the result in the Par monad.
+The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function which forks away the computation of a value and returns an IVar containing the result in the Par monad.
 \\\\

### 2017-01-12 paper/references.tex
added benchmarks to the paper
wip on the paper

@@
 	\url{https://jcp.org/en/jsr/detail?id=220}, November 10, 1998
-	%
+
+	\bibitem{marlow_parconc} Sample code for Simon Marlow's book "Parallel and Concurrent Programming in Haskell"
+	\url{https://github.com/simonmar/parconc-examples}, retrieved on 01/12/2017
+
+	\bibitem{multicore_hackage} Multicore's parallel package on Hackage
+	\url{https://hackage.haskell.org/package/parallel-3.2.1.0}, retrieved on 01/12/2017
+
+	\bibitem{monad_par_paper} A Monad for Deterministic Parallelism
+	\url{http://www.cs.indiana.edu/~rrnewton/papers/haskell2011_monad-par.pdf}, retrieved on 01/12/2017
+
+	\bibitem{monad_par_hackage} ParMonad (Control.Monad.Par) on Hackage
+	\url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}, retrieved on 01/12/2017
+
+	\bibitem{eden_hackage} Eden on Hackage
+	\url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}, retrieved on 01/12/2017
 \end{thebibliography}

### 2017-01-13 paper/conclusion.tex
first version of the complete paper

\section{Conclusion}
As we have seen in this paper, arrows are a useful tool for composing parallel programs. We have already seen that they have some benefits over existing parallel Haskells as they are, by nature, more extensible. Performancewise, parallel arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead. While we have seen that the parallel arrows can express computation on clusters, it is noteworthy that in its current state, the parallel arrow API doesn't have as much control over data-flow as for example Eden. In its current state this will end up in the root node being a possible bottleneck on clusters when parallel arrows are combined with \code{>>>}. This can probably be remedied by introducing some sort of a Future construct that can be passed between nodes so that nodes can communicate without going through the root node. This will have to be implemented in a follow up paper, though.

### 2017-01-13 paper/references.tex
first version of the complete paper

@@
 
-	\bibitem{hughes_arrows_98} Generalising Monads to Arrows
+	\bibitem{hughes_arrows_98} Hughes, Generalising Monads to Arrows
 	\url{https://jcp.org/en/jsr/detail?id=220}, November 10, 1998
@@
 
-	\bibitem{monad_par_paper} A Monad for Deterministic Parallelism
+	\bibitem{monad_par_paper} Marlow et al, A Monad for Deterministic Parallelism
 	\url{http://www.cs.indiana.edu/~rrnewton/papers/haskell2011_monad-par.pdf}, retrieved on 01/12/2017
@@
 
-	\bibitem{eden_hackage} Eden on Hackage
+	\bibitem{eden_hackage} Eden Modules on Hackage
 	\url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}, retrieved on 01/12/2017
+
+	\bibitem{programming_with_arrows} Hughes, Programming with Arrows
+	\url{http://www.cse.chalmers.se/~rjmh/afp-arrows.pdf}, retrieved on 02/12/2017
+
+	\bibitem{eden_skel_shuffle} Eden Skeletons' Control.Parallel.Eden.Map package source code
+	\url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}, retrieved on 02/12/2017
+
 \end{thebibliography}

### 2017-01-13 paper/skeletons.tex
first version of the complete paper

@@
 \subsection{farm}
-\code{parMap} spawns every single computation in a new thread (at least for the instances of ArrowParallel we gave in this paper).
+\code{parMap} spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
 This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
@@
 \end{lstlisting}
-(These were taken from Eden's source code.)
+(These were taken from Eden's source code. \cite{eden_skel_shuffle})
 

### 2017-01-13 paper/utilityFunctions.tex
first version of the complete paper

@@
 \\
-The map combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]},
+The map combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
 \begin{lstlisting}[frame=htrbl]
--- from http://www.cse.chalmers.se/~rjmh/afp-arrows.pdf
 mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]

### 2017-01-13 paper/abstract.tex
wip on the paper
minor

@@
 In the paper Hughes describes that arrows have one powerful additional property when compared to monads for API design: Extensibility.
-In this paper we will show how this property can be used to add parallelism capabilities to different arrows.
+In this paper we will show how this property can be used to add parallelism capabilities to different arrow-types.
 \\\\
@@
 After this we give the definition of several parallel skeletons.
-Then, we introduce some syntactic sugar to mimic the sequential arrow combinators introduced by Hughes in \citHughes but with parallelism added.
+Then, we introduce some syntactic sugar to mimic the sequential arrow combinators introduced by Hughes \citHughes ~but with parallelism added.
 We also give benchmarks of our newly created parallel Haskell based on arrows.

### 2017-01-14 paper/abstract.tex
minor

@@
 \begin{abstract}
-Arrows were introduced in John Hughes paper as an alternative to monads for API design \citHughes.
+Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
 In the paper Hughes describes that arrows have one powerful additional property when compared to monads for API design: Extensibility.
@@
 Then, we introduce some syntactic sugar to mimic the sequential arrow combinators introduced by Hughes \citHughes ~but with parallelism added.
-We also give benchmarks of our newly created parallel Haskell based on arrows.
+We also give benchmarks of our newly created parallel Haskell.
 Finally we give a short conclusion of what we managed to achieve.

### 2017-01-14 paper/arrows.tex
minor

@@
 \section{Arrows}
-
-John Hughes defined the arrow typeclass as follows\citHughes:
+Arrows were introduced by Hughes \citHughes ~as a general interface for computation.
+An arrow \code{arr a b} can be look at as a computation that converts an input \code{a} to an output \code{b}.
+Furthermore, he defined the arrow typeclass as follows:
 \begin{lstlisting}[frame=htrbl]
@@
 This can be thought of as analogous to the monadic \lstinline{return}.
-The \lstinline{>>>} operator, in a similar way, is analogous to the monadic composition operator \lstinline{>>=}.
-Ant lastly, the \lstinline{first} operator, which takes the input arrow from b to c and converts it into an arrow on pairs with the second argument untouched, is also needed for actual useful code as without it, we wouldn't have a way to save input across arrows.
+The \lstinline{>>>} operator, in a similar way, is analogous to the monadic composition operator \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}.
+And lastly, the \lstinline{first} operator, which takes the input arrow from b to c and converts it into an arrow on pairs with the second argument untouched, is also needed for actual useful code as without it, we wouldn't have a way to save input across arrows.
 \\\\

### 2017-01-18 paper/parallelHaskells.tex
minor

@@
 \section{Short introduction to parallel Haskells}
-There are already several ways write parallel programs in Haskell.
+There are already several ways to write parallel programs in Haskell.
 As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
@@
 For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
-This lazy list \code{[b]} is then evaluated in parallel with the \code{using} operator and a strategy \code{Strategy [b]}.
+This lazy list \code{[b]} is then forcibly evaluated in parallel with the strategy \code{Strategy [b]} by the \code{using} operator.
 This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.
@@
 \subsection{ParMonad}
-The \code{Par} monad introduced by Marlow et al \cite{monad_par_paper}, which can be found in the monad-par package on hackage \cite{monad_par_hackage}, is a monad designed for composition of parallel programs.
+The \code{Par} monad introduced by Marlow et al.
+\cite{monad_par_paper_2011}, which can be found in the monad-par package on hackage \cite{monad_par_hackage}, is a monad designed for composition of parallel programs.
 \\\\
@@
 \subsection{Eden}
-Eden \cite{eden_hackage} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backend.
+Eden  is a parallel Haskell for distributed memory and comes with a MPI and a PVM backend \cite{eden_hackage, Loogen2012, eden_homepage}.
 This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell.

### 2017-01-19 paper/arrows.tex
minor
wip on the paper, added some frege tests. all lazy versions have to be looked at

@@
 An arrow \code{arr a b} can be look at as a computation that converts an input \code{a} to an output \code{b}.
-Furthermore, he defined the arrow typeclass as follows:
+This is defined in the arrow typeclass:
 \begin{lstlisting}[frame=htrbl]
@@
 The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
-Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can immediately used together with every library that works on arrows.
+Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
 Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
@@
 \textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used.
-We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.
-We kept the original notation \code{a b c} for this section to not change too much from Hughes' original definitions.}
+We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}

### 2017-01-19 paper/conclusion.tex
wip on the paper, added some frege tests. all lazy versions have to be looked at

@@
 In its current state this will end up in the root node being a possible bottleneck on clusters when parallel arrows are combined with \code{>>>}.
-This can probably be remedied by introducing some sort of a Future construct that can be passed between nodes so that nodes can communicate without going through the root node.
+This can probably be remedied by introducing some sort of a Java-like Future construct that can be passed between nodes so that nodes can communicate without going through the root node.
 This will have to be implemented in a follow up paper, though.

### 2017-01-19 paper/parrows.tex
wip on the paper, added some frege tests. all lazy versions have to be looked at

--- 2017-01-12 paper/parrows.tex
@@
 \section{Parallel Arrows}
-We have seen what Arrows are and how they can be seen as a general interface to computation.
-In the following section we will discuss how Arrows can also be seen as a general interface not only to computation, but to \textbf{parallel computation} as well.
+We have seen what Arrows are and how they can be used as a general interface to computation.
+In the following section we will discuss how Arrows can also be looked at as a general interface not only to computation, but to \textbf{parallel computation} as well.
 We start by introducing the interface and explaining the reasonings behind it.
@@
 With this definition of \code{parEvalN}, parallel execution can be looked at as yet another arrow combinator.
-But as the implementation may differ depending on the actuall type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
+But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-Sometimes parallel Haskells require additional configuration parameters for information about the execution environment.
+Sometimes parallel Haskells require additional configuration parameters as information about the execution environment.
 This is why we also introduce an additional \code{conf} parameter to the function.
-We also do not want \code{conf} to be of a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}, so we add it to the type signature of the typeclass as well.
+We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}.
+So we add it to the type signature of the typeclass as well.
 \begin{lstlisting}[frame=htrbl]
@@
 We don't require the conf parameter in every implementation.
-If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN} as we will see in the implementation sections.
+If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN} as we will see in the implementation of the Multicore and the ParMonad backend.
 
 \subsection{Multicore Haskell}
-The Multicore Haskell implementation of this class is straightforward using listApp from \ref{utilfns} combined with the \code{using} operator from Multicore Haskell.
+The Multicore Haskell implementation of this class is straightforward using listApp from chapter \ref{utilfns} combined with the \code{using} operator from Multicore Haskell.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-We hardcode the \code{parList rdeepseq} strategy here as in this context it is the only one making sense as we usually want the output list to be fully evaluated to its normal form.
+We hardcode the \code{parList rdeepseq} strategy here, as in this context it is the only one making sense, since we usually want the output list to be fully evaluated to its normal form.
 
 \subsection{ParMonad}
-The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: a -> Par (IVar a)} function which forks away the computation of a value and returns an IVar containing the result in the Par monad.
+The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function, which forks away the computation of a value and returns an IVar containing the result in the Par monad.
 \\\\
@@
 \subsection{Eden}
-For the Multicore and ParMonad implementation we could use general implementations that just required the \code{ArrowApply} and \code{ArrowChoice} typeclasses.
+For the Multicore and ParMonad implementation we could use general instances of \code{ArrowParallel} that just require the \code{ArrowApply} and \code{ArrowChoice} typeclasses.
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
@@
 \end{lstlisting}
-we don't do this in this paper, as this seems too hacky. For now, we just implement \code{ArrowParallel} for normal functions
+, we don't do it in this paper, as this seems too hacky. For now, we just implement \code{ArrowParallel} for normal functions
 \begin{lstlisting}[frame=htrbl]

### 2017-01-19 paper/utilityFunctions.tex
wip on the paper, added some frege tests. all lazy versions have to be looked at

@@
 \\
-The map combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
+The map combinator, that lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and zipWith lift any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+and zipWith, which lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
 \begin{lstlisting}[frame=htrbl]

### 2017-01-26 paper/arrows.tex
wip on presentation, minor fix to paper

--- 2017-01-14 paper/arrows.tex
@@
 \end{lstlisting}
-and the Kleisli type.
+and the Kleisli type
 \begin{lstlisting}[frame=htrbl]
-instance Arrow (Kleisli m) where
+data Kleisli m a b = Kleisli { run :: a -> m b }
+\end{lstlisting}
+as well:
+\begin{lstlisting}[frame=htrbl]
+instance Monad m => Arrow (Kleisli m) where
 	arr f = Kleisli $ return . f

### 2017-01-26 paper/utilityFunctions.tex
wip on presentation, minor fix to paper

--- 2017-01-13 paper/utilityFunctions.tex
@@
 \\
-The map combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
+The mapArr combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and zipWith lift any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+and zipWithArr lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
 \begin{lstlisting}[frame=htrbl]

### 2017-01-26 presentation/utilityFunctions.tex
wip on presentation, minor fix to paper

--- 2017-01-13 paper/utilityFunctions.tex
@@
 \section{Utility Functions}\label{utilfns}
-Before we go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: map and zipWith on arrows.
-\\
-The map combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
+\subsection{mapArr}
+The mapArr combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and zipWith lift any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+with
+\begin{lstlisting}[frame=htrbl]
+(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
+\end{lstlisting}
+%It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
+\frbreak
+
+\subsection{zipWithArr}
+zipWithArr lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-These two combinators make use of the \code{ArrowChoice} typeclass, which allows us to use the \code{|||} combinator.
-It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
+\frbreak
+
+\subsection{listApp}
 \begin{lstlisting}[frame=htrbl]
-(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
-\end{lstlisting}
-With the zipWithArr combinator we can also write a combinator \code{listApp}, which lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]}.
-\begin{lstlisting}[frame=htrbl]
-listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
+listApp :: (ArrowChoice arr, ArrowApply arr) =>
+	[arr a b] -> arr [a] [b]
 listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app

### 2017-01-26 paper/benchmarks.tex
wip

@@
 \section{Benchmarks}
-To check performance of our new parallel arrow API, we conducted some benchmarks.
+To check the performance of our new parallel arrow API, we conducted some benchmarks.
 These are based on a sudoku solver from the examples for Simon Marlows book "Parallel and Concurrent Programming in Haskell" which can be found on his GitHub \cite{marlow_parconc}.
@@
 \\\\
-The Benchmarks were run on a Core i7-3970X CPU @ 3.5GHz with 6 cores and 12 threads. All versions were run 100 times. The resulting running-times were then averaged.
-\\
-\begin{tikzpicture}
+The Benchmarks were run on a Core i7-3970X CPU @ 3.5GHz with 6 cores and 12 threads.
+For sake of comparability with Simon Marlow's parallel version which uses the ParMonad, we use the ParMonad backend for the parallel arrow versions as well.
+\\\\
+\framebox{
+	\begin{tikzpicture}[thick, scale=0.7]
 	\begin{axis}[
@@
 	\end{tikzpicture}
+}
+\\\\
 \textit{Note: the bad result for parmonad-sudoku-parmap seems to be a artefact as it performs much better in the other benchmarks}
 \\\\
-\begin{tikzpicture}
+\framebox{
+	\begin{tikzpicture}[thick, scale=0.7]
 	\begin{axis}[
@@
 	\end{tikzpicture}
-\begin{tikzpicture}
+}
+\\\\
+\framebox{
+	\begin{tikzpicture}[thick, scale=0.7]
 	\begin{axis}[
@@
 	\end{tikzpicture}
+}
 \\\\
 As we can see, the parallel arrow versions of the program all have around the same speedup as the ParMonad based version.
-This means that using the \code{ArrowParallel} typeclass doesn't add any real overhead.
+This means that using the \code{ArrowParallel} typeclass doesn't add any real notable overhead.
 Also, the slightly worse results for "parrows-sudoku-farm" and "parrows-sudoku-farmChunk" can be explained by either imperfect parameters or by the fact that the benchmarks do not really require the scheduling introduced by the skeletons which in this case introduces unnecessary overhead.

### 2017-01-26 paper/parrows.tex

@@
 This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
-So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster using an Eden built, by just swapping out the actual \code{ArrowParallel} instance.
+So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just swapping out the actual \code{ArrowParallel} instance.
 \end{itemize}

### 2017-01-26 paper/utilityFunctions.tex

--- 2017-01-13 paper/utilityFunctions.tex
@@
 \\
-The map combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
+The mapArr combinator, that lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and zipWith lift any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+and zipWithArr, which lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
 \begin{lstlisting}[frame=htrbl]

### 2017-01-26 presentation/benchmarks.tex
copied all the relevant content from the paper to the presentation

--- 2017-01-26 paper/benchmarks.tex
@@
 \section{Benchmarks}
-To check the performance of our new parallel arrow API, we conducted some benchmarks.
-These are based on a sudoku solver from the examples for Simon Marlows book "Parallel and Concurrent Programming in Haskell" which can be found on his GitHub \cite{marlow_parconc}.
-The results are displayed in the following graphs which are ordered by problem size.
-\\\\
 The Benchmarks were run on a Core i7-3970X CPU @ 3.5GHz with 6 cores and 12 threads.
 For sake of comparability with Simon Marlow's parallel version which uses the ParMonad, we use the ParMonad backend for the parallel arrow versions as well.
-\\\\
+\frbreak
 \framebox{
@@
 }
-\\\\
-\textit{Note: the bad result for parmonad-sudoku-parmap seems to be a artefact as it performs much better in the other benchmarks}
-\\\\
+\frbreak
 \framebox{
@@
 }
-\\\\
+\frbreak
 \framebox{
@@
 }
-\\\\
-As we can see, the parallel arrow versions of the program all have around the same speedup as the ParMonad based version.
-This means that using the \code{ArrowParallel} typeclass doesn't add any real notable overhead.
-Also, the slightly worse results for "parrows-sudoku-farm" and "parrows-sudoku-farmChunk" can be explained by either imperfect parameters or by the fact that the benchmarks do not really require the scheduling introduced by the skeletons which in this case introduces unnecessary overhead.
-This would probably look different if the benchmark would use e.g.
-a divide and conquer approach to solve the sudoku puzzles.

### 2017-01-26 presentation/parrows.tex
copied all the relevant content from the paper to the presentation

--- 2017-01-12 paper/parrows.tex
@@
 \section{Parallel Arrows}
-We have seen what Arrows are and how they can be seen as a general interface to computation.
-In the following section we will discuss how Arrows can also be seen as a general interface not only to computation, but to \textbf{parallel computation} as well.
-We start by introducing the interface and explaining the reasonings behind it.
-Then, we discuss some implementations using exisiting parallel Haskells.
-Finally, we explain why using Arrows for expressing parallelism is beneficial.
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, that, in its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function listApp, but does parallel instead of serial evaluation).
+
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-With this definition of \code{parEvalN}, parallel execution can be looked at as yet another arrow combinator.
-But as the implementation may differ depending on the actuall type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
+
+\frbreak
+
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-Sometimes parallel Haskells require additional configuration parameters for information about the execution environment.
-This is why we also introduce an additional \code{conf} parameter to the function.
-We also do not want \code{conf} to be of a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}, so we add it to the type signature of the typeclass as well.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-We don't require the conf parameter in every implementation.
-If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN} as we will see in the implementation sections.
+
+\frbreak
 
 \subsection{Multicore Haskell}
-The Multicore Haskell implementation of this class is straightforward using listApp from \ref{utilfns} combined with the \code{using} operator from Multicore Haskell.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-We hardcode the \code{parList rdeepseq} strategy here as in this context it is the only one making sense as we usually want the output list to be fully evaluated to its normal form.
 
+\frbreak
 \subsection{ParMonad}
-The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function which forks away the computation of a value and returns an IVar containing the result in the Par monad.
-\\\\
-We therefore apply each function to its corresponding input value with \code{app} and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call.
-This yields a list \code{[Par (IVar b)]}, which we then convert into \code{Par [IVar b]} with \code{arr sequenceA}.
-In order to wait for the computation to finish, we map over the \code{IVar}s inside the ParMonad with \code{arr (>>= mapM get)}.
-The result of this operation is a \code{Par [b]} from which we can finally remove the monad again by running \code{arr runPar} to get our output of \code{[b]}.
 \begin{lstlisting}[frame=htrbl]
@@
 
+\frbreak
 \subsection{Eden}
-For the Multicore and ParMonad implementation we could use general implementations that just required the \code{ArrowApply} and \code{ArrowChoice} typeclasses.
-With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a typeclass like
-\begin{lstlisting}[frame=htrbl]
-class (Arrow arr) => ArrowUnwrap arr where
-	arr a b -> (a -> b)
-\end{lstlisting}
-we don't do this in this paper, as this seems too hacky. For now, we just implement \code{ArrowParallel} for normal functions
 \begin{lstlisting}[frame=htrbl]
@@
 
+\begin{lstlisting}[frame=htrbl]
+class (Arrow arr) => ArrowUnwrap arr where
+arr a b -> (a -> b)
+\end{lstlisting}
+
+\frbreak
+
 \subsection{Benefits of parallel Arrows}
-We have seen that we can wrap parallel Haskells inside of the \code{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
-\begin{itemize}
-	\item \textbf{Arrow API benefits}:
-	With the \code{ArrowParallel} typeclass we do not lose any benefits of using arrows as \code{parEvalN} is just yet another arrow combinator.
-The resulting arrow can be used in the same way its serial version can be used.
-This is a big advantage of this approach, especially compared to the monad solutions as all arrow combinators will continue to work instead of requiring special ones for each parallel monad.
-	\item \textbf{Abstraction}:
-	With the \code{ArrowParallel} typeclass, we abstracted all parallel implementation logic away from the business logic.
-This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
-So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster using an Eden built, by just swapping out the actual \code{ArrowParallel} instance.
-\end{itemize}

### 2017-01-26 presentation/skeletons.tex
copied all the relevant content from the paper to the presentation

--- 2017-01-13 paper/skeletons.tex
@@
 \subsection{parEvalNLazy}
-\code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
-Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
-\code{map (arr .
-(+)) [1..]} or just because we need the arrows evaluated in chunks.
-\code{parEvalNLazy} fixes this.
-It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{(arr \$ chunksOf chunkSize)}.
-These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}.
-The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\frbreak
 
 \subsection{parEval2}
-We have only talked about parallelizing the computation of arrows of the same type up until now.
-But sometimes we want to paralellize heterogenous types as well.
-For this, we introduce a helper combinator \code{arrMaybe} first, that converts an arrow \code{arr a b} to an arrow \code{arr (Maybe a) (Maybe b)}.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-With this, we can now easily write \code{parEval2} which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)}.
-We start by converting both arrows with \code{arrMaybe}, combining them with \code{***} into a new arrow \code{arr (Maybe a, Maybe c) (Maybe b, Maybe d)}.
-This is then replicated twice and fed into parEvalN to get a \code{arr [(Maybe a, Maybe c)] [(Maybe b, Maybe d)]}.
-We can then apply this arrow to the input \code{[(Just a, Nothing), (Nothing, Just c)]} and then extract the resulting values with \code{fromJust} and the \code{!!} operator on lists in the last step.
+\frbreak
+
 \begin{lstlisting}[frame=htrbl]
@@
 
+\frbreak
 \subsection{parMap}
-\code{parMap} is probably the most common skeleton for parallel programs.
-We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}.
-Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
 \begin{lstlisting}[frame=htrbl]
@@
 
+\frbreak
+
 \subsection{parMapStream}
-As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
-So it makes sense to also have a \code{parMapStream} which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
 \begin{lstlisting}[frame=htrbl]
@@
 
+\frbreak
 \subsection{farm}
-\code{parMap} spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
-This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
 \begin{lstlisting}[frame=htrbl]
@@
 
+\frbreak
+
 \subsection{farmChunk}
-As \code{farm} is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
-So we define \code{farmChunk} which uses \code{parEvalNLazy} instead of \code{parEvalN} like this:
 \begin{lstlisting}[frame=htrbl]

### 2017-01-28 paper/abstract.tex
minor changes in the paper, wip on the presentation

@@
 Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
-In the paper Hughes describes that arrows have one powerful additional property when compared to monads for API design: Extensibility.
-In this paper we will show how this property can be used to add parallelism capabilities to different arrow-types.
+In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
+In this paper we will use this concept to express parallelism.
 \\\\

### 2017-01-28 paper/conclusion.tex
minor changes in the paper, wip on the presentation

@@
 As we have seen in this paper, arrows are a useful tool for composing parallel programs.
-We have already seen that they have some benefits over existing parallel Haskells as they are, by nature, more extensible.
+By basing our parallel Haskell on them, we do not have to introduce new monadic types that wrap the computation and instead can use them just like they were regular sequential pure code.
 Performancewise, parallel arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.

### 2017-01-28 paper/parrows.tex
minor changes in the paper, wip on the presentation

@@
 	With the \code{ArrowParallel} typeclass we do not lose any benefits of using arrows as \code{parEvalN} is just yet another arrow combinator.
-The resulting arrow can be used in the same way its serial version can be used.
-This is a big advantage of this approach, especially compared to the monad solutions as all arrow combinators will continue to work instead of requiring special ones for each parallel monad.
+The resulting arrow can be used in the same way a potential serial version could be used.
+This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
+We can just "plug" in parallel parts into our sequential programs without having to change anything.
 	\item \textbf{Abstraction}:

### 2017-01-28 presentation/utilityFunctions.tex
minor changes in the paper, wip on the presentation

@@
 
-\subsection{zipWithArr}
-zipWithArr lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+\subsection{zipWithArr \& listApp}
+zipWithArr lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}:
 \begin{lstlisting}[frame=htrbl]
 zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
-zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
+zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>>
+	mapArr f
 \end{lstlisting}
-\frbreak
-
-\subsection{listApp}
+listApp converts a list of arrows \code{[arr a b]} to a new arrow \code{arr [a] [b]}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-This combinator also makes use of the \code{ArrowApply} typeclass which allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
+with the \code{ArrowApply} that defines \code{app :: arr (arr a b, a) c}.

### 2017-01-30 presentation/parrows.tex
wip on presentation

@@
-\section{Parallel Arrows}
-\subsection{The ArrowParallel typeclass}
+\subsection{Generalization to Arrows}
+
+\begin{frame}[fragile]{The ArrowParallel typeclass}
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-
-\frbreak
 
@@
 
-\frbreak
-
-\subsection{Multicore Haskell}
-\begin{lstlisting}[frame=htrbl]
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b conf where
-		parEvalN _ fs = listApp fs >>> arr (flip using $ parList rdeepseq)
-\end{lstlisting}
-
-\frbreak
-\subsection{ParMonad}
-\begin{lstlisting}[frame=htrbl]
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b conf where
-		parEvalN _ fs = 
-			(arr $ \as -> (fs, as)) >>>
-			zipWithArr (app >>> arr spawnP) >>>
-			arr sequenceA >>>
-			arr (>>= mapM get) >>>
-			arr runPar
-\end{lstlisting}
-
-\frbreak
-\subsection{Eden}
-\begin{lstlisting}[frame=htrbl]
-instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
-parEvalN _ fs as = spawnF fs as
-\end{lstlisting}
-and the Kleisli type.
-\begin{lstlisting}[frame=htrbl]
-instance (Monad m, Trans a, Trans b, Trans (m b)) =>
-	ArrowParallel (Kleisli m) a b conf where
-parEvalN conf fs =
-	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
-	(Kleisli $ sequence)
-\end{lstlisting}
-
-\begin{lstlisting}[frame=htrbl]
-class (Arrow arr) => ArrowUnwrap arr where
-arr a b -> (a -> b)
-\end{lstlisting}
-
-\frbreak
-
-\subsection{Benefits of parallel Arrows}
+\end{frame}

### 2017-01-30 presentation/utilityFunctions.tex
wip on presentation

@@
-\section{Utility Functions}\label{utilfns}
-\subsection{mapArr}
+\subsection{Utility Functions}\label{utilfns}
+\begin{frame}[fragile]{mapArr}
 The mapArr combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
@@
 %It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
-\frbreak
+\end{frame}
 
-\subsection{zipWithArr \& listApp}
+\begin{frame}[fragile]{zipWithArr \& listApp}
 zipWithArr lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}:
@@
 with the \code{ArrowApply} that defines \code{app :: arr (arr a b, a) c}.
+\end{frame}

### 2017-01-31 presentation/benchmarks.tex
wip on presentation
wip presentation

@@
 \section{Benchmarks}
-The Benchmarks were run on a Core i7-3970X CPU @ 3.5GHz with 6 cores and 12 threads.
-For sake of comparability with Simon Marlow's parallel version which uses the ParMonad, we use the ParMonad backend for the parallel arrow versions as well.
-\frbreak
+\begin{frame}
+\begin{itemize}
+	\item Run on: Core i7-3970X CPU @ 3.5GHz / 6C/12T.
+	\item compiled with ParMonad backend
+	\item used Sudoku Benchmark from ParMonad examples
+\end{itemize}
+\end{frame}
+\begin{frame}[fragile]
 \framebox{
@@
 }
-\frbreak
+\end{frame}
+\begin{frame}[fragile]
 \framebox{
@@
 }
+\end{frame}

### 2017-01-31 presentation/skeletons.tex
wip on presentation
wip presentation

@@
-\section{Skeletons}
-With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.
+\section{Usability}
+\subsection{Skeletons}
+\begin{frame}[fragile]{parEvalNLazy}
+\begin{lstlisting}[frame=htrbl]
+parEvalNLazy :: conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
+\end{lstlisting}
+\end{frame}
 
-\subsection{parEvalNLazy}
+\begin{frame}[fragile]{parEval2}
 \begin{lstlisting}[frame=htrbl]
-parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
-parEvalNLazy conf chunkSize fs =
-	arr (chunksOf chunkSize) >>>
-	listApp fchunks >>>
-	arr concat
-	where
-		fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
+parEval2 :: conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
 \end{lstlisting}
-\frbreak
+\end{frame}
 
-\subsection{parEval2}
+\begin{frame}[fragile]{parMap}
 \begin{lstlisting}[frame=htrbl]
-arrMaybe :: (ArrowApply arr) => (arr a b) -> arr (Maybe a) (Maybe b)
-arrMaybe fn = (arr $ go) >>> app
-	where 
-		go Nothing = (arr $ \Nothing -> Nothing, Nothing)
-		go (Just a) = ((arr $ \(Just x) -> (fn, x)) >>> app >>> arr Just, (Just a))
+parMap :: conf -> (arr a b) -> (arr [a] [b])
 \end{lstlisting}
-\frbreak
+\end{frame}
 
+\begin{frame}[fragile]{parMapStream}
 \begin{lstlisting}[frame=htrbl]
-parEval2 :: (ArrowParallel arr a b conf,
-	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf,
-	ArrowApply arr) =>
-	conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
-parEval2 conf f g =
-	(arr $ \(a, c) -> (f_g, [(Just a, Nothing), (Nothing, Just c)])) >>>
-	app >>>
-	(arr $ \comb -> (fromJust (fst (comb !! 0)), fromJust (snd (comb !! 1))))
-where
-	f_g = parEvalN conf $ replicate 2 $ arrMaybe f *** arrMaybe g
+parMapStream :: conf -> ChunkSize -> arr a b -> arr [a] [b]
 \end{lstlisting}
+\end{frame}
 
-\frbreak
-\subsection{parMap}
+\begin{frame}[fragile]{farm}
 \begin{lstlisting}[frame=htrbl]
-parMap :: (ArrowParallel arr a b conf, ArrowApply arr) =>
-	conf -> (arr a b) -> (arr [a] [b])
-parMap conf f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr repeat >>>
-		arr (parEvalN conf)) >>>
-	app
+farm :: conf -> NumCores -> arr a b -> arr [a] [b]
 \end{lstlisting}
+\end{frame}
 
-\frbreak
-
-\subsection{parMapStream}
+\begin{frame}[fragile]{farmChunk}
 \begin{lstlisting}[frame=htrbl]
-parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> arr a b -> arr [a] [b]
-parMapStream conf chunkSize f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr repeat >>>
-		arr (parEvalNLazy conf chunkSize)) >>>
-	app
+farmChunk ::
+	conf -> ChunkSize -> NumCores -> arr a b-> arr [a] [b]
 \end{lstlisting}
-
-\frbreak
-\subsection{farm}
-\begin{lstlisting}[frame=htrbl]
-farm :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
-	ArrowChoice arr, ArrowApply arr) =>
-	conf -> NumCores -> arr a b -> arr [a] [b]
-farm conf numCores f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr mapArr >>> arr repeat >>>
-		arr (parEvalN conf)) >>>
-	(second $ arr (unshuffle numCores)) >>>
-	app >>>
-	arr shuffle
-\end{lstlisting}
-The definition of \code{unshuffle} is
-\begin{lstlisting}[frame=htrbl]
-unshuffle :: Int
-	-> [a]
-	-> [[a]]
-unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
-\end{lstlisting}
-, while \code{shuffle} is defined as:
-\begin{lstlisting}[frame=htrbl]
-shuffle :: [[a]]
-	-> [a]
-shuffle = concat . transpose
-\end{lstlisting}
-(These were taken from Eden's source code. \cite{eden_skel_shuffle})
-
-\frbreak
-
-\subsection{farmChunk}
-\begin{lstlisting}[frame=htrbl]
-farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
-	ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
-farmChunk conf chunkSize numCores f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr mapArr >>> arr repeat >>>
-		arr (parEvalNLazy conf chunkSize)) >>>
-	(second $ arr (unshuffle numCores)) >>>
-	app >>>
-	arr shuffle
-\end{lstlisting}
+\end{frame}

### 2017-02-01 presentation/skeletons.tex
wip presentation
added diagrams
diagrams
minor
reduced length of presentation

@@
 \subsection{Skeletons}
-\begin{frame}[fragile]{parEvalNLazy}
+\begin{frame}[fragile]{Skeletons... (1)}
+parEvalN, but \textbf{chunky}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-\end{frame}
-
-\begin{frame}[fragile]{parEval2}
+\begin{center}
+\includegraphics[scale=0.5]{images/parEvalNLazy}
+\end{center}
+parallel evaluation of \textbf{different typed functions}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\begin{center}
+\includegraphics[scale=0.5]{images/parEval2}
+\end{center}
 \end{frame}
-
-\begin{frame}[fragile]{parMap}
+\begin{frame}[fragile]{Skeletons... (2)}
+map, but in \textbf{parallel}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\begin{center}
+\includegraphics[scale=0.5]{images/parMap}
+\end{center}
+%parMap, but \textbf{chunky}:
+%\begin{lstlisting}[frame=htrbl]
+%parMapStream :: conf -> ChunkSize -> arr a b -> arr [a] [b]
+%\end{lstlisting}
+%\begin{center}
+%\includegraphics[scale=0.5]{images/parMapStream}
+%\end{center}
+\begin{center}
+	and others...
+\end{center}
 \end{frame}
-
-\begin{frame}[fragile]{parMapStream}
-\begin{lstlisting}[frame=htrbl]
-parMapStream :: conf -> ChunkSize -> arr a b -> arr [a] [b]
-\end{lstlisting}
-\end{frame}
-
-\begin{frame}[fragile]{farm}
-\begin{lstlisting}[frame=htrbl]
-farm :: conf -> NumCores -> arr a b -> arr [a] [b]
-\end{lstlisting}
-\end{frame}
-
-\begin{frame}[fragile]{farmChunk}
-\begin{lstlisting}[frame=htrbl]
-farmChunk ::
-	conf -> ChunkSize -> NumCores -> arr a b-> arr [a] [b]
-\end{lstlisting}
-\end{frame}
+%\begin{frame}[fragile]{Skeletons... (3)}
+%parMap, but with \textbf{workload distribution}:
+%\begin{lstlisting}[frame=htrbl]
+%farm :: conf -> NumCores -> arr a b -> arr [a] [b]
+%\end{lstlisting}
+%\begin{center}
+%\includegraphics[scale=0.5]{images/farm}
+%\end{center}
+%farm, but \textbf{chunky}:
+%\begin{lstlisting}[frame=htrbl]
+%farmChunk ::
+%	conf -> ChunkSize -> NumCores -> arr a b-> arr [a] [b]
+%\end{lstlisting}
+%\begin{center}
+%\includegraphics[scale=0.5]{images/farmChunk}
+%\end{center}
+%\end{frame}

### 2017-02-01 presentation/skeletons.tex
minor

@@
 \begin{lstlisting}[frame=htrbl]
-parEvalNLazy :: conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
+parEvalNLazy :: conf -> ChunkSize -> [arr a b] -> arr [a] [b]
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-parEval2 :: conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
+parEval2 :: conf -> arr a b -> arr c d -> arr (a, c) (b, d)
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-parMap :: conf -> (arr a b) -> (arr [a] [b])
+parMap :: conf -> arr a b -> arr [a] [b]
 \end{lstlisting}

### 2017-02-20 paper/motivation.tex
minor

@@
 \section{Motivation}
-Arrows were introduced in John Hughes paper as an alternative to monads for API design \citHughes.
-In the paper Hughes describes that arrows have one powerful additional property when compared to monads for API design: Extensibility.
-In this paper we will show how this property can be used to add parallelism capabilities to different arrow-types.
+Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
+In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
+In this paper we will use this concept to express parallelism.

### 2017-02-20 presentation/utilityFunctions.tex

@@
 \subsection{Utility Functions}\label{utilfns}
-\begin{frame}[fragile]{mapArr}
-The mapArr combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
+\begin{frame}[fragile]{mapArr, zipWithArr, listApp}
+From \cite{programming_with_arrows}, we have:
 \begin{lstlisting}[frame=htrbl]
 mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
-mapArr f =
-	arr listcase >>>
-	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
-	where
-		listcase [] = Left ()
-		listcase (x:xs) = Right (x,xs)
 \end{lstlisting}
-with
-\begin{lstlisting}[frame=htrbl]
-(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
-\end{lstlisting}
-%It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
-\end{frame}
 
-\begin{frame}[fragile]{zipWithArr \& listApp}
 zipWithArr lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}:

### 2017-03-17 paper/parallelHaskells.tex
wip on presentation, minor fix to paper
wip on the paper

@@
 As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
-\\\\
+\begin{center}
+	\includegraphics[scale=0.7]{images/parEvalN}
+\end{center}
 In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
@@
 \end{lstlisting}
+We will now implement \code{parEvalN} with the different parallel Haskells.
 
@@
 This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.
+\begin{center}
+	\includegraphics[scale=0.5]{images/parEvalNMulticore}
+\end{center}
 \begin{lstlisting}[frame=htrbl]
@@
 We finally execute this process with \code{runPar} to finally get \code{[b]} again.
+\begin{center}
+	\includegraphics[scale=0.5]{images/parEvalNParMonad}
+\end{center}
 \begin{lstlisting}[frame=htrbl]
@@
 This allows us to define \code{parEvalN} quite easily:
+\begin{center}
+	\includegraphics[scale=0.5]{images/parEvalNEden}
+\end{center}
 \begin{lstlisting}[frame=htrbl]

### 2017-03-17 paper/topologySkeletons.tex
wip on the paper

\section{Topology Skeletons}
Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons. The Eden library leverages this problem and already comes with a \textit{ring} and a \textit{torus} implementation \cite{eden_cefp, eden_skel_topology}. These two seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that its feasibility.

\subsection{ring}

\subsection{torus}

### 2017-03-17 paper/mapSkeletons.tex
wip on the paper

--- 2017-01-13 paper/skeletons.tex
@@
-\section{Skeletons}
+\section{Map-based Skeletons}
 With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.
 
+\fixme{Extra Kapitel für die Basic Skeletons, vll den Syntactic Sugar dann danach machen statt erst nach den maps}
+
 \subsection{parEvalNLazy}
+\begin{center}
+	\includegraphics[scale=0.7]{images/parEvalNLazy}
+\end{center}
 \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
@@
 \subsection{parEval2}
-We have only talked about parallelizing the computation of arrows of the same type up until now.
+\begin{center}
+	\includegraphics[scale=0.7]{images/parEval2}
+\end{center}
+We have only talked about the paralellization arrows of the same type up until now.
 But sometimes we want to paralellize heterogenous types as well.
@@
 \subsection{parMap}
+\begin{center}
+	\includegraphics[scale=0.7]{images/parMap}
+\end{center}
 \code{parMap} is probably the most common skeleton for parallel programs.
@@
 \subsection{parMapStream}
+\begin{center}
+	\includegraphics[scale=0.7]{images/parMapStream}
+\end{center}
 As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
@@
 \subsection{farm}
+\begin{center}
+	\includegraphics[scale=0.7]{images/farm}
+\end{center}
 \code{parMap} spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
@@
 \subsection{farmChunk}
+\begin{center}
+	\includegraphics[scale=0.7]{images/farmChunk}
+\end{center}
 As \code{farm} is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
@@
 \end{lstlisting}
+
+\subsection{parMapReduce}

### 2017-03-19 paper/arrows.tex
wip on paper. added pictures from the presentation.

--- 2017-01-19 paper/arrows.tex
@@
 \section{Arrows}
+\begin{center}
+	\includegraphics[]{images/arrow.png}
+\end{center}
 Arrows were introduced by Hughes \citHughes ~as a general interface for computation.
@@
 This is defined in the arrow typeclass:
+
+\begin{minipage}{\textwidth}
+\begin{minipage}{0.5\textwidth}
 \begin{lstlisting}[frame=htrbl]
@@
 arr :: (a -> b) -> arr a b
+
 (>>>) :: arr a b -> arr b c -> arr a c
+
 first :: arr a b -> arr (a,c) (b,c)
 \end{lstlisting}
+\end{minipage}
+~~~~
+\begin{minipage}{0.25\textwidth}
+	~\\~\\~\\
+	\includegraphics[scale=0.6]{images/arr}~\\~\\
+	\includegraphics[scale=0.6]{images/compose}~\\~\\
+	\includegraphics[scale=0.6]{images/first}~\\~\\
+\end{minipage}
+\end{minipage}
 \lstinline{arr} is used to lift an ordinary function to an arrow type.
@@
 \end{lstlisting}
+\begin{figure}
+	\centering
+	\begin{tabular}{cc}
+		\subcaptionbox{first\label{1}}{\includegraphics[width = 1.5in]{images/first}} &
+		\subcaptionbox{second\label{2}}{\includegraphics[width = 1.5in]{images/second}} \\
+		\subcaptionbox{(***)\label{1}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+		\subcaptionbox{(\&\&\&)\label{2}}{\includegraphics[width = 1.5in]{images/andandand}}\
+	\end{tabular}
+	\caption{syntactic sugar}
+	\label{3}
+\end{figure}
 With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \lstinline{first}, called \lstinline{second},

### 2017-03-19 paper/basicSkeletons.tex
wip on paper. added pictures from the presentation.

--- 2017-01-12 paper/skeletons.tex
@@
-\section{Skeletons}
+\section{Basic Skeletons}
+
 With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.
@@
 \subsection{parEvalNLazy}
+\begin{center}
+	\includegraphics[scale=0.7]{images/parEvalNLazy}
+\end{center}
 \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
@@
 \end{lstlisting}
-
-\subsection{parMap}
-\code{parMap} is probably the most common skeleton for parallel programs.
-We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}.
-Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
-\begin{lstlisting}[frame=htrbl]
-parMap :: (ArrowParallel arr a b conf, ArrowApply arr) =>
-	conf -> (arr a b) -> (arr [a] [b])
-parMap conf f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr repeat >>>
-		arr (parEvalN conf)) >>>
-	app
-\end{lstlisting}
-
-\subsection{parMapStream}
-As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
-So it makes sense to also have a \code{parMapStream} which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
-\begin{lstlisting}[frame=htrbl]
-parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> arr a b -> arr [a] [b]
-parMapStream conf chunkSize f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr repeat >>>
-		arr (parEvalNLazy conf chunkSize)) >>>
-	app
-\end{lstlisting}
-
-\subsection{farm}
-\code{parMap} spawns every single computation in a new thread (at least for the instances of ArrowParallel we gave in this paper).
-This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
-\begin{lstlisting}[frame=htrbl]
-farm :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
-	ArrowChoice arr, ArrowApply arr) =>
-	conf -> NumCores -> arr a b -> arr [a] [b]
-farm conf numCores f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr mapArr >>> arr repeat >>>
-		arr (parEvalN conf)) >>>
-	(second $ arr (unshuffle numCores)) >>>
-	app >>>
-	arr shuffle
-\end{lstlisting}
-The definition of \code{unshuffle} is
-\begin{lstlisting}[frame=htrbl]
-unshuffle :: Int
-	-> [a]
-	-> [[a]]
-unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
-\end{lstlisting}
-, while \code{shuffle} is defined as:
-\begin{lstlisting}[frame=htrbl]
-shuffle :: [[a]]
-	-> [a]
-shuffle = concat . transpose
-\end{lstlisting}
-(These were taken from Eden's source code.)
-
-\subsection{farmChunk}
-As \code{farm} is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
-So we define \code{farmChunk} which uses \code{parEvalNLazy} instead of \code{parEvalN} like this:
-\begin{lstlisting}[frame=htrbl]
-farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
-	ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
-farmChunk conf chunkSize numCores f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr mapArr >>> arr repeat >>>
-		arr (parEvalNLazy conf chunkSize)) >>>
-	(second $ arr (unshuffle numCores)) >>>
-	app >>>
-	arr shuffle
-\end{lstlisting}

### 2017-03-19 paper/mapSkeletons.tex
wip on paper. added pictures from the presentation.

@@
 \section{Map-based Skeletons}
-With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.
-
-\fixme{Extra Kapitel für die Basic Skeletons, vll den Syntactic Sugar dann danach machen statt erst nach den maps}
-
-\subsection{parEvalNLazy}
-\begin{center}
-	\includegraphics[scale=0.7]{images/parEvalNLazy}
-\end{center}
-\code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
-Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
-\code{map (arr .
-(+)) [1..]} or just because we need the arrows evaluated in chunks.
-\code{parEvalNLazy} fixes this.
-It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{(arr \$ chunksOf chunkSize)}.
-These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}.
-The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
-\begin{lstlisting}[frame=htrbl]
-parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
-parEvalNLazy conf chunkSize fs =
-	arr (chunksOf chunkSize) >>>
-	listApp fchunks >>>
-	arr concat
-	where
-		fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
-\end{lstlisting}
-
-\subsection{parEval2}
-\begin{center}
-	\includegraphics[scale=0.7]{images/parEval2}
-\end{center}
-We have only talked about the paralellization arrows of the same type up until now.
-But sometimes we want to paralellize heterogenous types as well.
-For this, we introduce a helper combinator \code{arrMaybe} first, that converts an arrow \code{arr a b} to an arrow \code{arr (Maybe a) (Maybe b)}.
-\begin{lstlisting}[frame=htrbl]
-arrMaybe :: (ArrowApply arr) => (arr a b) -> arr (Maybe a) (Maybe b)
-arrMaybe fn = (arr $ go) >>> app
-	where 
-		go Nothing = (arr $ \Nothing -> Nothing, Nothing)
-		go (Just a) = ((arr $ \(Just x) -> (fn, x)) >>> app >>> arr Just, (Just a))
-\end{lstlisting}
-With this, we can now easily write \code{parEval2} which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)}.
-We start by converting both arrows with \code{arrMaybe}, combining them with \code{***} into a new arrow \code{arr (Maybe a, Maybe c) (Maybe b, Maybe d)}.
-This is then replicated twice and fed into parEvalN to get a \code{arr [(Maybe a, Maybe c)] [(Maybe b, Maybe d)]}.
-We can then apply this arrow to the input \code{[(Just a, Nothing), (Nothing, Just c)]} and then extract the resulting values with \code{fromJust} and the \code{!!} operator on lists in the last step.
-\begin{lstlisting}[frame=htrbl]
-parEval2 :: (ArrowParallel arr a b conf,
-	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf,
-	ArrowApply arr) =>
-	conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
-parEval2 conf f g =
-	(arr $ \(a, c) -> (f_g, [(Just a, Nothing), (Nothing, Just c)])) >>>
-	app >>>
-	(arr $ \comb -> (fromJust (fst (comb !! 0)), fromJust (snd (comb !! 1))))
-where
-	f_g = parEvalN conf $ replicate 2 $ arrMaybe f *** arrMaybe g
-\end{lstlisting}
-
 \subsection{parMap}

### 2017-03-19 paper/parallelHaskells.tex
wip on paper. added pictures from the presentation.

@@
 As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
+
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
+
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: [a -> b] -> [a] -> [b]
+\end{lstlisting}
 \begin{center}
@@
 \end{center}
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
-\begin{lstlisting}[frame=htrbl]
-parEvalN :: [a -> b] -> [a] -> [b]
-\end{lstlisting}
 We will now implement \code{parEvalN} with the different parallel Haskells.
@@
 This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.
-\begin{center}
-	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-\end{center}
+
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\begin{center}
+	\includegraphics[scale=0.5]{images/parEvalNMulticore}
+\end{center}
 
@@
 We finally execute this process with \code{runPar} to finally get \code{[b]} again.
-\begin{center}
-	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-\end{center}
+
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\begin{center}
+	\includegraphics[scale=0.5]{images/parEvalNParMonad}
+\end{center}
 
@@
 This allows us to define \code{parEvalN} quite easily:
-\begin{center}
-	\includegraphics[scale=0.5]{images/parEvalNEden}
-\end{center}
+
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\begin{center}
+	\includegraphics[scale=0.5]{images/parEvalNEden}
+\end{center}

### 2017-03-19 paper/parrows.tex
wip on the paper
wip on paper. added pictures from the presentation.

@@
 Finally, we explain why using Arrows for expressing parallelism is beneficial.
+\begin{center}
+	\includegraphics[scale=0.7]{images/parEvalN}
+\end{center}
 \subsection{The ArrowParallel typeclass}
@@
 \subsection{Benefits of parallel Arrows}
+\fixme{Okay so?}
+\\
 We have seen that we can wrap parallel Haskells inside of the \code{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?

### 2017-03-20 Definition/src/main/Parrows/Skeletons/Map.hs
minor

@@ -69,6 +69,6 @@ farmChunk conf chunkSize numCores f =
 -- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
 parMapReduceDirect :: (ArrowParallel arr [a] b conf, ArrowApply arr, ArrowChoice arr) => conf -> ChunkSize -> arr a b -> arr (b, b) b -> b -> arr [a] b
 parMapReduceDirect conf chunkSize mapfn foldfn neutral =
-                                   (arr $ chunksOf chunkSize) >>>
+                                   arr (chunksOf chunkSize) >>>
                                    parMap conf (mapArr mapfn >>> foldlArr foldfn neutral) >>>
-                                   (foldlArr foldfn neutral)
+                                   foldlArr foldfn neutral

### 2017-03-20 Definition/src/main/Parrows/Skeletons/Topology.hs

@@ -50,7 +50,7 @@ pipeFut conf fs = resolve (arr $ \(a, outs) -> lazy $ a : outs) (parEvalNFut con
         resolve :: (ArrowApply arr, ArrowLoop arr) => arr (a, b) c -> arr c b -> arr a b
         resolve transform f = loop $ (arr $ \(a, b) -> (b, (f, (transform, (a, b))))) >>> second (second app >>> app)
 
-ring :: (ArrowLoop arr, ArrowApply arr, Future fut r, (ArrowParallel arr (i, fut r) (o, fut r) conf)) =>
+ring :: (ArrowLoop arr, ArrowApply arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
     conf ->
     arr (i, r) (o, r) ->
     arr [i] [o]
@@ -58,8 +58,8 @@ ring conf f = loop $ second (arr rightRotate >>> arr lazy) >>> (arr $ uncurry zi
 
 -- similar to Eden's toRD
 toFut :: (Arrow arr, Future fut r) =>
-        (arr (i, r) (o, r))              -- ^ ring process function
-        -> (arr (i, fut r) (o, fut r))   -- ^ with remote data
+        arr (i, r) (o, r) ->
+        arr (i, fut r) (o, fut r)
 toFut f = (second $ arr get) >>> f >>> (second $ arr put)
 
 torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
@@ -67,8 +67,8 @@ torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
             ArrowParallel arr [(c, fut [a], fut [b])] [(d, fut [a], fut [b])] conf,
             Future fut [a], Future fut [b]) =>
          conf ->
-         arr (c, [a], [b]) (d, [a], [b]) -- ^ node function
-         -> arr [[c]] [[d]]                -- ^ input-output mapping
+         arr (c, [a], [b]) (d, [a], [b]) ->
+         arr [[c]] [[d]]
 torus conf f = loop $ second (arr (map rightRotate) *** arr rightRotate) >>>
                         arr (\ (inss, (inssA, inssB)) -> zipWith3 lazyzip3 inss (lazy inssA) (lazy inssB)) >>>
                         parEvalNM conf (repeat (repeat (ptorus f))) >>>

### 2017-03-20 Eden/examples/futures/withoutFutures.hs

@@ -1,0 +1,38 @@
+{-
+The MIT License (MIT)
+
+Copyright (c) 2016 Martin Braun
+
+Permission is hereby granted, free of charge, to any person obtaining
+a copy of this software and associated documentation files (the
+"Software"), to deal in the Software without restriction, including
+without limitation the rights to use, copy, modify, merge, publish,
+distribute, sublicense, and/or sell copies of the Software, and to
+permit persons to whom the Software is furnished to do so, subject to
+the following conditions:
+
+The above copyright notice and this permission notice shall be included
+in all copies or substantial portions of the Software.
+
+THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY
+CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
+TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
+SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+-}
+module Main where
+
+import Parrows.Definition
+import Parrows.Eden
+import Parrows.Skeletons.Topology
+import Parrows.Future
+
+import Control.DeepSeq
+
+import Control.Arrow
+
+main = print $ length $ (rnf val) `seq` val
+           where
+               val = (parEvalN () (replicate 1000 (+1))) >>> (parEvalN () (replicate 1000 (*3))) $ ([1..1000]::[Int])

### 2017-03-20 paper/mapSkeletons.tex
minor

@@
 \subsection{parMapReduce}
+-- this does not completely adhere to Google's definition of Map Reduce as it
+-- the mapping function does not allow for "reordering" of the output
+-- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
+
+\begin{lstlisting}[frame=htrbl]
+parMapReduceDirect :: (ArrowParallel arr [a] b conf,
+	ArrowApply arr, ArrowChoice arr) =>
+	conf -> ChunkSize -> arr a b -> arr (b, b) b -> b -> arr [a] b
+parMapReduceDirect conf chunkSize mapfn foldfn neutral =
+	arr (chunksOf chunkSize) >>>
+	parMap conf (mapArr mapfn >>> foldlArr foldfn neutral) >>>
+	foldlArr foldfn neutral
+\end{lstlisting}

### 2017-03-20 paper/topologySkeletons.tex
minor

@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with a \textit{ring} and a \textit{torus} implementation \cite{eden_cefp, eden_skel_topology}.
-These two seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that its feasibility.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them the \code{pipe}, \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
+These seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that we can express such skeletons with Arrows as well.
+
+\subsection{pipe}
+
+\begin{lstlisting}[frame=htrbl]
+pipe :: (ArrowLoop arr, ArrowApply arr,
+	ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
+	conf -> [arr a a] -> arr a a
+pipe conf fs = unliftFut $ pipeFut conf fs
+
+pipeFut :: (ArrowLoop arr, ArrowApply arr,
+	ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
+	conf -> [arr a a] -> arr (fut a) (fut a)
+pipeFut conf fs =
+	resolve (arr $ \(a, outs) -> lazy $ a : outs) (parEvalNFut conf fs) >>>
+	arr last
+    where
+        resolve :: (ArrowApply arr, ArrowLoop arr) =>
+					arr (a, b) c -> arr c b -> arr a b
+        resolve transform f =
+					loop $ (arr $ \(a, b) -> (b, (f, (transform, (a, b))))) >>>
+					second (second app >>> app)
+\end{lstlisting}
 
@@
 
+\begin{lstlisting}[frame=htrbl]
+ring :: (ArrowLoop arr, ArrowApply arr, Future fut r, 
+ArrowParallel arr (i, fut r) (o, fut r) conf) =>
+	conf ->
+	arr (i, r) (o, r) ->
+	arr [i] [o]
+ring conf f = loop $ second (arr rightRotate >>> arr lazy) >>>
+	(arr $ uncurry zip) >>> (parMap conf (toFut $ f)) >>> arr unzip
+
+-- similar to Eden's toRD
+toFut :: (Arrow arr, Future fut r) =>
+	arr (i, r) (o, r) ->
+	arr (i, fut r) (o, fut r)
+toFut f = (second $ arr get) >>> f >>> (second $ arr put)
+
+-- from Eden:
+rightRotate    :: [a] -> [a]
+rightRotate [] =  []
+rightRotate xs =  last xs : init xs
+\end{lstlisting}
+
 \subsection{torus}
+
+\begin{lstlisting}[frame=htrbl]
+torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
+    ArrowParallel arr (c, fut [a], fut [b]) (d, fut [a], fut [b]) conf,
+    ArrowParallel arr [(c, fut [a], fut [b])] [(d, fut [a], fut [b])] conf,
+    Future fut [a], Future fut [b]) =>
+	conf ->
+	arr (c, [a], [b]) (d, [a], [b])	->
+	arr [[c]] [[d]]
+torus conf f =
+	loop $ second (arr (map rightRotate) *** arr rightRotate) >>>
+	arr (\ (inss, (inssA, inssB)) ->
+		zipWith3 lazyzip3 inss (lazy inssA) (lazy inssB)) >>>
+	parEvalNM conf (repeat (repeat (ptorus f))) >>>
+	arr (map unzip3) >>> arr unzip3 >>> threetotwo
+
+lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
+lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
+
+ptorus :: (Arrow arr, Future fut [a], Future fut [b]) =>
+          arr (c, [a], [b]) (d, [a], [b]) ->
+          arr (c, fut [a], fut [b]) (d, fut [a], fut [b])
+ptorus f =
+	arr (\ ~(c, fas, fbs) -> (c, get fas, get fbs)) >>>
+	f >>>
+	arr (\ ~(c, as, bs) -> (c, put as, put bs))
+
+threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
+threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
+
+parEvalNM :: (ArrowChoice arr, ArrowApply arr, ArrowParallel arr a b conf,
+	ArrowParallel arr [a] [b] conf) =>
+	conf -> [[arr a b]] -> arr [[a]] [[b]]
+parEvalNM conf = listApp . (map (parEvalN conf))
+\end{lstlisting}

### 2017-03-20 paper/utilityFunctions.tex
minor

--- 2017-01-26 paper/utilityFunctions.tex
@@
 \section{Utility Functions}\label{utilfns}
-Before we go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: map and zipWith on arrows.
+Before we go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: map, foldl and zipWith on arrows.
 \\
-The mapArr combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows},
+The \code{mapArr} combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and zipWithArr lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+Similarly, we can also define \code{foldlArr} that lifts any arrow \code{arr (b, a) b} with a neutral element \code{b} to \code{arr [a] b}:
+\begin{lstlisting}[frame=htrbl]
+foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
+foldlArr f b =
+	arr listcase >>>
+	arr (const b) |||
+		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
+	where
+		listcase [] = Left []
+		listcase (x:xs) = Right (x,xs)
+\end{lstlisting}
+Finally, with the help of \code{mapArr}, we can define zipWithArr, which lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-These two combinators make use of the \code{ArrowChoice} typeclass, which allows us to use the \code{|||} combinator.
+These combinators make use of the \code{ArrowChoice} typeclass, which allows us to use the \code{|||} combinator.
 It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.

### 2017-03-20 paper/futures.tex
minor

\section{Futures}

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
	\caption[without Futures]{without Futures}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{with Futures}
\end{figure}

### 2017-03-23 paper/mapSkeletons.tex
updated definitions in tex files

@@
 \begin{lstlisting}[frame=htrbl]
-parMap :: (ArrowParallel arr a b conf, ArrowApply arr) =>
+parMap :: (ArrowParallel arr a b conf) =>
 	conf -> (arr a b) -> (arr [a] [b])
-parMap conf f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr repeat >>>
-		arr (parEvalN conf)) >>>
-	app
+parMap conf f = parEvalN conf (repeat f)
 \end{lstlisting}
@@
 	conf -> ChunkSize -> arr a b -> arr [a] [b]
-parMapStream conf chunkSize f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr repeat >>>
-		arr (parEvalNLazy conf chunkSize)) >>>
-	app
+parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-farm :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
-	ArrowChoice arr, ArrowApply arr) =>
+farm :: (ArrowParallel arr a b conf,
+	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
 	conf -> NumCores -> arr a b -> arr [a] [b]
 farm conf numCores f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr mapArr >>> arr repeat >>>
-		arr (parEvalN conf)) >>>
-	(second $ arr (unshuffle numCores)) >>>
-	app >>>
-	arr shuffle
+	unshuffle numCores >>>
+	parEvalN conf (repeat (mapArr f)) >>>
+	shuffle
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-unshuffle :: Int
-	-> [a]
-	-> [[a]]
-unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
+unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
+unshuffle n = arr (\xs -> [takeEach n (drop i xs) | i <- [0..n-1]])
+
+takeEach :: Int -> [a] -> [a]
+takeEach n [] = []
+takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-shuffle :: [[a]]
-	-> [a]
-shuffle = concat . transpose
+shuffle :: (Arrow arr) => arr [[a]] [a]
+shuffle = arr (concat . transpose)
 \end{lstlisting}
-(These were taken from Eden's source code. \cite{eden_skel_shuffle})
+(These were taken from Eden's source code \cite{eden_skel_shuffle} and translated into Arrows.)
 
@@
 farmChunk conf chunkSize numCores f =
-	(arr $ \as -> (f, as)) >>>
-	(first $ arr mapArr >>> arr repeat >>>
-		arr (parEvalNLazy conf chunkSize)) >>>
-	(second $ arr (unshuffle numCores)) >>>
-	app >>>
-	arr shuffle
+	unshuffle numCores >>>
+	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
+	shuffle
 \end{lstlisting}

### 2017-03-23 paper/topologySkeletons.tex
updated definitions in tex files

--- 2017-03-17 paper/topologySkeletons.tex
@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with a \textit{ring} and a \textit{torus} implementation \cite{eden_cefp, eden_skel_topology}.
-These two seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that its feasibility.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them the \code{pipe}, \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
+These seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that we can express such skeletons with Arrows as well.
+
+\subsection{pipe}
+
+\begin{lstlisting}[frame=htrbl]
+pipe :: (ArrowLoop arr,
+	ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
+	conf -> [arr a a] -> arr a a
+pipe conf fs = unliftFut (pipeFut conf fs)
+
+pipeFut :: (ArrowLoop arr,
+	ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
+	conf -> [arr a a] -> arr (fut a) (fut a)
+pipeFut conf fs =
+	loop (arr snd &&& (arr (uncurry (:) >>> lazy) >>>
+		parEvalNFut conf fs)) >>>
+	arr last
+\end{lstlisting}
 
@@
 
+\begin{lstlisting}[frame=htrbl]
+ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
+    conf ->
+    arr (i, r) (o, r) ->
+    arr [i] [o]
+ring conf f =
+	loop (second (rightRotate >>> lazy) >>>
+    arr (uncurry zip) >>>
+    parMap conf (second get >>> f >>> second put) >>>
+    arr unzip)
+
+-- from Eden, ported to Arrows:
+rightRotate :: (Arrow arr) => arr [a] [a]
+rightRotate = arr $ \list -> case
+	list of [] -> []
+			xs -> last xs : init xs
+\end{lstlisting}
+
 \subsection{torus}
+
+\begin{lstlisting}[frame=htrbl]
+torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
+            ArrowParallel arr (c, fut [a], fut [b]) (d, fut [a], fut [b]) conf,
+            Future fut [a], Future fut [b]) =>
+         conf ->
+         arr (c, [a], [b]) (d, [a], [b]) ->
+         arr [[c]] [[d]]
+torus conf f =
+	loop (second ((mapArr rightRotate >>> lazy) ***
+			(arr rightRotate >>> lazy)) >>>
+    arr (uncurry3 (zipWith3 lazyzip3)) >>>
+    (arr length >>> arr unshuffle) &&&
+        (shuffle >>> parEvalN conf (repeat (ptorus f))) >>>
+    app >>> arr (map unzip3) >>> arr unzip3 >>> threetotwo)
+
+uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
+uncurry3 f (a, (b, c)) = f a b c
+
+lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
+lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
+
+ptorus :: (Arrow arr, Future fut [a], Future fut [b]) =>
+          arr (c, [a], [b]) (d, [a], [b]) ->
+          arr (c, fut [a], fut [b]) (d, fut [a], fut [b])
+ptorus f =
+	arr (\ ~(c, fas, fbs) -> (c, get fas, get fbs)) >>>
+	f >>>
+	arr (\ ~(c, as, bs) -> (c, put as, put bs))
+
+threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
+threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
+\end{lstlisting}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows}
+	\caption[without Futures]{Matrix Multiplication with a torus (Parrows)}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden}
+	\caption[with Futures]{Matrix Multiplication with a torus (Eden)}
+\end{figure}

### 2017-03-23 paper/topologySkeletons.tex
minor
first draft of the Future chapter

@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons, among them the \code{pipe}, \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them are a \code{pipe}, \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
 These seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that we can express such skeletons with Arrows as well.
@@
 \begin{lstlisting}[frame=htrbl]
-ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
+ring :: (ArrowLoop arr, Future fut r,
+	ArrowParallel arr (i, fut r) (o, fut r) conf) =>
     conf ->

### 2017-03-23 Eden/examples/futures/withFutures.hs
minor

@@ -34,7 +34,7 @@ import Control.Arrow
 
 main = print $ length $ (rnf val) `seq` val
            where
-               val = (parEvalN () (replicate 4 $ put . (+1))) >>> rightRotate >>> (parEvalN () (replicate 4 $ (*3) . get)) $ ([1..16]::[Int])
+               val = (parEvalN () (map (>>> put) $ replicate 4 $ (+1))) >>> rightRotate >>> (parEvalN () (map (get >>>) $ replicate 4 $ (*3))) $ ([1..16]::[Int])
 
 -- from Eden, ported to Arrows:
 rightRotate :: (Arrow arr) => arr [a] [a]

### 2017-03-23 paper/basicSkeletons.tex
minor

@@
 \end{center}
-We have only talked about the paralellization arrows of the same type up until now.
+We have only talked about the paralellization arrows of the same type until now.
 But sometimes we want to paralellize heterogenous types as well.
@@
 \begin{lstlisting}[frame=htrbl]
-arrMaybe :: (ArrowApply arr) => (arr a b) -> arr (Maybe a) (Maybe b)
-arrMaybe fn = (arr $ go) >>> app
-	where 
-		go Nothing = (arr $ \Nothing -> Nothing, Nothing)
-		go (Just a) = ((arr $ \(Just x) -> (fn, x)) >>> app >>> arr Just, (Just a))
+arrMaybe :: (ArrowChoice arr) => (arr a b) -> arr (Maybe a) (Maybe b)
+arrMaybe fn = arr maybeCase >>> arr (const Nothing) ||| (fn >>> arr return)
+	where maybeCase Nothing = Left ()
+			maybeCase (Just val) = Right val
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=htrbl]
-parEval2 :: (ArrowParallel arr a b conf,
-	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf,
-	ArrowApply arr) =>
+parEval2 :: (ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf,
+	ArrowChoice arr) =>
 	conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
 parEval2 conf f g =
-	(arr $ \(a, c) -> (f_g, [(Just a, Nothing), (Nothing, Just c)])) >>>
-	app >>>
-	(arr $ \comb -> (fromJust (fst (comb !! 0)), fromJust (snd (comb !! 1))))
-where
-	f_g = parEvalN conf $ replicate 2 $ arrMaybe f *** arrMaybe g
+   (arr Just &&& arr (const Nothing)) *** 
+			((arr (const Nothing) &&& arr Just) >>> arr return) >>>
+   arr (uncurry (:)) >>>
+   parEvalN conf (replicate 2 (arrMaybe f *** arrMaybe g)) >>>
+   arr unzip >>>
+   (arr catMaybes >>> arr head) *** (arr catMaybes >>> arr head)
 \end{lstlisting}

### 2017-03-23 paper/futures.tex

@@
 \section{Futures}
-
+Consider the following arbitrary, but interesting, parallel arrow combinator:
+\begin{lstlisting}[frame=htrbl]
+someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
+someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
+\end{lstlisting}
+In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
+While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in figure \ref{fig:withoutFutures}.
+This can become a serious bottleneck in heavy threaded applications.
 \begin{figure}[ht]
@@
 	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
-	\caption[without Futures]{without Futures}
+	\caption[without Futures]{Communication without Futures}
+	\label{fig:withoutFutures}
 \end{figure}
+To fix this, we have to introduce a way that enables the nodes to communicate directly with each other.
+Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour.
+But as we want code written against our API to be implementation agnostic, we have to wrap this context.
+We do this with the \code{Future} typeclass:
+\begin{lstlisting}[frame=htrbl]
+class Future fut a | a -> fut where
+    put :: (Arrow arr) => arr a (fut a)
+    get :: (Arrow arr) => arr (fut a) a
+\end{lstlisting}
+As \code{RD} is only type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though:
+\begin{lstlisting}[frame=htrbl]
+data RemoteData a = RD { rd :: RD a }
 
+instance (Trans a) => Future RemoteData a where
+    put = arr (\a -> RD { rd = release a })
+    get = arr rd >>> arr fetch
+\end{lstlisting}
+For ParMonad and Multicore we can use a basic dummy wrapper as we have shared memory:
+\begin{lstlisting}[frame=htrbl]
+data BasicFuture a = BF { val :: a }
+
+instance (NFData a) => Future BasicFuture a where
+    put = arr (\a -> BF { val = a })
+    get = arr val
+\end{lstlisting}
+To fit the \code{ArrowParallel} instances we gave earlier, we also have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
+We need this implementation for our \code{RemoteData} wrapper
+\begin{lstlisting}[frame=htrbl]
+instance NFData (RemoteData a) where
+    rnf x = rnf $ rd x
+instance Trans (RemoteData a)
+\end{lstlisting}
+and the following for the \code{BasicFuture} dummy type:
+\begin{lstlisting}[frame=htrbl]
+instance (NFData a) => NFData (BasicFuture a) where
+    rnf bf = rnf $ val bf
+\end{lstlisting}
+Going back to our communication example we can use this new Future concept in order to enable direct communications between the nodes in the following way:
+\begin{lstlisting}[frame=htrbl]
+someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
+someCombinator fs1 fs2 =
+	parEvalN () (map (>>> put) fs1) >>>
+	rightRotate >>>
+	parEvalN () (map (get >>>) fs2)
+\end{lstlisting}
+This gives us a communication scheme as shown in figure \ref{fig:withFutures}
 \begin{figure}[ht]
@@
 	\includegraphics[width=0.9\textwidth]{images/withFutures}
-	\caption[with Futures]{with Futures}
+	\caption[with Futures]{Communication with Futures}
+	\label{fig:withFutures}
 \end{figure}

### 2017-03-23 paper/parrows.tex

@@
 \end{lstlisting}
-We don't require the conf parameter in every implementation.
-If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN} as we will see in the implementation of the Multicore and the ParMonad backend.
+Note that we don't require the conf parameter in every implementation.
+If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
 

### 2017-03-23 paper/syntacticSugar.tex

@@
 \section{Syntactic Sugar}
-To make using our new API a bit easier, we also introduce some syntactic sugar.
+To make our new API easier to use, we also introduce some syntactic sugar.
 We start with \code{|>>>|}, which is basically \code{>>>} on lists of arrows.
@@
 \begin{lstlisting}[frame=htrbl]
-(|***|) :: (ArrowParallel arr a b (),
-	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) (),
-	ArrowApply arr) =>
+(|***|) :: (ArrowChoice arr,
+	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf) =>
 	arr a b -> arr c d -> arr (a, c) (b, d)
@@
 \begin{lstlisting}[frame=htrbl]
-(|&&&|) :: (ArrowParallel arr a b (),
-	ArrowParallel arr (Maybe a, Maybe a) (Maybe b, Maybe c) (),
-	ArrowApply arr) =>
+(|&&&|) :: (ArrowChoice arr,
+	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf) =>
 	arr a b -> arr a c -> arr a (b, c)

### 2017-03-23 paper/futures.tex
minor

@@
 	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
-	\caption[without Futures]{Communication without Futures}
+	\caption[without Futures]{Communication between 4 threads without Futures}
 	\label{fig:withoutFutures}
@@
 \end{lstlisting}
-This gives us a communication scheme as shown in figure \ref{fig:withFutures}
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in figure \ref{fig:withFutures}.
 \begin{figure}[ht]
@@
 	\includegraphics[width=0.9\textwidth]{images/withFutures}
-	\caption[with Futures]{Communication with Futures}
+	\caption[with Futures]{Communication between 4 threads with Futures}
 	\label{fig:withFutures}

### 2017-03-25 paper/basicSkeletons.tex
minor. reworked parEval2

@@
 But sometimes we want to paralellize heterogenous types as well.
-For this, we introduce a helper combinator \code{arrMaybe} first, that converts an arrow \code{arr a b} to an arrow \code{arr (Maybe a) (Maybe b)}.
+However, we can implement such a \code{parEval2} combinator which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass.
+The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
+\\\\
+We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]}.
+Then, we feed this list into a parallel arrow constructed with \code{parEvalN} that uses \code{f +++ g}  we then feed into a 2-element parallel arrow that we created as described above.
+After the calculation is finished we convert the resulting \code{[Either b d]} into \code{([b] , [d])}.
+The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
 \begin{lstlisting}[frame=htrbl]
-arrMaybe :: (ArrowChoice arr) => (arr a b) -> arr (Maybe a) (Maybe b)
-arrMaybe fn = arr maybeCase >>> arr (const Nothing) ||| (fn >>> arr return)
-	where maybeCase Nothing = Left ()
-			maybeCase (Just val) = Right val
+parEval2 :: (ArrowChoice arr,
+	ArrowParallel arr (Either a c) (Either b d) conf) =>
+	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
+parEval2 conf f g = 
+	arr Left *** (arr Right >>> arr return) >>>
+	arr (uncurry (:)) >>>
+	parEvalN conf (replicate 2 (f +++ g)) >>>
+	arr partitionEithers >>>
+	arr head *** arr head
 \end{lstlisting}
-With this, we can now easily write \code{parEval2} which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)}.
-We start by converting both arrows with \code{arrMaybe}, combining them with \code{***} into a new arrow \code{arr (Maybe a, Maybe c) (Maybe b, Maybe d)}.
-This is then replicated twice and fed into parEvalN to get a \code{arr [(Maybe a, Maybe c)] [(Maybe b, Maybe d)]}.
-We can then apply this arrow to the input \code{[(Just a, Nothing), (Nothing, Just c)]} and then extract the resulting values with \code{fromJust} and the \code{!!} operator on lists in the last step.
-\begin{lstlisting}[frame=htrbl]
-parEval2 :: (ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf,
-	ArrowChoice arr) =>
-	conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
-parEval2 conf f g =
-   (arr Just &&& arr (const Nothing)) *** 
-			((arr (const Nothing) &&& arr Just) >>> arr return) >>>
-   arr (uncurry (:)) >>>
-   parEvalN conf (replicate 2 (arrMaybe f *** arrMaybe g)) >>>
-   arr unzip >>>
-   (arr catMaybes >>> arr head) *** (arr catMaybes >>> arr head)
-\end{lstlisting}

### 2017-03-25 paper/syntacticSugar.tex
minor. reworked parEval2

@@
 \section{Syntactic Sugar}
-To make our new API easier to use, we also introduce some syntactic sugar.
-We start with \code{|>>>|}, which is basically \code{>>>} on lists of arrows.
-\begin{lstlisting}[frame=htrbl]
-(|>>>|) :: (Arrow arr) => [arr a b] -> [arr b c] -> [arr a c]
-(|>>>|) = zipWith (>>>)
-\end{lstlisting}
 For basic arrows, we have the \code{***} combinator which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
@@
 \begin{lstlisting}[frame=htrbl]
-(|***|) :: (ArrowChoice arr,
-	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf) =>
+(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
 	arr a b -> arr c d -> arr (a, c) (b, d)
@@
 \begin{lstlisting}[frame=htrbl]
-(|&&&|) :: (ArrowChoice arr,
-	ArrowParallel arr (Maybe a, Maybe c) (Maybe b, Maybe d) conf) =>
+(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
 	arr a b -> arr a c -> arr a (b, c)

### 2017-03-25 paper/basicSkeletons.tex
added |>>>| as a combinator for pipe

@@
 \code{parEvalNLazy} fixes this.
-It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{(arr \$ chunksOf chunkSize)}.
+It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}.
 These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}.
@@
 \\\\
-We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]}.
-Then, we feed this list into a parallel arrow constructed with \code{parEvalN} that uses \code{f +++ g}  we then feed into a 2-element parallel arrow that we created as described above.
-After the calculation is finished we convert the resulting \code{[Either b d]} into \code{([b] , [d])}.
+We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]} by first tagging the two inputs with \code{Left} and \code{Right} and wrapping the right element in a singleton list with \code{return} so that we can combine them with \code{arr (uncurry (:))}.
+Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above.
+After the calculation is finished we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}.
 The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.

### 2017-03-25 paper/conclusion.tex
added |>>>| as a combinator for pipe

@@
 Performancewise, parallel arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
-While we have seen that the parallel arrows can express computation on clusters, it is noteworthy that in its current state, the parallel arrow API doesn't have as much control over data-flow as for example Eden.
-In its current state this will end up in the root node being a possible bottleneck on clusters when parallel arrows are combined with \code{>>>}.
-This can probably be remedied by introducing some sort of a Java-like Future construct that can be passed between nodes so that nodes can communicate without going through the root node.
-This will have to be implemented in a follow up paper, though.
+
+\fixme{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}

### 2017-03-25 paper/topologySkeletons.tex
added |>>>| as a combinator for pipe

@@
 	arr last
+\end{lstlisting}
+
+\begin{lstlisting}[frame=htrbl]
+(|>>>|) :: (ArrowLoop arr, ArrowChoice arr,
+	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
+	Future fut (([a], [b]), [c])) =>
+	arr a b -> arr b c -> arr a c
+	(|>>>|) f g =
+		arr (\a -> (([a], []), [])) >>>
+		pipe () (replicate 2 $ unify f g) >>> arr snd >>> arr head
+			where
+				unify :: (ArrowChoice arr) =>
+					arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
+				unify f g =
+					(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
+					arr (\((a, b), c) -> ((c, a), b))
 \end{lstlisting}

### 2017-03-25 paper/topologySkeletons.tex
wip on the paper

--- 2017-03-23 paper/topologySkeletons.tex
@@
 
+The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows don't have to reside on the same thread/machine.
+We implement this skeleton using the \code{ArrowLoop} typeclass which enables us to use \code{loop :: arr (a, b) (c, b) -> arr a c}.
+
 \begin{lstlisting}[frame=htrbl]
-pipe :: (ArrowLoop arr,
-	ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
+pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
 	conf -> [arr a a] -> arr a a
-pipe conf fs = unliftFut (pipeFut conf fs)
-
-pipeFut :: (ArrowLoop arr,
-	ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
-	conf -> [arr a a] -> arr (fut a) (fut a)
-pipeFut conf fs =
+pipeSimple conf fs =
 	loop (arr snd &&& (arr (uncurry (:) >>> lazy) >>>
-		parEvalNFut conf fs)) >>>
+		parEvalN conf fs)) >>>
 	arr last
+\end{lstlisting}
+where \code{lazy} is defined as:
+\begin{lstlisting}[frame=htrbl]
+lazy :: (Arrow arr) => arr [a] [a]
+lazy = arr (\ ~(x:xs) -> x : lazy xs)
+\end{lstlisting}
+However, using this definition directly, will result in the master node becoming a potential bottleneck in distributed environments as described in chapter \ref{futures}.
+Therefore, a more sophisticated version that uses Futures internally is a good idea:
+\begin{lstlisting}[frame=htrbl]
+pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf,
+	Future fut a) =>
+	conf -> [arr a a] -> arr a a
+pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
+\end{lstlisting}
+Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of different types to each other.
+\begin{lstlisting}[frame=htrbl]
+pipe2 :: (ArrowLoop arr, ArrowChoice arr,
+	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
+	Future fut (([a], [b]), [c])) =>
+	conf -> arr a b -> arr b c -> arr a c
+pipe2 conf f g = 
+	arr (\a -> (([a], []), [])) >>>
+	pipe conf (replicate 2 $ unify f g) >>>
+	arr snd >>>
+	arr head
+		where
+			unify :: (ArrowChoice arr) =>
+				arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
+			unify f g =
+				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
+				arr (\((a, b), c) -> ((c, a), b))
+\end{lstlisting}
+Note that extensive use of this combinator over \code{pipe} with a hand-written combination data-type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Nonetheless, we can define a parallel piping operator \code{|>>>|} which is semantically equivalent to \code{>>>} similar to the other parallel syntactic sugar from chapter \ref{syntacticSugar}:
+\begin{lstlisting}[frame=htrbl]
+(|>>>|) :: (ArrowLoop arr, ArrowChoice arr,
+	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
+	Future fut (([a], [b]), [c])) =>
+	arr a b -> arr b c -> arr a c
+(|>>>|) = pipe2 ()
 \end{lstlisting}

### 2017-03-25 Definition/src/main/Parrows/Skeletons/Topology.hs
minor

@@ -47,36 +47,43 @@ pipe2 :: (ArrowLoop arr, ArrowChoice arr,
             ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
             Future fut (([a], [b]), [c])) =>
             conf -> arr a b -> arr b c -> arr a c
-pipe2 conf f g = arr (\a -> (([a], []), [])) >>> pipe conf (replicate 2 $ unify f g) >>> arr snd >>> arr head
+pipe2 conf f g =
+    (arr return &&& arr (const [])) &&& arr (const []) >>>
+    pipe conf (replicate 2 (unify f g)) >>>
+    arr snd >>>
+    arr head
     where
         unify :: (ArrowChoice arr) => arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
-        unify f g = (mapArr f *** mapArr g) *** arr (\_ -> []) >>> arr (\((a, b), c) -> ((c, a), b))
+        unify f g = (mapArr f *** mapArr g) *** arr (const []) >>> arr (\((a, b), c) -> ((c, a), b))
 
 pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf, Future fut a) => conf -> [arr a a] -> arr a a
 pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
 
 pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) => conf -> [arr a a] -> arr a a
-pipeSimple conf fs = loop (arr snd &&& (arr (uncurry (:) >>> lazy) >>>
-                        parEvalN conf fs)) >>>
-                  arr last
+pipeSimple conf fs =
+    loop (arr snd &&& (arr (uncurry (:) >>> lazy) >>>
+        parEvalN conf fs)) >>>
+    arr last
 
 ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
     conf -> arr (i, r) (o, r) -> arr [i] [o]
-ring conf f = loop (second (rightRotate >>> lazy) >>>
-                    arr (uncurry zip) >>>
-                    parMap conf (second get >>> f >>> second put) >>>
-                    arr unzip)
+ring conf f =
+    loop (second (rightRotate >>> lazy) >>>
+        arr (uncurry zip) >>>
+        parMap conf (second get >>> f >>> second put) >>>
+        arr unzip)
 
[…]

### 2017-03-25 Eden/examples/torus/events.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+ghc torus_matrix_parrows -parmpi -rtsopts -eventlog
+ghc torus_matrix_eden -parmpi -rtsopts -eventlog
+
+./torus_matrix_parrows +RTS -N16 -ls
+./torus_matrix_eden +RTS -N16 -ls

### 2017-03-25 paper/syntacticSugar.tex

@@
-\section{Syntactic Sugar}
+\section{Syntactic Sugar} \label{syntacticSugar}
 For basic arrows, we have the \code{***} combinator which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.

### 2017-03-25 paper/topologySkeletons.tex
added ring and torus image for paper.

@@
 \end{lstlisting}
-Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of different types to each other.
+Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
+\code{arr a b} and \code{arr b c}.
+By wrapping these two arrows inside a common type
 \begin{lstlisting}[frame=htrbl]
@@
 pipe2 conf f g = 
-	arr (\a -> (([a], []), [])) >>>
-	pipe conf (replicate 2 $ unify f g) >>>
+	(arr return &&& arr (const [])) &&& arr (const []) >>>
+	pipe conf (replicate 2 (unify f g)) >>>
 	arr snd >>>
@@
 \subsection{ring}
-
+\begin{center}
+	\includegraphics[scale=0.75]{images/ring}
+\end{center}
 \begin{lstlisting}[frame=htrbl]
@@
 \subsection{torus}
-
+\begin{center}
+	\includegraphics[scale=0.75]{images/torus}
+\end{center}
 \begin{lstlisting}[frame=htrbl]
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows}
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
 	\caption[without Futures]{Matrix Multiplication with a torus (Parrows)}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden}
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
 	\caption[with Futures]{Matrix Multiplication with a torus (Eden)}

### 2017-03-25 paper/topologySkeletons.tex
wip on paper.

@@
 \subsection{ring}
+Eden's ring implementation allows a ring of functions to communicate over direct channels.
+We can rewrite its functionality easily again with the use of \code{loop}.
+In each loop we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
+We have to feed the intermediary input into \code{lazy} or the evaluation would hang.
+The reasoning was explained by \cite{eden_cefp}:
+\begin{quotation}
+Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
+Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.
+\end{quotation}
+Next, we zip the resulting \code{([i], [fut r])} to \code{[(i, fut r)]} with \code{arr (uncurry zip)} so we can feed that into a our input arrow \code{arr (i, r) (o, r)}, which we transform into \code{arr (i, fut r) (o, fut r)} before lifting it to \code{arr [(i, fut r)] [(o, fut r)]} to get a list \code{[(o, fut r)]}.
+Finally we unzip this list into \code{([o], [fut r])}.
+Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow.
 \begin{center}
@@
     arr unzip)
-
--- from Eden, ported to Arrows:
+\end{lstlisting}
+with \code{rightRotate}:
+\begin{lstlisting}[frame=htrbl]
 rightRotate :: (Arrow arr) => arr [a] [a]
@@
 \end{lstlisting}
-
+and lazy:
+\begin{lstlisting}[frame=htrbl]			
+lazy :: (Arrow arr) => arr [a] [a]
+lazy = arr (\ ~(x:xs) -> x : lazy xs
+\end{lstlisting}
+This combinator can be used for example to calculate the shortest paths in a graph using Warshall's algorithm. Further details on this can be found in \cite{eden_cefp}.
 \subsection{torus}
@@
 \end{lstlisting}
-
+In \cite{eden_cefp}, matrix multiplication using the gentlemans algorithm was given as a example of using this skeleton. Adapting this nodefunction to our arrow API gives us:
+\begin{lstlisting}[frame=htrbl]
+nodefunction :: Int ->
+	((Matrix, Matrix), [Matrix], [Matrix]) ->
+	([Matrix], [Matrix], [Matrix])
+nodefunction n ((bA, bB), rows, cols) =
+	([bSum], bA:nextAs , bB:nextBs)
+		where bSum =
+			foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
+			nextAs = take (n-1) rows
+			nextBs = take (n-1) cols
+\end{lstlisting}
+Comparing the trace from a call using our arrow definition of the torus (fig.
+\ref{fig:torus_parrows_trace}) with the Eden version (fig.
+\ref{fig:torus_eden_trace}) shows that the behaviour of the arrow version is comparable.
 \begin{figure}[ht]
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[without Futures]{Matrix Multiplication with a torus (Parrows)}
+	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (Parrows)}
+	\label{fig:torus_parrows_trace}
 \end{figure}
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
-	\caption[with Futures]{Matrix Multiplication with a torus (Eden)}
+	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden)}
+	\label{fig:torus_eden_trace}
 \end{figure}

### 2017-03-25 paper/futures.tex
wip

@@
-\section{Futures}
+\section{Futures} \label{futures}
 Consider the following arbitrary, but interesting, parallel arrow combinator:
@@
 \end{lstlisting}
-For ParMonad and Multicore we can use a basic dummy wrapper as we have shared memory:
+For ParMonad and Multicore we can use a basic dummy wrapper because we have shared memory in a single node:
 \begin{lstlisting}[frame=htrbl]
@@
 instance NFData (RemoteData a) where
-    rnf x = rnf $ rd x
+    rnf = rnf . rd
 instance Trans (RemoteData a)
@@
 instance (NFData a) => NFData (BasicFuture a) where
-    rnf bf = rnf $ val bf
+	rnf = rnf . val
 \end{lstlisting}
-Going back to our communication example we can use this new Future concept in order to enable direct communications between the nodes in the following way:
+Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
 \begin{lstlisting}[frame=htrbl]

### 2017-03-26 paper/topologySkeletons.tex
wip

@@
 \end{figure}
+
+%\begin{figure}[ht]
+%	\centering
+	\framebox{
+		\begin{tikzpicture}[thick, scale=0.7]
+		\begin{axis}[
+			xbar, xmin=0,
+			bar width=0.4cm,
+			width=12cm, height=7cm, enlarge y limits=0.1,
+			ylabel={amount of cores},
+			xlabel={mean runtime (seconds)},
+			symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
+			ytick=data,
+			nodes near coords, nodes near coords align={horizontal},
+			]
+			\addplot coordinates {(841,2) (248,4) (115,8) (63.9,16) (41.6,32) (34.6,48) (33.0,64) (32.9,96)};
+		\end{axis}
+		\end{tikzpicture}
+	}
+%	\caption[Parallel Matrix Multiplication using Eden (two 1024x1024 matrices)]
+%\end{figure}
+
+\framebox{
+	\begin{tikzpicture}[thick, scale=0.7]
+	\begin{axis}[
+	xbar, xmin=0,
+	bar width=0.4cm,
+	width=12cm, height=7cm, enlarge y limits=0.1,
+	ylabel={amount of cores},
+	xlabel={mean seconds run on torus\_matrix\_parrows (two 1024x1024 matrices)},
+	symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
+	ytick=data,
+	nodes near coords, nodes near coords align={horizontal},
+	]
+	\addplot coordinates {(663,2) (205,4) (113,8) (68.5,16) (41.9,32) (35.4,48) (36.3,64) (33.4,96)};
+	\end{axis}
+	\end{tikzpicture}
+}

### 2017-03-27 Definition/src/main/Parrows/Skeletons/Topology.hs
made torus easier to understand

@@ -75,14 +75,14 @@ ring conf f =
 
 --TODO: check whether this exchanges the futures the same way as Eden does it
 torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
-            ArrowParallel arr (c, fut [a], fut [b]) (d, fut [a], fut [b]) conf,
-            Future fut [a], Future fut [b]) =>
-         conf -> arr (c, [a], [b]) (d, [a], [b]) -> arr [[c]] [[d]]
+            ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
+            Future fut a, Future fut b) =>
+         conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
 torus conf f =
     loop (second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy)) >>>
         arr (uncurry3 (zipWith3 lazyzip3)) >>>
         (arr length >>> arr unshuffle) &&&
-            (shuffle >>> parEvalN conf (repeat (ptorus f))) >>>
+            (shuffle >>> parMap conf (ptorus f)) >>>
         app >>> arr (map unzip3) >>> arr unzip3 >>> threetotwo)
 
 uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
@@ -91,10 +91,10 @@ uncurry3 f (a, (b, c)) = f a b c
 lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
 lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
 
-ptorus :: (Arrow arr, Future fut [a], Future fut [b]) =>
-          arr (c, [a], [b]) (d, [a], [b]) ->
-          arr (c, fut [a], fut [b]) (d, fut [a], fut [b])
-ptorus f = arr (\ ~(c, fas, fbs) -> (c, get fas, get fbs)) >>> f >>> arr (\ ~(c, as, bs) -> (c, put as, put bs))
+ptorus :: (Arrow arr, Future fut a, Future fut b) =>
+          arr (c, a, b) (d, a, b) ->
+          arr (c, fut a, fut b) (d, fut a, fut b)
+ptorus f = arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>> arr (\ ~(d, a, b) -> (d, put a, put b))
 
[…]

### 2017-03-27 paper/futures.tex

@@
 In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
-While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in figure \ref{fig:withoutFutures}.
+While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in figure \ref{fig:withoutFutures}.
 This can become a serious bottleneck in heavy threaded applications.
@@
 \end{lstlisting}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in figure \ref{fig:withFutures}.
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in fig.
+\ref{fig:withFutures}.
 \begin{figure}[ht]

### 2017-03-27 Definition/src/main/Parrows/Skeletons/Topology.hs
minor

@@ -61,8 +61,8 @@ pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
 
 pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) => conf -> [arr a a] -> arr a a
 pipeSimple conf fs =
-    loop (arr snd &&& (arr (uncurry (:) >>> lazy) >>>
-        parEvalN conf fs)) >>>
+    loop (arr snd &&&
+        (arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
     arr last
 
 ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>

### 2017-03-27 paper/topologySkeletons.tex

@@
 \end{center}
+If we take the concept of a ring one dimension further, we get a torus.
+Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
+This gives us .
+With our parallel Arrows we implement a torus combinator yet again with the help of the \code{ArrowLoop} typeclass.
+\\\\
+Inside the loop we once again start by rotating the input, but this time not only in one direction, but in two.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple \code{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our input \code{([[c]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \code{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \code{[[fut a]]} horizontally and \code{[[fut b]]} vertically.
+We once again zip the inputs for the nodefunction with \code{arr (uncurry3 zipWith3 lazyzip3)} from \code{([[c]], ([[fut a]], [[fut b]]))} to \code{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
+\\\\
+This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
+We first have to \code{shuffle} all the inputs to then pass it into \code{parMap conf (ptorus f)} which yields us \code{[(d, fut a, fut b)]}.
+We can then unpack this shuffled list back to its original ordering by feeding this into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app} from the \code{ArrowApply} typeclass.
+Finally, we unpack this matrix \code{[[[(d, fut a, fut b)]]} with \code{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \code{([[d]], ([[fut a]], [[fut b]]))}.
+\\\\
+The complete definition of the \code{torus} combinator is:
 \begin{lstlisting}[frame=htrbl]
 torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
-            ArrowParallel arr (c, fut [a], fut [b]) (d, fut [a], fut [b]) conf,
-            Future fut [a], Future fut [b]) =>
-         conf ->
-         arr (c, [a], [b]) (d, [a], [b]) ->
-         arr [[c]] [[d]]
+	ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
+	Future fut a, Future fut b) =>
+	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
 torus conf f =
@@
 	(arr length >>> arr unshuffle) &&&
-        (shuffle >>> parEvalN conf (repeat (ptorus f))) >>>
-    app >>> arr (map unzip3) >>> arr unzip3 >>> threetotwo)
-
+		(shuffle >>> parMap conf (ptorus f)) >>>
+	app >>>
+	arr (map unzip3) >>> arr unzip3 >>> threetotwo)
+\end{lstlisting}
+with uncurry3,
+\begin{lstlisting}[frame=htrbl]
 uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
 uncurry3 f (a, (b, c)) = f a b c
-
+\end{lstlisting}
+lazyzip3,
+\begin{lstlisting}[frame=htrbl]
 lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
 lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
-
-ptorus :: (Arrow arr, Future fut [a], Future fut [b]) =>
-          arr (c, [a], [b]) (d, [a], [b]) ->
-          arr (c, fut [a], fut [b]) (d, fut [a], fut [b])
+\end{lstlisting}
+ptorus,
+\begin{lstlisting}[frame=htrbl]
+ptorus :: (Arrow arr, Future fut a, Future fut b) =>
+	arr (c, a, b) (d, a, b) ->
+	arr (c, fut a, fut b) (d, fut a, fut b)
 ptorus f =
-	arr (\ ~(c, fas, fbs) -> (c, get fas, get fbs)) >>>
+	arr (\ ~(c, a, b) -> (c, get a, get b)) >>>
 	f >>>
-	arr (\ ~(c, as, bs) -> (c, put as, put bs))
-
+	arr (\ ~(d, a, b) -> (d, put a, put b))
+\end{lstlisting}
+and threetotwo.
+\begin{lstlisting}[frame=htrbl]
 threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
@@
 \end{lstlisting}
-In \cite{eden_cefp}, matrix multiplication using the gentlemans algorithm was given as a example of using this skeleton. Adapting this nodefunction to our arrow API gives us:
+As an example of using this skeleton \cite{eden_cefp} gave the matrix multiplication using the gentlemans algorithm. Adapting this nodefunction to our arrow API gives us:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-Comparing the trace from a call using our arrow definition of the torus (fig.
+If we compare the trace from a call using our arrow definition of the torus (fig.
 \ref{fig:torus_parrows_trace}) with the Eden version (fig.
-\ref{fig:torus_eden_trace}) shows that the behaviour of the arrow version is comparable.
+\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.
 \begin{figure}[ht]

### 2017-03-27 paper/topologySkeletons.tex
reworked pipe and ring chapters for easier understandability

@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons, among them are a \code{pipe}, \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
 These seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that we can express such skeletons with Arrows as well.
@@
 The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows don't have to reside on the same thread/machine.
-We implement this skeleton using the \code{ArrowLoop} typeclass which enables us to use \code{loop :: arr (a, b) (c, b) -> arr a c}.
-
+We implement this skeleton using the \code{ArrowLoop} typeclass which gives us the \code{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express loop like computations.
+For example this
+\begin{lstlisting}[frame=htrbl]
+loop (arr (\(a, b) -> (b, a:b)))
+\end{lstlisting}
+,which is the same as
+\begin{lstlisting}[frame=htrbl]
+loop (arr snd &&& arr (uncurry (:)))
+\end{lstlisting}
+defines an arrow that takes its input \code{a} and converts it into an infinite stream \code{[a]} of it.
+Using this to our advantage gives us a first draft of a pipe implementation by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{&&&} and then only picking the first element of the resulting list with \code{arr last}:
 \begin{lstlisting}[frame=htrbl]
@@
 pipeSimple conf fs =
-	loop (arr snd &&& (arr (uncurry (:) >>> lazy) >>>
-		parEvalN conf fs)) >>>
+	loop (arr snd &&&
+		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
 	arr last
@@
 \end{lstlisting}
+Note that here the use of \code{lazy} is essential as without it programs using this definition would never halt.
+It is there so that the calculation of the input \code{[a]} halts before passing it into \code{parEvalN}.
+\\\\
 However, using this definition directly, will result in the master node becoming a potential bottleneck in distributed environments as described in chapter \ref{futures}.
-Therefore, a more sophisticated version that uses Futures internally is a good idea:
+Therefore, a more sophisticated version that internally uses Futures is a good idea:
 \begin{lstlisting}[frame=htrbl]
@@
 \subsection{ring}
-Eden's ring implementation allows a ring of functions to communicate over direct channels.
-We can rewrite its functionality easily again with the use of \code{loop}.
-In each loop we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
-We have to feed the intermediary input into \code{lazy} or the evaluation would hang.
-The reasoning was explained by \cite{eden_cefp}:
+\begin{center}
+	\includegraphics[scale=0.75]{images/ring}
+\end{center}
+Eden comes with a ring skeleton implementation that allows the computation of a function \code{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
+Its input is a node function \code{i -> r -> (o, r)} in which \code{r} serves as the intermediary output that gets send to the neighbour of each node.
+This data is sent over direct communication channels (remote data).
+The definition is as follows \cite{eden_skel_topology}:
+\begin{lstlisting}[frame=htrbl]
+ringSimple :: (Trans i, Trans o, Trans r) =>
+   (i -> r -> (o,r))
+   -> [i] -> [o]
+ringSimple f is =  os
+  where
+    (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
+                                   (zip is $ lazy ringIns))
+    ringIns = rightRotate ringOuts
+\end{lstlisting}
+with toRD (to make use of remote data)
+\begin{lstlisting}[frame=htrbl]
+toRD :: (Trans i, Trans o, Trans r) =>
+        ((i,r) -> (o,r))
+        -> ((i, RD r) -> (o, RD r))
+toRD  f (i, ringIn)  = (o, release ringOut)
+  where (o, ringOut) = f (i, fetch ringIn)
+\end{lstlisting}
+and rightRotate:
+\begin{lstlisting}[frame=htrbl]
+rightRotate    :: [a] -> [a]
+rightRotate [] =  []
+rightRotate xs =  last xs : init xs
+\end{lstlisting}
+We can rewrite its functionality easily with the use of \code{loop} as the definition of the node function, \code{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \code{loop}'s \code{arr (a, b) (c, b) -> arr a c}.
+In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
+Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.
+The reasoning is explained by \cite{eden_cefp}:
 \begin{quotation}
@@
 Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow.
-\begin{center}
-	\includegraphics[scale=0.75]{images/ring}
-\end{center}
+\\\\
+This gives us the following complete definition for the \code{ring} combinator:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-This combinator can be used for example to calculate the shortest paths in a graph using Warshall's algorithm. Further details on this can be found in \cite{eden_cefp}.
+This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm. Further details on this can be found in \cite{eden_cefp}.
 \subsection{torus}
@@
 \end{center}
-If we take the concept of a ring one dimension further, we get a torus.
-Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-This gives us .
+If we take the concept of a ring one dimension further, we get a torus. Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
+\\\\
 With our parallel Arrows we implement a torus combinator yet again with the help of the \code{ArrowLoop} typeclass.

### 2017-03-30 paper/topologySkeletons.tex
minor

@@
 \\\\
-With our parallel Arrows we implement a torus combinator yet again with the help of the \code{ArrowLoop} typeclass.
+In order to port this to parallel Arrows we have to use the torus combinator, yet again.
 \\\\
-Inside the loop we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple \code{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our input \code{([[c]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \code{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \code{[[fut a]]} horizontally and \code{[[fut b]]} vertically.
-We once again zip the inputs for the nodefunction with \code{arr (uncurry3 zipWith3 lazyzip3)} from \code{([[c]], ([[fut a]], [[fut b]]))} to \code{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
+Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple \code{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \code{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \code{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \code{[[fut a]]} horizontally and \code{[[fut b]]} vertically.
+Then, we once again zip the inputs for the input arrow with \code{arr (uncurry3 zipWith3 lazyzip3)} from \code{([[c]], ([[fut a]], [[fut b]]))} to \code{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
 \\\\
@@
 We first have to \code{shuffle} all the inputs to then pass it into \code{parMap conf (ptorus f)} which yields us \code{[(d, fut a, fut b)]}.
-We can then unpack this shuffled list back to its original ordering by feeding this into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app} from the \code{ArrowApply} typeclass.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app} from the \code{ArrowApply} typeclass.
 Finally, we unpack this matrix \code{[[[(d, fut a, fut b)]]} with \code{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \code{([[d]], ([[fut a]], [[fut b]]))}.

### 2017-03-30 paper/abstract.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
 \begin{abstract}
-Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
+Arrows form a general interface for computation and pose therefore as an alternative to monads for API design.
 In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
-In this paper we will use this concept to express parallelism.
-\\\\
-First, we give an introduction to some of the possible ways to add parallelism to Haskell programs.
-Then, we give the basic definition of Arrows, which is followed up by the introduction of some utility functions used in this paper.
-Next, we introduce the \code{ArrowParallel} typeclass together with backends for it written with the parallel Haskells introduced earlier, finishing up with the benefits of this new way of writing parallel programs.
-After this we give the definition of several parallel skeletons.
-Then, we introduce some syntactic sugar to mimic the sequential arrow combinators introduced by Hughes \citHughes ~but with parallelism added.
-We also give benchmarks of our newly created parallel Haskell.
-Finally we give a short conclusion of what we managed to achieve.
+We express parallelism using this concept.
+We define an Arrows-based interface for parallel computations and implement it using multiple parallel Haskells.
+
+This new way of writing parallel programs has a benefit of being portable across flavours of parallel Haskells used.
+With our framework we define  several parallel skeletons.
+We also  introduce some syntactic sugar to provide parallelism-aware arrow combinators similar to the traditional ones.
+Benchmarks shows that our framework does not induce too much overhead performance-wise.
 \end{abstract}

### 2017-03-30 paper/arrows.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
-\section{Arrows}
+\subsection{Arrows}
+\label{sec:arrows}
 \begin{center}
@@
 \end{center}
-Arrows were introduced by Hughes \citHughes ~as a general interface for computation.
-An arrow \code{arr a b} can be look at as a computation that converts an input \code{a} to an output \code{b}.
+Arrows were introduced by Hughes \citHughes\ as a general interface for computation.
+An arrow \code{arr a b} represents  a computation that converts an input \code{a} to an output \code{b}.
 This is defined in the arrow typeclass:
@@
 \end{minipage}
-\lstinline{arr} is used to lift an ordinary function to an arrow type.
-This can be thought of as analogous to the monadic \lstinline{return}.
-The \lstinline{>>>} operator, in a similar way, is analogous to the monadic composition operator \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}.
-And lastly, the \lstinline{first} operator, which takes the input arrow from b to c and converts it into an arrow on pairs with the second argument untouched, is also needed for actual useful code as without it, we wouldn't have a way to save input across arrows.
+\lstinline{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \lstinline{return}.
+The \lstinline{>>>} operator is analogous to the monadic composition  \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}.
+Lastly, the \lstinline{first} operator  takes the input arrow from \lstinline{b} to \lstinline{c} and converts it into an arrow on pairs with the second argument untouched.
+It allows us to to save input across arrows.
 \\\\
@@
 data Kleisli m a b = Kleisli { run :: a -> m b }
-\end{lstlisting}
-as well:
-\begin{lstlisting}[frame=htrbl]
+
 instance Monad m => Arrow (Kleisli m) where
@@
 	first f = Kleisli $ \(a,c) -> f a >>= \b -> return (b,c)
-\end{lstlisting}
+\end{lstlisting}%$
 \begin{figure}
@@
 	\begin{tabular}{cc}
-		\subcaptionbox{first\label{1}}{\includegraphics[width = 1.5in]{images/first}} &
-		\subcaptionbox{second\label{2}}{\includegraphics[width = 1.5in]{images/second}} \\
-		\subcaptionbox{(***)\label{1}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
-		\subcaptionbox{(\&\&\&)\label{2}}{\includegraphics[width = 1.5in]{images/andandand}}\
+		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
+		\subcaptionbox{second\label{t2}}{\includegraphics[width = 1.5in]{images/second}} \\
+		\subcaptionbox{(***)\label{c1}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+		\subcaptionbox{(\&\&\&)\label{c2}}{\includegraphics[width = 1.5in]{images/andandand}}\
 	\end{tabular}
-	\caption{syntactic sugar}
+	\caption{Syntactic sugar for arrows. \olcomment{FIXME: subreferences}}
 	\label{3}
@@
 \end{lstlisting}
-The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
+% The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
 Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
 Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
-\\\\
-\textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used.
+% \\\\
+% \textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used.
 We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}
+%
+The more restrictive interface of arrows (a monad can be \emph{anything}, an arrow is a process of doing something, a \emph{computation}) allows for more elaborate composition and transformation combinators.
+One of the major problems in parallel computing is composition of parallel processes.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-03-30 paper/basicSkeletons.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
-\section{Basic Skeletons}
+\subsection{Extending the Interface}
+\label{sec:extending-interface}
+With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
+These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
 
-With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.
-
-\subsection{parEvalNLazy}
+\subsection{Lazy \code{parEvalN}}
 \begin{center}
@@
 \end{center}
-\code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
+The function \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
 Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
@@
 	arr concat
-	where
-		fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
-\end{lstlisting}
+	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
+\end{lstlisting} %$ %% formatting
 
-\subsection{parEval2}
+\subsection{Heterogenous tasks}
 \begin{center}
@@
 \end{lstlisting}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-03-30 paper/conclusion.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
 \section{Conclusion}
-As we have seen in this paper, arrows are a useful tool for composing parallel programs.
-By basing our parallel Haskell on them, we do not have to introduce new monadic types that wrap the computation and instead can use them just like they were regular sequential pure code.
-Performancewise, parallel arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
+\label{sec:conclusion}
+Arrows are a generic concept that allows for powerful composition combinators.
+To our knowledge we are the first ones to represent parallel computation with arrows.\olcomment{that strange arrows-based robot interaction paper from 1993 or so!
+clearly discuss in related work}
 
-\fixme{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}
+Arrows turn out to be a useful tool for composing in parallel programs.
+We do not have to introduce new monadic types that wrap the computation.
+Instead use arrows just like regular sequential pure functions.
+%
+This work features multiple parallel backends: the already available parallel Haskell flavours.
+Parallel Arrows feature an implementation of the \code{ArrowParallel} interface for Multicore Haskell, \code{Par} monad, and Eden.
+With our approach parallel programs can be ported across these flavours with no effort.
+%
+%
+Performancewise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}
+
+\mbcomment{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}
+
+\olcomment{Remove websites from citations, put them into footnotes!}

### 2017-03-30 paper/futures.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

--- 2017-03-25 paper/futures.tex
@@
 \section{Futures} \label{futures}
-Consider the following arbitrary, but interesting, parallel arrow combinator:
+Consider the following  parallel arrow combinator:
 \begin{lstlisting}[frame=htrbl]
@@
 In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
-While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in figure \ref{fig:withoutFutures}.
-This can become a serious bottleneck in heavy threaded applications.
+Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,pact-fft}.
+
+While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+This can become a serious bottleneck %in heavy threaded applications.
+for larger amount of data and number of processes \citep[showcases][as, \eg]{Berthold2009b}.
 \begin{figure}[ht]
@@
 \end{figure}
-To fix this, we have to introduce a way that enables the nodes to communicate directly with each other.
-Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour.
+This motivates for an approach that allows the nodes to communicate directly with each other.
+Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour \cite{AlGo03a,dieterle2010remotedata}.
 But as we want code written against our API to be implementation agnostic, we have to wrap this context.
@@
 \end{lstlisting}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in figure \ref{fig:withFutures}.
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear.
+Do Figs 2-3 with a lot of load?}
 \begin{figure}[ht]

### 2017-03-30 paper/mapSkeletons.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
 \section{Map-based Skeletons}
-\subsection{parMap}
+\label{sec:map-skeletons}
+Now we have developed Parallel Arrows further enough to define some algorithmic skeletons useful to an application programmer.
+\subsection{Parallel map}
 \begin{center}
@@
 
-\subsection{parMapStream}
+\subsection{Lazy parallel map}
 \begin{center}
@@
 
-\subsection{farm}
+\subsection{Statically load-balancing parallel map}
 \begin{center}
@@
 \end{lstlisting}
-The definition of \code{unshuffle} is
-\begin{lstlisting}[frame=htrbl]
-unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
-unshuffle n = arr (\xs -> [takeEach n (drop i xs) | i <- [0..n-1]])
-
-takeEach :: Int -> [a] -> [a]
-takeEach n [] = []
-takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
-\end{lstlisting}
-, while \code{shuffle} is defined as:
-\begin{lstlisting}[frame=htrbl]
-shuffle :: (Arrow arr) => arr [[a]] [a]
-shuffle = arr (concat . transpose)
-\end{lstlisting}
-(These were taken from Eden's source code \cite{eden_skel_shuffle} and translated into Arrows.)
 
@@
 \subsection{parMapReduce}
+\olcomment{it appears STRANGE. are the data really left alone and noded after map and taken from there by reduce? It makes sense only is no communication through master takes place. ELSE: CUT!}
+
 -- this does not completely adhere to Google's definition of Map Reduce as it

### 2017-03-30 paper/motivation.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
-\section{Motivation}
-Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
+%\section{Motivation}
+%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
 In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
 In this paper we will use this concept to express parallelism.
+
+\section{Introduction}
+\label{sec:introduction}
+\olcomment{todo, reuse 5.5, "Impact" at the end and more}
+
+blablabla arrows, parallel, haskell.
+
+\paragraph{Contribution}
+
+HIT HERE REALLY STRONG
+
+\paragraph{Structure}
+The remaining text is structures as follows.
+Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
+Section~\ref{sec:related-work} discusses related work.
+Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface.
+Section~\ref{futures} defines futures for Parallel Arrows.
+Futures are a further concept that helps to define more advanced interfaces.
+Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons in our newly defined dialect.
+More advanced ones are showcased in Section~\ref{sec:topology-skeletons}.
+Section~\ref{sec:benchmarks} shows the benchmark results.
+Section~\ref{sec:conclusion} concludes.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-03-30 paper/parallelHaskells.tex
minor
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
-\section{Short introduction to parallel Haskells}
+\section{Background}
+\label{sec:background}
+\subsection{Short introduction to parallel Haskells}
 There are already several ways to write parallel programs in Haskell.
@@
 
-\subsection{Multicore Haskell}
-Multicore Haskell \cite{multicore_hackage} is the GHC native way to do parallel processing.
-It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}.
+\subsubsection{Multicore Haskell}
+Multicore Haskell \cite{Marlow2009} is way to do parallel processing found in standard GHC.\footnote{See \url{https://hackage.haskell.org/package/parallel-3.2.1.0}.} It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}.
 For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-\end{center}
+\end{center} %$ %% formatting
 
-\subsection{ParMonad}
-The \code{Par} monad introduced by Marlow et al.
-\cite{monad_par_paper_2011}, which can be found in the monad-par package on hackage \cite{monad_par_hackage}, is a monad designed for composition of parallel programs.
-\\\\
+\subsubsection{ParMonad}
+The \code{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
+
 Our parallel evaluation function \code{parEvalN} can be defined by zipping the list of \code{[a -> b]} with the list of inputs \code{[a]} with the application operator \code{\$} just like with Multicore Haskell.
@@
 
-\subsection{Eden}
-Eden  is a parallel Haskell for distributed memory and comes with a MPI and a PVM backend \cite{eden_hackage, Loogen2012, eden_homepage}.
-This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell.
-\\\\
-While it also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
+\subsubsection{Eden}
+Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{See also \url{http://www.mathematik.uni-marburg.de/~eden/} and \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell flavour.
+
+Eden was designed to work on clusters, but with a further simple backend it operates on multicores.
+However, in contrast to many other parallel Haskells, in Eden each process has its own heap.
+This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results also on multicores \cite{arcs-dc,no_pain}.
+
+While Eden also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
 \\
@@
 \\
-This allows us to define \code{parEvalN} quite easily:
+This allows us to define \code{parEvalN} directly:
 
@@
 parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = spawnF fs as
+parEvalN = spawnF 
 \end{lstlisting}
@@
 \end{center}
+
+\paragraph{Eden TraceViewer}
+To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
+While some large-scale solutions exist \cite{Geimer2010}, in parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See .....
+on hackage for the last available version of Eden TraceViewer.
+There was an effort to implement the TraceViewer using modern web technologies \cite{traceviewer-web}.} \cite{Berthold2007a}.
+In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.
+
+In a trace, the $x$ axis shows the time, the $y$ axis enumerates the machines and processes.
+A~trace shows a running process in green, a blocked process is red.
+If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow.
+The typical reason for then is GC.
+An inactive machine where no processes are started yet, or all are already terminated, is shows as a blue bar.
+A~comminication from one process to another is represented with a black arrow.
+A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-03-30 paper/parrows.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
 \section{Parallel Arrows}
+\label{sec:parallel-arrows}
 We have seen what Arrows are and how they can be used as a general interface to computation.
-In the following section we will discuss how Arrows can also be looked at as a general interface not only to computation, but to \textbf{parallel computation} as well.
+In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well.
 We start by introducing the interface and explaining the reasonings behind it.
@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function listApp, but does parallel instead of serial evaluation).
+Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function \lstinline{listApp}, but does parallel instead of serial evaluation).
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-With this definition of \code{parEvalN}, parallel execution can be looked at as yet another arrow combinator.
+With this definition of \code{parEvalN}, parallel execution is yet another arrow combinator.
 But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
@@
 \end{lstlisting}
-Sometimes parallel Haskells require additional configuration parameters as information about the execution environment.
-This is why we also introduce an additional \code{conf} parameter to the function.
+Sometimes parallel Haskells require additional configuration parameters, \eg an information about the execution environment.
+For this reason we also introduce an additional \code{conf} parameter to the function.
 We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}.
@@
 \end{lstlisting}
-Note that we don't require the conf parameter in every implementation.
+Note that we don't require the \code{conf} parameter in every implementation.
 If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
@@
 		parEvalN _ fs = listApp fs >>> arr (flip using $ parList rdeepseq)
-\end{lstlisting}
+\end{lstlisting}% $ %% formatting
 We hardcode the \code{parList rdeepseq} strategy here, as in this context it is the only one making sense, since we usually want the output list to be fully evaluated to its normal form.
@@
 \subsection{ParMonad}
-The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function, which forks away the computation of a value and returns an IVar containing the result in the Par monad.
-\\\\
+The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function.
+The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.
+
 We therefore apply each function to its corresponding input value with \code{app} and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call.
@@
 			arr runPar
-\end{lstlisting}
+\end{lstlisting} %$ %% formatting
 
@@
 	arr a b -> (a -> b)
-\end{lstlisting}
-, we don't do it in this paper, as this seems too hacky. For now, we just implement \code{ArrowParallel} for normal functions
+\end{lstlisting} %
+we don't do it here, for aesthetic resons. For now, we just implement \code{ArrowParallel} for normal functions
 \begin{lstlisting}[frame=htrbl]
@@
 
-\subsection{Benefits of parallel Arrows}
-\fixme{Okay so?}
-\\
+\subsection{Impact of parallel Arrows}
+\olcomment{move this to Contributions in the front or something}
 We have seen that we can wrap parallel Haskells inside of the \code{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
@@
 This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
-We can just "plug" in parallel parts into our sequential programs without having to change anything.
+We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
 	\item \textbf{Abstraction}:
@@
 This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
-So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just swapping out the actual \code{ArrowParallel} instance.
+So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \code{ArrowParallel} instance.
 \end{itemize}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-03-30 paper/syntacticSugar.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

--- 2017-03-25 paper/syntacticSugar.tex
@@
-\section{Syntactic Sugar}
+\subsection{Syntactic Sugar} \label{syntacticSugar}
 For basic arrows, we have the \code{***} combinator which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
@@
 \end{lstlisting}
-With this we can analogously to the serial \code{&&&} define the parallel \code{|&&&|}.
+% With this we can analogously to the serial \code{&&&}
+We define the parallel \code{|&&&|} in a similar manner to its sequential prototype.
 \begin{lstlisting}[frame=htrbl]
@@
 (|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
-\end{lstlisting}
+\end{lstlisting} % $ %% formatting

### 2017-03-30 paper/utilityFunctions.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

@@
 \section{Utility Functions}\label{utilfns}
-Before we go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: map, foldl and zipWith on arrows.
-\\
+To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \lstinline{map}, \lstinline{foldl} and \lstinline{zipWith} on Arrows.\olcomment{capitalisation!
+arrows vs.
+Arrows!}
+
 The \code{mapArr} combinator lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows}:
@@
 	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
-	where
-		listcase [] = Left ()
+	where listcase [] = Left ()
 	      listcase (x:xs) = Right (x,xs)
@@
 		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
-	where
-		listcase [] = Left []
+	where listcase [] = Left []
 	      listcase (x:xs) = Right (x,xs)
 \end{lstlisting}
-Finally, with the help of \code{mapArr}, we can define zipWithArr, which lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+Finally, with the help of \code{mapArr}, we can define \lstinline{zipWithArr} that lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
 \begin{lstlisting}[frame=htrbl]
@@
 zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
-\end{lstlisting}
-These combinators make use of the \code{ArrowChoice} typeclass, which allows us to use the \code{|||} combinator.
-It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
+\end{lstlisting} %$ %% formatting
+These combinators make use of the \code{ArrowChoice} type class, it provides the \code{|||} combinator.
+The latter takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
 \begin{lstlisting}[frame=htrbl]
@@
 listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
+\end{lstlisting}% $ %% formatting
+This combinator also makes use of the \code{ArrowApply} typeclass that allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
+
+The definition of \code{unshuffle} is
+\begin{lstlisting}[frame=htrbl]
+unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
+unshuffle n = arr (\xs -> [takeEach n (drop i xs) | i <- [0..n-1]])
+
+takeEach :: Int -> [a] -> [a]
+takeEach n [] = []
+takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
 \end{lstlisting}
-This combinator also makes use of the \code{ArrowApply} typeclass which allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
+while \code{shuffle} is defined as:
+\begin{lstlisting}[frame=htrbl]
+shuffle :: (Arrow arr) => arr [[a]] [a]
+shuffle = arr (concat . transpose)
+\end{lstlisting}
+These functions were taken from Eden skeleton source code \cite{eden_skel_shuffle} and lifted to Arrows.
+
+The helper functions for \code{torus} are
+\begin{lstlisting}[frame=htrbl]
+uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
+uncurry3 f (a, (b, c)) = f a b c
+
+lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
+lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
+
+threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
+threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
+\end{lstlisting} % $ %% formatting
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-03-30 paper/topologySkeletons.tex
topologySkeletons.tex merged with Olegs remarks

@@
-\section{Topology Skeletons}
+\section{Topological Skeletons}
+\label{sec:topology-skeletons}
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
 The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
-These seem like reasonable candidates to be ported to our arrow based parallel Haskell to prove that we can express such skeletons with Arrows as well.
+These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express such skeletons with Parallel Arrows as well.
 
-\subsection{pipe}
+\subsubsection{Parallel pipe}
 
-The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows don't have to reside on the same thread/machine.
+The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
 We implement this skeleton using the \code{ArrowLoop} typeclass which gives us the \code{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express loop like computations.
@@
 Note that here the use of \code{lazy} is essential as without it programs using this definition would never halt.
-It is there so that the calculation of the input \code{[a]} halts before passing it into \code{parEvalN}.
-\\\\
-However, using this definition directly, will result in the master node becoming a potential bottleneck in distributed environments as described in chapter \ref{futures}.
-Therefore, a more sophisticated version that internally uses Futures is a good idea:
+We need to enforce that the evaluation of the input \code{[a]} terminates before passing it into \code{parEvalN}.
+
+However, using this definition directly, will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
+Therefore, we introduce a more sophisticated version that internally uses Futures.
 \begin{lstlisting}[frame=htrbl]
@@
 \code{arr a b} and \code{arr b c}.
-By wrapping these two arrows inside a common type
+By wrapping these two arrows inside a common type we obtain
 \begin{lstlisting}[frame=htrbl]
@@
 Note that extensive use of this combinator over \code{pipe} with a hand-written combination data-type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \code{|>>>|} which is semantically equivalent to \code{>>>} similar to the other parallel syntactic sugar from chapter \ref{syntacticSugar}:
+Nonetheless, we can define a parallel piping operator \code{|>>>|} which is semantically equivalent to \code{>>>} in a similar manner to the other parallel syntactic sugar from Section~\ref{syntacticSugar}:
 \begin{lstlisting}[frame=htrbl]
@@
 
-\subsection{ring}
+\subsection{Ring skeleton}
 \begin{center}
@@
 This data is sent over direct communication channels (remote data).
-The definition is as follows \cite{eden_skel_topology}:
+They define it as \citep{eden_skel_topology}
 \begin{lstlisting}[frame=htrbl]
@@
 ringSimple f is =  os
-  where
-    (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
+  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
                                    (zip is $ lazy ringIns))
         ringIns = rightRotate ringOuts
-\end{lstlisting}
-with toRD (to make use of remote data)
-\begin{lstlisting}[frame=htrbl]
+
 toRD :: (Trans i, Trans o, Trans r) => 
@@
   where (o, ringOut) = f (i, fetch ringIn)
-\end{lstlisting}
-and rightRotate:
-\begin{lstlisting}[frame=htrbl]
+
 rightRotate    :: [a] -> [a]
@@
 \end{lstlisting}
+%\end{lstlisting}
+%with toRD (to make use of remote data)
+%\begin{lstlisting}[frame=htrbl]
+%\end{lstlisting}
+%and rightRotate:
+%\begin{lstlisting}[frame=htrbl]
+
 We can rewrite its functionality easily with the use of \code{loop} as the definition of the node function, \code{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \code{loop}'s \code{arr (a, b) (c, b) -> arr a c}.
 In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
-Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.
-The reasoning is explained by \cite{eden_cefp}:
+Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \cite{eden_cefp}:
 \begin{quotation}
@@
 Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow.
-\\\\
+
 This gives us the following complete definition for the \code{ring} combinator:
@@
 rightRotate :: (Arrow arr) => arr [a] [a]
-rightRotate = arr $ \list -> case
-	list of [] -> []
+rightRotate = arr $ \list -> case list of 
+  [] -> []
   xs -> last xs : init xs
-\end{lstlisting}
-and lazy:
-\begin{lstlisting}[frame=htrbl]			
-lazy :: (Arrow arr) => arr [a] [a]
-lazy = arr (\ ~(x:xs) -> x : lazy xs
-\end{lstlisting}
-This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm. Further details on this can be found in \cite{eden_cefp}.
-\subsection{torus}
+\end{lstlisting} %$ %% formatting
+% and lazy:
+% \begin{lstlisting}[frame=htrbl]			
+% lazy :: (Arrow arr) => arr [a] [a]
+% lazy = arr (\ ~(x:xs) -> x : lazy xs
+% \end{lstlisting}
+%% we have shown this already.
+This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm. 
+%Further details on this can be found in \cite{eden_cefp}.
+\olcomment{let's do it?}
+\subsection{Torus skeleton}
 \begin{center}
@@
 \\\\
-In order to port this to parallel Arrows we have to use the torus combinator, yet again.
+With our parallel Arrows we implement a torus combinator yet again with the help of the \code{ArrowLoop} typeclass.
 \\\\
@@
 \end{lstlisting}
-with uncurry3,
-\begin{lstlisting}[frame=htrbl]
-uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
-uncurry3 f (a, (b, c)) = f a b c
-\end{lstlisting}
-lazyzip3,
-\begin{lstlisting}[frame=htrbl]
-lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
-lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
-\end{lstlisting}
-ptorus,
+with uncurry3, lazyzip3, and threetotwo from the Appendix, and
 \begin{lstlisting}[frame=htrbl]
@@
 ptorus f =
-	arr (\ ~(c, a, b) -> (c, get a, get b)) >>>
-	f >>>
+	arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>>
 	arr (\ ~(d, a, b) -> (d, put a, put b))
 \end{lstlisting}
-and threetotwo.
-\begin{lstlisting}[frame=htrbl]
-threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
-threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
-\end{lstlisting}
-As an example of using this skeleton \cite{eden_cefp} gave the matrix multiplication using the gentlemans algorithm. Adapting this nodefunction to our arrow API gives us:
+As an example of using this skeleton \citet{eden_cefp} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}. Adapting this nodefunction to our Arrow API gives us:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-If we compare the trace from a call using our arrow definition of the torus (fig.
-\ref{fig:torus_parrows_trace}) with the Eden version (fig.
-\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.
+If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.\olcomment{much more details on this!}
 \begin{figure}[ht]
@@
 }
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2017-03-30 paper/topologySkeletons.tex
make it compile
Signed-off-by: Martin Braun <martinbraun123@aol.com>

@@
 
-\subsubsection{Parallel pipe}
+\subsection{Parallel pipe}
 
@@
 In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
-Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \cite{eden_cefp}:
-\begin{quotation}
-Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
-Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.
-\end{quotation}
+Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{eden_cefp}:
+%\begin{quotation}
+\enquote{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
+Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
+%\end{quotation}
 Next, we zip the resulting \code{([i], [fut r])} to \code{[(i, fut r)]} with \code{arr (uncurry zip)} so we can feed that into a our input arrow \code{arr (i, r) (o, r)}, which we transform into \code{arr (i, fut r) (o, fut r)} before lifting it to \code{arr [(i, fut r)] [(o, fut r)]} to get a list \code{[(o, fut r)]}.
@@
 
-%\begin{figure}[ht]
-%	\centering
-	\framebox{
-		\begin{tikzpicture}[thick, scale=0.7]
-		\begin{axis}[
-			xbar, xmin=0,
-			bar width=0.4cm,
-			width=12cm, height=7cm, enlarge y limits=0.1,
-			ylabel={amount of cores},
-			xlabel={mean runtime (seconds)},
-			symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
-			ytick=data,
-			nodes near coords, nodes near coords align={horizontal},
-			]
-			\addplot coordinates {(841,2) (248,4) (115,8) (63.9,16) (41.6,32) (34.6,48) (33.0,64) (32.9,96)};
-		\end{axis}
-		\end{tikzpicture}
-	}
-%	\caption[Parallel Matrix Multiplication using Eden (two 1024x1024 matrices)]
-%\end{figure}
-
-\framebox{
-	\begin{tikzpicture}[thick, scale=0.7]
-	\begin{axis}[
-	xbar, xmin=0,
-	bar width=0.4cm,
-	width=12cm, height=7cm, enlarge y limits=0.1,
-	ylabel={amount of cores},
-	xlabel={mean seconds run on torus\_matrix\_parrows (two 1024x1024 matrices)},
-	symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
-	ytick=data,
-	nodes near coords, nodes near coords align={horizontal},
-	]
-	\addplot coordinates {(663,2) (205,4) (113,8) (68.5,16) (41.9,32) (35.4,48) (36.3,64) (33.4,96)};
-	\end{axis}
-	\end{tikzpicture}
-}
-
 %%% Local Variables:

### 2017-05-08 paper/futures.tex
adaption of the paper to the new Multicore instance and the MVar Future.

@@
 \end{lstlisting}
-For ParMonad and Multicore we can use a basic dummy wrapper because we have shared memory in a single node:
+For ParMonad and Multicore we can simply use \code{MVar}s because we have shared memory in a single node:
 \begin{lstlisting}[frame=htrbl]
-data BasicFuture a = BF { val :: a }
+{-# NOINLINE putUnsafe #-}
+putUnsafe :: a -> MVar a
+putUnsafe a = unsafePerformIO $ do
+    mVar <- newEmptyMVar
+    putMVar mVar a
+    return mVar
 
-instance (NFData a) => Future BasicFuture a where
-    put = arr (\a -> BF { val = a })
-    get = arr val
+instance (NFData a) => Future MVar a where
+    put = arr putUnsafe
+    get = arr takeMVar >>> arr unsafePerformIO
 \end{lstlisting}
 To fit the \code{ArrowParallel} instances we gave earlier, we also have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
-We need this implementation for our \code{RemoteData} wrapper
+We need this implementation for our \code{RemoteData} wrapper,
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and the following for the \code{BasicFuture} dummy type:
-\begin{lstlisting}[frame=htrbl]
-instance (NFData a) => NFData (BasicFuture a) where
-	rnf = rnf . val
-\end{lstlisting}
+while \code{MVar} already has a suitable NFData implementation.
+\\\\
 Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:

### 2017-05-08 paper/mapSkeletons.tex
adaption of the paper to the new Multicore instance and the MVar Future.

@@
 \label{sec:map-skeletons}
-Now we have developed Parallel Arrows further enough to define some algorithmic skeletons useful to an application programmer.
+Now we have developed Parallel Arrows far enough to define some algorithmic skeletons useful to an application programmer.
 \subsection{Parallel map}

### 2017-05-08 paper/parallelHaskells.tex
adaption of the paper to the new Multicore instance and the MVar Future.

@@
 For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
-This lazy list \code{[b]} is then forcibly evaluated in parallel with the strategy \code{Strategy [b]} by the \code{using} operator.
-This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.
+This lazy list \code{[b]} is then according to a \code{Strategy [b]} by the \code{using :: a -> Strategy a -> a} operator.
+This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
+\textbf{\textcolor{red}{explain usage of pseq}}
 
@@
 parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = zipWith ($) fs as `using` parList rdeepseq
+parEvalN fs as = let bs = zipWith ($) fs as in
+	(bs `using` parList rdeepseq) `pseq` bs
 \end{lstlisting}
@@
 To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
-While some large-scale solutions exist \cite{Geimer2010}, in parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See .....
+While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See .....
 on hackage for the last available version of Eden TraceViewer.

### 2017-05-08 paper/parrows.tex
adaption of the paper to the new Multicore instance and the MVar Future.

@@
 \end{lstlisting}
-Sometimes parallel Haskells require additional configuration parameters, \eg an information about the execution environment.
+Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs.
+normalform).
 For this reason we also introduce an additional \code{conf} parameter to the function.
@@
 Note that we don't require the \code{conf} parameter in every implementation.
-If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
+If it is not needed, we usually just default the \code{conf} type parameter to () and even blank it out in the parameter list of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
 
 \subsection{Multicore Haskell}
-The Multicore Haskell implementation of this class is straightforward using listApp from chapter \ref{utilfns} combined with the \code{using} operator from Multicore Haskell.
+The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from chapter \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
 \begin{lstlisting}[frame=htrbl]
 instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b conf where
-		parEvalN _ fs = listApp fs >>> arr (flip using $ parList rdeepseq)
+	ArrowParallel arr a b () where
+    	parEvalN _ fs =
+       		listApp fs >>>
+        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
+        	arr (uncurry pseq)
 \end{lstlisting}% $ %% formatting
-We hardcode the \code{parList rdeepseq} strategy here, as in this context it is the only one making sense, since we usually want the output list to be fully evaluated to its normal form.
-
+For most cases this fully evaluating version would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \code{Strategy} interface, we want to the user not to lose this ability because of using our API.
+We therefore introduce the \code{Conf a} data-type that simply wraps a \code{Strategy a}.
+We can't directly use the \code{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
+\begin{lstlisting}[frame=htrbl]
+data Conf a = Conf (Strategy a)
+\end{lstlisting}
+We simply unwrap the strategy out of this wrapper and get the following (configurable) \code{ArrowParallel} instance:
+\begin{lstlisting}[frame=htrbl]
+instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
+    parEvalN (Conf strat) fs =
+        listApp fs >>>
+        arr (withStrategy (parList strat)) &&& arr id >>>
+        arr (uncurry pseq)
+\end{lstlisting}
 \subsection{ParMonad}

### 2017-05-10 paper/parallelHaskells.tex
minor things

@@
 \end{center}
-We will now implement \code{parEvalN} with the different parallel Haskells.
+Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \code{parEvalN} with several different parallel Haskells.
 
@@
 For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
-This lazy list \code{[b]} is then according to a \code{Strategy [b]} by the \code{using :: a -> Strategy a -> a} operator.
-This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
-\textbf{\textcolor{red}{explain usage of pseq}}
+We then evaluate this lazy list \code{[b]} according to a \code{Strategy [b]} with the \code{using :: a -> Strategy a -> a} operator.
+We construct this strategy with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
+To ensure that programs that use \code{parEvalN} have the correct evaluation order, we annotate the computation with \code{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \code{parEvalN} computations.
+This is particularly necessary in circular communication topologies like in the \code{torus} or \code{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \code{pseq} during testing for this paper.
 
@@
 We finally execute this process with \code{runPar} to finally get \code{[b]} again.
+
+\textbf{\textcolor{red}{explain problems with laziness here. Problems with torus}}
 

### 2017-05-10 paper/parrows.tex

@@
 \begin{lstlisting}[frame=htrbl]
-instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
+instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
+	ArrowParallel arr a b (Conf b) where
     	parEvalN (Conf strat) fs =
@@
 \subsection{ParMonad}
-The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function.
+The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function.
 The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.

### 2017-05-11 paper/abstract.tex
applied olegs patches

@@
 \begin{abstract}
-Arrows form a general interface for computation and pose therefore as an alternative to monads for API design.
-In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
+Arrows form\olcomment{are?} a general interface for computation and pose therefore as an alternative to monads for API design.
 We express parallelism using this concept.
-We define an Arrows-based interface for parallel computations and implement it using multiple parallel Haskells.
+This is a new way to represent parallel computation.
+We define an Arrows-based interface for parallelism and implement it using multiple parallel Haskells.
+\olcomment{Benefits:} In this manner we are able to bridge across various parallel Haskells with a common interface.
 
-This new way of writing parallel programs has a benefit of being portable across flavours of parallel Haskells used.
-With our framework we define  several parallel skeletons.
-We also  introduce some syntactic sugar to provide parallelism-aware arrow combinators similar to the traditional ones.
+This new way of writing parallel programs has a benefit of being portable across flavours of parallel Haskells.\olcomment{Wdh?}
+Each parallel computation is an arrow, they can be composed and transformed as such.
+We thus introduce some syntactic sugar to provide parallelism-aware arrow combinators.
+
+We also define  several parallel skeletons with our framework. 
 Benchmarks shows that our framework does not induce too much overhead performance-wise.

### 2017-05-11 paper/arrows.tex
applied olegs patches

@@
 \end{center}
-Arrows were introduced by Hughes \citHughes\ as a general interface for computation.
+Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
 An arrow \code{arr a b} represents  a computation that converts an input \code{a} to an output \code{b}.
@@
 \end{minipage}
+\olcomment{a figure?}
 \end{minipage}
@@
 	\end{tabular}
-	\caption{Syntactic sugar for arrows. \olcomment{FIXME: subreferences}}
+	\caption{Syntactic sugar for arrows. \olcomment{FIXME: subreferences? refer to the figure in text!}}
 	\label{3}
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \lstinline{first}, called \lstinline{second},
+With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second},
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and the \&\&\& combinator that constructs an arrow which outputs 2 different values like ***, but takes only one input.
+and the \code{&&&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input.
 \begin{lstlisting}[frame=htrbl]

### 2017-05-11 paper/futures.tex
applied olegs patches

@@
 In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
-Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,pact-fft}.
+Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
 
@@
 This can become a serious bottleneck %in heavy threaded applications.
-for larger amount of data and number of processes \citep[showcases][as, \eg]{Berthold2009b}.
+for larger amount of data and number of processes \citep[showcases][as, \eg]{Berthold2009-fft}.
 \begin{figure}[ht]
@@
 This motivates for an approach that allows the nodes to communicate directly with each other.
-Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour \cite{AlGo03a,dieterle2010remotedata}.
+Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
 But as we want code written against our API to be implementation agnostic, we have to wrap this context.

### 2017-05-11 paper/motivation.tex
applied olegs patches

@@
 The remaining text is structures as follows.
-Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
+Section~\ref{sec:background} briefly introduces known parallel Haskell flavours.
+and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
 Section~\ref{sec:related-work} discusses related work.
 Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface.
-Section~\ref{futures} defines futures for Parallel Arrows.
-Futures are a further concept that helps to define more advanced interfaces.
-Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons in our newly defined dialect.
-More advanced ones are showcased in Section~\ref{sec:topology-skeletons}.
+Section~\ref{futures} defines futures for Parallel Arrows, this concept enables better communication.
+Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \code{map} with and without load balancing, \code{map-reduce}) in our newly defined dialect.
+More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\code{pipe}, \code{ring}, \code{torus}).
 Section~\ref{sec:benchmarks} shows the benchmark results.
-Section~\ref{sec:conclusion} concludes.
+Section~\ref{sec:conclusion} discusses future work and concludes.
 

### 2017-05-11 paper/parallelHaskells.tex
applied olegs patches

@@
 	\includegraphics[scale=0.7]{images/parEvalN}
+\olcomment{make them to real figures? with environment, reference and such?}
 \end{center}
@@
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009} is way to do parallel processing found in standard GHC.\footnote{See \url{https://hackage.haskell.org/package/parallel-3.2.1.0}.} It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}.
+Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \code{using :: a -> Strategy a -> a}.
 For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
+\olcomment{swap $f_i$ and $a_i$! add ``evaluation'' to the last set of arrows?}
 \end{center} %$ %% formatting
@@
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
+\olcomment{add as figure? both code and image? reference then!}
 \end{center}
@@
 However, in contrast to many other parallel Haskells, in Eden each process has its own heap.
-This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results also on multicores \cite{arcs-dc,no_pain}.
+This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results also on multicores \cite{arcs-dc,aswad2009low}.
 
 While Eden also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
-\\
-\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
-\\
-This allows us to define \code{parEvalN} directly:
+%\\
+%\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
+%\\
+a \code{spawnF} function that
+%This 
+allows us to define \code{parEvalN} directly:
 
@@
 	\includegraphics[scale=0.5]{images/parEvalNEden}
+\olcomment{ditto mit figure}
 \end{center}
 
-\paragraph{Eden TraceViewer}
+\paragraph{Eden TraceViewer.}
 To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
@@
 
+\olcomment{show example trace or refer to a trace in later figures}
+
 %%% Local Variables:

### 2017-05-11 paper/topologySkeletons.tex
applied olegs patches

--- 2017-03-30 paper/topologySkeletons.tex
@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{eden_cefp, eden_skel_topology}.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{Loogen2012, eden_skel_topology}.
 These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express such skeletons with Parallel Arrows as well.
@@
 In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
-Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{eden_cefp}:
+Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
 %\begin{quotation}
@@
 \end{lstlisting}
-As an example of using this skeleton \citet{eden_cefp} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}. Adapting this nodefunction to our Arrow API gives us:
+As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}. Adapting this nodefunction to our Arrow API gives us:
 \begin{lstlisting}[frame=htrbl]

### 2017-05-11 paper/utilityFunctions.tex
replaced png pictures with pdf ones

@@
 \section{Utility Functions}\label{utilfns}
-To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \lstinline{map}, \lstinline{foldl} and \lstinline{zipWith} on Arrows.\olcomment{capitalisation!
-arrows vs.
-Arrows!}
+To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \lstinline{map}, \lstinline{foldl} and \lstinline{zipWith} on arrows.
 

### 2017-05-11 paper/arrows.tex
replaced png pictures with pdf ones
started making the pictures into figures.

@@
 \label{sec:arrows}
-\begin{center}
-	\includegraphics[]{images/arrow.png}
-\end{center}
+\begin{figure}[h]
+	\includegraphics{images/arrow}
+	\caption{schematic depiction of an arrow}
+\end{figure}
 Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
@@
 
+\begin{figure}[h]
 \begin{minipage}{\textwidth}
@@
 \begin{minipage}{0.25\textwidth}
-	~\\~\\~\\
-	\includegraphics[scale=0.6]{images/arr}~\\~\\
-	\includegraphics[scale=0.6]{images/compose}~\\~\\
-	\includegraphics[scale=0.6]{images/first}~\\~\\
+	\begin{center}
+	\subcaptionbox{arr}{\includegraphics[scale=0.6]{images/arr}}
+	\subcaptionbox{composition \code{(>>>)}}{\includegraphics[scale=0.6]{images/compose}}
+	\subcaptionbox{first}{\includegraphics[scale=0.6]{images/first}}
+	\end{center}
 \end{minipage}
-\olcomment{a figure?}
 \end{minipage}
+\caption{the arrow type class definition on the left with schematic depiction of its combinators on the right}
+\end{figure}
 \lstinline{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \lstinline{return}.
@@
 instance Monad m => Arrow (Kleisli m) where
-	arr f = Kleisli $ return . f
-	f >>> g = Kleisli $ \a -> f a >>= g
-	first f = Kleisli $ \(a,c) -> f a >>= \b -> return (b,c)
+	arr f = Kleisli (return . f)
+	f >>> g = Kleisli (\a -> f a >>= g)
+	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
 \end{lstlisting}%$

### 2017-05-11 paper/parallelHaskells.tex
started making the pictures into figures.

@@
 \end{lstlisting}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.7]{images/parEvalN}
+	\caption{schematic illustration of parEvalN}
+	\label{fig:parEvalN}
+\end{figure}
 \olcomment{make them to real figures? with environment, reference and such?}
-\end{center}
 Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \code{parEvalN} with several different parallel Haskells.
@@
 \end{lstlisting}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-\olcomment{swap $f_i$ and $a_i$! add ``evaluation'' to the last set of arrows?}
-\end{center} %$ %% formatting
+	\caption{dataflow of the Multicore Haskell parEvalN version}
+\end{figure} %$ %% formatting
 
@@
 \end{lstlisting}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-\olcomment{add as figure? both code and image? reference then!}
-\end{center}
+	\caption{dataflow of the Par Monad parEvalN version}
+\end{figure}
 
@@
 \end{lstlisting}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.5]{images/parEvalNEden}
-\olcomment{ditto mit figure}
-\end{center}
+	\caption{dataflow of the Eden parEvalN version}
+\end{figure}
 

### 2017-05-11 paper/parrows.tex
started making the pictures into figures.

@@
 Finally, we explain why using Arrows for expressing parallelism is beneficial.
-\begin{center}
-	\includegraphics[scale=0.7]{images/parEvalN}
-\end{center}
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel:
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalN}):
 \begin{lstlisting}[frame=htrbl]
@@
 \subsection{Multicore Haskell}
-The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from chapter \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
+The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
 \begin{lstlisting}[frame=htrbl]

### 2017-05-12 paper/arrows.tex
introduced more figures (arrows.tex, parallelHaskells.tex)

@@
 	\includegraphics{images/arrow}
-	\caption{schematic depiction of an arrow}
+	\caption{Schematic depiction of an arrow}
 \end{figure}
@@
 \begin{figure}[h]
-\begin{minipage}{\textwidth}
-\begin{minipage}{0.5\textwidth}
+\begin{subfigure}{0.5\textwidth}
+%\begin{minipage}{0.5\textwidth}
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-\end{minipage}
+\caption{Arrow class definition}
+%\end{minipage}
+\end{subfigure}
 ~~~~
-\begin{minipage}{0.25\textwidth}
+\begin{subfigure}{0.45\textwidth}
 	\begin{center}
@@
 	\end{center}
-\end{minipage}
-\end{minipage}
-\caption{the arrow type class definition on the left with schematic depiction of its combinators on the right}
+\end{subfigure}
+\caption{The arrow type class definition on the left with schematic depiction of its combinators on the right}
 \end{figure}
@@
 \\\\
-The most prominent instances of this interface are regular functions \lstinline{(->)},
+The most prominent instances of this interface are regular functions \lstinline{(->)} (Fig.~\ref{fig:arrowfn}),
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and the Kleisli type
+\caption{Arrow instance for regular functions}
+\label{fig:arrowfn}
+\end{figure}
+and the Kleisli type (Fig.~\ref{fig:arrowkleisli}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}%$
-\begin{figure}
+\caption{Definition of the Kleisli type and the corresponding arrow instance}
+\label{fig:arrowkleisli}
+\end{figure}
+\begin{figure}[h]
 	\centering
@@
 		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
-		\subcaptionbox{second\label{t2}}{\includegraphics[width = 1.5in]{images/second}} \\
-		\subcaptionbox{(***)\label{c1}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
-		\subcaptionbox{(\&\&\&)\label{c2}}{\includegraphics[width = 1.5in]{images/andandand}}\
+		\subcaptionbox{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
+		\subcaptionbox{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+		\subcaptionbox{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
 	\end{tabular}
-	\caption{Syntactic sugar for arrows. \olcomment{FIXME: subreferences? refer to the figure in text!}}
+	\caption{Syntactic sugar for arrows}
 	\label{3}
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second},
+With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second} (Fig.~\ref{fig:secondImg},\ref{fig:second}),
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-the *** combinator which combines first and second to handle two inputs in one arrow,
+\caption{The second combinator}
+\label{fig:second}
+\end{figure}
+the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},\ref{fig:***})
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and the \code{&&&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input.
+\caption{The (***) combinator}
+\label{fig:***}
+\end{figure}
+and the \code{\&\&\&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input (Fig.~\ref{fig:&&&Img},\ref{fig:&&&}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-A short example given by Hughes on how to use this is \lstinline{add} over arrows:
+\caption{The (\&\&\&) combinator}
+\label{fig:&&&}
+\end{figure}
+A short example given by Hughes on how to use this is \lstinline{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Add over arrows}
+\label{fig:addArrows}
+\end{figure}
 % The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
@@
 %
+
 The more restrictive interface of arrows (a monad can be \emph{anything}, an arrow is a process of doing something, a \emph{computation}) allows for more elaborate composition and transformation combinators.

### 2017-05-12 paper/parallelHaskells.tex
introduced more figures (arrows.tex, parallelHaskells.tex)

@@
 
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{parEvalN's type signature}
+\label{fig:parEvalNTypeSig}
+\end{figure}
 \begin{figure}[h]
 	\includegraphics[scale=0.7]{images/parEvalN}
-	\caption{schematic illustration of parEvalN}
+	\caption{Schematic illustration of parEvalN}
 	\label{fig:parEvalN}
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-	\caption{dataflow of the Multicore Haskell parEvalN version}
+	\caption{Dataflow of the Multicore Haskell parEvalN version}
 \end{figure} %$ %% formatting
@@
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-	\caption{dataflow of the Par Monad parEvalN version}
+	\caption{Dataflow of the Par Monad parEvalN version}
 \end{figure}
@@
 	\includegraphics[scale=0.5]{images/parEvalNEden}
-	\caption{dataflow of the Eden parEvalN version}
+	\caption{Dataflow of the Eden parEvalN version}
 \end{figure}

### 2017-05-12 paper/utilityFunctions.tex
minor fix in utilityFunctions.tex

@@
 \end{lstlisting}
-With the zipWithArr combinator we can also write a combinator \code{listApp}, which lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]}.
+With the zipWithArr combinator we can also write a combinator \code{listApp}, which lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]},
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}% $ %% formatting
-This combinator also makes use of the \code{ArrowApply} typeclass that allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
+which also makes use of the \code{ArrowApply} typeclass that allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
 

### 2017-05-12 paper/parrows.tex
introduced figures for the parrows Multicore chapter

@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalN}):
-\begin{lstlisting}[frame=htrbl]
-parEvalN :: [a -> b] -> [a] -> [b]
-\end{lstlisting}
-Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function \lstinline{listApp}, but does parallel instead of serial evaluation).
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},\ref{fig:parEvalN}).
+Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \lstinline{listApp} from Appendix \ref{utilfns}, but does parallel instead of serial evaluation).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{parEvalN Arrow combinator as a function}
+\label{fig:parEvalNArrowFn}
+\end{figure}
 With this definition of \code{parEvalN}, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
+But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we introduce a new typeclass \code{ArrowParallel arr a b} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
+\label{fig:parEvalNArrowTypeClass1}
+\end{figure}
 Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs.
@@
 We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}.
-So we add it to the type signature of the typeclass as well.
+So we add it to the type signature of the typeclass as well and get \code{ArrowParallel arr a b conf} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
+\label{fig:parEvalNArrowTypeClassFinal}
+\end{figure}
 Note that we don't require the \code{conf} parameter in every implementation.
@@
 The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}% $ %% formatting
-For most cases this fully evaluating version would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \code{Strategy} interface, we want to the user not to lose this ability because of using our API.
-We therefore introduce the \code{Conf a} data-type that simply wraps a \code{Strategy a}.
+\caption{Fully evaluating ArrowParallel instance for the Multicore backend}
+\label{fig:ArrowParallelMulticoreRdeepseq}
+\end{figure}
+For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \code{Strategy} interface, we want to the user not to lose this ability because of using our API.
+We therefore introduce the \code{Conf a} data-type that simply wraps a \code{Strategy a} (Fig.~\ref{fig:confa}).
 We can't directly use the \code{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-We simply unwrap the strategy out of this wrapper and get the following (configurable) \code{ArrowParallel} instance:
+\caption{Definition of Conf a}
+\label{fig:confa}
+\end{figure}
+To get our configurable \code{ArrowParallel} instance, we simply unwrap the strategy and pass it to \code{parList} as in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Configurable ArrowParallel instance for the Multicore backend}
+\label{fig:ArrowParallelMulticoreConfigurable}
+\end{figure}
 \subsection{ParMonad}

### 2017-05-12 paper/parrows.tex
added figures for the Par Monad chapter.

@@
 \end{figure}
-\subsection{ParMonad}
-The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function.
+\subsection{Par Monad}
+The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function.
 The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.
@@
 The result of this operation is a \code{Par [b]} from which we can finally remove the monad again by running \code{arr runPar} to get our output of \code{[b]}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting} %$ %% formatting
+\caption{ArrowParallel instance for the Par Monad backend}
+\label{fig:ArrowParallelParMonad}
+\end{figure}
 
@@
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a typeclass like
+While we could still manage to have only one class in the module by introducing a typeclass \code{ArrowUnwrap} (Fig.~\ref{fig:ArrowUnwrap}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 	arr a b -> (a -> b)
-\end{lstlisting} %
-we don't do it here, for aesthetic resons. For now, we just implement \code{ArrowParallel} for normal functions
+\end{lstlisting}
+\caption{possible ArrowUnwrap typeclass}
+\label{fig:ArrowUnwrap}
+\end{figure}
+We don't do it here for aesthetic resons, though. For now, we just implement \code{ArrowParallel} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-and the Kleisli type.
+\caption{ArrowParallel instance for functions in the Eden backend}
+\label{fig:ArrowParallelEdenFns}
+\end{figure}
+and the Kleisli type (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{ArrowParallel instance for the Kleisli type in the Eden backend}
+\label{fig:ArrowParallelEdenKleisli}
+\end{figure}
 

### 2017-05-12 paper/arrows.tex
spacing in figure enumerations.

@@
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second} (Fig.~\ref{fig:secondImg},\ref{fig:second}),
+With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
 \begin{figure}[h]
@@
 \end{figure}
-the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},\ref{fig:***})
+the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
 \begin{figure}[h]
@@
 \end{figure}
-and the \code{\&\&\&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input (Fig.~\ref{fig:&&&Img},\ref{fig:&&&}).
+and the \code{\&\&\&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
 \begin{figure}[h]

### 2017-05-12 paper/parrows.tex
spacing in figure enumerations.

@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},\ref{fig:parEvalN}).
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
 Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \lstinline{listApp} from Appendix \ref{utilfns}, but does parallel instead of serial evaluation).

### 2017-05-12 paper/basicSkeletons.tex
added figures for basicSkeletons.tex

@@
 \subsection{Lazy \code{parEvalN}}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.7]{images/parEvalNLazy}
-\end{center}
+	\caption{Schematic depiction of parEvalNLazy}
+	\label{fig:parEvalNLazyImg}
+\end{figure}
 The function \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
@@
 (+)) [1..]} or just because we need the arrows evaluated in chunks.
-\code{parEvalNLazy} fixes this.
+\code{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
 It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}.
@@
 The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting} %$ %% formatting
+\caption{Definition of parEvalNLazy}
+\label{fig:parEvalNLazy}
+\end{figure}
 
 \subsection{Heterogenous tasks}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.7]{images/parEval2}
-\end{center}
+	\caption{Schematic depiction of parEval2}
+	\label{fig:parEval2Img}
+\end{figure}
 We have only talked about the paralellization arrows of the same type until now.
 But sometimes we want to paralellize heterogenous types as well.
-However, we can implement such a \code{parEval2} combinator which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass.
+However, we can implement such a \code{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass.
 The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
@@
 Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above.
-After the calculation is finished we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}.
+After the calculation is finished, we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}.
 The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+	\caption{Definition of parEval2}
+	\label{fig:parEval2}
+\end{figure}
 

### 2017-05-12 paper/syntacticSugar.tex
added figures for syntacticSugar.tex

@@
 \subsection{Syntactic Sugar} \label{syntacticSugar}
-For basic arrows, we have the \code{***} combinator which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
-This can easily be translated into a parallel version with \code{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \code{()} as the conf parameter in the following code snippet).
+For basic arrows, we have the \code{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
+This can easily be translated into a parallel version \code{(|***|)} (Fig.~\ref{fig:|***|}) with the use of \code{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \code{()} as the conf parameter in Fig.~\ref{fig:|***|}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Definition of (|***|) - the parallel version of (***)}
+\label{fig:|***|}
+\end{figure}
 % With this we can analogously to the serial \code{&&&}
-We define the parallel \code{|&&&|} in a similar manner to its sequential prototype.
+We define the parallel \code{(|&&&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \code{(&&&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting} % $ %% formatting
+\caption{Definition of (|\&\&\&|) - the parallel version of (\&\&\&)}
+\label{fig:|&&&|}
+\end{figure}

### 2017-05-12 paper/parrows.tex
minor fix.

@@
 
-\subsection{Multicore Haskell}
+\subsection{Multicore Haskell} \label{sec:parrows:multicore}
 The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
@@
 \end{lstlisting}% $ %% formatting
-\caption{Fully evaluating ArrowParallel instance for the Multicore backend}
+\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
 \label{fig:ArrowParallelMulticoreRdeepseq}
@@
 \end{lstlisting}
-\caption{Configurable ArrowParallel instance for the Multicore backend}
+\caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
 \label{fig:ArrowParallelMulticoreConfigurable}

### 2017-05-12 paper/futures.tex
added figures in futures.tex. fixed some wording. added jones1996concurrent to bibliography.

@@
 \section{Futures} \label{futures}
-Consider the following  parallel arrow combinator:
+Consider the parallel arrow combinator in Fig.~\ref{fig:someCombinator}
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{An example parallel Arrow combinator without Futures}
+\label{fig:someCombinator}
+\end{figure}
 In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
@@
 
-While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
 This can become a serious bottleneck %in heavy threaded applications.
@@
 	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
-	\caption[without Futures]{Communication between 4 threads without Futures}
+	\caption[without Futures]{Communication between 4 threads without Futures.
+All communication goes through the master node.
+Each bar represents one process.
+Black lines between processes represent communication.
+Colors: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
 	\label{fig:withoutFutures}
 \end{figure}
+
 This motivates for an approach that allows the nodes to communicate directly with each other.
 Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
-But as we want code written against our API to be implementation agnostic, we have to wrap this context.
-We do this with the \code{Future} typeclass:
+
+But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \code{Future} typeclass (Fig.~\ref{fig:future}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-As \code{RD} is only type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though:
+\caption{Definition of the Future typeclass}
+\label{fig:future}
+\end{figure}
+Since \code{RD} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \code{Strategy a} in the Multicore Haskell implementation of \code{ArrowParallel} in chapter \ref{sec:parrows:multicore}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-For ParMonad and Multicore we can simply use \code{MVar}s because we have shared memory in a single node:
+\caption{RD based RemoteData version of Future for the Eden backend}
+\label{fig:RDFuture}
+\end{figure}
+
+For our Par Monad and Multicore Haskell backends, we can simply use \code{MVar}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
+\fixme{explain MVars}
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-To fit the \code{ArrowParallel} instances we gave earlier, we also have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
-We need this implementation for our \code{RemoteData} wrapper,
+\caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends}
+\label{fig:MVarFuture}
+\end{figure}
+
+Furthermore, in order for these \code{Future} types to fit with the \code{ArrowParallel} instances we gave earlier, we have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
+Because \code{MVar} already ships with a \code{NFData} instance, we only have to supply two simple instances for our \code{RemoteData} type.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-while \code{MVar} already has a suitable NFData implementation.
+\caption{NFData and Trans instances for the RemoteData type.
+The Trans instance does not have any functions declared as the default implementation suffices here.
+See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
+\end{figure}
 \\\\
@@
 	\includegraphics[width=0.9\textwidth]{images/withFutures}
-	\caption[with Futures]{Communication between 4 threads with Futures}
+	\caption[with Futures]{Communication between 4 threads with Futures.
+Other than in Fig.~\ref{fig:withoutFutures}, threads communicate directly (black lines between the bars) instead of always going through the master node (bottom bar).}
 	\label{fig:withFutures}

### 2017-05-12 paper/futures.tex
minor

@@
 \end{figure}
-\\\\
+
 Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel}
+\label{fig:someCombinatorParallel}
+\end{figure}
 In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear.

### 2017-05-12 paper/syntacticSugar.tex
minor

@@
 % With this we can analogously to the serial \code{&&&}
-We define the parallel \code{(|&&&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \code{(&&&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+We define the parallel \code{(|\&\&\&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \code{(\&\&\&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
 \begin{figure}[h]

### 2017-05-12 paper/topologySkeletons.tex
minor

@@
+\FloatBarrier
 \section{Topological Skeletons}
@@
 defines an arrow that takes its input \code{a} and converts it into an infinite stream \code{[a]} of it.
-Using this to our advantage gives us a first draft of a pipe implementation by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{&&&} and then only picking the first element of the resulting list with \code{arr last}:
+Using this to our advantage gives us a first draft of a pipe implementation by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{(\&\&\&)} and then only picking the first element of the resulting list with \code{arr last}:
 \begin{lstlisting}[frame=htrbl]

### 2017-05-12 paper/mapSkeletons.tex
added figures to mapSkeletons.tex

@@
+\FloatBarrier
 \section{Map-based Skeletons}
@@
 \subsection{Parallel map}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.7]{images/parMap}
-\end{center}
-\code{parMap} is probably the most common skeleton for parallel programs.
+	\caption{Schematic depiction of parMap}
+	\label{fig:parMapImg}
+\end{figure}
+\code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
 We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}.
 Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Definition of parMap}
+\label{fig:parMap}
+\end{figure}
 
 \subsection{Lazy parallel map}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.7]{images/parMapStream}
-\end{center}
-As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
-So it makes sense to also have a \code{parMapStream} which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
+	\caption{Schematic depiction of parMapStream}
+	\label{fig:parMapStreamImg}
+\end{figure}
+As \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
+So it makes sense to also have a \code{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Definition of parMapStream}
+\label{fig:parMapStream}
+\end{figure}
 
 \subsection{Statically load-balancing parallel map}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.7]{images/farm}
-\end{center}
-\code{parMap} spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
-This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
+	\caption{Schematic depiction of farm}
+	\label{fig:farmImg}
+\end{figure}
+A \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
+This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Definition of farm}
+\label{fig:farm}
+\end{figure}
 
 \subsection{farmChunk}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.7]{images/farmChunk}
-\end{center}
-As \code{farm} is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
-So we define \code{farmChunk} which uses \code{parEvalNLazy} instead of \code{parEvalN} like this:
+\caption{Schematic depiction of farmChunk}
+\label{fig:farmChunkImg}
+\end{figure}
+Since a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
+So we define \code{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \code{parEvalNLazy} instead of \code{parEvalN}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Definition of farmChunk}
+\label{fig:farmChunk}
+\end{figure}
 
@@
 
+\mbcomment{this requires some work. Either change this to use futures or cut, yes.}
 -- this does not completely adhere to Google's definition of Map Reduce as it
@@
 \end{lstlisting}
+\caption{Definition of parMapReduceDirect}
+\label{fig:parMapReduceDirect}
+\end{figure}
 

### 2017-05-12 paper/mapSkeletons.tex
wip on making things into figures (topologySkeletons.tex)

@@
 A \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
-This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
+This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
 \begin{figure}[h]

### 2017-05-12 paper/topologySkeletons.tex
wip on making things into figures (topologySkeletons.tex)

@@
 The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
-We implement this skeleton using the \code{ArrowLoop} typeclass which gives us the \code{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express loop like computations.
-For example this
+We implement this skeleton using the \code{ArrowLoop} typeclass which gives us the \code{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
+For example %this
+\mbcomment{das kann man hier so lassen, oder?}
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-,which is the same as
+, which is the same as
 \begin{lstlisting}[frame=htrbl]
@@
 defines an arrow that takes its input \code{a} and converts it into an infinite stream \code{[a]} of it.
-Using this to our advantage gives us a first draft of a pipe implementation by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{(\&\&\&)} and then only picking the first element of the resulting list with \code{arr last}:
+Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{(\&\&\&)} and then only picking the first element of the resulting list with \code{arr last}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 	arr last
-\end{lstlisting}
-where \code{lazy} is defined as:
-\begin{lstlisting}[frame=htrbl]
+
 lazy :: (Arrow arr) => arr [a] [a]
@@
 \end{lstlisting}
-Note that here the use of \code{lazy} is essential as without it programs using this definition would never halt.
-We need to enforce that the evaluation of the input \code{[a]} terminates before passing it into \code{parEvalN}.
+\caption{A first draft of the pipe skeleton expressed with parallel arrows.
+Note that the use of \code{lazy} is essential as without it programs using this definition would never halt.
+We need to enforce that the evaluation of the input \code{[a]} terminates before passing it into \code{parEvalN}.}
+\label{fig:pipeSimple}
+\end{figure}
 
-However, using this definition directly, will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures.
+However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
+Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \code{pipe} \ref{fig:pipe}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Final definition of the pipe skeleton which uses Futures}
+\label{fig:pipe}
+\end{figure}
+
 Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
 \code{arr a b} and \code{arr b c}.
-By wrapping these two arrows inside a common type we obtain
+By wrapping these two arrows inside a common type we obtain \code{pipe2} (Fig.~\ref{fig:pipe2}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-Note that extensive use of this combinator over \code{pipe} with a hand-written combination data-type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \code{|>>>|} which is semantically equivalent to \code{>>>} in a similar manner to the other parallel syntactic sugar from Section~\ref{syntacticSugar}:
+\caption{Definition of pipe2}
+\label{fig:pipe2}
+\end{figure}
+
+Note that extensive use of this combinator over \code{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Nonetheless, we can define a parallel piping operator \code{(|>>>|)} (Fig.~\ref{fig:|>>>|}, which is semantically equivalent to \code{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Definition of \texttt{(|>>>|)}}
+\label{fig:|>>>|}
+\end{figure}
 
 \subsection{Ring skeleton}
-\begin{center}
+\begin{figure}[h]
 	\includegraphics[scale=0.75]{images/ring}
-\end{center}
-Eden comes with a ring skeleton implementation that allows the computation of a function \code{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
+	\caption{Schematic depiction of a ring skeleton}
+	\label{fig:ringImg}
+\end{figure}
+Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \code{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
 Its input is a node function \code{i -> r -> (o, r)} in which \code{r} serves as the intermediary output that gets send to the neighbour of each node.
-This data is sent over direct communication channels (remote data).
-They define it as \citep{eden_skel_topology}
+This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Eden's definition of the ring skeleton \citep{eden_skel_topology}}
+\label{fig:ringEden}
+\end{figure}
 %\end{lstlisting}
@@
 In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
-Similarly to the pipe, we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
+Similarly to the \code{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
 %\begin{quotation}

### 2017-05-15 paper/topologySkeletons.tex
all the figures in topologySkeletons.tex

@@
 	\includegraphics[scale=0.75]{images/ring}
-	\caption{Schematic depiction of a ring skeleton}
+	\caption{Schematic depiction of the ring skeleton}
 	\label{fig:ringImg}
@@
 Finally we unzip this list into \code{([o], [fut r])}.
-Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow.
-
-This gives us the following complete definition for the \code{ring} combinator:
+Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
     arr unzip)
-\end{lstlisting}
-with \code{rightRotate}:
-\begin{lstlisting}[frame=htrbl]
+
 rightRotate :: (Arrow arr) => arr [a] [a]
@@
   xs -> last xs : init xs
-\end{lstlisting} %$ %% formatting
+\end{lstlisting}
+\caption{Final Definition of the ring Skeleton}
+\label{fig:ringFinal}
+\end{figure}
+ %$ %% formatting
 % and lazy:
@@
 \subsection{Torus skeleton}
-\begin{center}
+\begin{figure}
 	\includegraphics[scale=0.75]{images/torus}
-\end{center}
-If we take the concept of a ring one dimension further, we get a torus. Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
+	\caption{Schematic depiction of the torus skeleton}
+	\label{fig:ringTorusImg}
+\end{figure}
+If we take the concept of a ring one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
+Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
 \\\\
-With our parallel Arrows we implement a torus combinator yet again with the help of the \code{ArrowLoop} typeclass.
+With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \code{ArrowLoop} typeclass.
 \\\\
@@
 \\\\
-The complete definition of the \code{torus} combinator is:
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 	arr (map unzip3) >>> arr unzip3 >>> threetotwo)
-\end{lstlisting}
-with uncurry3, lazyzip3, and threetotwo from the Appendix, and
-\begin{lstlisting}[frame=htrbl]
+
 ptorus :: (Arrow arr, Future fut a, Future fut b) =>
@@
 \end{lstlisting}
-As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}. Adapting this nodefunction to our Arrow API gives us:
+\caption{Definition of the torus skeleton}
+\label{fig:torus}
+\end{figure}
+As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}.
+Their nodefunction can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
+\begin{figure}[h]
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\caption{Adapted nodefunction for matrix multiplication with the torus from Fig.~\ref{fig:torus}}
+\label{fig:torusMatMult}
+\end{figure}
 If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.\olcomment{much more details on this!}
@@
 
+\FloatBarrier
+
 %%% Local Variables:

### 2017-05-15 paper/basicSkeletons.tex
restructured some stuff. more cross references.

@@
 
-\subsection{Lazy \code{parEvalN}}
+\subsubsection{Lazy \code{parEvalN}}
 \begin{figure}[h]
@@
 It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}.
-These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}.
+These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp} (Fig.~\ref{fig:listApp}).
 The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
@@
 
-\subsection{Heterogenous tasks}
+\subsubsection{Heterogenous tasks}
 \begin{figure}[h]

### 2017-05-15 paper/mapSkeletons.tex
restructured some stuff. more cross references.

@@
 	shuffle
+
+unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
+unshuffle n = arr (\xs -> [takeEach n (drop i xs) | i <- [0..n-1]])
+
+takeEach :: Int -> [a] -> [a]
+takeEach n [] = []
+takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
+
+shuffle :: (Arrow arr) => arr [[a]] [a]
+shuffle = arr (concat . transpose)
 \end{lstlisting}
-\caption{Definition of farm}
+\caption{Definition of farm. \code{unshuffle}, \code{takeEach}, \code{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
 \label{fig:farm}

### 2017-05-15 paper/parrows.tex
minor
restructured some stuff. more cross references.

@@
 
-\subsection{Multicore Haskell} \label{sec:parrows:multicore}
+\subsection{ArrowParallel instances}
+
+\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
 The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
@@
 \end{figure}
-\subsection{Par Monad}
+\subsubsection{Par Monad}
 The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function.
@@
 
-We therefore apply each function to its corresponding input value with \code{app} and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call.
+We therefore apply each function to its corresponding input value with \code{listApp} (Fig.~\ref{fig:listApp}) and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call.
 This yields a list \code{[Par (IVar b)]}, which we then convert into \code{Par [IVar b]} with \code{arr sequenceA}.
@@
 
-\subsection{Eden}
+\subsubsection{Eden}
 For the Multicore and ParMonad implementation we could use general instances of \code{ArrowParallel} that just require the \code{ArrowApply} and \code{ArrowChoice} typeclasses.
@@
 
+\FloatBarrier
+
 \subsection{Impact of parallel Arrows}

### 2017-05-15 paper/topologySkeletons.tex
restructured some stuff. more cross references.

@@
 
-\subsection{Ring skeleton}
+\subsection{Ring skeleton} \label{sec:ring}
 \begin{figure}[h]
@@
 rightRotate xs =  last xs : init xs
+
+lazy :: [a] -> [a]
+lazy ~(x:xs) = x : lazy xs
 \end{lstlisting}
@@
   xs -> last xs : init xs
+
+lazy :: (Arrow arr) => arr [a] [a]
+lazy = arr (\ ~(x:xs) -> x : lazy xs)
 \end{lstlisting}
@@
 \end{figure}
-If we take the concept of a ring one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
+If we take the concept of a ring from \ref{sec:ring} one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
@@
 We first have to \code{shuffle} all the inputs to then pass it into \code{parMap conf (ptorus f)} which yields us \code{[(d, fut a, fut b)]}.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app} from the \code{ArrowApply} typeclass.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app :: arr (arr a b, a) c} from the \code{ArrowApply} typeclass.
 Finally, we unpack this matrix \code{[[[(d, fut a, fut b)]]} with \code{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \code{([[d]], ([[fut a]], [[fut b]]))}.
@@
 	arr (\ ~(d, a, b) -> (d, put a, put b))
+
+uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
+uncurry3 f (a, (b, c)) = f a b c
+
+lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
+lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
+
+threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
+threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
 \end{lstlisting}

### 2017-05-17 paper/content/parrows.tex
initial

--- 2017-05-15 paper/parrows.tex
@@
 \end{figure}
-To get our configurable \code{ArrowParallel} instance, we simply unwrap the strategy and pass it to \code{parList} as in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
+To get our configurable \code{ArrowParallel} instance, we simply unwrap the strategy and pass it to \code{parList} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
 \begin{figure}[h]

### 2017-05-17 paper/content/relwork.tex
initial

\section{Related Work}
\label{sec:related-work}

\olcomment{arrows or Arrows?}

\subsection{Parallel Haskells}
Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \code{Par} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work. We use these languages as backends: our library can switch from one to other at user's command.

HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \code{Par} monad to heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \code{Par} monad.
%
Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015} and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden. This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,brown2010ever,dieterle2010skeleton,delaEncina2011,Dieterle2013}.

More different approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582,}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
%
The Haskell--GPU bridge Accelerate \cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313} deserves a special mention. Accelerate is completely orthogonal to our approach. \citeauthor{marlow2013parallel} authored a recent book \citeyear{marlow2013parallel} on parallel Haskells.

\subsection{Algorithmic skeletons}

Algorithmic skeletons were introduced by \citet{Cole1989}.
Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,brown2010ever,lobachev-phd,Lobachev2012}, iteration skeletons \cite{Dieterle2013}. The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel. \citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeleton to stable process networks.

\subsection{Arrows}

Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\code{->}. \citet{Hughes2005} is a tutorial on arrows. Some theoretical details on arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But perhaps most prominent application of arrows is functional reactive programming \cite{Hudak2003}.\olcomment{cite more!}

\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of arrows that capsule the computation more than regular arrows do and thus enable optimizations. Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution. In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \code{parEvalN} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
\citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach. They basically use arrows to orchestrate several tasks in robotics. We propose a general interface for parallel programming, remaining completely in Haskell.

\subsection{Other languages}

Although this work is centred on Haskell implementation of arrows, it is applicable to any functional programming language where parallel evaluation and arrows can be defined. Our experiments with our approach in Frege language (which is basically Haskell on the JVM) were quite successful, we were able to use typical Java libraries for parallelism. However, it is beyond the scope of this work.

\citet{achten2004arrows,achten2007arrow} use an arrow implementation in Clean for better handling of typical GUI tasks. \citet{Dagand:2009:ORD:1481861.1481870} used arrows in OCaml in the implementation of a distributed system.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-05-17 paper/content/syntacticSugar.tex
initial

--- 2017-05-12 paper/syntacticSugar.tex
@@
-\subsection{Syntactic Sugar} \label{syntacticSugar}
+\subsubsection{Syntactic Sugar} \label{syntacticSugar}
 For basic arrows, we have the \code{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.

### 2017-05-17 paper/content/topologySkeletons.tex
initial

--- 2017-05-15 paper/topologySkeletons.tex
@@
 The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{Loogen2012, eden_skel_topology}.
-These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express such skeletons with Parallel Arrows as well.
+These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
 

### 2017-05-17 paper/content/utilityFunctions.tex
initial

\section{Utility Functions}\label{utilfns}
To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \code{map}, \code{foldl} and \code{zipWith} on arrows.

The \code{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows}:
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
mapArr f =
	arr listcase >>>
	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
	where listcase [] = Left ()
	      listcase (x:xs) = Right (x,xs)
\end{lstlisting}
\caption{\code{map} over arrows}
\label{fig:mapArr}
\end{figure}
Similarly, we can also define \code{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \code{arr (b, a) b} with a neutral element \code{b} to \code{arr [a] b}:
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
foldlArr f b =
	arr listcase >>>
	arr (const b) |||
		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
	where listcase [] = Left []
	      listcase (x:xs) = Right (x,xs)
\end{lstlisting}
\caption{\code{foldl} over arrows}
\label{fig:foldlArr}
\end{figure}
Finally, with the help of \code{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
\end{lstlisting}
\caption{\code{zipWith} over arrows}
\label{fig:zipWithArr}
\end{figure}
 %$ %% formatting
These combinators make use of the \code{ArrowChoice} type class which provides the \code{(|||)} combinator. It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
\end{lstlisting}
\caption{Type signature of \code{(|||)}}
\label{fig:codeSig|||}
\end{figure}

With the zipWithArr combinator we can also write a combinator \code{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
\end{lstlisting}
\caption{Definition of \code{listApp}}
\label{fig:listApp}
\end{figure}
% $ %% formatting
. Note that  this additionally makes use of the \code{ArrowApply} typeclass that allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.

% $ %% formatting

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-05-17 paper/content/examples/parallelism-eden.hs

@@ -1,0 +1,32 @@
+{-
+The MIT License (MIT)
+
+Copyright (c) 2016 Martin Braun
+
+Permission is hereby granted, free of charge, to any person obtaining
+a copy of this software and associated documentation files (the
+"Software"), to deal in the Software without restriction, including
+without limitation the rights to use, copy, modify, merge, publish,
+distribute, sublicense, and/or sell copies of the Software, and to
+permit persons to whom the Software is furnished to do so, subject to
+the following conditions:
+
+The above copyright notice and this permission notice shall be included
+in all copies or substantial portions of the Software.
+
+THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY
+CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
+TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
+SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+-}
+module Main where
+
+import Control.Parallel.Eden
+
+parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = spawnF fs as
+
+main = do print $ parEvalN [(*2), (+1)] [1::Int, 2]

### 2017-05-17 paper/content/examples/parallelism-multicore.hs

@@ -1,0 +1,32 @@
+{-
+The MIT License (MIT)
+
+Copyright (c) 2016 Martin Braun
+
+Permission is hereby granted, free of charge, to any person obtaining
+a copy of this software and associated documentation files (the
+"Software"), to deal in the Software without restriction, including
+without limitation the rights to use, copy, modify, merge, publish,
+distribute, sublicense, and/or sell copies of the Software, and to
+permit persons to whom the Software is furnished to do so, subject to
+the following conditions:
+
+The above copyright notice and this permission notice shall be included
+in all copies or substantial portions of the Software.
+
+THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY
+CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
+TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
+SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+-}
+module Main where
+
+import Control.Parallel.Strategies
+
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = zipWith ($) fs as `using` parList rdeepseq
+
+main = do print $ parEvalN [(*2), (+1)] [1::Int, 2]

### 2017-05-17 paper/content/examples/parallelism-parmonad.hs

@@ -1,0 +1,32 @@
+{-
+The MIT License (MIT)
+
+Copyright (c) 2016 Martin Braun
+
+Permission is hereby granted, free of charge, to any person obtaining
+a copy of this software and associated documentation files (the
+"Software"), to deal in the Software without restriction, including
+without limitation the rights to use, copy, modify, merge, publish,
+distribute, sublicense, and/or sell copies of the Software, and to
+permit persons to whom the Software is furnished to do so, subject to
+the following conditions:
+
+The above copyright notice and this permission notice shall be included
+in all copies or substantial portions of the Software.
+
+THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY
+CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
+TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
+SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+-}
+module Main where
+
+import Control.Monad.Par
+
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = runPar $ (sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+
+main = do print $ parEvalN [(*2), (+1)] [1::Int, 2]

### 2017-05-17 paper/content/arrows.tex
renamed \code to \inlinecode

--- 2017-05-12 paper/arrows.tex
@@
 Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
-An arrow \code{arr a b} represents  a computation that converts an input \code{a} to an output \code{b}.
+An arrow \inlinecode{arr a b} represents  a computation that converts an input \inlinecode{a} to an output \inlinecode{b}.
 This is defined in the arrow typeclass:
@@
 	\subcaptionbox{arr}{\includegraphics[scale=0.6]{images/arr}}
-	\subcaptionbox{composition \code{(>>>)}}{\includegraphics[scale=0.6]{images/compose}}
+	\subcaptionbox{composition \inlinecode{(>>>)}}{\includegraphics[scale=0.6]{images/compose}}
 	\subcaptionbox{first}{\includegraphics[scale=0.6]{images/first}}
@@
 \lstinline{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \lstinline{return}.
-The \lstinline{>>>} operator is analogous to the monadic composition  \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}.
+The \lstinline{>>>} operator is analogous to the monadic composition  \lstinline{>>=} and combines two arrows \inlinecode{arr a b} and \inlinecode{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \inlinecode{arr a c}.
 Lastly, the \lstinline{first} operator  takes the input arrow from \lstinline{b} to \lstinline{c} and converts it into an arrow on pairs with the second argument untouched.
@@
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
+With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \inlinecode{first}, called \inlinecode{second} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
 \begin{figure}[h]
@@
 \end{figure}
-and the \code{\&\&\&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+and the \inlinecode{\&\&\&} combinator that constructs an arrow which outputs two different values like \inlinecode{***}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
 \begin{figure}[h]
@@
 \end{figure}
-% The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
+% The benefit of using the \inlinecode{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
 Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
@@
 % \\\\
-% \textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used.
-We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}
+% \textit{Note: In the definitions Hughes gave in his paper, the notation \inlinecode{a b c} for an arrow from \inlinecode{b} to \inlinecode{c} is used.
+We use the equivalent definition \inlinecode{arr a b} for an arrow from \inlinecode{a} to \inlinecode{b} instead, to make it easier to find the arrow type in type signatures.}
 %

### 2017-05-17 paper/content/basicSkeletons.tex
renamed \code to \inlinecode

--- 2017-05-15 paper/basicSkeletons.tex
@@
 \label{sec:extending-interface}
-With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
+With the \inlinecode{ArrowParallel} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
 These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
 
-\subsubsection{Lazy \code{parEvalN}}
+\subsubsection{Lazy \inlinecode{parEvalN}}
 \begin{figure}[h]
@@
 \end{figure}
-The function \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
+The function \inlinecode{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
 Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
-\code{map (arr .
+\inlinecode{map (arr .
 (+)) [1..]} or just because we need the arrows evaluated in chunks.
-\code{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
-It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}.
-These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp} (Fig.~\ref{fig:listApp}).
-The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
+\inlinecode{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
+It works by first chunking the input from \inlinecode{[a]} to \inlinecode{[[a]]} with the given \inlinecode{ChunkSize} in \inlinecode{arr (chunksOf chunkSize)}.
+These chunks are then fed into a list \inlinecode{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \inlinecode{ChunkSize} into the regular parEvalN by using \inlinecode{listApp} (Fig.~\ref{fig:listApp}).
+The resulting \inlinecode{[[b]]} is lastly converted into \inlinecode{[b]} with \inlinecode{arr concat}.
 \begin{figure}[h]
@@
 But sometimes we want to paralellize heterogenous types as well.
-However, we can implement such a \code{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass.
-The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
+However, we can implement such a \inlinecode{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into a new parallel arrow \inlinecode{arr (a, c) (b, d)} quite easily with the help of the \inlinecode{ArrowChoice} typeclass.
+The idea is to use the \inlinecode{+++} combinator which combines two arrows \inlinecode{arr a b} and \inlinecode{arr c d} and transforms them into \inlinecode{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
 \\\\
-We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]} by first tagging the two inputs with \code{Left} and \code{Right} and wrapping the right element in a singleton list with \code{return} so that we can combine them with \code{arr (uncurry (:))}.
-Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above.
-After the calculation is finished, we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}.
-The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
+We start by transforming the \inlinecode{(a, c)} input into a 2-element list \inlinecode{[Either a c]} by first tagging the two inputs with \inlinecode{Left} and \inlinecode{Right} and wrapping the right element in a singleton list with \inlinecode{return} so that we can combine them with \inlinecode{arr (uncurry (:))}.
+Next, we feed this list into a parallel arrow running on 2 instances of \inlinecode{f +++ g} as described above.
+After the calculation is finished, we convert the resulting \inlinecode{[Either b d]} into \inlinecode{([b], [d])} with \inlinecode{arr partitionEithers}.
+The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \inlinecode{(b, d)} in the last step.
 \begin{figure}[h]

### 2017-05-17 paper/content/futures.tex
renamed \code to \inlinecode

--- 2017-05-12 paper/futures.tex
@@
 \end{figure}
-In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
+In a distributed environment, the resulting arrow of this combinator first evaluates all \inlinecode{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \inlinecode{[arr b c]} in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
 
-While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \inlinecode{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \inlinecode{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
 This can become a serious bottleneck %in heavy threaded applications.
@@
 This motivates for an approach that allows the nodes to communicate directly with each other.
-Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
+Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \inlinecode{RD} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
 
-But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \code{Future} typeclass (Fig.~\ref{fig:future}).
+But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \inlinecode{Future} typeclass (Fig.~\ref{fig:future}).
 \begin{figure}[h]
@@
 \end{figure}
-Since \code{RD} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \code{Strategy a} in the Multicore Haskell implementation of \code{ArrowParallel} in chapter \ref{sec:parrows:multicore}).
+Since \inlinecode{RD} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \inlinecode{Strategy a} in the Multicore Haskell implementation of \inlinecode{ArrowParallel} in chapter \ref{sec:parrows:multicore}).
 \begin{figure}[h]
@@
 
-For our Par Monad and Multicore Haskell backends, we can simply use \code{MVar}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
+For our Par Monad and Multicore Haskell backends, we can simply use \inlinecode{MVar}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
 \fixme{explain MVars}
@@
 
-Furthermore, in order for these \code{Future} types to fit with the \code{ArrowParallel} instances we gave earlier, we have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
-Because \code{MVar} already ships with a \code{NFData} instance, we only have to supply two simple instances for our \code{RemoteData} type.
+Furthermore, in order for these \inlinecode{Future} types to fit with the \inlinecode{ArrowParallel} instances we gave earlier, we have to give the necessary \inlinecode{NFData} and \inlinecode{Trans} instances - the latter only being needed in Eden.
+Because \inlinecode{MVar} already ships with a \inlinecode{NFData} instance, we only have to supply two simple instances for our \inlinecode{RemoteData} type.
 \begin{figure}[h]

### 2017-05-17 paper/content/mapSkeletons.tex
renamed \code to \inlinecode

--- 2017-05-15 paper/mapSkeletons.tex
@@
 \end{figure}
-\code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
-We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}.
-Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
+\inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
+We can implement it with \inlinecode{ArrowParallel} by repeating an arrow \inlinecode{arr a b} and then passing it into \inlinecode{parEvalN} to get an arrow \inlinecode{arr [a] [b]}.
+Just like \inlinecode{parEvalN}, \inlinecode{parMap} is 100 \% strict.
 \begin{figure}[h]
@@
 \end{figure}
-As \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
-So it makes sense to also have a \code{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
+As \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \inlinecode{parEvalN} compared to \inlinecode{parEvalNLazy}.
+So it makes sense to also have a \inlinecode{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \inlinecode{parMap}, but uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
 \begin{figure}[h]
@@
 \end{figure}
-A \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
-This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
+A \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \inlinecode{ArrowParallel} we gave in this paper).
+This can be quite wasteful and a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \inlinecode{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
 \begin{figure}[h]
@@
 \end{lstlisting}
-\caption{Definition of farm. \code{unshuffle}, \code{takeEach}, \code{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
+\caption{Definition of farm. \inlinecode{unshuffle}, \inlinecode{takeEach}, \inlinecode{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
 \label{fig:farm}
@@
 \end{figure}
-Since a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
-So we define \code{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \code{parEvalNLazy} instead of \code{parEvalN}.
+Since a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \inlinecode{parMap} with a different work distribution, it is, again, 100\% strict.
+So we define \inlinecode{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
 \begin{figure}[h]

### 2017-05-17 paper/content/motivation.tex
renamed \code to \inlinecode

--- 2017-05-11 paper/motivation.tex
@@
 Section~\ref{futures} defines futures for Parallel Arrows, this concept enables better communication.
-Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \code{map} with and without load balancing, \code{map-reduce}) in our newly defined dialect.
-More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\code{pipe}, \code{ring}, \code{torus}).
+Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \inlinecode{map} with and without load balancing, \inlinecode{map-reduce}) in our newly defined dialect.
+More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\inlinecode{pipe}, \inlinecode{ring}, \inlinecode{torus}).
 Section~\ref{sec:benchmarks} shows the benchmark results.

### 2017-05-17 paper/content/parallelHaskells.tex
renamed \code to \inlinecode

--- 2017-05-12 paper/parallelHaskells.tex
@@
 
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \inlinecode{a -> b} in parallel:
 
@@
 \olcomment{make them to real figures? with environment, reference and such?}
-Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \code{parEvalN} with several different parallel Haskells.
+Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \inlinecode{parEvalN} with several different parallel Haskells.
 
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \code{using :: a -> Strategy a -> a}.
-For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
-We then evaluate this lazy list \code{[b]} according to a \code{Strategy [b]} with the \code{using :: a -> Strategy a -> a} operator.
-We construct this strategy with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
-To ensure that programs that use \code{parEvalN} have the correct evaluation order, we annotate the computation with \code{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \code{parEvalN} computations.
-This is particularly necessary in circular communication topologies like in the \code{torus} or \code{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \code{pseq} during testing for this paper.
+Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \inlinecode{using :: a -> Strategy a -> a}.
+For \inlinecode{parEvalN} this means that we can just apply the list of functions \inlinecode{[a -> b]} to the list of inputs \inlinecode{[a]} by zipping them with the application operator \inlinecode{\$}.
+We then evaluate this lazy list \inlinecode{[b]} according to a \inlinecode{Strategy [b]} with the \inlinecode{using :: a -> Strategy a -> a} operator.
+We construct this strategy with \inlinecode{parList :: Strategy a -> Strategy [a]} and \inlinecode{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
+To ensure that programs that use \inlinecode{parEvalN} have the correct evaluation order, we annotate the computation with \inlinecode{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \inlinecode{parEvalN} computations.
+This is particularly necessary in circular communication topologies like in the \inlinecode{torus} or \inlinecode{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \inlinecode{pseq} during testing for this paper.
 
@@
 \subsubsection{ParMonad}
-The \code{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
+The \inlinecode{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
 
-Our parallel evaluation function \code{parEvalN} can be defined by zipping the list of \code{[a -> b]} with the list of inputs \code{[a]} with the application operator \code{\$} just like with Multicore Haskell.
-Then, we map over this not yet evaluated lazy list of results \code{[b]} with \code{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \code{[Par (IVar b)]}, which we convert to \code{Par [IVar b]} with \code{sequenceA}.
-We wait for the computations to finish by mapping over the \code{IVar b}'s inside the \code{Par} monad with \code{get}.
-This results in \code{Par [b]}.
-We finally execute this process with \code{runPar} to finally get \code{[b]} again.
+Our parallel evaluation function \inlinecode{parEvalN} can be defined by zipping the list of \inlinecode{[a -> b]} with the list of inputs \inlinecode{[a]} with the application operator \inlinecode{\$} just like with Multicore Haskell.
+Then, we map over this not yet evaluated lazy list of results \inlinecode{[b]} with \inlinecode{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \inlinecode{[Par (IVar b)]}, which we convert to \inlinecode{Par [IVar b]} with \inlinecode{sequenceA}.
+We wait for the computations to finish by mapping over the \inlinecode{IVar b}'s inside the \inlinecode{Par} monad with \inlinecode{get}.
+This results in \inlinecode{Par [b]}.
+We finally execute this process with \inlinecode{runPar} to finally get \inlinecode{[b]} again.
 
@@
 
-While Eden also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
+While Eden also comes with a monad \inlinecode{PA} for parallel evaluation, it also ships with a completely functional interface that includes
 %\\
-%\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
+%\inlinecode{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
 %\\
-a \code{spawnF} function that
+a \inlinecode{spawnF} function that
 %This 
-allows us to define \code{parEvalN} directly:
+allows us to define \inlinecode{parEvalN} directly:
 

### 2017-05-17 paper/content/syntacticSugar.tex
renamed \code to \inlinecode

--- 2017-05-12 paper/syntacticSugar.tex
@@
-\subsection{Syntactic Sugar} \label{syntacticSugar}
-For basic arrows, we have the \code{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
-This can easily be translated into a parallel version \code{(|***|)} (Fig.~\ref{fig:|***|}) with the use of \code{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \code{()} as the conf parameter in Fig.~\ref{fig:|***|}).
+\subsubsection{Syntactic Sugar} \label{syntacticSugar}
+For basic arrows, we have the \inlinecode{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into an arrow \inlinecode{arr (a, c) (b, d)} which does both computations at once.
+This can easily be translated into a parallel version \inlinecode{(|***|)} (Fig.~\ref{fig:|***|}) with the use of \inlinecode{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \inlinecode{()} as the conf parameter in Fig.~\ref{fig:|***|}).
 \begin{figure}[h]
@@
 \end{figure}
-% With this we can analogously to the serial \code{&&&}
-We define the parallel \code{(|&&&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \code{(&&&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+% With this we can analogously to the serial \inlinecode{&&&}
+We define the parallel \inlinecode{(|\&\&\&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \inlinecode{(\&\&\&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
 \begin{figure}[h]

### 2017-05-17 paper/content/topologySkeletons.tex
renamed \code to \inlinecode

@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{Loogen2012, eden_skel_topology}.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them a \inlinecode{pipe}, a \inlinecode{ring} and a \inlinecode{torus} implementation \cite{Loogen2012, eden_skel_topology}.
 These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
@@
 
-The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
-We implement this skeleton using the \code{ArrowLoop} typeclass which gives us the \code{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
+The parallel pipe skeleton is semantically equivalent to folding over a list \inlinecode{[arr a a]} of arrows with \inlinecode{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
+We implement this skeleton using the \inlinecode{ArrowLoop} typeclass which gives us the \inlinecode{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
 For example %this
@@
 \end{lstlisting}
-defines an arrow that takes its input \code{a} and converts it into an infinite stream \code{[a]} of it.
-Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{(\&\&\&)} and then only picking the first element of the resulting list with \code{arr last}.
+defines an arrow that takes its input \inlinecode{a} and converts it into an infinite stream \inlinecode{[a]} of it.
+Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \inlinecode{parEvalN conf fs} inside the second argument of \inlinecode{(\&\&\&)} and then only picking the first element of the resulting list with \inlinecode{arr last}.
 \begin{figure}[h]
@@
 \caption{A first draft of the pipe skeleton expressed with parallel arrows.
-Note that the use of \code{lazy} is essential as without it programs using this definition would never halt.
-We need to enforce that the evaluation of the input \code{[a]} terminates before passing it into \code{parEvalN}.}
+Note that the use of \inlinecode{lazy} is essential as without it programs using this definition would never halt.
+We need to enforce that the evaluation of the input \inlinecode{[a]} terminates before passing it into \inlinecode{parEvalN}.}
 \label{fig:pipeSimple}
@@
 However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \code{pipe} \ref{fig:pipe}.
+Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \inlinecode{pipe} \ref{fig:pipe}.
 \begin{figure}[h]
@@
 Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
-\code{arr a b} and \code{arr b c}.
-By wrapping these two arrows inside a common type we obtain \code{pipe2} (Fig.~\ref{fig:pipe2}).
+\inlinecode{arr a b} and \inlinecode{arr b c}.
+By wrapping these two arrows inside a common type we obtain \inlinecode{pipe2} (Fig.~\ref{fig:pipe2}).
 \begin{figure}[h]
@@
 
-Note that extensive use of this combinator over \code{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \code{(|>>>|)} (Fig.~\ref{fig:|>>>|}, which is semantically equivalent to \code{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+Note that extensive use of this combinator over \inlinecode{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Nonetheless, we can define a parallel piping operator \inlinecode{(|>>>|)} (Fig.~\ref{fig:|>>>|}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
 \begin{figure}[h]
@@
 \end{figure}
-Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \code{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
-Its input is a node function \code{i -> r -> (o, r)} in which \code{r} serves as the intermediary output that gets send to the neighbour of each node.
+Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \inlinecode{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
+Its input is a node function \inlinecode{i -> r -> (o, r)} in which \inlinecode{r} serves as the intermediary output that gets send to the neighbour of each node.
 This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
@@
 
-We can rewrite its functionality easily with the use of \code{loop} as the definition of the node function, \code{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \code{loop}'s \code{arr (a, b) (c, b) -> arr a c}.
-In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
-Similarly to the \code{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
+We can rewrite its functionality easily with the use of \inlinecode{loop} as the definition of the node function, \inlinecode{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \inlinecode{loop}'s \inlinecode{arr (a, b) (c, b) -> arr a c}.
+In each iteration we start by rotating the intermediary input from the nodes \inlinecode{[fut r]} with \inlinecode{second (rightRotate >>> lazy)}.
+Similarly to the \inlinecode{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \inlinecode{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
 %\begin{quotation}
@@
 %\end{quotation}
-Next, we zip the resulting \code{([i], [fut r])} to \code{[(i, fut r)]} with \code{arr (uncurry zip)} so we can feed that into a our input arrow \code{arr (i, r) (o, r)}, which we transform into \code{arr (i, fut r) (o, fut r)} before lifting it to \code{arr [(i, fut r)] [(o, fut r)]} to get a list \code{[(o, fut r)]}.
-Finally we unzip this list into \code{([o], [fut r])}.
-Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
+Next, we zip the resulting \inlinecode{([i], [fut r])} to \inlinecode{[(i, fut r)]} with \inlinecode{arr (uncurry zip)} so we can feed that into a our input arrow \inlinecode{arr (i, r) (o, r)}, which we transform into \inlinecode{arr (i, fut r) (o, fut r)} before lifting it to \inlinecode{arr [(i, fut r)] [(o, fut r)]} to get a list \inlinecode{[(o, fut r)]}.
+Finally we unzip this list into \inlinecode{([o], [fut r])}.
+Plugging this arrow \inlinecode{arr ([i], [fut r]) ([o], fut r)} into the definition of \inlinecode{loop} from earlier gives us \inlinecode{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
 \begin{figure}[h]
@@
 \\\\
-With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \code{ArrowLoop} typeclass.
+With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \inlinecode{ArrowLoop} typeclass.
 \\\\
 Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple \code{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \code{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \code{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \code{[[fut a]]} horizontally and \code{[[fut b]]} vertically.
-Then, we once again zip the inputs for the input arrow with \code{arr (uncurry3 zipWith3 lazyzip3)} from \code{([[c]], ([[fut a]], [[fut b]]))} to \code{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple \inlinecode{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \inlinecode{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \inlinecode{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \inlinecode{[[fut a]]} horizontally and \inlinecode{[[fut b]]} vertically.
+Then, we once again zip the inputs for the input arrow with \inlinecode{arr (uncurry3 zipWith3 lazyzip3)} from \inlinecode{([[c]], ([[fut a]], [[fut b]]))} to \inlinecode{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
 \\\\
 This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
-We first have to \code{shuffle} all the inputs to then pass it into \code{parMap conf (ptorus f)} which yields us \code{[(d, fut a, fut b)]}.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app :: arr (arr a b, a) c} from the \code{ArrowApply} typeclass.
-Finally, we unpack this matrix \code{[[[(d, fut a, fut b)]]} with \code{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \code{([[d]], ([[fut a]], [[fut b]]))}.
+We first have to \inlinecode{shuffle} all the inputs to then pass it into \inlinecode{parMap conf (ptorus f)} which yields us \inlinecode{[(d, fut a, fut b)]}.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \inlinecode{arr length >>> arr unshuffle} with the use of \inlinecode{app :: arr (arr a b, a) c} from the \inlinecode{ArrowApply} typeclass.
+Finally, we unpack this matrix \inlinecode{[[[(d, fut a, fut b)]]} with \inlinecode{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \inlinecode{([[d]], ([[fut a]], [[fut b]]))}.
 \\\\

### 2017-05-17 paper/content/utilityFunctions.tex
renamed \code to \inlinecode

@@
 \section{Utility Functions}\label{utilfns}
-To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \code{map}, \code{foldl} and \code{zipWith} on arrows.
+To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \inlinecode{map}, \inlinecode{foldl} and \inlinecode{zipWith} on arrows.
 
-The \code{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows}:
+The \inlinecode{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \inlinecode{arr a b} to an arrow \inlinecode{arr [a] [b]} \cite{programming_with_arrows}:
 \begin{figure}[h]
@@
 \end{lstlisting}
-\caption{\code{map} over arrows}
+\caption{\inlinecode{map} over arrows}
 \label{fig:mapArr}
 \end{figure}
-Similarly, we can also define \code{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \code{arr (b, a) b} with a neutral element \code{b} to \code{arr [a] b}:
+Similarly, we can also define \inlinecode{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \inlinecode{arr (b, a) b} with a neutral element \inlinecode{b} to \inlinecode{arr [a] b}:
 \begin{figure}[h]
@@
 \end{lstlisting}
-\caption{\code{foldl} over arrows}
+\caption{\inlinecode{foldl} over arrows}
 \label{fig:foldlArr}
 \end{figure}
-Finally, with the help of \code{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+Finally, with the help of \inlinecode{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \inlinecode{arr (a, b) c} to an arrow \inlinecode{arr ([a], [b]) [c]}.
 \begin{figure}[h]
@@
 \end{lstlisting}
-\caption{\code{zipWith} over arrows}
+\caption{\inlinecode{zipWith} over arrows}
 \label{fig:zipWithArr}
@@
  %$ %% formatting
-These combinators make use of the \code{ArrowChoice} type class which provides the \code{(|||)} combinator.
-It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
+These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \inlinecode{(|||)} combinator.
+It takes two arrows \inlinecode{arr a c} and \inlinecode{arr b c} and combines them into a new arrow \inlinecode{arr (Either a b) c} which pipes all \inlinecode{Left a}'s to the first arrow and all \inlinecode{Right b}'s to the second arrow.
 \begin{figure}[h]
@@
 \end{lstlisting}
-\caption{Type signature of \code{(|||)}}
+\caption{Type signature of \inlinecode{(|||)}}
 \label{fig:codeSig|||}
@@
 
-With the zipWithArr combinator we can also write a combinator \code{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]}.
+With the zipWithArr combinator we can also write a combinator \inlinecode{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \inlinecode{[arr a b]} to an arrow \inlinecode{arr [a] [b]}.
 \begin{figure}[h]
@@
 \end{lstlisting}
-\caption{Definition of \code{listApp}}
+\caption{Definition of \inlinecode{listApp}}
 \label{fig:listApp}
@@
 % $ %% formatting
-. Note that  this additionally makes use of the \code{ArrowApply} typeclass that allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
+. Note that  this additionally makes use of the \inlinecode{ArrowApply} typeclass that allows us to evaluate arrows with \inlinecode{app :: arr (arr a b, a) c}.
 

### 2017-05-17 paper/content/main.tex
fixed things to that lhs2TeX can run (without the code markers)

% This is our submission, modified from:
% the file JFP2egui.tex
% release v1.02, 27th September 2001
%   (based on JFPguide.tex v1.11 for LaTeX 2.09)
% Copyright (C) 2001 Cambridge University Press

\NeedsTeXFormat{LaTeX2e}

\documentclass{jfp1}

%include polycode.fmt

%%% Macros for the guide only %%%
%\providecommand\AMSLaTeX{AMS\,\LaTeX}
%\newcommand\eg{\emph{e.g.}\ }
%\newcommand\etc{\emph{etc.}}
%\newcommand\bcmdtab{\noindent\bgroup\tabcolsep=0pt%
%  \begin{tabular}{@{}p{10pc}@{}p{20pc}@{}}}
%\newcommand\ecmdtab{\end{tabular}\egroup}
%\newcommand\rch[1]{$\longrightarrow\rlap{$#1$}$\hspace{1em}}
%\newcommand\lra{\ensuremath{\quad\longrightarrow\quad}}

\jdate{April 2017}
\pubyear{2017}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\doi{...}

%\newtheorem{lemma}{Lemma}[section]

%include prelude.tex

\title{Arrows for Parallel Computations}
\ifthenelse{\boolean{anonymous}}{%
\author{Submission ID xxxxxx}
}{%
%\author{Martin Braun, Phil Trinder, and Oleg Lobachev}
%\affiliation{University Bayreuth, Germany and Glasgow University, UK}
 \author[M. Braun, P. Trinder and O. Lobachev]%
        {MARTIN BRAUN\\
         University Bayreuth, 95440 Bayreuth, Germany\\
         PHIL TRINDER\\
		 Glasgow University, Glasgow, G12 8QQ, Scotland\\
		 \and\ OLEG LOBACHEV\\
		 University Bayreuth, 95440 Bayreuth, Germany}
}% end ifthenelse

\begin{document}

\label{firstpage}

\maketitle

%% environment inside
%include abstract.tex

\tableofcontents

	%
	%%include abstract.tex
	%
	%\newpage
	\tableofcontents
	%\pagebreak
	%include motivation.tex
	%include parallelHaskells.tex
	%include arrows.tex
	%\pagebreak
	%include relwork.tex
	%\pagebreak
	%\pagebreak
	%include parrows.tex
	%\pagebreak
	%include basicSkeletons.tex
	%\pagebreak
	%include syntacticSugar.tex
	%\pagebreak
	%include futures.tex
	%\pagebreak
	%include mapSkeletons.tex
	%\pagebreak
	%include topologySkeletons.tex
	%\pagebreak
	%include benchmarks.tex
	%%\pagebreak
	%include conclusion.tex
	%\pagebreak
        \bibliographystyle{jfp}
	\bibliography{references,main}
        \appendix
	%include utilityFunctions.tex
\end{document}

### 2017-05-17 paper/content/topologySkeletons.tex
renamed fig references for |***| |&&&| and |>>>| so that they work with lhs2TeX

@@
 Note that extensive use of this combinator over \inlinecode{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \inlinecode{(|>>>|)} (Fig.~\ref{fig:|>>>|}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+Nonetheless, we can define a parallel piping operator \inlinecode{(|>>>|)} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
 \begin{figure}[h]
@@
 \caption{Definition of \texttt{(|>>>|)}}
-\label{fig:|>>>|}
+\label{fig:par>>>}
 \end{figure}

### 2017-05-17 paper/content/utilityFunctions.tex
intermediary

@@
 	arr listcase >>>
-	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
+	arr (const []) pipepipepipe (f *** mapArr f >>> arr (uncurry (:)))
 	where listcase [] = Left ()
@@
 	arr listcase >>>
-	arr (const b) |||
+	arr (const b) pipepipepipe
 		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
@@
 \end{figure}
+\fixme{pipepipepipe does not work with lhs2TeX}
 Finally, with the help of \inlinecode{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \inlinecode{arr (a, b) c} to an arrow \inlinecode{arr ([a], [b]) [c]}.
@@
  %$ %% formatting
-These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \inlinecode{(|||)} combinator.
+These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \inlinecode{(\textpipe\textpipe\textpipe)} combinator.
 It takes two arrows \inlinecode{arr a c} and \inlinecode{arr b c} and combines them into a new arrow \inlinecode{arr (Either a b) c} which pipes all \inlinecode{Left a}'s to the first arrow and all \inlinecode{Right b}'s to the second arrow.
@@
 \begin{lstlisting}[frame=htrbl]
-(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
+(@pipepipepipe@) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
 \end{lstlisting}
-\caption{Type signature of \inlinecode{(|||)}}
-\label{fig:codeSig|||}
+\caption{Type signature of \inlinecode{(\textpipe\textpipe\textpipe)}}
+\label{fig:codeSigPipePipePipe}
 \end{figure}
+\fixme{pipepipepipe does not work with lhs2TeX}
 

### 2017-05-18 paper/content/main.tex
it works with lhs2tex

@@
 % This is our submission, modified from:
-% the file JFP2egui.tex
+% the file JFP2egui.lhs
 % release v1.02, 27th September 2001
-%   (based on JFPguide.tex v1.11 for LaTeX 2.09)
+%   (based on JFPguide.lhs v1.11 for LaLhs 2.09)
 % Copyright (C) 2001 Cambridge University Press
@@
 
-%include polycode.fmt
+%% ODER: format ==         = "\mathrel{==}"
+%% ODER: format /=         = "\neq "
+%
+%
+\makeatletter
+\@ifundefined{lhs2tex.lhs2tex.sty.read}%
+  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
+   \newcommand\SkipToFmtEnd{}%
+   \newcommand\EndFmtInput{}%
+   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
+  }\SkipToFmtEnd
+
+\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
+\usepackage{amstext}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\DeclareFontFamily{OT1}{cmtex}{}
+\DeclareFontShape{OT1}{cmtex}{m}{n}
+  {<5><6><7><8>cmtex8
+   <9>cmtex9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
+\DeclareFontShape{OT1}{cmtex}{m}{it}
+  {<-> ssub * cmtt/m/it}{}
+\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
+\DeclareFontShape{OT1}{cmtt}{bx}{n}
+  {<5><6><7><8>cmtt8
+   <9>cmbtt9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
+\DeclareFontShape{OT1}{cmtex}{bx}{n}
+  {<-> ssub * cmtt/bx/n}{}
+\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU
+
+\newcommand{\Sp}{\hskip.33334em\relax}
+
+\newcommand{\Conid}[1]{\mathit{#1}}
+\newcommand{\Varid}[1]{\mathit{#1}}
+\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
+\newcommand{\plus}{\mathbin{+\!\!\!+}}
+\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
+\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
+\newcommand{\sequ}{\mathbin{>\!\!\!>}}
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\usepackage{polytable}
+
+%mathindent has to be defined
+\@ifundefined{mathindent}%
+  {\newdimen\mathindent\mathindent\leftmargini}%
+  {}%
+
+\def\resethooks{%
+  \global\let\SaveRestoreHook\empty
+  \global\let\ColumnHook\empty}
+\newcommand*{\savecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
+\newcommand*{\restorecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
+\newcommand*{\aligncolumn}[2]%
+  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}
+
+\resethooks
+
+\newcommand{\onelinecommentchars}{\quad-{}- }
+\newcommand{\commentbeginchars}{\enskip\{-}
+\newcommand{\commentendchars}{-\}\enskip}
+
+\newcommand{\visiblecomments}{%
+  \let\onelinecomment=\onelinecommentchars
+  \let\commentbegin=\commentbeginchars
+  \let\commentend=\commentendchars}
+
+\newcommand{\invisiblecomments}{%
+  \let\onelinecomment=\empty
+  \let\commentbegin=\empty
+  \let\commentend=\empty}
+
+\visiblecomments
+
+\newlength{\blanklineskip}
+\setlength{\blanklineskip}{0.66084ex}
+
+\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
+\let\hspre\empty
+\let\hspost\empty
+\newcommand{\NB}{\textbf{NB}}
+\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}
+
+\EndFmtInput
+\makeatother
+%
+%
+%
+%
+%
+%
+% This package provides two environments suitable to take the place
+% of hscode, called "plainhscode" and "arrayhscode". 
+%
+% The plain environment surrounds each code block by vertical space,
+% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
+% similar to formulas. Note that if these dimensions are changed,
+% the spacing around displayed math formulas changes as well.
+% All code is indented using \leftskip.
+%
+% Changed 19.08.2004 to reflect changes in colorcode. Should work with
+% CodeGroup.sty.
+%
+\ReadOnlyOnce{polycode.fmt}%
+\makeatletter
+
+\newcommand{\hsnewpar}[1]%
+  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}
+
+% can be used, for instance, to redefine the code size, by setting the
+% command to \small or something alike
+\newcommand{\hscodestyle}{}
+
+% The command \sethscode can be used to switch the code formatting
+% behaviour by mapping the hscode environment in the subst directive
+% to a new LaTeX environment.
+
+\newcommand{\sethscode}[1]%
+  {\expandafter\let\expandafter\hscode\csname #1\endcsname
+   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}
+
+% "compatibility" mode restores the non-polycode.fmt layout.
+
+\newenvironment{compathscode}%
+  {\par\noindent
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed\)%
+   \par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\compaths}{\sethscode{compathscode}}
+
+% "plain" mode is the proposed default.
+% It should now work with \centering.
+% This required some changes. The old version
+% is still available for reference as oldplainhscode.
+
+\newenvironment{plainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newenvironment{oldplainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+% Here, we make plainhscode the default environment.
+
+\newcommand{\plainhs}{\sethscode{plainhscode}}
+\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
+\plainhs
+
+% The arrayhscode is like plain, but makes use of polytable's
+% parray environment which disallows page breaks in code blocks.
+
+\newenvironment{arrayhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\parray}%
+  {\endparray\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newcommand{\arrayhs}{\sethscode{arrayhscode}}
+
+% The mathhscode environment also makes use of polytable's parray 
+% environment. It is supposed to be used only inside math mode 
+% (I used it to typeset the type rules in my thesis).
+
+\newenvironment{mathhscode}%
+  {\parray}{\endparray}
+
+\newcommand{\mathhs}{\sethscode{mathhscode}}
+
+% texths is similar to mathhs, but works in text mode.
+
+\newenvironment{texthscode}%
+  {\(\parray}{\endparray\)}
+
+\newcommand{\texths}{\sethscode{texthscode}}
+
+% The framed environment places code in a framed box.
+
+\def\codeframewidth{\arrayrulewidth}
+\RequirePackage{calc}
+
+\newenvironment{framedhscode}%
+  {\parskip=\abovedisplayskip\par\noindent
+   \hscodestyle
+   \arrayrulewidth=\codeframewidth
+   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
+   \hline\framedhslinecorrect\\{-1.5ex}%
+   \let\endoflinesave=\\
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \framedhslinecorrect\endoflinesave{.5ex}\hline
+   \endtabular
+   \parskip=\belowdisplayskip\par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\framedhslinecorrect}[2]%
+  {#1[#2]}
+
+\newcommand{\framedhs}{\sethscode{framedhscode}}
+
+% The inlinehscode environment is an experimental environment
+% that can be used to typeset displayed code inline.
+
+\newenvironment{inlinehscode}%
+  {\(\def\column##1##2{}%
+   \let\>\undefined\let\<\undefined\let\\\undefined
+   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
+   \def\fromto##1##2##3{##3}%
+   \def\nextline{}}{\) }%
+
+\newcommand{\inlinehs}{\sethscode{inlinehscode}}
+
+% The joincode environment is a separate environment that
+% can be used to surround and thereby connect multiple code
+% blocks.
+
+\newenvironment{joincode}%
+  {\let\orighscode=\hscode
+   \let\origendhscode=\endhscode
+   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
+   %\let\SaveRestoreHook=\empty
+   %\let\ColumnHook=\empty
+   %\let\resethooks=\empty
+   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
+  {\origendhscode
+   \global\let\hscode=\orighscode
+   \global\let\endhscode=\origendhscode}%
+
+\makeatother
+\EndFmtInput
+%
 
@@
 
-%include prelude.tex
+\usepackage{tikz}
+
+\usepackage{filecontents}
+\usepackage{pgfplots, pgfplotstable}
+\usepgfplotslibrary{statistics}
+
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+
+%\usepackage{color}
+\usepackage{hyperref}
+
+\usepackage{subcaption}
+
+\usepackage[round]{natbib}
+
+\usepackage{csquotes}
+
+\usepackage{placeins}
+
+\usepackage{microtype}
+\DisableLigatures[>,<]{encoding = T1,family=tt*}
+
+% für Listings
+\usepackage{listings}
+
+\renewcommand{\cite}[1]{\citep{#1}}
+
+\newcommand{\citHughes}{\citep{HughesArrows}}
+
+\newcommand{\inlinecode}[1]{\texttt{#1}}
+
+%\newcommand{\fixme}[1]{\colorbox{red}{#1}}
+
+%%%%%% more fun and typography
+
+\usepackage{xifthen}
+
+\newboolean{anonymous}
+\setboolean{anonymous}{False}
+
+\usepackage[british]{babel}
+
+\usepackage{siunitx}
+
+\usepackage{tipa}
+
+\usepackage{xspace}
+\usepackage{xcolor}
+\newcommand{\hl}[1]{\textcolor{red}{#1}}
+\newcommand{\comm}[2]{\textcolor{red}{\bfseries #1: #2}}
+%\newcommand{\comm}[2]{}
+\newcommand{\olcomment}[1]{\comm{OL}{#1}}
+\newcommand{\mbcomment}[1]{\comm{MB}{#1}}
+\newcommand{\ptcomment}[1]{\comm{Phil}{#1}}
+%\newcommand{\done}{\xspace\hl{done!}\xspace}
+\newcommand{\fixme}{\mbcomment}
+
+\DeclareRobustCommand{\xth}{\textsuperscript{th}\xspace}
+\DeclareRobustCommand{\st}{\textsuperscript{st}\xspace}
+\DeclareRobustCommand{\nd}{\textsuperscript{nd}\xspace}
+\DeclareRobustCommand{\xrd}{\textsuperscript{rd}\xspace}
+
+%% kuerzel
+%\newcommand{\todo}{\textcolor{red}{\bfseries TODO!}\xspace}
+\newcommand{\done}{\textcolor{green}{\bfseries done!}\xspace}
+\DeclareRobustCommand{\hairspn}{\hspace{1pt}\nolinebreak}% hair space with no break
+
+% \DeclareRobustCommand{\ie}{{i.\hairspn{}e.\nopagebreak[4] }}
+% \DeclareRobustCommand{\eg}{{e.\hairspn{}g.\nopagebreak[4] }}
+% \DeclareRobustCommand{\fe}{{f.\hairspn{}e.,}\nopagebreak[4] }
+\DeclareRobustCommand{\ie}{{i.\hairspn{}e.~}}
+\DeclareRobustCommand{\eg}{{e.\hairspn{}g.~}}
+\DeclareRobustCommand{\fe}{{f.\hairspn{}e.,~}}
+\DeclareRobustCommand{\ad}{{A.\hairspn{}D.\xspace}}
+\DeclareRobustCommand{\bcc}{{A.\hairspn{}C.\xspace}}
+\DeclareRobustCommand{\wrt}{w.\hairspn{}r.\hairspn{}t.~}
+\DeclareRobustCommand{\etc}{etc.\ }
+\DeclareRobustCommand{\cf}{\textit{cf.~}}
+\DeclareRobustCommand{\viz}{\textit{viz.~}}
+%\DeclareRobustCommand{\hof}{higher-or\-der function\xspace}
+
+% %% the numberings
+\DeclareRobustCommand{\xth}{\textsuperscript{th}\xspace}
+\DeclareRobustCommand{\st}{\textsuperscript{st}\xspace}
+\DeclareRobustCommand{\nd}{\textsuperscript{nd}\xspace}
+\DeclareRobustCommand{\xrd}{\textsuperscript{rd}\xspace}
+
+\newcommand{\tabsepbakup}{\tabcolsep}
+%% for narrower tables
+
+%% nice tables
+\usepackage{booktabs}
+%\usepackage{multirow}
+
+%\usepackage{ifpdf} %% broken?!!
+%\ifpdf
+%\usepackage[activate={true,nocompatibility},kerning=true,spacing=true,tracking=true]{microtype}
+%\else
+%% non-pdf latex can less, let microtype itself decide
+\usepackage{microtype}
+%\fi
+\microtypecontext{spacing=nonfrench} %% log said so
+
+%%% JFP requirements:
+%%% Harvard citing style, "(Curry 1933)".
+%%% code: identifies italic, keywords bold
+%%% figures: eps(!) or ps (can convert pdf), eventually provide a greyscale version, color NOT in cmyk
 
@@
 %% environment inside
-%include abstract.tex
+\begin{abstract}
+Arrows form\olcomment{are?} a general interface for computation and pose therefore as an alternative to monads for API design.
+We express parallelism using this concept.
+This is a new way to represent parallel computation.
+We define an Arrows-based interface for parallelism and implement it using multiple parallel Haskells.
+\olcomment{Benefits:} In this manner we are able to bridge across various parallel Haskells with a common interface.
+
+This new way of writing parallel programs has a benefit of being portable across flavours of parallel Haskells.\olcomment{Wdh?}
+Each parallel computation is an arrow, they can be composed and transformed as such.
+We thus introduce some syntactic sugar to provide parallelism-aware arrow combinators.
+
+We also define  several parallel skeletons with our framework. 
+Benchmarks shows that our framework does not induce too much overhead performance-wise.
+\end{abstract}
 
@@
 	%
-	%%include abstract.tex
+	%%include abstract.lhs
 	%
@@
 	%\pagebreak
-	%include motivation.tex
-	%include parallelHaskells.tex
-	%include arrows.tex
+	%\section{Motivation}
+%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
+In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
+In this paper we will use this concept to express parallelism.
+
+\section{Introduction}
+\label{sec:introduction}
+\olcomment{todo, reuse 5.5, "Impact" at the end and more}
+
+blablabla arrows, parallel, haskell.
+
+\paragraph{Contribution}
+
+HIT HERE REALLY STRONG
+
+\paragraph{Structure}
+The remaining text is structures as follows.
+Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
+Section~\ref{sec:related-work} discusses related work.
+Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface.
+Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication.
+Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \inlinecode{map} with and without load balancing, \inlinecode{map-reduce}) in our newly defined dialect.
+More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\inlinecode{pipe}, \inlinecode{ring}, \inlinecode{torus}).
+Section~\ref{sec:benchmarks} shows the benchmark results.
+Section~\ref{sec:conclusion} discusses future work and concludes.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
+	\section{Background}
+\label{sec:background}
+\subsection{Short introduction to parallel Haskells}
+There are already several ways to write parallel programs in Haskell.
+As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
+
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \inlinecode{a -> b} in parallel:
+
+\begin{figure}[h]
+%\begin{lstlisting}[frame=htrbl]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%\end{lstlisting}
+\caption{parEvalN's type signature}
+\label{fig:parEvalNTypeSig}
+\end{figure}
+\begin{figure}[h]
+	\includegraphics[scale=0.7]{images/parEvalN}
+	\caption{Schematic illustration of parEvalN}
+	\label{fig:parEvalN}
+\end{figure}
+\olcomment{make them to real figures? with environment, reference and such?}
+Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \inlinecode{parEvalN} with several different parallel Haskells.
+
+\subsubsection{Multicore Haskell}
+Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \inlinecode{using :: a -> Strategy a -> a}.
+For \inlinecode{parEvalN} this means that we can just apply the list of functions \inlinecode{[a -> b]} to the list of inputs \inlinecode{[a]} by zipping them with the application operator \inlinecode{\$}.
+We then evaluate this lazy list \inlinecode{[b]} according to a \inlinecode{Strategy [b]} with the \inlinecode{using :: a -> Strategy a -> a} operator.
+We construct this strategy with \inlinecode{parList :: Strategy a -> Strategy [a]} and \inlinecode{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
+To ensure that programs that use \inlinecode{parEvalN} have the correct evaluation order, we annotate the computation with \inlinecode{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \inlinecode{parEvalN} computations.
+This is particularly necessary in circular communication topologies like in the \inlinecode{torus} or \inlinecode{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \inlinecode{pseq} during testing for this paper.
+
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = let bs = zipWith ($) fs as in
+	(bs `using` parList rdeepseq) `pseq` bs
+\end{lstlisting}
+\begin{figure}[h]
+	\includegraphics[scale=0.5]{images/parEvalNMulticore}
+	\caption{Dataflow of the Multicore Haskell parEvalN version}
+\end{figure} %$ %% formatting
+
+\subsubsection{ParMonad}
+The \inlinecode{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
+
+Our parallel evaluation function \inlinecode{parEvalN} can be defined by zipping the list of \inlinecode{[a -> b]} with the list of inputs \inlinecode{[a]} with the application operator \inlinecode{\$} just like with Multicore Haskell.
+Then, we map over this not yet evaluated lazy list of results \inlinecode{[b]} with \inlinecode{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \inlinecode{[Par (IVar b)]}, which we convert to \inlinecode{Par [IVar b]} with \inlinecode{sequenceA}.
+We wait for the computations to finish by mapping over the \inlinecode{IVar b}'s inside the \inlinecode{Par} monad with \inlinecode{get}.
+This results in \inlinecode{Par [b]}.
+We finally execute this process with \inlinecode{runPar} to finally get \inlinecode{[b]} again.
+
+\textbf{\textcolor{red}{explain problems with laziness here. Problems with torus}}
+
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = runPar $ 
+	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+\end{lstlisting}
+\begin{figure}[h]
+	\includegraphics[scale=0.5]{images/parEvalNParMonad}
+	\caption{Dataflow of the Par Monad parEvalN version}
+\end{figure}
+
+\subsubsection{Eden}
+Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{See also \url{http://www.mathematik.uni-marburg.de/~eden/} and \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell flavour.
+
+Eden was designed to work on clusters, but with a further simple backend it operates on multicores.
+However, in contrast to many other parallel Haskells, in Eden each process has its own heap.
+This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results also on multicores \cite{arcs-dc,aswad2009low}.
+
+While Eden also comes with a monad \inlinecode{PA} for parallel evaluation, it also ships with a completely functional interface that includes
+%\\
+%\inlinecode{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
+%\\
+a \inlinecode{spawnF} function that
+%This 
+allows us to define \inlinecode{parEvalN} directly:
+
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
+parEvalN = spawnF 
+\end{lstlisting}
+\begin{figure}[h]
+	\includegraphics[scale=0.5]{images/parEvalNEden}
+	\caption{Dataflow of the Eden parEvalN version}
+\end{figure}
+
+\paragraph{Eden TraceViewer.}
+To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
+While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See .....
+on hackage for the last available version of Eden TraceViewer.
+There was an effort to implement the TraceViewer using modern web technologies \cite{traceviewer-web}.} \cite{Berthold2007a}.
+In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.
+
+In a trace, the $x$ axis shows the time, the $y$ axis enumerates the machines and processes.
+A~trace shows a running process in green, a blocked process is red.
+If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow.
+The typical reason for then is GC.
+An inactive machine where no processes are started yet, or all are already terminated, is shows as a blue bar.
+A~comminication from one process to another is represented with a black arrow.
+A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.
+
+\olcomment{show example trace or refer to a trace in later figures}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
+	\subsection{Arrows}
+\label{sec:arrows}
+\begin{figure}[h]
+	\includegraphics{images/arrow}
+	\caption{Schematic depiction of an arrow}
+\end{figure}
+Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
+An arrow \inlinecode{arr a b} represents  a computation that converts an input \inlinecode{a} to an output \inlinecode{b}.
+This is defined in the arrow typeclass:
+
+\begin{figure}[h]
+\begin{subfigure}{0.5\textwidth}
+%\begin{minipage}{0.5\textwidth}
+\begin{lstlisting}[frame=htrbl]
+class Arrow arr where
+arr :: (a -> b) -> arr a b
+
+(>>>) :: arr a b -> arr b c -> arr a c
+
+first :: arr a b -> arr (a,c) (b,c)
+\end{lstlisting}
+\caption{Arrow class definition}
+%\end{minipage}
+\end{subfigure}
+~~~~
+\begin{subfigure}{0.45\textwidth}
+	\begin{center}
+	\subcaptionbox{arr}{\includegraphics[scale=0.6]{images/arr}}
+	\subcaptionbox{composition \inlinecode{(>>>)}}{\includegraphics[scale=0.6]{images/compose}}
+	\subcaptionbox{first}{\includegraphics[scale=0.6]{images/first}}
+	\end{center}
+\end{subfigure}
+\caption{The arrow type class definition on the left with schematic depiction of its combinators on the right}
+\end{figure}
+\lstinline{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \lstinline{return}.
+The \lstinline{>>>} operator is analogous to the monadic composition  \lstinline{>>=} and combines two arrows \inlinecode{arr a b} and \inlinecode{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \inlinecode{arr a c}.
+Lastly, the \lstinline{first} operator  takes the input arrow from \lstinline{b} to \lstinline{c} and converts it into an arrow on pairs with the second argument untouched.
+It allows us to to save input across arrows.
+\\\\
+The most prominent instances of this interface are regular functions \lstinline{(->)} (Fig.~\ref{fig:arrowfn}),
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+instance Arrow (->) where
+	arr f = f
+	f >>> g = g . f
+	first f = \(a, c) -> (f a, c) 
+\end{lstlisting}
+\caption{Arrow instance for regular functions}
+\label{fig:arrowfn}
+\end{figure}
+and the Kleisli type (Fig.~\ref{fig:arrowkleisli}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+data Kleisli m a b = Kleisli { run :: a -> m b }
+
+instance Monad m => Arrow (Kleisli m) where
+	arr f = Kleisli (return . f)
+	f >>> g = Kleisli (\a -> f a >>= g)
+	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
+\end{lstlisting}%$
+\caption{Definition of the Kleisli type and the corresponding arrow instance}
+\label{fig:arrowkleisli}
+\end{figure}
+\begin{figure}[h]
+	\centering
+	\begin{tabular}{cc}
+		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
+		\subcaptionbox{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
+		\subcaptionbox{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+		\subcaptionbox{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
+	\end{tabular}
+	\caption{Syntactic sugar for arrows}
+	\label{3}
+\end{figure}
+With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \inlinecode{first}, called \inlinecode{second} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+second :: Arrow arr => arr a b -> arr (c, a) (c, b)
+second f = arr swap >>> first f >>> arr swap
+	where swap (x, y) = (y, x)
+\end{lstlisting}
+\caption{The second combinator}
+\label{fig:second}
+\end{figure}
+the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
+f *** g = first f >>> second g
+\end{lstlisting}
+\caption{The (***) combinator}
+\label{fig:***}
+\end{figure}
+and the \inlinecode{\&\&\&} combinator that constructs an arrow which outputs two different values like \inlinecode{***}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
+f &&& g = arr (\a -> (a, a)) >>> (f *** g)
+\end{lstlisting}
+\caption{The (\&\&\&) combinator}
+\label{fig:&&&}
+\end{figure}
+A short example given by Hughes on how to use this is \lstinline{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
+add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
+\end{lstlisting}
+\caption{Add over arrows}
+\label{fig:addArrows}
+\end{figure}
+% The benefit of using the \inlinecode{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
+Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
+Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
+% \\\\
+% \textit{Note: In the definitions Hughes gave in his paper, the notation \inlinecode{a b c} for an arrow from \inlinecode{b} to \inlinecode{c} is used.
+We use the equivalent definition \inlinecode{arr a b} for an arrow from \inlinecode{a} to \inlinecode{b} instead, to make it easier to find the arrow type in type signatures.}
+%
+
+The more restrictive interface of arrows (a monad can be \emph{anything}, an arrow is a process of doing something, a \emph{computation}) allows for more elaborate composition and transformation combinators.
+One of the major problems in parallel computing is composition of parallel processes.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 	%\pagebreak
-	%include relwork.tex
+	\section{Related Work}
+\label{sec:related-work}
+
+\olcomment{arrows or Arrows?}
+
+\subsection{Parallel Haskells}
+Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \inlinecode{Par} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
+We use these languages as backends: our library can switch from one to other at user's command.
+
+HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \inlinecode{Par} monad to heterogeneous clusters.
+LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \inlinecode{Par} monad.
+%
+Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015} and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}.
+Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden.
+This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,brown2010ever,dieterle2010skeleton,delaEncina2011,Dieterle2013}.
+
+More different approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582,}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
+%
+The Haskell--GPU bridge Accelerate \cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313} deserves a special mention.
+Accelerate is completely orthogonal to our approach.
+\citeauthor{marlow2013parallel} authored a recent book \citeyear{marlow2013parallel} on parallel Haskells.
+
+\subsection{Algorithmic skeletons}
+
+Algorithmic skeletons were introduced by \citet{Cole1989}.
+Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
+The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,brown2010ever,lobachev-phd,Lobachev2012}, iteration skeletons \cite{Dieterle2013}.
+The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel.
+\citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeleton to stable process networks.
+
+\subsection{Arrows}
+
+Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\inlinecode{->}.
+\citet{Hughes2005} is a tutorial on arrows.
+Some theoretical details on arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable.
+\citet{Paterson:2001:NNA:507669.507664} introduced a new notation for arrows.
+Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}.
+But perhaps most prominent application of arrows is functional reactive programming \cite{Hudak2003}.\olcomment{cite more!}
+
+\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of arrows that capsule the computation more than regular arrows do and thus enable optimizations.
+Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution.
+In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \inlinecode{parEvalN} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
+\citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach.
+They basically use arrows to orchestrate several tasks in robotics.
+We propose a general interface for parallel programming, remaining completely in Haskell.
+
+\subsection{Other languages}
+
+Although this work is centred on Haskell implementation of arrows, it is applicable to any functional programming language where parallel evaluation and arrows can be defined.
+Our experiments with our approach in Frege language (which is basically Haskell on the JVM) were quite successful, we were able to use typical Java libraries for parallelism.
+However, it is beyond the scope of this work.
+
+\citet{achten2004arrows,achten2007arrow} use an arrow implementation in Clean for better handling of typical GUI tasks.
+\citet{Dagand:2009:ORD:1481861.1481870} used arrows in OCaml in the implementation of a distributed system.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 	%\pagebreak
 	%\pagebreak
-	%include parrows.tex
+	\section{Parallel Arrows}
+\label{sec:parallel-arrows}
+We have seen what Arrows are and how they can be used as a general interface to computation.
+In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well.
+We start by introducing the interface and explaining the reasonings behind it.
+Then, we discuss some implementations using exisiting parallel Haskells.
+Finally, we explain why using Arrows for expressing parallelism is beneficial.
+\subsection{The ArrowParallel typeclass}
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \inlinecode{a -> b} in parallel, \inlinecode{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
+Translating this into arrow terms gives us a new operator \inlinecode{parEvalN} that lifts a list of arrows \inlinecode{[arr a b]} to a parallel arrow \inlinecode{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \lstinline{listApp} from Appendix \ref{utilfns}, but does parallel instead of serial evaluation).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
+\end{lstlisting}
+\caption{parEvalN Arrow combinator as a function}
+\label{fig:parEvalNArrowFn}
+\end{figure}
+With this definition of \inlinecode{parEvalN}, parallel execution is yet another arrow combinator.
+But as the implementation may differ depending on the actual type of the arrow \inlinecode{arr} and we want this to be an interface for different backends, we introduce a new typeclass \inlinecode{ArrowParallel arr a b} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+class Arrow arr => ArrowParallel arr a b where
+	parEvalN :: [arr a b] -> arr [a] [b]
+\end{lstlisting}
+\caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
+\label{fig:parEvalNArrowTypeClass1}
+\end{figure}
+Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs.
+normalform).
+For this reason we also introduce an additional \inlinecode{conf} parameter to the function.
+We also do not want \inlinecode{conf} to be a fixed type, as the configuration parameters can differ for different instances of \inlinecode{ArrowParallel}.
+So we add it to the type signature of the typeclass as well and get \inlinecode{ArrowParallel arr a b conf} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+class Arrow arr => ArrowParallel arr a b conf where
+	parEvalN :: conf -> [arr a b] -> arr [a] [b]
+\end{lstlisting}
+\caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
+\label{fig:parEvalNArrowTypeClassFinal}
+\end{figure}
+Note that we don't require the \inlinecode{conf} parameter in every implementation.
+If it is not needed, we usually just default the \inlinecode{conf} type parameter to () and even blank it out in the parameter list of the implemented \inlinecode{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
+
+\subsection{ArrowParallel instances}
+
+\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
+The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \inlinecode{withStrategy :: Strategy a -> a -> a} and \inlinecode{pseq :: a -> b -> b} combinators from Multicore Haskell, where \inlinecode{withStrategy} is the same as \inlinecode{using :: a -> Strategy a -> a} but with flipped parameters.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
+	ArrowParallel arr a b () where
+    	parEvalN _ fs =
+       		listApp fs >>>
+        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
+        	arr (uncurry pseq)
+\end{lstlisting}% $ %% formatting
+\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
+\label{fig:ArrowParallelMulticoreRdeepseq}
+\end{figure}
+For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \inlinecode{Strategy} interface, we want to the user not to lose this ability because of using our API.
+We therefore introduce the \inlinecode{Conf a} data-type that simply wraps a \inlinecode{Strategy a} (Fig.~\ref{fig:confa}).
+We can't directly use the \inlinecode{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+data Conf a = Conf (Strategy a)
+\end{lstlisting}
+\caption{Definition of Conf a}
+\label{fig:confa}
+\end{figure}
+To get our configurable \inlinecode{ArrowParallel} instance, we simply unwrap the strategy and pass it to \inlinecode{parList} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
+	ArrowParallel arr a b (Conf b) where
+    	parEvalN (Conf strat) fs =
+        	listApp fs >>>
+        	arr (withStrategy (parList strat)) &&& arr id >>>
+        	arr (uncurry pseq)
+\end{lstlisting}
+\caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
+\label{fig:ArrowParallelMulticoreConfigurable}
+\end{figure}
+\subsubsection{Par Monad}
+The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \inlinecode{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function.
+The latter forks away the computation of a value and returns an \inlinecode{IVar} containing the result in the \inlinecode{Par} monad.
+
+We therefore apply each function to its corresponding input value with \inlinecode{listApp} (Fig.~\ref{fig:listApp}) and then fork the computation away with \inlinecode{arr spawnP} inside a \inlinecode{zipWithArr} call.
+This yields a list \inlinecode{[Par (IVar b)]}, which we then convert into \inlinecode{Par [IVar b]} with \inlinecode{arr sequenceA}.
+In order to wait for the computation to finish, we map over the \inlinecode{IVar}s inside the ParMonad with \inlinecode{arr (>>= mapM get)}.
+The result of this operation is a \inlinecode{Par [b]} from which we can finally remove the monad again by running \inlinecode{arr runPar} to get our output of \inlinecode{[b]}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
+	ArrowParallel arr a b conf where
+		parEvalN _ fs = 
+			(arr $ \as -> (fs, as)) >>>
+			zipWithArr (app >>> arr spawnP) >>>
+			arr sequenceA >>>
+			arr (>>= mapM get) >>>
+			arr runPar
+\end{lstlisting} %$ %% formatting
+\caption{ArrowParallel instance for the Par Monad backend}
+\label{fig:ArrowParallelParMonad}
+\end{figure}
+
+\subsubsection{Eden}
+For the Multicore and ParMonad implementation we could use general instances of \inlinecode{ArrowParallel} that just require the \inlinecode{ArrowApply} and \inlinecode{ArrowChoice} typeclasses.
+With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
+While we could still manage to have only one class in the module by introducing a typeclass \inlinecode{ArrowUnwrap} (Fig.~\ref{fig:ArrowUnwrap}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+class (Arrow arr) => ArrowUnwrap arr where
+	arr a b -> (a -> b)
+\end{lstlisting}
+\caption{possible ArrowUnwrap typeclass}
+\label{fig:ArrowUnwrap}
+\end{figure}
+We don't do it here for aesthetic resons, though. For now, we just implement \inlinecode{ArrowParallel} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
+parEvalN _ fs as = spawnF fs as
+\end{lstlisting}
+\caption{ArrowParallel instance for functions in the Eden backend}
+\label{fig:ArrowParallelEdenFns}
+\end{figure}
+and the Kleisli type (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+instance (Monad m, Trans a, Trans b, Trans (m b)) =>
+	ArrowParallel (Kleisli m) a b conf where
+parEvalN conf fs =
+	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
+	(Kleisli $ sequence)
+\end{lstlisting}
+\caption{ArrowParallel instance for the Kleisli type in the Eden backend}
+\label{fig:ArrowParallelEdenKleisli}
+\end{figure}
+
+\FloatBarrier
+
+\subsection{Impact of parallel Arrows}
+\olcomment{move this to Contributions in the front or something}
+We have seen that we can wrap parallel Haskells inside of the \inlinecode{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
+\begin{itemize}
+	\item \textbf{Arrow API benefits}:
+	With the \inlinecode{ArrowParallel} typeclass we do not lose any benefits of using arrows as \inlinecode{parEvalN} is just yet another arrow combinator.
+The resulting arrow can be used in the same way a potential serial version could be used.
+This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
+We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
+	\item \textbf{Abstraction}:
+	With the \inlinecode{ArrowParallel} typeclass, we abstracted all parallel implementation logic away from the business logic.
+This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
+So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \inlinecode{ArrowParallel} instance.
+\end{itemize}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 	%\pagebreak
-	%include basicSkeletons.tex
+	\subsection{Extending the Interface}
+\label{sec:extending-interface}
+With the \inlinecode{ArrowParallel} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
+These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
+
+\subsubsection{Lazy \inlinecode{parEvalN}}
+\begin{figure}[h]
+	\includegraphics[scale=0.7]{images/parEvalNLazy}
+	\caption{Schematic depiction of parEvalNLazy}
+	\label{fig:parEvalNLazyImg}
+\end{figure}
+The function \inlinecode{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
+Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
+\inlinecode{map (arr .
+(+)) [1..]} or just because we need the arrows evaluated in chunks.
+\inlinecode{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
+It works by first chunking the input from \inlinecode{[a]} to \inlinecode{[[a]]} with the given \inlinecode{ChunkSize} in \inlinecode{arr (chunksOf chunkSize)}.
+These chunks are then fed into a list \inlinecode{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \inlinecode{ChunkSize} into the regular parEvalN by using \inlinecode{listApp} (Fig.~\ref{fig:listApp}).
+The resulting \inlinecode{[[b]]} is lastly converted into \inlinecode{[b]} with \inlinecode{arr concat}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
+	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
+parEvalNLazy conf chunkSize fs =
+	arr (chunksOf chunkSize) >>>
+	listApp fchunks >>>
+	arr concat
+	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
+\end{lstlisting} %$ %% formatting
+\caption{Definition of parEvalNLazy}
+\label{fig:parEvalNLazy}
+\end{figure}
+
+\subsubsection{Heterogenous tasks}
+\begin{figure}[h]
+	\includegraphics[scale=0.7]{images/parEval2}
+	\caption{Schematic depiction of parEval2}
+	\label{fig:parEval2Img}
+\end{figure}
+We have only talked about the paralellization arrows of the same type until now.
+But sometimes we want to paralellize heterogenous types as well.
+However, we can implement such a \inlinecode{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into a new parallel arrow \inlinecode{arr (a, c) (b, d)} quite easily with the help of the \inlinecode{ArrowChoice} typeclass.
+The idea is to use the \inlinecode{+++} combinator which combines two arrows \inlinecode{arr a b} and \inlinecode{arr c d} and transforms them into \inlinecode{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
+\\\\
+We start by transforming the \inlinecode{(a, c)} input into a 2-element list \inlinecode{[Either a c]} by first tagging the two inputs with \inlinecode{Left} and \inlinecode{Right} and wrapping the right element in a singleton list with \inlinecode{return} so that we can combine them with \inlinecode{arr (uncurry (:))}.
+Next, we feed this list into a parallel arrow running on 2 instances of \inlinecode{f +++ g} as described above.
+After the calculation is finished, we convert the resulting \inlinecode{[Either b d]} into \inlinecode{([b], [d])} with \inlinecode{arr partitionEithers}.
+The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \inlinecode{(b, d)} in the last step.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+parEval2 :: (ArrowChoice arr,
+	ArrowParallel arr (Either a c) (Either b d) conf) =>
+	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
+parEval2 conf f g =
+	arr Left *** (arr Right >>> arr return) >>>
+	arr (uncurry (:)) >>>
+	parEvalN conf (replicate 2 (f +++ g)) >>>
+	arr partitionEithers >>>
+	arr head *** arr head
+\end{lstlisting}
+	\caption{Definition of parEval2}
+	\label{fig:parEval2}
+\end{figure}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 	%\pagebreak
-	%include syntacticSugar.tex
+	\subsubsection{Syntactic Sugar} \label{syntacticSugar}
+For basic arrows, we have the \inlinecode{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into an arrow \inlinecode{arr (a, c) (b, d)} which does both computations at once.
+This can easily be translated into a parallel version \inlinecode{(\ensuremath{\mathbin{***}})} (Fig.~\ref{fig:par***}) with the use of \inlinecode{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \inlinecode{()} as the conf parameter in Fig.~\ref{fig:par***}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+(\ensuremath{\mathbin{***}}) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
+	arr a b -> arr c d -> arr (a, c) (b, d)
+(\ensuremath{\mathbin{***}}) = parEval2 ()
+\end{lstlisting}
+\caption{Definition of (\ensuremath{\mathbin{***}}) - the parallel version of (***)}
+\label{fig:par***}
+\end{figure}
+% With this we can analogously to the serial \inlinecode{&&&}
+We define the parallel \inlinecode{(\ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}})} (Fig.~\ref{fig:par&&&}) in a similar manner to its sequential pendant \inlinecode{(\&\&\&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+(\ensuremath{\mathbin{\&\&\&}}) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
+	arr a b -> arr a c -> arr a (b, c)
+(\ensuremath{\mathbin{\&\&\&}}) f g = (arr $ \a -> (a, a)) >>> f \ensuremath{\mathbin{***}} g
+\end{lstlisting} % $ %% formatting
+\caption{Definition of (\ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}}) - the parallel version of (\&\&\&)}
+\label{fig:par&&&}
+\end{figure}
 	%\pagebreak
-	%include futures.tex
+	\section{Futures} \label{futures}
+Consider the parallel arrow combinator in Fig.~\ref{fig:someCombinator}
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
+someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
+\end{lstlisting}
+\caption{An example parallel Arrow combinator without Futures}
+\label{fig:someCombinator}
+\end{figure}
+In a distributed environment, the resulting arrow of this combinator first evaluates all \inlinecode{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \inlinecode{[arr b c]} in parallel to then gather the input once again on the master node.
+Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
+
+While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \inlinecode{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \inlinecode{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+This can become a serious bottleneck %in heavy threaded applications.
+for larger amount of data and number of processes \citep[showcases][as, \eg]{Berthold2009-fft}.
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
+	\caption[without Futures]{Communication between 4 threads without Futures.
+All communication goes through the master node.
+Each bar represents one process.
+Black lines between processes represent communication.
+Colors: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
+	\label{fig:withoutFutures}
+\end{figure}
+
+This motivates for an approach that allows the nodes to communicate directly with each other.
+Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \inlinecode{RD} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
+
+But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \inlinecode{Future} typeclass (Fig.~\ref{fig:future}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+class Future fut a \text{\tt \char124{}} a -> fut where
+    put :: (Arrow arr) => arr a (fut a)
+    get :: (Arrow arr) => arr (fut a) a
+\end{lstlisting}
+\caption{Definition of the Future typeclass}
+\label{fig:future}
+\end{figure}
+Since \inlinecode{RD} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \inlinecode{Strategy a} in the Multicore Haskell implementation of \inlinecode{ArrowParallel} in chapter \ref{sec:parrows:multicore}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+data RemoteData a = RD { rd :: RD a }
+
+instance (Trans a) => Future RemoteData a where
+    put = arr (\a -> RD { rd = release a })
+    get = arr rd >>> arr fetch
+\end{lstlisting}
+\caption{RD based RemoteData version of Future for the Eden backend}
+\label{fig:RDFuture}
+\end{figure}
+
+For our Par Monad and Multicore Haskell backends, we can simply use \inlinecode{MVar}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
+\fixme{explain MVars}
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+{-# NOINLINE putUnsafe #-}
+putUnsafe :: a -> MVar a
+putUnsafe a = unsafePerformIO $ do
+    mVar <- newEmptyMVar
+    putMVar mVar a
+    return mVar
+
+instance (NFData a) => Future MVar a where
+    put = arr putUnsafe
+    get = arr takeMVar >>> arr unsafePerformIO
+\end{lstlisting}
+\caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends}
+\label{fig:MVarFuture}
+\end{figure}
+
+Furthermore, in order for these \inlinecode{Future} types to fit with the \inlinecode{ArrowParallel} instances we gave earlier, we have to give the necessary \inlinecode{NFData} and \inlinecode{Trans} instances - the latter only being needed in Eden.
+Because \inlinecode{MVar} already ships with a \inlinecode{NFData} instance, we only have to supply two simple instances for our \inlinecode{RemoteData} type.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+instance NFData (RemoteData a) where
+    rnf = rnf . rd
+instance Trans (RemoteData a)
+\end{lstlisting}
+\caption{NFData and Trans instances for the RemoteData type.
+The Trans instance does not have any functions declared as the default implementation suffices here.
+See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
+\end{figure}
+
+Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
+someCombinator fs1 fs2 =
+	parEvalN () (map (>>> put) fs1) >>>
+	rightRotate >>>
+	parEvalN () (map (get >>>) fs2)
+\end{lstlisting}
+\caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel}
+\label{fig:someCombinatorParallel}
+\end{figure}
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear.
+Do Figs 2-3 with a lot of load?}
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{images/withFutures}
+	\caption[with Futures]{Communication between 4 threads with Futures.
+Other than in Fig.~\ref{fig:withoutFutures}, threads communicate directly (black lines between the bars) instead of always going through the master node (bottom bar).}
+	\label{fig:withFutures}
+\end{figure}
 	%\pagebreak
-	%include mapSkeletons.tex
+
+\FloatBarrier
+\section{Map-based Skeletons}
+\label{sec:map-skeletons}
+Now we have developed Parallel Arrows far enough to define some algorithmic skeletons useful to an application programmer.
+\subsection{Parallel map}
+\begin{figure}[h]
+	\includegraphics[scale=0.7]{images/parMap}
+	\caption{Schematic depiction of parMap}
+	\label{fig:parMapImg}
+\end{figure}
+\inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
+We can implement it with \inlinecode{ArrowParallel} by repeating an arrow \inlinecode{arr a b} and then passing it into \inlinecode{parEvalN} to get an arrow \inlinecode{arr [a] [b]}.
+Just like \inlinecode{parEvalN}, \inlinecode{parMap} is 100 \% strict.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+parMap :: (ArrowParallel arr a b conf) =>
+	conf -> (arr a b) -> (arr [a] [b])
+parMap conf f = parEvalN conf (repeat f)
+\end{lstlisting}
+\caption{Definition of parMap}
+\label{fig:parMap}
+\end{figure}
+
+\subsection{Lazy parallel map}
+\begin{figure}[h]
+	\includegraphics[scale=0.7]{images/parMapStream}
+	\caption{Schematic depiction of parMapStream}
+	\label{fig:parMapStreamImg}
+\end{figure}
+As \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \inlinecode{parEvalN} compared to \inlinecode{parEvalNLazy}.
+So it makes sense to also have a \inlinecode{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \inlinecode{parMap}, but uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
+	conf -> ChunkSize -> arr a b -> arr [a] [b]
+parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
+\end{lstlisting}
+\caption{Definition of parMapStream}
+\label{fig:parMapStream}
+\end{figure}
+
+\subsection{Statically load-balancing parallel map}
+\begin{figure}[h]
+	\includegraphics[scale=0.7]{images/farm}
+	\caption{Schematic depiction of farm}
+	\label{fig:farmImg}
+\end{figure}
+A \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \inlinecode{ArrowParallel} we gave in this paper).
+This can be quite wasteful and a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \inlinecode{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+farm :: (ArrowParallel arr a b conf,
+	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
+	conf -> NumCores -> arr a b -> arr [a] [b]
+farm conf numCores f =
+	unshuffle numCores >>>
+	parEvalN conf (repeat (mapArr f)) >>>
+	shuffle
+
+unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
+unshuffle n = arr (\xs -> [takeEach n (drop i xs) \text{\tt \char124{}} i <- [0..n-1]])
+
+takeEach :: Int -> [a] -> [a]
+takeEach n [] = []
+takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
+
+shuffle :: (Arrow arr) => arr [[a]] [a]
+shuffle = arr (concat . transpose)
+\end{lstlisting}
+\caption{Definition of farm. \inlinecode{unshuffle}, \inlinecode{takeEach}, \inlinecode{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
+\label{fig:farm}
+\end{figure}
+
+\subsection{farmChunk}
+\begin{figure}[h]
+	\includegraphics[scale=0.7]{images/farmChunk}
+\caption{Schematic depiction of farmChunk}
+\label{fig:farmChunkImg}
+\end{figure}
+Since a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \inlinecode{parMap} with a different work distribution, it is, again, 100\% strict.
+So we define \inlinecode{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
+	ArrowChoice arr, ArrowApply arr) =>
+	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
+farmChunk conf chunkSize numCores f =
+	unshuffle numCores >>>
+	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
+	shuffle
+\end{lstlisting}
+\caption{Definition of farmChunk}
+\label{fig:farmChunk}
+\end{figure}
+
+\subsection{parMapReduce}
+\olcomment{it appears STRANGE. are the data really left alone and noded after map and taken from there by reduce? It makes sense only is no communication through master takes place. ELSE: CUT!}
+
+\mbcomment{this requires some work. Either change this to use futures or cut, yes.}
+-- this does not completely adhere to Google's definition of Map Reduce as it
+-- the mapping function does not allow for "reordering" of the output
+-- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
+
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+parMapReduceDirect :: (ArrowParallel arr [a] b conf,
+	ArrowApply arr, ArrowChoice arr) =>
+	conf -> ChunkSize -> arr a b -> arr (b, b) b -> b -> arr [a] b
+parMapReduceDirect conf chunkSize mapfn foldfn neutral =
+	arr (chunksOf chunkSize) >>>
+	parMap conf (mapArr mapfn >>> foldlArr foldfn neutral) >>>
+	foldlArr foldfn neutral
+\end{lstlisting}
+\caption{Definition of parMapReduceDirect}
+\label{fig:parMapReduceDirect}
+\end{figure}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 	%\pagebreak
-	%include topologySkeletons.tex
+	\FloatBarrier
+\section{Topological Skeletons}
+\label{sec:topology-skeletons}
+Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them a \inlinecode{pipe}, a \inlinecode{ring} and a \inlinecode{torus} implementation \cite{Loogen2012, eden_skel_topology}.
+These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
+
+\subsection{Parallel pipe}
+
+The parallel pipe skeleton is semantically equivalent to folding over a list \inlinecode{[arr a a]} of arrows with \inlinecode{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
+We implement this skeleton using the \inlinecode{ArrowLoop} typeclass which gives us the \inlinecode{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
+For example %this
+\mbcomment{das kann man hier so lassen, oder?}
+\begin{lstlisting}[frame=htrbl]
+loop (arr (\(a, b) -> (b, a:b)))
+\end{lstlisting}
+, which is the same as
+\begin{lstlisting}[frame=htrbl]
+loop (arr snd &&& arr (uncurry (:)))
+\end{lstlisting}
+defines an arrow that takes its input \inlinecode{a} and converts it into an infinite stream \inlinecode{[a]} of it.
+Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \inlinecode{parEvalN conf fs} inside the second argument of \inlinecode{(\&\&\&)} and then only picking the first element of the resulting list with \inlinecode{arr last}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
+	conf -> [arr a a] -> arr a a
+pipeSimple conf fs =
+	loop (arr snd &&&
+		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
+	arr last
+
+lazy :: (Arrow arr) => arr [a] [a]
+lazy = arr (\ ~(x:xs) -> x : lazy xs)
+\end{lstlisting}
+\caption{A first draft of the pipe skeleton expressed with parallel arrows.
+Note that the use of \inlinecode{lazy} is essential as without it programs using this definition would never halt.
+We need to enforce that the evaluation of the input \inlinecode{[a]} terminates before passing it into \inlinecode{parEvalN}.}
+\label{fig:pipeSimple}
+\end{figure}
+
+However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
+Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \inlinecode{pipe} \ref{fig:pipe}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf,
+	Future fut a) =>
+	conf -> [arr a a] -> arr a a
+pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
+\end{lstlisting}
+\caption{Final definition of the pipe skeleton which uses Futures}
+\label{fig:pipe}
+\end{figure}
+
+Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
+\inlinecode{arr a b} and \inlinecode{arr b c}.
+By wrapping these two arrows inside a common type we obtain \inlinecode{pipe2} (Fig.~\ref{fig:pipe2}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+pipe2 :: (ArrowLoop arr, ArrowChoice arr,
+	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
+	Future fut (([a], [b]), [c])) =>
+	conf -> arr a b -> arr b c -> arr a c
+pipe2 conf f g =
+	(arr return &&& arr (const [])) &&& arr (const []) >>>
+	pipe conf (replicate 2 (unify f g)) >>>
+	arr snd >>>
+	arr head
+		where
+			unify :: (ArrowChoice arr) =>
+				arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
+			unify f g =
+				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
+				arr (\((a, b), c) -> ((c, a), b))
+\end{lstlisting}
+\caption{Definition of pipe2}
+\label{fig:pipe2}
+\end{figure}
+
+Note that extensive use of this combinator over \inlinecode{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Nonetheless, we can define a parallel piping operator \inlinecode{(\ensuremath{\mathbin{>>>}})} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+(\ensuremath{\mathbin{>>>}}) :: (ArrowLoop arr, ArrowChoice arr,
+	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
+	Future fut (([a], [b]), [c])) =>
+	arr a b -> arr b c -> arr a c
+(\ensuremath{\mathbin{>>>}}) = pipe2 ()
+\end{lstlisting}
+\caption{Definition of \texttt{(\ensuremath{\mathbin{>>>}})}}
+\label{fig:par>>>}
+\end{figure}
+
+\subsection{Ring skeleton} \label{sec:ring}
+\begin{figure}[h]
+	\includegraphics[scale=0.75]{images/ring}
+	\caption{Schematic depiction of the ring skeleton}
+	\label{fig:ringImg}
+\end{figure}
+Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \inlinecode{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
+Its input is a node function \inlinecode{i -> r -> (o, r)} in which \inlinecode{r} serves as the intermediary output that gets send to the neighbour of each node.
+This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+ringSimple :: (Trans i, Trans o, Trans r) =>
+   (i -> r -> (o,r))
+   -> [i] -> [o]
+ringSimple f is =  os
+  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
+                                   (zip is $ lazy ringIns))
+        ringIns = rightRotate ringOuts
+
+toRD :: (Trans i, Trans o, Trans r) =>
+        ((i,r) -> (o,r))
+        -> ((i, RD r) -> (o, RD r))
+toRD  f (i, ringIn)  = (o, release ringOut)
+  where (o, ringOut) = f (i, fetch ringIn)
+
+rightRotate    :: [a] -> [a]
+rightRotate [] =  []
+rightRotate xs =  last xs : init xs
+
+lazy :: [a] -> [a]
+lazy ~(x:xs) = x : lazy xs
+\end{lstlisting}
+\caption{Eden's definition of the ring skeleton \citep{eden_skel_topology}}
+\label{fig:ringEden}
+\end{figure}
+%\end{lstlisting}
+%with toRD (to make use of remote data)
+%\begin{lstlisting}[frame=htrbl]
+%\end{lstlisting}
+%and rightRotate:
+%\begin{lstlisting}[frame=htrbl]
+
+We can rewrite its functionality easily with the use of \inlinecode{loop} as the definition of the node function, \inlinecode{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \inlinecode{loop}'s \inlinecode{arr (a, b) (c, b) -> arr a c}.
+In each iteration we start by rotating the intermediary input from the nodes \inlinecode{[fut r]} with \inlinecode{second (rightRotate >>> lazy)}.
+Similarly to the \inlinecode{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \inlinecode{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
+%\begin{quotation}
+\enquote{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
+Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
+%\end{quotation}
+Next, we zip the resulting \inlinecode{([i], [fut r])} to \inlinecode{[(i, fut r)]} with \inlinecode{arr (uncurry zip)} so we can feed that into a our input arrow \inlinecode{arr (i, r) (o, r)}, which we transform into \inlinecode{arr (i, fut r) (o, fut r)} before lifting it to \inlinecode{arr [(i, fut r)] [(o, fut r)]} to get a list \inlinecode{[(o, fut r)]}.
+Finally we unzip this list into \inlinecode{([o], [fut r])}.
+Plugging this arrow \inlinecode{arr ([i], [fut r]) ([o], fut r)} into the definition of \inlinecode{loop} from earlier gives us \inlinecode{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+ring :: (ArrowLoop arr, Future fut r,
+	ArrowParallel arr (i, fut r) (o, fut r) conf) =>
+    conf ->
+    arr (i, r) (o, r) ->
+    arr [i] [o]
+ring conf f =
+	loop (second (rightRotate >>> lazy) >>>
+    arr (uncurry zip) >>>
+    parMap conf (second get >>> f >>> second put) >>>
+    arr unzip)
+
+rightRotate :: (Arrow arr) => arr [a] [a]
+rightRotate = arr $ \list -> case list of
+  [] -> []
+  xs -> last xs : init xs
+
+lazy :: (Arrow arr) => arr [a] [a]
+lazy = arr (\ ~(x:xs) -> x : lazy xs)
+\end{lstlisting}
+\caption{Final Definition of the ring Skeleton}
+\label{fig:ringFinal}
+\end{figure}
+ %$ %% formatting
+% and lazy:
+% \begin{lstlisting}[frame=htrbl]
+% lazy :: (Arrow arr) => arr [a] [a]
+% lazy = arr (\ ~(x:xs) -> x : lazy xs
+% \end{lstlisting}
+%% we have shown this already.
+This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
+%Further details on this can be found in \cite{eden_cefp}.
+\olcomment{let's do it?}
+\subsection{Torus skeleton}
+\begin{figure}
+	\includegraphics[scale=0.75]{images/torus}
+	\caption{Schematic depiction of the torus skeleton}
+	\label{fig:ringTorusImg}
+\end{figure}
+If we take the concept of a ring from \ref{sec:ring} one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
+Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
+\\\\
+With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \inlinecode{ArrowLoop} typeclass.
+\\\\
+Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple \inlinecode{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \inlinecode{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \inlinecode{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \inlinecode{[[fut a]]} horizontally and \inlinecode{[[fut b]]} vertically.
+Then, we once again zip the inputs for the input arrow with \inlinecode{arr (uncurry3 zipWith3 lazyzip3)} from \inlinecode{([[c]], ([[fut a]], [[fut b]]))} to \inlinecode{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
+\\\\
+This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
+We first have to \inlinecode{shuffle} all the inputs to then pass it into \inlinecode{parMap conf (ptorus f)} which yields us \inlinecode{[(d, fut a, fut b)]}.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \inlinecode{arr length >>> arr unshuffle} with the use of \inlinecode{app :: arr (arr a b, a) c} from the \inlinecode{ArrowApply} typeclass.
+Finally, we unpack this matrix \inlinecode{[[[(d, fut a, fut b)]]} with \inlinecode{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \inlinecode{([[d]], ([[fut a]], [[fut b]]))}.
+\\\\
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
+	ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
+	Future fut a, Future fut b) =>
+	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
+torus conf f =
+	loop (second ((mapArr rightRotate >>> lazy) ***
+		(arr rightRotate >>> lazy)) >>>
+	arr (uncurry3 (zipWith3 lazyzip3)) >>>
+	(arr length >>> arr unshuffle) &&&
+		(shuffle >>> parMap conf (ptorus f)) >>>
+	app >>>
+	arr (map unzip3) >>> arr unzip3 >>> threetotwo)
+
+ptorus :: (Arrow arr, Future fut a, Future fut b) =>
+	arr (c, a, b) (d, a, b) ->
+	arr (c, fut a, fut b) (d, fut a, fut b)
+ptorus f =
+	arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>>
+	arr (\ ~(d, a, b) -> (d, put a, put b))
+
+uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
+uncurry3 f (a, (b, c)) = f a b c
+
+lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
+lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
+
+threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
+threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
+\end{lstlisting}
+\caption{Definition of the torus skeleton}
+\label{fig:torus}
+\end{figure}
+As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}.
+Their nodefunction can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+nodefunction :: Int ->
+	((Matrix, Matrix), [Matrix], [Matrix]) ->
+	([Matrix], [Matrix], [Matrix])
+nodefunction n ((bA, bB), rows, cols) =
+	([bSum], bA:nextAs , bB:nextBs)
+		where bSum =
+			foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
+			nextAs = take (n-1) rows
+			nextBs = take (n-1) cols
+\end{lstlisting}
+\caption{Adapted nodefunction for matrix multiplication with the torus from Fig.~\ref{fig:torus}}
+\label{fig:torusMatMult}
+\end{figure}
+If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.\olcomment{much more details on this!}
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
+	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (Parrows)}
+	\label{fig:torus_parrows_trace}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
+	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden)}
+	\label{fig:torus_eden_trace}
+\end{figure}
+
+\FloatBarrier
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 	%\pagebreak
-	%include benchmarks.tex
+	\section{Benchmarks}
+\label{sec:benchmarks}
+
+%\begin{figure}[ht]
+%	\centering
+	\framebox{
+		\begin{tikzpicture}[thick, scale=0.7]
+		\begin{axis}[
+			xbar, xmin=0,
+			bar width=0.4cm,
+			width=12cm, height=7cm, enlarge y limits=0.1,
+			ylabel={amount of cores},
+			xlabel={mean runtime (seconds)},
+			symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
+			ytick=data,
+			nodes near coords, nodes near coords align={horizontal},
+			]
+			\addplot coordinates {(841,2) (248,4) (115,8) (63.9,16) (41.6,32) (34.6,48) (33.0,64) (32.9,96)};
+		\end{axis}
+		\end{tikzpicture}
+	}
+%	\caption[Parallel Matrix Multiplication using Eden (two 1024x1024 matrices)]
+%\end{figure}
+
+\framebox{
+	\begin{tikzpicture}[thick, scale=0.7]
+	\begin{axis}[
+	xbar, xmin=0,
+	bar width=0.4cm,
+	width=12cm, height=7cm, enlarge y limits=0.1,
+	ylabel={amount of cores},
+	xlabel={mean seconds run on torus\_matrix\_parrows (two 1024x1024 matrices)},
+	symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
+	ytick=data,
+	nodes near coords, nodes near coords align={horizontal},
+	]
+	\addplot coordinates {(663,2) (205,4) (113,8) (68.5,16) (41.9,32) (35.4,48) (36.3,64) (33.4,96)};
+	\end{axis}
+	\end{tikzpicture}
+}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 	%%\pagebreak
-	%include conclusion.tex
+	\section{Conclusion}
+\label{sec:conclusion}
+Arrows are a generic concept that allows for powerful composition combinators.
+To our knowledge we are the first ones to represent parallel computation with arrows.\olcomment{that strange arrows-based robot interaction paper from 1993 or so!
+clearly discuss in related work}
+
+Arrows turn out to be a useful tool for composing in parallel programs.
+We do not have to introduce new monadic types that wrap the computation.
+Instead use arrows just like regular sequential pure functions.
+%
+This work features multiple parallel backends: the already available parallel Haskell flavours.
+Parallel Arrows feature an implementation of the \inlinecode{ArrowParallel} interface for Multicore Haskell, \inlinecode{Par} monad, and Eden.
+With our approach parallel programs can be ported across these flavours with no effort.
+%
+%
+Performancewise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}
+
+\mbcomment{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}
+
+\olcomment{Remove websites from citations, put them into footnotes!}
+
+\olcomment{Parrows + accelerate = love? Metion port to Frege. Mention the Par monad troubles.}
 	%\pagebreak
@@
         \appendix
-	%include utilityFunctions.tex
+	\section{Utility Functions}\label{utilfns}
+To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \inlinecode{map}, \inlinecode{foldl} and \inlinecode{zipWith} on arrows.
+
+The \inlinecode{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \inlinecode{arr a b} to an arrow \inlinecode{arr [a] [b]} \cite{programming_with_arrows}:
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
+mapArr f =
+	arr listcase >>>
+	arr (const []) pipepipepipe (f *** mapArr f >>> arr (uncurry (:)))
+	where listcase [] = Left ()
+	      listcase (x:xs) = Right (x,xs)
+\end{lstlisting}
+\caption{\inlinecode{map} over arrows}
+\label{fig:mapArr}
+\end{figure}
+Similarly, we can also define \inlinecode{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \inlinecode{arr (b, a) b} with a neutral element \inlinecode{b} to \inlinecode{arr [a] b}:
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
+foldlArr f b =
+	arr listcase >>>
+	arr (const b) pipepipepipe
+		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
+	where listcase [] = Left []
+	      listcase (x:xs) = Right (x,xs)
+\end{lstlisting}
+\caption{\inlinecode{foldl} over arrows}
+\label{fig:foldlArr}
+\end{figure}
+\fixme{pipepipepipe does not work with lhs2TeX}
+Finally, with the help of \inlinecode{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \inlinecode{arr (a, b) c} to an arrow \inlinecode{arr ([a], [b]) [c]}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
+zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
+\end{lstlisting}
+\caption{\inlinecode{zipWith} over arrows}
+\label{fig:zipWithArr}
+\end{figure}
+ %$ %% formatting
+These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \inlinecode{(\textpipe\textpipe\textpipe)} combinator.
+It takes two arrows \inlinecode{arr a c} and \inlinecode{arr b c} and combines them into a new arrow \inlinecode{arr (Either a b) c} which pipes all \inlinecode{Left a}'s to the first arrow and all \inlinecode{Right b}'s to the second arrow.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+(\text{\tt pipepipepipe}) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
+\end{lstlisting}
+\caption{Type signature of \inlinecode{(\textpipe\textpipe\textpipe)}}
+\label{fig:codeSigPipePipePipe}
+\end{figure}
+\fixme{pipepipepipe does not work with lhs2TeX}
+
+With the zipWithArr combinator we can also write a combinator \inlinecode{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \inlinecode{[arr a b]} to an arrow \inlinecode{arr [a] [b]}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
+listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
+\end{lstlisting}
+\caption{Definition of \inlinecode{listApp}}
+\label{fig:listApp}
+\end{figure}
+% $ %% formatting
+. Note that  this additionally makes use of the \inlinecode{ArrowApply} typeclass that allows us to evaluate arrows with \inlinecode{app :: arr (arr a b, a) c}.
+
+% $ %% formatting
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:
 \end{document}

### 2017-05-18 paper/content/main.tex
I broke lhs2tex, lol

@@
 \documentclass{jfp1}
+
+%% add a tweak to polycode, similar to literate in lstlistings, but uglier
+%% ^^^^ you want this. and similar for ***, +++, etc.
+%% a hack for list comprehensions' <-, typeset it as <--
+%%%%%format <--        = "\in "
+%fornat ` = "\`"
 
@@
 
-\usepackage{subcaption}
+%\usepackage{subcaption}
 
@@
 
-%\usepackage{ifpdf} %% broken?!!
-%\ifpdf
-%\usepackage[activate={true,nocompatibility},kerning=true,spacing=true,tracking=true]{microtype}
-%\else
-%% non-pdf latex can less, let microtype itself decide
-\usepackage{microtype}
-%\fi
+%% more microtype
 \microtypecontext{spacing=nonfrench} %% log said so
+
+\usepackage{subfig}
 
@@
 
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \inlinecode{a -> b} in parallel:
-
-\begin{figure}[h]
-%\begin{lstlisting}[frame=htrbl]
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \inlinecode{a -> b} in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:
+
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-%\end{lstlisting}
-\caption{parEvalN's type signature}
-\label{fig:parEvalNTypeSig}
-\end{figure}
-\begin{figure}[h]
+
+\begin{figure}[h]
+  \centering
 	\includegraphics[scale=0.7]{images/parEvalN}
-	\caption{Schematic illustration of parEvalN}
+	\caption{Schematic illustration of \texttt{parEvalN}.}
 	\label{fig:parEvalN}
 \end{figure}
-\olcomment{make them to real figures? with environment, reference and such?}
+
 Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \inlinecode{parEvalN} with several different parallel Haskells.
@@
 
-\begin{lstlisting}[frame=htrbl]
-parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = let bs = zipWith ($) fs as in
-	(bs `using` parList rdeepseq) `pseq` bs
-\end{lstlisting}
+% \begin{code}
+% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+% parEvalN fs as = let bs = zipWith ($) fs as 
+%                  in (bs `using` parList rdeepseq) `pseq` bs
+% \end{code}
+
 \begin{figure}[h]
@@
 
-\begin{lstlisting}[frame=htrbl]
-parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = runPar $ 
-	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{runPar}\mathbin{\$}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}(\Varid{sequenceA}\mathbin{\$}\Varid{map}\;(\Varid{spawnP})\mathbin{\$}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as})\bind \Varid{mapM}\;\Varid{get}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \begin{figure}[h]
@@
 
-\begin{lstlisting}[frame=htrbl]
-parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
-parEvalN = spawnF 
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{parEvalN}\mathrel{=}\Varid{spawnF}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \begin{figure}[h]
@@
 Note that extensive use of this combinator over \inlinecode{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \inlinecode{(\ensuremath{\mathbin{>>>}})} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(\ensuremath{\mathbin{>>>}}) :: (ArrowLoop arr, ArrowChoice arr,
+Nonetheless, we can define a parallel piping operator \inlinecode{(\ensuremath{\mathbin{>\!>\!>}})} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+\begin{figure}[h]
+\begin{lstlisting}[frame=htrbl]
+(\ensuremath{\mathbin{>\!>\!>}}) :: (ArrowLoop arr, ArrowChoice arr,
 	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
@@
 	arr a b -> arr b c -> arr a c
-(\ensuremath{\mathbin{>>>}}) = pipe2 ()
-\end{lstlisting}
-\caption{Definition of \texttt{(\ensuremath{\mathbin{>>>}})}}
+(\ensuremath{\mathbin{>\!>\!>}}) = pipe2 ()
+\end{lstlisting}
+\caption{Definition of \texttt{(\ensuremath{\mathbin{>\!>\!>}})}}
 \label{fig:par>>>}

### 2017-05-18 paper/content/main.tex
lhs2tex everywhere; problem with subfigures

@@
 % für Listings
-\usepackage{listings}
+%\usepackage{listings}
 
@@
 
-\usepackage{tipa}
+%%\usepackage{tipa} %% causes problems with lhs2tex?!
 
@@
 
-% \begin{code}
-% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-% parEvalN fs as = let bs = zipWith ($) fs as 
-%                  in (bs `using` parList rdeepseq) `pseq` bs
-% \end{code}
+%%%% BROKEN!
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\mathbf{let}\;\Varid{bs}\mathrel{=}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as}{}\<[E]%
+\\
+\>[B]{}\hsindent{18}{}\<[18]%
+\>[18]{}\mathbf{in}\;(\Varid{bs}\mathbin{`\Varid{using}`}\Varid{parList}\;\Varid{rdeepseq})\mathbin{`\Varid{pseq}`}\Varid{bs}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 
@@
 	\caption{Dataflow of the Multicore Haskell parEvalN version}
+\olcomment{Evaluation step explicitly shown?}
 \end{figure} %$ %% formatting
@@
 To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
-While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See .....
-on hackage for the last available version of Eden TraceViewer.
-There was an effort to implement the TraceViewer using modern web technologies \cite{traceviewer-web}.} \cite{Berthold2007a}.
+While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}.
 In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.
@@
 \begin{figure}[h]
-\begin{subfigure}{0.5\textwidth}
-%\begin{minipage}{0.5\textwidth}
-\begin{lstlisting}[frame=htrbl]
-class Arrow arr where
-arr :: (a -> b) -> arr a b
-
-(>>>) :: arr a b -> arr b c -> arr a c
-
-first :: arr a b -> arr (a,c) (b,c)
-\end{lstlisting}
-\caption{Arrow class definition}
-%\end{minipage}
-\end{subfigure}
-~~~~
-\begin{subfigure}{0.45\textwidth}
+\centering
+\subfloat[Arrow class definition][]{%
+% begin{minipage}{0.5\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{arr}\mathbin{::}(\Varid{a}\to \Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{b}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}(\mathbin{>\!>\!>})\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{first}\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{c}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+% caption{Arrow class definition}
+}%\hfill
+\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
 	\begin{center}
-	\subcaptionbox{arr}{\includegraphics[scale=0.6]{images/arr}}
-	\subcaptionbox{composition \inlinecode{(>>>)}}{\includegraphics[scale=0.6]{images/compose}}
-	\subcaptionbox{first}{\includegraphics[scale=0.6]{images/first}}
+	{\includegraphics[scale=0.6]{images/arr}}
+	{\includegraphics[scale=0.6]{images/compose}}
+	{\includegraphics[scale=0.6]{images/first}}
 	\end{center}
-\end{subfigure}
-\caption{The arrow type class definition on the left with schematic depiction of its combinators on the right}
+}%\hfill
+%\caption{The arrow type class definition on the left with schematic depiction of its combinators on the right}
 \end{figure}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance Arrow (->) where
-	arr f = f
-	f >>> g = g . f
-	first f = \(a, c) -> (f a, c) 
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;\Conid{Arrow}\;(\to )\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{f}\mathrel{=}\Varid{f}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{f}\mathbin{>\!>\!>}\Varid{g}\mathrel{=}\Varid{g}\mathbin{\circ}\Varid{f}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{first}\;\Varid{f}\mathrel{=}\lambda (\Varid{a},\Varid{c})\to (\Varid{f}\;\Varid{a},\Varid{c}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Arrow instance for regular functions}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-data Kleisli m a b = Kleisli { run :: a -> m b }
-
-instance Monad m => Arrow (Kleisli m) where
-	arr f = Kleisli (return . f)
-	f >>> g = Kleisli (\a -> f a >>= g)
-	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
-\end{lstlisting}%$
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{Kleisli}\;\Varid{m}\;\Varid{a}\;\Varid{b}\mathrel{=}\Conid{Kleisli}\;\{\mskip1.5mu \Varid{run}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{b}\mskip1.5mu\}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Varid{m}\Rightarrow \Conid{Arrow}\;(\Conid{Kleisli}\;\Varid{m})\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{f}\mathrel{=}\Conid{Kleisli}\;(\Varid{return}\mathbin{\circ}\Varid{f}){}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{f}\mathbin{>\!>\!>}\Varid{g}\mathrel{=}\Conid{Kleisli}\;(\lambda \Varid{a}\to \Varid{f}\;\Varid{a}\bind \Varid{g}){}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{first}\;\Varid{f}\mathrel{=}\Conid{Kleisli}\;(\lambda (\Varid{a},\Varid{c})\to \Varid{f}\;\Varid{a}\bind \lambda \Varid{b}\to \Varid{return}\;(\Varid{b},\Varid{c})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%$
 \caption{Definition of the Kleisli type and the corresponding arrow instance}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-second :: Arrow arr => arr a b -> arr (c, a) (c, b)
-second f = arr swap >>> first f >>> arr swap
-	where swap (x, y) = (y, x)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{second}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{c},\Varid{a})\;(\Varid{c},\Varid{b}){}\<[E]%
+\\
+\>[B]{}\Varid{second}\;\Varid{f}\mathrel{=}\Varid{arr}\;\Varid{swap}\mathbin{>\!>\!>}\Varid{first}\;\Varid{f}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{swap}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\mathbf{where}\;\Varid{swap}\;(\Varid{x},\Varid{y})\mathrel{=}(\Varid{y},\Varid{x}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{The second combinator}
@@
 \end{figure}
-the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
-f *** g = first f >>> second g
-\end{lstlisting}
+the \inlinecode{***} combinator which combines \includegraphics{first} and \inlinecode{second} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{***})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
+\\
+\>[B]{}\Varid{f}\mathbin{***}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!>\!>}\Varid{second}\;\Varid{g}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{The (***) combinator}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
-f &&& g = arr (\a -> (a, a)) >>> (f *** g)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{\&\&\&})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{a}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
+\\
+\>[B]{}\Varid{f}\mathbin{\&\&\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!>\!>}(\Varid{f}\mathbin{***}\Varid{g}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{The (\&\&\&) combinator}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
-add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{add}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}{}\<[E]%
+\\
+\>[B]{}\Varid{add}\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{f}\mathbin{\&\&\&}\Varid{g})\mathbin{>\!>\!>}\Varid{arr}\;(\lambda (\Varid{u},\Varid{v})\to \Varid{u}\mathbin{+}\Varid{v}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Add over arrows}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{parEvalN Arrow combinator as a function}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-class Arrow arr => ArrowParallel arr a b where
-	parEvalN :: [arr a b] -> arr [a] [b]
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-class Arrow arr => ArrowParallel arr a b conf where
-	parEvalN :: conf -> [arr a b] -> arr [a] [b]
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b () where
-    	parEvalN _ fs =
-       		listApp fs >>>
-        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
-        	arr (uncurry pseq)
-\end{lstlisting}% $ %% formatting
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{rdeepseq}))\mathbin{\&\&\&}\Varid{arr}\;\Varid{id}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{pseq}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+% $ %% formatting
 \caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-data Conf a = Conf (Strategy a)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{Conf}\;\Varid{a}\mathrel{=}\Conid{Conf}\;(\Conid{Strategy}\;\Varid{a}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of Conf a}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b (Conf b) where
-    	parEvalN (Conf strat) fs =
-        	listApp fs >>>
-        	arr (withStrategy (parList strat)) &&& arr id >>>
-        	arr (uncurry pseq)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;(\Conid{Conf}\;\Varid{strat})\;\Varid{fs}\mathrel{=}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{strat}))\mathbin{\&\&\&}\Varid{arr}\;\Varid{id}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{pseq}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b conf where
-		parEvalN _ fs = 
-			(arr $ \as -> (fs, as)) >>>
-			zipWithArr (app >>> arr spawnP) >>>
-			arr sequenceA >>>
-			arr (>>= mapM get) >>>
-			arr runPar
-\end{lstlisting} %$ %% formatting
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{zipWithArr}\;(\Varid{app}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{spawnP})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+ %$ %% formatting
 \caption{ArrowParallel instance for the Par Monad backend}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-class (Arrow arr) => ArrowUnwrap arr where
-	arr a b -> (a -> b)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Conid{ArrowUnwrap}\;\Varid{arr}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to (\Varid{a}\to \Varid{b}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{possible ArrowUnwrap typeclass}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
-parEvalN _ fs as = spawnF fs as
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow \Conid{ArrowParallel}\;(\to )\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{spawnF}\;\Varid{fs}\;\Varid{as}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{ArrowParallel instance for functions in the Eden backend}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance (Monad m, Trans a, Trans b, Trans (m b)) =>
-	ArrowParallel (Kleisli m) a b conf where
-parEvalN conf fs =
-	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
-	(Kleisli $ sequence)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;(\Conid{Monad}\;\Varid{m},\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b},\Conid{Trans}\;(\Varid{m}\;\Varid{b}))\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;(\Conid{Kleisli}\;\Varid{m})\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}(\Varid{arr}\mathbin{\$}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{map}\;(\lambda (\Conid{Kleisli}\;\Varid{f})\to \Varid{f})\;\Varid{fs}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}(\Conid{Kleisli}\mathbin{\$}\Varid{sequence}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
-parEvalNLazy conf chunkSize fs =
-	arr (chunksOf chunkSize) >>>
-	listApp fchunks >>>
-	arr concat
-	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
-\end{lstlisting} %$ %% formatting
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEvalNLazy}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
+\\
+\>[B]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{fs}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{listApp}\;\Varid{fchunks}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{concat}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\mathbf{where}\;\Varid{fchunks}\mathrel{=}\Varid{map}\;(\Varid{parEvalN}\;\Varid{conf})\mathbin{\$}\Varid{chunksOf}\;\Varid{chunkSize}\;\Varid{fs}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+ %$ %% formatting
 \caption{Definition of parEvalNLazy}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parEval2 :: (ArrowChoice arr,
-	ArrowParallel arr (Either a c) (Either b d) conf) =>
-	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
-parEval2 conf f g =
-	arr Left *** (arr Right >>> arr return) >>>
-	arr (uncurry (:)) >>>
-	parEvalN conf (replicate 2 (f +++ g)) >>>
-	arr partitionEithers >>>
-	arr head *** arr head
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEval2}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;\Varid{conf})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
+\\
+\>[B]{}\Varid{parEval2}\;\Varid{conf}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Conid{Left}\mathbin{***}(\Varid{arr}\;\Conid{Right}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{return})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{f}\mathbin{+++}\Varid{g}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{partitionEithers}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{head}\mathbin{***}\Varid{arr}\;\Varid{head}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 	\caption{Definition of parEval2}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(\ensuremath{\mathbin{***}}) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
-	arr a b -> arr c d -> arr (a, c) (b, d)
-(\ensuremath{\mathbin{***}}) = parEval2 ()
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{|***|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;()))\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
+\\
+\>[B]{}(\mathbin{|***|})\mathrel{=}\Varid{parEval2}\;(){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of (\ensuremath{\mathbin{***}}) - the parallel version of (***)}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(\ensuremath{\mathbin{\&\&\&}}) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
-	arr a b -> arr a c -> arr a (b, c)
-(\ensuremath{\mathbin{\&\&\&}}) f g = (arr $ \a -> (a, a)) >>> f \ensuremath{\mathbin{***}} g
-\end{lstlisting} % $ %% formatting
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{|\&\&\&|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{a})\;(\Conid{Either}\;\Varid{b}\;\Varid{c})\;())\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
+\\
+\>[B]{}(\mathbin{|\&\&\&|})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!>\!>}\Varid{f}\mathbin{|***|}\Varid{g}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+ % $ %% formatting
 \caption{Definition of (\ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}}) - the parallel version of (\&\&\&)}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
-someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}\Varid{parEvalN}\;()\;\Varid{fs1}\mathbin{>\!>\!>}\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{parEvalN}\;()\;\Varid{fs2}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{An example parallel Arrow combinator without Futures}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-class Future fut a \text{\tt \char124{}} a -> fut where
-    put :: (Arrow arr) => arr a (fut a)
-    get :: (Arrow arr) => arr (fut a) a
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{class}\;\Conid{Future}\;\Varid{fut}\;\Varid{a}\mathbin{@|@}\Varid{a}\to \Varid{fut}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{put}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{a}){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{get}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Varid{fut}\;\Varid{a})\;\Varid{a}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of the Future typeclass}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-data RemoteData a = RD { rd :: RD a }
-
-instance (Trans a) => Future RemoteData a where
-    put = arr (\a -> RD { rd = release a })
-    get = arr rd >>> arr fetch
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{RemoteData}\;\Varid{a}\mathrel{=}\Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathbin{::}\Conid{RD}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{RemoteData}\;\Varid{a}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{put}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to \Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathrel{=}\Varid{release}\;\Varid{a}\mskip1.5mu\}){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{rd}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{fetch}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{RD based RemoteData version of Future for the Eden backend}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-{-# NOINLINE putUnsafe #-}
-putUnsafe :: a -> MVar a
-putUnsafe a = unsafePerformIO $ do
-    mVar <- newEmptyMVar
-    putMVar mVar a
-    return mVar
-
-instance (NFData a) => Future MVar a where
-    put = arr putUnsafe
-    get = arr takeMVar >>> arr unsafePerformIO
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mbox{\enskip\{-\# NOINLINE putUnsafe  \#-\}\enskip}{}\<[E]%
+\\
+\>[B]{}\Varid{putUnsafe}\mathbin{::}\Varid{a}\to \Conid{MVar}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\Varid{putUnsafe}\;\Varid{a}\mathrel{=}\Varid{unsafePerformIO}\mathbin{\$}\mathbf{do}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{mVar}\leftarrow \Varid{newEmptyMVar}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{putMVar}\;\Varid{mVar}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{return}\;\Varid{mVar}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{MVar}\;\Varid{a}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{put}\mathrel{=}\Varid{arr}\;\Varid{putUnsafe}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{takeMVar}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{unsafePerformIO}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance NFData (RemoteData a) where
-    rnf = rnf . rd
-instance Trans (RemoteData a)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{instance}\;\Conid{NFData}\;(\Conid{RemoteData}\;\Varid{a})\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{rnf}\mathrel{=}\Varid{rnf}\mathbin{\circ}\Varid{rd}{}\<[E]%
+\\
+\>[B]{}\mathbf{instance}\;\Conid{Trans}\;(\Conid{RemoteData}\;\Varid{a}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{NFData and Trans instances for the RemoteData type.
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
-someCombinator fs1 fs2 =
-	parEvalN () (map (>>> put) fs1) >>>
-	rightRotate >>>
-	parEvalN () (map (get >>>) fs2)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\mathbin{>\!>\!>}\Varid{put})\;\Varid{fs1})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{rightRotate}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\Varid{get}\mathbin{>\!>\!>})\;\Varid{fs2}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parMap :: (ArrowParallel arr a b conf) =>
-	conf -> (arr a b) -> (arr [a] [b])
-parMap conf f = parEvalN conf (repeat f)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parMap}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to (\Varid{arr}\;\Varid{a}\;\Varid{b})\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
+\\
+\>[B]{}\Varid{parMap}\;\Varid{conf}\;\Varid{f}\mathrel{=}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;\Varid{f}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of parMap}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> arr a b -> arr [a] [b]
-parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parMapStream}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{parMapStream}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{f}\mathrel{=}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;\Varid{f}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of parMapStream}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-farm :: (ArrowParallel arr a b conf,
-	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
-	conf -> NumCores -> arr a b -> arr [a] [b]
-farm conf numCores f =
-	unshuffle numCores >>>
-	parEvalN conf (repeat (mapArr f)) >>>
-	shuffle
-
-unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
-unshuffle n = arr (\xs -> [takeEach n (drop i xs) \text{\tt \char124{}} i <- [0..n-1]])
-
-takeEach :: Int -> [a] -> [a]
-takeEach n [] = []
-takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
-
-shuffle :: (Arrow arr) => arr [[a]] [a]
-shuffle = arr (concat . transpose)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{farm}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{farm}\;\Varid{conf}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{shuffle}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{unshuffle}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Conid{Int}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{unshuffle}\;\Varid{n}\mathrel{=}\Varid{arr}\;(\lambda \Varid{xs}\to [\mskip1.5mu \Varid{takeEach}\;\Varid{n}\;(\Varid{drop}\;\Varid{i}\;\Varid{xs})\mathbin{@|@}\Varid{i}\leftarrow [\mskip1.5mu \mathrm{0}\mathinner{\ldotp\ldotp}\Varid{n}\mathbin{-}\mathrm{1}\mskip1.5mu]\mskip1.5mu]){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{takeEach}\mathbin{::}\Conid{Int}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{takeEach}\;\Varid{n}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{takeEach}\;\Varid{n}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{takeEach}\;\Varid{n}\;(\Varid{drop}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{xs}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{shuffle}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{shuffle}\mathrel{=}\Varid{arr}\;(\Varid{concat}\mathbin{\circ}\Varid{transpose}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of farm. \inlinecode{unshuffle}, \inlinecode{takeEach}, \inlinecode{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
-	ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
-farmChunk conf chunkSize numCores f =
-	unshuffle numCores >>>
-	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
-	shuffle
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{farmChunk}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{shuffle}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of farmChunk}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parMapReduceDirect :: (ArrowParallel arr [a] b conf,
-	ArrowApply arr, ArrowChoice arr) =>
-	conf -> ChunkSize -> arr a b -> arr (b, b) b -> b -> arr [a] b
-parMapReduceDirect conf chunkSize mapfn foldfn neutral =
-	arr (chunksOf chunkSize) >>>
-	parMap conf (mapArr mapfn >>> foldlArr foldfn neutral) >>>
-	foldlArr foldfn neutral
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parMapReduceDirect}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}\;\Varid{conf},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{b},\Varid{b})\;\Varid{b}\to \Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}{}\<[E]%
+\\
+\>[B]{}\Varid{parMapReduceDirect}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{mapfn}\;\Varid{foldfn}\;\Varid{neutral}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{mapArr}\;\Varid{mapfn}\mathbin{>\!>\!>}\Varid{foldlArr}\;\Varid{foldfn}\;\Varid{neutral})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{foldlArr}\;\Varid{foldfn}\;\Varid{neutral}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of parMapReduceDirect}
@@
 \mbcomment{das kann man hier so lassen, oder?}
-\begin{lstlisting}[frame=htrbl]
-loop (arr (\(a, b) -> (b, a:b)))
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{loop}\;(\Varid{arr}\;(\lambda (\Varid{a},\Varid{b})\to (\Varid{b},\Varid{a}\mathbin{:}\Varid{b}))){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 , which is the same as
-\begin{lstlisting}[frame=htrbl]
-loop (arr snd &&& arr (uncurry (:)))
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\&\&}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 defines an arrow that takes its input \inlinecode{a} and converts it into an infinite stream \inlinecode{[a]} of it.
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
-	conf -> [arr a a] -> arr a a
-pipeSimple conf fs =
-	loop (arr snd &&&
-		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
-	arr last
-
-lazy :: (Arrow arr) => arr [a] [a]
-lazy = arr (\ ~(x:xs) -> x : lazy xs)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{pipeSimple}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{a}\;\Varid{conf})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]\to \Varid{arr}\;\Varid{a}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\Varid{pipeSimple}\;\Varid{conf}\;\Varid{fs}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\&\&}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}(\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:})\mathbin{>\!>\!>}\Varid{lazy})\mathbin{>\!>\!>}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{last}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{lazy}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{lazy}\mathrel{=}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\to \Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{A first draft of the pipe skeleton expressed with parallel arrows.
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf,
-	Future fut a) =>
-	conf -> [arr a a] -> arr a a
-pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{pipe}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{a})\;(\Varid{fut}\;\Varid{a})\;\Varid{conf},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{a})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]\to \Varid{arr}\;\Varid{a}\;\Varid{a}{}\<[E]%
+\\
+\>[B]{}\Varid{pipe}\;\Varid{conf}\;\Varid{fs}\mathrel{=}\Varid{unliftFut}\;(\Varid{pipeSimple}\;\Varid{conf}\;(\Varid{map}\;\Varid{liftFut}\;\Varid{fs})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Final definition of the pipe skeleton which uses Futures}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-pipe2 :: (ArrowLoop arr, ArrowChoice arr,
-	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
-	Future fut (([a], [b]), [c])) =>
-	conf -> arr a b -> arr b c -> arr a c
-pipe2 conf f g =
-	(arr return &&& arr (const [])) &&& arr (const []) >>>
-	pipe conf (replicate 2 (unify f g)) >>>
-	arr snd >>>
-	arr head
-		where
-			unify :: (ArrowChoice arr) =>
-				arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
-			unify f g =
-				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
-				arr (\((a, b), c) -> ((c, a), b))
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{pipe2}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;\Varid{conf},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
+\\
+\>[B]{}\Varid{pipe2}\;\Varid{conf}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}(\Varid{arr}\;\Varid{return}\mathbin{\&\&\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu]))\mathbin{\&\&\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{pipe}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{unify}\;\Varid{f}\;\Varid{g}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{snd}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{head}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\mathbf{where}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{unify}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[25]{}\hsindent{8}{}\<[33]%
+\>[33]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]){}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{unify}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
+\\
+\>[25]{}\hsindent{8}{}\<[33]%
+\>[33]{}(\Varid{mapArr}\;\Varid{f}\mathbin{***}\Varid{mapArr}\;\Varid{g})\mathbin{***}\Varid{arr}\;(\mathbin{\char92 \char95 }\to [\mskip1.5mu \mskip1.5mu])\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[25]{}\hsindent{8}{}\<[33]%
+\>[33]{}\Varid{arr}\;(\lambda ((\Varid{a},\Varid{b}),\Varid{c})\to ((\Varid{c},\Varid{a}),\Varid{b})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of pipe2}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(\ensuremath{\mathbin{>\!>\!>}}) :: (ArrowLoop arr, ArrowChoice arr,
-	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
-	Future fut (([a], [b]), [c])) =>
-	arr a b -> arr b c -> arr a c
-(\ensuremath{\mathbin{>\!>\!>}}) = pipe2 ()
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{|\!>\!>\!>\!|})\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(),{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
+\\
+\>[B]{}(\mathbin{|\!>\!>\!>\!|})\mathrel{=}\Varid{pipe2}\;(){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of \texttt{(\ensuremath{\mathbin{>\!>\!>}})}}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-ringSimple :: (Trans i, Trans o, Trans r) =>
-   (i -> r -> (o,r))
-   -> [i] -> [o]
-ringSimple f is =  os
-  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
-                                   (zip is $ lazy ringIns))
-        ringIns = rightRotate ringOuts
-
-toRD :: (Trans i, Trans o, Trans r) =>
-        ((i,r) -> (o,r))
-        -> ((i, RD r) -> (o, RD r))
-toRD  f (i, ringIn)  = (o, release ringOut)
-  where (o, ringOut) = f (i, fetch ringIn)
-
-rightRotate    :: [a] -> [a]
-rightRotate [] =  []
-rightRotate xs =  last xs : init xs
-
-lazy :: [a] -> [a]
-lazy ~(x:xs) = x : lazy xs
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{ringSimple}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})){}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}\to [\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{ringSimple}\;\Varid{f}\;\Varid{is}\mathrel{=}{}\<[20]%
+\>[20]{}\Varid{os}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{where}\;(\Varid{os},\Varid{ringOuts})\mathrel{=}\Varid{unzip}\;(\Varid{parMap}\;(\Varid{toRD}\mathbin{\$}\Varid{uncurry}\;\Varid{f})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{33}{}\<[36]%
+\>[36]{}(\Varid{zip}\;\Varid{is}\mathbin{\$}\Varid{lazy}\;\Varid{ringIns})){}\<[E]%
+\\
+\>[3]{}\hsindent{6}{}\<[9]%
+\>[9]{}\Varid{ringIns}\mathrel{=}\Varid{rightRotate}\;\Varid{ringOuts}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{toRD}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}((\Varid{i},\Varid{r})\to (\Varid{o},\Varid{r})){}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\to ((\Varid{i},\Conid{RD}\;\Varid{r})\to (\Varid{o},\Conid{RD}\;\Varid{r})){}\<[E]%
+\\
+\>[B]{}\Varid{toRD}\;{}\<[7]%
+\>[7]{}\Varid{f}\;(\Varid{i},\Varid{ringIn}){}\<[22]%
+\>[22]{}\mathrel{=}(\Varid{o},\Varid{release}\;\Varid{ringOut}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{where}\;(\Varid{o},\Varid{ringOut})\mathrel{=}\Varid{f}\;(\Varid{i},\Varid{fetch}\;\Varid{ringIn}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{rightRotate}{}\<[16]%
+\>[16]{}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{rightRotate}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}{}\<[19]%
+\>[19]{}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{rightRotate}\;\Varid{xs}\mathrel{=}{}\<[19]%
+\>[19]{}\Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{lazy}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{lazy}\mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Eden's definition of the ring skeleton \citep{eden_skel_topology}}
@@
 \end{figure}
-%\end{lstlisting}
+%\end{code}
 %with toRD (to make use of remote data)
-%\begin{lstlisting}[frame=htrbl]
-%\end{lstlisting}
+%\begin{code}
+%\end{code}
 %and rightRotate:
-%\begin{lstlisting}[frame=htrbl]
+%\begin{code}
 
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-ring :: (ArrowLoop arr, Future fut r,
-	ArrowParallel arr (i, fut r) (o, fut r) conf) =>
-    conf ->
-    arr (i, r) (o, r) ->
-    arr [i] [o]
-ring conf f =
-	loop (second (rightRotate >>> lazy) >>>
-    arr (uncurry zip) >>>
-    parMap conf (second get >>> f >>> second put) >>>
-    arr unzip)
-
-rightRotate :: (Arrow arr) => arr [a] [a]
-rightRotate = arr $ \list -> case list of
-  [] -> []
-  xs -> last xs : init xs
-
-lazy :: (Arrow arr) => arr [a] [a]
-lazy = arr (\ ~(x:xs) -> x : lazy xs)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{ring}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{r},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})\;\Varid{conf})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{conf}\to {}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})\to {}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{ring}\;\Varid{conf}\;\Varid{f}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{loop}\;(\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{lazy})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!>\!>}\Varid{f}\mathbin{>\!>\!>}\Varid{second}\;\Varid{put})\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{arr}\;\Varid{unzip}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{rightRotate}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{rightRotate}\mathrel{=}\Varid{arr}\mathbin{\$}\lambda \Varid{list}\to \mathbf{case}\;\Varid{list}\;\mathbf{of}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}[\mskip1.5mu \mskip1.5mu]\to [\mskip1.5mu \mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{xs}\to \Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{lazy}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{lazy}\mathrel{=}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\to \Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Final Definition of the ring Skeleton}
@@
 % and lazy:
-% \begin{lstlisting}[frame=htrbl]
+% \begin{code}
 % lazy :: (Arrow arr) => arr [a] [a]
 % lazy = arr (\ ~(x:xs) -> x : lazy xs
-% \end{lstlisting}
+% \end{code}
 %% we have shown this already.
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
-	ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
-	Future fut a, Future fut b) =>
-	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
-torus conf f =
-	loop (second ((mapArr rightRotate >>> lazy) ***
-		(arr rightRotate >>> lazy)) >>>
-	arr (uncurry3 (zipWith3 lazyzip3)) >>>
-	(arr length >>> arr unshuffle) &&&
-		(shuffle >>> parMap conf (ptorus f)) >>>
-	app >>>
-	arr (map unzip3) >>> arr unzip3 >>> threetotwo)
-
-ptorus :: (Arrow arr, Future fut a, Future fut b) =>
-	arr (c, a, b) (d, a, b) ->
-	arr (c, fut a, fut b) (d, fut a, fut b)
-ptorus f =
-	arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>>
-	arr (\ ~(d, a, b) -> (d, put a, put b))
-
-uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
-uncurry3 f (a, (b, c)) = f a b c
-
-lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
-lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
-
-threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
-threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{torus}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;\Varid{conf},{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{a},\Conid{Future}\;\Varid{fut}\;\Varid{b})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to \Varid{arr}\;[\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu]\;[\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{torus}\;\Varid{conf}\;\Varid{f}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{lazy})\mathbin{***}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{lazy}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{uncurry3}\;(\Varid{zipWith3}\;\Varid{lazyzip3}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}(\Varid{arr}\;\Varid{length}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{unshuffle})\mathbin{\&\&\&}{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}(\Varid{shuffle}\mathbin{>\!>\!>}\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f}))\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{app}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!>\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!>\!>}\Varid{threetotwo}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{ptorus}\mathbin{::}(\Conid{Arrow}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{a},\Conid{Future}\;\Varid{fut}\;\Varid{b})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b}){}\<[E]%
+\\
+\>[B]{}\Varid{ptorus}\;\Varid{f}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{c},\Varid{a},\Varid{b})\to (\Varid{c},\Varid{get}\;\Varid{a},\Varid{get}\;\Varid{b}))\mathbin{>\!>\!>}\Varid{f}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{d},\Varid{a},\Varid{b})\to (\Varid{d},\Varid{put}\;\Varid{a},\Varid{put}\;\Varid{b})){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{uncurry3}\mathbin{::}(\Varid{a}\to \Varid{b}\to \Varid{c}\to \Varid{d})\to (\Varid{a},(\Varid{b},\Varid{c}))\to \Varid{d}{}\<[E]%
+\\
+\>[B]{}\Varid{uncurry3}\;\Varid{f}\;(\Varid{a},(\Varid{b},\Varid{c}))\mathrel{=}\Varid{f}\;\Varid{a}\;\Varid{b}\;\Varid{c}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{lazyzip3}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{c}\mskip1.5mu]\to [\mskip1.5mu (\Varid{a},\Varid{b},\Varid{c})\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{lazyzip3}\;\Varid{as}\;\Varid{bs}\;\Varid{cs}\mathrel{=}\Varid{zip3}\;\Varid{as}\;(\Varid{lazy}\;\Varid{bs})\;(\Varid{lazy}\;\Varid{cs}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{threetotwo}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Varid{a},\Varid{b},\Varid{c})\;(\Varid{a},(\Varid{b},\Varid{c})){}\<[E]%
+\\
+\>[B]{}\Varid{threetotwo}\mathrel{=}\Varid{arr}\mathbin{\$}\lambda \mathord{\sim}(\Varid{a},\Varid{b},\Varid{c})\to (\Varid{a},(\Varid{b},\Varid{c})){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of the torus skeleton}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-nodefunction :: Int ->
-	((Matrix, Matrix), [Matrix], [Matrix]) ->
-	([Matrix], [Matrix], [Matrix])
-nodefunction n ((bA, bB), rows, cols) =
-	([bSum], bA:nextAs , bB:nextBs)
-		where bSum =
-			foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
-			nextAs = take (n-1) rows
-			nextBs = take (n-1) cols
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{nodefunction}\mathbin{::}\Conid{Int}\to {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}((\Conid{Matrix},\Conid{Matrix}),[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu])\to {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}([\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu]){}\<[E]%
+\\
+\>[B]{}\Varid{nodefunction}\;\Varid{n}\;((\Varid{bA},\Varid{bB}),\Varid{rows},\Varid{cols})\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}([\mskip1.5mu \Varid{bSum}\mskip1.5mu],\Varid{bA}\mathbin{:}\Varid{nextAs},\Varid{bB}\mathbin{:}\Varid{nextBs}){}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}\mathbf{where}\;\Varid{bSum}\mathrel{=}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{foldl'}\;\Varid{matAdd}\;(\Varid{matMult}\;\Varid{bA}\;\Varid{bB})\;(\Varid{zipWith}\;\Varid{matMult}\;\Varid{nextAs}\;\Varid{nextBs}){}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{nextAs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{rows}{}\<[E]%
+\\
+\>[17]{}\hsindent{8}{}\<[25]%
+\>[25]{}\Varid{nextBs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{cols}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Adapted nodefunction for matrix multiplication with the torus from Fig.~\ref{fig:torus}}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
-mapArr f =
-	arr listcase >>>
-	arr (const []) pipepipepipe (f *** mapArr f >>> arr (uncurry (:)))
-	where listcase [] = Left ()
-	      listcase (x:xs) = Right (x,xs)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{mapArr}\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{mapArr}\;\Varid{f}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\;\Varid{pipepipepipe}\;(\Varid{f}\mathbin{***}\Varid{mapArr}\;\Varid{f}\mathbin{>\!>\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\mathbf{where}\;\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Conid{Left}\;(){}\<[E]%
+\\
+\>[9]{}\hsindent{6}{}\<[15]%
+\>[15]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{\inlinecode{map} over arrows}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
-foldlArr f b =
-	arr listcase >>>
-	arr (const b) pipepipepipe
-		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
-	where listcase [] = Left []
-	      listcase (x:xs) = Right (x,xs)
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{foldlArr}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Varid{b},\Varid{a})\;\Varid{b}\to \Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}{}\<[E]%
+\\
+\>[B]{}\Varid{foldlArr}\;\Varid{f}\;\Varid{b}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!>\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{const}\;\Varid{b})\;\Varid{pipepipepipe}\;{}\<[E]%
+\\
+\>[9]{}\hsindent{8}{}\<[17]%
+\>[17]{}(\Varid{first}\;(\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{b},\Varid{a}))\mathbin{>\!>\!>}\Varid{f}\mathbin{>\!>\!>}\Varid{arr}\;(\Varid{foldlArr}\;\Varid{f}))\mathbin{>\!>\!>}\Varid{app}){}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\mathbf{where}\;\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Conid{Left}\;[\mskip1.5mu \mskip1.5mu]{}\<[E]%
+\\
+\>[9]{}\hsindent{6}{}\<[15]%
+\>[15]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{\inlinecode{foldl} over arrows}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
-zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{zipWithArr}\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\Rightarrow \Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}\to \Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{zipWithArr}\;\Varid{f}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda (\Varid{as},\Varid{bs})\to \Varid{zipWith}\;(,)\;\Varid{as}\;\Varid{bs})\mathbin{>\!>\!>}\Varid{mapArr}\;\Varid{f}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{\inlinecode{zipWith} over arrows}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(\text{\tt pipepipepipe}) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathord{@}\Varid{pipepipepipe}\mathord{@})\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Type signature of \inlinecode{(\textpipe\textpipe\textpipe)}}
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
-listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
-\end{lstlisting}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{listApp}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{listApp}\;\Varid{fs}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!>\!>}\Varid{zipWithArr}\;\Varid{app}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
 \caption{Definition of \inlinecode{listApp}}

### 2017-05-18 paper/content/main.tex
it compiles

@@
 %%%%%format <--        = "\in "
-%fornat ` = "\`"
+%%%% %format ` = "\`"
+%%%% %% works fine without
 
@@
 
+%% fight for pipepipepipe
+\newcommand{\pipepipepipe}{\ensuremath{\mathbin{\mid\!\mid\!\mid}}\xspace}
+%%% ^^^^ keep consistent with definition at the top of main.lhs
+%%%\newcommand{\pipepipepipe}{\inlinecode{|||}\xspace}
+
 %%% JFP requirements:
@@
 \centering
-\subfloat[Arrow class definition][]{%
+%\subfloat[Arrow class definition][]{%
 % begin{minipage}{0.5\textwidth}
+\parbox{0.49\linewidth}{%
 \begin{hscode}\SaveRestoreHook
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}(\mathbin{>\!>\!>})\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
+\>[3]{}(\mathbin{>\!\!>\!\!>})\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-% caption{Arrow class definition}
-}%\hfill
-\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
-	\begin{center}
+\caption{Arrow class definition}
+}\qquad
+\begin{minipage}{0.49\linewidth}%
+%\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
+%	\begin{center}
+\centering
 	{\includegraphics[scale=0.6]{images/arr}}
@@
 	{\includegraphics[scale=0.6]{images/first}}
-	\end{center}
-}%\hfill
-%\caption{The arrow type class definition on the left with schematic depiction of its combinators on the right}
-\end{figure}
-\lstinline{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \lstinline{return}.
-The \lstinline{>>>} operator is analogous to the monadic composition  \lstinline{>>=} and combines two arrows \inlinecode{arr a b} and \inlinecode{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \inlinecode{arr a c}.
-Lastly, the \lstinline{first} operator  takes the input arrow from \lstinline{b} to \lstinline{c} and converts it into an arrow on pairs with the second argument untouched.
+%	\end{center}
+%}%\hfill
+\caption{The schematic depiction of \inlinecode{Arrow} combinators}
+\end{minipage}
+\end{figure}
+\inlinecode{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \inlinecode{return}.
+The \inlinecode{>>>} operator is analogous to the monadic composition  \inlinecode{>>=} and combines two arrows \inlinecode{arr a b} and \inlinecode{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \inlinecode{arr a c}.
+Lastly, the \inlinecode{first} operator  takes the input arrow from \inlinecode{b} to \inlinecode{c} and converts it into an arrow on pairs with the second argument untouched.
 It allows us to to save input across arrows.
 \\\\
-The most prominent instances of this interface are regular functions \lstinline{(->)} (Fig.~\ref{fig:arrowfn}),
+The most prominent instances of this interface are regular functions \inlinecode{(->)} (Fig.~\ref{fig:arrowfn}),
 \begin{figure}[h]
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{f}\mathbin{>\!>\!>}\Varid{g}\mathrel{=}\Varid{g}\mathbin{\circ}\Varid{f}{}\<[E]%
+\>[9]{}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{g}\mathrel{=}\Varid{g}\mathbin{\circ}\Varid{f}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{f}\mathbin{>\!>\!>}\Varid{g}\mathrel{=}\Conid{Kleisli}\;(\lambda \Varid{a}\to \Varid{f}\;\Varid{a}\bind \Varid{g}){}\<[E]%
+\>[9]{}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{g}\mathrel{=}\Conid{Kleisli}\;(\lambda \Varid{a}\to \Varid{f}\;\Varid{a}\bind \Varid{g}){}\<[E]%
 \\
@@
 	\begin{tabular}{cc}
-		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
-		\subcaptionbox{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
-		\subcaptionbox{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
-		\subcaptionbox{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
+		% \subcaptionbox
+{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
+		% \subcaptionbox
+{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
+		% \subcaptionbox
+{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+		% \subcaptionbox
+{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
 	\end{tabular}
@@
 \\
-\>[B]{}\Varid{second}\;\Varid{f}\mathrel{=}\Varid{arr}\;\Varid{swap}\mathbin{>\!>\!>}\Varid{first}\;\Varid{f}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{swap}{}\<[E]%
+\>[B]{}\Varid{second}\;\Varid{f}\mathrel{=}\Varid{arr}\;\Varid{swap}\mathbin{>\!\!>\!\!>}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{swap}{}\<[E]%
 \\
@@
 \end{figure}
-the \inlinecode{***} combinator which combines \includegraphics{first} and \inlinecode{second} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{***})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
-\\
-\>[B]{}\Varid{f}\mathbin{***}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!>\!>}\Varid{second}\;\Varid{g}{}\<[E]%
+the \inlinecode{***} combinator which combines \inlinecode{first} and \inlinecode{second} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{\*\!\*\!\*})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
+\\
+\>[B]{}\Varid{f}\mathbin{\*\!\*\!\*}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{g}{}\<[E]%
 \ColumnHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{\&\&\&})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{a}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
-\\
-\>[B]{}\Varid{f}\mathbin{\&\&\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!>\!>}(\Varid{f}\mathbin{***}\Varid{g}){}\<[E]%
+\>[B]{}(\mathbin{\&\!\&\!\&})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{a}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
+\\
+\>[B]{}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}(\Varid{f}\mathbin{\*\!\*\!\*}\Varid{g}){}\<[E]%
 \ColumnHook
@@
 \end{figure}
-A short example given by Hughes on how to use this is \lstinline{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
+A short example given by Hughes on how to use this is \inlinecode{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
 \begin{figure}[h]
@@
 \\
-\>[B]{}\Varid{add}\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{f}\mathbin{\&\&\&}\Varid{g})\mathbin{>\!>\!>}\Varid{arr}\;(\lambda (\Varid{u},\Varid{v})\to \Varid{u}\mathbin{+}\Varid{v}){}\<[E]%
+\>[B]{}\Varid{add}\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g})\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\lambda (\Varid{u},\Varid{v})\to \Varid{u}\mathbin{+}\Varid{v}){}\<[E]%
 \ColumnHook
@@
 \end{figure}
-% The benefit of using the \inlinecode{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
+% The benefit of using the \inlinecode{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \inlinecode{add} combinator.
 Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
@@
 As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \inlinecode{a -> b} in parallel, \inlinecode{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
-Translating this into arrow terms gives us a new operator \inlinecode{parEvalN} that lifts a list of arrows \inlinecode{[arr a b]} to a parallel arrow \inlinecode{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \lstinline{listApp} from Appendix \ref{utilfns}, but does parallel instead of serial evaluation).
+Translating this into arrow terms gives us a new operator \inlinecode{parEvalN} that lifts a list of arrows \inlinecode{[arr a b]} to a parallel arrow \inlinecode{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \inlinecode{listApp} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation).
 \begin{figure}[h]
@@
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!>\!>}{}\<[E]%
+\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{rdeepseq}))\mathbin{\&\&\&}\Varid{arr}\;\Varid{id}\mathbin{>\!>\!>}{}\<[E]%
+\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{rdeepseq}))\mathbin{\&\!\&\!\&}\Varid{arr}\;\Varid{id}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!>\!>}{}\<[E]%
+\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{strat}))\mathbin{\&\&\&}\Varid{arr}\;\Varid{id}\mathbin{>\!>\!>}{}\<[E]%
+\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{strat}))\mathbin{\&\!\&\!\&}\Varid{arr}\;\Varid{id}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!>\!>}{}\<[E]%
+\>[25]{}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
 \>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{zipWithArr}\;(\Varid{app}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{spawnP})\mathbin{>\!>\!>}{}\<[E]%
+\>[25]{}\Varid{zipWithArr}\;(\Varid{app}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{spawnP})\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
 \>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!>\!>}{}\<[E]%
+\>[25]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
 \>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})\mathbin{>\!>\!>}{}\<[E]%
+\>[25]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{arr}\mathbin{\$}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{map}\;(\lambda (\Conid{Kleisli}\;\Varid{f})\to \Varid{f})\;\Varid{fs}))\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}(\Varid{arr}\mathbin{\$}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{map}\;(\lambda (\Conid{Kleisli}\;\Varid{f})\to \Varid{f})\;\Varid{fs}))\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{listApp}\;\Varid{fchunks}\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{listApp}\;\Varid{fchunks}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Conid{Left}\mathbin{***}(\Varid{arr}\;\Conid{Right}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{return})\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{f}\mathbin{+++}\Varid{g}))\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{partitionEithers}\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{head}\mathbin{***}\Varid{arr}\;\Varid{head}{}\<[E]%
+\>[9]{}\Varid{arr}\;\Conid{Left}\mathbin{\*\!\*\!\*}(\Varid{arr}\;\Conid{Right}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{return})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{f}\mathbin{+++}\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{partitionEithers}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{head}\mathbin{\*\!\*\!\*}\Varid{arr}\;\Varid{head}{}\<[E]%
 \ColumnHook
@@
 For basic arrows, we have the \inlinecode{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into an arrow \inlinecode{arr (a, c) (b, d)} which does both computations at once.
-This can easily be translated into a parallel version \inlinecode{(\ensuremath{\mathbin{***}})} (Fig.~\ref{fig:par***}) with the use of \inlinecode{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \inlinecode{()} as the conf parameter in Fig.~\ref{fig:par***}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{|***|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;()))\Rightarrow {}\<[E]%
+This can easily be translated into a parallel version \inlinecode{(\ensuremath{\mathbin{\*\!\*\!\*}})} (Fig.~\ref{fig:par***}) with the use of \inlinecode{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \inlinecode{()} as the conf parameter in Fig.~\ref{fig:par***}).
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{|\!\*\!\*\!\*\!|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;()))\Rightarrow {}\<[E]%
 \\
@@
 \\
-\>[B]{}(\mathbin{|***|})\mathrel{=}\Varid{parEval2}\;(){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of (\ensuremath{\mathbin{***}}) - the parallel version of (***)}
+\>[B]{}(\mathbin{|\!\*\!\*\!\*\!|})\mathrel{=}\Varid{parEval2}\;(){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Definition of (\ensuremath{\mathbin{\*\!\*\!\*}}) - the parallel version of (***)}
 \label{fig:par***}
@@
 \\
-\>[B]{}(\mathbin{|\&\&\&|})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!>\!>}\Varid{f}\mathbin{|***|}\Varid{g}{}\<[E]%
+\>[B]{}(\mathbin{|\&\&\&|})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{|\!\*\!\*\!\*\!|}\Varid{g}{}\<[E]%
 \ColumnHook
@@
 \\
-\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}\Varid{parEvalN}\;()\;\Varid{fs1}\mathbin{>\!>\!>}\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{parEvalN}\;()\;\Varid{fs2}{}\<[E]%
+\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}\Varid{parEvalN}\;()\;\Varid{fs1}\mathbin{>\!\!>\!\!>}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{parEvalN}\;()\;\Varid{fs2}{}\<[E]%
 \ColumnHook
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{rd}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{fetch}{}\<[E]%
+\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{rd}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{fetch}{}\<[E]%
 \ColumnHook
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{takeMVar}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{unsafePerformIO}{}\<[E]%
+\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{takeMVar}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unsafePerformIO}{}\<[E]%
 \ColumnHook
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\mathbin{>\!>\!>}\Varid{put})\;\Varid{fs1})\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{rightRotate}\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\Varid{get}\mathbin{>\!>\!>})\;\Varid{fs2}){}\<[E]%
+\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{put})\;\Varid{fs1})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\Varid{get}\mathbin{>\!\!>\!\!>})\;\Varid{fs2}){}\<[E]%
 \ColumnHook
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{mapArr}\;\Varid{mapfn}\mathbin{>\!>\!>}\Varid{foldlArr}\;\Varid{foldfn}\;\Varid{neutral})\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{mapArr}\;\Varid{mapfn}\mathbin{>\!\!>\!\!>}\Varid{foldlArr}\;\Varid{foldfn}\;\Varid{neutral})\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\&\&}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
+\>[B]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
 \ColumnHook
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\&\&}{}\<[E]%
+\>[9]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\!\&\!\&}{}\<[E]%
 \\
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:})\mathbin{>\!>\!>}\Varid{lazy})\mathbin{>\!>\!>}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}))\mathbin{>\!>\!>}{}\<[E]%
+\>[17]{}(\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:})\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}))\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{arr}\;\Varid{return}\mathbin{\&\&\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu]))\mathbin{\&\&\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{pipe}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{unify}\;\Varid{f}\;\Varid{g}))\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{snd}\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}(\Varid{arr}\;\Varid{return}\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu]))\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{pipe}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{unify}\;\Varid{f}\;\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{snd}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[25]{}\hsindent{8}{}\<[33]%
-\>[33]{}(\Varid{mapArr}\;\Varid{f}\mathbin{***}\Varid{mapArr}\;\Varid{g})\mathbin{***}\Varid{arr}\;(\mathbin{\char92 \char95 }\to [\mskip1.5mu \mskip1.5mu])\mathbin{>\!>\!>}{}\<[E]%
+\>[33]{}(\Varid{mapArr}\;\Varid{f}\mathbin{\*\!\*\!\*}\Varid{mapArr}\;\Varid{g})\mathbin{\*\!\*\!\*}\Varid{arr}\;(\mathbin{\char92 \char95 }\to [\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 Note that extensive use of this combinator over \inlinecode{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \inlinecode{(\ensuremath{\mathbin{>\!>\!>}})} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{|\!>\!>\!>\!|})\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
+Nonetheless, we can define a parallel piping operator \inlinecode{(\ensuremath{\mathbin{>\!\!>\!\!>}})} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{|\!\!>\!\!>\!\!>\!\!|})\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
 \\
@@
 \\
-\>[B]{}(\mathbin{|\!>\!>\!>\!|})\mathrel{=}\Varid{pipe2}\;(){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of \texttt{(\ensuremath{\mathbin{>\!>\!>}})}}
+\>[B]{}(\mathbin{|\!\!>\!\!>\!\!>\!\!|})\mathrel{=}\Varid{pipe2}\;(){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Definition of \texttt{(\ensuremath{\mathbin{>\!\!>\!\!>}})}}
 \label{fig:par>>>}
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{lazy})\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}\Varid{loop}\;(\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!>\!>}{}\<[E]%
+\>[5]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!>\!>}\Varid{f}\mathbin{>\!>\!>}\Varid{second}\;\Varid{put})\mathbin{>\!>\!>}{}\<[E]%
+\>[5]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{put})\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{lazy})\mathbin{***}{}\<[E]%
+\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{\*\!\*\!\*}{}\<[E]%
 \\
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!>\!>}\Varid{lazy}))\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{uncurry3}\;(\Varid{zipWith3}\;\Varid{lazyzip3}))\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{arr}\;\Varid{length}\mathbin{>\!>\!>}\Varid{arr}\;\Varid{unshuffle})\mathbin{\&\&\&}{}\<[E]%
+\>[17]{}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{uncurry3}\;(\Varid{zipWith3}\;\Varid{lazyzip3}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}(\Varid{arr}\;\Varid{length}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unshuffle})\mathbin{\&\!\&\!\&}{}\<[E]%
 \\
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{shuffle}\mathbin{>\!>\!>}\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f}))\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{app}\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!>\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!>\!>}\Varid{threetotwo}){}\<[E]%
+\>[17]{}(\Varid{shuffle}\mathbin{>\!\!>\!\!>}\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{app}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!\!>\!\!>}\Varid{threetotwo}){}\<[E]%
 \\[\blanklineskip]%
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{c},\Varid{a},\Varid{b})\to (\Varid{c},\Varid{get}\;\Varid{a},\Varid{get}\;\Varid{b}))\mathbin{>\!>\!>}\Varid{f}\mathbin{>\!>\!>}{}\<[E]%
+\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{c},\Varid{a},\Varid{b})\to (\Varid{c},\Varid{get}\;\Varid{a},\Varid{get}\;\Varid{b}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\;\Varid{pipepipepipe}\;(\Varid{f}\mathbin{***}\Varid{mapArr}\;\Varid{f}\mathbin{>\!>\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
+\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{\mid\!\mid\!\mid}(\Varid{f}\mathbin{\*\!\*\!\*}\Varid{mapArr}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!>\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{const}\;\Varid{b})\;\Varid{pipepipepipe}\;{}\<[E]%
+\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{const}\;\Varid{b})\mathbin{\mid\!\mid\!\mid}{}\<[E]%
 \\
 \>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{first}\;(\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{b},\Varid{a}))\mathbin{>\!>\!>}\Varid{f}\mathbin{>\!>\!>}\Varid{arr}\;(\Varid{foldlArr}\;\Varid{f}))\mathbin{>\!>\!>}\Varid{app}){}\<[E]%
+\>[17]{}(\Varid{first}\;(\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{b},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{foldlArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}\Varid{app}){}\<[E]%
 \\
@@
 \fixme{pipepipepipe does not work with lhs2TeX}
-Finally, with the help of \inlinecode{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \inlinecode{arr (a, b) c} to an arrow \inlinecode{arr ([a], [b]) [c]}.
+Finally, with the help of \inlinecode{mapArr} (Fig.~\ref{fig:mapArr}), we can define \inlinecode{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \inlinecode{arr (a, b) c} to an arrow \inlinecode{arr ([a], [b]) [c]}.
 \begin{figure}[h]
@@
 \\
-\>[B]{}\Varid{zipWithArr}\;\Varid{f}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda (\Varid{as},\Varid{bs})\to \Varid{zipWith}\;(,)\;\Varid{as}\;\Varid{bs})\mathbin{>\!>\!>}\Varid{mapArr}\;\Varid{f}{}\<[E]%
+\>[B]{}\Varid{zipWithArr}\;\Varid{f}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda (\Varid{as},\Varid{bs})\to \Varid{zipWith}\;(,)\;\Varid{as}\;\Varid{bs})\mathbin{>\!\!>\!\!>}\Varid{mapArr}\;\Varid{f}{}\<[E]%
 \ColumnHook
@@
  %$ %% formatting
-These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \inlinecode{(\textpipe\textpipe\textpipe)} combinator.
+These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \pipepipepipe\olcomment{CHECK!} combinator.
 It takes two arrows \inlinecode{arr a c} and \inlinecode{arr b c} and combines them into a new arrow \inlinecode{arr (Either a b) c} which pipes all \inlinecode{Left a}'s to the first arrow and all \inlinecode{Right b}'s to the second arrow.
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathord{@}\Varid{pipepipepipe}\mathord{@})\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Type signature of \inlinecode{(\textpipe\textpipe\textpipe)}}
+\>[B]{}(\mathbin{\mid\!\mid\!\mid})\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Type signature of \pipepipepipe}
 \label{fig:codeSigPipePipePipe}
 \end{figure}
-\fixme{pipepipepipe does not work with lhs2TeX}
+\fixme{pipepipepipe does not work with lhs2TeX}\olcomment{trying
+  direct one}
 
@@
 \\
-\>[B]{}\Varid{listApp}\;\Varid{fs}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!>\!>}\Varid{zipWithArr}\;\Varid{app}{}\<[E]%
+\>[B]{}\Varid{listApp}\;\Varid{fs}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!\!>\!\!>}\Varid{zipWithArr}\;\Varid{app}{}\<[E]%
 \ColumnHook

### 2017-05-18 paper/content/main.tex
this is how inline works with lhs2tex

@@
 
-\newcommand{\inlinecode}[1]{\texttt{#1}}
+%\newcommand{\inlinecode}[1]{\texttt{#1}}
+\newcommand{\inlinecode}[1]{\emph{#1}} %%% booo
+%%% |code| is inline code in lhs2tex
 
@@
 	\section{Utility Functions}\label{utilfns}
-To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \inlinecode{map}, \inlinecode{foldl} and \inlinecode{zipWith} on arrows.
-
-The \inlinecode{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \inlinecode{arr a b} to an arrow \inlinecode{arr [a] [b]} \cite{programming_with_arrows}:
+To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \ensuremath{\Varid{map}}, \ensuremath{\Varid{foldl}} and \ensuremath{\Varid{zipWith}} on arrows.
+
+The \ensuremath{\Varid{mapArr}} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} \cite{programming_with_arrows}:
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{\inlinecode{map} over arrows}
+\caption{\ensuremath{\Varid{map}} over arrows}
 \label{fig:mapArr}
 \end{figure}
-Similarly, we can also define \inlinecode{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \inlinecode{arr (b, a) b} with a neutral element \inlinecode{b} to \inlinecode{arr [a] b}:
+Similarly, we can also define \ensuremath{\Varid{foldlArr}} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{b},\Varid{a})\;\Varid{b}} with a neutral element \ensuremath{\Varid{b}} to \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}}:
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{\inlinecode{foldl} over arrows}
+\caption{\ensuremath{\Varid{foldl}} over arrows}
 \label{fig:foldlArr}
@@
 \fixme{pipepipepipe does not work with lhs2TeX}
-Finally, with the help of \inlinecode{mapArr} (Fig.~\ref{fig:mapArr}), we can define \inlinecode{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \inlinecode{arr (a, b) c} to an arrow \inlinecode{arr ([a], [b]) [c]}.
+Finally, with the help of \ensuremath{\Varid{mapArr}} (Fig.~\ref{fig:mapArr}), we can define \ensuremath{\Varid{zipWithArr}} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}} to an arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]}.
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{\inlinecode{zipWith} over arrows}
+\caption{\ensuremath{\Varid{zipWith}} over arrows}
 \label{fig:zipWithArr}
@@
  %$ %% formatting
-These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \pipepipepipe\olcomment{CHECK!} combinator.
-It takes two arrows \inlinecode{arr a c} and \inlinecode{arr b c} and combines them into a new arrow \inlinecode{arr (Either a b) c} which pipes all \inlinecode{Left a}'s to the first arrow and all \inlinecode{Right b}'s to the second arrow.
+These combinators make use of the \ensuremath{\Conid{ArrowChoice}} type class which provides the \pipepipepipe\olcomment{CHECK!} combinator.
+It takes two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} and combines them into a new arrow \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}} which pipes all \ensuremath{\Conid{Left}\;\Varid{a}}'s to the first arrow and all \ensuremath{\Conid{Right}\;\Varid{b}}'s to the second arrow.
 \begin{figure}[h]
@@
 
-With the zipWithArr combinator we can also write a combinator \inlinecode{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \inlinecode{[arr a b]} to an arrow \inlinecode{arr [a] [b]}.
+With the zipWithArr combinator we can also write a combinator \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{Definition of \inlinecode{listApp}}
+\caption{Definition of \ensuremath{\Varid{listApp}}}
 \label{fig:listApp}
@@
 % $ %% formatting
-. Note that  this additionally makes use of the \inlinecode{ArrowApply} typeclass that allows us to evaluate arrows with \inlinecode{app :: arr (arr a b, a) c}.
+.
+Note that  this additionally makes use of the \ensuremath{\Conid{ArrowApply}} typeclass that allows us to evaluate arrows with \ensuremath{\Varid{app}\mathbin{::}\Varid{arr}\;(\Varid{arr}\;\Varid{a}\;\Varid{b},\Varid{a})\;\Varid{c}}.
 

### 2017-05-18 paper/content/main.tex
make it compile

@@
 
-\usepackage{placeins}
+% \usepackage{placeins} %% unneeded
 
@@
 Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication.
-Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \inlinecode{map} with and without load balancing, \inlinecode{map-reduce}) in our newly defined dialect.
-More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\inlinecode{pipe}, \inlinecode{ring}, \inlinecode{torus}).
+Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \ensuremath{\Varid{map}} with and without load balancing, \ensuremath{\Varid{map}\mathbin{-}\Varid{reduce}}) in our newly defined dialect.
+More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\ensuremath{\Varid{pipe}}, \ensuremath{\Varid{ring}}, \ensuremath{\Varid{torus}}).
 Section~\ref{sec:benchmarks} shows the benchmark results.
@@
 
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \inlinecode{a -> b} in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:
 
@@
 
-Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \inlinecode{parEvalN} with several different parallel Haskells.
+Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \ensuremath{\Varid{parEvalN}} with several different parallel Haskells.
 
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \inlinecode{using :: a -> Strategy a -> a}.
-For \inlinecode{parEvalN} this means that we can just apply the list of functions \inlinecode{[a -> b]} to the list of inputs \inlinecode{[a]} by zipping them with the application operator \inlinecode{\$}.
-We then evaluate this lazy list \inlinecode{[b]} according to a \inlinecode{Strategy [b]} with the \inlinecode{using :: a -> Strategy a -> a} operator.
-We construct this strategy with \inlinecode{parList :: Strategy a -> Strategy [a]} and \inlinecode{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
-To ensure that programs that use \inlinecode{parEvalN} have the correct evaluation order, we annotate the computation with \inlinecode{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \inlinecode{parEvalN} computations.
-This is particularly necessary in circular communication topologies like in the \inlinecode{torus} or \inlinecode{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \inlinecode{pseq} during testing for this paper.
+Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}.
+For \ensuremath{\Varid{parEvalN}} this means that we can just apply the list of functions \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} to the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} by zipping them with the application operator \ensuremath{\mathbin{\$}}.
+% $
+We then evaluate this lazy list \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} according to a \ensuremath{\Conid{Strategy}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} with the \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} operator.
+We construct this strategy with \ensuremath{\Varid{parList}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Conid{Strategy}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{\Varid{rdeepseq}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{Strategy}\;\Varid{a}} where the latter is a strategy which evalutes to normal form.
+To ensure that programs that use \ensuremath{\Varid{parEvalN}} have the correct evaluation order, we annotate the computation with \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} which forces the compiler to not reorder multiple \ensuremath{\Varid{parEvalN}} computations.
+This is particularly necessary in circular communication topologies like in the \ensuremath{\Varid{torus}} or \ensuremath{\Varid{ring}} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \ensuremath{\Varid{pseq}} during testing for this paper.
 
@@
 \subsubsection{ParMonad}
-The \inlinecode{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
-
-Our parallel evaluation function \inlinecode{parEvalN} can be defined by zipping the list of \inlinecode{[a -> b]} with the list of inputs \inlinecode{[a]} with the application operator \inlinecode{\$} just like with Multicore Haskell.
-Then, we map over this not yet evaluated lazy list of results \inlinecode{[b]} with \inlinecode{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \inlinecode{[Par (IVar b)]}, which we convert to \inlinecode{Par [IVar b]} with \inlinecode{sequenceA}.
-We wait for the computations to finish by mapping over the \inlinecode{IVar b}'s inside the \inlinecode{Par} monad with \inlinecode{get}.
-This results in \inlinecode{Par [b]}.
-We finally execute this process with \inlinecode{runPar} to finally get \inlinecode{[b]} again.
-
-\textbf{\textcolor{red}{explain problems with laziness here. Problems with torus}}
+The \ensuremath{\Conid{Par}} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
+
+Our parallel evaluation function \ensuremath{\Varid{parEvalN}} can be defined by zipping the list of \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} with the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} with the application operator \ensuremath{\mathbin{\$}} just like with Multicore Haskell.
+% $
+Then, we map over this not yet evaluated lazy list of results \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} to transform them to a list of not yet evaluated forked away computations \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we convert to \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{sequenceA}}.
+We wait for the computations to finish by mapping over the \ensuremath{\Conid{IVar}\;\Varid{b}} values inside the \ensuremath{\Conid{Par}} monad with \ensuremath{\Varid{get}}.
+This results in \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
+We finally execute this process with \ensuremath{\Varid{runPar}} to finally get \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} again.
+
+\mbcomment{explain problems with laziness here. Problems with torus}
 
@@
 
-While Eden also comes with a monad \inlinecode{PA} for parallel evaluation, it also ships with a completely functional interface that includes
+While Eden also comes with a monad \ensuremath{\Conid{PA}} for parallel evaluation, it also ships with a completely functional interface that includes
 %\\
-%\inlinecode{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
+a \ensuremath{\Varid{spawnF}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}
 %\\
-a \inlinecode{spawnF} function that
+%a |spawnF| 
+function that
 %This 
-allows us to define \inlinecode{parEvalN} directly:
+allows us to define \ensuremath{\Varid{parEvalN}} directly:
 
@@
 Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
-An arrow \inlinecode{arr a b} represents  a computation that converts an input \inlinecode{a} to an output \inlinecode{b}.
+An arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} represents  a computation that converts an input \ensuremath{\Varid{a}} to an output \ensuremath{\Varid{b}}.
 This is defined in the arrow typeclass:
@@
 %}%\hfill
-\caption{The schematic depiction of \inlinecode{Arrow} combinators}
+\caption{The schematic depiction of \ensuremath{\Conid{Arrow}} combinators}
 \end{minipage}
 \end{figure}
-\inlinecode{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \inlinecode{return}.
-The \inlinecode{>>>} operator is analogous to the monadic composition  \inlinecode{>>=} and combines two arrows \inlinecode{arr a b} and \inlinecode{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \inlinecode{arr a c}.
-Lastly, the \inlinecode{first} operator  takes the input arrow from \inlinecode{b} to \inlinecode{c} and converts it into an arrow on pairs with the second argument untouched.
+\ensuremath{\Varid{arr}} is used to lift an ordinary function to an arrow type, similarly to the monadic \ensuremath{\Varid{return}}.
+The \ensuremath{\mathbin{>\!\!>\!\!>}} operator is analogous to the monadic composition  \ensuremath{\bind } and combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}}.
+Lastly, the \ensuremath{\Varid{first}} operator  takes the input arrow from \ensuremath{\Varid{b}} to \ensuremath{\Varid{c}} and converts it into an arrow on pairs with the second argument untouched.
 It allows us to to save input across arrows.
 \\\\
-The most prominent instances of this interface are regular functions \inlinecode{(->)} (Fig.~\ref{fig:arrowfn}),
+The most prominent instances of this interface are regular functions \ensuremath{(\to )} (Fig.~\ref{fig:arrowfn}),
 \begin{figure}[h]
@@
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \inlinecode{first}, called \inlinecode{second} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
+With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \ensuremath{\Varid{first}}, called \ensuremath{\Varid{second}} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
 \begin{figure}[h]
@@
 \end{figure}
-the \inlinecode{***} combinator which combines \inlinecode{first} and \inlinecode{second} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
+the \ensuremath{\mathbin{\*\!\*\!\*}} combinator which combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
 \begin{figure}[h]
@@
 \end{figure}
-and the \inlinecode{\&\&\&} combinator that constructs an arrow which outputs two different values like \inlinecode{***}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+and the \ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{\*\!\*\!\*}}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
 \begin{figure}[h]
@@
 \end{figure}
-A short example given by Hughes on how to use this is \inlinecode{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
+A short example given by Hughes on how to use this is \ensuremath{\Varid{add}} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
 \begin{figure}[h]
@@
 \end{figure}
-% The benefit of using the \inlinecode{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \inlinecode{add} combinator.
+% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator.
 Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
@@
 % \\\\
-% \textit{Note: In the definitions Hughes gave in his paper, the notation \inlinecode{a b c} for an arrow from \inlinecode{b} to \inlinecode{c} is used.
-We use the equivalent definition \inlinecode{arr a b} for an arrow from \inlinecode{a} to \inlinecode{b} instead, to make it easier to find the arrow type in type signatures.}
+% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used.
+We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}
 %
@@
 \subsection{Parallel Haskells}
-Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \inlinecode{Par} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
+Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
 We use these languages as backends: our library can switch from one to other at user's command.
 
-HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \inlinecode{Par} monad to heterogeneous clusters.
-LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \inlinecode{Par} monad.
+HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} monad to heterogeneous clusters.
+LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \ensuremath{\Conid{Par}} monad.
 %
@@
 Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden.
-This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,brown2010ever,dieterle2010skeleton,delaEncina2011,Dieterle2013}.
+This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.
 
@@
 Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
-The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,brown2010ever,lobachev-phd,Lobachev2012}, iteration skeletons \cite{Dieterle2013}.
+The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,lobachev-phd,Lobachev2012,janjic2013space}, iteration skeletons \cite{Dieterle2013}.
 The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel.
@@
 
-Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\inlinecode{->}.
+Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\ensuremath{\to }.
 \citet{Hughes2005} is a tutorial on arrows.
@@
 Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution.
-In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \inlinecode{parEvalN} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
+In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
 \citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach.
@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \inlinecode{a -> b} in parallel, \inlinecode{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
-Translating this into arrow terms gives us a new operator \inlinecode{parEvalN} that lifts a list of arrows \inlinecode{[arr a b]} to a parallel arrow \inlinecode{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \inlinecode{listApp} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation).
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, \ensuremath{\Varid{parEvalN}} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
+Translating this into arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation).
 \begin{figure}[h]
@@
 \end{figure}
-With this definition of \inlinecode{parEvalN}, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow \inlinecode{arr} and we want this to be an interface for different backends, we introduce a new typeclass \inlinecode{ArrowParallel arr a b} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+With this definition of \ensuremath{\Varid{parEvalN}}, parallel execution is yet another arrow combinator.
+But as the implementation may differ depending on the actual type of the arrow \ensuremath{\Varid{arr}} and we want this to be an interface for different backends, we introduce a new typeclass \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
 \begin{figure}[h]
@@
 normalform).
-For this reason we also introduce an additional \inlinecode{conf} parameter to the function.
-We also do not want \inlinecode{conf} to be a fixed type, as the configuration parameters can differ for different instances of \inlinecode{ArrowParallel}.
-So we add it to the type signature of the typeclass as well and get \inlinecode{ArrowParallel arr a b conf} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
+For this reason we also introduce an additional \ensuremath{\Varid{conf}} parameter to the function.
+We also do not want \ensuremath{\Varid{conf}} to be a fixed type, as the configuration parameters can differ for different instances of \ensuremath{\Conid{ArrowParallel}}.
+So we add it to the type signature of the typeclass as well and get \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
 \begin{figure}[h]
@@
 \end{figure}
-Note that we don't require the \inlinecode{conf} parameter in every implementation.
-If it is not needed, we usually just default the \inlinecode{conf} type parameter to () and even blank it out in the parameter list of the implemented \inlinecode{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
+Note that we don't require the \ensuremath{\Varid{conf}} parameter in every implementation.
+If it is not needed, we usually just default the \ensuremath{\Varid{conf}} type parameter to () and even blank it out in the parameter list of the implemented \ensuremath{\Varid{parEvalN}}, as we will see in the implementation of the Multicore and the ParMonad backend.
 
@@
 \subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \inlinecode{withStrategy :: Strategy a -> a -> a} and \inlinecode{pseq :: a -> b -> b} combinators from Multicore Haskell, where \inlinecode{withStrategy} is the same as \inlinecode{using :: a -> Strategy a -> a} but with flipped parameters.
+The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} and \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} combinators from Multicore Haskell, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} but with flipped parameters.
 \begin{figure}[h]
@@
 \end{figure}
-For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \inlinecode{Strategy} interface, we want to the user not to lose this ability because of using our API.
-We therefore introduce the \inlinecode{Conf a} data-type that simply wraps a \inlinecode{Strategy a} (Fig.~\ref{fig:confa}).
-We can't directly use the \inlinecode{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
+For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \ensuremath{\Conid{Strategy}} interface, we want to the user not to lose this ability because of using our API.
+We therefore introduce the \ensuremath{\Conid{Conf}\;\Varid{a}} data-type that simply wraps a \ensuremath{\Conid{Strategy}\;\Varid{a}} (Fig.~\ref{fig:confa}).
+We can't directly use the \ensuremath{\Conid{Strategy}\;\Varid{a}} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
 \begin{figure}[h]
@@
 \end{figure}
-To get our configurable \inlinecode{ArrowParallel} instance, we simply unwrap the strategy and pass it to \inlinecode{parList} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
+To get our configurable \ensuremath{\Conid{ArrowParallel}} instance, we simply unwrap the strategy and pass it to \ensuremath{\Varid{parList}} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
 \begin{figure}[h]
@@
 \subsubsection{Par Monad}
-The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \inlinecode{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function.
-The latter forks away the computation of a value and returns an \inlinecode{IVar} containing the result in the \inlinecode{Par} monad.
-
-We therefore apply each function to its corresponding input value with \inlinecode{listApp} (Fig.~\ref{fig:listApp}) and then fork the computation away with \inlinecode{arr spawnP} inside a \inlinecode{zipWithArr} call.
-This yields a list \inlinecode{[Par (IVar b)]}, which we then convert into \inlinecode{Par [IVar b]} with \inlinecode{arr sequenceA}.
-In order to wait for the computation to finish, we map over the \inlinecode{IVar}s inside the ParMonad with \inlinecode{arr (>>= mapM get)}.
-The result of this operation is a \inlinecode{Par [b]} from which we can finally remove the monad again by running \inlinecode{arr runPar} to get our output of \inlinecode{[b]}.
+The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} function.
+The latter forks away the computation of a value and returns an \ensuremath{\Conid{IVar}} containing the result in the \ensuremath{\Conid{Par}} monad.
+
+We therefore apply each function to its corresponding input value with \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}) and then fork the computation away with \ensuremath{\Varid{arr}\;\Varid{spawnP}} inside a \ensuremath{\Varid{zipWithArr}} call.
+This yields a list \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we then convert into \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}.
+In order to wait for the computation to finish, we map over the \ensuremath{\Conid{IVar}}s inside the ParMonad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}.
+The result of this operation is a \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} from which we can finally remove the monad again by running \ensuremath{\Varid{arr}\;\Varid{runPar}} to get our output of \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
 \begin{figure}[h]
@@
 \subsubsection{Eden}
-For the Multicore and ParMonad implementation we could use general instances of \inlinecode{ArrowParallel} that just require the \inlinecode{ArrowApply} and \inlinecode{ArrowChoice} typeclasses.
+For the Multicore and ParMonad implementation we could use general instances of \ensuremath{\Conid{ArrowParallel}} that just require the \ensuremath{\Conid{ArrowApply}} and \ensuremath{\Conid{ArrowChoice}} typeclasses.
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a typeclass \inlinecode{ArrowUnwrap} (Fig.~\ref{fig:ArrowUnwrap}).
+While we could still manage to have only one class in the module by introducing a typeclass \ensuremath{\Conid{ArrowUnwrap}} (Fig.~\ref{fig:ArrowUnwrap}).
 \begin{figure}[h]
@@
 \end{figure}
-We don't do it here for aesthetic resons, though. For now, we just implement \inlinecode{ArrowParallel} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
+We don't do it here for aesthetic resons, though. For now, we just implement \ensuremath{\Conid{ArrowParallel}} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
 \begin{figure}[h]
@@
 
-\FloatBarrier
+%\FloatBarrier
 
@@
 \olcomment{move this to Contributions in the front or something}
-We have seen that we can wrap parallel Haskells inside of the \inlinecode{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
+We have seen that we can wrap parallel Haskells inside of the \ensuremath{\Conid{ArrowParallel}} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
 \begin{itemize}
 	\item \textbf{Arrow API benefits}:
-	With the \inlinecode{ArrowParallel} typeclass we do not lose any benefits of using arrows as \inlinecode{parEvalN} is just yet another arrow combinator.
+	With the \ensuremath{\Conid{ArrowParallel}} typeclass we do not lose any benefits of using arrows as \ensuremath{\Varid{parEvalN}} is just yet another arrow combinator.
 The resulting arrow can be used in the same way a potential serial version could be used.
@@
 	\item \textbf{Abstraction}:
-	With the \inlinecode{ArrowParallel} typeclass, we abstracted all parallel implementation logic away from the business logic.
+	With the \ensuremath{\Conid{ArrowParallel}} typeclass, we abstracted all parallel implementation logic away from the business logic.
 This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
-So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \inlinecode{ArrowParallel} instance.
+So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \ensuremath{\Conid{ArrowParallel}} instance.
 \end{itemize}
@@
 \label{sec:extending-interface}
-With the \inlinecode{ArrowParallel} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
+With the \ensuremath{\Conid{ArrowParallel}} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
 These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
 
-\subsubsection{Lazy \inlinecode{parEvalN}}
+\subsubsection{Lazy \ensuremath{\Varid{parEvalN}}}
 \begin{figure}[h]
@@
 \end{figure}
-The function \inlinecode{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
+The function \ensuremath{\Varid{parEvalN}} is 100\% strict, which means that it fully evaluates all passed arrows.
 Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
-\inlinecode{map (arr .
-(+)) [1..]} or just because we need the arrows evaluated in chunks.
-\inlinecode{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
-It works by first chunking the input from \inlinecode{[a]} to \inlinecode{[[a]]} with the given \inlinecode{ChunkSize} in \inlinecode{arr (chunksOf chunkSize)}.
-These chunks are then fed into a list \inlinecode{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \inlinecode{ChunkSize} into the regular parEvalN by using \inlinecode{listApp} (Fig.~\ref{fig:listApp}).
-The resulting \inlinecode{[[b]]} is lastly converted into \inlinecode{[b]} with \inlinecode{arr concat}.
+\ensuremath{\Varid{map}\;(\Varid{arr}\mathbin{\circ}(\mathbin{+}))\;[\mskip1.5mu \mathrm{1}\mathinner{\ldotp\ldotp}\mskip1.5mu]} or just because we need the arrows evaluated in chunks.
+\ensuremath{\Varid{parEvalNLazy}} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
+It works by first chunking the input from \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} to \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]} with the given \ensuremath{\Conid{ChunkSize}} in \ensuremath{\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})}.
+These chunks are then fed into a list \ensuremath{[\mskip1.5mu \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} of parallel arrows created by feeding chunks of the passed \ensuremath{\Conid{ChunkSize}} into the regular parEvalN by using \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}).
+The resulting \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} is lastly converted into \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{concat}}.
 \begin{figure}[h]
@@
 But sometimes we want to paralellize heterogenous types as well.
-However, we can implement such a \inlinecode{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into a new parallel arrow \inlinecode{arr (a, c) (b, d)} quite easily with the help of the \inlinecode{ArrowChoice} typeclass.
-The idea is to use the \inlinecode{+++} combinator which combines two arrows \inlinecode{arr a b} and \inlinecode{arr c d} and transforms them into \inlinecode{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
+However, we can implement such a \ensuremath{\Varid{parEval2}} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into a new parallel arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} quite easily with the help of the \ensuremath{\Conid{ArrowChoice}} typeclass.
+The idea is to use the \ensuremath{\mathbin{+++}} combinator which combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} and transforms them into \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})} to get a common arrow type that we can then feed into parEvalN.
 \\\\
-We start by transforming the \inlinecode{(a, c)} input into a 2-element list \inlinecode{[Either a c]} by first tagging the two inputs with \inlinecode{Left} and \inlinecode{Right} and wrapping the right element in a singleton list with \inlinecode{return} so that we can combine them with \inlinecode{arr (uncurry (:))}.
-Next, we feed this list into a parallel arrow running on 2 instances of \inlinecode{f +++ g} as described above.
-After the calculation is finished, we convert the resulting \inlinecode{[Either b d]} into \inlinecode{([b], [d])} with \inlinecode{arr partitionEithers}.
-The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \inlinecode{(b, d)} in the last step.
+We start by transforming the \ensuremath{(\Varid{a},\Varid{c})} input into a 2-element list \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{a}\;\Varid{c}\mskip1.5mu]} by first tagging the two inputs with \ensuremath{\Conid{Left}} and \ensuremath{\Conid{Right}} and wrapping the right element in a singleton list with \ensuremath{\Varid{return}} so that we can combine them with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))}.
+Next, we feed this list into a parallel arrow running on 2 instances of \ensuremath{\Varid{f}\mathbin{+++}\Varid{g}} as described above.
+After the calculation is finished, we convert the resulting \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{b}\;\Varid{d}\mskip1.5mu]} into \ensuremath{([\mskip1.5mu \Varid{b}\mskip1.5mu],[\mskip1.5mu \Varid{d}\mskip1.5mu])} with \ensuremath{\Varid{arr}\;\Varid{partitionEithers}}.
+The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \ensuremath{(\Varid{b},\Varid{d})} in the last step.
 \begin{figure}[h]
@@
 	\subsubsection{Syntactic Sugar} \label{syntacticSugar}
-For basic arrows, we have the \inlinecode{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \inlinecode{arr a b} and \inlinecode{arr c d} into an arrow \inlinecode{arr (a, c) (b, d)} which does both computations at once.
-This can easily be translated into a parallel version \inlinecode{(\ensuremath{\mathbin{\*\!\*\!\*}})} (Fig.~\ref{fig:par***}) with the use of \inlinecode{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \inlinecode{()} as the conf parameter in Fig.~\ref{fig:par***}).
+For basic arrows, we have the \ensuremath{(\mathbin{\*\!\*\!\*})} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into an arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} which does both computations at once.
+This can easily be translated into a parallel version \ensuremath{\mathbin{|\!\*\!\*\!\*\!|}} (Fig.~\ref{fig:par***}) with the use of \ensuremath{\Varid{parEval2}}, but for this we require a backend which has an implementation that does not require any configuration (hence the \ensuremath{()} as the conf parameter in Fig.~\ref{fig:par***}).
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{Definition of (\ensuremath{\mathbin{\*\!\*\!\*}}) - the parallel version of (***)}
+\caption{Definition of \ensuremath{\mathbin{|\!\*\!\*\!\*\!|}}, the parallel version of \ensuremath{\mathbin{\*\!\*\!\*}}.}
 \label{fig:par***}
 \end{figure}
-% With this we can analogously to the serial \inlinecode{&&&}
-We define the parallel \inlinecode{(\ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}})} (Fig.~\ref{fig:par&&&}) in a similar manner to its sequential pendant \inlinecode{(\&\&\&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{|\&\&\&|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{a})\;(\Conid{Either}\;\Varid{b}\;\Varid{c})\;())\Rightarrow {}\<[E]%
+% With this we can analogously to the serial |&&&|
+We define the parallel \ensuremath{\mathbin{|\!\&\!\&\!\&\!|}} (Fig.~\ref{fig:par&&&}) in a similar manner to its sequential pendant \ensuremath{\mathbin{\&\!\&\!\&}} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{|\!\&\!\&\!\&\!|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{a})\;(\Conid{Either}\;\Varid{b}\;\Varid{c})\;())\Rightarrow {}\<[E]%
 \\
@@
 \\
-\>[B]{}(\mathbin{|\&\&\&|})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{|\!\*\!\*\!\*\!|}\Varid{g}{}\<[E]%
+\>[B]{}(\mathbin{|\!\&\!\&\!\&\!|})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{|\!\*\!\*\!\*\!|}\Varid{g}{}\<[E]%
 \ColumnHook
@@
  % $ %% formatting
-\caption{Definition of (\ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}}) - the parallel version of (\&\&\&)}
+\caption{Definition of \ensuremath{\mathbin{|\!\&\!\&\!\&\!|}} - the parallel version of \ensuremath{\mathbin{\&\!\&\!\&}}.}
 \label{fig:par&&&}
@@
 \end{figure}
-In a distributed environment, the resulting arrow of this combinator first evaluates all \inlinecode{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \inlinecode{[arr b c]} in parallel to then gather the input once again on the master node.
+In a distributed environment, the resulting arrow of this combinator first evaluates all \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]} in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
 
-While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \inlinecode{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \inlinecode{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \ensuremath{\Varid{parEvalN}} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \ensuremath{\Conid{ArrowParallel}} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
 This can become a serious bottleneck %in heavy threaded applications.
@@
 This motivates for an approach that allows the nodes to communicate directly with each other.
-Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \inlinecode{RD} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
-
-But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \inlinecode{Future} typeclass (Fig.~\ref{fig:future}).
+Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \ensuremath{\Conid{RD}} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
+
+But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \ensuremath{\Conid{Future}} typeclass (Fig.~\ref{fig:future}).
 \begin{figure}[h]
@@
 \end{figure}
-Since \inlinecode{RD} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \inlinecode{Strategy a} in the Multicore Haskell implementation of \inlinecode{ArrowParallel} in chapter \ref{sec:parrows:multicore}).
+Since \ensuremath{\Conid{RD}} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \ensuremath{\Conid{Strategy}\;\Varid{a}} in the Multicore Haskell implementation of \ensuremath{\Conid{ArrowParallel}} in chapter \ref{sec:parrows:multicore}).
 \begin{figure}[h]
@@
 
-For our Par Monad and Multicore Haskell backends, we can simply use \inlinecode{MVar}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
+For our Par Monad and Multicore Haskell backends, we can simply use \ensuremath{\Conid{MVar}}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
 \fixme{explain MVars}
@@
 \label{fig:MVarFuture}
-\end{figure}
-
-Furthermore, in order for these \inlinecode{Future} types to fit with the \inlinecode{ArrowParallel} instances we gave earlier, we have to give the necessary \inlinecode{NFData} and \inlinecode{Trans} instances - the latter only being needed in Eden.
-Because \inlinecode{MVar} already ships with a \inlinecode{NFData} instance, we only have to supply two simple instances for our \inlinecode{RemoteData} type.
+\end{figure} % $
+
+Furthermore, in order for these \ensuremath{\Conid{Future}} types to fit with the \ensuremath{\Conid{ArrowParallel}} instances we gave earlier, we have to give the necessary \ensuremath{\Conid{NFData}} and \ensuremath{\Conid{Trans}} instances - the latter only being needed in Eden.
+Because \ensuremath{\Conid{MVar}} already ships with a \ensuremath{\Conid{NFData}} instance, we only have to supply two simple instances for our \ensuremath{\Conid{RemoteData}} type.
 \begin{figure}[h]
@@
 
-\FloatBarrier
+%%% \FloatBarrier
 \section{Map-based Skeletons}
@@
 \end{figure}
-\inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
-We can implement it with \inlinecode{ArrowParallel} by repeating an arrow \inlinecode{arr a b} and then passing it into \inlinecode{parEvalN} to get an arrow \inlinecode{arr [a] [b]}.
-Just like \inlinecode{parEvalN}, \inlinecode{parMap} is 100 \% strict.
+\ensuremath{\Varid{parMap}} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
+We can implement it with \ensuremath{\Conid{ArrowParallel}} by repeating an arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and then passing it into \ensuremath{\Varid{parEvalN}} to get an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
+Just like \ensuremath{\Varid{parEvalN}}, \ensuremath{\Varid{parMap}} is 100\% strict.
 \begin{figure}[h]
@@
 \end{figure}
-As \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \inlinecode{parEvalN} compared to \inlinecode{parEvalNLazy}.
-So it makes sense to also have a \inlinecode{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \inlinecode{parMap}, but uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
+As \ensuremath{\Varid{parMap}} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \ensuremath{\Varid{parEvalN}} compared to \ensuremath{\Varid{parEvalNLazy}}.
+So it makes sense to also have a \ensuremath{\Varid{parMapStream}} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \ensuremath{\Varid{parMap}}, but uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}.
 \begin{figure}[h]
@@
 \end{figure}
-A \inlinecode{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \inlinecode{ArrowParallel} we gave in this paper).
-This can be quite wasteful and a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \inlinecode{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
+A \ensuremath{\Varid{parMap}} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \ensuremath{\Conid{ArrowParallel}} we gave in this paper).
+This can be quite wasteful and a \ensuremath{\Varid{farm}} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \ensuremath{\Varid{numCores}} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{Definition of farm. \inlinecode{unshuffle}, \inlinecode{takeEach}, \inlinecode{shuffle} were taken from Eden's source code \cite{eden_skel_shuffle}}
+\caption{Definition of farm. \ensuremath{\Varid{unshuffle}}, \ensuremath{\Varid{takeEach}}, \ensuremath{\Varid{shuffle}} were taken from Eden's source code \cite{eden_skel_shuffle}}
 \label{fig:farm}
@@
 \end{figure}
-Since a \inlinecode{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \inlinecode{parMap} with a different work distribution, it is, again, 100\% strict.
-So we define \inlinecode{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \inlinecode{parEvalNLazy} instead of \inlinecode{parEvalN}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
+Since a \ensuremath{\Varid{farm}} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \ensuremath{\Varid{parMap}} with a different work distribution, it is, again, 100\% strict.
+So we define \ensuremath{\Varid{farmChunk}} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}.
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
 \\
@@
 	%\pagebreak
-	\FloatBarrier
+	%\FloatBarrier
 \section{Topological Skeletons}
@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons, among them a \inlinecode{pipe}, a \inlinecode{ring} and a \inlinecode{torus} implementation \cite{Loogen2012, eden_skel_topology}.
+The Eden library leverages this problem and already comes with more predefined skeletons, among them a \ensuremath{\Varid{pipe}}, a \ensuremath{\Varid{ring}} and a \ensuremath{\Varid{torus}} implementation \cite{Loogen2012, eden_skel_topology}.
 These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
@@
 
-The parallel pipe skeleton is semantically equivalent to folding over a list \inlinecode{[arr a a]} of arrows with \inlinecode{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
-We implement this skeleton using the \inlinecode{ArrowLoop} typeclass which gives us the \inlinecode{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
+The parallel pipe skeleton is semantically equivalent to folding over a list \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]} of arrows with \ensuremath{\mathbin{>\!\!>\!\!>}}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
+We implement this skeleton using the \ensuremath{\Conid{ArrowLoop}} typeclass which gives us the \ensuremath{\Varid{loop}\mathbin{::}\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
 For example %this
@@
 \end{hscode}\resethooks
-defines an arrow that takes its input \inlinecode{a} and converts it into an infinite stream \inlinecode{[a]} of it.
-Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \inlinecode{parEvalN conf fs} inside the second argument of \inlinecode{(\&\&\&)} and then only picking the first element of the resulting list with \inlinecode{arr last}.
+defines an arrow that takes its input \ensuremath{\Varid{a}} and converts it into an infinite stream \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} of it.
+Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \ensuremath{\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}} inside the second argument of \ensuremath{\mathbin{\&\!\&\!\&}} and then only picking the first element of the resulting list with \ensuremath{\Varid{arr}\;\Varid{last}}.
 \begin{figure}[h]
@@
 \caption{A first draft of the pipe skeleton expressed with parallel arrows.
-Note that the use of \inlinecode{lazy} is essential as without it programs using this definition would never halt.
-We need to enforce that the evaluation of the input \inlinecode{[a]} terminates before passing it into \inlinecode{parEvalN}.}
+Note that the use of \ensuremath{\Varid{lazy}} is essential as without it programs using this definition would never halt.
+We need to enforce that the evaluation of the input \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} terminates before passing it into \ensuremath{\Varid{parEvalN}}.}
 \label{fig:pipeSimple}
@@
 However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \inlinecode{pipe} \ref{fig:pipe}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{pipe}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{a})\;(\Varid{fut}\;\Varid{a})\;\Varid{conf},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{a})\Rightarrow {}\<[E]%
+Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \ensuremath{\Varid{pipe}} \ref{fig:pipe}.
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{pipe}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{a})\;(\Varid{fut}\;\Varid{a})\;\Varid{conf},\Conid{Future}\;\Varid{fut}\;\Varid{a})\Rightarrow {}\<[E]%
 \\
@@
 Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
-\inlinecode{arr a b} and \inlinecode{arr b c}.
-By wrapping these two arrows inside a common type we obtain \inlinecode{pipe2} (Fig.~\ref{fig:pipe2}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}}.
+By wrapping these two arrows inside a common type we obtain \ensuremath{\Varid{pipe2}} (Fig.~\ref{fig:pipe2}).
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{25}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{head}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\mathbf{where}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
+\>[9]{}\Varid{arr}\;\Varid{head}\;\mathbf{where}{}\<[E]%
+\\
+\>[9]{}\hsindent{16}{}\<[25]%
 \>[25]{}\Varid{unify}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
@@
 \\
-\>[17]{}\hsindent{8}{}\<[25]%
+\>[9]{}\hsindent{16}{}\<[25]%
 \>[25]{}\Varid{unify}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
@@
 
-Note that extensive use of this combinator over \inlinecode{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \inlinecode{(\ensuremath{\mathbin{>\!\!>\!\!>}})} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+Note that extensive use of this combinator over \ensuremath{\Varid{pipe}} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Nonetheless, we can define a parallel piping operator \ensuremath{(}>>>\ensuremath{)} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \ensuremath{(\mathbin{>\!\!>\!\!>})} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
 \begin{figure}[h]
@@
 \end{figure}
-Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \inlinecode{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
-Its input is a node function \inlinecode{i -> r -> (o, r)} in which \inlinecode{r} serves as the intermediary output that gets send to the neighbour of each node.
+Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate in a ring topology with each other.
+Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets send to the neighbour of each node.
 This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
@@
 
-We can rewrite its functionality easily with the use of \inlinecode{loop} as the definition of the node function, \inlinecode{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \inlinecode{loop}'s \inlinecode{arr (a, b) (c, b) -> arr a c}.
-In each iteration we start by rotating the intermediary input from the nodes \inlinecode{[fut r]} with \inlinecode{second (rightRotate >>> lazy)}.
-Similarly to the \inlinecode{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \inlinecode{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
-%\begin{quotation}
-\enquote{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
+We can rewrite its functionality easily with the use of \ensuremath{\Varid{loop}} as the definition of the node function, \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})}, after being transformed into an arrow, already fits quite neatly into the \ensuremath{\Varid{loop}}'s \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}}.
+In each iteration we start by rotating the intermediary input from the nodes \ensuremath{[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu]} with \ensuremath{\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})}.
+Similarly to the \ensuremath{\Varid{pipe}} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \ensuremath{\Varid{lazy}} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
+\begin{quote}
+{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
 Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
-%\end{quotation}
-Next, we zip the resulting \inlinecode{([i], [fut r])} to \inlinecode{[(i, fut r)]} with \inlinecode{arr (uncurry zip)} so we can feed that into a our input arrow \inlinecode{arr (i, r) (o, r)}, which we transform into \inlinecode{arr (i, fut r) (o, fut r)} before lifting it to \inlinecode{arr [(i, fut r)] [(o, fut r)]} to get a list \inlinecode{[(o, fut r)]}.
-Finally we unzip this list into \inlinecode{([o], [fut r])}.
-Plugging this arrow \inlinecode{arr ([i], [fut r]) ([o], fut r)} into the definition of \inlinecode{loop} from earlier gives us \inlinecode{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
+\end{quote}
+Next, we zip the resulting \ensuremath{([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])} to \ensuremath{[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})} so we can feed that into a our input arrow \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})}, which we transform into \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})} before lifting it to \ensuremath{\Varid{arr}\;[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]\;[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} to get a list \ensuremath{[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]}.
+Finally we unzip this list into \ensuremath{([\mskip1.5mu \Varid{o}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])}.
+Plugging this arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])\;([\mskip1.5mu \Varid{o}\mskip1.5mu],\Varid{fut}\;\Varid{r})} into the definition of \ensuremath{\Varid{loop}} from earlier gives us \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
 \begin{figure}[h]
@@
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-\\\\
-With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \inlinecode{ArrowLoop} typeclass.
-\\\\
+
+With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology}---yet again with the help of the \ensuremath{\Conid{ArrowLoop}} typeclass.
+
 Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple \inlinecode{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \inlinecode{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \inlinecode{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \inlinecode{[[fut a]]} horizontally and \inlinecode{[[fut b]]} vertically.
-Then, we once again zip the inputs for the input arrow with \inlinecode{arr (uncurry3 zipWith3 lazyzip3)} from \inlinecode{([[c]], ([[fut a]], [[fut b]]))} to \inlinecode{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu])} in the second argument (loop only allows for 2 arguments) of our looped arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))\;([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} and our rotation arrow becomes \ensuremath{\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{\*\!\*\!\*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))} instead of the singular rotation in the ring as we rotate \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu]} horizontally and \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]} vertically.
+Then, we once again zip the inputs for the input arrow with \ensuremath{\Varid{arr}\;(\Varid{uncurry3}\;\Varid{zipWith3}\;\Varid{lazyzip3})} from \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} to \ensuremath{[\mskip1.5mu [\mskip1.5mu (\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]}, which we then feed into our parallel execution.
 \\\\
 This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
-We first have to \inlinecode{shuffle} all the inputs to then pass it into \inlinecode{parMap conf (ptorus f)} which yields us \inlinecode{[(d, fut a, fut b)]}.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \inlinecode{arr length >>> arr unshuffle} with the use of \inlinecode{app :: arr (arr a b, a) c} from the \inlinecode{ArrowApply} typeclass.
-Finally, we unpack this matrix \inlinecode{[[[(d, fut a, fut b)]]} with \inlinecode{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \inlinecode{([[d]], ([[fut a]], [[fut b]]))}.
+We first have to \ensuremath{\Varid{shuffle}} all the inputs to then pass it into \ensuremath{\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f})} which yields us \ensuremath{[\mskip1.5mu (\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]}.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \ensuremath{\Varid{arr}\;\Varid{length}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unshuffle}} with the use of \ensuremath{\Varid{app}\mathbin{::}\Varid{arr}\;(\Varid{arr}\;\Varid{a}\;\Varid{b},\Varid{a})\;\Varid{c}} from the \ensuremath{\Conid{ArrowApply}} typeclass.
+Finally, we unpack this matrix \ensuremath{[\mskip1.5mu [\mskip1.5mu [\mskip1.5mu (\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!\!>\!\!>}\Varid{threetotwo}} to get  \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))}.
 \\\\
@@
 \end{hscode}\resethooks
+ % $
 \caption{Definition of the torus skeleton}
@@
 This work features multiple parallel backends: the already available parallel Haskell flavours.
-Parallel Arrows feature an implementation of the \inlinecode{ArrowParallel} interface for Multicore Haskell, \inlinecode{Par} monad, and Eden.
+Parallel Arrows feature an implementation of the \ensuremath{\Conid{ArrowParallel}} interface for Multicore Haskell, \ensuremath{\Conid{Par}} monad, and Eden.
 With our approach parallel programs can be ported across these flavours with no effort.

### 2017-05-18 paper/content/main.tex
kerning and looks and bugs

@@
 \end{figure}
-the \ensuremath{\mathbin{\*\!\*\!\*}} combinator which combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{\*\!\*\!\*})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
-\\
-\>[B]{}\Varid{f}\mathbin{\*\!\*\!\*}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{g}{}\<[E]%
+the \ensuremath{\mathbin{*\!*\!*}} combinator which combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{*\!*\!*})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
+\\
+\>[B]{}\Varid{f}\mathbin{*\!*\!*}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{g}{}\<[E]%
 \ColumnHook
@@
 \end{figure}
-and the \ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{\*\!\*\!\*}}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+and the \ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
 \begin{figure}[h]
@@
 \\
-\>[B]{}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}(\Varid{f}\mathbin{\*\!\*\!\*}\Varid{g}){}\<[E]%
+\>[B]{}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}(\Varid{f}\mathbin{*\!*\!*}\Varid{g}){}\<[E]%
 \ColumnHook
@@
 However, we can implement such a \ensuremath{\Varid{parEval2}} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into a new parallel arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} quite easily with the help of the \ensuremath{\Conid{ArrowChoice}} typeclass.
-The idea is to use the \ensuremath{\mathbin{+++}} combinator which combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} and transforms them into \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})} to get a common arrow type that we can then feed into parEvalN.
+The idea is to use the \ensuremath{\mathbin{+\!\!+\!\!+}} combinator which combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} and transforms them into \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})} to get a common arrow type that we can then feed into parEvalN.
 \\\\
 We start by transforming the \ensuremath{(\Varid{a},\Varid{c})} input into a 2-element list \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{a}\;\Varid{c}\mskip1.5mu]} by first tagging the two inputs with \ensuremath{\Conid{Left}} and \ensuremath{\Conid{Right}} and wrapping the right element in a singleton list with \ensuremath{\Varid{return}} so that we can combine them with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))}.
-Next, we feed this list into a parallel arrow running on 2 instances of \ensuremath{\Varid{f}\mathbin{+++}\Varid{g}} as described above.
+Next, we feed this list into a parallel arrow running on 2 instances of \ensuremath{\Varid{f}\mathbin{+\!\!+\!\!+}\Varid{g}} as described above.
 After the calculation is finished, we convert the resulting \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{b}\;\Varid{d}\mskip1.5mu]} into \ensuremath{([\mskip1.5mu \Varid{b}\mskip1.5mu],[\mskip1.5mu \Varid{d}\mskip1.5mu])} with \ensuremath{\Varid{arr}\;\Varid{partitionEithers}}.
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Conid{Left}\mathbin{\*\!\*\!\*}(\Varid{arr}\;\Conid{Right}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{return})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\>[9]{}\Varid{arr}\;\Conid{Left}\mathbin{*\!*\!*}(\Varid{arr}\;\Conid{Right}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{return})\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{f}\mathbin{+++}\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{f}\mathbin{+\!\!+\!\!+}\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{head}\mathbin{\*\!\*\!\*}\Varid{arr}\;\Varid{head}{}\<[E]%
+\>[9]{}\Varid{arr}\;\Varid{head}\mathbin{*\!*\!*}\Varid{arr}\;\Varid{head}{}\<[E]%
 \ColumnHook
@@
 	\subsubsection{Syntactic Sugar} \label{syntacticSugar}
-For basic arrows, we have the \ensuremath{(\mathbin{\*\!\*\!\*})} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into an arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} which does both computations at once.
-This can easily be translated into a parallel version \ensuremath{\mathbin{|\!\*\!\*\!\*\!|}} (Fig.~\ref{fig:par***}) with the use of \ensuremath{\Varid{parEval2}}, but for this we require a backend which has an implementation that does not require any configuration (hence the \ensuremath{()} as the conf parameter in Fig.~\ref{fig:par***}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{|\!\*\!\*\!\*\!|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;()))\Rightarrow {}\<[E]%
+For basic arrows, we have the \ensuremath{\mathbin{*\!*\!*}} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into an arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} which does both computations at once.
+This can easily be translated into a parallel version \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} (Fig.~\ref{fig:par***}) with the use of \ensuremath{\Varid{parEval2}}, but for this we require a backend which has an implementation that does not require any configuration (hence the \ensuremath{()} as the conf parameter in Fig.~\ref{fig:par***}).
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{\mid\!\!*\!*\!*\!\!\mid})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;()))\Rightarrow {}\<[E]%
 \\
@@
 \\
-\>[B]{}(\mathbin{|\!\*\!\*\!\*\!|})\mathrel{=}\Varid{parEval2}\;(){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of \ensuremath{\mathbin{|\!\*\!\*\!\*\!|}}, the parallel version of \ensuremath{\mathbin{\*\!\*\!\*}}.}
+\>[B]{}(\mathbin{\mid\!\!*\!*\!*\!\!\mid})\mathrel{=}\Varid{parEval2}\;(){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Definition of \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}}, the parallel version of \ensuremath{\mathbin{*\!*\!*}}.}
 \label{fig:par***}
@@
 % With this we can analogously to the serial |&&&|
-We define the parallel \ensuremath{\mathbin{|\!\&\!\&\!\&\!|}} (Fig.~\ref{fig:par&&&}) in a similar manner to its sequential pendant \ensuremath{\mathbin{\&\!\&\!\&}} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{|\!\&\!\&\!\&\!|})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{a})\;(\Conid{Either}\;\Varid{b}\;\Varid{c})\;())\Rightarrow {}\<[E]%
+We define the parallel \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}} (Fig.~\ref{fig:par&&&}) in a similar manner to its sequential pendant \ensuremath{\mathbin{\&\!\&\!\&}} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{a})\;(\Conid{Either}\;\Varid{b}\;\Varid{c})\;())\Rightarrow {}\<[E]%
 \\
@@
 \\
-\>[B]{}(\mathbin{|\!\&\!\&\!\&\!|})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{|\!\*\!\*\!\*\!|}\Varid{g}{}\<[E]%
+\>[B]{}(\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{\mid\!\!*\!*\!*\!\!\mid}\Varid{g}{}\<[E]%
 \ColumnHook
@@
  % $ %% formatting
-\caption{Definition of \ensuremath{\mathbin{|\!\&\!\&\!\&\!|}} - the parallel version of \ensuremath{\mathbin{\&\!\&\!\&}}.}
+\caption{Definition of \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}} - the parallel version of \ensuremath{\mathbin{\&\!\&\!\&}}.}
 \label{fig:par&&&}
@@
 \>[25]{}\hsindent{8}{}\<[33]%
-\>[33]{}(\Varid{mapArr}\;\Varid{f}\mathbin{\*\!\*\!\*}\Varid{mapArr}\;\Varid{g})\mathbin{\*\!\*\!\*}\Varid{arr}\;(\mathbin{\char92 \char95 }\to [\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
+\>[33]{}(\Varid{mapArr}\;\Varid{f}\mathbin{*\!*\!*}\Varid{mapArr}\;\Varid{g})\mathbin{*\!*\!*}\Varid{arr}\;(\mathbin{\char92 \char95 }\to [\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu])} in the second argument (loop only allows for 2 arguments) of our looped arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))\;([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} and our rotation arrow becomes \ensuremath{\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{\*\!\*\!\*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))} instead of the singular rotation in the ring as we rotate \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu]} horizontally and \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]} vertically.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu])} in the second argument (loop only allows for 2 arguments) of our looped arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))\;([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} and our rotation arrow becomes \ensuremath{\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))} instead of the singular rotation in the ring as we rotate \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu]} horizontally and \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]} vertically.
 Then, we once again zip the inputs for the input arrow with \ensuremath{\Varid{arr}\;(\Varid{uncurry3}\;\Varid{zipWith3}\;\Varid{lazyzip3})} from \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} to \ensuremath{[\mskip1.5mu [\mskip1.5mu (\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]}, which we then feed into our parallel execution.
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{\*\!\*\!\*}{}\<[E]%
+\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{\mid\!\mid\!\mid}(\Varid{f}\mathbin{\*\!\*\!\*}\Varid{mapArr}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
+\>[9]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{\mid\!\mid\!\mid}(\Varid{f}\mathbin{*\!*\!*}\Varid{mapArr}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
 \\

### 2017-05-19 paper/content/main.tex
generated pdf

@@
 \label{sec:introduction}
-\olcomment{todo, reuse 5.5, "Impact" at the end and more}
+\olcomment{todo, reuse 5.5, and more}
 
@@
 
-HIT HERE REALLY STRONG
+\olcomment{HIT HERE REALLY STRONG}
+
+%\subsection{Impact of parallel Arrows}
+%\olcomment{move this to Contributions in the front or something}
+We wrap parallel Haskells inside of our \ensuremath{\Conid{ArrowParallel}} interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
+\begin{itemize}
+	\item \textbf{Arrow API benefits}:
+	With the \ensuremath{\Conid{ArrowParallel}} typeclass we do not lose any benefits of using arrows as \ensuremath{\Varid{parEvalN}} is just yet another arrow combinator.
+The resulting arrow can be used in the same way a potential serial version could be used.
+This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
+We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
+	\item \textbf{Abstraction}:
+	With the \ensuremath{\Conid{ArrowParallel}} typeclass, we abstracted all parallel implementation logic away from the business logic.
+This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
+So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \ensuremath{\Conid{ArrowParallel}} instance.
+\end{itemize}
 
@@
 To ensure that programs that use \ensuremath{\Varid{parEvalN}} have the correct evaluation order, we annotate the computation with \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} which forces the compiler to not reorder multiple \ensuremath{\Varid{parEvalN}} computations.
-This is particularly necessary in circular communication topologies like in the \ensuremath{\Varid{torus}} or \ensuremath{\Varid{ring}} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \ensuremath{\Varid{pseq}} during testing for this paper.
-
-%%%% BROKEN!
+This is particularly necessary in circular communication topologies like in the \ensuremath{\Varid{torus}} or \ensuremath{\Varid{ring}} skeleton that we will see in Section~\ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \ensuremath{\Varid{pseq}} during testing for this paper.
+
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
+ %$
 
@@
 \\\\
-The most prominent instances of this interface are regular functions \ensuremath{(\to )} (Fig.~\ref{fig:arrowfn}),
-\begin{figure}[h]
+The most prominent instances of this interface are regular functions \ensuremath{(\to )}: %(Fig.~\ref{fig:arrowfn}),
+%\begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{Arrow instance for regular functions}
-\label{fig:arrowfn}
-\end{figure}
-and the Kleisli type (Fig.~\ref{fig:arrowkleisli}).
-\begin{figure}[h]
+% \caption{Arrow instance for regular functions}
+% \label{fig:arrowfn}
+% \end{figure}
+%%% on the verge of acceptable code
+and the Kleisli type: %(Fig.~\ref{fig:arrowkleisli}).
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 %$
-\caption{Definition of the Kleisli type and the corresponding arrow instance}
-\label{fig:arrowkleisli}
-\end{figure}
+% \caption{Definition of the Kleisli type and the corresponding arrow instance}
+% \label{fig:arrowkleisli}
+% \end{figure}
 \begin{figure}[h]
@@
 		% \subcaptionbox
-{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
+{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
 		% \subcaptionbox
-{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
+{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
+\ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
+\midrule
 		% \subcaptionbox
-{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
 		% \subcaptionbox
-{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
+{}{\includegraphics[width = 1.5in]{images/andandand}}\\
+\ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\
 	\end{tabular}
-	\caption{Syntactic sugar for arrows}
+	\caption{Visual depiction of syntactic sugar for Arrows.}
 	\label{3}
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \ensuremath{\Varid{first}}, called \ensuremath{\Varid{second}} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
-\begin{figure}[h]
+With this typeclass in place, Hughes also defined some syntactic sugar: the functions \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
+The mirrored version of \ensuremath{\Varid{first}}, called \ensuremath{\Varid{second}} (Fig.~\ref{fig:secondImg}% ,~\ref{fig:second}
+) is:
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{The second combinator}
-\label{fig:second}
-\end{figure}
-the \ensuremath{\mathbin{*\!*\!*}} combinator which combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
-\begin{figure}[h]
+% \caption{The second combinator}
+% \label{fig:second}
+% \end{figure}
+the \ensuremath{\mathbin{*\!*\!*}} combinator that combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img}%
+% ,~\ref{fig:***}
+) is defined as
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{The (***) combinator}
-\label{fig:***}
-\end{figure}
-and the \ensuremath{\mathbin{\char92 \&\char92 \&\char92 \&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
-\begin{figure}[h]
+% \caption{The (***) combinator}
+% \label{fig:***}
+% \end{figure}
+and the \ensuremath{\mathbin{\&\!\&\!\&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input (Fig.~\ref{fig:&&&Img}% ,~\ref{fig:&&&}
+) is:
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{The (\&\&\&) combinator}
-\label{fig:&&&}
-\end{figure}
-A short example given by Hughes on how to use this is \ensuremath{\Varid{add}} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
-\begin{figure}[h]
+% \caption{The (\&\&\&) combinator}
+% \label{fig:&&&}
+% \end{figure}
+A short example given by Hughes on how to use this is 
+addition with arrows:
+%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{Add over arrows}
-\label{fig:addArrows}
-\end{figure}
+% \caption{Add over arrows}
+% \label{fig:addArrows}
+% \end{figure}
 % The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator.
@@
 As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, \ensuremath{\Varid{parEvalN}} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
-Translating this into arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation).
-\begin{figure}[h]
+Translating this into arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
+% (Fig.~\ref{fig:parEvalNArrowFn}) (
+This combinator is similar to our utility function \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{parEvalN Arrow combinator as a function}
-\label{fig:parEvalNArrowFn}
-\end{figure}
+% \caption{parEvalN Arrow combinator as a function}
+% \label{fig:parEvalNArrowFn}
+% \end{figure}
 With this definition of \ensuremath{\Varid{parEvalN}}, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow \ensuremath{\Varid{arr}} and we want this to be an interface for different backends, we introduce a new typeclass \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
-\begin{figure}[h]
+But as the implementation may differ depending on the actual type of the arrow \ensuremath{\Varid{arr}} and we want this to be an interface for different backends, we introduce a new typeclass \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}} to host this combinator: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
-\label{fig:parEvalNArrowTypeClass1}
-\end{figure}
+% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
+% \label{fig:parEvalNArrowTypeClass1}
+% \end{figure}
 Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs.
@@
 We also do not want \ensuremath{\Varid{conf}} to be a fixed type, as the configuration parameters can differ for different instances of \ensuremath{\Conid{ArrowParallel}}.
-So we add it to the type signature of the typeclass as well and get \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
-\begin{figure}[h]
+So we add it to the type signature of the typeclass as well and get \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}}: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
+% \begin{figure}[h]
+\olcomment{\ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}} or \ensuremath{\Conid{ArrowParallel}\;\Varid{conf}\;\Varid{arr}\;\Varid{a}\;\Varid{b}}?}
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
-\label{fig:parEvalNArrowTypeClassFinal}
-\end{figure}
+% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
+% \label{fig:parEvalNArrowTypeClassFinal}
+% \end{figure}
 Note that we don't require the \ensuremath{\Varid{conf}} parameter in every implementation.
-If it is not needed, we usually just default the \ensuremath{\Varid{conf}} type parameter to () and even blank it out in the parameter list of the implemented \ensuremath{\Varid{parEvalN}}, as we will see in the implementation of the Multicore and the ParMonad backend.
+If it is not needed, we usually just default the \ensuremath{\Varid{conf}} type parameter to \ensuremath{()} and even blank it out in the parameter list of the implemented \ensuremath{\Varid{parEvalN}}, as we will see in the implementation of the Multicore and the \ensuremath{\Conid{Par}} Monad backend.
 
@@
 \subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} and \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} combinators from Multicore Haskell, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} but with flipped parameters.
+The Multicore Haskell implementation of this class is implemented in a straightforward manner by using \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns} combined with the \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} and \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} combinators from Multicore Haskell, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} but with flipped parameters.
 \begin{figure}[h]
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()\;\mathbf{where}{}\<[E]%
+\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()\;\mathbf{where}{}\<[E]%
 \\
@@
 We therefore introduce the \ensuremath{\Conid{Conf}\;\Varid{a}} data-type that simply wraps a \ensuremath{\Conid{Strategy}\;\Varid{a}} (Fig.~\ref{fig:confa}).
-We can't directly use the \ensuremath{\Conid{Strategy}\;\Varid{a}} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
-\begin{figure}[h]
+We can't directly use the \ensuremath{\Conid{Strategy}\;\Varid{a}} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances:
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{Definition of Conf a}
-\label{fig:confa}
-\end{figure}
+% \caption{Definition of Conf a}
+% \label{fig:confa}
+% \end{figure}
 To get our configurable \ensuremath{\Conid{ArrowParallel}} instance, we simply unwrap the strategy and pass it to \ensuremath{\Varid{parList}} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
@@
 \end{figure}
-\subsubsection{Par Monad}
-The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} function.
+\subsubsection{\ensuremath{\Conid{Par}} Monad}
+\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting}
+The \ensuremath{\Conid{Par}} monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and \ensuremath{\Conid{Par}} monad's \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} function.
 The latter forks away the computation of a value and returns an \ensuremath{\Conid{IVar}} containing the result in the \ensuremath{\Conid{Par}} monad.
@@
 This yields a list \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we then convert into \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}.
-In order to wait for the computation to finish, we map over the \ensuremath{\Conid{IVar}}s inside the ParMonad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}.
+In order to wait for the computation to finish, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}.
 The result of this operation is a \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} from which we can finally remove the monad again by running \ensuremath{\Varid{arr}\;\Varid{runPar}} to get our output of \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
@@
 \subsubsection{Eden}
-For the Multicore and ParMonad implementation we could use general instances of \ensuremath{\Conid{ArrowParallel}} that just require the \ensuremath{\Conid{ArrowApply}} and \ensuremath{\Conid{ArrowChoice}} typeclasses.
+For both the Multicore Haskell and \ensuremath{\Conid{Par}} Monad implementations we could use general instances of \ensuremath{\Conid{ArrowParallel}} that just require the \ensuremath{\Conid{ArrowApply}} and \ensuremath{\Conid{ArrowChoice}} typeclasses.
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a typeclass \ensuremath{\Conid{ArrowUnwrap}} (Fig.~\ref{fig:ArrowUnwrap}).
-\begin{figure}[h]
+While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{possible ArrowUnwrap typeclass}
-\label{fig:ArrowUnwrap}
-\end{figure}
-We don't do it here for aesthetic resons, though. For now, we just implement \ensuremath{\Conid{ArrowParallel}} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
-\begin{figure}[h]
+% \caption{possible ArrowUnwrap typeclass}
+% \label{fig:ArrowUnwrap}
+% \end{figure}
+We don't do it here for aesthetic resons, though. For now, we just implement \ensuremath{\Conid{ArrowParallel}} for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{ArrowParallel instance for functions in the Eden backend}
-\label{fig:ArrowParallelEdenFns}
-\end{figure}
-and the Kleisli type (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
-\begin{figure}[h]
+% \caption{ArrowParallel instance for functions in the Eden backend}
+% \label{fig:ArrowParallelEdenFns}
+% \end{figure}
+and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{ArrowParallel instance for the Kleisli type in the Eden backend}
-\label{fig:ArrowParallelEdenKleisli}
-\end{figure}
+% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
+% \label{fig:ArrowParallelEdenKleisli}
+% \end{figure}
 
 %\FloatBarrier
-
-\subsection{Impact of parallel Arrows}
-\olcomment{move this to Contributions in the front or something}
-We have seen that we can wrap parallel Haskells inside of the \ensuremath{\Conid{ArrowParallel}} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
-\begin{itemize}
-	\item \textbf{Arrow API benefits}:
-	With the \ensuremath{\Conid{ArrowParallel}} typeclass we do not lose any benefits of using arrows as \ensuremath{\Varid{parEvalN}} is just yet another arrow combinator.
-The resulting arrow can be used in the same way a potential serial version could be used.
-This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
-We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
-	\item \textbf{Abstraction}:
-	With the \ensuremath{\Conid{ArrowParallel}} typeclass, we abstracted all parallel implementation logic away from the business logic.
-This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
-So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \ensuremath{\Conid{ArrowParallel}} instance.
-\end{itemize}
 
@@
  %$ %% formatting
-\caption{Definition of parEvalNLazy}
+\caption{Definition of \ensuremath{\Varid{parEvalNLazy}}.}
 \label{fig:parEvalNLazy}
@@
 	\includegraphics[scale=0.7]{images/parEval2}
-	\caption{Schematic depiction of parEval2}
+	\caption{Schematic depiction of \ensuremath{\Varid{parEval2}}.}
 	\label{fig:parEval2Img}
@@
 For basic arrows, we have the \ensuremath{\mathbin{*\!*\!*}} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into an arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} which does both computations at once.
-This can easily be translated into a parallel version \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} (Fig.~\ref{fig:par***}) with the use of \ensuremath{\Varid{parEval2}}, but for this we require a backend which has an implementation that does not require any configuration (hence the \ensuremath{()} as the conf parameter in Fig.~\ref{fig:par***}).
-\begin{figure}[h]
+This can easily be translated into a parallel version \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} (Fig.~\ref{fig:par***}) with the use of \ensuremath{\Varid{parEval2}}, but for this we require a backend which has an implementation that does not require any configuration (hence the \ensuremath{()} as the \ensuremath{\Varid{conf}} parameter% in Fig.~\ref{fig:par***}
+):
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{Definition of \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}}, the parallel version of \ensuremath{\mathbin{*\!*\!*}}.}
-\label{fig:par***}
-\end{figure}
+% \caption{Definition of |parstar|, the parallel version of |***|.}
+% \label{fig:par***}
+% \end{figure}
 % With this we can analogously to the serial |&&&|
-We define the parallel \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}} (Fig.~\ref{fig:par&&&}) in a similar manner to its sequential pendant \ensuremath{\mathbin{\&\!\&\!\&}} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
-\begin{figure}[h]
+We define the parallel \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}} % (Fig.~\ref{fig:par&&&}) 
+in a similar manner to its sequential pendant \ensuremath{\mathbin{\&\!\&\!\&}} (Fig.~\ref{fig:&&&Img}% ,~\ref{fig:&&&}
+):
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
  % $ %% formatting
-\caption{Definition of \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}} - the parallel version of \ensuremath{\mathbin{\&\!\&\!\&}}.}
-\label{fig:par&&&}
-\end{figure}
+% \caption{Definition of |parand| - the parallel version of |&&&|.}
+% \label{fig:par&&&}
+% \end{figure}
 	%\pagebreak
@@
 	\label{fig:withoutFutures}
+\olcomment{more practical and heavy-weight example! fft (I have the code)?}
 \end{figure}
@@
 \end{figure}
-Since \ensuremath{\Conid{RD}} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \ensuremath{\Conid{Strategy}\;\Varid{a}} in the Multicore Haskell implementation of \ensuremath{\Conid{ArrowParallel}} in chapter \ref{sec:parrows:multicore}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{RemoteData}\;\Varid{a}\mathrel{=}\Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathbin{::}\Conid{RD}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{RemoteData}\;\Varid{a}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{put}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to \Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathrel{=}\Varid{release}\;\Varid{a}\mskip1.5mu\}){}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{rd}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{fetch}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{RD based RemoteData version of Future for the Eden backend}
-\label{fig:RDFuture}
-\end{figure}
+Since \ensuremath{\Conid{RD}} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Appendix in Fig.~\ref{fig:RDFuture}.
+This is due to the same reason we had to introduce a wrapper for \ensuremath{\Conid{Strategy}\;\Varid{a}} in the Multicore Haskell implementation of \ensuremath{\Conid{ArrowParallel}} in Section~\ref{sec:parrows:multicore}.
 
@@
 
-Furthermore, in order for these \ensuremath{\Conid{Future}} types to fit with the \ensuremath{\Conid{ArrowParallel}} instances we gave earlier, we have to give the necessary \ensuremath{\Conid{NFData}} and \ensuremath{\Conid{Trans}} instances - the latter only being needed in Eden.
-Because \ensuremath{\Conid{MVar}} already ships with a \ensuremath{\Conid{NFData}} instance, we only have to supply two simple instances for our \ensuremath{\Conid{RemoteData}} type.
-\begin{figure}[h]
+Furthermore, in order for these \ensuremath{\Conid{Future}} types to fit with the \ensuremath{\Conid{ArrowParallel}} instances we gave earlier, we have to give the necessary \ensuremath{\Conid{NFData}} and \ensuremath{\Conid{Trans}} instances, the latter are only needed in Eden.
+Because \ensuremath{\Conid{MVar}} already ships with a \ensuremath{\Conid{NFData}} instance, we only have to supply two simple instances for our \ensuremath{\Conid{RemoteData}} type:
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{NFData and Trans instances for the RemoteData type.
+% \caption{NFData and Trans instances for the RemoteData type.
 The Trans instance does not have any functions declared as the default implementation suffices here.
 See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
-\end{figure}
-
-Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
+% \end{figure}
+The \ensuremath{\Conid{Trans}} instance does not have any functions declared as the default implementation suffices here.
+
+Going back to our communication example we can use this \ensuremath{\Conid{Future}} concept in order to enable direct communications between the nodes in the following way:
 \begin{figure}[h]
@@
 	\includegraphics[scale=0.7]{images/parMap}
-	\caption{Schematic depiction of parMap}
+	\caption{Schematic depiction of \ensuremath{\Varid{parMap}}.}
 	\label{fig:parMapImg}
 \end{figure}
-\ensuremath{\Varid{parMap}} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
+The \ensuremath{\Varid{parMap}} skeleton (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
 We can implement it with \ensuremath{\Conid{ArrowParallel}} by repeating an arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and then passing it into \ensuremath{\Varid{parEvalN}} to get an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parMap}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to (\Varid{arr}\;\Varid{a}\;\Varid{b})\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parMap}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf})\Rightarrow \Varid{conf}\to (\Varid{arr}\;\Varid{a}\;\Varid{b})\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-\caption{Definition of parMapStream}
+\caption{Definition of \ensuremath{\Varid{parMapStream}}.}
 \label{fig:parMapStream}
@@
 	\includegraphics[scale=0.7]{images/farm}
-	\caption{Schematic depiction of farm}
+	\caption{Schematic depiction of a \ensuremath{\Varid{farm}}, a statically
+          load-balanced \ensuremath{\Varid{parMap}}.}
 	\label{fig:farmImg}
@@
 This can be quite wasteful and a \ensuremath{\Varid{farm}} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \ensuremath{\Varid{numCores}} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
+The definitions of helper functions \ensuremath{\Varid{unshuffle}}, \ensuremath{\Varid{takeEach}}, \ensuremath{\Varid{shuffle}} (shown in Appendix) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{Definition of farm. \ensuremath{\Varid{unshuffle}}, \ensuremath{\Varid{takeEach}}, \ensuremath{\Varid{shuffle}} were taken from Eden's source code \cite{eden_skel_shuffle}}
+\caption{The definition of \ensuremath{\Varid{farm}}.}
 \label{fig:farm}
@@
 
-\subsection{farmChunk}
+\subsection{The \ensuremath{\Varid{farmChunk}} Skeleton}
 \begin{figure}[h]
@@
 Since a \ensuremath{\Varid{farm}} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \ensuremath{\Varid{parMap}} with a different work distribution, it is, again, 100\% strict.
-So we define \ensuremath{\Varid{farmChunk}} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{farmChunk}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{shuffle}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of farmChunk}
-\label{fig:farmChunk}
-\end{figure}
-
-\subsection{parMapReduce}
+So we can define \ensuremath{\Varid{farmChunk}} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}.
+It is basically the same definition as for \ensuremath{\Varid{farm}}, with \ensuremath{\Varid{parEvalN}} replaced with \ensuremath{\Varid{parEvalNLazy}}, as Appendix shows.
+
+\subsection{Map and reduce}
+
+A~simple \ensuremath{\Varid{map}}--\ensuremath{\Varid{reduce}} can be written like in Figure~\ref{fig:parMapReduceDirect}.
+Notice that the performance of the \ensuremath{\mathbin{>\!\!>\!\!>}} combinator is essential for the performance of the skeleton.
+A~definitive version would use Futures.
+
  \olcomment{it appears STRANGE. are the data really left alone and noded after map and taken from there by reduce? It makes sense only is no communication through master takes place. ELSE: CUT!}
@@
  \mbcomment{this requires some work. Either change this to use futures or cut, yes.}
--- this does not completely adhere to Google's definition of Map Reduce as it
--- the mapping function does not allow for "reordering" of the output
--- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parMapReduceDirect}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}\;\Varid{conf},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
+% -- this does not completely adhere to Google's definition of Map Reduce as it
+% -- the mapping function does not allow for "reordering" of the output
+% -- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
+
+\olcomment{now rewritten as motivation for futures. maybe still cut?}
+
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parMapReduceDirect}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}\;\Varid{conf},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
 \\
@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons, among them a \ensuremath{\Varid{pipe}}, a \ensuremath{\Varid{ring}} and a \ensuremath{\Varid{torus}} implementation \cite{Loogen2012, eden_skel_topology}.
-These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
+The Eden library leverages this problem and already comes\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} with more predefined skeletons, among them a \ensuremath{\Varid{pipe}}, a \ensuremath{\Varid{ring}}, and a \ensuremath{\Varid{torus}} implementations \cite{Loogen2012}.
+These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell.
+We aim to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
 
@@
 
-The parallel pipe skeleton is semantically equivalent to folding over a list \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]} of arrows with \ensuremath{\mathbin{>\!\!>\!\!>}}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
+The parallel \ensuremath{\Varid{pipe}} skeleton is semantically equivalent to folding over a list \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]} of arrows with \ensuremath{\mathbin{>\!\!>\!\!>}}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
 We implement this skeleton using the \ensuremath{\Conid{ArrowLoop}} typeclass which gives us the \ensuremath{\Varid{loop}\mathbin{::}\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
 For example %this
-\mbcomment{das kann man hier so lassen, oder?}
+\mbcomment{das kann man hier so lassen, oder?}\olcomment{sicherlich!}
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-, which is the same as
+which is the same as
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{A first draft of the pipe skeleton expressed with parallel arrows.
+\caption{A~first implementation of the \ensuremath{\Varid{pipe}} skeleton expressed with Parallel Arrows.
 Note that the use of \ensuremath{\Varid{lazy}} is essential as without it programs using this definition would never halt.
@@
 However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \ensuremath{\Varid{pipe}} \ref{fig:pipe}.
-\begin{figure}[h]
+Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \ensuremath{\Varid{pipe}}: % ~\ref{fig:pipe}.
+% \begin{figure}[h]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{Final definition of the pipe skeleton which uses Futures}
-\label{fig:pipe}
-\end{figure}
-
-Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
+% \caption{Final definition of the pipe skeleton which uses Futures}
+% \label{fig:pipe}
+% \end{figure}
+
+Sometimes, this \ensuremath{\Varid{pipe}} definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
 \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}}.
@@
 \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\olcomment{I swapped the type classes here:}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{25}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{pipe2}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;\Varid{conf},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\Rightarrow {}\<[E]%
+\>[B]{}\Varid{pipe2}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]),{}\<[E]%
+\\
+\>[B]{}\hsindent{10}{}\<[10]%
+\>[10]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;\Varid{conf})\Rightarrow {}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{snd}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{head}\;\mathbf{where}{}\<[E]%
+\>[9]{}\Varid{arr}\;\Varid{snd}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{head}\;\mathbf{where}{}\<[E]%
 \\
 \>[9]{}\hsindent{16}{}\<[25]%
-\>[25]{}\Varid{unify}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[25]{}\hsindent{8}{}\<[33]%
-\>[33]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]){}\<[E]%
+\>[25]{}\Varid{unify}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]){}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-\caption{Definition of pipe2}
+\caption{Definition of \ensuremath{\Varid{pipe2}}.}
 \label{fig:pipe2}
@@
 Note that extensive use of this combinator over \ensuremath{\Varid{pipe}} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \ensuremath{(}>>>\ensuremath{)} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \ensuremath{(\mathbin{>\!\!>\!\!>})} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{|\!\!>\!\!>\!\!>\!\!|})\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(),{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\Rightarrow {}\<[E]%
+Nonetheless, we can define a parallel piping operator \ensuremath{\mathbin{|\!\!>\!\!>\!\!>\!\!|}} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \ensuremath{\mathbin{>\!\!>\!\!>}} similarly to other parallel syntactic sugar from Section~\ref{syntacticSugar}.
+\begin{figure}[h]
+\olcomment{swapped type classes}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathbin{|\!\!>\!\!>\!\!>\!\!|})\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]),{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;())\Rightarrow {}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-\caption{Definition of \texttt{(\ensuremath{\mathbin{>\!\!>\!\!>}})}}
+\caption{Definition of \ensuremath{\mathbin{|\!\!>\!\!>\!\!>\!\!|}}.}
 \label{fig:par>>>}
@@
 \end{figure}
-Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate in a ring topology with each other.
+Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate in a ring topology with each other.
 Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets send to the neighbour of each node.
-This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ringSimple}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}(\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})){}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}\to [\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{ringSimple}\;\Varid{f}\;\Varid{is}\mathrel{=}{}\<[20]%
-\>[20]{}\Varid{os}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathbf{where}\;(\Varid{os},\Varid{ringOuts})\mathrel{=}\Varid{unzip}\;(\Varid{parMap}\;(\Varid{toRD}\mathbin{\$}\Varid{uncurry}\;\Varid{f})\;{}\<[E]%
-\\
-\>[3]{}\hsindent{33}{}\<[36]%
-\>[36]{}(\Varid{zip}\;\Varid{is}\mathbin{\$}\Varid{lazy}\;\Varid{ringIns})){}\<[E]%
-\\
-\>[3]{}\hsindent{6}{}\<[9]%
-\>[9]{}\Varid{ringIns}\mathrel{=}\Varid{rightRotate}\;\Varid{ringOuts}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{toRD}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}((\Varid{i},\Varid{r})\to (\Varid{o},\Varid{r})){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\to ((\Varid{i},\Conid{RD}\;\Varid{r})\to (\Varid{o},\Conid{RD}\;\Varid{r})){}\<[E]%
-\\
-\>[B]{}\Varid{toRD}\;{}\<[7]%
-\>[7]{}\Varid{f}\;(\Varid{i},\Varid{ringIn}){}\<[22]%
-\>[22]{}\mathrel{=}(\Varid{o},\Varid{release}\;\Varid{ringOut}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathbf{where}\;(\Varid{o},\Varid{ringOut})\mathrel{=}\Varid{f}\;(\Varid{i},\Varid{fetch}\;\Varid{ringIn}){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{rightRotate}{}\<[16]%
-\>[16]{}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{rightRotate}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}{}\<[19]%
-\>[19]{}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{rightRotate}\;\Varid{xs}\mathrel{=}{}\<[19]%
-\>[19]{}\Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{lazy}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{lazy}\mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Eden's definition of the ring skeleton \citep{eden_skel_topology}}
-\label{fig:ringEden}
-\end{figure}
+This data is sent over direct communication channels, so called \enquote{remote data}.
+We depict it in Appendix, Fig.~\ref{fig:ringEden}.
 %\end{code}
@@
 
-We can rewrite its functionality easily with the use of \ensuremath{\Varid{loop}} as the definition of the node function, \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})}, after being transformed into an arrow, already fits quite neatly into the \ensuremath{\Varid{loop}}'s \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}}.
+We can rewrite this functionality easily with the use of \ensuremath{\Varid{loop}} as the definition of the node function, \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})}, after being transformed into an arrow, already fits quite neatly into the \ensuremath{\Varid{loop}}'s \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}}.
 In each iteration we start by rotating the intermediary input from the nodes \ensuremath{[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu]} with \ensuremath{\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})}.
@@
 \begin{quote}
-{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
+{Note that the list of ring inputs \ensuremath{\Varid{ringIns}} is the same as the list of ring outputs \ensuremath{\Varid{ringOuts}} rotated by one element to the right using the auxiliary function \ensuremath{\Varid{rightRotate}}.
 Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
@@
 Plugging this arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])\;([\mskip1.5mu \Varid{o}\mskip1.5mu],\Varid{fut}\;\Varid{r})} into the definition of \ensuremath{\Varid{loop}} from earlier gives us \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
+This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
+%Further details on this can be found in \cite{eden_cefp}.
+\olcomment{let's do it?}
+
 \begin{figure}[h]
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ring}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{r},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})\;\Varid{conf})\Rightarrow {}\<[E]%
+\>[B]{}\Varid{ring}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{r},\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})\;\Varid{conf})\Rightarrow {}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{conf}\to {}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})\to {}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
+\>[5]{}\Varid{conf}\to \Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})\to \Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
 \\
@@
 \\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{put})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{arr}\;\Varid{unzip}){}\<[E]%
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{put})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{arr}\;\Varid{unzip}){}\<[E]%
 \\[\blanklineskip]%
@@
 \end{hscode}\resethooks
-\caption{Final Definition of the ring Skeleton}
+\caption{Final definition of the \ensuremath{\Varid{ring}} skeleton.}
 \label{fig:ringFinal}
 \end{figure}
- %$ %% formatting
+ % $ %% formatting
 % and lazy:
@@
 %% we have shown this already.
-This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
-%Further details on this can be found in \cite{eden_cefp}.
-\olcomment{let's do it?}
+
 \subsection{Torus skeleton}
@@
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-
-With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology}---yet again with the help of the \ensuremath{\Conid{ArrowLoop}} typeclass.
-
-Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu])} in the second argument (loop only allows for 2 arguments) of our looped arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))\;([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} and our rotation arrow becomes \ensuremath{\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))} instead of the singular rotation in the ring as we rotate \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu]} horizontally and \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]} vertically.
+With our Parallel Arrows we re-implement the torus combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the \ensuremath{\Conid{ArrowLoop}} typeclass.
+
+Similar to the \ensuremath{\Varid{ring}}, we once again start by rotating the input, but this time not only in one direction, but in two.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu])} in the second argument (loop only allows for two arguments) of our looped arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))\;([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} and our rotation arrow becomes \ensuremath{\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))} instead of the singular rotation in the ring as we rotate \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu]} horizontally and \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]} vertically.
 Then, we once again zip the inputs for the input arrow with \ensuremath{\Varid{arr}\;(\Varid{uncurry3}\;\Varid{zipWith3}\;\Varid{lazyzip3})} from \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} to \ensuremath{[\mskip1.5mu [\mskip1.5mu (\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]}, which we then feed into our parallel execution.
-\\\\
+
 This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
@@
 Finally, we unpack this matrix \ensuremath{[\mskip1.5mu [\mskip1.5mu [\mskip1.5mu (\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!\!>\!\!>}\Varid{threetotwo}} to get  \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))}.
-\\\\
+
 \begin{figure}[h]
@@
  % $
-\caption{Definition of the torus skeleton}
+\caption{Definition of the \ensuremath{\Varid{torus}} skeleton.}
 \label{fig:torus}
@@
 As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}.
-Their nodefunction can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
+Their instantiation of the skeleton \ensuremath{\Varid{nodefunction}} can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
 \begin{figure}[h]
@@
 \end{hscode}\resethooks
-\caption{Adapted nodefunction for matrix multiplication with the torus from Fig.~\ref{fig:torus}}
+\caption{Adapted \ensuremath{\Varid{nodefunction}} for matrix multiplication with the \ensuremath{\Varid{torus}} from Fig.~\ref{fig:torus}}
 \label{fig:torusMatMult}
@@
 \begin{figure}[ht]
+\olcomment{more nodes!!!}
 	\centering
@@
 	%%\pagebreak
+
 \section{Conclusion}
@@
 This work features multiple parallel backends: the already available parallel Haskell flavours.
-Parallel Arrows feature an implementation of the \ensuremath{\Conid{ArrowParallel}} interface for Multicore Haskell, \ensuremath{\Conid{Par}} monad, and Eden.
+Parallel Arrows feature an implementation of the \ensuremath{\Conid{ArrowParallel}} interface for Multicore Haskell, \ensuremath{\Conid{Par}} Monad, and Eden.
 With our approach parallel programs can be ported across these flavours with no effort.
@@
 \olcomment{Parrows + accelerate = love? Metion port to Frege. Mention the Par monad troubles.}
+
+\subsection{Future Work}
+\label{sec:future-work}
+
+Our PArrows interface can be expanded to futher parallel Haskells.
+More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} a modern distributed Haskell that would benefit from our Arrows notation.
+Future-aware special versions of Arrow combinated can be extended and further improved.
+We would look into more transparency of the API, it should basically infuse as little overhead as possible.
+
+Of course, definitions of further skeletons are viable and needed. We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
+
+Accelerate~\cite{Chakravarty:2011:AHA:1926354.1926358} is not related to our approach. It would be interesting to see a hybrid of both APIs.
+
+\olcomment{replace API with DSL globally?}
 	%\pagebreak
@@
 
+\subsection{Omitted Funtion Definitions}
+
+We have omitted some function definitions in the main text for
+brevity, and redeem this here.
+%
+We warp Eden's build-in Futures in PArrows as in Figure~\ref{fig:RDFuture}.
+
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{RemoteData}\;\Varid{a}\mathrel{=}\Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathbin{::}\Conid{RD}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{RemoteData}\;\Varid{a}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{put}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to \Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathrel{=}\Varid{release}\;\Varid{a}\mskip1.5mu\}){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{rd}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{fetch}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{\ensuremath{\Conid{RD}}-based \ensuremath{\Conid{RemoteData}} version of \ensuremath{\Conid{Future}} for the Eden backend.}
+\label{fig:RDFuture}
+\end{figure}
+
+The full definition of \ensuremath{\Varid{farmChunk}} is in Figure~\ref{fig:farmChunk}.
+
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{farmChunk}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{shuffle}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Definition of \ensuremath{\Varid{farmChunk}}.}
+\label{fig:farmChunk}
+\end{figure}
+
+Eden definition of \ensuremath{\Varid{ring}} skeleton is in Figure~\ref{fig:ringEden}.
+
+\begin{figure}[h]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{ringSimple}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{4}{}\<[4]%
+\>[4]{}(\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r}))\to [\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{ringSimple}\;\Varid{f}\;\Varid{is}\mathrel{=}{}\<[20]%
+\>[20]{}\Varid{os}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{where}\;(\Varid{os},\Varid{ringOuts})\mathrel{=}\Varid{unzip}\;(\Varid{parMap}\;(\Varid{toRD}\mathbin{\$}\Varid{uncurry}\;\Varid{f})\;(\Varid{zip}\;\Varid{is}\mathbin{\$}\Varid{lazy}\;\Varid{ringIns})){}\<[E]%
+\\
+\>[3]{}\hsindent{6}{}\<[9]%
+\>[9]{}\Varid{ringIns}\mathrel{=}\Varid{rightRotate}\;\Varid{ringOuts}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{toRD}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}((\Varid{i},\Varid{r})\to (\Varid{o},\Varid{r}))\to ((\Varid{i},\Conid{RD}\;\Varid{r})\to (\Varid{o},\Conid{RD}\;\Varid{r})){}\<[E]%
+\\
+\>[B]{}\Varid{toRD}\;{}\<[7]%
+\>[7]{}\Varid{f}\;(\Varid{i},\Varid{ringIn}){}\<[22]%
+\>[22]{}\mathrel{=}(\Varid{o},\Varid{release}\;\Varid{ringOut}){}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{where}\;(\Varid{o},\Varid{ringOut})\mathrel{=}\Varid{f}\;(\Varid{i},\Varid{fetch}\;\Varid{ringIn}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{rightRotate}{}\<[16]%
+\>[16]{}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{rightRotate}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}{}\<[19]%
+\>[19]{}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{rightRotate}\;\Varid{xs}\mathrel{=}{}\<[19]%
+\>[19]{}\Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{lazy}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\Varid{lazy}\mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Eden's definition of the \ensuremath{\Varid{ring}} skeleton.}
+\label{fig:ringEden}
+\end{figure}
+
 %%% Local Variables:

### 2017-05-19 paper/content/main.tex
formatting of topo skels

@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{put})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{unzip}){}\<[E]%
+\>[9]{}\Varid{loop}\;(\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!\!>\!\!>}{}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{put})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip}){}\<[E]%
 \\[\blanklineskip]%
@@
 	\includegraphics[scale=0.75]{images/torus}
-	\caption{Schematic depiction of the torus skeleton}
+	\caption{Schematic depiction of the \ensuremath{\Varid{torus}} skeleton.}
 	\label{fig:ringTorusImg}
@@
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-With our Parallel Arrows we re-implement the torus combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the \ensuremath{\Conid{ArrowLoop}} typeclass.
+With our Parallel Arrows we re-implement the \ensuremath{\Varid{torus}} combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the \ensuremath{\Conid{ArrowLoop}} typeclass.
 
@@
 \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{torus}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;\Varid{conf},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{a},\Conid{Future}\;\Varid{fut}\;\Varid{b})\Rightarrow {}\<[E]%
+\olcomment{swapped type classes}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{torus}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{a},\Conid{Future}\;\Varid{fut}\;\Varid{b},{}\<[E]%
+\\
+\>[B]{}\hsindent{11}{}\<[11]%
+\>[11]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;\Varid{conf})\Rightarrow {}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))\mathbin{>\!\!>\!\!>}{}\<[E]%
+\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{arr}\;\Varid{length}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unshuffle})\mathbin{\&\!\&\!\&}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{shuffle}\mathbin{>\!\!>\!\!>}\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{app}\mathbin{>\!\!>\!\!>}{}\<[E]%
+\>[9]{}(\Varid{arr}\;\Varid{length}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unshuffle})\mathbin{\&\!\&\!\&}(\Varid{shuffle}\mathbin{>\!\!>\!\!>}\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f}))\mathbin{>\!\!>\!\!>}\Varid{app}\mathbin{>\!\!>\!\!>}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b}){}\<[E]%
-\\
-\>[B]{}\Varid{ptorus}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{c},\Varid{a},\Varid{b})\to (\Varid{c},\Varid{get}\;\Varid{a},\Varid{get}\;\Varid{b}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{d},\Varid{a},\Varid{b})\to (\Varid{d},\Varid{put}\;\Varid{a},\Varid{put}\;\Varid{b})){}\<[E]%
+\>[9]{}\Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to \Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b}){}\<[E]%
+\\
+\>[B]{}\Varid{ptorus}\;\Varid{f}\mathrel{=}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{c},\Varid{a},\Varid{b})\to (\Varid{c},\Varid{get}\;\Varid{a},\Varid{get}\;\Varid{b}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{d},\Varid{a},\Varid{b})\to (\Varid{d},\Varid{put}\;\Varid{a},\Varid{put}\;\Varid{b})){}\<[E]%
 \\[\blanklineskip]%
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{nodefunction}\mathbin{::}\Conid{Int}\to {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}((\Conid{Matrix},\Conid{Matrix}),[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu])\to {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}([\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu]){}\<[E]%
-\\
-\>[B]{}\Varid{nodefunction}\;\Varid{n}\;((\Varid{bA},\Varid{bB}),\Varid{rows},\Varid{cols})\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}([\mskip1.5mu \Varid{bSum}\mskip1.5mu],\Varid{bA}\mathbin{:}\Varid{nextAs},\Varid{bB}\mathbin{:}\Varid{nextBs}){}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\mathbf{where}\;\Varid{bSum}\mathrel{=}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{foldl'}\;\Varid{matAdd}\;(\Varid{matMult}\;\Varid{bA}\;\Varid{bB})\;(\Varid{zipWith}\;\Varid{matMult}\;\Varid{nextAs}\;\Varid{nextBs}){}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{nextAs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{rows}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{nextBs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{cols}{}\<[E]%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{nodefunction}\mathbin{::}\Conid{Int}\to ((\Conid{Matrix},\Conid{Matrix}),[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu])\to ([\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu]){}\<[E]%
+\\
+\>[B]{}\Varid{nodefunction}\;\Varid{n}\;((\Varid{bA},\Varid{bB}),\Varid{rows},\Varid{cols})\mathrel{=}([\mskip1.5mu \Varid{bSum}\mskip1.5mu],\Varid{bA}\mathbin{:}\Varid{nextAs},\Varid{bB}\mathbin{:}\Varid{nextBs}){}\<[E]%
+\\
+\>[B]{}\hsindent{17}{}\<[17]%
+\>[17]{}\mathbf{where}\;\Varid{bSum}\mathrel{=}\Varid{foldl'}\;\Varid{matAdd}\;(\Varid{matMult}\;\Varid{bA}\;\Varid{bB})\;(\Varid{zipWith}\;\Varid{matMult}\;\Varid{nextAs}\;\Varid{nextBs}){}\<[E]%
+\\
+\>[17]{}\hsindent{13}{}\<[30]%
+\>[30]{}\Varid{nextAs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{rows}{}\<[E]%
+\\
+\>[17]{}\hsindent{13}{}\<[30]%
+\>[30]{}\Varid{nextBs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{cols}{}\<[E]%
 \ColumnHook

### 2017-05-19 paper/content/main.tex
build pdf

@@
 
-\usepackage[round]{natbib}
+\usepackage[round,sectionbib]{natbib}
+%% https://tex.stackexchange.com/questions/148992/changing-the-heading-style-of-references-section
+%% and jfp1.cls
+\renewcommand\refname{References}
+\makeatletter
+\renewcommand\bibsection{%
+  \section*{\refname}%
+}%
+\makeatother
 
@@
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}.
+Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}.
 For \ensuremath{\Varid{parEvalN}} this means that we can just apply the list of functions \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} to the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} by zipping them with the application operator \ensuremath{\mathbin{\$}}.
@@
 \subsection{Parallel Haskells}
-Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
+Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
 We use these languages as backends: our library can switch from one to other at user's command.
@@
 To our knowledge we are the first ones to represent parallel computation with arrows.\olcomment{that strange arrows-based robot interaction paper from 1993 or so!
-clearly discuss in related work}
+clearly discuss in related work} \done
 
@@
 
-The \ensuremath{\Varid{mapArr}} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} \cite{programming_with_arrows}:
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+The \ensuremath{\Varid{mapArr}} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} \cite{programming_with_arrows}.
+Similarly, we can also define \ensuremath{\Varid{foldlArr}} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{b},\Varid{a})\;\Varid{b}} with a neutral element \ensuremath{\Varid{b}} to \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}}.
+
+\begin{figure}[htb]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[9]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{\mid\!\mid\!\mid}(\Varid{f}\mathbin{*\!*\!*}\Varid{mapArr}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\mathbf{where}\;\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Conid{Left}\;(){}\<[E]%
-\\
-\>[9]{}\hsindent{6}{}\<[15]%
-\>[15]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{\ensuremath{\Varid{map}} over arrows}
+\\[\blanklineskip]%
+\>[B]{}\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Conid{Left}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{The definition of \ensuremath{\Varid{map}} over Arrows and the \ensuremath{\Varid{listcase}} helper
+function.}
 \label{fig:mapArr}
 \end{figure}
-Similarly, we can also define \ensuremath{\Varid{foldlArr}} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{b},\Varid{a})\;\Varid{b}} with a neutral element \ensuremath{\Varid{b}} to \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}}:
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\begin{figure}[htb]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{17}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[17]{}(\Varid{first}\;(\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{b},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{foldlArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}\Varid{app}){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\mathbf{where}\;\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Conid{Left}\;[\mskip1.5mu \mskip1.5mu]{}\<[E]%
-\\
-\>[9]{}\hsindent{6}{}\<[15]%
-\>[15]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{\ensuremath{\Varid{foldl}} over arrows}
+\ColumnHook
+\end{hscode}\resethooks
+\caption{The definition of \ensuremath{\Varid{foldl}} over Arrows.}
 \label{fig:foldlArr}
 \end{figure}
-\fixme{pipepipepipe does not work with lhs2TeX}
-Finally, with the help of \ensuremath{\Varid{mapArr}} (Fig.~\ref{fig:mapArr}), we can define \ensuremath{\Varid{zipWithArr}} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}} to an arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]}.
-\begin{figure}[h]
+%\fixme{pipepipepipe does not work with lhs2TeX}
+
+Finally, with the help of \ensuremath{\Varid{mapArr}} (Fig.~\ref{fig:mapArr}), we can define \ensuremath{\Varid{zipWithArr}} that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}} to an arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]}.
+%\begin{figure}[htb]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{\ensuremath{\Varid{zipWith}} over arrows}
-\label{fig:zipWithArr}
-\end{figure}
+% \caption{|zipWith| over arrows}
+% \label{fig:zipWithArr}
+% \end{figure}
  %$ %% formatting
@@
 It takes two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} and combines them into a new arrow \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}} which pipes all \ensuremath{\Conid{Left}\;\Varid{a}}'s to the first arrow and all \ensuremath{\Conid{Right}\;\Varid{b}}'s to the second arrow.
-\begin{figure}[h]
+% \begin{figure}[htb]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{Type signature of \pipepipepipe}
-\label{fig:codeSigPipePipePipe}
-\end{figure}
-\fixme{pipepipepipe does not work with lhs2TeX}\olcomment{trying
-  direct one}
-
-With the zipWithArr combinator we can also write a combinator \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
-\begin{figure}[h]
+% \caption{Type signature of \pipepipepipe}
+% \label{fig:codeSigPipePipePipe}
+% \end{figure}
+
+With the zipWithArr combinator we can also write a combinator \ensuremath{\Varid{listApp}}, that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
+% \begin{figure}[htb]
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-\caption{Definition of \ensuremath{\Varid{listApp}}}
-\label{fig:listApp}
-\end{figure}
+% \caption{Definition of |listApp|}
+% \label{fig:listApp}
+% \end{figure}
 % $ %% formatting
-.
 Note that  this additionally makes use of the \ensuremath{\Conid{ArrowApply}} typeclass that allows us to evaluate arrows with \ensuremath{\Varid{app}\mathbin{::}\Varid{arr}\;(\Varid{arr}\;\Varid{a}\;\Varid{b},\Varid{a})\;\Varid{c}}.
@@
 
-\subsection{Omitted Funtion Definitions}
+\section{Omitted Funtion Definitions}
 
@@
 %
-We warp Eden's build-in Futures in PArrows as in Figure~\ref{fig:RDFuture}.
-
-\begin{figure}[h]
+We warp Eden's build-in Futures in PArrows as in
+Figure~\ref{fig:RDFuture}.
+The full definition of \ensuremath{\Varid{farmChunk}} is in Figure~\ref{fig:farmChunk}.
+Eden definition of \ensuremath{\Varid{ring}} skeleton following \citep{Loogen2012} is in Figure~\ref{fig:ringEden}.
+
+\begin{figure}[htb]
 \begin{hscode}\SaveRestoreHook
@@
 
-The full definition of \ensuremath{\Varid{farmChunk}} is in Figure~\ref{fig:farmChunk}.
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
+\begin{figure}[htb]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},{}\<[E]%
+\\
+\>[B]{}\hsindent{14}{}\<[14]%
+\>[14]{}\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
 \\
@@
 
-Eden definition of \ensuremath{\Varid{ring}} skeleton is in Figure~\ref{fig:ringEden}.
-
-\begin{figure}[h]
+\begin{figure}[htb]
 \begin{hscode}\SaveRestoreHook
@@
 \column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{7}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ringSimple}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}(\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r}))\to [\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
+\>[B]{}\Varid{ringSimple}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow (\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r}))\to [\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
 \\
@@
 \\[\blanklineskip]%
-\>[B]{}\Varid{toRD}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}((\Varid{i},\Varid{r})\to (\Varid{o},\Varid{r}))\to ((\Varid{i},\Conid{RD}\;\Varid{r})\to (\Varid{o},\Conid{RD}\;\Varid{r})){}\<[E]%
+\>[B]{}\Varid{toRD}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow ((\Varid{i},\Varid{r})\to (\Varid{o},\Varid{r}))\to ((\Varid{i},\Conid{RD}\;\Varid{r})\to (\Varid{o},\Conid{RD}\;\Varid{r})){}\<[E]%
 \\

### 2017-05-22 paper/content/main.tex
renamed *.lhs files to *.tex

--- 2017-05-19 paper/content/main.tex
@@
 %% add a tweak to polycode, similar to literate in lstlistings, but uglier
+%format |>>>|        = "\mathbin{|\!\!>\!\!>\!\!>\!\!|}"
+%format parcomp        = "\mathbin{|\!\!>\!\!>\!\!>\!\!|}"
+%format >>>        = "\mathbin{>\!\!>\!\!>}"
 %% ^^^^ you want this. and similar for ***, +++, etc.
@@
 %%%%%format <--        = "\in "
+%format DOLLAR = "\mathbin{\$}"
 %%%% %format ` = "\`"
 %%%% %% works fine without
+%format ||| = "\mathbin{\mid\!\mid\!\mid}"
+%format pipepipepipe = "\mathbin{\mid\!\mid\!\mid}"
+%format |&&&| = "\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}"
+%format parand = "\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}"
+%format &&& = "\mathbin{\&\!\&\!\&}"
+%format |***| = "\mathbin{\mid\!\!*\!*\!*\!\!\mid}"
+%format parstar = "\mathbin{\mid\!\!*\!*\!*\!\!\mid}"
+%format *** = "\mathbin{*\!*\!*}"
+%format |+++| = "\mathbin{\mid\!\!+\!\!+\!\!+\!\!\mid}"
+%format +++ = "\mathbin{+\!\!+\!\!+}"
 
-%% ODER: format ==         = "\mathrel{==}"
-%% ODER: format /=         = "\neq "
-%
-%
-\makeatletter
-\@ifundefined{lhs2tex.lhs2tex.sty.read}%
-  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
-   \newcommand\SkipToFmtEnd{}%
-   \newcommand\EndFmtInput{}%
-   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
-  }\SkipToFmtEnd
-
-\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
-\usepackage{amstext}
-\usepackage{amssymb}
-\usepackage{stmaryrd}
-\DeclareFontFamily{OT1}{cmtex}{}
-\DeclareFontShape{OT1}{cmtex}{m}{n}
-  {<5><6><7><8>cmtex8
-   <9>cmtex9
-   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
-\DeclareFontShape{OT1}{cmtex}{m}{it}
-  {<-> ssub * cmtt/m/it}{}
-\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
-\DeclareFontShape{OT1}{cmtt}{bx}{n}
-  {<5><6><7><8>cmtt8
-   <9>cmbtt9
-   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
-\DeclareFontShape{OT1}{cmtex}{bx}{n}
-  {<-> ssub * cmtt/bx/n}{}
-\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU
-
-\newcommand{\Sp}{\hskip.33334em\relax}
-
-\newcommand{\Conid}[1]{\mathit{#1}}
-\newcommand{\Varid}[1]{\mathit{#1}}
-\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
-\newcommand{\plus}{\mathbin{+\!\!\!+}}
-\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
-\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
-\newcommand{\sequ}{\mathbin{>\!\!\!>}}
-\renewcommand{\leq}{\leqslant}
-\renewcommand{\geq}{\geqslant}
-\usepackage{polytable}
-
-%mathindent has to be defined
-\@ifundefined{mathindent}%
-  {\newdimen\mathindent\mathindent\leftmargini}%
-  {}%
-
-\def\resethooks{%
-  \global\let\SaveRestoreHook\empty
-  \global\let\ColumnHook\empty}
-\newcommand*{\savecolumns}[1][default]%
-  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
-\newcommand*{\restorecolumns}[1][default]%
-  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
-\newcommand*{\aligncolumn}[2]%
-  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}
-
-\resethooks
-
-\newcommand{\onelinecommentchars}{\quad-{}- }
-\newcommand{\commentbeginchars}{\enskip\{-}
-\newcommand{\commentendchars}{-\}\enskip}
-
-\newcommand{\visiblecomments}{%
-  \let\onelinecomment=\onelinecommentchars
-  \let\commentbegin=\commentbeginchars
-  \let\commentend=\commentendchars}
-
-\newcommand{\invisiblecomments}{%
-  \let\onelinecomment=\empty
-  \let\commentbegin=\empty
-  \let\commentend=\empty}
-
-\visiblecomments
-
-\newlength{\blanklineskip}
-\setlength{\blanklineskip}{0.66084ex}
-
-\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
-\let\hspre\empty
-\let\hspost\empty
-\newcommand{\NB}{\textbf{NB}}
-\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}
-
-\EndFmtInput
-\makeatother
-%
-%
-%
-%
-%
-%
-% This package provides two environments suitable to take the place
-% of hscode, called "plainhscode" and "arrayhscode". 
-%
-% The plain environment surrounds each code block by vertical space,
-% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
-% similar to formulas. Note that if these dimensions are changed,
-% the spacing around displayed math formulas changes as well.
-% All code is indented using \leftskip.
-%
-% Changed 19.08.2004 to reflect changes in colorcode. Should work with
-% CodeGroup.sty.
-%
-\ReadOnlyOnce{polycode.fmt}%
-\makeatletter
-
-\newcommand{\hsnewpar}[1]%
-  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}
-
-% can be used, for instance, to redefine the code size, by setting the
-% command to \small or something alike
-\newcommand{\hscodestyle}{}
-
-% The command \sethscode can be used to switch the code formatting
-% behaviour by mapping the hscode environment in the subst directive
-% to a new LaTeX environment.
-
-\newcommand{\sethscode}[1]%
-  {\expandafter\let\expandafter\hscode\csname #1\endcsname
-   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}
-
-% "compatibility" mode restores the non-polycode.fmt layout.
-
-\newenvironment{compathscode}%
-  {\par\noindent
-   \advance\leftskip\mathindent
-   \hscodestyle
-   \let\\=\@normalcr
-   \let\hspre\(\let\hspost\)%
-   \pboxed}%
-  {\endpboxed\)%
-   \par\noindent
-   \ignorespacesafterend}
-
-\newcommand{\compaths}{\sethscode{compathscode}}
-
-% "plain" mode is the proposed default.
-% It should now work with \centering.
-% This required some changes. The old version
-% is still available for reference as oldplainhscode.
-
-\newenvironment{plainhscode}%
-  {\hsnewpar\abovedisplayskip
-   \advance\leftskip\mathindent
-   \hscodestyle
-   \let\hspre\(\let\hspost\)%
-   \pboxed}%
-  {\endpboxed%
-   \hsnewpar\belowdisplayskip
-   \ignorespacesafterend}
-
-\newenvironment{oldplainhscode}%
-  {\hsnewpar\abovedisplayskip
-   \advance\leftskip\mathindent
-   \hscodestyle
-   \let\\=\@normalcr
-   \(\pboxed}%
-  {\endpboxed\)%
-   \hsnewpar\belowdisplayskip
-   \ignorespacesafterend}
-
-% Here, we make plainhscode the default environment.
-
-\newcommand{\plainhs}{\sethscode{plainhscode}}
-\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
-\plainhs
-
-% The arrayhscode is like plain, but makes use of polytable's
-% parray environment which disallows page breaks in code blocks.
-
-\newenvironment{arrayhscode}%
-  {\hsnewpar\abovedisplayskip
-   \advance\leftskip\mathindent
-   \hscodestyle
-   \let\\=\@normalcr
-   \(\parray}%
-  {\endparray\)%
-   \hsnewpar\belowdisplayskip
-   \ignorespacesafterend}
-
-\newcommand{\arrayhs}{\sethscode{arrayhscode}}
-
-% The mathhscode environment also makes use of polytable's parray 
-% environment. It is supposed to be used only inside math mode 
-% (I used it to typeset the type rules in my thesis).
-
-\newenvironment{mathhscode}%
-  {\parray}{\endparray}
-
-\newcommand{\mathhs}{\sethscode{mathhscode}}
-
-% texths is similar to mathhs, but works in text mode.
-
-\newenvironment{texthscode}%
-  {\(\parray}{\endparray\)}
-
-\newcommand{\texths}{\sethscode{texthscode}}
-
-% The framed environment places code in a framed box.
-
-\def\codeframewidth{\arrayrulewidth}
-\RequirePackage{calc}
-
-\newenvironment{framedhscode}%
-  {\parskip=\abovedisplayskip\par\noindent
-   \hscodestyle
-   \arrayrulewidth=\codeframewidth
-   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
-   \hline\framedhslinecorrect\\{-1.5ex}%
-   \let\endoflinesave=\\
-   \let\\=\@normalcr
-   \(\pboxed}%
-  {\endpboxed\)%
-   \framedhslinecorrect\endoflinesave{.5ex}\hline
-   \endtabular
-   \parskip=\belowdisplayskip\par\noindent
-   \ignorespacesafterend}
-
-\newcommand{\framedhslinecorrect}[2]%
-  {#1[#2]}
-
-\newcommand{\framedhs}{\sethscode{framedhscode}}
-
-% The inlinehscode environment is an experimental environment
-% that can be used to typeset displayed code inline.
-
-\newenvironment{inlinehscode}%
-  {\(\def\column##1##2{}%
-   \let\>\undefined\let\<\undefined\let\\\undefined
-   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
-   \def\fromto##1##2##3{##3}%
-   \def\nextline{}}{\) }%
-
-\newcommand{\inlinehs}{\sethscode{inlinehscode}}
-
-% The joincode environment is a separate environment that
-% can be used to surround and thereby connect multiple code
-% blocks.
-
-\newenvironment{joincode}%
-  {\let\orighscode=\hscode
-   \let\origendhscode=\endhscode
-   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
-   %\let\SaveRestoreHook=\empty
-   %\let\ColumnHook=\empty
-   %\let\resethooks=\empty
-   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
-  {\origendhscode
-   \global\let\hscode=\orighscode
-   \global\let\endhscode=\origendhscode}%
-
-\makeatother
-\EndFmtInput
-%
+%include polycode.fmt
 
@@
 
-\usepackage{tikz}
-
-\usepackage{filecontents}
-\usepackage{pgfplots, pgfplotstable}
-\usepgfplotslibrary{statistics}
-
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-
-%\usepackage{color}
-\usepackage{hyperref}
-
-%\usepackage{subcaption}
-
-\usepackage[round]{natbib}
-
-\usepackage{csquotes}
-
-% \usepackage{placeins} %% unneeded
-
-\usepackage{microtype}
-\DisableLigatures[>,<]{encoding = T1,family=tt*}
-
-% für Listings
-%\usepackage{listings}
-
-\renewcommand{\cite}[1]{\citep{#1}}
-
-\newcommand{\citHughes}{\citep{HughesArrows}}
-
-%\newcommand{\inlinecode}[1]{\texttt{#1}}
-\newcommand{\inlinecode}[1]{\emph{#1}} %%% booo
-%%% |code| is inline code in lhs2tex
-
-%\newcommand{\fixme}[1]{\colorbox{red}{#1}}
-
-%%%%%% more fun and typography
-
-\usepackage{xifthen}
-
-\newboolean{anonymous}
-\setboolean{anonymous}{False}
-
-\usepackage[british]{babel}
-
-\usepackage{siunitx}
-
-%%\usepackage{tipa} %% causes problems with lhs2tex?!
-
-\usepackage{xspace}
-\usepackage{xcolor}
-\newcommand{\hl}[1]{\textcolor{red}{#1}}
-\newcommand{\comm}[2]{\textcolor{red}{\bfseries #1: #2}}
-%\newcommand{\comm}[2]{}
-\newcommand{\olcomment}[1]{\comm{OL}{#1}}
-\newcommand{\mbcomment}[1]{\comm{MB}{#1}}
-\newcommand{\ptcomment}[1]{\comm{Phil}{#1}}
-%\newcommand{\done}{\xspace\hl{done!}\xspace}
-\newcommand{\fixme}{\mbcomment}
-
-\DeclareRobustCommand{\xth}{\textsuperscript{th}\xspace}
-\DeclareRobustCommand{\st}{\textsuperscript{st}\xspace}
-\DeclareRobustCommand{\nd}{\textsuperscript{nd}\xspace}
-\DeclareRobustCommand{\xrd}{\textsuperscript{rd}\xspace}
-
-%% kuerzel
-%\newcommand{\todo}{\textcolor{red}{\bfseries TODO!}\xspace}
-\newcommand{\done}{\textcolor{green}{\bfseries done!}\xspace}
-\DeclareRobustCommand{\hairspn}{\hspace{1pt}\nolinebreak}% hair space with no break
-
-% \DeclareRobustCommand{\ie}{{i.\hairspn{}e.\nopagebreak[4] }}
-% \DeclareRobustCommand{\eg}{{e.\hairspn{}g.\nopagebreak[4] }}
-% \DeclareRobustCommand{\fe}{{f.\hairspn{}e.,}\nopagebreak[4] }
-\DeclareRobustCommand{\ie}{{i.\hairspn{}e.~}}
-\DeclareRobustCommand{\eg}{{e.\hairspn{}g.~}}
-\DeclareRobustCommand{\fe}{{f.\hairspn{}e.,~}}
-\DeclareRobustCommand{\ad}{{A.\hairspn{}D.\xspace}}
-\DeclareRobustCommand{\bcc}{{A.\hairspn{}C.\xspace}}
-\DeclareRobustCommand{\wrt}{w.\hairspn{}r.\hairspn{}t.~}
-\DeclareRobustCommand{\etc}{etc.\ }
-\DeclareRobustCommand{\cf}{\textit{cf.~}}
-\DeclareRobustCommand{\viz}{\textit{viz.~}}
-%\DeclareRobustCommand{\hof}{higher-or\-der function\xspace}
-
-% %% the numberings
-\DeclareRobustCommand{\xth}{\textsuperscript{th}\xspace}
-\DeclareRobustCommand{\st}{\textsuperscript{st}\xspace}
-\DeclareRobustCommand{\nd}{\textsuperscript{nd}\xspace}
-\DeclareRobustCommand{\xrd}{\textsuperscript{rd}\xspace}
-
-\newcommand{\tabsepbakup}{\tabcolsep}
-%% for narrower tables
-
-%% nice tables
-\usepackage{booktabs}
-%\usepackage{multirow}
-
-%% more microtype
-\microtypecontext{spacing=nonfrench} %% log said so
-
-\usepackage{subfig}
-
-%% fight for pipepipepipe
-\newcommand{\pipepipepipe}{\ensuremath{\mathbin{\mid\!\mid\!\mid}}\xspace}
-%%% ^^^^ keep consistent with definition at the top of main.lhs
-%%%\newcommand{\pipepipepipe}{\inlinecode{|||}\xspace}
-
-%%% JFP requirements:
-%%% Harvard citing style, "(Curry 1933)".
-%%% code: identifies italic, keywords bold
-%%% figures: eps(!) or ps (can convert pdf), eventually provide a greyscale version, color NOT in cmyk
+%include prelude.tex
 
@@
 %% environment inside
-\begin{abstract}
-Arrows form\olcomment{are?} a general interface for computation and pose therefore as an alternative to monads for API design.
-We express parallelism using this concept.
-This is a new way to represent parallel computation.
-We define an Arrows-based interface for parallelism and implement it using multiple parallel Haskells.
-\olcomment{Benefits:} In this manner we are able to bridge across various parallel Haskells with a common interface.
-
-This new way of writing parallel programs has a benefit of being portable across flavours of parallel Haskells.\olcomment{Wdh?}
-Each parallel computation is an arrow, they can be composed and transformed as such.
-We thus introduce some syntactic sugar to provide parallelism-aware arrow combinators.
-
-We also define  several parallel skeletons with our framework. 
-Benchmarks shows that our framework does not induce too much overhead performance-wise.
-\end{abstract}
+%include abstract.tex
 
@@
 	%\pagebreak
-	%\section{Motivation}
-%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
-In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
-In this paper we will use this concept to express parallelism.
-
-\section{Introduction}
-\label{sec:introduction}
-\olcomment{todo, reuse 5.5, and more}
-
-blablabla arrows, parallel, haskell.
-
-\paragraph{Contribution}
-
-\olcomment{HIT HERE REALLY STRONG}
-
-%\subsection{Impact of parallel Arrows}
-%\olcomment{move this to Contributions in the front or something}
-We wrap parallel Haskells inside of our \ensuremath{\Conid{ArrowParallel}} interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
-\begin{itemize}
-	\item \textbf{Arrow API benefits}:
-	With the \ensuremath{\Conid{ArrowParallel}} typeclass we do not lose any benefits of using arrows as \ensuremath{\Varid{parEvalN}} is just yet another arrow combinator.
-The resulting arrow can be used in the same way a potential serial version could be used.
-This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
-We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
-	\item \textbf{Abstraction}:
-	With the \ensuremath{\Conid{ArrowParallel}} typeclass, we abstracted all parallel implementation logic away from the business logic.
-This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
-So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \ensuremath{\Conid{ArrowParallel}} instance.
-\end{itemize}
-
-\paragraph{Structure}
-The remaining text is structures as follows.
-Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
-Section~\ref{sec:related-work} discusses related work.
-Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface.
-Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication.
-Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel \ensuremath{\Varid{map}} with and without load balancing, \ensuremath{\Varid{map}\mathbin{-}\Varid{reduce}}) in our newly defined dialect.
-More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (\ensuremath{\Varid{pipe}}, \ensuremath{\Varid{ring}}, \ensuremath{\Varid{torus}}).
-Section~\ref{sec:benchmarks} shows the benchmark results.
-Section~\ref{sec:conclusion} discusses future work and concludes.
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
-	\section{Background}
-\label{sec:background}
-\subsection{Short introduction to parallel Haskells}
-There are already several ways to write parallel programs in Haskell.
-As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
-
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-\begin{figure}[h]
-  \centering
-	\includegraphics[scale=0.7]{images/parEvalN}
-	\caption{Schematic illustration of \texttt{parEvalN}.}
-	\label{fig:parEvalN}
-\end{figure}
-
-Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \ensuremath{\Varid{parEvalN}} with several different parallel Haskells.
-
-\subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}.
-For \ensuremath{\Varid{parEvalN}} this means that we can just apply the list of functions \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} to the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} by zipping them with the application operator \ensuremath{\mathbin{\$}}.
-% $
-We then evaluate this lazy list \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} according to a \ensuremath{\Conid{Strategy}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} with the \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} operator.
-We construct this strategy with \ensuremath{\Varid{parList}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Conid{Strategy}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{\Varid{rdeepseq}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{Strategy}\;\Varid{a}} where the latter is a strategy which evalutes to normal form.
-To ensure that programs that use \ensuremath{\Varid{parEvalN}} have the correct evaluation order, we annotate the computation with \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} which forces the compiler to not reorder multiple \ensuremath{\Varid{parEvalN}} computations.
-This is particularly necessary in circular communication topologies like in the \ensuremath{\Varid{torus}} or \ensuremath{\Varid{ring}} skeleton that we will see in Section~\ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \ensuremath{\Varid{pseq}} during testing for this paper.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\mathbf{let}\;\Varid{bs}\mathrel{=}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as}{}\<[E]%
-\\
-\>[B]{}\hsindent{18}{}\<[18]%
-\>[18]{}\mathbf{in}\;(\Varid{bs}\mathbin{`\Varid{using}`}\Varid{parList}\;\Varid{rdeepseq})\mathbin{`\Varid{pseq}`}\Varid{bs}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
- %$
-
-\begin{figure}[h]
-	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-	\caption{Dataflow of the Multicore Haskell parEvalN version}
-\olcomment{Evaluation step explicitly shown?}
-\end{figure} %$ %% formatting
-
-\subsubsection{ParMonad}
-The \ensuremath{\Conid{Par}} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
-
-Our parallel evaluation function \ensuremath{\Varid{parEvalN}} can be defined by zipping the list of \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} with the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} with the application operator \ensuremath{\mathbin{\$}} just like with Multicore Haskell.
-% $
-Then, we map over this not yet evaluated lazy list of results \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} to transform them to a list of not yet evaluated forked away computations \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we convert to \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{sequenceA}}.
-We wait for the computations to finish by mapping over the \ensuremath{\Conid{IVar}\;\Varid{b}} values inside the \ensuremath{\Conid{Par}} monad with \ensuremath{\Varid{get}}.
-This results in \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
-We finally execute this process with \ensuremath{\Varid{runPar}} to finally get \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} again.
-
-\mbcomment{explain problems with laziness here. Problems with torus}
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{runPar}\mathbin{\$}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{sequenceA}\mathbin{\$}\Varid{map}\;(\Varid{spawnP})\mathbin{\$}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as})\bind \Varid{mapM}\;\Varid{get}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\begin{figure}[h]
-	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-	\caption{Dataflow of the Par Monad parEvalN version}
-\end{figure}
-
-\subsubsection{Eden}
-Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{See also \url{http://www.mathematik.uni-marburg.de/~eden/} and \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell flavour.
-
-Eden was designed to work on clusters, but with a further simple backend it operates on multicores.
-However, in contrast to many other parallel Haskells, in Eden each process has its own heap.
-This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results also on multicores \cite{arcs-dc,aswad2009low}.
-
-While Eden also comes with a monad \ensuremath{\Conid{PA}} for parallel evaluation, it also ships with a completely functional interface that includes
-%\\
-a \ensuremath{\Varid{spawnF}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}
-%\\
-%a |spawnF| 
-function that
-%This 
-allows us to define \ensuremath{\Varid{parEvalN}} directly:
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{parEvalN}\mathrel{=}\Varid{spawnF}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\begin{figure}[h]
-	\includegraphics[scale=0.5]{images/parEvalNEden}
-	\caption{Dataflow of the Eden parEvalN version}
-\end{figure}
-
-\paragraph{Eden TraceViewer.}
-To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
-While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}.
-In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.
-
-In a trace, the $x$ axis shows the time, the $y$ axis enumerates the machines and processes.
-A~trace shows a running process in green, a blocked process is red.
-If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow.
-The typical reason for then is GC.
-An inactive machine where no processes are started yet, or all are already terminated, is shows as a blue bar.
-A~comminication from one process to another is represented with a black arrow.
-A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.
-
-\olcomment{show example trace or refer to a trace in later figures}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
-	\subsection{Arrows}
-\label{sec:arrows}
-\begin{figure}[h]
-	\includegraphics{images/arrow}
-	\caption{Schematic depiction of an arrow}
-\end{figure}
-Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
-An arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} represents  a computation that converts an input \ensuremath{\Varid{a}} to an output \ensuremath{\Varid{b}}.
-This is defined in the arrow typeclass:
-
-\begin{figure}[h]
-\centering
-%\subfloat[Arrow class definition][]{%
-% begin{minipage}{0.5\textwidth}
-\parbox{0.49\linewidth}{%
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{arr}\mathbin{::}(\Varid{a}\to \Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{b}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}(\mathbin{>\!\!>\!\!>})\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{first}\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{c}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Arrow class definition}
-}\qquad
-\begin{minipage}{0.49\linewidth}%
-%\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
-%	\begin{center}
-\centering
-	{\includegraphics[scale=0.6]{images/arr}}
-	{\includegraphics[scale=0.6]{images/compose}}
-	{\includegraphics[scale=0.6]{images/first}}
-%	\end{center}
-%}%\hfill
-\caption{The schematic depiction of \ensuremath{\Conid{Arrow}} combinators}
-\end{minipage}
-\end{figure}
-\ensuremath{\Varid{arr}} is used to lift an ordinary function to an arrow type, similarly to the monadic \ensuremath{\Varid{return}}.
-The \ensuremath{\mathbin{>\!\!>\!\!>}} operator is analogous to the monadic composition  \ensuremath{\bind } and combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}}.
-Lastly, the \ensuremath{\Varid{first}} operator  takes the input arrow from \ensuremath{\Varid{b}} to \ensuremath{\Varid{c}} and converts it into an arrow on pairs with the second argument untouched.
-It allows us to to save input across arrows.
-\\\\
-The most prominent instances of this interface are regular functions \ensuremath{(\to )}: %(Fig.~\ref{fig:arrowfn}),
-%\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;\Conid{Arrow}\;(\to )\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{f}\mathrel{=}\Varid{f}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{g}\mathrel{=}\Varid{g}\mathbin{\circ}\Varid{f}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{first}\;\Varid{f}\mathrel{=}\lambda (\Varid{a},\Varid{c})\to (\Varid{f}\;\Varid{a},\Varid{c}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{Arrow instance for regular functions}
-% \label{fig:arrowfn}
-% \end{figure}
-%%% on the verge of acceptable code
-and the Kleisli type: %(Fig.~\ref{fig:arrowkleisli}).
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{Kleisli}\;\Varid{m}\;\Varid{a}\;\Varid{b}\mathrel{=}\Conid{Kleisli}\;\{\mskip1.5mu \Varid{run}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{b}\mskip1.5mu\}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Varid{m}\Rightarrow \Conid{Arrow}\;(\Conid{Kleisli}\;\Varid{m})\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{f}\mathrel{=}\Conid{Kleisli}\;(\Varid{return}\mathbin{\circ}\Varid{f}){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{g}\mathrel{=}\Conid{Kleisli}\;(\lambda \Varid{a}\to \Varid{f}\;\Varid{a}\bind \Varid{g}){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{first}\;\Varid{f}\mathrel{=}\Conid{Kleisli}\;(\lambda (\Varid{a},\Varid{c})\to \Varid{f}\;\Varid{a}\bind \lambda \Varid{b}\to \Varid{return}\;(\Varid{b},\Varid{c})){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%$
-% \caption{Definition of the Kleisli type and the corresponding arrow instance}
-% \label{fig:arrowkleisli}
-% \end{figure}
-\begin{figure}[h]
-	\centering
-	\begin{tabular}{cc}
-		% \subcaptionbox
-{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
-		% \subcaptionbox
-{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
-\ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
-\midrule
-		% \subcaptionbox
-{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
-		% \subcaptionbox
-{}{\includegraphics[width = 1.5in]{images/andandand}}\\
-\ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\
-	\end{tabular}
-	\caption{Visual depiction of syntactic sugar for Arrows.}
-	\label{3}
-\end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: the functions \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
-The mirrored version of \ensuremath{\Varid{first}}, called \ensuremath{\Varid{second}} (Fig.~\ref{fig:secondImg}% ,~\ref{fig:second}
-) is:
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{second}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{c},\Varid{a})\;(\Varid{c},\Varid{b}){}\<[E]%
-\\
-\>[B]{}\Varid{second}\;\Varid{f}\mathrel{=}\Varid{arr}\;\Varid{swap}\mathbin{>\!\!>\!\!>}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{swap}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\mathbf{where}\;\Varid{swap}\;(\Varid{x},\Varid{y})\mathrel{=}(\Varid{y},\Varid{x}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{The second combinator}
-% \label{fig:second}
-% \end{figure}
-the \ensuremath{\mathbin{*\!*\!*}} combinator that combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, (Fig.\ref{fig:***Img}%
-% ,~\ref{fig:***}
-) is defined as
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{*\!*\!*})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
-\\
-\>[B]{}\Varid{f}\mathbin{*\!*\!*}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{g}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{The (***) combinator}
-% \label{fig:***}
-% \end{figure}
-and the \ensuremath{\mathbin{\&\!\&\!\&}} combinator that constructs an arrow which outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input (Fig.~\ref{fig:&&&Img}% ,~\ref{fig:&&&}
-) is:
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{\&\!\&\!\&})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{a}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
-\\
-\>[B]{}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}(\Varid{f}\mathbin{*\!*\!*}\Varid{g}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{The (\&\&\&) combinator}
-% \label{fig:&&&}
-% \end{figure}
-A short example given by Hughes on how to use this is 
-addition with arrows:
-%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{add}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}{}\<[E]%
-\\
-\>[B]{}\Varid{add}\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g})\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\lambda (\Varid{u},\Varid{v})\to \Varid{u}\mathbin{+}\Varid{v}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{Add over arrows}
-% \label{fig:addArrows}
-% \end{figure}
-% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator.
-Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
-Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
-% \\\\
-% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used.
-We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}
-%
-
-The more restrictive interface of arrows (a monad can be \emph{anything}, an arrow is a process of doing something, a \emph{computation}) allows for more elaborate composition and transformation combinators.
-One of the major problems in parallel computing is composition of parallel processes.
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include motivation.tex
+	%include parallelHaskells.tex
+	%include arrows.tex
 	%\pagebreak
-	\section{Related Work}
-\label{sec:related-work}
-
-\olcomment{arrows or Arrows?}
-
-\subsection{Parallel Haskells}
-Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
-We use these languages as backends: our library can switch from one to other at user's command.
-
-HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} monad to heterogeneous clusters.
-LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \ensuremath{\Conid{Par}} monad.
-%
-Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015} and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}.
-Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden.
-This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.
-
-More different approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582,}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
-%
-The Haskell--GPU bridge Accelerate \cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313} deserves a special mention.
-Accelerate is completely orthogonal to our approach.
-\citeauthor{marlow2013parallel} authored a recent book \citeyear{marlow2013parallel} on parallel Haskells.
-
-\subsection{Algorithmic skeletons}
-
-Algorithmic skeletons were introduced by \citet{Cole1989}.
-Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
-The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,lobachev-phd,Lobachev2012,janjic2013space}, iteration skeletons \cite{Dieterle2013}.
-The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel.
-\citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeleton to stable process networks.
-
-\subsection{Arrows}
-
-Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\ensuremath{\to }.
-\citet{Hughes2005} is a tutorial on arrows.
-Some theoretical details on arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable.
-\citet{Paterson:2001:NNA:507669.507664} introduced a new notation for arrows.
-Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}.
-But perhaps most prominent application of arrows is functional reactive programming \cite{Hudak2003}.\olcomment{cite more!}
-
-\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of arrows that capsule the computation more than regular arrows do and thus enable optimizations.
-Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution.
-In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
-\citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach.
-They basically use arrows to orchestrate several tasks in robotics.
-We propose a general interface for parallel programming, remaining completely in Haskell.
-
-\subsection{Other languages}
-
-Although this work is centred on Haskell implementation of arrows, it is applicable to any functional programming language where parallel evaluation and arrows can be defined.
-Our experiments with our approach in Frege language (which is basically Haskell on the JVM) were quite successful, we were able to use typical Java libraries for parallelism.
-However, it is beyond the scope of this work.
-
-\citet{achten2004arrows,achten2007arrow} use an arrow implementation in Clean for better handling of typical GUI tasks.
-\citet{Dagand:2009:ORD:1481861.1481870} used arrows in OCaml in the implementation of a distributed system.
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include relwork.tex
 	%\pagebreak
 	%\pagebreak
-	\section{Parallel Arrows}
-\label{sec:parallel-arrows}
-We have seen what Arrows are and how they can be used as a general interface to computation.
-In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well.
-We start by introducing the interface and explaining the reasonings behind it.
-Then, we discuss some implementations using exisiting parallel Haskells.
-Finally, we explain why using Arrows for expressing parallelism is beneficial.
-\subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, \ensuremath{\Varid{parEvalN}} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
-Translating this into arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
-% (Fig.~\ref{fig:parEvalNArrowFn}) (
-This combinator is similar to our utility function \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{parEvalN Arrow combinator as a function}
-% \label{fig:parEvalNArrowFn}
-% \end{figure}
-With this definition of \ensuremath{\Varid{parEvalN}}, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow \ensuremath{\Varid{arr}} and we want this to be an interface for different backends, we introduce a new typeclass \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}} to host this combinator: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
-% \label{fig:parEvalNArrowTypeClass1}
-% \end{figure}
-Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs.
-normalform).
-For this reason we also introduce an additional \ensuremath{\Varid{conf}} parameter to the function.
-We also do not want \ensuremath{\Varid{conf}} to be a fixed type, as the configuration parameters can differ for different instances of \ensuremath{\Conid{ArrowParallel}}.
-So we add it to the type signature of the typeclass as well and get \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}}: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
-% \begin{figure}[h]
-\olcomment{\ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}} or \ensuremath{\Conid{ArrowParallel}\;\Varid{conf}\;\Varid{arr}\;\Varid{a}\;\Varid{b}}?}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
-% \label{fig:parEvalNArrowTypeClassFinal}
-% \end{figure}
-Note that we don't require the \ensuremath{\Varid{conf}} parameter in every implementation.
-If it is not needed, we usually just default the \ensuremath{\Varid{conf}} type parameter to \ensuremath{()} and even blank it out in the parameter list of the implemented \ensuremath{\Varid{parEvalN}}, as we will see in the implementation of the Multicore and the \ensuremath{\Conid{Par}} Monad backend.
-
-\subsection{ArrowParallel instances}
-
-\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of this class is implemented in a straightforward manner by using \ensuremath{\Varid{listApp}} from Appendix~\ref{utilfns} combined with the \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} and \ensuremath{\Varid{pseq}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}} combinators from Multicore Haskell, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} but with flipped parameters.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{rdeepseq}))\mathbin{\&\!\&\!\&}\Varid{arr}\;\Varid{id}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{pseq}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% $ %% formatting
-\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
-\label{fig:ArrowParallelMulticoreRdeepseq}
-\end{figure}
-For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \ensuremath{\Conid{Strategy}} interface, we want to the user not to lose this ability because of using our API.
-We therefore introduce the \ensuremath{\Conid{Conf}\;\Varid{a}} data-type that simply wraps a \ensuremath{\Conid{Strategy}\;\Varid{a}} (Fig.~\ref{fig:confa}).
-We can't directly use the \ensuremath{\Conid{Strategy}\;\Varid{a}} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances:
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{Conf}\;\Varid{a}\mathrel{=}\Conid{Conf}\;(\Conid{Strategy}\;\Varid{a}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{Definition of Conf a}
-% \label{fig:confa}
-% \end{figure}
-To get our configurable \ensuremath{\Conid{ArrowParallel}} instance, we simply unwrap the strategy and pass it to \ensuremath{\Varid{parList}} like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;(\Conid{Conf}\;\Varid{strat})\;\Varid{fs}\mathrel{=}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{listApp}\;\Varid{fs}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{strat}))\mathbin{\&\!\&\!\&}\Varid{arr}\;\Varid{id}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{pseq}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
-\label{fig:ArrowParallelMulticoreConfigurable}
-\end{figure}
-\subsubsection{\ensuremath{\Conid{Par}} Monad}
-\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting}
-The \ensuremath{\Conid{Par}} monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and \ensuremath{\Conid{Par}} monad's \ensuremath{\Varid{spawnP}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} function.
-The latter forks away the computation of a value and returns an \ensuremath{\Conid{IVar}} containing the result in the \ensuremath{\Conid{Par}} monad.
-
-We therefore apply each function to its corresponding input value with \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}) and then fork the computation away with \ensuremath{\Varid{arr}\;\Varid{spawnP}} inside a \ensuremath{\Varid{zipWithArr}} call.
-This yields a list \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we then convert into \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}.
-In order to wait for the computation to finish, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}.
-The result of this operation is a \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} from which we can finally remove the monad again by running \ensuremath{\Varid{arr}\;\Varid{runPar}} to get our output of \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{zipWithArr}\;(\Varid{app}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{spawnP})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
- %$ %% formatting
-\caption{ArrowParallel instance for the Par Monad backend}
-\label{fig:ArrowParallelParMonad}
-\end{figure}
-
-\subsubsection{Eden}
-For both the Multicore Haskell and \ensuremath{\Conid{Par}} Monad implementations we could use general instances of \ensuremath{\Conid{ArrowParallel}} that just require the \ensuremath{\Conid{ArrowApply}} and \ensuremath{\Conid{ArrowChoice}} typeclasses.
-With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Conid{ArrowUnwrap}\;\Varid{arr}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to (\Varid{a}\to \Varid{b}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{possible ArrowUnwrap typeclass}
-% \label{fig:ArrowUnwrap}
-% \end{figure}
-We don't do it here for aesthetic resons, though. For now, we just implement \ensuremath{\Conid{ArrowParallel}} for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow \Conid{ArrowParallel}\;(\to )\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{spawnF}\;\Varid{fs}\;\Varid{as}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{ArrowParallel instance for functions in the Eden backend}
-% \label{fig:ArrowParallelEdenFns}
-% \end{figure}
-and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;(\Conid{Monad}\;\Varid{m},\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b},\Conid{Trans}\;(\Varid{m}\;\Varid{b}))\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;(\Conid{Kleisli}\;\Varid{m})\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{arr}\mathbin{\$}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{map}\;(\lambda (\Conid{Kleisli}\;\Varid{f})\to \Varid{f})\;\Varid{fs}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Conid{Kleisli}\mathbin{\$}\Varid{sequence}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
-% \label{fig:ArrowParallelEdenKleisli}
-% \end{figure}
-
-%\FloatBarrier
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include parrows.tex
 	%\pagebreak
-	\subsection{Extending the Interface}
-\label{sec:extending-interface}
-With the \ensuremath{\Conid{ArrowParallel}} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
-These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
-
-\subsubsection{Lazy \ensuremath{\Varid{parEvalN}}}
-\begin{figure}[h]
-	\includegraphics[scale=0.7]{images/parEvalNLazy}
-	\caption{Schematic depiction of parEvalNLazy}
-	\label{fig:parEvalNLazyImg}
-\end{figure}
-The function \ensuremath{\Varid{parEvalN}} is 100\% strict, which means that it fully evaluates all passed arrows.
-Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
-\ensuremath{\Varid{map}\;(\Varid{arr}\mathbin{\circ}(\mathbin{+}))\;[\mskip1.5mu \mathrm{1}\mathinner{\ldotp\ldotp}\mskip1.5mu]} or just because we need the arrows evaluated in chunks.
-\ensuremath{\Varid{parEvalNLazy}} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
-It works by first chunking the input from \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} to \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]} with the given \ensuremath{\Conid{ChunkSize}} in \ensuremath{\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})}.
-These chunks are then fed into a list \ensuremath{[\mskip1.5mu \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} of parallel arrows created by feeding chunks of the passed \ensuremath{\Conid{ChunkSize}} into the regular parEvalN by using \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}).
-The resulting \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} is lastly converted into \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{concat}}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEvalNLazy}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
-\\
-\>[B]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{fs}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{listApp}\;\Varid{fchunks}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{concat}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\mathbf{where}\;\Varid{fchunks}\mathrel{=}\Varid{map}\;(\Varid{parEvalN}\;\Varid{conf})\mathbin{\$}\Varid{chunksOf}\;\Varid{chunkSize}\;\Varid{fs}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
- %$ %% formatting
-\caption{Definition of \ensuremath{\Varid{parEvalNLazy}}.}
-\label{fig:parEvalNLazy}
-\end{figure}
-
-\subsubsection{Heterogenous tasks}
-\begin{figure}[h]
-	\includegraphics[scale=0.7]{images/parEval2}
-	\caption{Schematic depiction of \ensuremath{\Varid{parEval2}}.}
-	\label{fig:parEval2Img}
-\end{figure}
-We have only talked about the paralellization arrows of the same type until now.
-But sometimes we want to paralellize heterogenous types as well.
-However, we can implement such a \ensuremath{\Varid{parEval2}} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into a new parallel arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} quite easily with the help of the \ensuremath{\Conid{ArrowChoice}} typeclass.
-The idea is to use the \ensuremath{\mathbin{+\!\!+\!\!+}} combinator which combines two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} and transforms them into \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})} to get a common arrow type that we can then feed into parEvalN.
-\\\\
-We start by transforming the \ensuremath{(\Varid{a},\Varid{c})} input into a 2-element list \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{a}\;\Varid{c}\mskip1.5mu]} by first tagging the two inputs with \ensuremath{\Conid{Left}} and \ensuremath{\Conid{Right}} and wrapping the right element in a singleton list with \ensuremath{\Varid{return}} so that we can combine them with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))}.
-Next, we feed this list into a parallel arrow running on 2 instances of \ensuremath{\Varid{f}\mathbin{+\!\!+\!\!+}\Varid{g}} as described above.
-After the calculation is finished, we convert the resulting \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{b}\;\Varid{d}\mskip1.5mu]} into \ensuremath{([\mskip1.5mu \Varid{b}\mskip1.5mu],[\mskip1.5mu \Varid{d}\mskip1.5mu])} with \ensuremath{\Varid{arr}\;\Varid{partitionEithers}}.
-The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \ensuremath{(\Varid{b},\Varid{d})} in the last step.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEval2}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;\Varid{conf})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
-\\
-\>[B]{}\Varid{parEval2}\;\Varid{conf}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Conid{Left}\mathbin{*\!*\!*}(\Varid{arr}\;\Conid{Right}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{return})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{f}\mathbin{+\!\!+\!\!+}\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{partitionEithers}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{head}\mathbin{*\!*\!*}\Varid{arr}\;\Varid{head}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-	\caption{Definition of parEval2}
-	\label{fig:parEval2}
-\end{figure}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include basicSkeletons.tex
 	%\pagebreak
-	\subsubsection{Syntactic Sugar} \label{syntacticSugar}
-For basic arrows, we have the \ensuremath{\mathbin{*\!*\!*}} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into an arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} which does both computations at once.
-This can easily be translated into a parallel version \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} (Fig.~\ref{fig:par***}) with the use of \ensuremath{\Varid{parEval2}}, but for this we require a backend which has an implementation that does not require any configuration (hence the \ensuremath{()} as the \ensuremath{\Varid{conf}} parameter% in Fig.~\ref{fig:par***}
-):
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{\mid\!\!*\!*\!*\!\!\mid})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;()))\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
-\\
-\>[B]{}(\mathbin{\mid\!\!*\!*\!*\!\!\mid})\mathrel{=}\Varid{parEval2}\;(){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{Definition of |parstar|, the parallel version of |***|.}
-% \label{fig:par***}
-% \end{figure}
-% With this we can analogously to the serial |&&&|
-We define the parallel \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}} % (Fig.~\ref{fig:par&&&}) 
-in a similar manner to its sequential pendant \ensuremath{\mathbin{\&\!\&\!\&}} (Fig.~\ref{fig:&&&Img}% ,~\ref{fig:&&&}
-):
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{a})\;(\Conid{Either}\;\Varid{b}\;\Varid{c})\;())\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
-\\
-\>[B]{}(\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{\mid\!\!*\!*\!*\!\!\mid}\Varid{g}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
- % $ %% formatting
-% \caption{Definition of |parand| - the parallel version of |&&&|.}
-% \label{fig:par&&&}
-% \end{figure}
+	%include syntacticSugar.tex
 	%\pagebreak
-	\section{Futures} \label{futures}
-Consider the parallel arrow combinator in Fig.~\ref{fig:someCombinator}
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}\Varid{parEvalN}\;()\;\Varid{fs1}\mathbin{>\!\!>\!\!>}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{parEvalN}\;()\;\Varid{fs2}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{An example parallel Arrow combinator without Futures}
-\label{fig:someCombinator}
-\end{figure}
-In a distributed environment, the resulting arrow of this combinator first evaluates all \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]} in parallel to then gather the input once again on the master node.
-Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
-
-While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \ensuremath{\Varid{parEvalN}} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \ensuremath{\Conid{ArrowParallel}} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
-This can become a serious bottleneck %in heavy threaded applications.
-for larger amount of data and number of processes \citep[showcases][as, \eg]{Berthold2009-fft}.
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
-	\caption[without Futures]{Communication between 4 threads without Futures.
-All communication goes through the master node.
-Each bar represents one process.
-Black lines between processes represent communication.
-Colors: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
-	\label{fig:withoutFutures}
-\olcomment{more practical and heavy-weight example! fft (I have the code)?}
-\end{figure}
-
-This motivates for an approach that allows the nodes to communicate directly with each other.
-Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \ensuremath{\Conid{RD}} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
-
-But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \ensuremath{\Conid{Future}} typeclass (Fig.~\ref{fig:future}).
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{Future}\;\Varid{fut}\;\Varid{a}\mathbin{@|@}\Varid{a}\to \Varid{fut}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{put}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{a}){}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{get}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Varid{fut}\;\Varid{a})\;\Varid{a}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of the Future typeclass}
-\label{fig:future}
-\end{figure}
-Since \ensuremath{\Conid{RD}} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Appendix in Fig.~\ref{fig:RDFuture}.
-This is due to the same reason we had to introduce a wrapper for \ensuremath{\Conid{Strategy}\;\Varid{a}} in the Multicore Haskell implementation of \ensuremath{\Conid{ArrowParallel}} in Section~\ref{sec:parrows:multicore}.
-
-For our Par Monad and Multicore Haskell backends, we can simply use \ensuremath{\Conid{MVar}}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
-\fixme{explain MVars}
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mbox{\enskip\{-\# NOINLINE putUnsafe  \#-\}\enskip}{}\<[E]%
-\\
-\>[B]{}\Varid{putUnsafe}\mathbin{::}\Varid{a}\to \Conid{MVar}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\Varid{putUnsafe}\;\Varid{a}\mathrel{=}\Varid{unsafePerformIO}\mathbin{\$}\mathbf{do}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{mVar}\leftarrow \Varid{newEmptyMVar}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{putMVar}\;\Varid{mVar}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{return}\;\Varid{mVar}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{MVar}\;\Varid{a}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{put}\mathrel{=}\Varid{arr}\;\Varid{putUnsafe}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{takeMVar}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unsafePerformIO}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends}
-\label{fig:MVarFuture}
-\end{figure} % $
-
-Furthermore, in order for these \ensuremath{\Conid{Future}} types to fit with the \ensuremath{\Conid{ArrowParallel}} instances we gave earlier, we have to give the necessary \ensuremath{\Conid{NFData}} and \ensuremath{\Conid{Trans}} instances, the latter are only needed in Eden.
-Because \ensuremath{\Conid{MVar}} already ships with a \ensuremath{\Conid{NFData}} instance, we only have to supply two simple instances for our \ensuremath{\Conid{RemoteData}} type:
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{instance}\;\Conid{NFData}\;(\Conid{RemoteData}\;\Varid{a})\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{rnf}\mathrel{=}\Varid{rnf}\mathbin{\circ}\Varid{rd}{}\<[E]%
-\\
-\>[B]{}\mathbf{instance}\;\Conid{Trans}\;(\Conid{RemoteData}\;\Varid{a}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{NFData and Trans instances for the RemoteData type.
-The Trans instance does not have any functions declared as the default implementation suffices here.
-See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
-% \end{figure}
-The \ensuremath{\Conid{Trans}} instance does not have any functions declared as the default implementation suffices here.
-
-Going back to our communication example we can use this \ensuremath{\Conid{Future}} concept in order to enable direct communications between the nodes in the following way:
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{put})\;\Varid{fs1})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\Varid{get}\mathbin{>\!\!>\!\!>})\;\Varid{fs2}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel}
-\label{fig:someCombinatorParallel}
-\end{figure}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear.
-Do Figs 2-3 with a lot of load?}
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.9\textwidth]{images/withFutures}
-	\caption[with Futures]{Communication between 4 threads with Futures.
-Other than in Fig.~\ref{fig:withoutFutures}, threads communicate directly (black lines between the bars) instead of always going through the master node (bottom bar).}
-	\label{fig:withFutures}
-\end{figure}
+	%include futures.tex
 	%\pagebreak
-
-%%% \FloatBarrier
-\section{Map-based Skeletons}
-\label{sec:map-skeletons}
-Now we have developed Parallel Arrows far enough to define some algorithmic skeletons useful to an application programmer.
-\subsection{Parallel map}
-\begin{figure}[h]
-	\includegraphics[scale=0.7]{images/parMap}
-	\caption{Schematic depiction of \ensuremath{\Varid{parMap}}.}
-	\label{fig:parMapImg}
-\end{figure}
-The \ensuremath{\Varid{parMap}} skeleton (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
-We can implement it with \ensuremath{\Conid{ArrowParallel}} by repeating an arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and then passing it into \ensuremath{\Varid{parEvalN}} to get an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
-Just like \ensuremath{\Varid{parEvalN}}, \ensuremath{\Varid{parMap}} is 100\% strict.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parMap}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf})\Rightarrow \Varid{conf}\to (\Varid{arr}\;\Varid{a}\;\Varid{b})\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
-\\
-\>[B]{}\Varid{parMap}\;\Varid{conf}\;\Varid{f}\mathrel{=}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;\Varid{f}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of parMap}
-\label{fig:parMap}
-\end{figure}
-
-\subsection{Lazy parallel map}
-\begin{figure}[h]
-	\includegraphics[scale=0.7]{images/parMapStream}
-	\caption{Schematic depiction of parMapStream}
-	\label{fig:parMapStreamImg}
-\end{figure}
-As \ensuremath{\Varid{parMap}} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \ensuremath{\Varid{parEvalN}} compared to \ensuremath{\Varid{parEvalNLazy}}.
-So it makes sense to also have a \ensuremath{\Varid{parMapStream}} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \ensuremath{\Varid{parMap}}, but uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parMapStream}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{parMapStream}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{f}\mathrel{=}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;\Varid{f}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of \ensuremath{\Varid{parMapStream}}.}
-\label{fig:parMapStream}
-\end{figure}
-
-\subsection{Statically load-balancing parallel map}
-\begin{figure}[h]
-	\includegraphics[scale=0.7]{images/farm}
-	\caption{Schematic depiction of a \ensuremath{\Varid{farm}}, a statically
-          load-balanced \ensuremath{\Varid{parMap}}.}
-	\label{fig:farmImg}
-\end{figure}
-A \ensuremath{\Varid{parMap}} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \ensuremath{\Conid{ArrowParallel}} we gave in this paper).
-This can be quite wasteful and a \ensuremath{\Varid{farm}} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \ensuremath{\Varid{numCores}} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
-The definitions of helper functions \ensuremath{\Varid{unshuffle}}, \ensuremath{\Varid{takeEach}}, \ensuremath{\Varid{shuffle}} (shown in Appendix) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{farm}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{farm}\;\Varid{conf}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{shuffle}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{unshuffle}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Conid{Int}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{unshuffle}\;\Varid{n}\mathrel{=}\Varid{arr}\;(\lambda \Varid{xs}\to [\mskip1.5mu \Varid{takeEach}\;\Varid{n}\;(\Varid{drop}\;\Varid{i}\;\Varid{xs})\mathbin{@|@}\Varid{i}\leftarrow [\mskip1.5mu \mathrm{0}\mathinner{\ldotp\ldotp}\Varid{n}\mathbin{-}\mathrm{1}\mskip1.5mu]\mskip1.5mu]){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{takeEach}\mathbin{::}\Conid{Int}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{takeEach}\;\Varid{n}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{takeEach}\;\Varid{n}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{takeEach}\;\Varid{n}\;(\Varid{drop}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{xs}){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{shuffle}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{shuffle}\mathrel{=}\Varid{arr}\;(\Varid{concat}\mathbin{\circ}\Varid{transpose}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{The definition of \ensuremath{\Varid{farm}}.}
-\label{fig:farm}
-\end{figure}
-
-\subsection{The \ensuremath{\Varid{farmChunk}} Skeleton}
-\begin{figure}[h]
-	\includegraphics[scale=0.7]{images/farmChunk}
-\caption{Schematic depiction of farmChunk}
-\label{fig:farmChunkImg}
-\end{figure}
-Since a \ensuremath{\Varid{farm}} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \ensuremath{\Varid{parMap}} with a different work distribution, it is, again, 100\% strict.
-So we can define \ensuremath{\Varid{farmChunk}} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}.
-It is basically the same definition as for \ensuremath{\Varid{farm}}, with \ensuremath{\Varid{parEvalN}} replaced with \ensuremath{\Varid{parEvalNLazy}}, as Appendix shows.
-
-\subsection{Map and reduce}
-
-A~simple \ensuremath{\Varid{map}}--\ensuremath{\Varid{reduce}} can be written like in Figure~\ref{fig:parMapReduceDirect}.
-Notice that the performance of the \ensuremath{\mathbin{>\!\!>\!\!>}} combinator is essential for the performance of the skeleton.
-A~definitive version would use Futures.
-
- \olcomment{it appears STRANGE. are the data really left alone and noded after map and taken from there by reduce? It makes sense only is no communication through master takes place. ELSE: CUT!}
-
- \mbcomment{this requires some work. Either change this to use futures or cut, yes.}
-% -- this does not completely adhere to Google's definition of Map Reduce as it
-% -- the mapping function does not allow for "reordering" of the output
-% -- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
-
-\olcomment{now rewritten as motivation for futures. maybe still cut?}
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parMapReduceDirect}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}\;\Varid{conf},\Conid{ArrowApply}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{b},\Varid{b})\;\Varid{b}\to \Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}{}\<[E]%
-\\
-\>[B]{}\Varid{parMapReduceDirect}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{mapfn}\;\Varid{foldfn}\;\Varid{neutral}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{mapArr}\;\Varid{mapfn}\mathbin{>\!\!>\!\!>}\Varid{foldlArr}\;\Varid{foldfn}\;\Varid{neutral})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{foldlArr}\;\Varid{foldfn}\;\Varid{neutral}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of parMapReduceDirect}
-\label{fig:parMapReduceDirect}
-\end{figure}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include mapSkeletons.tex
 	%\pagebreak
-	%\FloatBarrier
-\section{Topological Skeletons}
-\label{sec:topology-skeletons}
-Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} with more predefined skeletons, among them a \ensuremath{\Varid{pipe}}, a \ensuremath{\Varid{ring}}, and a \ensuremath{\Varid{torus}} implementations \cite{Loogen2012}.
-These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell.
-We aim to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
-
-\subsection{Parallel pipe}
-
-The parallel \ensuremath{\Varid{pipe}} skeleton is semantically equivalent to folding over a list \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]} of arrows with \ensuremath{\mathbin{>\!\!>\!\!>}}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
-We implement this skeleton using the \ensuremath{\Conid{ArrowLoop}} typeclass which gives us the \ensuremath{\Varid{loop}\mathbin{::}\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
-For example %this
-\mbcomment{das kann man hier so lassen, oder?}\olcomment{sicherlich!}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{loop}\;(\Varid{arr}\;(\lambda (\Varid{a},\Varid{b})\to (\Varid{b},\Varid{a}\mathbin{:}\Varid{b}))){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-which is the same as
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-defines an arrow that takes its input \ensuremath{\Varid{a}} and converts it into an infinite stream \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} of it.
-Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \ensuremath{\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}} inside the second argument of \ensuremath{\mathbin{\&\!\&\!\&}} and then only picking the first element of the resulting list with \ensuremath{\Varid{arr}\;\Varid{last}}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{pipeSimple}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{a}\;\Varid{conf})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]\to \Varid{arr}\;\Varid{a}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\Varid{pipeSimple}\;\Varid{conf}\;\Varid{fs}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\!\&\!\&}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:})\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{last}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{lazy}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{lazy}\mathrel{=}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\to \Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{A~first implementation of the \ensuremath{\Varid{pipe}} skeleton expressed with Parallel Arrows.
-Note that the use of \ensuremath{\Varid{lazy}} is essential as without it programs using this definition would never halt.
-We need to enforce that the evaluation of the input \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} terminates before passing it into \ensuremath{\Varid{parEvalN}}.}
-\label{fig:pipeSimple}
-\end{figure}
-
-However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \ensuremath{\Varid{pipe}}: % ~\ref{fig:pipe}.
-% \begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{pipe}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{a})\;(\Varid{fut}\;\Varid{a})\;\Varid{conf},\Conid{Future}\;\Varid{fut}\;\Varid{a})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]\to \Varid{arr}\;\Varid{a}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\Varid{pipe}\;\Varid{conf}\;\Varid{fs}\mathrel{=}\Varid{unliftFut}\;(\Varid{pipeSimple}\;\Varid{conf}\;(\Varid{map}\;\Varid{liftFut}\;\Varid{fs})){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-% \caption{Final definition of the pipe skeleton which uses Futures}
-% \label{fig:pipe}
-% \end{figure}
-
-Sometimes, this \ensuremath{\Varid{pipe}} definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
-\ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}}.
-By wrapping these two arrows inside a common type we obtain \ensuremath{\Varid{pipe2}} (Fig.~\ref{fig:pipe2}).
-\begin{figure}[h]
-\olcomment{I swapped the type classes here:}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{pipe2}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]),{}\<[E]%
-\\
-\>[B]{}\hsindent{10}{}\<[10]%
-\>[10]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;\Varid{conf})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
-\\
-\>[B]{}\Varid{pipe2}\;\Varid{conf}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{arr}\;\Varid{return}\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu]))\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{pipe}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{unify}\;\Varid{f}\;\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{snd}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{head}\;\mathbf{where}{}\<[E]%
-\\
-\>[9]{}\hsindent{16}{}\<[25]%
-\>[25]{}\Varid{unify}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]){}\<[E]%
-\\
-\>[9]{}\hsindent{16}{}\<[25]%
-\>[25]{}\Varid{unify}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
-\\
-\>[25]{}\hsindent{8}{}\<[33]%
-\>[33]{}(\Varid{mapArr}\;\Varid{f}\mathbin{*\!*\!*}\Varid{mapArr}\;\Varid{g})\mathbin{*\!*\!*}\Varid{arr}\;(\mathbin{\char92 \char95 }\to [\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[25]{}\hsindent{8}{}\<[33]%
-\>[33]{}\Varid{arr}\;(\lambda ((\Varid{a},\Varid{b}),\Varid{c})\to ((\Varid{c},\Varid{a}),\Varid{b})){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of \ensuremath{\Varid{pipe2}}.}
-\label{fig:pipe2}
-\end{figure}
-
-Note that extensive use of this combinator over \ensuremath{\Varid{pipe}} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \ensuremath{\mathbin{|\!\!>\!\!>\!\!>\!\!|}} (Fig.~\ref{fig:par>>>}, which is semantically equivalent to \ensuremath{\mathbin{>\!\!>\!\!>}} similarly to other parallel syntactic sugar from Section~\ref{syntacticSugar}.
-\begin{figure}[h]
-\olcomment{swapped type classes}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{|\!\!>\!\!>\!\!>\!\!|})\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]),{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;())\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
-\\
-\>[B]{}(\mathbin{|\!\!>\!\!>\!\!>\!\!|})\mathrel{=}\Varid{pipe2}\;(){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of \ensuremath{\mathbin{|\!\!>\!\!>\!\!>\!\!|}}.}
-\label{fig:par>>>}
-\end{figure}
-
-\subsection{Ring skeleton} \label{sec:ring}
-\begin{figure}[h]
-	\includegraphics[scale=0.75]{images/ring}
-	\caption{Schematic depiction of the ring skeleton}
-	\label{fig:ringImg}
-\end{figure}
-Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate in a ring topology with each other.
-Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets send to the neighbour of each node.
-This data is sent over direct communication channels, so called \enquote{remote data}.
-We depict it in Appendix, Fig.~\ref{fig:ringEden}.
-%\end{code}
-%with toRD (to make use of remote data)
-%\begin{code}
-%\end{code}
-%and rightRotate:
-%\begin{code}
-
-We can rewrite this functionality easily with the use of \ensuremath{\Varid{loop}} as the definition of the node function, \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})}, after being transformed into an arrow, already fits quite neatly into the \ensuremath{\Varid{loop}}'s \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}}.
-In each iteration we start by rotating the intermediary input from the nodes \ensuremath{[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu]} with \ensuremath{\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})}.
-Similarly to the \ensuremath{\Varid{pipe}} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \ensuremath{\Varid{lazy}} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
-\begin{quote}
-{Note that the list of ring inputs \ensuremath{\Varid{ringIns}} is the same as the list of ring outputs \ensuremath{\Varid{ringOuts}} rotated by one element to the right using the auxiliary function \ensuremath{\Varid{rightRotate}}.
-Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
-\end{quote}
-Next, we zip the resulting \ensuremath{([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])} to \ensuremath{[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})} so we can feed that into a our input arrow \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})}, which we transform into \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})} before lifting it to \ensuremath{\Varid{arr}\;[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]\;[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} to get a list \ensuremath{[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]}.
-Finally we unzip this list into \ensuremath{([\mskip1.5mu \Varid{o}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])}.
-Plugging this arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])\;([\mskip1.5mu \Varid{o}\mskip1.5mu],\Varid{fut}\;\Varid{r})} into the definition of \ensuremath{\Varid{loop}} from earlier gives us \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
-This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
-%Further details on this can be found in \cite{eden_cefp}.
-\olcomment{let's do it?}
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ring}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{r},\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})\;\Varid{conf})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{conf}\to \Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})\to \Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{ring}\;\Varid{conf}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parMap}\;\Varid{conf}\;(\Varid{second}\;\Varid{get}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{put})\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{unzip}){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{rightRotate}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{rightRotate}\mathrel{=}\Varid{arr}\mathbin{\$}\lambda \Varid{list}\to \mathbf{case}\;\Varid{list}\;\mathbf{of}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}[\mskip1.5mu \mskip1.5mu]\to [\mskip1.5mu \mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{xs}\to \Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{lazy}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{lazy}\mathrel{=}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\to \Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Final definition of the \ensuremath{\Varid{ring}} skeleton.}
-\label{fig:ringFinal}
-\end{figure}
- % $ %% formatting
-% and lazy:
-% \begin{code}
-% lazy :: (Arrow arr) => arr [a] [a]
-% lazy = arr (\ ~(x:xs) -> x : lazy xs
-% \end{code}
-%% we have shown this already.
-
-\subsection{Torus skeleton}
-\begin{figure}
-	\includegraphics[scale=0.75]{images/torus}
-	\caption{Schematic depiction of the torus skeleton}
-	\label{fig:ringTorusImg}
-\end{figure}
-If we take the concept of a ring from \ref{sec:ring} one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
-Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-With our Parallel Arrows we re-implement the torus combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the \ensuremath{\Conid{ArrowLoop}} typeclass.
-
-Similar to the \ensuremath{\Varid{ring}}, we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu])} in the second argument (loop only allows for two arguments) of our looped arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))\;([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} and our rotation arrow becomes \ensuremath{\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))} instead of the singular rotation in the ring as we rotate \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu]} horizontally and \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]} vertically.
-Then, we once again zip the inputs for the input arrow with \ensuremath{\Varid{arr}\;(\Varid{uncurry3}\;\Varid{zipWith3}\;\Varid{lazyzip3})} from \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} to \ensuremath{[\mskip1.5mu [\mskip1.5mu (\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]}, which we then feed into our parallel execution.
-
-This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
-We first have to \ensuremath{\Varid{shuffle}} all the inputs to then pass it into \ensuremath{\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f})} which yields us \ensuremath{[\mskip1.5mu (\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]}.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \ensuremath{\Varid{arr}\;\Varid{length}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unshuffle}} with the use of \ensuremath{\Varid{app}\mathbin{::}\Varid{arr}\;(\Varid{arr}\;\Varid{a}\;\Varid{b},\Varid{a})\;\Varid{c}} from the \ensuremath{\Conid{ArrowApply}} typeclass.
-Finally, we unpack this matrix \ensuremath{[\mskip1.5mu [\mskip1.5mu [\mskip1.5mu (\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!\!>\!\!>}\Varid{threetotwo}} to get  \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))}.
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{torus}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;\Varid{conf},{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{a},\Conid{Future}\;\Varid{fut}\;\Varid{b})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to \Varid{arr}\;[\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu]\;[\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{torus}\;\Varid{conf}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{uncurry3}\;(\Varid{zipWith3}\;\Varid{lazyzip3}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{arr}\;\Varid{length}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unshuffle})\mathbin{\&\!\&\!\&}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{shuffle}\mathbin{>\!\!>\!\!>}\Varid{parMap}\;\Varid{conf}\;(\Varid{ptorus}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{app}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!\!>\!\!>}\Varid{threetotwo}){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{ptorus}\mathbin{::}(\Conid{Arrow}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{a},\Conid{Future}\;\Varid{fut}\;\Varid{b})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b}){}\<[E]%
-\\
-\>[B]{}\Varid{ptorus}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{c},\Varid{a},\Varid{b})\to (\Varid{c},\Varid{get}\;\Varid{a},\Varid{get}\;\Varid{b}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{d},\Varid{a},\Varid{b})\to (\Varid{d},\Varid{put}\;\Varid{a},\Varid{put}\;\Varid{b})){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{uncurry3}\mathbin{::}(\Varid{a}\to \Varid{b}\to \Varid{c}\to \Varid{d})\to (\Varid{a},(\Varid{b},\Varid{c}))\to \Varid{d}{}\<[E]%
-\\
-\>[B]{}\Varid{uncurry3}\;\Varid{f}\;(\Varid{a},(\Varid{b},\Varid{c}))\mathrel{=}\Varid{f}\;\Varid{a}\;\Varid{b}\;\Varid{c}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{lazyzip3}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{c}\mskip1.5mu]\to [\mskip1.5mu (\Varid{a},\Varid{b},\Varid{c})\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{lazyzip3}\;\Varid{as}\;\Varid{bs}\;\Varid{cs}\mathrel{=}\Varid{zip3}\;\Varid{as}\;(\Varid{lazy}\;\Varid{bs})\;(\Varid{lazy}\;\Varid{cs}){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{threetotwo}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Varid{a},\Varid{b},\Varid{c})\;(\Varid{a},(\Varid{b},\Varid{c})){}\<[E]%
-\\
-\>[B]{}\Varid{threetotwo}\mathrel{=}\Varid{arr}\mathbin{\$}\lambda \mathord{\sim}(\Varid{a},\Varid{b},\Varid{c})\to (\Varid{a},(\Varid{b},\Varid{c})){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
- % $
-\caption{Definition of the \ensuremath{\Varid{torus}} skeleton.}
-\label{fig:torus}
-\end{figure}
-As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}.
-Their instantiation of the skeleton \ensuremath{\Varid{nodefunction}} can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{nodefunction}\mathbin{::}\Conid{Int}\to {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}((\Conid{Matrix},\Conid{Matrix}),[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu])\to {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}([\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu]){}\<[E]%
-\\
-\>[B]{}\Varid{nodefunction}\;\Varid{n}\;((\Varid{bA},\Varid{bB}),\Varid{rows},\Varid{cols})\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}([\mskip1.5mu \Varid{bSum}\mskip1.5mu],\Varid{bA}\mathbin{:}\Varid{nextAs},\Varid{bB}\mathbin{:}\Varid{nextBs}){}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}\mathbf{where}\;\Varid{bSum}\mathrel{=}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{foldl'}\;\Varid{matAdd}\;(\Varid{matMult}\;\Varid{bA}\;\Varid{bB})\;(\Varid{zipWith}\;\Varid{matMult}\;\Varid{nextAs}\;\Varid{nextBs}){}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{nextAs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{rows}{}\<[E]%
-\\
-\>[17]{}\hsindent{8}{}\<[25]%
-\>[25]{}\Varid{nextBs}\mathrel{=}\Varid{take}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{cols}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Adapted \ensuremath{\Varid{nodefunction}} for matrix multiplication with the \ensuremath{\Varid{torus}} from Fig.~\ref{fig:torus}}
-\label{fig:torusMatMult}
-\end{figure}
-If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.\olcomment{much more details on this!}
-\begin{figure}[ht]
-\olcomment{more nodes!!!}
-	\centering
-	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (Parrows)}
-	\label{fig:torus_parrows_trace}
-\end{figure}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
-	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden)}
-	\label{fig:torus_eden_trace}
-\end{figure}
-
-%\FloatBarrier
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include topologySkeletons.tex
 	%\pagebreak
-	\section{Benchmarks}
-\label{sec:benchmarks}
-
-%\begin{figure}[ht]
-%	\centering
-	\framebox{
-		\begin{tikzpicture}[thick, scale=0.7]
-		\begin{axis}[
-			xbar, xmin=0,
-			bar width=0.4cm,
-			width=12cm, height=7cm, enlarge y limits=0.1,
-			ylabel={amount of cores},
-			xlabel={mean runtime (seconds)},
-			symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
-			ytick=data,
-			nodes near coords, nodes near coords align={horizontal},
-			]
-			\addplot coordinates {(841,2) (248,4) (115,8) (63.9,16) (41.6,32) (34.6,48) (33.0,64) (32.9,96)};
-		\end{axis}
-		\end{tikzpicture}
-	}
-%	\caption[Parallel Matrix Multiplication using Eden (two 1024x1024 matrices)]
-%\end{figure}
-
-\framebox{
-	\begin{tikzpicture}[thick, scale=0.7]
-	\begin{axis}[
-	xbar, xmin=0,
-	bar width=0.4cm,
-	width=12cm, height=7cm, enlarge y limits=0.1,
-	ylabel={amount of cores},
-	xlabel={mean seconds run on torus\_matrix\_parrows (two 1024x1024 matrices)},
-	symbolic y coords={2, 4, 8, 16, 32, 48, 64, 96},
-	ytick=data,
-	nodes near coords, nodes near coords align={horizontal},
-	]
-	\addplot coordinates {(663,2) (205,4) (113,8) (68.5,16) (41.9,32) (35.4,48) (36.3,64) (33.4,96)};
-	\end{axis}
-	\end{tikzpicture}
-}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include benchmarks.tex
 	%%\pagebreak
-
-\section{Conclusion}
-\label{sec:conclusion}
-Arrows are a generic concept that allows for powerful composition combinators.
-To our knowledge we are the first ones to represent parallel computation with arrows.\olcomment{that strange arrows-based robot interaction paper from 1993 or so!
-clearly discuss in related work}
-
-Arrows turn out to be a useful tool for composing in parallel programs.
-We do not have to introduce new monadic types that wrap the computation.
-Instead use arrows just like regular sequential pure functions.
-%
-This work features multiple parallel backends: the already available parallel Haskell flavours.
-Parallel Arrows feature an implementation of the \ensuremath{\Conid{ArrowParallel}} interface for Multicore Haskell, \ensuremath{\Conid{Par}} Monad, and Eden.
-With our approach parallel programs can be ported across these flavours with no effort.
-%
-%
-Performancewise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}
-
-\mbcomment{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}
-
-\olcomment{Remove websites from citations, put them into footnotes!}
-
-\olcomment{Parrows + accelerate = love? Metion port to Frege. Mention the Par monad troubles.}
-
-\subsection{Future Work}
-\label{sec:future-work}
-
-Our PArrows interface can be expanded to futher parallel Haskells.
-More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} a modern distributed Haskell that would benefit from our Arrows notation.
-Future-aware special versions of Arrow combinated can be extended and further improved.
-We would look into more transparency of the API, it should basically infuse as little overhead as possible.
-
-Of course, definitions of further skeletons are viable and needed. We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
-
-Accelerate~\cite{Chakravarty:2011:AHA:1926354.1926358} is not related to our approach. It would be interesting to see a hybrid of both APIs.
-
-\olcomment{replace API with DSL globally?}
+	%include conclusion.tex
 	%\pagebreak
@@
         \appendix
-	\section{Utility Functions}\label{utilfns}
-To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \ensuremath{\Varid{map}}, \ensuremath{\Varid{foldl}} and \ensuremath{\Varid{zipWith}} on arrows.
-
-The \ensuremath{\Varid{mapArr}} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} \cite{programming_with_arrows}:
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{mapArr}\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{mapArr}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{\mid\!\mid\!\mid}(\Varid{f}\mathbin{*\!*\!*}\Varid{mapArr}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\mathbf{where}\;\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Conid{Left}\;(){}\<[E]%
-\\
-\>[9]{}\hsindent{6}{}\<[15]%
-\>[15]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{\ensuremath{\Varid{map}} over arrows}
-\label{fig:mapArr}
-\end{figure}
-Similarly, we can also define \ensuremath{\Varid{foldlArr}} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{b},\Varid{a})\;\Varid{b}} with a neutral element \ensuremath{\Varid{b}} to \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}}:
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{foldlArr}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Varid{b},\Varid{a})\;\Varid{b}\to \Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;\Varid{b}{}\<[E]%
-\\
-\>[B]{}\Varid{foldlArr}\;\Varid{f}\;\Varid{b}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;\Varid{listcase}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{arr}\;(\Varid{const}\;\Varid{b})\mathbin{\mid\!\mid\!\mid}{}\<[E]%
-\\
-\>[9]{}\hsindent{8}{}\<[17]%
-\>[17]{}(\Varid{first}\;(\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{b},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{foldlArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}\Varid{app}){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\mathbf{where}\;\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Conid{Left}\;[\mskip1.5mu \mskip1.5mu]{}\<[E]%
-\\
-\>[9]{}\hsindent{6}{}\<[15]%
-\>[15]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{\ensuremath{\Varid{foldl}} over arrows}
-\label{fig:foldlArr}
-\end{figure}
-\fixme{pipepipepipe does not work with lhs2TeX}
-Finally, with the help of \ensuremath{\Varid{mapArr}} (Fig.~\ref{fig:mapArr}), we can define \ensuremath{\Varid{zipWithArr}} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}} to an arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{zipWithArr}\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\Rightarrow \Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}\to \Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{zipWithArr}\;\Varid{f}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda (\Varid{as},\Varid{bs})\to \Varid{zipWith}\;(,)\;\Varid{as}\;\Varid{bs})\mathbin{>\!\!>\!\!>}\Varid{mapArr}\;\Varid{f}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{\ensuremath{\Varid{zipWith}} over arrows}
-\label{fig:zipWithArr}
-\end{figure}
- %$ %% formatting
-These combinators make use of the \ensuremath{\Conid{ArrowChoice}} type class which provides the \pipepipepipe\olcomment{CHECK!} combinator.
-It takes two arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} and combines them into a new arrow \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}} which pipes all \ensuremath{\Conid{Left}\;\Varid{a}}'s to the first arrow and all \ensuremath{\Conid{Right}\;\Varid{b}}'s to the second arrow.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\mathbin{\mid\!\mid\!\mid})\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Type signature of \pipepipepipe}
-\label{fig:codeSigPipePipePipe}
-\end{figure}
-\fixme{pipepipepipe does not work with lhs2TeX}\olcomment{trying
-  direct one}
-
-With the zipWithArr combinator we can also write a combinator \ensuremath{\Varid{listApp}} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to an arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{listApp}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{listApp}\;\Varid{fs}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{as}\to (\Varid{fs},\Varid{as}))\mathbin{>\!\!>\!\!>}\Varid{zipWithArr}\;\Varid{app}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of \ensuremath{\Varid{listApp}}}
-\label{fig:listApp}
-\end{figure}
-% $ %% formatting
-.
-Note that  this additionally makes use of the \ensuremath{\Conid{ArrowApply}} typeclass that allows us to evaluate arrows with \ensuremath{\Varid{app}\mathbin{::}\Varid{arr}\;(\Varid{arr}\;\Varid{a}\;\Varid{b},\Varid{a})\;\Varid{c}}.
-
-% $ %% formatting
-
-\subsection{Omitted Funtion Definitions}
-
-We have omitted some function definitions in the main text for
-brevity, and redeem this here.
-%
-We warp Eden's build-in Futures in PArrows as in Figure~\ref{fig:RDFuture}.
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{RemoteData}\;\Varid{a}\mathrel{=}\Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathbin{::}\Conid{RD}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{RemoteData}\;\Varid{a}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{put}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to \Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathrel{=}\Varid{release}\;\Varid{a}\mskip1.5mu\}){}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{get}\mathrel{=}\Varid{arr}\;\Varid{rd}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{fetch}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{\ensuremath{\Conid{RD}}-based \ensuremath{\Conid{RemoteData}} version of \ensuremath{\Conid{Future}} for the Eden backend.}
-\label{fig:RDFuture}
-\end{figure}
-
-The full definition of \ensuremath{\Varid{farmChunk}} is in Figure~\ref{fig:farmChunk}.
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{farmChunk}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Varid{shuffle}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Definition of \ensuremath{\Varid{farmChunk}}.}
-\label{fig:farmChunk}
-\end{figure}
-
-Eden definition of \ensuremath{\Varid{ring}} skeleton is in Figure~\ref{fig:ringEden}.
-
-\begin{figure}[h]
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ringSimple}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{4}{}\<[4]%
-\>[4]{}(\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r}))\to [\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{ringSimple}\;\Varid{f}\;\Varid{is}\mathrel{=}{}\<[20]%
-\>[20]{}\Varid{os}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathbf{where}\;(\Varid{os},\Varid{ringOuts})\mathrel{=}\Varid{unzip}\;(\Varid{parMap}\;(\Varid{toRD}\mathbin{\$}\Varid{uncurry}\;\Varid{f})\;(\Varid{zip}\;\Varid{is}\mathbin{\$}\Varid{lazy}\;\Varid{ringIns})){}\<[E]%
-\\
-\>[3]{}\hsindent{6}{}\<[9]%
-\>[9]{}\Varid{ringIns}\mathrel{=}\Varid{rightRotate}\;\Varid{ringOuts}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{toRD}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow {}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}((\Varid{i},\Varid{r})\to (\Varid{o},\Varid{r}))\to ((\Varid{i},\Conid{RD}\;\Varid{r})\to (\Varid{o},\Conid{RD}\;\Varid{r})){}\<[E]%
-\\
-\>[B]{}\Varid{toRD}\;{}\<[7]%
-\>[7]{}\Varid{f}\;(\Varid{i},\Varid{ringIn}){}\<[22]%
-\>[22]{}\mathrel{=}(\Varid{o},\Varid{release}\;\Varid{ringOut}){}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\mathbf{where}\;(\Varid{o},\Varid{ringOut})\mathrel{=}\Varid{f}\;(\Varid{i},\Varid{fetch}\;\Varid{ringIn}){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{rightRotate}{}\<[16]%
-\>[16]{}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{rightRotate}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}{}\<[19]%
-\>[19]{}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{rightRotate}\;\Varid{xs}\mathrel{=}{}\<[19]%
-\>[19]{}\Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{lazy}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{lazy}\mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\caption{Eden's definition of the \ensuremath{\Varid{ring}} skeleton.}
-\label{fig:ringEden}
-\end{figure}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+	%include utilityFunctions.tex
 \end{document}

### 2017-05-22 paper/content/mapSkeletons.tex
fixed things to that lhs2TeX can run (without the code markers)
renamed *.lhs files to *.tex

--- 2017-05-12 paper/mapSkeletons.tex
@@
-\FloatBarrier
+%%% \FloatBarrier
 \section{Map-based Skeletons}
@@
 	\includegraphics[scale=0.7]{images/parMap}
-	\caption{Schematic depiction of parMap}
+	\caption{Schematic depiction of |parMap|.}
 	\label{fig:parMapImg}
 \end{figure}
-\code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
-We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}.
-Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
+The |parMap| skeleton (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
+We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to get an arrow |arr [a] [b]|.
+Just like |parEvalN|, |parMap| is 100\% strict.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parMap :: (ArrowParallel arr a b conf) =>
-	conf -> (arr a b) -> (arr [a] [b])
+\begin{code}
+parMap :: (ArrowParallel arr a b conf) => conf -> (arr a b) -> (arr [a] [b])
 parMap conf f = parEvalN conf (repeat f)
-\end{lstlisting}
+\end{code}
 \caption{Definition of parMap}
@@
 \end{figure}
-As \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}.
-So it makes sense to also have a \code{parMapStream} (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
+As |parMap| (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as |parEvalN| compared to |parEvalNLazy|.
+So it makes sense to also have a |parMapStream| (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
@@
 parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
-\end{lstlisting}
-\caption{Definition of parMapStream}
+\end{code}
+\caption{Definition of |parMapStream|.}
 \label{fig:parMapStream}
@@
 	\includegraphics[scale=0.7]{images/farm}
-	\caption{Schematic depiction of farm}
+	\caption{Schematic depiction of a |farm|, a statically
+          load-balanced |parMap|.}
 	\label{fig:farmImg}
 \end{figure}
-A \code{parMap} (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper).
-This can be quite wasteful and a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \code{numCores} workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
+A |parMap| (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper).
+This can be quite wasteful and a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
+The definitions of helper functions |unshuffle|, |takeEach|, |shuffle| (shown in Appendix) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 farm :: (ArrowParallel arr a b conf,
@@
 	shuffle
-\end{lstlisting}
-\caption{Definition of farm}
+
+unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
+unshuffle n = arr (\xs -> [takeEach n (drop i xs) @|@ i <- [0..n-1]])
+
+takeEach :: Int -> [a] -> [a]
+takeEach n [] = []
+takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
+
+shuffle :: (Arrow arr) => arr [[a]] [a]
+shuffle = arr (concat . transpose)
+\end{code}
+\caption{The definition of |farm|.}
 \label{fig:farm}
@@
 
-\subsection{farmChunk}
+\subsection{The |farmChunk| Skeleton}
 \begin{figure}[h]
@@
 \end{figure}
-Since a \code{farm} (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just \code{parMap} with a different work distribution, it is, again, 100\% strict.
-So we define \code{farmChunk} (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \code{parEvalNLazy} instead of \code{parEvalN}.
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf,
-	ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
-farmChunk conf chunkSize numCores f =
-	unshuffle numCores >>>
-	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
-	shuffle
-\end{lstlisting}
-\caption{Definition of farmChunk}
-\label{fig:farmChunk}
-\end{figure}
+Since a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just |parMap| with a different work distribution, it is, again, 100\% strict.
+So we can define |farmChunk| (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|.
+It is basically the same definition as for |farm|, with |parEvalN| replaced with |parEvalNLazy|, as Appendix shows.
 
-\subsection{parMapReduce}
+\subsection{Map and reduce}
+
+A~simple |map|--|reduce| can be written like in Figure~\ref{fig:parMapReduceDirect}.
+Notice that the performance of the |>>>| combinator is essential for the performance of the skeleton.
+A~definitive version would use Futures.
+
  \olcomment{it appears STRANGE. are the data really left alone and noded after map and taken from there by reduce? It makes sense only is no communication through master takes place. ELSE: CUT!}
@@
  \mbcomment{this requires some work. Either change this to use futures or cut, yes.}
--- this does not completely adhere to Google's definition of Map Reduce as it
--- the mapping function does not allow for "reordering" of the output
--- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
+% -- this does not completely adhere to Google's definition of Map Reduce as it
+% -- the mapping function does not allow for "reordering" of the output
+% -- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
+
+\olcomment{now rewritten as motivation for futures. maybe still cut?}
 
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parMapReduceDirect :: (ArrowParallel arr [a] b conf,
-	ArrowApply arr, ArrowChoice arr) =>
+\begin{code}
+parMapReduceDirect :: (ArrowParallel arr [a] b conf, ArrowApply arr, ArrowChoice arr) =>
 	conf -> ChunkSize -> arr a b -> arr (b, b) b -> b -> arr [a] b
@@
 	foldlArr foldfn neutral
-\end{lstlisting}
+\end{code}
 \caption{Definition of parMapReduceDirect}

### 2017-05-22 paper/content/parrows.tex
renamed *.lhs files to *.tex

--- 2017-05-12 paper/parrows.tex
@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
-Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \lstinline{listApp} from Appendix \ref{utilfns}, but does parallel instead of serial evaluation).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions |a -> b| in parallel, |parEvalN| (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
+Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
+This combinator is similar to our utility function |listApp| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
+% \begin{figure}[h]
+\begin{code}
 parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
-\end{lstlisting}
-\caption{parEvalN Arrow combinator as a function}
-\label{fig:parEvalNArrowFn}
-\end{figure}
-With this definition of \code{parEvalN}, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we introduce a new typeclass \code{ArrowParallel arr a b} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{parEvalN Arrow combinator as a function}
+% \label{fig:parEvalNArrowFn}
+% \end{figure}
+With this definition of |parEvalN|, parallel execution is yet another arrow combinator.
+But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b| to host this combinator: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+% \begin{figure}[h]
+\begin{code}
 class Arrow arr => ArrowParallel arr a b where
 	parEvalN :: [arr a b] -> arr [a] [b]
-\end{lstlisting}
-\caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
-\label{fig:parEvalNArrowTypeClass1}
-\end{figure}
+\end{code}
+% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
+% \label{fig:parEvalNArrowTypeClass1}
+% \end{figure}
 Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs.
 normalform).
-For this reason we also introduce an additional \code{conf} parameter to the function.
-We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}.
-So we add it to the type signature of the typeclass as well and get \code{ArrowParallel arr a b conf} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+For this reason we also introduce an additional |conf| parameter to the function.
+We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|.
+So we add it to the type signature of the typeclass as well and get |ArrowParallel arr a b conf|: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
+% \begin{figure}[h]
+\olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?}
+\begin{code}
 class Arrow arr => ArrowParallel arr a b conf where
 	parEvalN :: conf -> [arr a b] -> arr [a] [b]
-\end{lstlisting}
-\caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
-\label{fig:parEvalNArrowTypeClassFinal}
-\end{figure}
-Note that we don't require the \code{conf} parameter in every implementation.
-If it is not needed, we usually just default the \code{conf} type parameter to () and even blank it out in the parameter list of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
+\end{code}
+% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
+% \label{fig:parEvalNArrowTypeClassFinal}
+% \end{figure}
+Note that we don't require the |conf| parameter in every implementation.
+If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|, as we will see in the implementation of the Multicore and the |Par| Monad backend.
 
-\subsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from appendix \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
+\subsection{ArrowParallel instances}
+
+\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
+The Multicore Haskell implementation of this class is implemented in a straightforward manner by using |listApp| from Appendix~\ref{utilfns} combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b () where
+\begin{code}
+instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
     	parEvalN _ fs =
@@
         	arr (uncurry pseq)
-\end{lstlisting}% $ %% formatting
+\end{code}% $ %% formatting
 \caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
@@
 \end{figure}
-For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \code{Strategy} interface, we want to the user not to lose this ability because of using our API.
-We therefore introduce the \code{Conf a} data-type that simply wraps a \code{Strategy a} (Fig.~\ref{fig:confa}).
-We can't directly use the \code{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
+For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, we want to the user not to lose this ability because of using our API.
+We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a| (Fig.~\ref{fig:confa}).
+We can't directly use the |Strategy a| type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances:
+% \begin{figure}[h]
+\begin{code}
+data Conf a = Conf (Strategy a)
+\end{code}
+% \caption{Definition of Conf a}
+% \label{fig:confa}
+% \end{figure}
+To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-data Conf a = Conf (Strategy a)
-\end{lstlisting}
-\caption{Definition of Conf a}
-\label{fig:confa}
-\end{figure}
-To get our configurable \code{ArrowParallel} instance, we simply unwrap the strategy and pass it to \code{parList} as in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
@@
         	arr (uncurry pseq)
-\end{lstlisting}
+\end{code}
 \caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
@@
 \end{figure}
-\subsection{Par Monad}
-The ParMonad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a =>\ \ a -> Par (IVar a)} function.
-The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.
+\subsubsection{|Par| Monad}
+\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting}
+The |Par| monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| monad's |spawnP :: NFData a => a -> Par (IVar a)| function.
+The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| monad.
 
-We therefore apply each function to its corresponding input value with \code{app} and then fork the computation away with \code{arr spawnP} inside a \code{zipWithArr} call.
-This yields a list \code{[Par (IVar b)]}, which we then convert into \code{Par [IVar b]} with \code{arr sequenceA}.
-In order to wait for the computation to finish, we map over the \code{IVar}s inside the ParMonad with \code{arr (>>= mapM get)}.
-The result of this operation is a \code{Par [b]} from which we can finally remove the monad again by running \code{arr runPar} to get our output of \code{[b]}.
+We therefore apply each function to its corresponding input value with |listApp| (Fig.~\ref{fig:listApp}) and then fork the computation away with |arr spawnP| inside a |zipWithArr| call.
+This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|.
+In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| monad with |arr (>>= mapM get)|.
+The result of this operation is a |Par [b]| from which we can finally remove the monad again by running |arr runPar| to get our output of |[b]|.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
@@
 			arr runPar
-\end{lstlisting} %$ %% formatting
+\end{code} %$ %% formatting
 \caption{ArrowParallel instance for the Par Monad backend}
@@
 
-\subsection{Eden}
-For the Multicore and ParMonad implementation we could use general instances of \code{ArrowParallel} that just require the \code{ArrowApply} and \code{ArrowChoice} typeclasses.
+\subsubsection{Eden}
+For both the Multicore Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses.
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a typeclass \code{ArrowUnwrap} (Fig.~\ref{fig:ArrowUnwrap}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
+% \begin{figure}[h]
+\begin{code}
 class (Arrow arr) => ArrowUnwrap arr where
 	arr a b -> (a -> b)
-\end{lstlisting}
-\caption{possible ArrowUnwrap typeclass}
-\label{fig:ArrowUnwrap}
-\end{figure}
-We don't do it here for aesthetic resons, though. For now, we just implement \code{ArrowParallel} for normal functions (Fig.~\ref{fig:ArrowParallelEdenFns})
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{possible ArrowUnwrap typeclass}
+% \label{fig:ArrowUnwrap}
+% \end{figure}
+We don't do it here for aesthetic resons, though. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
+% \begin{figure}[h]
+\begin{code}
 instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
 parEvalN _ fs as = spawnF fs as
-\end{lstlisting}
-\caption{ArrowParallel instance for functions in the Eden backend}
-\label{fig:ArrowParallelEdenFns}
-\end{figure}
-and the Kleisli type (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{ArrowParallel instance for functions in the Eden backend}
+% \label{fig:ArrowParallelEdenFns}
+% \end{figure}
+and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
+% \begin{figure}[h]
+\begin{code}
 instance (Monad m, Trans a, Trans b, Trans (m b)) =>
@@
 	(Kleisli $ sequence)
-\end{lstlisting}
-\caption{ArrowParallel instance for the Kleisli type in the Eden backend}
-\label{fig:ArrowParallelEdenKleisli}
-\end{figure}
+\end{code}
+% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
+% \label{fig:ArrowParallelEdenKleisli}
+% \end{figure}
 
-\subsection{Impact of parallel Arrows}
-\olcomment{move this to Contributions in the front or something}
-We have seen that we can wrap parallel Haskells inside of the \code{ArrowParallel} interface, but why do we abstract parallelism this way and what does this approach do better than the other parallel Haskells?
-\begin{itemize}
-	\item \textbf{Arrow API benefits}:
-	With the \code{ArrowParallel} typeclass we do not lose any benefits of using arrows as \code{parEvalN} is just yet another arrow combinator.
-The resulting arrow can be used in the same way a potential serial version could be used.
-This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
-We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
-	\item \textbf{Abstraction}:
-	With the \code{ArrowParallel} typeclass, we abstracted all parallel implementation logic away from the business logic.
-This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
-So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual \code{ArrowParallel} instance.
-\end{itemize}
+%\FloatBarrier
 

### 2017-05-25 paper/content/sources/arrows.tex
moving things around, reworking the makefile

\subsection{Arrows}
\label{sec:arrows}
\begin{figure}[h]
	\includegraphics{images/arrow}
	\caption{Schematic depiction of an arrow}
\end{figure}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation. An arrow |arr a b| represents  a computation that converts an input |a| to an output |b|. This is defined in the arrow typeclass:

\begin{figure}[h]
\centering
%\subfloat[Arrow class definition][]{%
% begin{minipage}{0.5\textwidth}
\parbox{0.49\linewidth}{%
\begin{code}
class Arrow arr where
  arr :: (a -> b) -> arr a b
  (>>>) :: arr a b -> arr b c -> arr a c
  first :: arr a b -> arr (a,c) (b,c)
\end{code}
\caption{Arrow class definition}
}\qquad
\begin{minipage}{0.49\linewidth}%
%\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
%	\begin{center}
\centering
	{\includegraphics[scale=0.6]{images/arr}}
	{\includegraphics[scale=0.6]{images/compose}}
	{\includegraphics[scale=0.6]{images/first}}
%	\end{center}
%}%\hfill
\caption{The schematic depiction of |Arrow| combinators}
\end{minipage}
\end{figure}
|arr| is used to lift an ordinary function to an arrow type, similarly to the monadic |return|. The |>>>| operator is analogous to the monadic composition  |>>=| and combines two arrows |arr a b| and |arr b c| by "wiring" the outputs of the first to the inputs to the second to get a new arrow |arr a c|. Lastly, the |first| operator  takes the input arrow from |b| to |c| and converts it into an arrow on pairs with the second argument untouched. It allows us to to save input across arrows.
\\\\
The most prominent instances of this interface are regular functions |(->)|: %(Fig.~\ref{fig:arrowfn}),
%\begin{figure}[h]
\begin{code}
instance Arrow (->) where
	arr f = f
	f >>> g = g . f
	first f = \(a, c) -> (f a, c) 
\end{code}
% \caption{Arrow instance for regular functions}
% \label{fig:arrowfn}
% \end{figure}
%%% on the verge of acceptable code
and the Kleisli type: %(Fig.~\ref{fig:arrowkleisli}).
% \begin{figure}[h]
\begin{code}
data Kleisli m a b = Kleisli { run :: a -> m b }

instance Monad m => Arrow (Kleisli m) where
	arr f = Kleisli (return . f)
	f >>> g = Kleisli (\a -> f a >>= g)
	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
\end{code}%$
% \caption{Definition of the Kleisli type and the corresponding arrow instance}
% \label{fig:arrowkleisli}
% \end{figure}
\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		% \subcaptionbox
{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
		% \subcaptionbox
{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
|first| & |second| \\
\midrule
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/andandand}}\\
|(***)|\label{fig:***Img} & |(&&&)| \label{fig:&&&Img} \\
	\end{tabular}
	\caption{Visual depiction of syntactic sugar for Arrows.}
	\label{3}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar: the functions |second|, |***| and |&&&|. 
The mirrored version of |first|, called |second| (Fig.~\ref{fig:secondImg}% ,~\ref{fig:second}
) is:
% \begin{figure}[h]
\begin{code}
second :: Arrow arr => arr a b -> arr (c, a) (c, b)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{code}
% \caption{The second combinator}
% \label{fig:second}
% \end{figure}
the |***| combinator that combines |first| and |second| to handle two inputs in one arrow, (Fig.\ref{fig:***Img}%
% ,~\ref{fig:***}
) is defined as
% \begin{figure}[h]
\begin{code}
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{code}
% \caption{The (***) combinator}
% \label{fig:***}
% \end{figure}
and the |&&&| combinator that constructs an arrow which outputs two different values like |***|, but takes only one input (Fig.~\ref{fig:&&&Img}% ,~\ref{fig:&&&}
) is:
% \begin{figure}[h]
\begin{code}
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{code}
% \caption{The (\&\&\&) combinator}
% \label{fig:&&&}
% \end{figure}
A short example given by Hughes on how to use this is 
addition with arrows:
%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
% \begin{figure}[h]
\begin{code}
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{code}
% \caption{Add over arrows}
% \label{fig:addArrows}
% \end{figure}
% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
% \\\\
% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used. We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}
%

The more restrictive interface of arrows (a monad can be \emph{anything}, an arrow is a process of doing something, a \emph{computation}) allows for more elaborate composition and transformation combinators. One of the major problems in parallel computing is composition of parallel processes.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-05-25 paper/content/sources/conclusion.tex
moving things around, reworking the makefile

--- 2017-03-30 paper/conclusion.tex
@@
 This work features multiple parallel backends: the already available parallel Haskell flavours.
-Parallel Arrows feature an implementation of the \code{ArrowParallel} interface for Multicore Haskell, \code{Par} monad, and Eden.
+Parallel Arrows feature an implementation of the |ArrowParallel| interface for Multicore Haskell, |Par| Monad, and Eden.
 With our approach parallel programs can be ported across these flavours with no effort.
@@
 \olcomment{Remove websites from citations, put them into footnotes!}
+
+\olcomment{Parrows + accelerate = love? Metion port to Frege. Mention the Par monad troubles.}
+
+\subsection{Future Work}
+\label{sec:future-work}
+
+Our PArrows interface can be expanded to futher parallel Haskells.
+More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} a modern distributed Haskell that would benefit from our Arrows notation.
+Future-aware special versions of Arrow combinated can be extended and further improved.
+We would look into more transparency of the API, it should basically infuse as little overhead as possible.
+
+Of course, definitions of further skeletons are viable and needed. We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
+
+Accelerate~\cite{Chakravarty:2011:AHA:1926354.1926358} is not related to our approach. It would be interesting to see a hybrid of both APIs.
+
+\olcomment{replace API with DSL globally?}

### 2017-05-25 paper/content/sources/futures.tex
moving things around, reworking the makefile

--- 2017-05-12 paper/futures.tex
@@
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
 someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
-\end{lstlisting}
+\end{code}
 \caption{An example parallel Arrow combinator without Futures}
@@
 \end{figure}
-In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
+In a distributed environment, the resulting arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
 
-While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one |parEvalN| call by directly wiring the arrows properly together, this example illustrates an important problem: When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
 This can become a serious bottleneck %in heavy threaded applications.
@@
 This motivates for an approach that allows the nodes to communicate directly with each other.
-Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
+Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of |RD| (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
 
-But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \code{Future} typeclass (Fig.~\ref{fig:future}).
+But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-class Future fut a | a -> fut where
+\begin{code}
+class Future fut a @|@ a -> fut where
     put :: (Arrow arr) => arr a (fut a)
     get :: (Arrow arr) => arr (fut a) a
-\end{lstlisting}
+\end{code}
 \caption{Definition of the Future typeclass}
@@
 \end{figure}
-Since \code{RD} is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture} (this is due to the same reason we had to introduce a wrapper for \code{Strategy a} in the Multicore Haskell implementation of \code{ArrowParallel} in chapter \ref{sec:parrows:multicore}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-data RemoteData a = RD { rd :: RD a }
+Since |RD| is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Appendix in Fig.~\ref{fig:RDFuture}.
+This is due to the same reason we had to introduce a wrapper for |Strategy a| in the Multicore Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
 
-instance (Trans a) => Future RemoteData a where
-    put = arr (\a -> RD { rd = release a })
-    get = arr rd >>> arr fetch
-\end{lstlisting}
-\caption{RD based RemoteData version of Future for the Eden backend}
-\label{fig:RDFuture}
-\end{figure}
-
-For our Par Monad and Multicore Haskell backends, we can simply use \code{MVar}s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
+For our Par Monad and Multicore Haskell backends, we can simply use |MVar|s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
 \fixme{explain MVars}
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 {-# NOINLINE putUnsafe #-}
@@
     get = arr takeMVar >>> arr unsafePerformIO
-\end{lstlisting}
+\end{code}
 \caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends}
 \label{fig:MVarFuture}
-\end{figure}
+\end{figure} % $
 
-Furthermore, in order for these \code{Future} types to fit with the \code{ArrowParallel} instances we gave earlier, we have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
-Because \code{MVar} already ships with a \code{NFData} instance, we only have to supply two simple instances for our \code{RemoteData} type.
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+Furthermore, in order for these |Future| types to fit with the |ArrowParallel| instances we gave earlier, we have to give the necessary |NFData| and |Trans| instances, the latter are only needed in Eden.
+Because |MVar| already ships with a |NFData| instance, we only have to supply two simple instances for our |RemoteData| type:
+% \begin{figure}[h]
+\begin{code}
 instance NFData (RemoteData a) where
@@
 instance Trans (RemoteData a)
-\end{lstlisting}
-\caption{NFData and Trans instances for the RemoteData type.
+\end{code}
+% \caption{NFData and Trans instances for the RemoteData type.
 The Trans instance does not have any functions declared as the default implementation suffices here.
 See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
-\end{figure}
+% \end{figure}
+The |Trans| instance does not have any functions declared as the default implementation suffices here.
 
-Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
+Going back to our communication example we can use this |Future| concept in order to enable direct communications between the nodes in the following way:
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
@@
 	parEvalN () (map (get >>>) fs2)
-\end{lstlisting}
+\end{code}
 \caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel}

### 2017-05-25 paper/content/sources/motivation.tex
moving things around, reworking the makefile

%\section{Motivation}
%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes. In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive. In this paper we will use this concept to express parallelism.

\section{Introduction}
\label{sec:introduction}
\olcomment{todo, reuse 5.5, and more}

blablabla arrows, parallel, haskell.

\paragraph{Contribution}

\olcomment{HIT HERE REALLY STRONG}

%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We wrap parallel Haskells inside of our |ArrowParallel| interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
\begin{itemize}
	\item \textbf{Arrow API benefits}:
	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is just yet another arrow combinator. The resulting arrow can be used in the same way a potential serial version could be used. This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types. We can just \enquote{plug} in parallel parts into our sequential programs without having to change anything.
	\item \textbf{Abstraction}:
	With the |ArrowParallel| typeclass, we abstracted all parallel implementation logic away from the business logic. This gives us the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell. So as an example, during development, we can run the code on the simple Multicore version and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
\end{itemize}

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel |map| with and without load balancing, |map-reduce|) in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|). Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} discusses future work and concludes.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-05-25 paper/content/sources/style/jfp2egui.tex
moving things around, reworking the makefile

% This is file JFP2egui.tex
% release v1.02, 27th September 2001
%   (based on JFPguide.tex v1.11 for LaTeX 2.09)
% Copyright (C) 2001 Cambridge University Press

\NeedsTeXFormat{LaTeX2e}

\documentclass{jfp1}

%%% Macros for the guide only %%%
\providecommand\AMSLaTeX{AMS\,\LaTeX}
\newcommand\eg{\emph{e.g.}\ }
\newcommand\etc{\emph{etc.}}
\newcommand\bcmdtab{\noindent\bgroup\tabcolsep=0pt%
  \begin{tabular}{@{}p{10pc}@{}p{20pc}@{}}}
\newcommand\ecmdtab{\end{tabular}\egroup}
\newcommand\rch[1]{$\longrightarrow\rlap{$#1$}$\hspace{1em}}
\newcommand\lra{\ensuremath{\quad\longrightarrow\quad}}

\title[Journal of Functional Programming]
      {\LaTeXe\ guide for the Journal of Functional Programming}

 \author[M. A. Reed]
        {MARK A. REED\\
         Cambridge University Press, Cambridge CB2 2BS, UK\\
         \email{texline@cambridge.org}}

\jdate{September 2001, update April 2007}
\pubyear{2001}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\doi{S0956796801004857}

\newtheorem{lemma}{Lemma}[section]

\begin{document}

\label{firstpage}

\maketitle

\begin{abstract}
This guide is for authors who are preparing papers for the \emph{Journal of
Functional Programming} using the \LaTeXe\ document-preparation system
and the Functional Programming class file (\texttt{jfp1.cls}).
\end{abstract}

\tableofcontents

\section{Introduction}

In addition to the standard submission of hardcopy from authors, the
journal now accepts machine-readable forms of papers
in \LaTeXe. The layout design for the \emph{Functional Programming} journal
has been implemented as a \LaTeXe\ class file, based on the \verb"article"
class as discussed in the \LaTeX\ manual (2nd edition) \cite{LaTeX}.
Commands which differ from the standard \LaTeXe\ interface, or which are
provided in addition to the standard interface, are explained in this
guide (which is \emph{not} a substitute for the \LaTeXe\ manual itself).

Note that the final printed version of papers will use the Monotype Times
typeface rather than the Computer Modern typeface available to authors. For
this reason line and page breaks will change and authors should not insert
hard breaks in their text.

Authors planning to submit their papers in \LaTeXe\ are advised to use
\verb"jfp1.cls" as early as possible in the creation of their files.

The older \LaTeX\ style file is no longer supported, but authors should
find it easy to convert to \LaTeXe\ (see Section~\ref{usingjfpclass}).

\subsection{Introduction to \LaTeX}

\LaTeX\ is constructed as a series of macros on top of the \TeX\ typesetting
program. \LaTeX\ adds to \TeX\ a collection of facilities which simplify
typesetting for authors by allowing them to concentrate on the logical
structure of the document rather than its visual layout. Careful use of the
\LaTeX\ mark-up philosophy results in uniform layout rather than the
\emph{ad hoc} results of some word-processing systems. Authors are advised to
let the defaults control font selection etc., rather than tinker themselves.

The \LaTeX\ system provides a consistent and comprehensive document preparation
interface. Among other things, \LaTeX\ can automatically number list
entries, equations, figures, tables and footnotes, as well as sections and
subsections. Using this numbering system, bibliographic citations, page
references and cross references to any other numbered entity (e.g. sections,
equations, figures) are straightforward.

\subsection{The JFP document class}

The use of document classes allows a simple change of style (or style option)
to transform the appearance of your document. The JFP class preserves the
standard \LaTeX\ interface such that any document which can be produced
using the standard \LaTeX\ \verb"article" class can also be produced with
the JFP class.
However, the measure (or width of text) is different from that
for ARTICLE; therefore line breaks will change and it is possible that
longer equations may need re-setting.
Authors are urged to use \verb"jfp1.cls" from the beginning of their document
preparation, otherwise longer lines may require re-formatting at a later stage.

\subsection{General style issues}

Use of \LaTeX\ defaults will result in a pleasing uniformity of layout
and font selection. Authors should resist the temptation to make
\emph{ad hoc} changes to these. Also avoid use of direct formatting unless
really necessary. Papers will be edited as usual, and this process may be
obstructed by the use of inserted line breaks, etc.

For general style issues, authors are referred to the `Preparation of
manuscripts' in the back cover of the journal. Authors who are interested in
the details of style are referred to \cite{Butcher} and \cite{Chicago}. The
language used in the journal is British English, and spelling should conform
to this.

Use should be made of symbolic references (\verb"\ref") in order to
protect against late changes of order, etc.

\subsection{Submission of \LaTeX\ articles}

Authors who intend to submit a \LaTeX\ article should obtain a copy of the
JFP class file. This is available by anonymous FTP from
%
\begin{verbatim}
  ftp.cup.cam.ac.uk
\end{verbatim}
%
You will find the class file and instructions contained in a
in the directory
%
\begin{verbatim}
  pub/texarchive/journals/latex/jfp-cls
\end{verbatim}
%
The \verb"readme.txt" (which is the same directory) gives more details.

When submitting the final article, ensure that  the following are included and
are clearly labelled.
\begin{enumerate}
  \item A hardcopy printout of the article.
  \item The input file (exactly matching the hardcopy).
  \item A copy of any user-defined macros.
  \item If you have used \textsc{Bib}\TeX, the \verb".bib", \verb".bbl"
        and \verb".bst" files that were used.
  \item Any other files necessary to prepare the article for typesetting.
\end{enumerate}
The source files for the \emph{final} article should be text-only with no
system-dependent control codes, via email as an attachment, along with
a pdf file that matches the source files exactly.

\section{Using the JFP1 class file}
\label{usingjfpclass}

First, copy the file \verb"jfp1.cls" (and \verb"jfp.bst" if you use Bib\TeX)
into an appropriate subdirectory on your system. The JFP class is implemented
as a complete document class, and \emph{not} as an class option.
In order to use the JFP class, replace \verb"article" by \verb"jfp1" in the
\verb"\documentclass" command at the beginning of your document: that is,
%
\begin{verbatim}
  \documentclass{article}
\end{verbatim}
%
is replaced by
%
\begin{verbatim}
  \documentclass{jfp1}
\end{verbatim}
%
Author-defined macros should be inserted before \verb"\begin{document}",
or in a separate file and should be included with the submission.
Authors must not change any of the macro definitions or parameters
in \verb"jfp1.cls".

If you have a document prepared using an old \verb"jfp.sty" file,
just change the line
\begin{verbatim}
  \documentstyle{jfp}
\end{verbatim}
to
\begin{verbatim}
  \documentclass{jfp1}
\end{verbatim}
The latter form uses the \verb"jfp1.cls" file.

\subsection{Document class options}\label{sec:ClassOp}

In general, the following standard document class options should \emph{not} be
used with the JFP class file:
%
\begin{itemize}
  \item \texttt{10pt}, \texttt{11pt} and \texttt{12pt} -- unavailable;
  \item \texttt{twoside} is the default (\texttt{oneside} is disabled);
  \item \texttt{onecolumn} is the default (\texttt{twocolumn} is disabled);
  \item \texttt{titlepage} is not required and is disabled;
  \item \texttt{fleqn} and \texttt{leqno} should not be used, and are disabled.
\end{itemize}
%
\ifprodtf
The following new class options are provided:
\begin{itemize}
  \item \texttt{prodtf} -- tells the class file that we want to use the
    production typeface. This automatically sets the odd, even and top
    margins.
\end{itemize}
\fi

\section{Additional facilities}

In addition to all the standard \LaTeX\ design elements, the JFP class
includes the following features.
%
\begin{itemize}
  \item Additional commands for typesetting the title page. Extended
        commands for specifying a short version of the title and author(s)
        for the running headlines.
  \item New options to the \verb"\maketitle" command to create Functional and
        Theoretical Pearl(s).
  \item A \verb"proof" environment.
  \item A \verb"capsule" environment for typesetting Capsule Reviews.
  \item Control of enumerated lists.
\end{itemize}
%
Once you have used these additional facilities in your document,
it can be processed only with \verb"jfp1.cls".

\subsection{Titles, authors' names and running headlines}

At the beginning of your article, the title should be generated in the
usual way using the \verb"\maketitle" command. Immediately following the
title you may include an abstract and/or capsule review. For example, the titles
for this guide were produced by the following source.
%
\begin{verbatim}
  \title[Journal of Functional Programming]
        {\LaTeXe\ guide for the Journal of Functional Programming}
  \author[M. A. Reed]
         {MARK A. REED\\
          Cambridge University Press, Cambridge CB2 2BS, UK\\
          \email{texline@cup.cam.ac.uk}}
\end{verbatim}

\verb"\begin{document}"\\
\indent\verb"\maketitle"

\begin{verbatim}
  \begin{abstract}
  This guide is for authors who are preparing papers...
  \end{abstract}
\end{verbatim}

In the JFP1 class, the title of the article and the author's name (or
authors' names) are used both at the beginning of the article for the main
title and throughout the article as running headlines at the top of every
page. The title is used on odd-numbered pages (rectos) and the author's name
appears on even-numbered pages (versos). The \verb"\pagestyle" and
\verb"\thispagestyle" commands should \emph{not} be used.  Similarly, the
commands \verb"\markright" and \verb"\markboth" should not be necessary.

Although the article title can run to several lines of text, the running
headline must be a single line. Moreover, the title can incorporate new-line
commands (e.g. \verb"\\"), but these are not acceptable in a running headline.
To enable you to specify an alternative short title, and an alternative short
author's name, the standard \verb"\title" and \verb"\author" commands have
been extended to take an optional argument to be used as the running headline.
%
\begin{verbatim}
  \title[Short title]
        {Full title which can be as long as necessary}
  \author[Author name]
         {AUTHOR NAME \\ Affiliation}
\end{verbatim}
%
Notice that the author name in the argument for the running head should
be in mixed case, and the author name for the title should be in upper
case only. The author affiliation is set in the normal way, after a
\verb"\\" in the argument to the \verb"\author" command.

Any `work supported by' or `authors current address' information should
be inserted via \verb"\thanks" commands, which should be positioned
after the appropriate `AUTHOR NAME' in the \verb"\author" command.

If there are four (or more) authors for the article, the author running head
should contain the first author name followed by `et~al.' only. e.g.
%
\begin{verbatim}
  \author[Author1 et al.]
         {AUTHOR1...}
\end{verbatim}

The previous examples show an article with one author, the normal
\LaTeX\ conventions have been extended to allow the author names and
their affiliations to be typeset in the correct JFP style. The following
examples should cover most possibilities:

\textit{Case 1.} Two authors with the same affiliation:
%
\begin{verbatim}
  \author[Author1 and Author2]
         {AUTHOR1 and AUTHOR2\\
          Affiliation for both authors}
\end{verbatim}
%
If the author names are too long to fit onto one line, it should be
broken into two or more lines using the \verb"\authorbreak" command.
Don't use \verb"\\" to linebreak the author names -- as this will not do what
you expect.

\textit{Case 2.} Two authors with different affiliations:
%
\begin{verbatim}
  \author[Author1 and Author2]
         {AUTHOR1\\
          Affiliation for Author1
          \and AUTHOR2\\
          Affiliation for Author2}
\end{verbatim}

\textit{Case 3.} Three (or more) authors, two with the same affiliation:
%
\begin{verbatim}
  \author[Author1, Author2 and Author3]
         {AUTHOR1, AUTHOR2\\
            Affiliation for Author1 and Author2
          \and AUTHOR3\\
            Affiliation for Author3}
\end{verbatim}

\subsection{Pearls}

To provide for the Theoretical Pearl and Functional Pearl styles,
\verb"\maketitle" can now take an optional argument, which can take the
values \verb"t" and \verb"f". Thus, \verb"\maketitle[t]" will make a
title page, and set up the running headline, for a \linebreak Theoretical
Pearls article; \verb"\maketitle[f]" will do the same for a Functional Pearls
article.
If you need to have a `Theoretical Pearl' or `Functional Pearl' (singular)
you can use the \verb"T" and \verb"F" options to obtain these.

\subsubsection{Other `Pearl' styles}

If you need to have a `Pearl' style which cannot be generated by the
standard \verb"t", \verb"f", \verb"T" or \verb"F" options, then the \verb"o"
option can be used.

The \verb"o" option allows you to define a custom `Pearl' style. \eg for
`CUSTOM PEARL' you would use:
%
\begin{verbatim}
  \begin{document}
  \renewcommand\otherpearl{C\ls U\ls S\ls T\ls O\ls M\ns
                        P\ls E\ls A\ls R\ls L}
  \maketitle[o]
  \shorttitle{Custom pearl}
\end{verbatim}
%
Notice the use of the \verb"\ls" and \verb"\ns" commands to letter-space the
words in the title.
You can use the \verb"\\" command to line-break the title as normal.
The \verb"\shorttitle" command should appear \emph{after} the
\verb"\maketitle" command (as above).

\subsection{Proofs}

A new environment exists for creating Proofs, e.g.
%
\begin{proof}
  Use $K_\lambda$ and $S_\lambda$ to translate combinators
  into $\lambda$-terms. For the converse, translate
  $\lambda x$ \ldots by [$x$] \ldots and use induction
  and the lemma.
\end{proof}
%
This was produced by the following code:
%
\begin{verbatim}
  \begin{proof}
    Use $K_\lambda$ and $S_\lambda$ to...
  \end{proof}
\end{verbatim}
%
The end of proof marker \proofbox\ is produced automatically. If you wish
to omit this, use the \verb"proof*" environment instead.

If a proof ends with a display equation, then it is customary for the proofbox
to be positioned at the end of equation finishing the proof. \eg
%
\begin{proof*}
  Use $K_\lambda$ and $S_\lambda$ to translate combinators
  into $\lambda$-terms. For the converse, translate
  $\lambda x$ \ldots by [$x$] \ldots and use induction
  and the lemma.
  \[ a_1 \equiv (2\Omega M^2/x) \mathproofbox \]
\end{proof*}
%
Was produced with:
%
\begin{verbatim}
  \begin{proof*}
    Use $K_\lambda$ and $S_\lambda$ to...
    \[ a_1 \equiv (2\Omega M^2/x) \mathproofbox \]
  \end{proof*}
\end{verbatim}
%
Notice the use of \verb"proof*" to turn off the automatic proofbox.

The proof environment will also take an optional argument which allows you
to produce `special' proofs. e.g.
%
\begin{proof}[Proof of Theorem 27]
  We define a linear isometry $A:H\rightarrow H$ by the nonlinear
  Schr\"{o}dinger equation. It would not be hard to modify the
  proof to obtain an analogous result for ellipsoids rather than
  spheres.
\end{proof}
%
Which was produced like this:
%
\begin{verbatim}
  \begin{proof}[Proof of Theorem 27]
    We define a linear isometry...
  \end{proof}
\end{verbatim}
%
Notice that once the optional argument is used, you have to type all of
the text which is to appear as the heading.

\subsection{Programs}\label{sec-programs}

JFP encourages authors to use one of two styles for typesetting programs,
\emph{mathematical} and \emph{verbatim}.

A program typeset in the mathematical style is shown in
Figure~\ref{mathfigure}, and the commands used to typeset this program
are shown in Figure~\ref{mathtypeset}.  This uses the ordinary
mathematics mode of \LaTeX: displayed programs are surrounded by
\verb|\[| and \verb|\]|, and the \verb|array| command is used for
alignment.  However, there are two important differences.  First, the
\verb|\programmath| command appears before the program text; this
causes math mode to use ordinary spacing for italic identifiers,
rather than math spacing.  The \verb|\unprogrammath| command returns
to normal math spacing.  Second, in math mode spaces are ignored, so a
tilde \verb|~| is used instead.  (In \LaTeX, a tilde generates a
``hard space'' that is never replaced by a line break.)  To include
program text in mathematics style inline, surround it with dollar
\verb|$| signs.  For example, the input
\begin{verbatim}
  See how \programmath $differ~x$
  differs from \unprogrammath $differ x$.
\end{verbatim}
produces the output
\begin{center}
  See how \programmath $differ~x$
  differs from \unprogrammath $differ x$.
\end{center}

A program typeset in the verbatim style is shown in
Figure~\ref{verbfigure}, and the commands used to typeset this program
are shown in Figure~\ref{verbtypeset}.  This uses the ordinary
verbatim mode of \LaTeX: displayed programs are surrounded by
\verb|\begin{verbatim}| and \verb|\end{verbatim}|, and alignment is
indicated with spaces in the source file (don't use tabs, which may
not be processed properly).  To include program
text in verbatim style inline, use the \verb|\verb| command.
For example, the input
\begin{center}
\begin{verbatim}
  On a terminal, this looks like \verb"differ x".
\end{verbatim}
\end{center}
produces the output
\begin{center}
  On a terminal, this looks like \verb"differ x".
\end{center}

It is recommended that programs in figures be offset from the text
using the \verb|\figrule| command, as shown in Figures~\ref{mathfigure}%
--\ref{verbtypeset}.

\begin{figure}
\figrule
\programmath
\[
\begin{array}{lcl}
exp                   & :: & Exp \rightarrow Arr \rightarrow Val \\
exp~(Var~i)~a         & =  & index~i~a \\
exp~(Const~v)~a       & =  & v \\
exp~(Plus~e_1~e_2)~a  & =  & exp~e_1~a+exp~e_2~a \\
\\
com                   & :: & Com \rightarrow Arr \rightarrow Arr \\
com~(Asgn~i~e)~a      & =  & update~i~(exp~e~a)~a \\
com~(Seq~c_1~c_2)~a   & =  & com~c_2~(com~c_1~a) \\
com~(If~e~c_1~c_2)~a  & =  & \textrm{if } exp~e~a \dequals 0 \textrm{ then }
                                      com~c_1~a \textrm{ else } com~c_2~a \\
\\
prog                  & :: & Prog \rightarrow Val\\
prog~(Prog~c~e)       & =  & exp~e~(com~c~(newarray~0))
\end{array}
\]
\unprogrammath
\caption{Example program in mathematical style.}\label{mathfigure}
\figrule
\begin{center}
\begin{verbatim}
\begin{figure}
\figrule
\programmath
\[
\begin{array}{lcl}
exp                   & :: & Exp \rightarrow Arr \rightarrow Val \\
exp~(Var~i)~a         & =  & index~i~a \\
exp~(Const~v)~a       & =  & v \\
exp~(Plus~e_1~e_2)~a  & =  & exp~e_1~a+exp~e_2~a \\
\\
com                   & :: & Com \rightarrow Arr \rightarrow Arr \\
com~(Asgn~i~e)~a      & =  & update~i~(exp~e~a)~a \\
com~(Seq~c_1~c_2)~a   & =  & com~c_2~(com~c_1~a) \\
com~(If~e~c_1~c_2)~a  & =  & \textrm{if } exp~e~a \dequals 0 \textrm{ then }
                                      com~c_1~a \textrm{ else } com~c_2~a \\
\\
prog                  & :: & Prog \rightarrow Val\\
prog~(Prog~c~e)       & =  & exp~e~(com~c~(newarray~0))
\end{array}
\]
\unprogrammath
\caption{Example program in mathematical style.}\label{mathfigure}
\figrule
\end{verbatim}
\end{center}
\caption{Typesetting the example program in mathematical style.}\label{mathtypeset}
\figrule
\end{figure}

\begin{figure}
\figrule
\begin{verbatim}
  exp                 :: Exp -> Arr -> Val
  exp (Var i) a       =  index i a
  exp (Const v) a     =  v
  exp (Plus e1 e2) a  =  exp e1 a + exp e2 a

  com                 :: Com -> Arr -> Arr
  com (Asgn i e) a    =  update i (exp e a) a
  com (Seq c1 c2) a   =  com c2 (com c1 a)
  com (If e c1 c2) a  =  if  exp e a == 0  then  com c1 a  else  com c2 a

  prog                :: Prog -> Val
  prog (Prog c e)     =  exp e (com c (newarray 0))
\end{verbatim}
\caption{Example program in verbatim style.}\label{verbfigure}
\figrule
\begin{tabular}{l}
\verb|\begin{figure}| \\
\verb|\figrule| \\
\verb|\begin{center}| \\
\verb|\begin{verbatim}| \\
\verb|exp                 :: Exp -> Arr -> Val| \\
\verb|exp (Var i) a       =  index i a| \\
\verb|exp (Const v) a     =  v| \\
\verb|exp (Plus e1 e2) a  =  exp e1 a + exp e2 a| \\
\\
\verb|com                 :: Com -> Arr -> Arr| \\
\verb|com (Asgn i e) a    =  update i (exp e a) a| \\
\verb|com (Seq c1 c2) a   =  com c2 (com c1 a)| \\
\verb|com (If e c1 c2) a  =  if  exp e a == 0  then  com c1 a  else  com c2 a| \\
\\
\verb|prog                :: Prog -> Val| \\
\verb|prog (Prog c e)     =  exp e (com c (newarray 0))| \\
\verb|\end{verbatim}| \\
\verb|\end{center}| \\
\verb|\caption{Example program in verbatim style.}\label{verbfigure}| \\
\verb|\figrule| \\
\verb|\end{figure}|
\end{tabular}
\caption{Typesetting the example program in verbatim style.}\label{verbtypeset}
\figrule
\end{figure}

Some new macros have been provided for a few convenient symbols in
math mode.  These are illustrated in Table~\ref{newcommands}.
\ifprodtf\newpage\fi

\begin{table}
\caption{New symbol macros}\label{newcommands}
\programmath
\begin{tabular}{lll}
\hline\hline
Symbol                & Usage                  & Keyed as\\
\hline
\verb"\dplus"         & $abc \dplus xyz$         & \verb"$abc \dplus xyz$"\\
\verb"\dequals"       & $abc \dequals xyz$       & \verb"$abc \dequals xyz$"\\
\verb"\dcolon"        & $abc \dcolon xyz$        & \verb"$abc \dcolon xyz$"\\
\verb"\dcolonequals"  & $abc \dcolonequals xyz$  & \verb"$abc \dcolonequals xyz$"\\
\hline\hline
\end{tabular}
\unprogrammath
\end{table}

\subsection{Abstract and Capsule reviews}

The JFP class provides for an abstract and/or a capsule review; the
abstract is produced by the following commands:
%
\begin{verbatim}
  \begin{abstract}
       :
  \end{abstract}
\end{verbatim}
%
whereas the capsule review is produced by:
%
\begin{verbatim}
  \begin{capsule}
       :
  \end{capsule}
\end{verbatim}
%
Either or both of these may be used, in either order, but it is assumed
that, if both are used, there will be no other material between them.
In JFP the abstract should precede any capsule review.

\subsection{Lists}

The JFP class provides the three standard list environments.
\begin{itemize}
  \item Numbered lists, created using the \verb"enumerate" environment;
  \item Bulleted lists, created using the \verb"itemize" environment;
  \item Labelled lists, created using the \verb"description" environment.
\end{itemize}
The \verb"enumerate" environment numbers each list item with an arabic numeral;
alternative styles can be achieved by inserting a redefinition of the
number labelling command after the \verb"\begin{enumerate}". For example, a
list numbered with roman numerals inside parentheses can be produced by the
following commands:
%
\begin{verbatim}
  \begin{enumerate}[(iii).]
    \renewcommand{\theenumi}{(\roman{enumi})}
    \item first item
          :
  \end{enumerate}
\end{verbatim}
%
This produces the following list:
%
\begin{enumerate}[(iii).]
  \renewcommand{\theenumi}{(\roman{enumi})}
  \item first item
  \item second item
  \item \etc
\end{enumerate}
%
Notice that an optional argument ``\verb"(iii)."'' has been given to the
\verb"enumerate" environment, specifying the \emph{widest label} used in the
list. This is because roman numerals are wider than the arabic numerals
normally used by \verb"enumerate", and so the labels would otherwise have been
pushed out into the margin.

\section{User-defined macros}

If you define your own macros, you must ensure that their names do not
conflict with any existing macros in \LaTeX\ (or \AMSLaTeX\ %
if you are using this). You should also place them in the preamble of
your input file, between the \verb"\documentclass" (but after any
\verb"\usepackage" commands) and before the \verb"\begin{document}" command.

Apart from scanning the indexes of the relevant manuals, you can check
whether a macro name is already used by using \verb"\newcommand", which
will check for the existence of the macro you are trying to define.
If the macro exists \LaTeX\ will respond with:
%
\begin{verbatim}
  ! LaTeX Error: Command ... already defined.
\end{verbatim}
%
In this case you should choose another name, and try again.

Such macros must be in a place where they can easily be found and
modified by the journal's editors or typesetter. They must be gathered
together in the preamble of your input file, or in a separate
\verb"macros.tex" file with the command \verb"\input{macros}" in the
preamble. Macro definitions must not be scattered about your document
where they are likely to be completely overlooked by the typesetter.

The same applies to font definitions that are based on Computer Modern
fonts. These must be changed by the typesetter to use the journal's
correct typeface. In this case, you should draw
attention to these font definitions on the hard copy that you submit for
publication and by placing a comment in your input file just before the
relevant definitions, for example \verb"% replace font!"

\section{Some guidelines for using standard facilities}

The following notes may help you achieve the best effects with the JFP class file.

\subsection{Sections}

\LaTeX\ provides five levels of section headings and they are all
defined in the JFP class file:
\begin{itemize}
  \item[] Heading A -- \verb"\section{...}"
  \item[] Heading B -- \verb"\subsection{...}"
  \item[] Heading C -- \verb"\subsubsection{...}"
  \item[] Heading D -- \verb"\paragraph{...}"
  \item[] Heading E -- \verb"\subparagraph{...}"
\end{itemize}
Section numbers are given for sections, subsection and subsubsection headings.

\subsection{Figures and tables}

The \texttt{figure} and \texttt{table} environments are implemented as described in
the \LaTeX\ Manual to
provide consecutively numbered floating inserts for illustrations and tables
respectively.
The standard inserts and their captions are formatted centred.
Line breaks in captions can be inserted as required using \verb"\\".

\subsubsection{Illustrations (or figures)}

The JFP class will cope with most positioning of your illustrations
and you should not normally use the optional positional qualifiers on
the \verb"figure" environment which would override these decisions.
Figure captions should be below the figure itself, therefore the
\verb"\caption" command should appear after the figure or space left
for an illustration.

Figures in JFP will frequently illustrate programs, as shown in
section~\ref{sec-programs} of this guide. Figure~\ref{sample-figure}
shows an example of space left above a caption for artwork to be pasted in.
This was produced with the following commands:
%
\begin{verbatim}
  \begin{figure}
    \vspace{5cm} % the vertical depth of the artwork
    \caption{An example figure with space for artwork.}
    \label{sample-figure}
  \end{figure}
\end{verbatim}
%
\begin{figure}
  \vspace{5cm} % the vertical depth of the artwork
  \caption{An example figure with space for artwork.}
  \label{sample-figure}
\end{figure}
%
The vertical depth should correspond roughly to the artwork you will submit;
it will be adjusted to fit the final artwork exactly.

If your illustration extends over two~pages, you can use the
\verb"\continuedfigure" facility. To use this, you key the figure
caption for the second figure as follows:
%
\begin{verbatim}
  \begin{figure}
    \continuedfigure
    \vspace{80pt}
    \caption{First figure, continued.}
    \label{continued}
  \end{figure}
\end{verbatim}
%
This ensures that the figure counter does not get incremented, and at the
same time adds the word (cont.) to the caption. You may still use labels
and references for this figure.

\subsubsection{Tables}

The JFP class file will cope with most positioning of your tables
and you should not normally use the optional positional qualifiers on the
\verb"table" environment which would override these decisions.
Normal journal style sets the table caption first, followed by a double
rule, the table body and a double rule at the bottom. Single rules and
spanner rules (\verb"\cline") can be used to separate headings from the
columns. For example, Table~\ref{sample-table} is produced using the
following commands:\par
%
{\small
\begin{verbatim}
\begin{table}
  \caption{Results of Overloading for 3 Experimental Setups}
  \label{sample-table}
  \begin{minipage}{\textwidth}
    \begin{tabular}{lcrrrrr}
      \hline\hline
      Program& Expt.&
      CPU\footnote{Seconds of elapsed time on an unloaded Sun 3/50.}&
      RelCPU\footnote{CPU Time relative to experiment (a).}& GC&
      Mem\footnote{Bytes of heap used over the duration of the program.}&
      RelMem\footnote{Memory usage relative to experient (a).}\\
      \hline
      8 Queens& (a)&   2.88&  1.00&    6&   1.7M&  1.00\\
      &         (b)&  32.51& 11.29&  193&  48.9M& 28.76\\
      &         (c)&   7.90&  2.74&   42&  11.3M&  6.65\\
      \noalign{\vspace {.5cm}}
      Primes&   (a)&   4.89&  1.00&   19&   5.3M&  1.00\\
      &         (b)&  47.54&  9.72&  204&  54.5M& 10.28\\
      &         (c)&  10.08&  2.06&   47&  13.0M&  2.45\\
      \noalign{\vspace {.5cm}}
      Nfib&     (a)&  21.65&  1.00&  161&  40.4M&  1.00\\
      &         (b)& 221.65& 10.24& 1382& 349.0M&  8.64\\
      &         (c)&  21.30&  0.98&  161&  42.0M&  1.03\\
      \noalign{\vspace {.5cm}}
      KWIC&     (a)&   7.07&  1.00&   15&   6.3M&  1.00\\
      &         (b)&  34.55&  4.89&  109&  47.8M&  7.59\\
      &         (c)&  31.62&  4.47&   53&  45.0M&  7.14\\
      \hline\hline
    \end{tabular}
    \vspace{-2\baselineskip}
  \end{minipage}
\end{table}
\end{verbatim}}
%
\begin{table}
  \caption{Results of Overloading for 3 Experimental Setups}
  \label{sample-table}
  \begin{minipage}{\textwidth}
    \begin{tabular}{lcrrrrr}
      \hline\hline
      Program& Expt.&
      CPU\footnote{Seconds of elapsed time on an unloaded Sun 3/50.}&
      RelCPU\footnote{CPU Time relative to experiment (a).}& GC&
      Mem\footnote{Bytes of heap used over the duration of the program.}&
      RelMem\footnote{Memory usage relative to experient (a).}\\
      \hline
      8 Queens& (a)&   2.88&  1.00&    6&   1.7M&  1.00\\
      &         (b)&  32.51& 11.29&  193&  48.9M& 28.76\\
      &         (c)&   7.90&  2.74&   42&  11.3M&  6.65\\
      \noalign{\vspace {.5cm}}
      Primes&   (a)&   4.89&  1.00&   19&   5.3M&  1.00\\
      &         (b)&  47.54&  9.72&  204&  54.5M& 10.28\\
      &         (c)&  10.08&  2.06&   47&  13.0M&  2.45\\
      \noalign{\vspace {.5cm}}
      Nfib&     (a)&  21.65&  1.00&  161&  40.4M&  1.00\\
      &         (b)& 221.65& 10.24& 1382& 349.0M&  8.64\\
      &         (c)&  21.30&  0.98&  161&  42.0M&  1.03\\
      \noalign{\vspace {.5cm}}
      KWIC&     (a)&   7.07&  1.00&   15&   6.3M&  1.00\\
      &         (b)&  34.55&  4.89&  109&  47.8M&  7.59\\
      &         (c)&  31.62&  4.47&   53&  45.0M&  7.14\\
      \hline\hline
    \end{tabular}
    \vspace{-2\baselineskip}
  \end{minipage}
\end{table}

Notice the use of the `\verb"\vspace{-2\baselineskip}"' command to remove the
unwanted vertical space from above the table footnotes in this example.

Captions for `continued' tables can be generated (in the same way as for figures)
using the \verb"\continuedtable" command. These should be positioned just before
the \verb"\caption" command in the appropriate table environment.

The \verb"tabular" environment should be used to produce ruled tables;
it has been modified for the JFP class in the following ways:
\begin{enumerate}
  \item Additional vertical space is inserted above and below a horizontal rule
        (produced by \verb"\hline");
  \item Tables are centred, and span the full width of the page; that is,
  they are similar to the tables that would be produced by
  \verb"\begin{minipage}{\textwidth}".
\end{enumerate}
Because of this reformatting, vertical rules should not be used;
furthermore, commands to
redefine quantities such as \verb"\arraystretch" should be omitted. If
the old tabular facilities are needed, there is a new environment,
\verb"oldtabular", which has none of the reformatting; it should be used
in exactly the same way.

\subsection{Appendices}

You should use the standard \LaTeX\ \verb"\appendix" command to place any
Appendices, normally, just before any references.
From that point on \verb"\section" will produce an appendix, which are
numbered A, B etc., equations as (A1), (B1) etc.\ Figures and
tables also number A\,1, B\,1 etc.

\subsection{References}

As with standard \LaTeX, there are two ways of producing a list of references;
either by using Bib\TeX\ with the JFP bibliography style \verb"jfp.bst",
or by compiling a list of references by hand (using a
\verb"thebibliography" environment).

\subsubsection{Using Bib\TeX}

If you have Bib\TeX\ installed on your system, the following is a brief description
on how to automatically generate a bibliography (\verb".bbl" file) for your
article. Your article should contain at least the following elements:\par
\vspace{6pt}\noindent
%
\verb"  % sample.tex"\\[3pt]
\verb"  \documentclass{jfp}"\\[3pt]
\verb"  \bibliographystyle{jfp}"\\[3pt]
\verb"  \begin{document}"\\[3pt]
\verb"    \cite{"\textit{citations}\verb"}"\\[3pt]
\verb"    \bibliography{"\textit{biblio database files}\verb"}"\\[3pt]
\verb"  \end{document}"\\[6pt]
%
Where `\emph{biblio database files}' may be one or more filenames of bibliographic
database files (without the \verb".bib" extension) separated by commas.
First, \LaTeX\ the file \verb"sample.tex". Second, run Bib\TeX\ by typing:
%
\begin{verbatim}
  bibtex sample
\end{verbatim}
%
This creates the file \verb"sample.bbl". Third, re-\LaTeX\ your
document, and the newly-created \verb"sample.bbl" will be read in
and typeset. You will then need to \LaTeX\ the document once more to resolve any
unresolved citation references.

\subsubsection{Typesetting the references by hand}

The following listing shows some references prepared in the style of the
journal; this code produces the references at the end of this guide.
%
\begin{verbatim}
\begin{thebibliography}{}
 \bibitem[\protect\citename{Augustsson and Johnsson, }1987]{AJ187}
   Augustsson,~L. and Johnsson,~T. (1987) LML users' manual. PMG
   Report, Department of Computer Science, Chalmers University of
   Technology, Goteborg, Sweden.
 \bibitem[\protect\citename{Butcher, }1981]{Butcher}
   Butcher,~J. (1981) Copy-editing: the Cambridge handbook.
   Cambridge University Press.
 \bibitem[\protect\citename{Chicago, }1982]{Chicago}
   The Chicago manual of style. University of Chicago Press.
 \bibitem[\protect\citename{Conklin, }1987]{JC87}
   Conklin,~J. (1987) Hypertext: an introduction and survey.
   \emph{IEEE Computer}, 20~(9): pp.~17--41.
 \bibitem[\protect\citename{Dijkstra, }1976]{EWD76}
   Dijkstra,~E.~W. (1976) \emph{A Discipline of Programming}.
   Prentice-Hall.
 \bibitem[\protect\citename{Knuth, }1984]{DEK84}
   Knuth,~D.~E. (1984) Literate programming. \emph{BCS Comput. J.}
   27~(2): 97--111 (May).
 \bibitem[\protect\citename{Lamport, }1986]{LaTeX}
   Lamport,~L. (1986) \LaTeX: a document preparation system
   (2nd edition). Addison-Wesley, New York.
 \bibitem[\protect\citename{Reynolds, }1969]{JCR69}
   Reynolds,~J.~C. (1969) Transformation systems and the
   algebraic structure of atomic formulas. In B. Meltzer and
   D. Michie (editors), \emph{Machine Intelligence 5}, pp.~135--151.
   Edinburgh University Press.
 \bibitem[\protect\citename{Toyn \emph{et al.}, }1987]{TDR87}
   Toyn,~I., Dix,~A. and Runciman,~C. (1987) Performance
   polymorphism. In \emph{Functional Programming Languages and
   Computer Architecture, Lecture Notes in Computer Science, 274},
   pp.~325--346. Springer-Verlag.
\end{thebibliography}
\end{verbatim}
%
The above list is typeset at the end of this guide.
Each entry takes the form
%
\begin{verbatim}
\bibitem[\protect\citename{Author(s), }Date]{tag}
  Bibliography entry
\end{verbatim}
%
where \verb"Author(s)"\ should be the author names as they are cited in
the text (note the space before the closing \verb"}" of the \verb"\citename"
command is vital),
\verb"Date" is the date to be cited in the text, and \verb"tag"
is the tag that is to be used as an argument for the \verb"\cite{}" and
\verb"\shortcite{}" commands. \verb"Bibliography entry" should be the
material that is to appear in the bibliography, suitably formatted. This
rather unwieldy scheme makes up for the lack of an author-date system in
\LaTeX.

\subsubsection{Multiple references}

References should be listed alphabetically by author name(s) and
then by year if the same author has several papers. If some papers by
the same author(s) also fall in the same year, their dates should be
in the form (1993a), (1993b), \etc

Formatting for italic etc.\ should be avoided unless you are sure you
understand the style of references; please concentrate on giving full and
clear information.

\subsubsection{References in the text}

References in the text are given by author and date. Whichever method
is used to produce the bibliography, the references in
the text are done in the same way. Each bibliographical entry has a key,
which is assigned by the author and used to refer to that entry in the
text.
There is one form of citation -- \verb"\cite{key}" -- to produce the
author and date, and another form -- \verb"\shortcite{key}" -- which
produces the date only. Thus,
Augustsson and Johnsson \shortcite{AJ187} is produced by
%
\begin{verbatim}
  Augustsson and Johnsson \shortcite{AJ187},
\end{verbatim}
%
while \cite{TDR87} is produced by
%
\begin{verbatim}
  \cite{TDR87}.
\end{verbatim}

To allow further flexibility of the \verb"\cite" (and \verb"\shortcite") commands
the function of the optional argument has been changed. For example, the output of
the command \verb"\cite{TDR87}" gives \cite{TDR87}, but you may want to list all
of the author names (instead of just the first author and \emph{et~al.}).
In this case we would use:
\[ \hbox{\verb"\cite[(Toyn, Dix and Runciman, 1987)]{TDR87}"} \]
in the text to produce the desired effect.

\newpage
\appendix
\section{Special commands in {\mdseries\texttt{jfp1.cls}}}

The following is a summary of the new commands, optional arguments and
environments which have been added to the standard \LaTeX\ user-interface
in creating the JFP class file.\par
\vspace{8pt}

\bcmdtab
\emph{New commands}    & \\[3pt]
\verb"\authorbreak"    & allows a list of authors to be broken in to separate
                         lines without starting an affiliation.\\
\verb"\continuedfigure" & adds the word (cont.) to the next figure caption, and
                          also stops the figure counter from being stepped.\\
\verb"\continuedtable"  & as for \verb"\continuedfigure", except this command
                          achieves the same effect for tables.\\
\verb"\email"          & used to typeset an authors e-mail address (should only be
                         used in the \verb"\author" command).\\
\verb"\figrule"        & adds a rule, used around programs set in figure environments.\\
\verb"\ls", \verb"\ns" & used to letter-space the heading in a alternate `Pearl'
                         style (when used with the \verb"\maketitle[o]" argument).\\
\verb"\programmath"    & gives normal spacing for verbatim math mode.\\
\verb"\proofbox"       & typesets a proof box \proofbox\ (this is normally
                         put in automatically at the end of the \verb"proof"
                         environment). If you need to insert a \verb"\proofbox"
                         manually, you should add a `\verb"\quad"' of
                         space before it in the output.\\
\verb"\removebrackets" & removes the `(\ )' brackets from the optional
                          argument of environments created by the
                          \verb"\newtheorem" command. Should be placed
                          just before the appropriate environment.\\
\verb"\unprogrammath"  & reverts to math spacing.\\
\ifprodtf
\verb"\removefullpoint" & removes the full point from the next \verb"\caption"
                          command, usually used for figures when they are split
                          from their captions.\\
\fi
\verb"\shortcite"      &  typesets the `year' part of the bibliographic entry only.
                          \eg (1987).\\
\verb"\mathproofbox"   & as \verb"\proofbox", except this version is intended
                         for use in equations ending proof's (it typesets the
                         proof box using \verb"\rlap" with 1em of extra space).\\[6pt]
\emph{New environments} & \\[3pt]
\ifprodtf
\verb"bottomfigure"     & for split figures and captions (on facing page).\\
\fi
\verb"capsule"          & used to typeset an articles Capsule Review.\\
\verb"oldtabular"       & preserves the orginal tabular environment, which
                          has been modified to insert additional space above
                          and below an \verb"\hrule". The body of the environment
                          is centred with rules full out across the text measure.\\
\verb"proof"            & to typeset mathematical proofs, the \verb"*"-form omits
                          the proof box.
\ecmdtab \newpage \bcmdtab
\emph{New optional arguments} & \\[3pt]
\verb"[<short title>]"  & in the \verb"\title" command: to define a right
                          running headline that is different from the article
                          title. The \verb"\shorttitle" command also
                          achieves the same effect.\\
\verb"[<short author>]" & in the \verb"\author" command: to define a left
                          running headline that is different from the
                          authors' names as typeset at the article opening.
                          The \verb"\shortauthor" command also
                          achieves the same effect.\\
\verb"[<pearl style>]"  & Optional argument to the \verb"\maketitle" command
                          which allows Functional or Theoretical pearls to be
                          typeset.\\
\verb"[<widest label>]" & in \verb"\begin{enumerate}": to ensure the correct
                          alignment of numbered lists.
\ecmdtab

\ifprodtf
%
\section{Notes for editors}

This appendix contains additional information which may be useful to
those who are involved with the final production stages of an article.
Authors, who are generally not typesetting the final pages in the
journal's typeface (Monotype Times), do not need this information.

\subsection{Setting the production typeface}

The global \verb"\documentclass" option `\verb"prodtf"' sets up
\verb"jfp1.cls" to typeset in the production typeface -- Monotype Times.
e.g.
%
\begin{verbatim}
  \documentclass[prodtf]{jfp1}
\end{verbatim}

\subsection{Catchline commands}

To be placed in the preamble:
\begin{itemize}
  \item \verb"\jdate{September 2001}"
  \item \verb"\pagerange{35--48}"
  \item \verb"\pubyear{2001}"
  \item \verb"\volume{\textbf{8} (1):}"
  \item \verb"\doi{S0956796801004857}"
\end{itemize}

\subsection{Landscape material}

The add on package \verb"JFPland" provides macros for landscape figures and
tables. See the \verb"JFPland" guide for further information.

\subsection{Figures with split artwork/captions}

When a figure is too large to fit on a page with its caption, you can use
the following procedure to place the figure, and then its caption at the foot
of the facing page. First set the figure with a short caption
using the normal \verb"figure" environment, e.g.
\begin{verbatim}
  \begin{figure}
    ...
    \caption{For caption see facing page.}
  \end{figure}
\end{verbatim}
Then set the correct (long) caption, so that it appears on the facing page:
\begin{verbatim}
  \begin{bottomfigure}
    \addtocounter{figure}{-1}
    \caption{This is the long caption...}
  \end{bottomfigure}
\end{verbatim}
If the figure falls on a recto, you may have to move the \verb"bottomfigure"
environment to before the \verb"figure" environment. In this case you need to
move the \verb"\addtocounter" command into the \verb"figure" environment
instead.

The \verb"bottomfigure" environment places a full measure rule above the
bottom-caption automatically.

\subsection{Editing citations (when the author has used the
 {\mdseries\textup{\texttt{cite}}} command)}

In the past when an automatic \verb"\cite" command produced text in the output
which needed to be changed, the argument (in [ ]) from the bibliography entry
was copied to the location of the \verb"\cite" command and then modified.
The \verb"\cite" command would then be removed as part of this process.

In the near future, we will probably have to supply \TeX\ output which will
need to contain `PDF marks' for interactive browsing.  Clearly by removing
the automatic link to the bibliographic entry (referenced by the \verb"\cite"),
we are making extra work for ourselves later on.

To avoid this, the function of the \verb"\cite" command's optional argument
has been changed. For example, the \verb"\cite" command for the
`\verb"JC87"' entry gives:
\[ \hbox{\cite{JC87}} \]
but you want the following to appear in the text:
\[ \hbox{\cite[(Conklin, 1987 -- see p.~22)]{JC87}} \]
you would then use:
\[ \hbox{\verb"\cite[(Conklin, 1987 -- see p.~22)]{JC87}"} \]
to obtain the desired result. Notice that you have to supply
the round brackets as well in the optional argument.
This modification also works for the \verb"\shortcite" command.

\section{Macros provided by {\mdseries\texttt{JFP2esym.sty}}}

\subsection{Automatic font/character changes}

\begin{itemize}\itemsep=6pt
\item The \verb|\le|, \verb|\leq|, \verb|\ge|, \verb|\geq| commands
use the equivalent AMS slanted symbols:
\[
\oldle \oldleq \oldge \oldgeq
 \lra
\le \leq \ge \geq
\]
The normal characters can be obtained by using the \verb|\old| form
(\eg \verb|\oldge|).
\end{itemize}

\subsection{Additional fonts}

\begin{itemize}\itemsep=6pt
\item The complete (v1) AMS symbols are available using the normal names:
\[
\hbox{\verb"\boxdot \boxplus \boxtimes"} \lra
  \boxdot \boxplus \boxtimes
\]

\item Blackboard bold:
\[
\hbox{\verb"$\mathbb{ABC}$"} \lra \mathbb{ABC}
\]

\item Fraktur/Gothic (bold math version available):
\[
   \hbox{\verb"$\mathfrak{ABC}$"} \lra \mathfrak{ABC}
\]

\item Bold math italic/symbols are provided by the \verb"\boldsymbol" macro
(from the \verb"amsbsy" package). The \verb"\bmath" macro is provided as
an alias.

\strut\hspace*{11pt}The \verb"JFP2esym" package also defines most of the
symbols from Appendix F of the \TeX book. These can be obtained by using
their normal (unbold) symbol name prefixed with a `b'. \eg \verb|\nabla|
becomes \verb|\bnabla|. The only exception to this rule is \verb|\eta|,
which whould lead to a clash with \verb|\beta|. In this case use
\verb|\boldeta| for bold eta.

\strut\hspace*{11pt}The problem with space disappearing around certain
bold math symbols (\verb"\bcdot") does not happen, when using the
\verb"amsbsy" package.

\item Sans serif symbols\\[6pt]
%
\verb"  \textsf{text}  " \lra \textsf{text}
  \qquad \verb"\mathsf{math}  " \lra $\mathsf{math}$\\
\verb"  \textsfi{text} " \lra \textsfi{text}
  \qquad \verb"\mathsfi{math} " \lra $\mathsfi{math}$\\
\verb"  \textsfb{text} " \lra \textsfb{text}
  \qquad \verb"\mathsfb{math} " \lra $\mathsfb{math}$\\
\verb"  \textsfbi{text}" \lra \textsfbi{text}
  \qquad \verb"\mathsfbi{math}" \lra $\mathsfbi{math}$
\end{itemize}

\subsection{Additional symbols}

\verb"  \colonequal   " \rch \colonequal \qquad \verb"\equalcolon     " \rch \equalcolon\\
\verb"  \colonequiv   " \rch \colonequiv \qquad \verb"\equivcolon     " \rch \equivcolon
%
\fi

\begin{thebibliography}{}
 \bibitem[\protect\citename{Augustsson and Johnsson, }1987]{AJ187}
   Augustsson,~L. and Johnsson,~T. (1987) LML users' manual. PMG
   Report, Department of Computer Science, Chalmers University of
   Technology, Goteborg, Sweden.
 \bibitem[\protect\citename{Butcher, }1981]{Butcher}
   Butcher,~J. (1981) Copy-editing: the Cambridge handbook.
   Cambridge University Press.
 \bibitem[\protect\citename{Chicago, }1982]{Chicago}
   The Chicago manual of style. University of Chicago Press.
 \bibitem[\protect\citename{Conklin, }1987]{JC87}
   Conklin,~J. (1987) Hypertext: an introduction and survey.
   \emph{IEEE Computer}, 20~(9): pp.~17--41.
 \bibitem[\protect\citename{Dijkstra, }1976]{EWD76}
   Dijkstra,~E.~W. (1976) \emph{A Discipline of Programming}.
   Prentice-Hall.
 \bibitem[\protect\citename{Knuth, }1984]{DEK84}
   Knuth,~D.~E. (1984) Literate programming. \emph{BCS Comput. J.}
   27~(2): 97--111 (May).
 \bibitem[\protect\citename{Lamport, }1986]{LaTeX}
   Lamport,~L. (1986) \LaTeX: a document preparation system
   (2nd edition). Addison-Wesley, New York.
 \bibitem[\protect\citename{Reynolds, }1969]{JCR69}
   Reynolds,~J.~C. (1969) Transformation systems and the
   algebraic structure of atomic formulas. In B. Meltzer and
   D. Michie (editors), \emph{Machine Intelligence 5}, pp.~135--151.
   Edinburgh University Press.
 \bibitem[\protect\citename{Toyn \emph{et al.}, }1987]{TDR87}
   Toyn,~I., Dix,~A. and Runciman,~C. (1987) Performance
   polymorphism. In \emph{Functional Programming Languages and
   Computer Architecture, Lecture Notes in Computer Science, 274},
   pp.~325--346. Springer-Verlag.
\end{thebibliography}

\label{lastpage}

\end{document}

% end of JFP2egui.tex

### 2017-05-25 paper/content/sources/topologySkeletons.tex
moving things around, reworking the makefile

%\FloatBarrier
\section{Topological Skeletons}
\label{sec:topology-skeletons}
Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons. The Eden library leverages this problem and already comes\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} with more predefined skeletons, among them a |pipe|, a |ring|, and a |torus| implementations \cite{Loogen2012}. These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell. We aim to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.

\subsection{Parallel pipe}

The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the |ArrowLoop| typeclass which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\mbcomment{das kann man hier so lassen, oder?}\olcomment{sicherlich!}
\begin{code}
loop (arr (\(a, b) -> (b, a:b)))
\end{code}
which is the same as
\begin{code}
loop (arr snd &&& arr (uncurry (:)))
\end{code}
defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it. Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |parEvalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
\begin{figure}[h]
\begin{code}
pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
	conf -> [arr a a] -> arr a a
pipeSimple conf fs =
	loop (arr snd &&&
		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
	arr last

lazy :: (Arrow arr) => arr [a] [a]
lazy = arr (\ ~(x:xs) -> x : lazy xs)
\end{code}
\caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows. Note that the use of |lazy| is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input |[a]| terminates before passing it into |parEvalN|.}
\label{fig:pipeSimple}
\end{figure}

However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}. Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of |pipe|: % ~\ref{fig:pipe}.
% \begin{figure}[h]
\begin{code}
pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
	conf -> [arr a a] -> arr a a
pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
\end{code}
% \caption{Final definition of the pipe skeleton which uses Futures}
% \label{fig:pipe}
% \end{figure}

Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. |arr a b| and |arr b c|. By wrapping these two arrows inside a common type we obtain |pipe2| (Fig.~\ref{fig:pipe2}).
\begin{figure}[h]
\olcomment{I swapped the type classes here:}
\begin{code}
pipe2 :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
         ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g =
	(arr return &&& arr (const [])) &&& arr (const []) >>>
	pipe conf (replicate 2 (unify f g)) >>>
	arr snd >>> arr head where
			unify :: (ArrowChoice arr) => arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
			unify f g =
				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
				arr (\((a, b), c) -> ((c, a), b))
\end{code}
\caption{Definition of |pipe2|.}
\label{fig:pipe2}
\end{figure}

Note that extensive use of this combinator over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN. Nonetheless, we can define a parallel piping operator |parcomp| (Fig.~\ref{fig:par>>>}, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Section~\ref{syntacticSugar}.
\begin{figure}[h]
\olcomment{swapped type classes}
\begin{code}
(|>>>|) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
	arr a b -> arr b c -> arr a c
(|>>>|) = pipe2 ()
\end{code}
\caption{Definition of |parcomp|.}
\label{fig:par>>>}
\end{figure}

\subsection{Ring skeleton} \label{sec:ring}
\begin{figure}[h]
	\includegraphics[scale=0.75]{images/ring}
	\caption{Schematic depiction of the ring skeleton}
	\label{fig:ringImg}
\end{figure}
Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function |[i] -> [o]| with a ring of nodes that communicate in a ring topology with each other. Its input is a node function |i -> r -> (o, r)| in which |r| serves as the intermediary output that gets send to the neighbour of each node. This data is sent over direct communication channels, so called \enquote{remote data}. We depict it in Appendix, Fig.~\ref{fig:ringEden}.
%\end{code}
%with toRD (to make use of remote data)
%\begin{code}
%\end{code}
%and rightRotate:
%\begin{code}

We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an arrow, already fits quite neatly into the |loop|'s |arr (a, b) (c, b) -> arr a c|. In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)|. Similarly to the |pipe| (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our |lazy| arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
\begin{quote}
{Note that the list of ring inputs |ringIns| is the same as the list of ring outputs |ringOuts| rotated by one element to the right using the auxiliary function |rightRotate|. Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
\end{quote}
Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|. Finally we unzip this list into |([o], [fut r])|. Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
%Further details on this can be found in \cite{eden_cefp}.
\olcomment{let's do it?}

\begin{figure}[h]
\begin{code}
ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
    conf -> arr (i, r) (o, r) -> arr [i] [o]
ring conf f =
	loop (second (rightRotate >>> lazy) >>> arr (uncurry zip) >>>
        parMap conf (second get >>> f >>> second put) >>> arr unzip)

rightRotate :: (Arrow arr) => arr [a] [a]
rightRotate = arr $ \list -> case list of
  [] -> []
  xs -> last xs : init xs

lazy :: (Arrow arr) => arr [a] [a]
lazy = arr (\ ~(x:xs) -> x : lazy xs)
\end{code}
\caption{Final definition of the |ring| skeleton.}
\label{fig:ringFinal}
\end{figure}
 % $ %% formatting
% and lazy:
% \begin{code}
% lazy :: (Arrow arr) => arr [a] [a]
% lazy = arr (\ ~(x:xs) -> x : lazy xs
% \end{code}
%% we have shown this already.

\subsection{Torus skeleton}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the |torus| skeleton.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a ring from \ref{sec:ring} one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.

Similar to the |ring|, we once again start by rotating the input, but this time not only in one direction, but in two. This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes |second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))| instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we once again zip the inputs for the input arrow with |arr (uncurry3 zipWith3 lazyzip3)| from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then feed into our parallel execution.

This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed. We first have to |shuffle| all the inputs to then pass it into |parMap conf (ptorus f)| which yields us |[(d, fut a, fut b)]|. We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass. Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.

\begin{figure}[h]
\olcomment{swapped type classes}
\begin{code}
torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr, Future fut a, Future fut b,
          ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf) =>
	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
torus conf f =
	loop (second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy)) >>>
	arr (uncurry3 (zipWith3 lazyzip3)) >>>
	(arr length >>> arr unshuffle) &&& (shuffle >>> parMap conf (ptorus f)) >>> app >>>
	arr (map unzip3) >>> arr unzip3 >>> threetotwo)

ptorus :: (Arrow arr, Future fut a, Future fut b) =>
	arr (c, a, b) (d, a, b) -> arr (c, fut a, fut b) (d, fut a, fut b)
ptorus f = arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>> arr (\ ~(d, a, b) -> (d, put a, put b))

uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
uncurry3 f (a, (b, c)) = f a b c

lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)

threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
\end{code} % $
\caption{Definition of the |torus| skeleton.}
\label{fig:torus}
\end{figure}
As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}. Their instantiation of the skeleton |nodefunction| can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
\begin{figure}[h]
\begin{code}
nodefunction :: Int -> ((Matrix, Matrix), [Matrix], [Matrix]) -> ([Matrix], [Matrix], [Matrix])
nodefunction n ((bA, bB), rows, cols) = ([bSum], bA:nextAs , bB:nextBs)
		where bSum = foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
			     nextAs = take (n-1) rows
			     nextBs = take (n-1) cols
\end{code}
\caption{Adapted |nodefunction| for matrix multiplication with the |torus| from Fig.~\ref{fig:torus}}
\label{fig:torusMatMult}
\end{figure}
If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.\olcomment{much more details on this!}
\begin{figure}[ht]
\olcomment{more nodes!!!}
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (Parrows)}
	\label{fig:torus_parrows_trace}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden)}
	\label{fig:torus_eden_trace}
\end{figure}

%\FloatBarrier

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-05-25 paper/content/sources/utilityFunctions.tex
moving things around, reworking the makefile

\section{Utility Functions}\label{utilfns}
To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: |map|, |foldl| and |zipWith| on arrows.

The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any arrow |arr a
b| to an arrow |arr [a] [b]| \cite{programming_with_arrows}. 
Similarly, we can also define |foldlArr| (Fig.~\ref{fig:foldlArr}) that lifts any arrow |arr (b, a) b| with a neutral element |b| to |arr [a] b|.

\begin{figure}[htb]
\begin{code}
mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
mapArr f =
	arr listcase >>>
	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))

listcase [] = Left ()
listcase (x:xs) = Right (x,xs)
\end{code}
\caption{The definition of |map| over Arrows and the |listcase| helper
function.}
\label{fig:mapArr}
\end{figure}
\begin{figure}[htb]
\begin{code}
foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
foldlArr f b =
	arr listcase >>>
	arr (const b) |||
		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
\end{code}
\caption{The definition of |foldl| over Arrows.}
\label{fig:foldlArr}
\end{figure}
%\fixme{pipepipepipe does not work with lhs2TeX}

Finally, with the help of |mapArr| (Fig.~\ref{fig:mapArr}), we can define |zipWithArr| that lifts any arrow |arr (a, b) c| to an arrow |arr ([a], [b]) [c]|.
%\begin{figure}[htb]
\begin{code}
zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
\end{code}
% \caption{|zipWith| over arrows}
% \label{fig:zipWithArr}
% \end{figure}
 %$ %% formatting
These combinators make use of the |ArrowChoice| type class which provides the \pipepipepipe\olcomment{CHECK!} combinator. It takes two arrows |arr a c| and |arr b c| and combines them into a new arrow |arr (Either a b) c| which pipes all |Left a|'s to the first arrow and all |Right b|'s to the second arrow.
% \begin{figure}[htb]
\begin{code}
(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
\end{code}
% \caption{Type signature of \pipepipepipe}
% \label{fig:codeSigPipePipePipe}
% \end{figure}

With the zipWithArr combinator we can also write a combinator |listApp|, that lifts a list of arrows |[arr a b]| to an arrow |arr [a] [b]|.
% \begin{figure}[htb]
\begin{code}
listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
\end{code}
% \caption{Definition of |listApp|}
% \label{fig:listApp}
% \end{figure}
% $ %% formatting
Note that  this additionally makes use of the |ArrowApply| typeclass that allows us to evaluate arrows with |app :: arr (arr a b, a) c|.

% $ %% formatting

\section{Omitted Funtion Definitions}

We have omitted some function definitions in the main text for
brevity, and redeem this here.
%
We warp Eden's build-in Futures in PArrows as in
Figure~\ref{fig:RDFuture}.
The full definition of |farmChunk| is in Figure~\ref{fig:farmChunk}.
Eden definition of |ring| skeleton following \citep{Loogen2012} is in Figure~\ref{fig:ringEden}.

\begin{figure}[htb]
\begin{code}
data RemoteData a = RD { rd :: RD a }

instance (Trans a) => Future RemoteData a where
    put = arr (\a -> RD { rd = release a })
    get = arr rd >>> arr fetch
\end{code}
\caption{|RD|-based |RemoteData| version of |Future| for the Eden backend.}
\label{fig:RDFuture}
\end{figure}

\begin{figure}[htb]
\begin{code}
farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf, 
             ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
farmChunk conf chunkSize numCores f =
	unshuffle numCores >>>
	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
	shuffle
\end{code}
\caption{Definition of |farmChunk|.}
\label{fig:farmChunk}
\end{figure}

\begin{figure}[htb]
\begin{code}
ringSimple :: (Trans i, Trans o, Trans r) => (i -> r -> (o,r)) -> [i] -> [o]
ringSimple f is =  os
  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f) (zip is $ lazy ringIns))
        ringIns = rightRotate ringOuts

toRD :: (Trans i, Trans o, Trans r) => ((i,r) -> (o,r)) -> ((i, RD r) -> (o, RD r))
toRD  f (i, ringIn)  = (o, release ringOut)
  where (o, ringOut) = f (i, fetch ringIn)

rightRotate    :: [a] -> [a]
rightRotate [] =  []
rightRotate xs =  last xs : init xs

lazy :: [a] -> [a]
lazy ~(x:xs) = x : lazy xs
\end{code}
\caption{Eden's definition of the |ring| skeleton.}
\label{fig:ringEden}
\end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-05-26 paper/content/arrows.tex
wip

--- 2017-05-12 paper/arrows.tex
@@
 Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
-An arrow \code{arr a b} represents  a computation that converts an input \code{a} to an output \code{b}.
+An arrow |arr a b| represents  a computation that converts an input |a| to an output |b|.
 This is defined in the arrow typeclass:
@@
 \begin{figure}[h]
-\begin{subfigure}{0.5\textwidth}
-%\begin{minipage}{0.5\textwidth}
-\begin{lstlisting}[frame=htrbl]
+\centering
+%\subfloat[Arrow class definition][]{%
+% begin{minipage}{0.5\textwidth}
+\parbox{0.49\linewidth}{%
+\begin{code}
 class Arrow arr where
   arr :: (a -> b) -> arr a b
-
   (>>>) :: arr a b -> arr b c -> arr a c
-
   first :: arr a b -> arr (a,c) (b,c)
-\end{lstlisting}
+\end{code}
 \caption{Arrow class definition}
-%\end{minipage}
-\end{subfigure}
-~~~~
-\begin{subfigure}{0.45\textwidth}
-	\begin{center}
-	\subcaptionbox{arr}{\includegraphics[scale=0.6]{images/arr}}
-	\subcaptionbox{composition \code{(>>>)}}{\includegraphics[scale=0.6]{images/compose}}
-	\subcaptionbox{first}{\includegraphics[scale=0.6]{images/first}}
-	\end{center}
-\end{subfigure}
-\caption{The arrow type class definition on the left with schematic depiction of its combinators on the right}
+}\qquad
+\begin{minipage}{0.49\linewidth}%
+%\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
+%	\begin{center}
+\centering
+	{\includegraphics[scale=0.6]{images/arr}}
+	{\includegraphics[scale=0.6]{images/compose}}
+	{\includegraphics[scale=0.6]{images/first}}
+%	\end{center}
+%}%\hfill
+\caption{The schematic depiction of |Arrow| combinators}
+\end{minipage}
 \end{figure}
-\lstinline{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \lstinline{return}.
-The \lstinline{>>>} operator is analogous to the monadic composition  \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}.
-Lastly, the \lstinline{first} operator  takes the input arrow from \lstinline{b} to \lstinline{c} and converts it into an arrow on pairs with the second argument untouched.
+|arr| is used to lift an ordinary function to an arrow type, similarly to the monadic |return|.
+The |>>>| operator is analogous to the monadic composition  |>>=| and combines two arrows |arr a b| and |arr b c| by "wiring" the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
+Lastly, the |first| operator  takes the input arrow from |b| to |c| and converts it into an arrow on pairs with the second argument untouched.
 It allows us to to save input across arrows.
 \\\\
-The most prominent instances of this interface are regular functions \lstinline{(->)} (Fig.~\ref{fig:arrowfn}),
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+The most prominent instances of this interface are regular functions |(->)|: %(Fig.~\ref{fig:arrowfn}),
+%\begin{figure}[h]
+\begin{code}
 instance Arrow (->) where
@@
 	first f = \(a, c) -> (f a, c) 
-\end{lstlisting}
-\caption{Arrow instance for regular functions}
-\label{fig:arrowfn}
-\end{figure}
-and the Kleisli type (Fig.~\ref{fig:arrowkleisli}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{Arrow instance for regular functions}
+% \label{fig:arrowfn}
+% \end{figure}
+%%% on the verge of acceptable code
+and the Kleisli type: %(Fig.~\ref{fig:arrowkleisli}).
+% \begin{figure}[h]
+\begin{code}
 data Kleisli m a b = Kleisli { run :: a -> m b }
@@
 	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
-\end{lstlisting}%$
-\caption{Definition of the Kleisli type and the corresponding arrow instance}
-\label{fig:arrowkleisli}
-\end{figure}
+\end{code}%$
+% \caption{Definition of the Kleisli type and the corresponding arrow instance}
+% \label{fig:arrowkleisli}
+% \end{figure}
 \begin{figure}[h]
@@
 	\begin{tabular}{cc}
-		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
-		\subcaptionbox{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
-		\subcaptionbox{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
-		\subcaptionbox{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
+		% \subcaptionbox
+{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
+		% \subcaptionbox
+{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
+|first| & |second| \\
+\midrule
+		% \subcaptionbox
+{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+		% \subcaptionbox
+{}{\includegraphics[width = 1.5in]{images/andandand}}\\
+|(***)|\label{fig:***Img} & |(&&&)| \label{fig:&&&Img} \\
 	\end{tabular}
-	\caption{Syntactic sugar for arrows}
-	\label{3}
+	\caption{Visual depiction of syntactic sugar for arrows.}
+	\label{fig:syntacticSugarArrows}
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second} (Fig.~\ref{fig:secondImg},\ref{fig:second}),
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+With this typeclass in place, Hughes also defined some syntactic sugar (see Fig.~\ref{fig:syntacticSugarArrows}): the functions |second|, |***| and |&&&|. 
+The mirrored version of |first|, called |second| is:
+% \begin{figure}[h]
+\begin{code}
 second :: Arrow arr => arr a b -> arr (c, a) (c, b)
@@
 	where swap (x, y) = (y, x)
-\end{lstlisting}
-\caption{The second combinator}
-\label{fig:second}
-\end{figure}
-the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},\ref{fig:***})
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{The second combinator}
+% \label{fig:second}
+% \end{figure}
+the |***| combinator that combines |first| and |second| to handle two inputs in one arrow, is defined as
+% \begin{figure}[h]
+\begin{code}
 (***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
 f *** g = first f >>> second g
-\end{lstlisting}
-\caption{The (***) combinator}
-\label{fig:***}
-\end{figure}
-and the \code{\&\&\&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input (Fig.~\ref{fig:&&&Img},\ref{fig:&&&}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{The (***) combinator}
+% \label{fig:***}
+% \end{figure}
+and the |&&&| combinator that constructs an arrow which outputs two different values like |***|, but takes only one input is:
+% \begin{figure}[h]
+\begin{code}
 (&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
 f &&& g = arr (\a -> (a, a)) >>> (f *** g)
-\end{lstlisting}
-\caption{The (\&\&\&) combinator}
-\label{fig:&&&}
-\end{figure}
-A short example given by Hughes on how to use this is \lstinline{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{The (\&\&\&) combinator}
+% \label{fig:&&&}
+% \end{figure}
+A short example given by Hughes on how to use this is 
+addition with arrows:
+%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
+% \begin{figure}[h]
+\begin{code}
 add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
 add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
-\end{lstlisting}
-\caption{Add over arrows}
-\label{fig:addArrows}
-\end{figure}
-% The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator.
+\end{code}
+% \caption{Add over arrows}
+% \label{fig:addArrows}
+% \end{figure}
+% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator.
 Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
@@
 % \\\\
-% \textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used.
-We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}
+% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used.
+We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}
 %

### 2017-05-26 paper/content/parallelHaskells.tex
wip

--- 2017-05-12 paper/parallelHaskells.tex
@@
 \subsection{Short introduction to parallel Haskells}
+\label{sec:parEvalNIntro}
 There are already several ways to write parallel programs in Haskell.
@@
 
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \code{a -> b} in parallel:
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:
+
+\begin{code}
+parEvalN :: [a -> b] -> [a] -> [b]
+\end{code}
 
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-parEvalN :: [a -> b] -> [a] -> [b]
-\end{lstlisting}
-\caption{parEvalN's type signature}
-\label{fig:parEvalNTypeSig}
-\end{figure}
-\begin{figure}[h]
+  \centering
 	\includegraphics[scale=0.7]{images/parEvalN}
-	\caption{Schematic illustration of parEvalN}
+	\caption{Schematic illustration of \texttt{parEvalN}.}
 	\label{fig:parEvalN}
 \end{figure}
-\olcomment{make them to real figures? with environment, reference and such?}
-Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement \code{parEvalN} with several different parallel Haskells.
+
+Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement |parEvalN| with several different parallel Haskells.
 
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1999} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with \code{using :: a -> Strategy a -> a}.
-For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}.
-We then evaluate this lazy list \code{[b]} according to a \code{Strategy [b]} with the \code{using :: a -> Strategy a -> a} operator.
-We construct this strategy with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
-To ensure that programs that use \code{parEvalN} have the correct evaluation order, we annotate the computation with \code{pseq :: a -> b -> b} which forces the compiler to not reorder multiple \code{parEvalN} computations.
-This is particularly necessary in circular communication topologies like in the \code{torus} or \code{ring} skeleton that we will see in chapter \ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without \code{pseq} during testing for this paper.
+Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with |using :: a -> Strategy a -> a|.
+For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|.
+% $
+We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator.
+We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form.
+To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations.
+This is particularly necessary in circular communication topologies like in the |torus| or |ring| skeleton that we will see in Section~\ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without |pseq| during testing for this paper.
 
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = let bs = zipWith ($) fs as in
-	(bs `using` parList rdeepseq) `pseq` bs
-\end{lstlisting}
+parEvalN fs as = let bs = zipWith ($) fs as 
+                 in (bs `using` parList rdeepseq) `pseq` bs
+\end{code} %$
+
 \begin{figure}[h]
@@
 \subsubsection{ParMonad}
-The \code{Par} monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
+The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
 
-Our parallel evaluation function \code{parEvalN} can be defined by zipping the list of \code{[a -> b]} with the list of inputs \code{[a]} with the application operator \code{\$} just like with Multicore Haskell.
-Then, we map over this not yet evaluated lazy list of results \code{[b]} with \code{spawnP :: NFData a => a -> Par (IVar a)} to transform them to a list of not yet evaluated forked away computations \code{[Par (IVar b)]}, which we convert to \code{Par [IVar b]} with \code{sequenceA}.
-We wait for the computations to finish by mapping over the \code{IVar b}'s inside the \code{Par} monad with \code{get}.
-This results in \code{Par [b]}.
-We finally execute this process with \code{runPar} to finally get \code{[b]} again.
+Our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |DOLLAR| just like with Multicore Haskell. % $
+Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|.
+We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| monad with |get|.
+This results in |Par [b]|.
+We finally execute this process with |runPar| to finally get |[b]| again.
 
-\textbf{\textcolor{red}{explain problems with laziness here. Problems with torus}}
+\mbcomment{explain problems with laziness here. Problems with torus}
 
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
@@
 	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
-\end{lstlisting}
+\end{code}
 \begin{figure}[h]
@@
 
-While Eden also comes with a monad \code{PA} for parallel evaluation, it also ships with a completely functional interface that includes
+While Eden also comes with a monad |PA| for parallel evaluation, it also ships with a completely functional interface that includes
 %\\
-%\code{spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]}.
+a |spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]|
 %\\
-a \code{spawnF} function that
+%a |spawnF| 
+function that
 %This 
-allows us to define \code{parEvalN} directly:
+allows us to define |parEvalN| directly:
 
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
 parEvalN = spawnF 
-\end{lstlisting}
+\end{code}
 \begin{figure}[h]

### 2017-05-26 paper/content/parrows.tex
wip

@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions |a -> b| in parallel, |parEvalN| (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions |a -> b| in parallel, |parEvalN| (Chap.~\ref{sec:parEvalNIntro}, Fig.~\ref{fig:parEvalN}).
 Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
@@
 For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, we want to the user not to lose this ability because of using our API.
-We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a| (Fig.~\ref{fig:confa}).
-We can't directly use the |Strategy a| type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances:
+We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|:
 % \begin{figure}[h]
@@
 % \end{figure}
+We can't directly use the |Strategy a| type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
 To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).

### 2017-05-26 paper/content/syntacticSugar.tex
wip

--- 2017-05-12 paper/syntacticSugar.tex
@@
-\subsection{Syntactic Sugar} \label{syntacticSugar}
-For basic arrows, we have the \code{(***)} combinator (Fig.~\ref{fig:***Img},~\ref{fig:***}) which allows us to combine two arrows \code{arr a b} and \code{arr c d} into an arrow \code{arr (a, c) (b, d)} which does both computations at once.
-This can easily be translated into a parallel version \code{(|***|)} (Fig.~\ref{fig:|***|}) with the use of \code{parEval2}, but for this we require a backend which has an implementation that does not require any configuration (hence the \code{()} as the conf parameter in Fig.~\ref{fig:|***|}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\subsubsection{Syntactic Sugar} \label{syntacticSugar}
+For basic arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two arrows |arr a b| and |arr c d| into an arrow |arr (a, c) (b, d)| which does both computations at once.
+This can easily be translated into a parallel version |parstar| (Fig.~\ref{fig:par***}) with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
+):
+% \begin{figure}[h]
+\begin{code}
 (|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
@@
 (|***|) = parEval2 ()
-\end{lstlisting}
-\caption{Definition of (|***|) - the parallel version of (***)}
-\label{fig:|***|}
-\end{figure}
-% With this we can analogously to the serial \code{&&&}
-We define the parallel \code{(|\&\&\&|)} (Fig.~\ref{fig:|&&&|}) in a similar manner to its sequential pendant \code{(\&\&\&)} (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+% \caption{Definition of |parstar|, the parallel version of |***|.}
+% \label{fig:par***}
+% \end{figure}
+% With this we can analogously to the serial |&&&|
+We define the parallel |parand| % (Fig.~\ref{fig:par&&&}) 
+in a similar manner to its sequential pendant |&&&| (Fig.~\ref{fig:syntacticSugarArrows}):
+% \begin{figure}[h]
+\begin{code}
 (|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
@@
 (|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
-\end{lstlisting} % $ %% formatting
-\caption{Definition of (|\&\&\&|) - the parallel version of (\&\&\&)}
-\label{fig:|&&&|}
-\end{figure}
+\end{code} % $ %% formatting
+% \caption{Definition of |parand| - the parallel version of |&&&|.}
+% \label{fig:par&&&}
+% \end{figure}

### 2017-05-26 paper/content/basicSkeletons.tex
fixed figure references

--- 2017-05-15 paper/basicSkeletons.tex
@@
 \label{sec:extending-interface}
-With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some further basic parallel interface functions.
+With the |ArrowParallel| typeclass in place and implemented, we can now implement some further basic parallel interface functions.
 These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
 
-\subsubsection{Lazy \code{parEvalN}}
+\subsubsection{Lazy |parEvalN|}
 \begin{figure}[h]
@@
 \end{figure}
-The function \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows.
+The function |parEvalN| is 100\% strict, which means that it fully evaluates all passed arrows.
 Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
-\code{map (arr .
-(+)) [1..]} or just because we need the arrows evaluated in chunks.
-\code{parEvalNLazy} (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
-It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}.
-These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp} (Fig.~\ref{fig:listApp}).
-The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
+|map (arr .
+(+)) [1..]| or just because we need the arrows evaluated in chunks.
+|parEvalNLazy| (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
+It works by first chunking the input from |[a]| to |[[a]]| with the given |ChunkSize| in |arr (chunksOf chunkSize)|.
+These chunks are then fed into a list |[arr [a] [b]]| of parallel arrows created by feeding chunks of the passed |ChunkSize| into the regular parEvalN by using |listApp|.
+The resulting |[[b]]| is lastly converted into |[b]| with |arr concat|.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
@@
 	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
-\end{lstlisting} %$ %% formatting
-\caption{Definition of parEvalNLazy}
+\end{code} %$ %% formatting
+\caption{Definition of |parEvalNLazy|.}
 \label{fig:parEvalNLazy}
@@
 	\includegraphics[scale=0.7]{images/parEval2}
-	\caption{Schematic depiction of parEval2}
+	\caption{Schematic depiction of |parEval2|.}
 	\label{fig:parEval2Img}
@@
 But sometimes we want to paralellize heterogenous types as well.
-However, we can implement such a \code{parEval2} combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass.
-The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
+However, we can implement such a |parEval2| combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| typeclass.
+The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into parEvalN.
 \\\\
-We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]} by first tagging the two inputs with \code{Left} and \code{Right} and wrapping the right element in a singleton list with \code{return} so that we can combine them with \code{arr (uncurry (:))}.
-Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above.
-After the calculation is finished, we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}.
-The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
+We start by transforming the |(a, c)| input into a 2-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|.
+Next, we feed this list into a parallel arrow running on 2 instances of |f +++ g| as described above.
+After the calculation is finished, we convert the resulting |[Either b d]| into |([b], [d])| with |arr partitionEithers|.
+The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to |(b, d)| in the last step.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 parEval2 :: (ArrowChoice arr,
@@
 	arr head *** arr head
-\end{lstlisting}
+\end{code}
 	\caption{Definition of parEval2}

### 2017-05-26 paper/content/parallelHaskells.tex
fixed figure references

@@
 
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel, as also Figure~\ref{fig:parevaln-init} symbolically shows:
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel, as also Figure~\ref{fig:parEvalN} symbolically shows:
 

### 2017-05-26 paper/content/parrows.tex
fixed figure references

@@
 
-We therefore apply each function to its corresponding input value with |listApp| (Fig.~\ref{fig:listApp}) and then fork the computation away with |arr spawnP| inside a |zipWithArr| call.
+We therefore apply each function to its corresponding input value with || and then fork the computation away with |arr spawnP| inside a |zipWithArr| call.
 This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|.

### 2017-05-26 paper/content/syntacticSugar.tex
fixed figure references

@@
 For basic arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two arrows |arr a b| and |arr c d| into an arrow |arr (a, c) (b, d)| which does both computations at once.
-This can easily be translated into a parallel version |parstar| (Fig.~\ref{fig:par***}) with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
+This can easily be translated into a parallel version |***| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
 ):

### 2017-05-26 paper/content/topologySkeletons.tex
fixed figure references

--- 2017-05-17 paper/content/topologySkeletons.tex
@@
-\FloatBarrier
+%\FloatBarrier
 \section{Topological Skeletons}
@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{Loogen2012, eden_skel_topology}.
-These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
+The Eden library leverages this problem and already comes\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} with more predefined skeletons, among them a |pipe|, a |ring|, and a |torus| implementations \cite{Loogen2012}.
+These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell.
+We aim to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
 
-\subsection{Parallel pipe}
+\subsection{Parallel pipe}\label{sec:pipe}
 
-The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
-We implement this skeleton using the \code{ArrowLoop} typeclass which gives us the \code{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
+The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
+We implement this skeleton using the |ArrowLoop| typeclass which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
 For example %this
-\mbcomment{das kann man hier so lassen, oder?}
-\begin{lstlisting}[frame=htrbl]
+\mbcomment{das kann man hier so lassen, oder?}\olcomment{sicherlich!}
+\begin{code}
 loop (arr (\(a, b) -> (b, a:b)))
-\end{lstlisting}
-, which is the same as
-\begin{lstlisting}[frame=htrbl]
+\end{code}
+which is the same as
+\begin{code}
 loop (arr snd &&& arr (uncurry (:)))
-\end{lstlisting}
-defines an arrow that takes its input \code{a} and converts it into an infinite stream \code{[a]} of it.
-Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{(\&\&\&)} and then only picking the first element of the resulting list with \code{arr last}.
+\end{code}
+defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it.
+Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |parEvalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
@@
 lazy = arr (\ ~(x:xs) -> x : lazy xs)
-\end{lstlisting}
-\caption{A first draft of the pipe skeleton expressed with parallel arrows.
-Note that the use of \code{lazy} is essential as without it programs using this definition would never halt.
-We need to enforce that the evaluation of the input \code{[a]} terminates before passing it into \code{parEvalN}.}
+\end{code}
+\caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows.
+Note that the use of |lazy| is essential as without it programs using this definition would never halt.
+We need to enforce that the evaluation of the input |[a]| terminates before passing it into |parEvalN|.}
 \label{fig:pipeSimple}
@@
 However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \code{pipe} \ref{fig:pipe}.
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf,
-	Future fut a) =>
+Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of |pipe|: % ~\ref{fig:pipe}.
+%\begin{figure}[H]
+\begin{code}
+pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
 	conf -> [arr a a] -> arr a a
 pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
-\end{lstlisting}
-\caption{Final definition of the pipe skeleton which uses Futures}
-\label{fig:pipe}
-\end{figure}
+\end{code}
+%\caption{Final definition of the pipe skeleton which uses Futures}
+%\label{fig:pipe}
+%\end{figure}
 
-Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
-\code{arr a b} and \code{arr b c}.
-By wrapping these two arrows inside a common type we obtain \code{pipe2} (Fig.~\ref{fig:pipe2}).
+Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
+|arr a b| and |arr b c|.
+By wrapping these two arrows inside a common type we obtain |pipe2| (Fig.~\ref{fig:pipe2}).
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-pipe2 :: (ArrowLoop arr, ArrowChoice arr,
-	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
-	Future fut (([a], [b]), [c])) =>
+\olcomment{I swapped the type classes here:}
+\begin{code}
+pipe2 :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
+         ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
 	conf -> arr a b -> arr b c -> arr a c
@@
 	pipe conf (replicate 2 (unify f g)) >>>
-	arr snd >>>
-	arr head
-		where
-			unify :: (ArrowChoice arr) =>
-				arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
+	arr snd >>> arr head where
+			unify :: (ArrowChoice arr) => arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
 			unify f g =
@@
 				arr (\((a, b), c) -> ((c, a), b))
-\end{lstlisting}
-\caption{Definition of pipe2}
+\end{code}
+\caption{Definition of |pipe2|.}
 \label{fig:pipe2}
@@
 
-Note that extensive use of this combinator over \code{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator \code{(|>>>|)} (Fig.~\ref{fig:|>>>|}, which is semantically equivalent to \code{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
+Note that extensive use of this combinator over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Nonetheless, we can define a parallel piping operator |parcomp| (Fig.~\ref{fig:par>>>}, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Section~\ref{syntacticSugar}.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(|>>>|) :: (ArrowLoop arr, ArrowChoice arr,
-	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
-	Future fut (([a], [b]), [c])) =>
+\olcomment{swapped type classes}
+\begin{code}
+(|>>>|) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
+	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
 	arr a b -> arr b c -> arr a c
 (|>>>|) = pipe2 ()
-\end{lstlisting}
-\caption{Definition of \texttt{(|>>>|)}}
-\label{fig:|>>>|}
+\end{code}
+\caption{Definition of |parcomp|.}
+\label{fig:par>>>}
 \end{figure}
@@
 \end{figure}
-Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \code{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other.
-Its input is a node function \code{i -> r -> (o, r)} in which \code{r} serves as the intermediary output that gets send to the neighbour of each node.
-This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
+Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function |[i] -> [o]| with a ring of nodes that communicate in a ring topology with each other.
+Its input is a node function |i -> r -> (o, r)| in which |r| serves as the intermediary output that gets send to the neighbour of each node.
+This data is sent over direct communication channels, so called \enquote{remote data}.
+We depict it in Appendix, Fig.~\ref{fig:ringEden}.
+%\end{code}
+%with toRD (to make use of remote data)
+%\begin{code}
+%\end{code}
+%and rightRotate:
+%\begin{code}
+
+We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an arrow, already fits quite neatly into the |loop|'s |arr (a, b) (c, b) -> arr a c|.
+In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)|.
+Similarly to the |pipe| from Chapter~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
+\begin{quote}
+{Note that the list of ring inputs |ringIns| is the same as the list of ring outputs |ringOuts| rotated by one element to the right using the auxiliary function |rightRotate|.
+Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
+\end{quote}
+Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|.
+Finally we unzip this list into |([o], [fut r])|.
+Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
+This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
+%Further details on this can be found in \cite{eden_cefp}.
+\olcomment{let's do it?}
+
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-ringSimple :: (Trans i, Trans o, Trans r) =>
-   (i -> r -> (o,r))
-   -> [i] -> [o]
-ringSimple f is =  os
-  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
-                                   (zip is $ lazy ringIns))
-        ringIns = rightRotate ringOuts
-
-toRD :: (Trans i, Trans o, Trans r) => 
-        ((i,r) -> (o,r))
-        -> ((i, RD r) -> (o, RD r))
-toRD  f (i, ringIn)  = (o, release ringOut)
-  where (o, ringOut) = f (i, fetch ringIn)
-
-rightRotate    :: [a] -> [a]
-rightRotate [] =  []
-rightRotate xs =  last xs : init xs
-
-lazy :: [a] -> [a]
-lazy ~(x:xs) = x : lazy xs
-\end{lstlisting}
-\caption{Eden's definition of the ring skeleton \citep{eden_skel_topology}}
-\label{fig:ringEden}
-\end{figure}
-%\end{lstlisting}
-%with toRD (to make use of remote data)
-%\begin{lstlisting}[frame=htrbl]
-%\end{lstlisting}
-%and rightRotate:
-%\begin{lstlisting}[frame=htrbl]
-
-We can rewrite its functionality easily with the use of \code{loop} as the definition of the node function, \code{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \code{loop}'s \code{arr (a, b) (c, b) -> arr a c}.
-In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}.
-Similarly to the \code{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
-%\begin{quotation}
-\enquote{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate.
-Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
-%\end{quotation}
-Next, we zip the resulting \code{([i], [fut r])} to \code{[(i, fut r)]} with \code{arr (uncurry zip)} so we can feed that into a our input arrow \code{arr (i, r) (o, r)}, which we transform into \code{arr (i, fut r) (o, fut r)} before lifting it to \code{arr [(i, fut r)] [(o, fut r)]} to get a list \code{[(o, fut r)]}.
-Finally we unzip this list into \code{([o], [fut r])}.
-Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-ring :: (ArrowLoop arr, Future fut r,
-	ArrowParallel arr (i, fut r) (o, fut r) conf) =>
-    conf ->
-    arr (i, r) (o, r) ->
-    arr [i] [o]
+\begin{code}
+ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
+    conf -> arr (i, r) (o, r) -> arr [i] [o]
 ring conf f =
-	loop (second (rightRotate >>> lazy) >>>
-    arr (uncurry zip) >>>
-    parMap conf (second get >>> f >>> second put) >>>
-    arr unzip)
+	loop (second (rightRotate >>> lazy) >>> arr (uncurry zip) >>>
+        parMap conf (second get >>> f >>> second put) >>> arr unzip)
 
@@
 lazy = arr (\ ~(x:xs) -> x : lazy xs)
-\end{lstlisting}
-\caption{Final Definition of the ring Skeleton}
+\end{code}
+\caption{Final definition of the |ring| skeleton.}
 \label{fig:ringFinal}
 \end{figure}
- %$ %% formatting
+ % $ %% formatting
 % and lazy:
-% \begin{lstlisting}[frame=htrbl]			
+% \begin{code}
 % lazy :: (Arrow arr) => arr [a] [a]
 % lazy = arr (\ ~(x:xs) -> x : lazy xs
-% \end{lstlisting}
+% \end{code}
 %% we have shown this already.
-This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm. 
-%Further details on this can be found in \cite{eden_cefp}.
-\olcomment{let's do it?}
+
 \subsection{Torus skeleton}
@@
 	\includegraphics[scale=0.75]{images/torus}
-	\caption{Schematic depiction of the torus skeleton}
+	\caption{Schematic depiction of the |torus| skeleton.}
 	\label{fig:ringTorusImg}
@@
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-\\\\
-With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \code{ArrowLoop} typeclass.
-\\\\
-Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple \code{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \code{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \code{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \code{[[fut a]]} horizontally and \code{[[fut b]]} vertically.
-Then, we once again zip the inputs for the input arrow with \code{arr (uncurry3 zipWith3 lazyzip3)} from \code{([[c]], ([[fut a]], [[fut b]]))} to \code{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
-\\\\
+With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.
+
+Similar to the |ring|, we once again start by rotating the input, but this time not only in one direction, but in two.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes |second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))| instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically.
+Then, we once again zip the inputs for the input arrow with |arr (uncurry3 zipWith3 lazyzip3)| from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then feed into our parallel execution.
+
 This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
-We first have to \code{shuffle} all the inputs to then pass it into \code{parMap conf (ptorus f)} which yields us \code{[(d, fut a, fut b)]}.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app :: arr (arr a b, a) c} from the \code{ArrowApply} typeclass.
-Finally, we unpack this matrix \code{[[[(d, fut a, fut b)]]} with \code{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \code{([[d]], ([[fut a]], [[fut b]]))}.
-\\\\
+We first have to |shuffle| all the inputs to then pass it into |parMap conf (ptorus f)| which yields us |[(d, fut a, fut b)]|.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass.
+Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.
+
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
-	ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
-	Future fut a, Future fut b) =>
+\olcomment{swapped type classes}
+\begin{code}
+torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr, Future fut a, Future fut b,
+          ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf) =>
 	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
 torus conf f =
-	loop (second ((mapArr rightRotate >>> lazy) ***
-		(arr rightRotate >>> lazy)) >>>
+	loop (second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy)) >>>
 	arr (uncurry3 (zipWith3 lazyzip3)) >>>
-	(arr length >>> arr unshuffle) &&&
-		(shuffle >>> parMap conf (ptorus f)) >>>
-	app >>>
+	(arr length >>> arr unshuffle) &&& (shuffle >>> parMap conf (ptorus f)) >>> app >>>
 	arr (map unzip3) >>> arr unzip3 >>> threetotwo)
@@
 ptorus :: (Arrow arr, Future fut a, Future fut b) =>
-	arr (c, a, b) (d, a, b) ->
-	arr (c, fut a, fut b) (d, fut a, fut b)
-ptorus f =
-	arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>>
-	arr (\ ~(d, a, b) -> (d, put a, put b))
+	arr (c, a, b) (d, a, b) -> arr (c, fut a, fut b) (d, fut a, fut b)
+ptorus f = arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>> arr (\ ~(d, a, b) -> (d, put a, put b))
 
@@
 threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
-\end{lstlisting}
-\caption{Definition of the torus skeleton}
+\end{code} % $
+\caption{Definition of the |torus| skeleton.}
 \label{fig:torus}
@@
 As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}.
-Their nodefunction can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
+Their instantiation of the skeleton |nodefunction| can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-nodefunction :: Int ->
-	((Matrix, Matrix), [Matrix], [Matrix]) ->
-	([Matrix], [Matrix], [Matrix])
-nodefunction n ((bA, bB), rows, cols) =
-	([bSum], bA:nextAs , bB:nextBs)
-		where bSum =
-			foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
+\begin{code}
+nodefunction :: Int -> ((Matrix, Matrix), [Matrix], [Matrix]) -> ([Matrix], [Matrix], [Matrix])
+nodefunction n ((bA, bB), rows, cols) = ([bSum], bA:nextAs , bB:nextBs)
+		where bSum = foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
 			     nextAs = take (n-1) rows
 			     nextBs = take (n-1) cols
-\end{lstlisting}
-\caption{Adapted nodefunction for matrix multiplication with the torus from Fig.~\ref{fig:torus}}
+\end{code}
+\caption{Adapted |nodefunction| for matrix multiplication with the |torus| from Fig.~\ref{fig:torus}}
 \label{fig:torusMatMult}

### 2017-05-26 paper/content/abstract.tex
oleg remarks for abstract.tex

\begin{abstract}
Arrows are a general interface for computation and therefore form an alternative to monads for API design. We express parallelism using this concept in a novel way: We define an arrows-based language for parallelism and implement it using multiple parallel Haskells.
In this manner we are able to bridge across various parallel Haskells.

Additionally, our way of writing parallel programs has the benefit of being portable across flavours of parallel Haskells.
Furthermore, as each parallel computation is an arrow, which means that they can be composed and transformed as such.
We introduce some syntactic sugar to provide parallelism-aware arrow combinators.

To show that our arrow-based language is on par with the existing parallel languages, we also define several parallel skeletons with our framework. 
Benchmarks show that our framework does not induce too much overhead performance-wise.
\end{abstract}

### 2017-05-26 paper/content/motivation.tex
minor

--- 2017-03-30 paper/motivation.tex
@@
 
-HIT HERE REALLY STRONG
+\olcomment{HIT HERE REALLY STRONG}
+
+%\subsection{Impact of parallel Arrows}
+%\olcomment{move this to Contributions in the front or something}
+\mbcomment{different, how?}
+We wrap parallel Haskells inside of our |ArrowParallel| interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
+\begin{itemize}
+	\item \textbf{Arrow DSL benefits}:
+	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is just yet another arrow combinator.
+The resulting arrow can be used in the same way a potential serial version could be used.
+This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
+We can just \enquote{plug} in parallel parts into sequential arrow-based programs without having to change anything.
+	\item \textbf{Abstraction}:
+	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic.
+This leaves us in the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
+So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
+\end{itemize}
 
@@
 Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface.
-Section~\ref{futures} defines futures for Parallel Arrows.
-Futures are a further concept that helps to define more advanced interfaces.
-Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons in our newly defined dialect.
-More advanced ones are showcased in Section~\ref{sec:topology-skeletons}.
+Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication.
+Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel |map| with and without load balancing, |map-reduce|) in our newly defined dialect.
+More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|).
 Section~\ref{sec:benchmarks} shows the benchmark results.
-Section~\ref{sec:conclusion} concludes.
+Section~\ref{sec:conclusion} discusses future work and concludes.
 

### 2017-05-26 paper/content/parallelHaskells.tex
minor

@@
 
-In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel, as also Figure~\ref{fig:parEvalN} symbolically shows:
-
-\begin{code}
-parEvalN :: [a -> b] -> [a] -> [b]
-\end{code}
-
-\begin{figure}[h]
+In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel or |parEvalN :: [a -> b] -> [a] -> [b]|, as also Figure~\ref{fig:parEvalN} symbolically shows.
+\begin{figure}[t]
   \centering
 	\includegraphics[scale=0.7]{images/parEvalN}
-	\caption{Schematic illustration of \texttt{parEvalN}.}
+	\caption{Schematic illustration of |parEvalN|.}
 	\label{fig:parEvalN}
 \end{figure}
-
 Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement |parEvalN| with several different parallel Haskells.
@@
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies \cite{Trinder1998a,Marlow2010} for several types which can be applied with |using :: a -> Strategy a -> a|.
+Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|.
 For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|.
@@
 To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations.
-This is particularly necessary in circular communication topologies like in the |torus| or |ring| skeleton that we will see in Section~\ref{sec:topology-skeletons} which resulted in deadlock scenarios when executed without |pseq| during testing for this paper.
+This is particularly necessary in circular communication topologies like in the |torus| or |ring| (Chap.~\ref{sec:topology-skeletons}), where a wrong execution order would result in deadlock scenarios when executed without |pseq|.
 
@@
 	\caption{Dataflow of the Multicore Haskell parEvalN version}
-\olcomment{Evaluation step explicitly shown?}
 \end{figure} %$ %% formatting
@@
 
-Our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |DOLLAR| just like with Multicore Haskell. % $
+Our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
 Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|.

### 2017-05-28 paper/content/futures.tex
fixed things to that lhs2TeX can run (without the code markers)
minor

--- 2017-05-12 paper/futures.tex
@@
 \end{figure}
-In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node.
+In a distributed environment, a resulting arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
 
-While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one |parEvalN| call by directly wiring the arrows properly together, this example illustrates an important problem: When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
 This can become a serious bottleneck %in heavy threaded applications.

### 2017-05-28 paper/content/parallelHaskells.tex
minor

@@
 Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|.
-For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|.
-% $
+
+For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
 We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator.
@@
 
+The resulting code and a graphical representation can be found in Fig.~\ref{fig:parEvalNMulticore} and Fig.~\ref{fig:parEvalNMulticoreImg}, respectively.
+
+\begin{figure}[h]
 \begin{code}
@@
                  in (bs `using` parList rdeepseq) `pseq` bs
-\end{code} %$
+\end{code}
+\caption{Multicore version of |parEvalN|}
+\label{fig:parEvalNMulticore}
+\end{figure}
 
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-	\caption{Dataflow of the Multicore Haskell parEvalN version}
+	\caption{Dataflow of the Multicore Haskell |parEvalN| version}
+	\label{fig:parEvalNMulticoreImg}
 \end{figure} %$ %% formatting
@@
 
-Our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
+The |Par| Monad version of our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell.
+% $
 Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|.
@@
 This results in |Par [b]|.
-We finally execute this process with |runPar| to finally get |[b]| again.
+We execute this process with |runPar| to finally get the final |[b]|.
+
+Again, the resulting code and a graphical representation can be found in Fig.~\ref{fig:parEvalNParMonad} and Fig.~\ref{fig:parEvalNParMonadImg}, respectively.
 
@@
 
+\begin{figure}[h]
 \begin{code}
@@
 \end{code}
+\caption{|Par| Monad version of |parEvalN|}
+\label{fig:parEvalNParMonad}
+\end{figure}
 \begin{figure}[h]
@@
 	\caption{Dataflow of the Par Monad parEvalN version}
+	\label{fig:parEvalNParMonadImg}
 \end{figure}

### 2017-05-28 paper/content/parallelHaskells.tex
parallelHaskells.tex fixes for olegs (analog) comments.

@@
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-	\caption{Dataflow of the Par Monad parEvalN version}
+	\caption{Dataflow of the |Par| Monad |parEvalN| version}
 	\label{fig:parEvalNParMonadImg}
@@
 \subsubsection{Eden}
-Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{See also \url{http://www.mathematik.uni-marburg.de/~eden/} and \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} This means that it works on clusters as well so it makes sense to have a Eden-based backend for our new parallel Haskell flavour.
-
-Eden was designed to work on clusters, but with a further simple backend it operates on multicores.
+Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{See also \url{http://www.mathematik.uni-marburg.de/~eden/} and \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} It is targeted towards clusters, but also functions well in a shared-memory setting with a further simple backend.
 However, in contrast to many other parallel Haskells, in Eden each process has its own heap.
@@
 	\includegraphics[scale=0.5]{images/parEvalNEden}
-	\caption{Dataflow of the Eden parEvalN version}
+	\caption{Dataflow of the Eden |parEvalN| version}
+	\label{fig:parEvalNEden}
 \end{figure}
+A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
 
 \paragraph{Eden TraceViewer.}
+\label{sec:edentv}
 To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
@@
 A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.
-
-\olcomment{show example trace or refer to a trace in later figures}
+An example trace can be found in Fig.~\ref{fig:withoutFutures}.
 

### 2017-05-28 paper/content/arrows.tex
fixed Figure 8 and 9

@@
 An arrow |arr a b| represents  a computation that converts an input |a| to an output |b|.
-This is defined in the arrow typeclass:
+This is defined in the |Arrow| shown in Fig.~\ref{fig:ArrowDefinition}.
 
@@
 % begin{minipage}{0.5\textwidth}
-\parbox{0.49\linewidth}{%
+\parbox[c][16.5em]{0.49\linewidth}{%
+\begin{center}
+\vfill
 \begin{code}
@@
 \end{code}
+\vfill
+\end{center}
 \caption{Arrow class definition}
-}\qquad
-\begin{minipage}{0.49\linewidth}%
+\label{fig:ArrowDefinition}
+}
+\parbox[c][16.5em]{0.49\linewidth}{%
+\vfill
 %\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
@@
 %}%\hfill
+\vfill
 \caption{The schematic depiction of |Arrow| combinators}
-\end{minipage}
+}
 \end{figure}
-|arr| is used to lift an ordinary function to an arrow type, similarly to the monadic |return|.
+Its |arr| operation is used to lift an ordinary function to the specified arrow type, similarly to the monadic |return|.
 The |>>>| operator is analogous to the monadic composition  |>>=| and combines two arrows |arr a b| and |arr b c| by "wiring" the outputs of the first to the inputs to the second to get a new arrow |arr a c|.

### 2017-05-28 paper/content/arrows.tex
fixed arrow chapter according to olegs analog remarks

@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 \centering
-%\subfloat[Arrow class definition][]{%
-% begin{minipage}{0.5\textwidth}
 \parbox[c][16.5em]{0.49\linewidth}{%
@@
 \end{center}
-\caption{Arrow class definition}
+\caption{|Arrow| class definition}
 \label{fig:ArrowDefinition}
@@
 \vfill
-%\subfloat[The arrow type class definition on the left with schematic depiction of its combinators on the right][]{%
-%	\begin{center}
 \centering
@@
 	{\includegraphics[scale=0.6]{images/first}}
-%	\end{center}
-%}%\hfill
 \vfill
@@
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar (see Fig.~\ref{fig:syntacticSugarArrows}): the functions |second|, |***| and |&&&|. 
-The mirrored version of |first|, called |second| is:
+With this typeclass in place, Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): The combinators |second|, |***| and |&&&|. 
+The definition of |second|, the mirrored version of |first|, is:
 % \begin{figure}[h]
@@
 % \end{figure}
-the |***| combinator that combines |first| and |second| to handle two inputs in one arrow, is defined as
+The |***| combinator that combines |first| and |second| to handle two inputs in one arrow, is defined as
 % \begin{figure}[h]
@@
 % \end{figure}
-and the |&&&| combinator that constructs an arrow which outputs two different values like |***|, but takes only one input is:
+, while the |&&&| combinator, that constructs an arrow which outputs two different values like |***|, but takes only one input, is:
 % \begin{figure}[h]
@@
 % \end{figure}
-A short example given by Hughes on how to use this is 
-addition with arrows:
+A first short example given by Hughes on how to use arrows is addition with arrows:
 %|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.

### 2017-06-01 paper/content/basicSkeletons.tex
olegs remarks up until (excluding) section about futures

@@
 \subsubsection{Lazy |parEvalN|}
-\begin{figure}[h]
+\begin{figure}[t]
 	\includegraphics[scale=0.7]{images/parEvalNLazy}
@@
 The resulting |[[b]]| is lastly converted into |[b]| with |arr concat|.
-\begin{figure}[h]
+\begin{figure}[t]
 \begin{code}
@@
 \end{code} %$ %% formatting
-\caption{Definition of |parEvalNLazy|.}
+\caption{Definition of |parEvalNLazy|}
 \label{fig:parEvalNLazy}
@@
 	\includegraphics[scale=0.7]{images/parEval2}
-	\caption{Schematic depiction of |parEval2|.}
+	\caption{Schematic depiction of |parEval2|}
 	\label{fig:parEval2Img}
@@
 The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into parEvalN.
-\\\\
-We start by transforming the |(a, c)| input into a 2-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|.
-Next, we feed this list into a parallel arrow running on 2 instances of |f +++ g| as described above.
-After the calculation is finished, we convert the resulting |[Either b d]| into |([b], [d])| with |arr partitionEithers|.
-The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to |(b, d)| in the last step.
-\begin{figure}[h]
-\begin{code}
-parEval2 :: (ArrowChoice arr,
-	ArrowParallel arr (Either a c) (Either b d) conf) =>
-	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
-parEval2 conf f g =
-	arr Left *** (arr Right >>> arr return) >>>
-	arr (uncurry (:)) >>>
-	parEvalN conf (replicate 2 (f +++ g)) >>>
-	arr partitionEithers >>>
-	arr head *** arr head
-\end{code}
-	\caption{Definition of parEval2}
-	\label{fig:parEval2}
-\end{figure}
 

### 2017-06-01 paper/content/futures.tex
parallelHaskells.tex fixes for olegs (analog) comments.
olegs remarks up until (excluding) section about futures

@@
 	\label{fig:withoutFutures}
-\olcomment{more practical and heavy-weight example! fft (I have the code)?}
+\olcomment{more practical and heavy-weight example! fft (I have the code)?}\\
+\mbcomment{Depends... Are the communications easy to read in such an example?}\\
+\mbcomment{Keep the description for the different colours, or link to the EdenTV description in \ref{sec:edentv}}
 \end{figure}
@@
 \begin{code}
-class Future fut a @|@ a -> fut where
+class Future fut a | a -> fut where
     put :: (Arrow arr) => arr a (fut a)

### 2017-06-01 paper/content/parrows.tex
olegs remarks up until (excluding) section about futures

@@
 We have seen what Arrows are and how they can be used as a general interface to computation.
-In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well.
+In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textit{parallel computation} as well.
 We start by introducing the interface and explaining the reasonings behind it.
@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions |a -> b| in parallel, |parEvalN| (Chap.~\ref{sec:parEvalNIntro}, Fig.~\ref{fig:parEvalN}).
+As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions |a -> b| in parallel, |parEvalN| (Chap.~\ref{sec:parEvalNIntro}).
 Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
@@
 With this definition of |parEvalN|, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b| to host this combinator: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
 % \begin{figure}[h]
@@
 % \end{figure}
-Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs.
-normalform).
+Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs.
+normal form).
 For this reason we also introduce an additional |conf| parameter to the function.
 We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|.
-So we add it to the type signature of the typeclass as well and get |ArrowParallel arr a b conf|: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
+So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
 % \begin{figure}[h]
@@
 Note that we don't require the |conf| parameter in every implementation.
-If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|, as we will see in the implementation of the Multicore and the |Par| Monad backend.
+If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|, as we will see in the implementation of the Multicore Haskell and the |Par| Monad backends.
 
@@
 \subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of this class is implemented in a straightforward manner by using |listApp| from Appendix~\ref{utilfns} combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
-\begin{figure}[h]
+The Multicore Haskell implementation of this class is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
+\begin{figure}[t]
 \begin{code}
@@
 \end{figure}
-For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, we want to the user not to lose this ability because of using our API.
+For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this.
 We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|:
@@
 % \end{figure}
-We can't directly use the |Strategy a| type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
+We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances.
 To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
@@
 
-We therefore apply each function to its corresponding input value with || and then fork the computation away with |arr spawnP| inside a |zipWithArr| call.
+We therefore apply each function to its corresponding input value with || and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call.
 This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|.

### 2017-06-01 paper/content/utilityFunctions.tex
olegs remarks up until (excluding) section about futures

--- 2017-05-17 paper/content/utilityFunctions.tex
@@
 \section{Utility Functions}\label{utilfns}
-To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \code{map}, \code{foldl} and \code{zipWith} on arrows.
+To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: |map|, |foldl| and |zipWith| on arrows.
 
-The \code{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows}:
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any arrow |arr a
+b| to an arrow |arr [a] [b]| \cite{programming_with_arrows}. 
+Similarly, we can also define |foldlArr| (Fig.~\ref{fig:foldlArr}) that lifts any arrow |arr (b, a) b| with a neutral element |b| to |arr [a] b|.
+
+\begin{figure}[htb]
+\begin{code}
 mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
@@
 	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
-	where listcase [] = Left ()
+
+listcase [] = Left ()
 listcase (x:xs) = Right (x,xs)
-\end{lstlisting}
-\caption{\code{map} over arrows}
+\end{code}
+\caption{The definition of |map| over Arrows and the |listcase| helper
+function.}
 \label{fig:mapArr}
 \end{figure}
-Similarly, we can also define \code{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \code{arr (b, a) b} with a neutral element \code{b} to \code{arr [a] b}:
-\begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{figure}[htb]
+\begin{code}
 foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
@@
 		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
-	where listcase [] = Left []
-	      listcase (x:xs) = Right (x,xs)
-\end{lstlisting}
-\caption{\code{foldl} over arrows}
+\end{code}
+\caption{The definition of |foldl| over Arrows.}
 \label{fig:foldlArr}
 \end{figure}
-Finally, with the help of \code{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
+%\fixme{pipepipepipe does not work with lhs2TeX}
+
+Finally, with the help of |mapArr| (Fig.~\ref{fig:mapArr}), we can define |zipWithArr|  (Fig.~\ref{fig:zipWithArr}) that lifts any arrow |arr (a, b) c| to an arrow |arr ([a], [b]) [c]|.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
+\begin{code}
 zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
 zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
-\end{lstlisting}
-\caption{\code{zipWith} over arrows}
+\end{code}
+\caption{|zipWith| over arrows}
 \label{fig:zipWithArr}
@@
  %$ %% formatting
-These combinators make use of the \code{ArrowChoice} type class which provides the \code{(|||)} combinator.
-It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
+
+These combinators make use of the |ArrowChoice| type class which provides the \pipepipepipe combinator.
+It takes two arrows |arr a c| and |arr b c| and combines them into a new arrow |arr (Either a b) c| which pipes all |Left a|'s to the first arrow and all |Right b|'s to the second arrow.
+% \begin{figure}[htb]
+\begin{code}
+(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
+\end{code}
+% \caption{Type signature of \pipepipepipe}
+% \label{fig:codeSigPipePipePipe}
+% \end{figure}
+
+With the zipWithArr combinator we can also write a combinator |listApp|, that lifts a list of arrows |[arr a b]| to an arrow |arr [a] [b]|.
+% \begin{figure}[htb]
+\begin{code}
+listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
+listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
+\end{code}
+% \caption{Definition of |listApp|}
+% \label{fig:listApp}
+% \end{figure}
+% $ %% formatting
+Note that  this additionally makes use of the |ArrowApply| typeclass that allows us to evaluate arrows with |app :: arr (arr a b, a) c|.
+
+% $ %% formatting
+
+\section{Omitted Funtion Definitions}
+
+We have omitted some function definitions in the main text for
+brevity, and redeem this here.
+%
+We warp Eden's build-in Futures in PArrows as in
+Figure~\ref{fig:RDFuture}.
+The full definition of |farmChunk| is in Figure~\ref{fig:farmChunk}.
+Eden definition of |ring| skeleton following \citep{Loogen2012} is in Figure~\ref{fig:ringEden}.
+
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
-\end{lstlisting}
-\caption{Type signature of \code{(|||)}}
-\label{fig:codeSig|||}
+\begin{code}
+data RemoteData a = RD { rd :: RD a }
+
+instance (Trans a) => Future RemoteData a where
+    put = arr (\a -> RD { rd = release a })
+    get = arr rd >>> arr fetch
+\end{code}
+\caption{|RD|-based |RemoteData| version of |Future| for the Eden backend.}
+\label{fig:RDFuture}
 \end{figure}
 
-With the zipWithArr combinator we can also write a combinator \code{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]}.
 \begin{figure}[h]
-\begin{lstlisting}[frame=htrbl]
-listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
-listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
-\end{lstlisting}
-\caption{Definition of \code{listApp}}
-\label{fig:listApp}
+\begin{code}
+farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf, 
+             ArrowChoice arr, ArrowApply arr) =>
+	conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
+farmChunk conf chunkSize numCores f =
+	unshuffle numCores >>>
+	parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
+	shuffle
+\end{code}
+\caption{Definition of |farmChunk|.}
+\label{fig:farmChunk}
 \end{figure}
-% $ %% formatting
-. Note that  this additionally makes use of the \code{ArrowApply} typeclass that allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.
 
-% $ %% formatting
+\begin{figure}[h]
+\begin{code}
+ringSimple :: (Trans i, Trans o, Trans r) => (i -> r -> (o,r)) -> [i] -> [o]
+ringSimple f is =  os
+  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f) (zip is $ lazy ringIns))
+        ringIns = rightRotate ringOuts
+
+toRD :: (Trans i, Trans o, Trans r) => ((i,r) -> (o,r)) -> ((i, RD r) -> (o, RD r))
+toRD  f (i, ringIn)  = (o, release ringOut)
+  where (o, ringOut) = f (i, fetch ringIn)
+
+rightRotate    :: [a] -> [a]
+rightRotate [] =  []
+rightRotate xs =  last xs : init xs
+
+lazy :: [a] -> [a]
+lazy ~(x:xs) = x : lazy xs
+\end{code}
+\caption{Eden's definition of the |ring| skeleton.}
+\label{fig:ringEden}
+\end{figure}
+
+Furthermore, |parEval2| (Fig.~\ref{fig:parEval2}) is achieved as follows:
+We start by transforming the |(a, c)| input into a two-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|.
+Next, we feed this list into a parallel arrow running on two instances of |f +++ g| as described above.
+After the calculation is finished, we convert the resulting |[Either b d]| into |([b], [d])| with |arr partitionEithers|.
+The two lists in this tuple contain only one element each by construction, so we can finally just convert the tuple to |(b, d)| in the last step.
+\begin{figure}[h]
+\begin{code}
+parEval2 :: (ArrowChoice arr,
+	ArrowParallel arr (Either a c) (Either b d) conf) =>
+	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
+parEval2 conf f g =
+	arr Left *** (arr Right >>> arr return) >>>
+	arr (uncurry (:)) >>>
+	parEvalN conf (replicate 2 (f +++ g)) >>>
+	arr partitionEithers >>>
+	arr head *** arr head
+\end{code}
+	\caption{Definition of parEval2}
+	\label{fig:parEval2}
+\end{figure}
+
+\section{Syntactic Sugar} \label{syntacticSugar}
+For basic arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two arrows |arr a b| and |arr c d| into an arrow |arr (a, c) (b, d)| which does both computations at once.
+This can easily be translated into a parallel version |***| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
+):
+% \begin{figure}[h]
+\begin{code}
+(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
+	arr a b -> arr c d -> arr (a, c) (b, d)
+(|***|) = parEval2 ()
+\end{code}
+% \caption{Definition of |parstar|, the parallel version of |***|.}
+% \label{fig:par***}
+% \end{figure}
+% With this we can analogously to the serial |&&&|
+We define the parallel |&&&| % (Fig.~\ref{fig:par&&&}) 
+in a similar manner to its sequential pendant |&&&| (Fig.~\ref{fig:syntacticSugarArrows}):
+% \begin{figure}[h]
+\begin{code}
+(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
+	arr a b -> arr a c -> arr a (b, c)
+(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
+\end{code} % $ %% formatting
+% \caption{Definition of |parand| - the parallel version of |&&&|.}
+% \label{fig:par&&&}
+% \end{figure}
 

### 2017-06-01 paper/content/futures.tex
added dots at the end of figure names.

@@
 \end{code}
-\caption{An example parallel Arrow combinator without Futures}
+\caption{An example parallel Arrow combinator without Futures.}
 \label{fig:someCombinator}
@@
 \end{code}
-\caption{Definition of the Future typeclass}
+\caption{Definition of the |Future| typeclass.}
 \label{fig:future}
@@
 \end{code}
-\caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends}
+\caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends.}
 \label{fig:MVarFuture}
@@
 \end{code}
-\caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel}
+\caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel.}
 \label{fig:someCombinatorParallel}

### 2017-06-01 paper/content/parallelHaskells.tex
added dots at the end of figure names.

@@
 \end{code}
-\caption{Multicore version of |parEvalN|}
+\caption{Multicore version of |parEvalN|.}
 \label{fig:parEvalNMulticore}
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-	\caption{Dataflow of the Multicore Haskell |parEvalN| version}
+	\caption{Dataflow of the Multicore Haskell |parEvalN| version.}
 	\label{fig:parEvalNMulticoreImg}
@@
 \end{code}
-\caption{|Par| Monad version of |parEvalN|}
+\caption{|Par| Monad version of |parEvalN|.}
 \label{fig:parEvalNParMonad}
@@
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-	\caption{Dataflow of the |Par| Monad |parEvalN| version}
+	\caption{Dataflow of the |Par| Monad |parEvalN| version.}
 	\label{fig:parEvalNParMonadImg}
@@
 	\includegraphics[scale=0.5]{images/parEvalNEden}
-	\caption{Dataflow of the Eden |parEvalN| version}
+	\caption{Dataflow of the Eden |parEvalN| version.}
 	\label{fig:parEvalNEden}

### 2017-06-01 paper/content/parrows.tex
added dots at the end of figure names.

@@
 \end{code}% $ %% formatting
-\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend}
+\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend.}
 \label{fig:ArrowParallelMulticoreRdeepseq}
@@
 \end{code}
-\caption{Configurable ArrowParallel instance for the Multicore Haskell backend}
+\caption{Configurable ArrowParallel instance for the Multicore Haskell backend.}
 \label{fig:ArrowParallelMulticoreConfigurable}
@@
 \end{code} %$ %% formatting
-\caption{ArrowParallel instance for the Par Monad backend}
+\caption{ArrowParallel instance for the Par Monad backend.}
 \label{fig:ArrowParallelParMonad}

### 2017-06-01 paper/content/topologySkeletons.tex
added dots at the end of figure names.

@@
 	\includegraphics[scale=0.75]{images/ring}
-	\caption{Schematic depiction of the ring skeleton}
+	\caption{Schematic depiction of the ring skeleton.}
 	\label{fig:ringImg}
@@
 \end{code}
-\caption{Adapted |nodefunction| for matrix multiplication with the |torus| from Fig.~\ref{fig:torus}}
+\caption{Adapted |nodefunction| for matrix multiplication with the |torus| from Fig.~\ref{fig:torus}.}
 \label{fig:torusMatMult}
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (Parrows)}
+	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (PArrows).}
 	\label{fig:torus_parrows_trace}
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
-	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden)}
+	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden).}
 	\label{fig:torus_eden_trace}

### 2017-06-01 paper/content/parrows.tex
minor

@@
 % \begin{figure}[h]
-\olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?}
+\olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
 \begin{code}
@@
 \subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of this class is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
+The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
 \begin{figure}[t]

### 2017-06-02 paper/content/futures.tex
cut parMapReduceDirect from paper. moved things around.

@@
 This can become a serious bottleneck %in heavy threaded applications.
-for larger amount of data and number of processes \citep[showcases][as, \eg]{Berthold2009-fft}.
+for larger amount of data and number of processes \citep[as \eg][showcases]{Berthold2009-fft}.
 \begin{figure}[ht]
@@
 	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
-	\caption[without Futures]{Communication between 4 threads without Futures.
+	\caption[without Futures]{Communication between 4 Eden processes without Futures.
 All communication goes through the master node.
@@
 
-This motivates for an approach that allows the nodes to communicate directly with each other.
-Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of |RD| (remote data) that enables this behaviour \cite{AlGo03a,Dieterle2010}.
+We should allow the nodes to communicate directly with each other. Eden already ships with "remote data" that enable this \cite{AlGo03a,Dieterle2010}.
 
-But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
+But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
 \begin{figure}[h]
@@
 \end{figure}
-Since |RD| is only a type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Appendix in Fig.~\ref{fig:RDFuture}.
+Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture}.
 This is due to the same reason we had to introduce a wrapper for |Strategy a| in the Multicore Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
 
-For our Par Monad and Multicore Haskell backends, we can simply use |MVar|s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), because we have shared memory in a single node and don't require Eden's sophisticated communication channels.
+For our |Par| Monad and Multicore Haskell backends, we can simply use |MVar|s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), as in a shared memory setting we do not require Eden's sophisticated communication channels.
 \fixme{explain MVars}
@@
 \end{code}
-\caption{MVar instance of the Future typeclass for the Par Monad and Multicore Haskell backends.}
+\caption{A |MVar| instance of the |Future| typeclass for the |Par| Monad and Multicore Haskell backends.}
 \label{fig:MVarFuture}
@@
 Furthermore, in order for these |Future| types to fit with the |ArrowParallel| instances we gave earlier, we have to give the necessary |NFData| and |Trans| instances, the latter are only needed in Eden.
-Because |MVar| already ships with a |NFData| instance, we only have to supply two simple instances for our |RemoteData| type:
+The |Trans| instance does not have any functions declared as the default implementation suffices here.
+Furthermore, because |MVar| already ships with a |NFData| instance, we only have to supply a simple delegating |NFData| instance for our |RemoteData| type, where |rd| simply unwraps |RD|:
 % \begin{figure}[h]
@@
 % \end{figure}
-The |Trans| instance does not have any functions declared as the default implementation suffices here.
 
-Going back to our communication example we can use this |Future| concept in order to enable direct communications between the nodes in the following way:
-\begin{figure}[h]
+In our communication example we can use this |Future| concept for direct communications between the nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
+\begin{figure}[t]
 \begin{code}
@@
 \end{figure}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear.
-Do Figs 2-3 with a lot of load?}
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.
+is not really clear.
+Do Figs with a lot of load?}
 \begin{figure}[ht]
@@
 	\includegraphics[width=0.9\textwidth]{images/withFutures}
-	\caption[with Futures]{Communication between 4 threads with Futures.
-Other than in Fig.~\ref{fig:withoutFutures}, threads communicate directly (black lines between the bars) instead of always going through the master node (bottom bar).}
+	\caption[with Futures]{Communication between 4 Eden processes with Futures.
+Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (black lines between the bars) instead of always going through the master node (bottom bar).}
 	\label{fig:withFutures}

### 2017-06-02 paper/content/mapSkeletons.tex
added dots at the end of figure names.
cut parMapReduceDirect from paper. moved things around.

@@
 \label{sec:map-skeletons}
-Now we have developed Parallel Arrows far enough to define some algorithmic skeletons useful to an application programmer.
+Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel tasks.
+
 \subsection{Parallel map}
+
 \begin{figure}[h]
+%parMap
 \includegraphics[scale=0.7]{images/parMap}
@@
 \label{fig:parMapImg}
-\end{figure}
-The |parMap| skeleton (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
-We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to get an arrow |arr [a] [b]|.
-Just like |parEvalN|, |parMap| is 100\% strict.
-\begin{figure}[h]
+
 \begin{code}
@@
 \end{code}
-\caption{Definition of parMap}
+\caption{Definition of parMap.}
 \label{fig:parMap}
-\end{figure}
 
-\subsection{Lazy parallel map}
-\begin{figure}[h]
+%parMapStream
 \includegraphics[scale=0.7]{images/parMapStream}
-	\caption{Schematic depiction of parMapStream}
+\caption{Schematic depiction of parMapStream.}
 \label{fig:parMapStreamImg}
-\end{figure}
-As |parMap| (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as |parEvalN| compared to |parEvalNLazy|.
-So it makes sense to also have a |parMapStream| (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
-\begin{figure}[h]
+
 \begin{code}
@@
 
+The |parMap| skeleton (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
+We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to get an arrow |arr [a] [b]|.
+Just like |parEvalN|, |parMap| is 100\% strict.
+
+\paragraph{Lazy parallel map}
+As |parMap| (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as |parEvalN| compared to |parEvalNLazy|.
+So it makes sense to also have a |parMapStream| (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
+
 \subsection{Statically load-balancing parallel map}
+
 \begin{figure}[h]
+%farm
 \includegraphics[scale=0.7]{images/farm}
@@
 \label{fig:farmImg}
-\end{figure}
-A |parMap| (Fig.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper).
-This can be quite wasteful and a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
-The definitions of helper functions |unshuffle|, |takeEach|, |shuffle| (shown in Appendix) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
-\begin{figure}[h]
+
 \begin{code}
@@
 	shuffle
-
-unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
-unshuffle n = arr (\xs -> [takeEach n (drop i xs) @|@ i <- [0..n-1]])
-
-takeEach :: Int -> [a] -> [a]
-takeEach n [] = []
-takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
-
-shuffle :: (Arrow arr) => arr [[a]] [a]
-shuffle = arr (concat . transpose)
 \end{code}
@@
 \label{fig:farm}
+
+%farmChunk
+\includegraphics[scale=0.7]{images/farmChunk}
+\caption{Schematic depiction of farmChunk.}
+\label{fig:farmChunkImg}
 \end{figure}
 
-\subsection{The |farmChunk| Skeleton}
-\begin{figure}[h]
-	\includegraphics[scale=0.7]{images/farmChunk}
-\caption{Schematic depiction of farmChunk}
-\label{fig:farmChunkImg}
-\end{figure}
+A |parMap| (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper).
+This can be quite wasteful and a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
+The definitions of helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
+
+\paragraph{Lazy statically load-balancing parallel map}
 Since a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just |parMap| with a different work distribution, it is, again, 100\% strict.
 So we can define |farmChunk| (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|.
-It is basically the same definition as for |farm|, with |parEvalN| replaced with |parEvalNLazy|, as Appendix shows.
-
-\subsection{Map and reduce}
-
-A~simple |map|--|reduce| can be written like in Figure~\ref{fig:parMapReduceDirect}.
-Notice that the performance of the |>>>| combinator is essential for the performance of the skeleton.
-A~definitive version would use Futures.
-
- \olcomment{it appears STRANGE. are the data really left alone and noded after map and taken from there by reduce? It makes sense only is no communication through master takes place. ELSE: CUT!}
-
- \mbcomment{this requires some work. Either change this to use futures or cut, yes.}
-% -- this does not completely adhere to Google's definition of Map Reduce as it
-% -- the mapping function does not allow for "reordering" of the output
-% -- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce
-
-\olcomment{now rewritten as motivation for futures. maybe still cut?}
-
-\begin{figure}[h]
-\begin{code}
-parMapReduceDirect :: (ArrowParallel arr [a] b conf, ArrowApply arr, ArrowChoice arr) =>
-	conf -> ChunkSize -> arr a b -> arr (b, b) b -> b -> arr [a] b
-parMapReduceDirect conf chunkSize mapfn foldfn neutral =
-	arr (chunksOf chunkSize) >>>
-	parMap conf (mapArr mapfn >>> foldlArr foldfn neutral) >>>
-	foldlArr foldfn neutral
-\end{code}
-\caption{Definition of parMapReduceDirect}
-\label{fig:parMapReduceDirect}
-\end{figure}
+It is basically the same definition as for |farm|, with |parEvalN| replaced with |parEvalNLazy|.
 

### 2017-06-02 paper/content/utilityFunctions.tex
added dots at the end of figure names.
cut parMapReduceDirect from paper. moved things around.

@@
 \end{code}
-\caption{|zipWith| over arrows}
+\caption{|zipWith| over arrows.}
 \label{fig:zipWithArr}
@@
 We warp Eden's build-in Futures in PArrows as in
-Figure~\ref{fig:RDFuture}.
+Figure~\ref{fig:RDFuture}. Arrow versions of Eden's |shuffle|, |unshuffle| and the definition of |takeEach| are in Figure~\ref{fig:edenshuffleetc}.
 The full definition of |farmChunk| is in Figure~\ref{fig:farmChunk}.
@@
 \label{fig:RDFuture}
+\end{figure}
+
+\begin{figure}[h]
+\begin{code}
+shuffle :: (Arrow arr) => arr [[a]] [a]
+shuffle = arr (concat . transpose)
+
+unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
+unshuffle n = arr (\xs -> [takeEach n (drop i xs) | i <- [0..n-1]])
+
+takeEach :: Int -> [a] -> [a]
+takeEach n [] = []
+takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
+\end{code}
+\caption{Definitions of |shuffle|, |unshuffle|, |takeEach|}
+\label{fig:edenshuffleetc}
 \end{figure}
@@
 \end{code}
-	\caption{Definition of parEval2}
+	\caption{Definition of parEval2.}
 	\label{fig:parEval2}

### 2017-06-02 paper/content/mapSkeletons.tex
.

@@
 This can be quite wasteful and a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
-The definitions of helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
+The definitions of the helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
 

### 2017-06-02 paper/content/topologySkeletons.tex
cut parMapReduceDirect from paper. moved things around.
minor

@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} with more predefined skeletons, among them a |pipe|, a |ring|, and a |torus| implementations \cite{Loogen2012}.
+The Eden library leverages this problem and already comes with more predefined skeletons\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.}, among them a |pipe|, a |ring|, and a |torus| implementations \cite{Loogen2012}.
 These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell.
@@
 For example %this
-\mbcomment{das kann man hier so lassen, oder?}\olcomment{sicherlich!}
 \begin{code}
@@
 \end{figure}
-Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function |[i] -> [o]| with a ring of nodes that communicate in a ring topology with each other.
+Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function |[i] -> [o]| with a ring of nodes that communicate in a ring topology with each other.
 Its input is a node function |i -> r -> (o, r)| in which |r| serves as the intermediary output that gets send to the neighbour of each node.
@@
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.0.0.2/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.
+With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.
 

### 2017-06-02 paper/content/topologySkeletons.tex
wip

@@
 \begin{figure}[h]
-\olcomment{I swapped the type classes here:}
 \begin{code}
@@
 				arr (\((a, b), c) -> ((c, a), b))
-\end{code}
-\caption{Definition of |pipe2|.}
-\label{fig:pipe2}
-\end{figure}
 
-Note that extensive use of this combinator over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator |parcomp| (Fig.~\ref{fig:par>>>}, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Section~\ref{syntacticSugar}.
-\begin{figure}[h]
-\olcomment{swapped type classes}
-\begin{code}
 (|>>>|) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
@@
 \end{code}
-\caption{Definition of |parcomp|.}
-\label{fig:par>>>}
+\caption{Definition of |pipe2| and parallel |>>>|.}
+\label{fig:pipe2}
 \end{figure}
+\mbcomment{parallel >>>!!!!!!}
+
+Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Nonetheless, we can define a parallel piping operator |>>>|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
 
@@
 In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)|.
-Similarly to the |pipe| from Chapter~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
-\begin{quote}
-{Note that the list of ring inputs |ringIns| is the same as the list of ring outputs |ringOuts| rotated by one element to the right using the auxiliary function |rightRotate|.
-Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
-\end{quote}
+Similarly to the |pipe| from Chapter~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012} as a demand problem.
+
 Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|.
@@
 Similar to the |ring|, we once again start by rotating the input, but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes |second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))| instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically.
-Then, we once again zip the inputs for the input arrow with |arr (uncurry3 zipWith3 lazyzip3)| from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then feed into our parallel execution.
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes
+\begin{code}
+second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))
+\end{code}
+instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we once again zip the inputs for the input arrow with 
+\begin{code}
+arr (uncurry3 zipWith3 lazyzip3)
+\end{code}
+from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then feed into our parallel execution.
 

### 2017-06-02 paper/content/topologySkeletons.tex
moving definitions into utilityFunctions.tex

@@
 Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |parEvalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
-\begin{figure}[h]
+
+However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
+Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of |pipe| in Fig.~\ref{fig:pipe}.
+
+\begin{figure}[t]
+
 \begin{code}
@@
 	arr last
-
-lazy :: (Arrow arr) => arr [a] [a]
-lazy = arr (\ ~(x:xs) -> x : lazy xs)
 \end{code}
 \caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows.
-Note that the use of |lazy| is essential as without it programs using this definition would never halt.
+Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt.
 We need to enforce that the evaluation of the input |[a]| terminates before passing it into |parEvalN|.}
 \label{fig:pipeSimple}
-\end{figure}
 
-However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of |pipe|: % ~\ref{fig:pipe}.
-%\begin{figure}[H]
 \begin{code}
@@
 \end{code}
-%\caption{Final definition of the pipe skeleton which uses Futures}
-%\label{fig:pipe}
-%\end{figure}
+\caption{Final definition of the pipe skeleton which uses Futures}
+\label{fig:pipe}
+\end{figure}
 
@@
 We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an arrow, already fits quite neatly into the |loop|'s |arr (a, b) (c, b) -> arr a c|.
-In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)|.
-Similarly to the |pipe| from Chapter~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012} as a demand problem.
+In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)| (Fig.~\ref{fig:edenlazyrightrotate}).
+Similarly to the |pipe| from Chapter~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012} as a demand problem.
 
@@
         parMap conf (second get >>> f >>> second put) >>> arr unzip)
-
-rightRotate :: (Arrow arr) => arr [a] [a]
-rightRotate = arr $ \list -> case list of
-  [] -> []
-  xs -> last xs : init xs
-
-lazy :: (Arrow arr) => arr [a] [a]
-lazy = arr (\ ~(x:xs) -> x : lazy xs)
 \end{code}
@@
 
-Similar to the |ring|, we once again start by rotating the input, but this time not only in one direction, but in two.
+Similar to the |ring|, we once again start by rotating the input (Fig.~\ref{fig:edenlazyrightrotate}), but this time not only in one direction, but in two.
 This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes
@@
 This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
-We first have to |shuffle| all the inputs to then pass it into |parMap conf (ptorus f)| which yields us |[(d, fut a, fut b)]|.
+We first have to |shuffle| all the inputs to then pass it into |parMap conf (ptorus f)| which yields |[(d, fut a, fut b)]|.
 We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass.
@@
 \begin{figure}[h]
-\olcomment{swapped type classes}
 \begin{code}
@@
 ptorus f = arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>> arr (\ ~(d, a, b) -> (d, put a, put b))
-
-uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
-uncurry3 f (a, (b, c)) = f a b c
-
-lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
-lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
-
-threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
-threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
 \end{code} % $
-\caption{Definition of the |torus| skeleton.}
+\caption{Definition of the |torus| skeleton. The definitions of |lazyzip3|, |uncurry3| and |threetotwo| have been omitted and can be found in Fig.~\ref{fig:lazyzip3etc}}.
 \label{fig:torus}

### 2017-06-02 paper/content/utilityFunctions.tex
moving definitions into utilityFunctions.tex

@@
-\section{Utility Functions}\label{utilfns}
+\section{Utility Arrows}\label{utilfns}
 To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: |map|, |foldl| and |zipWith| on arrows.
@@
 
-\begin{figure}[htb]
+\begin{figure}[h]
 \begin{code}
@@
 \end{figure}
-\begin{figure}[htb]
+
+\begin{figure}[h]
 \begin{code}
@@
 \end{figure}
-%\fixme{pipepipepipe does not work with lhs2TeX}
 
@@
 We warp Eden's build-in Futures in PArrows as in
-Figure~\ref{fig:RDFuture}. Arrow versions of Eden's |shuffle|, |unshuffle| and the definition of |takeEach| are in Figure~\ref{fig:edenshuffleetc}.
+Figure~\ref{fig:RDFuture}.
+Arrow versions of Eden's |shuffle|, |unshuffle| and the definition of |takeEach| are in Figure~\ref{fig:edenshuffleetc}.
+Similarly Figure~\ref{fig:edenlazyrightrotate} contains the definition of arrow versions of Eden's |lazy| and |rightRotate| utility functions.
+Fig.~\ref{fig:lazyzip3etc} contains Eden's definition of |lazyzip3| together with the utility functions |uncurry3| and |threetotwo|.
 The full definition of |farmChunk| is in Figure~\ref{fig:farmChunk}.
@@
 \end{code}
-\caption{Definitions of |shuffle|, |unshuffle|, |takeEach|}
+\caption{Definitions of |shuffle|, |unshuffle|, |takeEach|.}
 \label{fig:edenshuffleetc}
+\end{figure}
+
+\begin{figure}[h]
+\begin{code}
+lazy :: (Arrow arr) => arr [a] [a]
+lazy = arr (\ ~(x:xs) -> x : lazy xs)
+
+rightRotate :: (Arrow arr) => arr [a] [a]
+rightRotate = arr $ \list -> case list of
+  [] -> []
+  xs -> last xs : init xs
+\end{code}
+\caption{Definitions of |lazy| and |rightRotate|.}
+\label{fig:edenlazyrightrotate}
+\end{figure}
+
+\begin{figure}[h]
+\begin{code}
+lazyzip3 :: [a] -> [b] -> [c] -> [(a, b, c)]
+lazyzip3 as bs cs = zip3 as (lazy bs) (lazy cs)
+
+uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d
+uncurry3 f (a, (b, c)) = f a b c
+
+threetotwo :: (Arrow arr) => arr (a, b, c) (a, (b, c))
+threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
+\end{code}
+\caption{Definitions of |lazyzip3|, |uncurry3| and |threetotwo|.}
+\label{fig:lazyzip3etc}
 \end{figure}

### 2017-06-12 paper/content/conclusion.tex
initial
renamed \code to \inlinecode
minor stuff.

--- 2017-03-30 paper/conclusion.tex
@@
 %
-Performancewise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}
+Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}
 
@@
 \olcomment{Remove websites from citations, put them into footnotes!}
+
+\olcomment{Parrows + accelerate = love? Metion port to Frege. Mention the Par monad troubles.}
+
+\subsection{Future Work}
+\label{sec:future-work}
+
+Our PArrows interface can be expanded to futher parallel Haskells.
+More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} a modern distributed Haskell that would benefit from our Arrows notation.
+Future-aware special versions of Arrow combinated can be extended and further improved.
+We would look into more transparency of the DSL, it should basically infuse as little overhead as possible.
+
+Of course, definitions of further skeletons are viable and needed. We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
+
+Accelerate~\cite{Chakravarty:2011:AHA:1926354.1926358} is not related to our approach. It would be interesting to see a hybrid of both APIs.
+
+\olcomment{replace API with DSL globally?}

### 2017-06-12 paper/content/parallelHaskells.tex
minor stuff.

@@
-\section{Background}
+\begin{flushright}
+
+\end{flushright}\section{Background}
 \label{sec:background}
@@
 \subsubsection{ParMonad}
-The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{monad_par_paper_2011}, is a monad designed for composition of parallel programs.
+The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a monad designed for composition of parallel programs.
 

### 2017-06-12 paper/content/topologySkeletons.tex
minor stuff.

@@
 \end{figure}
-As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}.
+As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citeyear{Gentleman1978}.
 Their instantiation of the skeleton |nodefunction| can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
@@
 \end{figure}
-If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.\olcomment{much more details on this!}
+If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version and execution times are comparable.\olcomment{much more details on this!}
 \begin{figure}[ht]

### 2017-06-14 paper/build.sh
removed removed old paper tex files

@@ -1,15 +1,0 @@
-#!/bin/bash
-
-# works with:
-# texlive-base                                    install
-# texlive-binaries                                install
-# texlive-fonts-extra                             install
-# texlive-fonts-recommended                       install
-# texlive-generic-recommended                     install
-# texlive-latex-base                              install
-# texlive-latex-extra                             install
-# texlive-latex-recommended                       install
-# texlive-pictures                                install
-# texlive-science                                 install
-
-latexmk main.tex -pdf

### 2017-06-14 paper/examples/parallelism-parmonad.hs

@@ -1,32 +1,0 @@
-{-
-The MIT License (MIT)
-
-Copyright (c) 2016 Martin Braun
-
-Permission is hereby granted, free of charge, to any person obtaining
-a copy of this software and associated documentation files (the
-"Software"), to deal in the Software without restriction, including
-without limitation the rights to use, copy, modify, merge, publish,
-distribute, sublicense, and/or sell copies of the Software, and to
-permit persons to whom the Software is furnished to do so, subject to
-the following conditions:
-
-The above copyright notice and this permission notice shall be included
-in all copies or substantial portions of the Software.
-
-THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
-EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
-MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
-IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY
-CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT,
-TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE
-SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
--}
-module Main where
-
-import Control.Monad.Par
-
-parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = runPar $ (sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
-
-main = do print $ parEvalN [(*2), (+1)] [1::Int, 2]

### 2017-06-24 benchmarks/bench.sh
benchmarks

@@ -1,0 +1,45 @@
+#!/bin/bash
+
+procCount=640
+
+programs=(
+    "parrows-sudoku-parmap"
+    "parrows-sudoku-parmapstream"
+    "parrows-sudoku-farm"
+    "parrows-sudoku-farmChunk"
+)
+
+parameters=(
+    ""
+    "384"
+    "12"
+    "384 12"
+)
+
+puzzles=(
+    # "sudoku17.1000.txt"
+    # "sudoku17.16000.txt"
+    # "sudoku17.49151.txt"
+    "sudoku17.long.txt"
+)
+
+# get length of an array
+programCount=${#programs[@]}
+
+benchCmds=""
+
+for puzzle in "${puzzles[@]}"
+do
+    for (( i=0; i < ${programCount}; i++ ));
+    do
+        for (( j=${procCount}; j>=32; j=j/2 ));
+        do
+            progName=${programs[$i]}
+            cmd="\"./"${progName}" "${puzzle}" "${parameters[$i]}" +RTS -N"${j}"\""
+
+            benchCmds=${benchCmds}" "${cmd}
+        done
+    done
+done
+
+eval "bench"${benchCmds}" --csv bench.csv"

### 2017-06-24 benchmarks/fft/Main.hs

@@ -1,0 +1,81 @@
+{-# OPTIONS -cpp #-}
+module Main where
+
+-- import MathObj.FFT.Radix4Par as R4 (fft,fftDo2,fftDo3)
+import MathObj.FFT.Radix4Seq as R4 (fft)
+import MathObj.FFT.Reference as R2 (fft, ifft)
+import MathObj.FFT.MultiPar (twoDimFft, twoDimFftG, twoDimFftRD)
+import Prelude
+-- import System.CPUTime
+import Data.Complex
+import MathObj.FFT.Complex
+import Control.Concurrent
+#ifdef TEST
+import qualified MathObj.FFT.Reference as Ref
+#endif
+
+import System.Environment (getArgs)
+import System.IO
+import Control.Parallel.Eden (rnf, Trans)
+import Control.Parallel.Eden.Map (farmS)
+
+instance (RealFloat a, Trans a) => Trans (Complex a) -- ???
+
+main = do
+  hSetBuffering stdout NoBuffering
+  putStrLn "usage: testfftTest <size or base for power of 4(1024) of one side> <version 0,1,2,3,4(3)> <chunksize(512), no effect?> <prefetch(20)>"
+  args <- getArgs
+  let f = mkLists n
+--      n = if null args then 512 else read (head args)
+      n = if b > 15 then b else 4^b
+      b = if null args then 1024 else read (head args)
+      v = if length args < 2 then 3 else read (args!!1)
+      s = if length args < 3 then 512 else read (args!!2)
+      p = if length args < 4 then 20 else read (args!!3)
+--  let r = fftDo3 s (n, f)
+--      r2 = fftDo2 s (n,f)
+      r4 = twoDimFft n n f
+      r5 = twoDimFftG map R4.fft n n f
+      r6 = twoDimFftG (farmS p) R2.fft n n f
+      r7 = twoDimFftG (farmS p) R4.fft n n f
+      r8 = twoDimFftRD (farmS p) R4.fft n n f
+#ifdef TEST
+  let (len,resultlist) = r
+  putStrLn ("Length given as " ++ show len ++ 
+	    ", in reality " ++ show (length resultlist))
+  let ref = Ref.fft (n,f)
+  if (r /= ref) then putStrLn "WRONG RESULT" else putStrLn "CORRECT"
+  rnf r2 `seq` 
+--  (length (snd r)) `seq`
+--  (fst r ) `seq` 
[…]

### 2017-06-24 benchmarks/fft/Parallel/DivConN.hs

@@ -1,0 +1,161 @@
+-- divide and conquer skeleton for regular n-ary tree
+-- 2008, Jost Berthold, 
+-- Philipps-Universitaet Marburg, AG Eden
+--------------------------------------------------------
+module Parallel.DivConN where
+
+import Parallel.Eden
+
+import System.IO.Unsafe
+import Control.Monad
+import Control.Parallel.Strategies(seqList,Strategy, rwhnf)
+
+
+{-
+parMapEden :: (Trans a, Trans b) => (a -> b) -> [a] -> [b]
+parMapEden f xs = map deLift ([ createProcess (process f) x
+                               | x <- xs ] `using` whnfspine)
+-- demand control helper
+whnfspine :: Strategy [a]
+whnfspine [] = ()
+whnfspine (x:xs) = x `seq` whnfspine xs
+
+parMapEden2 :: (Trans a, Trans b) => (a -> b) -> [a] -> [b]
+parMapEden2 f xs = unsafePerformIO ( mapM (instantiate (process f)) xs )
+-}
+
+-- Eden parmap, placement:
+parMapAt :: (Trans a, Trans b) => [Int] -> (a -> b) -> [a] -> [b]
+parMapAt places f xs
+    = unsafePerformIO (
+         zipWithM (\pe x -> instantiateAt pe (process f) x)
+                  places xs
+        )
+
+spawnAt :: (Trans a, Trans b) => [Int] -> [Process a b] -> [a] -> [b]
+spawnAt places ps is = unsafePerformIO
+            (sequence
+             [instantiateAt st p i |
+              (st,p,i) <- zip3 (cycle places) ps is]
+            )
+
+          
+
+-- distributed expansion (tree-shaped process creation, one tasks stays local)
+dcN :: (Trans a, Trans b) => 
+       Int -> Int                 -- n (expect n children) / remaining depth
+       -> (a -> Bool) -> (a -> b) -- trivial? / solve
+       -> (a -> [a]) -> ([b] -> b) -- split / combine
+       ->  a -> b                 -- resulting mapping
+dcN n depth trivial solve split combine x 
+   = dcN_c n depth trivial solve split (\_ parts -> combine parts) x
+
+dcN_c :: (Trans a, Trans b) => 
+       Int -> Int                 -- n (expect n children) / remaining depth
+       -> (a -> Bool) -> (a -> b) -- trivial? / solve
[…]

### 2017-06-24 benchmarks/fft/Parallel/DivConRW.hs

@@ -1,0 +1,109 @@
+{-# OPTIONS -cpp #-}
+{- Divide and conquer skeleton on top of replicated workers.
+   The first parameter of divCon defines the depth at which
+   the divide&conquer tree is generated by the main tasks,
+   using function generateTasks. After generating enough tasks,
+   the replicated workers skeleton is used to solve them with
+   a good load balance. Finally, function combineTop combine
+   the partial results obtained.
+   The data type Tree is needed to pass information from 
+   generateTasks to combineTop, so that the subresults can
+   be combined in the proper way.   
+   
+                                              -- FR10 --
+-}                                              
+   
+
+module Parallel.DivConRW(divConRW,divConFarm,farm) where
+import Parallel.Eden
+-- import RW 
+
+-- workpools by MD
+-- import EdenMWToken
+
+-- edi workpool, sorting
+import Parallel.EdiWP
+
+import Data.List
+import System.IO.Unsafe
+import Control.Monad
+
+mw :: (Trans a, Trans b) => 
+      Int -> Int -> (a -> b) -> [a] -> [b]
+mw = ediWPf
+
+-- farm:
+unshuffle :: Int -> [a] -> [[a]]
+unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
+takeEach :: Int -> [a] -> [a] 
+takeEach n [] = []
+takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
+shuffle :: [[a]] -> [a]
+shuffle = concat . transpose
+
+farm :: (Trans a, Trans b) => Int -> (a -> b) -> [a] -> [b]
+farm np f ts = let inputss = unshuffle np ts 
+	       in shuffle (parMapAt (repeat 0) (map f) inputss)
+
+parMapAt :: (Trans a, Trans b) => [Int] -> (a -> b) -> [a] -> [b]
+parMapAt places f xs
+    = unsafePerformIO (
+         zipWithM (\pe x -> instantiateAt pe (process f) x)
+                  places xs
+        )
+
+       
+divConSeq :: (a->Bool) -> (a->b) -> (a->[a]) -> (a->[b]->b) -> a -> b
+divConSeq trivial solve split combine x 
+ | trivial x = solve x
+ | otherwise = combine x children
+ where children = map (divConSeq trivial solve split combine) (split x)
+
+
[…]

### 2017-06-24 benchmarks/fft/Parallel/Eden.hs

@@ -1,0 +1,418 @@
+{-# OPTIONS -cpp  -fglasgow-exts #-}
+
+-- Eden module, defining high-level coordination concepts via
+-- Prim.Op.s (which are wrapped inside ParPrim.hs. This version
+-- conforms exactly to the code presented at the IFL workshop,
+-- regarding functionality, types and function names.
+
+module Parallel.Eden(
+	-- reexported from Strategies ---
+	NFData(..), using, r0
+        ---------- basic Eden -----------
+	, noPe, selfPe
+	, Process, process, ( # )
+	, instantiate
+	, instantiateAt     -- explicit placement
+	, Trans(..)
+	------- dynamic channels --------
+	, ChanName          -- Communicator a -> IO(), abstract outside
+	, new, parfill      -- using unsafePerformIO
+	---------------------------------
+	, merge, mergeProc  -- merge, as specified in Eden language, but function!
+	---------------------------------
+	, Lift(..), deLift, createProcess, cpAt -- deprecated legacy code for Eden 5
+    )
+  where
+
+#ifndef __PARALLEL_HASKELL__
+#warning Compiling  a sequential version of Eden.hs
+#endif
+    
+-- load typecheck simulation if not compiling parallel
+#ifndef __SIMUL__
+import qualified Parallel.ParPrim as ParPrim
+import Parallel.ParPrim hiding(noPe,selfPe)
+#else
+import qualified Parallel.ParPrimSim as ParPrim
+import Parallel.ParPrimSim hiding(noPe,selfPe)
+#endif
+import Control.Concurrent      -- Instances only
+import System.IO.Unsafe(unsafePerformIO)
+
+import Control.Parallel.Strategies -- reexported!
+        (NFData(..),using, r0)
+
+--------------------------
+-- legacy code for Eden 5:
+
+{-# DEPRECATED deLift, Lift "Lift data type not needed in Eden 6 implementation" #-}
+data Lift a = Lift a
+deLift :: Lift a -> a
+deLift (Lift x) = x
+
+{-# DEPRECATED createProcess "better use instantiate :: Process a b -> a -> IO b instead" #-}
+createProcess :: (Trans a, Trans b) 
+		 => Process a b -> a -> Lift b
[…]

### 2017-06-24 benchmarks/fft/Parallel/EdiWP.hs

@@ -1,0 +1,114 @@
+{-# OPTIONS -cpp #-}
+module Parallel.EdiWP where 
+
+import Parallel.Edi
+import Parallel.ParPrim(connectToPort, sendData, Mode(Stream, Data))
+
+
+import Control.Parallel.Strategies
+import Control.Concurrent
+import Data.List
+import System.IO.Unsafe
+
+
+data Task t = Done --  implies worker termination 
+                   --  (task list ends with np times "Done")
+	    | T Int t -- tasks are tagged (for sorting results)
+type Res r  = (Int,r) -- results keep the original task tag
+
+#define SORTING
+
+--------------------------------------------
+-- difference to workpool type in EdenSkel.lhs: 
+--   1.  Trans / NFData
+--   2.  Process (on lists) replaced by function (on single items)
+--   3.  return type IO-monadic
+
+
+ediWP :: (NFData t, NFData r) => 
+          Int -> -- no of workers
+          Int -> -- prefetch
+          (t -> r) -> 
+          [t] -> 
+          IO [r]
+ediWP np prefetch workerF input 
+--    | np * prefetch > length input = ediWP np (prefetch - 1) workerF input
+--    | otherwise 
+          = do
+             (wInCCs, wInCs) <- createCs np
+             (wOutCs, wOuts) <- createCs np
+             sequence_ [ spawnProcessAt 0 (worker workerF wOutC wInCC)
+                         | (wOutC,wInCC) <- zip wOutCs wInCCs ]
+             -- input is written to a channel
+             taskChan <- newChan
+             fork $ writeList2Chan taskChan 
+                        ((zipWith T [1..] input) ++ (replicate np Done))
+             -- input senders supply tasks for workers
+             sequence_ [ fork (inputSender prefetch inC taskChan answers) 
+                         | (inC,answers) <- zip wInCs wOuts ]
+#ifdef SORTING
+             -- -- sort results
+             -- outs <- nMergeIO wOuts
+             -- return (sortByTag outs)
+             -- OR ELSE: use that all wOuts are ascending:
+             return (map snd (mergewOutsByTag wOuts))
+#else
+             -- results simply merged, unsorted
+             outs <- nmergeIO wOuts
+             return (map snd outs)
+#endif
+
+ediWPf :: (NFData t, NFData r) => 
[…]

### 2017-06-24 benchmarks/fft/Parallel/Skel/DivConN.hs

@@ -1,0 +1,71 @@
+-- divide and conquer skeleton for regular n-ary tree
+-- 2008, Jost Berthold, 
+-- Philipps-Universitaet Marburg, AG Eden
+--------------------------------------------------------
+module Parallel.Skel.DivConN where
+
+import Parallel.Eden
+
+import System.IO.Unsafe
+import Control.Monad
+import Control.Parallel.Strategies(seqList,Strategy)
+
+
+{-
+parMapEden :: (Trans a, Trans b) => (a -> b) -> [a] -> [b]
+parMapEden f xs = map deLift ([ createProcess (process f) x
+                               | x <- xs ] `using` whnfspine)
+-- demand control helper
+whnfspine :: Strategy [a]
+whnfspine [] = ()
+whnfspine (x:xs) = x `seq` whnfspine xs
+
+parMapEden2 :: (Trans a, Trans b) => (a -> b) -> [a] -> [b]
+parMapEden2 f xs = unsafePerformIO ( mapM (instantiate (process f)) xs )
+-}
+
+-- Eden parmap, placement:
+parMapAt :: (Trans a, Trans b) => [Int] -> (a -> b) -> [a] -> [b]
+parMapAt places f xs
+    = unsafePerformIO (
+         zipWithM (\pe x -> instantiateAt pe (process f) x)
+                  places xs
+        )
+
+
+-- distributed expansion (tree-shaped process creation, one tasks stays local)
+dcN :: (Trans a, Trans b) => 
+       Int -> Int                 -- n (expect n children) / remaining depth
+       -> (a -> Bool) -> (a -> b) -- trivial? / solve
+       -> (a -> [a]) -> ([b] -> b) -- split / combine
+       ->  a -> b                 -- resulting mapping
+dcN n depth trivial solve split combine x 
+   = if depth < 1 then seqDC x
+     else if trivial x then solve x 
+                       else childRs `seq` -- early demand on children list
+                            combine (myR : childRs)
+	    where myself = dcN n (depth - 1) trivial solve split combine
+	          seqDC x     = if trivial x then solve x 
+			                     else combine (map seqDC (split x))
+	          (mine:rest) = split x
+		  myR = myself mine
+		  childRs = parMapAt places myself rest
+			      `using` seqList r0 -- ???
+		  -- placement with stride for next children, round-robin
+		  places = map ((+1) . (`mod` noPe) . (+(-1))) shifts
[…]

### 2017-06-24 benchmarks/fft/Parallel/Strategies.hs

@@ -1,0 +1,602 @@
+{-# OPTIONS -cpp  #-}
+-----------------------------------------------------------------------------
+-- |
+-- Module      :  Control.Parallel.Strategies
+-- Copyright   :  (c) The University of Glasgow 2001
+-- License     :  BSD-style (see the file libraries/base/LICENSE)
+-- 
+-- Maintainer  :  libraries@haskell.org
+-- Stability   :  experimental
+-- Portability :  non-portable
+--
+-- Parallel strategy combinators. See
+-- <http://www.macs.hw.ac.uk/~dsg/gph/papers/html/Strategies/strategies.html>
+-- for more information.
+--
+-- Original authors:
+--	Phil Trinder, Hans-Wolfgang Loidl, Kevin Hammond et al. 
+--
+-----------------------------------------------------------------------------
+-- #ifdef __PARALLEL_HASKELL__
+-- module Control.Parallel.Strategies (
+-- #else
+module Parallel.Strategies (
+-- #endif
+   -- *	Strategy Type, Application and Semantics
+   Done, Strategy,
+   (>|), (>||),
+   using, demanding, sparking,
+   -- *	Basic Strategies				     
+   r0, rwhnf, NFData(..),
+   -- * Strategic Function Application
+   ($|), ($||),
+   (.|), (.||),
+   (-|), (-||),
+   -- * Tuples
+   seqPair, parPair,
+   seqTriple, parTriple,
+   -- * Lists: Parallel Strategies
+   parList, parListN, parListNth, parListChunk, 
+   parMap, parFlatMap, parZipWith,
+   -- * Lists: Sequential Strategies
+   seqList, seqListN, seqListNth, parBuffer,
+   -- *	Arrays
+   seqArr, parArr,
+   -- * Deprecated types and functions
+   sPar, sSeq,
+   Assoc(..),
+   fstPairFstList, force, sforce
+  ) where
+
+-- based on hslibs/concurrent/Strategies.lhs; see it for more detailed
+-- code comments. 
+#ifdef __PARALLEL_HASKELL__
+#else
+#warning "Using nonparallel GHC. The following line will fail."
+#endif
+import Control.Parallel as Parallel (pseq)
+import Data.Array
+import Data.Complex
+import Data.Int
+import qualified Data.IntMap (IntMap, toList)
+import qualified Data.IntSet (IntSet, toList)
+import qualified Data.Map (Map, toList)
[…]

### 2017-06-24 benchmarks/jacobi/Ringerweiterung.original.hs

@@ -1,0 +1,439 @@
+module Ringerweiterung where 
+-- Autor: Kent Kwee
+
+-- Realisierung des Polynomrings als Integerliste
+-- a + bx +cx^2 + .... < -- > [a,b,c,...] 
+type Poly = [Integer]
+
+-- Hilfsfunktionen fuer Listen
+
+-- | splittet eine Liste [a] in Listen der Laenge n
+splitlist :: [a]      -- ^ l: zu splittende Liste
+        -> Integer  -- ^ n: Laenge der Teillisten
+        -> [[a]]    -- ^ Liste der Listen der Laenge n
+splitlist [] _ = []
+splitlist l n = anfang:(splitlist rest n) 
+           where (anfang,rest) = zerlegebei l n   
+
+-- | splitat fuer Integer
+zerlegebei :: [a]       -- ^ l: zu splittende Liste
+         -> Integer   -- ^ n: Position der Zerlegung
+         -> ([a],[a]) -- ^ Anfang und Rest als Paar
+zerlegebei as 0 = ([],as)
+zerlegebei [] _ = ([],[])
+zerlegebei (a:as) n = (a:anfang,rest)
+                   where (anfang,rest) = zerlegebei as (n-1)
+
+-- | Addiert zwei Polynome
+plus :: Poly -- ^ erster Summand
+   -> Poly -- ^ zweiter Summand
+   -> Poly -- ^ Summe
+plus [] b = b
+plus a [] = a
+plus (a:as) (b:bs) = (a+b):(plus as bs)
+
+
+-- | Multipliziert zwei Polynome
+mal  :: Poly -- ^ erster Faktor
+   -> Poly -- ^ zweiter Faktor
+   -> Poly -- ^ Produkt
+mal [] b = []
+mal a [] = []
+mal (a:as) (b:bs) = (a*b):(plus (plus (mal [a] bs) (mal [b] as)) (0:(mal as bs)))
+
+-- | Bildet die n-te Potenz eine Polynoms
+polypot :: Poly     -- ^ Basis
+      -> Integer  -- ^ Exponent
+      -> Poly     -- ^ Potenz
+polypot polynom 0 = [1]
+polypot polynom 1 = polynom
+polypot polynom n = mal polynom (polypot polynom (n-1))  
+
+-- | Bildet die n-te Potenz eine Polynoms, das ein Element in Z[p^k-te Einheitswurzel] repraesentiert
+polypotsim :: Poly     -- ^ Basis
+         -> Integer  -- ^ Exponent
[…]

### 2017-06-24 benchmarks/rm/ModRabinMiller.hs

@@ -1,0 +1,156 @@
+{-# OPTIONS -cpp #-}
+module ModRabinMiller where 
+-- Original Autor: Kent Kwee
+-- Modifications by Oleg Lobachev
+
+import System.Random
+
+import System.IO.Unsafe (unsafePerformIO) -- HAAACK!
+
+import MathObj.Residue.IntMulti
+import MathObj.Residue.Modulo
+import IntMultiLift
+import Prelude hiding (div, mod, divMod)
+
+-- type MyInt = Integer -- so war das
+type MyInt = (Mod Int) -- a single residue. Now we can map!
+
+-- Hilfsfunktionen
+	   
+-- | Zerlege n in gerade und ungerade Anteile
+-- zerlege               :: (MyInt,MyInt)   -- ^ Eingabe (n,0)
+--                      -> (MyInt,MyInt)  -- ^ Ausgabe (q,t) mit n=2^t*q
+zerlege' :: (Int, Int) -> (Int, Int)
+zerlege' (q, t) 
+  | (r == 0) = zerlege' (b, t+1)
+  | otherwise        = (q,t)
+  where (b, r) = q `divMod` (2 :: Int)
+
+zerlege :: (MyInt, MyInt) -> (MyInt, MyInt)
+zerlege ((Z q qm), (Z t tm))
+    | qm /= tm = error "zerlege: different residues!"
+    | otherwise = (makeZ qr qm, makeZ tr tm)
+    where (qr, tr) = zerlege' (q, t)
+
+-- -- | Modulares Potenzieren: b^e mod m
+-- -- | TODO: A faster powering algorithm
+-- powermod             :: Integer    -- ^ b ist die Basis
+--                     -> Integer    -- ^ e ist der Exponent
+--                     -> Integer    -- ^ m der Modul
+--                     -> Integer    -- ^ Das Ergebnis b^e mod m
+-- powermod b e m 
+--   | (e==0)           = 1
+--   | (e `mod` 2 == 0) = (temp * temp) `mod` m 
+--   | otherwise        = b * (powermod b (e-1) m) `mod` m 
+--   where temp = (powermod b (e `div` 2) m) 
+
+powermod'             :: Int    -- ^ b ist die Basis
+                      -> Int    -- ^ e ist der Exponent
+                      -> Int    -- ^ m der Modul
+                      -> Int    -- ^ Das Ergebnis b^e mod m
+powermod' b e m 
+  | (e==0)           = 1
+  | (e `mod` 2 == 0) = (temp * temp) `mod` m 
[…]

### 2017-06-24 benchmarks/rm/OptRabinMiller.hs

@@ -1,0 +1,127 @@
+module OptRabinMiller where 
+-- Autor: Kent Kwee
+-- Optimised by Oleg Lobachev, 2011
+
+import System.Random
+
+import System.IO.Unsafe (unsafePerformIO) -- HAAACK!
+
+-- Hilfsfunktionen
+	   
+-- | Zerlege n in gerade und ungerade Anteile
+zerlege               ::(Integer,Integer)   -- ^ Eingabe (n,0)
+                     -> (Integer,Integer)  -- ^ Ausgabe (q,t) mit n=2^t*q
+{-
+zerlege (q,t) 
+  | (q `mod` 2 == 0) = zerlege(q `div` 2,t+1)
+  | otherwise        = (q,t)
+-}
+
+zerlege (q,t) 
+  | (r == 0) = zerlege (k, t+1)
+  | otherwise        = (q, t)
+  where (k, r) = divMod q 2
+
+zerlege1 (q,t) 
+  | (r == 0) = zerlege1(k, t+1)
+  | otherwise        = (q, t)
+  where (k, r) = divMod q 2
+
+
+-- | Modulares Potenzieren: b^e mod m
+-- | TODO: A faster powering algorithm
+powermod             :: Integer    -- ^ b ist die Basis
+                    -> Integer    -- ^ e ist der Exponent
+                    -> Integer    -- ^ m der Modul
+                    -> Integer    -- ^ Das Ergebnis b^e mod m
+{-
+powermod b e m 
+  | (e==0)           = 1
+  | (e `mod` 2 == 0) = (temp * temp) `mod` m 
+  | otherwise        = b * (powermod b (e-1) m) `mod` m 
+  where temp = (powermod b (e `div` 2) m) 
+-}
+powermod b e m 
+  | e==0      = 1
+  | r==0      = (binrec * binrec) `mod` m 
+  | otherwise = b * (powermod b (e-1) m) `mod` m 
+  where binrec = powermod b d m
+        (d, r) = divMod e 2
+
+{-
+-- bloedsinn?
+powermod b e m 
+  | e==0      = 1
+  | r==0      = (binrec * binrec) `mod` m 
+  | p==0      = (trerec * trerec * trerec) `mod` m
+  | otherwise = b * (powermod b (e-1) m) `mod` m 
+  where binrec = powermod b d m
+        (d, r) = divMod e 2
+        trerec = powermod b t m
+        (t, p) = divMod e 3
+-}
+
+-- | Zufallszahlenliste fuer die Basiswahl
[…]

### 2017-06-24 benchmarks/rm/ParMap.hs

@@ -1,0 +1,12 @@
+module ParMap(parMapF) where
+import Parallel.Eden
+
+
+parMapF :: (Trans a, Trans b) => (a->b) -> [a] -> [b]
+parMapF f xs = length result `seq` result  where result = [(pf f) # x | x <- xs]
+
+pf :: (Trans a, Trans b) => (a->b) -> Process a b
+pf f = process (\x -> f x)
+
+
+-- tl__ = createProcessTL pid (1::Int)

### 2017-06-24 benchmarks/rm/RabinMiller.hs

@@ -1,0 +1,92 @@
+module RabinMiller where 
+-- Autor: Kent Kwee
+
+import System.Random
+
+import System.IO.Unsafe (unsafePerformIO) -- HAAACK!
+
+-- Hilfsfunktionen
+	   
+-- | Zerlege n in gerade und ungerade Anteile
+zerlege               ::(Integer,Integer)   -- ^ Eingabe (n,0)
+                     -> (Integer,Integer)  -- ^ Ausgabe (q,t) mit n=2^t*q
+zerlege (q,t) 
+  | (q `mod` 2 == 0) = zerlege(q `div` 2,t+1)
+  | otherwise        = (q,t)
+
+-- | Modulares Potenzieren: b^e mod m
+-- | TODO: A faster powering algorithm
+powermod             :: Integer    -- ^ b ist die Basis
+                    -> Integer    -- ^ e ist der Exponent
+                    -> Integer    -- ^ m der Modul
+                    -> Integer    -- ^ Das Ergebnis b^e mod m
+powermod b e m 
+  | (e==0)           = 1
+  | (e `mod` 2 == 0) = (temp * temp) `mod` m 
+  | otherwise        = b * (powermod b (e-1) m) `mod` m 
+  where temp = (powermod b (e `div` 2) m) 
+
+-- | Zufallszahlenliste fuer die Basiswahl
+randomBaseList :: Integer      -- ^ n ist der Primzahlkandidat
+              -> IO [Integer] -- ^ die zugehoerige Basisliste
+randomBaseList n = do
+ gen <- getStdGen
+ let rs = randomRs (2, n-1) gen -- Zufallszahlen zwischen 2 und n-1
+ return rs
+
+-- RabinMiller für feste Basis a 
+singleRabinMillerBool             :: Integer  -- ^ n ist der Primzahlkandidat
+                                 -> Integer  -- ^ t kommt von der Zerlegung n=2^t*q
+                                 -> Integer  -- ^ e ist der aktuelle Wert von a^((2^e)q)
+                                 -> Integer  -- ^ b = a^q
+                                 -> Bool     -- ^ True: Moegliche Primzahl, False: keine Primzahl
+singleRabinMillerBool n t e b 
+     | (b==1)     = True
+     | (b==n-1)   = True
+     | (e<=t-2)   = singleRabinMillerBool n t (e+1) (b*b `mod` n)  
+     | otherwise  = False 
+
+
+-- | Rabin-Miller
[…]

### 2017-06-24 benchmarks/rm/SeqMain.hs

@@ -1,0 +1,15 @@
+module Main where
+
+import RabinMiller
+import System (getArgs)
+
+main = do
+  args <- getArgs
+  let k :: Integer
+      k = read $ head args
+      n :: Integer
+      n = 2^k-1
+  print n
+  res <- rabinMillerIO n
+  print res
+ 

### 2017-06-24 benchmarks/rm/gradiente.hs

@@ -1,0 +1,120 @@
+module Main(main) where
+import System
+import Parallel.Eden
+import IterUntil
+import Parallel.Skel.EdenSkel
+import System.IO.Unsafe (unsafePerformIO)
+
+pon m x = unsafePerformIO (print m) `seq` x
+
+main
+  = do ~[thr,size] <- getArgs
+       let size' = read size 
+           th = read thr
+           m = mA size'
+           b = v size'
+           np | noPe > th = noPe-1
+              | otherwise = noPe
+           a = gc np m b
+       --pon m (pon b (pon (prMV m a) print a))
+       rnf a `seq` putStrLn "done"
+
+v size = take size (drop 27 randomList)
+mA size = listToListList size (take (size*size) randomList)
+listToListList c m 
+ | length m <= c = [m]
+ | otherwise = c1 : listToListList c resto
+  where (c1,resto) = splitAt c m
+  
+  
+randomList :: [Double]
+--randomList = [8,3,2,4]++[1..23]++[19,8]
+randomList = map fromInteger (randomList' 30183 357 34713)
+randomList' x p m = x: randomList' (mod (x*p) m) p m
+
+------------------------
+type Matrix = [Vector]
+type Vector = [Double]
+
+-- instance (NFData a,NFData b) => NFData (Either a b) where
+--  rnf (Left v)  = rnf v
+-- rnf (Right v) = rnf v
+instance (Trans a,Trans b) => Trans (Either a b) 
+
+type Task = Either Vector Vector
+type SubResult = Either Vector Vector      
+type Local = (Matrix,Vector)
+type LocalM = (Vector,Vector,Vector,Double,Int)
+
+gc :: Int -> Matrix -> Vector -> Vector
+gc np a b = gc' np a b n0s b (map opp b)
+ where n0s = replicate (length b) 0
+       opp x = -x
+
+gc' np a b x d g = iterUntil (d,g,x,prVV g g,n) abs () split f_it comb
+ where as = splitIntoN np a
+       bs = splitIntoN np b
+--       gs = splitIntoN np g
+--       ds = splitIntoN np d
+--       xs = splitIntoN np x
+       n   = length b
+       abs = zip as bs
+--       abgdxs = zip6 as bs gs ds xs (repeat n)
+      
+       split :: () -> [Task] 
+       split () = replicate np (Left d)
+       
[…]

### 2017-06-24 paper/content/basicSkeletons.tex
added dots at the end of figure names.

@@
 	\includegraphics[scale=0.7]{images/parEvalNLazy}
-	\caption{Schematic depiction of parEvalNLazy}
+	\caption{Schematic depiction of |parEvalNLazy|.}
 	\label{fig:parEvalNLazyImg}
@@
 (+)) [1..]| or just because we need the arrows evaluated in chunks.
-|parEvalNLazy| (Fig.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
+|parEvalNLazy| (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
 It works by first chunking the input from |[a]| to |[[a]]| with the given |ChunkSize| in |arr (chunksOf chunkSize)|.
@@
 \end{code} %$ %% formatting
-\caption{Definition of |parEvalNLazy|}
+\caption{Definition of |parEvalNLazy|.}
 \label{fig:parEvalNLazy}
@@
 	\includegraphics[scale=0.7]{images/parEval2}
-	\caption{Schematic depiction of |parEval2|}
+	\caption{Schematic depiction of |parEval2|.}
 	\label{fig:parEval2Img}
@@
 But sometimes we want to paralellize heterogenous types as well.
-However, we can implement such a |parEval2| combinator (Fig.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| typeclass.
+However, we can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| typeclass.
 The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into parEvalN.

### 2017-06-24 paper/content/relwork.tex

@@
 \subsection{Parallel Haskells}
-Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1998a,Trinder1999} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \code{Par} monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
-We use these languages as backends: our library can switch from one to other at user's command.
+Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
+We use these languages as backends: our DSL can switch from one to other at user's command.
 
-HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \code{Par} monad to heterogeneous clusters.
-LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \code{Par} monad.
+HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of |Par| monad to heterogeneous clusters.
+LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of |Par| monad.
 %
-Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015} and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}.
+Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015}, and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}.
 Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden.
-This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,brown2010ever,dieterle2010skeleton,delaEncina2011,Dieterle2013}.
+This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.
 
-More different approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582,}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
+More different approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
 %
@@
 Accelerate is completely orthogonal to our approach.
-\citeauthor{marlow2013parallel} authored a recent book \citeyear{marlow2013parallel} on parallel Haskells.
+\citeauthor{marlow2013parallel} authored a recent book in \citeyear{marlow2013parallel} on parallel Haskells.
 
@@
 Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
-The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,brown2010ever,lobachev-phd,Lobachev2012}, iteration skeletons \cite{Dieterle2013}.
+The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,lobachev-phd,Lobachev2012,janjic2013space}, iteration skeletons \cite{Dieterle2013}.
 The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel.
-\citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeleton to stable process networks.
+\citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeletons to stable process networks.
 
@@
 
-Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~\code{->}.
+Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~|->|.
 \citet{Hughes2005} is a tutorial on arrows.
@@
 Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}.
-But perhaps most prominent application of arrows is functional reactive programming \cite{Hudak2003}.\olcomment{cite more!}
+But perhaps most prominent application of arrows is arrow-based functional reactive programming, AFRP \cite{Hudak2003}.\olcomment{cite more!}
 
@@
 Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution.
-In a contrast, we capture a whole parallel computation as a single entity: our main instantiation function \code{parEvalN} makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
+In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
 \citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach.
 They basically use arrows to orchestrate several tasks in robotics.
-We propose a general interface for parallel programming, remaining completely in Haskell.
+We, however, propose a general interface for parallel programming, remaining completely in Haskell.
 
@@
 Although this work is centred on Haskell implementation of arrows, it is applicable to any functional programming language where parallel evaluation and arrows can be defined.
-Our experiments with our approach in Frege language (which is basically Haskell on the JVM) were quite successful, we were able to use typical Java libraries for parallelism.
+Experiments with our approach in Frege language \footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM) were quite successful, we were able to use typical Java libraries for parallelism \mbcomment{not really tested completely.
+basic parEvalN worked, we didn't test for different Java ExecutorServices, though...}.
 However, it is beyond the scope of this work.

### 2017-06-24 paper/content/topologySkeletons.tex

@@
 Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator |>>>|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
+Nonetheless, we can define a parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
 

### 2017-07-03 paper/content/arrows.tex
olegs remarks up until (excluding) section about futures
added dots at the end of figure names.
added olegl changes

@@
-\subsection{Arrows}
-\label{sec:arrows}
-\begin{figure}[h]
-	\includegraphics{images/arrow}
-	\caption{Schematic depiction of an arrow}
-\end{figure}
-Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
-An arrow |arr a b| represents  a computation that converts an input |a| to an output |b|.
-This is defined in the |Arrow| shown in Fig.~\ref{fig:ArrowDefinition}.
-
 \begin{figure}[t]
 \centering
-\parbox[c][16.5em]{0.49\linewidth}{%
-\begin{center}
-\vfill
 \begin{code}
@@
   first :: arr a b -> arr (a,c) (b,c)
+
+instance Arrow (->) where
+	arr f = f
+	f >>> g = g . f
+	first f = \(a, c) -> (f a, c)
+
+data Kleisli m a b = Kleisli { run :: a -> m b }
+
+instance Monad m => Arrow (Kleisli m) where
+	arr f = Kleisli (return . f)
+	f >>> g = Kleisli (\a -> f a >>= g)
+	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
 \end{code}
 \vfill
-\end{center}
-\caption{|Arrow| class definition}
+\caption{The definition of |Arrow| type class and its two most typical instances.}
 \label{fig:ArrowDefinition}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\parbox[c][17em]{0.49\linewidth}{%
+\vfill
+\centering
+	\includegraphics{images/arrow}
+\vfill
+%	\caption{Schematic depiction of an Arrow.}
+%\label{fig:arrow-sch}
 }
-\parbox[c][16.5em]{0.49\linewidth}{%
+\parbox[c][17em]{0.49\linewidth}{%
 \vfill
@@
 \vfill
-\caption{The schematic depiction of |Arrow| combinators}
+%\caption{Schematic depiction of |Arrow| combinators |arr|, |>>>| and |first|.}
+%\label{fig:arrows-viz}
 }
+\caption{Schematic depiction of Arrow (left) and its basic
+  combinators |arr|, |>>>| and |first| (right).}
+\label{fig:arrow-sch}
+\label{fig:arrows-viz}
 \end{figure}
+
+\subsection{Arrows}
+\label{sec:arrows}
+Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
+An Arrow |arr a b| represents  a computation that converts an input |a| to an output |b|.
+This is defined in the |Arrow| type class shown in Fig.~\ref{fig:ArrowDefinition}.
+%
 Its |arr| operation is used to lift an ordinary function to the specified arrow type, similarly to the monadic |return|.
@@
 It allows us to to save input across arrows.
-\\\\
-The most prominent instances of this interface are regular functions |(->)|: %(Fig.~\ref{fig:arrowfn}),
-%\begin{figure}[h]
-\begin{code}
-instance Arrow (->) where
-	arr f = f
-	f >>> g = g . f
-	first f = \(a, c) -> (f a, c) 
-\end{code}
-% \caption{Arrow instance for regular functions}
-% \label{fig:arrowfn}
-% \end{figure}
-%%% on the verge of acceptable code
-and the Kleisli type: %(Fig.~\ref{fig:arrowkleisli}).
-% \begin{figure}[h]
-\begin{code}
-data Kleisli m a b = Kleisli { run :: a -> m b }
+Figure~\ref{fig:arrows-viz} shows a graphical representation of the basic Arrow combinators.
+The most prominent instances of this interface are regular functions |(->)|
+and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}).
 
-instance Monad m => Arrow (Kleisli m) where
-	arr f = Kleisli (return . f)
-	f >>> g = Kleisli (\a -> f a >>= g)
-	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
-\end{code}%$
-% \caption{Definition of the Kleisli type and the corresponding arrow instance}
-% \label{fig:arrowkleisli}
-% \end{figure}
 \begin{figure}[h]
@@
 	\end{tabular}
-	\caption{Visual depiction of syntactic sugar for arrows.}
+	\caption{Visual depiction of syntactic sugar for Arrows.}
 	\label{fig:syntacticSugarArrows}
@@
 With this typeclass in place, Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): The combinators |second|, |***| and |&&&|. 
-The definition of |second|, the mirrored version of |first|, is:
-% \begin{figure}[h]
-\begin{code}
-second :: Arrow arr => arr a b -> arr (c, a) (c, b)
-second f = arr swap >>> first f >>> arr swap
-	where swap (x, y) = (y, x)
-\end{code}
-% \caption{The second combinator}
-% \label{fig:second}
-% \end{figure}
-The |***| combinator that combines |first| and |second| to handle two inputs in one arrow, is defined as
+The combinator |second| is the mirrored version of |first| (Appendix~\ref{utilfns}).
+The |***| function combines |first| and |second| to handle two inputs in one arrow, is defined as
 % \begin{figure}[h]
@@
 % \end{figure}
-, while the |&&&| combinator, that constructs an arrow which outputs two different values like |***|, but takes only one input, is:
+while the |&&&| combinator, that constructs an arrow which outputs two different values like |***|, but takes only one input, is:
 % \begin{figure}[h]
@@
 % \end{figure}
-A first short example given by Hughes on how to use arrows is addition with arrows:
+A~first short example given by Hughes on how to use arrows is addition with arrows:
 %|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
@@
 
-The more restrictive interface of arrows (a monad can be \emph{anything}, an arrow is a process of doing something, a \emph{computation}) allows for more elaborate composition and transformation combinators.
-One of the major problems in parallel computing is composition of parallel processes.
+The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}.
+One of the major problems in parallel computing is, however, composition of parallel processes.
 

### 2017-07-03 paper/content/basicSkeletons.tex
added olegl changes

@@
 \subsubsection{Lazy |parEvalN|}
-\begin{figure}[t]
+\begin{figure}[tb]
 	\includegraphics[scale=0.7]{images/parEvalNLazy}
@@
 \subsubsection{Heterogenous tasks}
-\begin{figure}[h]
+\begin{figure}[tb]
 	\includegraphics[scale=0.7]{images/parEval2}
@@
 However, we can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| typeclass.
-The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into parEvalN.
+The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.
 

### 2017-07-03 paper/content/conclusion.tex
added olegl changes

@@
 
-\olcomment{Remove websites from citations, put them into footnotes!}
+\olcomment{Remove websites from citations, put them into footnotes! --
+\done}
 
@@
 
-Our PArrows interface can be expanded to futher parallel Haskells.
-More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} a modern distributed Haskell that would benefit from our Arrows notation.
-Future-aware special versions of Arrow combinated can be extended and further improved.
+Our PArrows DSL can be expanded to futher parallel Haskells.
+More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363}, a modern distributed Haskell that would benefit from our Arrows notation.
+More Future-aware versions of Arrow combinators can be defined and existing can be further improved.
 We would look into more transparency of the DSL, it should basically infuse as little overhead as possible.
 
-Of course, definitions of further skeletons are viable and needed. We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
+We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
+Of course, we are working on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
+It would also be interesting to see a hybrid of PArrows and Accelerate.
 
-Accelerate~\cite{Chakravarty:2011:AHA:1926354.1926358} is not related to our approach. It would be interesting to see a hybrid of both APIs.
-
-\olcomment{replace API with DSL globally?}
+\olcomment{replace API with DSL globally? \done}

### 2017-07-03 paper/content/futures.tex
added olegl changes

@@
 \section{Futures} \label{futures}
-Consider the parallel arrow combinator in Fig.~\ref{fig:someCombinator}
-\begin{figure}[h]
+Consider a mock-up parallel arrow combinator:
 \begin{code}
@@
 \end{code}
-\caption{An example parallel Arrow combinator without Futures.}
-\label{fig:someCombinator}
-\end{figure}
+
 In a distributed environment, a resulting arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
@@
 
-While the example in Fig.~\ref{fig:someCombinator} could be rewritten into only one |parEvalN| call by directly wiring the arrows properly together, this example illustrates an important problem: When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
+While the above example could be rewritten into only one |parEvalN| call by directly wiring the arrows properly together, it illustrates an important problem.
+When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
 This can become a serious bottleneck %in heavy threaded applications.
@@
 \mbcomment{Depends... Are the communications easy to read in such an example?}\\
-\mbcomment{Keep the description for the different colours, or link to the EdenTV description in \ref{sec:edentv}}
+\mbcomment{Keep the description for the different colours, or link to
+  the EdenTV description in \ref{sec:edentv}}
+\olcomment{ok as is}
+\olcomment{use the fft example (when it works)?}
 \end{figure}
@@
 We should allow the nodes to communicate directly with each other. Eden already ships with "remote data" that enable this \cite{AlGo03a,Dieterle2010}.
-
 But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
@@
 \end{figure}
-Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as seen in Fig.~\ref{fig:RDFuture}.
-This is due to the same reason we had to introduce a wrapper for |Strategy a| in the Multicore Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
+Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows.
+%This is due to the same reason we had to introduce a wrapper for |Strategy a| in the Multicore Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
+Technical details are in Appendix, in Section~\ref{app:omitted}.
 
@@
 \fixme{explain MVars}
-\begin{figure}[h]
+\begin{figure}[tb]
 \begin{code}
 {-# NOINLINE putUnsafe #-}
+
 putUnsafe :: a -> MVar a
@@
 
-Furthermore, in order for these |Future| types to fit with the |ArrowParallel| instances we gave earlier, we have to give the necessary |NFData| and |Trans| instances, the latter are only needed in Eden.
-The |Trans| instance does not have any functions declared as the default implementation suffices here.
-Furthermore, because |MVar| already ships with a |NFData| instance, we only have to supply a simple delegating |NFData| instance for our |RemoteData| type, where |rd| simply unwraps |RD|:
-% \begin{figure}[h]
-\begin{code}
-instance NFData (RemoteData a) where
-    rnf = rnf . rd
-instance Trans (RemoteData a)
-\end{code}
-% \caption{NFData and Trans instances for the RemoteData type.
-The Trans instance does not have any functions declared as the default implementation suffices here.
-See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
-% \end{figure}
-
 In our communication example we can use this |Future| concept for direct communications between the nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
-\begin{figure}[t]
+\begin{figure}[tbh]
 \begin{code}
@@
 \end{code}
-\caption{The combinator from Fig.~\ref{fig:someCombinator} in parallel.}
+\caption{The mock-up combinator in parallel.}
 \label{fig:someCombinatorParallel}
 \end{figure}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed---similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.\olcomment{Fig.
 is not really clear.
-Do Figs with a lot of load?}
+Do Figs with a lot of load?
+--- fft?}
 \begin{figure}[ht]

### 2017-07-03 paper/content/mapSkeletons.tex
added olegl changes

--- 2017-06-02 paper/content/mapSkeletons.tex
@@
 \label{sec:map-skeletons}
-Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel tasks.
+Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations.
 
-\subsection{Parallel map}
+\paragraph{Parallel |map| and laziness.}
+The |parMap| skeleton (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
+We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to obtain an arrow |arr [a] [b]|.
+Just like |parEvalN|, |parMap| is 100\% strict.
+As |parMap|  is 100\% strict it has the same restrictions as |parEvalN| compared to |parEvalNLazy|.
+So it makes sense to also have a |parMapStream| (Figs.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
+The code is quite straightforward, we show it in Appendix.
 
-\begin{figure}[h]
-%parMap
-\includegraphics[scale=0.7]{images/parMap}
-\caption{Schematic depiction of |parMap|.}
-\label{fig:parMapImg}
-
-\begin{code}
-parMap :: (ArrowParallel arr a b conf) => conf -> (arr a b) -> (arr [a] [b])
-parMap conf f = parEvalN conf (repeat f)
-\end{code}
-\caption{Definition of parMap.}
-\label{fig:parMap}
-
-%parMapStream
-\includegraphics[scale=0.7]{images/parMapStream}
-\caption{Schematic depiction of parMapStream.}
-\label{fig:parMapStreamImg}
-
-\begin{code}
-parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
-	conf -> ChunkSize -> arr a b -> arr [a] [b]
-parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
-\end{code}
-\caption{Definition of |parMapStream|.}
-\label{fig:parMapStream}
-\end{figure}
-
-The |parMap| skeleton (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
-We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to get an arrow |arr [a] [b]|.
-Just like |parEvalN|, |parMap| is 100\% strict.
-
-\paragraph{Lazy parallel map}
-As |parMap| (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is 100\% strict it has the same restrictions as |parEvalN| compared to |parEvalNLazy|.
-So it makes sense to also have a |parMapStream| (Fig.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
-
-\subsection{Statically load-balancing parallel map}
-
-\begin{figure}[h]
+\begin{figure}[thb]
 %farm
@@
 \includegraphics[scale=0.7]{images/farmChunk}
-\caption{Schematic depiction of farmChunk.}
+\caption{Schematic depiction of |farmChunk|.}
 \label{fig:farmChunkImg}
@@
 
-A |parMap| (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper).
-This can be quite wasteful and a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers (if numCores is greater than the actual processor count, the fastest processor(s) to finish will get more tasks) seems useful.
-The definitions of helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
+\paragraph{Statically load-balancing parallel |map|.}
+Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper).
+This can be quite wasteful and a |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful.
+\olcomment{it is a \emph{static} load balancing}
+The definitions of the helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.
 
-\paragraph{Lazy statically load-balancing parallel map}
-Since a |farm| (Fig.~\ref{fig:farmImg},~\ref{fig:farm}) is basically just |parMap| with a different work distribution, it is, again, 100\% strict.
-So we can define |farmChunk| (Fig.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|.
+%\paragraph{Lazy statically load-balancing parallel map}
+Since a |farm|  is basically just |parMap| with a different work distribution, it is, again, 100\% strict.
+So we can define |farmChunk| (Figs.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|.
 It is basically the same definition as for |farm|, with |parEvalN| replaced with |parEvalNLazy|.

### 2017-07-03 paper/content/motivation.tex
added olegl changes

@@
 	\item \textbf{Arrow DSL benefits}:
-	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is just yet another arrow combinator.
-The resulting arrow can be used in the same way a potential serial version could be used.
+	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator.
+The resulting Arrow can be used in the same way a potential serial version could be used.
 This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
-We can just \enquote{plug} in parallel parts into sequential arrow-based programs without having to change anything.
+We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
 	\item \textbf{Abstraction}:
 	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic.
-This leaves us in the beautiful situation of being able to write our code against the interface the typeclass gives us without being bound to any parallel Haskell.
+This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell.
 So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
@@
 The remaining text is structures as follows.
-Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
+Section~\ref{sec:background} briefly introduces known parallel Haskell flavours (Sec.~\ref{sec:parEvalNIntro}) and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
 Section~\ref{sec:related-work} discusses related work.
@@
 Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication.
-Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons (parallel |map| with and without load balancing, |map-reduce|) in our newly defined dialect.
-More advanced ones are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|).
+Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons  in our newly defined dialect: parallel |map| with and without load balancing.
+More advanced skeletons are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|).
 Section~\ref{sec:benchmarks} shows the benchmark results.

### 2017-07-03 paper/content/parallelHaskells.tex
added olegl changes

@@
 There are already several ways to write parallel programs in Haskell.
-As we will base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
+As we base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
 
@@
 \end{figure}
-Before we go into detail on how we can use this idea of parallelism for parallel Arrows, as a short introduction to parallelism in Haskell we will now implement |parEvalN| with several different parallel Haskells.
+%Before we go into detail on how we can use this idea of parallelism
+%for parallel Arrows, as a short introduction to parallelism in
+%Haskell we will now implement |parEvalN| with several different
+%parallel Haskells.
+As a demonstration, we implement here the non-Arrows |parEvalN| in multiple parallel Haskells.
 
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1998a} is way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|.
+Multicore Haskell \cite{Marlow2009,Trinder1998a} is a way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|.
+Let:
 
-For |parEvalN| this means that we can just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
+\begin{code}
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = let bs = zipWith ($) fs as 
+                 in (bs `using` parList rdeepseq) `pseq` bs
+\end{code} %$
+
+In the above definition of |parEvalN| we just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
 We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator.
@@
 This is particularly necessary in circular communication topologies like in the |torus| or |ring| (Chap.~\ref{sec:topology-skeletons}), where a wrong execution order would result in deadlock scenarios when executed without |pseq|.
+\olcomment{CAREFUL! is |pseq| REALLY about multiple instances or \emph{merely} order of execution?}
+%
+Fig.~\ref{fig:parEvalNMulticoreImg} shows a visual representation of this code.
 
-The resulting code and a graphical representation can be found in Fig.~\ref{fig:parEvalNMulticore} and Fig.~\ref{fig:parEvalNMulticoreImg}, respectively.
+% \begin{figure}[h]
+% \begin{code}
+% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+% parEvalN fs as = let bs = zipWith ($) fs as 
+%                  in (bs `using` parList rdeepseq) `pseq` bs
+% \end{code}
+% \caption{Multicore version of |parEvalN|.}
+% \label{fig:parEvalNMulticore}
+% \end{figure}
 
-\begin{figure}[h]
-\begin{code}
-parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = let bs = zipWith ($) fs as 
-                 in (bs `using` parList rdeepseq) `pseq` bs
-\end{code}
-\caption{Multicore version of |parEvalN|.}
-\label{fig:parEvalNMulticore}
-\end{figure}
-
-\begin{figure}[h]
+\begin{figure}[t]
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
@@
 The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a monad designed for composition of parallel programs.
+Let:
 
-The |Par| Monad version of our parallel evaluation function |parEvalN| can be defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell.
+\begin{code}
+parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+parEvalN fs as = runPar $ 
+	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+\end{code}
+
+The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell.
 % $
@@
 We execute this process with |runPar| to finally get the final |[b]|.
-
-Again, the resulting code and a graphical representation can be found in Fig.~\ref{fig:parEvalNParMonad} and Fig.~\ref{fig:parEvalNParMonadImg}, respectively.
+A graphical representation can be found Fig.~\ref{fig:parEvalNParMonadImg}.
 
@@
 
-\begin{figure}[h]
-\begin{code}
-parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
-parEvalN fs as = runPar $ 
-	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
-\end{code}
-\caption{|Par| Monad version of |parEvalN|.}
-\label{fig:parEvalNParMonad}
-\end{figure}
-\begin{figure}[h]
+% \begin{figure}[h]
+% \begin{code}
+% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
+% parEvalN fs as = runPar $ 
+% 	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+% \end{code}
+% \caption{|Par| Monad version of |parEvalN|.}
+% \label{fig:parEvalNParMonad}
+% \end{figure}
+\begin{figure}[t]
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
@@
 \end{code}
-\begin{figure}[h]
-	\includegraphics[scale=0.5]{images/parEvalNEden}
-	\caption{Dataflow of the Eden |parEvalN| version.}
-	\label{fig:parEvalNEden}
-\end{figure}
-A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
+% \begin{figure}[h]
+% 	\includegraphics[scale=0.5]{images/parEvalNEden}
+% 	\caption{Dataflow of the Eden |parEvalN| version.}
+% 	\label{fig:parEvalNEden}
+% \end{figure}
+% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
+\olcomment{removed the image as it shows nothing new}
 
@@
 While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}.
-In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.
+In the next sections we will present some \emph{trace visualizations}, the post-mortem process diagrams of Eden processes and their activity.
 
-In a trace, the $x$ axis shows the time, the $y$ axis enumerates the machines and processes.
-A~trace shows a running process in green, a blocked process is red.
+The trace visualizations are color-coded.
+In such a visualization (Fig.~\ref{fig:withoutFutures}), the $x$ axis shows the time, the $y$ axis enumerates the machines and processes.
+The visualization shows a running process in green, a blocked process is red.
 If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow.
-The typical reason for then is GC.
-An inactive machine where no processes are started yet, or all are already terminated, is shows as a blue bar.
+The typical reason for thus is GC.
+An inactive machine, where no processes are started yet, or all are already terminated, shows as a blue bar.
 A~comminication from one process to another is represented with a black arrow.
 A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.
-An example trace can be found in Fig.~\ref{fig:withoutFutures}.
 

### 2017-07-03 paper/content/parrows.tex
added olegl changes

@@
 \label{sec:parallel-arrows}
-We have seen what Arrows are and how they can be used as a general interface to computation.
-In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textit{parallel computation} as well.
-We start by introducing the interface and explaining the reasonings behind it.
+Arrows are a general interface to computation.
+Here we introduce special Arrows as general interface to \textit{parallel computations}.
+First, we present the interface and explain the reasonings behind it.
 Then, we discuss some implementations using exisiting parallel Haskells.
@@
 \subsection{The ArrowParallel typeclass}
-As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions |a -> b| in parallel, |parEvalN| (Chap.~\ref{sec:parEvalNIntro}).
+A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
 Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
@@
 We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|.
-So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
-% \begin{figure}[h]
-\olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
+% So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
+% % \begin{figure}[h]
+% \olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
 \begin{code}
@@
 % \end{figure}
-Note that we don't require the |conf| parameter in every implementation.
-If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|, as we will see in the implementation of the Multicore Haskell and the |Par| Monad backends.
+We do not require the |conf| parameter in every implementation.
+If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|.
+% , as we will see in the implementation of the Multicore Haskell and the |Par| Monad backends.
 
@@
 \begin{code}
+data Conf a = Conf (Strategy a)
+
 instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
@@
 \end{code}% $ %% formatting
-\caption{Fully evaluating ArrowParallel instance for the Multicore Haskell backend.}
+\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
 \label{fig:ArrowParallelMulticoreRdeepseq}
@@
 For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this.
-We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|:
+We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
 % \begin{figure}[h]
-\begin{code}
-data Conf a = Conf (Strategy a)
-\end{code}
 % \caption{Definition of Conf a}
@@
 % \end{figure}
-We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances.
+% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances.
 To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
-\begin{figure}[h]
+\begin{figure}[t]
 \begin{code}
@@
 \end{code}
-\caption{Configurable ArrowParallel instance for the Multicore Haskell backend.}
+\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
 \label{fig:ArrowParallelMulticoreConfigurable}
@@
 \subsubsection{|Par| Monad}
-\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting}
+\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"?}
 The |Par| monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| monad's |spawnP :: NFData a => a -> Par (IVar a)| function.
@@
 
-We therefore apply each function to its corresponding input value with || and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call.
+We therefore apply each function to its corresponding input value with and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call.
 This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|.
@@
 % \end{figure}
-We don't do it here for aesthetic resons, though. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
+% We don't do it here
+However, we avoid doing so for aesthetic resons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
 % \begin{figure}[h]

### 2017-07-03 paper/content/relwork.tex
added olegl changes

@@
 
-\subsection{Parallel Haskells}
+\paragraph{Parallel Haskells.}
 Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
@@
 
-\subsection{Algorithmic skeletons}
-
+\paragraph{Algorithmic skeletons.}
 Algorithmic skeletons were introduced by \citet{Cole1989}.
-Early efforts include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
+Early publications on this topic include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}.
+\citet{SkeletonBook} consolidated early reports on high-level programming approaches.
 The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,lobachev-phd,Lobachev2012,janjic2013space}, iteration skeletons \cite{Dieterle2013}.
@@
 
-\subsection{Arrows}
+\paragraph{Arrows.}
+Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~|->|.
+\citet{Hughes2005} presents a tutorial on Arrows.
+Some theoretical details on Arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable.
+\citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows.
+Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}.
+But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
 
-Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~|->|.
-\citet{Hughes2005} is a tutorial on arrows.
-Some theoretical details on arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable.
-\citet{Paterson:2001:NNA:507669.507664} introduced a new notation for arrows.
-Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}.
-But perhaps most prominent application of arrows is arrow-based functional reactive programming, AFRP \cite{Hudak2003}.\olcomment{cite more!}
+\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimizations.
+Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution.
+In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) Arrow out of list of Arrows.\olcomment{ugh, take care!}
+\citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach.
+They basically use Arrows to orchestrate several tasks in robotics.
+We, however, propose a general interface for parallel programming, while remaining completely in Haskell.
 
-\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of arrows that capsule the computation more than regular arrows do and thus enable optimizations.
-Their approach would allow parallel composition, as their special arrows would not interfere with each other in concurrent execution.
-In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) arrow out of list of arrows.\olcomment{ugh, take care!}
-\citet{Huang2007} utilise arrows for parallelism, but strikingly different from our approach.
-They basically use arrows to orchestrate several tasks in robotics.
-We, however, propose a general interface for parallel programming, remaining completely in Haskell.
-
-\subsection{Other languages}
-
+\paragraph{Other languages.}
 Although this work is centred on Haskell implementation of arrows, it is applicable to any functional programming language where parallel evaluation and arrows can be defined.
-Experiments with our approach in Frege language \footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM) were quite successful, we were able to use typical Java libraries for parallelism \mbcomment{not really tested completely.
-basic parEvalN worked, we didn't test for different Java ExecutorServices, though...}.
+Experiments with our approach in Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM) were quite successful.
 However, it is beyond the scope of this work.

### 2017-07-03 paper/content/topologySkeletons.tex
added olegl changes

--- 2017-06-02 paper/content/topologySkeletons.tex
@@
 However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of |pipe| in Fig.~\ref{fig:pipe}.
+Therefore, we introduce a more sophisticated version that internally uses Futures and obtain the final definition of |pipe| in Fig.~\ref{fig:pipe}.
 
 \begin{figure}[t]
-
 \begin{code}
@@
 \end{code}
-\caption{Final definition of the pipe skeleton which uses Futures}
+\caption{Final definition of the |pipe| skeleton with Futures.}
 \label{fig:pipe}
@@
 By wrapping these two arrows inside a common type we obtain |pipe2| (Fig.~\ref{fig:pipe2}).
-\begin{figure}[h]
+\begin{figure}[tb]
 \begin{code}
@@
 
-(|>>>|) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
+(parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
 	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
 	arr a b -> arr b c -> arr a c
-(|>>>|) = pipe2 ()
+(parcomp) = pipe2 ()
 \end{code}
-\caption{Definition of |pipe2| and parallel |>>>|.}
+\caption{Definition of |pipe2| and a parallel |>>>|.}
 \label{fig:pipe2}
@@
 Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
-Nonetheless, we can define a parallel piping operator |>>>|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
+Nonetheless, we can define a version of parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
+
+Another version of |>>>| is:
+\begin{code}
+f parparcomp g = (f . put) >>> (get . g)
+\end{code}
+It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data commuinication between them.
+Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.
 
 \subsection{Ring skeleton} \label{sec:ring}
-\begin{figure}[h]
+\begin{figure}[tb]
 	\includegraphics[scale=0.75]{images/ring}
@@
 Its input is a node function |i -> r -> (o, r)| in which |r| serves as the intermediary output that gets send to the neighbour of each node.
-This data is sent over direct communication channels, so called \enquote{remote data}.
+This data is sent over direct communication channels, the so called \enquote{remote data}.
 We depict it in Appendix, Fig.~\ref{fig:ringEden}.
@@
 In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)| (Fig.~\ref{fig:edenlazyrightrotate}).
-Similarly to the |pipe| from Chapter~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012} as a demand problem.
+Similarly to the |pipe| from Section~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) arrow here, or the evaluation would fail to terminate.
+The reasoning is explained by \citet{Loogen2012} as a demand problem.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[tb]
 \begin{code}
@@
 \end{figure}
-If we take the concept of a ring from \ref{sec:ring} one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
+If we take the concept of a ring from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
@@
 
-This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed.
+This action is, however,  more complicated than in the ring case as we have one more dimension of inputs to be transformed.
 We first have to |shuffle| all the inputs to then pass it into |parMap conf (ptorus f)| which yields |[(d, fut a, fut b)]|.
@@
 \end{figure}
-As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm \citep{Gentleman1978}.
+As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
 Their instantiation of the skeleton |nodefunction| can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
-\begin{figure}[h]
+\begin{figure}[tb]
 \begin{code}
@@
 \end{figure}
-If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version is comparable.\olcomment{much more details on this!}
-\begin{figure}[ht]
+If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version and execution times are comparable.
+We discuss further examples on larger clusters and in a more detail in the next section.
+\begin{figure}[tb]
 \olcomment{more nodes!!!}
@@
 
-\begin{figure}[ht]
+\begin{figure}[tb]
 	\centering

### 2017-07-03 paper/content/utilityFunctions.tex
removed removed old paper tex files
added olegl changes

@@
+\olcomment{more clear structure of the appendix? 1) Arrow utilities, 2) our things we omit for being too simple, 3) other's things we use.}
 \section{Utility Arrows}\label{utilfns}
-To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: |map|, |foldl| and |zipWith| on arrows.
+The 
+|second| combinator: 
+\begin{code}
+second :: Arrow arr => arr a b -> arr (c, a) (c, b)
+second f = arr swap >>> first f >>> arr swap
+	where swap (x, y) = (y, x)
+\end{code}
+is a mirrored version of |first|, used, \eg in
+definition of |***|.
 
+Utility combinators for Parallel Arrows follow. We define |map|, |foldl| and |zipWith| on Arrows.
 The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any arrow |arr a
@@
 
-These combinators make use of the |ArrowChoice| type class which provides the \pipepipepipe combinator.
-It takes two arrows |arr a c| and |arr b c| and combines them into a new arrow |arr (Either a b) c| which pipes all |Left a|'s to the first arrow and all |Right b|'s to the second arrow.
+These combinators make use of the |ArrowChoice| type class which provides the |pipepipepipe| combinator.
+It takes two arrows |arr a c| and |arr b c| and combines them into a new arrow |arr (Either a b) c| which pipes all |Left a|'s to the first arrow and all |Right b|'s to the second arrow:
 % \begin{figure}[htb]
 \begin{code}
-(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
+(pipepipepipe) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
 \end{code}
@@
 \section{Omitted Funtion Definitions}
-
+\label{app:omitted}
 We have omitted some function definitions in the main text for
@@
 We warp Eden's build-in Futures in PArrows as in
-Figure~\ref{fig:RDFuture}.
+Figure~\ref{fig:RDFuture}, where |rd| is the accessor function for the |RD| wrapped inside |RemoteData|.
+Furthermore, in order for these |Future| types to fit with the |ArrowParallel| instances we gave earlier, we have to give the necessary |NFData| and |Trans| instances, the latter are only needed in Eden.
+The |Trans| instance does not have any functions declared as the default implementation suffices here.
+Furthermore, because |MVar| already ships with a |NFData| instance, we only have to supply a simple delegating |NFData| instance for our |RemoteData| type, where |rd| simply unwraps |RD|.
+The |Trans| instance does not have any functions declared as the default implementation suffices:
+
+% \begin{figure}[h]
+\begin{code}
+instance NFData (RemoteData a) where
+    rnf = rnf . rd
+instance Trans (RemoteData a)
+\end{code}
+% \caption{NFData and Trans instances for the RemoteData type.
+The Trans instance does not have any functions declared as the default implementation suffices here.
+See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
+% \end{figure}
+
+\begin{figure}[thb]
+%parMap
+\includegraphics[scale=0.7]{images/parMap}
+\caption{Schematic depiction of |parMap|.}
+\label{fig:parMapImg}
+
+\begin{code}
+parMap :: (ArrowParallel arr a b conf) => conf -> (arr a b) -> (arr [a] [b])
+parMap conf f = parEvalN conf (repeat f)
+\end{code}
+\caption{Definition of parMap.}
+\label{fig:parMap}
+
+%parMapStream
+\includegraphics[scale=0.7]{images/parMapStream}
+\caption{Schematic depiction of |parMapStream|.}
+\label{fig:parMapStreamImg}
+
+\begin{code}
+parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
+	conf -> ChunkSize -> arr a b -> arr [a] [b]
+parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
+\end{code}
+\caption{Definition of |parMapStream|.}
+\label{fig:parMapStream}
+\end{figure}
+
+Figures~\ref{fig:parMapImg}--\ref{fig:parMapStream} show the definitions and a visualizations of two parallel |map| variants, defined using |parEvalN| and its lazy counterpart.
+
 Arrow versions of Eden's |shuffle|, |unshuffle| and the definition of |takeEach| are in Figure~\ref{fig:edenshuffleetc}.
-Similarly Figure~\ref{fig:edenlazyrightrotate} contains the definition of arrow versions of Eden's |lazy| and |rightRotate| utility functions.
+Similarly, Figure~\ref{fig:edenlazyrightrotate} contains the definition of arrow versions of Eden's |lazy| and |rightRotate| utility functions.
 Fig.~\ref{fig:lazyzip3etc} contains Eden's definition of |lazyzip3| together with the utility functions |uncurry3| and |threetotwo|.
 The full definition of |farmChunk| is in Figure~\ref{fig:farmChunk}.
-Eden definition of |ring| skeleton following \citep{Loogen2012} is in Figure~\ref{fig:ringEden}.
+Eden definition of |ring| skeleton is in Figure~\ref{fig:ringEden}. It
+follows \citet{Loogen2012}.
 
@@
 
-Furthermore, |parEval2| (Fig.~\ref{fig:parEval2}) is achieved as follows:
+The |parEval2| skeleton is defined in Figure~\ref{fig:parEval2}. 
 We start by transforming the |(a, c)| input into a two-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|.

### 2017-07-05 paper/content/arrows.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

@@
 Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
-Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
+Compared to simple Monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
 % \\\\
@@
 
-The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}.
+The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a Monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}.
 One of the major problems in parallel computing is, however, composition of parallel processes.

### 2017-07-05 paper/content/conclusion.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

@@
 Parallel Arrows feature an implementation of the |ArrowParallel| interface for Multicore Haskell, |Par| Monad, and Eden.
-With our approach parallel programs can be ported across these flavours with no effort.
+With our approach parallel programs can be ported across these flavours with little to no effort.
 %
 %
-Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}
+Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
 
-\mbcomment{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}
-
-\olcomment{Remove websites from citations, put them into footnotes! --
-\done}
-
-\olcomment{Parrows + accelerate = love? Metion port to Frege. Mention the Par monad troubles.}
+In order to fix the issues of the Par Monad backend with the |torus| and |ring| skeleton as explained in sections \ref{sec:ring} and \ref{sec:torus} - or any looping skeleton for that matter - additional work that fixes the strictness problems of the |ArrowParallel| instance is required.
+\olcomment{Parrows + accelerate = love? Metion port to Frege.}
 
@@
 It would also be interesting to see a hybrid of PArrows and Accelerate.
-
-\olcomment{replace API with DSL globally? \done}

### 2017-07-05 paper/content/mapSkeletons.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

@@
 Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper).
-This can be quite wasteful and a |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful.
-\olcomment{it is a \emph{static} load balancing}
+This can be quite wasteful and a statically load-balancing |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful.
 The definitions of the helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.

### 2017-07-05 paper/content/parallelHaskells.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

@@
 
-\subsubsection{ParMonad}
-The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a monad designed for composition of parallel programs.
+\subsubsection{Par Monad}
+The |Par| Monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs.
 Let:
@@
 Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|.
-We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| monad with |get|.
+We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| Monad with |get|.
 This results in |Par [b]|.
-We execute this process with |runPar| to finally get the final |[b]|.
-A graphical representation can be found Fig.~\ref{fig:parEvalNParMonadImg}.
-
-\mbcomment{explain problems with laziness here. Problems with torus}
+We execute this process with |runPar| to finally get |[b]|.
+A graphical representation can be found in Fig.~\ref{fig:parEvalNParMonadImg}.
 
@@
 
-While Eden also comes with a monad |PA| for parallel evaluation, it also ships with a completely functional interface that includes
+While Eden also comes with a Monad |PA| for parallel evaluation, it also ships with a completely functional interface that includes
 %\\
@@
 % A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
-\olcomment{removed the image as it shows nothing new}
 

### 2017-07-05 paper/content/parrows.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

@@
 \subsubsection{|Par| Monad}
-\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"?}
-The |Par| monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| monad's |spawnP :: NFData a => a -> Par (IVar a)| function.
-The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| monad.
+\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
+The |Par| Monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| Monad's |spawnP :: NFData a => a -> Par (IVar a)| function.
+The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| Monad.
 
@@
 This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|.
-In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| monad with |arr (>>= mapM get)|.
-The result of this operation is a |Par [b]| from which we can finally remove the monad again by running |arr runPar| to get our output of |[b]|.
+In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)|.
+The result of this operation is a |Par [b]| from which we can finally remove the Monad again by running |arr runPar| to get our output of |[b]|.
 \begin{figure}[h]

### 2017-07-05 paper/content/relwork.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

@@
 
-\olcomment{arrows or Arrows?}
-
 \paragraph{Parallel Haskells.}
-Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
+Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
 We use these languages as backends: our DSL can switch from one to other at user's command.
 
-HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of |Par| monad to heterogeneous clusters.
-LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of |Par| monad.
+HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of |Par| Monad to heterogeneous clusters.
+LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of |Par| Monad.
 %

### 2017-07-05 paper/content/topologySkeletons.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

@@
 \end{figure}
-\mbcomment{parallel >>>!!!!!!}
 
@@
 This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
+
+Note that the Par Monad version of the |ArrowParallel| version  presented in this paper (Fig.~\ref{fig:ArrowParallelParMonad}) does not work with the looping behaviour in the |ring| skeletons.
+Whenever |runPar| is invoked, strictness is enforced, which means that the |lazy| operation in the definition of |ring| will not work.
 %Further details on this can be found in \cite{eden_cefp}.
@@
 
-\subsection{Torus skeleton}
+\subsection{Torus skeleton}\label{sec:torus}
 \begin{figure}
@@
 
+Again, the implementation of the Par Monad version prevents this definition of the |torus| skeleton to work due to the reasons explained in the section about the |ring| skeleton in section \ref{sec:ring}.
+
 %\FloatBarrier

### 2017-07-06 paper/content/motivation.tex
mentioning Par Monad problems now. "Monad" instead of "monad"
minor

@@
 %\section{Motivation}
-%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes.
-In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive.
+%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to Monads for API design \citHughes.
+In the paper Hughes describes how arrows are a generalization of Monads and how they are not as restrictive.
 In this paper we will use this concept to express parallelism.
@@
 \olcomment{todo, reuse 5.5, and more}
+
+\mbcomment{
+Haskell is Spielwiese für Parallelität; verschiedene Ansätze (Par, Multicore, Eden); Orthogonale Ansätze; Verwenden höchstens eine Monade - manchmal auch nur intern; Wir wollen Parallelität mit Arrows abbilden, was noch niemand gemacht hat;
+Statt einer eigenen Implementierung definieren wir ein "shallow embedded DSL" (ACHTUNG, ist das der richtige Name? effektiv API);
+Umsetzung mit verschiedenen parallelen Haskells; We tame the zoo of parallel Haskells und vergewissern uns dass es nicht viel Overhead bringt
+}
 
@@
 The resulting Arrow can be used in the same way a potential serial version could be used.
-This is a big advantage of this approach, especially compared to the monad solutions as we do not introduce any new types.
+This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types.
 We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.

### 2017-07-06 paper/content/conclusion.tex
minor. added notes talk with oleg

@@
 
-In order to fix the issues of the Par Monad backend with the |torus| and |ring| skeleton as explained in sections \ref{sec:ring} and \ref{sec:torus} - or any looping skeleton for that matter - additional work that fixes the strictness problems of the |ArrowParallel| instance is required.
+The strictness problems of the |Par| Monad backend materializes with |ArrowLoop|. Additional work is required to ensure correct behaviour of the |Par| Monad in this context.
+
+\mbcomment{mention ArrowLoop in Torus and Ring chapters}
 \olcomment{Parrows + accelerate = love? Metion port to Frege.}
@@
 It would also be interesting to see a hybrid of PArrows and Accelerate.
+Ports of our approach to other languages like Frege or Java directly are in an early development stage.

### 2017-07-06 paper/content/relwork.tex
minor. added notes talk with oleg

@@
 But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
-
+\mbcomment{reference not found!!!}
 \citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimizations.
 Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution.
-In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) Arrow out of list of Arrows.\olcomment{ugh, take care!}
-\citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach.
+In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) Arrow out of list of Arrows.\citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach.
 They basically use Arrows to orchestrate several tasks in robotics.

### 2017-07-06 paper/content/topologySkeletons.tex
minor. added notes talk with oleg

@@
 %Further details on this can be found in \cite{eden_cefp}.
-\olcomment{let's do it?}
 
@@
 \end{figure}
-If we take the concept of a ring from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
+If we take the concept of a |ring| from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
 Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (PArrows).}
+	\caption[Matrix Multiplication with |torus| (Parrows)]{Matrix Multiplication with |torus| (PArrows).}
 	\label{fig:torus_parrows_trace}
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
-	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden).}
+	\caption[Matrix Multiplication with |torus| (Eden)]{Matrix Multiplication with |torus| (Eden).}
 	\label{fig:torus_eden_trace}

### 2017-07-08 paper/content/benchmarks (2).tex
wip on benchmark plotting

\documentclass[12pt,a4paper]{article}

\usepackage{microtype} %% looks better
\usepackage{pgfplots}

\usepackage{pgfplotstable} %% for regression

\begin{document}

\newcommand{\performanceplot}[5]{
\begin{tikzpicture}
\begin{axis}[title={Parallel performance of \\ #1},
title style={align=center},
xlabel=cores,
xtick={#4},
ylabel=time (seconds),
grid=major,
legend entries={#2},
legend pos=outer north east,
xmin=-1,
xmax=#3+1]
#5
\end{axis}
\end{tikzpicture}
}

\newcommand{\speedupplot}[5]{
\begin{tikzpicture}
\begin{axis}[title={Parallel speedup of \\ #1},
title style={align=center},
xlabel=cores,
xtick={#4},
ytick={#4},
ylabel=speedup,
grid=major,
legend entries={#2},
legend pos=outer north east,
ymin=-1,
xmin=-1,
ymax=#3 + 1,
xmax=#3 + 1]
\addplot [domain=0:#3]{x};
#5
\end{axis}
\end{tikzpicture}
}

\newcommand{\performanceSkelRMSM}[2]{
\performanceplot{SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}
}

\newcommand{\speedupSkelRMSM}[2]{
\speedupplot{SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}
}

\performanceSkelRMSM{4423}{32}

\speedupSkelRMSM{4423}{32}

\performanceSkelRMSM{4423}{64}

\speedupSkelRMSM{4423}{64}

\performanceSkelRMSM{9941}{32}

\speedupSkelRMSM{9941}{32}

\performanceSkelRMSM{9941}{64}

\speedupSkelRMSM{9941}{64}

\performanceSkelRMSM{11213}{32}

\speedupSkelRMSM{11213}{32}

\performanceSkelRMSM{11213}{64}

\speedupSkelRMSM{11213}{64}

\newcommand{\performanceSkelRMDist}[2]{
\performanceplot{SkelRM #1 #2}{Parrows, Eden}{256}{1,4,8,16,32,64,128,256}{
\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
}
}

\newcommand{\speedupSkelRMDist}[2]{
\speedupplot{SkelRM #1 #2}{Parrows, Eden}{256}{1,4,8,16,32,64,128,256}{
\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
}
}

\performanceSkelRMDist{9941}{256}

\speedupSkelRMDist{9941}{256}

\performanceSkelRMDist{11213}{256}

\speedupSkelRMDist{11213}{256}

\performanceSkelRMDist{44497}{256}

\speedupSkelRMDist{44497}{256}

\end{document}

### 2017-07-08 paper/content/benchmarks_tmp.tex
minor

--- 2017-07-08 paper/content/benchmarks (2).tex
@@
 \begin{tikzpicture}
-\begin{axis}[title={Parallel performance of \\ #1},
+\begin{axis}[title={#1},
 title style={align=center},
@@
 \begin{tikzpicture}
-\begin{axis}[title={Parallel speedup of \\ #1},
+\begin{axis}[title={#1},
 title style={align=center},
@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
+\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
 \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
+\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
 \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 
-\newcommand{\performanceSkelRMDist}[2]{
-\performanceplot{SkelRM #1 #2}{Parrows, Eden}{256}{1,4,8,16,32,64,128,256}{
+\newcommand{\performanceSkelRMDist}[3]{
+\performanceplot{Parallel performance of \\ SkelRM #1 #2}{PArrows, Eden}{256}{#3}{
 \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 
-\newcommand{\speedupSkelRMDist}[2]{
-\speedupplot{SkelRM #1 #2}{Parrows, Eden}{256}{1,4,8,16,32,64,128,256}{
+\newcommand{\performanceSkelRMDistDiff}[3]{
+\performanceplot{Parallel performance \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
+\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
+}
+}
+
+\newcommand{\speedupSkelRMDist}[3]{
+\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{PArrows, Eden}{256}{#3}{
 \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 
-\performanceSkelRMDist{9941}{256}
+\newcommand{\speedupSkelRMDistDiff}[3]{
+\performanceplot{Parallel speedup \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
+\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
+}
+}
 
-\speedupSkelRMDist{9941}{256}
+\performanceSkelRMDist{9941}{256}{16,32,64,128,256}
 
-\performanceSkelRMDist{11213}{256}
+\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256}
 
-\speedupSkelRMDist{11213}{256}
+\speedupSkelRMDist{9941}{256}{16,32,64,128,256}
 
-\performanceSkelRMDist{44497}{256}
+\speedupSkelRMDistDiff{9941}{256}{16,32,64,128,256}
 
-\speedupSkelRMDist{44497}{256}
+\performanceSkelRMDist{11213}{256}{16,32,64,128,256}
+
+\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256}
+
+\speedupSkelRMDist{11213}{256}{16,32,64,128,256}
+
+\speedupSkelRMDistDiff{11213}{256}{16,32,64,128,256}
+
+\performanceSkelRMDist{44497}{256}{32,64,128,256}
+
+\performanceSkelRMDistDiff{44497}{256}{32,64,128,256}
+
+\speedupSkelRMDist{44497}{256}{32,64,128,256}
+
+\speedupSkelRMDistDiff{44497}{256}{32,64,128,256}
 

### 2017-07-08 paper/content/benchmarks.tex

--- 2017-07-08 paper/content/benchmarks_tmp.tex
@@
-\documentclass[12pt,a4paper]{article}
-
-\usepackage{microtype} %% looks better
-\usepackage{pgfplots}
-
-\usepackage{pgfplotstable} %% for regression
-
-\begin{document}
+\section{Benchmarks}
+\label{sec:benchmarks}
 
@@
 title style={align=center},
-xlabel=cores,
+xlabel=threads,
 xtick={#4},
@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
+\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{1,4,8,16,32}{
 \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
+\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{1,4,8,16,32}{
 \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 \speedupSkelRMDistDiff{44497}{256}{32,64,128,256}
-
-\end{document}

### 2017-07-08 paper/content/benchmarks.tex
fixed legends in benchmarks

@@
 xmax=#3 + 1]
-\addplot [domain=0:#3]{x};
+\addplot [forget plot, domain=0:#3]{x};
 #5
@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{1,4,8,16,32}{
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{2,4,8,16,32}{
+\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{1,4,8,16,32}{
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{2,4,8,16,32}{
+\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};

### 2017-07-08 paper/content/benchmarks.tex
minor

@@
 
-\newcommand{\performanceplot}[5]{
+\newcommand{\performanceplot}[6]{
 \begin{tikzpicture}
@@
 legend entries={#2},
-legend pos=outer north east,
+legend style={at={(0.99,0.99)},anchor=north east},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
 xmin=-1,
-xmax=#3+1]
+xmax=#6]
 #5
@@
 
-\newcommand{\speedupplot}[5]{
+\newcommand{\performancediffplot}[7]{
+\begin{tikzpicture}
+\begin{axis}[title={#1},
+title style={align=center},
+xlabel=threads,
+xtick={#4},
+ytick distance=#6,
+minor tick num=9,
+ylabel=time (seconds),
+grid=both,
+legend entries={#2},
+legend style={at={(0.5,0.01)},anchor=south},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
+xmin=-1,
+xmax=#7]
+#5
+\end{axis}
+\end{tikzpicture}
+}
+
+\newcommand{\speedupplot}[6]{
 \begin{tikzpicture}
@@
 grid=major,
-legend entries={#2},
-legend pos=outer north east,
+legend entries={linear, #2},
+legend style={at={(0.01,0.99)},anchor=north west},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
 ymin=-1,
 xmin=-1,
-ymax=#3 + 1,
-xmax=#3 + 1]
-\addplot [forget plot, domain=0:#3]{x};
+ymax=#6,
+xmax=#6]
+\addplot [domain=0:#3, no markers,dotted,thick]{x};
+#5
+\end{axis}
+\end{tikzpicture}
+}
+
+\newcommand{\speedupdiffplot}[6]{
+\begin{tikzpicture}
+\begin{axis}[title={#1},
+title style={align=center},
+xlabel=threads,
+xtick={#4},
+ylabel=time (seconds),
+grid=major,
+legend entries={#2},
+legend style={at={(0.99,0.99)},anchor=north east},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
+xmin=-1,
+xmax=#6]
 #5
@@
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
-}
+}{33}
 }
@@
 
-\speedupSkelRMSM{9941}{32}
+\speedupSkelRMSM{9941}{32}{32}
 
@@
 
-\newcommand{\performanceSkelRMDist}[3]{
-\performanceplot{Parallel performance of \\ SkelRM #1 #2}{PArrows, Eden}{256}{#3}{
+\newcommand{\performanceSkelRMDist}[4]{
+\performanceplot{Parallel performance of \\ SkelRM #1 #2}{PArrows, Eden}{512}{#3}{
 \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-}
+}{#4}
 }
 
-\newcommand{\performanceSkelRMDistDiff}[3]{
-\performanceplot{Parallel performance \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\newcommand{\performanceSkelRMDistDiff}[5]{
+\performancediffplot{Parallel performance \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
+\addplot+[mark=*,thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
 smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-}
+}{#4}{#5}
 }
 
-\newcommand{\speedupSkelRMDist}[3]{
-\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{PArrows, Eden}{256}{#3}{
+\newcommand{\speedupSkelRMDist}[4]{
+\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{PArrows, Eden}{512}{#3}{
 \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-}
+}{#4}
 }
 
-\newcommand{\speedupSkelRMDistDiff}[3]{
-\performanceplot{Parallel speedup \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
+\newcommand{\speedupSkelRMDistDiff}[4]{
+\speedupdiffplot{Parallel speedup \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
 \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-}
+}{#4}
 }
 
-\performanceSkelRMDist{9941}{256}{16,32,64,128,256}
+\performanceSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
 
-\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256}
+\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{0.11}{528}
 
-\speedupSkelRMDist{9941}{256}{16,32,64,128,256}
+\speedupSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
 
-\speedupSkelRMDistDiff{9941}{256}{16,32,64,128,256}
+\speedupSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{528}
 
-\performanceSkelRMDist{11213}{256}{16,32,64,128,256}
+\performanceSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
 
-\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256}
+\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{0.1}{528}
 
-\speedupSkelRMDist{11213}{256}{16,32,64,128,256}
+\speedupSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
 
-\speedupSkelRMDistDiff{11213}{256}{16,32,64,128,256}
+\speedupSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{528}
 
-\performanceSkelRMDist{44497}{256}{32,64,128,256}
+\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
 
-\performanceSkelRMDistDiff{44497}{256}{32,64,128,256}
+\performanceSkelRMDistDiff{44497}{256}{32,64,128,256,512}{0.5}{544}
 
-\speedupSkelRMDist{44497}{256}{32,64,128,256}
+\speedupSkelRMDist{44497}{256}{32,64,128,256,512}{544}
 
-\speedupSkelRMDistDiff{44497}{256}{32,64,128,256}
+\speedupSkelRMDistDiff{44497}{256}{32,64,128,256,512}{544}

### 2017-07-08 paper/content/benchmarks_tmp.tex
minor

@@
 legend entries={#2},
-legend pos=outer north east,
+legend style={at={(0.99,0.99)},anchor=north east},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
+xmin=-1,
+xmax=#3+1]
+#5
+\end{axis}
+\end{tikzpicture}
+}
+
+\newcommand{\performancediffplot}[6]{
+\begin{tikzpicture}
+\begin{axis}[title={#1},
+title style={align=center},
+xlabel=threads,
+xtick={#4},
+ytick distance=#6,
+minor tick num=9,
+ylabel=time (seconds),
+grid=both,
+legend entries={#2},
+legend style={at={(0.93,0.01)},anchor=south east},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
 xmin=-1,
@@
 grid=major,
-legend entries={#2},
-legend pos=outer north east,
+legend entries={linear, #2},
+legend style={at={(0.01,0.99)},anchor=north west},
+max space between ticks=50pt,
+enlargelimits={abs=1 * #3/32},
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
 ymin=-1,
@@
 xmax=#3 + 1]
-\addplot [domain=0:#3]{x};
+\addplot [domain=0:#3, no markers,dotted,thick]{x};
+#5
+\end{axis}
+\end{tikzpicture}
+}
+
+\newcommand{\speedupdiffplot}[5]{
+\begin{tikzpicture}
+\begin{axis}[title={#1},
+title style={align=center},
+xlabel=threads,
+xtick={#4},
+ylabel=time (seconds),
+grid=major,
+legend entries={#2},
+legend style={at={(0.93,0.01)},anchor=south east},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
+xmin=-1,
+xmax=#3+1]
 #5
@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{2,4,8,16,32}{
+\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, ParMonad}{32}{1,4,8,16,32}{
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{2,4,8,16,32}{
+\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
@@
 
-\newcommand{\performanceSkelRMDistDiff}[3]{
-\performanceplot{Parallel performance \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\newcommand{\performanceSkelRMDistDiff}[4]{
+\performancediffplot{Parallel performance \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
+\addplot+[mark=*,thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
 smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-}
+}{#4}
 }
@@
 \newcommand{\speedupSkelRMDistDiff}[3]{
-\performanceplot{Parallel speedup \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
+\speedupdiffplot{Parallel speedup \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{256}{#3}{
 \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 
-\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256}
+\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256}{0.11}
 
@@
 
-\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256}
+\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256}{0.1}
 
@@
 
-\performanceSkelRMDistDiff{44497}{256}{32,64,128,256}
+\performanceSkelRMDistDiff{44497}{256}{32,64,128,256}{0.5}
 

### 2017-07-08 paper/content/benchmarks.tex
removed unnecessary benchmarks

@@
 legend entries={#2},
-legend style={at={(0.5,0.01)},anchor=south},
+legend style={at={(0.99,0.99)},anchor=north east},
 max space between ticks=50pt,
@@
 
-\performanceSkelRMSM{4423}{32}
+%\performanceSkelRMSM{4423}{32}
 
-\speedupSkelRMSM{4423}{32}
+%\speedupSkelRMSM{4423}{32}
 
-\performanceSkelRMSM{4423}{64}
+%\performanceSkelRMSM{4423}{64}
 
-\speedupSkelRMSM{4423}{64}
+%\speedupSkelRMSM{4423}{64}
 
-\performanceSkelRMSM{9941}{32}
+%\performanceSkelRMSM{9941}{32}
 
-\speedupSkelRMSM{9941}{32}{32}
+%\speedupSkelRMSM{9941}{32}{32}
 
-\performanceSkelRMSM{9941}{64}
+%\performanceSkelRMSM{9941}{64}
 
-\speedupSkelRMSM{9941}{64}
+%\speedupSkelRMSM{9941}{64}
 
@@
 
-\performanceSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
+%\performanceSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
 
-\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{0.11}{528}
+%\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{0.11}{528}
 
-\speedupSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
+%\speedupSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
 
-\speedupSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{528}
+%\speedupSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{528}
 
-\performanceSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
+%\performanceSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
 
-\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{0.1}{528}
+%\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{0.1}{528}
 
-\speedupSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
+%\speedupSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
 
-\speedupSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{528}
+%\speedupSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{528}
 

### 2017-07-08 paper/content/benchmarks.tex
minor

@@
 
-\performanceSkelRMSM{11213}{32}
+%\performanceSkelRMSM{11213}{32}
 
-\speedupSkelRMSM{11213}{32}
+%\speedupSkelRMSM{11213}{32}
 
@@
 \newcommand{\performanceSkelRMDistDiff}[5]{
-\performancediffplot{Parallel performance \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
+\performancediffplot{Parallel performance difference \\ of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
 \addplot+[mark=*,thick,error bars/.cd,
@@
 \newcommand{\speedupSkelRMDistDiff}[4]{
-\speedupdiffplot{Parallel speedup \\ difference of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
+\speedupdiffplot{Parallel speedup difference \\ of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
 \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,

### 2017-07-09 paper/content/futures.tex
removed notion of the MVar future from the paper

@@
 
-For our |Par| Monad and Multicore Haskell backends, we can simply use |MVar|s \cite{jones1996concurrent} (Fig.~\ref{fig:MVarFuture}), as in a shared memory setting we do not require Eden's sophisticated communication channels.
-\fixme{explain MVars}
+For our |Par| Monad and Multicore Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}) which are just simple wrappers around the actual data, as in a shared memory setting we do not require Eden's sophisticated communication channels.
 \begin{figure}[tb]
 \begin{code}
-{-# NOINLINE putUnsafe #-}
+data BasicFuture a = BF a
 
-putUnsafe :: a -> MVar a
-putUnsafe a = unsafePerformIO $ do
-    mVar <- newEmptyMVar
-    putMVar mVar a
-    return mVar
+instance (NFData a) => NFData (BasicFuture a) where
+    rnf (BF a) = rnf a
 
-instance (NFData a) => Future MVar a where
-    put = arr putUnsafe
-    get = arr takeMVar >>> arr unsafePerformIO
+instance (NFData a) => Future BasicFuture a where
+    put = arr BF
+    get = arr (\(BF a) -> a)
 \end{code}
-\caption{A |MVar| instance of the |Future| typeclass for the |Par| Monad and Multicore Haskell backends.}
-\label{fig:MVarFuture}
+\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and Multicore Haskell backends.}
+\label{fig:BasicFuture}
 \end{figure} % $

### 2017-07-11 paper/content/benchmarks.tex
removed notion of the MVar future from the paper
wip on the paper. troubles with Par monad are gone

@@
-\section{Benchmarks}
+%
+\section{Performance results}
 \label{sec:benchmarks}
 
-\newcommand{\performanceplot}[6]{
+\newcommand{\rmtest}{Rabin--Miller test\xspace}
+\newlength{\plotwidthSMP}
+\setlength{\plotwidthSMP}{0.39\textwidth}
+\newlength{\plotwidthDist}
+\setlength{\plotwidthDist}{0.6\textwidth}
+
+\newcommand{\performanceplot}[7]{
 \begin{tikzpicture}
@@
 title style={align=center},
-xlabel=threads,
-xtick={#4},
-ylabel=time (seconds),
+scale only axis, width=#7,
+xlabel=Cores,
+%xtick=data,
+xtick distance=#4,
+ylabel=Time (s),
+ylabel near ticks,
+minor tick num=2,
 grid=major,
@@
 
-\newcommand{\performancediffplot}[7]{
+\newcommand{\performancediffplot}[8]{
 \begin{tikzpicture}
@@
 title style={align=center},
-xlabel=threads,
-xtick={#4},
+scale only axis, width=#8,
+xlabel=Cores,
+%xtick=data,
 ytick distance=#6,
+xtick distance=#4,
 minor tick num=9,
-ylabel=time (seconds),
+ylabel=Absolute time difference (s),
+ylabel near ticks,
 grid=both,
@@
 
-\newcommand{\speedupplot}[6]{
+\newcommand{\speedupplot}[7]{
 \begin{tikzpicture}
@@
 title style={align=center},
-xlabel=threads,
-xtick={#4},
-ytick={#4},
-ylabel=speedup,
+scale only axis, width=#7,
+xlabel=Cores,
+%xtick=data,
+%ytick=data,
+xtick distance=#4,
+ytick distance=#4,
+ylabel=Speedup,
+ylabel near ticks,
 grid=major,
@@
 
-\newcommand{\speedupdiffplot}[6]{
+\newcommand{\speedupdiffplot}[7]{
 \begin{tikzpicture}
@@
 title style={align=center},
-xlabel=threads,
-xtick={#4},
-ylabel=time (seconds),
+scale only axis, width=#7,
+xlabel=Cores,
+%xtick=data,
+xtick distance=#4,
+ytick distance=0.5,
+ylabel=Absolute speedup difference,
+ylabel near ticks,
 grid=major,
@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel performance of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{2,4,8,16,32}{
-\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, Multicore, Par Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot+ table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
-}{33}
+}{17}{\plotwidthSMP}
 }
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{Eden CP, Multicore, Par Monad}{32}{2,4,8,16,32}{
-\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, Multicore, Par Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot+ table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
-}{33}
+}{17}{\plotwidthSMP}
 }
@@
 
-%\speedupSkelRMSM{11213}{32}
+\begin{figure}
+%\centering
+%\performanceSkelRMSM{11213}{64}\hfill%
+{\speedupSkelRMSM{11213}{32}}\hfill%
+{\speedupSkelRMSM{11213}{64}}
+\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
+  different backends on the same hardware. Measurements were performed on a single node of Glasgow
+  grid; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
+  tasks. The
+  closer to linear speedup the better.}
+\label{fig:bench-rm-sm}
+\end{figure}
 
-\performanceSkelRMSM{11213}{64}
+% \newcommand{\performanceSkelRMDist}[4]{
+% \performanceplot{Parallel run time \\ SkelRM #1 #2}{PArrows}{256}{#3}{
+% \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
+% % \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+% % smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
+% }{#4}
+% }
 
-\speedupSkelRMSM{11213}{64}
-
-\newcommand{\performanceSkelRMDist}[4]{
-\performanceplot{Parallel performance of \\ SkelRM #1 #2}{PArrows, Eden}{512}{#3}{
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+\newcommand{\speedupSkelRMDist}[4]{
+\speedupplot{Speedup of \rmtest ``#2''}{PArrows}{256}{#3}{
+% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
+\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
 smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-}{#4}
+% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
+}{#4}{\plotwidthDist}
 }
@@
 \newcommand{\performanceSkelRMDistDiff}[5]{
-\performancediffplot{Parallel performance difference \\ of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
-\addplot+[mark=*,thick,error bars/.cd,
+\performancediffplot{Run time differences\\for \rmtest ``#1, #2''}{(Eden $-$ PArrows)}{256}{#3}{
+\addplot+[mark=*,very thick,error bars/.cd,
     y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
 smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-}{#4}{#5}
+}{#4}{#5}{\plotwidthDist}
 }
 
-\newcommand{\speedupSkelRMDist}[4]{
-\speedupplot{Parallel speedup of \\ SkelRM #1 #2}{PArrows, Eden}{512}{#3}{
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-}{#4}
-}
-
-\newcommand{\speedupSkelRMDistDiff}[4]{
-\speedupdiffplot{Parallel speedup difference \\ of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
-\addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-}{#4}
-}
+% \newcommand{\speedupSkelRMDistDiff}[4]{
+% \speedupdiffplot{Parallel speedup difference \\ of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
+% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
+% }{#4}
+% }
 
@@
 
-\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupSkelRMDist{44497}{256}{32}{260}\label{subfig:rm-dist-a}}%
+%\hfill%
+{\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{260}\label{subfig:rm-dist-b}}
+\caption{Parallel performance of \rmtest on the Glasgow grid
+  consisting of 256 cores. Input was $2^{44497}-1$, we used 256
+  tasks. Top plot shows absolute speedup in a distributed memory setting. The
+  closer to linear speedup the better. Time
+  (and hence speedup) measurements for PArrows with Eden backend and
+  Eden almost coincide. Hence, bottom plot shows
+absolute time differences for this benchmark. The
+lower the value, the better for PArrows\olcomment{CHECKME}.}
+\label{fig:bench-rm-dist}
+\end{figure}
 
-\performanceSkelRMDistDiff{44497}{256}{32,64,128,256,512}{0.5}{544}
+%\speedupSkelRMDistDiff{44497}{256}{32,64,128,256,512}{544}
 
-\speedupSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Text
 
-\speedupSkelRMDistDiff{44497}{256}{32,64,128,256,512}{544}
+\subsection{Hardware}
+
+We have tested our parallel DSL and algorithmic skeletons implemented
+in it. Benchmarks were conducted both in a shared and in a distributed
+memory setting. The hardware was the ``Glasgow grid'', consisting of
+16 machines with 16 cores. \olcomment{insert typical blabla: consisting of
+  Intel XXXYYYY at XX~GHz with XXXX~GB RAM and FOOBAR Linux version
+  XXXX with kernel XXXXXXXX.}
+Each core had a hyperthreading capability,
+so the total number of reported processors per machine was
+32. Non-surprisingly, we found that hyperthreaded 32 cores do not
+behave in the same manner as real 16 cores (numbers here for a single
+machine). We disregarded the hyperthreading ability in most of the
+cases.
+
+We used a single machine with 16 real cores as a shared memory testbed
+and the whole grid with 256 real cores as a device to test our
+distributed memory software.
+
+\olcomment{only if we do it: list also our i7-6770K machine}
+
+\subsection{Test programs}
+
+We used multiple tests that originated from different
+sources. Most of them are parallel mathematical computations, initially
+implemented in Eden. Table~\ref{tab:benches} summarizes.
+
+\begin{table}
+\caption{The benchmarks we use in this paper.}
+\label{tab:benches}
+\centering
+%% something was wrong with separators in table
+\renewcommand{\tabcolsep}{1em}
+\begin{tabular}{lccll}
+\toprule
+Name & Area & Type & Origin & Citation \\
+\midrule
+\rmtest & Mathematics & |parMap+reduce| & Eden & \citet{Lobachev2012}\\
+Jacobi sum test & Mathematics & |workpool+reduce| & Eden & \citet{Lobachev2012}\\
+Gentleman & Mathematics & |torus| & Eden & \citet{Loogen2012}\\
+Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad}\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\rmtest is a probabilistic primality test that iterates multiple (32--256 here)
+``subtests''. Should a subtest fail, the input is definitely not a
+prime. If all $n$ subtest pass, the input is composite with the
+probability of $1/4^{n}$.
+
+Jacobi sum test or APRCL is also a primality test, that however,
+guarantees the correctness of the result. It is probabilistic in the
+sense that its run time is not certain. Unlike \rmtest, the subtests
+of Jacobi sum test have very different durations. \citet{lobachev-phd}
+discuss some optimisations of parallel APRCL. Generic parallel
+implementation of \rmtest and APRCL were presented in \citet{Lobachev2012}.
+
+``Gentleman'' is a standard Eden test program, developed
+for their |torus| skeleton. It implements a parallel matrix
+multiplication \citep{Gentleman1978}. We ported its latest Eden
+version \citep{Loogen2012} to PArrows.
+
+A~parallel Sudoku solver was used by \citet{par-monad} to compare |Par| Monad
+to Multicore Haskell. We ported it to PArrows.
+
+\subsection{What parallel Haskells run where}
+
+The \ensuremath{\Conid{Par}} monad and Multicore Haskell can be executed on a shared
+memory machines only. Although GpH is available on distributed memory
+clusters, and newer distributed memory Haskells such as HdpH exist,
+current support of distributed memory in PArrows is limited to
+Eden. We used the MPI backend of Eden in a distributed memory
+setting. However, for shared memory Eden features a ``CP'' backend
+that merely copies the memory blocks between distributed heaps. In
+this mode Eden still operates in the ``nothing shared'' setting, but
+is adapted better to multicore machines. We label this version of Eden
+in the plots as ``Eden~CP''.
+
+\subsection{Effect of hyperthreading}
+
+The PArrows version of \rmtest on a single node of the Glasgow grid
+showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
+of 64-task PArrows/Eden at 16 real cores version was 13.65, the
+efficiency was 85.3\%.  However, when we increase the number of
+requested cores to be 32---\ie when we use hyperthreading on 16 real
+cores---the speedup does not increase that well. It is merely 15.99
+for 32 tasks with PArrows/Eden. It is worse for other backends.  As
+for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
+hyperthreaded cores and only 13.55 with PArrows/Multicore
+Haskell. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
+version used here was Eden~CP, the ``share nothing'' SMP build.
+
+In the distributed memory setting same effect ensues. We obtain
+plummeting speedup of 124.31 at 512 hyperthreaded cores, whereas it was
+213.172 for 256 real cores. Apparently, hyperthreading in the Glasgow
+grid fails to execute two parallel Haskell processes with full-fledged
+parallelism. For this reason, we did not regard hyperthreaded cores in
+our speedup plots in Figs.~\ref{fig:bench-rm-sm}--XXXXXXXX.
+
+% rm 11213 32 32-sm speedup eden: 15.993037587283924
+% -"- multi: 15.09948017762912
+% -"- par: 14.909092857846693
+
+% -"- 64 32-sm speedup eden: 16.118040224478424
+% -"- multi: 13.545304115702333
+% -"- par: 15.155709987503396
+
+\subsection{Benchmark results}
+
+The difference between, say, PArrows with |Par| Monad backend and a
+genuine |Par|
+Monad benchmark is very small. To give an example, it is XXXXXXXX for
+XXXXXXXXX and XXXXXXXXXX in the shared memory setting. It is almost invisible in speedup and
+(non shown) run time plots. We thus show only the results for the
+PArrows-enabled versions.
+
+To showcase that the induced overhead of
+PArrows is small, we plot execution time differences between measurements for
+PArrows and the corresponding backend in a separate plot
+(Figs.~\ref{fig:bench-rm-dist}, XXXXXXX). The differences range in
+about 0.5~seconds for the execution time of 46~seconds on 256 cores
+for distributed \rmtest with PArrows and Eden. The plots show absolute
+time differences that are not relativated \wrt the total execution time.
+
+The error bars ends were computed from pointwise maximum of both standard
+deviations from both measurements for PArrows and non-PArrows
+versions. These are the values provided by |Criterion| package that we
+used for benchmarking. We call a difference between two versions
+significant when the border of the error bar of absolute time
+difference is above or below zero. In other words: the time
+difference is significant if it is above measurement error.
+
+\subsubsection{\rmtest}
+\olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
+  Why is it good?}
+The multicore version of our parallel \rmtest benchmark is in
+Figure~\ref{fig:bench-rm-sm}. We have executed the test with 32 and 64
+tasks. The plot shows the PArrows-enabled versions with corresponding backends.
+The performance of PArrows/Eden~CP in shared memory is slightly better than
+for SMP variants such as PArrows/Multicore Haskell and PArrows/|Par|
+Monad. The reason for this behavior lies in the backend in our reasoning. One
+possible explanation is garbage collection. It is easier to GC
+(independently) $n$ heaps than to GC one large. This agrees with known
+results\olcomment{WHICH EXACTLY?} that motivate the efforts for
+parallel GC in SMP-based Haskells.
+
+To show that PArrows induce very small overhead, we compare the ported
+versions of the benchmark to the PArrows-enabled ones with
+corresponding backend.\olcomment{THE RESULT?}
+
+In the distributed memory setting, we utilize PArrows/Eden. Eden in
+its turn used the MPI backend. We see an almost linear speedup of
+\rmtest with 256 tasks and input $2^444497-1$. The sequential run time
+was computed as mean of three consecutive executions on a single
+core---the single run took two hours 43 minutes. The zero difference  between
+PArrows/Eden and Eden almost always lies on the error bar of
+the measurement. The only exceptions, where PArrows version was slower
+and it was significant, were for 64 and
+512 cores, for 0.49 and 0.26~second, respectively. This corresponds to
+0.30\% and 0.33\% relative time difference, respectively. The 512 core
+version suffers, of course, from the aforementioned hyperthreading
+problem. The PArrows-induced overhead was merely 0.56\% for 256
+cores. The PArrows version was 0.30\% \emph{faster} for 128 cores.

### 2017-07-11 paper/content/topologySkeletons.tex
wip on the paper. troubles with Par monad are gone

@@
 We aim to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
+
+If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the Multicore and Par Monad backends which during runtime would result in the program crashing.
+This materializes with the usage of |loop| of the |ArrowLoop| typeclass and is probably due to the way their respective parallelism engines work internally.
+\mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| typeclass.
+This allows us for selective enabling and disabling of parallelism.
+\begin{code}
+class ArrowParallel arr a b conf => FutureEval arr a b conf where
+    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
+\end{code}
+As Eden has no problems with the looping skeletons, we declare a delegating instance:
+\begin{code}
+instance ArrowParallel arr a b conf => FutureEval arr a b conf where
+    evalN = parEvalN
+\end{code}
+The Par Monad and Multicore backends have parallelism disabled in their instance of |FutureEval|.
+This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
+\begin{code}
+instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
+    ArrowParallel arr a b conf) => FutureEval arr a b conf where
+    evalN _ = listApp
+\end{code}
 
@@
 defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it.
-Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |parEvalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
+Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
 
@@
 \begin{code}
-pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
+pipeSimple :: (ArrowLoop arr, FutureEval arr a a conf) =>
 	conf -> [arr a a] -> arr a a
@@
 	loop (arr snd &&&
-		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
+		(arr (uncurry (:) >>> lazy) >>> evalN conf fs)) >>>
 	arr last
@@
 Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt.
-We need to enforce that the evaluation of the input |[a]| terminates before passing it into |parEvalN|.}
+We need to enforce that the evaluation of the input |[a]| terminates before passing it into |evalN|.}
 \label{fig:pipeSimple}
@@
 \begin{code}
-pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
+pipe :: (ArrowLoop arr, FutureEval arr (fut a) (fut a) conf, Future fut a) =>
 	conf -> [arr a a] -> arr a a
@@
 pipe2 :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
-         ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
+         FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
 	conf -> arr a b -> arr b c -> arr a c
@@
 (parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
-	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
+	FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
 	arr a b -> arr b c -> arr a c
@@
 
-Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN.
+Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to |parEvalN| inside of |evalN|.
 Nonetheless, we can define a version of parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
@@
 This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
-
-Note that the Par Monad version of the |ArrowParallel| version  presented in this paper (Fig.~\ref{fig:ArrowParallelParMonad}) does not work with the looping behaviour in the |ring| skeletons.
-Whenever |runPar| is invoked, strictness is enforced, which means that the |lazy| operation in the definition of |ring| will not work.
 %Further details on this can be found in \cite{eden_cefp}.
@@
 \begin{code}
-ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
+ring :: (ArrowLoop arr, Future fut r, FutureEval arr (i, fut r) (o, fut r) conf) =>
     conf -> arr (i, r) (o, r) -> arr [i] [o]
@@
 	loop (second (rightRotate >>> lazy) >>> arr (uncurry zip) >>>
-        parMap conf (second get >>> f >>> second put) >>> arr unzip)
+        evalN conf (repeat (second get >>> f >>> second put)) >>> arr unzip)
 \end{code}
@@
 This action is, however,  more complicated than in the ring case as we have one more dimension of inputs to be transformed.
-We first have to |shuffle| all the inputs to then pass it into |parMap conf (ptorus f)| which yields |[(d, fut a, fut b)]|.
+We first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|.
 We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass.
@@
 torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr, Future fut a, Future fut b,
-          ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf) =>
+          FutureEval arr (c, fut a, fut b) (d, fut a, fut b) conf) =>
 	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
@@
 	arr (uncurry3 (zipWith3 lazyzip3)) >>>
-	(arr length >>> arr unshuffle) &&& (shuffle >>> parMap conf (ptorus f)) >>> app >>>
+	(arr length >>> arr unshuffle) &&& (shuffle >>> evalN conf (repeat (ptorus f)) >>> app >>>
 	arr (map unzip3) >>> arr unzip3 >>> threetotwo)
@@
 
-Again, the implementation of the Par Monad version prevents this definition of the |torus| skeleton to work due to the reasons explained in the section about the |ring| skeleton in section \ref{sec:ring}.
-
 %\FloatBarrier

### 2017-07-11 examples/sudoku/parrows-sudoku-parmap.hs

@@ -4,8 +4,7 @@ import Control.Exception
 import System.Environment
 import Data.Maybe
 import Parrows.Definition
-import Parrows.Skeletons.Map
-import Parrows.Multicore
+import Parrows.ParMonad
 import Control.DeepSeq
 
 main :: IO ()

### 2017-07-11 Definition/src/main/Parrows/Future.hs
minor

@@ -32,7 +32,7 @@ class Future fut a | a -> fut where
     put :: (Arrow arr) => arr a (fut a)
     get :: (Arrow arr) => arr (fut a) a
 
-class (ArrowParallel arr a b conf) => FutureEval arr a b conf where
+class ArrowParallel arr a b conf => FutureEval arr a b conf where
     evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
 
 liftFut :: (Arrow arr, Future fut a, Future fut b) => arr a b -> arr (fut a) (fut b)

### 2017-07-11 Definition/src/main/Parrows/Skeletons/Topology.hs

@@ -71,7 +71,7 @@ ring :: (ArrowLoop arr, FutureEval arr (i, fut r) (o, fut r) conf,
 ring conf f =
     loop (second (rightRotate >>> lazy) >>>
         arr (uncurry zip) >>>
-        evalN conf (repeat $ second get >>> f >>> second put) >>>
+        evalN conf (repeat (second get >>> f >>> second put)) >>>
         arr unzip)
 
 --TODO: check whether this exchanges the futures the same way as Eden does it

### 2017-07-11 Eden/src/main/Parrows/Eden.hs

@@ -42,7 +42,7 @@ instance (Trans a) => Future RemoteData a where
     put = arr (\a -> RD { rd = release a })
     get = arr rd >>> arr fetch
 
-instance (ArrowParallel arr a b conf) => FutureEval arr a b conf where
+instance ArrowParallel arr a b conf => FutureEval arr a b conf where
     evalN = parEvalN
 
 -- ArrowParallel Instances

### 2017-07-11 paper/content/parallelHaskells.tex
minor

@@
 parEvalN fs as = let bs = zipWith ($) fs as 
-                 in (bs `using` parList rdeepseq) `pseq` bs
+                 in bs `using` parList rdeepseq
 \end{code} %$
@@
 We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form.
-To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations.
-This is particularly necessary in circular communication topologies like in the |torus| or |ring| (Chap.~\ref{sec:topology-skeletons}), where a wrong execution order would result in deadlock scenarios when executed without |pseq|.
-\olcomment{CAREFUL! is |pseq| REALLY about multiple instances or \emph{merely} order of execution?}
 %
@@
 % parEvalN fs as = let bs = zipWith ($) fs as 
-%                  in (bs `using` parList rdeepseq) `pseq` bs
+%                  in (bs `using` parList rdeepseq)
 % \end{code}
@@
 We execute this process with |runPar| to finally get |[b]|.
-A graphical representation can be found in Fig.~\ref{fig:parEvalNParMonadImg}.
+Fig.~\ref{fig:parEvalNParMonadImg} shows a graphical representation.
 

### 2017-07-11 paper/content/parrows.tex

@@
 \subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
+The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
 \begin{figure}[t]
@@
        		listApp fs >>>
-        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
-        	arr (uncurry pseq)
+        	arr (withStrategy (parList rdeepseq))
 \end{code}% $ %% formatting

### 2017-07-21 paper/content/conclusion.tex
intro + conclusion: a first version

@@
 \label{sec:conclusion}
-Arrows are a generic concept that allows for powerful composition combinators.
-To our knowledge we are the first ones to represent parallel computation with arrows.\olcomment{that strange arrows-based robot interaction paper from 1993 or so!
-clearly discuss in related work} \done
-
-Arrows turn out to be a useful tool for composing in parallel programs.
-We do not have to introduce new monadic types that wrap the computation.
-Instead use arrows just like regular sequential pure functions.
+Arrows are a generic concept that allows for powerful composition
+combinators. To our knowledge we are the first ones to represent
+\emph{parallel} computation with Arrows.
+% \olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done
+%
+Arrows turn out to be a useful tool for composing in parallel
+programs. We do not have to introduce new monadic types that wrap the
+computation. Instead, we use Arrows in the same manner one uses sequential pure functions. 
 %
 This work features multiple parallel backends: the already available parallel Haskell flavours.
-Parallel Arrows feature an implementation of the |ArrowParallel| interface for Multicore Haskell, |Par| Monad, and Eden.
+Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| typeclass for Multicore Haskell, |Par| Monad, and Eden.
 With our approach parallel programs can be ported across these flavours with little to no effort.
+It is quite straightforward to add further backends.
 %
@@
 Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
-
-The strictness problems of the |Par| Monad backend materializes with |ArrowLoop|. Additional work is required to ensure correct behaviour of the |Par| Monad in this context.
+%
+The benefit is, however, the greatly increased portability of parallel
+programs.
 
 \mbcomment{mention ArrowLoop in Torus and Ring chapters}
-\olcomment{Parrows + accelerate = love? Metion port to Frege.}
+%\olcomment{Parrows + accelerate = love?} \olcomment{Metion port to Frege.}
 
@@
 
-Our PArrows DSL can be expanded to futher parallel Haskells.
-More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363}, a modern distributed Haskell that would benefit from our Arrows notation.
-More Future-aware versions of Arrow combinators can be defined and existing can be further improved.
-We would look into more transparency of the DSL, it should basically infuse as little overhead as possible.
+Our PArrows DSL can be expanded to further parallel Haskells.
+More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} for this future extension.
+HdpH is a modern distributed Haskell that would benefit from our Arrows notation.
+Further Future-aware versions of Arrow combinators can be defined.
+Existing combinators could also be improved.
+We would look into more transparency of our DSL,  basically it should infuse as little overhead as possible.
 
-We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
-Of course, we are working on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
-It would also be interesting to see a hybrid of PArrows and Accelerate.
+More experiences with seamless porting of parallel PArrows-based programs across the backends are welcome.
+Of course, we are working ourselves on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
+It would also be interesting to see a hybrid of PArrows and Accelerate \cite{McDonell:2015:TRC:2887747.2804313}.
 Ports of our approach to other languages like Frege or Java directly are in an early development stage.

### 2017-07-21 paper/content/motivation.tex
intro + conclusion: a first version

@@
 \label{sec:introduction}
-\olcomment{todo, reuse 5.5, and more}
+% \olcomment{todo, reuse 5.5, and more}
+%
+% \mbcomment{
+% Haskell is Spielwiese für Parallelität; verschiedene Ansätze (Par, Multicore, Eden); Orthogonale Ansätze; Verwenden höchstens eine Monade - manchmal auch nur intern; Wir wollen Parallelität mit Arrows abbilden, was noch niemand gemacht hat;
+% Statt einer eigenen Implementierung definieren wir ein "shallow embedded DSL" (ACHTUNG, ist das der richtige Name? effektiv API);
+% Umsetzung mit verschiedenen parallelen Haskells; We tame the zoo of parallel Haskells und vergewissern uns dass es nicht viel Overhead bringt
+% }
+%
+% blablabla arrows, parallel, haskell.
+%
+% \olcomment{An attempt ---->}
+%%%%%
 
-\mbcomment{
-Haskell is Spielwiese für Parallelität; verschiedene Ansätze (Par, Multicore, Eden); Orthogonale Ansätze; Verwenden höchstens eine Monade - manchmal auch nur intern; Wir wollen Parallelität mit Arrows abbilden, was noch niemand gemacht hat;
-Statt einer eigenen Implementierung definieren wir ein "shallow embedded DSL" (ACHTUNG, ist das der richtige Name? effektiv API);
-Umsetzung mit verschiedenen parallelen Haskells; We tame the zoo of parallel Haskells und vergewissern uns dass es nicht viel Overhead bringt
-}
+Pure functional programming languages have been used
+since longer time as a ``sand box'' for testing novel parallel
+programming approaches. Reasons for this include referential
+transparency and the ease of introducing new sophisticated language
+concepts. Specially for Haskell (on which this paper focuses), various
+approaches reached maturity. The definite parallel Haskell does not
+exist, there are multiple parallel Haskells. We regard here in-depth
+the Multicore Haskell (a SMP implementation of GpH language), the
+|Par| Monad, and Eden, a distributed memory parallel Haskell. These
+languages represent orthogonal approaches. Some use a monad, even if
+only for the internal representation. Some introduce additional
+language constructs. 
+%In this paper we introduce a notion of parallel
+%computations using Arrows.
+We use here Arrows to represent parallel computations. This has not
+been done before.
 
-blablabla arrows, parallel, haskell.
+%Our Arrows-based interface is a high-level one.
+We provide an Arrows-based typeclass and various implementations.
+Instead of 
+introducing a new low-level parallel backend in order to implement our
+Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
+is defined as a common interface and varying implementations in
+existing parallel Haskells: Multicore Haskell, |Par| Monad, and
+Eden. Thus, we not only define a parallel programming interface in a
+novel manner. We tame the zoo of parallel Haskells. We provide a
+common, very low-penalty programming interface that allows to switch
+the parallel Haskell backends at will. Further backends are viable, too.
 
 \paragraph{Contribution}
-
-\olcomment{HIT HERE REALLY STRONG}
-
+%
+%\olcomment{HIT HERE REALLY STRONG}
+%
 %\subsection{Impact of parallel Arrows}
@@
 \mbcomment{different, how?}
-We wrap parallel Haskells inside of our |ArrowParallel| interface, but why do we aim to abstract parallelism this way and what does this approach do better than the other parallel Haskells?
+We wrap parallel Haskells inside of our |ArrowParallel| interface, but
+% why do we aim to abstract parallelism this way and what does this
+% approach do better than the other parallel Haskells?
+ is such a parallelism abstraction of benefit and to what extent does
+ it improve existing approaches?
 \begin{itemize}
@@
 This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell.
-So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the actual |ArrowParallel| instance.
+So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
 \end{itemize}

### 2017-07-25 paper/content/benchmarks.tex
minor fix in benchmarks.tex (clarified that we have multiple Gentleman Multiplications at once) pdf for olegs changes

@@
 Jacobi sum test & Mathematics & |workpool+reduce| & Eden & \citet{Lobachev2012}\\
-Gentleman & Mathematics & |torus| & Eden & \citet{Loogen2012}\\
+Gentleman & Mathematics & |torus+parMap| & Eden & \citet{Loogen2012}\\
 Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad}\\
@@
 multiplication \citep{Gentleman1978}. We ported its latest Eden
-version \citep{Loogen2012} to PArrows.
+version \citep{Loogen2012} to PArrows and changed it so that it runs multiple |torus| computations in parallel with |parMap|.
 

### 2017-07-25 paper/content/benchmarks.tex
minor

@@
 of 64-task PArrows/Eden at 16 real cores version was 13.65, the
-efficiency was 85.3\%.  However, when we increase the number of
-requested cores to be 32---\ie when we use hyperthreading on 16 real
+efficiency was 85.3\%.  However, if we increase the number of
+requested cores to be 32---\ie if we use hyperthreading on 16 real
 cores---the speedup does not increase that well. It is merely 15.99
@@
 
-In the distributed memory setting same effect ensues. We obtain
+In the distributed memory setting the same effect ensues. We obtain
 plummeting speedup of 124.31 at 512 hyperthreaded cores, whereas it was
@@
 deviations from both measurements for PArrows and non-PArrows
-versions. These are the values provided by |Criterion| package that we
+versions. These are the values provided by the |bench| package that we
 used for benchmarking. We call a difference between two versions

### 2017-07-25 paper/content/motivation.tex
minor

@@
 Pure functional programming languages have been used
-since longer time as a ``sand box'' for testing novel parallel
+since a longer time as a ``sand box'' for testing novel parallel
 programming approaches. Reasons for this include referential
@@
 approaches reached maturity. The definite parallel Haskell does not
-exist, there are multiple parallel Haskells. We regard here in-depth
+exist, there are multiple parallel Haskells \mbcomment{Wortwiederholung? Oder okay so?}. We regard here in-depth
 the Multicore Haskell (a SMP implementation of GpH language), the
@@
 %computations using Arrows.
-We use here Arrows to represent parallel computations. This has not
+Here, we use Arrows to represent parallel computations. This has not
 been done before.
@@
 Eden. Thus, we not only define a parallel programming interface in a
-novel manner. We tame the zoo of parallel Haskells. We provide a
+novel manner - we tame the zoo of parallel Haskells. We provide a
 common, very low-penalty programming interface that allows to switch
@@
 \mbcomment{different, how?}
-We wrap parallel Haskells inside of our |ArrowParallel| interface, but
+We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
 % why do we aim to abstract parallelism this way and what does this

### 2017-08-15 paper/content/benchmarks.tex
wip on benchmarks for Torus in the paper

@@
 \newcommand{\rmtest}{Rabin--Miller test\xspace}
+\newcommand{\torustest}{Torus Matrix Multiplication test\xspace}
 \newlength{\plotwidthSMP}
@@
 \newcommand{\speedupSkelRMDist}[4]{
-\speedupplot{Speedup of \rmtest ``#2''}{PArrows}{256}{#3}{
+\speedupplot{Speedup of \rmtest ``#1,#2''}{PArrows}{256}{#3}{
 % \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
   consisting of 256 cores. Input was $2^{44497}-1$, we used 256
-  tasks. Top plot shows absolute speedup in a distributed memory setting. The
+  tasks. The top plot shows absolute speedup in a distributed memory setting. The
   closer to linear speedup the better. Time
@@
 \label{fig:bench-rm-dist}
+\end{figure}
+
+\newcommand{\speedupTorusDist}[3]{
+\speedupplot{Speedup of \torustest ``#1''}{PArrows}{256}{#2}{
+\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-#1.csv};
+}{#3}{\plotwidthDist}
+}
+
+\newcommand{\performanceTorusDistDiff}[4]{
+\performancediffplot{Run time differences\\for \torustest ``#1''}{(Eden $-$ PArrows)}{256}{#2}{
+\addplot+[mark=*,very thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+smooth]{benchmarks/distributed-torus/#1-diff.csv};
+}{#3}{#4}{\plotwidthDist}
+}
+
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupTorusDist{1024}{32}{260}\label{123123123}}%
+%\hfill%
+{\performanceTorusDistDiff{1024}{32}{0.5}{260}\label{12312312312}}
+\caption{Parallel performance of \torustest on the Glasgow grid
+  consisting of 256 cores. Input was a matrix size of $1024$. The top plot shows absolute speedup in a distributed memory setting. The
+  closer to linear speedup the better. Time
+  (and hence speedup) measurements for PArrows with Eden backend and
+  Eden almost coincide. Hence, bottom plot shows
+absolute time differences for this benchmark. The
+lower the value, the better for PArrows\olcomment{CHECKME}.}
+\label{123123123122}
 \end{figure}
@@
 Jacobi sum test & Mathematics & |workpool+reduce| & Eden & \citet{Lobachev2012}\\
-Gentleman & Mathematics & |torus+parMap| & Eden & \citet{Loogen2012}\\
+Gentleman & Mathematics & |torus| & Eden & \citet{Loogen2012}\\
 Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad}\\

### 2017-08-15 paper/content/benchmarks.tex
minor

@@
 of Jacobi sum test have very different durations. \citet{lobachev-phd}
-discuss some optimisations of parallel APRCL. Generic parallel
+\olcomment{discuss some optimisations of parallel APRCL.} Generic parallel
 implementation of \rmtest and APRCL were presented in \citet{Lobachev2012}.
@@
 multiplication \citep{Gentleman1978}. We ported its latest Eden
-version \citep{Loogen2012} to PArrows and changed it so that it runs multiple |torus| computations in parallel with |parMap|.
+version \citep{Loogen2012} to PArrows.
 

### 2017-08-15 paper/content/benchmarks.tex
added sudoku benchmarks (initial)

@@
 
+\newcommand{\performanceSudokuSM}[1]{
+\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, Par Monad, Par Monad (vanilla)}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
+}{17}{\plotwidthSMP}
+}
+
+\newcommand{\speedupSudokuSM}[1]{
+\speedupplot{Speedup of \rmtest ``#1''}{Eden CP, Multicore, Par Monad, Par Monad (vanilla)}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
+}{17}{\plotwidthSMP}
+}
+
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupSudokuSM{1000}\label{1231231232}}%
+%\hfill%
+{\performanceSudokuSM{1000}\label{1231231123122}}
+\caption{}
+\label{1231231233122}
+\end{figure}
+
 %\speedupSkelRMDistDiff{44497}{256}{32,64,128,256,512}{544}

### 2017-08-16 paper/content/benchmarks.tex
minor

@@
-%
 \section{Performance results}
@@
 \newcommand{\rmtest}{Rabin--Miller test\xspace}
+\newcommand{\sudokutest}{Sudoku test\xspace}
 \newcommand{\torustest}{Torus Matrix Multiplication test\xspace}
@@
 \caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware. Measurements were performed on a single node of Glasgow
+  different backends on the same hardware. Measurements were performed on a single node of the Glasgow
   grid; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
@@
 \newcommand{\speedupSudokuSM}[1]{
-\speedupplot{Speedup of \rmtest ``#1''}{Eden CP, Multicore, Par Monad, Par Monad (vanilla)}{16}{4}{
+\speedupplot{Speedup of \sudokutest ``#1''}{Eden CP, Multicore, Par Monad, Par Monad (vanilla)}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 {\performanceSudokuSM{1000}\label{1231231123122}}
-\caption{}
+\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
+  different backends on the same hardware and the |parMap| version from the |Par| Monad examples \citet{par-monad}\mbcomment{\\originally from http://community.haskell.org/~simonmar/par-tutorial.pdf}.
+Measurements were performed on a single node of the Glasgow
+  grid; it has 16 real cores. Input was a file of $1000$ Sudokus. The
+  closer to linear speedup the better.}
 \label{1231231233122}
@@
 Gentleman & Mathematics & |torus| & Eden & \citet{Loogen2012}\\
-Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad}\\
+Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad} \mbcomment{originally from: http://community.haskell.org/~simonmar/par-tutorial.pdf}\\
 \bottomrule

### 2017-08-16 paper/content/topologySkeletons.tex
minor
changed Torus Skeleton Example in Paper to the new one

@@
 \begin{code}
-f parparcomp g = (f . put) >>> (get . g)
+f parcomp g = (f . put) >>> (get . g)
 \end{code}
@@
 As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
-Their instantiation of the skeleton |nodefunction| can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
+We however used a later version from  \citep{Eden:SkeletonBookChapter02} and adapted it to the code seen in Fig.~\ref{fig:torusMatMult}.
 \begin{figure}[tb]
 \begin{code}
-nodefunction :: Int -> ((Matrix, Matrix), [Matrix], [Matrix]) -> ([Matrix], [Matrix], [Matrix])
-nodefunction n ((bA, bB), rows, cols) = ([bSum], bA:nextAs , bB:nextBs)
-		where bSum = foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
-			     nextAs = take (n-1) rows
-			     nextBs = take (n-1) cols
+type Matrix = [[Int]]
+
+prMM_torus :: Int -> Int -> Matrix -> Matrix -> Matrix
+prMM_torus numCores problemSizeVal m1 m2 =
+	combine $ torus () (mult torusSize) $ zipWith (zipWith (,)) (split m1) (split m2)
+    	where   torusSize = (floor . sqrt) $ fromIntegral numCores
+            	combine = concat . (map (foldr (zipWith (++)) (repeat [])))
+            	split = splitMatrix (problemSizeVal `div` torusSize)
+
+-- Function performed by each worker
+mult :: Int -> ((Matrix,Matrix),[Matrix],[Matrix]) -> (Matrix,[Matrix],[Matrix])
+mult size ((sm1,sm2),sm1s,sm2s) = (result,toRight,toBottom)
+    where toRight = take (size-1) (sm1:sm1s)
+          toBottom = take (size-1) (sm2':sm2s)
+          sm2' = transpose sm2
+          sms = zipWith prMMTr (sm1:sm1s) (sm2':sm2s)
+          result = foldl1' matAdd sms
+
+prMMTr m1 m2 = [[sum (zipWith (*) row col) | col <- m2 ] | row <- m1]
+
+splitMatrix :: Int -> Matrix -> [[Matrix]]
+splitMatrix size matrix = map (transpose . map (chunksOf size)) $ chunksOf size $ matrix
+
+matAdd = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
 \end{code}
-\caption{Adapted |nodefunction| for matrix multiplication with the |torus| from Fig.~\ref{fig:torus}.}
+\caption{Adapted matrix multiplication in Eden using a the |torus| skeleton.
+|prMM_torus| is the parallel matrix multiplication.
+|mult| is the function performed by each worker.
+|prMMTr| calculates $AB^T$ and is used for the (sequential) calculation in the chunks.
+|splitMatrix| splits the Matrix into chunks.
+|matAdd| calculates $A + B$.}
 \label{fig:torusMatMult}

### 2017-08-16 paper/content/utilityFunctions.tex
minor. added notes talk with oleg
moved trivial definitions of matrix multiplication to the appendix

@@
-\olcomment{more clear structure of the appendix? 1) Arrow utilities, 2) our things we omit for being too simple, 3) other's things we use.}
 \section{Utility Arrows}\label{utilfns}
+
+\mbcomment{Text außenrum bauen, siehe Omitted Function Definitions}
 The 
@@
 
-\section{Omitted Funtion Definitions}
+\section{Omitted Function Definitions}
 \label{app:omitted}
@@
 \end{code}
-	\caption{Definition of parEval2.}
+	\caption{Definition of |parEval2|.}
 	\label{fig:parEval2}
+\end{figure}
+Furthermore, Fig.~\ref{fig:torus_example_rest} contains the ommited definitions of |prMMTr|, - which calculates $AB^T$ for two matrices $A$ and $B$, |splitMatrix| - which splits the a matrix into chunks, and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
+\begin{figure}[h]
+\begin{code}
+prMMTr m1 m2 = [[sum (zipWith (*) row col) | col <- m2 ] | row <- m1]
+
+splitMatrix :: Int -> Matrix -> [[Matrix]]
+splitMatrix size matrix = map (transpose . map (chunksOf size)) $ chunksOf size $ matrix
+
+matAdd = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
+\end{code}
+	\caption{Definition of |prMMTr|, |splitMatrix| and |matAdd|.}
+	\label{fig:torus_example_rest}
 \end{figure}

### 2017-08-20 paper/content/benchmarks.tex
added info about the Glasgow Grid

@@
 in it. Benchmarks were conducted both in a shared and in a distributed
-memory setting. The hardware was the ``Glasgow grid'', consisting of
-16 machines with 16 cores. \olcomment{insert typical blabla: consisting of
-  Intel XXXYYYY at XX~GHz with XXXX~GB RAM and FOOBAR Linux version
-  XXXX with kernel XXXXXXXX.}
-Each core had a hyperthreading capability,
-so the total number of reported processors per machine was
-32. Non-surprisingly, we found that hyperthreaded 32 cores do not
-behave in the same manner as real 16 cores (numbers here for a single
-machine). We disregarded the hyperthreading ability in most of the
+memory setting. All benchmarks were done on the ``Glasgow grid'', consisting of
+16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each.
+Each processor has 8 cores and 16 (hyperthreaded) threads with a base frequency of 2 GHz and a turbo frequency of 2.50 GHz.
+This results in a total of 256 cores and 512 threads for the whole cluster.
+The operating system was Ubuntu 14.04 LTS with Kernel 3.19.0-33.
+Non-surprisingly, we found that hyperthreaded 32 cores do not behave in the same manner as real 16 cores (numbers here for a single machine).
+We disregarded the hyperthreading ability in most of the
 cases.
 
-We used a single machine with 16 real cores as a shared memory testbed
+We used a single node with 16 real cores as a shared memory testbed
 and the whole grid with 256 real cores as a device to test our
 distributed memory software.
-
-\olcomment{only if we do it: list also our i7-6770K machine}
 

### 2017-08-20 paper/content/benchmarks.tex
fixed citation of the torus code

@@
 Jacobi sum test & Mathematics & |workpool+reduce| & Eden & \citet{Lobachev2012}\\
-Gentleman & Mathematics & |torus| & Eden & \citet{Loogen2012}\\
-Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad} \mbcomment{originally from: http://community.haskell.org/~simonmar/par-tutorial.pdf}\\
+Gentleman & Mathematics & |torus| & Eden & \citet{Eden:SkeletonBookChapter02}\\
+Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad} 
+\tablefootnote{actual code from: http://community.haskell.org/~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}\\
 \bottomrule
@@
 for their |torus| skeleton. It implements a parallel matrix
-multiplication \citep{Gentleman1978}. We ported its latest Eden
-version \citep{Loogen2012} to PArrows.
+multiplication \citep{Gentleman1978}. We ported an Eden based version \citep{Eden:SkeletonBookChapter02} to PArrows.
 

### 2017-08-20 paper/content/benchmarks.tex
renamed figs (sudoku and torus)
filling in blanks in the benchmark text. now only the description for the benchmarks is needed for this chapter

@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, Multicore, Par Monad}{16}{4}{
+\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, Multicore, Par Monad}{16}{4}{
+\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 %
-{\speedupTorusDist{1024}{32}{260}\label{123123123}}%
+{\speedupTorusDist{1024}{32}{260}\label{subfig:speedupTorusDist}}%
 %\hfill%
-{\performanceTorusDistDiff{1024}{32}{0.5}{260}\label{12312312312}}
+{\performanceTorusDistDiff{1024}{32}{0.5}{260}\label{subfig:performancetorusDistDiff}}
 \caption{Parallel performance of \torustest on the Glasgow grid
@@
 lower the value, the better for PArrows\olcomment{CHECKME}.}
-\label{123123123122}
+\label{fig:torusBenchmark}
 \end{figure}
@@
 \newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, Par Monad, Par Monad (vanilla)}{16}{4}{
+\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
+%\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
 }{17}{\plotwidthSMP}
@@
 \newcommand{\speedupSudokuSM}[1]{
-\speedupplot{Speedup of \sudokutest ``#1''}{Eden CP, Multicore, Par Monad, Par Monad (vanilla)}{16}{4}{
+\speedupplot{Speedup of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
+%\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
 }{17}{\plotwidthSMP}
@@
 %
-{\speedupSudokuSM{1000}\label{1231231232}}%
+{\speedupSudokuSM{1000}\label{subfig:speedupSudokuSM}}%
 %\hfill%
-{\performanceSudokuSM{1000}\label{1231231123122}}
+{\performanceSudokuSM{1000}\label{subfig:performanceSudokuSM}}
 \caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware and the |parMap| version from the |Par| Monad examples \citet{par-monad}\mbcomment{\\originally from http://community.haskell.org/~simonmar/par-tutorial.pdf}.
-Measurements were performed on a single node of the Glasgow
+  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
   grid; it has 16 real cores. Input was a file of $1000$ Sudokus. The
   closer to linear speedup the better.}
-\label{1231231233122}
+\label{fig:sudokuSMBenchmark}
 \end{figure}
@@
 parallelism. For this reason, we did not regard hyperthreaded cores in
-our speedup plots in Figs.~\ref{fig:bench-rm-sm}--XXXXXXXX.
+our speedup plots in Figs.~\ref{fig:bench-rm-sm}-\ref{fig:sudokuSMBenchmark}.
 
@@
 genuine |Par|
-Monad benchmark is very small. To give an example, it is XXXXXXXX for
-XXXXXXXXX and XXXXXXXXXX in the shared memory setting. It is almost invisible in speedup and
+Monad benchmark is very small. To give an example, it is ~0.4s in favour of PArrows for 16 cores (10.8s vs. 11.2s) and ~-0.8s in favour of the |Par| monad for 8 cores (16.1s vs. 16.9s) for
+the Sudoku benchmark in the shared memory setting. It is almost invisible in speedup and
 (non shown) run time plots. We thus show only the results for the
@@
 PArrows and the corresponding backend in a separate plot
-(Figs.~\ref{fig:bench-rm-dist}, XXXXXXX). The differences range in
+(Figs.~\ref{fig:bench-rm-dist}, \ref{fig:torusBenchmark}). The differences range in
 about 0.5~seconds for the execution time of 46~seconds on 256 cores

### 2017-08-23 paper/content/benchmarks.tex
reordering code in benchmarks.tex for easier reading

@@
 \newcommand{\rmtest}{Rabin--Miller test\xspace}
-\newcommand{\sudokutest}{Sudoku test\xspace}
-\newcommand{\torustest}{Torus Matrix Multiplication test\xspace}
+\newcommand{\sudokutest}{Sudoku solver\xspace}
+\newcommand{\jacobitest}{Jacobi sum test\xsp}
+\newcommand{\torustest}{Gentleman\xspace}
 \newlength{\plotwidthSMP}
@@
 
-\newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
-}{17}{\plotwidthSMP}
-}
-
-\newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
-}{17}{\plotwidthSMP}
-}
-
-%\performanceSkelRMSM{4423}{32}
-
-%\speedupSkelRMSM{4423}{32}
-
-%\performanceSkelRMSM{4423}{64}
-
-%\speedupSkelRMSM{4423}{64}
-
-%\performanceSkelRMSM{9941}{32}
-
-%\speedupSkelRMSM{9941}{32}{32}
-
-%\performanceSkelRMSM{9941}{64}
-
-%\speedupSkelRMSM{9941}{64}
-
-%\performanceSkelRMSM{11213}{32}
-
-\begin{figure}
-%\centering
-%\performanceSkelRMSM{11213}{64}\hfill%
-{\speedupSkelRMSM{11213}{32}}\hfill%
-{\speedupSkelRMSM{11213}{64}}
-\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware. Measurements were performed on a single node of the Glasgow
-  grid; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
-  tasks. The
-  closer to linear speedup the better.}
-\label{fig:bench-rm-sm}
-\end{figure}
-
-% \newcommand{\performanceSkelRMDist}[4]{
-% \performanceplot{Parallel run time \\ SkelRM #1 #2}{PArrows}{256}{#3}{
-% \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
-% % \addplot table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-% % smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-% }{#4}
-% }
-
-\newcommand{\speedupSkelRMDist}[4]{
-\speedupplot{Speedup of \rmtest ``#1,#2''}{PArrows}{256}{#3}{
-% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
-\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
-% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-}{#4}{\plotwidthDist}
-}
-
-\newcommand{\performanceSkelRMDistDiff}[5]{
-\performancediffplot{Run time differences\\for \rmtest ``#1, #2''}{(Eden $-$ PArrows)}{256}{#3}{
-\addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
-smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-}{#4}{#5}{\plotwidthDist}
-}
-
-% \newcommand{\speedupSkelRMDistDiff}[4]{
-% \speedupdiffplot{Parallel speedup difference \\ of SkelRM #1 #2}{(Eden - PArrows)}{512}{#3}{
-% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-% }{#4}
-% }
-
-%\performanceSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
-
-%\performanceSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{0.11}{528}
-
-%\speedupSkelRMDist{9941}{256}{16,32,64,128,256,512}{528}
-
-%\speedupSkelRMDistDiff{9941}{256}{16,32,64,128,256,512}{528}
-
-%\performanceSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
-
-%\performanceSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{0.1}{528}
-
-%\speedupSkelRMDist{11213}{256}{16,32,64,128,256,512}{528}
-
-%\speedupSkelRMDistDiff{11213}{256}{16,32,64,128,256,512}{528}
-
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupSkelRMDist{44497}{256}{32}{260}\label{subfig:rm-dist-a}}%
-%\hfill%
-{\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{260}\label{subfig:rm-dist-b}}
-\caption{Parallel performance of \rmtest on the Glasgow grid
-  consisting of 256 cores. Input was $2^{44497}-1$, we used 256
-  tasks. The top plot shows absolute speedup in a distributed memory setting. The
-  closer to linear speedup the better. Time
-  (and hence speedup) measurements for PArrows with Eden backend and
-  Eden almost coincide. Hence, bottom plot shows
-absolute time differences for this benchmark. The
-lower the value, the better for PArrows\olcomment{CHECKME}.}
-\label{fig:bench-rm-dist}
-\end{figure}
-
-\newcommand{\speedupTorusDist}[3]{
-\speedupplot{Speedup of \torustest ``#1''}{PArrows}{256}{#2}{
-\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-#1.csv};
-}{#3}{\plotwidthDist}
-}
-
-\newcommand{\performanceTorusDistDiff}[4]{
-\performancediffplot{Run time differences\\for \torustest ``#1''}{(Eden $-$ PArrows)}{256}{#2}{
-\addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
-smooth]{benchmarks/distributed-torus/#1-diff.csv};
-}{#3}{#4}{\plotwidthDist}
-}
-
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupTorusDist{1024}{32}{260}\label{subfig:speedupTorusDist}}%
-%\hfill%
-{\performanceTorusDistDiff{1024}{32}{0.5}{260}\label{subfig:performancetorusDistDiff}}
-\caption{Parallel performance of \torustest on the Glasgow grid
-  consisting of 256 cores. Input was a matrix size of $1024$. The top plot shows absolute speedup in a distributed memory setting. The
-  closer to linear speedup the better. Time
-  (and hence speedup) measurements for PArrows with Eden backend and
-  Eden almost coincide. Hence, bottom plot shows
-absolute time differences for this benchmark. The
-lower the value, the better for PArrows\olcomment{CHECKME}.}
-\label{fig:torusBenchmark}
-\end{figure}
-
-\newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
-%\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
-}{17}{\plotwidthSMP}
-}
-
-\newcommand{\speedupSudokuSM}[1]{
-\speedupplot{Speedup of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
-%\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
-}{17}{\plotwidthSMP}
-}
-
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupSudokuSM{1000}\label{subfig:speedupSudokuSM}}%
-%\hfill%
-{\performanceSudokuSM{1000}\label{subfig:performanceSudokuSM}}
-\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
-  grid; it has 16 real cores. Input was a file of $1000$ Sudokus. The
-  closer to linear speedup the better.}
-\label{fig:sudokuSMBenchmark}
-\end{figure}
-
-%\speedupSkelRMDistDiff{44497}{256}{32,64,128,256,512}{544}
-
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Text
-
 \subsection{Hardware}
@@
 \rmtest & Mathematics & |parMap+reduce| & Eden & \citet{Lobachev2012}\\
-Jacobi sum test & Mathematics & |workpool+reduce| & Eden & \citet{Lobachev2012}\\
-Gentleman & Mathematics & |torus| & Eden & \citet{Eden:SkeletonBookChapter02}\\
-Sudoku & Puzzle & |parMap| & |Par| Monad & \citet{par-monad} 
+\jacobitest & Mathematics & |workpool+reduce| & Eden & \citet{Lobachev2012}\\
+\torustest & Mathematics & |torus| & Eden & \citet{Eden:SkeletonBookChapter02}\\
+\sudokutest & Puzzle & |parMap| & |Par| Monad & \citet{par-monad} 
 \tablefootnote{actual code from: http://community.haskell.org/~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}\\
@@
 \subsubsection{\rmtest}
+
+\newcommand{\performanceSkelRMSM}[2]{
+\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
+}{17}{\plotwidthSMP}
+}
+
+\newcommand{\speedupSkelRMSM}[2]{
+\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
+}{17}{\plotwidthSMP}
+}
+
+\newcommand{\speedupSkelRMDist}[4]{
+\speedupplot{Speedup of \rmtest ``#2''}{PArrows}{256}{#3}{
+% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
+\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
+% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
+}{#4}{\plotwidthDist}
+}
+
+\newcommand{\performanceSkelRMDistDiff}[5]{
+\performancediffplot{Run time differences\\for \rmtest ``#2''}{(Eden $-$ PArrows)}{256}{#3}{
+\addplot+[mark=*,very thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
+}{#4}{#5}{\plotwidthDist}
+}
+
+\begin{figure}
+%\centering
+%\performanceSkelRMSM{11213}{64}\hfill%
+{\speedupSkelRMSM{11213}{32}}\hfill%
+{\speedupSkelRMSM{11213}{64}}
+\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
+  different backends on the same hardware. Measurements were performed on a single node of the Glasgow
+  grid; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
+  tasks. The
+  closer to linear speedup the better.}
+\label{fig:bench-rm-sm}
+\end{figure}
+
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupSkelRMDist{44497}{256}{32}{260}\label{subfig:rm-dist-a}}%
+%\hfill%
+{\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{260}\label{subfig:rm-dist-b}}
+\caption{Parallel performance of \rmtest on the Glasgow grid
+  consisting of 256 cores. Input was $2^{44497}-1$, we used 256
+  tasks. The top plot shows absolute speedup in a distributed memory setting. The
+  closer to linear speedup the better. Time
+  (and hence speedup) measurements for PArrows with Eden backend and
+  Eden almost coincide. Hence, bottom plot shows
+absolute time differences for this benchmark. The
+lower the value, the better for PArrows\olcomment{CHECKME}.}
+\label{fig:bench-rm-dist}
+\end{figure}
+
 \olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
@@
 cores. The PArrows version was 0.30\% \emph{faster} for 128 cores.
+
+\subsubsection{\torustest}
+
+\newcommand{\speedupTorusDist}[3]{
+\speedupplot{Speedup of \torustest ``#1''}{PArrows}{256}{#2}{
+\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-#1.csv};
+}{#3}{\plotwidthDist}
+}
+
+\newcommand{\performanceTorusDistDiff}[4]{
+\performancediffplot{Run time differences\\for \torustest ``#1''}{(Eden $-$ PArrows)}{256}{#2}{
+\addplot+[mark=*,very thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+smooth]{benchmarks/distributed-torus/#1-diff.csv};
+}{#3}{#4}{\plotwidthDist}
+}
+
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupTorusDist{1024}{32}{260}\label{subfig:speedupTorusDist}}%
+%\hfill%
+{\performanceTorusDistDiff{1024}{32}{0.5}{260}\label{subfig:performancetorusDistDiff}}
+\caption{Parallel performance of \torustest on the Glasgow grid
+  consisting of 256 cores. Input was a matrix size of $1024$. The top plot shows absolute speedup in a distributed memory setting. The
+  closer to linear speedup the better. Time
+  (and hence speedup) measurements for PArrows with Eden backend and
+  Eden almost coincide. Hence, bottom plot shows
+absolute time differences for this benchmark. The
+lower the value, the better for PArrows\olcomment{CHECKME}.}
+\label{fig:torusBenchmark}
+\end{figure}
+
+\subsubsection{\sudokutest}
+
+\newcommand{\performanceSudokuSM}[1]{
+\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
+%\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
+}{17}{\plotwidthSMP}
+}
+
+\newcommand{\speedupSudokuSM}[1]{
+\speedupplot{ \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
+%\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
+}{17}{\plotwidthSMP}
+}
+
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupSudokuSM{1000}\label{subfig:speedupSudokuSM}}%
+%\hfill%
+{\performanceSudokuSM{1000}\label{subfig:performanceSudokuSM}}
+\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
+  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
+  grid; it has 16 real cores. Input was a file of $1000$ Sudokus. The
+  closer to linear speedup the better.}
+\label{fig:sudokuSMBenchmark}
+\end{figure}

### 2017-08-23 paper/content/benchmarks.tex
minor. added placeholder for real jacobi sum test

@@
 
+\subsubsection{\jacobitest}
+
+\newcommand{\speedupJacobiDist}[4]{
+\speedupplot{Speedup of \jacobitest ``#2''}{PArrows}{256}{#3}{
+% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
+\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#2.csv};
+% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
+}{#4}{\plotwidthDist}
+}
+
+\newcommand{\performanceJacobiDistDiff}[5]{
+\performancediffplot{Run time differences\\for \jacobitest ``#2''}{(Eden $-$ PArrows)}{256}{#3}{
+\addplot+[mark=*,very thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+smooth]{benchmarks/distributed-jacobi/#1-#2-diff.csv};
+}{#4}{#5}{\plotwidthDist}
+}
+
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupJacobiDist{3}{1279}{32}{260}\label{subfig:rm-dist-a}}%
+%\hfill%
+{\performanceJacobiDistDiff{3}{1279}{32}{0.5}{260}\label{subfig:rm-dist-b}}
+\caption{Parallel performance of \rmtest on the Glasgow grid
+  consisting of 256 cores. Input was $2^{1279}-1$, we used 256
+  tasks. The top plot shows absolute speedup in a distributed memory setting. The
+  closer to linear speedup the better. Time
+  (and hence speedup) measurements for PArrows with Eden backend and
+  Eden almost coincide. Hence, bottom plot shows
+absolute time differences for this benchmark. The
+lower the value, the better for PArrows\olcomment{CHECKME}.}
+\label{fig:bench-jacobi-dist}
+\end{figure}
+
+\mbcomment{FIXME: redo this test with $2^{44497}$}
+
 \subsubsection{\torustest}

### 2017-08-23 paper/content/benchmarks.tex
work in progress on the benchmark chapter. reordering things. SkelRM finished

@@
 
-To showcase that the induced overhead of
-PArrows is small, we plot execution time differences between measurements for
+To show that PArrows induce very small overhead in a distributed context as well, we compare the original Eden
+versions of the benchmark to its PArrows-enabled counterpart in the \rmtest, \torustest and \jacobitest benchmarks. We plot execution time differences between measurements for
 PArrows and the corresponding backend in a separate plot
-(Figs.~\ref{fig:bench-rm-dist}, \ref{fig:torusBenchmark}). The differences range in
+(Figs.~\ref{fig:bench-rm-dist}, \ref{fig:torusBenchmark}, \ref{fig:bench-jacobi-dist}). As an example, the differences range in
 about 0.5~seconds for the execution time of 46~seconds on 256 cores
-for distributed \rmtest with PArrows and Eden. The plots show absolute
-time differences that are not relativated \wrt the total execution time.
-
-The error bars ends were computed from pointwise maximum of both standard
+for distributed \rmtest with PArrows and Eden. For these comparisons, the plots show absolute
+time differences that are not relative \wrt the total execution time.
+Furthermore, the error bars ends were computed from pointwise maximum of both standard
 deviations from both measurements for PArrows and non-PArrows
@@
 The multicore version of our parallel \rmtest benchmark is in
-Figure~\ref{fig:bench-rm-sm}. We have executed the test with 32 and 64
+Figure~\ref{fig:bench-rm-sm}. We executed the test with 32 and 64
 tasks. The plot shows the PArrows-enabled versions with corresponding backends.
@@
 for SMP variants such as PArrows/Multicore Haskell and PArrows/|Par|
-Monad. The reason for this behavior lies in the backend in our reasoning. One
-possible explanation is garbage collection. It is easier to GC
-(independently) $n$ heaps than to GC one large. This agrees with known
-results\olcomment{WHICH EXACTLY?} that motivate the efforts for
-parallel GC in SMP-based Haskells.
+Monad but most of the time the performance is still comparable with the Multicore backend performing slightly worse than the other two in terms of speedup.
+In particular, the speedups for 16 cores fall behind quite noticeable with 11.69 for 32 tasks as Eden and the |Par| Monad both score 13.05 for 32 tasks.
+While the Multicore backend scores a bit better in the 64 tasks benchmark, its speedup of 12.70 is still behind the 13.65 and 13.33 of Eden and |Par| Monad, respectively.
 
-To show that PArrows induce very small overhead, we compare the ported
-versions of the benchmark to the PArrows-enabled ones with
-corresponding backend.\olcomment{THE RESULT?}
-
-In the distributed memory setting, we utilize PArrows/Eden. Eden in
-its turn used the MPI backend. We see an almost linear speedup of
-\rmtest with 256 tasks and input $2^444497-1$. The sequential run time
-was computed as mean of three consecutive executions on a single
+Comparing the PArrows version of the \rmtest with the original from Eden with the MPI backend in a distributed memory setting, we see an almost linear speedup of
+\rmtest with 256 tasks and input $2^444497-1$ in both versions. The sequential run time
+was computed as the mean of three consecutive executions on a single
 core---the single run took two hours 43 minutes. The zero difference  between

### 2017-08-23 paper/content/benchmarks.tex
minor

@@
 
-\subsubsection{\rmtest}
+\subsubsection{\rmtest}\label{sec:rmtest}
 
@@
 absolute time differences for this benchmark. The
-lower the value, the better for PArrows\olcomment{CHECKME}.}
+higher the value, the better for PArrows\olcomment{CHECKME}.}
 \label{fig:bench-rm-dist}
@@
   Why is it good?}
-The multicore version of our parallel \rmtest benchmark is in
+The multicore version of our parallel \rmtest benchmark is depicted in
 Figure~\ref{fig:bench-rm-sm}. We executed the test with 32 and 64
@@
 PArrows/Eden and Eden almost always lies on the error bar of
-the measurement. The only exceptions, where PArrows version was slower
-and it was significant, were for 64 and
-512 cores, for 0.49 and 0.26~second, respectively. This corresponds to
-0.30\% and 0.33\% relative time difference, respectively. The 512 core
-version suffers, of course, from the aforementioned hyperthreading
-problem. The PArrows-induced overhead was merely 0.56\% for 256
-cores. The PArrows version was 0.30\% \emph{faster} for 128 cores.
+the measurement. The only exception, where the PArrows and Eden versions differ and it was it was significant, was for 64 cores, where it performed 0.49s better. This corresponds to
+0.30\% relative time difference. Otherwise, PArrows was 0.23s (or 0.0007\%) slower for 32 cores, 0.26s (or 0.0030\%) slower for 128 cores, and 0.09s faster (or 0.0019\%) faster for 256 cores.
+
+%As the PArrows version uses Eden in the backend, these numbers suggest that there is no real performance difference between using PArrows or Eden for this task as they trade blows in this benchmark.
+Additionally, PArrows with an Eden-based backend performing better than what it is based upon suggests that any difference in runtime between the two is more of an anomaly than a real difference.
 
 \subsubsection{\jacobitest}
+
+Continuing, the results of the \jacobitest in \ref{fig:bench-jacobi-dist} are as follows:
 
@@
 absolute time differences for this benchmark. The
-lower the value, the better for PArrows\olcomment{CHECKME}.}
+higher the value, the better for PArrows\olcomment{CHECKME}.}
 \label{fig:bench-jacobi-dist}
@@
 \subsubsection{\torustest}
+
+Next is the \torustest benchmark.
+The results of the comparison of vanilla Eden to our PArrows-based version can be found in \ref{fig:torusBenchmark}.
+While we do not see too much speedup for more than 16 cores, we can prove that the difference between the Eden and PArrows version are again marginal.
+The difference between PArrows and Eden is only significant for 16 and 64 cores where it ran 1.7\% and 2.7\% slower which corresponds to a real-time difference of 0.12s and 0.13s.
+For 256 cores PArrows performed 0.2\% slower which corresponds to 0.01s overhead.
 
@@
 absolute time differences for this benchmark. The
-lower the value, the better for PArrows\olcomment{CHECKME}.}
+higher the value, the better for PArrows\olcomment{CHECKME}.}
 \label{fig:torusBenchmark}
@@
 \subsubsection{\sudokutest}
+
+As the last Benchmark in this paper we present the \sudokutest in \ref{fig:sudokuSMBenchmark} running in a shared memory setting.
+Here we see all three SM backends performing similar again like in the \rmtest SM benchmarks.
+\mbcomment{Benchmarks seem a bit off.
+redo}
 
@@
 {\performanceSudokuSM{1000}\label{subfig:performanceSudokuSM}}
-\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
+\caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
   different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow

### 2017-08-23 paper/content/utilityFunctions.tex
minor improvements in utilityfunctions.tex

@@
 \section{Utility Arrows}\label{utilfns}
-
-\mbcomment{Text außenrum bauen, siehe Omitted Function Definitions}
-The 
-|second| combinator: 
+Following are definitions of some utility Arrows used in this paper that have been lef out for brevity.
+We start with the |second| combinator from \citet{HughesArrows}, which is a mirrored version of |first|, which is for example used in the definition of |***|: 
 \begin{code}
@@
 \end{code}
-is a mirrored version of |first|, used, \eg in
-definition of |***|.
 
-Utility combinators for Parallel Arrows follow. We define |map|, |foldl| and |zipWith| on Arrows.
+Next, we also define |map|, |foldl| and |zipWith| on Arrows.
 The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any arrow |arr a
@@
 \section{Syntactic Sugar} \label{syntacticSugar}
+Finally, we also give the definitions for some syntactic sugar for PArrows, namely |***| and |&&&|.
 For basic arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two arrows |arr a b| and |arr c d| into an arrow |arr (a, c) (b, d)| which does both computations at once.

### 2017-08-26 paper/content/utilityFunctions.tex
minor update

@@
 brevity, and redeem this here.
+
+To begin with, we give the actual implementation of the non-configurable instance of the |ArrowParallel| instance:
+
+\begin{figure}
+\begin{code}
+instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
+    parEvalN _ fs = parEvalN (hack fs) fs
+                    where
+                        hack :: (NFData b) => [arr a b] -> Conf b
+                        hack _ = Conf rdeepseq
+\end{code}
+\caption{The actual default implementation of Multicore's |ArrowParallel|.}
+\label{fig:actualNonConfigMulticore}
+\end{figure}
+
 %
-We warp Eden's build-in Futures in PArrows as in
+Next, we warp Eden's build-in Futures in PArrows as in
 Figure~\ref{fig:RDFuture}, where |rd| is the accessor function for the |RD| wrapped inside |RemoteData|.

### 2017-08-26 paper/content/benchmarks.tex
minor

@@
 Monad but most of the time the performance is still comparable with the Multicore backend performing slightly worse than the other two in terms of speedup.
-In particular, the speedups for 16 cores fall behind quite noticeable with 11.69 for 32 tasks as Eden and the |Par| Monad both score 13.05 for 32 tasks.
+In particular, the speedups for 16 cores fall behind quite noticeably with 11.69 for 32 tasks as Eden and the |Par| Monad both score 13.05 for 32 tasks.
 While the Multicore backend scores a bit better in the 64 tasks benchmark, its speedup of 12.70 is still behind the 13.65 and 13.33 of Eden and |Par| Monad, respectively.
@@
 Comparing the PArrows version of the \rmtest with the original from Eden with the MPI backend in a distributed memory setting, we see an almost linear speedup of
-\rmtest with 256 tasks and input $2^444497-1$ in both versions. The sequential run time
+\rmtest with 256 tasks and input $2^{444497}-1$ in both versions. The sequential run time
 was computed as the mean of three consecutive executions on a single
@@
 \mbcomment{Benchmarks seem a bit off.
-redo}
+redoing atm}
 

### 2017-08-26 paper/content/topologySkeletons.tex
moved trivial definitions of matrix multiplication to the appendix

@@
 Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons.
-The Eden library leverages this problem and already comes with more predefined skeletons\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.}, among them a |pipe|, a |ring|, and a |torus| implementations \cite{Loogen2012}.
+The Eden library leverages this problem and already comes with more predefined skeletons\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.}, among them a |pipe|, a |ring|, and a |torus| implementations \citep{Eden:SkeletonBookChapter02}.
 These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell.
@@
 \end{figure}
-As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
-We however used a later version from  \citep{Eden:SkeletonBookChapter02} and adapted it to the code seen in Fig.~\ref{fig:torusMatMult}.
+As an example of using this skeleton \citep{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
+An adapted version can be found in Fig.~\ref{fig:torusMatMult}.
 \begin{figure}[tb]
@@
           result = foldl1' matAdd sms
-
-prMMTr m1 m2 = [[sum (zipWith (*) row col) | col <- m2 ] | row <- m1]
-
-splitMatrix :: Int -> Matrix -> [[Matrix]]
-splitMatrix size matrix = map (transpose . map (chunksOf size)) $ chunksOf size $ matrix
-
-matAdd = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
 \end{code}
@@
 |splitMatrix| splits the Matrix into chunks.
-|matAdd| calculates $A + B$.}
+|matAdd| calculates $A + B$.
+Omitted definitions can be found in \ref{fig:torus_example_rest}.
+}
 \label{fig:torusMatMult}

### 2017-08-26 paper/content/benchmarks.tex
sudoku benchmark finished

@@
 \newcommand{\rmtest}{Rabin--Miller test\xspace}
-\newcommand{\sudokutest}{Sudoku solver\xspace}
+\newcommand{\sudokutest}{Sudoku\xspace}
 \newcommand{\jacobitest}{Jacobi sum test\xsp}
@@
 scale only axis, width=#7,
-xlabel=Cores,
+xlabel=Threads,
 %xtick=data,
@@
 scale only axis, width=#8,
-xlabel=Cores,
+xlabel=Threads,
 %xtick=data,
@@
 As the last Benchmark in this paper we present the \sudokutest in \ref{fig:sudokuSMBenchmark} running in a shared memory setting.
-Here we see all three SM backends performing similar again like in the \rmtest SM benchmarks.
-\mbcomment{Benchmarks seem a bit off.
-redoing atm}
+Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark}-\ref{fig:sudokuSMBenchmark16000}.
+However, we notice that the Multicore backend seems to choke on a bigger input \ref{fig:sudokuSMBenchmark16000}.
+This is due to the benchmark only using |parMap| instead of a chunking variant (however we did not change that for simplicity's sake) and is reflected by debug output which shows, that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
+Another remarkable finding is that the Eden backend seems to lack behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.
 
 \newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{32}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 %smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
-}{17}{\plotwidthSMP}
+}{33}{\plotwidthSMP}
 }
@@
 \newcommand{\speedupSudokuSM}[1]{
-\speedupplot{ \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\speedupplot{ \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{32}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
   different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
-  grid; it has 16 real cores. Input was a file of $1000$ Sudokus. The
+  grid; it has 16 real cores and 32 threads. Input was a file of $1000$ Sudokus. The
   closer to linear speedup the better.}
@@
 \end{figure}
+
+\begin{figure}
+\centering
+%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
+%
+{\speedupSudokuSM{16000}\label{subfig:speedupSudokuSM16000}}%
+%\hfill%
+{\performanceSudokuSM{16000}\label{subfig:performanceSudokuSM16000}}
+\caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
+  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
+  grid; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
+  closer to linear speedup the better. The Multicore version shows signs of choking with too many sparks being created.}
+\label{fig:sudokuSMBenchmark16000}
+\end{figure}

### 2017-09-12 paper/content/abstract.tex
added olegl changes
mentioning Par Monad problems now. "Monad" instead of "monad"
minor
phil's abstract comments

@@
 \begin{abstract}
-Arrows are a general interface for computation and therefore form an alternative to monads for API design.
+Arrows are a general interface for computation and therefore form an alternative to Monads for API design.
 We express parallelism using this concept in a novel way: We define an arrows-based language for parallelism and implement it using multiple parallel Haskells.
@@
 
-Additionally, our way of writing parallel programs has the benefit of being portable across flavours of parallel Haskells.
-Furthermore, as each parallel computation is an arrow, which means that they can be composed and transformed as such.
+Additionally, our way of writing parallel programs has the benefit of being portable across multiple parallel Haskell implementations.
+Furthermore, as each parallel computation is an arrow, which means that they can be readily composed and transformed as such.
 We introduce some syntactic sugar to provide parallelism-aware arrow combinators.
 
-To show that our arrow-based language is on par with the existing parallel languages, we also define several parallel skeletons with our framework. 
-Benchmarks show that our framework does not induce too much overhead performance-wise.
+To show that our arrow-based language has similar expressiveness to existing parallel languages, we also define several parallel skeletons with our framework. 
+Benchmarks show that our framework does not induce too much overhead
+performance-wise.
+\olcomment{Summarize conclusions}
+
+\mbcomment{Jedes Kapitel soll einmal ins Abstract. Conclusions sollen mit ins Abstract}
 \end{abstract}

### 2017-09-12 paper/content/motivation.tex
Motivation.tex remarks Phil

@@
 
-Pure functional programming languages have been used
-since a longer time as a ``sand box'' for testing novel parallel
-programming approaches. Reasons for this include referential
-transparency and the ease of introducing new sophisticated language
-concepts. Specially for Haskell (on which this paper focuses), various
-approaches reached maturity. The definite parallel Haskell does not
-exist, there are multiple parallel Haskells \mbcomment{Wortwiederholung? Oder okay so?}. We regard here in-depth
+One particular reason for this is the ease of introducing new sophisticated language
+concepts.
+Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms including the expression of parallelism.
+Haskell, which we focus on in this paper, has  several mature implementations.
+We regard here in-depth
 the Multicore Haskell (a SMP implementation of GpH language), the
@@
 %computations using Arrows.
-Here, we use Arrows to represent parallel computations. This has not
-been done before.
+A key novelty here is to use Arrows to represent parallel computations. They seem a natural fit as they are a generalization of the function |->| and serve as general interface to computations.
 
 %Our Arrows-based interface is a high-level one.
-We provide an Arrows-based typeclass and various implementations.
+We provide an Arrows-based typeclass and implementations for three parallel Haskells.
 Instead of 
@@
 novel manner - we tame the zoo of parallel Haskells. We provide a
-common, very low-penalty programming interface that allows to switch
-the parallel Haskell backends at will. Further backends are viable, too.
+common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
+the parallel Haskell backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
 
-\paragraph{Contribution}
+\paragraph{Contributions}
 %

### 2017-09-12 paper/content/motivation.tex
motivation rewrite continued.

@@
 \mbcomment{different, how?}
-We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
+%We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
 % why do we aim to abstract parallelism this way and what does this
 % approach do better than the other parallel Haskells?
- is such a parallelism abstraction of benefit and to what extent does
- it improve existing approaches?
+% is such a parallelism abstraction of benefit and to what extent does
+% it improve existing approaches?
 \begin{itemize}
 	\item \textbf{Arrow DSL benefits}:
-	With the |ArrowParallel| typeclass we do not lose any benefits of using arrows as |parEvalN| is a yet another Arrow combinator.
+    To implement parallelism, we do not introduce any new types, but only rely on a typeclass that hosts |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of arrows into a new parallel arrow.
+Therefore, we do not lose any benefits of using arrows as parallelism is just encapsulated in yet another Arrow combinator.
 The resulting Arrow can be used in the same way a potential serial version could be used.
-This is a big advantage of this approach, especially compared to the Monad solutions as we do not introduce any new types.
+This is a big advantage of this approach, especially compared to Monad solutions like the |Par| Monad which require specialized Monad types.
 We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
@@
 	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic.
-This leaves us in the beautiful situation of being able to write our code against the interface of the typeclass without being bound to any parallel Haskell.
+This means it is possible to write our code against the interface of a common typeclass without being bound to any parallel Haskell.
 So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.

### 2017-09-12 paper/content/motivation.tex
wip on motivation

@@
 %\olcomment{move this to Contributions in the front or something}
-\mbcomment{different, how?}
+We propose a new Arrows-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of Arrows into a new parallel Arrow.
+Because of this, we do not lose any benefits of using Arrows as the parallelism is encapsulated as another combinator instead of a different type.
+The resulting Arrow can be used in the same way as a potential serial version.
+
+This is a big advantage as we do not introduce any new types like Monad solutions like the |Par| Monad do.
+We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
+
+We host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic in a way that the backend can be swapped easily.
+This means it is possible to write code against the interface of a common typeclass without being bound to any specific backend.
+So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance.
+
+\ptcomment{* We evaluate the expressive power of  arrow parallelism. Showing that 
+  + there is no need to introduce additional types
+  + that it is possible to define algorthmic skeletons (Section 6,7)
+  + ...
+
+* We demonstrate that Arrow Parallelism can exploit multiple parallel Haskell implementations, and with low overhead, i.e. less than X%. 
+... (Section 8) }
+
 %We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
@@
 % it improve existing approaches?
-\begin{itemize}
-	\item \textbf{Arrow DSL benefits}:
-    To implement parallelism, we do not introduce any new types, but only rely on a typeclass that hosts |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of arrows into a new parallel arrow.
+%\begin{itemize}
+%	\item \textbf{Arrow DSL benefits}:
+%    To implement parallelism, we do not introduce any new types, but only rely on a typeclass that hosts |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of arrows into a new parallel arrow.
 Therefore, we do not lose any benefits of using arrows as parallelism is just encapsulated in yet another Arrow combinator.
@@
 We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
-	\item \textbf{Abstraction}:
-	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic.
-This means it is possible to write our code against the interface of a common typeclass without being bound to any parallel Haskell.
+%	\item \textbf{Abstraction}:
+%	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic.
+This means it is possible to write code against the interface of a common typeclass without being bound to any parallel Haskell.
 So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
-\end{itemize}
+%\end{itemize}
 

### 2017-09-25 paper/content/motivation.tex
wip restructuring the paper, wip on motivation's contribution

@@
 %\olcomment{move this to Contributions in the front or something}
-We propose a new Arrows-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of Arrows into a new parallel Arrow.
+We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of Arrows into a new parallel Arrow.
 Because of this, we do not lose any benefits of using Arrows as the parallelism is encapsulated as another combinator instead of a different type.
@@
 
-This is a big advantage as we do not introduce any new types like Monad solutions like the |Par| Monad do.
-We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
+This is a big advantage as we do not introduce any new types as Monad solutions like the |Par| Monad do.
+We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change integral parts of the code.
 
-We host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic in a way that the backend can be swapped easily.
+We host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic in a way that the backend easily be swapped.
 This means it is possible to write code against the interface of a common typeclass without being bound to any specific backend.
@@
 
-\ptcomment{* We evaluate the expressive power of  arrow parallelism. Showing that 
-  + there is no need to introduce additional types
-  + that it is possible to define algorthmic skeletons (Section 6,7)
-  + ...
+We evaluate the expressive power of arrow parallelism further by showing that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}).
 
-* We demonstrate that Arrow Parallelism can exploit multiple parallel Haskell implementations, and with low overhead, i.e. less than X%. 
-... (Section 8) }
+Finally, we demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced, i.e.
+less than x\% \mbcomment{fix this after benchmarks are finally done} in Section \ref{sec:benchmarks}.
+
+%\ptcomment{* We evaluate the expressive power of  arrow parallelism. Showing that 
+%  + there is no need to introduce additional types
+%  + that it is possible to define algorthmic skeletons (Section 6,7)
+%  + ...
+
+%* We demonstrate that Arrow Parallelism can exploit multiple parallel %Haskell implementations, and with low overhead, i.e. less than X%. 
+%... (Section 8) }
 

### 2017-09-25 paper/content/parallelHaskells.tex
wip restructuring the paper, wip on motivation's contribution

@@
-\begin{flushright}
-
-\end{flushright}\section{Background}
-\label{sec:background}
 \subsection{Short introduction to parallel Haskells}
@@
 There are already several ways to write parallel programs in Haskell.
-As we base our parallel arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
+As we base our parallel Arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
 
@@
 %parallel Haskells.
-As a demonstration, we implement here the non-Arrows |parEvalN| in multiple parallel Haskells.
+We implement this non-Arrow version here which will later in this paper be adapted for usage in our Arrow based parallel Haskell.
+Furthermore, the implementation of |parEvalN| serves as a short demonstration of how the used parallel Haskell backends work.
 

### 2017-09-25 paper/content/motivation.tex
wip motivation

@@
 This is a big advantage as we do not introduce any new types as Monad solutions like the |Par| Monad do.
-We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change integral parts of the code.
+We can just \enquote{plug} in parallel parts into sequential Arrow-based programs and libraries without having to change integral parts of the code.
+\mbcomment{mention Functions here?
+We have everything Eden has, but more.}
 
@@
 
-We evaluate the expressive power of arrow parallelism further by showing that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}).
+We also show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced, i.e.
+less than x\% \mbcomment{fix this after benchmarks are finally done} as shown in Section \ref{sec:benchmarks}.
 
-Finally, we demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced, i.e.
-less than x\% \mbcomment{fix this after benchmarks are finally done} in Section \ref{sec:benchmarks}.
+\mbcomment{Mention Future work here?}
 
@@
 
-\paragraph{Structure}
-The remaining text is structures as follows.
+%\paragraph{Structure}
+%The remaining text is structures as follows.
 Section~\ref{sec:background} briefly introduces known parallel Haskell flavours (Sec.~\ref{sec:parEvalNIntro}) and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).

### 2017-09-25 paper/content/parallelHaskells.tex
tmp

@@
 \label{sec:parEvalNIntro}
-There are already several ways to write parallel programs in Haskell.
+There are already several ways to write parallel programs in Haskell:
+
+\begin{table}[]
+\centering
+\caption{Overview over some parallel Haskells}
+\label{my-label}
+\begin{tabular}{\pipe l \pipe l \pipe l \pipe} 
+Name		& Memory setting & Concepts used \\
+GpH         & shared \& distributed & |par|/|pseq| "hints" \\
+|Par| Monad & shared & deterministic parallel Monad \\
+Eden        & shared \& distributed & special compiler, pure \& monadic interface \\
+HdpH        & \mbcomment{shared?} \& distributed & parallel Monad, Template Haskell \\
+LVish       & shared & parallel Monad, parallel data-structures\\
+\end{tabular}
+\end{table}
+
 As we base our parallel Arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
@@
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1998a} is a way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|.
-Let:
+Multicore Haskell \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
 

### 2017-09-25 paper/content/lhs2TeXSafeCommands.tex
wip on parallelHaskells.tex (added overview table)

\newcommand{\parallelHaskellOverview}{
\begin{table}
\caption{Overview over some parallel Haskells}
\centering
%% something was wrong with separators in table
\renewcommand{\tabcolsep}{1em}
\begin{tabular}{ccc}
\toprule
Name & Memory setting & Concepts used \\
\midrule
GpH         & shared \& distributed & \ensuremath{par}/\ensuremath{pseq} "hints" \\
\ensuremath{Par} Monad & shared & deterministic parallel Monad \\
Eden        & shared \& distributed & special compiler, pure \& monadic interface \\
HdpH        & \mbcomment{shared?} \& distributed & parallel Monad, Template Haskell \\
LVish       & shared & parallel Monad, parallel data-structures \\
\bottomrule
\end{tabular}
\end{table}
}

### 2017-09-25 paper/content/parallelHaskells.tex
wip on parallelHaskells.tex (added overview table)

@@
 
-\begin{table}[]
-\centering
-\caption{Overview over some parallel Haskells}
-\label{my-label}
-\begin{tabular}{\pipe l \pipe l \pipe l \pipe} 
-Name		& Memory setting & Concepts used \\
-GpH         & shared \& distributed & |par|/|pseq| "hints" \\
-|Par| Monad & shared & deterministic parallel Monad \\
-Eden        & shared \& distributed & special compiler, pure \& monadic interface \\
-HdpH        & \mbcomment{shared?} \& distributed & parallel Monad, Template Haskell \\
-LVish       & shared & parallel Monad, parallel data-structures\\
-\end{tabular}
-\end{table}
+\parallelHaskellOverview
 

### 2017-09-25 paper/content/parallelHaskells.tex
wip on restructuring of the paper according to Phils first notes

--- 2017-09-25 paper/content/parallelHaskells.tex
@@
 \label{sec:parEvalNIntro}
-There are already several ways to write parallel programs in Haskell.
-As we base our parallel Arrows on existing parallel Haskells, we will now give a short introduction to the ones we use as backends in this paper.
-
 In its purest form, parallel computation (on functions) can be looked at as the execution of some functions |a -> b| in parallel or |parEvalN :: [a -> b] -> [a] -> [b]|, as also Figure~\ref{fig:parEvalN} symbolically shows.
@@
 \end{figure}
+
+In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow based parallel Haskell.
+
+There exist several ways to implement general parallelism in Haskell.
+Among the most important are probably GpH  (based on |par| \& |pseq| "hints") \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \cite{eden,Loogen2012} (a parallel Haskell for distributed memory), HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
+
+As the goal of this paper is not to reimplement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface.
+For this paper we chose GpH - for its simplicity, the |Par| Monad - to represent a monadic DSL, and Eden - as a distributed parallel Haskell.
+
+LVish and HdpH were not chosen as the former does not differ from the original |Par| Monad with regard to how we would have used it in this paper, while the latter (at least in its current form) does not comply with our representation of parallelism due to its heavy reliance on TemplateHaskell.
+
+We will now go into some detail on GpH, the |Par| Monad and Eden, and also give their respective implementations of the non-Arrow version of |parEvalN|.
+
 %Before we go into detail on how we can use this idea of parallelism
@@
 %parallel Haskells.
-We implement this non-Arrow version here which will later in this paper be adapted for usage in our Arrow based parallel Haskell.
-Furthermore, the implementation of |parEvalN| serves as a short demonstration of how the used parallel Haskell backends work.
 
 \subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1998a} is a way to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} It ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|.
-Let:
+Multicore Haskell \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
 

### 2017-09-26 paper/content/benchmarks.tex
minor
benchmarks missing for now.
replaced "Multicore Haskell" with "GpH"

@@
 \newcommand{\sudokutest}{Sudoku\xspace}
-\newcommand{\jacobitest}{Jacobi sum test\xsp}
+\newcommand{\jacobitest}{Jacobi sum test\xspace}
 \newcommand{\torustest}{Gentleman\xspace}
@@
 A~parallel Sudoku solver was used by \citet{par-monad} to compare |Par| Monad
-to Multicore Haskell. We ported it to PArrows.
+to GpH Haskell. We ported it to PArrows.
 
@@
 
-The \ensuremath{\Conid{Par}} monad and Multicore Haskell can be executed on a shared
+The \ensuremath{\Conid{Par}} monad and GpH Haskell can be executed on a shared
 memory machines only. Although GpH is available on distributed memory
@@
 for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
-hyperthreaded cores and only 13.55 with PArrows/Multicore
-Haskell. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
+hyperthreaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
 version used here was Eden~CP, the ``share nothing'' SMP build.
@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, Multicore, |Par| Monad}{16}{4}{
+\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 The performance of PArrows/Eden~CP in shared memory is slightly better than
-for SMP variants such as PArrows/Multicore Haskell and PArrows/|Par|
-Monad but most of the time the performance is still comparable with the Multicore backend performing slightly worse than the other two in terms of speedup.
+for SMP variants such as PArrows/GpH and PArrows/|Par|
+Monad but most of the time the performance is still comparable with the GpH backend performing slightly worse than the other two in terms of speedup.
 In particular, the speedups for 16 cores fall behind quite noticeably with 11.69 for 32 tasks as Eden and the |Par| Monad both score 13.05 for 32 tasks.
-While the Multicore backend scores a bit better in the 64 tasks benchmark, its speedup of 12.70 is still behind the 13.65 and 13.33 of Eden and |Par| Monad, respectively.
+While the GpH backend scores a bit better in the 64 tasks benchmark, its speedup of 12.70 is still behind the 13.65 and 13.33 of Eden and |Par| Monad, respectively.
 
@@
 
+\renewcommand{\speedupJacobiDist}[4]{
+	\textcolor{red}{benchmark missing. in the works}
+}
+
 \newcommand{\performanceJacobiDistDiff}[5]{
@@
 }{#4}{#5}{\plotwidthDist}
+}
+
+\renewcommand{\performanceJacobiDistDiff}[5]{
+	\textcolor{red}{benchmark missing. in the works}
 }
@@
 
+\renewcommand{\speedupTorusDist}[3]{
+	\textcolor{red}{benchmark missing. in the works}
+}
+
 \newcommand{\performanceTorusDistDiff}[4]{
@@
 }{#3}{#4}{\plotwidthDist}
+}
+
+\renewcommand{\performanceTorusDistDiff}[4]{
+		\textcolor{red}{benchmark missing. in the works}
 }
@@
 Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark}-\ref{fig:sudokuSMBenchmark16000}.
-However, we notice that the Multicore backend seems to choke on a bigger input \ref{fig:sudokuSMBenchmark16000}.
+However, we notice that the GpH backend seems to choke on a bigger input \ref{fig:sudokuSMBenchmark16000}.
 This is due to the benchmark only using |parMap| instead of a chunking variant (however we did not change that for simplicity's sake) and is reflected by debug output which shows, that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
@@
 \newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{32}{4}{
+\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 %smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
-}{33}{\plotwidthSMP}
+}{17}{\plotwidthSMP}
 }
@@
 \newcommand{\speedupSudokuSM}[1]{
-\speedupplot{ \sudokutest ``#1''}{Eden CP, Multicore, |Par| Monad}{32}{4}{
+\speedupplot{ Parallel speedup of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
   grid; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
-  closer to linear speedup the better. The Multicore version shows signs of choking with too many sparks being created.}
+  closer to linear speedup the better. The GpH version shows signs of choking with too many sparks being created.}
 \label{fig:sudokuSMBenchmark16000}

### 2017-09-26 paper/content/conclusion.tex
minor
replaced "Multicore Haskell" with "GpH"

@@
 Arrows are a generic concept that allows for powerful composition
-combinators. To our knowledge we are the first ones to represent
+combinators. To our knowledge we are first to represent
 \emph{parallel} computation with Arrows.
@@
 This work features multiple parallel backends: the already available parallel Haskell flavours.
-Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| typeclass for Multicore Haskell, |Par| Monad, and Eden.
+Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| typeclass for GpH Haskell, |Par| Monad, and Eden.
 With our approach parallel programs can be ported across these flavours with little to no effort.
@@
 %
-The benefit is, however, the greatly increased portability of parallel
-programs.
+The benefit is, however, the greatly increased portability of parallel programs.
 
-\mbcomment{mention ArrowLoop in Torus and Ring chapters}
+%\mbcomment{mention ArrowLoop in Torus and Ring chapters}
 %\olcomment{Parrows + accelerate = love?} \olcomment{Metion port to Frege.}

### 2017-09-26 paper/content/futures.tex
replaced "Multicore Haskell" with "GpH"

@@
 Since |RD| is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows.
-%This is due to the same reason we had to introduce a wrapper for |Strategy a| in the Multicore Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
+%This is due to the same reason we had to introduce a wrapper for |Strategy a| in the GpH Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
 Technical details are in Appendix, in Section~\ref{app:omitted}.
 
-For our |Par| Monad and Multicore Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}) which are just simple wrappers around the actual data, as in a shared memory setting we do not require Eden's sophisticated communication channels.
+For our |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}) which are just simple wrappers around the actual data, as in a shared memory setting we do not require Eden's sophisticated communication channels.
 \begin{figure}[tb]
@@
 \end{code}
-\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and Multicore Haskell backends.}
+\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and GpH Haskell backends.}
 \label{fig:BasicFuture}

### 2017-09-26 paper/content/motivation.tex
replaced "Multicore Haskell" with "GpH"

@@
 We regard here in-depth
-the Multicore Haskell (a SMP implementation of GpH language), the
+the GpH Haskell (a SMP implementation of GpH language), the
 |Par| Monad, and Eden, a distributed memory parallel Haskell. These
@@
 is defined as a common interface and varying implementations in
-existing parallel Haskells: Multicore Haskell, |Par| Monad, and
+existing parallel Haskells: GpH Haskell, |Par| Monad, and
 Eden. Thus, we not only define a parallel programming interface in a

### 2017-09-26 paper/content/parallelHaskells.tex
minor
replaced "Multicore Haskell" with "GpH"

@@
 	\includegraphics[scale=0.7]{images/parEvalN}
-	\caption{Schematic illustration of |parEvalN|.}
+	\caption{Schematic illustration of |parEvalN|. A list of inputs is transformed by different functions in parallel.}
 	\label{fig:parEvalN}
@@
 
-There exist several ways to implement general parallelism in Haskell.
+There exist several parallel Haskells already.
 Among the most important are probably GpH  (based on |par| \& |pseq| "hints") \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \cite{eden,Loogen2012} (a parallel Haskell for distributed memory), HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
@@
 
-\subsubsection{Multicore Haskell}
-Multicore Haskell \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{Multicore Haskell on Hackage is available under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
+\subsubsection{Glasgow parallel Haskell - GpH}
+GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{An implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
 
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-	\caption{Dataflow of the Multicore Haskell |parEvalN| version.}
+	\caption{Dataflow of the GpH |parEvalN| version.}
 	\label{fig:parEvalNMulticoreImg}
@@
 \subsubsection{Par Monad}
-The |Par| Monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs.
+The |Par| Monad\footnote{The |Par| monad can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs.
 Let:
@@
 
-The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell.
+The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with GpH.
 % $
@@
 If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow.
-The typical reason for thus is GC.
+The typical reason for this is GC.
 An inactive machine, where no processes are started yet, or all are already terminated, shows as a blue bar.

### 2017-09-26 paper/content/parrows.tex
minor update
replaced "Multicore Haskell" with "GpH"

@@
 If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|.
-% , as we will see in the implementation of the Multicore Haskell and the |Par| Monad backends.
+% , as we will see in the implementation of the GpH Haskell and the |Par| Monad backends.
 
@@
 
-\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
-The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
+\subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
+The GpH implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| combinators from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
 \begin{figure}[t]
@@
 \end{code}% $ %% formatting
-\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
+\caption{Fully evaluating |ArrowParallel| instance for the GpH Haskell backend.}
 \label{fig:ArrowParallelMulticoreRdeepseq}
 \end{figure}
-For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this.
+For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the GpH interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this.
 We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
+With this definition in place, we can provide a delegating version of the non-configurable instance.
+We show this in Fig.~\ref{fig:actualNonConfigMulticore}.
 % \begin{figure}[h]
@@
 \end{code}
-\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
+\caption{Configurable |ArrowParallel| instance for the GpH Haskell backend.}
 \label{fig:ArrowParallelMulticoreConfigurable}
@@
 \subsubsection{Eden}
-For both the Multicore Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses.
+For both the GpH Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses.
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.

### 2017-09-26 paper/content/topologySkeletons.tex
replaced "Multicore Haskell" with "GpH"

@@
 
-If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the Multicore and Par Monad backends which during runtime would result in the program crashing.
+If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing.
 This materializes with the usage of |loop| of the |ArrowLoop| typeclass and is probably due to the way their respective parallelism engines work internally.
@@
 \end{code}
-The Par Monad and Multicore backends have parallelism disabled in their instance of |FutureEval|.
+The Par Monad and GpH backends have parallelism disabled in their instance of |FutureEval|.
 This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.

### 2017-09-26 paper/content/arrows.tex
Arrows chapter. Phil's remarks

@@
 %
-Its |arr| operation is used to lift an ordinary function to the specified arrow type, similarly to the monadic |return|.
+Its |arr| operation is used to lift an ordinary function to the specified arrow type, analogous to the monadic |return|.
 The |>>>| operator is analogous to the monadic composition  |>>=| and combines two arrows |arr a b| and |arr b c| by "wiring" the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
@@
 The most prominent instances of this interface are regular functions |(->)|
-and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}).
+and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps functions that return monadic values, e.g.  |a -> m b| with |m| being a Monad.
 
@@
 \end{figure}
-With this typeclass in place, Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): The combinators |second|, |***| and |&&&|. 
+Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): The combinators |second|, |***| and |&&&|. 
 The combinator |second| is the mirrored version of |first| (Appendix~\ref{utilfns}).
@@
 \begin{code}
-(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
+(&&&) :: Arrow arr => arr a b -> arr a c -> arr a (b, c)
 f &&& g = arr (\a -> (a, a)) >>> (f *** g)
@@
 The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a Monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}.
-One of the major problems in parallel computing is, however, composition of parallel processes.
+This is exactly one of the key challenges in parallel computing.
 

### 2017-09-26 paper/content/motivation.tex
further restructuring according to Phil's remarks. moved related work to 2

@@
 We regard here in-depth
-the GpH Haskell (a SMP implementation of GpH language), the
+the GpH (a SMP implementation of GpH language), the
 |Par| Monad, and Eden, a distributed memory parallel Haskell. These
@@
 is defined as a common interface and varying implementations in
-existing parallel Haskells: GpH Haskell, |Par| Monad, and
+existing parallel Haskells: GpH, |Par| Monad, and
 Eden. Thus, we not only define a parallel programming interface in a
@@
 
-We host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic in a way that the backend easily be swapped.
-This means it is possible to write code against the interface of a common typeclass without being bound to any specific backend.
-So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance.
+\mbcomment{Überleitung auf Related Work hier machen}
 
-We also show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced, i.e.
+We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic.
+This way, we can have multiple backends - currently a GpH, a |Par| Monad and an Eden version are supported - backends can easily be swapped, so we are not bound to any specific one.
+So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.
+
+We show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced, i.e.
 less than x\% \mbcomment{fix this after benchmarks are finally done} as shown in Section \ref{sec:benchmarks}.

### 2017-09-26 paper/content/parallelHaskells.tex
further restructuring according to Phil's remarks. moved related work to 2

@@
 \subsubsection{Glasgow parallel Haskell - GpH}
-GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{An implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
+GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{The Multicore implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
 
@@
 \subsubsection{Par Monad}
-The |Par| Monad\footnote{The |Par| monad can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs.
+The |Par| Monad\footnote{The |Par| monad can be found in the \texttt{monad-par} package on Hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs.
 Let:
@@
 \subsubsection{Eden}
-Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{See also \url{http://www.mathematik.uni-marburg.de/~eden/} and \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} It is targeted towards clusters, but also functions well in a shared-memory setting with a further simple backend.
+Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{The projects homepage can be found at \url{http://www.mathematik.uni-marburg.de/~eden/}.
+The Hackage page is at \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} It is targeted towards clusters, but also functions well in a shared-memory setting with a further simple backend.
 However, in contrast to many other parallel Haskells, in Eden each process has its own heap.

### 2017-09-26 paper/content/relwork.tex
further restructuring according to Phil's remarks. moved related work to 2

@@
 \paragraph{Parallel Haskells.}
-Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
+Of course, the three parallel Haskell flavours we use as backends: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
 We use these languages as backends: our DSL can switch from one to other at user's command.
@@
 But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
-\mbcomment{reference not found!!!}
 \citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimizations.

### 2017-09-27 paper/utils/redo.sh
torus redone

@@ -16,14 +16,14 @@ cp bench-distributed.bench.* ../content/benchmarks/distributed-rm
 cp bench-torus.bench.* ../content/benchmarks/torus
 
 #./calculateDifferences bench-torus-distributed.bench.torus-matrix-eden-32-512-64-200.csv bench-torus-distributed.bench.torus-matrix-parrows-32-512-64-200.csv 32-512-64-200-diff.csv
-./calculateDifferences bench-torus-distributed.bench.torus-matrix-eden-1024.csv bench-torus-distributed.bench.torus-matrix-parrows-1024.csv 1024-diff.csv
+./calculateDifferences bench-torus-distributed.bench.torus-matrix-eden-4096.csv bench-torus-distributed.bench.torus-matrix-parrows-4096.csv 4096-diff.csv
 
 ./calculateDifferences bench-jacobi.bench.jacobi-eden-3-3217.csv bench-jacobi.bench.jacobi-parr-3-3217.csv 3-3217-diff.csv
 #./calculateDifferences bench-distributed.bench.skelrm-eden-11213-256.csv bench-distributed.bench.skelrm-parrows-11213-256.csv 11213-256-diff.csv
 
 ./calculateDifferences bench-distributed.bench.skelrm-eden-44497-256.csv bench-distributed.bench.skelrm-parrows-44497-256.csv 44497-256-diff.csv
 
-cp 1024-diff.csv ../content/benchmarks/distributed-torus
+cp 4096-diff.csv ../content/benchmarks/distributed-torus
 #cp 32-1024-64-200-diff.csv ../content/benchmarks/distributed-torus
 
 cp 3-3217-diff.csv ../content/benchmarks/distributed-jacobi

### 2017-09-28 paper/content/futures.tex
motivation and futures rework

@@
-\section{Futures} \label{futures}
+\section{Futures} \label{sec:futures}
 Consider a mock-up parallel arrow combinator:
@@
 
-While the above example could be rewritten into only one |parEvalN| call by directly wiring the arrows properly together, it illustrates an important problem.
+While the above example could be rewritten into only one |parEvalN| call by directly wiring the arrows together before spawning, it illustrates an important problem.
 When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
@@
 
-We should allow the nodes to communicate directly with each other. Eden already ships with "remote data" that enable this \cite{AlGo03a,Dieterle2010}.
+This is in fact only a problem in distributed memory (in the scope of this paper) and we should  allow nodes to communicate directly with each other.
+Eden already ships with "remote data" that enable this \cite{AlGo03a,Dieterle2010}.
 But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
@@
 
-For our |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}) which are just simple wrappers around the actual data, as in a shared memory setting we do not require Eden's sophisticated communication channels.
+For the |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the typeclass is satisfied.
+This is because the concept of a |Future| does not change anything for shared-memory execution as there are no communication problems to fix, but we require a common interface so the parallel Arrows are portable across backends.
 \begin{figure}[tb]
@@
 
-In our communication example we can use this |Future| concept for direct communications between the nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
+In our communication example we can use this |Future| concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
 \begin{figure}[tbh]
@@
 \end{figure}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed---similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.\olcomment{Fig.
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed---similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.
+\olcomment{Fig.
 is not really clear.

### 2017-09-28 paper/content/motivation.tex
motivation and futures rework

@@
 %\olcomment{move this to Contributions in the front or something}
-We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of Arrows into a new parallel Arrow.
+We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow.
 Because of this, we do not lose any benefits of using Arrows as the parallelism is encapsulated as another combinator instead of a different type.
@@
 
-We show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced, i.e.
-less than x\% \mbcomment{fix this after benchmarks are finally done} as shown in Section \ref{sec:benchmarks}.
+We introduce a |Future| typeclass in Section \ref{sec:futures} to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|).
+Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node.
+As the |Par| Monad and GpH backends (in their current form) only support shared memory execution and do not have to fix any communication problems, we only use a wrapping dummy |BasicFuture|.
+
+We show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced in Section \ref{sec:benchmarks}, i.e.
+less than x\% \mbcomment{fix this after benchmarks are finally done}.
 
 \mbcomment{Mention Future work here?}
+
+\mbcomment{Conclusion?}
 

### 2017-09-29 paper/content/motivation.tex
.
minor  changes in Introduction

@@
 
-One particular reason for this is the ease of introducing new sophisticated language
-concepts.
 Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms including the expression of parallelism.
@@
 We regard here in-depth
-the GpH (a SMP implementation of GpH language), the
+Glasgow parallel Haskell or short GpH (its Multicore SMP implementation, in particular), the
 |Par| Monad, and Eden, a distributed memory parallel Haskell. These
@@
 language constructs.
+
 %In this paper we introduce a notion of parallel
 %computations using Arrows.
-A key novelty here is to use Arrows to represent parallel computations. They seem a natural fit as they are a generalization of the function |->| and serve as general interface to computations.
+A key novelty in this paper is to use Arrows to represent parallel computations.
+They seem a natural fit as they are a generalization of the function |->| and serve as general interface to computations.
 
 %Our Arrows-based interface is a high-level one.
-We provide an Arrows-based typeclass and implementations for three parallel Haskells.
+We provide an Arrows-based typeclass and implementations for the three above mentioned parallel Haskells.
 Instead of 
@@
 Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
-is defined as a common interface and varying implementations in
-existing parallel Haskells: GpH, |Par| Monad, and
-Eden. Thus, we not only define a parallel programming interface in a
+is defined as a common interface with varying implementations in
+the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
 novel manner - we tame the zoo of parallel Haskells. We provide a
 common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
-the parallel Haskell backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
+the parallel backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
 

### 2017-09-29 paper/content/arrows.tex
reformulations, minor things

@@
 \vfill
-\caption{The definition of |Arrow| type class and its two most typical instances.}
+\caption{The definition of the |Arrow| type class and its two most typical instances.}
 \label{fig:ArrowDefinition}
@@
 }
-\caption{Schematic depiction of Arrow (left) and its basic
+\caption{Schematic depiction of  an Arrow (left) and its basic
   combinators |arr|, |>>>| and |first| (right).}
@@
 %
-Its |arr| operation is used to lift an ordinary function to the specified arrow type, analogous to the monadic |return|.
-The |>>>| operator is analogous to the monadic composition  |>>=| and combines two arrows |arr a b| and |arr b c| by "wiring" the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
-Lastly, the |first| operator  takes the input arrow from |b| to |c| and converts it into an arrow on pairs with the second argument untouched.
+To lift an ordinary function to an Arrow, |arr| is used, analogous to the monadic |return|.
+Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by "wiring" the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
+Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched.
 It allows us to to save input across arrows.
-Figure~\ref{fig:arrows-viz} shows a graphical representation of the basic Arrow combinators.
+Figure~\ref{fig:arrows-viz} shows a graphical representation of these basic Arrow combinators.
 The most prominent instances of this interface are regular functions |(->)|
-and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps functions that return monadic values, e.g.  |a -> m b| with |m| being a Monad.
+and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps monadic functions, e.g.  |a -> m b| (with |m| being a Monad).
 
@@
 \end{figure}
-Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): The combinators |second|, |***| and |&&&|. 
-The combinator |second| is the mirrored version of |first| (Appendix~\ref{utilfns}).
-The |***| function combines |first| and |second| to handle two inputs in one arrow, is defined as
+Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): |second|, |***| and |&&&|. 
+|second| is the mirrored version of |first| (Appendix~\ref{utilfns}).
+|***| combines |first| and |second| to handle two inputs in one arrow, and is defined as follows:
 % \begin{figure}[h]
@@
 % \end{figure}
-while the |&&&| combinator, that constructs an arrow which outputs two different values like |***|, but takes only one input, is:
+The |&&&| combinator, which constructs an Arrow that outputs two different values like |***|, but takes only one input, is:
 % \begin{figure}[h]

### 2017-09-29 paper/content/relwork.tex
reformulations, minor things

@@
 Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution.
-In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) Arrow out of list of Arrows.\citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach.
-They basically use Arrows to orchestrate several tasks in robotics.
+In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) Arrow out of list of Arrows.
+\citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach.
+They use Arrows to orchestrate several tasks in robotics.
 We, however, propose a general interface for parallel programming, while remaining completely in Haskell.
@@
 \paragraph{Other languages.}
-Although this work is centred on Haskell implementation of arrows, it is applicable to any functional programming language where parallel evaluation and arrows can be defined.
+Although this work is centered on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined.
 Experiments with our approach in Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM) were quite successful.

### 2017-10-12 paper/content/motivation.tex
Verbesserung bis relwork.tex

@@
 only for the internal representation. Some introduce additional
-language constructs.
+language constructs. Section \ref{sec:parallelHaskells} gives a short overview over these languages.
 
@@
 A key novelty in this paper is to use Arrows to represent parallel computations.
-They seem a natural fit as they are a generalization of the function |->| and serve as general interface to computations.
+They seem a natural fit as they are a generalization of the function arrow (|->|) and serve as general interface to computations.
+Section \ref{sec:arrows} gives a short introduction to Arrows.
 
@@
 We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic.
-This way, we can have multiple backends - currently a GpH, a |Par| Monad and an Eden version are supported - backends can easily be swapped, so we are not bound to any specific one.
+This way, we can have multiple backends - currently a GpH, a |Par| Monad and an Eden version are supported.
+The backends can easily be swapped, so we are not bound to any specific one.
 So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.

### 2017-10-12 paper/content/benchmarks.tex
minor
torus benchmarks updated
minor things

@@
 sources. Most of them are parallel mathematical computations, initially
-implemented in Eden. Table~\ref{tab:benches} summarizes.
+implemented in Eden. Table~\ref{tab:benches} summarises.
 
@@
 %% something was wrong with separators in table
-\renewcommand{\tabcolsep}{1em}
+\renewcommand{\tabcolsep}{0.5em}
 \begin{tabular}{lccll}
@@
 of Jacobi sum test have very different durations. \citet{lobachev-phd}
-\olcomment{discuss some optimisations of parallel APRCL.} Generic parallel
-implementation of \rmtest and APRCL were presented in \citet{Lobachev2012}.
+discusses some optimisations of parallel APRCL. Generic parallel
+implementations of \rmtest and APRCL were presented in \citet{Lobachev2012}.
 
@@
 
-The \ensuremath{\Conid{Par}} monad and GpH Haskell can be executed on a shared
+The \ensuremath{\Conid{Par}} monad and GpH Haskell - in its multicore version (\cite{Marlow2009}) -  can be executed on a shared
 memory machines only. Although GpH is available on distributed memory
@@
 that merely copies the memory blocks between distributed heaps. In
-this mode Eden still operates in the ``nothing shared'' setting, but
+this mode, Eden still operates in the ``nothing shared'' setting, but
 is adapted better to multicore machines. We label this version of Eden
@@
 parallelism. For this reason, we did not regard hyperthreaded cores in
-our speedup plots in Figs.~\ref{fig:bench-rm-sm}-\ref{fig:sudokuSMBenchmark}.
+our speedup plots in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark}.
 
@@
 PArrows and the corresponding backend in a separate plot
-(Figs.~\ref{fig:bench-rm-dist}, \ref{fig:torusBenchmark}, \ref{fig:bench-jacobi-dist}). As an example, the differences range in
+(Figs.~\ref{fig:bench-rm-dist}--\ref{fig:torusBenchmark}). As an example, the differences range in
 about 0.5~seconds for the execution time of 46~seconds on 256 cores
@@
 \newcommand{\speedupSkelRMDist}[4]{
-\speedupplot{Speedup of \rmtest ``#2''}{PArrows}{256}{#3}{
+\speedupplot{Speedup of \rmtest ``#1 #2''}{PArrows}{256}{#3}{
 % \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 \newcommand{\performanceSkelRMDistDiff}[5]{
-\performancediffplot{Run time differences\\for \rmtest ``#2''}{(Eden $-$ PArrows)}{256}{#3}{
+\performancediffplot{Run time differences\\for \rmtest ``#1 #2''}{(Eden $-$ PArrows)}{256}{#3}{
 \addplot+[mark=*,very thick,error bars/.cd,
@@
 %
-{\speedupSkelRMDist{44497}{256}{32}{260}\label{subfig:rm-dist-a}}%
+{\speedupSkelRMDist{44497}{256}{32}{272}\label{subfig:rm-dist-a}}%
 %\hfill%
-{\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{260}\label{subfig:rm-dist-b}}
+{\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{272}\label{subfig:rm-dist-b}}
 \caption{Parallel performance of \rmtest on the Glasgow grid
@@
 Monad but most of the time the performance is still comparable with the GpH backend performing slightly worse than the other two in terms of speedup.
-In particular, the speedups for 16 cores fall behind quite noticeably with 11.69 for 32 tasks as Eden and the |Par| Monad both score 13.05 for 32 tasks.
-While the GpH backend scores a bit better in the 64 tasks benchmark, its speedup of 12.70 is still behind the 13.65 and 13.33 of Eden and |Par| Monad, respectively.
 
 Comparing the PArrows version of the \rmtest with the original from Eden with the MPI backend in a distributed memory setting, we see an almost linear speedup of
-\rmtest with 256 tasks and input $2^{444497}-1$ in both versions. The sequential run time
+\rmtest with 256 tasks and input $2^{44497}-1$ in both versions. The sequential run time
 was computed as the mean of three consecutive executions on a single
-core---the single run took two hours 43 minutes. The zero difference  between
-PArrows/Eden and Eden almost always lies on the error bar of
-the measurement. The only exception, where the PArrows and Eden versions differ and it was it was significant, was for 64 cores, where it performed 0.49s better. This corresponds to
-0.30\% relative time difference. Otherwise, PArrows was 0.23s (or 0.0007\%) slower for 32 cores, 0.26s (or 0.0030\%) slower for 128 cores, and 0.09s faster (or 0.0019\%) faster for 256 cores.
+core---the single run took two hours 43 minutes. The difference between
+PArrows/Eden and Eden almost always lies inside the error bar of
+the measurement.
 
@@
 
-\renewcommand{\speedupJacobiDist}[4]{
-	\textcolor{red}{benchmark missing. in the works}
-}
-
 \newcommand{\performanceJacobiDistDiff}[5]{
@@
 
-\renewcommand{\performanceJacobiDistDiff}[5]{
-	\textcolor{red}{benchmark missing. in the works}
-}
-
 \begin{figure}
@@
 %
-{\speedupJacobiDist{3}{1279}{32}{260}\label{subfig:rm-dist-a}}%
+{\speedupJacobiDist{3}{3217}{32}{272}\label{subfig:rm-dist-a}}%
 %\hfill%
-{\performanceJacobiDistDiff{3}{1279}{32}{0.5}{260}\label{subfig:rm-dist-b}}
+{\performanceJacobiDistDiff{3}{3217}{32}{0.5}{272}\label{subfig:rm-dist-b}}
 \caption{Parallel performance of \rmtest on the Glasgow grid
-  consisting of 256 cores. Input was $2^{1279}-1$, we used 256
+  consisting of 256 cores. Input was $2^{3217}-1$, we used 256
   tasks. The top plot shows absolute speedup in a distributed memory setting. The
@@
 Next is the \torustest benchmark.
-The results of the comparison of vanilla Eden to our PArrows-based version can be found in \ref{fig:torusBenchmark}.
+The results of the comparison of vanilla Eden to our PArrows-based version can be found in Fig.~\ref{fig:torusBenchmark}.
 While we do not see too much speedup for more than 16 cores, we can prove that the difference between the Eden and PArrows version are again marginal.
-The difference between PArrows and Eden is only significant for 16 and 64 cores where it ran 1.7\% and 2.7\% slower which corresponds to a real-time difference of 0.12s and 0.13s.
+%The difference between PArrows and Eden is only significant for 16 and 64 cores where it ran 1.7\% and 2.7\% slower which corresponds to a real-time difference of 0.12s and 0.13s.
 For 256 cores PArrows performed 0.2\% slower which corresponds to 0.01s overhead.
@@
 }{#3}{\plotwidthDist}
-}
-
-\renewcommand{\speedupTorusDist}[3]{
-	\textcolor{red}{benchmark missing. in the works}
 }
@@
 
-\renewcommand{\performanceTorusDistDiff}[4]{
-		\textcolor{red}{benchmark missing. in the works}
-}
-
 \begin{figure}
@@
 %
-{\speedupTorusDist{1024}{32}{260}\label{subfig:speedupTorusDist}}%
+{\speedupTorusDist{4096}{32}{272}\label{subfig:speedupTorusDist}}%
 %\hfill%
-{\performanceTorusDistDiff{1024}{32}{0.5}{260}\label{subfig:performancetorusDistDiff}}
+{\performanceTorusDistDiff{4096}{32}{0.5}{272}\label{subfig:performancetorusDistDiff}}
 \caption{Parallel performance of \torustest on the Glasgow grid
-  consisting of 256 cores. Input was a matrix size of $1024$. The top plot shows absolute speedup in a distributed memory setting. The
+  consisting of 256 cores. Input was a matrix size of $4096$. The top plot shows absolute speedup in a distributed memory setting. The
   closer to linear speedup the better. Time
@@
 
-As the last Benchmark in this paper we present the \sudokutest in \ref{fig:sudokuSMBenchmark} running in a shared memory setting.
-Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark}-\ref{fig:sudokuSMBenchmark16000}.
-However, we notice that the GpH backend seems to choke on a bigger input \ref{fig:sudokuSMBenchmark16000}.
-This is due to the benchmark only using |parMap| instead of a chunking variant (however we did not change that for simplicity's sake) and is reflected by debug output which shows, that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
-Another remarkable finding is that the Eden backend seems to lack behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.
+As the last benchmark in this paper we present the \sudokutest in Fig.~\ref{fig:sudokuSMBenchmark} running in a shared memory setting.
+Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark} and \ref{fig:sudokuSMBenchmark16000}.
+However, we notice that the GpH backend seems to choke on a bigger input (Fig.~\ref{fig:sudokuSMBenchmark16000}).
+This is due to the benchmark only using |parMap| instead of a chunking variant -- however we did not change that for simplicity's sake.
+This issue is reflected by debug output which shows that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
+Another remarkable finding is that the Eden backend seems to lag behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.
 

### 2017-10-12 paper/content/futures.tex
minor things

@@
 This can become a serious bottleneck %in heavy threaded applications.
-for larger amount of data and number of processes \citep[as \eg][showcases]{Berthold2009-fft}.
+for larger amount of data and number of processes \citep[as e.g.][showcases]{Berthold2009-fft}.
 \begin{figure}[ht]
@@
 For the |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the typeclass is satisfied.
-This is because the concept of a |Future| does not change anything for shared-memory execution as there are no communication problems to fix, but we require a common interface so the parallel Arrows are portable across backends.
+This is because the concept of a |Future| does not change anything for shared-memory execution as there are no communication problems to fix.
+Nevertheless, we require a common interface so the parallel Arrows are portable across backends.
 \begin{figure}[tb]
@@
 \end{figure}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed---similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.
 \olcomment{Fig.

### 2017-10-12 paper/content/parallelHaskells.tex
Verbesserung bis relwork.tex
minor things

@@
 \subsection{Short introduction to parallel Haskells}
+\label{sec:parallelHaskells}
 \label{sec:parEvalNIntro}
@@
 
-In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow based parallel Haskell.
+In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow-based parallel Haskell.
 
 There exist several parallel Haskells already.
-Among the most important are probably GpH  (based on |par| \& |pseq| "hints") \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \cite{eden,Loogen2012} (a parallel Haskell for distributed memory), HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
+Among the most important are probably GpH  (based on |par| \& |pseq| "hints") \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden (a parallel Haskell for distributed memory) \cite{eden,Loogen2012}, HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
 
 As the goal of this paper is not to reimplement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface.
-For this paper we chose GpH - for its simplicity, the |Par| Monad - to represent a monadic DSL, and Eden - as a distributed parallel Haskell.
+For this paper we chose GpH for its simplicity, the |Par| Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.
 
@@
 However, in contrast to many other parallel Haskells, in Eden each process has its own heap.
-This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results also on multicores \cite{arcs-dc,aswad2009low}.
+This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results on multicores, as well \cite{arcs-dc,aswad2009low}.
 
-While Eden also comes with a Monad |PA| for parallel evaluation, it also ships with a completely functional interface that includes
+While Eden comes with a Monad |PA| for parallel evaluation, it also ships with a completely functional interface that includes
 %\\

### 2017-10-16 paper/content/conclusion.tex
minor
wip

@@
 %
-Arrows turn out to be a useful tool for composing in parallel
+Arrows turn out to be a useful tool for composing parallel
 programs. We do not have to introduce new monadic types that wrap the
@@
 Existing combinators could also be improved.
-We would look into more transparency of our DSL,  basically it should infuse as little overhead as possible.
+Arrow-based notation might enable further compiler optimizations.
 

### 2017-10-17 paper/content/benchmarks.tex
wip
added "linear" times to sudoku plots

@@
 \newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \sudokutest ``#1''}{linear, Eden CP, GpH, |Par| Monad}{16}{4}{
+\addplot [no markers,dotted,thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
+smooth] {benchmarks/sudoku-sm/bench-sudoku-sm.bench.fake-linear-sudoku-sudoku17.#1.txt.csv};
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSudokuSM}[1]{
-\speedupplot{ Parallel speedup of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\speedupplot{Parallel speedup of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 \end{figure}
+
+\subsection{Byline}
+
+In our distributed memory tests on the full grid we were unable to detect any performance deficiencies of PArrows \wrt Eden that were above of the error margin.
+In our distributed memory tests on the full grid the difference between PArrows and Eden was always below the error margin.

### 2017-10-17 paper/content/benchmarks.tex
wip

@@
 
-The \ensuremath{\Conid{Par}} monad and GpH Haskell - in its multicore version (\cite{Marlow2009}) -  can be executed on a shared
+The \ensuremath{\Conid{Par}} monad and GpH Haskell -- in its multicore version (\cite{Marlow2009}) --  can be executed on a shared
 memory machines only. Although GpH is available on distributed memory

### 2017-10-17 paper/content/parallelHaskells.tex
wip

@@
 
-\subsubsection{Glasgow parallel Haskell - GpH}
+\subsubsection{Glasgow parallel Haskell -- GpH}
 GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{The Multicore implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.

### 2017-10-17 paper/content/utilityFunctions.tex
replaced "Multicore Haskell" with "GpH"
wip

@@
 \end{code}
-\caption{The actual default implementation of Multicore's |ArrowParallel|.}
+\caption{The actual default implementation of GpH's |ArrowParallel|.}
 \label{fig:actualNonConfigMulticore}
@@
 \end{figure}
-Furthermore, Fig.~\ref{fig:torus_example_rest} contains the ommited definitions of |prMMTr|, - which calculates $AB^T$ for two matrices $A$ and $B$, |splitMatrix| - which splits the a matrix into chunks, and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
+Furthermore, Fig.~\ref{fig:torus_example_rest} contains the ommited definitions of |prMMTr| (which calculates $AB^T$ for two matrices $A$ and $B$), |splitMatrix| (which splits the a matrix into chunks), and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
 \begin{figure}[h]

### 2017-10-17 paper/content/arrows.tex
Enquotes

@@
 To lift an ordinary function to an Arrow, |arr| is used, analogous to the monadic |return|.
-Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by "wiring" the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
+Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by \enquote{wiring} the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
 Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched.

### 2017-10-17 paper/content/parallelHaskells.tex
Enquotes

@@
 There exist several parallel Haskells already.
-Among the most important are probably GpH  (based on |par| \& |pseq| "hints") \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden (a parallel Haskell for distributed memory) \cite{eden,Loogen2012}, HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
+Among the most important are probably GpH  (based on |par| \& |pseq| \enquote{hints}) \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden (a parallel Haskell for distributed memory) \cite{eden,Loogen2012}, HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
 

### 2017-10-17 paper/content/benchmarks.tex
minor fixes

@@
 efficiency was 85.3\%. However, if we increase the number of
-requested cores to be 32---\ie if we use hyperthreading on 16 real
-cores---the speedup does not increase that well. It is merely 15.99
+requested cores to be 32 -- \ie if we use hyperthreading on 16 real
+cores -- the speedup does not increase that well. It is merely 15.99
 for 32 tasks with PArrows/Eden. It is worse for other backends.  As
@@
 hyperthreaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
-version used here was Eden~CP, the ``share nothing'' SMP build.
+version used here was Eden~CP, the \enquote{share nothing} SMP build.
 
@@
 genuine |Par|
-Monad benchmark is very small. To give an example, it is ~0.4s in favour of PArrows for 16 cores (10.8s vs. 11.2s) and ~-0.8s in favour of the |Par| monad for 8 cores (16.1s vs. 16.9s) for
+Monad benchmark is very small.
+To give an example, it is ~0.4\seconds in favour of PArrows for 16 cores (10.8\seconds vs.
+11.2\seconds) and ~-0.8\seconds in favour of the |Par| monad for 8 cores (16.1\seconds vs.
+16.9\seconds) for
 the Sudoku benchmark in the shared memory setting. It is almost invisible in speedup and
@@
 (Figs.~\ref{fig:bench-rm-dist}--\ref{fig:torusBenchmark}). As an example, the differences range in
-about 0.5~seconds for the execution time of 46~seconds on 256 cores
+about 0.5\seconds for the execution time of 46\seconds on 256 cores
 for distributed \rmtest with PArrows and Eden. For these comparisons, the plots show absolute

### 2017-10-17 paper/content/futures.tex
Enquotes
minor fixes

@@
 This is in fact only a problem in distributed memory (in the scope of this paper) and we should  allow nodes to communicate directly with each other.
-Eden already ships with "remote data" that enable this \cite{AlGo03a,Dieterle2010}.
+Eden already ships with \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
 But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
@@
 	\caption[with Futures]{Communication between 4 Eden processes with Futures.
-Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (black lines between the bars) instead of always going through the master node (bottom bar).}
+Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (black lines between the bars, one example message is marked in the Figure) instead of always going through the master node (bottom bar).}
 	\label{fig:withFutures}

### 2017-10-24 paper/content/basicSkeletons.tex
spellchecked

@@
 \label{sec:extending-interface}
-With the |ArrowParallel| typeclass in place and implemented, we can now implement some further basic parallel interface functions.
+With the |ArrowParallel| type class in place and implemented, we can now implement some further basic parallel interface functions.
 These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
@@
 
-\subsubsection{Heterogenous tasks}
+\subsubsection{Heterogeneous tasks}
 \begin{figure}[tb]
@@
 \end{figure}
-We have only talked about the paralellization arrows of the same type until now.
-But sometimes we want to paralellize heterogenous types as well.
-However, we can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| typeclass.
+We have only talked about the parallelization arrows of the same type until now.
+But sometimes we want to parallelize heterogeneous types as well.
+However, we can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class.
 The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.

### 2017-10-24 paper/content/benchmarks.tex
spellchecked

@@
 16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each.
-Each processor has 8 cores and 16 (hyperthreaded) threads with a base frequency of 2 GHz and a turbo frequency of 2.50 GHz.
+Each processor has 8 cores and 16 (hyper-threaded) threads with a base frequency of 2 GHz and a turbo frequency of 2.50 GHz.
 This results in a total of 256 cores and 512 threads for the whole cluster.
 The operating system was Ubuntu 14.04 LTS with Kernel 3.19.0-33.
-Non-surprisingly, we found that hyperthreaded 32 cores do not behave in the same manner as real 16 cores (numbers here for a single machine).
-We disregarded the hyperthreading ability in most of the
+Non-surprisingly, we found that hyper-threaded 32 cores do not behave in the same manner as real 16 cores (numbers here for a single machine).
+We disregarded the hyper-threading ability in most of the
 cases.
@@
 
-\subsection{Effect of hyperthreading}
+\subsection{Effect of hyper-threading}
 
@@
 efficiency was 85.3\%. However, if we increase the number of
-requested cores to be 32 -- \ie if we use hyperthreading on 16 real
+requested cores to be 32 -- \ie if we use hyper-threading on 16 real
 cores -- the speedup does not increase that well. It is merely 15.99
@@
 for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
-hyperthreaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
+hyper-threaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
 version used here was Eden~CP, the \enquote{share nothing} SMP build.
@@
 In the distributed memory setting the same effect ensues. We obtain
-plummeting speedup of 124.31 at 512 hyperthreaded cores, whereas it was
-213.172 for 256 real cores. Apparently, hyperthreading in the Glasgow
+plummeting speedup of 124.31 at 512 hyper-threaded cores, whereas it was
+213.172 for 256 real cores. Apparently, hyper-threading in the Glasgow
 grid fails to execute two parallel Haskell processes with full-fledged
-parallelism. For this reason, we did not regard hyperthreaded cores in
+parallelism. For this reason, we did not regard hyper-threaded cores in
 our speedup plots in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark}.
@@
 time differences that are not relative \wrt the total execution time.
-Furthermore, the error bars ends were computed from pointwise maximum of both standard
+Furthermore, the error bars ends were computed from point-wise maximum of both standard
 deviations from both measurements for PArrows and non-PArrows
 versions. These are the values provided by the |bench| package that we
-used for benchmarking. We call a difference between two versions
+used for bench-marking. We call a difference between two versions
 significant when the border of the error bar of absolute time
@@
 This is due to the benchmark only using |parMap| instead of a chunking variant -- however we did not change that for simplicity's sake.
-This issue is reflected by debug output which shows that of 16000 sparks being created (one for each sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
+This issue is reflected by debug output which shows that of 16000 sparks being created (one for each Sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
 Another remarkable finding is that the Eden backend seems to lag behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.

### 2017-10-24 paper/content/futures.tex
spellchecked

@@
 Eden already ships with \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
-But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| typeclass (Fig.~\ref{fig:future}).
+But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| type class (Fig.~\ref{fig:future}).
 \begin{figure}[h]
@@
 \end{code}
-\caption{Definition of the |Future| typeclass.}
+\caption{Definition of the |Future| type class.}
 \label{fig:future}
@@
 
-For the |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the typeclass is satisfied.
+For the |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the type class is satisfied.
 This is because the concept of a |Future| does not change anything for shared-memory execution as there are no communication problems to fix.

### 2017-10-24 paper/content/motivation.tex
wip
spellchecked

@@
 %Our Arrows-based interface is a high-level one.
-We provide an Arrows-based typeclass and implementations for the three above mentioned parallel Haskells.
+We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
 Instead of 
@@
 
-We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| typeclass, which abstracts all parallel implementation logic.
-This way, we can have multiple backends - currently a GpH, a |Par| Monad and an Eden version are supported.
+We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
+This way, we can have multiple backends -- currently a GpH, a |Par| Monad and an Eden version are supported.
 The backends can easily be swapped, so we are not bound to any specific one.
@@
 
-We introduce a |Future| typeclass in Section \ref{sec:futures} to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|).
+We introduce a |Future| type class in Section \ref{sec:futures} to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|).
 Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node.

### 2017-10-24 paper/content/topologySkeletons.tex
spellchecked

@@
 If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing.
-This materializes with the usage of |loop| of the |ArrowLoop| typeclass and is probably due to the way their respective parallelism engines work internally.
-\mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| typeclass.
+This materializes with the usage of |loop| of the |ArrowLoop| type class and is probably due to the way their respective parallelism engines work internally.
+\mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| type class.
 This allows us for selective enabling and disabling of parallelism.
@@
 The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
-We implement this skeleton using the |ArrowLoop| typeclass which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
+We implement this skeleton using the |ArrowLoop| type class which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
 For example %this
@@
 \end{code}
-It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data commuinication between them.
+It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them.
 Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.
@@
 If we take the concept of a |ring| from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}).
-Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
-With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.
+Every node sends and receives data from horizontal and vertical neighbours in each communication round.
+With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| type class.
 
@@
 We first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| type class.
 Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.

### 2017-10-24 paper/content/benchmarks.tex
minor

@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel run time of \rmtest ``#2''}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \rmtest \enquote{#2}}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Speedup of \rmtest ``#2''}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\speedupplot{Speedup of \rmtest \enquote{#2}}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 \newcommand{\speedupSkelRMDist}[4]{
-\speedupplot{Speedup of \rmtest ``#1 #2''}{PArrows}{256}{#3}{
+\speedupplot{Speedup of \rmtest \enquote{#1 #2}}{PArrows}{256}{#3}{
 % \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 \newcommand{\performanceSkelRMDistDiff}[5]{
-\performancediffplot{Run time differences\\for \rmtest ``#1 #2''}{(Eden $-$ PArrows)}{256}{#3}{
+\performancediffplot{Run time differences\\for \rmtest \enquote{#1 #2}}{(Eden $-$ PArrows)}{256}{#3}{
 \addplot+[mark=*,very thick,error bars/.cd,
@@
 
-\newcommand{\speedupJacobiDist}[4]{
-\speedupplot{Speedup of \jacobitest ``#2''}{PArrows}{256}{#3}{
+\newcommand{\speedupJacobiDist}[5]{
+\speedupplot{Speedup of \jacobitest \enquote{#2} vs simulated \enquote{#5}}{PArrows #2, simulated PArrows #5}{256}{#3}{
 % \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#2.csv};
+\addplot [mark=x,very thick,red] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#5.csv};
 % \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 
-\newcommand{\performanceJacobiDistDiff}[5]{
-\performancediffplot{Run time differences\\for \jacobitest ``#2''}{(Eden $-$ PArrows)}{256}{#3}{
+\newcommand{\performanceJacobiDistDiff}[6]{
+\performancediffplot{Run time differences\\for \jacobitest \enquote{#2}}{(Eden $-$ PArrows)}{256}{#3}{
 \addplot+[mark=*,very thick,error bars/.cd,
@@
 smooth]{benchmarks/distributed-jacobi/#1-#2-diff.csv};
+\addplot+[mark=*,very thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+smooth]{benchmarks/distributed-jacobi/#1-#6-diff.csv};
 }{#4}{#5}{\plotwidthDist}
@@
 %
-{\speedupJacobiDist{3}{3217}{32}{272}\label{subfig:rm-dist-a}}%
+{\speedupJacobiDist{3}{3217}{32}{272}{4253}\label{subfig:rm-dist-a}}%
 %\hfill%
-{\performanceJacobiDistDiff{3}{3217}{32}{0.5}{272}\label{subfig:rm-dist-b}}
+{\performanceJacobiDistDiff{3}{3217}{32}{0.5}{272}{4253}\label{subfig:rm-dist-b}}
 \caption{Parallel performance of \rmtest on the Glasgow grid
@@
 \newcommand{\speedupTorusDist}[3]{
-\speedupplot{Speedup of \torustest ``#1''}{PArrows}{256}{#2}{
+\speedupplot{Speedup of \torustest \enquote{#1}}{PArrows}{256}{#2}{
 \addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 \newcommand{\performanceTorusDistDiff}[4]{
-\performancediffplot{Run time differences\\for \torustest ``#1''}{(Eden $-$ PArrows)}{256}{#2}{
+\performancediffplot{Run time differences\\for \torustest \enquote{#1}}{(Eden $-$ PArrows)}{256}{#2}{
 \addplot+[mark=*,very thick,error bars/.cd,
@@
 \newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest ``#1''}{linear, Eden CP, GpH, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \sudokutest \enquote{#1}}{linear, Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot [no markers,dotted,thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSudokuSM}[1]{
-\speedupplot{Parallel speedup of \sudokutest ``#1''}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\speedupplot{Parallel speedup of \sudokutest \enquote{#1}}{Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,

### 2017-10-24 paper/content/benchmarks.tex
minor

@@
 
+Not pictured: difference between Eden and PArrows with 4253
+
 \newcommand{\speedupJacobiDist}[5]{
@@
 
-\newcommand{\performanceJacobiDistDiff}[6]{
-\performancediffplot{Run time differences\\for \jacobitest \enquote{#2}}{(Eden $-$ PArrows)}{256}{#3}{
+\newcommand{\performanceJacobiDistDiff}[5]{
+\performancediffplot{Run time differences\\for \jacobitest \enquote{#2}}{(Eden $-$ PArrows) #2}{256}{#3}{
 \addplot+[mark=*,very thick,error bars/.cd,
@@
 smooth]{benchmarks/distributed-jacobi/#1-#2-diff.csv};
-\addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
-smooth]{benchmarks/distributed-jacobi/#1-#6-diff.csv};
 }{#4}{#5}{\plotwidthDist}
@@
 %\hfill%
-{\performanceJacobiDistDiff{3}{3217}{32}{0.5}{272}{4253}\label{subfig:rm-dist-b}}
+{\performanceJacobiDistDiff{3}{3217}{32}{5}{272}\label{subfig:rm-dist-b}}
 \caption{Parallel performance of \rmtest on the Glasgow grid
   consisting of 256 cores. Input was $2^{3217}-1$, we used 256
-  tasks. The top plot shows absolute speedup in a distributed memory setting. The
+  tasks. The top plot shows absolute speedup in a distributed memory setting compared to a simulated speedup for input $2^{4253}-1$. The
   closer to linear speedup the better. Time

### 2017-10-24 paper/content/benchmarks.tex
remarks done (oleg reread), only abstract missing

@@
 
-Continuing, the results of the \jacobitest in \ref{fig:bench-jacobi-dist} are as follows:
+Continuing, the results of the \jacobitest (MPI only) in Fig.~\ref{fig:bench-jacobi-dist} are as follows:
+The program does not seem to scale well beyond 16-32 tasks with input $2^{3217}-1$.
+Because of this, we also ran tests with input $2^{4253}-1$, but only for the 128 and 256 cores because of the long running time.
+For demonstration purposes in Fig.~\ref{fig:bench-jacobi-dist}, we chose a simulated speedup of 22 for 128 cores.
+For 256 cores, this results in a superlinear speedup of \~122.63, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
 
-Not pictured: difference between Eden and PArrows with 4253
+We once again compare the Eden version with the PArrrows version in Fig.~\ref{fig:bench-jacobi-dist}.
+We see similar behaviour to the MPI version of the \rmtest: The difference once again almost always remains within the bounds of the error bar of the measurement.
+
+Not pictured are the slightly bigger differences between Eden and PArrows for the larger input of $2^{4253}-1$: For 128 and 256 cores the differences are -1005.18\seconds and -94.33\seconds in favour of Eden.
+This maximum of 12.27\% slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.
 
@@
 \end{figure}
-
-\mbcomment{FIXME: redo this test with $2^{44497}$}
 
@@
 In our distributed memory tests on the full grid we were unable to detect any performance deficiencies of PArrows \wrt Eden that were above of the error margin.
-In our distributed memory tests on the full grid the difference between PArrows and Eden was always below the error margin.
+In our distributed memory tests on the full grid the difference between PArrows and Eden was always below the error margin if PArrows performed worse.
+Even the biggest difference of 12.27\% in the \jacobitest for input $2^{4253}-1$ and 128 cores remained in line with these findings.

### 2017-10-24 paper/content/benchmarks.tex
minor remarks

@@
 
-The \ensuremath{\Conid{Par}} monad and GpH Haskell -- in its multicore version (\cite{Marlow2009}) --  can be executed on a shared
+The |Par| monad and GpH Haskell -- in its multicore version \cite{Marlow2009} --  can be executed on a shared
 memory machines only. Although GpH is available on distributed memory
@@
 Monad benchmark is very small.
-To give an example, it is ~0.4\seconds in favour of PArrows for 16 cores (10.8\seconds vs.
-11.2\seconds) and ~-0.8\seconds in favour of the |Par| monad for 8 cores (16.1\seconds vs.
-16.9\seconds) for
+To give an example, it is $0.4\seconds$ in favour of PArrows for 16 cores ($10.8\seconds$ vs.
+$11.2\seconds$) and $-0.8\seconds$ in favour of the |Par| monad for 8 cores ($16.1\seconds$ vs.
+$16.9\seconds$) for
 the Sudoku benchmark in the shared memory setting. It is almost invisible in speedup and
@@
 (Figs.~\ref{fig:bench-rm-dist}--\ref{fig:torusBenchmark}). As an example, the differences range in
-about 0.5\seconds for the execution time of 46\seconds on 256 cores
+about $0.5\seconds$ for the execution time of $46\seconds$ on 256 cores
 for distributed \rmtest with PArrows and Eden. For these comparisons, the plots show absolute
@@
 {\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{272}\label{subfig:rm-dist-b}}
-\caption{Parallel performance of \rmtest on the Glasgow grid
+\caption{Parallel performance of the \rmtest on the Glasgow grid
   consisting of 256 cores. Input was $2^{44497}-1$, we used 256
@@
 Continuing, the results of the \jacobitest (MPI only) in Fig.~\ref{fig:bench-jacobi-dist} are as follows:
-The program does not seem to scale well beyond 16-32 tasks with input $2^{3217}-1$.
+The program does not seem to scale well beyond 64 cores with input $2^{3217}-1$.
 Because of this, we also ran tests with input $2^{4253}-1$, but only for the 128 and 256 cores because of the long running time.
-For demonstration purposes in Fig.~\ref{fig:bench-jacobi-dist}, we chose a simulated speedup of 22 for 128 cores.
-For 256 cores, this results in a superlinear speedup of \~122.63, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
+For demonstration purposes in Fig.~\ref{fig:bench-jacobi-dist}, we chose a \enquote{simulated} speedup of 22 for 128 cores.
+This was obtained by taking the speedup value of 64 cores (approx.
+11) of the $2^{3217}-1$ and assuming optimal \enquote{linear} speedup between 64 and 128 cores.
+For 256 cores, this results in a superlinear \enquote{speedup} of \~122.63 relative to 128 cores, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
 
@@
 
-Not pictured are the slightly bigger differences between Eden and PArrows for the larger input of $2^{4253}-1$: For 128 and 256 cores the differences are -1005.18\seconds and -94.33\seconds in favour of Eden.
-This maximum of 12.27\% slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.
+Not pictured are the slightly bigger differences between Eden and PArrows for the larger input of $2^{4253}-1$: For 128 and 256 cores the differences are $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden.
+This maximum of $12.27\%$ slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.
 
@@
 {\performanceJacobiDistDiff{3}{3217}{32}{5}{272}\label{subfig:rm-dist-b}}
-\caption{Parallel performance of \rmtest on the Glasgow grid
+\caption{Parallel performance of the \jacobitest on the Glasgow grid
   consisting of 256 cores. Input was $2^{3217}-1$, we used 256
@@
 \newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest \enquote{#1}}{linear, Eden CP, GpH, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \sudokutest \enquote{#1}}{linear speedup, Eden CP, GpH, |Par| Monad}{16}{4}{
 \addplot [no markers,dotted,thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,

### 2017-10-25 paper/content/abstract.tex
abstract rework

@@
 Arrows are a general interface for computation and therefore form an alternative to Monads for API design.
-We express parallelism using this concept in a novel way: We define an arrows-based language for parallelism and implement it using multiple parallel Haskells.
+We express parallelism using this concept in a novel way: We define an Arrow-based language for parallelism and implement it using multiple parallel Haskells.
 In this manner we are able to bridge across various parallel Haskells.
 
-Additionally, our way of writing parallel programs has the benefit of being portable across multiple parallel Haskell implementations.
-Furthermore, as each parallel computation is an arrow, which means that they can be readily composed and transformed as such.
-We introduce some syntactic sugar to provide parallelism-aware arrow combinators.
+Additionally, using these parallel Arrows (PArrows) has the benefit of being portable across multiple parallel Haskell implementations.
+Furthermore, as each parallel computation is an Arrow, PArrows can be readily composed and transformed as such.
+In order to allow for more sophisticated communication schemes between computation nodes in distributed systems we introduce the concept of Futures to wrap existing direct communication concepts in backends.
 
-To show that our arrow-based language has similar expressiveness to existing parallel languages, we also define several parallel skeletons with our framework. 
-Benchmarks show that our framework does not induce too much overhead
+To show that PArrows have similar expressiveness to existing parallel languages, we also implement several parallel skeletons. 
+Benchmarks show that our framework does not induce any notable overhead
 performance-wise.
+We finally conclude that Arrows turn out to be a useful tool for composing parallel programs and that our particular approach results in programs that are portable accross multiple backends.
 \olcomment{Summarize conclusions}

### 2017-10-25 paper/content/benchmarks.tex
fixed description of Gentleman Benchmark

@@
 The results of the comparison of vanilla Eden to our PArrows-based version can be found in Fig.~\ref{fig:torusBenchmark}.
-While we do not see too much speedup for more than 16 cores, we can prove that the difference between the Eden and PArrows version are again marginal.
+We see that the benchmark scales quite well with more cores until 64 cores.
+For more cores (>=96) we still have considerable speedup, but it increases slower.
+We also prove that the difference between the Eden and PArrows version are only again marginal with PArrows only being a maximum of $1.7\%$ slower -- for 160 cores -- when outside of the error bar.
 %The difference between PArrows and Eden is only significant for 16 and 64 cores where it ran 1.7\% and 2.7\% slower which corresponds to a real-time difference of 0.12s and 0.13s.

### 2017-10-25 paper/content/benchmarks.tex
more clarification for the benchmarks

--- 2017-10-24 paper/content/benchmarks.tex
@@
 The results of the comparison of vanilla Eden to our PArrows-based version can be found in Fig.~\ref{fig:torusBenchmark}.
-While we do not see too much speedup for more than 16 cores, we can prove that the difference between the Eden and PArrows version are again marginal.
+We see that the benchmark scales quite well with more cores until 64 cores.
+For $>=96$ cores, we still have considerable speedup, but with less slope.
+We also prove that the difference between the Eden and PArrows version are only marginal with PArrows only being a maximum of $1.7\%$ slower -- for 160 cores -- when outside of the error bar.
 %The difference between PArrows and Eden is only significant for 16 and 64 cores where it ran 1.7\% and 2.7\% slower which corresponds to a real-time difference of 0.12s and 0.13s.
@@
 
-In our distributed memory tests on the full grid we were unable to detect any performance deficiencies of PArrows \wrt Eden that were above of the error margin.
-In our distributed memory tests on the full grid the difference between PArrows and Eden was always below the error margin if PArrows performed worse.
+In the shared memory setting we naturally find that the backends perform differently.
+Furthermore, in our distributed memory tests on the full grid the difference between PArrows and Eden was almost always below the error margin if PArrows performed worse.
 Even the biggest difference of 12.27\% in the \jacobitest for input $2^{4253}-1$ and 128 cores remained in line with these findings.
+The only exception to this was the \torustest, which showed only a $1.7\%$ difference in favour of Eden which is only a marginal slowdown in our opinion.

### 2017-10-25 paper/content/conclusion.tex
spellchecked
more clarification for the benchmarks

@@
 This work features multiple parallel backends: the already available parallel Haskell flavours.
-Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| typeclass for GpH Haskell, |Par| Monad, and Eden.
+Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden.
 With our approach parallel programs can be ported across these flavours with little to no effort.
@@
 %
-Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.
+Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead in our benchmarks.
 %

### 2017-10-25 paper/content/motivation.tex
minor

@@
 novel manner - we tame the zoo of parallel Haskells. We provide a
-common \ptcomment{quantify, e.g. less than 10 \%}, very low-penalty programming interface that allows to switch
+common \ptcomment{quantify, e.g. less than 10 \%} \mbcomment{we do this in the contributions section of the Introduction, enough?}, very low-penalty programming interface that allows to switch
 the parallel backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
@@
 We show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced in Section \ref{sec:benchmarks}, i.e.
-less than x\% \mbcomment{fix this after benchmarks are finally done}.
+less than $1.7\%$ in cases outside of the error bar of the measurement.
+\mbcomment{fix this after benchmarks are finally done}.
 

### 2017-10-25 paper/content/benchmarks.tex
minor

@@
 
-We have tested our parallel DSL and algorithmic skeletons implemented
-in it. Benchmarks were conducted both in a shared and in a distributed
-memory setting. All benchmarks were done on the ``Glasgow grid'', consisting of
+In the following we describe the benchmarks of our parallel DSL and algorithmic skeletons we conducted. They were run both in a shared and in a distributed
+memory setting. All benchmarks were done on the \enquote{Glasgow grid}, consisting of
 16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each.
@@
 (non shown) run time plots. We thus show only the results for the
-PArrows-enabled versions.
+PArrows-enabled versions in the backend-comparison.
 
-To show that PArrows induce very small overhead in a distributed context as well, we compare the original Eden
+To show that PArrows induce very small overhead even in a distributed context (and in general as well), we compare the original Eden
 versions of the benchmark to its PArrows-enabled counterpart in the \rmtest, \torustest and \jacobitest benchmarks. We plot execution time differences between measurements for
@@
 difference is above or below zero. In other words: the time
-difference is significant if it is above measurement error.
+difference is significant if it is outside of the measurement error.
 
@@
 Even the biggest difference of 12.27\% in the \jacobitest for input $2^{4253}-1$ and 128 cores remained in line with these findings.
-The only exception to this was the \torustest, which showed only a $1.7\%$ difference in favour of Eden which is only a marginal slowdown in our opinion.
+The only exception to this was the \torustest  with only a $1.7\%$ difference in favour of Eden which is only a marginal slowdown in our opinion.

### 2017-11-02 paper/content/benchmarks.tex
minor rewording of the jacobi benchmark chapter, removed the "simulated" speedups

@@
+%
 \section{Performance results}
@@
 \midrule
-\rmtest & Mathematics & |parMap+reduce| & Eden & \citet{Lobachev2012}\\
-\jacobitest & Mathematics & |workpool+reduce| & Eden & \citet{Lobachev2012}\\
-\torustest & Mathematics & |torus| & Eden & \citet{Eden:SkeletonBookChapter02}\\
-\sudokutest & Puzzle & |parMap| & |Par| Monad & \citet{par-monad} 
+\rmtest & Mathematics & \ensuremath{\Varid{parMap}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
+\jacobitest & Mathematics & \ensuremath{\Varid{workpool}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
+\torustest & Mathematics & \ensuremath{\Varid{torus}} & Eden & \citet{Eden:SkeletonBookChapter02}\\
+\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad} 
 \tablefootnote{actual code from: http://community.haskell.org/~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}\\
@@
 ``Gentleman'' is a standard Eden test program, developed
-for their |torus| skeleton. It implements a parallel matrix
+for their \ensuremath{\Varid{torus}} skeleton. It implements a parallel matrix
 multiplication \citep{Gentleman1978}. We ported an Eden based version \citep{Eden:SkeletonBookChapter02} to PArrows.
 
-A~parallel Sudoku solver was used by \citet{par-monad} to compare |Par| Monad
+A~parallel Sudoku solver was used by \citet{par-monad} to compare \ensuremath{\Conid{Par}} Monad
 to GpH Haskell. We ported it to PArrows.
@@
 
-The |Par| monad and GpH Haskell -- in its multicore version \cite{Marlow2009} --  can be executed on a shared
+The \ensuremath{\Conid{Par}} monad and GpH Haskell -- in its multicore version \cite{Marlow2009} --  can be executed on a shared
 memory machines only. Although GpH is available on distributed memory
@@
 
-The difference between, say, PArrows with |Par| Monad backend and a
-genuine |Par|
+The difference between, say, PArrows with \ensuremath{\Conid{Par}} Monad backend and a
+genuine \ensuremath{\Conid{Par}}
 Monad benchmark is very small.
 To give an example, it is $0.4\seconds$ in favour of PArrows for 16 cores ($10.8\seconds$ vs.
-$11.2\seconds$) and $-0.8\seconds$ in favour of the |Par| monad for 8 cores ($16.1\seconds$ vs.
+$11.2\seconds$) and $-0.8\seconds$ in favour of the \ensuremath{\Conid{Par}} monad for 8 cores ($16.1\seconds$ vs.
 $16.9\seconds$) for
@@
 deviations from both measurements for PArrows and non-PArrows
-versions. These are the values provided by the |bench| package that we
+versions. These are the values provided by the \ensuremath{\Varid{bench}} package that we
 used for bench-marking. We call a difference between two versions
@@
 \newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel run time of \rmtest \enquote{#2}}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \rmtest \enquote{#2}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Speedup of \rmtest \enquote{#2}}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\speedupplot{Speedup of \rmtest \enquote{#2}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 
-\olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
-  Why is it good?}
+%\olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
+%  Why is it good?}
 The multicore version of our parallel \rmtest benchmark is depicted in
@@
 The performance of PArrows/Eden~CP in shared memory is slightly better than
-for SMP variants such as PArrows/GpH and PArrows/|Par|
+for SMP variants such as PArrows/GpH and PArrows/\ensuremath{\Conid{Par}}
 Monad but most of the time the performance is still comparable with the GpH backend performing slightly worse than the other two in terms of speedup.
@@
 The program does not seem to scale well beyond 64 cores with input $2^{3217}-1$.
-Because of this, we also ran tests with input $2^{4253}-1$, but only for the 128 and 256 cores because of the long running time.
-For demonstration purposes in Fig.~\ref{fig:bench-jacobi-dist}, we chose a \enquote{simulated} speedup of 22 for 128 cores.
-This was obtained by taking the speedup value of 64 cores (approx.
-11) of the $2^{3217}-1$ and assuming optimal \enquote{linear} speedup between 64 and 128 cores.
-For 256 cores, this results in a superlinear \enquote{speedup} of \~122.63 relative to 128 cores, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
+
+Because of this, we also ran tests with input $2^{4253}-1$, but only for the 128 and 256 cores (and therefore not pictured in the speedup figure) because of the long running time.
+For 128 threads, the benchmark took a mean of 9192.9\seconds, while with 256 threads, it only took a mean of 1649.1\seconds.
+This means that the 128 thread version ran more than 5.5 times slower than the 256 one, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
+However it shows that PArrows continue to scale, even if not perfectly for this test program.
 
@@
 
-Not pictured are the slightly bigger differences between Eden and PArrows for the larger input of $2^{4253}-1$: For 128 and 256 cores the differences are $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden.
+Not pictured again are the slightly bigger differences between Eden and PArrows for the larger input of $2^{4253}-1$: For 128 and 256 cores the differences are $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden.
 This maximum of $12.27\%$ slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.
@@
 smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#2.csv};
-\addplot [mark=x,very thick,red] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#5.csv};
+%\addplot [mark=x,very thick,red] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+%smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#5.csv};
 % \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 However, we notice that the GpH backend seems to choke on a bigger input (Fig.~\ref{fig:sudokuSMBenchmark16000}).
-This is due to the benchmark only using |parMap| instead of a chunking variant -- however we did not change that for simplicity's sake.
+This is due to the benchmark only using \ensuremath{\Varid{parMap}} instead of a chunking variant -- however we did not change that for simplicity's sake.
 This issue is reflected by debug output which shows that of 16000 sparks being created (one for each Sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
@@
 \newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest \enquote{#1}}{linear speedup, Eden CP, GpH, |Par| Monad}{16}{4}{
+\performanceplot{Parallel run time of \sudokutest \enquote{#1}}{linear speedup, Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
 \addplot [no markers,dotted,thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
@@
 \newcommand{\speedupSudokuSM}[1]{
-\speedupplot{Parallel speedup of \sudokutest \enquote{#1}}{Eden CP, GpH, |Par| Monad}{16}{4}{
+\speedupplot{Parallel speedup of \sudokutest \enquote{#1}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 \caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
+  different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
   grid; it has 16 real cores and 32 threads. Input was a file of $1000$ Sudokus. The
@@
 \caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware and the |parMap| version from the |Par| Monad examples. Measurements were performed on a single node of the Glasgow
+  different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
   grid; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The

### 2017-11-02 paper/content/benchmarks.tex
rewording

@@
 Continuing, the results of the \jacobitest (MPI only) in Fig.~\ref{fig:bench-jacobi-dist} are as follows:
-The program does not seem to scale well beyond 64 cores with input $2^{3217}-1$.
-
-Because of this, we also ran tests with input $2^{4253}-1$, but only for the 128 and 256 cores (and therefore not pictured in the speedup figure) because of the long running time.
-For 128 threads, the benchmark took a mean of 9192.9\seconds, while with 256 threads, it only took a mean of 1649.1\seconds.
-This means that the 128 thread version ran more than 5.5 times slower than the 256 one, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
-However it shows that PArrows continue to scale, even if not perfectly for this test program.
-
+The program does not seem to scale well beyond 64 threads with input $2^{3217}-1$.
 We once again compare the Eden version with the PArrrows version in Fig.~\ref{fig:bench-jacobi-dist}.
@@
 
-Not pictured again are the slightly bigger differences between Eden and PArrows for the larger input of $2^{4253}-1$: For 128 and 256 cores the differences are $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden.
+Because of the bad scaling behaviour for $2^{3217}-1$ beyond 64 threads, we also ran tests with input $2^{4253}-1$.
+Because of the long running time, we could do this only for the 128 and 256 threads.
+Therefore we do not show these results in Fig.~\ref{fig:bench-jacobi-dist} as we cannot properly compute a speedup without results for 1 thread.
+Nonetheless, for 128 threads, the benchmark took a mean of 9192.9\seconds, while with 256 threads, it only took a mean of 1649.1\seconds.
+This means that the 128 thread version ran more than 5.5 times slower than the 256 one, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
+However it still proves that PArrows continue to scale, even if not perfectly for this test program.
+Comparing PArrows with Eden, for the larger input of $2^{4253}-1$, we see slightly bigger differences between Eden and PArrows: The differences for 128 and 256 coresare $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden, respectively.
 This maximum of $12.27\%$ slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.

### 2017-11-03 benchmarks/jacobi-bench/jacobi/JacobiSum.hs
minor

@@ -64,6 +64,8 @@ bestimme_t n
 		 | ( n < 2^1566 ) = 720720
 		 | ( n < 2^2082 ) = 1663200
 		 | ( n < 2^3217 ) = 8648640 
+                 | ( n < 2^4253 ) = 69189120
+                 | ( n < 2^4423 ) = 69189120
                  | ( n < 2^44497) = 137944178989
 		 | otherwise = hilfsf  137944178989 n 
             where hilfsf = (\x w -> if ((e x)*(e x) > w) then x else (hilfsf (x + x) w))

### 2017-11-03 paper/content/benchmarks.tex

@@
 %
-{\speedupJacobiDist{3}{3217}{32}{272}{4253}\label{subfig:rm-dist-a}}%
+{\speedupJacobiDist{3}{3217}{32}{272}{4253}\label{subfig:jacobi-dist-a}}%
 %\hfill%
-{\performanceJacobiDistDiff{3}{3217}{32}{5}{272}\label{subfig:rm-dist-b}}
+{\performanceJacobiDistDiff{3}{3217}{32}{5}{272}\label{subfig:jacobi-dist-b}}
 \caption{Parallel performance of the \jacobitest on the Glasgow grid
   consisting of 256 cores. Input was $2^{3217}-1$, we used 256
-  tasks. The top plot shows absolute speedup in a distributed memory setting compared to a simulated speedup for input $2^{4253}-1$. The
+  tasks. The top plot shows relative speedup in a distributed memory setting compared to a simulated speedup for input $2^{4253}-1$. The
   closer to linear speedup the better. Time
@@
 {\performanceSudokuSM{1000}\label{subfig:performanceSudokuSM}}
-\caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
+\caption{Absolute speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
   different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
@@
 {\performanceSudokuSM{16000}\label{subfig:performanceSudokuSM16000}}
-\caption{Relative speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
+\caption{Absolute speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
   different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow

### 2017-11-03 paper/content/parallelHaskells.tex

@@
 We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator.
-We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form.
+We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evaluates to normal form.
 %
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-	\caption{Dataflow of the GpH |parEvalN| version.}
+	\caption{Data flow of the GpH |parEvalN| version.}
 	\label{fig:parEvalNMulticoreImg}
@@
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-	\caption{Dataflow of the |Par| Monad |parEvalN| version.}
+	\caption{Data flow of the |Par| Monad |parEvalN| version.}
 	\label{fig:parEvalNParMonadImg}
@@
 An inactive machine, where no processes are started yet, or all are already terminated, shows as a blue bar.
-A~comminication from one process to another is represented with a black arrow.
+A~communication from one process to another is represented with a black arrow.
 A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.

### 2017-11-03 paper/content/parrows.tex

@@
 First, we present the interface and explain the reasonings behind it.
-Then, we discuss some implementations using exisiting parallel Haskells.
+Then, we discuss some implementations using existing parallel Haskells.
 Finally, we explain why using Arrows for expressing parallelism is beneficial.
-\subsection{The ArrowParallel typeclass}
+\subsection{The ArrowParallel type class}
 A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
@@
 With this definition of |parEvalN|, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new typeclass |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new type class |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
 % \begin{figure}[h]
@@
 \subsubsection{Eden}
-For both the GpH Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| typeclasses.
+For both the GpH Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| type classes.
 With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a typeclass: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
+While we could still manage to have only one class in the module by introducing a type class: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
 % \begin{figure}[h]
@@
 % We don't do it here
-However, we avoid doing so for aesthetic resons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
+However, we avoid doing so for aesthetic reasons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
 % \begin{figure}[h]

### 2017-11-03 paper/content/relwork.tex

@@
 Experiments with our approach in Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM) were quite successful.
-However, it is beyond the scope of this work.
+A new approach to Haskell on the JVM is the Eta language\footnote{Eta project page at \url{eta-lang.org}}.
+However, both are beyond the scope of this work.
 
-\citet{achten2004arrows,achten2007arrow} use an arrow implementation in Clean for better handling of typical GUI tasks.
-\citet{Dagand:2009:ORD:1481861.1481870} used arrows in OCaml in the implementation of a distributed system.
+\citet{achten2004arrows,achten2007arrow} use an Arrow implementation in Clean for better handling of typical GUI tasks.
+\citet{Dagand:2009:ORD:1481861.1481870} used Arrows in OCaml in the implementation of a distributed system.
 

### 2017-11-03 paper/content/utilityFunctions.tex

@@
 \section{Utility Arrows}\label{utilfns}
-Following are definitions of some utility Arrows used in this paper that have been lef out for brevity.
+Following are definitions of some utility Arrows used in this paper that have been left out for brevity.
 We start with the |second| combinator from \citet{HughesArrows}, which is a mirrored version of |first|, which is for example used in the definition of |***|: 
@@
 % $ %% formatting
-Note that  this additionally makes use of the |ArrowApply| typeclass that allows us to evaluate arrows with |app :: arr (arr a b, a) c|.
+Note that  this additionally makes use of the |ArrowApply| type class that allows us to evaluate arrows with |app :: arr (arr a b, a) c|.
 
@@
 \end{figure}
-Furthermore, Fig.~\ref{fig:torus_example_rest} contains the ommited definitions of |prMMTr| (which calculates $AB^T$ for two matrices $A$ and $B$), |splitMatrix| (which splits the a matrix into chunks), and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
+Furthermore, Fig.~\ref{fig:torus_example_rest} contains the omitted definitions of |prMMTr| (which calculates $AB^T$ for two matrices $A$ and $B$), |splitMatrix| (which splits the a matrix into chunks), and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
 \begin{figure}[h]

### 2017-11-23 paper/content/benchmarks.tex
wip on tables

@@
 \rmtest is a probabilistic primality test that iterates multiple (32--256 here)
-``subtests''. Should a subtest fail, the input is definitely not a
+\enquote{subtests}. Should a subtest fail, the input is definitely not a
 prime. If all $n$ subtest pass, the input is composite with the
@@
 
-``Gentleman'' is a standard Eden test program, developed
+\enquote{Gentleman} is a standard Eden test program, developed
 for their \ensuremath{\Varid{torus}} skeleton. It implements a parallel matrix
@@
 to GpH Haskell. We ported it to PArrows.
+
+\begin{table}[]
+\centering
+\label{my-label}
+\begin{subtable}[t]{.5\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{l R{1cm} R{1.3cm} R{1.3cm}}
+		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+		vs.             & cores	& params   & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/sudoku-sm/worstSudoku.csv}{}{\vs & \nCores & \params & \factor \\}
+	\end{tabular}
+	\endgroup
+    \caption{Sudoku}\label{table1}
+\end{subtable}
+\caption{Worst Benchmark results}
+\end{table}
 
@@
 smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#2.csv};
-%\addplot [mark=x,very thick,red] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-%smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#5.csv};
+\addplot [mark=x,very thick,red] table [scatter, x="nCores", y="speedup"/, col sep=comma, mark=none,
+smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#5.csv};
 % \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,

### 2017-11-24 paper/utils/redoJacobi.sh
minor

@@ -38,15 +38,25 @@ params=(
 
 benchmark="\"Jacobi (Distributed)\""
 
-worstFileName="worstJacobi.csv"
-bestFileName="bestJacobi.csv"
+worstFileName=(
+    "worstJacobi-3217.csv"
+    "worstJacobi-4253.csv"
+)
+
+bestFileName=(
+    "bestJacobi-3217.csv"
+    "bestJacobi-4253.csv"
+)
 
 outputFolder="../content/benchmarks/distributed-jacobi"
 
 count=${#originalBenchmarks[@]}
 
-cp header.txt ${worstFileName}
-cp header.txt ${bestFileName}
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp header.txt ${worstFileName[i]}
+    cp header.txt ${bestFileName[i]}
+done
 
 for i in $(seq 0 $(expr ${count} - 1));
 do
@@ -60,7 +70,11 @@ do
     ./calculateDifferences ${originalBenchmarks[i]} ${parrowsBenchmarks[i]} ${bestFileName} True False
 done
 
-cp ${worstFileName} ${outputFolder}
-cp ${bestFileName} ${outputFolder}
+
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp ${worstFileName[i]} ${outputFolder}
+    cp ${bestFileName[i]} ${outputFolder}
+done
 
 rm *.csv

### 2017-11-24 paper/utils/redoRM.sh

@@ -31,15 +31,22 @@ params=(
 
 benchmark="\"Rabin-Miller (Distributed)\""
 
-worstFileName="worstRM.csv"
-bestFileName="bestRM.csv"
+worstFileName=(
+    "worstRM-44497-256.csv"
+)
+bestFileName=(
+    "bestRM-44497-256.csv"
+)
 
 outputFolder="../content/benchmarks/distributed-rm"
 
 count=${#originalBenchmarks[@]}
 
-cp header.txt ${worstFileName}
-cp header.txt ${bestFileName}
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp header.txt ${worstFileName[i]}
+    cp header.txt ${bestFileName[i]}
+done
 
 for i in $(seq 0 $(expr ${count} - 1));
 do
@@ -53,7 +60,11 @@ do
     ./calculateDifferences ${originalBenchmarks[i]} ${parrowsBenchmarks[i]} ${bestFileName} True False
 done
 
-cp ${worstFileName} ${outputFolder}
-cp ${bestFileName} ${outputFolder}
+
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp ${worstFileName[i]} ${outputFolder}
+    cp ${bestFileName[i]} ${outputFolder}
+done
 
 rm *.csv

### 2017-11-24 paper/utils/redoRMSM.sh

@@ -61,15 +61,33 @@ params=(
 
 benchmark="\"Rabin-Miller (Shared-Memory)\""
 
-worstFileName="worstRMSM.csv"
-bestFileName="bestRMSM.csv"
+worstFileName=(
+    "worstRMSM-11213-32.csv"
+    "worstRMSM-11213-64.csv"
+    "worstRMSM-11213-32.csv"
+    "worstRMSM-11213-64.csv"
+    "worstRMSM-11213-32.csv"
+    "worstRMSM-11213-64.csv"
+)
+
+bestFileName=(
+    "bestRMSM-11213-32.csv"
+    "bestRMSM-11213-64.csv"
+    "bestRMSM-11213-32.csv"
+    "bestRMSM-11213-64.csv"
+    "bestRMSM-11213-32.csv"
+    "bestRMSM-11213-64.csv"
+)
 
 outputFolder="../content/benchmarks/sm-rm"
 
 count=${#originalBenchmarks[@]}
 
-cp header.txt ${worstFileName}
-cp header.txt ${bestFileName}
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp header.txt ${worstFileName[i]}
+    cp header.txt ${bestFileName[i]}
+done
 
 for i in $(seq 0 $(expr ${count} - 1));
 do
@@ -83,7 +101,11 @@ do
     ./calculateDifferences ${originalBenchmarks[i]} ${parrowsBenchmarks[i]} ${bestFileName} True False
 done
 
-cp ${worstFileName} ${outputFolder}
-cp ${bestFileName} ${outputFolder}
+
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp ${worstFileName[i]} ${outputFolder}
+    cp ${bestFileName[i]} ${outputFolder}
+done
 
 rm *.csv

### 2017-11-24 paper/utils/redoSudokuSM.sh

@@ -61,15 +61,33 @@ params=(
 
 benchmark="\"Sudoku (Shared-Memory)\""
 
-worstFileName="worstSudoku.csv"
-bestFileName="bestSudoku.csv"
+worstFileName=(
+    "worstSudoku-1000.csv"
+    "worstSudoku-16000.csv"
+    "worstSudoku-1000.csv"
+    "worstSudoku-16000.csv"
+    "worstSudoku-1000.csv"
+    "worstSudoku-16000.csv"
+)
+
+bestFileName=(
+    "bestSudoku-1000.csv"
+    "bestSudoku-1600.csv"
+    "bestSudoku-1000.csv"
+    "bestSudoku-1600.csv"
+    "bestSudoku-1000.csv"
+    "bestSudoku-1600.csv"
+)
 
 outputFolder="../content/benchmarks/sudoku-sm"
 
 count=${#originalBenchmarks[@]}
 
-cp header.txt ${worstFileName}
-cp header.txt ${bestFileName}
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp header.txt ${worstFileName[i]}
+    cp header.txt ${bestFileName[i]}
+done
 
 for i in $(seq 0 $(expr ${count} - 1));
 do
@@ -83,7 +101,11 @@ do
     ./calculateDifferences ${originalBenchmarks[i]} ${parrowsBenchmarks[i]} ${bestFileName} True False
 done
 
-cp ${worstFileName} ${outputFolder}
-cp ${bestFileName} ${outputFolder}
+
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp ${worstFileName[i]} ${outputFolder}
+    cp ${bestFileName[i]} ${outputFolder}
+done
 
 rm *.csv

### 2017-11-24 paper/utils/redoTorus.sh

@@ -31,15 +31,23 @@ params=(
 
 benchmark="\"Torus (Distributed)\""
 
-worstFileName="worstTorus.csv"
-bestFileName="bestTorus.csv"
+worstFileName=(
+    "worstTorus-4096.csv"
+)
+
+bestFileName=(
+    "bestTorus-4096.csv"
+)
 
 outputFolder="../content/benchmarks/distributed-torus"
 
 count=${#originalBenchmarks[@]}
 
-cp header.txt ${worstFileName}
-cp header.txt ${bestFileName}
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp header.txt ${worstFileName[i]}
+    cp header.txt ${bestFileName[i]}
+done
 
 for i in $(seq 0 $(expr ${count} - 1));
 do
@@ -53,7 +61,11 @@ do
     ./calculateDifferences ${originalBenchmarks[i]} ${parrowsBenchmarks[i]} ${bestFileName} True False
 done
 
-cp ${worstFileName} ${outputFolder}
-cp ${bestFileName} ${outputFolder}
+
+for i in $(seq 0 $(expr ${count} - 1));
+do
+    cp ${worstFileName[i]} ${outputFolder}
+    cp ${bestFileName[i]} ${outputFolder}
+done
 
 rm *.csv

### 2017-11-24 paper/content/benchmarks.tex
initial version of all worst benchmarks

@@
 \centering
-\label{my-label}
-\begin{subtable}[t]{.5\linewidth}%
+\label{worstResults}
+\captionsetup[subtable]{justification=centering}
+\begin{subtable}[t]{.49\linewidth}%
     \centering%
     \begingroup\catcode`"=9
-	\begin{tabular}{l R{1cm} R{1.3cm} R{1.3cm}}
+	\begin{tabular}{l R{0.7cm} R{1.3cm} R{1.3cm}}
 		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
-		vs.             & cores	& params   & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/sudoku-sm/worstSudoku.csv}{}{\vs & \nCores & \params & \factor \\}
+		vs.             & cores & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/sudoku-sm/worstSudoku-1000.csv}{}{\vs & \nCores & \factor \\}
 	\end{tabular}
 	\endgroup
-    \caption{Sudoku}\label{table1}
+    \caption{\sudokutest \\ 1000}\label{sudokuWorst-1000}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{l R{0.7cm} R{1.3cm} R{1.3cm}}
+		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+		vs.             & cores & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/sudoku-sm/worstSudoku-16000.csv}{}{\vs & \nCores & \factor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\sudokutest \\ 16000}\label{sudokuWorst-16000}
+\end{subtable}
+\newline
+\vspace*{1 cm}
+\newline
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{l R{0.7cm} R{1.3cm} R{1.3cm}}
+		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+		vs.             & cores & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/sm-rm/worstRMSM-11213-32.csv}{}{\vs & \nCores & \factor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\rmtest \\ 11213 32}\label{sm-rmWorst-11213-32}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{l R{0.7cm} R{1.3cm} R{1.3cm}}
+		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+		vs.             & cores & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/sm-rm/worstRMSM-11213-64.csv}{}{\vs & \nCores & \factor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\rmtest \\ 11213 64}\label{sm-rmWorst-11213-64}
+\end{subtable}
+\newline
+\vspace*{1 cm}
+\newline
+\begin{subtable}[t]{.32\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{l R{0.7cm} R{1.3cm} R{1.3cm}}
+		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+		vs.             & cores & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-torus/worstTorus-4096.csv}{}{\vs & \nCores & \factor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\torustest (dist) 4096}\label{torusWorst-4096}
+\end{subtable}
+\begin{subtable}[t]{.32\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{l R{0.7cm} R{1.3cm} R{1.3cm}}
+		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+		vs.             & cores & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-rm/worstRM-44497-256.csv}{}{\vs & \nCores & \factor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\rmtest \\ (dist) 44497 256}\label{dist-rmWorst-44497}
+\end{subtable}
+\begin{subtable}[t]{.32\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{l R{0.7cm} R{1.3cm} R{1.3cm}}
+		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+		vs.             & cores & factor \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-jacobi/worstJacobi-4253.csv}{}{\vs & \nCores & \factor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\jacobitest \\ (dist) 4253}\label{jacobiWorst-4253}
 \end{subtable}

### 2017-11-24 paper/content/main.tex
moving things back to the parent folder. only synctex has to be fixed now.
wip
added olegl changes
wip on benchmarks for Torus in the paper
wip restructuring the paper, wip on motivation's contribution
tmp
further restructuring according to Phil's remarks. moved related work to 2

@@
 \documentclass{jfp1}
-
-%% add a tweak to polycode, similar to literate in lstlistings, but uglier
-%format |>>>|        = "\mathbin{|\!\!>\!\!>\!\!>\!\!|}"
-%format parcomp        = "\mathbin{|\!\!>\!\!>\!\!>\!\!|}"
-%format >>>        = "\mathbin{>\!\!>\!\!>}"
-%% ^^^^ you want this. and similar for ***, +++, etc.
-%% a hack for list comprehensions' <-, typeset it as <--
-%%%%%format <--        = "\in "
-%format DOLLAR = "\mathbin{\$}"
-%%%% %format ` = "\`"
-%%%% %% works fine without
-%format ||| = "\mathbin{\mid\!\mid\!\mid}"
-%format pipepipepipe = "\mathbin{\mid\!\mid\!\mid}"
-%format |&&&| = "\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}"
-%format parand = "\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}"
-%format &&& = "\mathbin{\&\!\&\!\&}"
-%format |***| = "\mathbin{\mid\!\!*\!*\!*\!\!\mid}"
-%format parstar = "\mathbin{\mid\!\!*\!*\!*\!\!\mid}"
-%format *** = "\mathbin{*\!*\!*}"
-%format |+++| = "\mathbin{\mid\!\!+\!\!+\!\!+\!\!\mid}"
-%format +++ = "\mathbin{+\!\!+\!\!+}"
-
-%include polycode.fmt
 
@@
 
-\jdate{April 2017}
+\jdate{August 2017}
 \pubyear{2017}
@@
 
-%include prelude.tex
+\input{lhs2TeXSafeCommands}
+\input{prelude}
 
@@
 %\affiliation{University Bayreuth, Germany and Glasgow University, UK}
- \author[M. Braun, P. Trinder and O. Lobachev]%
+ \author[M. Braun, O. Lobachev and P. Trinder]%
         {MARTIN BRAUN\\
          University Bayreuth, 95440 Bayreuth, Germany\\
-         PHIL TRINDER\\
-		 Glasgow University, Glasgow, G12 8QQ, Scotland\\
-		 \and\ OLEG LOBACHEV\\
-		 University Bayreuth, 95440 Bayreuth, Germany}
+		 OLEG LOBACHEV\\
+		 University Bayreuth, 95440 Bayreuth, Germany\\
+		 \and\ PHIL TRINDER\\
+		 Glasgow University, Glasgow, G12 8QQ, Scotland}
 }% end ifthenelse
@@
 
+\def\SymbReg{\textsuperscript{\textregistered}}
+
 \maketitle
@@
 %% environment inside
-%include abstract.tex
+\input{abstract.tex}
 
@@
 	%\newpage
-	\tableofcontents
 	%\pagebreak
-	%include motivation.tex
-	%include parallelHaskells.tex
-	%include arrows.tex
-	%\pagebreak
-	%include relwork.tex
+	\input{motivation}
+	\input{relwork}
+	\begin{flushright}
+
+	\end{flushright}\section{Background}
+	\label{sec:background}
+	\input{arrows}
+	\input{parallelHaskells}
 	%\pagebreak
 	%\pagebreak
-	%include parrows.tex
 	%\pagebreak
-	%include basicSkeletons.tex
+	\input{parrows}
 	%\pagebreak
-	%include syntacticSugar.tex
+	\input{basicSkeletons}
 	%\pagebreak
-	%include futures.tex
+	\input{syntacticSugar}
 	%\pagebreak
-	%include mapSkeletons.tex
+	\input{futures}
 	%\pagebreak
-	%include topologySkeletons.tex
+	\input{mapSkeletons}
 	%\pagebreak
-	%include benchmarks.tex
+	\input{topologySkeletons}
+	%\pagebreak
+	\input{benchmarks}
 	%%\pagebreak
-	%include conclusion.tex
+	\input{conclusion}
 	%\pagebreak
@@
         \appendix
-	%include utilityFunctions.tex
+	\input{utilityFunctions}
 \end{document}

### 2017-11-24 paper/content/bestAndWorstBenchmarks.tex
best and worst benchmarks redone in one table for easier lookup

\begin{table}[]
\centering
\label{bestResults}
\captionsetup[subtable]{justification=centering}
\begin{subtable}[t]{.49\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
		vs.             & cores & factor & cores & factor \\ \hline
		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
	\end{tabular}
	\endgroup
    \caption{\sudokutest \\ 1000}\label{sudokuBest-1000}
\end{subtable}
\begin{subtable}[t]{.49\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\	
		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
		vs.             & cores & factor & cores & factor \\ \hline
		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-16000.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
	\end{tabular}
	\endgroup
    \caption{\sudokutest \\ 16000}\label{sudokuBest-16000}
\end{subtable}
\newline
\vspace*{0.5 cm}
\newline
\begin{subtable}[t]{.49\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
		vs.             & cores & factor & cores & factor \\ \hline
		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
	\end{tabular}
	\endgroup
    \caption{\rmtest 11213 32}\label{sm-rmBest-11213-32}
\end{subtable}
\begin{subtable}[t]{.49\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\	
		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
		vs.             & cores & factor & cores & factor \\ \hline
		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
	\end{tabular}
	\endgroup
    \caption{\rmtest 11213 64}\label{sm-rmBest-11213-64}
\end{subtable}
\newline
\vspace*{0.5 cm}
\newline
\begin{subtable}[t]{.49\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\	
		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
		vs.             & cores & factor & cores & factor \\ \hline
		\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
	\end{tabular}
	\endgroup
    \caption{\torustest (dist) 4096}\label{torusBest-4096}
\end{subtable}
\begin{subtable}[t]{.49\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\	
		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
		vs.             & cores & factor & cores & factor \\ \hline
		\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
	\end{tabular}
	\endgroup
    \caption{\rmtest (dist) 44497 256}\label{dist-rmBest-44497}
\end{subtable}
\newline
\vspace*{0.5 cm}
\newline
\begin{subtable}[t]{.49\linewidth}%
    \centering%
    \begingroup\catcode`"=9
	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\	
		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
		vs.             & cores & factor & cores & factor \\ \hline
		\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-4253.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
	\end{tabular}
	\endgroup
    \caption{\jacobitest (dist) 4253}\label{jacobiBest-4253}
\end{subtable}
\caption{Best \& worst Benchmark results}
\end{table}

### 2017-11-25 paper/content/benchmarks.tex
minor
best and worst benchmarks redone in one table for easier lookup

--- 2017-11-23 paper/content/benchmarks.tex
@@
 
-\subsection{Hardware}
+\subsection{Hardware and Software used}
 
 In the following we describe the benchmarks of our parallel DSL and algorithmic skeletons we conducted. They were run both in a shared and in a distributed
-memory setting. All benchmarks were done on the \enquote{Glasgow grid}, consisting of
+memory setting. All benchmarks were done on \enquote{the Glasgow GPG Beowulf cluster}, consisting of
 16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each.
@@
 Non-surprisingly, we found that hyper-threaded 32 cores do not behave in the same manner as real 16 cores (numbers here for a single machine).
-We disregarded the hyper-threading ability in most of the
-cases.
+We disregarded the hyper-threading ability in most of the cases.
+
+Apart from Eden, all benchmarks and libaries were compiled with Stack's \footnote{see https://www.haskellstack.org/} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0.
+Stack itself was used in version 1.3.2.
+For GpH in its Multicore variant we used the the parallel cabal package in version 3.2.1.0, while for the |Par| monad we used monad-par in version 0.3.4.8.
+For all Eden tests we used its GHC-Eden compiler in version 7.8.2 together with OpenMPI 1.6.5.
 
@@
 \begin{table}
-\caption{The benchmarks we use in this paper.}
-\label{tab:benches}
 \centering
@@
 \sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad} 
-\tablefootnote{actual code from: http://community.haskell.org/~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}\\
+\tablefootnote{actual code from: http://community.haskell.org/\~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}\\
 \bottomrule
 \end{tabular}
+\label{tab:benches}
+\caption{The benchmarks we use in this paper.}
 \end{table}
 
-\rmtest is a probabilistic primality test that iterates multiple (32--256 here)
+\rmtest is a probabilistic primality test that iterates multiple (here: 32--256)
 \enquote{subtests}. Should a subtest fail, the input is definitely not a
@@
 
-\begin{table}[]
-\centering
-\label{my-label}
-\begin{subtable}[t]{.5\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l R{1cm} R{1.3cm} R{1.3cm}}
-		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
-		vs.             & cores	& params   & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/sudoku-sm/worstSudoku.csv}{}{\vs & \nCores & \params & \factor \\}
-	\end{tabular}
-	\endgroup
-    \caption{Sudoku}\label{table1}
-\end{subtable}
-\caption{Worst Benchmark results}
-\end{table}
+\input{bestAndWorstBenchmarks}
 
@@
 
-The PArrows version of \rmtest on a single node of the Glasgow grid
+The PArrows version of \rmtest on a single node of the Glasgow cluster
 showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
@@
 213.172 for 256 real cores. Apparently, hyper-threading in the Glasgow
-grid fails to execute two parallel Haskell processes with full-fledged
+cluster fails to execute two parallel Haskell processes with full-fledged
 parallelism. For this reason, we did not regard hyper-threaded cores in
@@
   different backends on the same hardware. Measurements were performed on a single node of the Glasgow
-  grid; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
+  cluster; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
   tasks. The
@@
 {\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{272}\label{subfig:rm-dist-b}}
-\caption{Parallel performance of the \rmtest on the Glasgow grid
+\caption{Parallel performance of the \rmtest on the Glasgow cluster
   consisting of 256 cores. Input was $2^{44497}-1$, we used 256
@@
 {\performanceJacobiDistDiff{3}{3217}{32}{5}{272}\label{subfig:jacobi-dist-b}}
-\caption{Parallel performance of the \jacobitest on the Glasgow grid
+\caption{Parallel performance of the \jacobitest on the Glasgow cluster
   consisting of 256 cores. Input was $2^{3217}-1$, we used 256
@@
 {\performanceTorusDistDiff{4096}{32}{0.5}{272}\label{subfig:performancetorusDistDiff}}
-\caption{Parallel performance of \torustest on the Glasgow grid
+\caption{Parallel performance of \torustest on the Glasgow cluster
   consisting of 256 cores. Input was a matrix size of $4096$. The top plot shows absolute speedup in a distributed memory setting. The
@@
   different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
-  grid; it has 16 real cores and 32 threads. Input was a file of $1000$ Sudokus. The
+  cluster; it has 16 real cores and 32 threads. Input was a file of $1000$ Sudokus. The
   closer to linear speedup the better.}
@@
   different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
-  grid; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
+  cluster; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
   closer to linear speedup the better. The GpH version shows signs of choking with too many sparks being created.}

### 2017-11-25 paper/content/benchmarks.tex
.

@@
 
-Apart from Eden, all benchmarks and libaries were compiled with Stack's \footnote{see https://www.haskellstack.org/} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0.
+Apart from Eden, all benchmarks and libaries were compiled with Stack's \footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0.
 Stack itself was used in version 1.3.2.
 For GpH in its Multicore variant we used the the parallel cabal package in version 3.2.1.0, while for the |Par| monad we used monad-par in version 0.3.4.8.
-For all Eden tests we used its GHC-Eden compiler in version 7.8.2 together with OpenMPI 1.6.5.
+For all Eden tests, we used its GHC-Eden compiler in version 7.8.2 together with OpenMPI 1.6.5.
+
+Furtermore, all benchmarks were done with help of the bench \footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2) internally.
+All runtime data (mean runtime, max stddev, etc.) was collected with this tool if not mentioned otherwise.
 
@@
 
-\subsection{Test programs}
+\subsection{Benchmarks}
 
@@
 \toprule
-Name & Area & Type & Origin & Citation \\
+Name & Area & Type & Origin & Source \\
 \midrule
@@
 \enquote{Gentleman} is a standard Eden test program, developed
-for their \ensuremath{\Varid{torus}} skeleton. It implements a parallel matrix
+for their \ensuremath{\Varid{torus}} skeleton. It implements a Gentleman's algorithm for parallel matrix
 multiplication \citep{Gentleman1978}. We ported an Eden based version \citep{Eden:SkeletonBookChapter02} to PArrows.
@@
 A~parallel Sudoku solver was used by \citet{par-monad} to compare \ensuremath{\Conid{Par}} Monad
-to GpH Haskell. We ported it to PArrows.
+to GpH, and we ported it to PArrows.
 
@@
 
-The \ensuremath{\Conid{Par}} monad and GpH Haskell -- in its multicore version \cite{Marlow2009} --  can be executed on a shared
+The \ensuremath{\Conid{Par}} monad and GpH -- in its multicore version \cite{Marlow2009} --  can be executed on a shared
 memory machines only. Although GpH is available on distributed memory
@@
 
-The PArrows version of \rmtest on a single node of the Glasgow cluster
-showed almost linear speedup (Fig.~\ref{fig:bench-rm-sm}). The speedup
-of 64-task PArrows/Eden at 16 real cores version was 13.65, the
-efficiency was 85.3\%. However, if we increase the number of
-requested cores to be 32 -- \ie if we use hyper-threading on 16 real
-cores -- the speedup does not increase that well. It is merely 15.99
-for 32 tasks with PArrows/Eden. It is worse for other backends.  As
-for 64 tasks, we obtain the speedup of 16.12 with PArrows/Eden at 32
-hyper-threaded cores and only 13.55 with PArrows/GpH. Efficiency is 50.4\% and 42.3\%, respectively. The Eden
+In preliminary tests, the PArrows version of \rmtest on a single node of the Glasgow cluster
+showed almost linear speedup on up to 16 shared-memory cores (Fig.~\ref{fig:bench-rm-sm}). The speedup
+of 64-task PArrows/Eden at 16 real cores version was 13.65 giving a parallel
+efficiency of 85.3\%. However, if we increased the number of
+requested cores to 32 -- \ie if we use hyper-threading on 16 real
+cores -- the speedup did not increase that well. It was merely 15.99
+for 32 tasks with PArrows/Eden. This was worse for other backends.  As
+for 64 tasks, we obtained a speedup of 16.12 with PArrows/Eden at 32
+hyper-threaded cores and only 13.55 with PArrows/GpH. The parallel eficiencies were 50.4\% and 42.3\%, respectively. The Eden
 version used here was Eden~CP, the \enquote{share nothing} SMP build.
@@
 cluster fails to execute two parallel Haskell processes with full-fledged
-parallelism. For this reason, we did not regard hyper-threaded cores in
-our speedup plots in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark}.
+parallelism.
+
+Hence, the performance measurements in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark} and table~\ref{tab:bestAndWorstResults} use only real cores as the purpose of this paper is to show the performance of PArrows and not to investigate parallel performance with hyper-threading in use.
 

### 2017-11-25 paper/utils/calculateDifferences.hs
calculating overheads

@@ -32,7 +32,9 @@ data Speedup = Speedup {
     speedup :: Double,
     stdDev :: Double,
     factor :: Double,
-    factorStdDev :: Double
+    factorStdDev :: Double,
+    overhead :: Double,
+    stdDevForOverhead :: Double
 } deriving (Show)
 
 
@@ -67,7 +69,9 @@ main = do
                                                                         speedup = speedup,
                                                                         stdDev = stdDev,
                                                                         factor = 1,
-                                                                        factorStdDev = 1
+                                                                        factorStdDev = 0,
+                                                                        overhead = 0,
+                                                                        stdDevForOverhead = 0
                                                                     }
                                                    | otherwise = Nothing
                 lineToSpeedup _ = Nothing
@@ -83,12 +87,14 @@ main = do
                             stdDev = max (stdDev x) (stdDev y),
                             -- care: different order than time. cause reasons
                             factor = (time y) / (time x),
-                            factorStdDev = (max (stdDev x) (stdDev y)) / (time x)
+                            factorStdDev = (max (stdDev x) (stdDev y)) / (time x),
+                            overhead = (time y - time x) / (time y),
+                            stdDevForOverhead = (max (stdDev x) (stdDev y)) / (time y)
                         }
                     | otherwise = Nothing
 
                 legend :: String
-                legend = "\"\",\"name\",\"time\",\"nCores\",\"speedup\",\"max stddev\",\"factor\"" ++ "\n"
+                legend = "\"\",\"name\",\"time\",\"nCores\",\"speedup\",\"max stddev\",\"factor\",\"factorStdDev\",\"overhead\",\"stdDevForOverhead\"" ++ "\n"
 
                 str :: String -> String
                 str st = "\"" ++ st ++ "\""
@@ -99,7 +105,8 @@ main = do
                 toString :: Speedup -> String
                 toString x = (str (num x)) ++ "," ++ (str $ name x) ++ "," ++  (show $ time x) ++ "," ++
                                     (nCores x) ++ "," ++ (show $ speedup x) ++ "," ++ (show $ stdDev x) ++ "," ++
-                                    (show $ factor x)
+                                    (show $ factor x) ++ "," ++ (show $ factorStdDev x) ++ "," ++ (show $ overhead x) ++
+                                    "," ++ (show $ stdDevForOverhead x)
 
                 handleParse :: Either ParseError [[String]] -> [Speedup]
[…]

### 2017-11-25 paper/content/bestAndWorstBenchmarks.tex
minor
.

@@
-\begin{table}[]
+\newcommand{\bestAndWorstBenchmarks}{\begin{table}[]
 \centering
-\label{bestResults}
+\label{tab:bestAndWorstResults}
 \captionsetup[subtable]{justification=centering}
@@
 						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
-		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
 		vs.             & cores & factor & cores & factor \\ \hline
@@
 	\endgroup
-    \caption{\sudokutest \\ 1000}\label{sudokuBest-1000}
+    \caption{\sudokutest 1000}\label{sudokuBest-1000}
 \end{subtable}
@@
 	\endgroup
-    \caption{\sudokutest \\ 16000}\label{sudokuBest-16000}
+    \caption{\sudokutest 16000}\label{sudokuBest-16000}
 \end{subtable}
@@
 	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\	
-		% \multicolumn{3}{c}{worst Sudoku Benchmarks} \\ \hline
+						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\ \hline
 		vs.             & cores & factor & cores & factor \\ \hline
@@
 \end{table}
+}
+
+%\bestAndWorstBenchmarks
+
+\newcommand{\meanFactor}{
+\begin{table}[]
+\centering
+\label{tab:meanFactors}
+\captionsetup[subtable]{justification=centering}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{2cm} C{2cm}}
+		vs.             & mean factor & maximum normalized stdDev \\ \hline
+		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \meanFactor & \maxStdDevFactor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\sudokutest 1000}\label{sudokuMeanFactor-1000}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{2cm} C{2cm}}
+		vs.             & mean factor & maximum normalized stdDev \\ \hline
+		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-16000.csv}{}{\vs & \meanFactor & \maxStdDevFactor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\sudokutest 16000}\label{sudokuMeanFactor-16000}
+\end{subtable}
+\newline
+\vspace*{0.5 cm}
+\newline
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{2cm} C{2cm}}
+		vs.             & mean factor & maximum normalized stdDev \\ \hline
+		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{\vs & \meanFactor & \maxStdDevFactor \\}
+	\end{tabular}
+	\endgroup
+	\caption{\rmtest 11213 32}\label{sm-rmMeanFactor-11213-32}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{2cm} C{2cm}}
+		vs.             & mean factor & maximum normalized stdDev \\ \hline
+		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{\vs & \meanFactor & \maxStdDevFactor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\rmtest 11213 64}\label{sm-rmMeanFactor-11213-64}
+\end{subtable}
+\newline
+\vspace*{0.5 cm}
+\newline
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{2cm} C{2cm}}
+		vs.             & mean factor & maximum normalized stdDev \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{\vs & \meanFactor & \maxStdDevFactor \\}
+	\end{tabular}
+	\endgroup
+	\caption{\torustest (dist) 4096}\label{torusMeanFactor-4096}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{2cm} C{2cm}}
+		vs.             & mean factor & maximum normalized stdDev \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{\vs & \meanFactor & \maxStdDevFactor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\rmtest (dist) 44497 256}\label{dist-rmMeanFactor-44497}
+\end{subtable}
+\newline
+\vspace*{0.5 cm}
+\newline
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{2cm} C{2cm}}
+		vs.             & mean factor & maximum normalized stdDev \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-4253.csv}{}{\vs & \meanFactor & \maxStdDevFactor \\}
+	\end{tabular}
+	\endgroup
+    \caption{\jacobitest (dist) 4253}\label{jacobiMeanFactor-4253}
+\end{subtable}
+\caption{Overhead in the benchmarks}
+\end{table}
+}
+
+\meanFactor

### 2017-11-27 paper/content/bestAndWorstBenchmarks.tex

--- 2017-11-24 paper/content/bestAndWorstBenchmarks.tex
@@
 \end{table}
+}
+
+%\bestAndWorstBenchmarks
+
+\newcommand{\meanOverhead}{
+\begin{table}[]
+\centering
+\label{tab:meanOverhead}
+\captionsetup[subtable]{justification=centering}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+    \caption{\sudokutest 1000}\label{sudokuMeanFactor-1000}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-16000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+    \caption{\sudokutest 16000}\label{sudokuMeanFactor-16000}
+\end{subtable}
+\newline
+\vspace*{0.5 cm}
+\newline
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+	\caption{\rmtest 11213 32}\label{sm-rmMeanFactor-11213-32}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+    \caption{\rmtest 11213 64}\label{sm-rmMeanFactor-11213-64}
+\end{subtable}
+\newline
+\vspace*{0.5 cm}
+\newline
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+	\caption{\torustest (dist) 4096}\label{torusMeanFactor-4096}
+\end{subtable}
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+    \caption{\rmtest (dist) 44497 256}\label{dist-rmMeanFactor-44497}
+\end{subtable}
+\newline
+\vspace*{0.5 cm}
+\newline
+\begin{subtable}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-4253.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+    \caption{\jacobitest (dist) 4253}\label{jacobiMeanFactor-4253}
+\end{subtable}
+\caption{Overhead in the benchmarks}
+\end{table}
+}
+
+\meanOverhead

### 2017-11-27 paper/content/benchmarks.tex
minor

@@
 However it still proves that PArrows continue to scale, even if not perfectly for this test program.
-Comparing PArrows with Eden, for the larger input of $2^{4253}-1$, we see slightly bigger differences between Eden and PArrows: The differences for 128 and 256 coresare $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden, respectively.
+Comparing PArrows with Eden, for the larger input of $2^{4253}-1$, we see slightly bigger differences between Eden and PArrows: The differences for 128 and 256 core sare $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden, respectively.
 This maximum of $12.27\%$ slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.

### 2017-11-27 paper/content/topologySkeletons.tex
minor

@@
 
-This action is, however,  more complicated than in the ring case as we have one more dimension of inputs to be transformed.
-We first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|.
+This action is, however,  more complicated than in the ring case as we have one more dimension of inputs that needs to be transformed.
+In order to achieve that, we first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|.
 We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| type class.
@@
 \end{figure}
-As an example of using this skeleton \citep{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
+As an example of using this skeleton \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
 An adapted version can be found in Fig.~\ref{fig:torusMatMult}.

### 2017-11-28 paper/content/benchmarks.tex
added proper introduction to benchmarks chapter

@@
 \label{sec:benchmarks}
+
+In the following section, we describe the benchmarks of our parallel DSL and algorithmic skeletons conducted.
+We start by explaining the Hardware \& Software stack and also elaborate on which Benchmarks programs were used
+and also which parallel Haskell were used in which setting. Before we go into detail on the benchmarks we
+also address the issue of poor performance gains when using hyper-threading and our reasoning behind not including
+hyper-threaded cores in our benchmarks.
+Finally, we show that PArrows hold up in terms of performance when compared to the original parallel Haskells used as backends in this paper, starting with the shared-memory variants (GpH, |Par| Monad and Eden CP) and concluding with Eden as a distributed backend.
 
@@
 
-In the following we describe the benchmarks of our parallel DSL and algorithmic skeletons we conducted. They were run both in a shared and in a distributed
+Benchmarks were run both in a shared and in a distributed
 memory setting. All benchmarks were done on \enquote{the Glasgow GPG Beowulf cluster}, consisting of

### 2017-11-28 paper/content/benchmarks.tex
minor

@@
 In the following section, we describe the benchmarks of our parallel DSL and algorithmic skeletons conducted.
-We start by explaining the Hardware \& Software stack and also elaborate on which Benchmarks programs were used
+We start by explaining the hardware and software stack and also elaborate on which benchmarks programs were used
 and also which parallel Haskell were used in which setting. Before we go into detail on the benchmarks we
@@
 Benchmarks were run both in a shared and in a distributed
-memory setting. All benchmarks were done on \enquote{the Glasgow GPG Beowulf cluster}, consisting of
+memory setting. All benchmarks were done on the Glasgow GPG Beowulf cluster, consisting of
 16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each.

### 2017-11-28 paper/content/bestAndWorstBenchmarks.tex
better table WIP
Signed-off-by: Martin Braun <martinbraun123@aol.com>

@@
-\newcommand{\bestAndWorstBenchmarks}{\begin{table}[]
-\centering
-\label{tab:bestAndWorstResults}
-\captionsetup[subtable]{justification=centering}
-\begin{subtable}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
-		vs.             & cores & factor & cores & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
+\renewcommand{\multirow}[3]{#3}
+\newcommand{\meanOverhead}{
+\begin{table}
+\caption{Best and worst benchmark results on shared memory multiprocessors.}
+\begin{tabular}{l c S S}
+\toprule
+	\multirow{4}{c}{\sudokutest 1000}    &	 
+    {backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+    \midrule
+    \multirow{4}{c}{\sudokutest 16000}    &	 
+	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+    \csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-16000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+    \midrule
+    \multirow{4}{c}{\rmtest 11213 32}    &	 
+	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+    \csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+    \midrule
+    \multirow{4}{c}{\rmtest 11213 64}    &	 
+	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+    \csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+    \bottomrule
     \end{tabular}
-	\endgroup
-    \caption{\sudokutest 1000}\label{sudokuBest-1000}
-\end{subtable}
-\begin{subtable}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
-		vs.             & cores & factor & cores & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-16000.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
+    \label{tab:overhead-smp}
+\end{table}
+\begin{table}
+\caption{Best and worst benchmark results on a distributed memory grid.}
+\begin{tabular}{l c S S}
+\toprule
+	\multirow{2}{c}{\torustest 4096}    &	 
+    {backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+		\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+    \midrule
+    \multirow{2}{c}{\rmtest 44497 256}    &	 
+	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+    \csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+        \midrule
+    \multirow{2}{c}{\jacobitest 3217}    &	 
+	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+    \csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-3217.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+    \midrule
+    \multirow{2}{c}{\jacobitest 4253}    &	 
+	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
+    \csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-4253.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+    \bottomrule
     \end{tabular}
-	\endgroup
-    \caption{\sudokutest 16000}\label{sudokuBest-16000}
-\end{subtable}
-\newline
-\vspace*{0.5 cm}
-\newline
-\begin{subtable}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
-		vs.             & cores & factor & cores & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
-	\end{tabular}
-	\endgroup
-    \caption{\rmtest 11213 32}\label{sm-rmBest-11213-32}
-\end{subtable}
-\begin{subtable}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
-		vs.             & cores & factor & cores & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
-	\end{tabular}
-	\endgroup
-    \caption{\rmtest 11213 64}\label{sm-rmBest-11213-64}
-\end{subtable}
-\newline
-\vspace*{0.5 cm}
-\newline
-\begin{subtable}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\ \hline
-		vs.             & cores & factor & cores & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
-	\end{tabular}
-	\endgroup
-    \caption{\torustest (dist) 4096}\label{torusBest-4096}
-\end{subtable}
-\begin{subtable}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
-		vs.             & cores & factor & cores & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
-	\end{tabular}
-	\endgroup
-    \caption{\rmtest (dist) 44497 256}\label{dist-rmBest-44497}
-\end{subtable}
-\newline
-\vspace*{0.5 cm}
-\newline
-\begin{subtable}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{l C{0.8cm} C{1.3cm} |@{\hspace{0.5em}} C{0.8cm} C{1.3cm}}
-						& \multicolumn{2}{c}{best} & \multicolumn{2}{c}{worst} \\
-		vs.             & cores & factor & cores & factor \\ \hline
-		\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-4253.csv}{}{\vs & \bestNCores & \bestFactor & \worstNCores & \worstFactor \\}
-	\end{tabular}
-	\endgroup
-    \caption{\jacobitest (dist) 4253}\label{jacobiBest-4253}
-\end{subtable}
-\caption{Best \& worst Benchmark results}
+    \label{tab:overhead-grid}
 \end{table}
@@
 
-\newcommand{\meanOverhead}{
+\newcommand{\meanOverheadOLD}{
 \begin{table}[]

### 2017-11-28 paper/content/bestAndWorstBenchmarks.tex
fixed tables again

@@
-\renewcommand{\multirow}[3]{#3}
-\newcommand{\meanOverhead}{
-\begin{table}
-\caption{Best and worst benchmark results on shared memory multiprocessors.}
-\begin{tabular}{l c S S}
-\toprule
-	\multirow{4}{c}{\sudokutest 1000}    &	 
-    {backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+%\bestAndWorstBenchmarks
+
+\newcommand{\meanOverheadTEST}{
+\begin{table}[]
+\centering
+\caption{Overhead in the benchmarks}
+\label{tab:meanOverhead}
+\captionsetup[subtable]{justification=centering}
+\subfloat[\sudokutest 1000]{
+	\begin{minipage}[t]{\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{3cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{Benchmark} & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \toprule
+		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \benchmark \params & \meanOverhead & \maxStdDevForOverhead \\}
 		\midrule
-    \multirow{4}{c}{\sudokutest 16000}    &	 
-	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-    \csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-16000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-    \midrule
-    \multirow{4}{c}{\rmtest 11213 32}    &	 
-	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-    \csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-    \midrule
-    \multirow{4}{c}{\rmtest 11213 64}    &	 
-	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-    \csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-    \bottomrule
 	\end{tabular}
-    \label{tab:overhead-smp}
-\end{table}
-\begin{table}
-\caption{Best and worst benchmark results on a distributed memory grid.}
-\begin{tabular}{l c S S}
-\toprule
-	\multirow{2}{c}{\torustest 4096}    &	 
-    {backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-		\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-    \midrule
-    \multirow{2}{c}{\rmtest 44497 256}    &	 
-	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-    \csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-        \midrule
-    \multirow{2}{c}{\jacobitest 3217}    &	 
-	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-    \csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-3217.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-    \midrule
-    \multirow{2}{c}{\jacobitest 4253}    &	 
-	{backend}             & {mean overhead} & {max.\ normalised stdDev} \\ \midrule
-    \csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-4253.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-    \bottomrule
-    \end{tabular}
-    \label{tab:overhead-grid}
+	\endgroup
+	\label{sudokuMeanFactor-1000}
+	\end{minipage}
+}
+\caption{Overhead in the benchmarks}
 \end{table}
@@
 
-%\bestAndWorstBenchmarks
-
-\newcommand{\meanOverheadOLD}{
+\newcommand{\meanOverhead}{
 \begin{table}[]
 \centering
+\caption{Overhead in the benchmarks}
 \label{tab:meanOverhead}
 \captionsetup[subtable]{justification=centering}
-\begin{subtable}[t]{.49\linewidth}%
+\subfloat[\sudokutest 1000]{
+	\begin{minipage}[t]{.49\linewidth}%
     \centering%
@@
 	\endgroup
-    \caption{\sudokutest 1000}\label{sudokuMeanFactor-1000}
-\end{subtable}
-\begin{subtable}[t]{.49\linewidth}%
+	\label{sudokuMeanFactor-1000}
+	\end{minipage}
+}
+\subfloat[\sudokutest 16000]{
+	\begin{minipage}[t]{.49\linewidth}%
     \centering%
@@
 	\endgroup
-    \caption{\sudokutest 16000}\label{sudokuMeanFactor-16000}
-\end{subtable}
+    \label{sudokuMeanFactor-16000}
+	\end{minipage}
+}
 \newline
@@
 \newline
-\begin{subtable}[t]{.49\linewidth}%
+\subfloat[\rmtest 11213 32]{
+	\begin{minipage}[t]{.49\linewidth}%
     \centering%
@@
 	\endgroup
-	\caption{\rmtest 11213 32}\label{sm-rmMeanFactor-11213-32}
-\end{subtable}
-\begin{subtable}[t]{.49\linewidth}%
+	\label{sm-rmMeanFactor-11213-32}
+	\end{minipage}
+}
+\subfloat[\rmtest 11213 64]{
+	\begin{minipage}[t]{.49\linewidth}%
     \centering%
@@
 	\endgroup
-    \caption{\rmtest 11213 64}\label{sm-rmMeanFactor-11213-64}
-\end{subtable}
+    \label{sm-rmMeanFactor-11213-64}
+	\end{minipage}
+}
 \newline
@@
 \newline
-\begin{subtable}[t]{.49\linewidth}%
+\subfloat[\torustest (dist) 4096]{
+	\begin{minipage}[t]{.49\linewidth}%
     \centering%
@@
 	\endgroup
-	\caption{\torustest (dist) 4096}\label{torusMeanFactor-4096}
-\end{subtable}
-\begin{subtable}[t]{.49\linewidth}%
+	\label{torusMeanFactor-4096}
+	\end{minipage}
+}
+\subfloat[\rmtest (dist) 44497 256]{
+	\begin{minipage}[t]{.49\linewidth}%
     \centering%
@@
 	\endgroup
-    \caption{\rmtest (dist) 44497 256}\label{dist-rmMeanFactor-44497}
-\end{subtable}
+    \label{dist-rmMeanFactor-44497}
+	\end{minipage}
+}
 \newline
@@
 \newline
-\begin{subtable}[t]{.49\linewidth}%
+\subfloat[\jacobitest (dist) 3217]{
+	\begin{minipage}[t]{.49\linewidth}%
+    \centering%
+    \begingroup\catcode`"=9
+	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
+		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+		\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-3217.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+	\end{tabular}
+	\endgroup
+    \label{jacobiMeanFactor-3217}
+	\end{minipage}
+}
+\subfloat[\jacobitest (dist) 4253]{
+	\begin{minipage}[t]{.49\linewidth}%
     \centering%
@@
 	\endgroup
-    \caption{\jacobitest (dist) 4253}\label{jacobiMeanFactor-4253}
-\end{subtable}
-\caption{Overhead in the benchmarks}
+    \label{jacobiMeanFactor-4253}
+	\end{minipage}
+}
 \end{table}

### 2017-11-28 paper/content/motivation.tex
minor
reworking contributions

@@
 common \ptcomment{quantify, e.g. less than 10 \%} \mbcomment{we do this in the contributions section of the Introduction, enough?}, very low-penalty programming interface that allows to switch
-the parallel backends at will. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
+the parallel backends at will. The induced penalty was max. 8.3 \% in our measurements. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
 
@@
 We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow.
-Because of this, we do not lose any benefits of using Arrows as the parallelism is encapsulated as another combinator instead of a different type.
-The resulting Arrow can be used in the same way as a potential serial version.
+A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.
 
-This is a big advantage as we do not introduce any new types as Monad solutions like the |Par| Monad do.
-We can just \enquote{plug} in parallel parts into sequential Arrow-based programs and libraries without having to change integral parts of the code.
-\mbcomment{mention Functions here?
-We have everything Eden has, but more.}
+We evaluate the expressive power of Arrow formalism in the context of parallel programming.
+
+\begin{itemize}
+\item We introduce a parallel evaluation formalism using Arrows.
+One big advantage is that we do not introduce any new types (Sec.~\ref{sec:parallel-arrows}) in contrast Monad solutions such as |Par| Monad.
+This behaviour encourages better composability.
+\item We utilitze multiple backends -- currently a GpH, a |Par| Monad, and Eden.
+We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
+The backends can easily be swapped, so we are not bound to any specific one.
+
+So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.
+(Sec.~\ref{sec:parallel-arrows})
+\item We extend our PArrows formalism with |Future|s.
+Our goal here is to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|).
+Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node.
+(Sec.~\ref{sec:futures})
+\item It is possible to define algorithmic skeletons with PArrows (Sec.~\ref{sec:skeletons})
+\item Finally, we demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low overhead (Sec.~\ref{sec:benchmarks}).
+\end{itemize}
 
 \mbcomment{Überleitung auf Related Work hier machen}
-
-We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
-This way, we can have multiple backends -- currently a GpH, a |Par| Monad and an Eden version are supported.
-The backends can easily be swapped, so we are not bound to any specific one.
-So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.
-
-We introduce a |Future| type class in Section \ref{sec:futures} to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|).
-Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node.
-As the |Par| Monad and GpH backends (in their current form) only support shared memory execution and do not have to fix any communication problems, we only use a wrapping dummy |BasicFuture|.
-
-We show that it is possible to define algorithmic skeletons with Arrows (Sections \ref{sec:map-skeletons},~\ref{sec:topology-skeletons}) and finally demonstrate that Arrow Parallelism is a viable alternative to existing approaches and prove that only low overhead is introduced in Section \ref{sec:benchmarks}, i.e.
-less than $1.7\%$ in cases outside of the error bar of the measurement.
-\mbcomment{fix this after benchmarks are finally done}.
 

### 2017-11-28 paper/content/abstract.tex
tmp

@@
 Furthermore, as each parallel computation is an Arrow, PArrows can be readily composed and transformed as such.
-In order to allow for more sophisticated communication schemes between computation nodes in distributed systems we introduce the concept of Futures to wrap existing direct communication concepts in backends.
+In order to allow for more sophisticated communication schemes between computation nodes in distributed systems we utilise the concept of Futures to wrap existing direct communication concepts in backends.
 
-To show that PArrows have similar expressiveness to existing parallel languages, we also implement several parallel skeletons. 
+To show that PArrows have similar expressive power as existing parallel languages, we also implement several parallel skeletons. 
 Benchmarks show that our framework does not induce any notable overhead

### 2017-11-28 paper/content/benchmarks.tex
minor
tmp

@@
 
-Apart from Eden, all benchmarks and libaries were compiled with Stack's \footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0.
+Apart from Eden, all benchmarks and libaries were compiled with Stack's\footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0.
 Stack itself was used in version 1.3.2.
-For GpH in its Multicore variant we used the the parallel cabal package in version 3.2.1.0, while for the |Par| monad we used monad-par in version 0.3.4.8.
-For all Eden tests, we used its GHC-Eden compiler in version 7.8.2 together with OpenMPI 1.6.5.
+For GpH in its Multicore variant we used the the parallel package in version 3.2.1.0\footnote{see \url{https://hackage.haskell.org/package/parallel-3.2.1.0}}, while for the |Par| monad we used monad-par in version 0.3.4.8\footnote{see \url{https://hackage.haskell.org/package/monad-par-0.3.4.8}}.
+For all Eden tests, we used its GHC-Eden compiler in version 7.8.2\footnote{see \url{http://www.mathematik.uni-marburg.de/~eden/?content=build_eden_7_&navi=build}} together with OpenMPI 1.6.5\footnote{see \url{https://www.open-mpi.org/software/ompi/v1.6/}}.
 
-Furtermore, all benchmarks were done with help of the bench \footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2) internally.
+Furtermore, all benchmarks were done with help of the bench\footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2)\footnote{see \url{https://hackage.haskell.org/package/criterion-1.1.1.0}} internally.
 All runtime data (mean runtime, max stddev, etc.) was collected with this tool if not mentioned otherwise.
@@
 \centering
+\caption{The benchmarks we use in this paper.}
 %% something was wrong with separators in table
@@
 \torustest & Mathematics & \ensuremath{\Varid{torus}} & Eden & \citet{Eden:SkeletonBookChapter02}\\
-\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad} 
-\tablefootnote{actual code from: http://community.haskell.org/\~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}\\
+\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad}\tablefootnote{actual code from: \url{http://community.haskell.org/\~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}}\\
 \bottomrule
@@
 \label{tab:benches}
-\caption{The benchmarks we use in this paper.}
 \end{table}

### 2017-11-28 paper/content/bestAndWorstBenchmarks.tex
tmp

@@
 
-\newcommand{\meanOverheadTEST}{
 \begin{table}[]
@@
 \captionsetup[subtable]{justification=centering}
-\subfloat[\sudokutest 1000]{
-	\begin{minipage}[t]{\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{3cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{Benchmark} & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \toprule
-		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \benchmark \params & \meanOverhead & \maxStdDevForOverhead \\}
-		\midrule
-	\end{tabular}
-	\endgroup
-	\label{sudokuMeanFactor-1000}
-	\end{minipage}
+\sisetup{
+	table-number-alignment = center
 }
-\caption{Overhead in the benchmarks}
-\end{table}
-}
-
-\newcommand{\meanOverhead}{
-\begin{table}[]
-\centering
-\caption{Overhead in the benchmarks}
-\label{tab:meanOverhead}
-\captionsetup[subtable]{justification=centering}
 \subfloat[\sudokutest 1000]{
@@
     \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
+	\begin{tabular}{S S[table-text-alignment = center] S[table-text-alignment = center]}
+		vs.             & mean overhead & maximum normalized stdDev \\ \hline
 		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
@@
 \end{table}
-}
-
-\meanOverhead

### 2017-11-28 paper/content/mapSkeletons.tex
reworking contributions
tmp

@@
+\section{Skeletons}
+\label{sec:skeletons}
+Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations.
+While there are many possible skeletons to implement, we regard in detail here only some |map|-based and topological skeletons to demonstrate the power of PArrows.
 %%% \FloatBarrier
-\section{Map-based Skeletons}
+\subsection{|map|-based Skeletons}
 \label{sec:map-skeletons}
-Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations.
+We start with |map|-based skeletons.
+The essential differences between the skeletons presented here are in terms of order of evaluation and work distribution but still provide the same output of a standard |map|.
 

### 2017-11-28 paper/content/motivation.tex
tmp

@@
 the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
-novel manner - we tame the zoo of parallel Haskells. We provide a
-common \ptcomment{quantify, e.g. less than 10 \%} \mbcomment{we do this in the contributions section of the Introduction, enough?}, very low-penalty programming interface that allows to switch
-the parallel backends at will. The induced penalty was max. 8.3 \% in our measurements. Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
+novel manner -- we tame the zoo of parallel Haskells. We provide a
+common, very low-penalty programming interface that allows to switch
+the parallel backends at will.
+The induced penalty was max.
+8.3 \% in our measurements \mbcomment{, and typically under X \%}.
+Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
 
@@
 %\olcomment{move this to Contributions in the front or something}
-We propose a new Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow.
+We propose an Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow.
 A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.
 
-We evaluate the expressive power of Arrow formalism in the context of parallel programming.
+In this paper evaluate the expressive power of our Arrow formalism in the context of parallel programming.
 
@@
 \item We introduce a parallel evaluation formalism using Arrows.
-One big advantage is that we do not introduce any new types (Sec.~\ref{sec:parallel-arrows}) in contrast Monad solutions such as |Par| Monad.
+One big advantage is that we do not have to introduce any new types (Sec.~\ref{sec:parallel-arrows}) in contrast to Monad solutions such as the |Par| Monad.
 This behaviour encourages better composability.
@@
 Our goal here is to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|).
-Direct communication is useful in a distributed memory setting because they allow for inter-node communication without blocking the master-node.
+Direct communication is useful in a distributed memory setting because it allows for inter-node communication without blocking the master-node.
 (Sec.~\ref{sec:futures})
-\item It is possible to define algorithmic skeletons with PArrows (Sec.~\ref{sec:skeletons})
-\item Finally, we demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low overhead (Sec.~\ref{sec:benchmarks}).
+\item It is possible to define algorithmic skeletons with PArrows (Sec.~\ref{sec:skeletons}). All our benchmarks were skeleton-based.
+\item Finally, we practically demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low performance overhead (Sec.~\ref{sec:benchmarks}).
 \end{itemize}

### 2017-11-28 paper/content/benchmarks.tex
minor

@@
 
-\subsection{Hardware and Software used}
+\subsection{Preliminaries}
+
+\subsubsection{Hardware and Software used}
 
@@
 
-\subsection{Benchmarks}
+\subsubsection{Benchmarks}
 
@@
 \caption{The benchmarks we use in this paper.}
+\label{tab:benches}
 %% something was wrong with separators in table
@@
 \end{tabular}
-\label{tab:benches}
 \end{table}
@@
 
-\subsection{What parallel Haskells run where}
+\subsubsection{What parallel Haskells run where}
 
@@
 
-\subsection{Effect of hyper-threading}
+\subsubsection{Effect of hyper-threading}
 
@@
 
-Hence, the performance measurements in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark} and table~\ref{tab:bestAndWorstResults} use only real cores as the purpose of this paper is to show the performance of PArrows and not to investigate parallel performance with hyper-threading in use.
+Hence, the performance measurements in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark} and Table~\ref{tab:meanOverhead} use only real cores as the purpose of this paper is to show the performance of PArrows and not to investigate parallel performance with hyper-threading in use.
 
@@
 
-\subsection{Byline}
+\subsection{Discussion}
 

### 2017-11-29 paper/content/benchmarks.tex
boldfacing in table

@@
 \subsection{Benchmark results}
-
-The difference between, say, PArrows with \ensuremath{\Conid{Par}} Monad backend and a
-genuine \ensuremath{\Conid{Par}}
-Monad benchmark is very small.
-To give an example, it is $0.4\seconds$ in favour of PArrows for 16 cores ($10.8\seconds$ vs.
-$11.2\seconds$) and $-0.8\seconds$ in favour of the \ensuremath{\Conid{Par}} monad for 8 cores ($16.1\seconds$ vs.
-$16.9\seconds$) for
-the Sudoku benchmark in the shared memory setting. It is almost invisible in speedup and
-(non shown) run time plots. We thus show only the results for the
-PArrows-enabled versions in the backend-comparison.
-
 To show that PArrows induce very small overhead even in a distributed context (and in general as well), we compare the original Eden
@@
 versions. These are the values provided by the \ensuremath{\Varid{bench}} package that we
-used for bench-marking. We call a difference between two versions
+used for benchmarking. We call a difference between two versions
 significant when the border of the error bar of absolute time
@@
 \newcommand{\performanceSkelRMDistDiff}[5]{
-\performancediffplot{Run time differences\\for \rmtest \enquote{#1 #2}}{(Eden $-$ PArrows)}{256}{#3}{
+\performancediffplot{Overhead\\for \rmtest \enquote{#1 #2}}{(Eden $-$ PArrows)}{256}{#3}{
 \addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
 smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
@@
 \newcommand{\performanceJacobiDistDiff}[5]{
-\performancediffplot{Run time differences\\for \jacobitest \enquote{#2}}{(Eden $-$ PArrows) #2}{256}{#3}{
+\performancediffplot{Overhead\\for \jacobitest \enquote{#2}}{(Eden $-$ PArrows) #2}{256}{#3}{
 \addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
 smooth]{benchmarks/distributed-jacobi/#1-#2-diff.csv};
@@
 \newcommand{\performanceTorusDistDiff}[4]{
-\performancediffplot{Run time differences\\for \torustest \enquote{#1}}{(Eden $-$ PArrows)}{256}{#2}{
+\performancediffplot{Overhead\\for \torustest \enquote{#1}}{(Eden $-$ PArrows)}{256}{#2}{
 \addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="time", y error="max stddev", col sep=comma, mark=dots,
+    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
 smooth]{benchmarks/distributed-torus/#1-diff.csv};

### 2017-11-29 paper/separate_plots/main.tex
wip on additional plots

--- 2017-11-24 paper/content/main.tex
@@
 
-\input{lhs2TeXSafeCommands}
 \input{prelude}
 
-\title{Arrows for Parallel Computations}
+\title{Plots for \enquote{Arrows for Parallel Computations}}
 \ifthenelse{\boolean{anonymous}}{%
@@
 
-%% environment inside
-\input{abstract.tex}
-
 \tableofcontents
 
-	%
-	%%include abstract.lhs
-	%
-	%\newpage
-	%\pagebreak
-	\input{motivation}
-	\input{relwork}
-	\begin{flushright}
+\section{Plots}
 
-	\end{flushright}\section{Background}
-	\label{sec:background}
-	\input{arrows}
-	\input{parallelHaskells}
-	%\pagebreak
-	%\pagebreak
-	%\pagebreak
-	\input{parrows}
-	%\pagebreak
-	\input{basicSkeletons}
-	%\pagebreak
-	\input{syntacticSugar}
-	%\pagebreak
-	\input{futures}
-	%\pagebreak
-	\input{mapSkeletons}
-	%\pagebreak
-	\input{topologySkeletons}
-	%\pagebreak
-	\input{benchmarks}
-	%%\pagebreak
-	\input{conclusion}
-	%\pagebreak
-        \bibliographystyle{jfp}
-	\bibliography{references,main}
-        \appendix
-	\input{utilityFunctions}
+\newlength{\plotwidthDist}
+\setlength{\plotwidthDist}{0.6\textwidth}
+
+\newcommand{\meanOverheadPlot}[5]{
+\begin{tikzpicture}
+\begin{axis}[title={#1},
+title style={align=center},
+scale only axis, width=\plotwidthDist,
+xlabel=Threads,
+%xtick=data,
+ytick distance=#2 / 4,
+xtick distance=#3,
+minor tick num=9,
+ylabel=Overhead (\%),
+ylabel near ticks,
+grid=both,
+legend entries={Mean Overhead},
+legend style={at={(0.99,0.99)},anchor=north east},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
+ymin=-#2,
+ymax=#2,
+xmin=-1,
+xmax=#4]
+\addplot+[mark=*,very thick,error bars/.cd,
+    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
+smooth]{#5};
+\end{axis}
+\end{tikzpicture}
+}
+
+\newcommand{\rmtest}{Rabin--Miller test\xspace}
+\newcommand{\sudokutest}{Sudoku\xspace}
+\newcommand{\jacobitest}{Jacobi sum test\xspace}
+\newcommand{\torustest}{Gentleman\xspace}
+
+\meanOverheadPlot{Mean Overhead for distributed \rmtest 44497 256}{0.02}{32}{272}{../content/benchmarks/distributed-rm/44497-256-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for distributed \jacobitest 4253}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for distributed \jacobitest 3217}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for distributed \torustest 4096}{0.5}{32}{272}{../content/benchmarks/distributed-torus/4096-diff.csv}
+
 \end{document}

### 2017-11-29 paper/separate_plots/main.tex
wip separate plots

@@
 
-\meanOverheadPlot{Mean Overhead for distributed \rmtest 44497 256}{0.02}{32}{272}{../content/benchmarks/distributed-rm/44497-256-diff.csv}
+\subsection{Distributed}
+
+\subsubsection{\rmtest}
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \rmtest 44497 256}{0.02}{32}{272}{../content/benchmarks/distributed-rm/44497-256-diff.csv}
+
+\subsubsection{\jacobitest}
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 4253}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for distributed \jacobitest 4253}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
+\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
+
+\subsubsection{\torustest}
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \torustest 4096}{0.5}{32}{272}{../content/benchmarks/distributed-torus/4096-diff.csv}
+
+\subsection{Shared-Memory}
+
+\subsubsection{\rmtest}
+
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Eden CP}{0.5}{4}{18}{../content/benchmarks/sm-rm/eden-cp-11213-32-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for distributed \jacobitest 3217}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Eden CP}{0.5}{4}{18}{../content/benchmarks/sm-rm/eden-cp-11213-64-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for distributed \torustest 4096}{0.5}{32}{272}{../content/benchmarks/distributed-torus/4096-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs GpH}{0.5}{4}{18}{../content/benchmarks/sm-rm/mult-11213-32-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs GpH}{0.5}{4}{18}{../content/benchmarks/sm-rm/mult-11213-32-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Par Monad}{0.5}{4}{18}{../content/benchmarks/sm-rm/par-11213-32-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Par Monad}{0.5}{4}{18}{../content/benchmarks/sm-rm/par-11213-32-diff.csv}
+
+    %benchmarks/sudoku-sm
+    %    "eden-cp-1000-diff.csv"
+    %"eden-cp-16000-diff.csv"
+    %"mult-1000-diff.csv"
+    %"mult-16000-diff.csv"
+    %"par-1000-diff.csv"
+    %"par-16000-diff.csv"
 

### 2017-12-07 paper/content/arrows.tex
reworked some stuff in the arrows chapter

@@
+\subsection{Arrows}
+\label{sec:arrows}
+Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads.
+\citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator |(>>=) :: m a -> (a -> m b) -> m b| (with |m| being a Monad).
+
+An Arrow |arr a b| represents a computation that converts an input |a| to an output |b|. This is defined in the |Arrow| type class shown in Fig.~\ref{fig:ArrowDefinition}.
+%
+To lift an ordinary function to an Arrow, |arr| is used, analogous to the monadic |return|.
+Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by \enquote{wiring} the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
+Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched.
+It allows us to to save input across arrows.
+Figure~\ref{fig:arrows-viz} shows a graphical representation of these basic Arrow combinators.
+The most prominent instances of this interface are regular functions |(->)|
+and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps monadic functions, e.g.  |a -> m b|.
+
+Note that with the Kleisli type we can define an alternative to |(>>=)| with the signature |m a -> Kleisli m a b -> m b|.
+The difference between the definitions of Arrows and Monads becomes more clear when we compare this definition with the Arrow compose operator |(>>>)|: Monads are types that allow composition of computations, while Arrows \emph{are} computations.
+
 \begin{figure}[t]
@@
 \end{figure}
-
-\subsection{Arrows}
-\label{sec:arrows}
-Arrows were introduced by \citet{HughesArrows} as a general interface for computation.
-An Arrow |arr a b| represents  a computation that converts an input |a| to an output |b|.
-This is defined in the |Arrow| type class shown in Fig.~\ref{fig:ArrowDefinition}.
-%
-To lift an ordinary function to an Arrow, |arr| is used, analogous to the monadic |return|.
-Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by \enquote{wiring} the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
-Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched.
-It allows us to to save input across arrows.
-Figure~\ref{fig:arrows-viz} shows a graphical representation of these basic Arrow combinators.
-The most prominent instances of this interface are regular functions |(->)|
-and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps monadic functions, e.g.  |a -> m b| (with |m| being a Monad).
 
@@
 
-The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a Monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}.
-This is exactly one of the key challenges in parallel computing.
-
 %%% Local Variables:

### 2017-12-07 paper/content/conclusion.tex
reworked some stuff in the arrows chapter

@@
 programs. We do not have to introduce new monadic types that wrap the
-computation. Instead, we use Arrows in the same manner one uses sequential pure functions. 
+computation.
+Instead, we use Arrows in the same manner one uses sequential pure functions.
+Even though we make heavy use of the |ArrowApply| typeclass - which only Arrows that are equivalent to Monads can satisfy - in this paper, we believe that the general idea of an Arrow, the process of doing something - a \emph{computation}, is the perfect fit to use in a library for parallel computation as it allows users to think about parallel programs more as a set of composed computations - like a data-flow graph.
+Furthermore, our approach allows for a more natural paralellization of any Arrow type (especially regular functions (|->|)), when compared to Monads as we do not have to change anything in order to run them in parallel.
 %
@@
 %
-Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead in our benchmarks.
+Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they only introduce minor overhead in some of our benchmarks.
 %

### 2017-12-07 paper/content/conclusion.tex
minor

@@
 Instead, we use Arrows in the same manner one uses sequential pure functions.
-Even though we make heavy use of the |ArrowApply| typeclass - which only Arrows that are equivalent to Monads can satisfy - in this paper, we believe that the general idea of an Arrow, the process of doing something - a \emph{computation}, is the perfect fit to use in a library for parallel computation as it allows users to think about parallel programs more as a set of composed computations - like a data-flow graph.
+Even though we make heavy use of the |ArrowApply| typeclass - which only Arrows that are equivalent to Monads can satisfy - in this paper, we believe that the general idea of an Arrow, the process of doing something - a composable \emph{computation}, is the perfect fit to use in a library for parallel computation as it allows users to think about parallel programs more as a set of composed computations - like a data-flow graph.
 Furthermore, our approach allows for a more natural paralellization of any Arrow type (especially regular functions (|->|)), when compared to Monads as we do not have to change anything in order to run them in parallel.

### 2017-12-07 Definition/src/main/Parrows/Definition.hs
removed dependency to ArrowApply from almost all combinators, only foldlArr and parMapReduceDirect require it now. Maybe this can be changed as well, though

@@ -69,7 +69,7 @@ toPar = return
 (<||||>) = (++)
 
 -- spawns the first n arrows to be evaluated in parallel. this works for infinite lists of arrows as well
-parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) => conf  -> ChunkSize -> [arr a b] -> arr [a] [b]
+parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr) => conf  -> ChunkSize -> [arr a b] -> arr [a] [b]
 parEvalNLazy conf chunkSize fs =
                -- chunk the functions, feed the function chunks into parEvalN, chunk the input accordingly
                -- evaluate the function chunks in parallel and concat the input to a single list again

### 2017-12-07 Definition/src/main/Parrows/Skeletons/Map.hs

@@ -41,7 +41,7 @@ parMap conf f = parEvalN conf (repeat f)
 
 -- contrary to parMap this schedules chunks of a given size (parMap has "chunks" of length = 1) to be
 -- evaluated on the same thread
-parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) => conf -> ChunkSize -> arr a b -> arr [a] [b]
+parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr) => conf -> ChunkSize -> arr a b -> arr [a] [b]
 parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
 
 -- similar to parMapStream, but divides the input list by the given number
@@ -51,7 +51,7 @@ farm conf numCores f = unshuffle numCores >>>
                        shuffle
 
 -- farm and parMapStream combined. divide the input list and inside work in chunks
-farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf, ArrowChoice arr, ArrowApply arr) => conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
+farmChunk :: (ArrowParallel arr a b conf, ArrowParallel arr [a] [b] conf, ArrowChoice arr) => conf -> ChunkSize -> NumCores -> arr a b -> arr [a] [b]
 farmChunk conf chunkSize numCores f = unshuffle numCores >>>
                                       parEvalNLazy conf chunkSize (repeat (mapArr f)) >>>
                                       shuffle

### 2017-12-07 Definition/src/main/Parrows/Skeletons/Topology.hs

@@ -77,16 +77,15 @@ ring conf f =
         arr unzip)
 
 --TODO: check whether this exchanges the futures the same way as Eden does it
-torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
+torus :: (ArrowLoop arr, ArrowChoice arr,
             FutureEval arr (c, fut a, fut b) (d, fut a, fut b) conf,
             Future fut a, Future fut b) =>
          conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
 torus conf f =
     loop (second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy)) >>>
         arr (uncurry3 (zipWith3 lazyzip3)) >>>
-        (arr length >>> arr unshuffle) &&&
-            (shuffle >>> evalN conf (repeat $ ptorus f)) >>>
-        app >>>
+        arr length &&& (shuffle >>> evalN conf (repeat (ptorus f))) >>>
+        arr (uncurry unshuffle) >>>
         arr (map unzip3) >>> arr unzip3 >>> threetotwo)
 
 uncurry3 :: (a -> b -> c -> d) -> (a, (b, c)) -> d

### 2017-12-07 Definition/src/main/Parrows/Util.hs

@@ -33,15 +33,16 @@ import           Data.List
 zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
 zipWithArr zipFn = arr (uncurry zip) >>> mapArr zipFn
 
-listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
-listApp fs = arr (\as -> (fs, as)) >>> zipWithArr app
+listApp :: (ArrowChoice arr) => [arr a b] -> arr [a] [b]
+listApp (f:fs) = arr listcase >>>
+         arr (const []) ||| (f *** listApp fs >>> arr (uncurry (:)))
+         where listcase []     = Left ()
+               listcase (x:xs) = Right (x,xs)
+listApp [] = arr (const [])
 
 -- from http://www.cse.chalmers.se/~rjmh/afp-arrows.pdf
 mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
-mapArr f = arr listcase >>>
-         arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
-         where listcase []     = Left ()
-               listcase (x:xs) = Right (x,xs)
+mapArr = listApp . repeat
 
 -- fold on Arrows inspired by mapArr
 foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b

### 2017-12-07 Multicore/src/main/Parrows/Multicore.hs

@@ -38,12 +38,12 @@ import           Control.Parallel.Strategies
 
 data Conf a = Conf (Strategy a)
 
-instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
+instance (NFData b, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
     parEvalN (Conf strat) fs =
         listApp fs >>>
         arr (withStrategy (parList strat))
 
-instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
+instance (NFData b, ArrowChoice arr) => ArrowParallel arr a b () where
     parEvalN _ fs = parEvalN (hack fs) fs
                     where
                         hack :: (NFData b) => [arr a b] -> Conf b
@@ -54,8 +54,8 @@ data BasicFuture a = BF a
 instance NFData a => NFData (BasicFuture a) where
     rnf (BF a) = rnf a
 
-instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
-    ArrowParallel arr a b conf) => FutureEval arr a b conf where
+instance (Arrow arr, ArrowChoice arr, ArrowParallel arr a b conf) 
+    => FutureEval arr a b conf where
     evalN _ = listApp
 
 instance Future BasicFuture a where

### 2017-12-07 ParMonad/src/main/Parrows/ParMonad.hs

@@ -35,9 +35,8 @@ import           Control.Arrow
 import           Control.DeepSeq
 import           Control.Monad.Par
 
-instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b conf where
-    parEvalN _ fs = arr (\as -> (fs, as)) >>>
-                    zipWithArr (app >>> arr spawnP) >>>
+instance (NFData b, ArrowChoice arr) => ArrowParallel arr a b conf where
+    parEvalN _ fs = listApp (map (>>> arr spawnP) fs) >>>
                     arr sequenceA >>>
                     arr (>>= mapM Control.Monad.Par.get) >>>
                     arr runPar
@@ -47,8 +46,7 @@ data BasicFuture a = BF a
 instance NFData a => NFData (BasicFuture a) where
     rnf (BF a) = rnf a
 
-instance (ArrowChoice arr, ArrowApply arr,
-    ArrowParallel arr a b conf) => FutureEval arr a b conf where
+instance (ArrowChoice arr, ArrowParallel arr a b conf) => FutureEval arr a b conf where
     evalN _ = listApp
 
 instance Future BasicFuture a where

### 2017-12-07 paper/content/arrows.tex

@@
 Note that with the Kleisli type we can define an alternative to |(>>=)| with the signature |m a -> Kleisli m a b -> m b|.
-The difference between the definitions of Arrows and Monads becomes more clear when we compare this definition with the Arrow compose operator |(>>>)|: Monads are types that allow composition of computations, while Arrows \emph{are} computations.
+The difference between the definitions of Arrows and Monads becomes more clear if we compare this with the signature of the Arrow compose operator |(>>>) :: arr a b -> arr b c -> arr a c|: Monads are types that allow composition of computations, while Arrows \emph{are} composable computations.
 

### 2017-12-07 paper/content/bestAndWorstBenchmarks.tex
minor

@@
 \centering
-\caption{Overhead in the benchmarks}
-\label{tab:meanOverhead}
-\captionsetup[subtable]{justification=centering}
-\sisetup{
-	table-number-alignment = center
-}
-\subfloat[\sudokutest 1000]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
+\caption{Overhead in the shared memory benchmarks}
+\label{tab:meanOverheadSharedMemory}
+\centering
 \begingroup\catcode`"=9
-	\begin{tabular}{S S[table-text-alignment = center] S[table-text-alignment = center]}
-		vs.             & mean overhead & maximum normalized stdDev \\ \hline
-		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+\begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
+	\thead{Benchmark} & \thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 16 cores} \\ \hline \\
+	\multirow{3}{*}{\sudokutest 1000}
+	\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{3}{*}{\rmtest 11213 32}
+	\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{3}{*}{\rmtest 11213 64}
+	\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 \end{tabular}
 \endgroup
-	\label{sudokuMeanFactor-1000}
-	\end{minipage}
-}
-\subfloat[\sudokutest 16000]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
+\end{table}
+
+\begin{table}[]
+\centering
+\caption{Overhead in the distributed memory benchmarks}
+\label{tab:meanOverHeadDistributedMemory}
+\centering
 \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
-		\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-16000.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
+\begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
+	\thead{Benchmark} & \thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 256 cores} \\ \hline \\
+	\multirow{1}{*}{\torustest 4096}
+	\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{1}{*}{\rmtest 44497 256}
+	\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{1}{*}{\jacobitest 3217}
+	\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-3217.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 \end{tabular}
 \endgroup
-    \label{sudokuMeanFactor-16000}
-	\end{minipage}
-}
-\newline
-\vspace*{0.5 cm}
-\newline
-\subfloat[\rmtest 11213 32]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
-		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-	\end{tabular}
-	\endgroup
-	\label{sm-rmMeanFactor-11213-32}
-	\end{minipage}
-}
-\subfloat[\rmtest 11213 64]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
-		\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-	\end{tabular}
-	\endgroup
-    \label{sm-rmMeanFactor-11213-64}
-	\end{minipage}
-}
-\newline
-\vspace*{0.5 cm}
-\newline
-\subfloat[\torustest (dist) 4096]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
-		\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-	\end{tabular}
-	\endgroup
-	\label{torusMeanFactor-4096}
-	\end{minipage}
-}
-\subfloat[\rmtest (dist) 44497 256]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
-		\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-	\end{tabular}
-	\endgroup
-    \label{dist-rmMeanFactor-44497}
-	\end{minipage}
-}
-\newline
-\vspace*{0.5 cm}
-\newline
-\subfloat[\jacobitest (dist) 3217]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
-		\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-3217.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-	\end{tabular}
-	\endgroup
-    \label{jacobiMeanFactor-3217}
-	\end{minipage}
-}
-\subfloat[\jacobitest (dist) 4253]{
-	\begin{minipage}[t]{.49\linewidth}%
-    \centering%
-    \begingroup\catcode`"=9
-	\begin{tabular}{C{2cm} C{1.7cm} C{2.3cm}}
-		\thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} \\ \hline
-		\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-4253.csv}{}{\vs & \meanOverhead & \maxStdDevForOverhead \\}
-	\end{tabular}
-	\endgroup
-    \label{jacobiMeanFactor-4253}
-	\end{minipage}
-}
 \end{table}

### 2017-12-12 paper/content/parallelHaskells.tex
wip rewrite parallelHaskells.tex parrows.tex, etc.

@@
 \subsubsection{Glasgow parallel Haskell -- GpH}
+\label{sec:GpHIntro}
 GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{The Multicore implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
@@
 parEvalN fs as = runPar $ 
-	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
+	(sequenceA (map (return . spawn) (zipWith ($) fs as))) >>= mapM get
 \end{code}
@@
 % $
-Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|.
+Then, we map over this not yet evaluated lazy list of results |[b]| with |spawn :: NFData a => Par a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|.
 We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| Monad with |get|.
@@
 We execute this process with |runPar| to finally get |[b]|.
+While we used |spawn| in the definition above, a head-strict variant can easily be defined by replacing |spawn| with |spawn_ :: Par a -> Par (IVar a)|.
 Fig.~\ref{fig:parEvalNParMonadImg} shows a graphical representation.

### 2017-12-12 paper/content/parrows.tex
wip rewrite parallelHaskells.tex parrows.tex, etc.

@@
 Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
-This combinator is similar to our utility function |listApp| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
+This combinator is similar to our utility function |evalN| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
 % \begin{figure}[h]
@@
 % \end{figure}
-We do not require the |conf| parameter in every implementation.
-If it is not needed, we usually just default the |conf| type parameter to |()| and even blank it out in the parameter list of the implemented |parEvalN|.
 % , as we will see in the implementation of the GpH Haskell and the |Par| Monad backends.
@@
 \subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
-The GpH implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| combinators from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
+The GpH implementation of |ArrowParallel| is implemented in a straightforward manner, but a bit different to the variant from Section \ref{sec:GpHIntro} by using |evalN :: [arr a b] -> arr [a] [b]| (think |zipWith ($) fs as| on Arrows, definition in Appendix~\ref{utilfns}) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
+Our |Conf a| datatype simply wraps a |Strategy a|, but could be extended in future versions of our DSL.
 \begin{figure}[t]
@@
 
-instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
-    	parEvalN _ fs =
-       		listApp fs >>>
-        	arr (withStrategy (parList rdeepseq))
-\end{code}% $ %% formatting
-\caption{Fully evaluating |ArrowParallel| instance for the GpH Haskell backend.}
-\label{fig:ArrowParallelMulticoreRdeepseq}
-\end{figure}
-For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the GpH interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this.
-We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
-With this definition in place, we can provide a delegating version of the non-configurable instance.
-We show this in Fig.~\ref{fig:actualNonConfigMulticore}.
-% \begin{figure}[h]
-% \caption{Definition of Conf a}
-% \label{fig:confa}
-% \end{figure}
-% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances.
-To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
-\begin{figure}[t]
-\begin{code}
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
+instance (NFData b, ArrowChoice arr) =>
   ArrowParallel arr a b (Conf b) where
     parEvalN (Conf strat) fs =
-        	listApp fs >>>
-        	arr (withStrategy (parList strat)) &&& arr id >>>
-        	arr (uncurry pseq)
-\end{code}
-\caption{Configurable |ArrowParallel| instance for the GpH Haskell backend.}
+        evalN fs >>>
+        arr (withStrategy (parList strat))
+\end{code}% $ %% formatting
+\caption{The |ArrowParallel| instance for the GpH Haskell backend.}
 \label{fig:ArrowParallelMulticoreConfigurable}
 \end{figure}
+
 \subsubsection{|Par| Monad}
 \olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
-The |Par| Monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| Monad's |spawnP :: NFData a => a -> Par (IVar a)| function.
-The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| Monad.
-
-We therefore apply each function to its corresponding input value with and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call.
-This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|.
-In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)|.
-The result of this operation is a |Par [b]| from which we can finally remove the Monad again by running |arr runPar| to get our output of |[b]|.
+Just like for GpH we can easily lift the definition of |parEvalN| to Arrows. To start off, we define the |Strategy a| and |Conf a| type so we can have a configurable instance of ArrowParallel:
+\begin{code}
+type Strategy a = a -> Par (IVar a)
+data Conf a = Conf (Strategy a)
+\end{code}
+Now we can once again define our |ArrowParallel| instance as follows: First, we convert our arrows |[arr a b]| with |evalN (map (>>> arr strat) fs)| into an arrow |arr [a] [(Par (IVar b))]| that yields composable computations in the |Par| monad, where |evalN| is once again of the type |evalN :: [arr a b] -> arr [a] [b]|.
+By combining the result of this arrow with |arr sequenceA|, we get an Arrow |arr [a] (Par [IVar b])|.
+Then, in order to fetch the results of the different threads, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)| - our intermediary Arrow is of type |arr [a] (Par [b])|.
+Finally, we execute the comptuation |Par [b]| by composing with |arr runPar| and get the final arrow |arr [a] [b]|.
 \begin{figure}[h]
 \begin{code}
-instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
-	ArrowParallel arr a b conf where
-		parEvalN _ fs = 
-			(arr $ \as -> (fs, as)) >>>
-			zipWithArr (app >>> arr spawnP) >>>
+instance (NFData b, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
+    parEvalN (Conf strat) fs = evalN (map (>>> arr strat) fs) >>>
                     arr sequenceA >>>
-			arr (>>= mapM get) >>>
+                    arr (>>= mapM Control.Monad.Par.get) >>>
                     arr runPar
@@
 \subsubsection{Eden}
-For both the GpH Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowApply| and |ArrowChoice| type classes.
-With Eden this is not the case as we can only spawn a list of functions and we cannot extract simple functions out of arrows.
-While we could still manage to have only one class in the module by introducing a type class: % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
+For both the GpH Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowChoice| type class.
+With Eden this is not the case as we can only spawn a list of functions, which we cannot extract from general arrows.
+While we could still manage to have only one instance in the module by introducing a type class % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
 % \begin{figure}[h]
@@
 % We don't do it here
-However, we avoid doing so for aesthetic reasons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
+, we avoid doing so for aesthetic reasons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
 % \begin{figure}[h]
 \begin{code}
-instance (Trans a, Trans b) => ArrowParallel (->) a b conf where
+instance (Trans a, Trans b) => ArrowParallel (->) a b (Conf b) where
     parEvalN _ fs as = spawnF fs as
@@
 \begin{code}
-instance (Monad m, Trans a, Trans b, Trans (m b)) =>
+instance (ArrowParallel (->) a (m b) (Conf b),
+  Monad m, Trans a, Trans b, Trans (m b)) =>
   ArrowParallel (Kleisli m) a b conf where
-parEvalN conf fs =
-	(arr $ parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
-	(Kleisli $ sequence)
+    parEvalN conf fs = arr (parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
+                       Kleisli sequence
 \end{code}
+where |Conf a| is simply defined as |data Conf a = Nil| since Eden does not have a configurable |spawnF| variant.
 % \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
@@
 % \end{figure}
+
+\subsubsection{Default configuration instances}
+While the configurability in the instances of the |ArrowParallel| instances above might be nice to have, users probably would prefer a proper default configuration for many parallel programs.
+These can easily be defined as well as we can see by the example of the default implementation of |ArrowParallel| for the GpH backend:
+
+\begin{code}
+instance (NFData b, ArrowChoice arr, ArrowParallel arr a b (Conf b)) => ArrowParallel arr a b () where
+    parEvalN _ fs = parEvalN (defaultConf fs) fs
+
+defaultConf :: (NFData b) => [arr a b] -> Conf b
+defaultConf fs = stratToConf fs rdeepseq
+
+stratToConf :: [arr a b] -> Strategy b -> Conf b
+stratToConf _ strat = Conf strat
+\end{code}
+
+The other backends have similarly structured implementations.
+Note that by restricting the original implementations of our backends to their specific |Conf| type, we also get interoperability between backends for free.
+We can parallelize one part of a program using one backend, and parallelize the next with another one.
+We can however only have one instance of |ArrowParallel arr a b ()| present at a time.
+
+\mbcomment{For convenience purposes, ArrowParallel arr a b () is defined as...}
+\mbcomment{Until now we described the arrow operations more literally, but in the following sectioins we only talk about the dataflow in general as the idea should be clear now.}
 

### 2017-12-12 paper/content/parrows.tex
minor

@@
 \end{code}
-Now we can once again define our |ArrowParallel| instance as follows: First, we convert our arrows |[arr a b]| with |evalN (map (>>> arr strat) fs)| into an arrow |arr [a] [(Par (IVar b))]| that yields composable computations in the |Par| monad, where |evalN| is once again of the type |evalN :: [arr a b] -> arr [a] [b]|.
-By combining the result of this arrow with |arr sequenceA|, we get an Arrow |arr [a] (Par [IVar b])|.
+Now we can once again define our |ArrowParallel| instance as follows: First, we convert our Arrows |[arr a b]| with |evalN (map (>>> arr strat) fs)| into an Arrow |arr [a] [(Par (IVar b))]| that yields composable computations in the |Par| monad, where |evalN| is once again of the type |evalN :: [arr a b] -> arr [a] [b]|.
+By combining the result of this Arrow with |arr sequenceA|, we get an Arrow |arr [a] (Par [IVar b])|.
 Then, in order to fetch the results of the different threads, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)| - our intermediary Arrow is of type |arr [a] (Par [b])|.
-Finally, we execute the comptuation |Par [b]| by composing with |arr runPar| and get the final arrow |arr [a] [b]|.
+Finally, we execute the comptuation |Par [b]| by composing with |arr runPar| and get the final Arrow |arr [a] [b]|.
 \begin{figure}[h]

### 2017-12-12 paper/content/parrows.tex
minor

@@
 \subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
-The GpH implementation of |ArrowParallel| is implemented in a straightforward manner, but a bit different to the variant from Section \ref{sec:GpHIntro} by using |evalN :: [arr a b] -> arr [a] [b]| (think |zipWith ($) fs as| on Arrows, definition in Appendix~\ref{utilfns}) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
+The GpH implementation of |ArrowParallel| is implemented in a straightforward manner, but a bit different compared to the variant from Section \ref{sec:GpHIntro}, by using |evalN :: [arr a b] -> arr [a] [b]| (think |zipWith ($) fs as| on Arrows, definition in Appendix~\ref{utilfns}) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
 Our |Conf a| datatype simply wraps a |Strategy a|, but could be extended in future versions of our DSL.

### 2017-12-12 Definition/src/main/Parrows/Future.hs
removed unnecessary comments

@@ -51,14 +51,6 @@ get' = arr (\(~(BF a)) -> a)
 instance NFData a => NFData (BasicFuture a) where
     rnf (BF a) = rnf a
 
-{-
-get :: (Arrow arr, Future fut a conf) => conf -> arr a (fut a)
-get conf = get' (typeToken [([]::a, [conf])])
-
-put :: (Arrow arr, Future fut a conf) => conf -> arr (fut a) a
-put conf = put' (typeToken _)
--}
-
 liftFut :: (Arrow arr, Future fut a conf, Future fut b conf) => conf -> arr a b -> arr (fut a) (fut b)
 liftFut conf f = get conf >>> f >>> put conf
 

### 2017-12-12 paper/content/arrows.tex

--- 2017-12-07 paper/content/arrows.tex
@@
 \citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator |(>>=) :: m a -> (a -> m b) -> m b| (with |m| being a Monad).
+In the example a parser of the type |Parser s a| with static meta information |s| and result |a| is shown to not be able to use the static information |s| without applying the monadic function |a -> m b|.
+With Arrows this is possible.
 
@@
 and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps monadic functions, e.g.  |a -> m b|.
-
-Note that with the Kleisli type we can define an alternative to |(>>=)| with the signature |m a -> Kleisli m a b -> m b|.
-The difference between the definitions of Arrows and Monads becomes more clear when we compare this definition with the Arrow compose operator |(>>>)|: Monads are types that allow composition of computations, while Arrows \emph{are} computations.
 
@@
 \end{code}
+
+As we can only rewrite the monadic bind operation |(>>=)| with the Kleisli type into |m a -> Kleisli m a b -> m b|, but not with a general Arrow |arr a b|, we can intuitively see that Arrows must be a generalisation of Monads.
+While this also means that a general Arrow can not express everything a Monad can, the Parser example shows that this trade-off is worth it in some cases.
+
+\mbcomment{fix this text...}
+In this paper we will show that we can express parallel computations with Arrows and do not require Monads to do so.
+We also do not restrict the applicable Arrows to ones which have |ArrowApply| instances as this would mean  We only use two additional typeclasses, |ArrowChoice| and |ArrowLoop| which allow us to express if-then-else like structures and loops with Arrows.
+We do however not require |ArrowApply| for anything in this paper meaning that we have a truly more generalised interface and do not restrict the type of Arrows applicable to the ones which are equal in expressive power to Monads.
+
 % \caption{Add over arrows}

### 2017-12-12 paper/content/parrows.tex

--- 2017-12-12 paper/content/parrows.tex
@@
 \subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
-The GpH implementation of |ArrowParallel| is implemented in a straightforward manner, but a bit different to the variant from Section \ref{sec:GpHIntro} by using |evalN :: [arr a b] -> arr [a] [b]| (think |zipWith ($) fs as| on Arrows, definition in Appendix~\ref{utilfns}) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
+The GpH implementation of |ArrowParallel| is implemented in a straightforward manner, but a bit different compared to the variant from Section \ref{sec:GpHIntro}, by using |evalN :: [arr a b] -> arr [a] [b]| (definition in Appendix~\ref{utilfns}, think |zipWith ($)| on Arrows) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
 Our |Conf a| datatype simply wraps a |Strategy a|, but could be extended in future versions of our DSL.
@@
 \end{code}% $ %% formatting
-\caption{The |ArrowParallel| instance for the GpH Haskell backend.}
+\caption{The |ArrowParallel| instance for the GpH backend.}
 \label{fig:ArrowParallelMulticoreConfigurable}
@@
 \begin{code}
-instance (NFData b, ArrowChoice arr, ArrowParallel arr a b (Conf b)) => ArrowParallel arr a b () where
+instance (NFData b, ArrowChoice arr, ArrowParallel arr a b (Conf b)) =>
+  ArrowParallel arr a b () where
     parEvalN _ fs = parEvalN (defaultConf fs) fs

### 2017-12-12 paper/content/parrows.tex
minor

@@
 \subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
-The GpH implementation of |ArrowParallel| is implemented in a straightforward manner, but a bit different compared to the variant from Section \ref{sec:GpHIntro}, by using |evalN :: [arr a b] -> arr [a] [b]| (definition in Appendix~\ref{utilfns}, think |zipWith ($)| on Arrows) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
+The GpH implementation of |ArrowParallel| is implemented in a straightforward manner in Fig.~\ref{fig:ArrowParallelMulticore}, but a bit different compared to the variant from Section \ref{sec:GpHIntro}, by using |evalN :: [arr a b] -> arr [a] [b]| (definition in Appendix~\ref{utilfns}, think |zipWith ($)| on Arrows) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
 Our |Conf a| datatype simply wraps a |Strategy a|, but could be extended in future versions of our DSL.
@@
 \caption{The |ArrowParallel| instance for the GpH backend.}
-\label{fig:ArrowParallelMulticoreConfigurable}
+\label{fig:ArrowParallelMulticore}
 \end{figure}
@@
 \olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
-Just like for GpH we can easily lift the definition of |parEvalN| to Arrows. To start off, we define the |Strategy a| and |Conf a| type so we can have a configurable instance of ArrowParallel:
+Just like for GpH we can easily lift the definition of |parEvalN| for the |Par| Monad to Arrows in Fig.~\ref{fig:ArrowParallelParMonad}.
+To start off, we define the |Strategy a| and |Conf a| type so we can have a configurable instance of ArrowParallel:
 \begin{code}

### 2017-12-12 Definition/src/main/Parrows/Util.hs
clean up dead code

@@ -40,7 +40,7 @@ evalN (f:fs) = arr listcase >>>
                listcase (x:xs) = Right (x,xs)
 evalN [] = arr (const [])
 
--- from http://www.cse.chalmers.se/~rjmh/afp-arrows.pdf
+-- inspiration from http://www.cse.chalmers.se/~rjmh/afp-arrows.pdf
 mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
 mapArr = evalN . repeat
 

### 2017-12-12 HdpH/examples/karatsuba_new/DivConRW.hs

@@ -1,318 +1,0 @@
-{-# OPTIONS -cpp #-}
-{-# LANGUAGE FlexibleContexts #-}
-{- Divide and conquer skeleton on top of replicated workers.
-   The first parameter of divCon defines the depth at which
-   the divide&conquer tree is generated by the main tasks,
-   using function generateTasks. After generating enough tasks,
-   the replicated workers skeleton is used to solve them with
-   a good load balance. Finally, function combineTop combine
-   the partial results obtained.
-   The data type Tree is needed to pass information from 
-   generateTasks to combineTop, so that the subresults can
-   be combined in the proper way.   
-   
-                                              -- FR10 --
--}                                              
-   
-
-module DivConRW where
--- (divConRW,divConDM,divConRW2,divConDM_N,divConDM_N_NoRes) where
-import Prelude 
---import Control.Parallel (pseq)
---import Control.Parallel.Strategies (parList,using,NFData)  
---import Control.Parallel.Strategies as M (parMap)
---import Control.Parallel.Strategies (rdeepseq,Strategy)
-
-import Control.DeepSeq
-import Control.Parallel.HdpH
-import Control.Parallel.HdpH.Strategies
-
---import ParallelSplit.ParMonad
-import Control.Arrow
-import Parrows.Definition as P
-import Parrows.HdpH
-import Data.Typeable
-
--- import Eden
--- import RW 
-
--- workpools by MD
--- import EdenMWToken (mwNest')
-
--- edi workpool, sorting
--- import EdiWP
-
-import Data.List
-import System.IO.Unsafe
-import Control.Monad
-import Data.List.Split
-import Data.Maybe
--- import Observe 
-
-rnf :: NFData a => Strategy a
-rnf = rdeepseq -- incorrect, rnf is old signature, but works this way with `using`
-
-
---seq = pseq 
-{-
-mw :: (Trans a, Trans b) => 
-      Int -> Int -> (a -> b) -> [a] -> [b]
-mw = ediWPf
-
--- farm:
-unshuffle :: Int -> [a] -> [[a]]
-unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
-takeEach :: Int -> [a] -> [a] 
-takeEach n [] = []
[…]

### 2017-12-12 HdpH/examples/karatsuba_new/newkara.hs

@@ -1,199 +1,0 @@
-{-# LANGUAGE FlexibleInstances #-}
-{-# LANGUAGE TemplateHaskell #-}
-{-# LANGUAGE FlexibleInstances #-}
-{-# LANGUAGE BangPatterns #-}
-{-# OPTIONS -cpp #-}
-module Main where
--- import Eden
-import System.Environment
--- import DivConN
-import DivConRW hiding (rnf)
-
-import Control.DeepSeq
-
-import Control.Monad
-import Data.Serialize
-
-import System.Environment (getArgs)
-import System.IO (stdout, stderr, hSetBuffering, BufferMode(..))
-
-
-import Control.Parallel.HdpH hiding (put, get)
-import Control.Parallel.HdpH.Strategies
-
-import Prelude hiding (seq) -- exported in DC
---import Control.Parallel.Eden (Trans)
-
--- parse runtime system config options; abort if there is an error
-parseOpts :: [String] -> IO (RTSConf, [String])
-parseOpts args = do
-  either_conf <- updateConf args defaultRTSConf
-  case either_conf of
-    Left err_msg                 -> error $ "parseOpts: " ++ err_msg
-    Right (conf, remaining_args) -> return (conf, remaining_args)
-
-
-
-usage = "I need 3 parameters!, parallel depth, "
-    ++ "no. of digits (for number 1 and 2)"
-
-main = let
-       i1' = read "8000"
-       i2' = read "8000"
-       d'  = read "4"
-       is  = concat (repeat [1..9])
-       mi1 = take i1' is
-       mi2 = take i2' is
-       tmp conf = karat conf 2 mi1 mi2
-       tmpSeq conf = karat conf (-1) mi1 mi2
-       in do 
-          hSetBuffering stdout LineBuffering
-  	  hSetBuffering stderr LineBuffering
-  	  opts_args <- getArgs
-          print opts_args
-          (conf, args) <- parseOpts opts_args 
-          print $ length (fromD (tmp conf d'))
-
-           --print $ length (fromD (tmp d'))
-           --defaultMain [ bgroup ((show i1') ++ ", " ++ (show i2') ++ ", " ++ (show d') ++ " depth") [
-           --                                 bench "par" $ nf tmp d'
-           --                            ]
-           --             ]
[…]

### 2017-12-12 examples/karatsuba/NewDC.hs

@@ -1,129 +1,0 @@
-module NewDC where
-
-import ParallelSplit.ParMonad
-import ParallelSplit.Definition
-import Control.Parallel.Strategies hiding (parMap)
-import Data.List
-import Control.Parallel
-import Control.DeepSeq (rnf)
-import Debug.Trace
-
-parMap' :: (NFData b) => (a -> b) -> [a] -> [b]
-parMap' f as = parMap f as
-
--- old Eden code below, has Trans context removed, but needs NFData context. added
-
--- Code from MapHacks.hs, new farm from https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html#farm might be better, but I doubt it
-farmBase :: NFData b => ([a] -> [[a]]) -- ^ input distribution function
-         -> ([[b]] -> [b]) -- ^ result combination function
-         -> (a ->  b)      -- ^ worker function
-         -> [a]            -- ^ input
-         -> [b]            -- ^ output
-
-farmBase distr comb f tasks = comb $ parMap' (map f) $ distr tasks
-
-farm' :: NFData b => Int     -- ^ number of child processes
-     -> ((a -> b) -> [a] -> [b]) -- ^ map type signature
-farm' n = farmBase (unshuffle n) shuffle
-
--- hack!!!
-num_cores = 8
-farm :: NFData b => (a -> b) -> [a] -> [b]
-farm = farm' num_cores
-
--- helpers
--- see also https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Auxiliary.html#unshuffle
--- | Round robin distribution - inverse to shuffle
---
-unshuffle :: Int      -- ^number of sublists
-             -> [a]   -- ^input list
-             -> [[a]] -- ^distributed output
-unshuffle n xs = [takeEach n (drop i xs) | i <- [0..n-1]]
-
-takeEach :: Int -> [a] -> [a]
-takeEach n [] = []
-takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
-
-
--- | Simple shuffling - inverse to round robin distribution
-shuffle :: [[a]]  -- ^ sublists
-           -> [a] -- ^ shuffled sublists
-shuffle = concat . transpose
-
--- Code from https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-DivConq.html
[…]

### 2017-12-12 examples/matrix_seq.hs

@@ -1,89 +1,0 @@
-{-# LANGUAGE FlexibleInstances, FlexibleContexts, UndecidableInstances,
-MultiParamTypeClasses #-}
-
-module Main where
-
-import Data.List
-
---import GHC.Conc
-
-import Control.Arrow
-import Control.DeepSeq
-
-import Control.DeepSeq
-
-import Control.Applicative
-import Data.Functor
-import Data.List.Split
-
-import System.Environment
-
-import System.Random
-
-import Debug.Trace
-
-type Vector = [Int]
-type Matrix = [Vector]
-
-dimX :: Matrix -> Int
-dimX = length
-
-dimY :: Matrix -> Int
-dimY = length . head
-
---from: https://rosettacode.org/wiki/Matrix_multiplication#Haskell
-foldlZipWith::(a -> b -> c) -> (d -> c -> d) -> d -> [a] -> [b]  -> d
-foldlZipWith _ _ u [] _          = u
-foldlZipWith _ _ u _ []          = u
-foldlZipWith f g u (x:xs) (y:ys) = foldlZipWith f g (g u (f x y)) xs ys
-
-foldl1ZipWith::(a -> b -> c) -> (c -> c -> c) -> [a] -> [b] -> c
-foldl1ZipWith _ _ [] _          = error "First list is empty"
-foldl1ZipWith _ _ _ []          = error "Second list is empty"
-foldl1ZipWith f g (x:xs) (y:ys) = foldlZipWith f g (f x y) xs ys
-
-multAdd::(a -> b -> c) -> (c -> c -> c) -> [[a]] -> [[b]] -> [[c]]
-multAdd f g xs ys = map (\us -> foldl1ZipWith (\u vs -> map (f u) vs) (zipWith g) us ys) xs
-
-matMult:: Num a => [[a]] -> [[a]] -> [[a]]
-matMult xs ys = multAdd (*) (+) xs ys
-
-matAdd :: Matrix -> Matrix -> Matrix
-matAdd x y
-    | dimX x /= dimX y = error "dimX x not equal to dimX y"
-    | dimY x /= dimY y = error "dimY x not equal to dimY y"
-    | otherwise = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
-
--- from: http://www.mathematik.uni-marburg.de/~eden/paper/edenCEFP.pdf (page 38)
-nodefunction :: Int                         -- torus dimension
-    -> ((Matrix, Matrix), [Matrix], [Matrix]) -- process input
[…]

### 2017-12-12 examples/sudoku/Sudoku.hs

@@ -1,150 +1,0 @@
---
--- Sudoku solver using constraint propagation.  The algorithm is by
--- Peter Norvig http://norvig.com/sudoku.html; the Haskell
--- implementation is by Manu and Daniel Fischer, and can be found on
--- the Haskell Wiki http://www.haskell.org/haskellwiki/Sudoku
---
--- The Haskell wiki license applies to this code:
---
--- Permission is hereby granted, free of charge, to any person obtaining
--- this work (the "Work"), to deal in the Work without restriction,
--- including without limitation the rights to use, copy, modify, merge,
--- publish, distribute, sublicense, and/or sell copies of the Work, and
--- to permit persons to whom the Work is furnished to do so.
--- 
--- THE WORK IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
--- IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
--- MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
--- NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE
--- LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION
--- OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION
--- WITH THE WORK OR THE USE OR OTHER DEALINGS IN THE WORK.
-
-module Sudoku (solve, printGrid) where
-
-import Data.List hiding (lookup)
-import Data.Array
-import Control.Monad
-import Data.Maybe
-
--- Types
-type Digit  = Char
-type Square = (Char,Char)
-type Unit   = [Square]
- 
--- We represent our grid as an array
-type Grid = Array Square [Digit]
- 
- 
--- Setting Up the Problem
-rows = "ABCDEFGHI"
-cols = "123456789"
-digits = "123456789"
-box = (('A','1'),('I','9'))
- 
-cross :: String -> String -> [Square]
-cross rows cols = [ (r,c) | r <- rows, c <- cols ]
- 
-squares :: [Square]
-squares = cross rows cols  -- [('A','1'),('A','2'),('A','3'),...]
- 
-peers :: Array Square [Square]
-peers = array box [(s, set (units!s)) | s <- squares ]
-      where
[…]

### 2017-12-12 examples/sudoku/bench.sh

@@ -1,72 +1,0 @@
-#!/bin/bash
-
-rtsOpts="+RTS -N12 -s"
-
-## declare an array variable
-programs=(
-	"sudoku1"
-	"sudoku2"
-	"sudoku3"
-	"sudoku4"
-	"sudoku-par4"
-    "parrows-sudoku-parmap"
-    "parrows-sudoku-parmapstream"
-    "parrows-sudoku-farm"
-    "parrows-sudoku-farmChunk"
-)
-
-parameters=(
-	""
-	""
-	""
-	""
-	"384"
-    ""
-    "384"
-    "12"
-    "384 12"
-)
-
-puzzles=(
-    "sudoku17.1000.txt"
-    "sudoku17.16000.txt"
-    "sudoku17.49151.txt"
-)
-
-runsPerConfig=100
-
-echo "#runsPerConfig: $runsPerConfig"
-
-# get length of an array
-programCount=${#programs[@]}
-
-for puzzle in "${puzzles[@]}"
-do
-    echo $puzzle
-    for (( i=0; i < ${programCount}; i++ ));
-    do
-        progName=${programs[$i]}
-        cmd="./"$progName" "$puzzle" "${parameters[$i]}" "$rtsOpts
-        #echo "#running: " $cmd
-        grepCmd="2>&1 >/dev/null | grep Total | sed \"s/Total[a-z ]*[0-9.s]*[ ]*[( ]*\([0-9.]*\)[ a-z)]*/\\1/\" | awk '{\$1=\$1};1\'"
-
-        totalTime=0
-
-        for (( j = 0; j < $runsPerConfig; j++ ));
-        do
-            curTime=$(eval $cmd" "$grepCmd)
-            #echo $curTime
-
-            totalTime=`echo $totalTime + $curTime | bc`
-        done
-
-        average=`echo "scale=3;$totalTime / $runsPerConfig" | bc`
-
-        echo $progName" "$average
-    done
-
-    echo ""
-    echo ""
-    echo ""
-done
-

### 2017-12-12 examples/sudoku/parrows-sudoku-farm.hs

@@ -1,17 +1,0 @@
-{-# LANGUAGE FlexibleContexts #-}
-import Sudoku
-import Control.Exception
-import System.Environment
-import Data.Maybe
-import Parrows.Definition
-import Parrows.ParMonad
-import Control.DeepSeq
-
-main :: IO ()
-main = do
-    [f,n] <- getArgs
-    grids <- fmap lines $ readFile f
-    print (length (filter isJust (farm' (read n) solve grids)))
-
-farm' :: NFData b => Int -> (a -> b) -> [a] -> [b]
-farm' = farm ()

### 2017-12-12 examples/sudoku/parrows-sudoku-farmChunk.hs

@@ -1,17 +1,0 @@
-{-# LANGUAGE FlexibleContexts #-}
-import Sudoku
-import Control.Exception
-import System.Environment
-import Data.Maybe
-import Parrows.Definition
-import Parrows.ParMonad
-import Control.DeepSeq
-
-main :: IO ()
-main = do
-    [f,chunkSize, cores] <- getArgs
-    grids <- fmap lines $ readFile f
-    print (length (filter isJust (farmChunk' (read chunkSize) (read cores) solve grids)))
-
-farmChunk' :: NFData b => Int -> Int -> (a -> b) -> [a] -> [b]
-farmChunk' = farmChunk ()

### 2017-12-12 examples/sudoku/parrows-sudoku-parmap-mult.hs

@@ -1,15 +1,0 @@
-{-# LANGUAGE FlexibleContexts #-}
-import Sudoku
-import Control.Exception
-import System.Environment
-import Data.Maybe
-import Parrows.Definition
-import Parrows.Skeletons.Map
-import Parrows.Multicore
-import Control.DeepSeq
-
-main :: IO ()
-main = do
-    [f] <- getArgs
-    grids <- fmap lines $ readFile f
-    print (length (filter isJust (parMap () solve grids)))

### 2017-12-12 examples/sudoku/parrows-sudoku-parmapstream.hs

@@ -1,17 +1,0 @@
-{-# LANGUAGE FlexibleContexts #-}
-import Sudoku
-import Control.Exception
-import System.Environment
-import Data.Maybe
-import Parrows.Definition
-import Parrows.ParMonad
-import Control.DeepSeq
-
-main :: IO ()
-main = do
-    [f,n] <- getArgs
-    grids <- fmap lines $ readFile f
-    print (length (filter isJust (parMapChunk (read n) solve grids)))
-
-parMapChunk :: NFData b => Int -> (a -> b) -> [a] -> [b]
-parMapChunk = parMapStream ()

### 2017-12-12 examples/sudoku/sudoku-par1.hs

@@ -1,10 +1,0 @@
-import Sudoku
-import Control.Exception
-import System.Environment
-import Data.Maybe
-
-main :: IO ()
-main = do
-    [f] <- getArgs
-    grids <- fmap lines $ readFile f
-    print (length (filter isJust (map solve grids)))

### 2017-12-12 examples/sudoku/sudoku-par2.hs

@@ -1,21 +1,0 @@
-import Sudoku
-import Control.Exception
-import System.Environment
-import Data.Maybe
-import Control.Monad.Par.Scheds.Trace
-
-main :: IO ()
-main = do
-    [f] <- getArgs
-    grids <- fmap lines $ readFile f
-
-    let (as,bs) = splitAt (length grids `div` 2) grids
-
-    print $ length $ filter isJust $ runPar $ do
-       i1 <- new
-       i2 <- new
-       fork $ put i1 (map solve as)
-       fork $ put i2 (map solve bs)
-       as' <- get i1
-       bs' <- get i2
-       return (as' ++ bs')

### 2017-12-12 examples/sudoku/sudoku1.hs

@@ -1,16 +1,0 @@
--- <<all
-import Sudoku
-import Control.Exception
-import System.Environment
-import Data.Maybe
-
-main :: IO ()
-main = do
-  [f] <- getArgs                           -- <1>
-  file <- readFile f                       -- <2>
-
-  let puzzles   = lines file               -- <3>
-      solutions = map solve puzzles        -- <4>
-
-  print (length (filter isJust solutions)) -- <5>
--- >>

### 2017-12-12 examples/sudoku/sudoku2.hs

@@ -1,27 +1,0 @@
-import Sudoku
-import Control.Exception
-import System.Environment
-import Control.Parallel.Strategies
-import Control.DeepSeq
-import Data.Maybe
-
--- <<main
-main :: IO ()
-main = do
-  [f] <- getArgs
-  file <- readFile f
-
-  let puzzles = lines file
-
-      (as,bs) = splitAt (length puzzles `div` 2) puzzles -- <1>
-
-      solutions = runEval $ do
-                    as' <- rpar (force (map solve as))   -- <2>
-                    bs' <- rpar (force (map solve bs))   -- <2>
-                    rseq as'                             -- <3>
-                    rseq bs'                             -- <3>
-                    return (as' ++ bs')                  -- <4>
-
-  print (length (filter isJust solutions))
--- >>
-

### 2017-12-12 examples/sudoku/sudoku3.hs

@@ -1,26 +1,0 @@
-import Sudoku
-import Control.Exception
-import System.Environment
-import Control.Parallel.Strategies hiding (parMap)
-import Data.Maybe
-
--- <<main
-main :: IO ()
-main = do
-  [f] <- getArgs
-  file <- readFile f
-
-  let puzzles   = lines file
-      solutions = runEval (parMap solve puzzles)
-
-  print (length (filter isJust solutions))
--- >>
-
--- <<parMap
-parMap :: (a -> b) -> [a] -> Eval [b]
-parMap f [] = return []
-parMap f (a:as) = do
-   b <- rpar (f a)
-   bs <- parMap f as
-   return (b:bs)
--- >>

### 2017-12-12 examples/sudoku/sudoku4.hs

@@ -1,25 +1,0 @@
-import Sudoku
-import Control.Exception
-import System.Environment
-import Control.Parallel.Strategies hiding (parMap)
-import Data.Maybe
-
--- <<main
-main :: IO ()
-main = do
-  [f] <- getArgs
-  file <- readFile f
-
-  let puzzles   = lines file
-      solutions = runEval (parMap solve puzzles)
-
-  evaluate (length puzzles)
-  print (length (filter isJust solutions))
--- >>
-
-parMap :: (a -> b) -> [a] -> Eval [b]
-parMap f [] = return []
-parMap f (a:as) = do
-   b <- rpar (f a)
-   bs <- parMap f as
-   return (b:bs)

### 2017-12-12 examples/talk/add.hs

@@ -1,14 +1,0 @@
-{-# LANGUAGE MultiParamTypeClasses, FlexibleContexts #-}
-module Main where
-
-import Control.Arrow
-import Parrows.Definition
-import Parrows.ParMonad
-
-add :: (ArrowParallel arr a Int (),
-        ArrowApply arr,
-        ArrowChoice arr)
-    => arr a Int -> arr a Int -> arr a Int
-add f g = f |&&&| g >>> (arr $ \(u, v) -> u + v)
-
-main = print $ (add (+1) (+2)) (2::Int)

### 2017-12-12 examples/torus_matrix_multicore.hs

@@ -1,99 +1,0 @@
-{-# LANGUAGE FlexibleInstances, FlexibleContexts, UndecidableInstances,
-MultiParamTypeClasses #-}
-
-module Main where
-
-import Parrows.Multicore
-import Parrows.Definition
-import Parrows.Future
-import Parrows.Util
-import Parrows.Skeletons.Topology
-import Data.List
-
---import GHC.Conc
-
-import Control.Arrow
---import Control.Monad.Par
-import Control.Parallel.Strategies
-import Control.Parallel
-import Control.DeepSeq
-
-import Control.DeepSeq
-
-import Control.Applicative
-import Data.Functor
-import Data.List.Split
-
-import System.Environment
-
-import System.Random
-
-import Debug.Trace
-
-type Vector = [Int]
-type Matrix = [Vector]
-
-dimX :: Matrix -> Int
-dimX = length
-
-dimY :: Matrix -> Int
-dimY = length . head
-
---from: https://rosettacode.org/wiki/Matrix_multiplication#Haskell
-foldlZipWith::(a -> b -> c) -> (d -> c -> d) -> d -> [a] -> [b]  -> d
-foldlZipWith _ _ u [] _          = u
-foldlZipWith _ _ u _ []          = u
-foldlZipWith f g u (x:xs) (y:ys) = foldlZipWith f g (g u (f x y)) xs ys
-
-foldl1ZipWith::(a -> b -> c) -> (c -> c -> c) -> [a] -> [b] -> c
-foldl1ZipWith _ _ [] _          = error "First list is empty"
-foldl1ZipWith _ _ _ []          = error "Second list is empty"
-foldl1ZipWith f g (x:xs) (y:ys) = foldlZipWith f g (f x y) xs ys
-
-multAdd::(a -> b -> c) -> (c -> c -> c) -> [[a]] -> [[b]] -> [[c]]
-multAdd f g xs ys = map (\us -> foldl1ZipWith (\u vs -> map (f u) vs) (zipWith g) us ys) xs
-
-matMult:: Num a => [[a]] -> [[a]] -> [[a]]
-matMult xs ys = multAdd (*) (+) xs ys
-
-matAdd :: Matrix -> Matrix -> Matrix
-matAdd x y
-    | dimX x /= dimX y = error "dimX x not equal to dimX y"
-    | dimY x /= dimY y = error "dimY x not equal to dimY y"
-    | otherwise = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
-
[…]

### 2017-12-12 examples/torus_matrix_parrows_mult.hs

@@ -1,118 +1,0 @@
-{-# LANGUAGE FlexibleInstances, FlexibleContexts, UndecidableInstances,
-MultiParamTypeClasses #-}
-
-module Main where
-
-import Parrows.Multicore
-import Parrows.Definition
-import Parrows.Future
-import Parrows.Util
-import Parrows.Skeletons.Topology
-import Parrows.Skeletons.Map
-
-import Control.Parallel.Strategies
-
-import Data.List
-
-import GHC.Conc
-
-import Control.Arrow
-import Control.DeepSeq
-
-import Control.DeepSeq
-
-import Control.Applicative
-import Data.Functor
-import Data.List.Split
-
-import System.Environment
-
-import System.Random
-
-import Debug.Trace
-
-type Vector = [Int]
-type Matrix = [Vector]
-
-dimX :: Matrix -> Int
-dimX = length
-
-dimY :: Matrix -> Int
-dimY = length . head
-
---from: https://rosettacode.org/wiki/Matrix_multiplication#Haskell
-foldlZipWith::(a -> b -> c) -> (d -> c -> d) -> d -> [a] -> [b]  -> d
-foldlZipWith _ _ u [] _          = u
-foldlZipWith _ _ u _ []          = u
-foldlZipWith f g u (x:xs) (y:ys) = foldlZipWith f g (g u (f x y)) xs ys
-
-foldl1ZipWith::(a -> b -> c) -> (c -> c -> c) -> [a] -> [b] -> c
-foldl1ZipWith _ _ [] _          = error "First list is empty"
-foldl1ZipWith _ _ _ []          = error "Second list is empty"
-foldl1ZipWith f g (x:xs) (y:ys) = foldlZipWith f g (f x y) xs ys
-
-multAdd::(a -> b -> c) -> (c -> c -> c) -> [[a]] -> [[b]] -> [[c]]
-multAdd f g xs ys = map (\us -> foldl1ZipWith (\u vs -> map (f u) vs) (zipWith g) us ys) xs
-
-matMult:: Num a => [[a]] -> [[a]] -> [[a]]
-matMult xs ys = multAdd (*) (+) xs ys
-
-matAdd :: Matrix -> Matrix -> Matrix
-matAdd x y
-    | dimX x /= dimX y = error "dimX x not equal to dimX y"
-    | dimY x /= dimY y = error "dimY x not equal to dimY y"
-    | otherwise = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
-
[…]

### 2017-12-12 paper/content/utilityFunctions.tex

@@
 
-Next, we also define |map|, |foldl| and |zipWith| on Arrows.
+Next, we give the definition of |evalN| which also helps us to define |map|, and |zipWith| on Arrows.
+The |evalN| combinator in Fig.~\ref{fig:evalN} converts a list of arrows |[arr a b]| into an arrow |arr [a] [b]|.
+
+\begin{figure}[h]
+\begin{code}
+evalN :: (ArrowChoice arr) => [arr a b] -> arr [a] [b]
+evalN (f:fs) = arr listcase >>>
+         arr (const []) ||| (f *** evalN fs >>> arr (uncurry (:)))
+         where listcase []     = Left ()
+               listcase (x:xs) = Right (x,xs)
+evalN [] = arr (const [])
+\end{code}
+\caption{The definition of |evalN|}
+\label{fig:evalN}
+\end{figure}
+
 The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any arrow |arr a
-b| to an arrow |arr [a] [b]| \cite{programming_with_arrows}. 
-Similarly, we can also define |foldlArr| (Fig.~\ref{fig:foldlArr}) that lifts any arrow |arr (b, a) b| with a neutral element |b| to |arr [a] b|.
+b| to an arrow |arr [a] [b]|. The original inspiration was from \cite{programming_with_arrows},
+but the definition as then unified with |evalN|.
 
@@
 mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
-mapArr f =
-	arr listcase >>>
-	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
-
-listcase [] = Left ()
-listcase (x:xs) = Right (x,xs)
+mapArr = evalN . repeat
 \end{code}
-\caption{The definition of |map| over Arrows and the |listcase| helper
-function.}
+\caption{The definition of |map| over Arrows.}
 \label{fig:mapArr}
-\end{figure}
-
-\begin{figure}[h]
-\begin{code}
-foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
-foldlArr f b =
-	arr listcase >>>
-	arr (const b) |||
-		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
-\end{code}
-\caption{The definition of |foldl| over Arrows.}
-\label{fig:foldlArr}
 \end{figure}
@@
 \end{code}
-% \caption{Type signature of \pipepipepipe}
-% \label{fig:codeSigPipePipePipe}
-% \end{figure}
-
-With the zipWithArr combinator we can also write a combinator |listApp|, that lifts a list of arrows |[arr a b]| to an arrow |arr [a] [b]|.
-% \begin{figure}[htb]
-\begin{code}
-listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
-listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
-\end{code}
-% \caption{Definition of |listApp|}
-% \label{fig:listApp}
-% \end{figure}
-% $ %% formatting
-Note that  this additionally makes use of the |ArrowApply| type class that allows us to evaluate arrows with |app :: arr (arr a b, a) c|.
-
-% $ %% formatting
 
@@
 
-To begin with, we give the actual implementation of the non-configurable instance of the |ArrowParallel| instance:
-
-\begin{figure}
-\begin{code}
-instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
-    parEvalN _ fs = parEvalN (hack fs) fs
-                    where
-                        hack :: (NFData b) => [arr a b] -> Conf b
-                        hack _ = Conf rdeepseq
-\end{code}
-\caption{The actual default implementation of GpH's |ArrowParallel|.}
-\label{fig:actualNonConfigMulticore}
-\end{figure}
-
 %
-Next, we warp Eden's build-in Futures in PArrows as in
+We begin with warping Eden's build-in Futures in PArrows as in
 Figure~\ref{fig:RDFuture}, where |rd| is the accessor function for the |RD| wrapped inside |RemoteData|.
@@
 % \end{figure}
+
+Figures~\ref{fig:parMapImg}--\ref{fig:parMapStream} show the definitions and a visualizations of two parallel |map| variants, defined using |parEvalN| and its lazy counterpart.
 
@@
 \end{figure}
-
-Figures~\ref{fig:parMapImg}--\ref{fig:parMapStream} show the definitions and a visualizations of two parallel |map| variants, defined using |parEvalN| and its lazy counterpart.
 

### 2017-12-13 paper/content/arrows.tex
fixes in arrows.tex. reasoning why we chose Arrows over Profunctors

--- 2017-12-07 paper/content/arrows.tex
@@
 \end{code}
+
+As we can rewrite the monadic bind operation |(>>=)| with only the Kleisli type into |m a -> Kleisli m a b -> m b|, but not with a general Arrow |arr a b|, we can intuitively get an idea of why Arrows must be a generalisation of Monads.
+While this also means that a general Arrow can not express everything a Monad can, the Parser example shows that this trade-off is worth it in some cases.
+
+In this paper we will show that we can express parallel computations with Arrows and do not require Monads to do so.
+We also do not restrict the compatible Arrows to ones which have |ArrowApply| instances but only require the |ArrowChoice| and |ArrowLoop|.
+Because of this, we have a truly more general interface as compared to a monadic one.
+
+Also note that, while we could have based our DSL on Profunctors as well, we chose Arrows.
+for this paper since they they are a more direct translation of pure functions than Profunctors.
+Note that Profunctors can easily be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
+However, they are a promising candidate for future improvement of our DSL.
+
 % \caption{Add over arrows}

### 2017-12-13 paper/content/utilityFunctions.tex
fixes in arrows.tex. reasoning why we chose Arrows over Profunctors

@@
 (pipepipepipe) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
+\end{code}
+
+\section{Profunctor Arrows}
+\label{app:profunctorArrows}
+
+Following is a theoretical way of using Profunctors together with our DSL.
+\footnote{for additional information on the typeclasses, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html}}
+
+\begin{code}
+instance (Category p, Strong p) => Arrow p where
+  arr f = dimap id f id
+  first = first'
+
+instance (Category p, Strong p, Costrong p) => ArrowLoop p where
+  loop = loop'
+
+instance (Category p, Strong p, Choice p) => ArrowChoice p where
+  left = left'
 \end{code}

### 2017-12-13 paper/content/arrows.tex
review of chapters 1-4 done

@@
 In this paper we will show that we can express parallel computations with Arrows and do not require Monads to do so.
-We also do not restrict the compatible Arrows to ones which have |ArrowApply| instances but only require the |ArrowChoice| and |ArrowLoop|.
+We also do not restrict the compatible Arrows to ones which have |ArrowApply| instances but instead only require instances for |ArrowChoice| and |ArrowLoop|.
 Because of this, we have a truly more general interface as compared to a monadic one.
@@
 Note that Profunctors can easily be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
-However, they are a promising candidate for future improvement of our DSL.
-
-% \caption{Add over arrows}
-% \label{fig:addArrows}
-% \end{figure}
-% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator.
-Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows.
-Compared to simple Monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
-% \\\\
-% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used.
-We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}
-%
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+However, they are a promising candidate for future improvements of our DSL.

### 2017-12-13 paper/content/basicSkeletons.tex
review of chapters 1-4 done

@@
 	arr (chunksOf chunkSize) >>>
-	listApp fchunks >>>
+    evalN fchunks >>>
     arr concat
-	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
+    where
+      fchunks = map (parEvalN conf) (chunksOf chunkSize fs)
 \end{code} %$ %% formatting
@@
 \end{figure}
-We have only talked about the parallelization arrows of the same type until now.
+We have only talked about the parallelisation of arrows of the same type until now.
 But sometimes we want to parallelize heterogeneous types as well.
-However, we can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class.
+We can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class.
 The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.

### 2017-12-13 paper/content/utilityFunctions.tex
review of chapters 1-4 done

@@
 
-Following is a theoretical way of using Profunctors together with our DSL.
-\footnote{for additional information on the typeclasses, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html}}
+In Fig.~\ref{fig:profunctorArrow} we show how specific Profunctors can be turned into Arrows.
 
+\begin{figure}[h]
 \begin{code}
@@
 \end{code}
+\caption{Profunctors as Arrows}
+\footnote{for additional information on the typeclasses see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html}}}
+\label{fig:profunctorArrow}
+\end{figure}
 

### 2017-12-13 Definition/src/main/Parrows/Definition.hs
minor

@@ -66,7 +66,7 @@ parEvalNLazy conf chunkSize fs =
                evalN fchunks >>>
                arr concat
                where
-                fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
+                fchunks = map (parEvalN conf) (chunksOf chunkSize fs)
 
 -- evaluate two functions with different types in parallel
 parEval2 :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) conf) => conf -> arr a b -> arr c d -> arr (a, c) (b, d)

### 2017-12-13 Definition/src/main/Parrows/Skeletons/Topology.hs

@@ -107,8 +107,3 @@ threetotwo = arr $ \ ~(a, b, c) -> (a, (b, c))
 
 twotothree :: (Arrow arr) => arr (a, (b, c)) (a, b, c)
 twotothree = arr $ \ ~(a, (b, c)) -> (a, b, c)
-
--- from Eden, ported to Arrows:
-rightRotate :: (Arrow arr) => arr [a] [a]
-rightRotate = arr $ \list -> case list of [] -> []
-                                          xs -> last xs : init xs

### 2017-12-13 Definition/src/main/Parrows/Util.hs

@@ -52,6 +52,11 @@ foldlArr f = arr listcase >>>
                    listcase (b, x:xs) = Right (b, (x,xs))
 
 
+-- from Eden, ported to Arrows:
+rightRotate :: (Arrow arr) => arr [a] [a]
+rightRotate = arr $ \list -> case list of [] -> []
+                                          xs -> last xs : init xs
+
 assoc :: ((a, b), c) -> (a,(b,c))
 assoc ((a,b),c) = (a,(b,c))
 

### 2017-12-13 Eden/src/main/Parrows/Eden.hs

@@ -29,13 +29,16 @@ SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
 {-# LANGUAGE UndecidableInstances   #-}
 module Parrows.Eden(
   Conf(..),
+  RemoteData(..),
   defaultConf,
+  put',
+  get',
   module Parrows.Definition,
   module Parrows.Future
 ) where
 
 import           Parrows.Definition
-import           Parrows.Future
+import           Parrows.Future hiding (put', get')
 import           Parrows.Util
 
 import           Control.Arrow
@@ -52,9 +55,15 @@ data Conf a = Nil
 defaultConf :: [arr a b] -> Conf b
 defaultConf _ = Nil
 
+put' :: (Arrow arr, Trans a) => arr a (RemoteData a)
+put' = arr (\a -> RD { rd = release a })
+
+get' :: (Arrow arr, Trans a) => arr (RemoteData a) a
+get' = arr rd >>> arr fetch
+
 instance (Trans a) => Future RemoteData a (Conf a) where
-    put _ = arr (\a -> RD { rd = release a })
-    get _ = arr rd >>> arr fetch
+    put _ = put'
+    get _ = get'
 
 instance (ArrowChoice arr, ArrowParallel arr a b (Conf b)) => FutureEval arr a b (Conf b) where
     headStrictEvalN = parEvalN

### 2017-12-13 Eden/src/main/Parrows/Eden/Simple.hs

@@ -39,6 +39,6 @@ instance (Trans b, ArrowChoice arr, ArrowParallel arr a b (), FutureEval arr a b
     headStrictEvalN _ fs = headStrictEvalN (defaultConf fs) fs
     postHeadStrictEvalN _  fs = postHeadStrictEvalN (defaultConf fs) fs
 
-instance Future BasicFuture b () where
+instance (Trans a) => Future RemoteData a () where
     put _ = put'
     get _ = get'

### 2017-12-13 paper/content/parrows.tex

@@
 \subsubsection{Default configuration instances}
-While the configurability in the instances of the |ArrowParallel| instances above might be nice to have, users probably would prefer a proper default configuration for many parallel programs.
-These can easily be defined as well as we can see by the example of the default implementation of |ArrowParallel| for the GpH backend:
+While the configurability in the instances of the |ArrowParallel| instances above is nice, users probably would like to have a proper default configuration for many parallel programs as well.
+These can also easily be defined as we can see by the example of the default implementation of |ArrowParallel| for the GpH backend:
 
@@
 
-The other backends have similarly structured implementations.
+The other backends have similarly structured implementations which we do not discuss here for brevity reasons.
 Note that by restricting the original implementations of our backends to their specific |Conf| type, we also get interoperability between backends for free.
@@
 
-\mbcomment{For convenience purposes, ArrowParallel arr a b () is defined as...}
-\mbcomment{Until now we described the arrow operations more literally, but in the following sectioins we only talk about the dataflow in general as the idea should be clear now.}
-
-%\FloatBarrier
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:
+Up until now we discussed Arrow operations more in detail, but in the following sections we focus on the dataflow instead of how the Arrows are obtained so we can focus on the important parts now that we have seen that Arrows are capable of expressing parallelism.

### 2017-12-13 paper/content/futures.tex
Futures chapter

@@
 \begin{code}
-someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
+someCombinator:: (ArrowChoice arr, NFData c, ArrowParallel arr a b ()) => [arr a b] -> [arr b c] -> arr [a] [c]
 someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
@@
 
-In a distributed environment, a resulting arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
+In a distributed environment, a resulting arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
@@
 When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
-This can become a serious bottleneck %in heavy threaded applications.
+This can become a serious bottleneck
 for larger amount of data and number of processes \citep[as e.g.][showcases]{Berthold2009-fft}.
@@
 Eden already ships with \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
-But as we want code with our DSL to be implementation agnostic, we have to wrap this context. We do this with the |Future| type class (Fig.~\ref{fig:future}).
+But as we want code with our DSL to be implementation agnostic, we have to wrap this concept.
+We do this with the |Future| type class (Fig.~\ref{fig:future}).
+Note that we require a conf parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the |ArrowParallel| typeclass earlier.
 \begin{figure}[h]
 \begin{code}
-class Future fut a | a -> fut where
-    put :: (Arrow arr) => arr a (fut a)
-    get :: (Arrow arr) => arr (fut a) a
+class Future fut a conf | a conf -> fut where
+    put :: (Arrow arr) => conf -> arr a (fut a)
+    get :: (Arrow arr) => conf -> arr (fut a) a
 \end{code}
@@
 Nevertheless, we require a common interface so the parallel Arrows are portable across backends.
-\begin{figure}[tb]
-\begin{code}
-data BasicFuture a = BF a
-
-instance (NFData a) => NFData (BasicFuture a) where
-    rnf (BF a) = rnf a
-
-instance (NFData a) => Future BasicFuture a where
-    put = arr BF
-    get = arr (\(BF a) -> a)
-\end{code}
-\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and GpH Haskell backends.}
-\label{fig:BasicFuture}
-\end{figure} % $
+The implementation can also be found in Section ~\ref{app:omitted}.
 
 In our communication example we can use this |Future| concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
+Note that with our PArrows DSL we can also define default instances |Future fut a ()| for each backend similar to how |ArrowParallel arr a b ()| was defined in Section~\ref{sec:parallel-arrows}.
+Details can be found in Section~\ref{app:omitted}.
 \begin{figure}[tbh]
 \begin{code}
-someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
+someCombinator :: (ArrowChoice arr, NFData b, NFData c,
+	ArrowParallel arr a (fut b) (), 
+	ArrowParallel arr (fut b) c (),
+	Future fut b ()) =>
+	[arr a b] -> [arr b c] -> arr [a] [c]
 someCombinator fs1 fs2 =
-	parEvalN () (map (>>> put) fs1) >>>
+	parEvalN () (map (>>> put ()) fs1) >>>
 	rightRotate >>>
-	parEvalN () (map (get >>>) fs2)
+	parEvalN () (map (get () >>>) fs2)
 \end{code}
@@
 In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.
+What's especially elegant about this solution is that we can specify the type of |Future| to be used per backend, without even requring to know about the actual type used for communication.
+We only specify that there has to be a compatible Future and do not care about any specifics.
 \olcomment{Fig.

### 2017-12-13 paper/content/utilityFunctions.tex
Futures chapter

@@
 %
-We begin with warping Eden's build-in Futures in PArrows as in
-Figure~\ref{fig:RDFuture}, where |rd| is the accessor function for the |RD| wrapped inside |RemoteData|.
-Furthermore, in order for these |Future| types to fit with the |ArrowParallel| instances we gave earlier, we have to give the necessary |NFData| and |Trans| instances, the latter are only needed in Eden.
-The |Trans| instance does not have any functions declared as the default implementation suffices here.
-Furthermore, because |MVar| already ships with a |NFData| instance, we only have to supply a simple delegating |NFData| instance for our |RemoteData| type, where |rd| simply unwraps |RD|.
-The |Trans| instance does not have any functions declared as the default implementation suffices:
+We begin with warping Eden's build-in |RemoteData| to |Future| in
+Figure~\ref{fig:RDFuture}
 
-% \begin{figure}[h]
+\begin{figure}[h]
 \begin{code}
+data RemoteData a = RD { rd :: RD a }
+
+put' :: (Arrow arr) => arr a (BasicFuture a)
+put' = arr BF
+
+get' :: (Arrow arr) => arr (BasicFuture a) a
+get' = arr (\(~(BF a)) -> a)
+
 instance NFData (RemoteData a) where
@@
 instance Trans (RemoteData a)
+
+instance (Trans a) => Future RemoteData a (Conf a) where
+    put _ = put'
+    get _ = get'
+
+instance (Trans a) => Future RemoteData a () where
+    put _ = put'
+    get _ = get'
 \end{code}
+\caption{|RD|-based |RemoteData| version of |Future| for the Eden backend.}
+\label{fig:RDFuture}
+\end{figure}
+
+Next, we have the definition of |BasicFuture| in Fig.~\ref{fig:BasicFuture} and the corresponding |Future| instances.
+
+\begin{figure}[h]
+\begin{code}
+data BasicFuture a = BF a
+
+put' :: (Arrow arr) => arr a (BasicFuture a)
+put' = arr BF
+
+get' :: (Arrow arr) => arr (BasicFuture a) a
+get' = arr (\(~(BF a)) -> a)
+
+instance NFData a => NFData (BasicFuture a) where
+    rnf (BF a) = rnf a
+
+instance Future BasicFuture a (Conf a) where
+    put _ = put'
+    get _ = get'
+
+instance Future BasicFuture a () where
+    put _ = put'
+    get _ = get'
+\end{code}
+\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and GpH Haskell backends.}
+\label{fig:BasicFuture}
+\end{figure}
+
 % \caption{NFData and Trans instances for the RemoteData type.
@@
 follows \citet{Loogen2012}.
-
-\begin{figure}[h]
-\begin{code}
-data RemoteData a = RD { rd :: RD a }
-
-instance (Trans a) => Future RemoteData a where
-    put = arr (\a -> RD { rd = release a })
-    get = arr rd >>> arr fetch
-\end{code}
-\caption{|RD|-based |RemoteData| version of |Future| for the Eden backend.}
-\label{fig:RDFuture}
-\end{figure}
 

### 2017-12-13 paper/content/futures.tex
minorino

@@
 \section{Futures} \label{sec:futures}
-Consider a mock-up parallel arrow combinator:
+Consider a mock-up parallel Arrow combinator:
 \begin{code}
-someCombinator:: (ArrowChoice arr, NFData c, ArrowParallel arr a b ()) => [arr a b] -> [arr b c] -> arr [a] [c]
-someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
+someCombinator:: (ArrowChoice arr,
+	ArrowParallel arr a b (),
+	ArrowParallel arr b c ()) =>
+	[arr a b] -> [arr b c] -> arr [a] [c]
+someCombinator fs1 fs2 =
+	parEvalN () fs1 >>>
+	rightRotate >>>
+	parEvalN () fs2
 \end{code}
@@
 In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.
-What's especially elegant about this solution is that we can specify the type of |Future| to be used per backend, without even requring to know about the actual type used for communication.
+One especially elegant aspect of this solution is that we can specify the type of |Future| to be used per backend with full interoperability between code using different backends, without even requring to know about the actual type used for communication.
 We only specify that there has to be a compatible Future and do not care about any specifics.

### 2017-12-14 Eden/src/main/Parrows/Eden.hs
minor

@@ -50,10 +50,10 @@ instance NFData (RemoteData a) where
     rnf = rnf . rd
 instance Trans (RemoteData a)
 
-data Conf a = Nil
+data Conf = Nil
 
-defaultConf :: [arr a b] -> Conf b
-defaultConf _ = Nil
+defaultConf :: Conf
+defaultConf = Nil
 
 put' :: (Arrow arr, Trans a) => arr a (RemoteData a)
 put' = arr (\a -> RD { rd = release a })
@@ -61,17 +61,17 @@ put' = arr (\a -> RD { rd = release a })
 get' :: (Arrow arr, Trans a) => arr (RemoteData a) a
 get' = arr rd >>> arr fetch
 
-instance (Trans a) => Future RemoteData a (Conf a) where
+instance (Trans a) => Future RemoteData a Conf where
     put _ = put'
     get _ = get'
 
-instance (ArrowChoice arr, ArrowParallel arr a b (Conf b)) => FutureEval arr a b (Conf b) where
+instance (ArrowChoice arr, ArrowParallel arr a b Conf) => FutureEval arr a b Conf where
     headStrictEvalN = parEvalN
     postHeadStrictEvalN _ = evalN
 
-instance (Trans a, Trans b) => ArrowParallel (->) a b (Conf b) where
+instance (Trans a, Trans b) => ArrowParallel (->) a b Conf where
     parEvalN _ = spawnF
 
-instance (ArrowParallel (->) a (m b) (Conf b), Monad m, Trans a, Trans b, Trans (m b)) => ArrowParallel (Kleisli m) a b (Conf b) where
+instance (ArrowParallel (->) a (m b) Conf, Monad m, Trans a, Trans b, Trans (m b)) => ArrowParallel (Kleisli m) a b Conf where
     parEvalN conf fs = arr (parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
                        Kleisli sequence

### 2017-12-14 Eden/src/main/Parrows/Eden/Simple.hs

@@ -32,12 +32,12 @@ import Parrows.Eden
 import Control.Arrow
 import Control.Parallel.Eden (Trans)
 
-instance (Trans b, ArrowChoice arr, ArrowParallel arr a b (Conf b)) => ArrowParallel arr a b () where
-    parEvalN _ fs = parEvalN (defaultConf fs) fs
+instance (Trans b, ArrowChoice arr, ArrowParallel arr a b Conf) => ArrowParallel arr a b () where
+    parEvalN _ fs = parEvalN defaultConf fs
 
-instance (Trans b, ArrowChoice arr, ArrowParallel arr a b (), FutureEval arr a b (Conf b)) => FutureEval arr a b () where
-    headStrictEvalN _ fs = headStrictEvalN (defaultConf fs) fs
-    postHeadStrictEvalN _  fs = postHeadStrictEvalN (defaultConf fs) fs
+instance (Trans b, ArrowChoice arr, ArrowParallel arr a b (), FutureEval arr a b Conf) => FutureEval arr a b () where
+    headStrictEvalN _ fs = headStrictEvalN defaultConf fs
+    postHeadStrictEvalN _  fs = postHeadStrictEvalN defaultConf fs
 
 instance (Trans a) => Future RemoteData a () where
     put _ = put'

### 2017-12-14 paper/content/mapSkeletons.tex

@@
 Just like |parEvalN|, |parMap| is 100\% strict.
-As |parMap|  is 100\% strict it has the same restrictions as |parEvalN| compared to |parEvalNLazy|.
+Because of this, it has the same restrictions as |parEvalN| as compared to |parEvalNLazy|.
 So it makes sense to also have a |parMapStream| (Figs.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
-The code is quite straightforward, we show it in Appendix.
+The code of these two skeletons is quite straightforward, we show it in Appendix.
 

### 2017-12-14 paper/content/utilityFunctions.tex
added clarification for (>>>) being hosted on Category in Standard GHC
minor

@@
 In Fig.~\ref{fig:profunctorArrow} we show how specific Profunctors can be turned into Arrows.
+Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| typeclass.
 
@@
 \caption{Profunctors as Arrows}
-\footnote{for additional information on the typeclasses see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html}}}
+\footnote{for additional information on the typeclasses see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}}
 \label{fig:profunctorArrow}
@@
 
-instance (Trans a) => Future RemoteData a (Conf a) where
+instance (Trans a) => Future RemoteData a Conf where
     put _ = put'

### 2017-12-14 paper/content/utilityFunctions.tex
minor

@@
 \end{code}
-\caption{Profunctors as Arrows}
-\footnote{for additional information on the typeclasses see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}}
+\caption{Profunctors as Arrows\protect\footnotemark}
 \label{fig:profunctorArrow}
 \end{figure}
+
+\footnotetext{for additional information on the typeclasses used, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}}
 

### 2017-12-15 benchmarks/jacobi-bench/jacobi-parrows/JacobiSum.hs
redone RM dist bench (again)

@@ -64,8 +64,7 @@ bestimme_t n
 		 | ( n < 2^1566 ) = 720720
 		 | ( n < 2^2082 ) = 1663200
 		 | ( n < 2^3217 ) = 8648640 
-                 | ( n < 2^4253 ) = 69189120
-                 | ( n < 2^4423 ) = 69189120
+                 | ( n < 2^4253 ) = 691891200
                  | ( n < 2^44497) = 137944178989
 		 | otherwise = hilfsf  137944178989 n 
             where hilfsf = (\x w -> if ((e x)*(e x) > w) then x else (hilfsf (x + x) w))

### 2017-12-15 benchmarks/rm/bench.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 
-procCount=512
+procCount=256
 
 programs=(
     "skelrm-parrows"

### 2017-12-15 benchmarks/sudoku/bench.sh

@@ -3,13 +3,13 @@
 procCount=16
 
 programs=(
-#    "parmonad-sudoku"
-#    "parrows-sudoku-parmap-eden"
-#    "parrows-sudoku-parmap-mult"
-#    "parrows-sudoku-parmap-par"
-#    "sudoku-seq"
-     "multicore-sudoku"
-     "eden-sudoku"
+    "parmonad-sudoku"
+    "parrows-sudoku-parmap-eden"
+    "parrows-sudoku-parmap-mult"
+    "parrows-sudoku-parmap-par"
+    "sudoku-seq"
+    "multicore-sudoku"
+    "eden-sudoku"
 )
 
 parameters=(

### 2017-12-15 paper/content/arrows.tex

@@
 Also note that, while we could have based our DSL on Profunctors as well, we chose Arrows.
-for this paper since they they are a more direct translation of pure functions than Profunctors.
-Note that Profunctors can easily be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
+for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
 However, they are a promising candidate for future improvements of our DSL.
+Note that some Profunctors, especially ones supporting a composition operation among other things, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.

### 2017-12-15 paper/content/parrows.tex

@@
 \begin{code}
-instance (Trans a, Trans b) => ArrowParallel (->) a b (Conf b) where
-    parEvalN _ fs as = spawnF fs as
+instance (Trans a, Trans b) => ArrowParallel (->) a b Conf where
+    parEvalN _ = spawnF
 \end{code}
@@
 \begin{code}
-instance (ArrowParallel (->) a (m b) (Conf b),
+instance (ArrowParallel (->) a (m b) Conf,
   Monad m, Trans a, Trans b, Trans (m b)) =>
@@
 \end{code}
-where |Conf a| is simply defined as |data Conf a = Nil| since Eden does not have a configurable |spawnF| variant.
+where |Conf| is simply defined as |data Conf = Nil| since Eden does not have a configurable |spawnF| variant.
 % \caption{ArrowParallel instance for the Kleisli type in the Eden backend}

### 2017-12-15 paper/content/topologySkeletons.tex
reworking contributions
minor

@@
 %\FloatBarrier
-\section{Topological Skeletons}
+\subsection{Topological Skeletons}
 \label{sec:topology-skeletons}
@@
 
-\subsection{Parallel pipe}\label{sec:pipe}
+\subsubsection{Parallel pipe}\label{sec:pipe}
 
@@
 |arr a b| and |arr b c|.
-By wrapping these two arrows inside a common type we obtain |pipe2| (Fig.~\ref{fig:pipe2}).
+By wrapping these two arrows inside a bigger arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
 \begin{figure}[tb]
@@
 
-\subsection{Ring skeleton} \label{sec:ring}
+\subsubsection{Ring skeleton} \label{sec:ring}
 \begin{figure}[tb]
@@
 
-Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|.
+Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|.
 Finally we unzip this list into |([o], [fut r])|.
@@
 
-\subsection{Torus skeleton}\label{sec:torus}
+\subsubsection{Torus skeleton}\label{sec:torus}
 \begin{figure}

### 2017-12-15 paper/content/utilityFunctions.tex
wip

@@
 These combinators make use of the |ArrowChoice| type class which provides the |pipepipepipe| combinator.
-It takes two arrows |arr a c| and |arr b c| and combines them into a new arrow |arr (Either a b) c| which pipes all |Left a|'s to the first arrow and all |Right b|'s to the second arrow:
+It takes two Arrows |arr a c| and |arr b c| and combines them into a new Arrow |arr (Either a b) c| which pipes all |Left a|'s to the first arrow and all |Right b|'s to the second arrow:
 % \begin{figure}[htb]
@@
 In Fig.~\ref{fig:profunctorArrow} we show how specific Profunctors can be turned into Arrows.
-Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| typeclass.
+This works because Arrows are strong Monads in the bicategory Prof of Profunctors as shown by \citet{Asada:2010:ASM:1863597.1863607}.
+Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| typeclass like presented in this paper.
 

### 2017-12-16 Definition/src/main/Parrows/Skeletons/Topology.hs
minor

@@ -65,8 +65,9 @@ pipeSimple conf fs =
         (arr (uncurry (:) >>> lazy) >>> loopParEvalN conf fs)) >>>
     arr last
 
-ring :: (ArrowLoop arr, ArrowLoopParallel arr (i, fut r) (o, fut r) conf,
-    Future fut r conf,
+ring :: (Future fut r conf,
+    ArrowLoop arr,
+    ArrowLoopParallel arr (i, fut r) (o, fut r) conf,
     ArrowLoopParallel arr o o conf) =>
     conf -> arr (i, r) (o, r) -> arr [i] [o]
 ring conf f =
@@ -77,11 +78,11 @@ ring conf f =
     postLoopParEvalN conf (repeat (arr id))
 
 --TODO: check whether this exchanges the futures the same way as Eden does it
-torus :: (ArrowLoop arr, ArrowChoice arr,
-            ArrowLoopParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
-            Future fut a conf, Future fut b conf,
-            ArrowLoopParallel arr [d] [d] conf) =>
-         conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
+torus :: (Future fut a conf, Future fut b conf,
+      ArrowLoop arr, ArrowChoice arr,
+      ArrowLoopParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
+      ArrowLoopParallel arr [d] [d] conf) =>
+      conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
 torus conf f =
     loop (second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy)) >>>
         arr (uncurry3 (zipWith3 lazyzip3)) >>>

### 2017-12-16 paper/content/topologySkeletons.tex

@@
 If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing.
-This materializes with the usage of |loop| of the |ArrowLoop| type class and is probably due to the way their respective parallelism engines work internally.
-\mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| type class.
-This allows us for selective enabling and disabling of parallelism.
+This materializes with the usage of |loop| of the |ArrowLoop| type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden.
+An investigation of where this difference comes from is beyond the scope of this work, we only provide a workaround for these types of skeletons as such skeletons probably are not of much importance outside of a distributed memory environment.
+However our workaround enables users of the DSL to test their code within a shared memory setting even though other skeletons would be better suited to be run with them.
+The idea of the fix is to provide a |ArrowLoopParallel| typeclass that has two functions - |loopParEvalN| and |postLoopParEvalN| where the first is to be used inside an |loop| construct while the latter will be used right outside of the |loop|.
+This way we can delegate to the actual |parEvalN| in the spot where the backend supports it.
 \begin{code}
-class ArrowParallel arr a b conf => FutureEval arr a b conf where
-    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
+class ArrowParallel arr a b conf => ArrowLoopParallel arr a b conf where
+    loopParEvalN :: conf -> [arr a b] -> arr [a] [b]
+    postLoopParEvalN :: conf -> [arr a b] -> arr [a] [b]
 \end{code}
-As Eden has no problems with the looping skeletons, we declare a delegating instance:
+As Eden has no problems with the looping skeletons, we use this instance:
 \begin{code}
-instance ArrowParallel arr a b conf => FutureEval arr a b conf where
-    evalN = parEvalN
+instance (ArrowChoice arr, ArrowParallel arr a b Conf) => ArrowLoopParallel arr a b Conf where
+    loopParEvalN = parEvalN
+    postLoopParEvalN _ = evalN
 \end{code}
-The Par Monad and GpH backends have parallelism disabled in their instance of |FutureEval|.
-This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
+As the backends for the |Par| Monad and GpH have problems with |parEvalN| inside of |loop| their respective instances for |ArrowLoopParallel| look like this:
 \begin{code}
-instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
-    ArrowParallel arr a b conf) => FutureEval arr a b conf where
-    evalN _ = listApp
+instance (ArrowChoice arr, ArrowParallel arr a b (Conf b)) => ArrowLoopParallel arr a b (Conf b) where
+    loopParEvalN _ = evalN
+    postLoopParEvalN = parEvalN
 \end{code}
@@
 
-However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
-Therefore, we introduce a more sophisticated version that internally uses Futures and obtain the final definition of |pipe| in Fig.~\ref{fig:pipe}.
-
 \begin{figure}[t]
 \begin{code}
-pipeSimple :: (ArrowLoop arr, FutureEval arr a a conf) =>
+pipeSimple :: (ArrowLoop arr, ArrowLoopParallel arr a a conf) =>
 	conf -> [arr a a] -> arr a a
@@
     loop (arr snd &&&
-		(arr (uncurry (:) >>> lazy) >>> evalN conf fs)) >>>
+        (arr (uncurry (:) >>> lazy) >>> loopParEvalN conf fs)) >>>
     arr last
@@
 
+However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}.
+Therefore, we introduce a more sophisticated version that internally uses Futures and obtain the final definition of |pipe| in Fig.~\ref{fig:pipe}.
+
 \begin{code}
-pipe :: (ArrowLoop arr, FutureEval arr (fut a) (fut a) conf, Future fut a) =>
+pipe :: (ArrowLoop arr, ArrowLoopParallel arr (fut a) (fut a) conf,
+	Future fut a conf) =>
 	conf -> [arr a a] -> arr a a
-pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
+pipe conf fs = unliftFut conf (pipeSimple conf (map (liftFut conf) fs))
+
+liftFut :: (Arrow arr, Future fut a conf, Future fut b conf) =>
+	conf -> arr a b -> arr (fut a) (fut b)
+liftFut conf f = get conf >>> f >>> put conf
+
+unliftFut :: (Arrow arr, Future fut a conf, Future fut b conf) =>
+	conf -> arr (fut a) (fut b) -> arr a b
+unliftFut conf f = put conf >>> f >>> get conf
 \end{code}
@@
 \begin{code}
-pipe2 :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
-         FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
+pipe2 :: (ArrowLoop arr, ArrowChoice arr,
+    ArrowLoopParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
+    Future fut (([a], [b]), [c]) conf) =>
 	conf -> arr a b -> arr b c -> arr a c
@@
     pipe conf (replicate 2 (unify f g)) >>>
-	arr snd >>> arr head where
-			unify :: (ArrowChoice arr) => arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
-			unify f g =
-				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
-				arr (\((a, b), c) -> ((c, a), b))
+    arr snd >>>
+    arr head
+    where
+        unify :: (ArrowChoice arr) => 
+			arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
+        unify f' g' =
+			(mapArr f' *** mapArr g') *** arr (const []) >>>
+			arr (\((b, c), a) -> ((a, b), c))
 
-(parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
-	FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
+(parcomp) :: (ArrowLoop arr, ArrowChoice arr,
+    ArrowLoopParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
+    Future fut (([a], [b]), [c]) ()) =>
     arr a b -> arr b c -> arr a c
@@
 \end{code}
-\caption{Definition of |pipe2| and a parallel |>>>|.}
+\caption{Definition of |pipe2| and |(parcomp)|, a parallel |>>>|.}
 \label{fig:pipe2}
@@
 
-Another version of |>>>| is:
-\begin{code}
-f parcomp g = (f . put) >>> (get . g)
-\end{code}
-It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them.
+%Another version of |>>>| is:
+%\begin{code}
+%f parcomp g = (f . put) >>> (get . g)
+%\end{code}
+%It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them.
 Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.

### 2017-12-16 paper/content/topologySkeletons.tex
minor

@@
 \end{figure}
-Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function |[i] -> [o]| with a ring of nodes that communicate in a ring topology with each other.
+Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function |[i] -> [o]| with a ring of nodes that communicate with each other.
 Its input is a node function |i -> r -> (o, r)| in which |r| serves as the intermediary output that gets send to the neighbour of each node.
@@
 
-Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|.
+Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| with |second get >>> f >>> second put| before converting it into a parallel arrow |arr [(i, fut r)] [(o, fut r)]| with the help of |repeat| and |loopParEvalN| to get an output list |[(o, fut r)]|.
 Finally we unzip this list into |([o], [fut r])|.
 Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
+To make sure this algorithm has speedup on shared-memory machines as well, we pass the result of this arrow to |postLoopParEvalN conf (repeat (arr id))|.
 This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
@@
 \begin{code}
-ring :: (ArrowLoop arr, Future fut r, FutureEval arr (i, fut r) (o, fut r) conf) =>
+ring :: (Future fut r conf,
+    ArrowLoop arr,
+    ArrowLoopParallel arr (i, fut r) (o, fut r) conf,
+    ArrowLoopParallel arr o o conf) =>
     conf -> arr (i, r) (o, r) -> arr [i] [o]
 ring conf f =
-	loop (second (rightRotate >>> lazy) >>> arr (uncurry zip) >>>
-        evalN conf (repeat (second get >>> f >>> second put)) >>> arr unzip)
+    loop (second (rightRotate >>> lazy) >>>
+        arr (uncurry zip) >>>
+        loopParEvalN conf (repeat (second (get conf) >>> f >>> second (put conf))) >>>
+        arr unzip) >>>
+    postLoopParEvalN conf (repeat (arr id))
 \end{code}
-\caption{Final definition of the |ring| skeleton.}
+\caption{Definition of the |ring| skeleton.}
 \label{fig:ringFinal}
 \end{figure}
- % $ %% formatting
-% and lazy:
-% \begin{code}
-% lazy :: (Arrow arr) => arr [a] [a]
-% lazy = arr (\ ~(x:xs) -> x : lazy xs
-% \end{code}
-%% we have shown this already.
 
@@
 
-Similar to the |ring|, we once again start by rotating the input (Fig.~\ref{fig:edenlazyrightrotate}), but this time not only in one direction, but in two.
+Similar to the |ring|, we start by rotating the input (Fig.~\ref{fig:edenlazyrightrotate}), but this time not only in one direction, but in two.
 This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes
@@
 \end{code}
-instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we once again zip the inputs for the input arrow with 
+instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we zip the inputs for the input arrow with 
 \begin{code}
@@
 from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then feed into our parallel execution.
-
+\mbcomment{fix this. transition is bumpy}
 This action is, however,  more complicated than in the ring case as we have one more dimension of inputs that needs to be transformed.
-In order to achieve that, we first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| type class.
+In order to achieve this, we first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|.
+We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffling arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| type class.
 Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.
@@
 \begin{code}
-torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr, Future fut a, Future fut b,
-          FutureEval arr (c, fut a, fut b) (d, fut a, fut b) conf) =>
+torus :: (Future fut a conf, Future fut b conf,
+      ArrowLoop arr, ArrowChoice arr,
+      ArrowLoopParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
+      ArrowLoopParallel arr [d] [d] conf) =>
       conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
@@
         arr (uncurry3 (zipWith3 lazyzip3)) >>>
-	(arr length >>> arr unshuffle) &&& (shuffle >>> evalN conf (repeat (ptorus f)) >>> app >>>
-	arr (map unzip3) >>> arr unzip3 >>> threetotwo)
+        arr length &&& (shuffle >>> loopParEvalN conf (repeat (ptorus conf f))) >>>
+        arr (uncurry unshuffle) >>>
+        arr (map unzip3) >>> arr unzip3 >>> threetotwo) >>>
+    postLoopParEvalN conf (repeat (arr id))
 

### 2017-12-17 Eden/src/test/EdenSpec.hs
fixed testsuite for Eden

@@ -3,7 +3,8 @@ module EdenSpec (spec) where
 import Parrows.Definition
 import Parrows.Skeletons.Topology
 import Parrows.Skeletons.Map
-import Parrows.Eden()
+import Parrows.Eden
+import Parrows.Eden.Simple
 import Parrows.Future
 
 import Control.Arrow
@@ -33,7 +34,7 @@ parEvalSpec = describe "Basic Parrow Functionality Eden" $ do
         parEvalNLazyInt xs = parEvalNLazy () 4 (repeat (+1)) xs == map (+1) xs
 
         parEvalNFutInt :: [Int] -> Bool
-        parEvalNFutInt xs = (map (get) $ parEvalNFut () (repeat (+1)) (map put xs)) == map (+1) xs
+        parEvalNFutInt xs = (map (get ()) (parEvalNFut () (repeat (+1)) (map (put ()) xs))) == map (+1) xs
 
         parEval2Int :: (Int, Int) -> Bool
         parEval2Int (x, y) = (parEval2 () (+1) (*2) (x, y)) == (x+1,y*2)
@@ -87,4 +88,4 @@ mapReduceSpec = describe "parMapReduceDirect Test" $
     prop "Basic parMapReduceDirect Test" $ parMapReduceDirectTest
       where
         parMapReduceDirectTest :: [Int] -> Bool
-        parMapReduceDirectTest xs = (parMapReduceDirect () 4 (+1) (uncurry (*)) 0 xs) == (foldl (*) 0 $ map (+1) xs)
+        parMapReduceDirectTest xs = (parMapReduceDirect () 4 (+1) (uncurry (*)) (0, xs)) == (foldl (*) 0 $ map (+1) xs)

### 2017-12-17 paper/content/parrows.tex

@@
 
-Up until now we discussed Arrow operations more in detail, but in the following sections we focus on the dataflow instead of how the Arrows are obtained so we can focus on the important parts now that we have seen that Arrows are capable of expressing parallelism.
+Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the dataflow between the arrows, now that we have seen that Arrows are capable of expressing parallelism.
+We do however explain new concepts with more details if required for easier understanding.

### 2017-12-17 paper/content/topologySkeletons.tex
topology chapter fixed

@@
 \end{code}
-from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then feed into our parallel execution.
-\mbcomment{fix this. transition is bumpy}
-This action is, however,  more complicated than in the ring case as we have one more dimension of inputs that needs to be transformed.
-In order to achieve this, we first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|.
-We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffling arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| type class.
-Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.
+from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then evaluate in parallel.
+
+This, however, is more complicated than in the ring case as we have one more dimension of inputs that needs to be transformed.
+We first have to |shuffle| all the inputs to then pass them into |loopParEvalN conf (repeat (ptorus conf f))| to get an output of |[(d, fut a, fut b)]|.
+We then unshuffle this list back to its original ordering by feeding it into |arr (uncurry unshuffle)| which takes the input length we saved one step earlier as additional input to get a result matrix |[[[(d, fut a, fut b)]]|.
+Finally, we unpack this matrix  with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get |([[d]], ([[fut a]], [[fut b]]))|.
+
+This internal looping computation is once again fed into |loop| and we also compose a final |postLoopParEvalN conf (repeat (arr id))| for the same reasons as explained for the |ring| skeleton.
 
@@
 \end{figure}
-As an example of using this skeleton \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
+As an example of using this skeleton, \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
 An adapted version can be found in Fig.~\ref{fig:torusMatMult}.
@@
 If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version and execution times are comparable.
-We discuss further examples on larger clusters and in a more detail in the next section.
+We discuss further benchmarks on larger clusters and in a more detail in the next section.
 \begin{figure}[tb]

### 2017-12-17 paper/content/topologySkeletons.tex
minor

@@
 
-ptorus :: (Arrow arr, Future fut a, Future fut b) =>
-	arr (c, a, b) (d, a, b) -> arr (c, fut a, fut b) (d, fut a, fut b)
-ptorus f = arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>> arr (\ ~(d, a, b) -> (d, put a, put b))
+ptorus :: (Arrow arr, Future fut a conf, Future fut b conf) =>
+          conf ->
+          arr (c, a, b) (d, a, b) ->
+          arr (c, fut a, fut b) (d, fut a, fut b)
+ptorus conf f =
+	arr (\ ~(c, a, b) -> (c, get conf a, get conf b)) >>>
+	f >>>
+	arr (\ ~(d, a, b) -> (d, put conf a, put conf b))
 \end{code} % $

### 2017-12-17 paper/content/motivation.tex
minor
minor fix in motivation.tex

@@
 \label{sec:introduction}
-% \olcomment{todo, reuse 5.5, and more}
-%
-% \mbcomment{
-% Haskell is Spielwiese für Parallelität; verschiedene Ansätze (Par, Multicore, Eden); Orthogonale Ansätze; Verwenden höchstens eine Monade - manchmal auch nur intern; Wir wollen Parallelität mit Arrows abbilden, was noch niemand gemacht hat;
-% Statt einer eigenen Implementierung definieren wir ein "shallow embedded DSL" (ACHTUNG, ist das der richtige Name? effektiv API);
-% Umsetzung mit verschiedenen parallelen Haskells; We tame the zoo of parallel Haskells und vergewissern uns dass es nicht viel Overhead bringt
-% }
-%
-% blablabla arrows, parallel, haskell.
-%
-% \olcomment{An attempt ---->}
-%%%%%
 
@@
 
-%In this paper we introduce a notion of parallel
-%computations using Arrows.
 A key novelty in this paper is to use Arrows to represent parallel computations.
-They seem a natural fit as they are a generalization of the function arrow (|->|) and serve as general interface to computations.
+They seem a natural fit as they can be thought of as a more general function arrow (|->|) and serve as general interface to computations while not being as restrictive as Monads \citep{HughesArrows}.
 Section \ref{sec:arrows} gives a short introduction to Arrows.
 
-%Our Arrows-based interface is a high-level one.
 We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
@@
 %\olcomment{move this to Contributions in the front or something}
-We propose an Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]| in Section \ref{sec:parallel-arrows}, which converts a list of Arrows into a new parallel Arrow.
+We propose an Arrow-based encoding for parallelism based on a new Arrow combinator |parEvalN :: [arr a b] -> arr [a] [b]|.
 A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.
+In this paper we evaluate the expressive power of such a formalism in the context of parallel programming.
 
-In this paper evaluate the expressive power of our Arrow formalism in the context of parallel programming.
+We structure this paper as follows:
 
@@
 
-\mbcomment{Überleitung auf Related Work hier machen}
-
-\mbcomment{Mention Future work here?}
-
-\mbcomment{Conclusion?}
+PArrows are open source and available from \url{https://github.com/s4ke/Parrows}.
 

### 2017-12-17 paper/content/benchmarks.tex
wip on benchmarks (once again)

@@
 %
-\section{Performance results}
+\section{Performance results and discussion}
 \label{sec:benchmarks}
@@
 We start by explaining the hardware and software stack and also elaborate on which benchmarks programs were used
-and also which parallel Haskell were used in which setting. Before we go into detail on the benchmarks we
+and also which parallel Haskell were used in which setting and why. Before we go into detail on the benchmarks we
 also address the issue of poor performance gains when using hyper-threading and our reasoning behind not including
@@
 
-The \ensuremath{\Conid{Par}} monad and GpH -- in its multicore version \cite{Marlow2009} --  can be executed on a shared
-memory machines only. Although GpH is available on distributed memory
+The \ensuremath{\Conid{Par}} monad and GpH -- in its multicore version \cite{Marlow2009} --  can be executed on shared memory machines only.
+Although GpH is available on distributed memory
 clusters, and newer distributed memory Haskells such as HdpH exist,
@@
 
+\mbcomment{fix this when benchmarks are redone}
 Hence, the performance measurements in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark} and Table~\ref{tab:meanOverhead} use only real cores as the purpose of this paper is to show the performance of PArrows and not to investigate parallel performance with hyper-threading in use.
@@
 \subsection{Benchmark results}
-To show that PArrows induce very small overhead even in a distributed context (and in general as well), we compare the original Eden
-versions of the benchmark to its PArrows-enabled counterpart in the \rmtest, \torustest and \jacobitest benchmarks. We plot execution time differences between measurements for
-PArrows and the corresponding backend in a separate plot
-(Figs.~\ref{fig:bench-rm-dist}--\ref{fig:torusBenchmark}). As an example, the differences range in
-about $0.5\seconds$ for the execution time of $46\seconds$ on 256 cores
-for distributed \rmtest with PArrows and Eden. For these comparisons, the plots show absolute
-time differences that are not relative \wrt the total execution time.
-Furthermore, the error bars ends were computed from point-wise maximum of both standard
-deviations from both measurements for PArrows and non-PArrows
-versions. These are the values provided by the \ensuremath{\Varid{bench}} package that we
-used for benchmarking. We call a difference between two versions
-significant when the border of the error bar of absolute time
-difference is above or below zero. In other words: the time
-difference is significant if it is outside of the measurement error.
 
-\subsubsection{\rmtest}\label{sec:rmtest}
+In the following paragraphs we will go into detail on how well PArrows perform when compared to versions of the benchmark programs implemented with the original parallel Haskells that were used in the backends.
+We start by comparing Speedups and Overheads of for the shared memory backends and conclude with OpenMPI variants of the Eden-enabled backend as a representative of a distributed memory backend.
 
-\newcommand{\performanceSkelRMSM}[2]{
-\performanceplot{Parallel run time of \rmtest \enquote{#2}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
-}{17}{\plotwidthSMP}
-}
+\subsubsection{Shared memory}
 
-\newcommand{\speedupSkelRMSM}[2]{
-\speedupplot{Speedup of \rmtest \enquote{#2}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
-}{17}{\plotwidthSMP}
-}
+\paragraph{Speedup}
 
-\newcommand{\speedupSkelRMDist}[4]{
-\speedupplot{Speedup of \rmtest \enquote{#1 #2}}{PArrows}{256}{#3}{
-% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
-\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
-% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-}{#4}{\plotwidthDist}
-}
+\paragraph{Overhead}
 
-\newcommand{\performanceSkelRMDistDiff}[5]{
-\performancediffplot{Overhead\\for \rmtest \enquote{#1 #2}}{(Eden $-$ PArrows)}{256}{#3}{
-\addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
-smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
-}{#4}{#5}{\plotwidthDist}
-}
+\subsubsection{Distributed Memory}
 
-\begin{figure}
-%\centering
-%\performanceSkelRMSM{11213}{64}\hfill%
-{\speedupSkelRMSM{11213}{32}}\hfill%
-{\speedupSkelRMSM{11213}{64}}
-\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware. Measurements were performed on a single node of the Glasgow
-  cluster; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
-  tasks. The
-  closer to linear speedup the better.}
-\label{fig:bench-rm-sm}
-\end{figure}
+\paragraph{Speedup}
 
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupSkelRMDist{44497}{256}{32}{272}\label{subfig:rm-dist-a}}%
-%\hfill%
-{\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{272}\label{subfig:rm-dist-b}}
-\caption{Parallel performance of the \rmtest on the Glasgow cluster
-  consisting of 256 cores. Input was $2^{44497}-1$, we used 256
-  tasks. The top plot shows absolute speedup in a distributed memory setting. The
-  closer to linear speedup the better. Time
-  (and hence speedup) measurements for PArrows with Eden backend and
-  Eden almost coincide. Hence, bottom plot shows
-absolute time differences for this benchmark. The
-higher the value, the better for PArrows\olcomment{CHECKME}.}
-\label{fig:bench-rm-dist}
-\end{figure}
-
-%\olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
-%  Why is it good?}
-The multicore version of our parallel \rmtest benchmark is depicted in
-Figure~\ref{fig:bench-rm-sm}. We executed the test with 32 and 64
-tasks. The plot shows the PArrows-enabled versions with corresponding backends.
-The performance of PArrows/Eden~CP in shared memory is slightly better than
-for SMP variants such as PArrows/GpH and PArrows/\ensuremath{\Conid{Par}}
-Monad but most of the time the performance is still comparable with the GpH backend performing slightly worse than the other two in terms of speedup.
-
-Comparing the PArrows version of the \rmtest with the original from Eden with the MPI backend in a distributed memory setting, we see an almost linear speedup of
-\rmtest with 256 tasks and input $2^{44497}-1$ in both versions. The sequential run time
-was computed as the mean of three consecutive executions on a single
-core---the single run took two hours 43 minutes. The difference between
-PArrows/Eden and Eden almost always lies inside the error bar of
-the measurement.
-
-%As the PArrows version uses Eden in the backend, these numbers suggest that there is no real performance difference between using PArrows or Eden for this task as they trade blows in this benchmark.
-Additionally, PArrows with an Eden-based backend performing better than what it is based upon suggests that any difference in runtime between the two is more of an anomaly than a real difference.
-
-\subsubsection{\jacobitest}
-
-Continuing, the results of the \jacobitest (MPI only) in Fig.~\ref{fig:bench-jacobi-dist} are as follows:
-The program does not seem to scale well beyond 64 threads with input $2^{3217}-1$.
-We once again compare the Eden version with the PArrrows version in Fig.~\ref{fig:bench-jacobi-dist}.
-We see similar behaviour to the MPI version of the \rmtest: The difference once again almost always remains within the bounds of the error bar of the measurement.
-
-Because of the bad scaling behaviour for $2^{3217}-1$ beyond 64 threads, we also ran tests with input $2^{4253}-1$.
-Because of the long running time, we could do this only for the 128 and 256 threads.
-Therefore we do not show these results in Fig.~\ref{fig:bench-jacobi-dist} as we cannot properly compute a speedup without results for 1 thread.
-Nonetheless, for 128 threads, the benchmark took a mean of 9192.9\seconds, while with 256 threads, it only took a mean of 1649.1\seconds.
-This means that the 128 thread version ran more than 5.5 times slower than the 256 one, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores.
-However it still proves that PArrows continue to scale, even if not perfectly for this test program.
-Comparing PArrows with Eden, for the larger input of $2^{4253}-1$, we see slightly bigger differences between Eden and PArrows: The differences for 128 and 256 core sare $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden, respectively.
-This maximum of $12.27\%$ slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.
-
-\newcommand{\speedupJacobiDist}[5]{
-\speedupplot{Speedup of \jacobitest \enquote{#2} vs simulated \enquote{#5}}{PArrows #2, simulated PArrows #5}{256}{#3}{
-% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
-\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#2.csv};
-\addplot [mark=x,very thick,red] table [scatter, x="nCores", y="speedup"/, col sep=comma, mark=none,
-smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#5.csv};
-% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
-}{#4}{\plotwidthDist}
-}
-
-\newcommand{\performanceJacobiDistDiff}[5]{
-\performancediffplot{Overhead\\for \jacobitest \enquote{#2}}{(Eden $-$ PArrows) #2}{256}{#3}{
-\addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
-smooth]{benchmarks/distributed-jacobi/#1-#2-diff.csv};
-}{#4}{#5}{\plotwidthDist}
-}
-
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupJacobiDist{3}{3217}{32}{272}{4253}\label{subfig:jacobi-dist-a}}%
-%\hfill%
-{\performanceJacobiDistDiff{3}{3217}{32}{5}{272}\label{subfig:jacobi-dist-b}}
-\caption{Parallel performance of the \jacobitest on the Glasgow cluster
-  consisting of 256 cores. Input was $2^{3217}-1$, we used 256
-  tasks. The top plot shows relative speedup in a distributed memory setting compared to a simulated speedup for input $2^{4253}-1$. The
-  closer to linear speedup the better. Time
-  (and hence speedup) measurements for PArrows with Eden backend and
-  Eden almost coincide. Hence, bottom plot shows
-absolute time differences for this benchmark. The
-higher the value, the better for PArrows\olcomment{CHECKME}.}
-\label{fig:bench-jacobi-dist}
-\end{figure}
-
-\subsubsection{\torustest}
-
-Next is the \torustest benchmark.
-The results of the comparison of vanilla Eden to our PArrows-based version can be found in Fig.~\ref{fig:torusBenchmark}.
-We see that the benchmark scales quite well with more cores until 64 cores.
-For $>=96$ cores, we still have considerable speedup, but with less slope.
-We also prove that the difference between the Eden and PArrows version are only marginal with PArrows only being a maximum of $1.7\%$ slower -- for 160 cores -- when outside of the error bar.
-%The difference between PArrows and Eden is only significant for 16 and 64 cores where it ran 1.7\% and 2.7\% slower which corresponds to a real-time difference of 0.12s and 0.13s.
-For 256 cores PArrows performed 0.2\% slower which corresponds to 0.01s overhead.
-
-\newcommand{\speedupTorusDist}[3]{
-\speedupplot{Speedup of \torustest \enquote{#1}}{PArrows}{256}{#2}{
-\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-#1.csv};
-}{#3}{\plotwidthDist}
-}
-
-\newcommand{\performanceTorusDistDiff}[4]{
-\performancediffplot{Overhead\\for \torustest \enquote{#1}}{(Eden $-$ PArrows)}{256}{#2}{
-\addplot+[mark=*,very thick,error bars/.cd,
-    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
-smooth]{benchmarks/distributed-torus/#1-diff.csv};
-}{#3}{#4}{\plotwidthDist}
-}
-
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupTorusDist{4096}{32}{272}\label{subfig:speedupTorusDist}}%
-%\hfill%
-{\performanceTorusDistDiff{4096}{32}{0.5}{272}\label{subfig:performancetorusDistDiff}}
-\caption{Parallel performance of \torustest on the Glasgow cluster
-  consisting of 256 cores. Input was a matrix size of $4096$. The top plot shows absolute speedup in a distributed memory setting. The
-  closer to linear speedup the better. Time
-  (and hence speedup) measurements for PArrows with Eden backend and
-  Eden almost coincide. Hence, bottom plot shows
-absolute time differences for this benchmark. The
-higher the value, the better for PArrows\olcomment{CHECKME}.}
-\label{fig:torusBenchmark}
-\end{figure}
-
-\subsubsection{\sudokutest}
-
-As the last benchmark in this paper we present the \sudokutest in Fig.~\ref{fig:sudokuSMBenchmark} running in a shared memory setting.
-Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark} and \ref{fig:sudokuSMBenchmark16000}.
-However, we notice that the GpH backend seems to choke on a bigger input (Fig.~\ref{fig:sudokuSMBenchmark16000}).
-This is due to the benchmark only using \ensuremath{\Varid{parMap}} instead of a chunking variant -- however we did not change that for simplicity's sake.
-This issue is reflected by debug output which shows that of 16000 sparks being created (one for each Sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool.
-Another remarkable finding is that the Eden backend seems to lag behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.
-
-\newcommand{\performanceSudokuSM}[1]{
-\performanceplot{Parallel run time of \sudokutest \enquote{#1}}{linear speedup, Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
-\addplot [no markers,dotted,thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth] {benchmarks/sudoku-sm/bench-sudoku-sm.bench.fake-linear-sudoku-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
-%\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
-%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
-}{17}{\plotwidthSMP}
-}
-
-\newcommand{\speedupSudokuSM}[1]{
-\speedupplot{Parallel speedup of \sudokutest \enquote{#1}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
-%\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
-}{17}{\plotwidthSMP}
-}
-
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupSudokuSM{1000}\label{subfig:speedupSudokuSM}}%
-%\hfill%
-{\performanceSudokuSM{1000}\label{subfig:performanceSudokuSM}}
-\caption{Absolute speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
-  cluster; it has 16 real cores and 32 threads. Input was a file of $1000$ Sudokus. The
-  closer to linear speedup the better.}
-\label{fig:sudokuSMBenchmark}
-\end{figure}
-
-\begin{figure}
-\centering
-%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
-%
-{\speedupSudokuSM{16000}\label{subfig:speedupSudokuSM16000}}%
-%\hfill%
-{\performanceSudokuSM{16000}\label{subfig:performanceSudokuSM16000}}
-\caption{Absolute speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
-  different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
-  cluster; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
-  closer to linear speedup the better. The GpH version shows signs of choking with too many sparks being created.}
-\label{fig:sudokuSMBenchmark16000}
-\end{figure}
+\paragraph{Overhead}
 

### 2017-12-19 paper/content/benchmarks_old.tex

To show that PArrows induce very small overhead even in a distributed context (and in general as well), we compare the original Eden
versions of the benchmark to its PArrows-enabled counterpart in the \rmtest, \torustest and \jacobitest benchmarks. We plot execution time differences between measurements for
PArrows and the corresponding backend in a separate plot
(Figs.~\ref{fig:bench-rm-dist}--\ref{fig:torusBenchmark}). As an example, the differences range in
about $0.5\seconds$ for the execution time of $46\seconds$ on 256 cores
for distributed \rmtest with PArrows and Eden. For these comparisons, the plots show absolute
time differences that are not relative \wrt the total execution time.
Furthermore, the error bars ends were computed from point-wise maximum of both standard
deviations from both measurements for PArrows and non-PArrows
versions. These are the values provided by the \ensuremath{\Varid{bench}} package that we
used for benchmarking. We call a difference between two versions
significant when the border of the error bar of absolute time
difference is above or below zero. In other words: the time
difference is significant if it is outside of the measurement error.

\subsubsection{\rmtest}\label{sec:rmtest}

\newcommand{\performanceSkelRMSM}[2]{
\performanceplot{Parallel run time of \rmtest \enquote{#2}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}{17}{\plotwidthSMP}
}

\newcommand{\speedupSkelRMSM}[2]{
\speedupplot{Speedup of \rmtest \enquote{#2}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-#1-#2.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sm-rm/bench-sm-rm.bench.skelrm-parr-par-#1-#2.csv};
}{17}{\plotwidthSMP}
}

\newcommand{\speedupSkelRMDist}[4]{
\speedupplot{Speedup of \rmtest \enquote{#1 #2}}{PArrows}{256}{#3}{
% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-#1-#2.csv};
% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
}{#4}{\plotwidthDist}
}

\newcommand{\performanceSkelRMDistDiff}[5]{
\performancediffplot{Overhead\\for \rmtest \enquote{#1 #2}}{(Eden $-$ PArrows)}{256}{#3}{
\addplot+[mark=*,very thick,error bars/.cd,
    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
smooth]{benchmarks/distributed-rm/#1-#2-diff.csv};
}{#4}{#5}{\plotwidthDist}
}

\begin{figure}
%\centering
%\performanceSkelRMSM{11213}{64}\hfill%
{\speedupSkelRMSM{11213}{32}}\hfill%
{\speedupSkelRMSM{11213}{64}}
\caption{Relative speedup of \rmtest on a multicore machine. We used the same PArrows-based implementation with
  different backends on the same hardware. Measurements were performed on a single node of the Glasgow
  cluster; it has 16 real cores. Input was $2^{11213}-1$, we used 32 (left) or 64 (right)
  tasks. The
  closer to linear speedup the better.}
\label{fig:bench-rm-sm}
\end{figure}

\begin{figure}
\centering
%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
%
{\speedupSkelRMDist{44497}{256}{32}{272}\label{subfig:rm-dist-a}}%
%\hfill%
{\performanceSkelRMDistDiff{44497}{256}{32}{0.5}{272}\label{subfig:rm-dist-b}}
\caption{Parallel performance of the \rmtest on the Glasgow cluster
  consisting of 256 cores. Input was $2^{44497}-1$, we used 256
  tasks. The top plot shows absolute speedup in a distributed memory setting. The
  closer to linear speedup the better. Time
  (and hence speedup) measurements for PArrows with Eden backend and
  Eden almost coincide. Hence, bottom plot shows
absolute time differences for this benchmark. The
higher the value, the better for PArrows\olcomment{CHECKME}.}
\label{fig:bench-rm-dist}
\end{figure}

%\olcomment{THE ACTUAL TEXT IS MISSING. What do we see in the plots?
%  Why is it good?}
The multicore version of our parallel \rmtest benchmark is depicted in
Figure~\ref{fig:bench-rm-sm}. We executed the test with 32 and 64
tasks. The plot shows the PArrows-enabled versions with corresponding backends.
The performance of PArrows/Eden~CP in shared memory is slightly better than
for SMP variants such as PArrows/GpH and PArrows/\ensuremath{\Conid{Par}}
Monad but most of the time the performance is still comparable with the GpH backend performing slightly worse than the other two in terms of speedup.

Comparing the PArrows version of the \rmtest with the original from Eden with the MPI backend in a distributed memory setting, we see an almost linear speedup of
\rmtest with 256 tasks and input $2^{44497}-1$ in both versions. The sequential run time
was computed as the mean of three consecutive executions on a single
core---the single run took two hours 43 minutes. The difference between
PArrows/Eden and Eden almost always lies inside the error bar of
the measurement.

%As the PArrows version uses Eden in the backend, these numbers suggest that there is no real performance difference between using PArrows or Eden for this task as they trade blows in this benchmark. Additionally, PArrows with an Eden-based backend performing better than what it is based upon suggests that any difference in runtime between the two is more of an anomaly than a real difference.

\subsubsection{\jacobitest}

Continuing, the results of the \jacobitest (MPI only) in Fig.~\ref{fig:bench-jacobi-dist} are as follows:
The program does not seem to scale well beyond 64 threads with input $2^{3217}-1$. We once again compare the Eden version with the PArrrows version in Fig.~\ref{fig:bench-jacobi-dist}. We see similar behaviour to the MPI version of the \rmtest: The difference once again almost always remains within the bounds of the error bar of the measurement.

Because of the bad scaling behaviour for $2^{3217}-1$ beyond 64 threads, we also ran tests with input $2^{4253}-1$. Because of the long running time, we could do this only for the 128 and 256 threads. Therefore we do not show these results in Fig.~\ref{fig:bench-jacobi-dist} as we cannot properly compute a speedup without results for 1 thread. Nonetheless, for 128 threads, the benchmark took a mean of 9192.9\seconds, while with 256 threads, it only took a mean of 1649.1\seconds. This means that the 128 thread version ran more than 5.5 times slower than the 256 one, which suggests an IO limitation for this big input that is somewhat mitigated by adding more cores. However it still proves that PArrows continue to scale, even if not perfectly for this test program.
Comparing PArrows with Eden, for the larger input of $2^{4253}-1$, we see slightly bigger differences between Eden and PArrows: The differences for 128 and 256 core sare $-1005.18\seconds$ and $-94.33\seconds$ in favour of Eden, respectively. This maximum of $12.27\%$ slower PArrows runtime, however, was still in the (quite big) error bar of our measurements.

\newcommand{\speedupJacobiDist}[5]{
\speedupplot{Speedup of \jacobitest \enquote{#2} vs simulated \enquote{#5}}{PArrows #2, simulated PArrows #5}{256}{#3}{
% \addplot [mark=*,very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-parrows-11213-#2.csv};
\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#2.csv};
\addplot [mark=x,very thick,red] table [scatter, x="nCores", y="speedup"/, col sep=comma, mark=none,
smooth]{benchmarks/distributed-jacobi/bench-jacobi.bench.jacobi-parr-#1-#5.csv};
% \addplot table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
% smooth]{benchmarks/distributed-rm/bench-distributed.bench.skelrm-eden-#1-#2.csv};
}{#4}{\plotwidthDist}
}

\newcommand{\performanceJacobiDistDiff}[5]{
\performancediffplot{Overhead\\for \jacobitest \enquote{#2}}{(Eden $-$ PArrows) #2}{256}{#3}{
\addplot+[mark=*,very thick,error bars/.cd,
    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
smooth]{benchmarks/distributed-jacobi/#1-#2-diff.csv};
}{#4}{#5}{\plotwidthDist}
}

\begin{figure}
\centering
%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
%
{\speedupJacobiDist{3}{3217}{32}{272}{4253}\label{subfig:jacobi-dist-a}}%
%\hfill%
{\performanceJacobiDistDiff{3}{3217}{32}{5}{272}\label{subfig:jacobi-dist-b}}
\caption{Parallel performance of the \jacobitest on the Glasgow cluster
  consisting of 256 cores. Input was $2^{3217}-1$, we used 256
  tasks. The top plot shows relative speedup in a distributed memory setting compared to a simulated speedup for input $2^{4253}-1$. The
  closer to linear speedup the better. Time
  (and hence speedup) measurements for PArrows with Eden backend and
  Eden almost coincide. Hence, bottom plot shows
absolute time differences for this benchmark. The
higher the value, the better for PArrows\olcomment{CHECKME}.}
\label{fig:bench-jacobi-dist}
\end{figure}

\subsubsection{\torustest}

Next is the \torustest benchmark. The results of the comparison of vanilla Eden to our PArrows-based version can be found in Fig.~\ref{fig:torusBenchmark}. We see that the benchmark scales quite well with more cores until 64 cores. For $>=96$ cores, we still have considerable speedup, but with less slope. We also prove that the difference between the Eden and PArrows version are only marginal with PArrows only being a maximum of $1.7\%$ slower -- for 160 cores -- when outside of the error bar.  %The difference between PArrows and Eden is only significant for 16 and 64 cores where it ran 1.7\% and 2.7\% slower which corresponds to a real-time difference of 0.12s and 0.13s. For 256 cores PArrows performed 0.2\% slower which corresponds to 0.01s overhead.

\newcommand{\speedupTorusDist}[3]{
\speedupplot{Speedup of \torustest \enquote{#1}}{PArrows}{256}{#2}{
\addplot [mark=*,very thick,blue] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-#1.csv};
}{#3}{\plotwidthDist}
}

\newcommand{\performanceTorusDistDiff}[4]{
\performancediffplot{Overhead\\for \torustest \enquote{#1}}{(Eden $-$ PArrows)}{256}{#2}{
\addplot+[mark=*,very thick,error bars/.cd,
    y dir=both,y explicit] table [x="nCores", y="overhead", y error="stdDevForOverhead", col sep=comma, mark=dots,
smooth]{benchmarks/distributed-torus/#1-diff.csv};
}{#3}{#4}{\plotwidthDist}
}

\begin{figure}
\centering
%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
%
{\speedupTorusDist{4096}{32}{272}\label{subfig:speedupTorusDist}}%
%\hfill%
{\performanceTorusDistDiff{4096}{32}{0.5}{272}\label{subfig:performancetorusDistDiff}}
\caption{Parallel performance of \torustest on the Glasgow cluster
  consisting of 256 cores. Input was a matrix size of $4096$. The top plot shows absolute speedup in a distributed memory setting. The
  closer to linear speedup the better. Time
  (and hence speedup) measurements for PArrows with Eden backend and
  Eden almost coincide. Hence, bottom plot shows
absolute time differences for this benchmark. The
higher the value, the better for PArrows\olcomment{CHECKME}.}
\label{fig:torusBenchmark}
\end{figure}

\subsubsection{\sudokutest}

As the last benchmark in this paper we present the \sudokutest in Fig.~\ref{fig:sudokuSMBenchmark} running in a shared memory setting. Here we see all three SM backends performing similarly again like in the \rmtest SM benchmarks in Figs.~\ref{fig:sudokuSMBenchmark} and \ref{fig:sudokuSMBenchmark16000}. However, we notice that the GpH backend seems to choke on a bigger input (Fig.~\ref{fig:sudokuSMBenchmark16000}). This is due to the benchmark only using \ensuremath{\Varid{parMap}} instead of a chunking variant -- however we did not change that for simplicity's sake. This issue is reflected by debug output which shows that of 16000 sparks being created (one for each Sudoku) only 8365 were converted (executed) with the rest (7635) overflowing the runtime spark pool. Another remarkable finding is that the Eden backend seems to lag behind for $\leq$16 threads, but manages to pull ahead noticeably with all 32 threads of the system in use.

\newcommand{\performanceSudokuSM}[1]{
\performanceplot{Parallel run time of \sudokutest \enquote{#1}}{linear speedup, Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
\addplot [no markers,dotted,thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth] {benchmarks/sudoku-sm/bench-sudoku-sm.bench.fake-linear-sudoku-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
%\addplot+ [very thick] table [scatter, x="nCores", y="time", col sep=comma, mark=none,
%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
}{17}{\plotwidthSMP}
}

\newcommand{\speedupSudokuSM}[1]{
\speedupplot{Parallel speedup of \sudokutest \enquote{#1}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.#1.txt.csv};
\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.#1.txt.csv};
%\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
%smooth]{benchmarks/sudoku-sm/bench-sudoku-sm.bench.parmonad-sudoku-sudoku17.#1.txt.csv};
}{17}{\plotwidthSMP}
}

\begin{figure}
\centering
%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
%
{\speedupSudokuSM{1000}\label{subfig:speedupSudokuSM}}%
%\hfill%
{\performanceSudokuSM{1000}\label{subfig:performanceSudokuSM}}
\caption{Absolute speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
  different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
  cluster; it has 16 real cores and 32 threads. Input was a file of $1000$ Sudokus. The
  closer to linear speedup the better.}
\label{fig:sudokuSMBenchmark}
\end{figure}

\begin{figure}
\centering
%\performanceSkelRMDist{44497}{256}{32,64,128,256,512}{544}
%
{\speedupSudokuSM{16000}\label{subfig:speedupSudokuSM16000}}%
%\hfill%
{\performanceSudokuSM{16000}\label{subfig:performanceSudokuSM16000}}
\caption{Absolute speedup of \sudokutest on a multicore machine. We used the same PArrows-based implementation with
  different backends on the same hardware and the \ensuremath{\Varid{parMap}} version from the \ensuremath{\Conid{Par}} Monad examples. Measurements were performed on a single node of the Glasgow
  cluster; it has 16 real cores and 32 threads. Input was a file of $16000$ Sudokus. The
  closer to linear speedup the better. The GpH version shows signs of choking with too many sparks being created.}
\label{fig:sudokuSMBenchmark16000}
\end{figure}

### 2017-12-19 paper/content/conclusion.tex

--- 2017-10-25 paper/content/conclusion.tex
@@
 Arrows turn out to be a useful tool for composing parallel
-programs. We do not have to introduce new monadic types that wrap the
-computation. Instead, we use Arrows in the same manner one uses sequential pure functions. 
+programs.
+We have shown, that, in order to have a generic and extensible parallel Haskell, we do not have to restrict ourselves to a monadic interface.
+Instead, we use Arrows which are a more general concept than Monads, but still allow for powerful composition.
+Furthermore, we think Arrows are a better fit to parallelize pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way.
+We manage to retain this nice property of parallel Haskells like Eden or GpH (which use pure constructs) while still having a generic and composable interface.
 %
-This work features multiple parallel backends: the already available parallel Haskell flavours.
+The DSL natively allows for parallelization of monadic code via the Kleisli type and additionally allows to parallelize any Arrow type that has an instance for |ArrowChoice| (note that some skeletons require an additional |ArrowLoop| instance).
+%
 Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden.
@@
 More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} for this future extension.
-HdpH is a modern distributed Haskell that would benefit from our Arrows notation.
+HdpH is a modern distributed Haskell that would benefit from our Arrow notation.
 Further Future-aware versions of Arrow combinators can be defined.
@@
 More experiences with seamless porting of parallel PArrows-based programs across the backends are welcome.
-Of course, we are working ourselves on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
+Of course, we are ourselves working on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
 It would also be interesting to see a hybrid of PArrows and Accelerate \cite{McDonell:2015:TRC:2887747.2804313}.
-Ports of our approach to other languages like Frege or Java directly are in an early development stage.
+Ports of our approach to other languages like Frege, Eta or Java directly are in an early development stage.

### 2018-01-03 paper/separate_plots/main.tex
wip benchmark plots

--- 2017-11-29 paper/separate_plots/main.tex
@@
 
+% hack, but we do not need 100\% perfect rendering
+\newcommand{\Conid}[1]{#1}
+
 \begin{document}
@@
 \setlength{\plotwidthDist}{0.6\textwidth}
+\newlength{\plotwidthSMP}
+\setlength{\plotwidthSMP}{0.39\textwidth}
 
@@
 
+\newcommand{\speedupplot}[7]{
+\begin{tikzpicture}
+\begin{axis}[title={#1},
+title style={align=center},
+scale only axis, width=#7,
+xlabel=Threads,
+%xtick=data,
+%ytick=data,
+xtick distance=#4,
+ytick distance=#4,
+ylabel=Speedup,
+ylabel near ticks,
+grid=major,
+legend entries={linear, #2},
+legend style={at={(0.01,0.99)},anchor=north west},
+max space between ticks=50pt,
+grid style={line width=.1pt, draw=gray!10},
+major grid style={line width=.2pt,draw=gray!50},
+ymin=-1,
+xmin=-1,
+ymax=#6,
+xmax=#6]
+\addplot [domain=0:#3, no markers,dotted,thick]{x};
+#5
+\end{axis}
+\end{tikzpicture}
+}
+
 \newcommand{\rmtest}{Rabin--Miller test\xspace}
@@
 
-\meanOverheadPlot{Mean Overhead for distributed \rmtest 44497 256}{0.02}{32}{272}{../content/benchmarks/distributed-rm/44497-256-diff.csv}
+\newcommand{\benchmarkDir}{../content/benchmarks}
+
+\subsection{Distributed}
+
+\subsubsection{\rmtest}
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \rmtest 44497 256}{0.02}{32}{272}{\benchmarkDir/distributed-rm/44497-256-diff.csv}
+
+\subsubsection{\jacobitest}
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 4253}{0.5}{32}{272}{\benchmarkDir/distributed-jacobi/3-4253-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for distributed \jacobitest 4253}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
+\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.5}{32}{272}{\benchmarkDir/distributed-jacobi/3-4253-diff.csv}
+
+\subsubsection{\torustest}
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \torustest 4096}{0.5}{32}{272}{\benchmarkDir/distributed-torus/4096-diff.csv}
+
+\subsection{Shared-Memory}
+
+\subsubsection{\rmtest}
+
+\paragraph{Speedup}~\\
+
+\speedupplot{Speedup of \rmtest \enquote{32}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-11213-32.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-11213-32.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-par-11213-32.csv};
+}{17}{\plotwidthSMP}
+
+\speedupplot{Speedup of \rmtest \enquote{64}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-eden-cp-11213-64.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-mult-11213-64.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-par-11213-64.csv};
+}{17}{\plotwidthSMP}
+
+\paragraph{Overhead}~\\
+
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-32-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for distributed \jacobitest 3217}{0.5}{32}{272}{../content/benchmarks/distributed-jacobi/3-4253-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-64-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for distributed \torustest 4096}{0.5}{32}{272}{../content/benchmarks/distributed-torus/4096-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs GpH}{0.5}{4}{18}{\benchmarkDir/sm-rm/mult-11213-32-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs GpH}{0.5}{4}{18}{\benchmarkDir/sm-rm/mult-11213-32-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Par Monad}{0.5}{4}{18}{\benchmarkDir/sm-rm/par-11213-32-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Par Monad}{0.5}{4}{18}{\benchmarkDir/sm-rm/par-11213-32-diff.csv}
+
+\subsubsection{\sudokutest}
+
+\paragraph{Speedup}~\\
+
+\speedupplot{Parallel speedup of \sudokutest \enquote{1000}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-eden-sudoku17.1000.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-mult-sudoku17.1000.txt.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.1000.txt.csv};
+}{17}{\plotwidthSMP}
+
+\paragraph{Overhead}~\\
+
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/sudoku-sm/eden-cp-1000-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs GpH}{0.5}{4}{18}{\benchmarkDir/sudoku-sm/mult-1000-diff.csv}
+\\
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Par Monad}{0.5}{4}{18}{\benchmarkDir/sudoku-sm/par-1000-diff.csv}
+
+\subsubsection{\torustest}
+
+\paragraph{Speedup}~\\
+
+\speedupplot{Parallel speedup of \sudokutest \enquote{512}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/torus-sm/bench-torus-sm.bench.torus-matrix-parrows-eden-sm-512.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/torus-sm/bench-torus-sm.bench.torus-matrix-parrows-multicore-512.csv};
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/torus-sm/bench-torus-sm.bench.torus-matrix-parrows-parmonad-512.csv};
+}{17}{\plotwidthSMP}
+
+\paragraph{Overhead}~\\
 

### 2018-01-03 paper/separate_plots/main.tex
all relevant benchmarks added to separate plots pdf file, however scaling has still to be improved

@@
 
-\newcommand{\speedupplot}[7]{
+\newcommand{\speedupplot}[8]{
 \begin{tikzpicture}
@@
 xmin=-1,
-ymax=#6,
+ymax=#8,
 xmax=#6]
@@
 
+\paragraph{Overhead}~\\
+
 \meanOverheadPlot{Mean Overhead for distributed \\ \rmtest 44497 256}{0.02}{32}{272}{\benchmarkDir/distributed-rm/44497-256-diff.csv}
+
+\paragraph{Speedup}~\\
+
+\speedupplot{Speedup of distributed \rmtest \enquote{44497 256}}{PArrows Eden}{256}{32}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/distributed-rm/bench-distributed.bench.skelrm-parrows-44497-256.csv};
+}{272}{\plotwidthDist}{272}
 
@@
 
+\paragraph{Overhead}~\\
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.5}{32}{272}{\benchmarkDir/distributed-jacobi/3-4253-diff.csv}
+\\
 \meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 4253}{0.5}{32}{272}{\benchmarkDir/distributed-jacobi/3-4253-diff.csv}
-\\
-\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.5}{32}{272}{\benchmarkDir/distributed-jacobi/3-4253-diff.csv}
+
+\paragraph{Speedup}~\\
+
+\speedupplot{Speedup of distributed \jacobitest \enquote{3217}}{PArrows Eden}{256}{32}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/distributed-jacobi/bench-jacobi.bench.jacobi-parr-3-4253.csv};
+}{272}{\plotwidthDist}{272}
+
+\speedupplot{Speedup of distributed \jacobitest \enquote{4253}}{PArrows Eden}{256}{32}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/distributed-jacobi/bench-jacobi.bench.jacobi-parr-3-4253.csv};
+}{272}{\plotwidthDist}{272}
 
@@
 \meanOverheadPlot{Mean Overhead for distributed \\ \torustest 4096}{0.5}{32}{272}{\benchmarkDir/distributed-torus/4096-diff.csv}
+
+\paragraph{Speedup}~\\
+
+\speedupplot{Speedup of distributed \torustest \enquote{4096}}{PArrows Eden}{256}{32}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-eden-4096.csv};
+}{272}{\plotwidthDist}{160}
 
@@
 
-\speedupplot{Speedup of \rmtest \enquote{32}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
+\speedupplot{Speedup of \rmtest \enquote{11213 32}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-par-11213-32.csv};
-}{17}{\plotwidthSMP}
+}{17}{\plotwidthSMP}{17}
 
-\speedupplot{Speedup of \rmtest \enquote{64}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
+\speedupplot{Speedup of \rmtest \enquote{11213 64}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
@@
 smooth]{\benchmarkDir/sm-rm/bench-sm-rm.bench.skelrm-parr-par-11213-64.csv};
-}{17}{\plotwidthSMP}
+}{17}{\plotwidthSMP}{17}
 
@@
 smooth]{\benchmarkDir/sudoku-sm/bench-sudoku-sm.bench.parrows-sudoku-parmap-par-sudoku17.1000.txt.csv};
-}{17}{\plotwidthSMP}
+}{17}{\plotwidthSMP}{17}
 
@@
 smooth]{\benchmarkDir/torus-sm/bench-torus-sm.bench.torus-matrix-parrows-parmonad-512.csv};
-}{17}{\plotwidthSMP}
+}{17}{\plotwidthSMP}{17}
 
@@
 
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \torustest 512 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/torus-sm/eden-cp-512-diff.csv}
+
 \end{document}

### 2018-01-03 paper/separate_plots/main.tex
scaling done

@@
 
-\meanOverheadPlot{Mean Overhead for distributed \\ \rmtest 44497 256}{0.02}{32}{272}{\benchmarkDir/distributed-rm/44497-256-diff.csv}
+\meanOverheadPlot{Mean Overhead for distributed \\ \rmtest 44497 256}{0.04}{32}{272}{\benchmarkDir/distributed-rm/44497-256-diff.csv}
 
@@
 
-\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.5}{32}{272}{\benchmarkDir/distributed-jacobi/3-4253-diff.csv}
-\\
-\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 4253}{0.5}{32}{272}{\benchmarkDir/distributed-jacobi/3-4253-diff.csv}
+\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.04}{32}{272}{\benchmarkDir/distributed-jacobi/3-3217-diff.csv}
 
@@
 \addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{\benchmarkDir/distributed-jacobi/bench-jacobi.bench.jacobi-parr-3-4253.csv};
-}{272}{\plotwidthDist}{272}
-
-\speedupplot{Speedup of distributed \jacobitest \enquote{4253}}{PArrows Eden}{256}{32}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{\benchmarkDir/distributed-jacobi/bench-jacobi.bench.jacobi-parr-3-4253.csv};
-}{272}{\plotwidthDist}{272}
+smooth]{\benchmarkDir/distributed-jacobi/bench-jacobi.bench.jacobi-parr-3-3217.csv};
+}{272}{\plotwidthDist}{32}
 
@@
 
-\meanOverheadPlot{Mean Overhead for distributed \\ \torustest 4096}{0.5}{32}{272}{\benchmarkDir/distributed-torus/4096-diff.csv}
+\meanOverheadPlot{Mean Overhead for distributed \\ \torustest 4096}{0.04}{32}{272}{\benchmarkDir/distributed-torus/4096-diff.csv}
 
@@
 
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-32-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Eden CP}{0.04}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-32-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-64-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Eden CP}{0.03}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-64-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs GpH}{0.5}{4}{18}{\benchmarkDir/sm-rm/mult-11213-32-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs GpH}{0.2}{4}{18}{\benchmarkDir/sm-rm/mult-11213-32-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs GpH}{0.5}{4}{18}{\benchmarkDir/sm-rm/mult-11213-32-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs GpH}{0.05}{4}{18}{\benchmarkDir/sm-rm/mult-11213-64-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Par Monad}{0.5}{4}{18}{\benchmarkDir/sm-rm/par-11213-32-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Par Monad}{0.2}{4}{18}{\benchmarkDir/sm-rm/par-11213-32-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Par Monad}{0.5}{4}{18}{\benchmarkDir/sm-rm/par-11213-32-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Par Monad}{0.2}{4}{18}{\benchmarkDir/sm-rm/par-11213-64-diff.csv}
 
@@
 
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/sudoku-sm/eden-cp-1000-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Eden CP}{0.05}{4}{18}{\benchmarkDir/sudoku-sm/eden-cp-1000-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs GpH}{0.5}{4}{18}{\benchmarkDir/sudoku-sm/mult-1000-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs GpH}{0.05}{4}{18}{\benchmarkDir/sudoku-sm/mult-1000-diff.csv}
 \\
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Par Monad}{0.5}{4}{18}{\benchmarkDir/sudoku-sm/par-1000-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Par Monad}{0.05}{4}{18}{\benchmarkDir/sudoku-sm/par-1000-diff.csv}
 
@@
 
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \torustest 512 vs Eden CP}{0.5}{4}{18}{\benchmarkDir/torus-sm/eden-cp-512-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \torustest 512 vs Eden CP}{0.05}{4}{18}{\benchmarkDir/torus-sm/eden-cp-512-diff.csv}
 

### 2018-01-03 paper/separate_plots/main.tex
minor

@@
 
-\section{Plots}
-
 \newlength{\plotwidthDist}
@@
 minor tick num=9,
-ylabel=Overhead (\%),
+ylabel=Overhead,
 ylabel near ticks,
@@
 
-\subsection{Distributed}
+\section{Distributed Plots}
 
-\subsubsection{\rmtest}
+\subsection{\rmtest}
 
-\paragraph{Overhead}~\\
+\subsubsection{Overhead}~\\
 
@@
 
-\paragraph{Speedup}~\\
+\subsubsection{Speedup}~\\
 
@@
 
-\subsubsection{\jacobitest}
+\subsection{\jacobitest}
 
-\paragraph{Overhead}~\\
+\subsubsection{Overhead}~\\
 
@@
 
-\subsubsection{\torustest}
+\subsection{\torustest}
 
@@
 
-\subsection{Shared-Memory}
+\section{Shared-Memory Plots}
 
-\subsubsection{\rmtest}
+\subsection{\rmtest}
 
-\paragraph{Speedup}~\\
+\subsubsection{Speedup}~\\
 
@@
 
-\paragraph{Overhead}~\\
+\subsubsection{Overhead}~\\
 
@@
 
-\subsubsection{\sudokutest}
+\subsection{\sudokutest}
 
-\paragraph{Speedup}~\\
+\subsubsection{Speedup}~\\
 
@@
 
-\subsubsection{\torustest}
+\subsection{\torustest}
 
-\paragraph{Speedup}~\\
+\subsubsection{Speedup}~\\
 
@@
 
-\paragraph{Overhead}~\\
+\subsubsection{Overhead}~\\
 
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \torustest 512 vs Eden CP}{0.05}{4}{18}{\benchmarkDir/torus-sm/eden-cp-512-diff.csv}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \torustest 512 vs Eden CP}{0.08}{4}{18}{\benchmarkDir/torus-sm/eden-cp-512-diff.csv}
 

### 2018-01-04 paper/content/benchmarks.tex
benchmark chapter

@@
 
-\newcommand{\performanceplot}[7]{
-\begin{tikzpicture}
-\begin{axis}[title={#1},
-title style={align=center},
-scale only axis, width=#7,
-xlabel=Threads,
-%xtick=data,
-xtick distance=#4,
-ylabel=Time (s),
-ylabel near ticks,
-minor tick num=2,
-grid=major,
-legend entries={#2},
-legend style={at={(0.99,0.99)},anchor=north east},
-max space between ticks=50pt,
-grid style={line width=.1pt, draw=gray!10},
-major grid style={line width=.2pt,draw=gray!50},
-xmin=-1,
-xmax=#6]
-#5
-\end{axis}
-\end{tikzpicture}
-}
+\newcommand{\benchmarkDir}{benchmarks}
 
-\newcommand{\performancediffplot}[8]{
-\begin{tikzpicture}
-\begin{axis}[title={#1},
-title style={align=center},
-scale only axis, width=#8,
-xlabel=Threads,
-%xtick=data,
-ytick distance=#6,
-xtick distance=#4,
-minor tick num=9,
-ylabel=Absolute time difference (s),
-ylabel near ticks,
-grid=both,
-legend entries={#2},
-legend style={at={(0.99,0.99)},anchor=north east},
-max space between ticks=50pt,
-grid style={line width=.1pt, draw=gray!10},
-major grid style={line width=.2pt,draw=gray!50},
-xmin=-1,
-xmax=#7]
-#5
-\end{axis}
-\end{tikzpicture}
-}
-
-\newcommand{\speedupplot}[7]{
+\newcommand{\speedupplot}[8]{
 \begin{tikzpicture}
@@
 xmin=-1,
-ymax=#6,
+ymax=#8,
 xmax=#6]
 \addplot [domain=0:#3, no markers,dotted,thick]{x};
-#5
-\end{axis}
-\end{tikzpicture}
-}
-
-\newcommand{\speedupdiffplot}[7]{
-\begin{tikzpicture}
-\begin{axis}[title={#1},
-title style={align=center},
-scale only axis, width=#7,
-xlabel=Threads,
-%xtick=data,
-xtick distance=#4,
-ytick distance=0.5,
-ylabel=Absolute speedup difference,
-ylabel near ticks,
-grid=major,
-legend entries={#2},
-legend style={at={(0.99,0.99)},anchor=north east},
-max space between ticks=50pt,
-grid style={line width=.1pt, draw=gray!10},
-major grid style={line width=.2pt,draw=gray!50},
-xmin=-1,
-xmax=#6]
 #5
@@
 
-\input{bestAndWorstBenchmarks}
-
 \subsubsection{What parallel Haskells run where}
@@
 \mbcomment{fix this when benchmarks are redone}
-Hence, the performance measurements in Figs.~\ref{fig:bench-rm-sm}--\ref{fig:sudokuSMBenchmark} and Table~\ref{tab:meanOverhead} use only real cores as the purpose of this paper is to show the performance of PArrows and not to investigate parallel performance with hyper-threading in use.
+With this in mind, the performance measurements in the Section~\ref{sec:benchmarkResults} use only real cores as the purpose of this paper is to show the performance of PArrows and not to investigate parallel performance with hyper-threading in more detail.
 
@@
 
-\subsection{Benchmark results}
+\subsection{Benchmark results}\label{sec:benchmarkResults}
 
@@
 \paragraph{Speedup}
+The \rmtest benchmark showed almost linear speedup for both 32 and 64 tasks, the performance is slightly better in the latter case: 13.7 at 16 cores for input $2^{11213}-1$ and 64 tasks in the best case scenario with Eden CP.
+The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (Par Monad), 8.14 (Eden CP) for 16 cores and 1000 Sudokus.
+In contrast to Rabin–Miller, here the |GpH| Monad backend seems to be the best of all, while Rabin–Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
+Gentleman on shared memory has a plummeting speedup curve with GpH and |Par| Monad and logarithmically increasing speedup for Eden-based version.
+The latter reached a speedup of 6.56 at 16 cores.
+
+\mbcomment{Ich hab ein bisschen mehr Daten hinzugefügt für Sudoku}
 
 \paragraph{Overhead}
+We compared the overhead of our PArrows implementations to the original parallel Haskells.
+We basically implemented the test programs both in PArrows and in traditional style, benchmarked both and compare the mean overhead, i.e.
+the \textit{relative} difference between both benchmarks with same settings.
+The error margin of the benchmarks, supplied by criterion package, goes into the mean overhead computation.
+
+For the shared memory \rmtest benchmark, implemented with PArrows both with Eden CP and with |Par| Monad backends, \mbcomment{stimmt das so noch?
+Ich bin die Daten da nicht zu 100\% durch gegangen, aber die Aussage always finde ich etwas viel, almost always?} the zero \mbcomment{almost} always lay in the error margin of the mean overhead.
+This means that though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the benchmarks and the difference might not be existent.
+In the following, we call the contrary case “significant”.
+In other words, we notice no significant slow-down through PArrows here.
+For GpH backend there was only a small significant overhead of $8\%$ at 16 cores with 32 tasks and input value $2^{11213}-1$.
+For Sudoku the situation is slightly different.
+There is a minimal ($1.4\%$) significant speed improvement with PArrows Eden CP version when compared with the base Eden CP benchmark.
+However, with increasing number of cores the error margin reaches zero again.
+The |Par| Monad shows a similar development.
+The GpH version shows a small significant speed \textit{improvement} from PArrows of $4.2\%$.
+The Gentleman multiplication with Eden CP shows a minor significant overhead of $2.6\%$ at 8 cores and an insignificant improvement at 16 cores.
+
+We observe a low (if significant at all) overhead, induced by PArrows in the shared memory setting.
 
@@
 \paragraph{Speedup}
+The speedup of distributed memory Rabin–Miller benchmark with PArrows and distributed Eden backend showed an almost linear speedup sans some issues in the middle because of unfortunate task distribution.
+As seen in Fig.~\ref{fig:rabinMillerDistSpeedup}, we reached a speedup of 213.4 with PArrrows at 256 cores (vs.
+207.7 with pure Eden).
+We managed to measure the speedup of Jacobi sum test for sufficient large inputs only in a massively distributed setting, because of memory limitations.
+We improved there from 9193 seconds (128 cores) to 1649 seconds (256 cores) in our PArrows version, so we only include a graph for a run $2^{3217}-1$ in the supplementary material.
+This speedup at about 11 for both PArrows and Eden for more than 64 cores.
+There is apparently not enough work for that many cores.
+The Gentleman test with input 4096 had an almost linear speedup first, then plummeted between 128 and 224 cores, and recovered at 256 cores with speedup of 129.
+
+\begin{figure}[ht]
+	\centering
+	\speedupplot{Speedup of distributed \rmtest \enquote{44497 256}}{PArrows Eden}{256}{32}{
+	\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+	smooth]{\benchmarkDir/distributed-rm/bench-distributed.bench.skelrm-parrows-44497-256.csv};
+	}{272}{\plotwidthDist}{272}
+	\caption[Speedup distributed Rabin Miller]{Speedup of the distributed \rmtest benchmark with the Eden backend}
+	\label{fig:rabinMillerDistSpeedup}
+\end{figure}
 
 \paragraph{Overhead}
+We use our above mean overhead quality measure and the notion of significance also for distributed memory benchmarks.
+The mean overhead of Rabin–Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favor of PArrows), but these values are not significant.
+A sole significant (by a very low margin) overhead is $0.35\%$ at 64 cores.
+We measured the mean overhead for Jacobi benchmark for an input of $2^{3217}-1$ for up to 256 cores.
+We reached a maximum overhead of $0.32\%$ at 256 cores.
+In all cases it was not significant, the zero was inside the error margin.
+Mean overhead for distributed Gentleman multiplication was also low.
+Significant values include $1.23\%$ at 64 cores, $6.4\%$ at 128 cores, and $2.4\%$ at 256 cores.
+It took PArrows 64.2 seconds at 256 cores to complete the benchmark.
+
+Similar to the shared memory setting, PArrows only imply a very low penalty with distributed memory that is outweighed by their convenience and usefulness to the user.
 
@@
 
-In the shared memory setting we naturally find that the backends perform differently.
-Furthermore, in our distributed memory tests on the full grid the difference between PArrows and Eden was almost always below the error margin if PArrows performed worse.
-Even the biggest difference of 12.27\% in the \jacobitest for input $2^{4253}-1$ and 128 cores remained in line with these findings.
-The only exception to this was the \torustest  with only a $1.7\%$ difference in favour of Eden which is only a marginal slowdown in our opinion.
+\input{bestAndWorstBenchmarks}
+
+\mbcomment{Schau hier vll nochmal drüber, das ist etwas unsauber formuliert evtl.?}
+Summing up, PArrows performed with little to no overhead.
+Tables \ref{tab:meanOverheadSharedMemory} and \ref{tab:meanOverHeadDistributedMemory} clarify this once more: The PArrow enabled versions trade blows with the vanilla counterparts when comparing the mean of the mean overheads.
+If we combine these findings with the convenience of our DSL, PArrows seem like a great improvement over the respective backends with.

### 2018-01-04 paper/content/bestAndWorstBenchmarks.tex
scaling done
benchmark chapter

@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 16 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{vs.}             & \thead{mean of \\ mean \\overheads} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 16 cores} \\ \hline \\
 	\multirow{3}{*}{\sudokutest 1000}
 	\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{1}{*}{\torustest 512}
+	\csvreader[head to column names]{benchmarks/torus-sm/bestAndWorstTorusSM-512.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 	\\ \hline \\
@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{vs.}             & \thead{mean \\ overhead} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 256 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{vs.}             & \thead{mean of \\ mean \\overheads} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 256 cores} \\ \hline \\
 	\multirow{1}{*}{\torustest 4096}

### 2018-01-04 paper/content/benchmarks.tex
benchmarks chapter cnt

@@
 207.7 with pure Eden).
-We managed to measure the speedup of Jacobi sum test for sufficient large inputs only in a massively distributed setting, because of memory limitations.
+We managed to measure the speedup of Jacobi sum test for sufficient large inputs like $2^{4253}-1$ only in a massively distributed setting, because of memory limitations.
 We improved there from 9193 seconds (128 cores) to 1649 seconds (256 cores) in our PArrows version, so we only include a graph for a run $2^{3217}-1$ in the supplementary material.

### 2018-01-05 paper/content/abstract.tex
oleg patches

@@
 performance-wise.
-We finally conclude that Arrows turn out to be a useful tool for composing parallel programs and that our particular approach results in programs that are portable accross multiple backends.
+We finally conclude that Arrows turn out to be a useful tool for composing parallel programs and that our particular approach results in programs that are portable across multiple backends.
 \olcomment{Summarize conclusions}

### 2018-01-05 paper/content/benchmarks.tex
oleg patches

@@
 
-Apart from Eden, all benchmarks and libaries were compiled with Stack's\footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0.
+Apart from Eden, all benchmarks and libraries were compiled with Stack's\footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0.
 Stack itself was used in version 1.3.2.
@@
 
-Furtermore, all benchmarks were done with help of the bench\footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2)\footnote{see \url{https://hackage.haskell.org/package/criterion-1.1.1.0}} internally.
+Furthermore, all benchmarks were done with help of the bench\footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2)\footnote{see \url{https://hackage.haskell.org/package/criterion-1.1.1.0}} internally.
 All runtime data (mean runtime, max stddev, etc.) was collected with this tool if not mentioned otherwise.
@@
 In preliminary tests, the PArrows version of \rmtest on a single node of the Glasgow cluster
-showed almost linear speedup on up to 16 shared-memory cores (Fig.~\ref{fig:bench-rm-sm}). The speedup
+showed almost linear speedup on up to 16 shared-memory cores (as supplementary materials show). The speedup
 of 64-task PArrows/Eden at 16 real cores version was 13.65 giving a parallel
@@
 for 64 tasks, we obtained a speedup of 16.12 with PArrows/Eden at 32
-hyper-threaded cores and only 13.55 with PArrows/GpH. The parallel eficiencies were 50.4\% and 42.3\%, respectively. The Eden
+hyper-threaded cores and only 13.55 with PArrows/GpH. The parallel efficiencies were 50.4\% and 42.3\%, respectively. The Eden
 version used here was Eden~CP, the \enquote{share nothing} SMP build.
@@
 In the following paragraphs we will go into detail on how well PArrows perform when compared to versions of the benchmark programs implemented with the original parallel Haskells that were used in the backends.
-We start by comparing Speedups and Overheads of for the shared memory backends and conclude with OpenMPI variants of the Eden-enabled backend as a representative of a distributed memory backend.
+We start with the definition of an Overhead to compare both PArrows-enabled and standard benchmark implementations.
+We continue comparing speedups and overheads for the shared memory backends and then study OpenMPI variants of the Eden-enabled backend as a representative of a distributed memory backend.
+We plot all speedup curves and all overhead values in the supplementary materials.
+
+\subsubsection{Defining overhead}
+
+We compared the overhead of our PArrows implementations to the original parallel Haskells.
+We basically implemented the test programs both in PArrows and in traditional style, benchmarked both and compare the mean overhead, i.e.
+the \textit{relative} difference between both benchmarks with same settings.
+The error margin of the time measurements, supplied by criterion package, is computed together with the mean overhead computation.
+
+Quite often the zero value lies in the error margin of the mean overhead.
+This means that though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the benchmarks and the difference might not be existent.
+We are mostly interested in the cases where above issue does not persist, we call them \emph{significant}.
+We often denote the error margin with $\pm$ after the mean overhead value.
 
@@
 
-\mbcomment{Ich hab ein bisschen mehr Daten hinzugefügt für Sudoku}
+\paragraph{Overhead}
 
-\paragraph{Overhead}
-We compared the overhead of our PArrows implementations to the original parallel Haskells.
-We basically implemented the test programs both in PArrows and in traditional style, benchmarked both and compare the mean overhead, i.e.
-the \textit{relative} difference between both benchmarks with same settings.
-The error margin of the benchmarks, supplied by criterion package, goes into the mean overhead computation.
+For the shared memory \rmtest benchmark, implemented with PArrows using Eden CP, GpH, and |Par| Monad backends, the overhead values are within single percents range, but also negative overhead (\ie PArrows are better) and larger error margins happen.
+To give a few examples, the overhead for Eden~CP with input value $2^{11213}-1$, 32 tasks, and 16 cores is $1.5\%$, but the error margin is around $5.2\%$!
+Same backend in the same setting with 64 tasks reaches $-0.2\%$ overhead, PArrows apparently fare better than Eden -- but the error margin of $1.9\%$ disallows this interpretation.
+We focus now on significant overhead values.
+To name a few: $0.41\%\; \pm 7\cdot 10^{-2}\%$ for Eden~CP and 64 tasks at 4 cores, $4.7\% \; \pm 0.72\%$ for GpH, 32 tasks, 8 cores, $0.34\% \; \pm 0.31\%$ for |Par| Monad at 4 cores with 64 tasks.
+The worst significant overhead was  GpH backend with $8\% \; \pm 6.9\%$ at 16 cores with 32 tasks and input value $2^{11213}-1$.
+In other words, we notice no major slow-down through PArrows here.
 
-For the shared memory \rmtest benchmark, implemented with PArrows both with Eden CP and with |Par| Monad backends, \mbcomment{stimmt das so noch?
-Ich bin die Daten da nicht zu 100\% durch gegangen, aber die Aussage always finde ich etwas viel, almost always?} the zero \mbcomment{almost} always lay in the error margin of the mean overhead.
-This means that though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the benchmarks and the difference might not be existent.
-In the following, we call the contrary case “significant”.
-In other words, we notice no significant slow-down through PArrows here.
-For GpH backend there was only a small significant overhead of $8\%$ at 16 cores with 32 tasks and input value $2^{11213}-1$.
 For Sudoku the situation is slightly different.
-There is a minimal ($1.4\%$) significant speed improvement with PArrows Eden CP version when compared with the base Eden CP benchmark.
-However, with increasing number of cores the error margin reaches zero again.
-The |Par| Monad shows a similar development.
-The GpH version shows a small significant speed \textit{improvement} from PArrows of $4.2\%$.
-The Gentleman multiplication with Eden CP shows a minor significant overhead of $2.6\%$ at 8 cores and an insignificant improvement at 16 cores.
+There is a minimal significant ($-1.4\% \; \pm 1.2\%$ at 8 cores) speed improvement with PArrows Eden~CP version when compared with the base Eden CP benchmark.
+However, with increasing number of cores the error margin reaches zero again: $-1.6\% \; \pm 5.0\%$ at 16 cores.
+The |Par| Monad shows a similar development, \eg with $-1.95\% \; \pm 0.64\%$ at 8 cores.
+The GpH version shows both a significant speed \textit{improvement} from PArrows of $-4.2\% \; \pm 0.26\%$ and a minor overhead of $0.87\% \; \pm 0.70\%$.
 
-We observe a low (if significant at all) overhead, induced by PArrows in the shared memory setting.
+The Gentleman multiplication with Eden CP shows a minor significant overhead of $2.6\% \; \pm 1.0\%$ at 8 cores and an insignificant improvement at 16 cores.
+Summarising, we observe a low (if significant at all) overhead, induced by PArrows in the shared memory setting.
 
@@
 We managed to measure the speedup of Jacobi sum test for sufficient large inputs like $2^{4253}-1$ only in a massively distributed setting, because of memory limitations.
-We improved there from 9193 seconds (128 cores) to 1649 seconds (256 cores) in our PArrows version, so we only include a graph for a run $2^{3217}-1$ in the supplementary material.
-This speedup at about 11 for both PArrows and Eden for more than 64 cores.
+We improved there from 9193 seconds (128 cores) to 1649 seconds (256 cores) in our PArrows version.
+A~scaled-down version with input $2^{3217}-1$ stagnates the speedup at about 11 for both PArrows and Eden for more than 64 cores.
 There is apparently not enough work for that many cores.
@@
 	}{272}{\plotwidthDist}{272}
-	\caption[Speedup distributed Rabin Miller]{Speedup of the distributed \rmtest benchmark with the Eden backend}
+	\caption[Speedup distributed Rabin--Miller]{Speedup of the distributed \rmtest benchmark using PArrows with Eden backend.}
 	\label{fig:rabinMillerDistSpeedup}
@@
 \paragraph{Overhead}
-We use our above mean overhead quality measure and the notion of significance also for distributed memory benchmarks.
-The mean overhead of Rabin–Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favor of PArrows), but these values are not significant.
-A sole significant (by a very low margin) overhead is $0.35\%$ at 64 cores.
+We use our mean overhead quality measure and the notion of significance also for distributed memory benchmarks.
+The mean overhead of Rabin–Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favour of PArrows), but these values are not significant with error margins $\pm 0.8\%$ and $\pm 2.9\%$ correspondingly.
+A sole significant (by a very low margin) overhead is $0.35\% \; \pm 0.33\%$ at 64 cores.
 We measured the mean overhead for Jacobi benchmark for an input of $2^{3217}-1$ for up to 256 cores.
-We reached a maximum overhead of $0.32\%$ at 256 cores.
-In all cases it was not significant, the zero was inside the error margin.
+We reach the flattering value $-3.8\% \; \pm 0.93\%$ at 16 cores in favour of PArrows, it was the sole significant overhead value. The  value for 256 cores was $0.31\% \; \pm 0.39\%$.
 Mean overhead for distributed Gentleman multiplication was also low.
-Significant values include $1.23\%$ at 64 cores, $6.4\%$ at 128 cores, and $2.4\%$ at 256 cores.
+Significant values include $1.23\% \; \pm 1.20\%$ at 64 cores and $2.4\% \; \pm 0.97\%$ at 256 cores.
 It took PArrows 64.2 seconds at 256 cores to complete the benchmark.
 
-Similar to the shared memory setting, PArrows only imply a very low penalty with distributed memory that is outweighed by their convenience and usefulness to the user.
+Similar to the shared memory setting, PArrows only imply a very low penalty with distributed memory that lies in lower single-percent digits at most.
 
@@
 
-\mbcomment{Schau hier vll nochmal drüber, das ist etwas unsauber formuliert evtl.?}
-Summing up, PArrows performed with little to no overhead.
-Tables \ref{tab:meanOverheadSharedMemory} and \ref{tab:meanOverHeadDistributedMemory} clarify this once more: The PArrow enabled versions trade blows with the vanilla counterparts when comparing the mean of the mean overheads.
-If we combine these findings with the convenience of our DSL, PArrows seem like a great improvement over the respective backends with.
+PArrows performed in our benchmarks with little to no overhead.
+Tables~\ref{tab:meanOverheadSharedMemory} and \ref{tab:meanOverHeadDistributedMemory} clarify this once more: The PArrows-enabled versions trade blows with their vanilla counterparts when comparing the means over all cores of the mean overheads.
+If we combine these findings with the usability of our DSL,
+the minor overhead induced by PArrows is outweighed by their convenience and usefulness to the user.
+
+PArrows is an extendable formalism, they can be easily ported to further parallel Haskells while still maintaining interchangeability.
+It is straightforward to provide further implementations of algorithmic skeletons in PArrows.

### 2018-01-05 paper/content/bestAndWorstBenchmarks.tex
oleg patches

@@
 \centering
-\caption{Overhead in the shared memory benchmarks}
+\caption{Overhead in the shared memory benchmarks. Bold marks values
+  in favour of PArrows. Overhead is unitless, multiply by 100 to
+  obtain percent. Runtime is in seconds.}
 \label{tab:meanOverheadSharedMemory}
@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{vs.}             & \thead{mean of \\ mean \\overheads} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 16 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalized \\ stdDev} & \thead{Runtime for \\ 16 cores} \\ \hline \\
 	\multirow{3}{*}{\sudokutest 1000}
@@
 \centering
-\caption{Overhead in the distributed memory benchmarks}
+\caption{Overhead in the distributed memory benchmarks. Bold marks values
+  in favour of PArrows. Overhead is unitless, multiply by 100 to
+  obtain percent. Runtime is in seconds.}
+\olcomment{Vielleicht gleich durchmultiplizieren und \% angeben? Nicht
+zu viele leere Zeilen in der Tabelle?}
 \label{tab:meanOverHeadDistributedMemory}
@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{vs.}             & \thead{mean of \\ mean \\overheads} & \thead{maximum \\ normalized \\ stdDev} & \thead{runtime for \\ 256 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalized \\ stdDev} & \thead{Runtime for \\ 256 cores} \\ \hline \\
 	\multirow{1}{*}{\torustest 4096}

### 2018-01-05 paper/content/conclusion.tex
oleg patches

@@
 Furthermore, we think Arrows are a better fit to parallelize pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way.
-We manage to retain this nice property of parallel Haskells like Eden or GpH (which use pure constructs) while still having a generic and composable interface.
+We manage to retain this nice property of parallel Haskells such as Eden or GpH (which use pure constructs) while still having a generic and composable interface.
 %
@@
 It would also be interesting to see a hybrid of PArrows and Accelerate \cite{McDonell:2015:TRC:2887747.2804313}.
-Ports of our approach to other languages like Frege, Eta or Java directly are in an early development stage.
+Ports of our approach to other languages such as Frege, Eta, or Java directly are in an early development stage.

### 2018-01-05 paper/content/futures.tex
oleg patches

@@
-\section{Futures} \label{sec:futures}
+\section{Futures} \label{sec:futures}\label{futures}
 Consider a mock-up parallel Arrow combinator:
@@
 Black lines between processes represent communication.
-Colors: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
+Colours: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
 	\label{fig:withoutFutures}
-\olcomment{more practical and heavy-weight example! fft (I have the code)?}\\
-\mbcomment{Depends... Are the communications easy to read in such an example?}\\
-\mbcomment{Keep the description for the different colours, or link to
-  the EdenTV description in \ref{sec:edentv}}
-\olcomment{ok as is}
-\olcomment{use the fft example (when it works)?}
+% \olcomment{more practical and heavy-weight example! fft (I have the code)?}
+% \mbcomment{Depends... Are the communications easy to read in such an example?}
+% \mbcomment{Keep the description for the different colours, or link to
+%   the EdenTV description in \ref{sec:edentv}}
+% \olcomment{ok as is}
+% \olcomment{use the fft example (when it works)?}
 \end{figure}
@@
 We do this with the |Future| type class (Fig.~\ref{fig:future}).
-Note that we require a conf parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the |ArrowParallel| typeclass earlier.
+Note that we require a |conf| parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the |ArrowParallel| typeclass earlier.
 \begin{figure}[h]
@@
 \end{figure}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualization in Fig.~\ref{fig:withFutures}.
-One especially elegant aspect of this solution is that we can specify the type of |Future| to be used per backend with full interoperability between code using different backends, without even requring to know about the actual type used for communication.
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}.
+One especially elegant aspect of this solution is that we can specify the type of |Future| to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication.
 We only specify that there has to be a compatible Future and do not care about any specifics.

### 2018-01-05 paper/content/motivation.tex
better argumentation why Arrows are a better fit for parallel evaluation than monads or even pure variants
oleg patches

@@
 the parallel backends at will.
-The induced penalty was max.
-8.3 \% in our measurements \mbcomment{, and typically under X \%}.
+The induced penalty was in the single-digit percent range, with means over the varying cores configuration typically under 1\% overhead in our measurements (Section~\ref{sec:benchmarks}).
 Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
@@
 We structure this paper as follows:
-
 \begin{itemize}
 \item We introduce a parallel evaluation formalism using Arrows.
-One big advantage is that we do not have to introduce any new types (Sec.~\ref{sec:parallel-arrows}) in contrast to Monad solutions such as the |Par| Monad.
+One big advantage of our specific approach is that we do not have to introduce any new types (Sec.~\ref{sec:parallel-arrows}).
 This behaviour encourages better composability.
-\item We utilitze multiple backends -- currently a GpH, a |Par| Monad, and Eden.
+\item We utilise multiple backends -- currently a GpH, a |Par| Monad, and Eden.
 We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.

### 2018-01-05 paper/content/parallelHaskells.tex
oleg patches

@@
 There exist several parallel Haskells already.
-Among the most important are probably GpH  (based on |par| \& |pseq| \enquote{hints}) \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden (a parallel Haskell for distributed memory) \cite{eden,Loogen2012}, HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
+Among the most important are probably GpH  (based on |par| and |pseq| \enquote{hints}) \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden (a parallel Haskell for distributed memory) \cite{eden,Loogen2012}, HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
 
@@
 \label{sec:GpHIntro}
-GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{The Multicore implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| \& |pseq| hints), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
+GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{The Multicore implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (|par| and |pseq|), it ships with parallel evaluation strategies for several types which can be applied with |using :: a -> Strategy a -> a|, which is exactly what is required for an implementation of |parEvalN|.
 

### 2018-01-05 paper/content/relwork.tex
oleg patches

@@
 \paragraph{Arrows.}
-Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~|->|.
+Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, basically they are a generalised function arrow~|->|.
 \citet{Hughes2005} presents a tutorial on Arrows.
@@
 \paragraph{Other languages.}
-Although this work is centered on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined.
+Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined.
 Experiments with our approach in Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM) were quite successful.

### 2018-01-05 paper/content/topologySkeletons.tex
oleg patches

@@
 If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing.
-This materializes with the usage of |loop| of the |ArrowLoop| type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden.
+This materialises with the usage of |loop| of the |ArrowLoop| type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden.
 An investigation of where this difference comes from is beyond the scope of this work, we only provide a workaround for these types of skeletons as such skeletons probably are not of much importance outside of a distributed memory environment.

### 2018-01-05 paper/content/utilityFunctions.tex
oleg patches

--- 2017-12-14 paper/content/utilityFunctions.tex
@@
 Next, we give the definition of |evalN| which also helps us to define |map|, and |zipWith| on Arrows.
-The |evalN| combinator in Fig.~\ref{fig:evalN} converts a list of arrows |[arr a b]| into an arrow |arr [a] [b]|.
+The |evalN| combinator in Fig.~\ref{fig:evalN} converts a list of Arrows |[arr a b]| into an Arrow |arr [a] [b]|.
 
@@
 
-The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any arrow |arr a
-b| to an arrow |arr [a] [b]|. The original inspiration was from \cite{programming_with_arrows},
+The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any Arrow |arr a
+b| to an Arrow |arr [a] [b]|. The original inspiration was from \citet{programming_with_arrows},
 but the definition as then unified with |evalN|.
@@
 
-Finally, with the help of |mapArr| (Fig.~\ref{fig:mapArr}), we can define |zipWithArr|  (Fig.~\ref{fig:zipWithArr}) that lifts any arrow |arr (a, b) c| to an arrow |arr ([a], [b]) [c]|.
+Finally, with the help of |mapArr| (Fig.~\ref{fig:mapArr}), we can define |zipWithArr|  (Fig.~\ref{fig:zipWithArr}) that lifts any Arrow |arr (a, b) c| to an Arrow |arr ([a], [b]) [c]|.
 \begin{figure}[h]
@@
 \end{code}
-\caption{|zipWith| over arrows.}
+\caption{|zipWith| over Arrows.} %$
 \label{fig:zipWithArr}
@@
 These combinators make use of the |ArrowChoice| type class which provides the |pipepipepipe| combinator.
-It takes two arrows |arr a c| and |arr b c| and combines them into a new arrow |arr (Either a b) c| which pipes all |Left a|'s to the first arrow and all |Right b|'s to the second arrow:
+It takes two Arrows |arr a c| and |arr b c| and combines them into a new Arrow |arr (Either a b) c| which pipes all |Left a|'s to the first Arrow and all |Right b|'s to the second Arrow:
 % \begin{figure}[htb]
@@
 In Fig.~\ref{fig:profunctorArrow} we show how specific Profunctors can be turned into Arrows.
-Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| typeclass.
+This works because Arrows are strong Monads in the bicategory |Prof| of Profunctors as shown by \citet{Asada:2010:ASM:1863597.1863607}.
+Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| typeclass like presented in this paper.\footnote{For additional information on the typeclasses used, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}.}
 
@@
 \end{code}
-\caption{Profunctors as Arrows}
-\footnote{for additional information on the typeclasses see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}}
+\caption{Profunctors as Arrows.}
 \label{fig:profunctorArrow}
@@
 brevity, and redeem this here.
-
+%
+%
 %
@@
 We start by transforming the |(a, c)| input into a two-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|.
-Next, we feed this list into a parallel arrow running on two instances of |f +++ g| as described above.
+Next, we feed this list into a parallel arrow running on two instances of |f +++ g| as described in the paper.
 After the calculation is finished, we convert the resulting |[Either b d]| into |([b], [d])| with |arr partitionEithers|.

### 2018-01-05 paper/content/arrows.tex
wip
own remarks until page 17.

@@
 Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads.
-\citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator |(>>=) :: m a -> (a -> m b) -> m b| (with |m| being a Monad).
-In the example a parser of the type |Parser s a| with static meta information |s| and result |a| is shown to not be able to use the static information |s| without applying the monadic function |a -> m b|.
-With Arrows this is possible.
+\citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator |(>>=) :: m a -> (a -> m b) -> m b| (with |m| being a Monad)\footnote{In the example a parser of the type |Parser s a| with static meta information |s| and result |a| is shown to not be able to use the static information |s| without applying the monadic function |a -> m b|.
+With Arrows this is possible.}.
 
@@
 
-In this paper we will show that we can express parallel computations with Arrows and do not require Monads to do so.
-We also do not restrict the compatible Arrows to ones which have |ArrowApply| instances but instead only require instances for |ArrowChoice| and |ArrowLoop|.
+In this paper we will show that parallel computations can be expressed with this more general interface of Arrows without requiring Monads.
+We also do not restrict the compatible Arrows to ones which have |ArrowApply| instances but instead only require instances for |ArrowChoice| (for if-then-else constructs) and |ArrowLoop| (for looping).
 Because of this, we have a truly more general interface as compared to a monadic one.
 
-Also note that, while we could have based our DSL on Profunctors as well, we chose Arrows.
+Also note that, while we could have based our DSL on Profunctors as well, we chose Arrows \mbcomment{cite missing?}.
 for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.

### 2018-01-05 paper/content/basicSkeletons.tex
own remarks until page 17.

@@
 \label{sec:extending-interface}
-With the |ArrowParallel| type class in place and implemented, we can now implement some further basic parallel interface functions.
+With the |ArrowParallel| type class in place and implemented, we can now define some further basic parallel interface functions.
 These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
@@
 \end{figure}
-The function |parEvalN| is 100\% strict, which means that it fully evaluates all passed arrows.
+|parEvalN| fully traverses the list of passed Arrows as well as their inputs.
 Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
@@
 |parEvalNLazy| (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
-It works by first chunking the input from |[a]| to |[[a]]| with the given |ChunkSize| in |arr (chunksOf chunkSize)|.
-These chunks are then fed into a list |[arr [a] [b]]| of parallel arrows created by feeding chunks of the passed |ChunkSize| into the regular parEvalN by using |listApp|.
+It works by first chunking the input from |[a]| to |[[a]]| with the given |chunkSize| in |arr (chunksOf chunkSize)|.
+These chunks are then fed into a list |[arr [a] [b]]| of chunk-wise parallel Arrows with the help of our lazy and sequential |evalN|.
 The resulting |[[b]]| is lastly converted into |[b]| with |arr concat|.
@@
 \end{figure}
-We have only talked about the parallelisation of arrows of the same type until now.
+We have only talked about the parallelisation of arrows of the same set of input and output types until now.
 But sometimes we want to parallelize heterogeneous types as well.
 We can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class.
-The idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.
+Here, the general idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.
 

### 2018-01-05 paper/content/futures.tex
own remarks until page 17.

@@
 This can become a serious bottleneck
-for larger amount of data and number of processes \citep[as e.g.][showcases]{Berthold2009-fft}.
+for a larger amount of data and number of processes \citep[as e.g.][showcases]{Berthold2009-fft}.
 \begin{figure}[ht]
@@
 In our communication example we can use this |Future| concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
-Note that with our PArrows DSL we can also define default instances |Future fut a ()| for each backend similar to how |ArrowParallel arr a b ()| was defined in Section~\ref{sec:parallel-arrows}.
-Details can be found in Section~\ref{app:omitted}.
 \begin{figure}[tbh]
@@
 In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}.
-One especially elegant aspect of this solution is that we can specify the type of |Future| to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication.
-We only specify that there has to be a compatible Future and do not care about any specifics.
+One especially elegant aspect of the definition in Fig.~\ref{fig:future} is that we can specify the type of |Future| to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication.
+We only specify that there has to be a compatible Future and do not care about any specifics as can be seen in Fig.~\ref{fig:someCombinatorParallel}.
+Note that with our PArrows DSL we can also define default instances |Future fut a ()| for each backend similar to how |ArrowParallel arr a b ()| was defined in Section~\ref{sec:parallel-arrows}.
+Details can be found in Section~\ref{app:omitted}.
 \olcomment{Fig.

### 2018-01-05 paper/content/mapSkeletons.tex
own remarks until page 17.

@@
 Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations.
-While there are many possible skeletons to implement, we regard in detail here only some |map|-based and topological skeletons to demonstrate the power of PArrows.
+While there are many possible skeletons to implement, here we only regard in detail some |map|-based and topological skeletons to demonstrate the power of PArrows.
 %%% \FloatBarrier
@@
 We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to obtain an arrow |arr [a] [b]|.
-Just like |parEvalN|, |parMap| is 100\% strict.
+Just like |parEvalN|, |parMap| traverses all input Arrows as well as the inputs.
 Because of this, it has the same restrictions as |parEvalN| as compared to |parEvalNLazy|.
 So it makes sense to also have a |parMapStream| (Figs.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
-The code of these two skeletons is quite straightforward, we show it in Appendix.
+The code of these two skeletons is quite straightforward, we show it in Appendix \ref{app:omitted} in Figs.\ref{fig:parMap} and \ref{fig:parMapStream}.
 
@@
 \paragraph{Statically load-balancing parallel |map|.}
-Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper).
+Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we presented in this paper).
 This can be quite wasteful and a statically load-balancing |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful.
@@
 %\paragraph{Lazy statically load-balancing parallel map}
-Since a |farm|  is basically just |parMap| with a different work distribution, it is, again, 100\% strict.
-So we can define |farmChunk| (Figs.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|.
-It is basically the same definition as for |farm|, with |parEvalN| replaced with |parEvalNLazy|.
+Since a |farm|  is basically just |parMap| with a different work distribution, it has the same restrictions as |parEvalN| and |parMap|.
+We can, however, define |farmChunk| (Figs.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses |parEvalNLazy| instead of |parEvalN|.
+It is basically the same definition as for |farm|, but with |parEvalNLazy| instead of |parEvalN|.
 

### 2018-01-05 paper/content/motivation.tex
own remarks until page 17.

@@
 
-Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms including the expression of parallelism.
+Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms.
 Haskell, which we focus on in this paper, has  several mature implementations.
@@
 Instead of 
-introducing a new low-level parallel backend in order to implement our
+introducing a new low-level parallel backend to implement our
 Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL

### 2018-01-05 paper/content/parallelHaskells.tex
own remarks until page 17.

@@
 We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evaluates to normal form.
+Other strategies like e.g.
+evaluation to weak head normal form are available as well.
+It also allows for custom |Strategy| implementations to be used.
 %
@@
 \subsubsection{Eden}
-Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with a MPI and a PVM backends.\footnote{The projects homepage can be found at \url{http://www.mathematik.uni-marburg.de/~eden/}.
+Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with MPI and PVM as distributed backends.\footnote{The projects homepage can be found at \url{http://www.mathematik.uni-marburg.de/~eden/}.
 The Hackage page is at \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} It is targeted towards clusters, but also functions well in a shared-memory setting with a further simple backend.

### 2018-01-05 paper/content/parrows.tex
oleg patches
own remarks until page 17.

@@
 Arrows are a general interface to computation.
-Here we introduce special Arrows as general interface to \textit{parallel computations}.
-First, we present the interface and explain the reasonings behind it.
+Here we introduce special Arrows as a general interface to \textit{parallel computations}.
+First, we present the |ArrowParallel| type class and explain the reasoning behind it.
 Then, we discuss some implementations using existing parallel Haskells.
@@
 Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
-This combinator is similar to our utility function |evalN| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
+This combinator is similar to |evalN| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
 % \begin{figure}[h]
@@
 With this definition of |parEvalN|, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow |arr| and we want this to be an interface for different backends, we introduce a new type class |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+But as the implementation may differ depending on the actual type of the arrow |arr| - or even the input |a| and output |b| - and we want this to be an interface for different backends, we introduce a new type class |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
 % \begin{figure}[h]
@@
 normal form).
-For this reason we also introduce an additional |conf| parameter to the function.
-We also do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|.
+For this reason we introduce an additional |conf| parameter as we do not want |conf| to be a fixed type, as the configuration parameters can differ for different instances of |ArrowParallel|.
 % So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
@@
 \end{code}
+Note that by restricting the implementations of our backends to a specific |conf| type, we also get interoperability between backends for free.
+We can parallelize one part of a program using one backend, and parallelize the next with another one.
 % \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
@@
 \subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
-The GpH implementation of |ArrowParallel| is implemented in a straightforward manner in Fig.~\ref{fig:ArrowParallelMulticore}, but a bit different compared to the variant from Section \ref{sec:GpHIntro}, by using |evalN :: [arr a b] -> arr [a] [b]| (definition in Appendix~\ref{utilfns}, think |zipWith ($)| on Arrows) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
+The GpH implementation of |ArrowParallel| is implemented in a straightforward manner in Fig.~\ref{fig:ArrowParallelMulticore}, but a bit different compared to the variant from Section \ref{sec:GpHIntro}.
+We use |evalN :: [arr a b] -> arr [a] [b]| (definition in Appendix~\ref{utilfns}, think |zipWith ($)| on Arrows) combined with |withStrategy :: Strategy a -> a -> a| from GpH, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
 Our |Conf a| datatype simply wraps a |Strategy a|, but could be extended in future versions of our DSL.
@@
 \end{code}
-Now we can once again define our |ArrowParallel| instance as follows: First, we convert our Arrows |[arr a b]| with |evalN (map (>>> arr strat) fs)| into an Arrow |arr [a] [(Par (IVar b))]| that yields composable computations in the |Par| monad, where |evalN| is once again of the type |evalN :: [arr a b] -> arr [a] [b]|.
+Now we can once again define our |ArrowParallel| instance as follows: First, we convert our Arrows |[arr a b]| with |evalN (map (>>> arr strat) fs)| into an Arrow |arr [a] [(Par (IVar b))]| that yields composable computations in the |Par| monad.
 By combining the result of this Arrow with |arr sequenceA|, we get an Arrow |arr [a] (Par [IVar b])|.
@@
 \subsubsection{Default configuration instances}
-While the configurability in the instances of the |ArrowParallel| instances above is nice, users probably would like to have a proper default configuration for many parallel programs as well.
+While the configurability in the instances of the |ArrowParallel| instances above is nice, users probably would like to have proper default configurations for many parallel programs as well.
 These can also easily be defined as we can see by the example of the default implementation of |ArrowParallel| for the GpH backend:
@@
 
-The other backends have similarly structured implementations which we do not discuss here for brevity reasons.
-Note that by restricting the original implementations of our backends to their specific |Conf| type, we also get interoperability between backends for free.
-We can parallelize one part of a program using one backend, and parallelize the next with another one.
-We can however only have one instance of |ArrowParallel arr a b ()| present at a time.
+The other backends have similarly structured implementations which we do not discuss here for the sake of brevity.
+Also note that we can only have one instance of |ArrowParallel arr a b ()| present at a time, which should not be a problem, though.
 
 Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the dataflow between the arrows, now that we have seen that Arrows are capable of expressing parallelism.
-We do however explain new concepts with more details if required for easier understanding.
+We do however explain new concepts with more details if required for better understanding.

### 2018-01-05 paper/content/benchmarks.tex
martin remarks

@@
 
-In the following section, we describe the benchmarks of our parallel DSL and algorithmic skeletons conducted.
-We start by explaining the hardware and software stack and also elaborate on which benchmarks programs were used
-and also which parallel Haskell were used in which setting and why. Before we go into detail on the benchmarks we
-also address the issue of poor performance gains when using hyper-threading and our reasoning behind not including
-hyper-threaded cores in our benchmarks.
+In the following section, we describe the benchmarks we conducted of our parallel DSL and algorithmic skeletons.
+We start by explaining the hardware and software stack and also elaborate on which benchmarks programs and parallel Haskells were used in which setting and why.
+Before we go into detail on the benchmarks we
+also shortly address hyper-threading and why we do not use it in our benchmarks.
 Finally, we show that PArrows hold up in terms of performance when compared to the original parallel Haskells used as backends in this paper, starting with the shared-memory variants (GpH, |Par| Monad and Eden CP) and concluding with Eden as a distributed backend.
@@
 Eden. We used the MPI backend of Eden in a distributed memory
-setting. However, for shared memory Eden features a ``CP'' backend
+setting. However, for shared memory Eden features a \enquote{CP} backend
 that merely copies the memory blocks between distributed heaps. In
-this mode, Eden still operates in the ``nothing shared'' setting, but
-is adapted better to multicore machines. We label this version of Eden
-in the plots as ``Eden~CP''.
+this mode, Eden still operates in the \enquote{nothing shared} setting, but
+is adapted better to multicore machines. We call this version of Eden
+\enquote{Eden~CP}.
 
@@
 for 64 tasks, we obtained a speedup of 16.12 with PArrows/Eden at 32
-hyper-threaded cores and only 13.55 with PArrows/GpH. The parallel efficiencies were 50.4\% and 42.3\%, respectively. The Eden
-version used here was Eden~CP, the \enquote{share nothing} SMP build.
+hyper-threaded cores and only 13.55 with PArrows/GpH.
 
-In the distributed memory setting the same effect ensues. We obtain
-plummeting speedup of 124.31 at 512 hyper-threaded cores, whereas it was
-213.172 for 256 real cores. Apparently, hyper-threading in the Glasgow
-cluster fails to execute two parallel Haskell processes with full-fledged
-parallelism.
-
-\mbcomment{fix this when benchmarks are redone}
-With this in mind, the performance measurements in the Section~\ref{sec:benchmarkResults} use only real cores as the purpose of this paper is to show the performance of PArrows and not to investigate parallel performance with hyper-threading in more detail.
+While this shows that hyper-threading can be of benefit in scenarios similar to the ones presented in the benchmarks, we only use real cores for the performance measurements in Section~\ref{sec:benchmarkResults} as the purpose of this paper is to show the performance of PArrows and not to investigate parallel behaviour with hyper-threading.
 
@@
 We compared the overhead of our PArrows implementations to the original parallel Haskells.
-We basically implemented the test programs both in PArrows and in traditional style, benchmarked both and compare the mean overhead, i.e.
+We basically implemented the test programs both in PArrows and in traditional style, benchmarked both and will now compare the mean overhead, i.e.
 the \textit{relative} difference between both benchmarks with same settings.
@@
 Quite often the zero value lies in the error margin of the mean overhead.
-This means that though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the benchmarks and the difference might not be existent.
+This means that even though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the benchmarks and the difference might not be existent.
 We are mostly interested in the cases where above issue does not persist, we call them \emph{significant}.
@@
 The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (Par Monad), 8.14 (Eden CP) for 16 cores and 1000 Sudokus.
-In contrast to Rabin–Miller, here the |GpH| Monad backend seems to be the best of all, while Rabin–Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
-Gentleman on shared memory has a plummeting speedup curve with GpH and |Par| Monad and logarithmically increasing speedup for Eden-based version.
+In contrast to Rabin–Miller, here the |GpH| backend seems to be the best of all, while Rabin–Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
+Gentleman on shared memory has a plummeting speedup curve with GpH and |Par| Monad and logarithmically increasing speedup for the Eden-based version.
 The latter reached a speedup of 6.56 at 16 cores.
@@
 The GpH version shows both a significant speed \textit{improvement} from PArrows of $-4.2\% \; \pm 0.26\%$ and a minor overhead of $0.87\% \; \pm 0.70\%$.
+\mbcomment{Widerspricht sich das nicht?}
 

### 2018-01-05 paper/content/topologySkeletons.tex
martin remarks

@@
 These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell.
-We aim to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.
+We aim to showcase that we can express more sophisticated skeletons with parallel Arrows as well.
 
@@
 This materialises with the usage of |loop| of the |ArrowLoop| type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden.
-An investigation of where this difference comes from is beyond the scope of this work, we only provide a workaround for these types of skeletons as such skeletons probably are not of much importance outside of a distributed memory environment.
+An investigation of why this difference exists is beyond the scope of this work, we only provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment.
 However our workaround enables users of the DSL to test their code within a shared memory setting even though other skeletons would be better suited to be run with them.
@@
 \begin{code}
-class ArrowParallel arr a b conf => ArrowLoopParallel arr a b conf where
+class ArrowParallel arr a b conf =>
+	ArrowLoopParallel arr a b conf where
     loopParEvalN :: conf -> [arr a b] -> arr [a] [b]
@@
 \begin{code}
-instance (ArrowChoice arr, ArrowParallel arr a b Conf) => ArrowLoopParallel arr a b Conf where
+instance (ArrowChoice arr, ArrowParallel arr a b Conf) =>
+	ArrowLoopParallel arr a b Conf where
     loopParEvalN = parEvalN
@@
 \begin{code}
-instance (ArrowChoice arr, ArrowParallel arr a b (Conf b)) => ArrowLoopParallel arr a b (Conf b) where
+instance (ArrowChoice arr, ArrowParallel arr a b (Conf b)) =>
+	ArrowLoopParallel arr a b (Conf b) where
     loopParEvalN _ = evalN
@@
 defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it.
-Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
+Using |loop| to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
 
@@
 \end{code}
-\caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows.
+\caption{A~first implementation of the |pipe| skeleton expressed with parallel Arrows.
 Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt.
@@
 \label{fig:pipeSimple}
+\end{figure}
 
@@
 
+\begin{figure}[t]
 \begin{code}
@@
 Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to |parEvalN| inside of |evalN|.
-Nonetheless, we can define a version of parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
+Nonetheless, we can define a parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
 
@@
 
-We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an arrow, already fits quite neatly into the |loop|'s |arr (a, b) (c, b) -> arr a c|.
+We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an arrow, already fits quite neatly into |loop|'s signature: |arr (a, b) (c, b) -> arr a c|.
 In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)| (Fig.~\ref{fig:edenlazyrightrotate}).
@@
 
-Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| with |second get >>> f >>> second put| before converting it into a parallel arrow |arr [(i, fut r)] [(o, fut r)]| with the help of |repeat| and |loopParEvalN| to get an output list |[(o, fut r)]|.
-Finally we unzip this list into |([o], [fut r])|.
+Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)|.
+We then feed this into our parallel Arrow |arr [(i, fut r)] [(o, fut r)]| obtained by transforming our ipnut arrow |f :: arr (i, r) (o, r)| into |arr (i, fut r) (o, fut r)| before |repeat|ing and lifting it with |loopParEvalN|.
+Finally we unzip the output list |[(o, fut r)]| list into |([o], [fut r])|.
+
 Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).

### 2018-01-05 paper/content/utilityFunctions.tex
martin remarks

@@
 zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
-zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
+zipWithArr f = (arr (\(as, bs) -> zipWith (,) as bs)) >>> mapArr f
 \end{code}
@@
 \section{Syntactic Sugar} \label{syntacticSugar}
-Finally, we also give the definitions for some syntactic sugar for PArrows, namely |***| and |&&&|.
+Finally, we also give the definitions for some syntactic sugar for PArrows, namely |parstar| and |parand|.
 For basic arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two arrows |arr a b| and |arr c d| into an arrow |arr (a, c) (b, d)| which does both computations at once.
-This can easily be translated into a parallel version |***| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
+This can easily be translated into a parallel version |parstar| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
 ):
@@
 % With this we can analogously to the serial |&&&|
-We define the parallel |&&&| % (Fig.~\ref{fig:par&&&}) 
+We define the parallel |parand| % (Fig.~\ref{fig:par&&&}) 
 in a similar manner to its sequential pendant |&&&| (Fig.~\ref{fig:syntacticSugarArrows}):

### 2018-01-05 paper/content/benchmarks.tex
fixed minor unclear statement in benchmark.tex

@@
 The |Par| Monad shows a similar development, \eg with $-1.95\% \; \pm 0.64\%$ at 8 cores.
-The GpH version shows both a significant speed \textit{improvement} from PArrows of $-4.2\% \; \pm 0.26\%$ and a minor overhead of $0.87\% \; \pm 0.70\%$.
-\mbcomment{Widerspricht sich das nicht?}
+The GpH version shows both a significant speed \textit{improvement} from PArrows of $-4.2\% \; \pm 0.26\%$ (for 16 cores) and a minor overhead of $0.87\% \; \pm 0.70\%$ (4 cores).
 

### 2018-01-05 paper/content/benchmarks.tex
spellchecked

@@
 Stack itself was used in version 1.3.2.
-For GpH in its Multicore variant we used the the parallel package in version 3.2.1.0\footnote{see \url{https://hackage.haskell.org/package/parallel-3.2.1.0}}, while for the |Par| monad we used monad-par in version 0.3.4.8\footnote{see \url{https://hackage.haskell.org/package/monad-par-0.3.4.8}}.
+For GpH in its Multicore variant we used the the parallel package in version 3.2.1.0\footnote{see \url{https://hackage.haskell.org/package/parallel-3.2.1.0}}, while for the |Par| Monad we used monad-par in version 0.3.4.8\footnote{see \url{https://hackage.haskell.org/package/monad-par-0.3.4.8}}.
 For all Eden tests, we used its GHC-Eden compiler in version 7.8.2\footnote{see \url{http://www.mathematik.uni-marburg.de/~eden/?content=build_eden_7_&navi=build}} together with OpenMPI 1.6.5\footnote{see \url{https://www.open-mpi.org/software/ompi/v1.6/}}.
@@
 
-We used a single node with 16 real cores as a shared memory testbed
+We used a single node with 16 real cores as a shared memory test-bed
 and the whole grid with 256 real cores as a device to test our
@@
 The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (Par Monad), 8.14 (Eden CP) for 16 cores and 1000 Sudokus.
-In contrast to Rabin–Miller, here the |GpH| backend seems to be the best of all, while Rabin–Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
+In contrast to Rabin€“Miller, here the |GpH| backend seems to be the best of all, while Rabin€“Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
 Gentleman on shared memory has a plummeting speedup curve with GpH and |Par| Monad and logarithmically increasing speedup for the Eden-based version.
@@
 \paragraph{Speedup}
-The speedup of distributed memory Rabin–Miller benchmark with PArrows and distributed Eden backend showed an almost linear speedup sans some issues in the middle because of unfortunate task distribution.
+The speedup of distributed memory Rabin€“Miller benchmark with PArrows and distributed Eden backend showed an almost linear speedup sans some issues in the middle because of unfortunate task distribution.
 As seen in Fig.~\ref{fig:rabinMillerDistSpeedup}, we reached a speedup of 213.4 with PArrrows at 256 cores (vs.
@@
 We use our mean overhead quality measure and the notion of significance also for distributed memory benchmarks.
-The mean overhead of Rabin–Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favour of PArrows), but these values are not significant with error margins $\pm 0.8\%$ and $\pm 2.9\%$ correspondingly.
+The mean overhead of Rabin€“Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favour of PArrows), but these values are not significant with error margins $\pm 0.8\%$ and $\pm 2.9\%$ correspondingly.
 A sole significant (by a very low margin) overhead is $0.35\% \; \pm 0.33\%$ at 64 cores.

### 2018-01-05 paper/content/bestAndWorstBenchmarks.tex
spellchecked

@@
 \caption{Overhead in the shared memory benchmarks. Bold marks values
-  in favour of PArrows. Overhead is unitless, multiply by 100 to
+  in favour of PArrows. Overhead is unit-less, multiply by 100 to
   obtain percent. Runtime is in seconds.}
@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalized \\ stdDev} & \thead{Runtime for \\ 16 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 16 cores} \\ \hline \\
 	\multirow{3}{*}{\sudokutest 1000}
@@
 \caption{Overhead in the distributed memory benchmarks. Bold marks values
-  in favour of PArrows. Overhead is unitless, multiply by 100 to
+  in favour of PArrows. Overhead is unit-less, multiply by 100 to
   obtain percent. Runtime is in seconds.}
@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalized \\ stdDev} & \thead{Runtime for \\ 256 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 256 cores} \\ \hline \\
 	\multirow{1}{*}{\torustest 4096}

### 2018-01-05 paper/content/futures.tex
spellchecked

@@
 We do this with the |Future| type class (Fig.~\ref{fig:future}).
-Note that we require a |conf| parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the |ArrowParallel| typeclass earlier.
+Note that we require a |conf| parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the |ArrowParallel| type class earlier.
 \begin{figure}[h]

### 2018-01-05 paper/content/motivation.tex
spellchecked

@@
 \item We utilise multiple backends -- currently a GpH, a |Par| Monad, and Eden.
-We do not reimplement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
+We do not re-implement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
 The backends can easily be swapped, so we are not bound to any specific one.

### 2018-01-05 paper/content/parallelHaskells.tex
spellchecked

@@
 
-As the goal of this paper is not to reimplement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface.
+As the goal of this paper is not to re-implement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface.
 For this paper we chose GpH for its simplicity, the |Par| Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.
@@
 While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}.
-In the next sections we will present some \emph{trace visualizations}, the post-mortem process diagrams of Eden processes and their activity.
+In the next sections we will present some \emph{trace visualisations}, the post-mortem process diagrams of Eden processes and their activity.
 
-The trace visualizations are color-coded.
-In such a visualization (Fig.~\ref{fig:withoutFutures}), the $x$ axis shows the time, the $y$ axis enumerates the machines and processes.
-The visualization shows a running process in green, a blocked process is red.
+The trace visualisations are colour-coded.
+In such a visualisation (Fig.~\ref{fig:withoutFutures}), the $x$ axis shows the time, the $y$ axis enumerates the machines and processes.
+The visualisation shows a running process in green, a blocked process is red.
 If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow.

### 2018-01-05 paper/content/parrows.tex
spellchecked

@@
 Then, in order to fetch the results of the different threads, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)| - our intermediary Arrow is of type |arr [a] (Par [b])|.
-Finally, we execute the comptuation |Par [b]| by composing with |arr runPar| and get the final Arrow |arr [a] [b]|.
+Finally, we execute the computation |Par [b]| by composing with |arr runPar| and get the final Arrow |arr [a] [b]|.
 \begin{figure}[h]
@@
 
-Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the dataflow between the arrows, now that we have seen that Arrows are capable of expressing parallelism.
+Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the data-flow between the arrows, now that we have seen that Arrows are capable of expressing parallelism.
 We do however explain new concepts with more details if required for better understanding.

### 2018-01-05 paper/content/relwork.tex
own remarks until page 17.
spellchecked

@@
 Of course, the three parallel Haskell flavours we use as backends: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
-We use these languages as backends: our DSL can switch from one to other at user's command.
+We use these languages as backends: our DSL can switch from one to another at user's command.
 
@@
 But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
-\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimizations.
+\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimisations.
 Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution.

### 2018-01-05 paper/content/topologySkeletons.tex
spellchecked

@@
 However our workaround enables users of the DSL to test their code within a shared memory setting even though other skeletons would be better suited to be run with them.
-The idea of the fix is to provide a |ArrowLoopParallel| typeclass that has two functions - |loopParEvalN| and |postLoopParEvalN| where the first is to be used inside an |loop| construct while the latter will be used right outside of the |loop|.
+The idea of the fix is to provide a |ArrowLoopParallel| type class that has two functions - |loopParEvalN| and |postLoopParEvalN| where the first is to be used inside an |loop| construct while the latter will be used right outside of the |loop|.
 This way we can delegate to the actual |parEvalN| in the spot where the backend supports it.
@@
 Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)|.
-We then feed this into our parallel Arrow |arr [(i, fut r)] [(o, fut r)]| obtained by transforming our ipnut arrow |f :: arr (i, r) (o, r)| into |arr (i, fut r) (o, fut r)| before |repeat|ing and lifting it with |loopParEvalN|.
+We then feed this into our parallel Arrow |arr [(i, fut r)] [(o, fut r)]| obtained by transforming our input arrow |f :: arr (i, r) (o, r)| into |arr (i, fut r) (o, fut r)| before |repeat|ing and lifting it with |loopParEvalN|.
 Finally we unzip the output list |[(o, fut r)]| list into |([o], [fut r])|.

### 2018-01-05 paper/content/utilityFunctions.tex
spellchecked

@@
 This works because Arrows are strong Monads in the bicategory |Prof| of Profunctors as shown by \citet{Asada:2010:ASM:1863597.1863607}.
-Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| typeclass like presented in this paper.\footnote{For additional information on the typeclasses used, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}.}
+Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| type class like presented in this paper.\footnote{For additional information on the type classes used, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}.}
 
@@
 
-Figures~\ref{fig:parMapImg}--\ref{fig:parMapStream} show the definitions and a visualizations of two parallel |map| variants, defined using |parEvalN| and its lazy counterpart.
+Figures~\ref{fig:parMapImg}--\ref{fig:parMapStream} show the definitions and a visualisations of two parallel |map| variants, defined using |parEvalN| and its lazy counterpart.
 

### 2018-01-05 paper/content/benchmarks.tex
minor fix with encoding

@@
 The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (Par Monad), 8.14 (Eden CP) for 16 cores and 1000 Sudokus.
-In contrast to Rabin€“Miller, here the |GpH| backend seems to be the best of all, while Rabin€“Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
+In contrast to Rabin--Miller, here the |GpH| backend seems to be the best of all, while Rabin--Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
 Gentleman on shared memory has a plummeting speedup curve with GpH and |Par| Monad and logarithmically increasing speedup for the Eden-based version.
@@
 \paragraph{Speedup}
-The speedup of distributed memory Rabin€“Miller benchmark with PArrows and distributed Eden backend showed an almost linear speedup sans some issues in the middle because of unfortunate task distribution.
+The speedup of distributed memory Rabin--Miller benchmark with PArrows and distributed Eden backend showed an almost linear speedup sans some issues in the middle because of unfortunate task distribution.
 As seen in Fig.~\ref{fig:rabinMillerDistSpeedup}, we reached a speedup of 213.4 with PArrrows at 256 cores (vs.
@@
 We use our mean overhead quality measure and the notion of significance also for distributed memory benchmarks.
-The mean overhead of Rabin€“Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favour of PArrows), but these values are not significant with error margins $\pm 0.8\%$ and $\pm 2.9\%$ correspondingly.
+The mean overhead of Rabin-Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favour of PArrows), but these values are not significant with error margins $\pm 0.8\%$ and $\pm 2.9\%$ correspondingly.
 A sole significant (by a very low margin) overhead is $0.35\% \; \pm 0.33\%$ at 64 cores.

### 2018-01-05 paper/content/arrows.tex
arrow -> Arrow

@@
 To lift an ordinary function to an Arrow, |arr| is used, analogous to the monadic |return|.
-Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by \enquote{wiring} the outputs of the first to the inputs to the second to get a new arrow |arr a c|.
-Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched.
-It allows us to to save input across arrows.
+Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two Arrows |arr a b| and |arr b c| by \enquote{wiring} the outputs of the first to the inputs to the second to get a new Arrow |arr a c|.
+Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an Arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched.
+It allows us to to save input across Arrows.
 Figure~\ref{fig:arrows-viz} shows a graphical representation of these basic Arrow combinators.
@@
 % \end{figure}
-A~first short example given by Hughes on how to use arrows is addition with arrows:
-%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
+A~first short example given by Hughes on how to use Arrows is addition with Arrows:
+%|add| over Arrows, which can be seen in Fig.~\ref{fig:addArrows}.
 % \begin{figure}[h]

### 2018-01-05 paper/content/basicSkeletons.tex
spellchecked
arrow -> Arrow

@@
 |map (arr .
-(+)) [1..]| or just because we need the arrows evaluated in chunks.
+(+)) [1..]| or just because we need the Arrows evaluated in chunks.
 |parEvalNLazy| (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this.
@@
 \end{figure}
-We have only talked about the parallelisation of arrows of the same set of input and output types until now.
+We have only talked about the parallelization of Arrows of the same set of input and output types until now.
 But sometimes we want to parallelize heterogeneous types as well.
-We can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two arrows |arr a b| and |arr c d| into a new parallel arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class.
-Here, the general idea is to use the |+++| combinator which combines two arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common arrow type that we can then feed into |parEvalN|.
+We can implement such a |parEval2| combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two Arrows |arr a b| and |arr c d| into a new parallel Arrow |arr (a, c) (b, d)| quite easily with the help of the |ArrowChoice| type class.
+Here, the general idea is to use the |+++| combinator which combines two Arrows |arr a b| and |arr c d| and transforms them into |arr (Either a c) (Either b d)| to get a common Arrow type that we can then feed into |parEvalN|.
 

### 2018-01-05 paper/content/futures.tex
arrow -> Arrow

@@
 
-In a distributed environment, a resulting arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
+In a distributed environment, a resulting Arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
 
-While the above example could be rewritten into only one |parEvalN| call by directly wiring the arrows together before spawning, it illustrates an important problem.
+While the above example could be rewritten into only one |parEvalN| call by directly wiring the Arrows together before spawning, it illustrates an important problem.
 When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.

### 2018-01-05 paper/content/parrows.tex
arrow -> Arrow

@@
 A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
-Translating this into arrow terms gives us a new operator |parEvalN| that lifts a list of arrows |[arr a b]| to a parallel arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
+Translating this into Arrow terms gives us a new operator |parEvalN| that lifts a list of Arrows |[arr a b]| to a parallel Arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
 This combinator is similar to |evalN| from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
@@
 % \end{figure}
-With this definition of |parEvalN|, parallel execution is yet another arrow combinator.
-But as the implementation may differ depending on the actual type of the arrow |arr| - or even the input |a| and output |b| - and we want this to be an interface for different backends, we introduce a new type class |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
+With this definition of |parEvalN|, parallel execution is yet another Arrow combinator.
+But as the implementation may differ depending on the actual type of the Arrow |arr| - or even the input |a| and output |b| - and we want this to be an interface for different backends, we introduce a new type class |ArrowParallel arr a b|: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
 % \begin{figure}[h]
@@
 \subsubsection{|Par| Monad}
-\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
+\olcomment{introduce a newcommand for par-monad, "Arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
 Just like for GpH we can easily lift the definition of |parEvalN| for the |Par| Monad to Arrows in Fig.~\ref{fig:ArrowParallelParMonad}.
@@
 For both the GpH Haskell and |Par| Monad implementations we could use general instances of |ArrowParallel| that just require the |ArrowChoice| type class.
-With Eden this is not the case as we can only spawn a list of functions, which we cannot extract from general arrows.
+With Eden this is not the case as we can only spawn a list of functions, which we cannot extract from general Arrows.
 While we could still manage to have only one instance in the module by introducing a type class % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
@@
 
-Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the data-flow between the arrows, now that we have seen that Arrows are capable of expressing parallelism.
+Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the data-flow between the Arrows, now that we have seen that Arrows are capable of expressing parallelism.
 We do however explain new concepts with more details if required for better understanding.

### 2018-01-05 paper/content/relwork.tex
arrow -> Arrow

@@
 But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
-\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimisations.
+\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations.
 Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution.

### 2018-01-05 paper/content/topologySkeletons.tex
arrow -> Arrow

@@
 
-The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine.
+The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of Arrows with |>>>|, but does this in parallel, meaning that the Arrows do not have to reside on the same thread/machine.
 We implement this skeleton using the |ArrowLoop| type class which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input.
@@
 \end{code}
-defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it.
+defines an Arrow that takes its input |a| and converts it into an infinite stream |[a]| of it.
 Using |loop| to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
@@
 
-Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e.
+Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe Arrows of mixed types together, i.e.
 |arr a b| and |arr b c|.
-By wrapping these two arrows inside a bigger arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
+By wrapping these two Arrows inside a bigger Arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
 \begin{figure}[tb]
@@
 %\end{code}
-%It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them.
+%It does not launch both Arrows |f| and |g| in parallel, but allows for more smooth data communication between them.
 Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.
@@
 
-We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an arrow, already fits quite neatly into |loop|'s signature: |arr (a, b) (c, b) -> arr a c|.
+We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an Arrow, already fits quite neatly into |loop|'s signature: |arr (a, b) (c, b) -> arr a c|.
 In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)| (Fig.~\ref{fig:edenlazyrightrotate}).
-Similarly to the |pipe| from Section~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) arrow here, or the evaluation would fail to terminate.
+Similarly to the |pipe| from Section~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) Arrow here, or the evaluation would fail to terminate.
 The reasoning is explained by \citet{Loogen2012} as a demand problem.
@@
 Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)|.
-We then feed this into our parallel Arrow |arr [(i, fut r)] [(o, fut r)]| obtained by transforming our input arrow |f :: arr (i, r) (o, r)| into |arr (i, fut r) (o, fut r)| before |repeat|ing and lifting it with |loopParEvalN|.
+We then feed this into our parallel Arrow |arr [(i, fut r)] [(o, fut r)]| obtained by transforming our input Arrow |f :: arr (i, r) (o, r)| into |arr (i, fut r) (o, fut r)| before |repeat|ing and lifting it with |loopParEvalN|.
 Finally we unzip the output list |[(o, fut r)]| list into |([o], [fut r])|.
 
-Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
-To make sure this algorithm has speedup on shared-memory machines as well, we pass the result of this arrow to |postLoopParEvalN conf (repeat (arr id))|.
+Plugging this Arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring Arrow (Fig.~\ref{fig:ringFinal}).
+To make sure this algorithm has speedup on shared-memory machines as well, we pass the result of this Arrow to |postLoopParEvalN conf (repeat (arr id))|.
 This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
@@
 Similar to the |ring|, we start by rotating the input (Fig.~\ref{fig:edenlazyrightrotate}), but this time not only in one direction, but in two.
-This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes
+This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped Arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation Arrow becomes
 \begin{code}
@@
 \end{code}
-instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we zip the inputs for the input arrow with 
+instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we zip the inputs for the input Arrow with 
 \begin{code}
@@
 \end{figure}
-If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version and execution times are comparable.
+If we compare the trace from a call using our Arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the Arrow version and execution times are comparable.
 We discuss further benchmarks on larger clusters and in a more detail in the next section.
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[Matrix Multiplication with |torus| (Parrows)]{Matrix Multiplication with |torus| (PArrows).}
+	\caption[Matrix Multiplication with |torus| (PArrows)]{Matrix Multiplication with |torus| (PArrows).}
 	\label{fig:torus_parrows_trace}

### 2018-01-05 paper/content/utilityFunctions.tex
arrow -> Arrow

@@
 Arrow versions of Eden's |shuffle|, |unshuffle| and the definition of |takeEach| are in Figure~\ref{fig:edenshuffleetc}.
-Similarly, Figure~\ref{fig:edenlazyrightrotate} contains the definition of arrow versions of Eden's |lazy| and |rightRotate| utility functions.
+Similarly, Figure~\ref{fig:edenlazyrightrotate} contains the definition of Arrow versions of Eden's |lazy| and |rightRotate| utility functions.
 Fig.~\ref{fig:lazyzip3etc} contains Eden's definition of |lazyzip3| together with the utility functions |uncurry3| and |threetotwo|.
@@
 We start by transforming the |(a, c)| input into a two-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|.
-Next, we feed this list into a parallel arrow running on two instances of |f +++ g| as described in the paper.
+Next, we feed this list into a parallel Arrow running on two instances of |f +++ g| as described in the paper.
 After the calculation is finished, we convert the resulting |[Either b d]| into |([b], [d])| with |arr partitionEithers|.
@@
 Finally, we also give the definitions for some syntactic sugar for PArrows, namely |parstar| and |parand|.
-For basic arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two arrows |arr a b| and |arr c d| into an arrow |arr (a, c) (b, d)| which does both computations at once.
+For basic Arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two Arrows |arr a b| and |arr c d| into an Arrow |arr (a, c) (b, d)| which does both computations at once.
 This can easily be translated into a parallel version |parstar| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}

### 2018-01-06 paper/content/motivation.tex
arrow -> Arrow
Sec -> Section

@@
 %Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to Monads for API design \citHughes.
-In the paper Hughes describes how arrows are a generalization of Monads and how they are not as restrictive.
+In the paper Hughes describes how Arrows are a generalization of Monads and how they are not as restrictive.
 In this paper we will use this concept to express parallelism.
@@
 \item We introduce a parallel evaluation formalism using Arrows.
-One big advantage of our specific approach is that we do not have to introduce any new types (Sec.~\ref{sec:parallel-arrows}).
+One big advantage of our specific approach is that we do not have to introduce any new types (Section~\ref{sec:parallel-arrows}).
 This behaviour encourages better composability.
@@
 So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.
-(Sec.~\ref{sec:parallel-arrows})
+(Section~\ref{sec:parallel-arrows})
 \item We extend our PArrows formalism with |Future|s.
@@
 Direct communication is useful in a distributed memory setting because it allows for inter-node communication without blocking the master-node.
-(Sec.~\ref{sec:futures})
-\item It is possible to define algorithmic skeletons with PArrows (Sec.~\ref{sec:skeletons}). All our benchmarks were skeleton-based.
-\item Finally, we practically demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low performance overhead (Sec.~\ref{sec:benchmarks}).
+(Section~\ref{sec:futures})
+\item It is possible to define algorithmic skeletons with PArrows (Section~\ref{sec:skeletons}). All our benchmarks were skeleton-based.
+\item Finally, we practically demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low performance overhead (Section~\ref{sec:benchmarks}).
 \end{itemize}
@@
 
-%\ptcomment{* We evaluate the expressive power of  arrow parallelism. Showing that 
+%\ptcomment{* We evaluate the expressive power of  Arrow parallelism. Showing that 
 %  + there is no need to introduce additional types
@@
 %	\item \textbf{Arrow DSL benefits}:
-%    To implement parallelism, we do not introduce any new types, but only rely on a typeclass that hosts |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of arrows into a new parallel arrow.
-Therefore, we do not lose any benefits of using arrows as parallelism is just encapsulated in yet another Arrow combinator.
+%    To implement parallelism, we do not introduce any new types, but only rely on a typeclass that hosts |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of Arrows into a new parallel Arrow.
+Therefore, we do not lose any benefits of using Arrows as parallelism is just encapsulated in yet another Arrow combinator.
 The resulting Arrow can be used in the same way a potential serial version could be used.
@@
 %The remaining text is structures as follows.
-Section~\ref{sec:background} briefly introduces known parallel Haskell flavours (Sec.~\ref{sec:parEvalNIntro}) and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}).
+Section~\ref{sec:background} briefly introduces known parallel Haskell flavours (Section~\ref{sec:parEvalNIntro}) and gives an overview of Arrows to the reader (Section~\ref{sec:arrows}).
 Section~\ref{sec:related-work} discusses related work.

### 2018-01-06 paper/content/parrows.tex
Sec -> Section

@@
 \subsection{The ArrowParallel type class}
-A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
+A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Section~\ref{sec:parEvalNIntro}).
 Translating this into Arrow terms gives us a new operator |parEvalN| that lifts a list of Arrows |[arr a b]| to a parallel Arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (

### 2018-01-06 paper/content/abstract.tex
spellchecked
Phil remarks Abstract (v1)

@@
 \begin{abstract}
-Arrows are a general interface for computation and therefore form an alternative to Monads for API design.
-We express parallelism using this concept in a novel way: We define an Arrow-based language for parallelism and implement it using multiple parallel Haskells.
-In this manner we are able to bridge across various parallel Haskells.
+Arrows are a general interface for computation and an alternative to Monads for API design.
+In contrast to Monad based parallelism, we explore the use of Arrows for specifying generalised parallelism.
+Specifically, we define an Arrow-based language and implement it using multiple parallel Haskells.
 
-Additionally, using these parallel Arrows (PArrows) has the benefit of being portable across multiple parallel Haskell implementations.
-Furthermore, as each parallel computation is an Arrow, PArrows can be readily composed and transformed as such.
-In order to allow for more sophisticated communication schemes between computation nodes in distributed systems we utilise the concept of Futures to wrap existing direct communication concepts in backends.
+As each parallel computation is an Arrow, such parallel Arrows (PArrows) can be readily composed and transformed as such.
+To allow for more sophisticated communication schemes between computation nodes in distributed systems we utilise the concept of Futures to wrap direct communication.
 
-To show that PArrows have similar expressive power as existing parallel languages, we also implement several parallel skeletons. 
-Benchmarks show that our framework does not induce any notable overhead
-performance-wise.
-We finally conclude that Arrows turn out to be a useful tool for composing parallel programs and that our particular approach results in programs that are portable across multiple backends.
-\olcomment{Summarize conclusions}
+To show that PArrows have similar expressive power as existing parallel languages, we implement several skeletons and 4 benchmarks. 
+Benchmarks show that our framework does not induce any notable performance overhead.
+We conclude that Arrows have considerable potential for composing parallel programs, and for producing programs that can execute on multiple parallel language implementations.
+
+\olcomment{Summarise conclusions}
 

### 2018-01-06 paper/content/benchmarks.tex
OL: Phil's comments, part 1

@@
 
-In the following section, we describe the benchmarks we conducted of our parallel DSL and algorithmic skeletons.
-We start by explaining the hardware and software stack and also elaborate on which benchmarks programs and parallel Haskells were used in which setting and why.
-Before we go into detail on the benchmarks we
-also shortly address hyper-threading and why we do not use it in our benchmarks.
-Finally, we show that PArrows hold up in terms of performance when compared to the original parallel Haskells used as backends in this paper, starting with the shared-memory variants (GpH, |Par| Monad and Eden CP) and concluding with Eden as a distributed backend.
+The preceding section has shown that PArrows are expressive.
+This section evaluates the performance overhead of this compositional abstraction in comparison to  GpH and the |Par| Monad on shared memory architectures and Eden on  a distributed memory cluster.
+We start by explaining the hardware and software stack and outline the benchmark programs, and motivations for choosing them. Before we go into detail on the benchmarks we
+also shortly address hyper-threading and why we do not use it in our
+benchmarks.
+Our results show that PArrows hold up in terms of performance when compared to the original parallel Haskells, starting with the shared-memory variants (GpH, |Par| Monad and Eden~CP) and concluding with Eden in a distributed memory setting.
 
@@
 
-\subsection{Preliminaries}
+\subsection{Measurement Platform}
 
-\subsubsection{Hardware and Software used}
+\subsubsection{Hardware and Software}
 
-Benchmarks were run both in a shared and in a distributed
-memory setting. All benchmarks were done on the Glasgow GPG Beowulf cluster, consisting of
+The benchmarks are executed  both in a shared and in a distributed memory setting using the Glasgow GPG Beowulf cluster, consisting of
 16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each.
@@
 
-We used multiple tests that originated from different
+We measure four benchmarks from different
 sources. Most of them are parallel mathematical computations, initially
@@
 \torustest & Mathematics & \ensuremath{\Varid{torus}} & Eden & \citet{Eden:SkeletonBookChapter02}\\
-\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad}\tablefootnote{actual code from: \url{http://community.haskell.org/\~simonmar/par-tutorial.pdf and https://github.com/simonmar/parconc-examples}}\\
+\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad}\tablefootnote{actual code from: \url{http://community.haskell.org/\~simonmar/par-tutorial.pdf} and \url{https://github.com/simonmar/parconc-examples}}\\
 \bottomrule
@@
 
-In the following paragraphs we will go into detail on how well PArrows perform when compared to versions of the benchmark programs implemented with the original parallel Haskells that were used in the backends.
-We start with the definition of an Overhead to compare both PArrows-enabled and standard benchmark implementations.
+We compare the PArrow performance with direct implementations of the benchmarks in Eden, GpH and the |Par| Monad.
+We start with the definition of mean overhead to compare both PArrows-enabled and standard benchmark implementations.
 We continue comparing speedups and overheads for the shared memory backends and then study OpenMPI variants of the Eden-enabled backend as a representative of a distributed memory backend.
@@
 
-We compared the overhead of our PArrows implementations to the original parallel Haskells.
-We basically implemented the test programs both in PArrows and in traditional style, benchmarked both and will now compare the mean overhead, i.e.
-the \textit{relative} difference between both benchmarks with same settings.
-The error margin of the time measurements, supplied by criterion package, is computed together with the mean overhead computation.
+We compare the mean overhead, \ie the mean of relative differences between the PArrow and direct benchmark implementations executed multiple times with the same settings.
+The error margins of the time measurements, supplied by criterion package\footnote{\url{https://hackage.haskell.org/package/criterion-1.1.1.0}}, yield the error margin of the mean overhead.
 
@@
 \paragraph{Speedup}
-The speedup of distributed memory Rabin--Miller benchmark with PArrows and distributed Eden backend showed an almost linear speedup sans some issues in the middle because of unfortunate task distribution.
+The speedup of distributed memory Rabin--Miller benchmark with PArrows and distributed Eden backend showed an almost 
+linear speedup excepting around 192 cores where an unfortunate task distribution reduces performance.
 As seen in Fig.~\ref{fig:rabinMillerDistSpeedup}, we reached a speedup of 213.4 with PArrrows at 256 cores (vs.
 207.7 with pure Eden).
-We managed to measure the speedup of Jacobi sum test for sufficient large inputs like $2^{4253}-1$ only in a massively distributed setting, because of memory limitations.
-We improved there from 9193 seconds (128 cores) to 1649 seconds (256 cores) in our PArrows version.
+Because of memory limitations, the speedup of Jacobi sum test for large inputs (such as $2^{4253}-1$) could be measured only in a massively distributed setting.
+PArrows improved there from \SI{9193}{\second} (at 128 cores) to \SI{1649}{\second} (at 256 cores).
 A~scaled-down version with input $2^{3217}-1$ stagnates the speedup at about 11 for both PArrows and Eden for more than 64 cores.
 There is apparently not enough work for that many cores.
-The Gentleman test with input 4096 had an almost linear speedup first, then plummeted between 128 and 224 cores, and recovered at 256 cores with speedup of 129.
+The Gentleman test with input 4096 had an almost linear speedup first, then plummeted between 128 and 224 cores, and recovered at 256 cores with speedup of~129.
 

### 2018-01-06 paper/content/bestAndWorstBenchmarks.tex
OL: Phil's comments, part 1

--- 2018-01-05 paper/content/bestAndWorstBenchmarks.tex
@@
 \caption{Overhead in the shared memory benchmarks. Bold marks values
-  in favour of PArrows. Overhead is unitless, multiply by 100 to
-  obtain percent. Runtime is in seconds.}
+  in favour of PArrows. Overhead is unit-less. Runtime is in seconds.}
 \label{tab:meanOverheadSharedMemory}
@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalized \\ stdDev} & \thead{Runtime for \\ 16 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 16 cores (s)} \\ \hline \\
 	\multirow{3}{*}{\sudokutest 1000}
@@
 \caption{Overhead in the distributed memory benchmarks. Bold marks values
-  in favour of PArrows. Overhead is unitless, multiply by 100 to
-  obtain percent. Runtime is in seconds.}
+  in favour of PArrows. Overhead is unit-less. Runtime is in seconds.}
 \olcomment{Vielleicht gleich durchmultiplizieren und \% angeben? Nicht
@@
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
-	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalized \\ stdDev} & \thead{Runtime for \\ 256 cores} \\ \hline \\
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 256 cores (s)} \\ \hline \\
 	\multirow{1}{*}{\torustest 4096}

### 2018-01-06 paper/content/topologySkeletons.tex
OL: Phil's comments, part 1

@@
 \end{code}
-\caption{A~first implementation of the |pipe| skeleton expressed with parallel Arrows.
-Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt.
+\caption{Simple |pipe| skeleton.
+The use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt.
 We need to enforce that the evaluation of the input |[a]| terminates before passing it into |evalN|.}
@@
 \end{code}
-\caption{Final definition of the |pipe| skeleton with Futures.}
+\caption{|pipe| skeleton definition with Futures.}
 \label{fig:pipe}
@@
 	\includegraphics[scale=0.75]{images/ring}
-	\caption{Schematic depiction of the ring skeleton.}
+	\caption{|ring| skeleton depiction.}
 	\label{fig:ringImg}
@@
 \end{code}
-\caption{Definition of the |ring| skeleton.}
+\caption{|ring| skeleton definition.}
 \label{fig:ringFinal}
@@
 	\includegraphics[scale=0.75]{images/torus}
-	\caption{Schematic depiction of the |torus| skeleton.}
+	\caption{|torus| skeleton depiction.}
 	\label{fig:ringTorusImg}
@@
 \end{code} % $
-\caption{Definition of the |torus| skeleton. The definitions of |lazyzip3|, |uncurry3| and |threetotwo| have been omitted and can be found in Fig.~\ref{fig:lazyzip3etc}}.
+\caption{|torus| skeleton definition. |lazyzip3|, |uncurry3| and |threetotwo| definitions are in Fig.~\ref{fig:lazyzip3etc}}.
 \label{fig:torus}
@@
           result = foldl1' matAdd sms
-\end{code}
+\end{code} %$
 \caption{Adapted matrix multiplication in Eden using a the |torus| skeleton.
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[Matrix Multiplication with |torus| (PArrows)]{Matrix Multiplication with |torus| (PArrows).}
+	\caption[Matrix multiplication with |torus| (PArrows)]{Matrix Multiplication with |torus| (PArrows).}
 	\label{fig:torus_parrows_trace}
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
-	\caption[Matrix Multiplication with |torus| (Eden)]{Matrix Multiplication with |torus| (Eden).}
+	\caption[Matrix multiplication with |torus| (Eden)]{Matrix Multiplication with |torus| (Eden).}
 	\label{fig:torus_eden_trace}

### 2018-01-06 paper/content/motivation.tex
phil first email remarks

@@
 \item We introduce a parallel evaluation formalism using Arrows.
-One big advantage of our specific approach is that we do not have to introduce any new types (Section~\ref{sec:parallel-arrows}).
-This behaviour encourages better composability.
-\item We utilise multiple backends -- currently a GpH, a |Par| Monad, and Eden.
+One big advantage of our specific approach is that we do not have to introduce any new types, facilitating composability (Section~\ref{sec:parallel-arrows}).
+\item We show that PArrow programs can readily exploit multiple parallel language implementations.
+We demonstrate the use of GpH, a |Par| Monad, and Eden.
 We do not re-implement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
@@
 
-So as an example, during development, we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.
+This has many practical advantages.
+For example, during development we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.
 (Section~\ref{sec:parallel-arrows})
-\item We extend our PArrows formalism with |Future|s.
-Our goal here is to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|).
+\item We extend the PArrows formalism with |Future|s to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|) \citet{Dieterle2010}.
 Direct communication is useful in a distributed memory setting because it allows for inter-node communication without blocking the master-node.
 (Section~\ref{sec:futures})
-\item It is possible to define algorithmic skeletons with PArrows (Section~\ref{sec:skeletons}). All our benchmarks were skeleton-based.
-\item Finally, we practically demonstrate that Arrow parallelism is a viable alternative to existing approaches. It introduces only low performance overhead (Section~\ref{sec:benchmarks}).
+\item We demonstrate the expressiveness of PArrows by using them to define common algorithmic skeletons ((Section~\ref{sec:skeletons}), and by using these skeletons to implement four benchmarks (Section~\ref{sec:benchmarks}).
+\item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg with mean of mean overhead of less than $3.5\%, 0.8\%$ for all benchmarks with GpH and Eden respectively.
+Compared to |Par| Monad we virtually noticed no overhead with a mean of mean overheads of less than $0\%$ in all benchmarks.
+(Section~\ref{sec:benchmarks}).
 \end{itemize}

### 2018-01-06 paper/content/arrows.tex
phil remarks pages 1-22 (ish)

@@
 
-Also note that, while we could have based our DSL on Profunctors as well, we chose Arrows \mbcomment{cite missing?}.
+While we could have based our DSL on Profunctors as well, we chose Arrows \mbcomment{cite missing?}.
 for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
 However, they are a promising candidate for future improvements of our DSL.
-Note that some Profunctors, especially ones supporting a composition operation among other things, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
+Some Profunctors, especially ones supporting a composition operation, choice, and looping, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.

### 2018-01-06 paper/content/basicSkeletons.tex
phil remarks pages 1-22 (ish)

@@
 \label{sec:extending-interface}
-With the |ArrowParallel| type class in place and implemented, we can now define some further basic parallel interface functions.
-These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.
+With the |ArrowParallel| type class in place and implemented, we can now define other parallel interface functions.
+These are basic algorithmic skeletons that are used to define more sophisticated  skeletons.
 

### 2018-01-06 paper/content/futures.tex
phil remarks pages 1-22 (ish)

--- 2018-01-05 paper/content/futures.tex
@@
 \section{Futures} \label{sec:futures}\label{futures}
-Consider a mock-up parallel Arrow combinator:
+Consider an outline parallel Arrow combinator:
 \begin{code}
@@
 
-In a distributed environment, a resulting arrow of this combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
+In a distributed environment such a combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
 
-While the above example could be rewritten into only one |parEvalN| call by directly wiring the arrows together before spawning, it illustrates an important problem.
+While the example could be rewritten into a single |parEvalN| call by directly wiring the Arrows together before spawning, it illustrates an important problem.
 When using a |ArrowParallel| backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}.
@@
 Each bar represents one process.
-Black lines between processes represent communication.
+Black lines represent communication.
 Colours: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
@@
 
-This is in fact only a problem in distributed memory (in the scope of this paper) and we should allow nodes to communicate directly with each other.
-Eden already ships with \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
+This is only a problem in distributed memory (in the scope of this paper) and we should allow nodes to communicate directly with each other.
+Eden already provides \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
 But as we want code with our DSL to be implementation agnostic, we have to wrap this concept.
 We do this with the |Future| type class (Fig.~\ref{fig:future}).
-Note that we require a |conf| parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the |ArrowParallel| typeclass earlier.
+We require a |conf| parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the |ArrowParallel| type class earlier.
 \begin{figure}[h]
@@
 
-For the |Par| Monad and GpH Haskell backends, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the type class is satisfied.
+For GpH and |Par| Monad, we can simply use |BasicFuture|s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the type class is satisfied.
 This is because the concept of a |Future| does not change anything for shared-memory execution as there are no communication problems to fix.
@@
 	\caption[with Futures]{Communication between 4 Eden processes with Futures.
-Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (black lines between the bars, one example message is marked in the Figure) instead of always going through the master node (bottom bar).}
+Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is marked in the Figure) instead of always going through the master node (bottom bar).}
 	\label{fig:withFutures}

### 2018-01-06 paper/content/mapSkeletons.tex
arrow -> Arrow
phil remarks pages 1-22 (ish)

@@
 Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations.
-While there are many possible skeletons to implement, here we only regard in detail some |map|-based and topological skeletons to demonstrate the power of PArrows.
+While there are many possible skeletons to implement, we demonstrate the expressive power of PArrows here using four |map|-based and three toplogical skeletons.
 %%% \FloatBarrier
@@
 \label{sec:map-skeletons}
-We start with |map|-based skeletons.
-The essential differences between the skeletons presented here are in terms of order of evaluation and work distribution but still provide the same output of a standard |map|.
+The essential differences between the mapping skeletons presented here are in terms of order of evaluation and work distribution but still provide the same semantics as a sequential |map|.
 
@@
 The |parMap| skeleton (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs.
-We can implement it with |ArrowParallel| by repeating an arrow |arr a b| and then passing it into |parEvalN| to obtain an arrow |arr [a] [b]|.
+We can implement it with |ArrowParallel| by repeating an Arrow |arr a b| and then passing it into |parEvalN| to obtain an Arrow |arr [a] [b]|.
 Just like |parEvalN|, |parMap| traverses all input Arrows as well as the inputs.
@@
 So it makes sense to also have a |parMapStream| (Figs.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like |parMap|, but uses |parEvalNLazy| instead of |parEvalN|.
-The code of these two skeletons is quite straightforward, we show it in Appendix \ref{app:omitted} in Figs.\ref{fig:parMap} and \ref{fig:parMapStream}.
+Implementing these skeletons is straightforward as in Appendix \ref{app:omitted} in Figs.\ref{fig:parMap} and \ref{fig:parMapStream}.
 
@@
 \includegraphics[scale=0.7]{images/farm}
-\caption{Schematic depiction of a |farm|, a statically
-      load-balanced |parMap|.}
+\caption{|farm| depiction.}
 \label{fig:farmImg}
@@
 \end{code}
-\caption{The definition of |farm|.}
+\caption{|farm| definition.}
 \label{fig:farm}
@@
 \includegraphics[scale=0.7]{images/farmChunk}
-\caption{Schematic depiction of |farmChunk|.}
+\caption{|farmChunk| depiction.}
 \label{fig:farmChunkImg}

### 2018-01-06 paper/content/parrows.tex
phil remarks pages 1-22 (ish)

--- 2018-01-05 paper/content/parrows.tex
@@
 \label{sec:parallel-arrows}
-Arrows are a general interface to computation.
-Here we introduce special Arrows as a general interface to \textit{parallel computations}.
-First, we present the |ArrowParallel| type class and explain the reasoning behind it.
-Then, we discuss some implementations using existing parallel Haskells.
-Finally, we explain why using Arrows for expressing parallelism is beneficial.
+While Arrows are a general interface to computation, we introduce here specialised Arrows as a general interface to \textit{parallel computations}.
+We present the |ArrowParallel| type class and explain the reasoning behind it before discussing some parallel Haskell implementations and basic extensions.
 \subsection{The ArrowParallel type class}
-A~parallel computation (on functions) in its purest form can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Sec.~\ref{sec:parEvalNIntro}).
+A~parallel computation (on functions) can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Section~\ref{sec:parEvalNIntro}).
 Translating this into Arrow terms gives us a new operator |parEvalN| that lifts a list of Arrows |[arr a b]| to a parallel Arrow |arr [a] [b]|. % (Fig.~\ref{fig:parEvalNArrowFn}) (
@@
 \end{code}
-Note that by restricting the implementations of our backends to a specific |conf| type, we also get interoperability between backends for free.
+By restricting the implementations of our backends to a specific |conf| type, we also get interoperability between backends for free.
 We can parallelize one part of a program using one backend, and parallelize the next with another one.
@@
 \end{code}% $ %% formatting
-\caption{The |ArrowParallel| instance for the GpH backend.}
+\caption{GpH |ArrowParallel| instance.}
 \label{fig:ArrowParallelMulticore}
@@
 \olcomment{introduce a newcommand for par-monad, "Arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
-Just like for GpH we can easily lift the definition of |parEvalN| for the |Par| Monad to Arrows in Fig.~\ref{fig:ArrowParallelParMonad}.
+As for GpH we can easily lift the definition of |parEvalN| for the |Par| Monad to Arrows in Fig.~\ref{fig:ArrowParallelParMonad}.
 To start off, we define the |Strategy a| and |Conf a| type so we can have a configurable instance of ArrowParallel:
@@
 \end{code} %$ %% formatting
-\caption{ArrowParallel instance for the Par Monad backend.}
+\caption{|Par| Monad |ArrowParallel| instance.}
 \label{fig:ArrowParallelParMonad}
@@
 % We don't do it here
-, we avoid doing so for aesthetic reasons. For now, we just implement |ArrowParallel| for normal functions: % (Fig.~\ref{fig:ArrowParallelEdenFns})
-% \begin{figure}[h]
+we avoid doing so for aesthetic reasons.
+For now, we just implement |ArrowParallel| for normal functions and the Kleisli type in Fig.~\ref{fig:ArrowParallelEden}, where |Conf| is simply defined as |data Conf = Nil| since Eden does not have a configurable |spawnF| variant.
+% (Fig.~\ref{fig:ArrowParallelEdenFns})
+\begin{figure}[t]
 \begin{code}
@@
     parEvalN _ = spawnF
-\end{code}
-% \caption{ArrowParallel instance for functions in the Eden backend}
-% \label{fig:ArrowParallelEdenFns}
-% \end{figure}
-and the Kleisli type: % (Fig.~\ref{fig:ArrowParallelEdenKleisli}).
-% \begin{figure}[h]
-\begin{code}
+
 instance (ArrowParallel (->) a (m b) Conf,
@@
                        Kleisli sequence
-\end{code}
-where |Conf| is simply defined as |data Conf = Nil| since Eden does not have a configurable |spawnF| variant.
-% \caption{ArrowParallel instance for the Kleisli type in the Eden backend}
-% \label{fig:ArrowParallelEdenKleisli}
-% \end{figure}
+\end{code} %$ %% formatting
+\caption{Eden |ArrowParallel| instance.}
+\label{fig:ArrowParallelEden}
+\end{figure}
 
@@
 The other backends have similarly structured implementations which we do not discuss here for the sake of brevity.
-Also note that we can only have one instance of |ArrowParallel arr a b ()| present at a time, which should not be a problem, though.
+We can, however, only have one instance of |ArrowParallel arr a b ()| present at a time, which should not be a problem, though.
 
 Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the data-flow between the Arrows, now that we have seen that Arrows are capable of expressing parallelism.
-We do however explain new concepts with more details if required for better understanding.
+We do explain new concepts with more details if required for better understanding, though.

### 2018-01-06 paper/content/topologySkeletons.tex
phil remarks pages 1-22 (ish)

@@
           result = foldl1' matAdd sms
-\end{code} %$
+\end{code}
 \caption{Adapted matrix multiplication in Eden using a the |torus| skeleton.
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[Matrix multiplication with |torus| (PArrows)]{Matrix Multiplication with |torus| (PArrows).}
+	\caption[Matrix Multiplication with |torus| (PArrows)]{Matrix Multiplication with |torus| (PArrows).}
 	\label{fig:torus_parrows_trace}
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
-	\caption[Matrix multiplication with |torus| (Eden)]{Matrix Multiplication with |torus| (Eden).}
+	\caption[Matrix Multiplication with |torus| (Eden)]{Matrix Multiplication with |torus| (Eden).}
 	\label{fig:torus_eden_trace}

### 2018-01-06 paper/content/conclusion.tex
martin remarks
spellchecked
OL: Phil's comments, part 2

--- 2017-12-19 paper/content/conclusion.tex
@@
 combinators. To our knowledge we are first to represent
-\emph{parallel} computation with Arrows.
-% \olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done
+\emph{parallel} computation with
+Arrows, and hence to show their usefulness for
+% Arrows are a useful tool for
+composing parallel
+programs.
+We have shown that for a generic and extensible parallel Haskell, we do not have to restrict ourselves to a monadic interface.
+% Instead, we use Arrows which are a more general concept than Monads, but still allow for powerful composition.
+Furthermore, we think
+We argue that Arrows are better suited to  parallelise
+pure functions than monads,  as the functions are already Arrows and can be used
+directly in our DSL.
+Arrows are a better fit to parallelise pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way.
+% We manage to retain this nice property of parallel Haskells such as Eden or GpH (which use pure constructs)
+% while still having a generic and composable interface.
+We use a non-monadic interface (similar to Eden or GpH) and retain composibility.
 %
-Arrows turn out to be a useful tool for composing parallel
-programs.
-We have shown, that, in order to have a generic and extensible parallel Haskell, we do not have to restrict ourselves to a monadic interface.
-Instead, we use Arrows which are a more general concept than Monads, but still allow for powerful composition.
-Furthermore, we think Arrows are a better fit to parallelize pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way.
-We manage to retain this nice property of parallel Haskells like Eden or GpH (which use pure constructs) while still having a generic and composable interface.
-%
-The DSL natively allows for parallelization of monadic code via the Kleisli type and additionally allows to parallelize any Arrow type that has an instance for |ArrowChoice| (note that some skeletons require an additional |ArrowLoop| instance).
-%
-Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden.
+The DSL allows for a direct parallelisation of monadic code via the Kleisli type and additionally allows to parallelise any Arrow type that has an instance for |ArrowChoice| (but some skeletons require an additional |ArrowLoop| instance).
+
+We have demonstrated the generality of the approach by exhibiting PArrow implementations for Eden, GpH, and the |Par| Monad.
+Hence, paralle programs can be ported between task parallel Haskell implementations with little or no effort.
+We are confident that it will be straightforward to add other task-parallel backends.
+% Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden.
 With our approach parallel programs can be ported across these flavours with little to no effort.
 It is quite straightforward to add further backends.
+% %
+% %
+% Performance-wise,
+%Parallel Arrows are on par with existing parallel Haskells, as they only introduce minor overhead in some of our benchmarks.
+%
+%The benefit is, however, the
+In other words, PArrows greatly increase portability of parallel Haskell programs.
 %
 %
-Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they only introduce minor overhead in some of our benchmarks.
-%
-The benefit is, however, the greatly increased portability of parallel programs.
+Our measurements of four benchmarks on both shared and distributed memory platforms shows that the generality and portability of PArrows has very low performance overheads, \ie never more than $8\% \; \pm 6.9\%$ and typically under $2\%$.
 
@@
 
-Our PArrows DSL can be expanded to further parallel Haskells.
+Our PArrows DSL can be expanded % to further parallel Haskells.
 More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} for this future extension.
 HdpH is a modern distributed Haskell that would benefit from our Arrow notation.
-Further Future-aware versions of Arrow combinators can be defined.
-Existing combinators could also be improved.
-Arrow-based notation might enable further compiler optimizations.
+to other task parallel Haskells, and a specific target is HdpH \cite{Maier:2014:HDS:2775050.2633363}.
+Further Future-aware versions of Arrow combinators can be defined. Existing combinators could also be improved, for example a more special versions of |>>>| and |***| combinators are viable.
+%Arrow-based notation might enable further compiler optimisations, such as
 
-More experiences with seamless porting of parallel PArrows-based programs across the backends are welcome.
-Of course, we are ourselves working on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
+% More experiences with seamless porting of parallel PArrows-based programs across the backends are welcome.
+% Of course, we are ourselves working on expanding
+In ongoing work we are expanding 
+both our skeleton library and the number of skeleton-based parallel programs that use our DSL.
+% to be portable across flavours of parallel Haskells.
 It would also be interesting to see a hybrid of PArrows and Accelerate \cite{McDonell:2015:TRC:2887747.2804313}.
-Ports of our approach to other languages like Frege, Eta or Java directly are in an early development stage.
+Ports of our approach to other languages such as Frege, Eta, or Java directly are at an early development stage.

### 2018-01-06 paper/content/benchmarks.tex
minor: dots in paragraphs

@@
 
-\paragraph{Speedup}
+\paragraph{Speedup.}
 The \rmtest benchmark showed almost linear speedup for both 32 and 64 tasks, the performance is slightly better in the latter case: 13.7 at 16 cores for input $2^{11213}-1$ and 64 tasks in the best case scenario with Eden CP.
@@
 
-\paragraph{Overhead}
+\paragraph{Overhead.}
 
@@
 
-\paragraph{Speedup}
+\paragraph{Speedup.}
 The speedup of distributed memory Rabin--Miller benchmark with PArrows and distributed Eden backend showed an almost 
@@
 
-\paragraph{Overhead}
+\paragraph{Overhead.}
 We use our mean overhead quality measure and the notion of significance also for distributed memory benchmarks.

### 2018-01-06 paper/content/relwork.tex
OL: Phil's remarks in RelWork, part 1

@@
 
+\ptcomment{The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations.
+A range of approaches have been explored, including data parallelism (Chakravarty et al., 2007; Keller et al.,
+2010), GPU-based approaches (Mainland & Morrisett, 2010; Svensson, 2011), software
+transactional memory (Harris et al., 2005; Perfumo et al., 2008). The Haskell–GPU bridge
+Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
+deserves a special mention. <<Why does it deserve special mention?>>  Accelerate is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.
+
+Our PArrow implementation uses three task parallel languages as backends: the
+GpH (Trinder et al., 1996, 1998) parallel Haskell dialect and its multicore version (Marlow
+et al., 2009), the Par Monad (Marlow et al., 2011; Foltzer et al., 2012), and Eden (Loogen
+et al., 2005; Loogen, 2012). These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available (Aljabri et al., 2014,
+2015). Research on Eden includes a low-level  implementation (Berthold, 2008; Berthold et al., 2016), work on skeleton
+composition (Dieterle et al., 2016), communication (Dieterle et al., 2010a), and generation
+of process networks (Horstmeyer & Loogen, 2013). The definitions of new Eden skeletons is a specific focus (Hammond et al., 2003; Berthold
+& Loogen, 2006; Berthold et al., 2009b,c; Dieterle et al., 2010b; de la Encina et al., 2011;
+Dieterle et al., 2013; Janjic et al., 2013).
+
+Other task parallel Haskells related to Eden, GpH, and the Par Monad include the following.  
+HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of Par Monad to
+heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
+of Par Monad.}
+
+\paragraph{Parallel Haskells.}
+The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations.
+A range of approaches have been explored, including data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
+The Haskell--GPU bridge Accelerate
+\cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313}
+is completely orthogonal to our approach.
+A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.
+
+Our PArrow implementation uses three task parallel languages as backends: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012}.
+These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available \cite{Aljabri:2013:DIG:2620678.2620682,Aljabri2015}.
+Research on Eden includes low-level  implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016},
+work on skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013}.
+The definitions of new Eden skeletons is a specific focus \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.
+
+Other task parallel Haskells related to Eden, GpH, and the |Par| Monad include the following.  
+HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of |Par| Monad to
+heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension
+of |Par| Monad.
+
+\olcomment{OLD:
 \paragraph{Parallel Haskells.}
@@
 
-More different approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
+Alternative approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
 %
-The Haskell--GPU bridge Accelerate \cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313} deserves a special mention.
-Accelerate is completely orthogonal to our approach.
-\citeauthor{marlow2013parallel} authored a recent book in \citeyear{marlow2013parallel} on parallel Haskells.
+The Haskell--GPU bridge Accelerate
+\cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313}
+deserves a special mention. Accelerate is completely orthogonal to our
+approach. \citeauthor{marlow2013parallel} authored a recent book in
+\citeyear{marlow2013parallel} on parallel Haskells.
+}
 

### 2018-01-06 paper/content/relwork.tex
OL: types of skeletons, as Phil suggested

@@
 \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
+Typical types of algorithmic skeletons include |map|-, |fold|-, and |scan|-based parallel programming patterns, special applications such as divide-and-conquer or topological skeletons.
+
+The |farm| skeleton \citep{Hey1990185,Eden:PPDP01,Kuchen05} is statically task-balanced parallel |map|.
+When tasks' durations cannot be foreseen, a dynamic load balancing (|workpool|) brings a lot of improvement \citep{Rudolph:1991:SLB:113379.113401,Eden:PPL03,Hippold2006,PADL08HMWS,Marlow2009}.
+For special tasks |workpool| skeletons can be extended with dynamic task creation \cite{WPEuropar06,Dinan:2009:SWS:1654059.1654113,brown2010ever}.
+%
+The |fold| (or |reduce|) skeleton was implemented in various skeleton libraries \cite{Kuchen2002,BUONO20102095,Dastgeer:2011:ASM:1984693.1984697,5361825}, as also its inverse, |scan|, \cite{Bischof2002,harris2007parallel}.
+%
+Google |map|--|reduce| is more special than just a composition of the two skeletons \cite{LAMMEL20081,Berthold2009-mr}.
+
 The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,lobachev-phd,Lobachev2012,janjic2013space}, iteration skeletons \cite{Dieterle2013}.

### 2018-01-06 paper/content/relwork.tex
OL: work stealing and whatnot for workpools
OL: Phil's remarks in RelWork, part 2

@@
 For special tasks |workpool| skeletons can be extended with dynamic task creation \cite{WPEuropar06,Dinan:2009:SWS:1654059.1654113,brown2010ever}.
+Efficient load-balancing schemes for |workpool|s are subject of research \cite{Acar:2000:DLW:341800.341801,Blumofe:1999:SMC:324133.324234,vanNieuwpoort:2001:ELB:568014.379563,Chase:2005:DCW:1073970.1073974,4625841,Michael:2009:IWS:1594835.1504186}.
 %
@@
 \paragraph{Arrows.}
-Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, basically they are a generalised function arrow~|->|.
+Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~|->|.
 \citet{Hughes2005} presents a tutorial on Arrows.
-Some theoretical details on Arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable.
+\citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows.
 \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows.
@@
 
-\paragraph{Other languages.}
+\paragraph{Arrows in other languages.}
 Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined.
-Experiments with our approach in Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM) were quite successful.
-A new approach to Haskell on the JVM is the Eta language\footnote{Eta project page at \url{eta-lang.org}}.
-However, both are beyond the scope of this work.
+Basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM).
+However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{eta-lang.org}}
+a new approach to Haskell on the JVM.
 

### 2018-01-06 paper/content/relwork.tex
make it compile

@@
 is completely orthogonal to our approach.
-A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.
+A~good survey of parallel Haskells can be found in~\citet{marlow2013parallel}.
 
@@
 These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available \cite{Aljabri:2013:DIG:2620678.2620682,Aljabri2015}.
-Research on Eden includes low-level  implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016},
-work on skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013}.
+Research on Eden includes low-level  implementation
+\cite{JostThesis,berthold_loidl_hammond_2016}, skeleton composition
+\cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013}.
 The definitions of new Eden skeletons is a specific focus \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.

### 2018-01-06 paper/content/relwork.tex
sorting out references

@@
 The |farm| skeleton \citep{Hey1990185,Eden:PPDP01,Kuchen05} is statically task-balanced parallel |map|.
-When tasks' durations cannot be foreseen, a dynamic load balancing (|workpool|) brings a lot of improvement \citep{Rudolph:1991:SLB:113379.113401,Eden:PPL03,Hippold2006,PADL08HMWS,Marlow2009}.
+When tasks' durations cannot be foreseen, a dynamic load balancing (|workpool|) brings a lot of improvement \citep{Rudolph:1991:SLB:113379.113401,doi:10.1142/S0129626403001380,Hippold2006,PADL08HMWS,Marlow2009}.
 For special tasks |workpool| skeletons can be extended with dynamic task creation \cite{WPEuropar06,Dinan:2009:SWS:1654059.1654113,brown2010ever}.

### 2018-01-06 paper/content/relwork.tex
polishing relwork

@@
 Algorithmic skeletons were introduced by \citet{Cole1989}.
-Early publications on this topic include \cite{darlington1993parallel,botorog1996efficient,p3l97,Gorlatch1998,Lengauer1997}.
+Early publications on this topic include \cite{DANELUTTO1992205,darlington1993parallel,botorog1996efficient,Lengauer1997,Gorlatch1998}.
 \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
@@
 
-The |farm| skeleton \citep{Hey1990185,Eden:PPDP01,Kuchen05} is statically task-balanced parallel |map|.
+The |farm| skeleton \citep{Hey1990185,Eden:PPDP01,Kuchen05} is a statically task-balanced parallel |map|.
 When tasks' durations cannot be foreseen, a dynamic load balancing (|workpool|) brings a lot of improvement \citep{Rudolph:1991:SLB:113379.113401,doi:10.1142/S0129626403001380,Hippold2006,PADL08HMWS,Marlow2009}.
 For special tasks |workpool| skeletons can be extended with dynamic task creation \cite{WPEuropar06,Dinan:2009:SWS:1654059.1654113,brown2010ever}.
-Efficient load-balancing schemes for |workpool|s are subject of research \cite{Acar:2000:DLW:341800.341801,Blumofe:1999:SMC:324133.324234,vanNieuwpoort:2001:ELB:568014.379563,Chase:2005:DCW:1073970.1073974,4625841,Michael:2009:IWS:1594835.1504186}.
+Efficient load-balancing schemes for |workpool|s are subject of research \cite{Blumofe:1999:SMC:324133.324234,Acar:2000:DLW:341800.341801,vanNieuwpoort:2001:ELB:568014.379563,Chase:2005:DCW:1073970.1073974,4625841,Michael:2009:IWS:1594835.1504186}.
 %
-The |fold| (or |reduce|) skeleton was implemented in various skeleton libraries \cite{Kuchen2002,BUONO20102095,Dastgeer:2011:ASM:1984693.1984697,5361825}, as also its inverse, |scan|, \cite{Bischof2002,harris2007parallel}.
+The |fold| (or |reduce|) skeleton was implemented in various skeleton libraries \cite{Kuchen2002,5361825,BUONO20102095,Dastgeer:2011:ASM:1984693.1984697}, as also its inverse, |scan| \cite{Bischof2002,harris2007parallel}.
 %
-Google |map|--|reduce| is more special than just a composition of the two skeletons \cite{LAMMEL20081,Berthold2009-mr}.
+Google |map|--|reduce| \cite{Dean:2008:MSD:1327452.1327492,Dean:2010:MFD:1629175.1629198} is more special than just a composition of the two skeletons \cite{LAMMEL20081,Berthold2009-mr}.
 
@@
 Basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM).
-However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{eta-lang.org}}
+However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{eta-lang.org}},
 a new approach to Haskell on the JVM.

### 2018-01-06 paper/content/motivation.tex
phil remarks pages 1-22 (ish)
minor

@@
 the parallel backends at will.
-The induced penalty was in the single-digit percent range, with means over the varying cores configuration typically under 1\% overhead in our measurements (Section~\ref{sec:benchmarks}).
+The induced penalty was in the single-digit percent range, with means over the varying cores configuration typically under 2\% overhead in our measurements (Section~\ref{sec:benchmarks}).
 Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
@@
 
-We structure this paper as follows:
 \begin{itemize}
@@
 PArrows are open source and available from \url{https://github.com/s4ke/Parrows}.
-
-%\ptcomment{* We evaluate the expressive power of  Arrow parallelism. Showing that 
-%  + there is no need to introduce additional types
-%  + that it is possible to define algorthmic skeletons (Section 6,7)
-%  + ...
-
-%* We demonstrate that Arrow Parallelism can exploit multiple parallel %Haskell implementations, and with low overhead, i.e. less than X%. 
-%... (Section 8) }
-
-%We wrap parallel Haskells inside of our |ArrowParallel| typeclass, but
-% why do we aim to abstract parallelism this way and what does this
-% approach do better than the other parallel Haskells?
-% is such a parallelism abstraction of benefit and to what extent does
-% it improve existing approaches?
-%\begin{itemize}
-%	\item \textbf{Arrow DSL benefits}:
-%    To implement parallelism, we do not introduce any new types, but only rely on a typeclass that hosts |parEvalN :: [arr a b] -> arr [a] [b]|, which converts a list of Arrows into a new parallel Arrow.
-Therefore, we do not lose any benefits of using Arrows as parallelism is just encapsulated in yet another Arrow combinator.
-The resulting Arrow can be used in the same way a potential serial version could be used.
-This is a big advantage of this approach, especially compared to Monad solutions like the |Par| Monad which require specialized Monad types.
-We can just \enquote{plug} in parallel parts into sequential Arrow-based programs without having to change anything.
-%	\item \textbf{Abstraction}:
-%	With the |ArrowParallel| typeclass, we abstract all parallel implementation logic away from the business logic.
-This means it is possible to write code against the interface of a common typeclass without being bound to any parallel Haskell.
-So as an example, during development, we can run the program in a simple GHC-compiled variant and afterwards deploy it on a cluster by converting it into an Eden version, by just replacing the current |ArrowParallel| instance.
-%\end{itemize}
-
-%\paragraph{Structure}
-%The remaining text is structures as follows.
-Section~\ref{sec:background} briefly introduces known parallel Haskell flavours (Section~\ref{sec:parEvalNIntro}) and gives an overview of Arrows to the reader (Section~\ref{sec:arrows}).
-Section~\ref{sec:related-work} discusses related work.
-Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface.
-Section~\ref{futures} defines Futures for Parallel Arrows, this concept enables better communication.
-Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons  in our newly defined dialect: parallel |map| with and without load balancing.
-More advanced skeletons are showcased in Section~\ref{sec:topology-skeletons} (|pipe|, |ring|, |torus|).
-Section~\ref{sec:benchmarks} shows the benchmark results.
-Section~\ref{sec:conclusion} discusses future work and concludes.
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: "main"
-%%% End:

### 2018-01-06 paper/content/abstract.tex
phil first email remarks
minor things

@@
 Arrows are a general interface for computation and an alternative to Monads for API design.
-In contrast to Monad based parallelism, we explore the use of Arrows for specifying generalised parallelism.
+In contrast to Monad-based parallelism, we explore the use of Arrows for specifying generalised parallelism.
 Specifically, we define an Arrow-based language and implement it using multiple parallel Haskells.
@@
 
-To show that PArrows have similar expressive power as existing parallel languages, we implement several skeletons and 4 benchmarks. 
+To show that PArrows have similar expressive power as existing parallel languages, we implement several skeletons and four benchmarks. 
 Benchmarks show that our framework does not induce any notable performance overhead.

### 2018-01-06 paper/content/motivation.tex
minor things

@@
 
-\paragraph{Contributions}
+\paragraph{Contributions.}
 %
@@
 (Section~\ref{sec:parallel-arrows})
-\item We extend the PArrows formalism with |Future|s to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data (|RD|) \citet{Dieterle2010}.
+\item We extend the PArrows formalism with |Future|s to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data \citep[|RD|,][]{Dieterle2010}.
 Direct communication is useful in a distributed memory setting because it allows for inter-node communication without blocking the master-node.
@@
 \item We demonstrate the expressiveness of PArrows by using them to define common algorithmic skeletons ((Section~\ref{sec:skeletons}), and by using these skeletons to implement four benchmarks (Section~\ref{sec:benchmarks}).
-\item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg with mean of mean overhead of less than $3.5\%, 0.8\%$ for all benchmarks with GpH and Eden respectively.
-Compared to |Par| Monad we virtually noticed no overhead with a mean of mean overheads of less than $0\%$ in all benchmarks.
-(Section~\ref{sec:benchmarks}).
+\item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg with mean of mean overhead of less than $3.5\%$ and less than $0.8\%$ for all benchmarks with GpH and Eden, respectively.
+Compared to |Par| Monad we noticed no overhead: the mean of mean overheads was in our favour in all benchmarks (Section~\ref{sec:benchmarks}).
 \end{itemize}
 
-PArrows are open source and available from \url{https://github.com/s4ke/Parrows}.
+PArrows are open source and are available from \url{https://github.com/s4ke/Parrows}.

### 2018-01-06 paper/content/relwork.tex
minor things

@@
 Basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM).
-However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{eta-lang.org}},
+However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}},
 a new approach to Haskell on the JVM.

### 2018-01-06 paper/content/parallelHaskells.tex
cite parenthesis trick

@@
 There exist several parallel Haskells already.
-Among the most important are probably GpH  (based on |par| and |pseq| \enquote{hints}) \cite{Trinder1996,Trinder1998a}, the |Par| Monad (a monad for deterministic parallelism) \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden (a parallel Haskell for distributed memory) \cite{eden,Loogen2012}, HdpH (a Template Haskell based parallel Haskell for distributed memory) \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish (a |Par| extension with focus on communication) \cite{Kuper:2014:TPE:2666356.2594312}.
+Among the most important are probably GpH \citep[based on |par| and |pseq| \enquote{hints},][]{Trinder1996,Trinder1998a}, the |Par| Monad \citep[a monad for deterministic parallelism,][]{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \citep[a parallel Haskell for distributed memory,][]{eden,Loogen2012}, HdpH \citep[a Template Haskell-based parallel Haskell for distributed memory,][]{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish \citep[a |Par| extension with focus on communication,][]{Kuper:2014:TPE:2666356.2594312}.
 
@@
 
-LVish and HdpH were not chosen as the former does not differ from the original |Par| Monad with regard to how we would have used it in this paper, while the latter (at least in its current form) does not comply with our representation of parallelism due to its heavy reliance on TemplateHaskell.
+LVish and HdpH were not chosen as the former does not differ from the original |Par| Monad with regard to how we would have used it in this paper, while the latter (at least in its current form) does not comply with our representation of parallelism due to its heavy reliance on Template Haskell.
 
@@
 	\includegraphics[scale=0.5]{images/parEvalNMulticore}
-	\caption{Data flow of the GpH |parEvalN| version.}
+	\caption{|parEvalN| (GpH).}
 	\label{fig:parEvalNMulticoreImg}
@@
 	\includegraphics[scale=0.5]{images/parEvalNParMonad}
-	\caption{Data flow of the |Par| Monad |parEvalN| version.}
+	\caption{|parEvalN| (|Par| Monad).}
 	\label{fig:parEvalNParMonadImg}

### 2018-01-06 paper/content/main.tex
removed trailing s from paper name
letterspacing in title and other minor thingies

@@
 
-\title{Arrows for Parallel Computations}
+\title{Arrows for Parallel Computation}
 \ifthenelse{\boolean{anonymous}}{%
@@
 %\affiliation{University Bayreuth, Germany and Glasgow University, UK}
- \author[M. Braun, O. Lobachev and P. Trinder]%
-        {MARTIN BRAUN\\
+ \author[M. Braun, O. Lobachev, and P. Trinder]%
+        {\textls*{MARTIN BRAUN}\\
          University Bayreuth, 95440 Bayreuth, Germany\\
-		 OLEG LOBACHEV\\
+		 \textls*{OLEG LOBACHEV}\\
 		 University Bayreuth, 95440 Bayreuth, Germany\\
-		 \and\ PHIL TRINDER\\
+		 \and\ \textls*{PHIL TRINDER}\\
 		 Glasgow University, Glasgow, G12 8QQ, Scotland}
@@
 	\label{sec:background}
+	As a introduction, we here give a short overview over Arrows and GpH, the |Par| Monad, and Eden, the three parallel Haskells which we base our DSL on.
 	\input{arrows}

### 2018-01-06 paper/content/utilityFunctions.tex
sorting out references
shorter captions for figs

@@
 The |mapArr| combinator (Fig.~\ref{fig:mapArr}) lifts any Arrow |arr a
-b| to an Arrow |arr [a] [b]|. The original inspiration was from \citet{programming_with_arrows},
+b| to an Arrow |arr [a] [b]|. The original inspiration was from \citet{Hughes2005},
 but the definition as then unified with |evalN|.
@@
 \end{code}
-\caption{The |BasicFuture| type and its |Future| instance for the |Par| Monad and GpH Haskell backends.}
+\caption{|BasicFuture| type and its |Future| instance for the |Par| Monad and GpH Haskell backends.}
 \label{fig:BasicFuture}
@@
 \includegraphics[scale=0.7]{images/parMap}
-\caption{Schematic depiction of |parMap|.}
+\caption{|parMap| depiction.}
 \label{fig:parMapImg}
@@
 \includegraphics[scale=0.7]{images/parMapStream}
-\caption{Schematic depiction of |parMapStream|.}
+\caption{|parMapStream| depiction.}
 \label{fig:parMapStreamImg}
@@
 \end{code}
-\caption{Definition of |parMapStream|.}
+\caption{|parMapStream| definition.}
 \label{fig:parMapStream}
@@
 \end{code}
-\caption{Definitions of |shuffle|, |unshuffle|, |takeEach|.}
+\caption{|shuffle|, |unshuffle|, |takeEach| definition.}
 \label{fig:edenshuffleetc}
@@
 \end{code}
-\caption{Definitions of |lazy| and |rightRotate|.}
+\caption{|lazy| and |rightRotate| definitions.}
 \label{fig:edenlazyrightrotate}
@@
 \end{code}
-\caption{Definitions of |lazyzip3|, |uncurry3| and |threetotwo|.}
+\caption{|lazyzip3|, |uncurry3| and |threetotwo| definitions.}
 \label{fig:lazyzip3etc}
@@
 \end{code}
-\caption{Definition of |farmChunk|.}
+\caption{|farmChunk| definition.}
 \label{fig:farmChunk}
@@
 \end{code}
-	\caption{Definition of |parEval2|.}
+	\caption{|parEval2| definition.}
 	\label{fig:parEval2}
@@
 \end{code}
-	\caption{Definition of |prMMTr|, |splitMatrix| and |matAdd|.}
+	\caption{|prMMTr|, |splitMatrix| and |matAdd| definition.}
 	\label{fig:torus_example_rest}

### 2018-01-07 paper/content/arrows.tex
shorter captions for figs
minor wording

@@
 \vfill
-\caption{The definition of the |Arrow| type class and its two most typical instances.}
+\caption{The |Arrow| type class and its two most typical instances.}
 \label{fig:ArrowDefinition}
@@
 |second| is the mirrored version of |first| (Appendix~\ref{utilfns}).
-|***| combines |first| and |second| to handle two inputs in one arrow, and is defined as follows:
+The |***| function combines |first| and |second| to handle two inputs in one arrow, and is defined as follows:
 % \begin{figure}[h]
@@
 (&&&) :: Arrow arr => arr a b -> arr a c -> arr a (b, c)
-f &&& g = arr (\a -> (a, a)) >>> (f *** g)
+f &&& g = arr (\a -> (a, a)) >>> f *** g
 \end{code}
@@
 add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
-add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
+add f g = f &&& g >>> arr (\(u, v) -> u + v)
 \end{code}
@@
 As we can rewrite the monadic bind operation |(>>=)| with only the Kleisli type into |m a -> Kleisli m a b -> m b|, but not with a general Arrow |arr a b|, we can intuitively get an idea of why Arrows must be a generalisation of Monads.
-While this also means that a general Arrow can not express everything a Monad can, the Parser example shows that this trade-off is worth it in some cases.
+While this also means that a general Arrow can not express everything a Monad can, \citet{HughesArrows} shows in his parser example that this trade-off is worth it in some cases.
 
@@
 
-While we could have based our DSL on Profunctors as well, we chose Arrows \mbcomment{cite missing?}.
-for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
+While we could have based our DSL on Profunctors as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
 However, they are a promising candidate for future improvements of our DSL.

### 2018-01-07 paper/content/benchmarks.tex
every section has text now

@@
 This section evaluates the performance overhead of this compositional abstraction in comparison to  GpH and the |Par| Monad on shared memory architectures and Eden on  a distributed memory cluster.
-We start by explaining the hardware and software stack and outline the benchmark programs, and motivations for choosing them. Before we go into detail on the benchmarks we
-also shortly address hyper-threading and why we do not use it in our
-benchmarks.
-Our results show that PArrows hold up in terms of performance when compared to the original parallel Haskells, starting with the shared-memory variants (GpH, |Par| Monad and Eden~CP) and concluding with Eden in a distributed memory setting.
+We describe our measurement platform, the benchmark results -- the shared-memory variants (GpH, |Par| Monad and Eden~CP) followed by Eden in a distributed memory setting, and conclude that PArrows hold up in terms of performance when compared to the original parallel Haskells.
 
@@
 \subsection{Measurement Platform}
+We start by explaining the hardware and software stack and outline the benchmark programs and motivation for choosing them.
+We also shortly address hyper-threading and why we do not use it in our benchmarks.
 

### 2018-01-07 paper/content/futures.tex
shorter captions for figs
Note that removal

@@
 \end{code}
-\caption{Definition of the |Future| type class.}
+\caption{The |Future| type class.}
 \label{fig:future}
@@
 \end{code}
-\caption{The mock-up combinator in parallel.}
+\caption{The outline parallel combinator.}
 \label{fig:someCombinatorParallel}
@@
 We only specify that there has to be a compatible Future and do not care about any specifics as can be seen in Fig.~\ref{fig:someCombinatorParallel}.
-Note that with our PArrows DSL we can also define default instances |Future fut a ()| for each backend similar to how |ArrowParallel arr a b ()| was defined in Section~\ref{sec:parallel-arrows}.
+With the PArrows DSL we can also define default instances |Future fut a ()| for each backend similar to how |ArrowParallel arr a b ()| was defined in Section~\ref{sec:parallel-arrows}.
 Details can be found in Section~\ref{app:omitted}.

### 2018-01-07 paper/content/utilityFunctions.tex
Note that removal

@@
 This works because Arrows are strong Monads in the bicategory |Prof| of Profunctors as shown by \citet{Asada:2010:ASM:1863597.1863607}.
-Note that in Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| type class like presented in this paper.\footnote{For additional information on the type classes used, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}.}
+In Standard GHC |(>>>)| has the type |(>>>) :: Category cat => cat a b -> cat b c -> cat a c| and is therefore not part of the |Arrow| type class like presented in this paper.\footnote{For additional information on the type classes used, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}.}
 

### 2018-01-07 paper/content/main.tex
added short intro to section 3
typo

@@
 	\label{sec:background}
-	As a introduction, we here give a short overview over Arrows and GpH, the |Par| Monad, and Eden, the three parallel Haskells which we base our DSL on.
+	As an introduction, we here give a short overview over Arrows and GpH, the |Par| Monad, and Eden, the three parallel Haskells which we base our DSL on.
 	\input{arrows}

### 2018-01-07 paper/content/motivation.tex
wording

@@
 \item We demonstrate the expressiveness of PArrows by using them to define common algorithmic skeletons ((Section~\ref{sec:skeletons}), and by using these skeletons to implement four benchmarks (Section~\ref{sec:benchmarks}).
-\item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg with mean of mean overhead of less than $3.5\%$ and less than $0.8\%$ for all benchmarks with GpH and Eden, respectively.
-Compared to |Par| Monad we noticed no overhead: the mean of mean overheads was in our favour in all benchmarks (Section~\ref{sec:benchmarks}).
+\item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg the mean over all cores of relative mean overhead was less than $3.5\%$ and less than $0.8\%$ for all benchmarks with GpH and Eden, respectively.
+As for |Par| Monad, the mean of mean overheads was in our favour in all benchmarks (Section~\ref{sec:benchmarks}).
 \end{itemize}

### 2018-01-07 paper/content/main.tex
oops
that text in Sec.3

--- 2018-01-06 paper/content/main.tex
@@
 	\input{relwork}
-	\begin{flushright}
-
-	\end{flushright}\section{Background}
+	%\begin{flushright}
+	%
+	%\end{flushright}
+        \section{Background}
 	\label{sec:background}
-	As a introduction, we here give a short overview over Arrows and GpH, the |Par| Monad, and Eden, the three parallel Haskells which we base our DSL on.
+	This section gives a short overview of Arrows
+        (Section~\ref{sec:arrows}) and of GpH, the |Par| Monad, and
+        Eden, the three parallel Haskells which we base our DSL on
+        (Section~\ref{sec:parallelHaskells}).
         \input{arrows}

### 2018-01-07 ParMonad/src/main/Parrows/ParMonad.hs
removed NFData b from Par monad instance

@@ -61,7 +61,7 @@ defaultConf fs = stratToConf fs strict
 stratToConf :: [arr a b] -> Strategy b -> Conf b
 stratToConf _ strat = Conf strat
 
-instance (NFData b, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
+instance (ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
     parEvalN (Conf strat) fs = evalN (map (>>> arr strat) fs) >>>
                     arr sequenceA >>>
                     arr (>>= mapM Control.Monad.Par.get) >>>

### 2018-01-07 paper/content/parrows.tex
every section has text now
removed NFData b constraint from GpH and Par Monad instances in parrows.tex

@@
 \subsection{ArrowParallel instances}
-
+With the type class defined, we will now give implementations of it with GpH, the |Par| Monad and Eden.
 \subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
@@
 
-instance (NFData b, ArrowChoice arr) =>
+instance (ArrowChoice arr) =>
   ArrowParallel arr a b (Conf b) where
@@
 \begin{code}
-instance (NFData b, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
+instance (ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
     parEvalN (Conf strat) fs = evalN (map (>>> arr strat) fs) >>>

### 2018-01-07 paper/content/futures.tex
Figure fix for someCombinator

@@
 \section{Futures} \label{sec:futures}\label{futures}
-Consider an outline parallel Arrow combinator:
+\begin{figure}[t]
 \begin{code}
@@
 \end{code}
+\caption{The outline combinator.}
+\label{fig:someCombinator}
+\end{figure}
 
-In a distributed environment such a combinator first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
+Consider an outline parallel Arrow combinator in Fig.~\ref{fig:someCombinator}.
+In a distributed environment this first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
 Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.
@@
 In our communication example we can use this |Future| concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
-\begin{figure}[tbh]
+\begin{figure}[t]
 \begin{code}
-someCombinator :: (ArrowChoice arr, NFData b, NFData c,
+someCombinator :: (ArrowChoice arr,
 	ArrowParallel arr a (fut b) (), 
@@
 \end{code}
-\caption{The outline parallel combinator.}
+\caption{The outline combinator in parallel.}
 \label{fig:someCombinatorParallel}

### 2018-01-07 paper/content/parrows.tex
formatting in code listings

@@
 By combining the result of this Arrow with |arr sequenceA|, we get an Arrow |arr [a] (Par [IVar b])|.
-Then, in order to fetch the results of the different threads, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)| - our intermediary Arrow is of type |arr [a] (Par [b])|.
+Then, in order to fetch the results of the different threads, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)| -- our intermediary Arrow is of type |arr [a] (Par [b])|.
 Finally, we execute the computation |Par [b]| by composing with |arr runPar| and get the final Arrow |arr [a] [b]|.
@@
 instance (ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
-    parEvalN (Conf strat) fs = evalN (map (>>> arr strat) fs) >>>
+    parEvalN (Conf strat) fs =
+        evalN (map (>>> arr strat) fs) >>>
         arr sequenceA >>>
@@
   ArrowParallel (Kleisli m) a b conf where
-    parEvalN conf fs = arr (parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
+    parEvalN conf fs = 
+      arr (parEvalN conf (map (\(Kleisli f) -> f) fs)) >>>
       Kleisli sequence

### 2018-01-07 paper/content/futures.tex
minor
minor wording

@@
 
-Consider an outline parallel Arrow combinator in Fig.~\ref{fig:someCombinator}.
+Consider the outline parallel Arrow combinator in Fig.~\ref{fig:someCombinator}.
 In a distributed environment this first evaluates all |[arr a b]| in parallel, sends the results back to the master node, rotates the input once (in the example we require |ArrowChoice| for this) and then evaluates the |[arr b c]| in parallel to then gather the input once again on the master node.
@@
 	\caption[with Futures]{Communication between 4 Eden processes with Futures.
-Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is marked in the Figure) instead of always going through the master node (bottom bar).}
+Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is highlighted) instead of always going through the master node (bottom bar).}
 	\label{fig:withFutures}

### 2018-01-07 paper/content/abstract.tex
proofreading abstract

@@
 As each parallel computation is an Arrow, such parallel Arrows (PArrows) can be readily composed and transformed as such.
-To allow for more sophisticated communication schemes between computation nodes in distributed systems we utilise the concept of Futures to wrap direct communication.
+To allow for more sophisticated communication schemes between computation nodes in distributed systems, we utilise the concept of Futures to wrap direct communication.
 
-To show that PArrows have similar expressive power as existing parallel languages, we implement several skeletons and four benchmarks. 
+To show that PArrows have similar expressive power as existing parallel languages, we implement several algorithmic skeletons and four benchmarks. 
 Benchmarks show that our framework does not induce any notable performance overhead.
-We conclude that Arrows have considerable potential for composing parallel programs, and for producing programs that can execute on multiple parallel language implementations.
-
-\olcomment{Summarise conclusions}
-
-\mbcomment{Jedes Kapitel soll einmal ins Abstract. Conclusions sollen mit ins Abstract}
+We conclude that Arrows have considerable potential for composing parallel programs and for producing programs that can execute on multiple parallel language implementations.
+%
+% \olcomment{Summarise conclusions}
+%
+% \mbcomment{Jedes Kapitel soll einmal ins Abstract. Conclusions sollen mit ins Abstract}
 \end{abstract}

### 2018-01-07 paper/content/benchmarks.tex
headings consistent

@@
 
-\subsection{Measurement Platform}
+\subsection{Measurement platform}
 We start by explaining the hardware and software stack and outline the benchmark programs and motivation for choosing them.
@@
 
-\subsubsection{Hardware and Software}
+\subsubsection{Hardware and software}
 
@@
 
-\subsubsection{Distributed Memory}
+\subsubsection{Distributed memory}
 

### 2018-01-07 paper/content/parrows.tex
headings consistent

@@
 We present the |ArrowParallel| type class and explain the reasoning behind it before discussing some parallel Haskell implementations and basic extensions.
-\subsection{The ArrowParallel type class}
+\subsection{The |ArrowParallel| type class}
 A~parallel computation (on functions) can be seen as execution of some functions |a -> b| in parallel, as our |parEvalN| prototype shows (Section~\ref{sec:parEvalNIntro}).
@@
 
-\subsection{ArrowParallel instances}
+\subsection{|ArrowParallel| instances}
 With the type class defined, we will now give implementations of it with GpH, the |Par| Monad and Eden.

### 2018-01-07 paper/content/utilityFunctions.tex
headings consistent

@@
 
-\section{Omitted Function Definitions}
+\section{Additional function definitions}
 \label{app:omitted}
@@
 	\label{fig:torus_example_rest}
-\end{figure}
+\end{figure} %$
 
-\section{Syntactic Sugar} \label{syntacticSugar}
+\section{Syntactic sugar} \label{syntacticSugar}
 Finally, we also give the definitions for some syntactic sugar for PArrows, namely |parstar| and |parand|.

### 2018-01-07 paper/content/motivation.tex
intro part

@@
 |Par| Monad, and Eden, a distributed memory parallel Haskell. These
-languages represent orthogonal approaches. Some use a monad, even if
+languages represent orthogonal approaches. Some use a Monad, even if
 only for the internal representation. Some introduce additional
@@
 the parallel backends at will.
-The induced penalty was in the single-digit percent range, with means over the varying cores configuration typically under 2\% overhead in our measurements (Section~\ref{sec:benchmarks}).
+The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}).
 Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
@@
 This has many practical advantages.
-For example, during development we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant.
-(Section~\ref{sec:parallel-arrows})
+For example, during development we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant (Section~\ref{sec:parallel-arrows}).
 \item We extend the PArrows formalism with |Future|s to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data \citep[|RD|,][]{Dieterle2010}.
@@
 (Section~\ref{sec:futures})
-\item We demonstrate the expressiveness of PArrows by using them to define common algorithmic skeletons ((Section~\ref{sec:skeletons}), and by using these skeletons to implement four benchmarks (Section~\ref{sec:benchmarks}).
+\item We demonstrate the expressiveness of PArrows by using them to define common algorithmic skeletons (Section~\ref{sec:skeletons}), and by using these skeletons to implement four benchmarks (Section~\ref{sec:benchmarks}).
 \item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg the mean over all cores of relative mean overhead was less than $3.5\%$ and less than $0.8\%$ for all benchmarks with GpH and Eden, respectively.

### 2018-01-07 paper/content/relwork.tex
intro part

@@
 GpH (Trinder et al., 1996, 1998) parallel Haskell dialect and its multicore version (Marlow
-et al., 2009), the Par Monad (Marlow et al., 2011; Foltzer et al., 2012), and Eden (Loogen
+et al., 2009), the |Par| Monad (Marlow et al., 2011; Foltzer et al., 2012), and Eden (Loogen
 et al., 2005; Loogen, 2012). These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available (Aljabri et al., 2014,
@@
 
-Other task parallel Haskells related to Eden, GpH, and the Par Monad include the following.  
-HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of Par Monad to
+Other task parallel Haskells, related to Eden, GpH, and the |Par| Monad, include the following.  
+HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of |Par| Monad to
 heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
-of Par Monad.}
+of |Par| Monad.}
 
@@
 \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
-Typical types of algorithmic skeletons include |map|-, |fold|-, and |scan|-based parallel programming patterns, special applications such as divide-and-conquer or topological skeletons.
+Types of algorithmic skeletons include |map|-, |fold|-, and |scan|-based parallel programming patterns, special applications such as divide-and-conquer or topological skeletons.
 

### 2018-01-07 paper/content/benchmarks.tex
|Par| everywhere, wording in topology

@@
 \paragraph{Speedup.}
-The \rmtest benchmark showed almost linear speedup for both 32 and 64 tasks, the performance is slightly better in the latter case: 13.7 at 16 cores for input $2^{11213}-1$ and 64 tasks in the best case scenario with Eden CP.
-The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (Par Monad), 8.14 (Eden CP) for 16 cores and 1000 Sudokus.
+The \rmtest benchmark showed almost linear speedup for both 32 and 64 tasks, the performance is slightly better in the latter case: 13.7 at 16 cores for input $2^{11213}-1$ and 64 tasks in the best case scenario with Eden~CP.
+The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (|Par| Monad), 8.14 (Eden~CP) for 16 cores and 1000 Sudokus.
 In contrast to Rabin--Miller, here the |GpH| backend seems to be the best of all, while Rabin--Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.

### 2018-01-07 paper/content/parallelHaskells.tex
|Par| everywhere, wording in topology

@@
 
-\subsubsection{Par Monad}
+\subsubsection{|Par| Monad}
 The |Par| Monad\footnote{The |Par| monad can be found in the \texttt{monad-par} package on Hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs.

### 2018-01-07 paper/content/topologySkeletons.tex
formatting in code listings
headings consistent
|Par| everywhere, wording in topology

--- 2018-01-06 paper/content/topologySkeletons.tex
@@
 %\FloatBarrier
-\subsection{Topological Skeletons}
+\subsection{Topological skeletons}
 \label{sec:topology-skeletons}
@@
 
-If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and Par Monad backends which during runtime would result in the program crashing.
+If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the GpH and |Par| Monad which during runtime would result in the program crash.
 This materialises with the usage of |loop| of the |ArrowLoop| type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden.
 An investigation of why this difference exists is beyond the scope of this work, we only provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment.
-However our workaround enables users of the DSL to test their code within a shared memory setting even though other skeletons would be better suited to be run with them.
-The idea of the fix is to provide a |ArrowLoopParallel| type class that has two functions - |loopParEvalN| and |postLoopParEvalN| where the first is to be used inside an |loop| construct while the latter will be used right outside of the |loop|.
+However our workaround enables users of the DSL to test their code within a shared memory setting.
+% even though other skeletons would be better suited to be run with them. --- OL: I don't get this sentence part
+The idea of the fix is to provide a |ArrowLoopParallel| type class that has two functions -- |loopParEvalN| and |postLoopParEvalN|, where the first is to be used inside an |loop| construct while the latter will be used right outside of the |loop|.
 This way we can delegate to the actual |parEvalN| in the spot where the backend supports it.
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
-	\caption[Matrix multiplication with |torus| (PArrows)]{Matrix Multiplication with |torus| (PArrows).}
+	\caption[Matrix multiplication with |torus| (PArrows)]{Matrix multiplication with |torus| (PArrows).}
 	\label{fig:torus_parrows_trace}
@@
 	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
-	\caption[Matrix multiplication with |torus| (Eden)]{Matrix Multiplication with |torus| (Eden).}
+	\caption[Matrix multiplication with |torus| (Eden)]{Matrix multiplication with |torus| (Eden).}
 	\label{fig:torus_eden_trace}

### 2018-01-07 paper/content/benchmarks.tex
proof-reading benchmarks section

@@
 Stack itself was used in version 1.3.2.
-For GpH in its Multicore variant we used the the parallel package in version 3.2.1.0\footnote{see \url{https://hackage.haskell.org/package/parallel-3.2.1.0}}, while for the |Par| Monad we used monad-par in version 0.3.4.8\footnote{see \url{https://hackage.haskell.org/package/monad-par-0.3.4.8}}.
+For GpH in its Multicore variant we used the parallel package in version 3.2.1.0\footnote{see \url{https://hackage.haskell.org/package/parallel-3.2.1.0}}, while for the |Par| Monad we used monad-par in version 0.3.4.8\footnote{see \url{https://hackage.haskell.org/package/monad-par-0.3.4.8}}.
 For all Eden tests, we used its GHC-Eden compiler in version 7.8.2\footnote{see \url{http://www.mathematik.uni-marburg.de/~eden/?content=build_eden_7_&navi=build}} together with OpenMPI 1.6.5\footnote{see \url{https://www.open-mpi.org/software/ompi/v1.6/}}.
@@
 Furthermore, all benchmarks were done with help of the bench\footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2)\footnote{see \url{https://hackage.haskell.org/package/criterion-1.1.1.0}} internally.
-All runtime data (mean runtime, max stddev, etc.) was collected with this tool if not mentioned otherwise.
+All runtime data (mean runtime, max stddev, etc.) was collected with this tool.
+% if not mentioned otherwise. -- OL: nowhere is it otherwise
 
@@
 for their \ensuremath{\Varid{torus}} skeleton. It implements a Gentleman's algorithm for parallel matrix
-multiplication \citep{Gentleman1978}. We ported an Eden based version \citep{Eden:SkeletonBookChapter02} to PArrows.
+multiplication \citep{Gentleman1978}. We ported an Eden-based version \citep{Eden:SkeletonBookChapter02} to PArrows.
 
 A~parallel Sudoku solver was used by \citet{par-monad} to compare \ensuremath{\Conid{Par}} Monad
-to GpH, and we ported it to PArrows.
+to GpH, we ported it to PArrows.
 
@@
 setting. However, for shared memory Eden features a \enquote{CP} backend
-that merely copies the memory blocks between distributed heaps. In
+that merely copies the memory blocks between disjoint heaps. In
 this mode, Eden still operates in the \enquote{nothing shared} setting, but
@@
 
-We compare the mean overhead, \ie the mean of relative differences between the PArrow and direct benchmark implementations executed multiple times with the same settings.
+We compare the mean overhead, \ie the mean of relative wall-clock run time differences between the PArrow and direct benchmark implementations executed multiple times with the same settings.
 The error margins of the time measurements, supplied by criterion package\footnote{\url{https://hackage.haskell.org/package/criterion-1.1.1.0}}, yield the error margin of the mean overhead.
@@
 Quite often the zero value lies in the error margin of the mean overhead.
-This means that even though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the benchmarks and the difference might not be existent.
+This means that even though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the measurement and the difference might not be existent.
 We are mostly interested in the cases where above issue does not persist, we call them \emph{significant}.
@@
 We focus now on significant overhead values.
-To name a few: $0.41\%\; \pm 7\cdot 10^{-2}\%$ for Eden~CP and 64 tasks at 4 cores, $4.7\% \; \pm 0.72\%$ for GpH, 32 tasks, 8 cores, $0.34\% \; \pm 0.31\%$ for |Par| Monad at 4 cores with 64 tasks.
+To name a few: $0.41\%\; \pm 7\cdot 10^{-2}\%$ for Eden~CP and 64 tasks at 4 cores; $4.7\% \; \pm 0.72\%$ for GpH, 32 tasks, 8 cores; $0.34\% \; \pm 0.31\%$ for |Par| Monad at 4 cores with 64 tasks.
 The worst significant overhead was  GpH backend with $8\% \; \pm 6.9\%$ at 16 cores with 32 tasks and input value $2^{11213}-1$.
@@
 For Sudoku the situation is slightly different.
-There is a minimal significant ($-1.4\% \; \pm 1.2\%$ at 8 cores) speed improvement with PArrows Eden~CP version when compared with the base Eden CP benchmark.
+There is a minimal significant ($-1.4\% \; \pm 1.2\%$ at 8 cores) speed \emph{improvement} with PArrows Eden~CP version when compared with the base Eden CP benchmark.
 However, with increasing number of cores the error margin reaches zero again: $-1.6\% \; \pm 5.0\%$ at 16 cores.
 The |Par| Monad shows a similar development, \eg with $-1.95\% \; \pm 0.64\%$ at 8 cores.
-The GpH version shows both a significant speed \textit{improvement} from PArrows of $-4.2\% \; \pm 0.26\%$ (for 16 cores) and a minor overhead of $0.87\% \; \pm 0.70\%$ (4 cores).
+The GpH version shows both a significant speed improvement of $-4.2\% \; \pm 0.26\%$ (for 16 cores) with PArrows and a minor overhead of $0.87\% \; \pm 0.70\%$ (4 cores).
 

### 2018-01-07 paper/content/conclusion.tex
headings consistent
proofread conclusions

@@
 We argue that Arrows are better suited to  parallelise
-pure functions than monads,  as the functions are already Arrows and can be used
+pure functions than Monads,  as the functions are already Arrows and can be used
 directly in our DSL.
@@
 %
-The DSL allows for a direct parallelisation of monadic code via the Kleisli type and additionally allows to parallelise any Arrow type that has an instance for |ArrowChoice| (but some skeletons require an additional |ArrowLoop| instance).
+The DSL allows for a direct parallelisation of monadic code via the Kleisli type and additionally allows to parallelise any Arrow type that has an instance for |ArrowChoice|.
+(Some skeletons require an additional |ArrowLoop| instance.)
 
 We have demonstrated the generality of the approach by exhibiting PArrow implementations for Eden, GpH, and the |Par| Monad.
-Hence, paralle programs can be ported between task parallel Haskell implementations with little or no effort.
+Hence, parallel programs can be ported between task parallel Haskell implementations with little or no effort.
 We are confident that it will be straightforward to add other task-parallel backends.
@@
 
-\subsection{Future Work}
+\subsection{Future work}
 \label{sec:future-work}

### 2018-01-07 paper/content/benchmarks.tex
Phil's comment: backends

@@
 cores -- the speedup did not increase that well. It was merely 15.99
-for 32 tasks with PArrows/Eden. This was worse for other backends.  As
+for 32 tasks with PArrows/Eden. This was worse for other implementations.  As
 for 64 tasks, we obtained a speedup of 16.12 with PArrows/Eden at 32
@@
 We start with the definition of mean overhead to compare both PArrows-enabled and standard benchmark implementations.
-We continue comparing speedups and overheads for the shared memory backends and then study OpenMPI variants of the Eden-enabled backend as a representative of a distributed memory backend.
+We continue comparing speedups and overheads for the shared memory implementations and then study OpenMPI variants of the Eden-enabled PArrows as a representative of a distributed memory backend.
 We plot all speedup curves and all overhead values in the supplementary materials.
@@
 The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (|Par| Monad), 8.14 (Eden~CP) for 16 cores and 1000 Sudokus.
-In contrast to Rabin--Miller, here the |GpH| backend seems to be the best of all, while Rabin--Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
+In contrast to Rabin--Miller, here the |GpH| seems to be the best of all, while Rabin--Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
 Gentleman on shared memory has a plummeting speedup curve with GpH and |Par| Monad and logarithmically increasing speedup for the Eden-based version.
@@
 
-For the shared memory \rmtest benchmark, implemented with PArrows using Eden CP, GpH, and |Par| Monad backends, the overhead values are within single percents range, but also negative overhead (\ie PArrows are better) and larger error margins happen.
+For the shared memory \rmtest benchmark, implemented with PArrows using Eden~CP, GpH, and |Par| Monad, the overhead values are within single percents range, but also negative overhead (\ie PArrows are better) and larger error margins happen.
 To give a few examples, the overhead for Eden~CP with input value $2^{11213}-1$, 32 tasks, and 16 cores is $1.5\%$, but the error margin is around $5.2\%$!
-Same backend in the same setting with 64 tasks reaches $-0.2\%$ overhead, PArrows apparently fare better than Eden -- but the error margin of $1.9\%$ disallows this interpretation.
+Same implementation in the same setting with 64 tasks reaches $-0.2\%$ overhead, PArrows apparently fare better than Eden -- but the error margin of $1.9\%$ disallows this interpretation.
 We focus now on significant overhead values.
 To name a few: $0.41\%\; \pm 7\cdot 10^{-2}\%$ for Eden~CP and 64 tasks at 4 cores; $4.7\% \; \pm 0.72\%$ for GpH, 32 tasks, 8 cores; $0.34\% \; \pm 0.31\%$ for |Par| Monad at 4 cores with 64 tasks.
-The worst significant overhead was  GpH backend with $8\% \; \pm 6.9\%$ at 16 cores with 32 tasks and input value $2^{11213}-1$.
+The worst significant overhead was in case of GpH  with $8\% \; \pm 6.9\%$ at 16 cores with 32 tasks and input value $2^{11213}-1$.
 In other words, we notice no major slow-down through PArrows here.
@@
 \paragraph{Speedup.}
-The speedup of distributed memory Rabin--Miller benchmark with PArrows and distributed Eden backend showed an almost 
+The speedup of distributed memory Rabin--Miller benchmark with PArrows and Eden showed an almost 
 linear speedup excepting around 192 cores where an unfortunate task distribution reduces performance.
@@
 	}{272}{\plotwidthDist}{272}
-	\caption[Speedup distributed Rabin--Miller]{Speedup of the distributed \rmtest benchmark using PArrows with Eden backend.}
+	\caption[Speedup distributed Rabin--Miller]{Speedup of the distributed \rmtest benchmark using PArrows with Eden.}
 	\label{fig:rabinMillerDistSpeedup}

### 2018-01-07 paper/content/motivation.tex
Phil's comment: backends

@@
 common, very low-penalty programming interface that allows to switch
-the parallel backends at will.
+the parallel implementations at will.
 The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}).
-Further backends based on HdpH or a Frege implementation (on the Java Virtual Machine) are viable, too.
+Further implementations, based on HdpH or a Frege implementation (on the Java Virtual Machine), are viable, too.
 
@@
 We do not re-implement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
-The backends can easily be swapped, so we are not bound to any specific one.
+The implementations can easily be swapped, so we are not bound to any specific one.
 
 This has many practical advantages.
-For example, during development we can run the program in a simple GHC-compiled variant using a GpH backend and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant (Section~\ref{sec:parallel-arrows}).
+For example, during development we can run the program in a simple GHC-compiled variant using GpH and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the |ArrowParallel| instance and compiling with Eden's GHC variant (Section~\ref{sec:parallel-arrows}).
 \item We extend the PArrows formalism with |Future|s to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data \citep[|RD|,][]{Dieterle2010}.

### 2018-01-07 paper/content/topologySkeletons.tex
Phil's comment: backends

@@
 \end{code}
-As the backends for the |Par| Monad and GpH have problems with |parEvalN| inside of |loop| their respective instances for |ArrowLoopParallel| look like this:
+As |Par| Monad and GpH have problems with |parEvalN| inside of |loop| their respective instances for |ArrowLoopParallel| look like this:
 \begin{code}
@@
 
-Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to |parEvalN| inside of |evalN|.
+Extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to |parEvalN| inside of |evalN|.
 Nonetheless, we can define a parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.

### 2018-01-07 paper/content/basicSkeletons.tex
headings consistent
minor

@@
-\subsection{Extending the Interface}
+\subsection{Extending the interface}
 \label{sec:extending-interface}
@@
 	\includegraphics[scale=0.7]{images/parEvalNLazy}
-	\caption{Schematic depiction of |parEvalNLazy|.}
+	\caption{|parEvalNLazy| depiction.}
 	\label{fig:parEvalNLazyImg}
 \end{figure}
-|parEvalN| fully traverses the list of passed Arrows as well as their inputs.
-Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g.
-|map (arr .
+The function |parEvalN| fully traverses the list of passed Arrows as well as their inputs.
+Sometimes this might not be feasible, as it will not work on infinite lists of functions like \eg |map (arr .
 (+)) [1..]| or just because we need the Arrows evaluated in chunks.
@@
 	\includegraphics[scale=0.7]{images/parEval2}
-	\caption{Schematic depiction of |parEval2|.}
+	\caption{|parEval2| depiction.}
 	\label{fig:parEval2Img}

### 2018-01-07 paper/content/benchmarks.tex
i.e.

@@
 The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (|Par| Monad), 8.14 (Eden~CP) for 16 cores and 1000 Sudokus.
-In contrast to Rabin--Miller, here the |GpH| seems to be the best of all, while Rabin--Miller profited most from Eden CP (i.e., Eden with direct memory copy) implementation of PArrows.
+In contrast to Rabin--Miller, here the |GpH| seems to be the best of all, while Rabin--Miller profited most from Eden~CP (\ie Eden with direct memory copy) implementation of PArrows.
 Gentleman on shared memory has a plummeting speedup curve with GpH and |Par| Monad and logarithmically increasing speedup for the Eden-based version.

### 2018-01-07 paper/content/conclusion.tex
Phil's comment: backends
spellcheck

@@
 % while still having a generic and composable interface.
-We use a non-monadic interface (similar to Eden or GpH) and retain composibility.
+We use a non-monadic interface (similar to Eden or GpH) and retain composability.
 %
@@
 Hence, parallel programs can be ported between task parallel Haskell implementations with little or no effort.
-We are confident that it will be straightforward to add other task-parallel backends.
+We are confident that it will be straightforward to add other task-parallel Haskells.
 % Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden.

### 2018-01-07 paper/content/ms.tex
getting ready for arXiv

% This is our submission, modified from:
% the file JFP2egui.lhs
% release v1.02, 27th September 2001
%   (based on JFPguide.lhs v1.11 for LaLhs 2.09)
% Copyright (C) 2001 Cambridge University Press

\NeedsTeXFormat{LaTeX2e}

\documentclass{jfp1}

%% add a tweak to polycode, similar to literate in lstlistings, but uglier
%% ^^^^ you want this. and similar for ***, +++, etc.
%% a hack for list comprehensions' <-, typeset it as <--
%%%%%format <--        = "\in "
%%%% %format ` = "\`"
%%%% %% works fine without

%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%

%%% Macros for the guide only %%%
%\providecommand\AMSLaTeX{AMS\,\LaTeX}
%\newcommand\eg{\emph{e.g.}\ }
%\newcommand\etc{\emph{etc.}}
%\newcommand\bcmdtab{\noindent\bgroup\tabcolsep=0pt%
%  \begin{tabular}{@{}p{10pc}@{}p{20pc}@{}}}
%\newcommand\ecmdtab{\end{tabular}\egroup}
%\newcommand\rch[1]{$\longrightarrow\rlap{$#1$}$\hspace{1em}}
%\newcommand\lra{\ensuremath{\quad\longrightarrow\quad}}

\jdate{January 2018}
\pubyear{2018}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\doi{...}

%\newtheorem{lemma}{Lemma}[section]

\usepackage{tikz}

\usepackage{filecontents}
\usepackage{pgfplots, pgfplotstable}
\usepgfplotslibrary{statistics}

\usepackage{makecell}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

%\usepackage{color}
\usepackage{hyperref}

%\usepackage{subcaption}

\usepackage[round,sectionbib]{natbib}
%% https://tex.stackexchange.com/questions/148992/changing-the-heading-style-of-references-section
%% and jfp1.cls
\renewcommand\refname{References}
\makeatletter
\renewcommand\bibsection{%
  \section*{\refname}%
}%
\makeatother

\usepackage{csquotes}

% \usepackage{placeins} %% unneeded

\usepackage{microtype}
\DisableLigatures[>,<]{encoding = T1,family=tt*}

% für Listings
%\usepackage{listings}

\renewcommand{\cite}[1]{\citep{#1}}

\newcommand{\citHughes}{\citep{HughesArrows}}

%\newcommand{\inlinecode}[1]{\texttt{#1}}
\newcommand{\inlinecode}[1]{\emph{#1}} %%% booo
%%% |code| is inline code in lhs2tex

%\newcommand{\fixme}[1]{\colorbox{red}{#1}}

%%%%%% more fun and typography

\usepackage{xifthen}

\newboolean{anonymous}
\setboolean{anonymous}{False}

\usepackage[british]{babel}

\usepackage{siunitx}
\sisetup{detect-weight=true,detect-inline-weight=math}

\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

%%\usepackage{tipa} %% causes problems with lhs2tex?!

\usepackage{xspace}
\usepackage{xcolor}
\newcommand{\hl}[1]{\textcolor{red}{#1}}
%\newcommand{\comm}[2]{\textcolor{red}{\bfseries #1: #2}}
\newcommand{\comm}[2]{}
\newcommand{\olcomment}[1]{\comm{OL}{#1}}
\newcommand{\mbcomment}[1]{\comm{MB}{#1}}
\newcommand{\ptcomment}[1]{\comm{Phil}{#1}}
%\newcommand{\done}{\xspace\hl{done!}\xspace}
\newcommand{\fixme}{\mbcomment}

%\renewcommand{\olcomment}[1]{}
%\renewcommand{\mbcomment}[1]{}

\DeclareRobustCommand{\xth}{\textsuperscript{th}\xspace}
\DeclareRobustCommand{\st}{\textsuperscript{st}\xspace}
\DeclareRobustCommand{\nd}{\textsuperscript{nd}\xspace}
\DeclareRobustCommand{\xrd}{\textsuperscript{rd}\xspace}

%% kuerzel
%\newcommand{\todo}{\textcolor{red}{\bfseries TODO!}\xspace}
\newcommand{\done}{\textcolor{green}{\bfseries done!}\xspace}
\DeclareRobustCommand{\hairspn}{\hspace{1pt}\nolinebreak}% hair space with no break

\newcommand{\seconds}{\ensuremath{s}}

% \DeclareRobustCommand{\ie}{{i.\hairspn{}e.\nopagebreak[4] }}
% \DeclareRobustCommand{\eg}{{e.\hairspn{}g.\nopagebreak[4] }}
% \DeclareRobustCommand{\fe}{{f.\hairspn{}e.,}\nopagebreak[4] }
\DeclareRobustCommand{\ie}{{i.\hairspn{}e.~}}
\DeclareRobustCommand{\eg}{{e.\hairspn{}g.~}}
\DeclareRobustCommand{\fe}{{f.\hairspn{}e.,~}}
\DeclareRobustCommand{\ad}{{A.\hairspn{}D.\xspace}}
\DeclareRobustCommand{\bcc}{{A.\hairspn{}C.\xspace}}
\DeclareRobustCommand{\wrt}{w.\hairspn{}r.\hairspn{}t.~}
\DeclareRobustCommand{\etc}{etc.\ }
\DeclareRobustCommand{\cf}{\textit{cf.~}}
\DeclareRobustCommand{\viz}{\textit{viz.~}}
%\DeclareRobustCommand{\hof}{higher-or\-der function\xspace}

% %% the numberings
\DeclareRobustCommand{\xth}{\textsuperscript{th}\xspace}
\DeclareRobustCommand{\st}{\textsuperscript{st}\xspace}
\DeclareRobustCommand{\nd}{\textsuperscript{nd}\xspace}
\DeclareRobustCommand{\xrd}{\textsuperscript{rd}\xspace}

\newcommand{\tabsepbakup}{\tabcolsep}
%% for narrower tables

%% nice tables
\usepackage{booktabs}
\usepackage{multirow}

\usepackage{csvsimple}

\usepackage{tablefootnote}

%% more microtype
\microtypecontext{spacing=nonfrench} %% log said so

\usepackage{subfig}

%% fight for pipepipepipe
\newcommand{\pipepipepipe}{\ensuremath{\mathbin{\mid\!\mid\!\mid}}\xspace}
%%% ^^^^ keep consistent with definition at the top of main.lhs
%%%\newcommand{\pipepipepipe}{\inlinecode{|||}\xspace}

%%% JFP requirements:
%%% Harvard citing style, "(Curry 1933)".
%%% code: identifies italic, keywords bold
%%% figures: eps(!) or ps (can convert pdf), eventually provide a greyscale version, color NOT in cmyk

\title{Arrows for Parallel Computation}
\ifthenelse{\boolean{anonymous}}{%
\author{Submission ID xxxxxx}
}{%
%\author{Martin Braun, Phil Trinder, and Oleg Lobachev}
%\affiliation{University Bayreuth, Germany and Glasgow University, UK}
 \author[M. Braun, O. Lobachev, and P. Trinder]%
        {\textls*{MARTIN BRAUN}\\
         University Bayreuth, 95440 Bayreuth, Germany\\
		 \textls*{OLEG LOBACHEV}\\
		 University Bayreuth, 95440 Bayreuth, Germany\\
		 \and\ \textls*{PHIL TRINDER}\\
		 Glasgow University, Glasgow, G12 8QQ, Scotland}
}% end ifthenelse

\begin{document}

\label{firstpage}

\def\SymbReg{\textsuperscript{\textregistered}}

\maketitle

%% environment inside
\begin{abstract}
Arrows are a general interface for computation and an alternative to Monads for API design. In contrast to Monad-based parallelism, we explore the use of Arrows for specifying generalised parallelism. Specifically, we define an Arrow-based language and implement it using multiple parallel Haskells.

As each parallel computation is an Arrow, such parallel Arrows (PArrows) can be readily composed and transformed as such.
To allow for more sophisticated communication schemes between computation nodes in distributed systems, we utilise the concept of Futures to wrap direct communication.

To show that PArrows have similar expressive power as existing parallel languages, we implement several algorithmic skeletons and four benchmarks. 
Benchmarks show that our framework does not induce any notable performance overhead. We conclude that Arrows have considerable potential for composing parallel programs and for producing programs that can execute on multiple parallel language implementations. 
%
% \olcomment{Summarise conclusions}
%
% \mbcomment{Jedes Kapitel soll einmal ins Abstract. Conclusions sollen mit ins Abstract}
\end{abstract}

\tableofcontents

	%
	%%---include abstract.lhs
	%
	%\newpage
	%\pagebreak
	%\section{Motivation}
%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to Monads for API design \citHughes. In the paper Hughes describes how Arrows are a generalization of Monads and how they are not as restrictive. In this paper we will use this concept to express parallelism.

\section{Introduction}
\label{sec:introduction}

Parallel functional languages have a long history of being used for experimenting with novel parallel programming paradigms. Haskell, which we focus on in this paper, has  several mature implementations. We regard here in-depth
Glasgow parallel Haskell or short GpH (its Multicore SMP implementation, in particular), the
\ensuremath{\Conid{Par}} Monad, and Eden, a distributed memory parallel Haskell. These
languages represent orthogonal approaches. Some use a Monad, even if
only for the internal representation. Some introduce additional
language constructs. Section \ref{sec:parallelHaskells} gives a short overview over these languages.

A key novelty in this paper is to use Arrows to represent parallel computations. They seem a natural fit as they can be thought of as a more general function arrow (\ensuremath{\to }) and serve as general interface to computations while not being as restrictive as Monads \citep{HughesArrows}. Section \ref{sec:arrows} gives a short introduction to Arrows.

We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
Instead of 
introducing a new low-level parallel backend to implement our
Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
is defined as a common interface with varying implementations in
the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
novel manner -- we tame the zoo of parallel Haskells. We provide a
common, very low-penalty programming interface that allows to switch
the parallel implementations at will. The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}). Further implementations, based on HdpH or a Frege implementation (on the Java Virtual Machine), are viable, too.

\paragraph{Contributions.}
%
%\olcomment{HIT HERE REALLY STRONG}
%
%\subsection{Impact of parallel Arrows}
%\olcomment{move this to Contributions in the front or something}
We propose an Arrow-based encoding for parallelism based on a new Arrow combinator \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}. A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version. In this paper we evaluate the expressive power of such a formalism in the context of parallel programming.

\begin{itemize}
\item We introduce a parallel evaluation formalism using Arrows. One big advantage of our specific approach is that we do not have to introduce any new types, facilitating composability (Section~\ref{sec:parallel-arrows}).
\item We show that PArrow programs can readily exploit multiple parallel language implementations. We demonstrate the use of GpH, a \ensuremath{\Conid{Par}} Monad, and Eden. We do not re-implement all the parallel internals, as we host this functionality in the \ensuremath{\Conid{ArrowParallel}} type class, which abstracts all parallel implementation logic. The implementations can easily be swapped, so we are not bound to any specific one.

This has many practical advantages. For example, during development we can run the program in a simple GHC-compiled variant using GpH and afterwards deploy it on a cluster by converting it into an Eden program, by just replacing the \ensuremath{\Conid{ArrowParallel}} instance and compiling with Eden's GHC variant (Section~\ref{sec:parallel-arrows}).
\item We extend the PArrows formalism with \ensuremath{\Conid{Future}}s to enable direct communication of data between nodes in a distributed memory setting similar to Eden's Remote Data \citep[\ensuremath{\Conid{RD}},][]{Dieterle2010}. Direct communication is useful in a distributed memory setting because it allows for inter-node communication without blocking the master-node. (Section~\ref{sec:futures})
\item We demonstrate the expressiveness of PArrows by using them to define common algorithmic skeletons (Section~\ref{sec:skeletons}), and by using these skeletons to implement four benchmarks (Section~\ref{sec:benchmarks}).
\item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg the mean over all cores of relative mean overhead was less than $3.5\%$ and less than $0.8\%$ for all benchmarks with GpH and Eden, respectively. As for \ensuremath{\Conid{Par}} Monad, the mean of mean overheads was in our favour in all benchmarks (Section~\ref{sec:benchmarks}).
\end{itemize}

PArrows are open source and are available from \url{https://github.com/s4ke/Parrows}.
	\section{Related Work}
\label{sec:related-work}

\ptcomment{The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations. A range of approaches have been explored, including data parallelism (Chakravarty et al., 2007; Keller et al.,
2010), GPU-based approaches (Mainland & Morrisett, 2010; Svensson, 2011), software
transactional memory (Harris et al., 2005; Perfumo et al., 2008). The Haskell–GPU bridge
Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
deserves a special mention. <<Why does it deserve special mention?>>  Accelerate is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.

Our PArrow implementation uses three task parallel languages as backends: the
GpH (Trinder et al., 1996, 1998) parallel Haskell dialect and its multicore version (Marlow
et al., 2009), the \ensuremath{\Conid{Par}} Monad (Marlow et al., 2011; Foltzer et al., 2012), and Eden (Loogen
et al., 2005; Loogen, 2012). These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available (Aljabri et al., 2014,
2015). Research on Eden includes a low-level  implementation (Berthold, 2008; Berthold et al., 2016), work on skeleton
composition (Dieterle et al., 2016), communication (Dieterle et al., 2010a), and generation
of process networks (Horstmeyer & Loogen, 2013). The definitions of new Eden skeletons is a specific focus (Hammond et al., 2003; Berthold
& Loogen, 2006; Berthold et al., 2009b,c; Dieterle et al., 2010b; de la Encina et al., 2011;
Dieterle et al., 2013; Janjic et al., 2013).

Other task parallel Haskells, related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad, include the following.  
HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of \ensuremath{\Conid{Par}} Monad to
heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
of \ensuremath{\Conid{Par}} Monad.}

\paragraph{Parallel Haskells.}
The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations. A range of approaches have been explored, including data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
The Haskell--GPU bridge Accelerate
\cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313}
is completely orthogonal to our approach.
A~good survey of parallel Haskells can be found in~\citet{marlow2013parallel}.

Our PArrow implementation uses three task parallel languages as backends: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012}. These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available \cite{Aljabri:2013:DIG:2620678.2620682,Aljabri2015}.
Research on Eden includes low-level  implementation
\cite{JostThesis,berthold_loidl_hammond_2016}, skeleton composition
\cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013}. The definitions of new Eden skeletons is a specific focus \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.

Other task parallel Haskells related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad include the following.  
HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} Monad to
heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension
of \ensuremath{\Conid{Par}} Monad.

\olcomment{OLD:
\paragraph{Parallel Haskells.}
Of course, the three parallel Haskell flavours we use as backends: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work. We use these languages as backends: our DSL can switch from one to another at user's command.

HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} Monad to heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \ensuremath{\Conid{Par}} Monad.
%
Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015}, and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden. This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.

Alternative approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
%
The Haskell--GPU bridge Accelerate
\cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313}
deserves a special mention. Accelerate is completely orthogonal to our
approach. \citeauthor{marlow2013parallel} authored a recent book in
\citeyear{marlow2013parallel} on parallel Haskells.
}

\paragraph{Algorithmic skeletons.}
Algorithmic skeletons were introduced by \citet{Cole1989}.
Early publications on this topic include \cite{DANELUTTO1992205,darlington1993parallel,botorog1996efficient,Lengauer1997,Gorlatch1998}. \citet{SkeletonBook} consolidated early reports on high-level programming approaches.
Types of algorithmic skeletons include \ensuremath{\Varid{map}}-, \ensuremath{\Varid{fold}}-, and \ensuremath{\Varid{scan}}-based parallel programming patterns, special applications such as divide-and-conquer or topological skeletons.

The \ensuremath{\Varid{farm}} skeleton \citep{Hey1990185,Eden:PPDP01,Kuchen05} is a statically task-balanced parallel \ensuremath{\Varid{map}}. When tasks' durations cannot be foreseen, a dynamic load balancing (\ensuremath{\Varid{workpool}}) brings a lot of improvement \citep{Rudolph:1991:SLB:113379.113401,doi:10.1142/S0129626403001380,Hippold2006,PADL08HMWS,Marlow2009}. For special tasks \ensuremath{\Varid{workpool}} skeletons can be extended with dynamic task creation \cite{WPEuropar06,Dinan:2009:SWS:1654059.1654113,brown2010ever}. Efficient load-balancing schemes for \ensuremath{\Varid{workpool}}s are subject of research \cite{Blumofe:1999:SMC:324133.324234,Acar:2000:DLW:341800.341801,vanNieuwpoort:2001:ELB:568014.379563,Chase:2005:DCW:1073970.1073974,4625841,Michael:2009:IWS:1594835.1504186}.
%
The \ensuremath{\Varid{fold}} (or \ensuremath{\Varid{reduce}}) skeleton was implemented in various skeleton libraries \cite{Kuchen2002,5361825,BUONO20102095,Dastgeer:2011:ASM:1984693.1984697}, as also its inverse, \ensuremath{\Varid{scan}} \cite{Bischof2002,harris2007parallel}.
%
Google \ensuremath{\Varid{map}}--\ensuremath{\Varid{reduce}} \cite{Dean:2008:MSD:1327452.1327492,Dean:2010:MFD:1629175.1629198} is more special than just a composition of the two skeletons \cite{LAMMEL20081,Berthold2009-mr}.

The effort is ongoing, including topological skeletons \cite{Eden:PARCO05}, special-purpose skeletons for computer algebra \cite{Berthold2009-fft,lobachev-phd,Lobachev2012,janjic2013space}, iteration skeletons \cite{Dieterle2013}. The idea of \citet{scscp} is to use a parallel Haskell to orchestrate further software systems to run in parallel. \citet{dieterle_horstmeyer_loogen_berthold_2016} compare the composition of skeletons to stable process networks.

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows} as a less restrictive alternative to Monads, in essence they are a generalised function arrow~\ensuremath{\to }. \citet{Hughes2005} presents a tutorial on Arrows. \citet{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} develop theoretical background of Arrows. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of list of Arrows. \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.

\paragraph{Arrows in other languages.}
Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined. Basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM). However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}}, 
a new approach to Haskell on the JVM.

\citet{achten2004arrows,achten2007arrow} use an Arrow implementation in Clean for better handling of typical GUI tasks. \citet{Dagand:2009:ORD:1481861.1481870} used Arrows in OCaml in the implementation of a distributed system.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
        \section{Background}
	\label{sec:background}
	This section gives a short overview of Arrows
        (Section~\ref{sec:arrows}) and of GpH, the \ensuremath{\Conid{Par}} Monad, and
        Eden, the three parallel Haskells which we base our DSL on
        (Section~\ref{sec:parallelHaskells}).
	\subsection{Arrows}
\label{sec:arrows}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads. \citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator \ensuremath{(\bind )\mathbin{::}\Varid{m}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{b})\to \Varid{m}\;\Varid{b}} (with \ensuremath{\Varid{m}} being a Monad)\footnote{In the example a parser of the type \ensuremath{\Conid{Parser}\;\Varid{s}\;\Varid{a}} with static meta information \ensuremath{\Varid{s}} and result \ensuremath{\Varid{a}} is shown to not be able to use the static information \ensuremath{\Varid{s}} without applying the monadic function \ensuremath{\Varid{a}\to \Varid{m}\;\Varid{b}}. With Arrows this is possible.}.

An Arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} represents a computation that converts an input \ensuremath{\Varid{a}} to an output \ensuremath{\Varid{b}}. This is defined in the \ensuremath{\Conid{Arrow}} type class shown in Fig.~\ref{fig:ArrowDefinition}.
%
To lift an ordinary function to an Arrow, \ensuremath{\Varid{arr}} is used, analogous to the monadic \ensuremath{\Varid{return}}. Similarly, the composition operator \ensuremath{\mathbin{>\!\!>\!\!>}} is analogous to the monadic composition \ensuremath{\bind } and combines two Arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} by \enquote{wiring} the outputs of the first to the inputs to the second to get a new Arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}}. Lastly, the \ensuremath{\Varid{first}} operator takes the input Arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and converts it into an Arrow on pairs \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{c})} that leaves the second argument untouched. It allows us to to save input across Arrows. Figure~\ref{fig:arrows-viz} shows a graphical representation of these basic Arrow combinators.
The most prominent instances of this interface are regular functions \ensuremath{(\to )}
and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps monadic functions, e.g.  \ensuremath{\Varid{a}\to \Varid{m}\;\Varid{b}}.

\begin{figure}[t]
\centering
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{arr}\mathbin{::}(\Varid{a}\to \Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{b}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\mathbin{>\!\!>\!\!>})\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{first}\mathbin{::}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{c}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{Arrow}\;(\to )\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{f}\mathrel{=}\Varid{f}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{g}\mathrel{=}\Varid{g}\mathbin{\circ}\Varid{f}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{first}\;\Varid{f}\mathrel{=}\lambda (\Varid{a},\Varid{c})\to (\Varid{f}\;\Varid{a},\Varid{c}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{data}\;\Conid{Kleisli}\;\Varid{m}\;\Varid{a}\;\Varid{b}\mathrel{=}\Conid{Kleisli}\;\{\mskip1.5mu \Varid{run}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{b}\mskip1.5mu\}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Varid{m}\Rightarrow \Conid{Arrow}\;(\Conid{Kleisli}\;\Varid{m})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{f}\mathrel{=}\Conid{Kleisli}\;(\Varid{return}\mathbin{\circ}\Varid{f}){}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{g}\mathrel{=}\Conid{Kleisli}\;(\lambda \Varid{a}\to \Varid{f}\;\Varid{a}\bind \Varid{g}){}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{first}\;\Varid{f}\mathrel{=}\Conid{Kleisli}\;(\lambda (\Varid{a},\Varid{c})\to \Varid{f}\;\Varid{a}\bind \lambda \Varid{b}\to \Varid{return}\;(\Varid{b},\Varid{c})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\vfill
\caption{The \ensuremath{\Conid{Arrow}} type class and its two most typical instances.}
\label{fig:ArrowDefinition}
\end{figure}

\begin{figure}[t]
\centering
\parbox[c][17em]{0.49\linewidth}{%
\vfill
\centering
	\includegraphics{images/arrow}
\vfill
%	\caption{Schematic depiction of an Arrow.}
%\label{fig:arrow-sch}
}
\parbox[c][17em]{0.49\linewidth}{%
\vfill
\centering
	{\includegraphics[scale=0.6]{images/arr}}
	{\includegraphics[scale=0.6]{images/compose}}
	{\includegraphics[scale=0.6]{images/first}}
\vfill
%\caption{Schematic depiction of |Arrow| combinators |arr|, |>>>| and |first|.}
%\label{fig:arrows-viz}
}
\caption{Schematic depiction of  an Arrow (left) and its basic
  combinators \ensuremath{\Varid{arr}}, \ensuremath{\mathbin{>\!\!>\!\!>}} and \ensuremath{\Varid{first}} (right).}
\label{fig:arrow-sch}
\label{fig:arrows-viz}
\end{figure}

\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		% \subcaptionbox
{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
		% \subcaptionbox
{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
\ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
\midrule
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/andandand}}\\
\ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\
	\end{tabular}
	\caption{Visual depiction of syntactic sugar for Arrows.}
	\label{fig:syntacticSugarArrows}
\end{figure}
Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
\ensuremath{\Varid{second}} is the mirrored version of \ensuremath{\Varid{first}} (Appendix~\ref{utilfns}).
The \ensuremath{\mathbin{*\!*\!*}} function combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, and is defined as follows:
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{*\!*\!*})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
\\
\>[B]{}\Varid{f}\mathbin{*\!*\!*}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{g}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{The (***) combinator}
% \label{fig:***}
% \end{figure}
The \ensuremath{\mathbin{\&\!\&\!\&}} combinator, which constructs an Arrow that outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input, is:
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{\&\!\&\!\&})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
\\
\>[B]{}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{*\!*\!*}\Varid{g}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{The (\&\&\&) combinator}
% \label{fig:&&&}
% \end{figure}
A~first short example given by Hughes on how to use Arrows is addition with Arrows:
%|add| over Arrows, which can be seen in Fig.~\ref{fig:addArrows}.
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{add}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}{}\<[E]%
\\
\>[B]{}\Varid{add}\;\Varid{f}\;\Varid{g}\mathrel{=}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\lambda (\Varid{u},\Varid{v})\to \Varid{u}\mathbin{+}\Varid{v}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

As we can rewrite the monadic bind operation \ensuremath{(\bind )} with only the Kleisli type into \ensuremath{\Varid{m}\;\Varid{a}\to \Conid{Kleisli}\;\Varid{m}\;\Varid{a}\;\Varid{b}\to \Varid{m}\;\Varid{b}}, but not with a general Arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}}, we can intuitively get an idea of why Arrows must be a generalisation of Monads. While this also means that a general Arrow can not express everything a Monad can, \citet{HughesArrows} shows in his parser example that this trade-off is worth it in some cases.

In this paper we will show that parallel computations can be expressed with this more general interface of Arrows without requiring Monads. We also do not restrict the compatible Arrows to ones which have \ensuremath{\Conid{ArrowApply}} instances but instead only require instances for \ensuremath{\Conid{ArrowChoice}} (for if-then-else constructs) and \ensuremath{\Conid{ArrowLoop}} (for looping). Because of this, we have a truly more general interface as compared to a monadic one.

While we could have based our DSL on Profunctors as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability. However, they are a promising candidate for future improvements of our DSL. Some Profunctors, especially ones supporting a composition operation, choice, and looping, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
	\subsection{Short introduction to parallel Haskells}
\label{sec:parallelHaskells}
\label{sec:parEvalNIntro}
In its purest form, parallel computation (on functions) can be looked at as the execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel or \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}, as also Figure~\ref{fig:parEvalN} symbolically shows.
\begin{figure}[t]
  \centering
	\includegraphics[scale=0.7]{images/parEvalN}
	\caption{Schematic illustration of \ensuremath{\Varid{parEvalN}}. A list of inputs is transformed by different functions in parallel.}
	\label{fig:parEvalN}
\end{figure}

In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow-based parallel Haskell.

There exist several parallel Haskells already. Among the most important are probably GpH \citep[based on \ensuremath{\Varid{par}} and \ensuremath{\Varid{pseq}} \enquote{hints},][]{Trinder1996,Trinder1998a}, the \ensuremath{\Conid{Par}} Monad \citep[a monad for deterministic parallelism,][]{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \citep[a parallel Haskell for distributed memory,][]{eden,Loogen2012}, HdpH \citep[a Template Haskell-based parallel Haskell for distributed memory,][]{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish \citep[a \ensuremath{\Conid{Par}} extension with focus on communication,][]{Kuper:2014:TPE:2666356.2594312}.

As the goal of this paper is not to re-implement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface. For this paper we chose GpH for its simplicity, the \ensuremath{\Conid{Par}} Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.

LVish and HdpH were not chosen as the former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper, while the latter (at least in its current form) does not comply with our representation of parallelism due to its heavy reliance on Template Haskell.

We will now go into some detail on GpH, the \ensuremath{\Conid{Par}} Monad and Eden, and also give their respective implementations of the non-Arrow version of \ensuremath{\Varid{parEvalN}}.

%Before we go into detail on how we can use this idea of parallelism
%for parallel Arrows, as a short introduction to parallelism in
%Haskell we will now implement |parEvalN| with several different
%parallel Haskells.

\subsubsection{Glasgow parallel Haskell -- GpH}
\label{sec:GpHIntro}
GpH \cite{Marlow2009,Trinder1998a} is one of the simplest ways to do parallel processing found in standard GHC.\footnote{The Multicore implementation of GpH is available on Hackage under \url{https://hackage.haskell.org/package/parallel-3.2.1.0}, compiler support is integrated in the stock GHC.} Besides some basic primitives (\ensuremath{\Varid{par}} and \ensuremath{\Varid{pseq}}), it ships with parallel evaluation strategies for several types which can be applied with \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}, which is exactly what is required for an implementation of \ensuremath{\Varid{parEvalN}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\mathbf{let}\;\Varid{bs}\mathrel{=}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as}{}\<[E]%
\\
\>[B]{}\hsindent{18}{}\<[18]%
\>[18]{}\mathbf{in}\;\Varid{bs}\mathbin{`\Varid{using}`}\Varid{parList}\;\Varid{rdeepseq}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$

In the above definition of \ensuremath{\Varid{parEvalN}} we just apply the list of functions \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} to the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} by zipping them with the application operator \ensuremath{\mathbin{\$}}. % $
We then evaluate this lazy list \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} according to a \ensuremath{\Conid{Strategy}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} with the \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}} operator. We construct this strategy with \ensuremath{\Varid{parList}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Conid{Strategy}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{\Varid{rdeepseq}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{Strategy}\;\Varid{a}} where the latter is a strategy which evaluates to normal form. Other strategies like e.g. evaluation to weak head normal form are available as well. It also allows for custom \ensuremath{\Conid{Strategy}} implementations to be used.
%
Fig.~\ref{fig:parEvalNMulticoreImg} shows a visual representation of this code.

% \begin{figure}[h]
% \begin{code}
% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
% parEvalN fs as = let bs = zipWith ($) fs as 
%                  in (bs `using` parList rdeepseq)
% \end{code}
% \caption{Multicore version of |parEvalN|.}
% \label{fig:parEvalNMulticore}
% \end{figure}

\begin{figure}[t]
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
	\caption{\ensuremath{\Varid{parEvalN}} (GpH).}
	\label{fig:parEvalNMulticoreImg}
\end{figure} %$ %% formatting

\subsubsection{\ensuremath{\Conid{Par}} Monad}
The \ensuremath{\Conid{Par}} Monad\footnote{The \ensuremath{\Conid{Par}} monad can be found in the \texttt{monad-par} package on Hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs. Let:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{runPar}\mathbin{\$}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}(\Varid{sequenceA}\;(\Varid{map}\;(\Varid{return}\mathbin{\circ}\Varid{spawn})\;(\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as})))\bind \Varid{mapM}\;\Varid{get}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The \ensuremath{\Conid{Par}} Monad version of our parallel evaluation function \ensuremath{\Varid{parEvalN}} is defined by zipping the list of \ensuremath{[\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]} with the list of inputs \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} with the application operator \ensuremath{\mathbin{\$}} just like with GpH. % $
Then, we map over this not yet evaluated lazy list of results \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{spawn}\mathbin{::}\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{Par}\;\Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})} to transform them to a list of not yet evaluated forked away computations \ensuremath{[\mskip1.5mu \Conid{Par}\;(\Conid{IVar}\;\Varid{b})\mskip1.5mu]}, which we convert to \ensuremath{\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{sequenceA}}. We wait for the computations to finish by mapping over the \ensuremath{\Conid{IVar}\;\Varid{b}} values inside the \ensuremath{\Conid{Par}} Monad with \ensuremath{\Varid{get}}. This results in \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}. We execute this process with \ensuremath{\Varid{runPar}} to finally get \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]}. While we used \ensuremath{\Varid{spawn}} in the definition above, a head-strict variant can easily be defined by replacing \ensuremath{\Varid{spawn}} with \ensuremath{\Varid{spawn\char95 }\mathbin{::}\Conid{Par}\;\Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a})}.
Fig.~\ref{fig:parEvalNParMonadImg} shows a graphical representation.

% \begin{figure}[h]
% \begin{code}
% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
% parEvalN fs as = runPar $ 
% 	(sequenceA $ map (spawnP) $ zipWith ($) fs as) >>= mapM get
% \end{code}
% \caption{|Par| Monad version of |parEvalN|.}
% \label{fig:parEvalNParMonad}
% \end{figure}
\begin{figure}[t]
	\includegraphics[scale=0.5]{images/parEvalNParMonad}
	\caption{\ensuremath{\Varid{parEvalN}} (\ensuremath{\Conid{Par}} Monad).}
	\label{fig:parEvalNParMonadImg}
\end{figure}

\subsubsection{Eden}
Eden \cite{eden,Loogen2012} is a parallel Haskell for distributed memory and comes with MPI and PVM as distributed backends.\footnote{The projects homepage can be found at \url{http://www.mathematik.uni-marburg.de/~eden/}. The Hackage page is at \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/}.} It is targeted towards clusters, but also functions well in a shared-memory setting with a further simple backend. However, in contrast to many other parallel Haskells, in Eden each process has its own heap. This seems to be a waste of memory, but with distributed programming paradigm and individual GC per process, Eden yields good performance results on multicores, as well \cite{arcs-dc,aswad2009low}.

While Eden comes with a Monad \ensuremath{\Conid{PA}} for parallel evaluation, it also ships with a completely functional interface that includes
%\\
a \ensuremath{\Varid{spawnF}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]}
%\\
%a |spawnF| 
function that
%This 
allows us to define \ensuremath{\Varid{parEvalN}} directly:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\mathrel{=}\Varid{spawnF}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{images/parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.

\paragraph{Eden TraceViewer.}
\label{sec:edentv}
To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}. In the next sections we will present some \emph{trace visualisations}, the post-mortem process diagrams of Eden processes and their activity.

The trace visualisations are colour-coded. In such a visualisation (Fig.~\ref{fig:withoutFutures}), the $x$ axis shows the time, the $y$ axis enumerates the machines and processes. The visualisation shows a running process in green, a blocked process is red. If the process is \enquote{runnable}, \ie it may run, but does not, it is yellow. The typical reason for this is GC. An inactive machine, where no processes are started yet, or all are already terminated, shows as a blue bar. A~communication from one process to another is represented with a black arrow. A~stream of communications, \eg a transmitted list is shows as a dark shading between sender and receiver processes.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
	%\pagebreak
	%\pagebreak
	%\pagebreak
	\section{Parallel Arrows}
\label{sec:parallel-arrows}
While Arrows are a general interface to computation, we introduce here specialised Arrows as a general interface to \textit{parallel computations}. We present the \ensuremath{\Conid{ArrowParallel}} type class and explain the reasoning behind it before discussing some parallel Haskell implementations and basic extensions.
\subsection{The \ensuremath{\Conid{ArrowParallel}} type class}
A~parallel computation (on functions) can be seen as execution of some functions \ensuremath{\Varid{a}\to \Varid{b}} in parallel, as our \ensuremath{\Varid{parEvalN}} prototype shows (Section~\ref{sec:parEvalNIntro}).
Translating this into Arrow terms gives us a new operator \ensuremath{\Varid{parEvalN}} that lifts a list of Arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} to a parallel Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}. % (Fig.~\ref{fig:parEvalNArrowFn}) (
This combinator is similar to \ensuremath{\Varid{evalN}} from Appendix~\ref{utilfns}, but does parallel instead of serial evaluation.
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{parEvalN Arrow combinator as a function}
% \label{fig:parEvalNArrowFn}
% \end{figure}
With this definition of \ensuremath{\Varid{parEvalN}}, parallel execution is yet another Arrow combinator. But as the implementation may differ depending on the actual type of the Arrow \ensuremath{\Varid{arr}} - or even the input \ensuremath{\Varid{a}} and output \ensuremath{\Varid{b}} - and we want this to be an interface for different backends, we introduce a new type class \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}}: %(Fig.~\ref{fig:parEvalNArrowTypeClass1}).
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClass1}
% \end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak head normal form vs. normal form). For this reason we introduce an additional \ensuremath{\Varid{conf}} parameter as we do not want \ensuremath{\Varid{conf}} to be a fixed type, as the configuration parameters can differ for different instances of \ensuremath{\Conid{ArrowParallel}}. % So we add it to the type signature of the typeclass as well: %(Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
% % \begin{figure}[h]
% \olcomment{|ArrowParallel arr a b conf| or |ArrowParallel conf arr a b|?} \mbcomment{does it really matter?}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{Arrow}\;\Varid{arr}\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
By restricting the implementations of our backends to a specific \ensuremath{\Varid{conf}} type, we also get interoperability between backends for free. We can parallelize one part of a program using one backend, and parallelize the next with another one.
% \caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
% \label{fig:parEvalNArrowTypeClassFinal}
% \end{figure}
% , as we will see in the implementation of the GpH Haskell and the |Par| Monad backends.

\subsection{\ensuremath{\Conid{ArrowParallel}} instances}
With the type class defined, we will now give implementations of it with GpH, the \ensuremath{\Conid{Par}} Monad and Eden.
\subsubsection{Glasgow parallel Haskell} \label{sec:parrows:multicore}
The GpH implementation of \ensuremath{\Conid{ArrowParallel}} is implemented in a straightforward manner in Fig.~\ref{fig:ArrowParallelMulticore}, but a bit different compared to the variant from Section \ref{sec:GpHIntro}. We use \ensuremath{\Varid{evalN}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} (definition in Appendix~\ref{utilfns}, think \ensuremath{\Varid{zipWith}\;(\mathbin{\$})} on Arrows) combined with \ensuremath{\Varid{withStrategy}\mathbin{::}\Conid{Strategy}\;\Varid{a}\to \Varid{a}\to \Varid{a}} from GpH, where \ensuremath{\Varid{withStrategy}} is the same as \ensuremath{\Varid{using}\mathbin{::}\Varid{a}\to \Conid{Strategy}\;\Varid{a}\to \Varid{a}}, but with flipped parameters. Our \ensuremath{\Conid{Conf}\;\Varid{a}} datatype simply wraps a \ensuremath{\Conid{Strategy}\;\Varid{a}}, but could be extended in future versions of our DSL.
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Conf}\;\Varid{a}\mathrel{=}\Conid{Conf}\;(\Conid{Strategy}\;\Varid{a}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{parEvalN}\;(\Conid{Conf}\;\Varid{strat})\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{evalN}\;\Varid{fs}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{withStrategy}\;(\Varid{parList}\;\Varid{strat})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% $ %% formatting
\caption{GpH \ensuremath{\Conid{ArrowParallel}} instance.}
\label{fig:ArrowParallelMulticore}
\end{figure}

\subsubsection{\ensuremath{\Conid{Par}} Monad}
\olcomment{introduce a newcommand for par-monad, "Arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
As for GpH we can easily lift the definition of \ensuremath{\Varid{parEvalN}} for the \ensuremath{\Conid{Par}} Monad to Arrows in Fig.~\ref{fig:ArrowParallelParMonad}. To start off, we define the \ensuremath{\Conid{Strategy}\;\Varid{a}} and \ensuremath{\Conid{Conf}\;\Varid{a}} type so we can have a configurable instance of ArrowParallel:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{Strategy}\;\Varid{a}\mathrel{=}\Varid{a}\to \Conid{Par}\;(\Conid{IVar}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{Conf}\;\Varid{a}\mathrel{=}\Conid{Conf}\;(\Conid{Strategy}\;\Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Now we can once again define our \ensuremath{\Conid{ArrowParallel}} instance as follows: First, we convert our Arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{evalN}\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{strat})\;\Varid{fs})} into an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu (\Conid{Par}\;(\Conid{IVar}\;\Varid{b}))\mskip1.5mu]} that yields composable computations in the \ensuremath{\Conid{Par}} monad. By combining the result of this Arrow with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}, we get an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu])}. Then, in order to fetch the results of the different threads, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} Monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})} -- our intermediary Arrow is of type \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu])}. Finally, we execute the computation \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} by composing with \ensuremath{\Varid{arr}\;\Varid{runPar}} and get the final Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow \Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{parEvalN}\;(\Conid{Conf}\;\Varid{strat})\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{evalN}\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{strat})\;\Varid{fs})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{sequenceA}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{\Conid{Control}.\Conid{Monad}.\Conid{Par}.get})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{runPar}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{\ensuremath{\Conid{Par}} Monad \ensuremath{\Conid{ArrowParallel}} instance.}
\label{fig:ArrowParallelParMonad}
\end{figure}

\subsubsection{Eden}
For both the GpH Haskell and \ensuremath{\Conid{Par}} Monad implementations we could use general instances of \ensuremath{\Conid{ArrowParallel}} that just require the \ensuremath{\Conid{ArrowChoice}} type class. With Eden this is not the case as we can only spawn a list of functions, which we cannot extract from general Arrows. While we could still manage to have only one instance in the module by introducing a type class % |ArrowUnwrap| (Fig.~\ref{fig:ArrowUnwrap}).
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Conid{ArrowUnwrap}\;\Varid{arr}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to (\Varid{a}\to \Varid{b}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{possible ArrowUnwrap typeclass}
% \label{fig:ArrowUnwrap}
% \end{figure}
% We don't do it here
we avoid doing so for aesthetic reasons. For now, we just implement \ensuremath{\Conid{ArrowParallel}} for normal functions and the Kleisli type in Fig.~\ref{fig:ArrowParallelEden}, where \ensuremath{\Conid{Conf}} is simply defined as \ensuremath{\mathbf{data}\;\Conid{Conf}\mathrel{=}\Conid{Nil}} since Eden does not have a configurable \ensuremath{\Varid{spawnF}} variant. % (Fig.~\ref{fig:ArrowParallelEdenFns})
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow \Conid{ArrowParallel}\;(\to )\;\Varid{a}\;\Varid{b}\;\Conid{Conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{parEvalN}\;\anonymous \mathrel{=}\Varid{spawnF}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowParallel}\;(\to )\;\Varid{a}\;(\Varid{m}\;\Varid{b})\;\Conid{Conf},{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{Monad}\;\Varid{m},\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b},\Conid{Trans}\;(\Varid{m}\;\Varid{b}))\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{ArrowParallel}\;(\Conid{Kleisli}\;\Varid{m})\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{parEvalN}\;\Varid{conf}\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{arr}\;(\Varid{parEvalN}\;\Varid{conf}\;(\Varid{map}\;(\lambda (\Conid{Kleisli}\;\Varid{f})\to \Varid{f})\;\Varid{fs}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Conid{Kleisli}\;\Varid{sequence}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{Eden \ensuremath{\Conid{ArrowParallel}} instance.}
\label{fig:ArrowParallelEden}
\end{figure}

\subsubsection{Default configuration instances}
While the configurability in the instances of the \ensuremath{\Conid{ArrowParallel}} instances above is nice, users probably would like to have proper default configurations for many parallel programs as well. These can also easily be defined as we can see by the example of the default implementation of \ensuremath{\Conid{ArrowParallel}} for the GpH:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{NFData}\;\Varid{b},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b}))\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()\;\mathbf{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{parEvalN}\;\anonymous \;\Varid{fs}\mathrel{=}\Varid{parEvalN}\;(\Varid{defaultConf}\;\Varid{fs})\;\Varid{fs}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{defaultConf}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Conid{Conf}\;\Varid{b}{}\<[E]%
\\
\>[B]{}\Varid{defaultConf}\;\Varid{fs}\mathrel{=}\Varid{stratToConf}\;\Varid{fs}\;\Varid{rdeepseq}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{stratToConf}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Conid{Strategy}\;\Varid{b}\to \Conid{Conf}\;\Varid{b}{}\<[E]%
\\
\>[B]{}\Varid{stratToConf}\;\anonymous \;\Varid{strat}\mathrel{=}\Conid{Conf}\;\Varid{strat}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The other backends have similarly structured implementations which we do not discuss here for the sake of brevity. We can, however, only have one instance of \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()} present at a time, which should not be a problem, though.

Up until now we discussed Arrow operations more in detail, but in the following sections we focus more on the data-flow between the Arrows, now that we have seen that Arrows are capable of expressing parallelism. We do explain new concepts with more details if required for better understanding, though.
	%\pagebreak
	\subsection{Extending the interface}
\label{sec:extending-interface}
With the \ensuremath{\Conid{ArrowParallel}} type class in place and implemented, we can now define other parallel interface functions. These are basic algorithmic skeletons that are used to define more sophisticated  skeletons.

\subsubsection{Lazy \ensuremath{\Varid{parEvalN}}}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEvalNLazy}
	\caption{\ensuremath{\Varid{parEvalNLazy}} depiction.}
	\label{fig:parEvalNLazyImg}
\end{figure}
The function \ensuremath{\Varid{parEvalN}} fully traverses the list of passed Arrows as well as their inputs. Sometimes this might not be feasible, as it will not work on infinite lists of functions like \eg \ensuremath{\Varid{map}\;(\Varid{arr}\mathbin{\circ}(\mathbin{+}))\;[\mskip1.5mu \mathrm{1}\mathinner{\ldotp\ldotp}\mskip1.5mu]} or just because we need the Arrows evaluated in chunks. \ensuremath{\Varid{parEvalNLazy}} (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this. It works by first chunking the input from \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} to \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]} with the given \ensuremath{\Varid{chunkSize}} in \ensuremath{\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})}. These chunks are then fed into a list \ensuremath{[\mskip1.5mu \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} of chunk-wise parallel Arrows with the help of our lazy and sequential \ensuremath{\Varid{evalN}}. The resulting \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} is lastly converted into \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{concat}}.
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalNLazy}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{evalN}\;\Varid{fchunks}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{arr}\;\Varid{concat}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathbf{where}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{fchunks}\mathrel{=}\Varid{map}\;(\Varid{parEvalN}\;\Varid{conf})\;(\Varid{chunksOf}\;\Varid{chunkSize}\;\Varid{fs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{Definition of \ensuremath{\Varid{parEvalNLazy}}.}
\label{fig:parEvalNLazy}
\end{figure}

\subsubsection{Heterogeneous tasks}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{images/parEval2}
	\caption{\ensuremath{\Varid{parEval2}} depiction.}
	\label{fig:parEval2Img}
\end{figure}
We have only talked about the parallelization of Arrows of the same set of input and output types until now. But sometimes we want to parallelize heterogeneous types as well. We can implement such a \ensuremath{\Varid{parEval2}} combinator (Figs.~\ref{fig:parEval2Img},~\ref{fig:parEval2}) which combines two Arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into a new parallel Arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} quite easily with the help of the \ensuremath{\Conid{ArrowChoice}} type class. Here, the general idea is to use the \ensuremath{\mathbin{+\!\!+\!\!+}} combinator which combines two Arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} and transforms them into \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})} to get a common Arrow type that we can then feed into \ensuremath{\Varid{parEvalN}}.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
	%\pagebreak
		%\pagebreak
	\section{Futures} \label{sec:futures}\label{futures}
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(),{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{b}\;\Varid{c}\;())\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;\Varid{fs1}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;\Varid{fs2}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The outline combinator.}
\label{fig:someCombinator}
\end{figure}

Consider the outline parallel Arrow combinator in Fig.~\ref{fig:someCombinator}. In a distributed environment this first evaluates all \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} in parallel, sends the results back to the master node, rotates the input once (in the example we require \ensuremath{\Conid{ArrowChoice}} for this) and then evaluates the \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]} in parallel to then gather the input once again on the master node.
Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,Berthold2009-fft}.

While the example could be rewritten into a single \ensuremath{\Varid{parEvalN}} call by directly wiring the Arrows together before spawning, it illustrates an important problem. When using a \ensuremath{\Conid{ArrowParallel}} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}. This can become a serious bottleneck
for a larger amount of data and number of processes \citep[as e.g.][showcases]{Berthold2009-fft}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
	\caption[without Futures]{Communication between 4 Eden processes without Futures. All communication goes through the master node. Each bar represents one process. Black lines represent communication. Colours: blue $\hat{=}$ idle, green $\hat{=}$ running, red  $\hat{=}$ blocked, yellow $\hat{=}$ suspended.}
	\label{fig:withoutFutures}
% \olcomment{more practical and heavy-weight example! fft (I have the code)?}
% \mbcomment{Depends... Are the communications easy to read in such an example?}
% \mbcomment{Keep the description for the different colours, or link to
%   the EdenTV description in \ref{sec:edentv}}
% \olcomment{ok as is}
% \olcomment{use the fft example (when it works)?}
\end{figure}

This is only a problem in distributed memory (in the scope of this paper) and we should allow nodes to communicate directly with each other. Eden already provides \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
But as we want code with our DSL to be implementation agnostic, we have to wrap this concept. We do this with the \ensuremath{\Conid{Future}} type class (Fig.~\ref{fig:future}). We require a \ensuremath{\Varid{conf}} parameter here as well, but only so that Haskells type system allows us to have multiple Future implementations imported at once without breaking any dependencies similar to what we did with the \ensuremath{\Conid{ArrowParallel}} type class earlier.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf}\mid \Varid{a}\;\Varid{conf}\to \Varid{fut}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{put}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{conf}\to \Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{get}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{conf}\to \Varid{arr}\;(\Varid{fut}\;\Varid{a})\;\Varid{a}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The \ensuremath{\Conid{Future}} type class.}
\label{fig:future}
\end{figure}
Since \ensuremath{\Conid{RD}} is only a type synonym for a communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though, as Fig.~\ref{fig:RDFuture} shows. %This is due to the same reason we had to introduce a wrapper for |Strategy a| in the GpH Haskell implementation of |ArrowParallel| in Section~\ref{sec:parrows:multicore}.
Technical details are in Appendix, in Section~\ref{app:omitted}.

For GpH and \ensuremath{\Conid{Par}} Monad, we can simply use \ensuremath{\Conid{BasicFuture}}s (Fig.~\ref{fig:BasicFuture}), which are just simple wrappers around the actual data with boiler-plate logic so that the type class is satisfied. This is because the concept of a \ensuremath{\Conid{Future}} does not change anything for shared-memory execution as there are no communication problems to fix. Nevertheless, we require a common interface so the parallel Arrows are portable across backends. The implementation can also be found in Section ~\ref{app:omitted}.

In our communication example we can use this \ensuremath{\Conid{Future}} concept for direct communication between nodes as shown in Fig.~\ref{fig:someCombinatorParallel}.
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{someCombinator}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;(\Varid{fut}\;\Varid{b})\;(),{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{b})\;\Varid{c}\;(),{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{b}\;())\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{arr}\;\Varid{b}\;\Varid{c}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{someCombinator}\;\Varid{fs1}\;\Varid{fs2}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\mathbin{>\!\!>\!\!>}\Varid{put}\;())\;\Varid{fs1})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{rightRotate}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;()\;(\Varid{map}\;(\Varid{get}\;()\mathbin{>\!\!>\!\!>})\;\Varid{fs2}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The outline combinator in parallel.}
\label{fig:someCombinatorParallel}
\end{figure}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}. One especially elegant aspect of the definition in Fig.~\ref{fig:future} is that we can specify the type of \ensuremath{\Conid{Future}} to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication. We only specify that there has to be a compatible Future and do not care about any specifics as can be seen in Fig.~\ref{fig:someCombinatorParallel}. With the PArrows DSL we can also define default instances \ensuremath{\Conid{Future}\;\Varid{fut}\;\Varid{a}\;()} for each backend similar to how \ensuremath{\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;()} was defined in Section~\ref{sec:parallel-arrows}. Details can be found in Section~\ref{app:omitted}. \olcomment{Fig. is not really clear. Do Figs with a lot of load? --- fft?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{Communication between 4 Eden processes with Futures. Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is highlighted) instead of always going through the master node (bottom bar).}
	\label{fig:withFutures}
\end{figure}
	%\pagebreak
	\section{Skeletons}
\label{sec:skeletons}
Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations. While there are many possible skeletons to implement, we demonstrate the expressive power of PArrows here using four \ensuremath{\Varid{map}}-based and three toplogical skeletons.
%%% \FloatBarrier
\subsection{\ensuremath{\Varid{map}}-based skeletons}
\label{sec:map-skeletons}
The essential differences between the mapping skeletons presented here are in terms of order of evaluation and work distribution but still provide the same semantics as a sequential \ensuremath{\Varid{map}}.

\paragraph{Parallel \ensuremath{\Varid{map}} and laziness.}
The \ensuremath{\Varid{parMap}} skeleton (Figs.~\ref{fig:parMapImg},~\ref{fig:parMap}) is probably the most common skeleton for parallel programs. We can implement it with \ensuremath{\Conid{ArrowParallel}} by repeating an Arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and then passing it into \ensuremath{\Varid{parEvalN}} to obtain an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
Just like \ensuremath{\Varid{parEvalN}}, \ensuremath{\Varid{parMap}} traverses all input Arrows as well as the inputs.
Because of this, it has the same restrictions as \ensuremath{\Varid{parEvalN}} as compared to \ensuremath{\Varid{parEvalNLazy}}. So it makes sense to also have a \ensuremath{\Varid{parMapStream}} (Figs.~\ref{fig:parMapStreamImg},~\ref{fig:parMapStream}) which behaves like \ensuremath{\Varid{parMap}}, but uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}. Implementing these skeletons is straightforward as in Appendix \ref{app:omitted} in Figs.\ref{fig:parMap} and \ref{fig:parMapStream}.

\begin{figure}[thb]
%farm
\includegraphics[scale=0.7]{images/farm}
\caption{\ensuremath{\Varid{farm}} depiction.}
\label{fig:farmImg}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{farm}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{farm}\;\Varid{conf}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{shuffle}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{farm}} definition.}
\label{fig:farm}

%farmChunk
\includegraphics[scale=0.7]{images/farmChunk}
\caption{\ensuremath{\Varid{farmChunk}} depiction.}
\label{fig:farmChunkImg}
\end{figure}

\paragraph{Statically load-balancing parallel \ensuremath{\Varid{map}}.}
Our \ensuremath{\Varid{parMap}} spawns every single computation in a new thread (at least for the instances of \ensuremath{\Conid{ArrowParallel}} we presented in this paper). This can be quite wasteful and a statically load-balancing \ensuremath{\Varid{farm}} (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over \ensuremath{\Varid{numCores}} workers seems useful.
The definitions of the helper functions \ensuremath{\Varid{unshuffle}}, \ensuremath{\Varid{takeEach}}, \ensuremath{\Varid{shuffle}} (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.

%\paragraph{Lazy statically load-balancing parallel map}
Since a \ensuremath{\Varid{farm}}  is basically just \ensuremath{\Varid{parMap}} with a different work distribution, it has the same restrictions as \ensuremath{\Varid{parEvalN}} and \ensuremath{\Varid{parMap}}. We can, however, define \ensuremath{\Varid{farmChunk}} (Figs.~\ref{fig:farmChunkImg},~\ref{fig:farmChunk}) which uses \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}. It is basically the same definition as for \ensuremath{\Varid{farm}}, but with \ensuremath{\Varid{parEvalNLazy}} instead of \ensuremath{\Varid{parEvalN}}.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
	%\pagebreak
	%\FloatBarrier
\subsection{Topological skeletons}
\label{sec:topology-skeletons}
Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons. The Eden library leverages this problem and already comes with more predefined skeletons\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.}, among them a \ensuremath{\Varid{pipe}}, a \ensuremath{\Varid{ring}}, and a \ensuremath{\Varid{torus}} implementations \citep{Eden:SkeletonBookChapter02}. These seem like reasonable candidates to be ported to our Arrow-based parallel Haskell. We aim to showcase that we can express more sophisticated skeletons with parallel Arrows as well.

If we used the original definition of \ensuremath{\Varid{parEvalN}}, however, these skeletons would produce an infinite loop with the GpH and \ensuremath{\Conid{Par}} Monad which during runtime would result in the program crash. This materialises with the usage of \ensuremath{\Varid{loop}} of the \ensuremath{\Conid{ArrowLoop}} type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden. An investigation of why this difference exists is beyond the scope of this work, we only provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment. However our workaround enables users of the DSL to test their code within a shared memory setting.
% even though other skeletons would be better suited to be run with them. --- OL: I don't get this sentence part
The idea of the fix is to provide a \ensuremath{\Conid{ArrowLoopParallel}} type class that has two functions -- \ensuremath{\Varid{loopParEvalN}} and \ensuremath{\Varid{postLoopParEvalN}}, where the first is to be used inside an \ensuremath{\Varid{loop}} construct while the latter will be used right outside of the \ensuremath{\Varid{loop}}. This way we can delegate to the actual \ensuremath{\Varid{parEvalN}} in the spot where the backend supports it.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loopParEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\mathbin{::}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As Eden has no problems with the looping skeletons, we use this instance:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Conid{Conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Conid{Conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loopParEvalN}\mathrel{=}\Varid{parEvalN}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\;\anonymous \mathrel{=}\Varid{evalN}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As \ensuremath{\Conid{Par}} Monad and GpH have problems with \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{loop}} their respective instances for \ensuremath{\Conid{ArrowLoopParallel}} look like this:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b}))\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;(\Conid{Conf}\;\Varid{b})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loopParEvalN}\;\anonymous \mathrel{=}\Varid{evalN}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\mathrel{=}\Varid{parEvalN}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\subsubsection{Parallel pipe}\label{sec:pipe}

The parallel \ensuremath{\Varid{pipe}} skeleton is semantically equivalent to folding over a list \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]} of Arrows with \ensuremath{\mathbin{>\!\!>\!\!>}}, but does this in parallel, meaning that the Arrows do not have to reside on the same thread/machine. We implement this skeleton using the \ensuremath{\Conid{ArrowLoop}} type class which gives us the \ensuremath{\Varid{loop}\mathbin{::}\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}} combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{loop}\;(\Varid{arr}\;(\lambda (\Varid{a},\Varid{b})\to (\Varid{b},\Varid{a}\mathbin{:}\Varid{b}))){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
which is the same as
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
defines an Arrow that takes its input \ensuremath{\Varid{a}} and converts it into an infinite stream \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} of it. Using \ensuremath{\Varid{loop}} to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \ensuremath{\Varid{evalN}\;\Varid{conf}\;\Varid{fs}} inside the second argument of \ensuremath{\mathbin{\&\!\&\!\&}} and then only picking the first element of the resulting list with \ensuremath{\Varid{arr}\;\Varid{last}}.

\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{pipeSimple}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{a}\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]\to \Varid{arr}\;\Varid{a}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\Varid{pipeSimple}\;\Varid{conf}\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loop}\;(\Varid{arr}\;\Varid{snd}\mathbin{\&\!\&\!\&}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}(\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:})\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}\Varid{loopParEvalN}\;\Varid{conf}\;\Varid{fs}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{arr}\;\Varid{last}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Simple \ensuremath{\Varid{pipe}} skeleton. The use of \ensuremath{\Varid{lazy}} (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} terminates before passing it into \ensuremath{\Varid{evalN}}.}
\label{fig:pipeSimple}
\end{figure}

However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}. Therefore, we introduce a more sophisticated version that internally uses Futures and obtain the final definition of \ensuremath{\Varid{pipe}} in Fig.~\ref{fig:pipe}.

\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{pipe}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowLoopParallel}\;\Varid{arr}\;(\Varid{fut}\;\Varid{a})\;(\Varid{fut}\;\Varid{a})\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{a}\mskip1.5mu]\to \Varid{arr}\;\Varid{a}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\Varid{pipe}\;\Varid{conf}\;\Varid{fs}\mathrel{=}\Varid{unliftFut}\;\Varid{conf}\;(\Varid{pipeSimple}\;\Varid{conf}\;(\Varid{map}\;(\Varid{liftFut}\;\Varid{conf})\;\Varid{fs})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{liftFut}\mathbin{::}(\Conid{Arrow}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf},\Conid{Future}\;\Varid{fut}\;\Varid{b}\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{fut}\;\Varid{a})\;(\Varid{fut}\;\Varid{b}){}\<[E]%
\\
\>[B]{}\Varid{liftFut}\;\Varid{conf}\;\Varid{f}\mathrel{=}\Varid{get}\;\Varid{conf}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{put}\;\Varid{conf}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{unliftFut}\mathbin{::}(\Conid{Arrow}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf},\Conid{Future}\;\Varid{fut}\;\Varid{b}\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Varid{arr}\;(\Varid{fut}\;\Varid{a})\;(\Varid{fut}\;\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{b}{}\<[E]%
\\
\>[B]{}\Varid{unliftFut}\;\Varid{conf}\;\Varid{f}\mathrel{=}\Varid{put}\;\Varid{conf}\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{get}\;\Varid{conf}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{pipe}} skeleton definition with Futures.}
\label{fig:pipe}
\end{figure}

Sometimes, this \ensuremath{\Varid{pipe}} definition can be a bit inconvenient, especially if we want to pipe Arrows of mixed types together, \ie  \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}}. By wrapping these two Arrows inside a bigger Arrow \ensuremath{\Varid{arr}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])} suitable for \ensuremath{\Varid{pipe}}, we can define \ensuremath{\Varid{pipe2}} as in Fig.~\ref{fig:pipe2}.
\begin{figure}[tb]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{pipe2}\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{conf}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
\\
\>[B]{}\Varid{pipe2}\;\Varid{conf}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}(\Varid{arr}\;\Varid{return}\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu]))\mathbin{\&\!\&\!\&}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{pipe}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{unify}\;\Varid{f}\;\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{arr}\;\Varid{snd}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{arr}\;\Varid{head}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathbf{where}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{unify}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[9]{}\hsindent{16}{}\<[25]%
\>[25]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{unify}\;\Varid{f'}\;\Varid{g'}\mathrel{=}{}\<[E]%
\\
\>[9]{}\hsindent{16}{}\<[25]%
\>[25]{}(\Varid{mapArr}\;\Varid{f'}\mathbin{*\!*\!*}\Varid{mapArr}\;\Varid{g'})\mathbin{*\!*\!*}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[9]{}\hsindent{16}{}\<[25]%
\>[25]{}\Varid{arr}\;(\lambda ((\Varid{b},\Varid{c}),\Varid{a})\to ((\Varid{a},\Varid{b}),\Varid{c})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}(\mathbin{|\!>\!\!>\!\!>\!|})\mathbin{::}(\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu]))\;(),{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Future}\;\Varid{fut}\;(([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu]),[\mskip1.5mu \Varid{c}\mskip1.5mu])\;())\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;\Varid{c}{}\<[E]%
\\
\>[B]{}(\mathbin{|\!>\!\!>\!\!>\!|})\mathrel{=}\Varid{pipe2}\;(){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Definition of \ensuremath{\Varid{pipe2}} and \ensuremath{(\mathbin{|\!>\!\!>\!\!>\!|})}, a parallel \ensuremath{\mathbin{>\!\!>\!\!>}}.}
\label{fig:pipe2}
\end{figure}

Extensive use of \ensuremath{\Varid{pipe2}} over \ensuremath{\Varid{pipe}} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{evalN}}. Nonetheless, we can define a parallel piping operator \ensuremath{\mathbin{|\!>\!\!>\!\!>\!|}}, which is semantically equivalent to \ensuremath{\mathbin{>\!\!>\!\!>}} similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.

%Another version of |>>>| is:
%\begin{code}
%f parcomp g = (f . put) >>> (get . g)
%\end{code}
%It does not launch both Arrows |f| and |g| in parallel, but allows for more smooth data communication between them. Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.

\subsubsection{Ring skeleton} \label{sec:ring}
\begin{figure}[tb]
	\includegraphics[scale=0.75]{images/ring}
	\caption{\ensuremath{\Varid{ring}} skeleton depiction.}
	\label{fig:ringImg}
\end{figure}
Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate with each other. Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets send to the neighbour of each node. This data is sent over direct communication channels, the so called \enquote{remote data}. We depict it in Appendix, Fig.~\ref{fig:ringEden}.
%\end{code}
%with toRD (to make use of remote data)
%\begin{code}
%\end{code}
%and rightRotate:
%\begin{code}

We can rewrite this functionality easily with the use of \ensuremath{\Varid{loop}} as the definition of the node function, \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})}, after being transformed into an Arrow, already fits quite neatly into \ensuremath{\Varid{loop}}'s signature: \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;(\Varid{c},\Varid{b})\to \Varid{arr}\;\Varid{a}\;\Varid{c}}. In each iteration we start by rotating the intermediary input from the nodes \ensuremath{[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu]} with \ensuremath{\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})} (Fig.~\ref{fig:edenlazyrightrotate}). Similarly to the \ensuremath{\Varid{pipe}} from Section~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our \ensuremath{\Varid{lazy}} (Fig.~\ref{fig:edenlazyrightrotate}) Arrow here, or the evaluation would fail to terminate. The reasoning is explained by \citet{Loogen2012} as a demand problem.

Next, we zip the resulting \ensuremath{([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])} to \ensuremath{[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})}. We then feed this into our parallel Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]\;[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} obtained by transforming our input Arrow \ensuremath{\Varid{f}\mathbin{::}\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})} into \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})} before \ensuremath{\Varid{repeat}}ing and lifting it with \ensuremath{\Varid{loopParEvalN}}. Finally we unzip the output list \ensuremath{[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} list into \ensuremath{([\mskip1.5mu \Varid{o}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])}.

Plugging this Arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])\;([\mskip1.5mu \Varid{o}\mskip1.5mu],\Varid{fut}\;\Varid{r})} into the definition of \ensuremath{\Varid{loop}} from earlier gives us \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]}, our ring Arrow (Fig.~\ref{fig:ringFinal}). To make sure this algorithm has speedup on shared-memory machines as well, we pass the result of this Arrow to \ensuremath{\Varid{postLoopParEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{arr}\;\Varid{id}))}.
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
%Further details on this can be found in \cite{eden_cefp}.

\begin{figure}[tb]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ring}\mathbin{::}(\Conid{Future}\;\Varid{fut}\;\Varid{r}\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{ArrowLoop}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;\Varid{o}\;\Varid{o}\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{conf}\to \Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})\to \Varid{arr}\;[\mskip1.5mu \Varid{i}\mskip1.5mu]\;[\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{ring}\;\Varid{conf}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loop}\;(\Varid{second}\;(\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{loopParEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{second}\;(\Varid{get}\;\Varid{conf})\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;(\Varid{put}\;\Varid{conf})))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{unzip})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{arr}\;\Varid{id})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{ring}} skeleton definition.}
\label{fig:ringFinal}
\end{figure}

\subsubsection{Torus skeleton}\label{sec:torus}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{\ensuremath{\Varid{torus}} skeleton depiction.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a \ensuremath{\Varid{ring}} from Section~\ref{sec:ring} one dimension further, we obtain a \ensuremath{\Varid{torus}} skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends and receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the \ensuremath{\Varid{torus}} combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the \ensuremath{\Conid{ArrowLoop}} type class.

Similar to the \ensuremath{\Varid{ring}}, we start by rotating the input (Fig.~\ref{fig:edenlazyrightrotate}), but this time not only in one direction, but in two. This means that the intermediary input from the neighbour nodes has to be stored in a tuple \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu])} in the second argument (loop only allows for two arguments) of our looped Arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))\;([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} and our rotation Arrow becomes 
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
instead of the singular rotation in the ring as we rotate \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu]} horizontally and \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]} vertically. Then, we zip the inputs for the input Arrow with 
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{arr}\;(\Varid{uncurry3}\;\Varid{zipWith3}\;\Varid{lazyzip3}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
from \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))} to \ensuremath{[\mskip1.5mu [\mskip1.5mu (\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]}, which we then evaluate in parallel.

This, however, is more complicated than in the ring case as we have one more dimension of inputs that needs to be transformed. We first have to \ensuremath{\Varid{shuffle}} all the inputs to then pass them into \ensuremath{\Varid{loopParEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{ptorus}\;\Varid{conf}\;\Varid{f}))} to get an output of \ensuremath{[\mskip1.5mu (\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]}. We then unshuffle this list back to its original ordering by feeding it into \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;\Varid{unshuffle})} which takes the input length we saved one step earlier as additional input to get a result matrix \ensuremath{[\mskip1.5mu [\mskip1.5mu [\mskip1.5mu (\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\mskip1.5mu]\mskip1.5mu]}. Finally, we unpack this matrix  with \ensuremath{\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!\!>\!\!>}\Varid{threetotwo}} to get \ensuremath{([\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu],([\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{a}\mskip1.5mu]\mskip1.5mu],[\mskip1.5mu [\mskip1.5mu \Varid{fut}\;\Varid{b}\mskip1.5mu]\mskip1.5mu]))}.

This internal looping computation is once again fed into \ensuremath{\Varid{loop}} and we also compose a final \ensuremath{\Varid{postLoopParEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{arr}\;\Varid{id}))} for the same reasons as explained for the \ensuremath{\Varid{ring}} skeleton.

\begin{figure}[tb]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{torus}\mathbin{::}(\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf},\Conid{Future}\;\Varid{fut}\;\Varid{b}\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{7}{}\<[7]%
\>[7]{}\Conid{ArrowLoop}\;\Varid{arr},\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{7}{}\<[7]%
\>[7]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{7}{}\<[7]%
\>[7]{}\Conid{ArrowLoopParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{d}\mskip1.5mu]\;[\mskip1.5mu \Varid{d}\mskip1.5mu]\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{7}{}\<[7]%
\>[7]{}\Varid{conf}\to \Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to \Varid{arr}\;[\mskip1.5mu [\mskip1.5mu \Varid{c}\mskip1.5mu]\mskip1.5mu]\;[\mskip1.5mu [\mskip1.5mu \Varid{d}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{torus}\;\Varid{conf}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{loop}\;(\Varid{second}\;((\Varid{mapArr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy})\mathbin{*\!*\!*}(\Varid{arr}\;\Varid{rightRotate}\mathbin{>\!\!>\!\!>}\Varid{lazy}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{uncurry3}\;(\Varid{zipWith3}\;\Varid{lazyzip3}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{length}\mathbin{\&\!\&\!\&}(\Varid{shuffle}\mathbin{>\!\!>\!\!>}\Varid{loopParEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{ptorus}\;\Varid{conf}\;\Varid{f})))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;\Varid{unshuffle})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{map}\;\Varid{unzip3})\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{unzip3}\mathbin{>\!\!>\!\!>}\Varid{threetotwo})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{postLoopParEvalN}\;\Varid{conf}\;(\Varid{repeat}\;(\Varid{arr}\;\Varid{id})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{ptorus}\mathbin{::}(\Conid{Arrow}\;\Varid{arr},\Conid{Future}\;\Varid{fut}\;\Varid{a}\;\Varid{conf},\Conid{Future}\;\Varid{fut}\;\Varid{b}\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{11}{}\<[11]%
\>[11]{}\Varid{conf}\to {}\<[E]%
\\
\>[B]{}\hsindent{11}{}\<[11]%
\>[11]{}\Varid{arr}\;(\Varid{c},\Varid{a},\Varid{b})\;(\Varid{d},\Varid{a},\Varid{b})\to {}\<[E]%
\\
\>[B]{}\hsindent{11}{}\<[11]%
\>[11]{}\Varid{arr}\;(\Varid{c},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b})\;(\Varid{d},\Varid{fut}\;\Varid{a},\Varid{fut}\;\Varid{b}){}\<[E]%
\\
\>[B]{}\Varid{ptorus}\;\Varid{conf}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{c},\Varid{a},\Varid{b})\to (\Varid{c},\Varid{get}\;\Varid{conf}\;\Varid{a},\Varid{get}\;\Varid{conf}\;\Varid{b}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{f}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{d},\Varid{a},\Varid{b})\to (\Varid{d},\Varid{put}\;\Varid{conf}\;\Varid{a},\Varid{put}\;\Varid{conf}\;\Varid{b})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 % $
\caption{\ensuremath{\Varid{torus}} skeleton definition. \ensuremath{\Varid{lazyzip3}}, \ensuremath{\Varid{uncurry3}} and \ensuremath{\Varid{threetotwo}} definitions are in Fig.~\ref{fig:lazyzip3etc}}.
\label{fig:torus}
\end{figure}
As an example of using this skeleton, \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}). An adapted version can be found in Fig.~\ref{fig:torusMatMult}.
\begin{figure}[tb]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{Matrix}\mathrel{=}[\mskip1.5mu [\mskip1.5mu \Conid{Int}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{prMM\char95 torus}\mathbin{::}\Conid{Int}\to \Conid{Int}\to \Conid{Matrix}\to \Conid{Matrix}\to \Conid{Matrix}{}\<[E]%
\\
\>[B]{}\Varid{prMM\char95 torus}\;\Varid{numCores}\;\Varid{problemSizeVal}\;\Varid{m1}\;\Varid{m2}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{combine}\mathbin{\$}\Varid{torus}\;()\;(\Varid{mult}\;\Varid{torusSize})\mathbin{\$}\Varid{zipWith}\;(\Varid{zipWith}\;(,))\;(\Varid{split}\;\Varid{m1})\;(\Varid{split}\;\Varid{m2}){}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\mathbf{where}\;{}\<[17]%
\>[17]{}\Varid{torusSize}\mathrel{=}(\Varid{floor}\mathbin{\circ}\Varid{sqrt})\mathbin{\$}\Varid{fromIntegral}\;\Varid{numCores}{}\<[E]%
\\
\>[17]{}\Varid{combine}\mathrel{=}\Varid{concat}\mathbin{\circ}(\Varid{map}\;(\Varid{foldr}\;(\Varid{zipWith}\;(\plus ))\;(\Varid{repeat}\;[\mskip1.5mu \mskip1.5mu]))){}\<[E]%
\\
\>[17]{}\Varid{split}\mathrel{=}\Varid{splitMatrix}\;(\Varid{problemSizeVal}\mathbin{\Varid{`div`}}\Varid{torusSize}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mbox{\onelinecomment  Function performed by each worker}{}\<[E]%
\\
\>[B]{}\Varid{mult}\mathbin{::}\Conid{Int}\to ((\Conid{Matrix},\Conid{Matrix}),[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu])\to (\Conid{Matrix},[\mskip1.5mu \Conid{Matrix}\mskip1.5mu],[\mskip1.5mu \Conid{Matrix}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{mult}\;\Varid{size}\;((\Varid{sm1},\Varid{sm2}),\Varid{sm1s},\Varid{sm2s})\mathrel{=}(\Varid{result},\Varid{toRight},\Varid{toBottom}){}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathbf{where}\;\Varid{toRight}\mathrel{=}\Varid{take}\;(\Varid{size}\mathbin{-}\mathrm{1})\;(\Varid{sm1}\mathbin{:}\Varid{sm1s}){}\<[E]%
\\
\>[5]{}\hsindent{6}{}\<[11]%
\>[11]{}\Varid{toBottom}\mathrel{=}\Varid{take}\;(\Varid{size}\mathbin{-}\mathrm{1})\;(\Varid{sm2'}\mathbin{:}\Varid{sm2s}){}\<[E]%
\\
\>[5]{}\hsindent{6}{}\<[11]%
\>[11]{}\Varid{sm2'}\mathrel{=}\Varid{transpose}\;\Varid{sm2}{}\<[E]%
\\
\>[5]{}\hsindent{6}{}\<[11]%
\>[11]{}\Varid{sms}\mathrel{=}\Varid{zipWith}\;\Varid{prMMTr}\;(\Varid{sm1}\mathbin{:}\Varid{sm1s})\;(\Varid{sm2'}\mathbin{:}\Varid{sm2s}){}\<[E]%
\\
\>[5]{}\hsindent{6}{}\<[11]%
\>[11]{}\Varid{result}\mathrel{=}\Varid{foldl1'}\;\Varid{matAdd}\;\Varid{sms}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$
\caption{Adapted matrix multiplication in Eden using a the \ensuremath{\Varid{torus}} skeleton. \ensuremath{\Varid{prMM\char95 torus}} is the parallel matrix multiplication. \ensuremath{\Varid{mult}} is the function performed by each worker. \ensuremath{\Varid{prMMTr}} calculates $AB^T$ and is used for the (sequential) calculation in the chunks. \ensuremath{\Varid{splitMatrix}} splits the Matrix into chunks. \ensuremath{\Varid{matAdd}} calculates $A + B$. Omitted definitions can be found in \ref{fig:torus_example_rest}. }
\label{fig:torusMatMult}
\end{figure}
If we compare the trace from a call using our Arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the Arrow version and execution times are comparable. We discuss further benchmarks on larger clusters and in a more detail in the next section.
\begin{figure}[tb]
\olcomment{more nodes!!!}
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (PArrows)]{Matrix multiplication with \ensuremath{\Varid{torus}} (PArrows).}
	\label{fig:torus_parrows_trace}
\end{figure}

\begin{figure}[tb]
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (Eden)]{Matrix multiplication with \ensuremath{\Varid{torus}} (Eden).}
	\label{fig:torus_eden_trace}
\end{figure}

%\FloatBarrier

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
	%\pagebreak

%
\section{Performance results and discussion}
\label{sec:benchmarks}

The preceding section has shown that PArrows are expressive. This section evaluates the performance overhead of this compositional abstraction in comparison to  GpH and the \ensuremath{\Conid{Par}} Monad on shared memory architectures and Eden on  a distributed memory cluster.
We describe our measurement platform, the benchmark results -- the shared-memory variants (GpH, \ensuremath{\Conid{Par}} Monad and Eden~CP) followed by Eden in a distributed memory setting, and conclude that PArrows hold up in terms of performance when compared to the original parallel Haskells.

\newcommand{\rmtest}{Rabin--Miller test\xspace}
\newcommand{\sudokutest}{Sudoku\xspace}
\newcommand{\jacobitest}{Jacobi sum test\xspace}
\newcommand{\torustest}{Gentleman\xspace}
\newlength{\plotwidthSMP}
\setlength{\plotwidthSMP}{0.39\textwidth}
\newlength{\plotwidthDist}
\setlength{\plotwidthDist}{0.6\textwidth}

\newcommand{\benchmarkDir}{benchmarks}

\newcommand{\speedupplot}[8]{
\begin{tikzpicture}
\begin{axis}[title={#1},
title style={align=center},
scale only axis, width=#7,
xlabel=Threads,
%xtick=data,
%ytick=data,
xtick distance=#4,
ytick distance=#4,
ylabel=Speedup,
ylabel near ticks,
grid=major,
legend entries={linear, #2},
legend style={at={(0.01,0.99)},anchor=north west},
max space between ticks=50pt,
grid style={line width=.1pt, draw=gray!10},
major grid style={line width=.2pt,draw=gray!50},
ymin=-1,
xmin=-1,
ymax=#8,
xmax=#6]
\addplot [domain=0:#3, no markers,dotted,thick]{x};
#5
\end{axis}
\end{tikzpicture}
}

\subsection{Measurement platform}
We start by explaining the hardware and software stack and outline the benchmark programs and motivation for choosing them. We also shortly address hyper-threading and why we do not use it in our benchmarks.

\subsubsection{Hardware and software}

The benchmarks are executed  both in a shared and in a distributed memory setting using the Glasgow GPG Beowulf cluster, consisting of
16 machines with 2 Intel\SymbReg~Xeon\SymbReg~E5-2640 v2 and 64 GB of DDR3 RAM each. Each processor has 8 cores and 16 (hyper-threaded) threads with a base frequency of 2 GHz and a turbo frequency of 2.50 GHz. This results in a total of 256 cores and 512 threads for the whole cluster. The operating system was Ubuntu 14.04 LTS with Kernel 3.19.0-33. Non-surprisingly, we found that hyper-threaded 32 cores do not behave in the same manner as real 16 cores (numbers here for a single machine). We disregarded the hyper-threading ability in most of the cases.

Apart from Eden, all benchmarks and libraries were compiled with Stack's\footnote{see \url{https://www.haskellstack.org/}} lts-7.1 GHC compiler which is equivalent to a standard GHC 8.0.1 with the base package in version 4.9.0.0. Stack itself was used in version 1.3.2. For GpH in its Multicore variant we used the parallel package in version 3.2.1.0\footnote{see \url{https://hackage.haskell.org/package/parallel-3.2.1.0}}, while for the \ensuremath{\Conid{Par}} Monad we used monad-par in version 0.3.4.8\footnote{see \url{https://hackage.haskell.org/package/monad-par-0.3.4.8}}. For all Eden tests, we used its GHC-Eden compiler in version 7.8.2\footnote{see \url{http://www.mathematik.uni-marburg.de/~eden/?content=build_eden_7_&navi=build}} together with OpenMPI 1.6.5\footnote{see \url{https://www.open-mpi.org/software/ompi/v1.6/}}.

Furthermore, all benchmarks were done with help of the bench\footnote{see \url{https://hackage.haskell.org/package/bench}} tool in version 1.0.2 which uses criterion (>=1.1.1.0 \&\& < 1.2)\footnote{see \url{https://hackage.haskell.org/package/criterion-1.1.1.0}} internally. All runtime data (mean runtime, max stddev, etc.) was collected with this tool.
% if not mentioned otherwise. -- OL: nowhere is it otherwise

We used a single node with 16 real cores as a shared memory test-bed
and the whole grid with 256 real cores as a device to test our
distributed memory software.

\subsubsection{Benchmarks}

We measure four benchmarks from different
sources. Most of them are parallel mathematical computations, initially
implemented in Eden. Table~\ref{tab:benches} summarises.

\begin{table}
\centering
\caption{The benchmarks we use in this paper.}
\label{tab:benches}
%% something was wrong with separators in table
\renewcommand{\tabcolsep}{0.5em}
\begin{tabular}{lccll}
\toprule
Name & Area & Type & Origin & Source \\
\midrule
\rmtest & Mathematics & \ensuremath{\Varid{parMap}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
\jacobitest & Mathematics & \ensuremath{\Varid{workpool}\mathbin{+}\Varid{reduce}} & Eden & \citet{Lobachev2012}\\
\torustest & Mathematics & \ensuremath{\Varid{torus}} & Eden & \citet{Eden:SkeletonBookChapter02}\\
\sudokutest & Puzzle & \ensuremath{\Varid{parMap}} & \ensuremath{\Conid{Par}} Monad & \citet{par-monad}\tablefootnote{actual code from: \url{http://community.haskell.org/\~simonmar/par-tutorial.pdf} and \url{https://github.com/simonmar/parconc-examples}}\\
\bottomrule
\end{tabular}
\end{table}

\rmtest is a probabilistic primality test that iterates multiple (here: 32--256)
\enquote{subtests}. Should a subtest fail, the input is definitely not a
prime. If all $n$ subtest pass, the input is composite with the
probability of $1/4^{n}$.

Jacobi sum test or APRCL is also a primality test, that however,
guarantees the correctness of the result. It is probabilistic in the
sense that its run time is not certain. Unlike \rmtest, the subtests
of Jacobi sum test have very different durations. \citet{lobachev-phd}
discusses some optimisations of parallel APRCL. Generic parallel
implementations of \rmtest and APRCL were presented in \citet{Lobachev2012}.

\enquote{Gentleman} is a standard Eden test program, developed
for their \ensuremath{\Varid{torus}} skeleton. It implements a Gentleman's algorithm for parallel matrix
multiplication \citep{Gentleman1978}. We ported an Eden-based version \citep{Eden:SkeletonBookChapter02} to PArrows.

A~parallel Sudoku solver was used by \citet{par-monad} to compare \ensuremath{\Conid{Par}} Monad
to GpH, we ported it to PArrows.

\subsubsection{What parallel Haskells run where}

The \ensuremath{\Conid{Par}} monad and GpH -- in its multicore version \cite{Marlow2009} --  can be executed on shared memory machines only.
Although GpH is available on distributed memory
clusters, and newer distributed memory Haskells such as HdpH exist,
current support of distributed memory in PArrows is limited to
Eden. We used the MPI backend of Eden in a distributed memory
setting. However, for shared memory Eden features a \enquote{CP} backend
that merely copies the memory blocks between disjoint heaps. In
this mode, Eden still operates in the \enquote{nothing shared} setting, but
is adapted better to multicore machines. We call this version of Eden
\enquote{Eden~CP}.

\subsubsection{Effect of hyper-threading}

In preliminary tests, the PArrows version of \rmtest on a single node of the Glasgow cluster
showed almost linear speedup on up to 16 shared-memory cores (as supplementary materials show). The speedup
of 64-task PArrows/Eden at 16 real cores version was 13.65 giving a parallel
efficiency of 85.3\%. However, if we increased the number of
requested cores to 32 -- \ie if we use hyper-threading on 16 real
cores -- the speedup did not increase that well. It was merely 15.99
for 32 tasks with PArrows/Eden. This was worse for other implementations.  As
for 64 tasks, we obtained a speedup of 16.12 with PArrows/Eden at 32
hyper-threaded cores and only 13.55 with PArrows/GpH.

While this shows that hyper-threading can be of benefit in scenarios similar to the ones presented in the benchmarks, we only use real cores for the performance measurements in Section~\ref{sec:benchmarkResults} as the purpose of this paper is to show the performance of PArrows and not to investigate parallel behaviour with hyper-threading.

% rm 11213 32 32-sm speedup eden: 15.993037587283924
% -"- multi: 15.09948017762912
% -"- par: 14.909092857846693

% -"- 64 32-sm speedup eden: 16.118040224478424
% -"- multi: 13.545304115702333
% -"- par: 15.155709987503396

\subsection{Benchmark results}\label{sec:benchmarkResults}

We compare the PArrow performance with direct implementations of the benchmarks in Eden, GpH and the \ensuremath{\Conid{Par}} Monad.
We start with the definition of mean overhead to compare both PArrows-enabled and standard benchmark implementations. We continue comparing speedups and overheads for the shared memory implementations and then study OpenMPI variants of the Eden-enabled PArrows as a representative of a distributed memory backend. We plot all speedup curves and all overhead values in the supplementary materials.

\subsubsection{Defining overhead}

We compare the mean overhead, \ie the mean of relative wall-clock run time differences between the PArrow and direct benchmark implementations executed multiple times with the same settings.
The error margins of the time measurements, supplied by criterion package\footnote{\url{https://hackage.haskell.org/package/criterion-1.1.1.0}}, yield the error margin of the mean overhead.

Quite often the zero value lies in the error margin of the mean overhead. This means that even though we have measured some difference (against or even in favour of PArrows), it could be merely the error margin of the measurement and the difference might not be existent. We are mostly interested in the cases where above issue does not persist, we call them \emph{significant}. We often denote the error margin with $\pm$ after the mean overhead value.

\subsubsection{Shared memory}

\paragraph{Speedup.}
The \rmtest benchmark showed almost linear speedup for both 32 and 64 tasks, the performance is slightly better in the latter case: 13.7 at 16 cores for input $2^{11213}-1$ and 64 tasks in the best case scenario with Eden~CP. The performance of the \sudokutest benchmark merely reaches a speedup of 9.19 (GpH), 8.78 (\ensuremath{\Conid{Par}} Monad), 8.14 (Eden~CP) for 16 cores and 1000 Sudokus. In contrast to Rabin--Miller, here the \ensuremath{\Conid{GpH}} seems to be the best of all, while Rabin--Miller profited most from Eden~CP (\ie Eden with direct memory copy) implementation of PArrows. Gentleman on shared memory has a plummeting speedup curve with GpH and \ensuremath{\Conid{Par}} Monad and logarithmically increasing speedup for the Eden-based version. The latter reached a speedup of 6.56 at 16 cores.

\paragraph{Overhead.}

For the shared memory \rmtest benchmark, implemented with PArrows using Eden~CP, GpH, and \ensuremath{\Conid{Par}} Monad, the overhead values are within single percents range, but also negative overhead (\ie PArrows are better) and larger error margins happen. To give a few examples, the overhead for Eden~CP with input value $2^{11213}-1$, 32 tasks, and 16 cores is $1.5\%$, but the error margin is around $5.2\%$! Same implementation in the same setting with 64 tasks reaches $-0.2\%$ overhead, PArrows apparently fare better than Eden -- but the error margin of $1.9\%$ disallows this interpretation. We focus now on significant overhead values. To name a few: $0.41\%\; \pm 7\cdot 10^{-2}\%$ for Eden~CP and 64 tasks at 4 cores; $4.7\% \; \pm 0.72\%$ for GpH, 32 tasks, 8 cores; $0.34\% \; \pm 0.31\%$ for \ensuremath{\Conid{Par}} Monad at 4 cores with 64 tasks. The worst significant overhead was in case of GpH  with $8\% \; \pm 6.9\%$ at 16 cores with 32 tasks and input value $2^{11213}-1$. In other words, we notice no major slow-down through PArrows here.

For Sudoku the situation is slightly different. There is a minimal significant ($-1.4\% \; \pm 1.2\%$ at 8 cores) speed \emph{improvement} with PArrows Eden~CP version when compared with the base Eden CP benchmark. However, with increasing number of cores the error margin reaches zero again: $-1.6\% \; \pm 5.0\%$ at 16 cores. The \ensuremath{\Conid{Par}} Monad shows a similar development, \eg with $-1.95\% \; \pm 0.64\%$ at 8 cores. The GpH version shows both a significant speed improvement of $-4.2\% \; \pm 0.26\%$ (for 16 cores) with PArrows and a minor overhead of $0.87\% \; \pm 0.70\%$ (4 cores).

The Gentleman multiplication with Eden CP shows a minor significant overhead of $2.6\% \; \pm 1.0\%$ at 8 cores and an insignificant improvement at 16 cores. Summarising, we observe a low (if significant at all) overhead, induced by PArrows in the shared memory setting.

\subsubsection{Distributed memory}

\paragraph{Speedup.}
The speedup of distributed memory Rabin--Miller benchmark with PArrows and Eden showed an almost 
linear speedup excepting around 192 cores where an unfortunate task distribution reduces performance.
As seen in Fig.~\ref{fig:rabinMillerDistSpeedup}, we reached a speedup of 213.4 with PArrrows at 256 cores (vs. 207.7 with pure Eden). Because of memory limitations, the speedup of Jacobi sum test for large inputs (such as $2^{4253}-1$) could be measured only in a massively distributed setting. PArrows improved there from \SI{9193}{\second} (at 128 cores) to \SI{1649}{\second} (at 256 cores). A~scaled-down version with input $2^{3217}-1$ stagnates the speedup at about 11 for both PArrows and Eden for more than 64 cores. There is apparently not enough work for that many cores. The Gentleman test with input 4096 had an almost linear speedup first, then plummeted between 128 and 224 cores, and recovered at 256 cores with speedup of~129.

\begin{figure}[ht]
	\centering
	\speedupplot{Speedup of distributed \rmtest \enquote{44497 256}}{PArrows Eden}{256}{32}{
	\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
	smooth]{\benchmarkDir/distributed-rm/bench-distributed.bench.skelrm-parrows-44497-256.csv};
	}{272}{\plotwidthDist}{272}
	\caption[Speedup distributed Rabin--Miller]{Speedup of the distributed \rmtest benchmark using PArrows with Eden.}
	\label{fig:rabinMillerDistSpeedup}
\end{figure}

\paragraph{Overhead.}
We use our mean overhead quality measure and the notion of significance also for distributed memory benchmarks. The mean overhead of Rabin-Miller test in the distributed memory setting ranges from $0.29\%$ to $-2.8\%$ (last value in favour of PArrows), but these values are not significant with error margins $\pm 0.8\%$ and $\pm 2.9\%$ correspondingly. A sole significant (by a very low margin) overhead is $0.35\% \; \pm 0.33\%$ at 64 cores.
We measured the mean overhead for Jacobi benchmark for an input of $2^{3217}-1$ for up to 256 cores.
We reach the flattering value $-3.8\% \; \pm 0.93\%$ at 16 cores in favour of PArrows, it was the sole significant overhead value. The  value for 256 cores was $0.31\% \; \pm 0.39\%$.
Mean overhead for distributed Gentleman multiplication was also low. Significant values include $1.23\% \; \pm 1.20\%$ at 64 cores and $2.4\% \; \pm 0.97\%$ at 256 cores. It took PArrows 64.2 seconds at 256 cores to complete the benchmark.

Similar to the shared memory setting, PArrows only imply a very low penalty with distributed memory that lies in lower single-percent digits at most.

\subsection{Discussion}

\input{bestAndWorstBenchmarks}

PArrows performed in our benchmarks with little to no overhead. Tables~\ref{tab:meanOverheadSharedMemory} and \ref{tab:meanOverHeadDistributedMemory} clarify this once more: The PArrows-enabled versions trade blows with their vanilla counterparts when comparing the means over all cores of the mean overheads. If we combine these findings with the usability of our DSL,
the minor overhead induced by PArrows is outweighed by their convenience and usefulness to the user.

PArrows is an extendable formalism, they can be easily ported to further parallel Haskells while still maintaining interchangeability. It is straightforward to provide further implementations of algorithmic skeletons in PArrows.
	%%\pagebreak

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are first to represent
\emph{parallel} computation with
Arrows, and hence to show their usefulness for
% Arrows are a useful tool for
composing parallel
programs. We have shown that for a generic and extensible parallel Haskell, we do not have to restrict ourselves to a monadic interface. % Instead, we use Arrows which are a more general concept than Monads, but still allow for powerful composition. Furthermore, we think
We argue that Arrows are better suited to  parallelise
pure functions than Monads,  as the functions are already Arrows and can be used
directly in our DSL.
Arrows are a better fit to parallelise pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way. % We manage to retain this nice property of parallel Haskells such as Eden or GpH (which use pure constructs)
% while still having a generic and composable interface.
We use a non-monadic interface (similar to Eden or GpH) and retain composability.
%
The DSL allows for a direct parallelisation of monadic code via the Kleisli type and additionally allows to parallelise any Arrow type that has an instance for \ensuremath{\Conid{ArrowChoice}}. (Some skeletons require an additional \ensuremath{\Conid{ArrowLoop}} instance.)

We have demonstrated the generality of the approach by exhibiting PArrow implementations for Eden, GpH, and the \ensuremath{\Conid{Par}} Monad. Hence, parallel programs can be ported between task parallel Haskell implementations with little or no effort. We are confident that it will be straightforward to add other task-parallel Haskells.
% Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort. It is quite straightforward to add further backends.
% %
% %
% Performance-wise,
%Parallel Arrows are on par with existing parallel Haskells, as they only introduce minor overhead in some of our benchmarks.
%
%The benefit is, however, the
In other words, PArrows greatly increase portability of parallel Haskell programs.
%
%
Our measurements of four benchmarks on both shared and distributed memory platforms shows that the generality and portability of PArrows has very low performance overheads, \ie never more than $8\% \; \pm 6.9\%$ and typically under $2\%$.

%\mbcomment{mention ArrowLoop in Torus and Ring chapters}
%\olcomment{Parrows + accelerate = love?} \olcomment{Metion port to Frege.}

\subsection{Future work}
\label{sec:future-work}

Our PArrows DSL can be expanded % to further parallel Haskells.  More specifically we target HdpH \cite{Maier:2014:HDS:2775050.2633363} for this future extension. HdpH is a modern distributed Haskell that would benefit from our Arrow notation.
to other task parallel Haskells, and a specific target is HdpH \cite{Maier:2014:HDS:2775050.2633363}.
Further Future-aware versions of Arrow combinators can be defined. Existing combinators could also be improved, for example a more special versions of \ensuremath{\mathbin{>\!\!>\!\!>}} and \ensuremath{\mathbin{*\!*\!*}} combinators are viable.
%Arrow-based notation might enable further compiler optimisations, such as

% More experiences with seamless porting of parallel PArrows-based programs across the backends are welcome.
% Of course, we are ourselves working on expanding
In ongoing work we are expanding 
both our skeleton library and the number of skeleton-based parallel programs that use our DSL.
% to be portable across flavours of parallel Haskells.
It would also be interesting to see a hybrid of PArrows and Accelerate \cite{McDonell:2015:TRC:2887747.2804313}.
Ports of our approach to other languages such as Frege, Eta, or Java directly are at an early development stage.
	%\pagebreak
        %\bibliographystyle{jfp}
	%\bibliography{references,main}
        \providecommand{\noopsort}[1]{}
\begin{thebibliography}{}

\bibitem[\protect\citename{Acar {\em et~al.}\relax,
  }2000]{Acar:2000:DLW:341800.341801}
Acar, Umut~A., Blelloch, Guy~E., \& Blumofe, Robert~D. (2000).
\newblock The data locality of work stealing.
\newblock {\em Pages  1--12 of:} {\em Proceedings of the 12\xth Annual ACM
  Symposium on Parallel Algorithms and Architectures}.
\newblock SPAA '00.
\newblock ACM.

\bibitem[\protect\citename{Achten {\em et~al.}\relax, }2007]{achten2007arrow}
Achten, Peter, van Eekelen, Marko, de~Mol, Maarten, \& Plasmeijer, Rinus.
  (2007).
\newblock An arrow based semantics for interactive applications.
\newblock  {\em Draft Proceedings of the Symposium on Trends in Functional
  Programming}.
\newblock TFP '07.

\bibitem[\protect\citename{Achten {\em et~al.}\relax, }2004]{achten2004arrows}
Achten, PM, van Eekelen, Marko~CJD, Plasmeijer, MJ, \& Weelden, A~van. (2004).
\newblock {\em Arrows for generic graphical editor components}.
\newblock Tech. rept. Nijmegen Institute for Computing and Information
  Sciences, Faculty of Science, University of Nijmegen, The Netherlands.
\newblock Technical Report NIII-R0416.

\bibitem[\protect\citename{Alimarine {\em et~al.}\relax,
  }2005]{Alimarine:2005:BAA:1088348.1088357}
Alimarine, Artem, Smetsers, Sjaak, van Weelden, Arjen, van Eekelen, Marko, \&
  Plasmeijer, Rinus. (2005).
\newblock There and back again: Arrows for invertible programming.
\newblock {\em Pages  86--97 of:} {\em Proceedings of the {2005} {ACM}
  {SIGPLAN} Workshop on {Haskell}}.
\newblock Haskell '05.
\newblock ACM.

\bibitem[\protect\citename{Aljabri {\em et~al.}\relax,
  }2014]{Aljabri:2013:DIG:2620678.2620682}
Aljabri, Malak, Loidl, Hans-Wolfgang, \& Trinder, Phil~W. (2014).
\newblock The design and implementation of {GUMSMP}: A multilevel parallel
  haskell implementation.
\newblock {\em Pages  37:37--37:48 of:} {\em Proceedings of the 25\xth
  Symposium on Implementation and Application of Functional Languages}.
\newblock IFL '13.
\newblock ACM.

\bibitem[\protect\citename{Aljabri {\em et~al.}\relax, }2015]{Aljabri2015}
Aljabri, Malak, Loidl, Hans-Wolfgang, \& Trinder, Phil. (2015).
\newblock {\em Balancing shared and distributed heaps on {NUMA} architectures}.
\newblock Springer.
\newblock Pages  1--17.

\bibitem[\protect\citename{Alt \& Gorlatch, }2003]{AlGo03a}
Alt, Martin, \& Gorlatch, Sergei. (2003).
\newblock {F}uture-{B}ased {RMI}: Optimizing compositions of remote method
  calls on the {G}rid.
\newblock {\em Pages  682--693 of:} Kosch, Harald, B{\"o}sz{\"o}rm{\'e}nyi,
  L{\'a}szl{\'o}, \& Hellwagner, Hermann (eds), {\em Euro-Par~2003}.
\newblock LNCS 2790.
\newblock Springer-Verlag.

\bibitem[\protect\citename{Asada, }2010]{Asada:2010:ASM:1863597.1863607}
Asada, Kazuyuki. (2010).
\newblock Arrows are strong monads.
\newblock {\em Pages  33--42 of:} {\em Proceedings of the Third ACM SIGPLAN
  Workshop on Mathematically Structured Functional Programming}.
\newblock MSFP '10.
\newblock New York, NY, USA: ACM.

\bibitem[\protect\citename{Aswad {\em et~al.}\relax, }2009]{aswad2009low}
Aswad, Mustafa, Trinder, Phil, Al~Zain, Abdallah, Michaelson, Greg, \&
  Berthold, Jost. (2009).
\newblock Low pain vs no pain multi-core {Haskells}.
\newblock {\em Pages  49--64 of:} {\em Trends in Functional Programming}.

\bibitem[\protect\citename{Atkey, }2011]{ATKEY201119}
Atkey, Robert. (2011).
\newblock What is a categorical model of arrows?
\newblock {\em Electronic notes in theoretical computer science}, {\bf 229}(5),
  19--37.

\bibitem[\protect\citename{Berthold, }2008]{JostThesis}
Berthold, Jost. (2008).
\newblock {\em Explicit and implicit parallel functional programming ---
  concepts and implementation}.
\newblock Ph.D. thesis, Philipps-Universit\"{a}t Marburg.

\bibitem[\protect\citename{Berthold \& Loogen, }2006]{Eden:PARCO05}
Berthold, Jost, \& Loogen, Rita. (2006).
\newblock Skeletons for recursively unfolding process topologies.
\newblock  Joubert, Gerhard~R., Nagel, Wolfgang~E., Peters, Frans~J., Plata,
  Oscar~G., Tirado, P., \& Zapata, Emilio~L. (eds), {\em Parallel Computing:
  Current \& Future Issues of High-End Computing, ParCo 2005, Malaga, Spain}.
\newblock NIC Series 33.
\newblock Central Institute for Applied Mathematics, J{\"u}lich, Germany.

\bibitem[\protect\citename{Berthold \& Loogen, }2007]{Berthold2007a}
Berthold, Jost, \& Loogen, Rita. (2007).
\newblock {Visualizing Parallel Functional Program Executions: Case Studies
  with the Eden Trace Viewer}.
\newblock  {\em ParCo '07. Parallel Computing: Architectures, Algorithms and
  Applications}.
\newblock IOS Press.

\bibitem[\protect\citename{Berthold {\em et~al.}\relax, }2008]{PADL08HMWS}
Berthold, Jost, Dieterle, Mischa, Loogen, Rita, \& Priebe, Steffen. (2008).
\newblock Hierarchical master-worker skeletons.
\newblock  Warren, David~S., \& Hudak, Paul (eds), {\em Practical Aspects of
  Declarative Languages (PADL'08)}.
\newblock LNCS 4902.
\newblock San Francisco (CA), USA: Springer-Verlag.

\bibitem[\protect\citename{Berthold {\em et~al.}\relax, }2009a]{arcs-dc}
Berthold, Jost, Dieterle, Mischa, Lobachev, Oleg, \& Loogen, Rita. (2009a).
\newblock {Distributed Memory Programming on Many-Cores -- A Case Study Using
  {Eden} Divide-\&-Conquer Skeletons}.
\newblock {\em Pages  47--55 of:} Gro\ss{}pitsch, K.-E., Henkersdorf, A.,
  Uhrig, S., Ungerer, T., \& H\"{a}hner, J. (eds), {\em Workshop on Many-Cores
  at ARCS '09 -- 22\nd International Conference on Architecture of Computing
  Systems 2009}.
\newblock VDE-Verlag.

\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
  }2009b]{Berthold2009-mr}
Berthold, Jost, Dieterle, Mischa, \& Loogen, Rita. (2009b).
\newblock Implementing parallel {Google} map-reduce in {Eden}.
\newblock {\em Pages  990--1002 of:} Sips, Henk, Epema, Dick, \& Lin, Hai-Xiang
  (eds), {\em Euro-Par 2009 Parallel Processing}.
\newblock LNCS 5704.
\newblock Springer Berlin Heidelberg.

\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
  }2009c]{Berthold2009-fft}
Berthold, Jost, Dieterle, Mischa, Lobachev, Oleg, \& Loogen, Rita. (2009c).
\newblock {\em Parallel {FFT} with {Eden} skeletons}.
\newblock PaCT '09.
\newblock Springer.
\newblock Pages  73--83.

\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
  }2016]{berthold_loidl_hammond_2016}
Berthold, Jost, Loidl, Hans-Wolfgang, \& Hammond, Kevin. (2016).
\newblock {PAEAN}: Portable and scalable runtime support for parallel {Haskell}
  dialects.
\newblock {\em Journal of functional programming}, {\bf 26}.

\bibitem[\protect\citename{Bischof \& Gorlatch, }2002]{Bischof2002}
Bischof, Holger, \& Gorlatch, Sergei. (2002).
\newblock {\em Double-scan: Introducing and implementing a new data-parallel
  skeleton}.
\newblock Euro-Par '02.
\newblock Springer.
\newblock Pages  640--647.

\bibitem[\protect\citename{Blumofe \& Leiserson,
  }1999]{Blumofe:1999:SMC:324133.324234}
Blumofe, Robert~D., \& Leiserson, Charles~E. (1999).
\newblock Scheduling multithreaded computations by work stealing.
\newblock {\em J. acm}, {\bf 46}(5), 720--748.

\bibitem[\protect\citename{Botorog \& Kuchen, }1996]{botorog1996efficient}
Botorog, G.~H., \& Kuchen, H. (1996).
\newblock {\em {Euro-Par'96 Parallel Processing}}.
\newblock LNCS 1123.
\newblock Springer-Verlag.
\newblock Chap. Efficient parallel programming with algorithmic skeletons,
  pages  718--731.

\bibitem[\protect\citename{Brown \& Hammond, }2010]{brown2010ever}
Brown, C., \& Hammond, K. (2010).
\newblock Ever-decreasing circles: a skeleton for parallel orbit calculations
  in {Eden}.
\newblock  {\em Draft Proceedings of the Symposium on Trends in Functional
  Programming}.
\newblock TFP '10.

\bibitem[\protect\citename{Buono {\em et~al.}\relax, }2010]{BUONO20102095}
Buono, D., Danelutto, M., \& Lametti, S. (2010).
\newblock Map, reduce and mapreduce, the skeleton way.
\newblock {\em Procedia computer science}, {\bf 1}(1), 2095--2103.
\newblock ICCS 2010.

\bibitem[\protect\citename{{C}hakravarty {\em et~al.}\relax,
  }2007]{Chakravarty2007}
{C}hakravarty, Manuel M.~T., {L}eshchinskiy, Roman, {P}eyton {J}ones, Simon~L.,
  {K}eller, Gabriele, \& {M}arlow, Simon. (2007).
\newblock {Data Parallel Haskell}: a status report.
\newblock {\em Pages  10--18 of:} {\em DAMP '07}.
\newblock ACM Press.

\bibitem[\protect\citename{Chakravarty {\em et~al.}\relax,
  }2011]{Chakravarty:2011:AHA:1926354.1926358}
Chakravarty, Manuel~M.T., Keller, Gabriele, Lee, Sean, McDonell, Trevor~L., \&
  Grover, Vinod. (2011).
\newblock Accelerating {Haskell} array codes with multicore {GPUs}.
\newblock {\em Pages  3--14 of:} {\em Proceedings of the 6\xth Workshop on
  Declarative Aspects of Multicore Programming}.
\newblock DAMP '11.
\newblock ACM.

\bibitem[\protect\citename{Chase \& Lev, }2005]{Chase:2005:DCW:1073970.1073974}
Chase, David, \& Lev, Yossi. (2005).
\newblock Dynamic circular work-stealing deque.
\newblock {\em Pages  21--28 of:} {\em Proceedings of the 17\xth Annual ACM
  Symposium on Parallelism in Algorithms and Architectures}.
\newblock SPAA '05.
\newblock ACM.

\bibitem[\protect\citename{Clifton-Everest {\em et~al.}\relax, }2014]{CMCK14}
Clifton-Everest, Robert, McDonell, Trevor~L, Chakravarty, Manuel M~T, \&
  Keller, Gabriele. (2014).
\newblock {Embedding Foreign Code}.
\newblock  {\em PADL '14: The 16th International Symposium on Practical Aspects
  of Declarative Languages}.
\newblock LNCS.
\newblock Springer-Verlag.

\bibitem[\protect\citename{{C}ole, }1989]{Cole1989}
{C}ole, M.~I. (1989).
\newblock Algorithmic skeletons: Structured management of parallel computation.
\newblock  {\em Research Monographs in Parallel and Distributed Computing}.
\newblock Pitman.

\bibitem[\protect\citename{Czaplicki \& Chong,
  }2013]{Czaplicki:2013:AFR:2499370.2462161}
Czaplicki, Evan, \& Chong, Stephen. (2013).
\newblock Asynchronous functional reactive programming for guis.
\newblock {\em Sigplan not.}, {\bf 48}(6), 411--422.

\bibitem[\protect\citename{Dagand {\em et~al.}\relax,
  }2009]{Dagand:2009:ORD:1481861.1481870}
Dagand, Pierre-\'{E}variste, Kosti\'{c}, Dejan, \& Kuncak, Viktor. (2009).
\newblock Opis: Reliable distributed systems in {OCaml}.
\newblock {\em Pages  65--78 of:} {\em Proceedings of the 4\xth International
  Workshop on Types in Language Design and Implementation}.
\newblock TLDI '09.
\newblock ACM.

\bibitem[\protect\citename{Danelutto {\em et~al.}\relax,
  }1992]{DANELUTTO1992205}
Danelutto, Marco, Meglio, Roberto~Di, Orlando, Salvatore, Pelagatti, Susanna,
  \& Vanneschi, Marco. (1992).
\newblock A methodology for the development and the support of massively
  parallel programs.
\newblock {\em Future generation computer systems}, {\bf 8}(1), 205--220.

\bibitem[\protect\citename{Darlington {\em et~al.}\relax,
  }1993]{darlington1993parallel}
Darlington, J., Field, AJ, Harrison, PG, Kelly, PHJ, Sharp, DWN, Wu, Q., \&
  While, RL. (1993).
\newblock Parallel programming using skeleton functions.
\newblock {\em Pages  146--160 of:} {\em Parallel architectures and languages
  Europe}.
\newblock Springer-Verlag.

\bibitem[\protect\citename{Dastgeer {\em et~al.}\relax,
  }2011]{Dastgeer:2011:ASM:1984693.1984697}
Dastgeer, Usman, Enmyren, Johan, \& Kessler, Christoph~W. (2011).
\newblock Auto-tuning {SkePU}: A multi-backend skeleton programming framework
  for multi-{GPU} systems.
\newblock {\em Pages  25--32 of:} {\em Proceedings of the 4\xth International
  Workshop on Multicore Software Engineering}.
\newblock IWMSE '11.
\newblock ACM.

\bibitem[\protect\citename{de~la Encina {\em et~al.}\relax,
  }2011]{delaEncina2011}
de~la Encina, Alberto, Hidalgo-Herrero, Mercedes, Rabanal, Pablo, \& Rubio,
  Fernando. (2011).
\newblock {\em A parallel skeleton for genetic algorithms}.
\newblock IWANN '11.
\newblock Springer.
\newblock Pages  388--395.

\bibitem[\protect\citename{Dean \& Ghemawat,
  }2008]{Dean:2008:MSD:1327452.1327492}
Dean, Jeffrey, \& Ghemawat, Sanjay. (2008).
\newblock {MapReduce}: simplified data processing on large clusters.
\newblock {\em Communications of the acm}, {\bf 51}, 107--113.

\bibitem[\protect\citename{Dean \& Ghemawat,
  }2010]{Dean:2010:MFD:1629175.1629198}
Dean, Jeffrey, \& Ghemawat, Sanjay. (2010).
\newblock {MapReduce}: a flexible data processing tool.
\newblock {\em Communications of the acm}, {\bf 53}, 72--77.

\bibitem[\protect\citename{Dieterle {\em et~al.}\relax, }2010a]{Dieterle2010}
Dieterle, M., Horstmeyer, T., \& Loogen, R. (2010a).
\newblock Skeleton composition using remote data.
\newblock {\em Pages  73--87 of:} Carro, M., \& Pe\~{n}a, R. (eds), {\em 12\xth
  International Symposium on Practical Aspects of Declarative Languages}.
\newblock PADL '10, vol. 5937.
\newblock Springer-Verlag.

\bibitem[\protect\citename{Dieterle {\em et~al.}\relax,
  }2016]{dieterle_horstmeyer_loogen_berthold_2016}
Dieterle, M., Horstmeyer, T., Loogen, R., \& Berthold, J. (2016).
\newblock Skeleton composition versus stable process systems in {Eden}.
\newblock {\em Journal of functional programming}, {\bf 26}.

\bibitem[\protect\citename{Dieterle {\em et~al.}\relax,
  }2010b]{dieterle2010skeleton}
Dieterle, Mischa, Berthold, Jost, \& Loogen, Rita. (2010b).
\newblock {\em A skeleton for distributed work pools in {Eden}}.
\newblock FLOPS '10.
\newblock Springer.
\newblock Pages  337--353.

\bibitem[\protect\citename{Dieterle {\em et~al.}\relax, }2013]{Dieterle2013}
Dieterle, Mischa, Horstmeyer, Thomas, Berthold, Jost, \& Loogen, Rita. (2013).
\newblock {\em Iterating skeletons}.
\newblock IFL '12.
\newblock Springer.
\newblock Pages  18--36.

\bibitem[\protect\citename{Dinan {\em et~al.}\relax,
  }2009]{Dinan:2009:SWS:1654059.1654113}
Dinan, James, Larkins, D.~Brian, Sadayappan, P., Krishnamoorthy, Sriram, \&
  Nieplocha, Jarek. (2009).
\newblock Scalable work stealing.
\newblock {\em Pages  53:1--53:11 of:} {\em Proceedings of the Conference on
  High Performance Computing Networking, Storage and Analysis}.
\newblock SC '09.
\newblock ACM.

\bibitem[\protect\citename{Foltzer {\em et~al.}\relax,
  }2012]{Foltzer:2012:MPC:2398856.2364562}
Foltzer, Adam, Kulkarni, Abhishek, Swords, Rebecca, Sasidharan, Sajith, Jiang,
  Eric, \& Newton, Ryan. (2012).
\newblock A meta-scheduler for the {Par}-monad: Composable scheduling for the
  heterogeneous cloud.
\newblock {\em Sigplan not.}, {\bf 47}(9), 235--246.

\bibitem[\protect\citename{Geimer {\em et~al.}\relax, }2010]{Geimer2010}
Geimer, M., Wolf, F., Wylie, B. J.~N., {\'A}brah{\'a}m, E., Becker, D., \&
  Mohr, B. (2010).
\newblock The {Scalasca} performance toolset architecture.
\newblock {\em Concurrency and computation: Practice and experience}, {\bf
  22}(6).

\bibitem[\protect\citename{Gentleman, }1978]{Gentleman1978}
Gentleman, W.~Morven. (1978).
\newblock Some complexity results for matrix computations on parallel
  processors.
\newblock {\em Journal of the acm}, {\bf 25}(1), 112--115.

\bibitem[\protect\citename{Gorlatch, }1998]{Gorlatch1998}
Gorlatch, Sergei. (1998).
\newblock Programming with divide-and-conquer skeletons: A case study of {FFT}.
\newblock {\em Journal of supercomputing}, {\bf 12}(1-2), 85--97.

\bibitem[\protect\citename{Gorlatch \& Bischof, }1998]{Gorlatch}
Gorlatch, Sergei, \& Bischof, Holger. (1998).
\newblock A generic {MPI} implementation for a data-parallel skeleton: Formal
  derivation and application to {FFT}.
\newblock {\em Parallel processing letters}, {\bf 8}(4).

\bibitem[\protect\citename{Hammond {\em et~al.}\relax,
  }2003]{doi:10.1142/S0129626403001380}
Hammond, Kevin, Berthold, Jost, \& Loogen, Rita. (2003).
\newblock Automatic skeletons in {Template} {Haskell}.
\newblock {\em Parallel processing letters}, {\bf 13}(03), 413--424.

\bibitem[\protect\citename{Harris {\em et~al.}\relax,
  }2007]{harris2007parallel}
Harris, Mark, Sengupta, Shubhabrata, \& Owens, John~D. (2007).
\newblock Parallel prefix sum (scan) with {CUDA}.
\newblock {\em {GPU} gems}, {\bf 3}(39), 851--876.

\bibitem[\protect\citename{Harris {\em et~al.}\relax,
  }2005]{Harris:2005:CMT:1065944.1065952}
Harris, Tim, Marlow, Simon, {Peyton Jones}, Simon, \& Herlihy, Maurice. (2005).
\newblock Composable memory transactions.
\newblock {\em Pages  48--60 of:} {\em Proceedings of the 10\xth ACM SIGPLAN
  Symposium on Principles and Practice of Parallel Programming}.
\newblock PPoPP '05.
\newblock ACM.

\bibitem[\protect\citename{Hey, }1990]{Hey1990185}
Hey, Anthony J.~G. (1990).
\newblock Experiments in {MIMD} parallelism.
\newblock {\em Future generation computer systems}, {\bf 6}(3), 185--196.

\bibitem[\protect\citename{Hippold \& R{\"u}nger, }2006]{Hippold2006}
Hippold, J., \& R{\"u}nger, G. (2006).
\newblock Task pool teams: A hybrid programming environment for irregular
  algorithms on {SMP} clusters.
\newblock {\em Concurrency and computation: Practice and experience}, {\bf 18},
  1575--1594.

\bibitem[\protect\citename{Horstmeyer \& Loogen, }2013]{Horstmeyer2013}
Horstmeyer, Thomas, \& Loogen, Rita. (2013).
\newblock Graph-based communication in {Eden}.
\newblock {\em Higher-order and symbolic computation}, {\bf 26}(1), 3--28.

\bibitem[\protect\citename{Huang {\em et~al.}\relax, }2007]{Huang2007}
Huang, Liwen, Hudak, Paul, \& Peterson, John. (2007).
\newblock {\em {HPorter}: Using arrows to compose parallel processes}.
\newblock Berlin, Heidelberg: Springer Berlin Heidelberg.
\newblock Pages  275--289.

\bibitem[\protect\citename{Hudak {\em et~al.}\relax, }2003]{Hudak2003}
Hudak, Paul, Courtney, Antony, Nilsson, Henrik, \& Peterson, John. (2003).
\newblock {\em Arrows, robots, and functional reactive programming}.
\newblock Springer.
\newblock Pages  159--187.

\bibitem[\protect\citename{Hughes, }2000]{HughesArrows}
Hughes, John. (2000).
\newblock Generalising monads to arrows.
\newblock {\em Science of computer programming}, {\bf 37}(1–3), 67--111.

\bibitem[\protect\citename{Hughes, }2005]{Hughes2005}
Hughes, John. (2005).
\newblock {\em Programming with arrows}.
\newblock AFP '04.
\newblock Springer.
\newblock Pages  73--129.

\bibitem[\protect\citename{Jacobs {\em et~al.}\relax,
  }2009]{jacobs_heunen_hasuo_2009}
Jacobs, Bart, Heunen, Chris, \& Hasuo, Ichiro. (2009).
\newblock Categorical semantics for arrows.
\newblock {\em Journal of functional programming}, {\bf 19}(3-4), 403--438.

\bibitem[\protect\citename{Janjic {\em et~al.}\relax, }2013]{janjic2013space}
Janjic, Vladimir, Brown, Christopher~Mark, Neunhoeffer, Max, Hammond, Kevin,
  Linton, Stephen~Alexander, \& Loidl, Hans-Wolfgang. (2013).
\newblock Space exploration using parallel orbits: a study in parallel symbolic
  computing.
\newblock {\em Parallel computing}.

\bibitem[\protect\citename{Karasawa \& Iwasaki, }2009]{5361825}
Karasawa, Y., \& Iwasaki, H. (2009).
\newblock A parallel skeleton library for multi-core clusters.
\newblock {\em Pages  84--91 of:} {\em International Conference on Parallel
  Processing 2009}.

\bibitem[\protect\citename{Keller {\em et~al.}\relax,
  }2010]{Keller:2010:RSP:1932681.1863582}
Keller, Gabriele, Chakravarty, Manuel~M.T., Leshchinskiy, Roman, {Peyton
  Jones}, Simon, \& Lippmeier, Ben. (2010).
\newblock Regular, shape-polymorphic, parallel arrays in haskell.
\newblock {\em Sigplan not.}, {\bf 45}(9), 261--272.

\bibitem[\protect\citename{Kuchen, }2002]{Kuchen2002}
Kuchen, Herbert. (2002).
\newblock A skeleton library.
\newblock {\em Pages  620--629 of:} Monien, Burkhard, \& Feldmann, Rainer
  (eds), {\em Parallel Processing}.
\newblock Euro-Par '02.
\newblock Springer.

\bibitem[\protect\citename{Kuper {\em et~al.}\relax,
  }2014]{Kuper:2014:TPE:2666356.2594312}
Kuper, Lindsey, Todd, Aaron, Tobin-Hochstadt, Sam, \& Newton, Ryan~R. (2014).
\newblock Taming the parallel effect zoo: Extensible deterministic parallelism
  with {LVish}.
\newblock {\em Sigplan not.}, {\bf 49}(6), 2--14.

\bibitem[\protect\citename{Lengauer {\em et~al.}\relax, }1997]{Lengauer1997}
Lengauer, Christian, Gorlatch, Sergei, \& Herrmann, Christoph. (1997).
\newblock The static parallelization of loops and recursions.
\newblock {\em The journal of supercomputing}, {\bf 11}(4), 333--353.

\bibitem[\protect\citename{Li \& Zdancewic, }2006]{1648705}
Li, Peng, \& Zdancewic, S. (2006).
\newblock Encoding information flow in {Haskell}.
\newblock {\em Pages  12--16 of:} {\em 19\xth IEEE Computer Security
  Foundations Workshop}.
\newblock CSFW '06.

\bibitem[\protect\citename{Li \& Zdancewic, }2010]{LI20101974}
Li, Peng, \& Zdancewic, Steve. (2010).
\newblock Arrows for secure information flow.
\newblock {\em Theoretical computer science}, {\bf 411}(19), 1974--1994.

\bibitem[\protect\citename{Lindley {\em et~al.}\relax, }2011]{LINDLEY201197}
Lindley, Sam, Wadler, Philip, \& Yallop, Jeremy. (2011).
\newblock Idioms are oblivious, arrows are meticulous, monads are promiscuous.
\newblock {\em Electronic notes in theoretical computer science}, {\bf 229}(5),
  97--117.

\bibitem[\protect\citename{Linton {\em et~al.}\relax, }2010]{scscp}
Linton, S., Hammond, K., Konovalov, A., Al~Zain, A.~D., Trinder, P., Horn, P.,
  \& Roozemond, D. (2010).
\newblock Easy composition of symbolic computation software: a new lingua
  franca for symbolic computation.
\newblock {\em Pages  339--346 of:} {\em Proceedings of the 2010 International
  Symposium on Symbolic and Algebraic Computation}.
\newblock ISSAC '10.
\newblock ACM Press.

\bibitem[\protect\citename{Liu {\em et~al.}\relax,
  }2009]{Liu:2009:CCA:1631687.1596559}
Liu, Hai, Cheng, Eric, \& Hudak, Paul. (2009).
\newblock Causal commutative arrows and their optimization.
\newblock {\em Sigplan not.}, {\bf 44}(9), 35--46.

\bibitem[\protect\citename{Lobachev, }2011]{lobachev-phd}
Lobachev, Oleg. (2011).
\newblock {\em Implementation and evaluation of algorithmic skeletons:
  Parallelisation of computer algebra algorithms}.
\newblock Ph.D. thesis, Philipps-Universit{\"a}t Marburg.

\bibitem[\protect\citename{Lobachev, }2012]{Lobachev2012}
Lobachev, Oleg. (2012).
\newblock {\em Parallel computation skeletons with premature termination
  property}.
\newblock FLOPS 2012.
\newblock Springer.
\newblock Pages  197--212.

\bibitem[\protect\citename{Loogen {\em et~al.}\relax,
  }2003]{Eden:SkeletonBookChapter02}
Loogen, R., {Ortega-Mall\'{e}n}, Y., {Pe\~{n}a}, R., Priebe, S., \& Rubio, F.
  (2003).
\newblock {Parallelism Abstractions in Eden}.
\newblock {\em Pages  71--88 of:} Rabhi, F.~A., \& Gorlatch, S. (eds), {\em
  {Patterns and Skeletons for Parallel and Distributed Computing}}.
\newblock Springer-Verlag.

\bibitem[\protect\citename{Loogen, }2012]{Loogen2012}
Loogen, Rita. (2012).
\newblock {\em Eden -- parallel functional programming with {Haskell}}.
\newblock Springer.
\newblock Pages  142--206.

\bibitem[\protect\citename{Loogen {\em et~al.}\relax, }2005]{eden}
Loogen, Rita, Ortega-Mall\'{e}n, Yolanda, \& {Pe\~{n}a-Mar\'{\i}}, Ricardo.
  (2005).
\newblock {Parallel Functional Programming in Eden}.
\newblock {\em {Journal of Functional Programming}}, {\bf 15}(3), 431--475.
\newblock Special Issue on Functional Approaches to High-Performance Parallel
  Programming.

\bibitem[\protect\citename{Lämmel, }2008]{LAMMEL20081}
Lämmel, Ralf. (2008).
\newblock Google’s mapreduce programming model — revisited.
\newblock {\em Science of computer programming}, {\bf 70}(1), 1--30.

\bibitem[\protect\citename{Maier {\em et~al.}\relax,
  }2014]{Maier:2014:HDS:2775050.2633363}
Maier, Patrick, Stewart, Robert, \& Trinder, Phil. (2014).
\newblock The {HdpH} {DSLs} for scalable reliable computation.
\newblock {\em Sigplan not.}, {\bf 49}(12), 65--76.

\bibitem[\protect\citename{Mainland \& Morrisett,
  }2010]{Mainland:2010:NEC:2088456.1863533}
Mainland, Geoffrey, \& Morrisett, Greg. (2010).
\newblock Nikola: Embedding compiled {GPU} functions in {Haskell}.
\newblock {\em Sigplan not.}, {\bf 45}(11), 67--78.

\bibitem[\protect\citename{Marlow {\em et~al.}\relax, }2009]{Marlow2009}
Marlow, S., {Peyton Jones}, S., \& Singh, S. (2009).
\newblock Runtime support for multicore {Haskell}.
\newblock {\em Acm sigplan notices}, {\bf 44}(9), 65--78.

\bibitem[\protect\citename{Marlow, }2013]{marlow2013parallel}
Marlow, Simon. (2013).
\newblock {\em Parallel and concurrent programming in {Haskell}: Techniques for
  multicore and multithreaded programming}.
\newblock "O'Reilly Media, Inc.".

\bibitem[\protect\citename{Marlow {\em et~al.}\relax, }2011]{par-monad}
Marlow, Simon, Newton, Ryan, \& {Peyton Jones}, Simon. (2011).
\newblock A monad for deterministic parallelism.
\newblock {\em Sigplan not.}, {\bf 46}(12), 71--82.

\bibitem[\protect\citename{McDonell {\em et~al.}\relax,
  }2015]{McDonell:2015:TRC:2887747.2804313}
McDonell, Trevor~L., Chakravarty, Manuel M.~T., Grover, Vinod, \& Newton,
  Ryan~R. (2015).
\newblock Type-safe runtime code generation: Accelerate to {LLVM}.
\newblock {\em Sigplan not.}, {\bf 50}(12), 201--212.

\bibitem[\protect\citename{Michael {\em et~al.}\relax,
  }2009]{Michael:2009:IWS:1594835.1504186}
Michael, Maged~M., Vechev, Martin~T., \& Saraswat, Vijay~A. (2009).
\newblock Idempotent work stealing.
\newblock {\em Sigplan not.}, {\bf 44}(4), 45--54.

\bibitem[\protect\citename{Nilsson {\em et~al.}\relax,
  }2002]{Nilsson:2002:FRP:581690.581695}
Nilsson, Henrik, Courtney, Antony, \& Peterson, John. (2002).
\newblock Functional reactive programming, continued.
\newblock {\em Pages  51--64 of:} {\em Proceedings of the 2002 {ACM} {SIGPLAN}
  Workshop on {Haskell}}.
\newblock Haskell '02.
\newblock New York, NY, USA: ACM.

\bibitem[\protect\citename{Olivier \& Prins, }2008]{4625841}
Olivier, S., \& Prins, J. (2008).
\newblock Scalable dynamic load balancing using {UPC}.
\newblock {\em Pages  123--131 of:} {\em 37\xth International Conference on
  Parallel Processing}.

\bibitem[\protect\citename{Paterson, }2001]{Paterson:2001:NNA:507669.507664}
Paterson, Ross. (2001).
\newblock A new notation for arrows.
\newblock {\em Sigplan not.}, {\bf 36}(10), 229--240.

\bibitem[\protect\citename{Pe{\~n}a \& Rubio, }2001]{Eden:PPDP01}
Pe{\~n}a, R., \& Rubio, F. (2001).
\newblock {Parallel Functional Programming at Two Levels of Abstraction}.
\newblock {\em Pages  187--198 of:} {\em PPDP'01 --- Intl.\ Conf.\ on
  Principles and Practice of Declarative Programming}.

\bibitem[\protect\citename{Perfumo {\em et~al.}\relax,
  }2008]{Perfumo:2008:LST:1366230.1366241}
Perfumo, Cristian, S\"{o}nmez, Nehir, Stipic, Srdjan, Unsal, Osman, Cristal,
  Adri\'{a}n, Harris, Tim, \& Valero, Mateo. (2008).
\newblock The limits of software transactional memory ({STM}): Dissecting
  {Haskell} {STM} applications on a many-core environment.
\newblock {\em Pages  67--78 of:} {\em Proceedings of the 5\xth Conference on
  Computing Frontiers}.
\newblock CF '08.
\newblock ACM.

\bibitem[\protect\citename{Poldner \& Kuchen, }2005]{Kuchen05}
Poldner, Michael, \& Kuchen, Herbert. (2005).
\newblock Scalable farms.
\newblock {\em Pages  795--802 of:} Joubert, Gerhard~R., Nagel, Wolfgang~E.,
  Peters, Frans~J., Plata, Oscar~G., Tirado, P., \& Zapata, Emilio~L. (eds),
  {\em PARCO}.
\newblock John von Neumann Institute for Computing Series, vol. 33.
\newblock Central Institute for Applied Mathematics, J{\"u}lich, Germany.

\bibitem[\protect\citename{Priebe, }2006]{WPEuropar06}
Priebe, Steffen. (2006).
\newblock Dynamic task generation and transformation within a nestable workpool
  skeleton.
\newblock  {\em Euro-Par}.
\newblock LNCS 4128.

\bibitem[\protect\citename{Rabhi \& Gorlatch, }2003]{SkeletonBook}
Rabhi, F.~A., \& Gorlatch, S. (eds). (2003).
\newblock {\em {Patterns and Skeletons for Parallel and Distributed
  Computing}}.
\newblock Springer-Verlag.

\bibitem[\protect\citename{Rudolph {\em et~al.}\relax,
  }1991]{Rudolph:1991:SLB:113379.113401}
Rudolph, Larry, Slivkin-Allalouf, Miriam, \& Upfal, Eli. (1991).
\newblock A simple load balancing scheme for task allocation in parallel
  machines.
\newblock {\em Pages  237--245 of:} {\em Proceedings of the 3\xrd Annual ACM
  Symposium on Parallel Algorithms and Architectures}.
\newblock SPAA '91.
\newblock ACM.

\bibitem[\protect\citename{Russo {\em et~al.}\relax,
  }2008]{Russo:2008:LLI:1411286.1411289}
Russo, Alejandro, Claessen, Koen, \& Hughes, John. (2008).
\newblock A library for light-weight information-flow security in {Haskell}.
\newblock {\em Pages  13--24 of:} {\em Proceedings of the 1\st ACM SIGPLAN
  Symposium on Haskell}.
\newblock Haskell '08.
\newblock ACM.

\bibitem[\protect\citename{Stewart {\em et~al.}\relax,
  }2016]{stewart_maier_trinder_2016}
Stewart, Robert, Maier, Patrick, \& Trinder, Phil. (2016).
\newblock Transparent fault tolerance for scalable functional computation.
\newblock {\em Journal of functional programming}, {\bf 26}.

\bibitem[\protect\citename{Svensson, }2011]{obsidian-phd}
Svensson, Joel. (2011).
\newblock {\em Obsidian: Gpu kernel programming in haskell}.
\newblock Ph.D. thesis, Chalmers University of Technology.

\bibitem[\protect\citename{Trinder {\em et~al.}\relax, }1996]{Trinder1996}
Trinder, Phil~W., Hammond, Kevin, {Mattson Jr.}, James~S., Partridge,
  Andrew~S., \& {Peyton Jones}, Simon~L. (1996).
\newblock {GUM: a Portable Parallel Implementation of Haskell}.
\newblock  {\em {PLDI'96}}.
\newblock ACM Press.

\bibitem[\protect\citename{Trinder {\em et~al.}\relax, }1998]{Trinder1998a}
Trinder, P.W., Hammond, K., Loidl, H-W., \& {Peyton Jones}, S. (1998).
\newblock {Algorithm + Strategy = Parallelism}.
\newblock {\em Journal of functional programming}, {\bf 8}(1), 23--60.

\bibitem[\protect\citename{van Nieuwpoort {\em et~al.}\relax,
  }2001]{vanNieuwpoort:2001:ELB:568014.379563}
van Nieuwpoort, Rob~V., Kielmann, Thilo, \& Bal, Henri~E. (2001).
\newblock Efficient load balancing for wide-area divide-and-conquer
  applications.
\newblock {\em Sigplan not.}, {\bf 36}(7), 34--43.

\bibitem[\protect\citename{Vizzotto {\em et~al.}\relax,
  }2006]{vizzotto_altenkirch_sabry_2006}
Vizzotto, Juliana, Altenkirch, Thorsten, \& Sabry, Amr. (2006).
\newblock Structuring quantum effects: superoperators as arrows.
\newblock {\em Mathematical structures in computer science}, {\bf 16}(3),
  453--468.

\bibitem[\protect\citename{Wheeler \& Thain, }2009]{Wheeler2009}
Wheeler, K.~B., \& Thain, D. (2009).
\newblock Visualizing massively multithreaded applications with {ThreadScope}.
\newblock {\em Concurrency and computation: Practice and experience}, {\bf
  22}(1), 45--67.

\end{thebibliography}
        \appendix
	\section{Utility Arrows}\label{utilfns}
Following are definitions of some utility Arrows used in this paper that have been left out for brevity.
We start with the \ensuremath{\Varid{second}} combinator from \citet{HughesArrows}, which is a mirrored version of \ensuremath{\Varid{first}}, which is for example used in the definition of \ensuremath{\mathbin{*\!*\!*}}: 
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{second}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;(\Varid{c},\Varid{a})\;(\Varid{c},\Varid{b}){}\<[E]%
\\
\>[B]{}\Varid{second}\;\Varid{f}\mathrel{=}\Varid{arr}\;\Varid{swap}\mathbin{>\!\!>\!\!>}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{swap}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\mathbf{where}\;\Varid{swap}\;(\Varid{x},\Varid{y})\mathrel{=}(\Varid{y},\Varid{x}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Next, we give the definition of \ensuremath{\Varid{evalN}} which also helps us to define \ensuremath{\Varid{map}}, and \ensuremath{\Varid{zipWith}} on Arrows.
The \ensuremath{\Varid{evalN}} combinator in Fig.~\ref{fig:evalN} converts a list of Arrows \ensuremath{[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]} into an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{evalN}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{evalN}\;(\Varid{f}\mathbin{:}\Varid{fs})\mathrel{=}\Varid{arr}\;\Varid{listcase}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{10}{}\<[10]%
\>[10]{}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu])\mathbin{\mid\!\mid\!\mid}(\Varid{f}\mathbin{*\!*\!*}\Varid{evalN}\;\Varid{fs}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))){}\<[E]%
\\
\>[B]{}\hsindent{10}{}\<[10]%
\>[10]{}\mathbf{where}\;\Varid{listcase}\;[\mskip1.5mu \mskip1.5mu]{}\<[32]%
\>[32]{}\mathrel{=}\Conid{Left}\;(){}\<[E]%
\\
\>[10]{}\hsindent{6}{}\<[16]%
\>[16]{}\Varid{listcase}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Conid{Right}\;(\Varid{x},\Varid{xs}){}\<[E]%
\\
\>[B]{}\Varid{evalN}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}\Varid{arr}\;(\Varid{const}\;[\mskip1.5mu \mskip1.5mu]){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The definition of \ensuremath{\Varid{evalN}}}
\label{fig:evalN}
\end{figure}

The \ensuremath{\Varid{mapArr}} combinator (Fig.~\ref{fig:mapArr}) lifts any Arrow \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} to an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}. The original inspiration was from \citet{Hughes2005},
but the definition as then unified with \ensuremath{\Varid{evalN}}.

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{mapArr}\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{mapArr}\mathrel{=}\Varid{evalN}\mathbin{\circ}\Varid{repeat}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The definition of \ensuremath{\Varid{map}} over Arrows.}
\label{fig:mapArr}
\end{figure}

Finally, with the help of \ensuremath{\Varid{mapArr}} (Fig.~\ref{fig:mapArr}), we can define \ensuremath{\Varid{zipWithArr}}  (Fig.~\ref{fig:zipWithArr}) that lifts any Arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}} to an Arrow \ensuremath{\Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]}.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{zipWithArr}\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\Rightarrow \Varid{arr}\;(\Varid{a},\Varid{b})\;\Varid{c}\to \Varid{arr}\;([\mskip1.5mu \Varid{a}\mskip1.5mu],[\mskip1.5mu \Varid{b}\mskip1.5mu])\;[\mskip1.5mu \Varid{c}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{zipWithArr}\;\Varid{f}\mathrel{=}(\Varid{arr}\;(\lambda (\Varid{as},\Varid{bs})\to \Varid{zipWith}\;(,)\;\Varid{as}\;\Varid{bs}))\mathbin{>\!\!>\!\!>}\Varid{mapArr}\;\Varid{f}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{zipWith}} over Arrows.} %$
\label{fig:zipWithArr}
\end{figure}
 %$ %% formatting

These combinators make use of the \ensuremath{\Conid{ArrowChoice}} type class which provides the \ensuremath{\mathbin{\mid\!\mid\!\mid}} combinator. It takes two Arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{c}} and \ensuremath{\Varid{arr}\;\Varid{b}\;\Varid{c}} and combines them into a new Arrow \ensuremath{\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}} which pipes all \ensuremath{\Conid{Left}\;\Varid{a}}'s to the first Arrow and all \ensuremath{\Conid{Right}\;\Varid{b}}'s to the second Arrow:
% \begin{figure}[htb]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{\mid\!\mid\!\mid})\mathbin{::}\Conid{ArrowChoice}\;\Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{b}\;\Varid{c}\to \Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{b})\;\Varid{c}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\section{Profunctor Arrows}
\label{app:profunctorArrows}

In Fig.~\ref{fig:profunctorArrow} we show how specific Profunctors can be turned into Arrows. This works because Arrows are strong Monads in the bicategory \ensuremath{\Conid{Prof}} of Profunctors as shown by \citet{Asada:2010:ASM:1863597.1863607}. In Standard GHC \ensuremath{(\mathbin{>\!\!>\!\!>})} has the type \ensuremath{(\mathbin{>\!\!>\!\!>})\mathbin{::}\Conid{Category}\;\Varid{cat}\Rightarrow \Varid{cat}\;\Varid{a}\;\Varid{b}\to \Varid{cat}\;\Varid{b}\;\Varid{c}\to \Varid{cat}\;\Varid{a}\;\Varid{c}} and is therefore not part of the \ensuremath{\Conid{Arrow}} type class like presented in this paper.\footnote{For additional information on the type classes used, see: \url{https://hackage.haskell.org/package/profunctors-5.2.1/docs/Data-Profunctor.html} and \url{https://hackage.haskell.org/package/base-4.9.1.0/docs/Control-Category.html}.}

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;(\Conid{Category}\;\Varid{p},\Conid{Strong}\;\Varid{p})\Rightarrow \Conid{Arrow}\;\Varid{p}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{arr}\;\Varid{f}\mathrel{=}\Varid{dimap}\;\Varid{id}\;\Varid{f}\;\Varid{id}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{first}\mathrel{=}\Varid{first'}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\Conid{Category}\;\Varid{p},\Conid{Strong}\;\Varid{p},\Conid{Costrong}\;\Varid{p})\Rightarrow \Conid{ArrowLoop}\;\Varid{p}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{loop}\mathrel{=}\Varid{loop'}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\Conid{Category}\;\Varid{p},\Conid{Strong}\;\Varid{p},\Conid{Choice}\;\Varid{p})\Rightarrow \Conid{ArrowChoice}\;\Varid{p}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{left}\mathrel{=}\Varid{left'}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Profunctors as Arrows.}
\label{fig:profunctorArrow}
\end{figure}

\section{Additional function definitions}
\label{app:omitted}
We have omitted some function definitions in the main text for
brevity, and redeem this here.
%
%
%
We begin with warping Eden's build-in \ensuremath{\Conid{RemoteData}} to \ensuremath{\Conid{Future}} in
Figure~\ref{fig:RDFuture}

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{RemoteData}\;\Varid{a}\mathrel{=}\Conid{RD}\;\{\mskip1.5mu \Varid{rd}\mathbin{::}\Conid{RD}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{put'}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;\Varid{a}\;(\Conid{BasicFuture}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\Varid{put'}\mathrel{=}\Varid{arr}\;\Conid{BF}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{get'}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Conid{BasicFuture}\;\Varid{a})\;\Varid{a}{}\<[E]%
\\
\>[B]{}\Varid{get'}\mathrel{=}\Varid{arr}\;(\lambda (\mathord{\sim}(\Conid{BF}\;\Varid{a}))\to \Varid{a}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{NFData}\;(\Conid{RemoteData}\;\Varid{a})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{rnf}\mathrel{=}\Varid{rnf}\mathbin{\circ}\Varid{rd}{}\<[E]%
\\
\>[B]{}\mathbf{instance}\;\Conid{Trans}\;(\Conid{RemoteData}\;\Varid{a}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{RemoteData}\;\Varid{a}\;\Conid{Conf}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{put}\;\anonymous \mathrel{=}\Varid{put'}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{get}\;\anonymous \mathrel{=}\Varid{get'}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;(\Conid{Trans}\;\Varid{a})\Rightarrow \Conid{Future}\;\Conid{RemoteData}\;\Varid{a}\;()\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{put}\;\anonymous \mathrel{=}\Varid{put'}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{get}\;\anonymous \mathrel{=}\Varid{get'}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Conid{RD}}-based \ensuremath{\Conid{RemoteData}} version of \ensuremath{\Conid{Future}} for the Eden backend.}
\label{fig:RDFuture}
\end{figure}

Next, we have the definition of \ensuremath{\Conid{BasicFuture}} in Fig.~\ref{fig:BasicFuture} and the corresponding \ensuremath{\Conid{Future}} instances.

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{BasicFuture}\;\Varid{a}\mathrel{=}\Conid{BF}\;\Varid{a}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{put'}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;\Varid{a}\;(\Conid{BasicFuture}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\Varid{put'}\mathrel{=}\Varid{arr}\;\Conid{BF}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{get'}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Conid{BasicFuture}\;\Varid{a})\;\Varid{a}{}\<[E]%
\\
\>[B]{}\Varid{get'}\mathrel{=}\Varid{arr}\;(\lambda (\mathord{\sim}(\Conid{BF}\;\Varid{a}))\to \Varid{a}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{NFData}\;\Varid{a}\Rightarrow \Conid{NFData}\;(\Conid{BasicFuture}\;\Varid{a})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{rnf}\;(\Conid{BF}\;\Varid{a})\mathrel{=}\Varid{rnf}\;\Varid{a}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{Future}\;\Conid{BasicFuture}\;\Varid{a}\;(\Conid{Conf}\;\Varid{a})\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{put}\;\anonymous \mathrel{=}\Varid{put'}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{get}\;\anonymous \mathrel{=}\Varid{get'}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{Future}\;\Conid{BasicFuture}\;\Varid{a}\;()\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{put}\;\anonymous \mathrel{=}\Varid{put'}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{get}\;\anonymous \mathrel{=}\Varid{get'}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Conid{BasicFuture}} type and its \ensuremath{\Conid{Future}} instance for the \ensuremath{\Conid{Par}} Monad and GpH.}
\label{fig:BasicFuture}
\end{figure}

% \caption{NFData and Trans instances for the RemoteData type. The Trans instance does not have any functions declared as the default implementation suffices here. See \url{https://hackage.haskell.org/package/edenmodules-1.2.0.0/docs/Control-Parallel-Eden.html\#g:5} for more information.}
% \end{figure}

Figures~\ref{fig:parMapImg}--\ref{fig:parMapStream} show the definitions and a visualisations of two parallel \ensuremath{\Varid{map}} variants, defined using \ensuremath{\Varid{parEvalN}} and its lazy counterpart.

\begin{figure}[thb]
%parMap
\includegraphics[scale=0.7]{images/parMap}
\caption{\ensuremath{\Varid{parMap}} depiction.}
\label{fig:parMapImg}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parMap}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf})\Rightarrow \Varid{conf}\to (\Varid{arr}\;\Varid{a}\;\Varid{b})\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{parMap}\;\Varid{conf}\;\Varid{f}\mathrel{=}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{repeat}\;\Varid{f}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Definition of parMap.}
\label{fig:parMap}

%parMapStream
\includegraphics[scale=0.7]{images/parMapStream}
\caption{\ensuremath{\Varid{parMapStream}} depiction.}
\label{fig:parMapStreamImg}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parMapStream}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parMapStream}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{f}\mathrel{=}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;\Varid{f}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{parMapStream}} definition.}
\label{fig:parMapStream}
\end{figure}

Arrow versions of Eden's \ensuremath{\Varid{shuffle}}, \ensuremath{\Varid{unshuffle}} and the definition of \ensuremath{\Varid{takeEach}} are in Figure~\ref{fig:edenshuffleetc}. Similarly, Figure~\ref{fig:edenlazyrightrotate} contains the definition of Arrow versions of Eden's \ensuremath{\Varid{lazy}} and \ensuremath{\Varid{rightRotate}} utility functions. Fig.~\ref{fig:lazyzip3etc} contains Eden's definition of \ensuremath{\Varid{lazyzip3}} together with the utility functions \ensuremath{\Varid{uncurry3}} and \ensuremath{\Varid{threetotwo}}.
The full definition of \ensuremath{\Varid{farmChunk}} is in Figure~\ref{fig:farmChunk}.
Eden definition of \ensuremath{\Varid{ring}} skeleton is in Figure~\ref{fig:ringEden}. It
follows \citet{Loogen2012}.

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{shuffle}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{shuffle}\mathrel{=}\Varid{arr}\;(\Varid{concat}\mathbin{\circ}\Varid{transpose}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{unshuffle}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Conid{Int}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{unshuffle}\;\Varid{n}\mathrel{=}\Varid{arr}\;(\lambda \Varid{xs}\to [\mskip1.5mu \Varid{takeEach}\;\Varid{n}\;(\Varid{drop}\;\Varid{i}\;\Varid{xs})\mid \Varid{i}\leftarrow [\mskip1.5mu \mathrm{0}\mathinner{\ldotp\ldotp}\Varid{n}\mathbin{-}\mathrm{1}\mskip1.5mu]\mskip1.5mu]){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{takeEach}\mathbin{::}\Conid{Int}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{takeEach}\;\Varid{n}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{takeEach}\;\Varid{n}\;(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{takeEach}\;\Varid{n}\;(\Varid{drop}\;(\Varid{n}\mathbin{-}\mathrm{1})\;\Varid{xs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{shuffle}}, \ensuremath{\Varid{unshuffle}}, \ensuremath{\Varid{takeEach}} definition.}
\label{fig:edenshuffleetc}
\end{figure}

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{lazy}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{lazy}\mathrel{=}\Varid{arr}\;(\lambda \mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\to \Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{rightRotate}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{rightRotate}\mathrel{=}\Varid{arr}\mathbin{\$}\lambda \Varid{list}\to \mathbf{case}\;\Varid{list}\;\mathbf{of}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}[\mskip1.5mu \mskip1.5mu]\to [\mskip1.5mu \mskip1.5mu]{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{xs}\to \Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{lazy}} and \ensuremath{\Varid{rightRotate}} definitions.}
\label{fig:edenlazyrightrotate}
\end{figure}

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{lazyzip3}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{c}\mskip1.5mu]\to [\mskip1.5mu (\Varid{a},\Varid{b},\Varid{c})\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{lazyzip3}\;\Varid{as}\;\Varid{bs}\;\Varid{cs}\mathrel{=}\Varid{zip3}\;\Varid{as}\;(\Varid{lazy}\;\Varid{bs})\;(\Varid{lazy}\;\Varid{cs}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{uncurry3}\mathbin{::}(\Varid{a}\to \Varid{b}\to \Varid{c}\to \Varid{d})\to (\Varid{a},(\Varid{b},\Varid{c}))\to \Varid{d}{}\<[E]%
\\
\>[B]{}\Varid{uncurry3}\;\Varid{f}\;(\Varid{a},(\Varid{b},\Varid{c}))\mathrel{=}\Varid{f}\;\Varid{a}\;\Varid{b}\;\Varid{c}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{threetotwo}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow \Varid{arr}\;(\Varid{a},\Varid{b},\Varid{c})\;(\Varid{a},(\Varid{b},\Varid{c})){}\<[E]%
\\
\>[B]{}\Varid{threetotwo}\mathrel{=}\Varid{arr}\mathbin{\$}\lambda \mathord{\sim}(\Varid{a},\Varid{b},\Varid{c})\to (\Varid{a},(\Varid{b},\Varid{c})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{lazyzip3}}, \ensuremath{\Varid{uncurry3}} and \ensuremath{\Varid{threetotwo}} definitions.}
\label{fig:lazyzip3etc}
\end{figure}

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{farmChunk}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowParallel}\;\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\;\Varid{conf},{}\<[E]%
\\
\>[B]{}\hsindent{14}{}\<[14]%
\>[14]{}\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to \Conid{NumCores}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{farmChunk}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{numCores}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{unshuffle}\;\Varid{numCores}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;(\Varid{repeat}\;(\Varid{mapArr}\;\Varid{f}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{shuffle}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{\ensuremath{\Varid{farmChunk}} definition.}
\label{fig:farmChunk}
\end{figure}

\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ringSimple}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow (\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r}))\to [\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{ringSimple}\;\Varid{f}\;\Varid{is}\mathrel{=}{}\<[20]%
\>[20]{}\Varid{os}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathbf{where}\;(\Varid{os},\Varid{ringOuts})\mathrel{=}\Varid{unzip}\;(\Varid{parMap}\;(\Varid{toRD}\mathbin{\$}\Varid{uncurry}\;\Varid{f})\;(\Varid{zip}\;\Varid{is}\mathbin{\$}\Varid{lazy}\;\Varid{ringIns})){}\<[E]%
\\
\>[3]{}\hsindent{6}{}\<[9]%
\>[9]{}\Varid{ringIns}\mathrel{=}\Varid{rightRotate}\;\Varid{ringOuts}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{toRD}\mathbin{::}(\Conid{Trans}\;\Varid{i},\Conid{Trans}\;\Varid{o},\Conid{Trans}\;\Varid{r})\Rightarrow ((\Varid{i},\Varid{r})\to (\Varid{o},\Varid{r}))\to ((\Varid{i},\Conid{RD}\;\Varid{r})\to (\Varid{o},\Conid{RD}\;\Varid{r})){}\<[E]%
\\
\>[B]{}\Varid{toRD}\;{}\<[7]%
\>[7]{}\Varid{f}\;(\Varid{i},\Varid{ringIn}){}\<[22]%
\>[22]{}\mathrel{=}(\Varid{o},\Varid{release}\;\Varid{ringOut}){}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathbf{where}\;(\Varid{o},\Varid{ringOut})\mathrel{=}\Varid{f}\;(\Varid{i},\Varid{fetch}\;\Varid{ringIn}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{rightRotate}{}\<[16]%
\>[16]{}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{rightRotate}\;[\mskip1.5mu \mskip1.5mu]\mathrel{=}{}\<[19]%
\>[19]{}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{rightRotate}\;\Varid{xs}\mathrel{=}{}\<[19]%
\>[19]{}\Varid{last}\;\Varid{xs}\mathbin{:}\Varid{init}\;\Varid{xs}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{lazy}\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{lazy}\mathord{\sim}(\Varid{x}\mathbin{:}\Varid{xs})\mathrel{=}\Varid{x}\mathbin{:}\Varid{lazy}\;\Varid{xs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Eden's definition of the \ensuremath{\Varid{ring}} skeleton.}
\label{fig:ringEden}
\end{figure}

The \ensuremath{\Varid{parEval2}} skeleton is defined in Figure~\ref{fig:parEval2}. 
We start by transforming the \ensuremath{(\Varid{a},\Varid{c})} input into a two-element list \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{a}\;\Varid{c}\mskip1.5mu]} by first tagging the two inputs with \ensuremath{\Conid{Left}} and \ensuremath{\Conid{Right}} and wrapping the right element in a singleton list with \ensuremath{\Varid{return}} so that we can combine them with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))}. Next, we feed this list into a parallel Arrow running on two instances of \ensuremath{\Varid{f}\mathbin{+\!\!+\!\!+}\Varid{g}} as described in the paper. After the calculation is finished, we convert the resulting \ensuremath{[\mskip1.5mu \Conid{Either}\;\Varid{b}\;\Varid{d}\mskip1.5mu]} into \ensuremath{([\mskip1.5mu \Varid{b}\mskip1.5mu],[\mskip1.5mu \Varid{d}\mskip1.5mu])} with \ensuremath{\Varid{arr}\;\Varid{partitionEithers}}. The two lists in this tuple contain only one element each by construction, so we can finally just convert the tuple to \ensuremath{(\Varid{b},\Varid{d})} in the last step.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEval2}\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;\Varid{conf})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
\\
\>[B]{}\Varid{parEval2}\;\Varid{conf}\;\Varid{f}\;\Varid{g}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Conid{Left}\mathbin{*\!*\!*}(\Varid{arr}\;\Conid{Right}\mathbin{>\!\!>\!\!>}\Varid{arr}\;\Varid{return})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{uncurry}\;(\mathbin{:}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{parEvalN}\;\Varid{conf}\;(\Varid{replicate}\;\mathrm{2}\;(\Varid{f}\mathbin{+\!\!+\!\!+}\Varid{g}))\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{partitionEithers}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{head}\mathbin{*\!*\!*}\Varid{arr}\;\Varid{head}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
	\caption{\ensuremath{\Varid{parEval2}} definition.}
	\label{fig:parEval2}
\end{figure}
Furthermore, Fig.~\ref{fig:torus_example_rest} contains the omitted definitions of \ensuremath{\Varid{prMMTr}} (which calculates $AB^T$ for two matrices $A$ and $B$), \ensuremath{\Varid{splitMatrix}} (which splits the a matrix into chunks), and lastly \ensuremath{\Varid{matAdd}}, that calculates $A + B$ for two matrices $A$ and $B$.
\begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{prMMTr}\;\Varid{m1}\;\Varid{m2}\mathrel{=}[\mskip1.5mu [\mskip1.5mu \Varid{sum}\;(\Varid{zipWith}\;(\mathbin{*})\;\Varid{row}\;\Varid{col})\mid \Varid{col}\leftarrow \Varid{m2}\mskip1.5mu]\mid \Varid{row}\leftarrow \Varid{m1}\mskip1.5mu]{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{splitMatrix}\mathbin{::}\Conid{Int}\to \Conid{Matrix}\to [\mskip1.5mu [\mskip1.5mu \Conid{Matrix}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{splitMatrix}\;\Varid{size}\;\Varid{matrix}\mathrel{=}\Varid{map}\;(\Varid{transpose}\mathbin{\circ}\Varid{map}\;(\Varid{chunksOf}\;\Varid{size}))\mathbin{\$}\Varid{chunksOf}\;\Varid{size}\mathbin{\$}\Varid{matrix}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{matAdd}\mathrel{=}\Varid{chunksOf}\;(\Varid{dimX}\;\Varid{x})\mathbin{\$}\Varid{zipWith}\;(\mathbin{+})\;(\Varid{concat}\;\Varid{x})\;(\Varid{concat}\;\Varid{y}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
	\caption{\ensuremath{\Varid{prMMTr}}, \ensuremath{\Varid{splitMatrix}} and \ensuremath{\Varid{matAdd}} definition.}
	\label{fig:torus_example_rest}
\end{figure} %$

\section{Syntactic sugar} \label{syntacticSugar}
Finally, we also give the definitions for some syntactic sugar for PArrows, namely \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} and \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}}.
For basic Arrows, we have the \ensuremath{\mathbin{*\!*\!*}} combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two Arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into an Arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} which does both computations at once. This can easily be translated into a parallel version \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} with the use of \ensuremath{\Varid{parEval2}}, but for this we require a backend which has an implementation that does not require any configuration (hence the \ensuremath{()} as the \ensuremath{\Varid{conf}} parameter% in Fig.~\ref{fig:par***}
):
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{\mid\!\!*\!*\!*\!\!\mid})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{c})\;(\Conid{Either}\;\Varid{b}\;\Varid{d})\;()))\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
\\
\>[B]{}(\mathbin{\mid\!\!*\!*\!*\!\!\mid})\mathrel{=}\Varid{parEval2}\;(){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{Definition of |parstar|, the parallel version of |***|.}
% \label{fig:par***}
% \end{figure}
% With this we can analogously to the serial |&&&|
We define the parallel \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}} % (Fig.~\ref{fig:par&&&}) 
in a similar manner to its sequential pendant \ensuremath{\mathbin{\&\!\&\!\&}} (Fig.~\ref{fig:syntacticSugarArrows}):
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid})\mathbin{::}(\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowParallel}\;\Varid{arr}\;(\Conid{Either}\;\Varid{a}\;\Varid{a})\;(\Conid{Either}\;\Varid{b}\;\Varid{c})\;())\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
\\
\>[B]{}(\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid})\;\Varid{f}\;\Varid{g}\mathrel{=}(\Varid{arr}\mathbin{\$}\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{\mid\!\!*\!*\!*\!\!\mid}\Varid{g}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 % $ %% formatting
% \caption{Definition of |parand| - the parallel version of |&&&|.}
% \label{fig:par&&&}
% \end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
\end{document}

### 2018-01-07 paper/content/ms.tex
image references

@@
 \centering
-	\includegraphics{images/arrow}
+	\includegraphics{arrow}
 \vfill
@@
 \centering
-	{\includegraphics[scale=0.6]{images/arr}}
-	{\includegraphics[scale=0.6]{images/compose}}
-	{\includegraphics[scale=0.6]{images/first}}
+	{\includegraphics[scale=0.6]{arr}}
+	{\includegraphics[scale=0.6]{compose}}
+	{\includegraphics[scale=0.6]{first}}
 \vfill
@@
 		% \subcaptionbox
-{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
+{\label{t1}}{\includegraphics[width = 1.5in]{first}} &
 		% \subcaptionbox
-{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
+{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{second}} \\
 \ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
@@
 		% \subcaptionbox
-{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
+{}{\includegraphics[width = 1.5in]{starstarstar}} &
 		% \subcaptionbox
-{}{\includegraphics[width = 1.5in]{images/andandand}}\\
+{}{\includegraphics[width = 1.5in]{andandand}}\\
 \ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\
@@
   \centering
-	\includegraphics[scale=0.7]{images/parEvalN}
+	\includegraphics[scale=0.7]{parEvalN}
 	\caption{Schematic illustration of \ensuremath{\Varid{parEvalN}}. A list of inputs is transformed by different functions in parallel.}
@@
 \begin{figure}[t]
-	\includegraphics[scale=0.5]{images/parEvalNMulticore}
+	\includegraphics[scale=0.5]{parEvalNMulticore}
 	\caption{\ensuremath{\Varid{parEvalN}} (GpH).}
@@
 \begin{figure}[t]
-	\includegraphics[scale=0.5]{images/parEvalNParMonad}
+	\includegraphics[scale=0.5]{parEvalNParMonad}
 	\caption{\ensuremath{\Varid{parEvalN}} (\ensuremath{\Conid{Par}} Monad).}
@@
 \begin{figure}[tb]
-	\includegraphics[scale=0.7]{images/parEvalNLazy}
+	\includegraphics[scale=0.7]{parEvalNLazy}
 	\caption{\ensuremath{\Varid{parEvalNLazy}} depiction.}
@@
 \begin{figure}[tb]
-	\includegraphics[scale=0.7]{images/parEval2}
+	\includegraphics[scale=0.7]{parEval2}
 	\caption{\ensuremath{\Varid{parEval2}} depiction.}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
+	\includegraphics[width=0.9\textwidth]{withoutFutures}
 	\caption[without Futures]{Communication between 4 Eden processes without Futures.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{images/withFutures}
+	\includegraphics[width=0.9\textwidth]{withFutures}
 	\caption[with Futures]{Communication between 4 Eden processes with Futures.
@@
 %farm
-\includegraphics[scale=0.7]{images/farm}
+\includegraphics[scale=0.7]{farm}
 \caption{\ensuremath{\Varid{farm}} depiction.}
@@
 %farmChunk
-\includegraphics[scale=0.7]{images/farmChunk}
+\includegraphics[scale=0.7]{farmChunk}
 \caption{\ensuremath{\Varid{farmChunk}} depiction.}
@@
 \begin{figure}[tb]
-	\includegraphics[scale=0.75]{images/ring}
+	\includegraphics[scale=0.75]{ring}
 	\caption{\ensuremath{\Varid{ring}} skeleton depiction.}
@@
 \begin{figure}
-	\includegraphics[scale=0.75]{images/torus}
+	\includegraphics[scale=0.75]{torus}
 	\caption{\ensuremath{\Varid{torus}} skeleton depiction.}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
+	\includegraphics[width=0.9\textwidth]{torus_matrix_parrows_scale}
 	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (PArrows)]{Matrix multiplication with \ensuremath{\Varid{torus}} (PArrows).}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
+	\includegraphics[width=0.9\textwidth]{torus_matrix_eden_scale}
 	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (Eden)]{Matrix multiplication with \ensuremath{\Varid{torus}} (Eden).}
@@
 %parMap
-\includegraphics[scale=0.7]{images/parMap}
+\includegraphics[scale=0.7]{parMap}
 \caption{\ensuremath{\Varid{parMap}} depiction.}
@@
 %parMapStream
-\includegraphics[scale=0.7]{images/parMapStream}
+\includegraphics[scale=0.7]{parMapStream}
 \caption{\ensuremath{\Varid{parMapStream}} depiction.}

### 2018-01-07 paper/content/arxiv/bestAndWorstBenchmarks.tex
oops, one reference still there: csv

--- 2018-01-06 paper/content/bestAndWorstBenchmarks.tex
@@
 \centering
-\caption{Overhead in the shared memory benchmarks. Bold marks values
-  in favour of PArrows. Overhead is unit-less. Runtime is in seconds.}
+\caption{Overhead in the shared memory benchmarks. Bold marks values in favour of PArrows.}
 \label{tab:meanOverheadSharedMemory}
@@
 \begingroup\catcode`"=9
+\sisetup{round-mode=places, round-precision=3}
 \begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
@@
 	\multirow{3}{*}{\sudokutest 1000}
-	\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\csvreader[head to column names]{bestAndWorstSudoku-1000.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 	\\ \hline \\
 	\multirow{1}{*}{\torustest 512}
-	\csvreader[head to column names]{benchmarks/torus-sm/bestAndWorstTorusSM-512.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\csvreader[head to column names]{bestAndWorstTorusSM-512.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 	\\ \hline \\
 	\multirow{3}{*}{\rmtest 11213 32}
-	\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\csvreader[head to column names]{bestAndWorstRMSM-11213-32.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 	\\ \hline \\
 	\multirow{3}{*}{\rmtest 11213 64}
-	\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\csvreader[head to column names]{bestAndWorstRMSM-11213-64.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 \end{tabular}
@@
 \caption{Overhead in the distributed memory benchmarks. Bold marks values
-  in favour of PArrows. Overhead is unit-less. Runtime is in seconds.}
-\olcomment{Vielleicht gleich durchmultiplizieren und \% angeben? Nicht
-zu viele leere Zeilen in der Tabelle?}
+  in favour of PArrows.}
+% \olcomment{Vielleicht gleich durchmultiplizieren und \% angeben? Nicht
+% zu viele leere Zeilen in der Tabelle?}
 \label{tab:meanOverHeadDistributedMemory}
@@
 	\multirow{1}{*}{\torustest 4096}
-	\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\csvreader[head to column names]{bestAndWorstTorus-4096.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 	\\ \hline \\
   	\multirow{1}{*}{\rmtest 44497 256}
-	\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\csvreader[head to column names]{bestAndWorstRM-44497-256.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 	\\ \hline \\
 	\multirow{1}{*}{\jacobitest 3217}
-	\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-3217.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\csvreader[head to column names]{bestAndWorstJacobi-3217.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
 \end{tabular}

### 2018-01-07 paper/content/arxiv/ms.tex
oops, one reference still there: csv

--- 2018-01-07 paper/content/ms.tex
@@
 	\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-	smooth]{\benchmarkDir/distributed-rm/bench-distributed.bench.skelrm-parrows-44497-256.csv};
+	smooth]{bench-distributed.bench.skelrm-parrows-44497-256.csv};
 	}{272}{\plotwidthDist}{272}

### 2018-01-07 paper/content/arxiv/ms.tex
minor edits in biblio

@@
 \newblock The design and implementation of {GUMSMP}: A multilevel parallel
-  haskell implementation.
+  Haskell implementation.
 \newblock {\em Pages  37:37--37:48 of:} {\em Proceedings of the 25\xth
@@
   Jones}, Simon, \& Lippmeier, Ben. (2010).
-\newblock Regular, shape-polymorphic, parallel arrays in haskell.
+\newblock Regular, shape-polymorphic, parallel arrays in Haskell.
 \newblock {\em Sigplan not.}, {\bf 45}(9), 261--272.
@@
 Svensson, Joel. (2011).
-\newblock {\em Obsidian: Gpu kernel programming in haskell}.
+\newblock {\em Obsidian: GPU kernel programming in Haskell}.
 \newblock Ph.D. thesis, Chalmers University of Technology.

### 2018-01-07 paper/content/motivation.tex
minor minor things

@@
 the parallel implementations at will.
-The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}).
+The induced penalty is in the single-digit percent range, with means typically under 2\% overhead in measurements over the varying cores configuration (Section~\ref{sec:benchmarks}).
 Further implementations, based on HdpH or a Frege implementation (on the Java Virtual Machine), are viable, too.
@@
 \item We introduce a parallel evaluation formalism using Arrows.
-One big advantage of our specific approach is that we do not have to introduce any new types, facilitating composability (Section~\ref{sec:parallel-arrows}).
+One big advantage of this specific approach is that we do not have to introduce any new types, facilitating composability (Section~\ref{sec:parallel-arrows}).
 \item We show that PArrow programs can readily exploit multiple parallel language implementations.
 We demonstrate the use of GpH, a |Par| Monad, and Eden.
-We do not re-implement all the parallel internals, as we host this functionality in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
+We do not re-implement all the parallel internals, as this functionality is hosted in the |ArrowParallel| type class, which abstracts all parallel implementation logic.
 The implementations can easily be swapped, so we are not bound to any specific one.
@@
 \item We practically demonstrate that Arrow parallelism has a low performance overhead compared with existing approaches, \eg the mean over all cores of relative mean overhead was less than $3.5\%$ and less than $0.8\%$ for all benchmarks with GpH and Eden, respectively.
-As for |Par| Monad, the mean of mean overheads was in our favour in all benchmarks (Section~\ref{sec:benchmarks}).
+As for |Par| Monad, the mean of mean overheads was in favour of PArrows in all benchmarks (Section~\ref{sec:benchmarks}).
 \end{itemize}

### 2018-01-07 paper/content/arxiv/ms.tex
re-generating arxiv base file

--- 2018-01-07 paper/content/ms.tex
@@
 \centering
-	\includegraphics{arrow}
+	\includegraphics{images/arrow}
 \vfill
@@
 \centering
-	{\includegraphics[scale=0.6]{arr}}
-	{\includegraphics[scale=0.6]{compose}}
-	{\includegraphics[scale=0.6]{first}}
+	{\includegraphics[scale=0.6]{images/arr}}
+	{\includegraphics[scale=0.6]{images/compose}}
+	{\includegraphics[scale=0.6]{images/first}}
 \vfill
@@
 		% \subcaptionbox
-{\label{t1}}{\includegraphics[width = 1.5in]{first}} &
+{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
 		% \subcaptionbox
-{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{second}} \\
+{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
 \ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
@@
 		% \subcaptionbox
-{}{\includegraphics[width = 1.5in]{starstarstar}} &
+{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
 		% \subcaptionbox
-{}{\includegraphics[width = 1.5in]{andandand}}\\
+{}{\includegraphics[width = 1.5in]{images/andandand}}\\
 \ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\
@@
   \centering
-	\includegraphics[scale=0.7]{parEvalN}
+	\includegraphics[scale=0.7]{images/parEvalN}
 	\caption{Schematic illustration of \ensuremath{\Varid{parEvalN}}. A list of inputs is transformed by different functions in parallel.}
@@
 \begin{figure}[t]
-	\includegraphics[scale=0.5]{parEvalNMulticore}
+	\includegraphics[scale=0.5]{images/parEvalNMulticore}
 	\caption{\ensuremath{\Varid{parEvalN}} (GpH).}
@@
 \begin{figure}[t]
-	\includegraphics[scale=0.5]{parEvalNParMonad}
+	\includegraphics[scale=0.5]{images/parEvalNParMonad}
 	\caption{\ensuremath{\Varid{parEvalN}} (\ensuremath{\Conid{Par}} Monad).}
@@
 \begin{figure}[tb]
-	\includegraphics[scale=0.7]{parEvalNLazy}
+	\includegraphics[scale=0.7]{images/parEvalNLazy}
 	\caption{\ensuremath{\Varid{parEvalNLazy}} depiction.}
@@
 \begin{figure}[tb]
-	\includegraphics[scale=0.7]{parEval2}
+	\includegraphics[scale=0.7]{images/parEval2}
 	\caption{\ensuremath{\Varid{parEval2}} depiction.}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{withoutFutures}
+	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
 	\caption[without Futures]{Communication between 4 Eden processes without Futures.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{withFutures}
+	\includegraphics[width=0.9\textwidth]{images/withFutures}
 	\caption[with Futures]{Communication between 4 Eden processes with Futures.
@@
 %farm
-\includegraphics[scale=0.7]{farm}
+\includegraphics[scale=0.7]{images/farm}
 \caption{\ensuremath{\Varid{farm}} depiction.}
@@
 %farmChunk
-\includegraphics[scale=0.7]{farmChunk}
+\includegraphics[scale=0.7]{images/farmChunk}
 \caption{\ensuremath{\Varid{farmChunk}} depiction.}
@@
 \begin{figure}[tb]
-	\includegraphics[scale=0.75]{ring}
+	\includegraphics[scale=0.75]{images/ring}
 	\caption{\ensuremath{\Varid{ring}} skeleton depiction.}
@@
 \begin{figure}
-	\includegraphics[scale=0.75]{torus}
+	\includegraphics[scale=0.75]{images/torus}
 	\caption{\ensuremath{\Varid{torus}} skeleton depiction.}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{torus_matrix_parrows_scale}
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
 	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (PArrows)]{Matrix multiplication with \ensuremath{\Varid{torus}} (PArrows).}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{torus_matrix_eden_scale}
+	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
 	\caption[Matrix multiplication with \ensuremath{\Varid{torus}} (Eden)]{Matrix multiplication with \ensuremath{\Varid{torus}} (Eden).}
@@
 
-\input{bestAndWorstBenchmarks}
+%input bestAndWorstBenchmarks.tex
 
@@
 Svensson, Joel. (2011).
-\newblock {\em Obsidian: Gpu kernel programming in haskell}.
+\newblock {\em Obsidian: {GPU} kernel programming in {Haskell}}.
 \newblock Ph.D. thesis, Chalmers University of Technology.
@@
 %parMap
-\includegraphics[scale=0.7]{parMap}
+\includegraphics[scale=0.7]{images/parMap}
 \caption{\ensuremath{\Varid{parMap}} depiction.}
@@
 %parMapStream
-\includegraphics[scale=0.7]{parMapStream}
+\includegraphics[scale=0.7]{images/parMapStream}
 \caption{\ensuremath{\Varid{parMapStream}} depiction.}

### 2018-01-07 paper/content/arxiv/ms.tex
arXiv images

--- 2018-01-07 paper/content/arxiv/ms.tex
@@
 % \begin{figure}[h]
-% 	\includegraphics[scale=0.5]{images/parEvalNEden}
+% 	\includegraphics[scale=0.5]{parEvalNEden}
 % 	\caption{Dataflow of the Eden |parEvalN| version.}
@@
 	\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-	smooth]{bench-distributed.bench.skelrm-parrows-44497-256.csv};
+	smooth]{\benchmarkDir/distributed-rm/bench-distributed.bench.skelrm-parrows-44497-256.csv};
 	}{272}{\plotwidthDist}{272}
@@
 \newblock Arrows are strong monads.
-\newblock {\em Pages  33--42 of:} {\em Proceedings of the Third ACM SIGPLAN
+\newblock {\em Pages  33--42 of:} {\em Proceedings of the Third {ACM SIGPLAN}
   Workshop on Mathematically Structured Functional Programming}.
@@
 \newblock Asynchronous functional reactive programming for guis.
-\newblock {\em Sigplan not.}, {\bf 48}(6), 411--422.
+\newblock {\em {SIGPLAN} not.}, {\bf 48}(6), 411--422.
 
@@
   heterogeneous cloud.
-\newblock {\em Sigplan not.}, {\bf 47}(9), 235--246.
+\newblock {\em {SIGPLAN} not.}, {\bf 47}(9), 235--246.
 
@@
 \newblock Composable memory transactions.
-\newblock {\em Pages  48--60 of:} {\em Proceedings of the 10\xth ACM SIGPLAN
+\newblock {\em Pages  48--60 of:} {\em Proceedings of the 10\xth {ACM SIGPLAN}
   Symposium on Principles and Practice of Parallel Programming}.
@@
 \newblock Regular, shape-polymorphic, parallel arrays in haskell.
-\newblock {\em Sigplan not.}, {\bf 45}(9), 261--272.
+\newblock {\em {SIGPLAN} not.}, {\bf 45}(9), 261--272.
 
@@
   with {LVish}.
-\newblock {\em Sigplan not.}, {\bf 49}(6), 2--14.
+\newblock {\em {SIGPLAN} not.}, {\bf 49}(6), 2--14.
 
@@
 \newblock Causal commutative arrows and their optimization.
-\newblock {\em Sigplan not.}, {\bf 44}(9), 35--46.
+\newblock {\em {SIGPLAN} not.}, {\bf 44}(9), 35--46.
 
@@
 \newblock The {HdpH} {DSLs} for scalable reliable computation.
-\newblock {\em Sigplan not.}, {\bf 49}(12), 65--76.
+\newblock {\em {SIGPLAN} not.}, {\bf 49}(12), 65--76.
 
@@
 \newblock Nikola: Embedding compiled {GPU} functions in {Haskell}.
-\newblock {\em Sigplan not.}, {\bf 45}(11), 67--78.
+\newblock {\em {SIGPLAN} not.}, {\bf 45}(11), 67--78.
 
@@
 \newblock Runtime support for multicore {Haskell}.
-\newblock {\em Acm sigplan notices}, {\bf 44}(9), 65--78.
+\newblock {\em {SIGPLAN} not.}, {\bf 44}(9), 65--78.
 
@@
 \newblock A monad for deterministic parallelism.
-\newblock {\em Sigplan not.}, {\bf 46}(12), 71--82.
+\newblock {\em {SIGPLAN} not.}, {\bf 46}(12), 71--82.
 
@@
 \newblock Type-safe runtime code generation: Accelerate to {LLVM}.
-\newblock {\em Sigplan not.}, {\bf 50}(12), 201--212.
+\newblock {\em {SIGPLAN} not.}, {\bf 50}(12), 201--212.
 
@@
 \newblock Idempotent work stealing.
-\newblock {\em Sigplan not.}, {\bf 44}(4), 45--54.
+\newblock {\em {SIGPLAN} not.}, {\bf 44}(4), 45--54.
 
@@
 \newblock A new notation for arrows.
-\newblock {\em Sigplan not.}, {\bf 36}(10), 229--240.
+\newblock {\em {SIGPLAN} not.}, {\bf 36}(10), 229--240.
 
@@
 \newblock A library for light-weight information-flow security in {Haskell}.
-\newblock {\em Pages  13--24 of:} {\em Proceedings of the 1\st ACM SIGPLAN
-  Symposium on Haskell}.
+\newblock {\em Pages  13--24 of:} {\em Proceedings of the 1\st {ACM SIGPLAN}
+  Symposium on {Haskell}}.
 \newblock Haskell '08.
@@
   applications.
-\newblock {\em Sigplan not.}, {\bf 36}(7), 34--43.
+\newblock {\em {SIGPLAN} not.}, {\bf 36}(7), 34--43.
 

### 2018-01-07 paper/content/arxiv/ms.tex
argh, benchmarks

@@
 
-%input bestAndWorstBenchmarks.tex
+%\bestAndWorstBenchmarks
+
+\begin{table}[]
+\centering
+\caption{Overhead in the shared memory benchmarks. Bold marks values in favour of PArrows.}
+\label{tab:meanOverheadSharedMemory}
+\centering
+\begingroup\catcode`"=9
+\sisetup{round-mode=places, round-precision=3}
+\begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 16 cores (s)} \\ \hline \\
+	\multirow{3}{*}{\sudokutest 1000}
+	\csvreader[head to column names]{benchmarks/sudoku-sm/bestAndWorstSudoku-1000.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{1}{*}{\torustest 512}
+	\csvreader[head to column names]{benchmarks/torus-sm/bestAndWorstTorusSM-512.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{3}{*}{\rmtest 11213 32}
+	\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-32.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{3}{*}{\rmtest 11213 64}
+	\csvreader[head to column names]{benchmarks/sm-rm/bestAndWorstRMSM-11213-64.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+\end{tabular}
+\endgroup
+\end{table}
+
+\begin{table}[]
+\centering
+\caption{Overhead in the distributed memory benchmarks. Bold marks values
+  in favour of PArrows.}
+% \olcomment{Vielleicht gleich durchmultiplizieren und \% angeben? Nicht
+% zu viele leere Zeilen in der Tabelle?}
+\label{tab:meanOverHeadDistributedMemory}
+\centering
+\begingroup\catcode`"=9
+\begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 256 cores (s)} \\ \hline \\
+	\multirow{1}{*}{\torustest 4096}
+	\csvreader[head to column names]{benchmarks/distributed-torus/bestAndWorstTorus-4096.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+  	\multirow{1}{*}{\rmtest 44497 256}
+	\csvreader[head to column names]{benchmarks/distributed-rm/bestAndWorstRM-44497-256.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{1}{*}{\jacobitest 3217}
+	\csvreader[head to column names]{benchmarks/distributed-jacobi/bestAndWorstJacobi-3217.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+\end{tabular}
+\endgroup
+\end{table}
 

### 2018-01-07 paper/separate_plots/main.tex
rudimentary layout for separate plots added
olegs fixes for supplementary material added
removed trailing s from paper name
supp

@@
 
-\documentclass{jfp1}
+\documentclass[final]{jfp1}
 
@@
 
-\jdate{August 2017}
-\pubyear{2017}
+\jdate{January 2018}
+\pubyear{2018}
 \pagerange{\pageref{firstpage}--\pageref{lastpage}}
@@
 
-\title{Plots for \enquote{Arrows for Parallel Computations}}
+\title{Arrows for Parallel Computation}
+\subtitle{Supplementary Material}
 \ifthenelse{\boolean{anonymous}}{%
@@
 
+\begin{abstract}
+  This is supplementary material to our paper \enquote{Arrows for
+    parallel computation}. It shows the speedup and mean overhead
+  plots for all benchmarks.
+\end{abstract}
+
 \tableofcontents
+%
+% \newpage
 
@@
 \newlength{\plotwidthSMP}
-\setlength{\plotwidthSMP}{0.39\textwidth}
+% \setlength{\plotwidthSMP}{0.39\textwidth}
+\setlength{\plotwidthSMP}{\plotwidthDist}
 
@@
 
-\section{Distributed Plots}
-
-\subsection{\rmtest}
-
-\subsubsection{Overhead}~\\
-
-\meanOverheadPlot{Mean Overhead for distributed \\ \rmtest 44497 256}{0.04}{32}{272}{\benchmarkDir/distributed-rm/44497-256-diff.csv}
-
-\subsubsection{Speedup}~\\
-
-\speedupplot{Speedup of distributed \rmtest \enquote{44497 256}}{PArrows Eden}{256}{32}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{\benchmarkDir/distributed-rm/bench-distributed.bench.skelrm-parrows-44497-256.csv};
-}{272}{\plotwidthDist}{272}
-
-\subsection{\jacobitest}
-
-\subsubsection{Overhead}~\\
-
-\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.04}{32}{272}{\benchmarkDir/distributed-jacobi/3-3217-diff.csv}
-
-\subsubsection{Speedup}~\\
-
-\speedupplot{Speedup of distributed \jacobitest \enquote{3217}}{PArrows Eden}{256}{32}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{\benchmarkDir/distributed-jacobi/bench-jacobi.bench.jacobi-parr-3-3217.csv};
-}{272}{\plotwidthDist}{32}
-
-\subsection{\torustest}
-
-\meanOverheadPlot{Mean Overhead for distributed \\ \torustest 4096}{0.04}{32}{272}{\benchmarkDir/distributed-torus/4096-diff.csv}
-
-\subsubsection{Speedup}~\\
-
-\speedupplot{Speedup of distributed \torustest \enquote{4096}}{PArrows Eden}{256}{32}{
-\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
-smooth]{\benchmarkDir/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-eden-4096.csv};
-}{272}{\plotwidthDist}{160}
+%%%%% hack: center all!
+\centering
 
 \section{Shared-Memory Plots}
-
 \subsection{\rmtest}
-
-\subsubsection{Speedup}~\\
-
+\subsubsection{Speedup}
 \speedupplot{Speedup of \rmtest \enquote{11213 32}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
@@
 
-\subsubsection{Overhead}~\\
+\subsubsection{Overhead}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Eden CP}{0.04}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-32-diff.csv}
 
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Eden CP}{0.04}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-32-diff.csv}
-\\
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Eden CP}{0.03}{4}{18}{\benchmarkDir/sm-rm/eden-cp-11213-64-diff.csv}
-\\
+
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs GpH}{0.2}{4}{18}{\benchmarkDir/sm-rm/mult-11213-32-diff.csv}
-\\
+
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs GpH}{0.05}{4}{18}{\benchmarkDir/sm-rm/mult-11213-64-diff.csv}
-\\
+
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 32 vs Par Monad}{0.2}{4}{18}{\benchmarkDir/sm-rm/par-11213-32-diff.csv}
-\\
+
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \rmtest 11213 64 vs Par Monad}{0.2}{4}{18}{\benchmarkDir/sm-rm/par-11213-64-diff.csv}
@@
 \subsection{\sudokutest}
-
-\subsubsection{Speedup}~\\
-
+\subsubsection{Speedup}
 \speedupplot{Parallel speedup of \sudokutest \enquote{1000}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
@@
 
-\subsubsection{Overhead}~\\
+\subsubsection{Overhead}
+\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Eden CP}{0.05}{4}{18}{\benchmarkDir/sudoku-sm/eden-cp-1000-diff.csv}
 
-\meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Eden CP}{0.05}{4}{18}{\benchmarkDir/sudoku-sm/eden-cp-1000-diff.csv}
-\\
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs GpH}{0.05}{4}{18}{\benchmarkDir/sudoku-sm/mult-1000-diff.csv}
-\\
+
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \sudokutest 1000 vs Par Monad}{0.05}{4}{18}{\benchmarkDir/sudoku-sm/par-1000-diff.csv}
@@
 \subsection{\torustest}
-
-\subsubsection{Speedup}~\\
-
+\subsubsection{Speedup}
 \speedupplot{Parallel speedup of \sudokutest \enquote{512}}{Eden CP, GpH, \ensuremath{\Conid{Par}} Monad}{16}{4}{
@@
 
-\subsubsection{Overhead}~\\
-
+\subsubsection{Overhead}
 \meanOverheadPlot{Mean Overhead for shared-memory \\ \torustest 512 vs Eden CP}{0.08}{4}{18}{\benchmarkDir/torus-sm/eden-cp-512-diff.csv}
 
+\section{Distributed Plots}
+\subsection{\rmtest}
+
+\subsubsection{Speedup}
+\speedupplot{Speedup of distributed \rmtest \enquote{44497 256}}{PArrows Eden}{256}{32}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/distributed-rm/bench-distributed.bench.skelrm-parrows-44497-256.csv};
+}{272}{\plotwidthDist}{272}
+
+\subsubsection{Overhead}
+
+\meanOverheadPlot{Mean Overhead for distributed \\ \rmtest 44497 256}{0.04}{32}{272}{\benchmarkDir/distributed-rm/44497-256-diff.csv}
+
+\subsection{\jacobitest}
+
+\subsubsection{Speedup}
+\speedupplot{Speedup of distributed \jacobitest \enquote{3217}}{PArrows Eden}{256}{32}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/distributed-jacobi/bench-jacobi.bench.jacobi-parr-3-3217.csv};
+}{272}{\plotwidthDist}{32}
+
+\subsubsection{Overhead}
+\meanOverheadPlot{Mean Overhead for distributed \\ \jacobitest 3217}{0.04}{32}{272}{\benchmarkDir/distributed-jacobi/3-3217-diff.csv}
+
+\subsection{\torustest}
+\meanOverheadPlot{Mean Overhead for distributed \\ \torustest 4096}{0.04}{32}{272}{\benchmarkDir/distributed-torus/4096-diff.csv}
+
+\subsubsection{Speedup}
+\speedupplot{Speedup of distributed \torustest \enquote{4096}}{PArrows Eden}{256}{32}{
+\addplot+ [very thick] table [scatter, x="nCores", y="speedup", col sep=comma, mark=none,
+smooth]{\benchmarkDir/distributed-torus/bench-torus-distributed.bench.torus-matrix-parrows-eden-4096.csv};
+}{272}{\plotwidthDist}{160}
+
 \end{document}

### 2018-01-08 paper/content/parrows.tex
Phil's comment: backends
missing unwrap :: added to ArrowUnwrap

@@
 class (Arrow arr) => ArrowUnwrap arr where
-	arr a b -> (a -> b)
+	unwrap :: arr a b -> (a -> b)
 \end{code}
@@
 While the configurability in the instances of the |ArrowParallel| instances above is nice, users probably would like to have proper default configurations for many parallel programs as well.
-These can also easily be defined as we can see by the example of the default implementation of |ArrowParallel| for the GpH backend:
+These can also easily be defined as we can see by the example of the default implementation of |ArrowParallel| for the GpH:
 

### 2018-01-12 paper/content/utilityFunctions.tex
Phil's comment: backends
fixed paper matrix multiplication

@@
 \end{code}
-\caption{|BasicFuture| type and its |Future| instance for the |Par| Monad and GpH Haskell backends.}
+\caption{|BasicFuture| type and its |Future| instance for the |Par| Monad and GpH.}
 \label{fig:BasicFuture}
@@
 \end{figure}
-Furthermore, Fig.~\ref{fig:torus_example_rest} contains the omitted definitions of |prMMTr| (which calculates $AB^T$ for two matrices $A$ and $B$), |splitMatrix| (which splits the a matrix into chunks), and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
+Furthermore, Fig.~\ref{fig:torus_example_rest} contains the omitted definitions of |prMM| (sequential matrix multiplication), |splitMatrix| (which splits the a matrix into chunks), |staggerHorizontally| and |staggerVertically| (to pre-rotate the matrices), and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
 \begin{figure}[h]
 \begin{code}
-prMMTr m1 m2 = [[sum (zipWith (*) row col) | col <- m2 ] | row <- m1]
+prMM :: Matrix -> Matrix -> Matrix
+prMM m1 m2 = prMMTr m1 (transpose m2)
+  where
+    prMMTr m1' m2' = [[sum (zipWith (*) row col) | col <- m2' ] | row <- m1']
 
@@
 
+staggerHorizontally :: [[a]] -> [[a]]
+staggerHorizontally matrix = zipWith leftRotate [0..] matrix
+
+staggerVertically :: [[a]] -> [[a]]
+staggerVertically matrix = transpose $ zipWith leftRotate [0..] (transpose matrix)
+
+leftRotate :: Int -> [a] -> [a]
+leftRotate i xs = xs2 ++ xs1 where
+    (xs1,xs2) = splitAt i xs
+
 matAdd = chunksOf (dimX x) $ zipWith (+) (concat x) (concat y)
 \end{code}
-	\caption{|prMMTr|, |splitMatrix| and |matAdd| definition.}
+	\caption{|prMMTr|, |splitMatrix|, |staggerHorizontally|, |staggerVertically| and |matAdd| definition.}
 	\label{fig:torus_example_rest}

### 2018-01-12 benchmarks/torus/matrix_seq.hs
same input for all torus benchmarks

@@ -29,10 +29,10 @@ import System.Environment
 import Debug.Trace
 
 randoms1 :: [Int]
-randoms1 = [1..]
+randoms1 = cycle [4,5,6]
 
 randoms2 :: [Int]
-randoms2 = [1..]
+randoms2 = cycle [7,8,9]
 
 type Vector = [Int]
 type Matrix = [Vector]

### 2018-01-12 benchmarks/torus/torus_matrix_parrows_eden.hs

@@ -28,10 +28,10 @@ import System.Environment
 import Debug.Trace
 
 randoms1 :: [Int]
-randoms1 = [1..]
+randoms1 = cycle [4,5,6]
 
 randoms2 :: [Int]
-randoms2 = [1..]
+randoms2 = cycle [7,8,9]
 
 type Vector = [Int]
 type Matrix = [Vector]
@@ -114,7 +114,6 @@ main = do
         let problemSizeVal = read problemSize
         let numCores = noPe
         let matrixA = toMatrix problemSizeVal randoms1
-        --let matrixB = identity problemSizeVal
         let matrixB = toMatrix problemSizeVal randoms2
 
         let matrixC = prMM_torus numCores problemSizeVal matrixA matrixB

### 2018-01-12 benchmarks/torus/torus_matrix_parrows_multicore.hs

@@ -31,10 +31,10 @@ import Debug.Trace
 noPe = 4
 
 randoms1 :: [Int]
-randoms1 = [1..]
+randoms1 = cycle [4,5,6]
 
 randoms2 :: [Int]
-randoms2 = [1..]
+randoms2 = cycle [7,8,9]
 
 type Vector = [Int]
 type Matrix = [Vector]

### 2018-01-12 benchmarks/torus/torus_matrix_parrows_parmonad.hs

@@ -30,10 +30,10 @@ import Debug.Trace
 noPe = 4
 
 randoms1 :: [Int]
-randoms1 = [1..]
+randoms1 = cycle [4,5,6]
 
 randoms2 :: [Int]
-randoms2 = [1..]
+randoms2 = cycle [7,8,9]
 
 type Vector = [Int]
 type Matrix = [Vector]

### 2018-01-12 paper/content/topologySkeletons.tex
i.e.

@@
 
-Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe Arrows of mixed types together, i.e.
-|arr a b| and |arr b c|.
+Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe Arrows of mixed types together, \ie  |arr a b| and |arr b c|.
 By wrapping these two Arrows inside a bigger Arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
@@
 prMM_torus numCores problemSizeVal m1 m2 = 
-	combine $ torus () (mult torusSize) $ zipWith (zipWith (,)) (split m1) (split m2)
-    	where   torusSize = (floor . sqrt) $ fromIntegral numCores
-            	combine = concat . (map (foldr (zipWith (++)) (repeat [])))
-            	split = splitMatrix (problemSizeVal `div` torusSize)
+	combine $ torus () (mult torusSize) $ zipWith zip (split1 m1) (split2 m2)
+    where   torusSize = (floor . sqrt) $ fromIntegral $ numCoreCalc numCores
+            combine x = concat (map ((map (concat)) . transpose) x)
+            split1 x = staggerHorizontally (splitMatrix (problemSizeVal `div` torusSize) x)
+            split2 x = staggerVertically (splitMatrix (problemSizeVal `div` torusSize) x)
 
@@
     where 	toRight = take (size-1) (sm1:sm1s)
-          toBottom = take (size-1) (sm2':sm2s)
-          sm2' = transpose sm2
-          sms = zipWith prMMTr (sm1:sm1s) (sm2':sm2s)
+            toBottom = take (size-1) (sm2:sm2s)
+            sms = zipWith prMM (sm1:sm1s) (sm2:sm2s)
             result = foldl1' matAdd sms
@@
 |mult| is the function performed by each worker.
-|prMMTr| calculates $AB^T$ and is used for the (sequential) calculation in the chunks.
+|prMM| is the sequential matrix multiplication in the chunks.
 |splitMatrix| splits the Matrix into chunks.
+|staggerHorizontally| and |staggerVertically| pre-rotate the matrices.
 |matAdd| calculates $A + B$.
@@
 If we compare the trace from a call using our Arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the Arrow version and execution times are comparable.
-We discuss further benchmarks on larger clusters and in a more detail in the next section.
+We discuss further benchmarks on larger clusters in more detail in the next section.
 \begin{figure}[tb]

### 2018-01-21 paper/content/futures.tex
wip

@@
 \begin{code}
-someCombinator:: (ArrowChoice arr,
+someCombinator :: (ArrowChoice arr,
 	ArrowParallel arr a b (),

### 2018-01-22 BaseSpec/src/main/BaseSpec/CompleteSpecBase.hs
minor

@@ -25,7 +25,7 @@ specBase :: (
   ArrowLoopParallel (->) (fut (([Int], [Int]), [Int])) (fut (([Int], [Int]), [Int])) (),
   ArrowLoopParallel (->) ((Matrix, Matrix), fut [Matrix], fut [Matrix]) (Matrix, fut [Matrix], fut [Matrix]) ()) =>
   Spec
-specBase = describe "Dummy Parrow Check" $ do
+specBase = do
     basicSpec
     pipeSpec
     ringSpec

### 2018-01-22 presentation/skeletons.tex

--- 2017-01-31 presentation/skeletons.tex
@@
-\section{Usability}
-\subsection{Skeletons}
-\begin{frame}[fragile]{parEvalNLazy}
+\subsection{Map Skeletons}
+\begin{frame}[fragile]{Map Skeletons (1)}
+parEvalN, but \textbf{chunky}:
 \begin{lstlisting}[frame=htrbl]
-parEvalNLazy :: conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
+parEvalNLazy :: conf -> ChunkSize -> [arr a b] -> arr [a] [b]
 \end{lstlisting}
+\begin{center}
+\includegraphics[scale=0.5]{images/parEvalNLazy}
+\end{center}
+parallel evaluation of \textbf{different typed functions}:
+\begin{lstlisting}[frame=htrbl]
+parEval2 :: conf -> arr a b -> arr c d -> arr (a, c) (b, d)
+\end{lstlisting}
+\begin{center}
+\includegraphics[scale=0.5]{images/parEval2}
+\end{center}
 \end{frame}
-
-\begin{frame}[fragile]{parEval2}
+\begin{frame}[fragile] {Map Skeletons (2)}
+map, but in \textbf{parallel}:
 \begin{lstlisting}[frame=htrbl]
-parEval2 :: conf -> arr a b -> arr c d -> (arr (a, c) (b, d))
+parMap :: conf -> arr a b -> arr [a] [b]
 \end{lstlisting}
-\end{frame}
-
-\begin{frame}[fragile]{parMap}
-\begin{lstlisting}[frame=htrbl]
-parMap :: conf -> (arr a b) -> (arr [a] [b])
-\end{lstlisting}
-\end{frame}
-
-\begin{frame}[fragile]{parMapStream}
+\begin{center}
+\includegraphics[scale=0.5]{images/parMap}
+\end{center}
+parMap, but \textbf{chunky}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\begin{center}
+\includegraphics[scale=0.5]{images/parMapStream}
+\end{center}
 \end{frame}
-
-\begin{frame}[fragile]{farm}
+\begin{frame}[fragile]{Map Skeletons (3)}
+parMap, but with \textbf{workload distribution}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
-\end{frame}
-
-\begin{frame}[fragile]{farmChunk}
+\begin{center}
+\includegraphics[scale=0.5]{images/farm}
+\end{center}
+farm, but \textbf{chunky}:
 \begin{lstlisting}[frame=htrbl]
@@
 \end{lstlisting}
+\begin{center}
+\includegraphics[scale=0.5]{images/farmChunk}
+\end{center}
 \end{frame}

### 2018-07-26 paper/content/topologySkeletons.tex

@@
 defines an Arrow that takes its input |a| and converts it into an infinite stream |[a]| of it.
-Using |loop| to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
+Using |loop| to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |loopParEvalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.
 
@@
 The use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt.
-We need to enforce that the evaluation of the input |[a]| terminates before passing it into |evalN|.}
+We need to ensure that the evaluation of the input |[a]| is not forced fully before passing it into |loopParEvalN|.}
 \label{fig:pipeSimple}
@@
 \begin{code}
-pipe :: (ArrowLoop arr, ArrowLoopParallel arr (fut a) (fut a) conf,
+pipe :: (ArrowLoop arr,
+    ArrowLoopParallel arr (fut a) (fut a) conf,
 	Future fut a conf) =>

### 2019-12-27 paper/content/arxiv/ms.tex
reviewer's comments (1)

@@
 the existing parallel Haskells. Thus, we not only define a parallel programming interface in a
-novel manner -- we tame the zoo of parallel Haskells. We provide a
+novel manner---we tame the zoo of parallel Haskells. We provide a
 common, very low-penalty programming interface that allows to switch
@@
 The induced penalty was in the single-digit percent range, with means typically under 2\% overhead in our measurements over the varying cores configuration (Section~\ref{sec:benchmarks}).
-Further implementations, based on HdpH or a Frege implementation (on the Java Virtual Machine), are viable, too.
+We have also conducted some initial experiments with HdpH and Frege; those experiments inducate that our parallel Arrows DSL can be relatively easily extended to further languages.
 
@@
 Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
-deserves a special mention. <<Why does it deserve special mention?>>  Accelerate is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.
+is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.
 
@@
 of \ensuremath{\Conid{Par}} Monad.}
+We chose to not extend our implementation to LVish and HdpH at the moment.
+The former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper.
+The latter heavy relies on Template Haskell, which adds complexity to the implementation.
 
@@
 Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution.
-In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of list of Arrows.
+In contrast, we capture a whole parallel computation as a single entity: our main instantiation function \ensuremath{\Varid{parEvalN}} makes a single (parallel) Arrow out of a list of Arrows.
 \citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach.
@@
 Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined.
-Basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM).
+Preliminary experiments show that basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM).
 However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}},
@@
 
-While we could have based our DSL on Profunctors as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
+While we could have based our DSL on Profunctors \cite{boisseau_what_2018} as well, we chose Arrows\mbcomment{cite missing?} for this paper since they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
 However, they are a promising candidate for future improvements of our DSL.
@@
 In this section, we will implement this non-Arrow version which will later be adapted for usage in our Arrow-based parallel Haskell.
-
-There exist several parallel Haskells already.
-Among the most important are probably GpH \citep[based on \ensuremath{\Varid{par}} and \ensuremath{\Varid{pseq}} \enquote{hints},][]{Trinder1996,Trinder1998a}, the \ensuremath{\Conid{Par}} Monad \citep[a monad for deterministic parallelism,][]{par-monad,Foltzer:2012:MPC:2398856.2364562}, Eden \citep[a parallel Haskell for distributed memory,][]{eden,Loogen2012}, HdpH \citep[a Template Haskell-based parallel Haskell for distributed memory,][]{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} and LVish \citep[a \ensuremath{\Conid{Par}} extension with focus on communication,][]{Kuper:2014:TPE:2666356.2594312}.
-
 As the goal of this paper is not to re-implement yet another parallel runtime, but to represent parallelism with Arrows, we base our efforts on existing work which we wrap as backends behind a common interface.
 For this paper we chose GpH for its simplicity, the \ensuremath{\Conid{Par}} Monad to represent a monadic DSL, and Eden as a distributed parallel Haskell.
-
-LVish and HdpH were not chosen as the former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper, while the latter (at least in its current form) does not comply with our representation of parallelism due to its heavy reliance on Template Haskell.
-
-We will now go into some detail on GpH, the \ensuremath{\Conid{Par}} Monad and Eden, and also give their respective implementations of the non-Arrow version of \ensuremath{\Varid{parEvalN}}.
-
-%Before we go into detail on how we can use this idea of parallelism
-%for parallel Arrows, as a short introduction to parallelism in
-%Haskell we will now implement |parEvalN| with several different
-%parallel Haskells.
 
@@
 \label{sec:edentv}
-To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
+To comprehend the efficiency or the lack thereof in a parallel program, an inspection of its execution is extremely helpful.
 While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See \url{http://hackage.haskell.org/package/edentv} on Hackage for the last available version of Eden TraceViewer.} \cite{Berthold2007a}.
@@
 By combining the result of this Arrow with \ensuremath{\Varid{arr}\;\Varid{sequenceA}}, we get an Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Conid{IVar}\;\Varid{b}\mskip1.5mu])}.
-Then, in order to fetch the results of the different threads, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} Monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})} -- our intermediary Arrow is of type \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu])}.
+Then, in order to fetch the results of the different threads, we map over the \ensuremath{\Conid{IVar}}s inside the \ensuremath{\Conid{Par}} Monad with \ensuremath{\Varid{arr}\;(\bind \Varid{mapM}\;\Varid{get})}---our intermediary Arrow is of type \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;(\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu])}.
 Finally, we execute the computation \ensuremath{\Conid{Par}\;[\mskip1.5mu \Varid{b}\mskip1.5mu]} by composing with \ensuremath{\Varid{arr}\;\Varid{runPar}} and get the final Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
@@
 \subsubsection{Default configuration instances}
-While the configurability in the instances of the \ensuremath{\Conid{ArrowParallel}} instances above is nice, users probably would like to have proper default configurations for many parallel programs as well.
+While the configurability in the instances of the \ensuremath{\Conid{ArrowParallel}} interface above is nice, users probably would like to have proper default configurations for many parallel programs as well.
 These can also easily be defined as we can see by the example of the default implementation of \ensuremath{\Conid{ArrowParallel}} for the GpH:
@@
 Eden already provides \enquote{remote data} that enable this \cite{AlGo03a,Dieterle2010}.
-But as we want code with our DSL to be implementation agnostic, we have to wrap this concept.
+But as we want code with our DSL to be portable and implementation agnostic, we have to wrap this concept.
 We do this with the \ensuremath{\Conid{Future}} type class (Fig.~\ref{fig:future}).
@@
 \end{figure}
-In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed -- similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}.
+In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed---similar to what is shown in the trace visualisation in Fig.~\ref{fig:withFutures}.
 One especially elegant aspect of the definition in Fig.~\ref{fig:future} is that we can specify the type of \ensuremath{\Conid{Future}} to be used per backend with full interoperability between code using different backends, without even requiring to know about the actual type used for communication.
@@
 Now we have developed Parallel Arrows far enough to define some useful algorithmic skeletons that abstract typical parallel computations.
-While there are many possible skeletons to implement, we demonstrate the expressive power of PArrows here using four \ensuremath{\Varid{map}}-based and three toplogical skeletons.
+While there are many possible skeletons to implement, we demonstrate the expressive power of PArrows here using four \ensuremath{\Varid{map}}-based and three topological skeletons.
 %%% \FloatBarrier
@@
 This materialises with the usage of \ensuremath{\Varid{loop}} of the \ensuremath{\Conid{ArrowLoop}} type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden.
-An investigation of why this difference exists is beyond the scope of this work, we only provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment.
+We provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment.
 However our workaround enables users of the DSL to test their code within a shared memory setting.
 % even though other skeletons would be better suited to be run with them. --- OL: I don't get this sentence part
-The idea of the fix is to provide a \ensuremath{\Conid{ArrowLoopParallel}} type class that has two functions -- \ensuremath{\Varid{loopParEvalN}} and \ensuremath{\Varid{postLoopParEvalN}}, where the first is to be used inside an \ensuremath{\Varid{loop}} construct while the latter will be used right outside of the \ensuremath{\Varid{loop}}.
+The idea of the fix is to provide a \ensuremath{\Conid{ArrowLoopParallel}} type class that has two functions---\ensuremath{\Varid{loopParEvalN}} and \ensuremath{\Varid{postLoopParEvalN}}, where the first is to be used inside an \ensuremath{\Varid{loop}} construct while the latter will be used right outside of the \ensuremath{\Varid{loop}}.
 This way we can delegate to the actual \ensuremath{\Varid{parEvalN}} in the spot where the backend supports it.
@@
 Next, we zip the resulting \ensuremath{([\mskip1.5mu \Varid{i}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])} to \ensuremath{[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]} with \ensuremath{\Varid{arr}\;(\Varid{uncurry}\;\Varid{zip})}.
-We then feed this into our parallel Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]\;[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} obtained by transforming our input Arrow \ensuremath{\Varid{f}\mathbin{::}\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})} into \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})} before \ensuremath{\Varid{repeat}}ing and lifting it with \ensuremath{\Varid{loopParEvalN}}.
+We then feed this into our parallel Arrow \ensuremath{\Varid{arr}\;[\mskip1.5mu (\Varid{i},\Varid{fut}\;\Varid{r})\mskip1.5mu]\;[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} obtained by transforming our input Arrow \ensuremath{\Varid{f}\mathbin{::}\Varid{arr}\;(\Varid{i},\Varid{r})\;(\Varid{o},\Varid{r})} into \ensuremath{\Varid{arr}\;(\Varid{i},\Varid{fut}\;\Varid{r})\;(\Varid{o},\Varid{fut}\;\Varid{r})} before \ensuremath{\Varid{repeat}}'ing and lifting it with \ensuremath{\Varid{loopParEvalN}}.
 Finally we unzip the output list \ensuremath{[\mskip1.5mu (\Varid{o},\Varid{fut}\;\Varid{r})\mskip1.5mu]} list into \ensuremath{([\mskip1.5mu \Varid{o}\mskip1.5mu],[\mskip1.5mu \Varid{fut}\;\Varid{r}\mskip1.5mu])}.
@@
 \end{figure}
-As an example of using this skeleton, \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}).
+As an example of using this skeleton, \citet{Eden:SkeletonBookChapter02} showed the matrix multiplication using Gentleman's algorithm (\citeyear{Gentleman1978}).
 An adapted version can be found in Fig.~\ref{fig:torusMatMult}.
@@
 	\label{fig:torus_parrows_trace}
-\end{figure}
-
-\begin{figure}[tb]
+% \end{figure}
+%
+% \begin{figure}[tb]
 	\centering
@@
 This section evaluates the performance overhead of this compositional abstraction in comparison to  GpH and the \ensuremath{\Conid{Par}} Monad on shared memory architectures and Eden on  a distributed memory cluster.
-We describe our measurement platform, the benchmark results -- the shared-memory variants (GpH, \ensuremath{\Conid{Par}} Monad and Eden~CP) followed by Eden in a distributed memory setting, and conclude that PArrows hold up in terms of performance when compared to the original parallel Haskells.
+We describe our measurement platform, the benchmark results---the shared-memory variants (GpH, \ensuremath{\Conid{Par}} Monad and Eden~CP) followed by Eden in a distributed memory setting, and conclude that PArrows hold up in terms of performance when compared to the original parallel Haskells.
 
@@
 
-The \ensuremath{\Conid{Par}} monad and GpH -- in its multicore version \cite{Marlow2009} --  can be executed on shared memory machines only.
+The \ensuremath{\Conid{Par}} monad and GpH---in its multicore version \cite{Marlow2009}--- can be executed on shared memory machines only.
 Although GpH is available on distributed memory
@@
 efficiency of 85.3\%. However, if we increased the number of
-requested cores to 32 -- \ie if we use hyper-threading on 16 real
-cores -- the speedup did not increase that well. It was merely 15.99
+requested cores to 32---\ie if we use hyper-threading on 16 real
+cores---the speedup did not increase that well. It was merely 15.99
 for 32 tasks with PArrows/Eden. This was worse for other implementations.  As
@@
 To give a few examples, the overhead for Eden~CP with input value $2^{11213}-1$, 32 tasks, and 16 cores is $1.5\%$, but the error margin is around $5.2\%$!
-Same implementation in the same setting with 64 tasks reaches $-0.2\%$ overhead, PArrows apparently fare better than Eden -- but the error margin of $1.9\%$ disallows this interpretation.
+Same implementation in the same setting with 64 tasks reaches $-0.2\%$ overhead, PArrows apparently fare better than Eden---but the error margin of $1.9\%$ disallows this interpretation.
 We focus now on significant overhead values.

### 2019-12-27 paper/content/arrows.tex
cabal

--- 2018-01-07 paper/content/arrows.tex
@@
 
-While we could have based our DSL on Profunctors as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
+While we could have based our DSL on Profunctors \cite{boisseau_what_2018} as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
 However, they are a promising candidate for future improvements of our DSL.

### 2019-12-27 paper/content/arxiv/ms.tex
it compiles

@@
 	%\pagebreak
-        %\bibliographystyle{jfp}
-	%\bibliography{references,main}
-        \providecommand{\noopsort}[1]{}
-\begin{thebibliography}{}
-
-\bibitem[\protect\citename{Acar {\em et~al.}\relax,
-  }2000]{Acar:2000:DLW:341800.341801}
-Acar, Umut~A., Blelloch, Guy~E., \& Blumofe, Robert~D. (2000).
-\newblock The data locality of work stealing.
-\newblock {\em Pages  1--12 of:} {\em Proceedings of the 12\xth Annual ACM
-  Symposium on Parallel Algorithms and Architectures}.
-\newblock SPAA '00.
-\newblock ACM.
-
-\bibitem[\protect\citename{Achten {\em et~al.}\relax, }2007]{achten2007arrow}
-Achten, Peter, van Eekelen, Marko, de~Mol, Maarten, \& Plasmeijer, Rinus.
-  (2007).
-\newblock An arrow based semantics for interactive applications.
-\newblock  {\em Draft Proceedings of the Symposium on Trends in Functional
-  Programming}.
-\newblock TFP '07.
-
-\bibitem[\protect\citename{Achten {\em et~al.}\relax, }2004]{achten2004arrows}
-Achten, PM, van Eekelen, Marko~CJD, Plasmeijer, MJ, \& Weelden, A~van. (2004).
-\newblock {\em Arrows for generic graphical editor components}.
-\newblock Tech. rept. Nijmegen Institute for Computing and Information
-  Sciences, Faculty of Science, University of Nijmegen, The Netherlands.
-\newblock Technical Report NIII-R0416.
-
-\bibitem[\protect\citename{Alimarine {\em et~al.}\relax,
-  }2005]{Alimarine:2005:BAA:1088348.1088357}
-Alimarine, Artem, Smetsers, Sjaak, van Weelden, Arjen, van Eekelen, Marko, \&
-  Plasmeijer, Rinus. (2005).
-\newblock There and back again: Arrows for invertible programming.
-\newblock {\em Pages  86--97 of:} {\em Proceedings of the {2005} {ACM}
-  {SIGPLAN} Workshop on {Haskell}}.
-\newblock Haskell '05.
-\newblock ACM.
-
-\bibitem[\protect\citename{Aljabri {\em et~al.}\relax,
-  }2014]{Aljabri:2013:DIG:2620678.2620682}
-Aljabri, Malak, Loidl, Hans-Wolfgang, \& Trinder, Phil~W. (2014).
-\newblock The design and implementation of {GUMSMP}: A multilevel parallel
-  haskell implementation.
-\newblock {\em Pages  37:37--37:48 of:} {\em Proceedings of the 25\xth
-  Symposium on Implementation and Application of Functional Languages}.
-\newblock IFL '13.
-\newblock ACM.
-
-\bibitem[\protect\citename{Aljabri {\em et~al.}\relax, }2015]{Aljabri2015}
-Aljabri, Malak, Loidl, Hans-Wolfgang, \& Trinder, Phil. (2015).
-\newblock {\em Balancing shared and distributed heaps on {NUMA} architectures}.
-\newblock Springer.
-\newblock Pages  1--17.
-
-\bibitem[\protect\citename{Alt \& Gorlatch, }2003]{AlGo03a}
-Alt, Martin, \& Gorlatch, Sergei. (2003).
-\newblock {F}uture-{B}ased {RMI}: Optimizing compositions of remote method
-  calls on the {G}rid.
-\newblock {\em Pages  682--693 of:} Kosch, Harald, B{\"o}sz{\"o}rm{\'e}nyi,
-  L{\'a}szl{\'o}, \& Hellwagner, Hermann (eds), {\em Euro-Par~2003}.
-\newblock LNCS 2790.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Asada, }2010]{Asada:2010:ASM:1863597.1863607}
-Asada, Kazuyuki. (2010).
-\newblock Arrows are strong monads.
-\newblock {\em Pages  33--42 of:} {\em Proceedings of the Third {ACM SIGPLAN}
-  Workshop on Mathematically Structured Functional Programming}.
-\newblock MSFP '10.
-\newblock New York, NY, USA: ACM.
-
-\bibitem[\protect\citename{Aswad {\em et~al.}\relax, }2009]{aswad2009low}
-Aswad, Mustafa, Trinder, Phil, Al~Zain, Abdallah, Michaelson, Greg, \&
-  Berthold, Jost. (2009).
-\newblock Low pain vs no pain multi-core {Haskells}.
-\newblock {\em Pages  49--64 of:} {\em Trends in Functional Programming}.
-
-\bibitem[\protect\citename{Atkey, }2011]{ATKEY201119}
-Atkey, Robert. (2011).
-\newblock What is a categorical model of arrows?
-\newblock {\em Electronic notes in theoretical computer science}, {\bf 229}(5),
-  19--37.
-
-\bibitem[\protect\citename{Berthold, }2008]{JostThesis}
-Berthold, Jost. (2008).
-\newblock {\em Explicit and implicit parallel functional programming ---
-  concepts and implementation}.
-\newblock Ph.D. thesis, Philipps-Universit\"{a}t Marburg.
-
-\bibitem[\protect\citename{Berthold \& Loogen, }2006]{Eden:PARCO05}
-Berthold, Jost, \& Loogen, Rita. (2006).
-\newblock Skeletons for recursively unfolding process topologies.
-\newblock  Joubert, Gerhard~R., Nagel, Wolfgang~E., Peters, Frans~J., Plata,
-  Oscar~G., Tirado, P., \& Zapata, Emilio~L. (eds), {\em Parallel Computing:
-  Current \& Future Issues of High-End Computing, ParCo 2005, Malaga, Spain}.
-\newblock NIC Series 33.
-\newblock Central Institute for Applied Mathematics, J{\"u}lich, Germany.
-
-\bibitem[\protect\citename{Berthold \& Loogen, }2007]{Berthold2007a}
-Berthold, Jost, \& Loogen, Rita. (2007).
-\newblock {Visualizing Parallel Functional Program Executions: Case Studies
-  with the Eden Trace Viewer}.
-\newblock  {\em ParCo '07. Parallel Computing: Architectures, Algorithms and
-  Applications}.
-\newblock IOS Press.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax, }2008]{PADL08HMWS}
-Berthold, Jost, Dieterle, Mischa, Loogen, Rita, \& Priebe, Steffen. (2008).
-\newblock Hierarchical master-worker skeletons.
-\newblock  Warren, David~S., \& Hudak, Paul (eds), {\em Practical Aspects of
-  Declarative Languages (PADL'08)}.
-\newblock LNCS 4902.
-\newblock San Francisco (CA), USA: Springer-Verlag.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax, }2009a]{arcs-dc}
-Berthold, Jost, Dieterle, Mischa, Lobachev, Oleg, \& Loogen, Rita. (2009a).
-\newblock {Distributed Memory Programming on Many-Cores -- A Case Study Using
-  {Eden} Divide-\&-Conquer Skeletons}.
-\newblock {\em Pages  47--55 of:} Gro\ss{}pitsch, K.-E., Henkersdorf, A.,
-  Uhrig, S., Ungerer, T., \& H\"{a}hner, J. (eds), {\em Workshop on Many-Cores
-  at ARCS '09 -- 22\nd International Conference on Architecture of Computing
-  Systems 2009}.
-\newblock VDE-Verlag.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
-  }2009b]{Berthold2009-mr}
-Berthold, Jost, Dieterle, Mischa, \& Loogen, Rita. (2009b).
-\newblock Implementing parallel {Google} map-reduce in {Eden}.
-\newblock {\em Pages  990--1002 of:} Sips, Henk, Epema, Dick, \& Lin, Hai-Xiang
-  (eds), {\em Euro-Par 2009 Parallel Processing}.
-\newblock LNCS 5704.
-\newblock Springer Berlin Heidelberg.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
-  }2009c]{Berthold2009-fft}
-Berthold, Jost, Dieterle, Mischa, Lobachev, Oleg, \& Loogen, Rita. (2009c).
-\newblock {\em Parallel {FFT} with {Eden} skeletons}.
-\newblock PaCT '09.
-\newblock Springer.
-\newblock Pages  73--83.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
-  }2016]{berthold_loidl_hammond_2016}
-Berthold, Jost, Loidl, Hans-Wolfgang, \& Hammond, Kevin. (2016).
-\newblock {PAEAN}: Portable and scalable runtime support for parallel {Haskell}
-  dialects.
-\newblock {\em Journal of functional programming}, {\bf 26}.
-
-\bibitem[\protect\citename{Bischof \& Gorlatch, }2002]{Bischof2002}
-Bischof, Holger, \& Gorlatch, Sergei. (2002).
-\newblock {\em Double-scan: Introducing and implementing a new data-parallel
-  skeleton}.
-\newblock Euro-Par '02.
-\newblock Springer.
-\newblock Pages  640--647.
-
-\bibitem[\protect\citename{Blumofe \& Leiserson,
-  }1999]{Blumofe:1999:SMC:324133.324234}
-Blumofe, Robert~D., \& Leiserson, Charles~E. (1999).
-\newblock Scheduling multithreaded computations by work stealing.
-\newblock {\em J. acm}, {\bf 46}(5), 720--748.
-
-\bibitem[\protect\citename{Botorog \& Kuchen, }1996]{botorog1996efficient}
-Botorog, G.~H., \& Kuchen, H. (1996).
-\newblock {\em {Euro-Par'96 Parallel Processing}}.
-\newblock LNCS 1123.
-\newblock Springer-Verlag.
-\newblock Chap. Efficient parallel programming with algorithmic skeletons,
-  pages  718--731.
-
-\bibitem[\protect\citename{Brown \& Hammond, }2010]{brown2010ever}
-Brown, C., \& Hammond, K. (2010).
-\newblock Ever-decreasing circles: a skeleton for parallel orbit calculations
-  in {Eden}.
-\newblock  {\em Draft Proceedings of the Symposium on Trends in Functional
-  Programming}.
-\newblock TFP '10.
-
-\bibitem[\protect\citename{Buono {\em et~al.}\relax, }2010]{BUONO20102095}
-Buono, D., Danelutto, M., \& Lametti, S. (2010).
-\newblock Map, reduce and mapreduce, the skeleton way.
-\newblock {\em Procedia computer science}, {\bf 1}(1), 2095--2103.
-\newblock ICCS 2010.
-
-\bibitem[\protect\citename{{C}hakravarty {\em et~al.}\relax,
-  }2007]{Chakravarty2007}
-{C}hakravarty, Manuel M.~T., {L}eshchinskiy, Roman, {P}eyton {J}ones, Simon~L.,
-  {K}eller, Gabriele, \& {M}arlow, Simon. (2007).
-\newblock {Data Parallel Haskell}: a status report.
-\newblock {\em Pages  10--18 of:} {\em DAMP '07}.
-\newblock ACM Press.
-
-\bibitem[\protect\citename{Chakravarty {\em et~al.}\relax,
-  }2011]{Chakravarty:2011:AHA:1926354.1926358}
-Chakravarty, Manuel~M.T., Keller, Gabriele, Lee, Sean, McDonell, Trevor~L., \&
-  Grover, Vinod. (2011).
-\newblock Accelerating {Haskell} array codes with multicore {GPUs}.
-\newblock {\em Pages  3--14 of:} {\em Proceedings of the 6\xth Workshop on
-  Declarative Aspects of Multicore Programming}.
-\newblock DAMP '11.
-\newblock ACM.
-
-\bibitem[\protect\citename{Chase \& Lev, }2005]{Chase:2005:DCW:1073970.1073974}
-Chase, David, \& Lev, Yossi. (2005).
-\newblock Dynamic circular work-stealing deque.
-\newblock {\em Pages  21--28 of:} {\em Proceedings of the 17\xth Annual ACM
-  Symposium on Parallelism in Algorithms and Architectures}.
-\newblock SPAA '05.
-\newblock ACM.
-
-\bibitem[\protect\citename{Clifton-Everest {\em et~al.}\relax, }2014]{CMCK14}
-Clifton-Everest, Robert, McDonell, Trevor~L, Chakravarty, Manuel M~T, \&
-  Keller, Gabriele. (2014).
-\newblock {Embedding Foreign Code}.
-\newblock  {\em PADL '14: The 16th International Symposium on Practical Aspects
-  of Declarative Languages}.
-\newblock LNCS.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{{C}ole, }1989]{Cole1989}
-{C}ole, M.~I. (1989).
-\newblock Algorithmic skeletons: Structured management of parallel computation.
-\newblock  {\em Research Monographs in Parallel and Distributed Computing}.
-\newblock Pitman.
-
-\bibitem[\protect\citename{Czaplicki \& Chong,
-  }2013]{Czaplicki:2013:AFR:2499370.2462161}
-Czaplicki, Evan, \& Chong, Stephen. (2013).
-\newblock Asynchronous functional reactive programming for guis.
-\newblock {\em {SIGPLAN} not.}, {\bf 48}(6), 411--422.
-
-\bibitem[\protect\citename{Dagand {\em et~al.}\relax,
-  }2009]{Dagand:2009:ORD:1481861.1481870}
-Dagand, Pierre-\'{E}variste, Kosti\'{c}, Dejan, \& Kuncak, Viktor. (2009).
-\newblock Opis: Reliable distributed systems in {OCaml}.
-\newblock {\em Pages  65--78 of:} {\em Proceedings of the 4\xth International
-  Workshop on Types in Language Design and Implementation}.
-\newblock TLDI '09.
-\newblock ACM.
-
-\bibitem[\protect\citename{Danelutto {\em et~al.}\relax,
-  }1992]{DANELUTTO1992205}
-Danelutto, Marco, Meglio, Roberto~Di, Orlando, Salvatore, Pelagatti, Susanna,
-  \& Vanneschi, Marco. (1992).
-\newblock A methodology for the development and the support of massively
-  parallel programs.
-\newblock {\em Future generation computer systems}, {\bf 8}(1), 205--220.
-
-\bibitem[\protect\citename{Darlington {\em et~al.}\relax,
-  }1993]{darlington1993parallel}
-Darlington, J., Field, AJ, Harrison, PG, Kelly, PHJ, Sharp, DWN, Wu, Q., \&
-  While, RL. (1993).
-\newblock Parallel programming using skeleton functions.
-\newblock {\em Pages  146--160 of:} {\em Parallel architectures and languages
-  Europe}.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Dastgeer {\em et~al.}\relax,
-  }2011]{Dastgeer:2011:ASM:1984693.1984697}
-Dastgeer, Usman, Enmyren, Johan, \& Kessler, Christoph~W. (2011).
-\newblock Auto-tuning {SkePU}: A multi-backend skeleton programming framework
-  for multi-{GPU} systems.
-\newblock {\em Pages  25--32 of:} {\em Proceedings of the 4\xth International
-  Workshop on Multicore Software Engineering}.
-\newblock IWMSE '11.
-\newblock ACM.
-
-\bibitem[\protect\citename{de~la Encina {\em et~al.}\relax,
-  }2011]{delaEncina2011}
-de~la Encina, Alberto, Hidalgo-Herrero, Mercedes, Rabanal, Pablo, \& Rubio,
-  Fernando. (2011).
-\newblock {\em A parallel skeleton for genetic algorithms}.
-\newblock IWANN '11.
-\newblock Springer.
-\newblock Pages  388--395.
-
-\bibitem[\protect\citename{Dean \& Ghemawat,
-  }2008]{Dean:2008:MSD:1327452.1327492}
-Dean, Jeffrey, \& Ghemawat, Sanjay. (2008).
-\newblock {MapReduce}: simplified data processing on large clusters.
-\newblock {\em Communications of the acm}, {\bf 51}, 107--113.
-
-\bibitem[\protect\citename{Dean \& Ghemawat,
-  }2010]{Dean:2010:MFD:1629175.1629198}
-Dean, Jeffrey, \& Ghemawat, Sanjay. (2010).
-\newblock {MapReduce}: a flexible data processing tool.
-\newblock {\em Communications of the acm}, {\bf 53}, 72--77.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax, }2010a]{Dieterle2010}
-Dieterle, M., Horstmeyer, T., \& Loogen, R. (2010a).
-\newblock Skeleton composition using remote data.
-\newblock {\em Pages  73--87 of:} Carro, M., \& Pe\~{n}a, R. (eds), {\em 12\xth
-  International Symposium on Practical Aspects of Declarative Languages}.
-\newblock PADL '10, vol. 5937.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax,
-  }2016]{dieterle_horstmeyer_loogen_berthold_2016}
-Dieterle, M., Horstmeyer, T., Loogen, R., \& Berthold, J. (2016).
-\newblock Skeleton composition versus stable process systems in {Eden}.
-\newblock {\em Journal of functional programming}, {\bf 26}.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax,
-  }2010b]{dieterle2010skeleton}
-Dieterle, Mischa, Berthold, Jost, \& Loogen, Rita. (2010b).
-\newblock {\em A skeleton for distributed work pools in {Eden}}.
-\newblock FLOPS '10.
-\newblock Springer.
-\newblock Pages  337--353.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax, }2013]{Dieterle2013}
-Dieterle, Mischa, Horstmeyer, Thomas, Berthold, Jost, \& Loogen, Rita. (2013).
-\newblock {\em Iterating skeletons}.
-\newblock IFL '12.
-\newblock Springer.
-\newblock Pages  18--36.
-
-\bibitem[\protect\citename{Dinan {\em et~al.}\relax,
-  }2009]{Dinan:2009:SWS:1654059.1654113}
-Dinan, James, Larkins, D.~Brian, Sadayappan, P., Krishnamoorthy, Sriram, \&
-  Nieplocha, Jarek. (2009).
-\newblock Scalable work stealing.
-\newblock {\em Pages  53:1--53:11 of:} {\em Proceedings of the Conference on
-  High Performance Computing Networking, Storage and Analysis}.
-\newblock SC '09.
-\newblock ACM.
-
-\bibitem[\protect\citename{Foltzer {\em et~al.}\relax,
-  }2012]{Foltzer:2012:MPC:2398856.2364562}
-Foltzer, Adam, Kulkarni, Abhishek, Swords, Rebecca, Sasidharan, Sajith, Jiang,
-  Eric, \& Newton, Ryan. (2012).
-\newblock A meta-scheduler for the {Par}-monad: Composable scheduling for the
-  heterogeneous cloud.
-\newblock {\em {SIGPLAN} not.}, {\bf 47}(9), 235--246.
-
-\bibitem[\protect\citename{Geimer {\em et~al.}\relax, }2010]{Geimer2010}
-Geimer, M., Wolf, F., Wylie, B. J.~N., {\'A}brah{\'a}m, E., Becker, D., \&
-  Mohr, B. (2010).
-\newblock The {Scalasca} performance toolset architecture.
-\newblock {\em Concurrency and computation: Practice and experience}, {\bf
-  22}(6).
-
-\bibitem[\protect\citename{Gentleman, }1978]{Gentleman1978}
-Gentleman, W.~Morven. (1978).
-\newblock Some complexity results for matrix computations on parallel
-  processors.
-\newblock {\em Journal of the acm}, {\bf 25}(1), 112--115.
-
-\bibitem[\protect\citename{Gorlatch, }1998]{Gorlatch1998}
-Gorlatch, Sergei. (1998).
-\newblock Programming with divide-and-conquer skeletons: A case study of {FFT}.
-\newblock {\em Journal of supercomputing}, {\bf 12}(1-2), 85--97.
-
-\bibitem[\protect\citename{Gorlatch \& Bischof, }1998]{Gorlatch}
-Gorlatch, Sergei, \& Bischof, Holger. (1998).
-\newblock A generic {MPI} implementation for a data-parallel skeleton: Formal
-  derivation and application to {FFT}.
-\newblock {\em Parallel processing letters}, {\bf 8}(4).
-
-\bibitem[\protect\citename{Hammond {\em et~al.}\relax,
-  }2003]{doi:10.1142/S0129626403001380}
-Hammond, Kevin, Berthold, Jost, \& Loogen, Rita. (2003).
-\newblock Automatic skeletons in {Template} {Haskell}.
-\newblock {\em Parallel processing letters}, {\bf 13}(03), 413--424.
-
-\bibitem[\protect\citename{Harris {\em et~al.}\relax,
-  }2007]{harris2007parallel}
-Harris, Mark, Sengupta, Shubhabrata, \& Owens, John~D. (2007).
-\newblock Parallel prefix sum (scan) with {CUDA}.
-\newblock {\em {GPU} gems}, {\bf 3}(39), 851--876.
-
-\bibitem[\protect\citename{Harris {\em et~al.}\relax,
-  }2005]{Harris:2005:CMT:1065944.1065952}
-Harris, Tim, Marlow, Simon, {Peyton Jones}, Simon, \& Herlihy, Maurice. (2005).
-\newblock Composable memory transactions.
-\newblock {\em Pages  48--60 of:} {\em Proceedings of the 10\xth {ACM SIGPLAN}
-  Symposium on Principles and Practice of Parallel Programming}.
-\newblock PPoPP '05.
-\newblock ACM.
-
-\bibitem[\protect\citename{Hey, }1990]{Hey1990185}
-Hey, Anthony J.~G. (1990).
-\newblock Experiments in {MIMD} parallelism.
-\newblock {\em Future generation computer systems}, {\bf 6}(3), 185--196.
-
-\bibitem[\protect\citename{Hippold \& R{\"u}nger, }2006]{Hippold2006}
-Hippold, J., \& R{\"u}nger, G. (2006).
-\newblock Task pool teams: A hybrid programming environment for irregular
-  algorithms on {SMP} clusters.
-\newblock {\em Concurrency and computation: Practice and experience}, {\bf 18},
-  1575--1594.
-
-\bibitem[\protect\citename{Horstmeyer \& Loogen, }2013]{Horstmeyer2013}
-Horstmeyer, Thomas, \& Loogen, Rita. (2013).
-\newblock Graph-based communication in {Eden}.
-\newblock {\em Higher-order and symbolic computation}, {\bf 26}(1), 3--28.
-
-\bibitem[\protect\citename{Huang {\em et~al.}\relax, }2007]{Huang2007}
-Huang, Liwen, Hudak, Paul, \& Peterson, John. (2007).
-\newblock {\em {HPorter}: Using arrows to compose parallel processes}.
-\newblock Berlin, Heidelberg: Springer Berlin Heidelberg.
-\newblock Pages  275--289.
-
-\bibitem[\protect\citename{Hudak {\em et~al.}\relax, }2003]{Hudak2003}
-Hudak, Paul, Courtney, Antony, Nilsson, Henrik, \& Peterson, John. (2003).
-\newblock {\em Arrows, robots, and functional reactive programming}.
-\newblock Springer.
-\newblock Pages  159--187.
-
-\bibitem[\protect\citename{Hughes, }2000]{HughesArrows}
-Hughes, John. (2000).
-\newblock Generalising monads to arrows.
-\newblock {\em Science of computer programming}, {\bf 37}(1–3), 67--111.
-
-\bibitem[\protect\citename{Hughes, }2005]{Hughes2005}
-Hughes, John. (2005).
-\newblock {\em Programming with arrows}.
-\newblock AFP '04.
-\newblock Springer.
-\newblock Pages  73--129.
-
-\bibitem[\protect\citename{Jacobs {\em et~al.}\relax,
-  }2009]{jacobs_heunen_hasuo_2009}
-Jacobs, Bart, Heunen, Chris, \& Hasuo, Ichiro. (2009).
-\newblock Categorical semantics for arrows.
-\newblock {\em Journal of functional programming}, {\bf 19}(3-4), 403--438.
-
-\bibitem[\protect\citename{Janjic {\em et~al.}\relax, }2013]{janjic2013space}
-Janjic, Vladimir, Brown, Christopher~Mark, Neunhoeffer, Max, Hammond, Kevin,
-  Linton, Stephen~Alexander, \& Loidl, Hans-Wolfgang. (2013).
-\newblock Space exploration using parallel orbits: a study in parallel symbolic
-  computing.
-\newblock {\em Parallel computing}.
-
-\bibitem[\protect\citename{Karasawa \& Iwasaki, }2009]{5361825}
-Karasawa, Y., \& Iwasaki, H. (2009).
-\newblock A parallel skeleton library for multi-core clusters.
-\newblock {\em Pages  84--91 of:} {\em International Conference on Parallel
-  Processing 2009}.
-
-\bibitem[\protect\citename{Keller {\em et~al.}\relax,
-  }2010]{Keller:2010:RSP:1932681.1863582}
-Keller, Gabriele, Chakravarty, Manuel~M.T., Leshchinskiy, Roman, {Peyton
-  Jones}, Simon, \& Lippmeier, Ben. (2010).
-\newblock Regular, shape-polymorphic, parallel arrays in haskell.
-\newblock {\em {SIGPLAN} not.}, {\bf 45}(9), 261--272.
-
-\bibitem[\protect\citename{Kuchen, }2002]{Kuchen2002}
-Kuchen, Herbert. (2002).
-\newblock A skeleton library.
-\newblock {\em Pages  620--629 of:} Monien, Burkhard, \& Feldmann, Rainer
-  (eds), {\em Parallel Processing}.
-\newblock Euro-Par '02.
-\newblock Springer.
-
-\bibitem[\protect\citename{Kuper {\em et~al.}\relax,
-  }2014]{Kuper:2014:TPE:2666356.2594312}
-Kuper, Lindsey, Todd, Aaron, Tobin-Hochstadt, Sam, \& Newton, Ryan~R. (2014).
-\newblock Taming the parallel effect zoo: Extensible deterministic parallelism
-  with {LVish}.
-\newblock {\em {SIGPLAN} not.}, {\bf 49}(6), 2--14.
-
-\bibitem[\protect\citename{Lengauer {\em et~al.}\relax, }1997]{Lengauer1997}
-Lengauer, Christian, Gorlatch, Sergei, \& Herrmann, Christoph. (1997).
-\newblock The static parallelization of loops and recursions.
-\newblock {\em The journal of supercomputing}, {\bf 11}(4), 333--353.
-
-\bibitem[\protect\citename{Li \& Zdancewic, }2006]{1648705}
-Li, Peng, \& Zdancewic, S. (2006).
-\newblock Encoding information flow in {Haskell}.
-\newblock {\em Pages  12--16 of:} {\em 19\xth IEEE Computer Security
-  Foundations Workshop}.
-\newblock CSFW '06.
-
-\bibitem[\protect\citename{Li \& Zdancewic, }2010]{LI20101974}
-Li, Peng, \& Zdancewic, Steve. (2010).
-\newblock Arrows for secure information flow.
-\newblock {\em Theoretical computer science}, {\bf 411}(19), 1974--1994.
-
-\bibitem[\protect\citename{Lindley {\em et~al.}\relax, }2011]{LINDLEY201197}
-Lindley, Sam, Wadler, Philip, \& Yallop, Jeremy. (2011).
-\newblock Idioms are oblivious, arrows are meticulous, monads are promiscuous.
-\newblock {\em Electronic notes in theoretical computer science}, {\bf 229}(5),
-  97--117.
-
-\bibitem[\protect\citename{Linton {\em et~al.}\relax, }2010]{scscp}
-Linton, S., Hammond, K., Konovalov, A., Al~Zain, A.~D., Trinder, P., Horn, P.,
-  \& Roozemond, D. (2010).
-\newblock Easy composition of symbolic computation software: a new lingua
-  franca for symbolic computation.
-\newblock {\em Pages  339--346 of:} {\em Proceedings of the 2010 International
-  Symposium on Symbolic and Algebraic Computation}.
-\newblock ISSAC '10.
-\newblock ACM Press.
-
-\bibitem[\protect\citename{Liu {\em et~al.}\relax,
-  }2009]{Liu:2009:CCA:1631687.1596559}
-Liu, Hai, Cheng, Eric, \& Hudak, Paul. (2009).
-\newblock Causal commutative arrows and their optimization.
-\newblock {\em {SIGPLAN} not.}, {\bf 44}(9), 35--46.
-
-\bibitem[\protect\citename{Lobachev, }2011]{lobachev-phd}
-Lobachev, Oleg. (2011).
-\newblock {\em Implementation and evaluation of algorithmic skeletons:
-  Parallelisation of computer algebra algorithms}.
-\newblock Ph.D. thesis, Philipps-Universit{\"a}t Marburg.
-
-\bibitem[\protect\citename{Lobachev, }2012]{Lobachev2012}
-Lobachev, Oleg. (2012).
-\newblock {\em Parallel computation skeletons with premature termination
-  property}.
-\newblock FLOPS 2012.
-\newblock Springer.
-\newblock Pages  197--212.
-
-\bibitem[\protect\citename{Loogen {\em et~al.}\relax,
-  }2003]{Eden:SkeletonBookChapter02}
-Loogen, R., {Ortega-Mall\'{e}n}, Y., {Pe\~{n}a}, R., Priebe, S., \& Rubio, F.
-  (2003).
-\newblock {Parallelism Abstractions in Eden}.
-\newblock {\em Pages  71--88 of:} Rabhi, F.~A., \& Gorlatch, S. (eds), {\em
-  {Patterns and Skeletons for Parallel and Distributed Computing}}.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Loogen, }2012]{Loogen2012}
-Loogen, Rita. (2012).
-\newblock {\em Eden -- parallel functional programming with {Haskell}}.
-\newblock Springer.
-\newblock Pages  142--206.
-
-\bibitem[\protect\citename{Loogen {\em et~al.}\relax, }2005]{eden}
-Loogen, Rita, Ortega-Mall\'{e}n, Yolanda, \& {Pe\~{n}a-Mar\'{\i}}, Ricardo.
-  (2005).
-\newblock {Parallel Functional Programming in Eden}.
-\newblock {\em {Journal of Functional Programming}}, {\bf 15}(3), 431--475.
-\newblock Special Issue on Functional Approaches to High-Performance Parallel
-  Programming.
-
-\bibitem[\protect\citename{Lämmel, }2008]{LAMMEL20081}
-Lämmel, Ralf. (2008).
-\newblock Google’s mapreduce programming model — revisited.
-\newblock {\em Science of computer programming}, {\bf 70}(1), 1--30.
-
-\bibitem[\protect\citename{Maier {\em et~al.}\relax,
-  }2014]{Maier:2014:HDS:2775050.2633363}
-Maier, Patrick, Stewart, Robert, \& Trinder, Phil. (2014).
-\newblock The {HdpH} {DSLs} for scalable reliable computation.
-\newblock {\em {SIGPLAN} not.}, {\bf 49}(12), 65--76.
-
-\bibitem[\protect\citename{Mainland \& Morrisett,
-  }2010]{Mainland:2010:NEC:2088456.1863533}
-Mainland, Geoffrey, \& Morrisett, Greg. (2010).
-\newblock Nikola: Embedding compiled {GPU} functions in {Haskell}.
-\newblock {\em {SIGPLAN} not.}, {\bf 45}(11), 67--78.
-
-\bibitem[\protect\citename{Marlow {\em et~al.}\relax, }2009]{Marlow2009}
-Marlow, S., {Peyton Jones}, S., \& Singh, S. (2009).
-\newblock Runtime support for multicore {Haskell}.
-\newblock {\em {SIGPLAN} not.}, {\bf 44}(9), 65--78.
-
-\bibitem[\protect\citename{Marlow, }2013]{marlow2013parallel}
-Marlow, Simon. (2013).
-\newblock {\em Parallel and concurrent programming in {Haskell}: Techniques for
-  multicore and multithreaded programming}.
-\newblock "O'Reilly Media, Inc.".
-
-\bibitem[\protect\citename{Marlow {\em et~al.}\relax, }2011]{par-monad}
-Marlow, Simon, Newton, Ryan, \& {Peyton Jones}, Simon. (2011).
-\newblock A monad for deterministic parallelism.
-\newblock {\em {SIGPLAN} not.}, {\bf 46}(12), 71--82.
-
-\bibitem[\protect\citename{McDonell {\em et~al.}\relax,
-  }2015]{McDonell:2015:TRC:2887747.2804313}
-McDonell, Trevor~L., Chakravarty, Manuel M.~T., Grover, Vinod, \& Newton,
-  Ryan~R. (2015).
-\newblock Type-safe runtime code generation: Accelerate to {LLVM}.
-\newblock {\em {SIGPLAN} not.}, {\bf 50}(12), 201--212.
-
-\bibitem[\protect\citename{Michael {\em et~al.}\relax,
-  }2009]{Michael:2009:IWS:1594835.1504186}
-Michael, Maged~M., Vechev, Martin~T., \& Saraswat, Vijay~A. (2009).
-\newblock Idempotent work stealing.
-\newblock {\em {SIGPLAN} not.}, {\bf 44}(4), 45--54.
-
-\bibitem[\protect\citename{Nilsson {\em et~al.}\relax,
-  }2002]{Nilsson:2002:FRP:581690.581695}
-Nilsson, Henrik, Courtney, Antony, \& Peterson, John. (2002).
-\newblock Functional reactive programming, continued.
-\newblock {\em Pages  51--64 of:} {\em Proceedings of the 2002 {ACM} {SIGPLAN}
-  Workshop on {Haskell}}.
-\newblock Haskell '02.
-\newblock New York, NY, USA: ACM.
-
-\bibitem[\protect\citename{Olivier \& Prins, }2008]{4625841}
-Olivier, S., \& Prins, J. (2008).
-\newblock Scalable dynamic load balancing using {UPC}.
-\newblock {\em Pages  123--131 of:} {\em 37\xth International Conference on
-  Parallel Processing}.
-
-\bibitem[\protect\citename{Paterson, }2001]{Paterson:2001:NNA:507669.507664}
-Paterson, Ross. (2001).
-\newblock A new notation for arrows.
-\newblock {\em {SIGPLAN} not.}, {\bf 36}(10), 229--240.
-
-\bibitem[\protect\citename{Pe{\~n}a \& Rubio, }2001]{Eden:PPDP01}
-Pe{\~n}a, R., \& Rubio, F. (2001).
-\newblock {Parallel Functional Programming at Two Levels of Abstraction}.
-\newblock {\em Pages  187--198 of:} {\em PPDP'01 --- Intl.\ Conf.\ on
-  Principles and Practice of Declarative Programming}.
-
-\bibitem[\protect\citename{Perfumo {\em et~al.}\relax,
-  }2008]{Perfumo:2008:LST:1366230.1366241}
-Perfumo, Cristian, S\"{o}nmez, Nehir, Stipic, Srdjan, Unsal, Osman, Cristal,
-  Adri\'{a}n, Harris, Tim, \& Valero, Mateo. (2008).
-\newblock The limits of software transactional memory ({STM}): Dissecting
-  {Haskell} {STM} applications on a many-core environment.
-\newblock {\em Pages  67--78 of:} {\em Proceedings of the 5\xth Conference on
-  Computing Frontiers}.
-\newblock CF '08.
-\newblock ACM.
-
-\bibitem[\protect\citename{Poldner \& Kuchen, }2005]{Kuchen05}
-Poldner, Michael, \& Kuchen, Herbert. (2005).
-\newblock Scalable farms.
-\newblock {\em Pages  795--802 of:} Joubert, Gerhard~R., Nagel, Wolfgang~E.,
-  Peters, Frans~J., Plata, Oscar~G., Tirado, P., \& Zapata, Emilio~L. (eds),
-  {\em PARCO}.
-\newblock John von Neumann Institute for Computing Series, vol. 33.
-\newblock Central Institute for Applied Mathematics, J{\"u}lich, Germany.
-
-\bibitem[\protect\citename{Priebe, }2006]{WPEuropar06}
-Priebe, Steffen. (2006).
-\newblock Dynamic task generation and transformation within a nestable workpool
-  skeleton.
-\newblock  {\em Euro-Par}.
-\newblock LNCS 4128.
-
-\bibitem[\protect\citename{Rabhi \& Gorlatch, }2003]{SkeletonBook}
-Rabhi, F.~A., \& Gorlatch, S. (eds). (2003).
-\newblock {\em {Patterns and Skeletons for Parallel and Distributed
-  Computing}}.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Rudolph {\em et~al.}\relax,
-  }1991]{Rudolph:1991:SLB:113379.113401}
-Rudolph, Larry, Slivkin-Allalouf, Miriam, \& Upfal, Eli. (1991).
-\newblock A simple load balancing scheme for task allocation in parallel
-  machines.
-\newblock {\em Pages  237--245 of:} {\em Proceedings of the 3\xrd Annual ACM
-  Symposium on Parallel Algorithms and Architectures}.
-\newblock SPAA '91.
-\newblock ACM.
-
-\bibitem[\protect\citename{Russo {\em et~al.}\relax,
-  }2008]{Russo:2008:LLI:1411286.1411289}
-Russo, Alejandro, Claessen, Koen, \& Hughes, John. (2008).
-\newblock A library for light-weight information-flow security in {Haskell}.
-\newblock {\em Pages  13--24 of:} {\em Proceedings of the 1\st {ACM SIGPLAN}
-  Symposium on {Haskell}}.
-\newblock Haskell '08.
-\newblock ACM.
-
-\bibitem[\protect\citename{Stewart {\em et~al.}\relax,
-  }2016]{stewart_maier_trinder_2016}
-Stewart, Robert, Maier, Patrick, \& Trinder, Phil. (2016).
-\newblock Transparent fault tolerance for scalable functional computation.
-\newblock {\em Journal of functional programming}, {\bf 26}.
-
-\bibitem[\protect\citename{Svensson, }2011]{obsidian-phd}
-Svensson, Joel. (2011).
-\newblock {\em Obsidian: {GPU} kernel programming in {Haskell}}.
-\newblock Ph.D. thesis, Chalmers University of Technology.
-
-\bibitem[\protect\citename{Trinder {\em et~al.}\relax, }1996]{Trinder1996}
-Trinder, Phil~W., Hammond, Kevin, {Mattson Jr.}, James~S., Partridge,
-  Andrew~S., \& {Peyton Jones}, Simon~L. (1996).
-\newblock {GUM: a Portable Parallel Implementation of Haskell}.
-\newblock  {\em {PLDI'96}}.
-\newblock ACM Press.
-
-\bibitem[\protect\citename{Trinder {\em et~al.}\relax, }1998]{Trinder1998a}
-Trinder, P.W., Hammond, K., Loidl, H-W., \& {Peyton Jones}, S. (1998).
-\newblock {Algorithm + Strategy = Parallelism}.
-\newblock {\em Journal of functional programming}, {\bf 8}(1), 23--60.
-
-\bibitem[\protect\citename{van Nieuwpoort {\em et~al.}\relax,
-  }2001]{vanNieuwpoort:2001:ELB:568014.379563}
-van Nieuwpoort, Rob~V., Kielmann, Thilo, \& Bal, Henri~E. (2001).
-\newblock Efficient load balancing for wide-area divide-and-conquer
-  applications.
-\newblock {\em {SIGPLAN} not.}, {\bf 36}(7), 34--43.
-
-\bibitem[\protect\citename{Vizzotto {\em et~al.}\relax,
-  }2006]{vizzotto_altenkirch_sabry_2006}
-Vizzotto, Juliana, Altenkirch, Thorsten, \& Sabry, Amr. (2006).
-\newblock Structuring quantum effects: superoperators as arrows.
-\newblock {\em Mathematical structures in computer science}, {\bf 16}(3),
-  453--468.
-
-\bibitem[\protect\citename{Wheeler \& Thain, }2009]{Wheeler2009}
-Wheeler, K.~B., \& Thain, D. (2009).
-\newblock Visualizing massively multithreaded applications with {ThreadScope}.
-\newblock {\em Concurrency and computation: Practice and experience}, {\bf
-  22}(1), 45--67.
-
-\end{thebibliography}
+        \bibliographystyle{jfp}
+	\bibliography{references,main}
         \appendix

### 2019-12-27 paper/content/response.md
review response

Dear Editor,

thank you for the thorough reviews. We would like to use this document to provide a line-by-line response to the reviewers' comments.

> The reviewers generally agree that some technically sound work has been done, but they seriously question the necessity of having an arrow interface: one reviewer say it's not surprising, the other reviewer observes that the examples mostly force use of the pure function or Kleisli instance of arrows (meaning that they are just syntax on top of a monadic structure). 
> Moreover, IMHO the textual parts also need improvement: sometimes they are very terse, sometimes the line of attention jumps forth and back between the main text and the appendix, as also indicated by one of the reviewers.
> 
> Hence, I contend that significant work is needed on carving out the motivation for having the arrow interface (just "look, I can do it" is not enough) and also on smoothing out the nooks in the text. The reviewers suggest you resubmit and are all willing to review a reworked version of the paper.
> 
> Please note that resubmitting your manuscript does not guarantee eventual acceptance, and that your resubmission will be subject to re-review by the referees before a decision is rendered.

> __Referee: 1__
> Comments:
> 
> Overall the paper leaves a mixed impression. 
> 
> Of course, it is an interesting exercise to use arrows for parallel programming.
> In this sense the paper does a good job in explaining the background from the
> concept of arrows to the three parallel Haskell variants and then step by step
> lays out its story. On the way so many papers are referenced that a side 
> contribution of the paper would be the annotated bibliograpyhy for the topics
> concerned.
>

Thank you.

> But when approaching the final steps of the paper one question became 
> increasingly pending:  So what?
> 
> What is the real underlying contribution of the paper, in the sense of novel 
> insight, beyond the foreground contributions summarised in the paper?
>

TODO

> After all, it wasn't particularly difficult or challenging to use arrows for 
> parallel computing. Assuming the authors are right that no-one tried before
> them, and I have no reason to doubt this, maybe simply no-one has found it
> worthwhile so far. At least, I found the pure "it can be done" view on the
> contribution not entirely convincing. A relevant question, for me, would be
> whether the arrow-based approach makes parallel computing with Haskell
> in any sense simpler or more accessible to a larger audience than the 
> existing solutions, preferably beyond a pure matter of taste. This aspect is,
> unless I have overlooked it, not discussed in the paper, and so the whole 
> story gets a bit lost in technicalities.
> 
> One argument that the proposed interface is non-monadic could be considered
> an advantage over the Par monad, but neither over GpH nor over Eden, as 
> rightfully discussed in the conclusion. 
> 
> I agree that a common API across parallel Haskell variants has the obvious
> principle advantages, but that would hold for any sort of common API no matter
> how good or bad, with arrows or without.

There are two sides to our contribution. Firstly, we provide a common API to parallel Haskells, regardless its nature. Such an interface makes various approaches to parallelism in Haskell more comparable. Programs and benchmarks written on top of the common API would be more portable. Secondly, we did an intended design choice for such an API and used Arrows for it. Such approach appled more to us than an implementation of a further parallel Haskell using Arrows.

> The skeletons and benchmarks are all adopted from existing publications about
> the three parallel Haskell variants and merely ported to the arrow-based API.

We had multiple goals when designing the benchmarks. One of them was to be able to benchmark the Arrow-based API itself, i.e., to compare the performance of an Arrow-based benchmark with a non-Arrow benchmark. Naturally, we chose existing non-Arrow implementations to achieve this goal.

> The final observation that the proposed approach comes with marginal if any
> measurable overhead is at first glance good news. In fact, the opposite would
> surely be bad news and obviously question the entire approach. However, the
> results also demonstrate what the proposed solution effectively is: hardly 
> more than a thin layer of abstraction over three well established and 
> carefully engineered parallel implementations of Haskell.

If we are allowed to summarize, here you basically critique the shallowness of the API. However, as performance was vigilantly kept in mind during the design and implementation of PArrows, we consider this shallow layer a valuable contribution. With an abstraction combining Kleisli functions and pure functions in one notion, we are able to bridge conceptually very different flavors of parallel Haskells. And do so efficiently.

> ***************************************************************************
> Suggestions:
> 
> The paper is generally well written, but I nonetheless have a few 
> suggestions for improvement and minor corrections.
> 
> page 2:
> 
> "Further implementations ... are viable, too": What is this supposed to 
> precisely mean? I would assume that many implementations would be somewhat
> "viable". If you have accomplished concrete steps in the directions of
> HdpH or Frege, I suggest to explain that in concrete terms. As is, the text
> is too hand-waiving.

We adjusted the wording and made more clear, that we conducted some experiments with those further implementations, but do not focus on them in this work.

> 
> page 4:
> 
> "out of A list of Arrows"

Adjusted.

> 
> page 5:
> 
> Like on page 2, the diversion towards Frege remains cryptic.
>

We clarified the wording.

> page 7:
> 
> Provide a brief characterisation of profunctors, or at least a reference.
> "since they they allow"

We provided an additional reference explaining Profunctors in the main text, (Boisseau & Gibbons). It is possible to define Arrows deriving from Profunctors, as Appendix B shows. There we also cite Asada (2010).

> 
> The introductory text of section 3.2 (before 3.2.1) is mostly re-iterating 
> aspects already discussed in previous sections, partially with identical
> phrasing. I suggest to cut this short.
> 
> Other side issues like LVish or HdpH should have been discussed earlier.
> The background section should really focus on the relevant background,
> whereas the selection of background topics should be clear at this point.
> 
> page 10:
> 
> "efficiency OR the lack thereof" maybe?
> 
> page 11:
> 
> The dashes should be longer, or in Latex terms: "---".
> 
> page 13:
> 
> "instances of the ArrowParallel instances" ?
> 
> page 15:
> 
> What is meant be the term "implementation agnostic"? Any (high-level) 
> programming approach is somewhat implementation-agnostic.
> 
> page 16ff:
> 
> The text flow continuously jumps to the appendix and back; numerous forward
> references to figures in the appendix are increasingly annoying for the 
> reader. I believe that the authors should make a choice: if some example,
> implementation, illustration, etc, is relevant for the reader then it should
> appear immediately and not in the appendix. Everything else should either be
> skipped or the remaining contents of the appendix should be explained once.
> 
> page 16:
> 
> "toplogical"
> 
> page 19:
> 
> The question arising on the top of page 19 should be investigated. The statement
> that such differences would be beyond the scope of the paper appears a bit lame.
> 
> page 22:
> 
> "repeating": why is the front part of the word set in italics? 
> 
> page 23:
> 
> "using the Gentleman algorithm" -> "using Gentleman's algorithm" 
> 
> page 24:
> 
> If the purpose is to compare Figures 29 and 30, I strongly suggest to make
> sure they appear on the same page, or better the graphs are combined in a 
> single figure. A related question is: why should there be any observable
> difference? After all the exact same infrastructure and almost the same
> implementation are used.
> 
> page 25:
> 
> I find the statement that "PArrows hold up in terms of performance when 
> compared with the original parallel Haskells" a bit strong. At the end
> of the day, PArrows is not a new parallel Haskell, but merely an abstraction
> wrapper around the existing parallel Haskells. 
> 
> It is of little surprise that 32 hyperthreaded cores do not behave exactly as
> 16 real cores. Still the conclusion to disregard hyperthreading because it
> behaves differently appears illogic. I'm inclined to agree that hyperthreading
> might not provide any particular interesting insights for the questions
> concerned, but the justification should be rephrased.
> 
> page 27:
> 
> The text in 7.1.4 and the beginning of 7.2 should be condensed.
> 
> page 28:
> 
> A graphical illustration of the experimental findings appears mandatory.
> The mentioning of a zillion number is pretty unreadable and complicates
> comprehension of the results. The text should be used to explain the most
> interesting numbers and discuss the findings.
> 
> The term "speedup" is used extensively without ever explaining the base line.
> 
> page 30:
> 
> The absolute runtimes on 16 cores are fairly small. This raises the question
> whether larger problem sizes or simply repeated computing of the existing
> problem sizes could contribute to noise reduction and might possibly lead to
> more reliable experimental results.
> 
> "extensible formalism THAT can be easily ported"
> 
> The sentence "We argue that ..." is pretty much repeated right thereafter.
> 
> page 32ff:
> 
> The references contain numerous capitalisation errors. For conferences
> there is an unmotivated but consistent space between the acronym and the year. 
> 
> 
> ***************************************************************************
> 
> 
> 
> 
> Referee: 2
> 
> Comments to the Author
> Summary:
> 
> This paper introduces the use of Arrows for specifying parallelism. In
> particular, it defines an Arrow-based domain specific language
> (library) and implements it using multiple parallel Haskells.
> 
> More specifically, it introduces specialized Arrows as a general
> interface to parallel computations, by defining the ArrowParallel 
> type class and giving instance implementations for GpH, the Par Monad 
> and Eden. To demonstrate its expressiveness, it shows that the
> Parallel Arrows can be used to define some useful algorithmic
> skeletons that abstract typical parallel computations. The
> experimental results (on four examples) shows the minor overhead 
> induced by PArrows, which is outweighed by their uniform and general
> interface. 
> 
> Overall Evaluation:
> 
> This paper shows a useful development of a general arrow-based
> interface for specifying parallel computation in Haskell. This
> development is definitely worth doing (and I would wish an earlier
> development of such arrow-based interface).
> 
> The whole paper is clearly written and well structured. It explains
> not only the background and related work to make the paper
> self-contained, but also the motivation and the details of the
> arrow-based language for parallel programming in Haskell and the three
> instance implementations.
> 
> I like the paper and enjoyed reading it, but the *technical*
> contributions of this paper sounds limited with the standard of JFP.
> 
> I agree with the contributions listed in the introduction, but I am
> wondering how challenging it is to have these contributions. Following
> the research work on arrow-interfaces for GUI, inverse computation or
> functional reactive computation, it does not look difficult to define
> such an arrow-based interface for parallel programming for functional
> parallel programming. It is my hope that the authors can clarify the
> challenges and how these challenges are tacked in the later revision.
> 
> Detailed Comments:
> 
> Section 1:
> 
> On attractive point of this arrow-based general interface is that it
> allows to parallel implementation at will. It seems that it would be
> impossible for this switching unless every parallel computing
> primitive is wrapped with the general interface. Is the arrow-based
> language defined in this paper sufficient?  Or in other words, is
> there any parallelism in the current parallel Haskell system (say GhP)
> that cannot be described in this arrow-based interface?
> 
> Section 3:
> 
> For GpH, can you describe the two primitives par/seq in terms of
> PArrows? 
> 
> Section 4:
> 
> How general is the new interface language? I see that "parEvalN" in
> Section 4 needs to be extended to deal with "loop" later in Section
> 4. Do we need to extend it again to deal with other cases? Can you
> give a clear definition of the interface language, together with its
> application scope (including discussion what it cannot do)?
> 
> Section 5:
> 
> I got a bit lost when reading the "Future" class, and I cannot fully
> understand why it can support direct communications between
> nodes. What is the relationship between "Future" and "parEvalN"? Is
> "Future" part of PArrows?
> 
> Section 6:
> 
> At the beginning of this section, it says "Now we have developed
> Parallel Arrows far enough to .... While there are many possible
> skeletons to implement, we demonstrate the expressive power of PArrows
> here using four map-based and three toplogical skeletons."  Later in
> Section 6.2, it shows that the original developed PArrows are not good
> enough and should to be extended to deal with "loop". Does this mean
> that PArrows defined in this paper can be the basis for the future
> extension?
> 
> Referee: 3
> 
> Comments to the Author
> # Paper summary
> 
> There are now a large number of Haskell APIs and runtimes for parallel
> computing, from shared memory on a single machine to distributed
> computation. Programmers currently must choose a particular API and
> program to it, making it difficult to experiment with alternative
> parallelism solutions.
> 
> This paper proposes a common API, based on Hughes'
> Arrows. Implementations of this API for GpH, the Par Monad, and Eden
> are given. The authors demonstrate a number of programs written using
> this API, and present some benchmarks to show that using a common API
> does not introduce any significant overhead.
> 
> # Assessment
> 
> The authors present a (mostly) nice looking API for writing parallel
> programs in Haskell, that can be supported by at least three different
> backends, and has low overhead.
> 
> My problem with this paper is that I can't see why Arrows are
> essential. Since the title is "Arrows for Parallel Computation", this
> is a major problem.
> 
> Why don't I think Arrows are essential to this work? The authors
> present the ArrowParallel type class (Sec 4.1, pg 11):
> 
>    class Arrow arr => ArrowParallel arr a b conf where
>       parEvalN :: conf -> [arr a b] -> arr [a] [b]
> 
> and then give instances for this class for GpH, the Par monad, and
> Eden. The former two permit any Arrow instance that is also an
> ArrowChoice, but Eden only permits the (->) and Kleisli
> instances.
> 
> However, as far as I can tell, all the actual example parallel
> programs presented only use the (->) instance (Figure 28 seems to be
> the only serious program presented). Are there any examples of using
> other Arrow instances? Is there any point in using anything other than
> the (->) and Kleisli instances if they don't work with Eden?
> 
> The code in Figure 9 shows that a 'parEvalN' for monadic computations
> is derivable from one for pure arrows. So one could present the
> library as:
> 
>    class ParListEval arr a b conf where
>      parEvalN :: conf -> [a -> b] -> [a] -> [b]
> 
> and use the 'conf' parameter as a way of selecting between
> implementations as well as passing in configuration parameters. The
> Kleisli instance, if it is needed, can now be written as (following
> Fig 9):
> 
>    parEvalNM :: (Monad m, ParListEval arr a (m b) conf) => conf -> [a -> m b] -> [a] -> m [b]
>    parEvalNM conf fs as = sequence (parEvalN conf fs as)
> 
> Just using plain functions rather than the Arrow combinators would
> probably also lead to much easier to read code, especially when it
> comes to writing the topological skeletons in Section 6.
> 
> Now, it would be an interesting observation that the three parallelism
> libraries considered here can be unified behind this
> interface. However, given that the whole set up that this paper
> proposes seems unnecessary, I recommend reject and resubmit.
> 
> If there were compelling examples of uses of non `(->)` arrows for
> writing parallel programs, then I'd be happier with this paper.
> 
> ## Other comments
> 
> - The `parEvalN` function is the central contribution of this paper,
>  but very little is specified about how it should work, beyond "in
>  parallel" and what is specified by the types. For instance, what
>  laziness behaviour is expected of this function? (this is very
>  important for the looping examples later). What happens if the two
>  input lists have different lengths? Since the Eden implementation is
>  precisely Eden's `spawnF`, then aren't you in effect just settling
>  for whatever specification Eden provides?
> 
> - It is necessary to add a facility for 'Futures' to organise
>  inter-node communication more effectively. This is perhaps
>  necessary, but it spoils the "backend agnostic" approach -- if one
>  is testing on GpH and then deploying to Eden then adding Futures to
>  a GpH targeted implementation is just noise. On the other hand, it
>  is good that they are semantically "the identity", and do not affect
>  the overall result.
> 
>  I found the introduction of Futures in Section 5 quite difficult to
>  understand until I guessed that they are effectively 'handles' that
>  are passed around until 'get' retrieves the actual data -- this
>  isn't explained! There is also no explanation of how Futures fix the
>  problem with the combinator in Figure 13 (why is it called
>  'someCombinator' and not 'outlineCombinator'?)
> 
> - The ArrowLoopParallel type class on page 19. The only reason for
>  this type class is to paper over differences in evaluation order
>  between GpH and the Par Monad, and Eden. For a journal paper, I
>  would have expected some more investigation into the cause of the
>  discrepancy between the backends. Does the Eden implementation force
>  more of the input lists in one go? Is there a smallest example that
>  shows the problem?
> 
>    - Since you are in control of the implementations of parEvalN for GpH
>      and Par, then is it not possible to alter them to match Eden?
> 
>    - The fact that the programmer has to remember where to use
>      `parEvalN`, `loopParEvalN` and `postLoopParEvalN` is a burden,
>      and it seems that if they mix it up, it will depend on the
>      backend whether or not the program will crash. It also affects
>      the composability of the library. What if someone wants to use,
>      e.g., the `torus` skeleton inside another `loop`?
> 
>    - Am I right in thinking that `postLoopParEvalN` is always used
>      with `(repeat (arr id))` (except in `simplePipe`)? So it is
>      being used to insert parallelism after the loop? If this is the
>      pattern always being used, then why not provide a special `loop`
>      combinator that does the right thing for parallel arrows? For
>      example:
> 
>            loopPar conf f = loop f >>> parEvalN conf (repeat (arr id))
> 
>      for GpH and the Par monad, and
> 
>            loopPar conf f = loop f
> 
>      for Eden (this would remove any overhead of mapping `id` over
>      everything afterwards in Eden). The programmer would still have
>      to remember to use `loopParEvalN` inside the loop though.
> 
> - The further examples of using parEvalN to reconstruct known
>  parallelism strategies and skeletons are also unsatisfying. The
>  Heterogenous tasks in 4.3.2 rely on the use of partial functions
>  like 'head' to work, and indicate that the parEvalN interface is
>  perhaps too lax with types and that perhaps an interface like:
> 
>        parEval2 :: (a -> b) -> (c -> d) -> (a,c) -> (b,d)
> 
>  could be provided as well as parEvalN?
> 
> - The implementations of topological skeletons in Section 6.2 start to
>  get quite involved quite quickly, due to the use of laziness, and
>  I'm not convinced that the Arrow interface is really helpful
>  here. If they were written just using functions they'd look like
>  normal Haskell functions, albeit ones that have quite subtle
>  laziness behaviour. For instance, pipeSimple can be rewritten as the
>  considerably simpler looking:
> 
>        pipeSimple conf fs a =
>           let xs = parEvalN conf fs (lazy (a : tail xs)) in last (a:xs)
> 
>  this works with the following definition of `parEvalN` for GpH:
> 
>        parEvalN fs xs = evalN fs xs `using` parList rseq
> 
>  (which also has the advantage of (a) working on the empty list of
>  functions; and (b) running the pipeline in parallel, unlike in the
>  GpH instance in the paper due to its instance of
>  `ArrowLoopParallel`). I think the other examples, `pipe2`, `ring`, and
>  `torus` would also benefit from not using the Arrow
>  interface. (unless I'm wrong that there are other useful Arrow
>  instances for ParallelArrow?)
> 
>  I think the subtle uses of laziness in these definitions also
>  requires some discussion -- for instance the dataflow going in
>  `pipeSimple`, why is this loop OK?
> 
> - The benchmarks don't cover all the skeletons discussed. In
>  particular, the pipe and ring skeletons are not covered by any of
>  the four benchmarks. Since the whole point of parallelisation is to
>  make programs go faster, it would seem to me that a particular
>  construct in a paper about parallelism should only appear if it can
>  be shown to actually make programs go faster.
> 
> - A frustrating amount of the code is placed in the appendix (mainly
>  appendix C) and referenced from the main text. It would be much
>  easier to read the paper if all of the code was inlined in the
>  text. In the present state, the reader has to keep flipping to the
>  end to find the implementations of crucial bits of code (e.g.,
>  `lazy`).
> 
> - Throughout, the writing is a bit sloppy. I've highlighted some
>  examples below.
> 
> ## Text comments
> 
> ### Page 2
> 
> Para 1, line 3: "short" => "for short"
> 
> Para 2, line 1: "A key" => "The key"
> 
> Para 3: line 1: "Arrows" => "Arrow"
> 
> Para 3: line 3: "we define a shallow-embedded DSL for Arrows". Aren't
> you defining a shallowly embedded DSL for parallel programming,
> *using* Arrows?
> 
> Para 3: line 6: "very" is vague.
> 
> Para 4: line 2: is there a missing 'Arrow arr' constraint on the type
> signature?
> 
> Para 6: line 1: this is the first mention of "PArrow" -- it should be
> defined earlier.
> 
> ### Page 3
> 
> Para 1: line 1: "This has many practical advantages". Are there any
> others beyond being able to switch backend easily?
> 
> Para 6: line 2: "parallelisations" => "execution strategies"
> 
> Para 6: line 7: "is completely orthogonal to our approach". This seems
> a bit defensive. But since you bring it up, then why not explain why
> it is orthogonal?
> 
> Para 8: line 2: Should be a colon after "following"?
> 
> Para 8: line 3: I'm not sure what you mean by "communication-centred"
> in relation to LVars?
> 
> ### Page 4
> 
> The section on Algorithmic skeletons needs a forward reference to
> later in the paper where you discuss how to express skeletons using
> your formalism.
> 
> "Arrows" para: "less restrictive alternative", well, it is less
> restrictive for the writer of the arrow, but not for the user.
> 
> "Arrow": line 10: "capsule the computation", do you mean "encapsulate
> the computation"? This whole sentence seems quite vague anyway. I'd
> characterise Liu et al's work as providing a specific implementation
> of arrows that allows for optimisation of arrows with feedback by
> retaining some intensional information about the computation. The
> comparison with you work confuses me: you say that their arrows allow
> parallel composition, but then seem to say that yours is different?
> What is a thing that you can express but they can't?
> 
> "Arrow": line 14: missing "a" before "list".
> 
> "Arrow": line 15: "different" => "differently"
> 
> "Arrow": line 16: "several tasks in robotics". This is ambiguous, do
> you mean several different sorts of tasks that a robot might be asked
> to do separately, or orchestrate the tasks at runtime?
> 
> Another potential piece of related work is Turon's Reagents, that are
> based on an Arrow-like interface (albeit for concurrency, not
> parallelism): https://people.mpi-sws.org/~turon/reagents.pdf
> 
> ### Page 5
> 
> Para 1: Line 2: "which is basically"; better "A Haskell-like language
> for the JVM"
> 
> Para 2; line 2: "typical GUI tasks" => you could be less vague.
> 
> Sec 3.1, para 1, last line: why not use the typeclass constraint
> "Monad m =>" instead of saying "where m is a Monad"?
> 
> Sec 3.1, para 2, first line: What does the "This" at the end of the
> line refer to? Also, there is no mention of the Arrow laws.
> 
> Sec 3.1, para 3: "syntactic sugar". I don't think these derived
> combinators should be called syntactic sugar, because you aren't
> defining the syntax of a language anywhere. They are either "derived
> combinators", or more prosaically, "utility functions".
> 
> ### Page 6
> 
> The diagramatic representations of arrows on this page and the next
> are nice, but they don't seem to be used anywhere else. The diagrams
> in the text seem to be a mixture of communication topologies, mappings
> from input to output locations, and these arrow diagrams.
> 
> `ArrowApply`, `ArrowLoop` and `ArrowChoice` aren't defined anywhere -- how much
> background knowledge about Arrows do you expect the reader to have?
> 
> "we have a truly more general interface as compared to a monadic one"
> -- but this isn't true for Eden right? And there are no examples in
> the paper of using non `(->)` arrows to write programs, unless I've
> missed one?
> 
> ### Page 7
> 
> You need some reference to where Profunctors are defined, or at least
> a short introduction to them.
> 
> "Among the most important are probably" ... "probably" is a vague. You
> could just say "the ones we focus on are...". Also, this sentence is
> sort of claiming that you chose these ones because they are the "most
> important", but the first sentence on the next page gives reasons in
> terms of the features of each one.
> 
> ### Page 9
> 
> You could write `parEvalN` for the Par monad more concisely as:
> 
>    parEvalN fs as = runPar $ mapM (return >>> spawn >>= get) $ zipWith ($) fs as
> 
> ### Page 10
> 
> You don't explain what the `Trans` typeclass is. `Trans`portable?
> 
> ### Page 11
> 
> Why not put the `evalN` definition in the text for the reader?
> 
> ### Page 12
> 
> Figure 7: why not use a `newtype` for `Conf`?
> 
> Why is `Conf` for GpH in the figure (Fig 7), but not for the Par Monad
> instance (Fig 8)?
> 
> ### Page 13
> 
> Figure 9: does the second instance need the `Trans b` constraint?
> Should the `conf` before the `where` be a `Conf`?
> 
> ### Page 14
> 
> Figure 11: what is the `ChunkSize` type? `Int`?
> 
> Where is the `ArrowApply` constraint used? Is it in `chunksOf` -- I
> can't find a definition of this function in the paper.
> 
> Section 4.3.2: the readbility of this subsection would be greatly
> improved by having the code of the function you are talking about!
> 
> ### Page 16
> 
> As I said above, Section 5 doesn't really explain how Futures fix the
> problem.
> 
> ### Page 17
> 
> Fig 17, caption: "Other than" => "Unlike"
> 
> ### Page 18
> 
> second last line: remove the final "the"
> 
> ### Page 19
> 
> first line: remove "when"
> 
> ### Page 20
> 
> paragraph 3: "probably": do some benchmarks?
> 
> last line: "send" => "sent"
> 
> ### Page 22
> 
> "This combinator can, for example, be used to calculate the shortest
> paths in a graph using Warshall's algorithm." Have you done this?
> Could it be included in the benchmarks?
> 
> ### Page 23
> 
> Figure 27 is where the code starts to get really unreadable. If you
> are going to stick with Arrows, then why not use Paterson's arrow
> notation? Or give a diagramatic representation to help the reader?
> 
> ### Page 25
> 
> "We disregarded the hyper-threading ability in most of the cases."
> There should be a forward reference to the section discussing hyper
> threading specifically.
> 
> ### Page 37
> 
> last line: "as" => "was"
> 
> ### Page 38
> 
> The type signature for `(|||)` seems messed up (the ArrowChoice
> typeclass constraint is muddled up with the type of the first
> argument).

### 2019-12-27 paper/content/arxiv/ms.tex
remaining files

@@
 
-\ptcomment{The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations.
-A range of approaches have been explored, including data parallelism (Chakravarty et al., 2007; Keller et al.,
-2010), GPU-based approaches (Mainland & Morrisett, 2010; Svensson, 2011), software
-transactional memory (Harris et al., 2005; Perfumo et al., 2008). The Haskell–GPU bridge
-Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
-is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.
-
-Our PArrow implementation uses three task parallel languages as backends: the
-GpH (Trinder et al., 1996, 1998) parallel Haskell dialect and its multicore version (Marlow
-et al., 2009), the \ensuremath{\Conid{Par}} Monad (Marlow et al., 2011; Foltzer et al., 2012), and Eden (Loogen
-et al., 2005; Loogen, 2012). These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available (Aljabri et al., 2014,
-2015). Research on Eden includes a low-level  implementation (Berthold, 2008; Berthold et al., 2016), work on skeleton
-composition (Dieterle et al., 2016), communication (Dieterle et al., 2010a), and generation
-of process networks (Horstmeyer & Loogen, 2013). The definitions of new Eden skeletons is a specific focus (Hammond et al., 2003; Berthold
-& Loogen, 2006; Berthold et al., 2009b,c; Dieterle et al., 2010b; de la Encina et al., 2011;
-Dieterle et al., 2013; Janjic et al., 2013).
-
-Other task parallel Haskells, related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad, include the following.  
-HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of \ensuremath{\Conid{Par}} Monad to
-heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
-of \ensuremath{\Conid{Par}} Monad.}
-We chose to not extend our implementation to LVish and HdpH at the moment.
-The former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper.
-The latter heavy relies on Template Haskell, which adds complexity to the implementation.
-
 \paragraph{Parallel Haskells.}
@@
 of \ensuremath{\Conid{Par}} Monad.
-
-\olcomment{OLD:
-\paragraph{Parallel Haskells.}
-Of course, the three parallel Haskell flavours we use as backends: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the \ensuremath{\Conid{Par}} Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work.
-We use these languages as backends: our DSL can switch from one to another at user's command.
-
-HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} Monad to heterogeneous clusters.
-LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of \ensuremath{\Conid{Par}} Monad.
-%
-Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015}, and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}.
-Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden.
-This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.
-
-Alternative approaches include data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
-%
-The Haskell--GPU bridge Accelerate
-\cite{Chakravarty:2011:AHA:1926354.1926358,CMCK14,McDonell:2015:TRC:2887747.2804313}
-deserves a special mention. Accelerate is completely orthogonal to our
-approach. \citeauthor{marlow2013parallel} authored a recent book in
-\citeyear{marlow2013parallel} on parallel Haskells.
-}
+We chose to not extend our implementation to LVish and HdpH at the moment.
+The former does not differ from the original \ensuremath{\Conid{Par}} Monad with regard to how we would have used it in this paper.
+The latter heavy relies on Template Haskell, which adds complexity to the implementation.
 

### 2019-12-27 paper/content/ms.tex
shortening; explaining speedup

--- 2018-01-07 paper/content/ms.tex
@@
 
-\ptcomment{The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations.
-A range of approaches have been explored, including data parallelism (Chakravarty et al., 2007; Keller et al.,
-2010), GPU-based approaches (Mainland & Morrisett, 2010; Svensson, 2011), software
-transactional memory (Harris et al., 2005; Perfumo et al., 2008). The Haskell–GPU bridge
-Accelerate (Chakravarty et al., 2011; Clifton-Everest et al., 2014; McDonell et al., 2015)
-deserves a special mention. <<Why does it deserve special mention?>>  Accelerate is completely orthogonal to our approach. A good survey of parallel Haskells can be found in ~\cite{Marlow2013}.
-
-Our PArrow implementation uses three task parallel languages as backends: the
-GpH (Trinder et al., 1996, 1998) parallel Haskell dialect and its multicore version (Marlow
-et al., 2009), the \ensuremath{\Conid{Par}} Monad (Marlow et al., 2011; Foltzer et al., 2012), and Eden (Loogen
-et al., 2005; Loogen, 2012). These languages are under active development, for example a combined shared and distributed memory implementation of GpH is available (Aljabri et al., 2014,
-2015). Research on Eden includes a low-level  implementation (Berthold, 2008; Berthold et al., 2016), work on skeleton
-composition (Dieterle et al., 2016), communication (Dieterle et al., 2010a), and generation
-of process networks (Horstmeyer & Loogen, 2013). The definitions of new Eden skeletons is a specific focus (Hammond et al., 2003; Berthold
-& Loogen, 2006; Berthold et al., 2009b,c; Dieterle et al., 2010b; de la Encina et al., 2011;
-Dieterle et al., 2013; Janjic et al., 2013).
-
-Other task parallel Haskells, related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad, include the following.  
-HdpH (Maier et al., 2014; Stewart et al., 2016) is an extension of \ensuremath{\Conid{Par}} Monad to
-heterogeneous clusters. LVish (Kuper et al., 2014) is a communication-centred extension
-of \ensuremath{\Conid{Par}} Monad.}
-
 \paragraph{Parallel Haskells.}
@@
 We, however, propose a general interface for parallel programming, while remaining completely in Haskell.
+\citet{zhang_session_2019} introduced a session-type framework that generates parallel code. As they stated, their approach to express dataflow with Arrows was inspired by our formalism.
 
@@
 
-While we could have based our DSL on Profunctors as well, we chose Arrows\mbcomment{cite missing?} for this paper since they they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
-However, they are a promising candidate for future improvements of our DSL.
-Some Profunctors, especially ones supporting a composition operation, choice, and looping, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
+While our DSL could have been based on Profunctors \citep[see, e.g.][]{boisseau_what_2018} directly, we chose Arrows for this paper for the easier representation of composability.
+Special Profunctors supporting a composition operation, choice, and looping, can already be used as a foundation of our interface as shown in Appendix~\ref{app:profunctorArrows}.
 	\subsection{Short introduction to parallel Haskells}
@@
 hyper-threaded cores and only 13.55 with PArrows/GpH. 
-
-While this shows that hyper-threading can be of benefit in scenarios similar to the ones presented in the benchmarks, we only use real cores for the performance measurements in Section~\ref{sec:benchmarkResults} as the purpose of this paper is to show the performance of PArrows and not to investigate parallel behaviour with hyper-threading.
+Although above demonstrates that hyper-threading can be of benefit, we only use real cores for the performance measurements in Section~\ref{sec:benchmarkResults}.
 
@@
 We compare the PArrow performance with direct implementations of the benchmarks in Eden, GpH and the \ensuremath{\Conid{Par}} Monad.
-We start with the definition of mean overhead to compare both PArrows-enabled and standard benchmark implementations.
-We continue comparing speedups and overheads for the shared memory implementations and then study OpenMPI variants of the Eden-enabled PArrows as a representative of a distributed memory backend.
+We start with the definition of mean overhead in Section~\ref{sec:def-overhead}.
+We continue comparing speedups and overheads for the shared memory implementations (Section~\ref{sec:pref-shm}) and  OpenMPI variants (Section~\ref{sec:pref-dm}).
 We plot all speedup curves and all overhead values in the supplementary materials.
 
-\subsubsection{Defining overhead}
+\subsubsection{Defining overhead and speedup}
+\label{sec:def-overhead}
 
@@
 
+The speedup is a measure of parallel quality.
+Basically, it is a quotient of sequential execution time to the parallel execution time; speedup measures the speed factor from running a parallel program compared to a sequential one.
+The ideal speedup is linear speedup, e.g.
+32 for 32 cores.
+
 \subsubsection{Shared memory}
+\label{sec:pref-shm}
 
@@
 \subsubsection{Distributed memory}
+\label{sec:pref-dm}
 
@@
 
-\input{bestAndWorstBenchmarks}
+%\bestAndWorstBenchmarks
+
+\begin{table}[]
+\centering
+\caption{Overhead in the shared memory benchmarks. Bold marks values in favour of PArrows.}
+\label{tab:meanOverheadSharedMemory}
+\centering
+\begingroup\catcode`"=9
+\sisetup{round-mode=places, round-precision=3}
+\begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 16 cores (s)} \\ \hline \\
+	\multirow{3}{*}{\sudokutest 1000}
+	\csvreader[head to column names]{bestAndWorstSudoku-1000.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{1}{*}{\torustest 512}
+	\csvreader[head to column names]{bestAndWorstTorusSM-512.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{3}{*}{\rmtest 11213 32}
+	\csvreader[head to column names]{bestAndWorstRMSM-11213-32.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{3}{*}{\rmtest 11213 64}
+	\csvreader[head to column names]{bestAndWorstRMSM-11213-64.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+\end{tabular}
+\endgroup
+\end{table}
+
+\begin{table}[]
+\centering
+\caption{Overhead in the distributed memory benchmarks. Bold marks values
+  in favour of PArrows.}
+% \olcomment{Vielleicht gleich durchmultiplizieren und \% angeben? Nicht
+% zu viele leere Zeilen in der Tabelle?}
+\label{tab:meanOverHeadDistributedMemory}
+\centering
+\begingroup\catcode`"=9
+\begin{tabular}{C{4cm} C{2cm} C{1.7cm} C{2.3cm} C{1.7cm}}
+	\thead{Benchmark} & \thead{Base}             & \thead{Mean of \\ mean \\overheads} & \thead{Maximum \\ normalised \\ stdDev} & \thead{Runtime for \\ 256 cores (s)} \\ \hline \\
+	\multirow{1}{*}{\torustest 4096}
+	\csvreader[head to column names]{bestAndWorstTorus-4096.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+  	\multirow{1}{*}{\rmtest 44497 256}
+	\csvreader[head to column names]{bestAndWorstRM-44497-256.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+	\\ \hline \\
+	\multirow{1}{*}{\jacobitest 3217}
+	\csvreader[head to column names]{bestAndWorstJacobi-3217.csv}{}{& \vs & \meanOverhead & \maxStdDevForOverhead & \runtimeMaxCores \\}
+\end{tabular}
+\endgroup
+\end{table}
 
@@
 	%\pagebreak
-        %\bibliographystyle{jfp}
-	%\bibliography{references,main}
-        \providecommand{\noopsort}[1]{}
-\begin{thebibliography}{}
-
-\bibitem[\protect\citename{Acar {\em et~al.}\relax,
-  }2000]{Acar:2000:DLW:341800.341801}
-Acar, Umut~A., Blelloch, Guy~E., \& Blumofe, Robert~D. (2000).
-\newblock The data locality of work stealing.
-\newblock {\em Pages  1--12 of:} {\em Proceedings of the 12\xth Annual ACM
-  Symposium on Parallel Algorithms and Architectures}.
-\newblock SPAA '00.
-\newblock ACM.
-
-\bibitem[\protect\citename{Achten {\em et~al.}\relax, }2007]{achten2007arrow}
-Achten, Peter, van Eekelen, Marko, de~Mol, Maarten, \& Plasmeijer, Rinus.
-  (2007).
-\newblock An arrow based semantics for interactive applications.
-\newblock  {\em Draft Proceedings of the Symposium on Trends in Functional
-  Programming}.
-\newblock TFP '07.
-
-\bibitem[\protect\citename{Achten {\em et~al.}\relax, }2004]{achten2004arrows}
-Achten, PM, van Eekelen, Marko~CJD, Plasmeijer, MJ, \& Weelden, A~van. (2004).
-\newblock {\em Arrows for generic graphical editor components}.
-\newblock Tech. rept. Nijmegen Institute for Computing and Information
-  Sciences, Faculty of Science, University of Nijmegen, The Netherlands.
-\newblock Technical Report NIII-R0416.
-
-\bibitem[\protect\citename{Alimarine {\em et~al.}\relax,
-  }2005]{Alimarine:2005:BAA:1088348.1088357}
-Alimarine, Artem, Smetsers, Sjaak, van Weelden, Arjen, van Eekelen, Marko, \&
-  Plasmeijer, Rinus. (2005).
-\newblock There and back again: Arrows for invertible programming.
-\newblock {\em Pages  86--97 of:} {\em Proceedings of the {2005} {ACM}
-  {SIGPLAN} Workshop on {Haskell}}.
-\newblock Haskell '05.
-\newblock ACM.
-
-\bibitem[\protect\citename{Aljabri {\em et~al.}\relax,
-  }2014]{Aljabri:2013:DIG:2620678.2620682}
-Aljabri, Malak, Loidl, Hans-Wolfgang, \& Trinder, Phil~W. (2014).
-\newblock The design and implementation of {GUMSMP}: A multilevel parallel
-  haskell implementation.
-\newblock {\em Pages  37:37--37:48 of:} {\em Proceedings of the 25\xth
-  Symposium on Implementation and Application of Functional Languages}.
-\newblock IFL '13.
-\newblock ACM.
-
-\bibitem[\protect\citename{Aljabri {\em et~al.}\relax, }2015]{Aljabri2015}
-Aljabri, Malak, Loidl, Hans-Wolfgang, \& Trinder, Phil. (2015).
-\newblock {\em Balancing shared and distributed heaps on {NUMA} architectures}.
-\newblock Springer.
-\newblock Pages  1--17.
-
-\bibitem[\protect\citename{Alt \& Gorlatch, }2003]{AlGo03a}
-Alt, Martin, \& Gorlatch, Sergei. (2003).
-\newblock {F}uture-{B}ased {RMI}: Optimizing compositions of remote method
-  calls on the {G}rid.
-\newblock {\em Pages  682--693 of:} Kosch, Harald, B{\"o}sz{\"o}rm{\'e}nyi,
-  L{\'a}szl{\'o}, \& Hellwagner, Hermann (eds), {\em Euro-Par~2003}.
-\newblock LNCS 2790.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Asada, }2010]{Asada:2010:ASM:1863597.1863607}
-Asada, Kazuyuki. (2010).
-\newblock Arrows are strong monads.
-\newblock {\em Pages  33--42 of:} {\em Proceedings of the Third ACM SIGPLAN
-  Workshop on Mathematically Structured Functional Programming}.
-\newblock MSFP '10.
-\newblock New York, NY, USA: ACM.
-
-\bibitem[\protect\citename{Aswad {\em et~al.}\relax, }2009]{aswad2009low}
-Aswad, Mustafa, Trinder, Phil, Al~Zain, Abdallah, Michaelson, Greg, \&
-  Berthold, Jost. (2009).
-\newblock Low pain vs no pain multi-core {Haskells}.
-\newblock {\em Pages  49--64 of:} {\em Trends in Functional Programming}.
-
-\bibitem[\protect\citename{Atkey, }2011]{ATKEY201119}
-Atkey, Robert. (2011).
-\newblock What is a categorical model of arrows?
-\newblock {\em Electronic notes in theoretical computer science}, {\bf 229}(5),
-  19--37.
-
-\bibitem[\protect\citename{Berthold, }2008]{JostThesis}
-Berthold, Jost. (2008).
-\newblock {\em Explicit and implicit parallel functional programming ---
-  concepts and implementation}.
-\newblock Ph.D. thesis, Philipps-Universit\"{a}t Marburg.
-
-\bibitem[\protect\citename{Berthold \& Loogen, }2006]{Eden:PARCO05}
-Berthold, Jost, \& Loogen, Rita. (2006).
-\newblock Skeletons for recursively unfolding process topologies.
-\newblock  Joubert, Gerhard~R., Nagel, Wolfgang~E., Peters, Frans~J., Plata,
-  Oscar~G., Tirado, P., \& Zapata, Emilio~L. (eds), {\em Parallel Computing:
-  Current \& Future Issues of High-End Computing, ParCo 2005, Malaga, Spain}.
-\newblock NIC Series 33.
-\newblock Central Institute for Applied Mathematics, J{\"u}lich, Germany.
-
-\bibitem[\protect\citename{Berthold \& Loogen, }2007]{Berthold2007a}
-Berthold, Jost, \& Loogen, Rita. (2007).
-\newblock {Visualizing Parallel Functional Program Executions: Case Studies
-  with the Eden Trace Viewer}.
-\newblock  {\em ParCo '07. Parallel Computing: Architectures, Algorithms and
-  Applications}.
-\newblock IOS Press.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax, }2008]{PADL08HMWS}
-Berthold, Jost, Dieterle, Mischa, Loogen, Rita, \& Priebe, Steffen. (2008).
-\newblock Hierarchical master-worker skeletons.
-\newblock  Warren, David~S., \& Hudak, Paul (eds), {\em Practical Aspects of
-  Declarative Languages (PADL'08)}.
-\newblock LNCS 4902.
-\newblock San Francisco (CA), USA: Springer-Verlag.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax, }2009a]{arcs-dc}
-Berthold, Jost, Dieterle, Mischa, Lobachev, Oleg, \& Loogen, Rita. (2009a).
-\newblock {Distributed Memory Programming on Many-Cores -- A Case Study Using
-  {Eden} Divide-\&-Conquer Skeletons}.
-\newblock {\em Pages  47--55 of:} Gro\ss{}pitsch, K.-E., Henkersdorf, A.,
-  Uhrig, S., Ungerer, T., \& H\"{a}hner, J. (eds), {\em Workshop on Many-Cores
-  at ARCS '09 -- 22\nd International Conference on Architecture of Computing
-  Systems 2009}.
-\newblock VDE-Verlag.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
-  }2009b]{Berthold2009-mr}
-Berthold, Jost, Dieterle, Mischa, \& Loogen, Rita. (2009b).
-\newblock Implementing parallel {Google} map-reduce in {Eden}.
-\newblock {\em Pages  990--1002 of:} Sips, Henk, Epema, Dick, \& Lin, Hai-Xiang
-  (eds), {\em Euro-Par 2009 Parallel Processing}.
-\newblock LNCS 5704.
-\newblock Springer Berlin Heidelberg.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
-  }2009c]{Berthold2009-fft}
-Berthold, Jost, Dieterle, Mischa, Lobachev, Oleg, \& Loogen, Rita. (2009c).
-\newblock {\em Parallel {FFT} with {Eden} skeletons}.
-\newblock PaCT '09.
-\newblock Springer.
-\newblock Pages  73--83.
-
-\bibitem[\protect\citename{Berthold {\em et~al.}\relax,
-  }2016]{berthold_loidl_hammond_2016}
-Berthold, Jost, Loidl, Hans-Wolfgang, \& Hammond, Kevin. (2016).
-\newblock {PAEAN}: Portable and scalable runtime support for parallel {Haskell}
-  dialects.
-\newblock {\em Journal of functional programming}, {\bf 26}.
-
-\bibitem[\protect\citename{Bischof \& Gorlatch, }2002]{Bischof2002}
-Bischof, Holger, \& Gorlatch, Sergei. (2002).
-\newblock {\em Double-scan: Introducing and implementing a new data-parallel
-  skeleton}.
-\newblock Euro-Par '02.
-\newblock Springer.
-\newblock Pages  640--647.
-
-\bibitem[\protect\citename{Blumofe \& Leiserson,
-  }1999]{Blumofe:1999:SMC:324133.324234}
-Blumofe, Robert~D., \& Leiserson, Charles~E. (1999).
-\newblock Scheduling multithreaded computations by work stealing.
-\newblock {\em J. acm}, {\bf 46}(5), 720--748.
-
-\bibitem[\protect\citename{Botorog \& Kuchen, }1996]{botorog1996efficient}
-Botorog, G.~H., \& Kuchen, H. (1996).
-\newblock {\em {Euro-Par'96 Parallel Processing}}.
-\newblock LNCS 1123.
-\newblock Springer-Verlag.
-\newblock Chap. Efficient parallel programming with algorithmic skeletons,
-  pages  718--731.
-
-\bibitem[\protect\citename{Brown \& Hammond, }2010]{brown2010ever}
-Brown, C., \& Hammond, K. (2010).
-\newblock Ever-decreasing circles: a skeleton for parallel orbit calculations
-  in {Eden}.
-\newblock  {\em Draft Proceedings of the Symposium on Trends in Functional
-  Programming}.
-\newblock TFP '10.
-
-\bibitem[\protect\citename{Buono {\em et~al.}\relax, }2010]{BUONO20102095}
-Buono, D., Danelutto, M., \& Lametti, S. (2010).
-\newblock Map, reduce and mapreduce, the skeleton way.
-\newblock {\em Procedia computer science}, {\bf 1}(1), 2095--2103.
-\newblock ICCS 2010.
-
-\bibitem[\protect\citename{{C}hakravarty {\em et~al.}\relax,
-  }2007]{Chakravarty2007}
-{C}hakravarty, Manuel M.~T., {L}eshchinskiy, Roman, {P}eyton {J}ones, Simon~L.,
-  {K}eller, Gabriele, \& {M}arlow, Simon. (2007).
-\newblock {Data Parallel Haskell}: a status report.
-\newblock {\em Pages  10--18 of:} {\em DAMP '07}.
-\newblock ACM Press.
-
-\bibitem[\protect\citename{Chakravarty {\em et~al.}\relax,
-  }2011]{Chakravarty:2011:AHA:1926354.1926358}
-Chakravarty, Manuel~M.T., Keller, Gabriele, Lee, Sean, McDonell, Trevor~L., \&
-  Grover, Vinod. (2011).
-\newblock Accelerating {Haskell} array codes with multicore {GPUs}.
-\newblock {\em Pages  3--14 of:} {\em Proceedings of the 6\xth Workshop on
-  Declarative Aspects of Multicore Programming}.
-\newblock DAMP '11.
-\newblock ACM.
-
-\bibitem[\protect\citename{Chase \& Lev, }2005]{Chase:2005:DCW:1073970.1073974}
-Chase, David, \& Lev, Yossi. (2005).
-\newblock Dynamic circular work-stealing deque.
-\newblock {\em Pages  21--28 of:} {\em Proceedings of the 17\xth Annual ACM
-  Symposium on Parallelism in Algorithms and Architectures}.
-\newblock SPAA '05.
-\newblock ACM.
-
-\bibitem[\protect\citename{Clifton-Everest {\em et~al.}\relax, }2014]{CMCK14}
-Clifton-Everest, Robert, McDonell, Trevor~L, Chakravarty, Manuel M~T, \&
-  Keller, Gabriele. (2014).
-\newblock {Embedding Foreign Code}.
-\newblock  {\em PADL '14: The 16th International Symposium on Practical Aspects
-  of Declarative Languages}.
-\newblock LNCS.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{{C}ole, }1989]{Cole1989}
-{C}ole, M.~I. (1989).
-\newblock Algorithmic skeletons: Structured management of parallel computation.
-\newblock  {\em Research Monographs in Parallel and Distributed Computing}.
-\newblock Pitman.
-
-\bibitem[\protect\citename{Czaplicki \& Chong,
-  }2013]{Czaplicki:2013:AFR:2499370.2462161}
-Czaplicki, Evan, \& Chong, Stephen. (2013).
-\newblock Asynchronous functional reactive programming for guis.
-\newblock {\em Sigplan not.}, {\bf 48}(6), 411--422.
-
-\bibitem[\protect\citename{Dagand {\em et~al.}\relax,
-  }2009]{Dagand:2009:ORD:1481861.1481870}
-Dagand, Pierre-\'{E}variste, Kosti\'{c}, Dejan, \& Kuncak, Viktor. (2009).
-\newblock Opis: Reliable distributed systems in {OCaml}.
-\newblock {\em Pages  65--78 of:} {\em Proceedings of the 4\xth International
-  Workshop on Types in Language Design and Implementation}.
-\newblock TLDI '09.
-\newblock ACM.
-
-\bibitem[\protect\citename{Danelutto {\em et~al.}\relax,
-  }1992]{DANELUTTO1992205}
-Danelutto, Marco, Meglio, Roberto~Di, Orlando, Salvatore, Pelagatti, Susanna,
-  \& Vanneschi, Marco. (1992).
-\newblock A methodology for the development and the support of massively
-  parallel programs.
-\newblock {\em Future generation computer systems}, {\bf 8}(1), 205--220.
-
-\bibitem[\protect\citename{Darlington {\em et~al.}\relax,
-  }1993]{darlington1993parallel}
-Darlington, J., Field, AJ, Harrison, PG, Kelly, PHJ, Sharp, DWN, Wu, Q., \&
-  While, RL. (1993).
-\newblock Parallel programming using skeleton functions.
-\newblock {\em Pages  146--160 of:} {\em Parallel architectures and languages
-  Europe}.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Dastgeer {\em et~al.}\relax,
-  }2011]{Dastgeer:2011:ASM:1984693.1984697}
-Dastgeer, Usman, Enmyren, Johan, \& Kessler, Christoph~W. (2011).
-\newblock Auto-tuning {SkePU}: A multi-backend skeleton programming framework
-  for multi-{GPU} systems.
-\newblock {\em Pages  25--32 of:} {\em Proceedings of the 4\xth International
-  Workshop on Multicore Software Engineering}.
-\newblock IWMSE '11.
-\newblock ACM.
-
-\bibitem[\protect\citename{de~la Encina {\em et~al.}\relax,
-  }2011]{delaEncina2011}
-de~la Encina, Alberto, Hidalgo-Herrero, Mercedes, Rabanal, Pablo, \& Rubio,
-  Fernando. (2011).
-\newblock {\em A parallel skeleton for genetic algorithms}.
-\newblock IWANN '11.
-\newblock Springer.
-\newblock Pages  388--395.
-
-\bibitem[\protect\citename{Dean \& Ghemawat,
-  }2008]{Dean:2008:MSD:1327452.1327492}
-Dean, Jeffrey, \& Ghemawat, Sanjay. (2008).
-\newblock {MapReduce}: simplified data processing on large clusters.
-\newblock {\em Communications of the acm}, {\bf 51}, 107--113.
-
-\bibitem[\protect\citename{Dean \& Ghemawat,
-  }2010]{Dean:2010:MFD:1629175.1629198}
-Dean, Jeffrey, \& Ghemawat, Sanjay. (2010).
-\newblock {MapReduce}: a flexible data processing tool.
-\newblock {\em Communications of the acm}, {\bf 53}, 72--77.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax, }2010a]{Dieterle2010}
-Dieterle, M., Horstmeyer, T., \& Loogen, R. (2010a).
-\newblock Skeleton composition using remote data.
-\newblock {\em Pages  73--87 of:} Carro, M., \& Pe\~{n}a, R. (eds), {\em 12\xth
-  International Symposium on Practical Aspects of Declarative Languages}.
-\newblock PADL '10, vol. 5937.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax,
-  }2016]{dieterle_horstmeyer_loogen_berthold_2016}
-Dieterle, M., Horstmeyer, T., Loogen, R., \& Berthold, J. (2016).
-\newblock Skeleton composition versus stable process systems in {Eden}.
-\newblock {\em Journal of functional programming}, {\bf 26}.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax,
-  }2010b]{dieterle2010skeleton}
-Dieterle, Mischa, Berthold, Jost, \& Loogen, Rita. (2010b).
-\newblock {\em A skeleton for distributed work pools in {Eden}}.
-\newblock FLOPS '10.
-\newblock Springer.
-\newblock Pages  337--353.
-
-\bibitem[\protect\citename{Dieterle {\em et~al.}\relax, }2013]{Dieterle2013}
-Dieterle, Mischa, Horstmeyer, Thomas, Berthold, Jost, \& Loogen, Rita. (2013).
-\newblock {\em Iterating skeletons}.
-\newblock IFL '12.
-\newblock Springer.
-\newblock Pages  18--36.
-
-\bibitem[\protect\citename{Dinan {\em et~al.}\relax,
-  }2009]{Dinan:2009:SWS:1654059.1654113}
-Dinan, James, Larkins, D.~Brian, Sadayappan, P., Krishnamoorthy, Sriram, \&
-  Nieplocha, Jarek. (2009).
-\newblock Scalable work stealing.
-\newblock {\em Pages  53:1--53:11 of:} {\em Proceedings of the Conference on
-  High Performance Computing Networking, Storage and Analysis}.
-\newblock SC '09.
-\newblock ACM.
-
-\bibitem[\protect\citename{Foltzer {\em et~al.}\relax,
-  }2012]{Foltzer:2012:MPC:2398856.2364562}
-Foltzer, Adam, Kulkarni, Abhishek, Swords, Rebecca, Sasidharan, Sajith, Jiang,
-  Eric, \& Newton, Ryan. (2012).
-\newblock A meta-scheduler for the {Par}-monad: Composable scheduling for the
-  heterogeneous cloud.
-\newblock {\em Sigplan not.}, {\bf 47}(9), 235--246.
-
-\bibitem[\protect\citename{Geimer {\em et~al.}\relax, }2010]{Geimer2010}
-Geimer, M., Wolf, F., Wylie, B. J.~N., {\'A}brah{\'a}m, E., Becker, D., \&
-  Mohr, B. (2010).
-\newblock The {Scalasca} performance toolset architecture.
-\newblock {\em Concurrency and computation: Practice and experience}, {\bf
-  22}(6).
-
-\bibitem[\protect\citename{Gentleman, }1978]{Gentleman1978}
-Gentleman, W.~Morven. (1978).
-\newblock Some complexity results for matrix computations on parallel
-  processors.
-\newblock {\em Journal of the acm}, {\bf 25}(1), 112--115.
-
-\bibitem[\protect\citename{Gorlatch, }1998]{Gorlatch1998}
-Gorlatch, Sergei. (1998).
-\newblock Programming with divide-and-conquer skeletons: A case study of {FFT}.
-\newblock {\em Journal of supercomputing}, {\bf 12}(1-2), 85--97.
-
-\bibitem[\protect\citename{Gorlatch \& Bischof, }1998]{Gorlatch}
-Gorlatch, Sergei, \& Bischof, Holger. (1998).
-\newblock A generic {MPI} implementation for a data-parallel skeleton: Formal
-  derivation and application to {FFT}.
-\newblock {\em Parallel processing letters}, {\bf 8}(4).
-
-\bibitem[\protect\citename{Hammond {\em et~al.}\relax,
-  }2003]{doi:10.1142/S0129626403001380}
-Hammond, Kevin, Berthold, Jost, \& Loogen, Rita. (2003).
-\newblock Automatic skeletons in {Template} {Haskell}.
-\newblock {\em Parallel processing letters}, {\bf 13}(03), 413--424.
-
-\bibitem[\protect\citename{Harris {\em et~al.}\relax,
-  }2007]{harris2007parallel}
-Harris, Mark, Sengupta, Shubhabrata, \& Owens, John~D. (2007).
-\newblock Parallel prefix sum (scan) with {CUDA}.
-\newblock {\em {GPU} gems}, {\bf 3}(39), 851--876.
-
-\bibitem[\protect\citename{Harris {\em et~al.}\relax,
-  }2005]{Harris:2005:CMT:1065944.1065952}
-Harris, Tim, Marlow, Simon, {Peyton Jones}, Simon, \& Herlihy, Maurice. (2005).
-\newblock Composable memory transactions.
-\newblock {\em Pages  48--60 of:} {\em Proceedings of the 10\xth ACM SIGPLAN
-  Symposium on Principles and Practice of Parallel Programming}.
-\newblock PPoPP '05.
-\newblock ACM.
-
-\bibitem[\protect\citename{Hey, }1990]{Hey1990185}
-Hey, Anthony J.~G. (1990).
-\newblock Experiments in {MIMD} parallelism.
-\newblock {\em Future generation computer systems}, {\bf 6}(3), 185--196.
-
-\bibitem[\protect\citename{Hippold \& R{\"u}nger, }2006]{Hippold2006}
-Hippold, J., \& R{\"u}nger, G. (2006).
-\newblock Task pool teams: A hybrid programming environment for irregular
-  algorithms on {SMP} clusters.
-\newblock {\em Concurrency and computation: Practice and experience}, {\bf 18},
-  1575--1594.
-
-\bibitem[\protect\citename{Horstmeyer \& Loogen, }2013]{Horstmeyer2013}
-Horstmeyer, Thomas, \& Loogen, Rita. (2013).
-\newblock Graph-based communication in {Eden}.
-\newblock {\em Higher-order and symbolic computation}, {\bf 26}(1), 3--28.
-
-\bibitem[\protect\citename{Huang {\em et~al.}\relax, }2007]{Huang2007}
-Huang, Liwen, Hudak, Paul, \& Peterson, John. (2007).
-\newblock {\em {HPorter}: Using arrows to compose parallel processes}.
-\newblock Berlin, Heidelberg: Springer Berlin Heidelberg.
-\newblock Pages  275--289.
-
-\bibitem[\protect\citename{Hudak {\em et~al.}\relax, }2003]{Hudak2003}
-Hudak, Paul, Courtney, Antony, Nilsson, Henrik, \& Peterson, John. (2003).
-\newblock {\em Arrows, robots, and functional reactive programming}.
-\newblock Springer.
-\newblock Pages  159--187.
-
-\bibitem[\protect\citename{Hughes, }2000]{HughesArrows}
-Hughes, John. (2000).
-\newblock Generalising monads to arrows.
-\newblock {\em Science of computer programming}, {\bf 37}(1–3), 67--111.
-
-\bibitem[\protect\citename{Hughes, }2005]{Hughes2005}
-Hughes, John. (2005).
-\newblock {\em Programming with arrows}.
-\newblock AFP '04.
-\newblock Springer.
-\newblock Pages  73--129.
-
-\bibitem[\protect\citename{Jacobs {\em et~al.}\relax,
-  }2009]{jacobs_heunen_hasuo_2009}
-Jacobs, Bart, Heunen, Chris, \& Hasuo, Ichiro. (2009).
-\newblock Categorical semantics for arrows.
-\newblock {\em Journal of functional programming}, {\bf 19}(3-4), 403--438.
-
-\bibitem[\protect\citename{Janjic {\em et~al.}\relax, }2013]{janjic2013space}
-Janjic, Vladimir, Brown, Christopher~Mark, Neunhoeffer, Max, Hammond, Kevin,
-  Linton, Stephen~Alexander, \& Loidl, Hans-Wolfgang. (2013).
-\newblock Space exploration using parallel orbits: a study in parallel symbolic
-  computing.
-\newblock {\em Parallel computing}.
-
-\bibitem[\protect\citename{Karasawa \& Iwasaki, }2009]{5361825}
-Karasawa, Y., \& Iwasaki, H. (2009).
-\newblock A parallel skeleton library for multi-core clusters.
-\newblock {\em Pages  84--91 of:} {\em International Conference on Parallel
-  Processing 2009}.
-
-\bibitem[\protect\citename{Keller {\em et~al.}\relax,
-  }2010]{Keller:2010:RSP:1932681.1863582}
-Keller, Gabriele, Chakravarty, Manuel~M.T., Leshchinskiy, Roman, {Peyton
-  Jones}, Simon, \& Lippmeier, Ben. (2010).
-\newblock Regular, shape-polymorphic, parallel arrays in haskell.
-\newblock {\em Sigplan not.}, {\bf 45}(9), 261--272.
-
-\bibitem[\protect\citename{Kuchen, }2002]{Kuchen2002}
-Kuchen, Herbert. (2002).
-\newblock A skeleton library.
-\newblock {\em Pages  620--629 of:} Monien, Burkhard, \& Feldmann, Rainer
-  (eds), {\em Parallel Processing}.
-\newblock Euro-Par '02.
-\newblock Springer.
-
-\bibitem[\protect\citename{Kuper {\em et~al.}\relax,
-  }2014]{Kuper:2014:TPE:2666356.2594312}
-Kuper, Lindsey, Todd, Aaron, Tobin-Hochstadt, Sam, \& Newton, Ryan~R. (2014).
-\newblock Taming the parallel effect zoo: Extensible deterministic parallelism
-  with {LVish}.
-\newblock {\em Sigplan not.}, {\bf 49}(6), 2--14.
-
-\bibitem[\protect\citename{Lengauer {\em et~al.}\relax, }1997]{Lengauer1997}
-Lengauer, Christian, Gorlatch, Sergei, \& Herrmann, Christoph. (1997).
-\newblock The static parallelization of loops and recursions.
-\newblock {\em The journal of supercomputing}, {\bf 11}(4), 333--353.
-
-\bibitem[\protect\citename{Li \& Zdancewic, }2006]{1648705}
-Li, Peng, \& Zdancewic, S. (2006).
-\newblock Encoding information flow in {Haskell}.
-\newblock {\em Pages  12--16 of:} {\em 19\xth IEEE Computer Security
-  Foundations Workshop}.
-\newblock CSFW '06.
-
-\bibitem[\protect\citename{Li \& Zdancewic, }2010]{LI20101974}
-Li, Peng, \& Zdancewic, Steve. (2010).
-\newblock Arrows for secure information flow.
-\newblock {\em Theoretical computer science}, {\bf 411}(19), 1974--1994.
-
-\bibitem[\protect\citename{Lindley {\em et~al.}\relax, }2011]{LINDLEY201197}
-Lindley, Sam, Wadler, Philip, \& Yallop, Jeremy. (2011).
-\newblock Idioms are oblivious, arrows are meticulous, monads are promiscuous.
-\newblock {\em Electronic notes in theoretical computer science}, {\bf 229}(5),
-  97--117.
-
-\bibitem[\protect\citename{Linton {\em et~al.}\relax, }2010]{scscp}
-Linton, S., Hammond, K., Konovalov, A., Al~Zain, A.~D., Trinder, P., Horn, P.,
-  \& Roozemond, D. (2010).
-\newblock Easy composition of symbolic computation software: a new lingua
-  franca for symbolic computation.
-\newblock {\em Pages  339--346 of:} {\em Proceedings of the 2010 International
-  Symposium on Symbolic and Algebraic Computation}.
-\newblock ISSAC '10.
-\newblock ACM Press.
-
-\bibitem[\protect\citename{Liu {\em et~al.}\relax,
-  }2009]{Liu:2009:CCA:1631687.1596559}
-Liu, Hai, Cheng, Eric, \& Hudak, Paul. (2009).
-\newblock Causal commutative arrows and their optimization.
-\newblock {\em Sigplan not.}, {\bf 44}(9), 35--46.
-
-\bibitem[\protect\citename{Lobachev, }2011]{lobachev-phd}
-Lobachev, Oleg. (2011).
-\newblock {\em Implementation and evaluation of algorithmic skeletons:
-  Parallelisation of computer algebra algorithms}.
-\newblock Ph.D. thesis, Philipps-Universit{\"a}t Marburg.
-
-\bibitem[\protect\citename{Lobachev, }2012]{Lobachev2012}
-Lobachev, Oleg. (2012).
-\newblock {\em Parallel computation skeletons with premature termination
-  property}.
-\newblock FLOPS 2012.
-\newblock Springer.
-\newblock Pages  197--212.
-
-\bibitem[\protect\citename{Loogen {\em et~al.}\relax,
-  }2003]{Eden:SkeletonBookChapter02}
-Loogen, R., {Ortega-Mall\'{e}n}, Y., {Pe\~{n}a}, R., Priebe, S., \& Rubio, F.
-  (2003).
-\newblock {Parallelism Abstractions in Eden}.
-\newblock {\em Pages  71--88 of:} Rabhi, F.~A., \& Gorlatch, S. (eds), {\em
-  {Patterns and Skeletons for Parallel and Distributed Computing}}.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Loogen, }2012]{Loogen2012}
-Loogen, Rita. (2012).
-\newblock {\em Eden -- parallel functional programming with {Haskell}}.
-\newblock Springer.
-\newblock Pages  142--206.
-
-\bibitem[\protect\citename{Loogen {\em et~al.}\relax, }2005]{eden}
-Loogen, Rita, Ortega-Mall\'{e}n, Yolanda, \& {Pe\~{n}a-Mar\'{\i}}, Ricardo.
-  (2005).
-\newblock {Parallel Functional Programming in Eden}.
-\newblock {\em {Journal of Functional Programming}}, {\bf 15}(3), 431--475.
-\newblock Special Issue on Functional Approaches to High-Performance Parallel
-  Programming.
-
-\bibitem[\protect\citename{Lämmel, }2008]{LAMMEL20081}
-Lämmel, Ralf. (2008).
-\newblock Google’s mapreduce programming model — revisited.
-\newblock {\em Science of computer programming}, {\bf 70}(1), 1--30.
-
-\bibitem[\protect\citename{Maier {\em et~al.}\relax,
-  }2014]{Maier:2014:HDS:2775050.2633363}
-Maier, Patrick, Stewart, Robert, \& Trinder, Phil. (2014).
-\newblock The {HdpH} {DSLs} for scalable reliable computation.
-\newblock {\em Sigplan not.}, {\bf 49}(12), 65--76.
-
-\bibitem[\protect\citename{Mainland \& Morrisett,
-  }2010]{Mainland:2010:NEC:2088456.1863533}
-Mainland, Geoffrey, \& Morrisett, Greg. (2010).
-\newblock Nikola: Embedding compiled {GPU} functions in {Haskell}.
-\newblock {\em Sigplan not.}, {\bf 45}(11), 67--78.
-
-\bibitem[\protect\citename{Marlow {\em et~al.}\relax, }2009]{Marlow2009}
-Marlow, S., {Peyton Jones}, S., \& Singh, S. (2009).
-\newblock Runtime support for multicore {Haskell}.
-\newblock {\em Acm sigplan notices}, {\bf 44}(9), 65--78.
-
-\bibitem[\protect\citename{Marlow, }2013]{marlow2013parallel}
-Marlow, Simon. (2013).
-\newblock {\em Parallel and concurrent programming in {Haskell}: Techniques for
-  multicore and multithreaded programming}.
-\newblock "O'Reilly Media, Inc.".
-
-\bibitem[\protect\citename{Marlow {\em et~al.}\relax, }2011]{par-monad}
-Marlow, Simon, Newton, Ryan, \& {Peyton Jones}, Simon. (2011).
-\newblock A monad for deterministic parallelism.
-\newblock {\em Sigplan not.}, {\bf 46}(12), 71--82.
-
-\bibitem[\protect\citename{McDonell {\em et~al.}\relax,
-  }2015]{McDonell:2015:TRC:2887747.2804313}
-McDonell, Trevor~L., Chakravarty, Manuel M.~T., Grover, Vinod, \& Newton,
-  Ryan~R. (2015).
-\newblock Type-safe runtime code generation: Accelerate to {LLVM}.
-\newblock {\em Sigplan not.}, {\bf 50}(12), 201--212.
-
-\bibitem[\protect\citename{Michael {\em et~al.}\relax,
-  }2009]{Michael:2009:IWS:1594835.1504186}
-Michael, Maged~M., Vechev, Martin~T., \& Saraswat, Vijay~A. (2009).
-\newblock Idempotent work stealing.
-\newblock {\em Sigplan not.}, {\bf 44}(4), 45--54.
-
-\bibitem[\protect\citename{Nilsson {\em et~al.}\relax,
-  }2002]{Nilsson:2002:FRP:581690.581695}
-Nilsson, Henrik, Courtney, Antony, \& Peterson, John. (2002).
-\newblock Functional reactive programming, continued.
-\newblock {\em Pages  51--64 of:} {\em Proceedings of the 2002 {ACM} {SIGPLAN}
-  Workshop on {Haskell}}.
-\newblock Haskell '02.
-\newblock New York, NY, USA: ACM.
-
-\bibitem[\protect\citename{Olivier \& Prins, }2008]{4625841}
-Olivier, S., \& Prins, J. (2008).
-\newblock Scalable dynamic load balancing using {UPC}.
-\newblock {\em Pages  123--131 of:} {\em 37\xth International Conference on
-  Parallel Processing}.
-
-\bibitem[\protect\citename{Paterson, }2001]{Paterson:2001:NNA:507669.507664}
-Paterson, Ross. (2001).
-\newblock A new notation for arrows.
-\newblock {\em Sigplan not.}, {\bf 36}(10), 229--240.
-
-\bibitem[\protect\citename{Pe{\~n}a \& Rubio, }2001]{Eden:PPDP01}
-Pe{\~n}a, R., \& Rubio, F. (2001).
-\newblock {Parallel Functional Programming at Two Levels of Abstraction}.
-\newblock {\em Pages  187--198 of:} {\em PPDP'01 --- Intl.\ Conf.\ on
-  Principles and Practice of Declarative Programming}.
-
-\bibitem[\protect\citename{Perfumo {\em et~al.}\relax,
-  }2008]{Perfumo:2008:LST:1366230.1366241}
-Perfumo, Cristian, S\"{o}nmez, Nehir, Stipic, Srdjan, Unsal, Osman, Cristal,
-  Adri\'{a}n, Harris, Tim, \& Valero, Mateo. (2008).
-\newblock The limits of software transactional memory ({STM}): Dissecting
-  {Haskell} {STM} applications on a many-core environment.
-\newblock {\em Pages  67--78 of:} {\em Proceedings of the 5\xth Conference on
-  Computing Frontiers}.
-\newblock CF '08.
-\newblock ACM.
-
-\bibitem[\protect\citename{Poldner \& Kuchen, }2005]{Kuchen05}
-Poldner, Michael, \& Kuchen, Herbert. (2005).
-\newblock Scalable farms.
-\newblock {\em Pages  795--802 of:} Joubert, Gerhard~R., Nagel, Wolfgang~E.,
-  Peters, Frans~J., Plata, Oscar~G., Tirado, P., \& Zapata, Emilio~L. (eds),
-  {\em PARCO}.
-\newblock John von Neumann Institute for Computing Series, vol. 33.
-\newblock Central Institute for Applied Mathematics, J{\"u}lich, Germany.
-
-\bibitem[\protect\citename{Priebe, }2006]{WPEuropar06}
-Priebe, Steffen. (2006).
-\newblock Dynamic task generation and transformation within a nestable workpool
-  skeleton.
-\newblock  {\em Euro-Par}.
-\newblock LNCS 4128.
-
-\bibitem[\protect\citename{Rabhi \& Gorlatch, }2003]{SkeletonBook}
-Rabhi, F.~A., \& Gorlatch, S. (eds). (2003).
-\newblock {\em {Patterns and Skeletons for Parallel and Distributed
-  Computing}}.
-\newblock Springer-Verlag.
-
-\bibitem[\protect\citename{Rudolph {\em et~al.}\relax,
-  }1991]{Rudolph:1991:SLB:113379.113401}
-Rudolph, Larry, Slivkin-Allalouf, Miriam, \& Upfal, Eli. (1991).
-\newblock A simple load balancing scheme for task allocation in parallel
-  machines.
-\newblock {\em Pages  237--245 of:} {\em Proceedings of the 3\xrd Annual ACM
-  Symposium on Parallel Algorithms and Architectures}.
-\newblock SPAA '91.
-\newblock ACM.
-
-\bibitem[\protect\citename{Russo {\em et~al.}\relax,
-  }2008]{Russo:2008:LLI:1411286.1411289}
-Russo, Alejandro, Claessen, Koen, \& Hughes, John. (2008).
-\newblock A library for light-weight information-flow security in {Haskell}.
-\newblock {\em Pages  13--24 of:} {\em Proceedings of the 1\st ACM SIGPLAN
-  Symposium on Haskell}.
-\newblock Haskell '08.
-\newblock ACM.
-
-\bibitem[\protect\citename{Stewart {\em et~al.}\relax,
-  }2016]{stewart_maier_trinder_2016}
-Stewart, Robert, Maier, Patrick, \& Trinder, Phil. (2016).
-\newblock Transparent fault tolerance for scalable functional computation.
-\newblock {\em Journal of functional programming}, {\bf 26}.
-
-\bibitem[\protect\citename{Svensson, }2011]{obsidian-phd}
-Svensson, Joel. (2011).
-\newblock {\em Obsidian: Gpu kernel programming in haskell}.
-\newblock Ph.D. thesis, Chalmers University of Technology.
-
-\bibitem[\protect\citename{Trinder {\em et~al.}\relax, }1996]{Trinder1996}
-Trinder, Phil~W., Hammond, Kevin, {Mattson Jr.}, James~S., Partridge,
-  Andrew~S., \& {Peyton Jones}, Simon~L. (1996).
-\newblock {GUM: a Portable Parallel Implementation of Haskell}.
-\newblock  {\em {PLDI'96}}.
-\newblock ACM Press.
-
-\bibitem[\protect\citename{Trinder {\em et~al.}\relax, }1998]{Trinder1998a}
-Trinder, P.W., Hammond, K., Loidl, H-W., \& {Peyton Jones}, S. (1998).
-\newblock {Algorithm + Strategy = Parallelism}.
-\newblock {\em Journal of functional programming}, {\bf 8}(1), 23--60.
-
-\bibitem[\protect\citename{van Nieuwpoort {\em et~al.}\relax,
-  }2001]{vanNieuwpoort:2001:ELB:568014.379563}
-van Nieuwpoort, Rob~V., Kielmann, Thilo, \& Bal, Henri~E. (2001).
-\newblock Efficient load balancing for wide-area divide-and-conquer
-  applications.
-\newblock {\em Sigplan not.}, {\bf 36}(7), 34--43.
-
-\bibitem[\protect\citename{Vizzotto {\em et~al.}\relax,
-  }2006]{vizzotto_altenkirch_sabry_2006}
-Vizzotto, Juliana, Altenkirch, Thorsten, \& Sabry, Amr. (2006).
-\newblock Structuring quantum effects: superoperators as arrows.
-\newblock {\em Mathematical structures in computer science}, {\bf 16}(3),
-  453--468.
-
-\bibitem[\protect\citename{Wheeler \& Thain, }2009]{Wheeler2009}
-Wheeler, K.~B., \& Thain, D. (2009).
-\newblock Visualizing massively multithreaded applications with {ThreadScope}.
-\newblock {\em Concurrency and computation: Practice and experience}, {\bf
-  22}(1), 45--67.
-
-\end{thebibliography}
+        \bibliographystyle{jfp}
+	\bibliography{references,main}
         \appendix

### 2019-12-27 paper/content/response.md
reviewer's comments (2)

@@
 TODO
+
+Our formalism inspired S. Zhang's Session Arrows (Zhang, 2019).
 
@@
 
-We provided an additional reference explaining Profunctors in the main text, (Boisseau & Gibbons).
-It is possible to define Arrows deriving from Profunctors, as Appendix B shows.
+We rephrased the offending paragraph, providing an additional reference explaining Profunctors (Boisseau & Gibbons, 2018).
+The text emphases now that it is possible to define Arrows deriving from Profunctors, as Appendix B shows.
 There we also cite Asada (2010).
@@
 > phrasing. I suggest to cut this short.
+
+The text was removed.
+
 > 
@@
 >
+
+We moved the text ob LVish and HdpH to the related work.
+
 > page 10:
@@
 > "efficiency OR the lack thereof" maybe?
+
+Fixed.
+
 > 
@@
 > The dashes should be longer, or in Latex terms: "---".
+
+Fixed globally.
+
 > 
@@
 > "instances of the ArrowParallel instances" ?
+
+This was obviously a writing mishap, fixed.
+
 > 
@@
 > programming approach is somewhat implementation-agnostic.
+
+We clarified the wording and emphasised portability across parallel Haskells.
+
 > 
@@
 >
+
+TODO
+
 > page 16:
@@
 > "toplogical"
+
+Fixed.
+
 > 
@@
 >
+
+We changed the wording.
+
+Our reasoning is that the behaviour in question arises from the subtle details of laziness and demand in Eden standard library.
+We argue that such details are not of major interest to the generic JFP audience.
+Our paper focuses on establishing a common interface between different parallel Haskells, with a minor workaround the above demand issue becomes irrelevant.
+
 > page 22:
@@
 > "repeating": why is the front part of the word set in italics?
+
+We were referencing the `repeat` function. We now use an apostrophe for more clarity.
+
 > 
@@
 > "using the Gentleman algorithm" -> "using Gentleman's algorithm"
+
+Fixed.
+
 > 
@@
 > implementation are used.
+
+We forced both figures on a single page.
+
+The point of those figures is exactly to showcase that there is no observable difference, despite using a further intermediate abstraction layer in form of our API.
+
 > 
@@
 > The text in 7.1.4 and the beginning of 7.2 should be condensed.
+
+We implemented this change request.
+
 > 
@@
 > interesting numbers and discuss the findings.
+
+TODO
+
 > 
 > The term "speedup" is used extensively without ever explaining the base line.
+
+We added a definition of speedup.
+
 > 
@@
 > "extensible formalism THAT can be easily ported"
+
+Fixed.
+
 > 
 > The sentence "We argue that ..." is pretty much repeated right thereafter.
+
+We removed this sentence all together.
+
 > 
@@
 > there is an unmotivated but consistent space between the acronym and the year.
+
+The space is removed.
+
 > 
@@
 >
+
+We would expect some at least technical difficulties with Accelerate.
+However, our interface seems to be fairly universal.
+A very different approach is data parallel programming.
+Our approach is task parallel.
+
 > Section 3:
@@
 > PArrows?
+
+Our approach is directly reversed: we define PArrows in terms of par/seq. However, TODO TODO
+
 > 
@@
 > "Future" part of PArrows?
+
+It is not part of the class holding parEvalN, but it extends the general interface of PArrows.
+
 > 
@@
 > is a major problem.
+
+TODO
+
 > 
@@
 > the (->) and Kleisli instances if they don't work with Eden?
+
+It would be possible to creature further special instances for Eden, if those are of interest.
+
 > 
@@
 >  be shown to actually make programs go faster.
+
+The `pipe` and `ring` skeletons are needed, as they allow for a more explainable and didactic introduction of the `torus`.
+Basically, `ring` is a warped `pipe`, and `torus` is a `ring` of `ring`s.
+Hence, the `torus` benchmark is the ultimate test for the whole family of `pipe`, `ring`, and `torus`.
+
 > 
@@
 > Para 3: line 6: "very" is vague.
+
+We fixed the above spelling problems, thank you very much.
+
 > 
@@
 >
+
+Indeed, we fixed this mistake and apologize.
+
 > Para 6: line 1: this is the first mention of "PArrow" -- it should be
 > defined earlier.
+
+We fixed this.
+
 > 
@@
 > others beyond being able to switch backend easily?
+
+TODO
+
 > 
 > Para 6: line 2: "parallelisations" => "execution strategies"
-> 
+
+Thank you for this suggestion. We accepted it, the final wording is: "... relatively easy to define alternate execution strategies, including parallel evaluation."
+
 > Para 6: line 7: "is completely orthogonal to our approach". This seems
@@
 > it is orthogonal?
+
+Accelerate focuses on accessing the GPU and adapts to the notions of Cuda language.
+We found it hard to combine the concepts of PArrows and Accelerate under one hood.
+However, we are considering to use PArrows to orchestrate multiple Accelerate instances.
+
+We reason that such an in-depth explanation is going to far in the paper text, hence we omit it.
+
 > 
 > Para 8: line 2: Should be a colon after "following"?
+
+Fixed.
+
 > 
@@
 > in relation to LVars?
+
+TODO
+
 > 
@@
 >
+
+TODO
+
 > "Arrow": line 10: "capsule the computation", do you mean "encapsulate
@@
 >
+
+We adjusted the description of Lue et al.'s work.
+
+The difference between our work and Lue et al.
+is that they would allow for composition of their arrows in parallel.
+We provide operators to execute a set of arrows in parallel (yielding an arrow again).
+
+WHAT'S THE DIFFERENCE
+
 > "Arrow": line 14: missing "a" before "list".

### 2019-12-27 paper/content/ms.tex
reviewer's comments (2)

@@
 We regard here in-depth
-Glasgow parallel Haskell or short GpH (its Multicore SMP implementation, in particular), the
+Glasgow parallel Haskell or for short GpH (its Multicore SMP implementation, in particular), the
 \ensuremath{\Conid{Par}} Monad, and Eden, a distributed memory parallel Haskell. These
@@
 
-A key novelty in this paper is to use Arrows to represent parallel computations.
+The key novelty in this paper is to use Arrows to represent parallel computations.
 They seem a natural fit as they can be thought of as a more general function arrow (\ensuremath{\to }) and serve as general interface to computations while not being as restrictive as Monads \citep{HughesArrows}.
@@
 
-We provide an Arrows-based type class and implementations for the three above mentioned parallel Haskells.
+We provide an Arrow-based type class and implementations for the three above mentioned parallel Haskells.
 Instead of 
 introducing a new low-level parallel backend to implement our
-Arrows-based interface, we define a shallow-embedded DSL for Arrows. This DSL
+Arrow-based interface, we define a shallow-embedded DSL using Arrows. This DSL
 is defined as a common interface with varying implementations in
@@
 novel manner---we tame the zoo of parallel Haskells. We provide a
-common, very low-penalty programming interface that allows to switch
+common, low-penalty programming interface that allows to switch
 the parallel implementations at will.
@@
 %\olcomment{move this to Contributions in the front or something}
-We propose an Arrow-based encoding for parallelism based on a new Arrow combinator \ensuremath{\Varid{parEvalN}\mathbin{::}[\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]}.
-A parallel Arrow is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.
+We propose an Arrow-based encoding for parallelism based on a new Arrow combinator \ensuremath{\Varid{parEvalN}\mathbin{::}(\Conid{Arrow}\;\Varid{arr})\Rightarrow [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]{}}.
+A parallel Arrow (short: PArrow) is still an Arrow, hence the resulting parallel Arrow can still be used in the same way as a potential sequential version.
 In this paper we evaluate the expressive power of such a formalism in the context of parallel programming.
@@
 \paragraph{Parallel Haskells.}
-The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate parallelisations.
+The non-strict semantics of Haskell, and the fact that reduction encapsulates computations as closures, makes it relatively easy to define alternate execution strategies, including parallel evaluation.
 A range of approaches have been explored, including data parallelism \cite{Chakravarty2007,Keller:2010:RSP:1932681.1863582}, GPU-based approaches \cite{Mainland:2010:NEC:2088456.1863533,obsidian-phd}, software transactional memory \cite{Harris:2005:CMT:1065944.1065952,Perfumo:2008:LST:1366230.1366241}.
@@
 
-Other task parallel Haskells related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad include the following.  
+Other task parallel Haskells related to Eden, GpH, and the \ensuremath{\Conid{Par}} Monad include the following: 
 HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of \ensuremath{\Conid{Par}} Monad to
@@
 But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
-\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular Arrows do and thus enable optimisations.
+\citet{Liu:2009:CCA:1631687.1596559} formally define a more special
+% kind of Arrows that encapsulate the computation more than regular
+% Arrows do and thus enable optimisations.
+implementation of Arrows that allows for better optimisations.
 Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution.
@@
 
-PArrows is an extendable formalism, they can be easily ported to further parallel Haskells while still maintaining interchangeability.
+PArrows is an extendable formalism that can be easily ported to further parallel Haskells while still maintaining interchangeability.
 It is straightforward to provide further implementations of algorithmic skeletons in PArrows.
@@
 Furthermore, we think
-We argue that Arrows are better suited to  parallelise
-pure functions than Monads,  as the functions are already Arrows and can be used
-directly in our DSL.
+% We argue that Arrows are better suited to  parallelise
+% pure functions than Monads,  as the functions are already Arrows and can be used
+% directly in our DSL.
 Arrows are a better fit to parallelise pure code than a monadic solution as regular functions are already Arrows and can be used with our DSL in a more natural way.

### 2019-12-28 paper/content/response.md
affiliations

@@
 >
+
+XXXXXXXXXXXXXXXXX
+
 > Another potential piece of related work is Turon's Reagents, that are
@@
 >
+
+Thank you for this reference, we include it now.
+
 > ### Page 5

### 2019-12-28 paper/content/arxiv/ms.tex
affiliations

@@
         {\textls*{MARTIN BRAUN}\\
-         University Bayreuth, 95440 Bayreuth, Germany\\
+          University of Bayreuth, 95440 Bayreuth, Germany 
+          and\\
+          NeuroForge GmbH \& Co.\ KG,
+          Nibelungenstr.~7,
+          95444 Bayreuth,
+          Germany\\
 		 \textls*{OLEG LOBACHEV}\\
-		 University Bayreuth, 95440 Bayreuth, Germany\\
+		 University of Bayreuth, 95440 Bayreuth, Germany 
+                 and\\
+                 Hannover Medical School, OE~4120, Carl-Neuberg-Straße~1, 30625 Hannover, Germany\\
 		 \and\ \textls*{PHIL TRINDER}\\
@@
 \citet{Dagand:2009:ORD:1481861.1481870} used Arrows in OCaml in the implementation of a distributed system.
+
+\citet{turon_reagents:_2012} has somewhat similar goals to our work. They deriving a formalism for concurrency, including commits, choice, channels, with composable components.
 
@@
 
-While we could have based our DSL on Profunctors \cite{boisseau_what_2018} as well, we chose Arrows\mbcomment{cite missing?} for this paper since they allow for a more direct way of thinking about parallelism than general Profunctors because of their composability.
-However, they are a promising candidate for future improvements of our DSL.
-Some Profunctors, especially ones supporting a composition operation, choice, and looping, can already be adapted to our interface as shown in Appendix~\ref{app:profunctorArrows}.
+While our DSL could have been based on Profunctors \citep[see, e.g.][]{boisseau_what_2018} directly, we chose Arrows for this paper for the easier representation of composability.
+Special Profunctors supporting a composition operation, choice, and looping, can already be used as a foundation of our interface as shown in Appendix~\ref{app:profunctorArrows}.
 	\subsection{Short introduction to parallel Haskells}

### 2019-12-28 paper/content/response.md
reviewer's comments (3)

@@
 > for the JVM"
+
+We adapted your wording.
+
 > 
 > Para 2; line 2: "typical GUI tasks" => you could be less vague.
+
+Fixed.
+
 > 
@@
 > "Monad m =>" instead of saying "where m is a Monad"?
-> 
+
+Fixed.
+
 > Sec 3.1, para 2, first line: What does the "This" at the end of the
@@
 >
+
+XXXXXXXXXXXXXXXXX
+
 > Sec 3.1, para 3: "syntactic sugar". I don't think these derived
@@
 > combinators", or more prosaically, "utility functions".
+
+We changed the way we reference those combinators throughout the text.
 > 
@@
 > background knowledge about Arrows do you expect the reader to have?
-> 
+
+XXXXXXXXXXXXXXXXXXX
+
 > "we have a truly more general interface as compared to a monadic one"
@@
 > missed one?
+
+TODO: monadic RabinMiller!
+
 > 
@@
 > a short introduction to them.
+
+Fixed, as mentioned above.
+
 > 
@@
 >
+
+We have severely shortened and rewritten this paragraph alltogether, now there is no mention of importance at this point.
+
 > ### Page 9
@@
 >    parEvalN fs as = runPar $ mapM (return >>> spawn >>= get) $ zipWith ($) fs as
+
+Thank you very much, we replace the definition.
+
 > 
@@
 > You don't explain what the `Trans` typeclass is. `Trans`portable?
+
+Yes, specialization of packing and unpacking and network transmission, e.g. streams for lists. We added two sentences explaining this issue.
+
 > 
@@
 >
+
+XXXXXXXXXXXXXXXXXXX
+
 > ### Page 14
@@
 > Figure 11: what is the `ChunkSize` type? `Int`?
+
+Yes. We added an explanation.
+
 > 
@@
 > problem.
+
+XXXXXXXXXXXXXX
+
 > 
@@
 > Fig 17, caption: "Other than" => "Unlike"
+
+Fixed.
+
 > 
@@
 > second last line: remove the final "the"
+
+Fixed.
+
 > 
@@
 > first line: remove "when"
+
+Fixed.
+
 > 
@@
 > paragraph 3: "probably": do some benchmarks?
+
+This was more of an speculation on what is the better in the API than a solid performance claim. We adjusted the wording correspondingly.
+
 > 
 > last line: "send" => "sent"
+
+Fixed.
+
 > 
@@
 >
+
+XXXXXXXXXXXXXXXXXX
+
 > ### Page 37
@@
 > last line: "as" => "was"
+
+Fixed.
+
 > 
@@
 > argument).
+
+TODO TODO TODO

### 2019-12-28 paper/content/arxiv/ms.tex
reviewer's comments (3)

@@
 Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined.
-Preliminary experiments show that basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM).
+Preliminary experiments show that basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (a Haskell-like language for the JVM).
 However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}},
@@
 
-\citet{achten2004arrows,achten2007arrow} use an Arrow implementation in Clean for better handling of typical GUI tasks.
+\citet{achten2004arrows,achten2007arrow} use an Arrow implementation in Clean for graphical user interfaces.
 \citet{Dagand:2009:ORD:1481861.1481870} used Arrows in OCaml in the implementation of a distributed system.
@@
 Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads.
-\citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator \ensuremath{(\bind )\mathbin{::}\Varid{m}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{b})\to \Varid{m}\;\Varid{b}} (with \ensuremath{\Varid{m}} being a Monad)\footnote{In the example a parser of the type \ensuremath{\Conid{Parser}\;\Varid{s}\;\Varid{a}} with static meta information \ensuremath{\Varid{s}} and result \ensuremath{\Varid{a}} is shown to not be able to use the static information \ensuremath{\Varid{s}} without applying the monadic function \ensuremath{\Varid{a}\to \Varid{m}\;\Varid{b}}.
+\citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator \ensuremath{(\bind )\mathbin{::}\Conid{Monad}\;\Varid{m}\Rightarrow \Varid{m}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{b})\to \Varid{m}\;\Varid{b}}.\footnote{In the example a parser of the type \ensuremath{\Conid{Parser}\;\Varid{s}\;\Varid{a}} with static meta information \ensuremath{\Varid{s}} and result \ensuremath{\Varid{a}} is shown to not be able to use the static information \ensuremath{\Varid{s}} without applying the monadic function \ensuremath{\Varid{a}\to \Varid{m}\;\Varid{b}}.
 With Arrows this is possible.}.
@@
 	\end{tabular}
-	\caption{Visual depiction of syntactic sugar for Arrows.}
+	\caption{Visual depiction of derived combinators for Arrows.}
 	\label{fig:syntacticSugarArrows}
 \end{figure}
-Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
+Hughes also derived some further combinators (Fig.~\ref{fig:syntacticSugarArrows}): \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
 \ensuremath{\Varid{second}} is the mirrored version of \ensuremath{\Varid{first}} (Appendix~\ref{utilfns}).
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
-\\
-\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{runPar}\mathbin{\$}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}(\Varid{sequenceA}\;(\Varid{map}\;(\Varid{return}\mathbin{\circ}\Varid{spawn})\;(\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as})))\bind \Varid{mapM}\;\Varid{get}{}\<[E]%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{runPar}\mathbin{\$}\Varid{mapM}\;(\Varid{return}\mathbin{>>>}\Varid{spawn}\bind \Varid{get})\mathbin{\$}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as}{}\<[E]%
 \ColumnHook
@@
 % A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
+Here, the \ensuremath{\Conid{Trans}} typeclass originates from the Eden standard library and details the implementation of packing and of the transmission.
+A~default instance of \ensuremath{\Conid{Trans}} exists.
 
@@
  %$ %% formatting
-\caption{Definition of \ensuremath{\Varid{parEvalNLazy}}.}
+\caption{Definition of \ensuremath{\Varid{parEvalNLazy}}. The type \ensuremath{\Conid{ChunkSize}} is an alias to \ensuremath{\Conid{Int}}.}
 \label{fig:parEvalNLazy}
@@
 	\caption[with Futures]{Communication between 4 Eden processes with Futures.
-Other than in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is highlighted) instead of always going through the master node (bottom bar).}
+Unlike in Fig.~\ref{fig:withoutFutures}, processes communicate directly (one example message is highlighted) instead of always going through the master node (bottom bar).}
 	\label{fig:withFutures}
@@
 
-If we used the original definition of \ensuremath{\Varid{parEvalN}}, however, these skeletons would produce an infinite loop with the GpH and \ensuremath{\Conid{Par}} Monad which during runtime would result in the program crash.
-This materialises with the usage of \ensuremath{\Varid{loop}} of the \ensuremath{\Conid{ArrowLoop}} type class and we think that this is due to difference of how evaluation is done in these backends when compared to Eden.
+If we used the original definition of \ensuremath{\Varid{parEvalN}}, however, these skeletons would produce an infinite loop with the GpH and \ensuremath{\Conid{Par}} Monad which during runtime would result in program crash.
+This materialises with the usage of \ensuremath{\Varid{loop}} of the \ensuremath{\Conid{ArrowLoop}} type class and we think that this is due to difference of how evaluation is done in these backends compared to Eden.
 We provide a workaround for these types of skeletons as such they probably are not of much importance outside of a distributed memory environment.
@@
 
-Extensive use of \ensuremath{\Varid{pipe2}} over \ensuremath{\Varid{pipe}} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{evalN}}.
-Nonetheless, we can define a parallel piping operator \ensuremath{\mathbin{|\!>\!\!>\!\!>\!|}}, which is semantically equivalent to \ensuremath{\mathbin{>\!\!>\!\!>}} similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.
+We speculate that extensive use of \ensuremath{\Varid{pipe2}} over \ensuremath{\Varid{pipe}} with a hand-written combination data type would result in worse performance because of more communication overhead from the many calls to \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{evalN}}.
+Nonetheless, we can define a parallel piping operator \ensuremath{\mathbin{|\!>\!\!>\!\!>\!|}}, which is semantically equivalent to \ensuremath{\mathbin{>\!\!>\!\!>}} similarly to other parallel derived combinators from Appendix~\ref{syntacticSugar}.
 
@@
 Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate with each other.
-Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets send to the neighbour of each node.
+Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets sent to the neighbour of each node.
 This data is sent over direct communication channels, the so called \enquote{remote data}.
@@
 The original inspiration was from \citet{Hughes2005},
-but the definition as then unified with \ensuremath{\Varid{evalN}}.
+but the definition was then unified with \ensuremath{\Varid{evalN}}.
 
@@
 
-\section{Syntactic sugar} \label{syntacticSugar}
-Finally, we also give the definitions for some syntactic sugar for PArrows, namely \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} and \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}}.
+\section{Derived combinators} \label{syntacticSugar}
+Finally, we also give the definitions for some derived combinators for PArrows, namely \ensuremath{\mathbin{\mid\!\!*\!*\!*\!\!\mid}} and \ensuremath{\mathbin{\mid\!\!\&\!\&\!\&\!\!\mid}}.
 For basic Arrows, we have the \ensuremath{\mathbin{*\!*\!*}} combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two Arrows \ensuremath{\Varid{arr}\;\Varid{a}\;\Varid{b}} and \ensuremath{\Varid{arr}\;\Varid{c}\;\Varid{d}} into an Arrow \ensuremath{\Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d})} which does both computations at once.
